### 2022-05-29 manuscript.Rmd
Uploaded the manuscript
At this stage there are no changes

---
title             : "Best Practices in Mixture Modeling using Free Open Source Software"
shorttitle        : "BEST PRACTICES MIXTURE MODELING"

author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
    email         : "c.j.vanlissa@gmail.com"

affiliation:
  - id            : "1"
    institution   : "Utrecht University, Methodology & Statistics"
  - id            : "2"
    institution   : "Open Science Community Utrecht"

author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands

abstract: |
  Latent class analysis is a popular technique for identifying groups in data based
  on a parametric model. Examples of this technique are known as mixture models, latent profile
  analysis, latent class analysis, growth mixture modeling, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture mixture
  models in the `tidySEM` R-package. These functions rely on estimation using
  the OpenMx R-package.

keywords          : "keywords"
wordcount         : "X"

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
mask              : no

class             : "man"
output            : papaja::apa6_pdf
---

```{r load_packages, include = FALSE}
library("papaja")
#bibliography      : ["r-references.bib"]
library(worcs)
```

Latent class analysis has become quite popular across scientific fields,
and under a number of different names.
<!-- Vermunt and Magidson (2004) defined it more generally as virtually any statistical model where “some of the parameters … differ across unobserved subgroups” (Vermunt, J.K. and Magidson, J., 2004. Latent class analysis. In: M.S. Lewis-Beck, A. Bryman, and T.F. Liao (eds.), The Sage Encyclopedia of Social Sciences Research Methods, 549-553. Thousand Oaks, CA: Sage Publications). -->
The purpose of this technique is to estimate unobserved group membership
based on a parametric model of one or more observed indicators of group membership.
Despite the popularity of the method, there is a noted absence of 
guidelines for estimating and reporting latent class analyses [].
<!-- Prior work resulted in the "GRoLTS-Checklist"; reporting guidelines for a specific subcategory of latent class analysis known as latent growth models. vermuntj.k.LatentClassAnalysis2004-->
<!-- General reporting guidelines for latent class analysis are still lacking, however. -->
This complicates manuscript review and assessment of the quality of published studies,
and introduces a risk of misapplications of the technique.
The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on best practice.

Throughout this paper, examples make use of free, open source software for latent class analysis in R. 
As the open science movement is gaining momentum,
researchers are becoming increasingly aware
how important it is that analyses can be reproduced and audited.
In line with open science principles, one of the suggested reporting standards relates to reproducible code.
In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., *Mplus* and *Latent GOLD*).
A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
and prevents auditing research because the underlying source code is proprietary.
To overcome these limitations, the present paper introduces new user-friendly functions in the `tidySEM` R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package `OpenMx`.
The reporting guidelines described in this paper are adopted in `tidySEM` by default.
The `tidySEM` R-package thus makes advanced mixture modeling based on best practices widely accessible,
and facilitates the adoption of the estimation and reporting guidelines described in this paper.

<!-- In practice, this technique is often used to  -->

## Defining latent class analysis

Latent class analysis can be understood as a method for estimating unobserved groups
based on a parametric model of observed indicators of group membership.
The concept of latent class analysis can be understood in different ways.
Generally speaking, it can be said that a mixture model assumes that the study population
is composed of $K$ subpopulations, or classes.
It further assumes that the observed data are a mixture of data generated by class-specific models.
The simplest univariate "model" is a normal distribution, which can be described with two parameters:
the mean and the variance.
Commonly, the same model is estimated accross all classes, but with different parameters for each class (i.e., class-specific means and variances).
Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.

As an illustrative example, imagine that a detective wants to know if it would be
possible to use mixture modeling to identify the sex of a suspect,
based on footprints found at the crime scene.
To test the feasibility of this approach,
the detective records the shoe sizes and sex of 100 volunteers.
The resulting observed data look like this:

```{r shoedens, echo = FALSE, fig.cap="Kernel density plot of shoe sizes."}
library(tidySEM)
library(ggplot2)
set.seed(1)
n = 100
C <- sample(c("Man", "Woman"), n, replace = TRUE)
means <- c(Man = 9, Woman = 7)
X <- rnorm(n, mean = means[C], sd = 1)
X <- round(X*2)/2
obsparam <- tapply(X, C, mean)
est <- mx_profiles(data.frame(X), 2)
estparam <- table_results(est, c("label", "est", "matrix"))
estparam <- estparam$est[estparam$matrix == "M"]
props <- class_prob(est, "individual")$individual
props <- apply(props, 1, which.max)

ggplot(data.frame(Shoesize = X), aes(x = Shoesize)) + geom_density() + theme_bw()
```

The distribution is evidently bimodal, which bodes well for the intended mixture model.
In this case, the number of classes is known a-priori.
When estimating a two-class mixture model, the detective observes that the model estimates
the mean shoesize of the two groups are equal to `r report(estparam[1], equals = FALSE)` and  `r report(estparam[2], equals = FALSE)`,
which is close to the true means of the two groups, namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
When tabulating estimated group membership against observed (known) group membership, it can be seen that women are classified with a high degree of accuracy, but men are not:

```{r tabshoe, results="asis"}
library(kableExtra)
tab <- table(C, props)
tab <- data.frame(Observed = rownames(tab), as.data.frame.array(tab))
names(tab)[2:3] <- c("Class 1", "Class 2")
rownames(tab) <- NULL
kbl(tab, caption = "Observed group membership by estimated class membership.")
```

One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibsonTenFrequentlyAsked2018].
In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.

When the focus is on the model parameters in each group,
then latent class analysis can be thought of as similar to a multi-group structural equation model.
The main distinction is that group membership is not known a-priori,
but is instead estimated - with measurement error - based on the data.
Whereas in a multigroup model, the data are split by group and treated as independent samples,
in a mixture model, all cases contribute to the estimation of all parameters in all groups.
The relative contribution of each case to the parameters of each group is determined by that case's posterior probability of belonging to that group.

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as "model-based
clustering" (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017).
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that latent class analysis 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, latent class analysis can be considered as an
*unsupervised classification* problem [@figueiredoUnsupervisedLearningFinite2002].
The term *unsupervised* refers to the fact that the outcome variable, true class membership,
is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome (class membership).

## A taxonomy of latent class analyses

In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermuntj.k.LatentClassAnalysis2004].
Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
The term "latent class analysis" seems most appropriate as an umbrella term for this broader class of models,
as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.

One common type of latent class analyses is the *finite Gaussian mixture model*;
a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
In the multivatiate case, with more than one indicator variable,
the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
These parameters can be estimated freely, or constrained, across classes.

The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, 
after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for example, when the cohesion between indicators is expected to differ between classes.
As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
A class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.

It is also possible to estimate a mixture model based on latent indicators.
This means that, within each class, one or more continuous latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these continuous latent variables.
A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
Examples of this approach include *growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model to describe individual trajectories over time.
The growth mixture model is a latent class model where the parameters that indicate class membership 
are the intercepts and variances (and typically covariances) of the latent growth variables, e.g., a latent intercept and slope.
This technique assumes that individuals within a class can have heterogenous trajectories.
If the variance of the growth parameters is fixed to zero, it is known as a latent class growth analysis.
This latter approach assumes that all individuals within a class share the same identical trajectory,
and that any variance in the indicators not explained by the class-specific latent trajectories is due to residual error variance.

When a latent class analysis relies on categorical indicators,
typically of a binary or ordinal measurement level,
it has a different parameterization than a mixture model.
One common assumption is that each categorical variable reflects an underlying standard normal distribution.
The parameters in such a model are then "thresholds", corresponding to the quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
For example, a binary indicator has a single threshold that distinguishes the two response categories.
If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.
This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to latent class analyses.

<!-- f a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis. Modified to handle discrete data, this constrained analysis is known as LCA. Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit. -->

<!-- ecause LPA is model-based, a number of different model parameterizations can be -->
<!-- estimated. These models differ in terms of whether--and how--parameters are -->
<!-- estimated across the profiles. These parameters are the means for the different -->
<!-- profiles, which, in this approach, always are estimated freely across the -->
<!-- profiles; the variances for the variables used to create the profiles, which can -->
<!-- be estimated freely or can be estimated to be the same, or equal, across -->
<!-- profiles; and the covariances of the variables used to create the profiles, -->
<!-- which can be freely-estimated, estimated to be equal, or fixed to be zero. -->

<!-- In general, as more parameters are estimated (i.e., those that are fixed to zero -->
<!-- are estimated as being equal across profiles; or those estimated as being equal -->
<!-- across profiles are freely-estimated across them), the model becomes more -->
<!-- complex; the model may fit better, but also be overfit, meaning that the -->
<!-- profiles identified may be challenging to replicate with another, separate data -->
<!-- set. Even still, flexibility in terms of which models can be estimated also has -->
<!-- affordances. For example, the varying means, equal variances, and covariances -->
<!-- fixed to 0. A researcher might choose this model specification if she wants to -->
<!-- model the variables to be used to create profiles that are independent. This -->
<!-- model is very simple, as no covariances are estimated and the variances are -->
<!-- estimated to be the same across profiles. As we estimate more parameters (and -->
<!-- decrease the degrees of freedom), we are more likely to fit the data, but less -->
<!-- likely to be able to replicate the model with a second set of data. As we -->
<!-- progress toward more complex models (with increasingly complex -->
<!-- parameterization), then we are more likely to fit the data better. -->
<!-- In all, this flexibility associated with LPA also has a more pragmatic cost. It -->
<!-- can be very difficult to efficiently---and reproducibly---estimate and compare a -->
<!-- number of models. This cost means that analysts may focus on a subset of models. -->
<!-- It may also mean that unintentional errors are introduced through copying and -->
<!-- pasting the output of model estimations. Part of why we developed tidyLPA was to -->
<!-- make the model estimation process more efficient and more reproducible. As we -->
<!-- describe later, we do this using R, which has a number of benefits, including -->
<!-- being open-source, powerful (in terms of pre-processing and using the output of -->
<!-- models), and increasingly widely-used in the psychological sciences. We also -->
<!-- provide an interface to perhaps the most widely-used software for LPA, MPlus. In -->
<!-- this way, tidyLPA can both allow researchers to estimate models they already -->
<!-- estimate using MPlus more efficiently and reproducibly, and, for those without -->
<!-- access to this software, to do so entirely using R and the open-source R package -->
<!-- mclust. -->

## Use cases for latent class analysis

There are several use cases for which latent class analyses are suitable.
One example is to test a theory that postulates the existence of a categorical latent variable.
For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses.
Mixture modeling can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
If results indicate that the data are better described by a different number of classes,
or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, then the theory may be called into question.

Another use case is unsupervised learning;
when the goal is to restore unobserved class membership based on observed indicators,
or to classify individuals.
For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughmanMixtureModelAnalysis2006] or mental [@wuAbuseDependencePrescription2011] health problems.
The example of shoe size is a rudimentary illustration of this type of application.

<!-- Inductively identify number of latent classes -->
<!-- Classify individuals -->
<!-- Identify items (observed characteristics) that -->
<!-- indicate these classes well -->
<!-- Predict class membership from covariates -->
<!-- Predict outcomes from class membership -->
<!-- Dimension reduction -->
<!-- Violation of assumptions -->

<!-- In medical and psychological science, LPA can be useful when -->
<!-- considerable between-subject heterogeneity exists in scores on a range of variables and when -->
<!-- this variation cannot be explained by known, manifest variables (e.g., Wolfe 1970; Sterba -->
<!-- 2013). Here, LPA can help to identify or approximate possibly meaningful sub-grouping of -->
<!-- subjects that may help to better understand sample heterogeneity (Sterba 2013). -->
<!-- Generally, LPA works under the assumption that sample (residual) variance can be reduced -->
<!-- by assuming a categorical latent variable that effectively subdivides the sample into >=2 -->
<!-- subgroups that are more homogeneous in terms of their patterns of variable means and -->
<!-- (co)variances. In case an LPA model is found to fit a dataset well, it is often the case that -->
<!-- subjects in each class resemble each other relatively well in terms of their scores on the input -->
<!-- variables. Depending on the model configuration, the identified classes can show different -->
<!-- class-specific patterns of means and class-specific or class-varying variances. -->

<!--                                                1 -->
<!-- A -->
<!-- downside is that these models are more complex (much more parameters to be estimated). -->
<!-- Using criteria, such as the Bayesian Information criterion (BIC) helps to find a model that -->
<!-- strikes a good balance between model-fit and model-complexity. When doing an LPA, most -->
<!-- applied researchers will usually be most interested in the differences and/or overlap between -->
<!-- the classes’ specific patterns of parameter estimates. These can be used to characterize the -->
<!-- classes and, possibly, provide clues about underlying mechanisms (Sterba 2013). -->

<!-- This tutorial -->
<!-- LPA is less widely used than other latent variable models and, possibly due to that, has -->
<!-- long been only available in specialized software packages such as Mplus. Luckily, ongoing -->
<!-- developments in many different scientific fields (e.g., ecology, econometrics) have yielded a -->
<!-- number of packages that also allow users to conduct LPA in the open-source R-platform. -->
<!-- However, the use of R does require experience and documentation of packages can be rather -->
<!-- limited or technical, making it less easily accessible for applied researchers. Therefore, this -->
<!-- tutorial is aimed to help applied researchers to get going with LPA in R, illustrating the use -->
<!-- of several packages and, for reference, providing a comparison of the results obtained in R -->
<!-- with results obtained with Mplus. -->

# Best practices

## In estimation

The best practices in estimation, as outlined in Table \@ref(tab:checkest),
are inspired by prior recommendations for best practices for estimating
specific sub-types of latent class analyses,
including latent class growth analysis [@schootGRoLTSChecklistGuidelinesReporting2017]
and latent class analysis with ordinal indicators [e.g., @nylund-gibsonTenFrequentlyAsked2018].
These were generalized to be more relevant to all types of latent class analyses,
and updated to current best practices, as explained below.

```{r checkest}
tab_check <- data.frame(Item = c(
  "Examine observed data",
  "Handling missing data",
  "Alternative model specifications",
  "Starting values",
  "Algorithm"
))
tab_check$`#` = paste0(1:length(tab_check$Item), ".")
kbl(tab_check[, c("#", "Item")])
```
### Examine observed data

Examining observed data is essential for any analysis,
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Pay special attention to level of measurement of the indicators.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions [see @vermuntKmeansMayPerform2011].
Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is perhaps the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @normanLikertScalesLevels2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
The problem becomes eggregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
a latent class analysis with ordinal indicators may be more appropriate.

Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function `tidySEM::descriptives(data)`.
Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
Plots (density plots for continuous variables, and bar charts for categorical ones) may be more diagnostic.
Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @vanlissaWORCSWorkflowOpen2020].

### Missing data

Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled [@schootGRoLTSChecklistGuidelinesReporting2017].
Three types of missingness have been distinguished in the literature [@rubinInferenceMissingData1976]:
Missing completely at random (MCAR), which means that missingness is random;
missing at random (MAR), which means that missingness is contingent on the *observed* data (and can thus be accounted for);
and finally missing not at random (MNAR), which means that missingness is related to unobserved factors.
It is possible to conduct a so-called "MCAR" test, 
for example the non-parametric MCAR test [@jamshidianTestsHomoscedasticityNormality2010].
But note that the name "MCAR test" is somewhat misleading,
as the null-hypothesis of this test is that the data are not MAR,
and a significant test statistic indicates that missingness is related to the observed data (MAR).
A non-significant test statistic does not distinguish between MCAR or MNAR.
As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@littleTestMissingCompletely1988].
This assumption is tenuous in the context of latent class analysis.
A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable [-@jamshidianTestsHomoscedasticityNormality2010].
Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
For this tutorial, I have re-implemented it in the `mice` package as `mice::MCAR()`,
with a fast backend in C++ and new printing and plotting methods.

While we concur that investigating missingness is due dilligence,
it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as `OpenMx` (and e.g., Mplus).
These packages use Full Information Maximum Likelihood (FIML) estimation,
which makes use of all available information without imputing missing values.
FIML is a best-practice solution for handling missing data; on par with multiple imputation [@leeComparisonFullInformation2021].
FIML estimation assumes that missingness is either MCAR or MAR.
Thus, one would typically proceed with FIML regardless of the outcome of an MCAR test.
Although FIML does not, by default, handle missingness in exogenous variables - all indicator variables in latent class analysis are endogenous, so this is not a concern.

Multiple imputation is less suitable to latent class analyses for two reasons.
First, because latent class analyses are often computationally expensive,
and conducting them on multiple imputed datasets may be unfeasible.
Second, because there is no straightforward way to integrate latent class analysis results across multiple datasets.
To conclude; our recommendation is to inspect missingness (e.g., using `mice::MCAR()`)
and report the number of missings per variable (e.g., using `tidySEM::descriptives()`),
before proceeding with FIML.
One minor concern is that the K-means algorithm,
which `tidySEM` uses for determining starting values,
is *not* robust to missing values.
When it fails, `tidySEM` automatically switches to hierarchical clustering,
unless the user specifies a different
clustering algorithm or uses manual starting values.

### Alternative model specifications

The different types of latent class models have different parameters.
For example, mixture models and latent profile analyses typically have class-specific means, variances, and covariances.
Latent growth analyses have the same parameters, but with respect to the latent growth variables.
Latent class analyses with ordinal indicators have thresholds.
All of these parameters can be freely estimated across classes, or constrained,
or fixed (e.g., to zero).
The total number of parameters thus scales with the number of estimated classes.
Consequently, latent class analyses have a potentially very high number of parameters.
As any of these parameters could be mis-specified,
it is important to consider alternative model specifications.
However, alternative model specifications may be approached differently depending on whether an analysis is data driven (exploratory), or theoretically driven (confirmatory).
This distinction has remained underemphasized in prior writing.

Prior literature on latent class analysis has emphasized exploratory applications of the method [see @nylundDecidingNumberClasses2007].
In exploratory analyses, a large number of models are typically estimated in batch,
with varying numbers of classes and model specifications.
The "correct" model specification is then determined based on a combination of fit indices,
significance tests,
and interpretability.
For latent profile analysis, the function `tidySEM::mx_profiles(classes, variances, covariances)`
largely automates this process.
The argument `classes` indicates which class solutions should be estimated (e.g., 1 through 6).
The argument `variances` specifies whether variances should be `"equal"` or `"varying"` across classes.
The argument `covariances` specifies whether covariances should be constrained to `"zero"`, `"equal"` or `"varying"` across classes.
The means are free to vary across classes by default, although the more general function `tidySEM::mx_mixture()` could be used to circumvent this.
After all models have been estimated, the function `tidySEM::table_fit()` can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
because this test is very computationally expensive.
It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
Fit indices typically used for determining the optimal number of classes include
the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC).
Both information criteria are based on the -2 log likelihood (which is lower for better fitting models),
and add a penalty for the number of parameters (thus incentivizing simpler models).
This helps balance fit and model complexity.
The BIC usually applies a stronger penalty for complexity that scales logarithmically with the sample size.

Fit indices may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
One option is to select a specific fit index before analyzing the data.
Another option is to always prefer the most parsimoneous model that has best fit according to any of the available fit indices.
Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogulComparisonInformationCriteria2016].

Confirmatory analyses typically require less comprehensive alternative model specifications.
For example, in the context of preregistered analyses, the main models of interest may have been specified a priori.

Van de Schoot and colleagues [-@vandeschootGRoLTSChecklistGuidelinesReporting2017] argue that there are many choices to be made in the specification of latent trajectory models,
and that alternative

	* Consider alternative parameterizations; e.g., free variances/covariances 
	* See Nylund
1. Starting values [@nylund-gibsonTenFrequentlyAsked2018; ]
    In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different. Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm. [@biernackiChoosingStartingValues2003]
1. Algorithm
    EM is a local optimizer. Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator (ignoring the variance for now). The EM algorithm will attempt to find a combination of values for these two parameters that maximizes the likelihood ($LL$) of all observed data. In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. We can think of this optimization problem as a three-dimensional landscape: The X and Y dimensions are determined by the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. The optimizer must find the deepest "valley" in this landscape, which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
The EM optimizer behaves somewhat like a marble, dropped in this landscape.
It is dropped at some random point in space, and will roll into the nearest valley.
The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley (this is known as "convergence").
It cannot climb out again.
Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley (a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
This is the "random starts" approach.

One problem with the random starts approach is that it is computationally expensive to run this many replications.
Moreover, because the algorithm begins with random starting values,
many of the marbles are likely to be very far away from a "good enough" solution.
Two innovations may improve the estimation procedure.
The first is that, instead of picking random starting values, a "reasonable solution" may be used for the starting values.
For example, if we assume that the different classes are likely to have different mean values on the indicators, then the K-means clustering algorithm can be used to determine these cluster centroids.
We can compute the expected values of all model parameters by treating the K-means solution as a known class solution, and use these as starting values for a mixture model.
One remaining concern is that this approach may result in starting values close to a local optimum,
and that the EM algorithm will thus never find the global optimum.
A second innovation addresses this concern.
Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
Simulated annealing iteratively considers some "destination" in the landscape, and compares its likelihood to the current one. If the destination likelihood is higher, the estimator moves there.
If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
This latter property allows it to escape local optima.

By default, `tidySEM` eploys this solution of deriving starting values using K-means clustering,
and identifying the global optimum solution using simulated annealing.
Once a solution has been found,
simulated annealing is followed up with a short run of the EM algorithm,
as EM inherently produces an asymptotic covariance matrix for the parameters that can be used to compute standard errors.
Note that these defaults can be manually overridden.

## Software

Many software packages are available for the estimation of latent class analyses.
Some of these packages have limited functionality, or implement particular innovations.
Other packages implement latent class analyses in the context of a more flexible structural equation modeling framework.
The most notable examples of the latter are the commercial programs Mplus and Latent GOLD, 
and the free open source R-package OpenMx.
The commercial packages stand out because they offer relatively user-friendly interfaces
and implement sensible defaults for complex analyses, including latent class analysis.
This somewhat lowers the threshold for applied researchers to adopt such methods.
Commercial software also has several downsides, however.
One such downside is that use of the software is restricted to those individuals and institutions who can afford a license.
A second downside is that the source code, being proprietary, cannot be audited, debugged, or enhanced by third parties.
This incurs the risk that mistakes in the source code may go unnoticed, and curbs progress as software developers cannot add new functionality.

Conversely, the free open source program OpenMx is very flexible,
but not very user-friendly.

New functionality in the R-package `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
It adheres to best practices in estimation and reporting, as described in this paper.
The user interface is simple, making use of the model syntax of the widely used `lavaan` R-package.
This syntax offers a human-readable way to specify latent variable models.
Minor enhancements are made to simplify the specification of latent class analysis.

Because of the limitations in ex tools, we set out to develop a tool that a)
provided sensible defaults and were easy to use, but provided the option to
access and modify all of the inputs to the model (i.e., low barrier, high
ceiling), b) interfaced to existing tools, and are able to translate between
what existing tools are capable of and what researchers and analysts
carrying-out person-oriented analyses would like to specify,

fully-reproducible analyses and

<!-- 3. Packages -->
<!-- 3.1 R-packages -->
<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
<!-- analytic functionality. However, the majority of packages is focused on analyses with discrete -->
<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->

<!-- Although ‘tidyLPA’ is easy to use, this comes at the expense of restricting modeling options -->
<!-- to a few, oft-used options. For completeness, both the ‘mclust’ and ‘tidyLPA’ approach will -->
<!-- be illustrated. -->

## Best practices in estimation

### Starting values

### Algorithm

### Label switching

<!-- Initialization around a Single Mode -->
<!-- Another common approach is to run a single chain or to initialize the parameters near realistic values.36 -->

<!-- This can work better than the hard constraint approach if reasonable initial values can be found and the labels do not switch within a Markov chain. The result is that all chains are glued to a neighborhood of a particular mode in the posterior. -->
<!-- https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/1467-9868.00265?casa_token=Tjk8gR4w8AEAAAAA:RygJL5OgdBaVWY6oaB8XlnyMOnzQgOVDBd2EHDyQD6TH-DzxRxpmU1J4AG8N-HpL5dHHNINRybiaMl6w -->

## Best practices in reporting

The most relevant

1. Open science
	* Use comprehensive citation; this includes referencing the software used.
	* Share code and (if possible) data. Van Lissa and colleagues suggest sharing synthetic data in case the original data cannot be shared, and provide functions to generate such synthetic data.
	* Ideally, make the entire research project reproducible so that others may download it and reproduce it with just one click; for guidance, see Van Lissa and colleagues.

```{r}
grolts <- structure(list(Number = c("1.", "2.", "3a.", "3b.", "3c.", "4.", 
"5.", "6a.", "6b.", "7.", "8.", "9.", "10.", "11.", "12.", "13.", 
"14a.", "14b.", "14c.", "15.", "16."), Item = c("Is the metric of time used in the statistical model reported?", 
"Is information presented about the mean and variance of time within a wave?", 
"Is the missing data mechanism reported?", "Is a description provided of what variables are related to attrition/missing data?", 
"Is a description provided of how missing data in the analyses were dealt with?", 
"Is information about the distribution of the observed variables included?", 
"Is the software mentioned?", "Are alternative specifications of within-class heterogeneity considered (e.g., LGCA vs. LGMM) and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative shape/functional forms of the trajectories described?", 
"If covariates have been used, can analyses still be replicated?", 
"Is information reported about the number of random start values and final iterations included?", 
"Are the model comparison (and selection) tools described from a statistical perspective?", 
"Are the total number of fitted models reported, including a one-class solution?", 
"Are the number of cases per class reported for each model (absolute sample size, or proportion)?", 
"If classification of cases in a trajectory is the goal, is entropy reported?", 
"Is a plot included with the estimated mean trajectories of the final solution?", 
"Are plots included with the estimated mean trajectories for each model?", 
"Is a plot included of the combination of estimated means of the final model and the observed individual trajectories split out for each latent class?", 
"Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?", 
"Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?"
), relevant = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, TRUE, TRUE)), row.names = c(NA, -21L), class = "data.frame")
```

## Best practices in visualization

# Tutorial

<!-- 5. Tutorial -->
<!-- 5.1 ‘mclust’ -->
<!-- To run LPA using the ‘mclust’ package, the Mclust() function can be used. This function -->
<!-- requires data as its minimum input. In addition, the number of clusters to fit (G) and model -->
<!-- variants to fit and select from (modelNames) can be entered (defaults are: G=1-9 and all -->
<!-- fourteen model variants). In many cases, we may want to consider a more limited range of -->
<!-- model variants to with different numbers of classes. For multivariate data, a total of fourteen -->
<!-- options are available (see help("mclustModelNames") for an overview). Each model variant -->
<!-- is estimated for each value of the specified range of G using Expectation Maximization (EM), -->
<!-- using the Bayesian Information Criterion (BIC) to compare different model variants and -->
<!-- select the optimal one. -->
<!-- Conducting a latent variable model or mixture analysis with Mplus or comparable software -->
<!-- usually entails fitting models with different numbers of classes (that are otherwise configured -->
<!-- similarly) and to compare the fit between these models to select the best model. The -->
<!-- default approach taken in the ‘mclust’ package is slightly different in that the main modeling -->
<!-- command Mclust() fits multiple models for all possible combinations of the specified number -->
<!-- of classes and model configurations . The eventual output of ‘mclust’ is the BIC-selected best -->
<!-- model variant with the optimal combination of number of classes and model configuration. -->
<!-- An advantage of the ‘mclust’ approach is that it is efficient and flexible: given a number of -->
<!-- classes, the best model configuration is selected straight away and we do not have to run all -->
<!-- model combinations one by one. This makes the approach especially suitable for exploratory -->
<!-- analyses. However, like stated above, this approach is somewhat less usual in some fields, -->
<!-- where estimation of LPA is approached in a more confirmatory fashion. In addition, it does -->
<!-- take some control away from the researcher, who may want to primarily focus on selecting -->
<!-- the number of classes, given a particular , theory-based model configuration that is kept -->
<!-- constant throughout analyses, which is often the chosen strategy in software like Mplus and -->
<!-- made easily available through the ‘tidyLPA’ package. -->
<!-- Here, both approaches are shown: (1) the ‘mclust’ aproach and (2) the ‘tidyLPA’ approach. -->

<!--                                               5 -->
<!-- 5.1.1 The ‘mclust’ approach -->
<!-- In this approach, we run consecutive models with increasing numbers of classes (G=1 to -->
<!-- G=9), for each model we let the package fit the four above described model variants (“EEI”, -->
<!-- “EEE”, “VVI” and “VVV”) and select the best-fitting one. We start by loading the package -->
<!-- (library(mclust)) and creating an object mnames that contains the names of the four models -->
<!-- we want to be fitted. Next, we use the Mclust() function to fit the models: -->
<!-- library(mclust) -->

<!-- ## Package 'mclust' version 5.4.7 -->
<!-- ## Type 'citation("mclust")' for citing this R package in publications. -->
<!-- mnames <- c("EEI", "EEE", "VVI", "VVV") -->

<!-- # Fit 1-5 class model -->
<!-- mod_g1_9 <- Mclust(dat1[, 1:10], modelNames = mnames) -->

<!-- We can first look at the optimal number of classes and the optimal model variant that were -->
<!-- selected, using the following code: -->
<!-- # Opimal number of classes -->
<!-- mod_g1_9$G -->

<!-- ## [1] 3 -->
<!-- # Optimal model variant -->
<!-- mod_g1_9$modelName -->

<!-- ## [1] "EEI" -->
<!-- This shows us that a 3-class EEI model was selected as the best model based on the BIC. We -->
<!-- can get a better perspective of this model’s performance if we compare it to the other fitted -->
<!-- models. We can do this by taking a closer look at the other models’ BIC values: -->
<!-- mod_g1_9$BIC -->

<!-- ##   Bayesian Information Criterion (BIC): -->
<!-- ##           EEI       EEE       VVI       VVV -->
<!-- ##   1 -14661.61 -12986.57 -14661.61 -12986.57 -->
<!-- ##   2 -13275.65 -12611.60 -13095.82 -12678.12 -->
<!-- ##   3 -12142.11 -12290.68 -12163.08 -12731.18 -->
<!-- ##   4 -12178.44 -12330.04 -12241.21 -12970.57 -->
<!-- ##   5 -12223.14 -12379.96 -12320.47 -13231.35 -->
<!-- ##   6 -12264.04 -12423.72 -12406.59 -13522.40 -->
<!-- ##   7 -12269.39 -12451.25 -12497.03 -13748.91 -->
<!-- ##   8 -12298.30 -12483.35 -12568.62 -13980.82 -->
<!-- ##   9 -12347.11 -12529.40 -12645.50 -14269.34 -->
<!-- ## -->

<!--                                              6 -->
<!-- ## Top 3 models based on the BIC criterion: -->
<!-- ##     EEI,3     VVI,3     EEI,4 -->
<!-- ## -12142.11 -12163.08 -12178.44 -->
<!-- mod_g1_9$loglik -->

<!-- ## [1] -5951.275 -->
<!-- mod_g1_9$df -->

<!-- ## [1] 42 -->
<!-- Here, we can see the BICs for all fitted models (in this case 1-9 classes and 4 model variants: -->
<!-- 36 models in total). We can see that the closest contenders were a 3-class model with a VVI -->
<!-- configuration (with variances allowed to vary both within and between classes) and a 4-class -->
<!-- EEI model. -->
<!-- Note that the BIC values are negative, and that values that are more negative are considered -->
<!-- to fit poorer than values closer to 0. This may strike some users as odd, given that in -->
<!-- many modeling applications, we are used to consider lower BIC values to indicate better -->
<!-- fit. However, the ‘mclust’ approach of maximizing the BIC is correct for the used modeling -->
<!-- approach (see e.g., Fraley & Raftery, 2003; Banfield & Raftery, 1993). For theoretical reasons, -->
<!-- the BIC is calculated in’mclust’ as: -->

<!--                            BIC = 2(Loglikelihood) - df (log(n)) -->

<!-- where df is the number of parameters (degrees of freedom) and n is the sample size. We can -->
<!-- see, that here, higher loglikelihood values lead to a higher BIC. For the best fitting model in -->
<!-- our example, the loglikelihood is -5951.275, the number of parameters is 42 and the sample -->
<!-- size is 300. If we plug these into the formula, we get the following BIC: -->

<!--                      BIC = 2(-5951.275) - 42(log(300)) = -12142.11 -->

<!-- As expected, the obtained value indeed corresponds to the BIC we got in the model output -->
<!-- (see above). Other software packages, such as Mplus and ‘tidyLPA’ (see below) use a slightly -->
<!-- different formula to calculate the BIC, where higher loglikelihood values lead to lower BIC -->
<!-- values: -->

<!--                            BIC = df (log(n)) - 2(Loglikelihood) -->

<!-- When we plug the best-model values into this formula, we get: -->

<!--                       BIC = 42(log(300)) - 2(-5951.275) = 12142.11 -->

<!-- We can see that the absolute BIC value is the same, irrespective of the used formula, with -->
<!-- only the sign differing between the two approaches. -->

<!--                                               7 -->
<!-- Now, let us continue by taking a closer look at the class sizes and the mean values and -->
<!-- variances of the ten variables for the three identified classes. Because the best model has and -->
<!-- EEI configuration, printing out a single variance matrix from one class will suffice, because -->
<!-- the variances are the same across classes (this is different in cases where the optimal model -->
<!-- allows for between-class differences in (co)variances: -->
<!-- # tabulate class-membership numbers -->
<!-- table(summary(mod_g1_9)$classification) -->

<!-- ## -->
<!-- ##   1   2      3 -->
<!-- ## 100 110     90 -->
<!-- # display the means per class -->
<!-- mod_g1_9$parameters$mean -->

<!-- ##             [,1]         [,2]        [,3] -->
<!-- ##   var1 2.023572      6.083019   2.0304980 -->
<!-- ##   var2 3.056837      6.937785   4.8997993 -->
<!-- ##   var3 1.718166      1.107873   7.9385872 -->
<!-- ##   var4 6.086727      2.854402   6.9657536 -->
<!-- ##   var5 7.101109      2.225675   4.7416266 -->
<!-- ##   var6 3.691936      8.081570   2.2469977 -->
<!-- ##   var7 5.009819      2.817889   0.8862725 -->
<!-- ##   var8 8.316894      8.080891   3.9350536 -->
<!-- ##   var9 1.978136      1.896294   9.1537131 -->
<!-- ##   var10 1.191249     5.142590   8.0332628 -->
<!-- # select the first of 3 10-by-10 variance matrices -->
<!-- mod_g1_9$parameters$variance$sigma[1:10, 1:10, 1] -->

<!-- ##               var1       var2       var3       var4       var5       var6       var7       var8 -->
<!-- ##   var1    2.539537   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
<!-- ##   var2    0.000000   2.117941   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
<!-- ##   var3    0.000000   0.000000   2.509035   0.000000   0.000000   0.000000   0.000000   0.000000 -->
<!-- ##   var4    0.000000   0.000000   0.000000   2.812796   0.000000   0.000000   0.000000   0.000000 -->
<!-- ##   var5    0.000000   0.000000   0.000000   0.000000   2.647399   0.000000   0.000000   0.000000 -->
<!-- ##   var6    0.000000   0.000000   0.000000   0.000000   0.000000   2.694156   0.000000   0.000000 -->
<!-- ##   var7    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.368379   0.000000 -->
<!-- ##   var8    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.540704 -->
<!-- ##   var9    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
<!-- ##   var10   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
<!-- ##               var9      var10 -->
<!-- ##   var1    0.000000   0.000000 -->
<!-- ##   var2    0.000000   0.000000 -->
<!-- ##   var3    0.000000   0.000000 -->
<!-- ##   var4    0.000000   0.000000 -->

<!--                                                8 -->
<!-- ##   var5    0.000000   0.000000 -->
<!-- ##   var6    0.000000   0.000000 -->
<!-- ##   var7    0.000000   0.000000 -->
<!-- ##   var8    0.000000   0.000000 -->
<!-- ##   var9    2.509146   0.000000 -->
<!-- ##   var10   0.000000   2.204457 -->
<!-- From this output, we can see that the three classes have class-sizes of 100, 110 and 90, -->
<!-- respectively. These sizes along with the patterns of class-specific mean variable scores -->
<!-- correspond with those that served as the input for the simulation. As specified, the variances -->
<!-- vary slightly across variables. -->
<!-- The class allocation in LPA is probabilistic in nature: each subject in the data is assigned a -->
<!-- probability for each of the estimated classes, based on their pattern of scores on the input -->
<!-- variables. These probabilities can be inspected in the z-matrix (here: mod_g1_9$z). Subjects -->
<!-- can be allocated to one of the classes based on their highest class-probability. These posterior -->
<!-- allocations can be found in the classification matrix (here: mod_g1_9$classification). -->
<!-- These class-allocations were tabulated above to evaluate class sizes. -->
<!-- Note that class-allocation in this way only yields useful classifications if the patterns of -->
<!-- class-probabilities allow for allocation of each subject to a single class with sufficient certainty. -->
<!-- In case of too much uncertainty, using the classification is not advised. We can inspect the -->
<!-- uncertainty of allocation for all subjects (here: mod_g1_9$uncertainty). This gives us a list -->
<!-- of uncertainties for all subjects. Next, we can evaluate the extent of uncertainty by looking -->
<!-- at the maximum uncertainty or, for instance, the averaged uncertainty across subjects: -->
<!-- max(mod_g1_9$uncertainty) -->

<!-- ## [1] 0.003852604 -->
<!-- mean(mod_g1_9$uncertainty) -->

<!-- ## [1] 2.607299e-05 -->
<!-- Here, the uncertainty is atypically low because of the used idealized data. In many cases, -->
<!-- uncertainty is higher and it may be of interest to investigate it further, for instance, by -->
<!-- looking at (e.g., plotting or tabulating) the uncertainty per class: -->
<!-- cprob <- cbind(mod_g1_9$z, mod_g1_9$classification) -->
<!-- cprob <- as.data.frame(cprob) -->
<!-- colnames(cprob) <- c("prob (class 1)", "prob (class 2)", "prob (class 3)", "class") -->
<!-- aggregate(cprob[, 1:3], list(cprob$class), mean) -->

<!-- ##   Group.1 prob (class 1) prob (class 2) prob (class 3) -->
<!-- ## 1       1   9.999824e-01   1.761611e-05   3.350070e-10 -->
<!-- ## 2       2   5.486300e-05   9.999451e-01   7.192849e-08 -->
<!-- ## 3       3   2.591358e-08   1.675400e-07   9.999998e-01 -->
<!-- Here we can see that for each class, the mean probability of membership of that class was -->

<!--                                                  9 -->
<!-- very high and probabilities for the other classes were very low. Again, these values reflect the -->
<!-- idealized data; in many cases, more differentiated class-probability patterns may be observed. -->

<!-- 5.1.2 ‘tidyLPA’ -->
<!-- When running ‘mclust’ models from the ‘tidyLPA’ package, the estimations are approached -->
<!-- a little differently. By default, the number of models that can be estimated is restricted to -->
<!-- six relatively common variants (see help(tidyLPA::estimate_profiles) for details). Of -->
<!-- these models, four can be estimated with ‘mclust’. These models differ in terms of how the -->
<!-- variances and covariances are allowed to vary or constrained to be equal across classes and -->
<!-- whether covariances are or are not fixed to zero within classes. These models have each been -->
<!-- allocated a number (1-6) in ‘tidyLPA’. See the table below how these numbers correspond to -->
<!-- the above mentioned model configurations. -->
<!--  Model variant    model number -->
<!--  EEI              1 -->
<!--  EEE              3 -->
<!--  VVI              2 -->
<!--  VVV              6 -->
<!-- Models four and five are different and can only be fit when R is interfacing with Mplus using -->
<!-- the ‘MplusAutomation’ package. This is outside the scope of this tutorial. -->
<!-- To fit an LPA model in ‘tidyLPA’, we use the estimate_profiles() function. Here, we need -->
<!-- to enter the data.frame to be used (df) and the number of classes to estimate (n_profiles). -->
<!-- To determine what kind of model configuration to estimate, the authors have provided two -->
<!-- different ways in the estimate_profiles() command. In the first approach, we simply tell what -->
<!-- model number (see Table 2) we want to estimate using the models= argument For instance, if -->
<!-- we want to estimate LPA models with 1 to 9 classes, with an EEI configuration, we can use: -->
<!-- <!-- suppressMessages(library(tidyLPA)) --> -->
<!-- <!-- suppressMessages(mod_1c_v1 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, --> -->
<!-- <!--     models = 1)) --> -->

<!-- The package issues a standard message by default notifying us that we use the models -->
<!-- argument and that the variances and covariances arguments are ignored. Here, we suppress -->
<!-- the message. The output shows us the model-fit information for the 1 to 9-class EEI models: -->
<!-- <!-- mod_1c_v1 --> -->

<!-- ## tidyLPA analysis using mclust: -->
<!-- ## -->
<!-- ## Model Classes AIC       BIC      Entropy            prob_min   prob_max   n_min   n_max   BLRT_p -->
<!-- ## 1      1       14587.53 14661.61 1.00               1.00       1.00       1.00    1.00 -->
<!-- ## 1      2       13160.84 13275.65 1.00               1.00       1.00       0.30    0.70    0.01 -->
<!-- ## 1      3       11986.55 12142.11 1.00               1.00       1.00       0.30    0.37    0.01 -->
<!-- ## 1      4       11982.14 12178.44 0.91               0.79       1.00       0.14    0.33    0.05 -->
<!-- ## 1      5       11986.09 12223.14 0.89               0.75       1.00       0.06    0.33    0.30 -->

<!--                                               10 -->
<!-- ##   1      6         11986.25    12264.04   0.84      0.78       1.00        0.08   0.30    0.16 -->
<!-- ##   1      7         11950.86    12269.39   0.81      0.77       0.91        0.08   0.17    0.01 -->
<!-- ##   1      8         11939.04    12298.30   0.82      0.79       0.92        0.08   0.17    0.02 -->
<!-- ##   1      9         11947.10    12347.11   0.82      0.76       0.92        0.06   0.16    0.57 -->
<!-- Here, we can see that the BIC takes on different values compared to ‘mclust’, with lower -->
<!-- rather than higher values indicating best fit (see explanation above). Instead of drawing the -->
<!-- fit indices (e.g., BIC, AIC) directly from the ‘mclust’ package, ‘tidyLPA’ only draws the ‘raw’ -->
<!-- loglikelihood values and posterior probabilities from the ‘mclust’ output and (re)calculates -->
<!-- the BIC, AIC, entropy etc. to mirror as well as possible those provided in Mplus. In addition, -->
<!-- p-values of the the bootstrapped likelihood ratio test (BLRT) are given by default. These -->
<!-- p-values indicate for each k-class model whether adding the kth class adds significantly to -->
<!-- model fit. For instance, the BLRT_p of 0.04 for the 4-class model indicates that adding -->
<!-- a fourth class led to an improvement of model fit compared to the 3-class model that was -->
<!-- just significant when using an alpha of 0.05. When using ’mclust’ as a stand-alone package, -->
<!-- the BLRT is not calculated by default, but can be obtained with the mclustBootstrapLRT -->
<!-- function. -->
<!-- Next, we can run similar commands, with different values for the models argument. It is -->
<!-- also possible to estimate more than one model variant in a single run by providing a vector -->
<!-- of model numbers (e.g., models=c(1,6)). -->
<!-- The second approach to determine the model variant to be estimated is by using the variances -->
<!-- and covariances arguments in the estimate_profiles command. The variances argument -->
<!-- can have the values "equal" (i.e. variable variances constrained to be equal across classes) or -->
<!-- "varying" (i.e. variances allowed to vary across classes). The covariances argument can -->
<!-- have the values: "zero" (i.e. all variable covariances fixed to zero), "equal" (i.e. covariances -->
<!-- constrained to be equal across classes) or "varying" (i.e. covariances allowed to vary across -->
<!-- classes). Now, if we again want to estimate LPA models with 1 to 9 classes, with an EEI -->
<!-- configuration, we can use: -->
<!-- <!-- mod_1c_v2 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, variances = "equal", --> -->
<!-- <!--     covariances = "zero") --> -->
<!-- <!-- mod_1c_v2$model_1_class_2$fit --> -->

<!-- <!-- ##         Model           Classes        LogLik                   AIC              AWE --> -->
<!-- <!-- ## 1.000000e+00       2.000000e+00 -6.549418e+03          1.316084e+04     1.354347e+04 --> -->
<!-- <!-- ##           BIC              CAIC           CLC                   KIC            SABIC --> -->
<!-- <!-- ## 1.327565e+04       1.330665e+04 1.310084e+04           1.319484e+04     1.317734e+04 --> -->
<!-- <!-- ##           ICL           Entropy      prob_min              prob_max            n_min --> -->
<!-- <!-- ## -1.327568e+04      9.996750e-01 9.999484e-01           9.999990e-01     3.000000e-01 --> -->
<!-- <!-- ##         n_max          BLRT_val        BLRT_p --> -->
<!-- <!-- ## 7.000000e-01       1.448699e+03 9.900990e-03 --> -->
<!-- As expected, the output is exactly the same as the one we got when using the models -->
<!-- argument. Now, if we want to, we can continue fitting other model variants using either -->
<!-- approach. Next, we can manually compare the models in terms of their BIC and AIC values -->

<!--                                               11 -->
<!-- or we can use compare_solutions to do this in an automated fashion. Interestingly, the -->
<!-- best model can be selected based on the integrated information about several fit indices -->
<!-- (analytic hierarchy process; see help(tidyLPA::AHP)) for more details). We obtain the model -->
<!-- comparison with the following code. -->
<!-- <!-- comp <- suppressWarnings(compare_solutions(mod_1c_v1)) --> -->
<!-- <!-- comp$fits --> -->

<!-- <!-- ##   # A tibble: 9 x 18 --> -->
<!-- <!-- ##     Model Classes LogLik    AIC    AWE    BIC   CAIC    CLC    KIC SABIC      ICL --> -->
<!-- <!-- ##     <dbl>   <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>           <dbl> --> -->
<!-- <!-- ##   1     1       1 -7274. 14588. 14834. 14662. 14682. 14550. 14611. 14598. -14662. --> -->
<!-- <!-- ##   2     1       2 -6549. 13161. 13543. 13276. 13307. 13101. 13195. 13177. -13276. --> -->
<!-- <!-- ##   3     1       3 -5951. 11987. 12506. 12142. 12184. 11905. 12032. 12009. -12142. --> -->
<!-- <!-- ##   4     1       4 -5938. 11982. 12638. 12178. 12231. 11878. 12038. 12010. -12218. --> -->
<!-- <!-- ##   5     1       5 -5929. 11986. 12778. 12223. 12287. 11860. 12053. 12020. -12273. --> -->
<!-- <!-- ##   6     1       6 -5918. 11986. 12915. 12264. 12339. 11838. 12064. 12026. -12353. --> -->
<!-- <!-- ##   7     1       7 -5889. 11951. 13016. 12269. 12355. 11780. 12040. 11997. -12376. --> -->
<!-- <!-- ##   8     1       8 -5873. 11939. 13141. 12298. 12395. 11747. 12039. 11991. -12409. --> -->
<!-- <!-- ##   9     1       9 -5866. 11947. 13285. 12347. 12455. 11733. 12058. 12005. -12458. --> -->
<!-- <!-- ##   # ... with 7 more variables: Entropy <dbl>, prob_min <dbl>, prob_max <dbl>, --> -->
<!-- <!-- ##   #   n_min <dbl>, n_max <dbl>, BLRT_val <dbl>, BLRT_p <dbl> --> -->
<!-- <!-- comp$best --> -->

<!-- <!-- ## LogLik       AIC      AWE      BIC     CAIC        CLC    KIC   SABIC       ICL --> -->
<!-- <!-- ##      9         8        3        3        3          9      3       8         3 --> -->
<!-- As shown in the ‘mclust’ examples above, the EEI model (models=1) with 3 classes fit the -->
<!-- data best, not only according to the BIC, but according to the integrated information from a -->
<!-- broader range of fit indices. Note that we did not fit any competing EEE, VVV and VVI -->
<!-- models here. If we would have wanted to compare fit to these model variants too, we should -->
<!-- have actively told the package to fit these too, either by specifying a vector of different types -->
<!-- of models to fit in the models argument or by repeating the analysis with different values for -->
<!-- the variances and covariances arguments. -->

<!-- 5.2 Comparison to Mplus -->
<!-- For comparison, we fit the equivalent EEI model variants in Mplus. For this, we can mostly -->
<!-- rely on the defaults of Mplus. We only need to specify that we want to estimate class-specific -->
<!-- variable means. The variances are automatically constrained to be equal across classes and -->
<!-- within class variable covariances are fixed to 0. Here we have the separate codes for fitting -->
<!-- the first 4 models wit 1-4 class models (code for the 5-9 class models not shown). -->
<!-- <!-- !! 1-class model --> -->
<!-- <!--   DATA: file='dat1.dat'; --> -->
<!-- <!--   VARIABLE: --> -->

<!-- <!--                                                  12 --> -->
<!-- <!--     names=id v1-v10; --> -->
<!-- <!--     usevariables= v1-v10; --> -->
<!-- <!--     classes = c(1); --> -->
<!-- <!--   ANALYSIS: --> -->
<!-- <!--     type = mixture; --> -->
<!-- <!--     starts= 250 50; --> -->
<!-- <!--   MODEL: --> -->
<!-- <!--   %overall% --> -->
<!-- <!--   %C#1% --> -->
<!-- <!--   [v1-v10]; --> -->

<!-- <!-- !! 2-class model --> -->
<!-- <!--   DATA: file='dat1.dat'; --> -->
<!-- <!--   VARIABLE: --> -->
<!-- <!--     names=id v1-v10; --> -->
<!-- <!--     usevariables= v1-v10; --> -->
<!-- <!--     classes = c(2); --> -->
<!-- <!--   ANALYSIS: --> -->
<!-- <!--     type = mixture; --> -->
<!-- <!--     starts= 250 50; --> -->
<!-- <!--   MODEL: --> -->
<!-- <!--   %overall% --> -->
<!-- <!--   %C#1% --> -->
<!-- <!--   [v1-v10]; --> -->
<!-- <!--   %C#2% --> -->
<!-- <!--   [v1-v10]; --> -->

<!-- <!-- !! 3-class model --> -->
<!-- <!--   DATA: file='dat1.dat'; --> -->
<!-- <!--   VARIABLE: --> -->
<!-- <!--     names=id v1-v10; --> -->
<!-- <!--     usevariables= v1-v10; --> -->
<!-- <!--     classes = c(3); --> -->
<!-- <!--   ANALYSIS: --> -->
<!-- <!--     type = mixture; --> -->
<!-- <!--     starts= 250 50; --> -->
<!-- <!--   MODEL: --> -->
<!-- <!--   %overall% --> -->
<!-- <!--   %C#1% --> -->
<!-- <!--   [v1-v10]; --> -->
<!-- <!--   %C#2% --> -->
<!-- <!--   [v1-v10]; --> -->
<!-- <!--   %C#3% --> -->
<!-- <!--   [v1-v10]; --> -->

<!-- <!--                             13 --> -->
<!-- <!-- !! 3-class model --> -->
<!-- <!--   DATA: file='dat1.dat'; --> -->
<!-- <!--   VARIABLE: --> -->
<!-- <!--     names=id v1-v10; --> -->
<!-- <!--     usevariables= v1-v10; --> -->
<!-- <!--     classes = c(4); --> -->
<!-- <!--   ANALYSIS: --> -->
<!-- <!--     type = mixture; --> -->
<!-- <!--     starts= 250 50; --> -->
<!-- <!--   MODEL: --> -->
<!-- <!--   %overall% --> -->
<!-- <!--   %C#1% --> -->
<!-- <!--   [v1-v10]; --> -->
<!-- <!--   %C#2% --> -->
<!-- <!--   [v1-v10]; --> -->
<!-- <!--   %C#3% --> -->
<!-- <!--   [v1-v10]; --> -->
<!-- <!--   %C#4% --> -->
<!-- <!--   [v1-v10]; --> -->
<!-- Each of the models was fitted with multiple initial stage random starts and multiple final -->
<!-- stage optimizations to reduce the risk of finding solutions on a local maximum. For the 1-6 -->
<!-- class models, 250 iniitial and 50 final starts were used. For the 7-class model 1000 initial -->
<!-- starts and 200 final optimizations were used and for the 8- and 9-class models, a replicable -->
<!-- global solution could only be obtained with 5000 initial stage starts and 1000 final stage -->
<!-- optimizations. The fit indices for each of the fitted models is displayed in the table below. -->
<!-- For now, we only look at the estimated fit indices. -->
<!--                 EEI models -->
<!--  Classes   AIC       BIC -->
<!--  1         14587.535 14587.535 -->
<!--  2         13160.836 13275.653 -->
<!--  3         11986.551 12142.11 -->
<!--  4         11957.71  12154.01 -->
<!--  5         11951.766 12188.808 -->
<!--  6         11944.632 12222.415 -->
<!--  7         11935.501 12254.027 -->
<!--  8         11929.428 12288.695 -->
<!--  9         11925.203 12325.212 -->
<!-- If we compare the BIC values for the EEI models to those obtained with ‘mclust’ directly -->
<!-- and via ‘tidyLPA’, we can see that the results are quite similar for the less complex models, -->
<!-- with the BIC-values for the 2- and 3-class models being exactly the same in terms of their -->
<!-- absolute values. When taking a closer look at the 3-class model that was previously found to -->
<!-- be the best model, we can also see that that the Mplus-based classifications (n=100, n=110, -->
<!-- and n=90) and entropy estimates (entropy=1.0) are similar to those obtained with ‘mclust’ -->

<!--                                              14 -->
<!-- and/or ‘tidyLPA’. -->
<!--  If we look at the larger picture and evaluate the overlap between the BIC values obtained -->
<!--  for all models with class numbers ranging from 1 to 9, we can see that the estimated BIC -->
<!-- values show more differences between the R-packages and Mplus for the models with larger -->
<!--  numbers of classes. This can be explained by the different methods used by Mplus and -->
<!-- ‘mclust’/’tidyLPA’ to generate the start values for estimation (see below). These differences -->
<!--  can cause the packages to yield different results for more complex models and/or models that -->
<!--  are increasingly misspecified (as the models with more than 3 classes were in our example). -->

<!-- 6. Final comments -->
<!-- In this tutorial we have seen that it is relatively easy to run an LPA in R. As with all -->
<!-- data-driven analyses, care should be taken not to over-interpret the results. In the end, LPA -->
<!-- identifies classes in such a way as to optimally explain variance on a range of variables and -->
<!-- not to optimize interpretability or usefulness. -->
<!-- Another remark with regard to LPA results is that class-membership is probabilistic in nature: -->
<!-- each subject in an analyzed sample has a probability of being in each of the model’s classes. -->
<!-- Subjects can be allocated to a class based on their highest class-probability. Importantly, -->
<!-- this can only be done with enough certainty if separation between classes is sufficient, which -->
<!-- means that we see that each subject has a clear highest probability (e.g., p=0.9) for one of -->
<!-- the classes. The entropy statistic is often used to quantify this separation, with values <0.8 -->
<!-- being taken to indicate insufficient separation to allocate subjects to one class. In such cases, -->
<!-- the class probabilities themselves can still be used in further investigations. -->
<!-- Depending on one’s preferences, one can choose either the ‘mclust’ or the ‘tidyLPA’ package. -->
<!-- The latter may be especially attractive for persons that already have experience with Mplus -->
<!-- or comparable software. It is important to note, though, that differences exist between the -->
<!-- estimation approaches taken by both R-packages and by Mplus. -->
<!-- Mixture models such as LPA need to be estimated in an iterative fashion, using an EM -->
<!-- algorithm that is very sensitive to the used start values, where poor start values can lead the -->
<!-- estimation process to a poor solution. In addition, there is often a chance that the iterative -->
<!-- EM process leads to a solution that is at a local, rather than a global maximum in the -->
<!-- likelihood ‘landscape’. -->

### Starting values

<!-- Different methods have been developed to initialize the estimation (get starting values) in -->
<!-- such a way as to optimize the chance of arriving at an accurate model solution (Biernacki et -->
<!-- al., 2003; Shireman et al., 2016). Mplus and ‘mclust’ take two different approaches. Mplus -->
<!-- uses a ‘brute force’ approach and reruns the model with multiple sets of random start values, -->
<!-- each generated from uniform distributions of values with ranges that are based on the data -->
<!-- (Muthén & Muthén, 1998-2015). The best solution is the model with the highest loglikelihood -->
<!-- that was arrived upon from at least two different starting points. In contrast, ‘mclust’ uses -->
<!-- the data itself to generate a set of plausible start values using hierarchical clustering. This -->
<!-- means that the start values are informed by the (hierarchical structure of) the data, which -->
<!-- can work well. A downside is that only a single set of start values is used, so the risk of -->
<!-- arriving at a local optimum is not addressed. These approaches show clear differences and -->
<!-- have been shown previously to not always arrive at the same solutions, with some authors -->
<!-- being especially critical of the ‘mclust’ approach (Shireman et al., 2017). In our example, -->
<!-- we did not see much difference between the two approaches, but it should be noted that -->
<!-- many real-world data sets do not have such a clear latent structure, making it much more -->
<!-- challenging to arrive at an accurate solution, given the data. -->
<!-- Based on the provided examples and these technical differences between ‘mclust’ and Mplus, -->
<!-- it is probably better to see ‘mclust’ and ‘tidyLPA’ as useful alternatives for Mplus, rather -->
<!-- than as tools to exactly mimic what can be done with the latter software. -->

<!-- References -->
<!-- Banfield, J., Raftery, A.E. (1993). Model-based Gaussian and non-Gaussian clustering. -->
<!-- Biometrics. 49: 803–821. -->
<!-- Biernacki,C.,Celeux,G., Govaert,G.(2003).Choosing starting values for the EM algorithm -->
<!-- for getting the highest likelihood in multivariate Gaussian mixture models. Computational -->
<!-- Studies and Data Analysis, 41, 561–575. -->
<!-- Celeux, G., Govaert, G. (1995) Gaussian parsimonious clustering models. Pattern Recognition. -->
<!-- 28:781–793 -->
<!-- Fraley, C., Raftery, A.E. (1998) How many clusters? Which clustering method? Answers via -->
<!-- model-based cluster analysis. The Computer Journal 41: 578-588. -->
<!-- Gibson, W. A. (1959). Three multivariate models: Factor analysis, latent structure analysis, -->
<!-- and latent profile analysis. Psychometrika, 24, 229–252. -->
<!-- Linzer, D.A., Lewis J. (2011). “poLCA: an R Package for Polytomous Variable Latent Class -->
<!-- Analysis.” Journal of Statistical Software. 42(10): 1-29 -->
<!-- Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., Leisch, F. (2019). e1071: Misc -->
<!-- Functions of the Department of Statistics, Probability Theory Group (Formerly: E1071), TU -->
<!-- Wien. R package version 1.7-3. https://CRAN.R-project.org/package=e1071 -->
<!-- Muthén, L., Muthén, B., 1998–2015. Mplus User’s Guide, 7th ed., Muthén & Muthén, Los -->
<!-- Angeles. -->
<!-- Neale, M.C., Hunter, M.D., Pritikin, J.N., Zahery, M., Brick, T.R., Kirkpatrick, R.M., -->
<!-- Estabrook, R., Bates, T.C., Maes, H.H., Boker, S.M. (2016). “OpenMx 2.0: Extended -->
<!-- structural equation and statistical modeling.” Psychometrika 81(2): 535-549. -->
<!-- Nylund-Gibson, K., Choi, A. Y. (2018). Ten frequently asked questions about latent class -->
<!-- analysis. Translational Issues in Psychological Sciences, 4, 440–461. -->
<!-- Oberski, D. (2016). Mixture models: Latent profile and latent class analysis. In J. Robertson -->
<!-- & M. Kaptein (Eds.), Modern statistical methods for HCI (pp. 275–287). Cham, Switzerland: -->
<!-- Springer Inter-national Publishing. -->

<!--                                              16 -->
<!-- Pastor, D.A., Barron, K.E., Miller, B.J., Davis, S.L. (2007). A latent profile analysis of -->
<!-- college students’ achievement goal orientation. Contemporary Educational Psychology 32(1): -->
<!-- 8-47. -->
<!-- Rosenberg, J.M., Beymer, P.N., Anderson, D.J., Van Lissa, C.J., & Schmidt, J.A. (2018). -->
<!-- tidyLPA: An R Package to Easily Carry Out Latent Profile Analysis (LPA) Using Open-Source -->
<!-- or Commercial Software. Journal of Open Source Software 3(30): 978. -->
<!-- Scrucca, L., Fop M., Murphy, T.B., Raftery A.E. (2016) mclust 5: clustering, classification -->
<!-- and density estimation using Gaussian finite mixture models The R Journal 8/1: 289-317. -->
<!-- Shireman, E. M., Steinley, D., & Brusco, M. J. (2016). Local Optima in Mixture Modeling. -->
<!-- Multivariate Behavioral Research 51(4): 466–481. -->
<!-- Shireman, E., Steinley, D., Brusco, M.J. (2017). Examining the effect of initialization -->
<!-- strategies on the performance of Gaussian mixture modeling. Behavioral Research Methods -->
<!-- 49(1):282-293. -->
<!-- Sterba, S.K. (2013). Understanding Linkages Among Mixture Models. Multivariate Behav -->
<!-- Res. 2013 48(6): 775-815. -->
<!-- Wolfe, J. (1970). Pattern clustering by multivariate mixture analysis. Multivariate Behavioral -->
<!-- Research, 5: 329–350. -->

<!-- Shireman 2016: -->
<!-- https://www.tandfonline.com/doi/full/10.1080/00273171.2016.1160359?casa_token=YMvI52JWPq0AAAAA%3A20nlLRukoCkUmIil2lRWnsUS-33AlU6HmDRGsu07aJ0nTzS-skjvHnroJBtBVHM5HE0X8iStM1eyqw -->

<!-- mixture models have two possible uses: either (a) classifying observations into groups through the estimation of distinct, underlying distributions or (b) using a set of well-defined parametric distributions (often times the Gaussian distribution) to approximate a messy data space that cannot be approximated by standard models that assume a single distribution (whether univariate or multivariate). While both are legitimate uses of mixture modeling, the only necessary and sufficient condition for fitting a mixture model is nonnormality in the data (see Bauer & Curran, 2003). It is certainly true that if there are distinct groups arising from distinct distributions, it is necessary to have a set of parameters to model each of those groups. Unfortunately, it may often be the case (in fact, it is probably likely) we are in the converse situation where the full data space is either nonlinear, asymmetric, or both, and there exist no distinct groups in the population. In this case, the mixture model is capturing the departure from “clean” structure, but the groups themselves are not meaningful in the sense that they are giving rise to distinct subpopulations. As such, their boundaries are going to be amorphous, and their parameters, regardless of their stability, lose meaning in terms of reflecting an underlying data-generative mechanism, making the interpretation of the means, covariances, or a specific individuals' membership an exercise in folly. Obviously, knowing which situation a data set falls in is paramount, and using the prevalence of local optima as a potential indicator is a crucial first step. -->

<!-- The optimal number of latent classes (i.e. identity configurations) was determined using the BIC (Fraley & Raftery, 1998); with lower values indicating a better statistical fit. Solutions were considered to be admissible if they consisted of at least 5% of the sample (Depaoli, 2013), if the entropy (a measure of class separability) exceeded .8 (Wang, Deng, Yang, 2017), and if the solution was theoretically interpretable and qualitatively distinct from the other class solutions (Howard & Hoffman, 2018). -->

### 2022-05-29 manuscript.Rmd
First Changes
Moved Algorithm part below. Added the starting sentence to that paragraph. Wrote a short paragraph on starting values.

@@
 
-Latent class analysis has become quite popular across scientific fields,
-and under a number of different names.
+Latent class analysis has become quite popular across scientific fields,  and under a number of different names.
 <!-- Vermunt and Magidson (2004) defined it more generally as virtually any statistical model where “some of the parameters … differ across unobserved subgroups” (Vermunt, J.K.
@@
 based on a parametric model of one or more observed indicators of group membership.
-Despite the popularity of the method, there is a noted absence of 
-guidelines for estimating and reporting latent class analyses [].
+Despite the popularity of the method, there is a noted absence of guidelines for estimating and reporting latent class analyses [].
 <!-- Prior work resulted in the "GRoLTS-Checklist"; reporting guidelines for a specific subcategory of latent class analysis known as latent growth models. vermuntj.k.LatentClassAnalysis2004-->
@@
 	* See Nylund
-1. Starting values [@nylund-gibsonTenFrequentlyAsked2018; ]
-    In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
-Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
-[@biernackiChoosingStartingValues2003]
-1. Algorithm
-    EM is a local optimizer.
-Imagine we are estimating two parameters, e.g.
-the class-specific means $\mu_c$ on a continuous indicator (ignoring the variance for now).
-The EM algorithm will attempt to find a combination of values for these two parameters that maximizes the likelihood ($LL$) of all observed data.
-In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages.
-We can think of this optimization problem as a three-dimensional landscape: The X and Y dimensions are determined by the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$.
-The optimizer must find the deepest "valley" in this landscape, which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble, dropped in this landscape.
-It is dropped at some random point in space, and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley (this is known as "convergence").
-It cannot climb out again.
-Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley (a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
-One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
-choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
-This is the "random starts" approach.
-
-One problem with the random starts approach is that it is computationally expensive to run this many replications.
-Moreover, because the algorithm begins with random starting values,
-many of the marbles are likely to be very far away from a "good enough" solution.
-Two innovations may improve the estimation procedure.
-The first is that, instead of picking random starting values, a "reasonable solution" may be used for the starting values.
-For example, if we assume that the different classes are likely to have different mean values on the indicators, then the K-means clustering algorithm can be used to determine these cluster centroids.
-We can compute the expected values of all model parameters by treating the K-means solution as a known class solution, and use these as starting values for a mixture model.
-One remaining concern is that this approach may result in starting values close to a local optimum,
-and that the EM algorithm will thus never find the global optimum.
-A second innovation addresses this concern.
-Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some "destination" in the landscape, and compares its likelihood to the current one. If the destination likelihood is higher, the estimator moves there.
-If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
-This latter property allows it to escape local optima.
-
-By default, `tidySEM` eploys this solution of deriving starting values using K-means clustering,
-and identifying the global optimum solution using simulated annealing.
-Once a solution has been found,
-simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for the parameters that can be used to compute standard errors.
-Note that these defaults can be manually overridden.
 
@@
 
+### Algorithm
+
+    Mixture model parameters are estimated through maximum likelihood (ML) estimation with expectation-maximization (EM) as a local optimizer.
+Imagine we are estimating two parameters, e.g.
+the class-specific means $\mu_c$ on a continuous indicator (ignoring the variance for now).
+The EM algorithm will attempt to find a combination of values for these two parameters that maximizes the likelihood ($LL$) of all observed data.
+In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages.
+We can think of this optimization problem as a three-dimensional landscape: The X and Y dimensions are determined by the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$.
+The optimizer must find the deepest "valley" in this landscape, which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
+The EM optimizer behaves somewhat like a marble, dropped in this landscape.
+It is dropped at some random point in space, and will roll into the nearest valley.
+The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley (this is known as "convergence").
+It cannot climb out again.
+Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley (a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
+One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
+choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
+This is the "random starts" approach.
+
+One problem with the random starts approach is that it is computationally expensive to run this many replications.
+Moreover, because the algorithm begins with random starting values,
+many of the marbles are likely to be very far away from a "good enough" solution.
+Two innovations may improve the estimation procedure.
+The first is that, instead of picking random starting values, a "reasonable solution" may be used for the starting values.
+For example, if we assume that the different classes are likely to have different mean values on the indicators, then the K-means clustering algorithm can be used to determine these cluster centroids.
+We can compute the expected values of all model parameters by treating the K-means solution as a known class solution, and use these as starting values for a mixture model.
+One remaining concern is that this approach may result in starting values close to a local optimum,
+and that the EM algorithm will thus never find the global optimum.
+A second innovation addresses this concern.
+Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
+Simulated annealing iteratively considers some "destination" in the landscape, and compares its likelihood to the current one. If the destination likelihood is higher, the estimator moves there.
+If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
+This latter property allows it to escape local optima.
+
+By default, `tidySEM` eploys this solution of deriving starting values using K-means clustering,
+and identifying the global optimum solution using simulated annealing.
+Once a solution has been found,
+simulated annealing is followed up with a short run of the EM algorithm,
+as EM inherently produces an asymptotic covariance matrix for the parameters that can be used to compute standard errors.
+Note that these defaults can be manually overridden.
+
 ### Starting values
 
-### Algorithm
+The algorithm needs starting values for each parameter it wants to estimate.
+If it were provided a single starting value for each parameter, the  EM optimizer would be at risk of finding a local optimum instead of the global optimum.
+In other words, it would converge on a suboptimal solution and the resulting parameter estimates would be biased.
+To prevent this, we should give the algorithm a sufficient set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (Geiser, 2013).
+
+[@nylund-gibsonTenFrequentlyAsked2018; ]
+
+    In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
+Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
+[@biernackiChoosingStartingValues2003]
 

### 2022-06-04 manuscript.Rmd
Changed Spacing
Update manuscript.Rmd
Added to Starting Values
Added Assessing Estimation Results

@@
 
-    Mixture model parameters are estimated through maximum likelihood (ML) estimation with expectation-maximization (EM) as a local optimizer.
+    Mixture model parameters and model fit statistics are estimated by maximum likelihood (ML) estimation with the expectation-maximization (EM) algorithm as a local optimizer.
 Imagine we are estimating two parameters, e.g.
@@
 
-By default, `tidySEM` eploys this solution of deriving starting values using K-means clustering,
+By default, `tidySEM` employs this solution of deriving starting values using K-means clustering,
 and identifying the global optimum solution using simulated annealing.
@@
 
-The algorithm needs starting values for each parameter it wants to estimate.
-If it were provided a single starting value for each parameter, the  EM optimizer would be at risk of finding a local optimum instead of the global optimum.
-In other words, it would converge on a suboptimal solution and the resulting parameter estimates would be biased.
-To prevent this, we should give the algorithm a sufficient set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (Geiser, 2013).
+The EM algorithm needs starting values for each parameter it wants to estimate. 
+If it were provided a single starting value for each parameter, 
+the EM optimizer would be at risk of finding a local optimum instead of the global optimum. 
+In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates. 
+To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2000).
+Furthermore, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2013).
+
+Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2000).
+The key is to use a large number of starting value sets.
+Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, and a greater number of sets for more complex models.
+A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately.
 
@@
 [@biernackiChoosingStartingValues2003]
+
+### Assessing Estimation Results
+
+Each set of starting values will result in a model with a log likelihood value.
+Ideally, the largest log likelihood value will be replicated a large number of times (each time is generated by a different starting value set).
+Successful replications of the largest log likelihood value across different starting values instill confidence that the estimation was successful.
+The model with the highest replicated likelihood will be chosen as our best-fit.
+If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2013).
+
+[to add: Nylund-Gibson & Choi  p.22. about 3-10% and Geiser p. 237 -240]
 

### 2022-06-05 manuscript.Rmd
Update manuscript.Rmd
Added to Starting Values
Added Assessing Estimation Results

@@
 To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2000).
-Furthermore, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2013).
+Furthermore, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
 
@@
 Each set of starting values will result in a model with a log likelihood value.
-Ideally, the largest log likelihood value will be replicated a large number of times (each time is generated by a different starting value set).
+Ideally, the largest log likelihood value will be replicated a large number of times with each log likelihood value generated by a different starting value set.
 Successful replications of the largest log likelihood value across different starting values instill confidence that the estimation was successful.
 The model with the highest replicated likelihood will be chosen as our best-fit.
-If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2013).
-
-[to add: Nylund-Gibson & Choi  p.22. about 3-10% and Geiser p. 237 -240]
+If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012).
+Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated in at least 3-10% of the models initialized with different starting values in the first step of the optimization.
+
+Another consideration has to do with conditional item probabilities which indicate the probability of a positive score on an item given that the person belongs to a particular latent class.
+If many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1), we are observing boundary parameter estimates.
+This can be a sign of an invalid solution, warn us that too many classes were extracted, or indicate a local optimum.
+Boundary parameter estimates are more likely to happen when working with sparse data and should be interpreted with caution (Geiser, 2012).
 

### 2022-06-05 manuscript.Rmd
Update manuscript.Rmd
Adjusting a citation

@@
 In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates. 
-To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2000).
+To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004).
 Furthermore, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
 
-Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2000).
+Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004).
 The key is to use a large number of starting value sets.

### 2022-06-19 manuscript.Rmd
Update manuscript.Rmd
Added a paragraph under ### Assessing Estimation Results

@@
 To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004).
-Furthermore, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
+Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
 
@@
 
+An important outcome of LCA are unconditional class probabilities.
+These communicate the probability that an individual belongs to each class.
+A limitation is that our final class solution need not reflect true class membership.
+Once we do accept a particular class solution, special care also needs to be taken when naming the classes.
+Class names should be chosen to accurately reflect group membership.
+Overly simplified and generalized class names may prove misleading to both audiences and researches alike leading to what is known as a naming fallacy (Weller, Bowen & Faubert, 2020).
+
 ### Label switching
@@
 ## Best practices in reporting
+
+Among studies using LCA, reporting practices vary significantly (Weller, Bowen & Faubert, 2020).
 

### 2022-06-26 manuscript.Rmd
Update manuscript.Rmd
Adjusted spacing of some paragraphs for ease of reading on GitHub.

Changes begin under "Assessing Estimation". Added the part about information criteria and model comparison using elbow plots.

@@
 
-    Mixture model parameters and model fit statistics are estimated by maximum likelihood (ML) estimation with the expectation-maximization (EM) algorithm as a local optimizer.
-Imagine we are estimating two parameters, e.g.
-the class-specific means $\mu_c$ on a continuous indicator (ignoring the variance for now).
-The EM algorithm will attempt to find a combination of values for these two parameters that maximizes the likelihood ($LL$) of all observed data.
+Mixture model parameters and model fit statistics can be estimated in a variety of ways.
+The choice of the estimator depends on the presence of missing values,
+sample size, number of indicators, and available computational resources (Weller, Bowen & Faubert, 2020).
+A commonly used technique is maximum likelihood (ML) estimation
+with the expectation-maximization (EM) algorithm as a local optimizer. 
+Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator
+(ignoring the variance for now). The EM algorithm will attempt to find a combination of values
+for these two parameters that maximizes the likelihood ($LL$) of all observed data. 
 In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. 
-We can think of this optimization problem as a three-dimensional landscape: The X and Y dimensions are determined by the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$.
-The optimizer must find the deepest "valley" in this landscape, which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
+We can think of this optimization problem as a three-dimensional landscape: 
+The X and Y dimensions are determined by the class-specific means, 
+so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. 
+The optimizer must find the deepest "valley" in this landscape, 
+which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
 The EM optimizer behaves somewhat like a marble, dropped in this landscape.
 It is dropped at some random point in space, and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley (this is known as "convergence").
+The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
+(this is known as "convergence").
 It cannot climb out again.
-Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley (a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
+Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley 
+(a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
 One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
@@
 
-One problem with the random starts approach is that it is computationally expensive to run this many replications.
+One problem with the random starts approach is that 
+it is computationally expensive to run this many replications.
 Moreover, because the algorithm begins with random starting values,
@@
 Two innovations may improve the estimation procedure.
-The first is that, instead of picking random starting values, a "reasonable solution" may be used for the starting values.
-For example, if we assume that the different classes are likely to have different mean values on the indicators, then the K-means clustering algorithm can be used to determine these cluster centroids.
-We can compute the expected values of all model parameters by treating the K-means solution as a known class solution, and use these as starting values for a mixture model.
+The first is that, instead of picking random starting values, 
+a "reasonable solution" may be used for the starting values.
+For example, if we assume that the different classes are likely to have different mean values on the indicators, 
+then the K-means clustering algorithm can be used to determine these cluster centroids.
+We can compute the expected values of all model parameters by treating the K-means solution as a known class solution,
+and use these as starting values for a mixture model.
 One remaining concern is that this approach may result in starting values close to a local optimum,
@@
 A second innovation addresses this concern.
+
 Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some "destination" in the landscape, and compares its likelihood to the current one. If the destination likelihood is higher, the estimator moves there.
+Simulated annealing iteratively considers some "destination" in the landscape, 
+and compares its likelihood to the current one. If the destination likelihood is higher, 
+the estimator moves there.
 If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
@@
 In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates. 
-To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004).
+To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter
+in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004). 
 Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
 
-Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004).
+Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence
+or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004).
 The key is to use a large number of starting value sets. 
-Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, and a greater number of sets for more complex models.
+Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, 
+and a greater number of sets for more complex models. 
 A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately.
@@
 
-Each set of starting values will result in a model with a log likelihood value.
-Ideally, the largest log likelihood value will be replicated a large number of times with each log likelihood value generated by a different starting value set.
-Successful replications of the largest log likelihood value across different starting values instill confidence that the estimation was successful.
+In LCA, a sequence of models is fitted to the data with each additional model having one more class than the previous model.
+The final model called the final class solution is chosen based on both theoretical and statistical criteria.
+Theory should inform the choice of indicator variables ... 
+With regards to statistical criteria, model fit is often assessed through multiple information criteria.
+Examples include the Akaike information criterion (AIC), the Consistent Akaike Information Criterion (CAIC),
+the Bayesian information criterion (BIC), and the Sample-size Adjusted Bayesian Information Criterion (SABIC). 
+As a general rule, the lower the value of an information criterion, the better the model fits the data.
+Multiple information criteria should be compared when choosing the final class solution. 
+This can be done by means of an elbow plot (for an example see Nylund-Gibson and Choi, 2018).
+Other statistical criteria include the Bayes factor and likelihood ratio tests
+which can also be used to assess model fit (Weller, Bowen & Faubert, 2020).
+
+When estimators which require sets of starting values are used,
+each fitted model will have a corresponding log likelihood value. 
+Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times 
+with each log likelihood value generated by a different starting value set. 
+Successful replications of the largest log likelihood value across different starting values
+instill confidence that the estimation was successful. 
 The model with the highest replicated likelihood will be chosen as our best-fit. 
 If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012). 
-Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated in at least 3-10% of the models initialized with different starting values in the first step of the optimization.
-
-Another consideration has to do with conditional item probabilities which indicate the probability of a positive score on an item given that the person belongs to a particular latent class.
-If many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1), we are observing boundary parameter estimates.
-This can be a sign of an invalid solution, warn us that too many classes were extracted, or indicate a local optimum.
-Boundary parameter estimates are more likely to happen when working with sparse data and should be interpreted with caution (Geiser, 2012).
+Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated 
+in at least 3-10% of the models initialized with different starting values in the first step of the optimization. 
+This consideration does not apply in case of models which use simulated annealing since these do not have starting values.
+
+Once we have selected the final class solution, 
+<!-- Continue the paragraph with info about classification diagnostics, e.g. the average latent class posterior probability. -->
+
+Another consideration has to do with conditional item probabilities which indicate 
+the probability of a positive score on an item given that the person belongs to a particular latent class. 
+If many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1), 
+we are observing boundary parameter estimates. This can be a sign of an invalid solution, 
+warn us that too many classes were extracted, or indicate a local optimum. 
+Boundary parameter estimates are more likely to happen when working with sparse data 
+and should be interpreted with caution (Geiser, 2012).
 
@@
 Class names should be chosen to accurately reflect group membership. 
-Overly simplified and generalized class names may prove misleading to both audiences and researches alike leading to what is known as a naming fallacy (Weller, Bowen & Faubert, 2020).
+Overly simplified and generalized class names may prove misleading to both audiences and researches alike
+leading to what is known as a naming fallacy (Weller, Bowen & Faubert, 2020).
 

### 2022-07-10 manuscript.Rmd
Week 27 June and Week 4 July
Added content to Assessing Estimation Results and to Label Switching

@@
 in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004). 
+Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values.
+<!-- add more on this -->
 Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
@@
 
-In LCA, a sequence of models is fitted to the data with each additional model having one more class than the previous model.
+In LCA, a sequence of models is fitted to the data 
+with each additional model estimating one more class than the previous model.
 The final model called the final class solution is chosen based on both theoretical and statistical criteria.
-Theory should inform the choice of indicator variables ... 
-With regards to statistical criteria, model fit is often assessed through multiple information criteria.
+Theory should drive the selection of indicator variables, inform the expectations and reflect on the findings.
+There are several statistical criteria to consider in model selection.
+These include but are not limited to information criteria, likelihood ratio tests, 
+and the Bayes factor (Weller, Bowen & Faubert, 2020).
+
+Model fit can be assessed through a comparison of multiple information criteria.
 Examples include the Akaike information criterion (AIC), the Consistent Akaike Information Criterion (CAIC),
@@
 As a general rule, the lower the value of an information criterion, the better the model fits the data.
+We expect the information criteria values to drop with each successive model until the final class solution is reached
+Further models with additional classes should show worse fit. 
+Therefore, information criteria reach an optimum before their values start to rise with increasing number of classes.
 Multiple information criteria should be compared when choosing the final class solution. 
 This can be done by means of an elbow plot (for an example see Nylund-Gibson and Choi, 2018).
-Other statistical criteria include the Bayes factor and likelihood ratio tests
-which can also be used to assess model fit (Weller, Bowen & Faubert, 2020).
+Other statistical criteria to assess model fit include the likelihood ratio test and the Bayes factor (Weller, Bowen & Faubert, 2020).
+
+The likelihood ratio (LR) $\chi^2$ test is commonly used to assess goodness-of-fit. 
+It can be used to compare two nested models.
+Two models are said to be nested when one model is based on a subset of indicators of the other model.
+Typically, we compare our final class solution, a constrained model to an unconstrained one.
+The perfect fit would mean the $\chi^2\ =\ 0$ and $p\ =\ 1$.
+We want to the LR test to be non-significant as this implies
+that our model is more parsimonious yet it fits the data as well as the unconstrained model.
+A downside of the LR $\chi^2$ test statistic is that it is highly dependent on sample size (Masyn, 2013).
+Large samples will often effortlessly to reject the null even when the final class solution is a good fit,
+and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic.
+Special caution is advised when interpreting p-values derived from models with small sample sizes 
+or sparse data (Masyn, 2013).
+<!-- A potential solution to small sample size biased p values: parametric bootstrapping and posterior predictive checking see Collins & Lanza cited in Masyn 2013) -->
+
+<!-- Add a paragraph about using the Bayes factor -->
 
@@
 Successful replications of the largest log likelihood value across different starting values
-instill confidence that the estimation was successful. 
+instill confidence that the estimation was successful and we indeed found the global maximum (Masyn, 2013). 
 The model with the highest replicated likelihood will be chosen as our best-fit. 
 If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012).
+Such a model does not have a unique solution. For a discussion on model identification for LCA, see Masyn (2013).
 Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated 
@@
 This consideration does not apply in case of models which use simulated annealing since these do not have starting values.
-
-Once we have selected the final class solution, 
+<!-- Masyn 2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
+<!-- to ensure that a global (rather than a local) maximum has been reached.". For discussion see Hipp & Bauer, 2006 -->
+
+Once we have selected the final class solution, we have to inspect classification diagnostics.
+Classification diagnostics answer the question of how accurate
+
+Entropy is a measure of accuracy with regards to modeled classes. 
+Values above .80 and close to 1 are considered ideal.
+
+The average latent class posterior probability [see Masyn, 2013]
 <!-- Continue the paragraph with info about classification diagnostics, e.g. the average latent class posterior probability. -->
 
-Another consideration has to do with conditional item probabilities which indicate 
-the probability of a positive score on an item given that the person belongs to a particular latent class. 
-If many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1), 
+Conditional item probabilities indicate the probability of a positive score on an item 
+given that the person belongs to a particular latent class. 
+<!-- If a particular item is endorsed by two or more classes at markedly different rates, -->
+<!-- it is said to discriminate well between classes and is consequently considered a good indicator. -->
+<!-- The classes are considered highly homogeneous with respect to an item when for a particular item -->
+<!-- there are a distinct difference in conditional item probabilities of two or more classes. -->
+<!-- If an item is endorsed below 30% for one class and above 70% for another class, 
+the classes have high homogeneity with respect to this item (Masyn, 2013). MY OWN WRITING BUT NOT FLOWY ENOUGH-->
+However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1), 
 we are observing boundary parameter estimates. This can be a sign of an invalid solution, 
@@
 ### Label switching
+
+The final class solution will usually discover and enumerate several classes. 
+The class ordering however is completely arbitrary.
+The class labeled as Class 1 in one solution may become Class 2 or Class 3 in another model,
+even when the only difference between the models is in their starting values. 
+Label switching is something to be mindful of when comparing different LCA models (Masyn, 2013).
+
+<!-- Article discussing label switching: see Chung, Loken & Schafer -->
 

### 2022-07-25 manuscript.Rmd

@@
 ### Assessing Estimation Results
+<!-- maybe change to class enumeration -->
 
@@
 Theory should drive the selection of indicator variables, inform the expectations and reflect on the findings.
-There are several statistical criteria to consider in model selection.
-These include but are not limited to information criteria, likelihood ratio tests, 
+In addition to this, there are several statistical criteria to consider in model selection.
+These include but are not limited to likelihood ratio tests, information criteria, 
 and the Bayes factor (Weller, Bowen & Faubert, 2020).
 
-Model fit can be assessed through a comparison of multiple information criteria.
+Absolute model fit can be assessed using the likelihood ratio (LR) $\chi^2$ goodness-of-fit test. 
+It can compare two nested models when they have the same number of classes (Lanza et al., 2003).
+Typically, we compare our final class solution (a constrained model) 
+to an unconstrained one (a model characterized by the perfect fit).
+The former model is a special, less complex version of the latter.
+In this case, we compare our model-based response pattern frequencies to those in the dataset.
+Therefore, we make a statement about how well our model fits the data.
+The perfect fit would be indicated by the $\chi^2\ =\ 0$ and $p\ =\ 1$.
+Ideally, the LR $\chi^2$ test will be non-significant as this implies
+that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model.
+The underlying assumption of this procedure is that the test statistic is asymptotically distributed
+as a $\chi^2$ meaning that it can be approximated using this distribution.
+This is the case when the sample size is adequate (Lanza et al., 2003). 
+The dependence of the LR $\chi^2$ test statistic on sample size has its downsides (Masyn, 2013). 
+Very large samples will often effortlessly reject the null even when the final class solution is a good fit,
+and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic.
+In fact, small samples might not be well approximated using the $\chi^2$ distribution,
+which is a problem shared with large yet sparse datasets.
+In such cases, special caution is advised when interpreting the resulting p-values (Masyn, 2013).
+
+Relative model fit can be examined using the likelihood ratio test. 
+This is only appropriate when the two models we wish to compare are nested.
+The likelihood ratio test statistic is computed as the difference in maximum log likelihoods of the two models,
+with the new degrees of freedom being the difference in their degrees of freedom.
+This statistic also follows the $\chi^2$ distribution.
+Similar to the LR $\chi^2$ goodness-of-fit test, we want the test statistic to be non-significant
+in order to give support to the simpler model. 
+The likelihood ratio test can only compare two models at a time (Lanza et al., 2003).
+
+If we wish to simultaneously compare multiple models based on their relative fit,
+this can be done through a comparison of multiple information criteria.
 Examples include the Akaike information criterion (AIC), the Consistent Akaike Information Criterion (CAIC),
 the Bayesian information criterion (BIC), and the Sample-size Adjusted Bayesian Information Criterion (SABIC).
+Information criteria are a sum of a measure of fit (usually a form of the converged maximum log likelihood value) 
+and a penalty for model complexity (combination of sample size and number of modeled parameters).
 As a general rule, the lower the value of an information criterion, the better the model fits the data.
-We expect the information criteria values to drop with each successive model until the final class solution is reached
-Further models with additional classes should show worse fit. 
-Therefore, information criteria reach an optimum before their values start to rise with increasing number of classes.
+When examining several competing models, each with differing number of classes,
+we expect the information criteria values to drop with each successive model until the final class solution is reached.
+Further models with additional classes should show worse fit as information criteria reach an optimum 
+before their values rise again with an increasing number of classes.
 Multiple information criteria should be compared when choosing the final class solution. 
 This can be done by means of an elbow plot (for an example see Nylund-Gibson and Choi, 2018).
-Other statistical criteria to assess model fit include the likelihood ratio test and the Bayes factor (Weller, Bowen & Faubert, 2020).
-
-The likelihood ratio (LR) $\chi^2$ test is commonly used to assess goodness-of-fit. 
-It can be used to compare two nested models.
-Two models are said to be nested when one model is based on a subset of indicators of the other model.
-Typically, we compare our final class solution, a constrained model to an unconstrained one.
-The perfect fit would mean the $\chi^2\ =\ 0$ and $p\ =\ 1$.
-We want to the LR test to be non-significant as this implies
-that our model is more parsimonious yet it fits the data as well as the unconstrained model.
-A downside of the LR $\chi^2$ test statistic is that it is highly dependent on sample size (Masyn, 2013).
-Large samples will often effortlessly to reject the null even when the final class solution is a good fit,
-and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic.
-Special caution is advised when interpreting p-values derived from models with small sample sizes 
-or sparse data (Masyn, 2013).
-<!-- A potential solution to small sample size biased p values: parametric bootstrapping and posterior predictive checking see Collins & Lanza cited in Masyn 2013) -->
-
-<!-- Add a paragraph about using the Bayes factor -->
-
+Advantages of using information criteria are that we can compare multiple models at a time,
+and these models need not be nested (Masyn, 2013).
+
+<!-- Add a paragraph about using the Bayes factor see Weller, Bowen & Faubert, 2020-->
+
+<!-- decide whether to keep the following paragraph, if so, where to place it -->
 When estimators which require sets of starting values are used,
@@
 
-Once we have selected the final class solution, we have to inspect classification diagnostics.
-Classification diagnostics answer the question of how accurate
-
-Entropy is a measure of accuracy with regards to modeled classes. 
-Values above .80 and close to 1 are considered ideal.
-
-The average latent class posterior probability [see Masyn, 2013]
-<!-- Continue the paragraph with info about classification diagnostics, e.g. the average latent class posterior probability. -->
+### Classification Diagnostics
+
+Best models will divide the sample into subgroups which are internally homogeneous and externally distinct.
+Classification diagnostics give us a way to assess the degree to which this is the case.
+They are separate from the absolute and relative goodness-of-fit 
+as a model can fit the data well but show poor latent class separation (Masyn, 2013).
+A fundamental concept when examining classification precision and accuracy are the posterior class probabilities.
+A probability of belonging to each latent class is computed for each individual.
+The highest posterior class probability is then determined and the individual is assigned to the corresponding class.
+We want each individual's posterior class probabilities to be high for one and low for the remaining latent classes.
+This is considered a high classification accuracy and means that the classes are distinct.
+<!-- see if you find content on homogeneity of classes -->
+Three important classification diagnostic measures are entropy, the average posterior probability and the modal class assignment proportion.
+
+Entropy is a summary measure of posterior class probabilities across classes and individuals.
+It ranges from 0 (model classification no better than random chance) to 1 (perfect classification).
+As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
+<!-- Masyn 2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
+<!-- add content about entropy -->
+
+The average posterior probability is a measure of classification uncertainty for each latent class.
+
+The modal class assignment proportion
 
@@
 Among studies using LCA, reporting practices vary significantly (Weller, Bowen & Faubert, 2020).
-
-The most relevant
-
-1. Open science
+Various authors have tried to better and standardize ways of reporting LCA (e.g. Masyn, 2013; Weller, Bowen & Faubert, 2020). 
+Building on their work, we provide guidelines with a strong emphasis on scientific reproducibility and transparency
+better known as the open science framework (OSF). Open science framework promotes scientific openness, reproducibility and integrity 
+through an establishment of guidelines for the scientific process (Foster & Deardorff, 2017)
+
 	* Use comprehensive citation; this includes referencing the software used.
@@
 <!-- where df is the number of parameters (degrees of freedom) and n is the sample size. We can -->
-<!-- see, that here, higher loglikelihood values lead to a higher BIC. For the best fitting model in -->
-<!-- our example, the loglikelihood is -5951.275, the number of parameters is 42 and the sample -->
+<!-- see, that here, higher log likelihood values lead to a higher BIC. For the best fitting model in -->
+<!-- our example, the log likelihood is -5951.275, the number of parameters is 42 and the sample -->
 <!-- size is 300. If we plug these into the formula, we get the following BIC: -->
@@
 <!-- (see above). Other software packages, such as Mplus and ‘tidyLPA’ (see below) use a slightly -->
-<!-- different formula to calculate the BIC, where higher loglikelihood values lead to lower BIC -->
+<!-- different formula to calculate the BIC, where higher log likelihood values lead to lower BIC -->
 <!-- values: -->
@@
 <!-- fit indices (e.g., BIC, AIC) directly from the ‘mclust’ package, ‘tidyLPA’ only draws the ‘raw’ -->
-<!-- loglikelihood values and posterior probabilities from the ‘mclust’ output and (re)calculates -->
+<!-- log likelihood values and posterior probabilities from the ‘mclust’ output and (re)calculates -->
 <!-- the BIC, AIC, entropy etc. to mirror as well as possible those provided in Mplus. In addition, -->
@@
 <!-- each generated from uniform distributions of values with ranges that are based on the data -->
-<!-- (Muthén & Muthén, 1998-2015). The best solution is the model with the highest loglikelihood -->
+<!-- (Muthén & Muthén, 1998-2015). The best solution is the model with the highest log likelihood -->
 <!-- that was arrived upon from at least two different starting points. In contrast, ‘mclust’ uses -->

### 2022-08-01 manuscript.Rmd
Update manuscript.Rmd

@@
 
+In the case of Latent Profile Analysis, it has been suggested that when tests of multivariate normality indicate a violation, 
+maximum likelihood with robust standard errors should be used (Spurk et al., 2020). 
+This statement is problematic for two reasons. 
+Firstly, multivariate normality will always be violated in mixture models by definition. 
+We assume that the population consists of several groups each characterized by its own normal distribution. 
+When two or more normal distributions with different means and standard deviations comprise a mixed population distribution, 
+this combined distribution is no longer normal. 
+For example, you can imagine a bimodal distribution when dealing with two latent classes. 
+If this were not the case and our mixed population distribution were normal, 
+there would be no latent classes to extract as the population would consists of a single class only. 
+For this reason, multivariate normality will always be violated when the population is comprised of several latent groups. 
+Secondly, the claim that maximum likelihood with robust standard errors should be preferred 
+implicitly implies that only software packages with this estimation technique can be used.
+
 ### Starting values
@@
 As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
+SHould not be used to select a particural class solution i fclass separability is too low
+if the group is too small to be meaningful or to calculate descriptive statistics
+could be used to disqualify certain solutions
 <!-- Masyn 2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
@@
 through an establishment of guidelines for the scientific process (Foster & Deardorff, 2017)
+To this end, van Lissa and colleagues (2020) developed WORCS, a workflow for open reproducible code in science.
+WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by Nosek and colleagues (2015).
+It can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
+version control, citation and formatting, among others (Van Lissa et al., 2020)
 

### 2022-08-01 manuscript.Rmd

@@
 As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
-SHould not be used to select a particural class solution i fclass separability is too low
-if the group is too small to be meaningful or to calculate descriptive statistics
-could be used to disqualify certain solutions
+Entropy should not be used to select a particular class solution. 
+An appropriate use of entropy is that it can disqualify certain solutions if class separability is too low 
+or if one of the latent classes is too small to be meaningful or to calculate descriptive statistics.
+
 <!-- Masyn 2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->

### 2022-08-01 manuscript.Rmd
add rstudio project, move open science paragraph, fix shoesize

@@
 library(worcs)
+run_everything = FALSE
 ```
@@
 Throughout this paper, examples make use of free, open source software for latent class analysis in R.
-As the open science movement is gaining momentum,
-researchers are becoming increasingly aware
-how important it is that analyses can be reproduced and audited.
-In line with open science principles, one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., *Mplus* and *Latent GOLD*).
-A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because the underlying source code is proprietary.
-To overcome these limitations, the present paper introduces new user-friendly functions in the `tidySEM` R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package `OpenMx`.
-The reporting guidelines described in this paper are adopted in `tidySEM` by default.
-The `tidySEM` R-package thus makes advanced mixture modeling based on best practices widely accessible,
-and facilitates the adoption of the estimation and reporting guidelines described in this paper.
 
@@
 
-```{r shoedens, echo = FALSE, fig.cap="Kernel density plot of shoe sizes."}
+```{r echo = FALSE}
 library(tidySEM)
@@
 obsparam <- tapply(X, C, mean)
+```
+
+```{r echo = FALSE, eval = run_everything}
 est <- mx_profiles(data.frame(X), 2)
 estparam <- table_results(est, c("label", "est", "matrix"))
-estparam <- estparam$est[estparam$matrix == "M"]
-props <- class_prob(est, "individual")$individual
-props <- apply(props, 1, which.max)
-
-ggplot(data.frame(Shoesize = X), aes(x = Shoesize)) + geom_density() + theme_bw()
+estparam <- as.numeric(estparam$est[estparam$matrix == "M"])
+props <- class_prob(est, "individual")$individual[ , "predicted"]
+dput(props, "props.txt")
+dput(estparam, "estparam.txt")
+
+p <- ggplot(data.frame(Shoesize = X), aes(x = Shoesize)) + geom_density() + theme_bw()
+ggsave("shoedens.png", p, device = "png")
+```
+
+```{r shoedens, echo = FALSE, fig.cap="Kernel density plot of shoe sizes."}
+props <- eval(parse("props.txt"))
+estparam <- eval(parse("estparam.txt"))
+knitr::include_graphics("shoedens.png")
 ```
@@
 	* Ideally, make the entire research project reproducible so that others may download it and reproduce it with just one click; for guidance, see Van Lissa and colleagues.
+
+<!-- CJ: I think this belongs here -->
+As the open science movement is gaining momentum,
+researchers are becoming increasingly aware
+how important it is that analyses can be reproduced and audited.
+In line with open science principles, one of the suggested reporting standards relates to reproducible code.
+In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., *Mplus* and *Latent GOLD*).
+A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because the underlying source code is proprietary.
+To overcome these limitations, the present paper introduces new user-friendly functions in the `tidySEM` R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package `OpenMx`.
+The reporting guidelines described in this paper are adopted in `tidySEM` by default.
+The `tidySEM` R-package thus makes advanced mixture modeling based on best practices widely accessible,
+and facilitates the adoption of the estimation and reporting guidelines described in this paper.
 

### 2022-08-01 manuscript.tex
add rstudio project, move open science paragraph, fix shoesize

--- 2022-08-01 manuscript.Rmd
@@
----
-title             : "Best Practices in Mixture Modeling using Free Open Source Software"
-shorttitle        : "BEST PRACTICES MIXTURE MODELING"
-
-author: 
-  - name          : "Caspar J. van Lissa"
-    affiliation   : "1,2"
-    corresponding : yes    # Define only one corresponding author
-    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
-    email         : "c.j.vanlissa@gmail.com"
-
-affiliation:
-  - id            : "1"
-    institution   : "Utrecht University, Methodology & Statistics"
-  - id            : "2"
-    institution   : "Open Science Community Utrecht"
-
-author_note: |
-  Methodology & Statistics
-  Padualaan 14
-  3582CH Utrecht, The Netherlands
-
-abstract: |
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+%
+\documentclass[
+  ,man]{apa6}
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\hypersetup{
+  pdftitle={Best Practices in Mixture Modeling using Free Open Source Software},
+  pdfauthor={Caspar J. van Lissa1,2},
+  pdflang={en-EN},
+  pdfkeywords={keywords},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+\urlstyle{same} % disable monospaced font for URLs
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\ifLuaTeX
+\usepackage[bidi=basic]{babel}
+\else
+\usepackage[bidi=default]{babel}
+\fi
+\babelprovide[main,import]{english}
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
+% Manuscript styling
+\usepackage{upgreek}
+\captionsetup{font=singlespacing,justification=justified}
+
+% Table formatting
+\usepackage{longtable}
+\usepackage{lscape}
+% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
+\usepackage{multirow}		% Table styling
+\usepackage{tabularx}		% Control Column width
+\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
+\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+
+% Create new environments so endfloat can handle them
+% \newenvironment{ltable}
+%   {\begin{landscape}\centering\begin{threeparttable}}
+%   {\end{threeparttable}\end{landscape}}
+\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
+
+% Enables adjusting longtable caption width to table width
+% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
+\makeatletter
+\newcommand\LastLTentrywidth{1em}
+\newlength\longtablewidth
+\setlength{\longtablewidth}{1in}
+\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
+
+% \setlength{\parindent}{0.5in}
+% \setlength{\parskip}{0pt plus 0pt minus 0pt}
+
+% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
+% See https://github.com/crsh/papaja/issues/292
+\makeatletter
+\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-1em}%
+  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
+
+\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-\z@\relax}%
+  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
+\makeatother
+
+% \usepackage{etoolbox}
+\makeatletter
+\patchcmd{\HyOrg@maketitle}
+  {\section{\normalfont\normalsize\abstractname}}
+  {\section*{\normalfont\normalsize\abstractname}}
+  {}{\typeout{Failed to patch abstract.}}
+\patchcmd{\HyOrg@maketitle}
+  {\section{\protect\normalfont{\@title}}}
+  {\section*{\protect\normalfont{\@title}}}
+  {}{\typeout{Failed to patch title.}}
+\makeatother
+
+\usepackage{xpatch}
+\makeatletter
+\xapptocmd\appendix
+  {\xapptocmd\section
+    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
+    {}{\InnerPatchFailed}%
+  }
+{}{\PatchFailed}
+\keywords{keywords\newline\indent Word count: X}
+\DeclareDelayedFloatFlavor{ThreePartTable}{table}
+\DeclareDelayedFloatFlavor{lltable}{table}
+\DeclareDelayedFloatFlavor*{longtable}{table}
+\makeatletter
+\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
+\makeatother
+\usepackage{lineno}
+
+\linenumbers
+\usepackage{csquotes}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+
+\title{Best Practices in Mixture Modeling using Free Open Source Software}
+\author{Caspar J. van Lissa\textsuperscript{1,2}}
+\date{}
+
+\shorttitle{BEST PRACTICES MIXTURE MODELING}
+
+\authornote{
+
+Correspondence concerning this article should be addressed to Caspar J.
+van Lissa, Padualaan 14, 3584CH Utrecht, The Netherlands.
+E-mail: \href{mailto:c.j.vanlissa@gmail.com}{\nolinkurl{c.j.vanlissa@gmail.com}}
+
+}
+
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht}
+
+\abstract{%
 Latent class analysis is a popular technique for identifying groups in data based
@@
 introduces new functionality for estimating and reporting mixture mixture
-  models in the `tidySEM` R-package. These functions rely on estimation using
+models in the \texttt{tidySEM} R-package. These functions rely on estimation using
 the OpenMx R-package.
-
-keywords          : "keywords"
-wordcount         : "X"
-
-figsintext        : no
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-lineno            : yes
-mask              : no
-
-class             : "man"
-output            : papaja::apa6_pdf
----
-
-```{r load_packages, include = FALSE}
-library("papaja")
-#bibliography      : ["r-references.bib"]
-library(worcs)
-```
+}
+
+\begin{document}
+\maketitle
 
 Latent class analysis has become quite popular across scientific fields, and under a number of different names.
-<!-- Vermunt and Magidson (2004) defined it more generally as virtually any statistical model where “some of the parameters … differ across unobserved subgroups” (Vermunt, J.K.
-and Magidson, J., 2004.
-Latent class analysis.
-In: M.S.
-Lewis-Beck, A.
-Bryman, and T.F.
-Liao (eds.), The Sage Encyclopedia of Social Sciences Research Methods, 549-553.
-Thousand Oaks, CA: Sage Publications).
--->
 The purpose of this technique is to estimate unobserved group membership
 based on a parametric model of one or more observed indicators of group membership.
-Despite the popularity of the method, there is a noted absence of guidelines for estimating and reporting latent class analyses [].
-<!-- Prior work resulted in the "GRoLTS-Checklist"; reporting guidelines for a specific subcategory of latent class analysis known as latent growth models. vermuntj.k.LatentClassAnalysis2004-->
-<!-- General reporting guidelines for latent class analysis are still lacking, however. -->
+Despite the popularity of the method, there is a noted absence of guidelines for estimating and reporting latent class analyses {[}{]}.
 This complicates manuscript review and assessment of the quality of published studies,
@@
 Throughout this paper, examples make use of free, open source software for latent class analysis in R.
-As the open science movement is gaining momentum,
-researchers are becoming increasingly aware
-how important it is that analyses can be reproduced and audited.
-In line with open science principles, one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., *Mplus* and *Latent GOLD*).
-A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because the underlying source code is proprietary.
-To overcome these limitations, the present paper introduces new user-friendly functions in the `tidySEM` R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package `OpenMx`.
-The reporting guidelines described in this paper are adopted in `tidySEM` by default.
-The `tidySEM` R-package thus makes advanced mixture modeling based on best practices widely accessible,
-and facilitates the adoption of the estimation and reporting guidelines described in this paper.
-
-<!-- In practice, this technique is often used to  -->
-
-## Defining latent class analysis
+
+\hypertarget{defining-latent-class-analysis}{%
+\subsection{Defining latent class analysis}\label{defining-latent-class-analysis}}
 
@@
 Generally speaking, it can be said that a mixture model assumes that the study population
-is composed of $K$ subpopulations, or classes.
+is composed of \(K\) subpopulations, or classes.
 It further assumes that the observed data are a mixture of data generated by class-specific models.
-The simplest univariate "model" is a normal distribution, which can be described with two parameters:
+The simplest univariate ``model'' is a normal distribution, which can be described with two parameters:
 the mean and the variance.
@@
 
-```{r shoedens, echo = FALSE, fig.cap="Kernel density plot of shoe sizes."}
-library(tidySEM)
-library(ggplot2)
-set.seed(1)
-n = 100
-C <- sample(c("Man", "Woman"), n, replace = TRUE)
-means <- c(Man = 9, Woman = 7)
-X <- rnorm(n, mean = means[C], sd = 1)
-X <- round(X*2)/2
-obsparam <- tapply(X, C, mean)
-est <- mx_profiles(data.frame(X), 2)
-estparam <- table_results(est, c("label", "est", "matrix"))
-estparam <- estparam$est[estparam$matrix == "M"]
-props <- class_prob(est, "individual")$individual
-props <- apply(props, 1, which.max)
-
-ggplot(data.frame(Shoesize = X), aes(x = Shoesize)) + geom_density() + theme_bw()
-```
+\begin{figure}
+\includegraphics[width=7in]{shoedens} \caption{Kernel density plot of shoe sizes.}\label{fig:shoedens}
+\end{figure}
 
@@
 When estimating a two-class mixture model, the detective observes that the model estimates
-the mean shoesize of the two groups are equal to `r report(estparam[1], equals = FALSE)` and  `r report(estparam[2], equals = FALSE)`,
-which is close to the true means of the two groups, namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
+the mean shoesize of the two groups are equal to 7.25 and 9.22,
+which is close to the true means of the two groups, namely 9.04 and 6.93.
 When tabulating estimated group membership against observed (known) group membership, it can be seen that women are classified with a high degree of accuracy, but men are not:
 
-```{r tabshoe, results="asis"}
-library(kableExtra)
-tab <- table(C, props)
-tab <- data.frame(Observed = rownames(tab), as.data.frame.array(tab))
-names(tab)[2:3] <- c("Class 1", "Class 2")
-rownames(tab) <- NULL
-kbl(tab, caption = "Observed group membership by estimated class membership.")
-```
+\begin{table}
+
+\caption{\label{tab:tabshoe}Observed group membership by estimated class membership.}
+\centering
+\begin{tabular}[t]{l|r|r}
+\hline
+Observed & Class 1 & Class 2\\
+\hline
+Man & 21 & 28\\
+\hline
+Woman & 51 & 0\\
+\hline
+\end{tabular}
+\end{table}
 
@@
 A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
-One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
-By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibsonTenFrequentlyAsked2018].
-In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.
+One difference between the two techniques is that factor analysis can be considered as a way to group observed \emph{variables} into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
+By contrast, mixture modeling groups \emph{individuals} into classes (see \textbf{nylund-gibsonTenFrequentlyAsked2018?}).
+In line with this distinction, latent class analysis is sometimes referred to as a ``person-centered'' technique, and factor analysis as a ``variable-centered'' technique.
 
@@
 latent class analysis can be thought of as a type of clustering algorithm.
-In line with this perspective, mixture modeling is sometimes described as "model-based
-clustering" (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
+In line with this perspective, mixture modeling is sometimes described as ``model-based
+clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \&
 Raftery, 2017).
 Many clustering algorithms apply some recursive splitting algorithm to the data.
-By contrast "model-based" clustering refers to the fact that latent class analysis 
+By contrast ``model-based'' clustering refers to the fact that latent class analysis
 estimates cluster membership based on a parametric model.
@@
 Finally, in the context of machine learning, latent class analysis can be considered as an
-*unsupervised classification* problem [@figueiredoUnsupervisedLearningFinite2002].
-The term *unsupervised* refers to the fact that the outcome variable, true class membership,
-is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome (class membership).
-
-## A taxonomy of latent class analyses
+\emph{unsupervised classification} problem (\textbf{figueiredoUnsupervisedLearningFinite2002?}).
+The term \emph{unsupervised} refers to the fact that the outcome variable, true class membership,
+is not known, and the term \emph{classification} refers to the fact that the algorithm is predicting a categorical outcome (class membership).
+
+\hypertarget{a-taxonomy-of-latent-class-analyses}{%
+\subsection{A taxonomy of latent class analyses}\label{a-taxonomy-of-latent-class-analyses}}
 
 In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
-From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermuntj.k.LatentClassAnalysis2004].
+From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators (\textbf{vermuntj.k.LatentClassAnalysis2004?}).
 Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
-The term "latent class analysis" seems most appropriate as an umbrella term for this broader class of models,
+The term ``latent class analysis'' seems most appropriate as an umbrella term for this broader class of models,
 as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
@@
 
-One common type of latent class analyses is the *finite Gaussian mixture model*;
+One common type of latent class analyses is the \emph{finite Gaussian mixture model};
 a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
@@
 
-The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
+The technique known as \emph{latent profile analysis (LPA)} is a special case of the mixture model,
 which assumes conditional independence of the indicators.
@@
 A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
-Examples of this approach include *growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
+Examples of this approach include \emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA).
 These techniques estimate a latent growth model to describe individual trajectories over time.
@@
 One common assumption is that each categorical variable reflects an underlying standard normal distribution.
-The parameters in such a model are then "thresholds", corresponding to the quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
+The parameters in such a model are then ``thresholds'', corresponding to the quantiles of a standard normal distribution (with \(N(\mu = 0, \sigma = 1)\)).
 These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
 For example, a binary indicator has a single threshold that distinguishes the two response categories.
-If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
-If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.
+If responses are distributed 50/50, then the corresponding threshold would be \(t_1 = 0.00\).
+If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).
 This paper will primarily focus on mixture models and special cases thereof,
@@
 
-<!-- f a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis.
-Modified to handle discrete data, this constrained analysis is known as LCA.
-Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit.
--->
-
-<!-- ecause LPA is model-based, a number of different model parameterizations can be -->
-<!-- estimated. These models differ in terms of whether--and how--parameters are -->
-<!-- estimated across the profiles. These parameters are the means for the different -->
-<!-- profiles, which, in this approach, always are estimated freely across the -->
-<!-- profiles; the variances for the variables used to create the profiles, which can -->
-<!-- be estimated freely or can be estimated to be the same, or equal, across -->
-<!-- profiles; and the covariances of the variables used to create the profiles, -->
-<!-- which can be freely-estimated, estimated to be equal, or fixed to be zero. -->
-
-<!-- In general, as more parameters are estimated (i.e., those that are fixed to zero -->
-<!-- are estimated as being equal across profiles; or those estimated as being equal -->
-<!-- across profiles are freely-estimated across them), the model becomes more -->
-<!-- complex; the model may fit better, but also be overfit, meaning that the -->
-<!-- profiles identified may be challenging to replicate with another, separate data -->
-<!-- set. Even still, flexibility in terms of which models can be estimated also has -->
-<!-- affordances. For example, the varying means, equal variances, and covariances -->
-<!-- fixed to 0. A researcher might choose this model specification if she wants to -->
-<!-- model the variables to be used to create profiles that are independent. This -->
-<!-- model is very simple, as no covariances are estimated and the variances are -->
-<!-- estimated to be the same across profiles. As we estimate more parameters (and -->
-<!-- decrease the degrees of freedom), we are more likely to fit the data, but less -->
-<!-- likely to be able to replicate the model with a second set of data. As we -->
-<!-- progress toward more complex models (with increasingly complex -->
-<!-- parameterization), then we are more likely to fit the data better. -->
-<!-- In all, this flexibility associated with LPA also has a more pragmatic cost. It -->
-<!-- can be very difficult to efficiently---and reproducibly---estimate and compare a -->
-<!-- number of models. This cost means that analysts may focus on a subset of models. -->
-<!-- It may also mean that unintentional errors are introduced through copying and -->
-<!-- pasting the output of model estimations. Part of why we developed tidyLPA was to -->
-<!-- make the model estimation process more efficient and more reproducible. As we -->
-<!-- describe later, we do this using R, which has a number of benefits, including -->
-<!-- being open-source, powerful (in terms of pre-processing and using the output of -->
-<!-- models), and increasingly widely-used in the psychological sciences. We also -->
-<!-- provide an interface to perhaps the most widely-used software for LPA, MPlus. In -->
-<!-- this way, tidyLPA can both allow researchers to estimate models they already -->
-<!-- estimate using MPlus more efficiently and reproducibly, and, for those without -->
-<!-- access to this software, to do so entirely using R and the open-source R package -->
-<!-- mclust. -->
-
-## Use cases for latent class analysis
+\hypertarget{use-cases-for-latent-class-analysis}{%
+\subsection{Use cases for latent class analysis}\label{use-cases-for-latent-class-analysis}}
 
@@
 One example is to test a theory that postulates the existence of a categorical latent variable.
-For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses.
+For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses.
 Mixture modeling can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
@@
 or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughmanMixtureModelAnalysis2006] or mental [@wuAbuseDependencePrescription2011] health problems.
+For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical (\textbf{baughmanMixtureModelAnalysis2006?}) or mental (\textbf{wuAbuseDependencePrescription2011?}) health problems.
 The example of shoe size is a rudimentary illustration of this type of application.
 
-<!-- Inductively identify number of latent classes -->
-<!-- Classify individuals -->
-<!-- Identify items (observed characteristics) that -->
-<!-- indicate these classes well -->
-<!-- Predict class membership from covariates -->
-<!-- Predict outcomes from class membership -->
-<!-- Dimension reduction -->
-<!-- Violation of assumptions -->
-
-<!-- In medical and psychological science, LPA can be useful when -->
-<!-- considerable between-subject heterogeneity exists in scores on a range of variables and when -->
-<!-- this variation cannot be explained by known, manifest variables (e.g., Wolfe 1970; Sterba -->
-<!-- 2013). Here, LPA can help to identify or approximate possibly meaningful sub-grouping of -->
-<!-- subjects that may help to better understand sample heterogeneity (Sterba 2013). -->
-<!-- Generally, LPA works under the assumption that sample (residual) variance can be reduced -->
-<!-- by assuming a categorical latent variable that effectively subdivides the sample into >=2 -->
-<!-- subgroups that are more homogeneous in terms of their patterns of variable means and -->
-<!-- (co)variances. In case an LPA model is found to fit a dataset well, it is often the case that -->
-<!-- subjects in each class resemble each other relatively well in terms of their scores on the input -->
-<!-- variables. Depending on the model configuration, the identified classes can show different -->
-<!-- class-specific patterns of means and class-specific or class-varying variances. -->
-
-<!--                                                1 -->
-<!-- A -->
-<!-- downside is that these models are more complex (much more parameters to be estimated). -->
-<!-- Using criteria, such as the Bayesian Information criterion (BIC) helps to find a model that -->
-<!-- strikes a good balance between model-fit and model-complexity. When doing an LPA, most -->
-<!-- applied researchers will usually be most interested in the differences and/or overlap between -->
-<!-- the classes’ specific patterns of parameter estimates. These can be used to characterize the -->
-<!-- classes and, possibly, provide clues about underlying mechanisms (Sterba 2013). -->
-
-<!-- This tutorial -->
-<!-- LPA is less widely used than other latent variable models and, possibly due to that, has -->
-<!-- long been only available in specialized software packages such as Mplus. Luckily, ongoing -->
-<!-- developments in many different scientific fields (e.g., ecology, econometrics) have yielded a -->
-<!-- number of packages that also allow users to conduct LPA in the open-source R-platform. -->
-<!-- However, the use of R does require experience and documentation of packages can be rather -->
-<!-- limited or technical, making it less easily accessible for applied researchers. Therefore, this -->
-<!-- tutorial is aimed to help applied researchers to get going with LPA in R, illustrating the use -->
-<!-- of several packages and, for reference, providing a comparison of the results obtained in R -->
-<!-- with results obtained with Mplus. -->
-
-# Best practices
-
-## In estimation
-
-The best practices in estimation, as outlined in Table \@ref(tab:checkest),
+\hypertarget{best-practices}{%
+\section{Best practices}\label{best-practices}}
+
+\hypertarget{in-estimation}{%
+\subsection{In estimation}\label{in-estimation}}
+
+The best practices in estimation, as outlined in Table \ref{tab:checkest},
 are inspired by prior recommendations for best practices for estimating
 specific sub-types of latent class analyses,
-including latent class growth analysis [@schootGRoLTSChecklistGuidelinesReporting2017]
-and latent class analysis with ordinal indicators [e.g., @nylund-gibsonTenFrequentlyAsked2018].
+including latent class growth analysis (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?})
+and latent class analysis with ordinal indicators (e.g., \textbf{nylund-gibsonTenFrequentlyAsked2018?}).
 These were generalized to be more relevant to all types of latent class analyses,
@@
 
-```{r checkest}
-tab_check <- data.frame(Item = c(
-  "Examine observed data",
-  "Handling missing data",
-  "Alternative model specifications",
-  "Starting values",
-  "Algorithm"
-))
-tab_check$`#` = paste0(1:length(tab_check$Item), ".")
-kbl(tab_check[, c("#", "Item")])
-```
-### Examine observed data
+\begin{tabular}[t]{l|l}
+\hline
+\# & Item\\
+\hline
+1. & Examine observed data\\
+\hline
+2. & Handling missing data\\
+\hline
+3. & Alternative model specifications\\
+\hline
+4. & Starting values\\
+\hline
+5. & Algorithm\\
+\hline
+\end{tabular}
+
+\hypertarget{examine-observed-data}{%
+\subsubsection{Examine observed data}\label{examine-observed-data}}
 
@@
 Pay special attention to level of measurement of the indicators.
-Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions [see @vermuntKmeansMayPerform2011].
-Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
+Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions (see \textbf{vermuntKmeansMayPerform2011?}).
+Personal experience consulting on latent class analyses and moderating the \texttt{tidyLPA} Google group
 suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is perhaps the most common source of user error.
-Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @normanLikertScalesLevels2010],
+Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous (e.g., \textbf{normanLikertScalesLevels2010?}),
 this certainly does not imply that all methods are.
-It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
+It is certainly unlikely that such ordinal variables can be treated as a \emph{mixture} of multiple normal distributions.
 The problem becomes eggregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
@@
 
-Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function `tidySEM::descriptives(data)`.
+Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function \texttt{tidySEM::descriptives(data)}.
 Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
@@
 Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @vanlissaWORCSWorkflowOpen2020].
-
-### Missing data
-
-Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled [@schootGRoLTSChecklistGuidelinesReporting2017].
-Three types of missingness have been distinguished in the literature [@rubinInferenceMissingData1976]:
+Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible (consider using a GitHub repository as a comprehensive public research archive, as explained in \textbf{vanlissaWORCSWorkflowOpen2020?}).
+
+\hypertarget{missing-data}{%
+\subsubsection{Missing data}\label{missing-data}}
+
+Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?}).
+Three types of missingness have been distinguished in the literature (\textbf{rubinInferenceMissingData1976?}):
 Missing completely at random (MCAR), which means that missingness is random;
-missing at random (MAR), which means that missingness is contingent on the *observed* data (and can thus be accounted for);
+missing at random (MAR), which means that missingness is contingent on the \emph{observed} data (and can thus be accounted for);
 and finally missing not at random (MNAR), which means that missingness is related to unobserved factors.
-It is possible to conduct a so-called "MCAR" test, 
-for example the non-parametric MCAR test [@jamshidianTestsHomoscedasticityNormality2010].
-But note that the name "MCAR test" is somewhat misleading,
+It is possible to conduct a so-called ``MCAR'' test,
+for example the non-parametric MCAR test (\textbf{jamshidianTestsHomoscedasticityNormality2010?}).
+But note that the name ``MCAR test'' is somewhat misleading,
 as the null-hypothesis of this test is that the data are not MAR,
@@
 A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@littleTestMissingCompletely1988].
+As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (\textbf{littleTestMissingCompletely1988?}).
 This assumption is tenuous in the context of latent class analysis.
-A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable [-@jamshidianTestsHomoscedasticityNormality2010].
+A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable (\textbf{jamshidianTestsHomoscedasticityNormality2010?}).
 Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
-For this tutorial, I have re-implemented it in the `mice` package as `mice::MCAR()`,
+For this tutorial, I have re-implemented it in the \texttt{mice} package as \texttt{mice::MCAR()},
 with a fast backend in C++ and new printing and plotting methods.
@@
 While we concur that investigating missingness is due dilligence,
-it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as `OpenMx` (and e.g., Mplus).
+it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as \texttt{OpenMx} (and e.g., Mplus).
 These packages use Full Information Maximum Likelihood (FIML) estimation,
 which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data; on par with multiple imputation [@leeComparisonFullInformation2021].
+FIML is a best-practice solution for handling missing data; on par with multiple imputation (\textbf{leeComparisonFullInformation2021?}).
 FIML estimation assumes that missingness is either MCAR or MAR.
@@
 Second, because there is no straightforward way to integrate latent class analysis results across multiple datasets.
-To conclude; our recommendation is to inspect missingness (e.g., using `mice::MCAR()`)
-and report the number of missings per variable (e.g., using `tidySEM::descriptives()`),
+To conclude; our recommendation is to inspect missingness (e.g., using \texttt{mice::MCAR()})
+and report the number of missings per variable (e.g., using \texttt{tidySEM::descriptives()}),
 before proceeding with FIML.
 One minor concern is that the K-means algorithm,
-which `tidySEM` uses for determining starting values,
-is *not* robust to missing values.
-When it fails, `tidySEM` automatically switches to hierarchical clustering,
+which \texttt{tidySEM} uses for determining starting values,
+is \emph{not} robust to missing values.
+When it fails, \texttt{tidySEM} automatically switches to hierarchical clustering,
 unless the user specifies a different
@@
 
-### Alternative model specifications
+\hypertarget{alternative-model-specifications}{%
+\subsubsection{Alternative model specifications}\label{alternative-model-specifications}}
 
@@
 
-Prior literature on latent class analysis has emphasized exploratory applications of the method [see @nylundDecidingNumberClasses2007].
+Prior literature on latent class analysis has emphasized exploratory applications of the method (see \textbf{nylundDecidingNumberClasses2007?}).
 In exploratory analyses, a large number of models are typically estimated in batch,
 with varying numbers of classes and model specifications.
-The "correct" model specification is then determined based on a combination of fit indices,
+The ``correct'' model specification is then determined based on a combination of fit indices,
 significance tests,
 and interpretability.
-For latent profile analysis, the function `tidySEM::mx_profiles(classes, variances, covariances)`
+For latent profile analysis, the function \texttt{tidySEM::mx\_profiles(classes,\ variances,\ covariances)}
 largely automates this process.
-The argument `classes` indicates which class solutions should be estimated (e.g., 1 through 6).
-The argument `variances` specifies whether variances should be `"equal"` or `"varying"` across classes.
-The argument `covariances` specifies whether covariances should be constrained to `"zero"`, `"equal"` or `"varying"` across classes.
-The means are free to vary across classes by default, although the more general function `tidySEM::mx_mixture()` could be used to circumvent this.
-After all models have been estimated, the function `tidySEM::table_fit()` can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
+The argument \texttt{classes} indicates which class solutions should be estimated (e.g., 1 through 6).
+The argument \texttt{variances} specifies whether variances should be \texttt{"equal"} or \texttt{"varying"} across classes.
+The argument \texttt{covariances} specifies whether covariances should be constrained to \texttt{"zero"}, \texttt{"equal"} or \texttt{"varying"} across classes.
+The means are free to vary across classes by default, although the more general function \texttt{tidySEM::mx\_mixture()} could be used to circumvent this.
+After all models have been estimated, the function \texttt{tidySEM::table\_fit()} can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
 Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
 because this test is very computationally expensive.
-It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
+It is recommended to use the function \texttt{tidySEM::BLRT()} to compare a shortlist of likely candidate models based on other fit indices.
 Fit indices typically used for determining the optimal number of classes include
@@
 Another option is to always prefer the most parsimoneous model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogulComparisonInformationCriteria2016].
+Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (\textbf{akogulComparisonInformationCriteria2016?}).
 
@@
 
-Van de Schoot and colleagues [-@vandeschootGRoLTSChecklistGuidelinesReporting2017] argue that there are many choices to be made in the specification of latent trajectory models,
+Van de Schoot and colleagues (\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}) argue that there are many choices to be made in the specification of latent trajectory models,
 and that alternative
 
+\begin{verbatim}
 * Consider alternative parameterizations; e.g., free variances/covariances 
 * See Nylund
-
-## Software
+\end{verbatim}
+
+\hypertarget{software}{%
+\subsection{Software}\label{software}}
 
@@
 
-New functionality in the R-package `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
+New functionality in the R-package \texttt{tidySEM} seeks to lower the threshold for latent class analysis using \texttt{OpenMx}.
 It adheres to best practices in estimation and reporting, as described in this paper.
-The user interface is simple, making use of the model syntax of the widely used `lavaan` R-package.
+The user interface is simple, making use of the model syntax of the widely used \texttt{lavaan} R-package.
 This syntax offers a human-readable way to specify latent variable models.
@@
 
-<!-- 3. Packages -->
-<!-- 3.1 R-packages -->
-<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
-<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
-<!-- analytic functionality. However, the majority of packages is focused on analyses with discrete -->
-<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
-<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
-<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
-<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
-<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
-<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
-<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
-<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
-<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
-<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
-<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->
-
-<!-- Although ‘tidyLPA’ is easy to use, this comes at the expense of restricting modeling options -->
-<!-- to a few, oft-used options. For completeness, both the ‘mclust’ and ‘tidyLPA’ approach will -->
-<!-- be illustrated. -->
-
-## Best practices in estimation
-
-### Algorithm
+\hypertarget{best-practices-in-estimation}{%
+\subsection{Best practices in estimation}\label{best-practices-in-estimation}}
+
+\hypertarget{algorithm}{%
+\subsubsection{Algorithm}\label{algorithm}}
 
@@
 The choice of the estimator depends on the presence of missing values,
-sample size, number of indicators, and available computational resources (Weller, Bowen & Faubert, 2020).
+sample size, number of indicators, and available computational resources (Weller, Bowen \& Faubert, 2020).
 A commonly used technique is maximum likelihood (ML) estimation
 with the expectation-maximization (EM) algorithm as a local optimizer.
-Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator
+Imagine we are estimating two parameters, e.g.~the class-specific means \(\mu_c\) on a continuous indicator
 (ignoring the variance for now). The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes the likelihood ($LL$) of all observed data. 
-In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. 
+for these two parameters that maximizes the likelihood (\(LL\)) of all observed data.
+In practice, instead of maximizing \(LL\), often \(-2*LL\) is minimized, as this offers computational advantages.
 We can think of this optimization problem as a three-dimensional landscape:
 The X and Y dimensions are determined by the class-specific means,
-so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. 
-The optimizer must find the deepest "valley" in this landscape, 
-which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
+so \(X = \mu_1\) and \(Y = \mu_2\) - and the Z-dimension is determined by \(Z = -2*LL\).
+The optimizer must find the deepest ``valley'' in this landscape,
+which reflects the combination of \(\mu_1\) and \(\mu_2\) that maximizes the likelihood of the data.
 The EM optimizer behaves somewhat like a marble, dropped in this landscape.
@@
 The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
-(this is known as "convergence").
+(this is known as ``convergence'').
 It cannot climb out again.
 Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley
-(a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
-One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
-choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
-This is the "random starts" approach.
+(a ``local optimum''), and never discovers the deepest valley (the ``global optimum'', or best solution).
+One solution to this problem is to drop many marbles at random places, compare their final \(-2*LL\) values,
+choose the solution with the lowest \(-2*LL\), and make sure that several marbles replicated this solution.
+This is the ``random starts'' approach.
 
@@
 Moreover, because the algorithm begins with random starting values,
-many of the marbles are likely to be very far away from a "good enough" solution.
+many of the marbles are likely to be very far away from a ``good enough'' solution.
 Two innovations may improve the estimation procedure.
 The first is that, instead of picking random starting values,
-a "reasonable solution" may be used for the starting values.
+a ``reasonable solution'' may be used for the starting values.
 For example, if we assume that the different classes are likely to have different mean values on the indicators,
@@
 Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some "destination" in the landscape, 
+Simulated annealing iteratively considers some ``destination'' in the landscape,
 and compares its likelihood to the current one. If the destination likelihood is higher,
 the estimator moves there.
-If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
+If the destination likelihood is \emph{lower}, the estimator still moves there occasionally, based on probability.
 This latter property allows it to escape local optima.
 
-By default, `tidySEM` employs this solution of deriving starting values using K-means clustering,
+By default, \texttt{tidySEM} employs this solution of deriving starting values using K-means clustering,
 and identifying the global optimum solution using simulated annealing.
@@
 
-### Starting values
+\hypertarget{starting-values}{%
+\subsubsection{Starting values}\label{starting-values}}
 
@@
 To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter
-in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004). 
+in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan \& Peel, 2004).
 Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values.
-<!-- add more on this -->
 Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
@@
 Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence
-or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004).
+or in case of having poor initial values the algorithm might diverge altogether (McLachlan \& Peel, 2004).
 The key is to use a large number of starting value sets.
@@
 
-[@nylund-gibsonTenFrequentlyAsked2018; ]
-
+{[}(\textbf{nylund-gibsonTenFrequentlyAsked2018?}); {]}
+
+\begin{verbatim}
 In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
@@
 [@biernackiChoosingStartingValues2003]
-
-### Assessing Estimation Results
-<!-- maybe change to class enumeration -->
+\end{verbatim}
+
+\hypertarget{assessing-estimation-results}{%
+\subsubsection{Assessing Estimation Results}\label{assessing-estimation-results}}
 
@@
 These include but are not limited to likelihood ratio tests, information criteria,
-and the Bayes factor (Weller, Bowen & Faubert, 2020).
-
-Absolute model fit can be assessed using the likelihood ratio (LR) $\chi^2$ goodness-of-fit test. 
+and the Bayes factor (Weller, Bowen \& Faubert, 2020).
+
+Absolute model fit can be assessed using the likelihood ratio (LR) \(\chi^2\) goodness-of-fit test.
 It can compare two nested models when they have the same number of classes (Lanza et al., 2003).
@@
 Therefore, we make a statement about how well our model fits the data.
-The perfect fit would be indicated by the $\chi^2\ =\ 0$ and $p\ =\ 1$.
-Ideally, the LR $\chi^2$ test will be non-significant as this implies
+The perfect fit would be indicated by the \(\chi^2\ =\ 0\) and \(p\ =\ 1\).
+Ideally, the LR \(\chi^2\) test will be non-significant as this implies
 that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model.
 The underlying assumption of this procedure is that the test statistic is asymptotically distributed
-as a $\chi^2$ meaning that it can be approximated using this distribution.
+as a \(\chi^2\) meaning that it can be approximated using this distribution.
 This is the case when the sample size is adequate (Lanza et al., 2003).
-The dependence of the LR $\chi^2$ test statistic on sample size has its downsides (Masyn, 2013). 
+The dependence of the LR \(\chi^2\) test statistic on sample size has its downsides (Masyn, 2013).
 Very large samples will often effortlessly reject the null even when the final class solution is a good fit,
-and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic.
-In fact, small samples might not be well approximated using the $\chi^2$ distribution,
+and small samples will often be underpowered resulting in non-significant \(\chi^2\) statistic.
+In fact, small samples might not be well approximated using the \(\chi^2\) distribution,
 which is a problem shared with large yet sparse datasets.
@@
 with the new degrees of freedom being the difference in their degrees of freedom.
-This statistic also follows the $\chi^2$ distribution.
-Similar to the LR $\chi^2$ goodness-of-fit test, we want the test statistic to be non-significant
+This statistic also follows the \(\chi^2\) distribution.
+Similar to the LR \(\chi^2\) goodness-of-fit test, we want the test statistic to be non-significant
 in order to give support to the simpler model.
@@
 
-<!-- Add a paragraph about using the Bayes factor see Weller, Bowen & Faubert, 2020-->
-
-<!-- decide whether to keep the following paragraph, if so, where to place it -->
 When estimators which require sets of starting values are used,
@@
 Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated
-in at least 3-10% of the models initialized with different starting values in the first step of the optimization. 
+in at least 3-10\% of the models initialized with different starting values in the first step of the optimization.
 This consideration does not apply in case of models which use simulated annealing since these do not have starting values.
-<!-- Masyn 2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
-<!-- to ensure that a global (rather than a local) maximum has been reached.". For discussion see Hipp & Bauer, 2006 -->
-
-### Classification Diagnostics
+
+\hypertarget{classification-diagnostics}{%
+\subsubsection{Classification Diagnostics}\label{classification-diagnostics}}
 
@@
 This is considered a high classification accuracy and means that the classes are distinct.
-<!-- see if you find content on homogeneity of classes -->
 Three important classification diagnostic measures are entropy, the average posterior probability and the modal class assignment proportion.
@@
 
-<!-- Masyn 2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
-<!-- add content about entropy -->
-
 The average posterior probability is a measure of classification uncertainty for each latent class.
@@
 given that the person belongs to a particular latent class.
-<!-- If a particular item is endorsed by two or more classes at markedly different rates, -->
-<!-- it is said to discriminate well between classes and is consequently considered a good indicator. -->
-<!-- The classes are considered highly homogeneous with respect to an item when for a particular item -->
-<!-- there are a distinct difference in conditional item probabilities of two or more classes. -->
-<!-- If an item is endorsed below 30% for one class and above 70% for another class, 
-the classes have high homogeneity with respect to this item (Masyn, 2013). MY OWN WRITING BUT NOT FLOWY ENOUGH-->
 However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1),
@@
 Overly simplified and generalized class names may prove misleading to both audiences and researches alike
-leading to what is known as a naming fallacy (Weller, Bowen & Faubert, 2020).
-
-### Label switching
+leading to what is known as a naming fallacy (Weller, Bowen \& Faubert, 2020).
+
+\hypertarget{label-switching}{%
+\subsubsection{Label switching}\label{label-switching}}
 
@@
 
-<!-- Article discussing label switching: see Chung, Loken & Schafer -->
-
-<!-- Initialization around a Single Mode -->
-<!-- Another common approach is to run a single chain or to initialize the parameters near realistic values.36 -->
-
-<!-- This can work better than the hard constraint approach if reasonable initial values can be found and the labels do not switch within a Markov chain.
-The result is that all chains are glued to a neighborhood of a particular mode in the posterior.
--->
-<!-- https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/1467-9868.00265?casa_token=Tjk8gR4w8AEAAAAA:RygJL5OgdBaVWY6oaB8XlnyMOnzQgOVDBd2EHDyQD6TH-DzxRxpmU1J4AG8N-HpL5dHHNINRybiaMl6w -->
-
-## Best practices in reporting
-
-Among studies using LCA, reporting practices vary significantly (Weller, Bowen & Faubert, 2020).
-Various authors have tried to better and standardize ways of reporting LCA (e.g. Masyn, 2013; Weller, Bowen & Faubert, 2020). 
+\hypertarget{best-practices-in-reporting}{%
+\subsection{Best practices in reporting}\label{best-practices-in-reporting}}
+
+Among studies using LCA, reporting practices vary significantly (Weller, Bowen \& Faubert, 2020).
+Various authors have tried to better and standardize ways of reporting LCA (e.g.~Masyn, 2013; Weller, Bowen \& Faubert, 2020).
 Building on their work, we provide guidelines with a strong emphasis on scientific reproducibility and transparency
 better known as the open science framework (OSF). Open science framework promotes scientific openness, reproducibility and integrity
-through an establishment of guidelines for the scientific process (Foster & Deardorff, 2017)
+through an establishment of guidelines for the scientific process (Foster \& Deardorff, 2017)
 To this end, van Lissa and colleagues (2020) developed WORCS, a workflow for open reproducible code in science.
@@
 
+\begin{verbatim}
 * Use comprehensive citation; this includes referencing the software used.
@@
 * Ideally, make the entire research project reproducible so that others may download it and reproduce it with just one click; for guidance, see Van Lissa and colleagues.
-
-```{r}
-grolts <- structure(list(Number = c("1.", "2.", "3a.", "3b.", "3c.", "4.", 
-"5.", "6a.", "6b.", "7.", "8.", "9.", "10.", "11.", "12.", "13.", 
-"14a.", "14b.", "14c.", "15.", "16."), Item = c("Is the metric of time used in the statistical model reported?", 
-"Is information presented about the mean and variance of time within a wave?", 
-"Is the missing data mechanism reported?", "Is a description provided of what variables are related to attrition/missing data?", 
-"Is a description provided of how missing data in the analyses were dealt with?", 
-"Is information about the distribution of the observed variables included?", 
-"Is the software mentioned?", "Are alternative specifications of within-class heterogeneity considered (e.g., LGCA vs.
-LGMM) and clearly documented?
-If not, was sufficient justification provided as to eliminate certain specifications from consideration?",
-"Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented?
-If not, was sufficient justification provided as to eliminate certain specifications from consideration?",
-"Are alternative shape/functional forms of the trajectories described?", 
-"If covariates have been used, can analyses still be replicated?", 
-"Is information reported about the number of random start values and final iterations included?", 
-"Are the model comparison (and selection) tools described from a statistical perspective?", 
-"Are the total number of fitted models reported, including a one-class solution?", 
-"Are the number of cases per class reported for each model (absolute sample size, or proportion)?", 
-"If classification of cases in a trajectory is the goal, is entropy reported?", 
-"Is a plot included with the estimated mean trajectories of the final solution?", 
-"Are plots included with the estimated mean trajectories for each model?", 
-"Is a plot included of the combination of estimated means of the final model and the observed individual trajectories split out for each latent class?", 
-"Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?", 
-"Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?"
-), relevant = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
-TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
-TRUE, TRUE, TRUE)), row.names = c(NA, -21L), class = "data.frame")
-```
-
-## Best practices in visualization
-
-# Tutorial
-
-<!-- 5. Tutorial -->
-<!-- 5.1 ‘mclust’ -->
-<!-- To run LPA using the ‘mclust’ package, the Mclust() function can be used. This function -->
-<!-- requires data as its minimum input. In addition, the number of clusters to fit (G) and model -->
-<!-- variants to fit and select from (modelNames) can be entered (defaults are: G=1-9 and all -->
-<!-- fourteen model variants). In many cases, we may want to consider a more limited range of -->
-<!-- model variants to with different numbers of classes. For multivariate data, a total of fourteen -->
-<!-- options are available (see help("mclustModelNames") for an overview). Each model variant -->
-<!-- is estimated for each value of the specified range of G using Expectation Maximization (EM), -->
-<!-- using the Bayesian Information Criterion (BIC) to compare different model variants and -->
-<!-- select the optimal one. -->
-<!-- Conducting a latent variable model or mixture analysis with Mplus or comparable software -->
-<!-- usually entails fitting models with different numbers of classes (that are otherwise configured -->
-<!-- similarly) and to compare the fit between these models to select the best model. The -->
-<!-- default approach taken in the ‘mclust’ package is slightly different in that the main modeling -->
-<!-- command Mclust() fits multiple models for all possible combinations of the specified number -->
-<!-- of classes and model configurations . The eventual output of ‘mclust’ is the BIC-selected best -->
-<!-- model variant with the optimal combination of number of classes and model configuration. -->
-<!-- An advantage of the ‘mclust’ approach is that it is efficient and flexible: given a number of -->
-<!-- classes, the best model configuration is selected straight away and we do not have to run all -->
-<!-- model combinations one by one. This makes the approach especially suitable for exploratory -->
-<!-- analyses. However, like stated above, this approach is somewhat less usual in some fields, -->
-<!-- where estimation of LPA is approached in a more confirmatory fashion. In addition, it does -->
-<!-- take some control away from the researcher, who may want to primarily focus on selecting -->
-<!-- the number of classes, given a particular , theory-based model configuration that is kept -->
-<!-- constant throughout analyses, which is often the chosen strategy in software like Mplus and -->
-<!-- made easily available through the ‘tidyLPA’ package. -->
-<!-- Here, both approaches are shown: (1) the ‘mclust’ aproach and (2) the ‘tidyLPA’ approach. -->
-
-<!--                                               5 -->
-<!-- 5.1.1 The ‘mclust’ approach -->
-<!-- In this approach, we run consecutive models with increasing numbers of classes (G=1 to -->
-<!-- G=9), for each model we let the package fit the four above described model variants (“EEI”, -->
-<!-- “EEE”, “VVI” and “VVV”) and select the best-fitting one. We start by loading the package -->
-<!-- (library(mclust)) and creating an object mnames that contains the names of the four models -->
-<!-- we want to be fitted. Next, we use the Mclust() function to fit the models: -->
-<!-- library(mclust) -->
-
-<!-- ## Package 'mclust' version 5.4.7 -->
-<!-- ## Type 'citation("mclust")' for citing this R package in publications. -->
-<!-- mnames <- c("EEI", "EEE", "VVI", "VVV") -->
-
-<!-- # Fit 1-5 class model -->
-<!-- mod_g1_9 <- Mclust(dat1[, 1:10], modelNames = mnames) -->
-
-<!-- We can first look at the optimal number of classes and the optimal model variant that were -->
-<!-- selected, using the following code: -->
-<!-- # Opimal number of classes -->
-<!-- mod_g1_9$G -->
-
-<!-- ## [1] 3 -->
-<!-- # Optimal model variant -->
-<!-- mod_g1_9$modelName -->
-
-<!-- ## [1] "EEI" -->
-<!-- This shows us that a 3-class EEI model was selected as the best model based on the BIC. We -->
-<!-- can get a better perspective of this model’s performance if we compare it to the other fitted -->
-<!-- models. We can do this by taking a closer look at the other models’ BIC values: -->
-<!-- mod_g1_9$BIC -->
-
-<!-- ##   Bayesian Information Criterion (BIC): -->
-<!-- ##           EEI       EEE       VVI       VVV -->
-<!-- ##   1 -14661.61 -12986.57 -14661.61 -12986.57 -->
-<!-- ##   2 -13275.65 -12611.60 -13095.82 -12678.12 -->
-<!-- ##   3 -12142.11 -12290.68 -12163.08 -12731.18 -->
-<!-- ##   4 -12178.44 -12330.04 -12241.21 -12970.57 -->
-<!-- ##   5 -12223.14 -12379.96 -12320.47 -13231.35 -->
-<!-- ##   6 -12264.04 -12423.72 -12406.59 -13522.40 -->
-<!-- ##   7 -12269.39 -12451.25 -12497.03 -13748.91 -->
-<!-- ##   8 -12298.30 -12483.35 -12568.62 -13980.82 -->
-<!-- ##   9 -12347.11 -12529.40 -12645.50 -14269.34 -->
-<!-- ## -->
-
-<!--                                              6 -->
-<!-- ## Top 3 models based on the BIC criterion: -->
-<!-- ##     EEI,3     VVI,3     EEI,4 -->
-<!-- ## -12142.11 -12163.08 -12178.44 -->
-<!-- mod_g1_9$loglik -->
-
-<!-- ## [1] -5951.275 -->
-<!-- mod_g1_9$df -->
-
-<!-- ## [1] 42 -->
-<!-- Here, we can see the BICs for all fitted models (in this case 1-9 classes and 4 model variants: -->
-<!-- 36 models in total). We can see that the closest contenders were a 3-class model with a VVI -->
-<!-- configuration (with variances allowed to vary both within and between classes) and a 4-class -->
-<!-- EEI model. -->
-<!-- Note that the BIC values are negative, and that values that are more negative are considered -->
-<!-- to fit poorer than values closer to 0. This may strike some users as odd, given that in -->
-<!-- many modeling applications, we are used to consider lower BIC values to indicate better -->
-<!-- fit. However, the ‘mclust’ approach of maximizing the BIC is correct for the used modeling -->
-<!-- approach (see e.g., Fraley & Raftery, 2003; Banfield & Raftery, 1993). For theoretical reasons, -->
-<!-- the BIC is calculated in’mclust’ as: -->
-
-<!--                            BIC = 2(Loglikelihood) - df (log(n)) -->
-
-<!-- where df is the number of parameters (degrees of freedom) and n is the sample size. We can -->
-<!-- see, that here, higher log likelihood values lead to a higher BIC. For the best fitting model in -->
-<!-- our example, the log likelihood is -5951.275, the number of parameters is 42 and the sample -->
-<!-- size is 300. If we plug these into the formula, we get the following BIC: -->
-
-<!--                      BIC = 2(-5951.275) - 42(log(300)) = -12142.11 -->
-
-<!-- As expected, the obtained value indeed corresponds to the BIC we got in the model output -->
-<!-- (see above). Other software packages, such as Mplus and ‘tidyLPA’ (see below) use a slightly -->
-<!-- different formula to calculate the BIC, where higher log likelihood values lead to lower BIC -->
-<!-- values: -->
-
-<!--                            BIC = df (log(n)) - 2(Loglikelihood) -->
-
-<!-- When we plug the best-model values into this formula, we get: -->
-
-<!--                       BIC = 42(log(300)) - 2(-5951.275) = 12142.11 -->
-
-<!-- We can see that the absolute BIC value is the same, irrespective of the used formula, with -->
-<!-- only the sign differing between the two approaches. -->
-
-<!--                                               7 -->
-<!-- Now, let us continue by taking a closer look at the class sizes and the mean values and -->
-<!-- variances of the ten variables for the three identified classes. Because the best model has and -->
-<!-- EEI configuration, printing out a single variance matrix from one class will suffice, because -->
-<!-- the variances are the same across classes (this is different in cases where the optimal model -->
-<!-- allows for between-class differences in (co)variances: -->
-<!-- # tabulate class-membership numbers -->
-<!-- table(summary(mod_g1_9)$classification) -->
-
-<!-- ## -->
-<!-- ##   1   2      3 -->
-<!-- ## 100 110     90 -->
-<!-- # display the means per class -->
-<!-- mod_g1_9$parameters$mean -->
-
-<!-- ##             [,1]         [,2]        [,3] -->
-<!-- ##   var1 2.023572      6.083019   2.0304980 -->
-<!-- ##   var2 3.056837      6.937785   4.8997993 -->
-<!-- ##   var3 1.718166      1.107873   7.9385872 -->
-<!-- ##   var4 6.086727      2.854402   6.9657536 -->
-<!-- ##   var5 7.101109      2.225675   4.7416266 -->
-<!-- ##   var6 3.691936      8.081570   2.2469977 -->
-<!-- ##   var7 5.009819      2.817889   0.8862725 -->
-<!-- ##   var8 8.316894      8.080891   3.9350536 -->
-<!-- ##   var9 1.978136      1.896294   9.1537131 -->
-<!-- ##   var10 1.191249     5.142590   8.0332628 -->
-<!-- # select the first of 3 10-by-10 variance matrices -->
-<!-- mod_g1_9$parameters$variance$sigma[1:10, 1:10, 1] -->
-
-<!-- ##               var1       var2       var3       var4       var5       var6       var7       var8 -->
-<!-- ##   var1    2.539537   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var2    0.000000   2.117941   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var3    0.000000   0.000000   2.509035   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var4    0.000000   0.000000   0.000000   2.812796   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var5    0.000000   0.000000   0.000000   0.000000   2.647399   0.000000   0.000000   0.000000 -->
-<!-- ##   var6    0.000000   0.000000   0.000000   0.000000   0.000000   2.694156   0.000000   0.000000 -->
-<!-- ##   var7    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.368379   0.000000 -->
-<!-- ##   var8    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.540704 -->
-<!-- ##   var9    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var10   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##               var9      var10 -->
-<!-- ##   var1    0.000000   0.000000 -->
-<!-- ##   var2    0.000000   0.000000 -->
-<!-- ##   var3    0.000000   0.000000 -->
-<!-- ##   var4    0.000000   0.000000 -->
-
-<!--                                                8 -->
-<!-- ##   var5    0.000000   0.000000 -->
-<!-- ##   var6    0.000000   0.000000 -->
-<!-- ##   var7    0.000000   0.000000 -->
-<!-- ##   var8    0.000000   0.000000 -->
-<!-- ##   var9    2.509146   0.000000 -->
-<!-- ##   var10   0.000000   2.204457 -->
-<!-- From this output, we can see that the three classes have class-sizes of 100, 110 and 90, -->
-<!-- respectively. These sizes along with the patterns of class-specific mean variable scores -->
-<!-- correspond with those that served as the input for the simulation. As specified, the variances -->
-<!-- vary slightly across variables. -->
-<!-- The class allocation in LPA is probabilistic in nature: each subject in the data is assigned a -->
-<!-- probability for each of the estimated classes, based on their pattern of scores on the input -->
-<!-- variables. These probabilities can be inspected in the z-matrix (here: mod_g1_9$z). Subjects -->
-<!-- can be allocated to one of the classes based on their highest class-probability. These posterior -->
-<!-- allocations can be found in the classification matrix (here: mod_g1_9$classification). -->
-<!-- These class-allocations were tabulated above to evaluate class sizes. -->
-<!-- Note that class-allocation in this way only yields useful classifications if the patterns of -->
-<!-- class-probabilities allow for allocation of each subject to a single class with sufficient certainty. -->
-<!-- In case of too much uncertainty, using the classification is not advised. We can inspect the -->
-<!-- uncertainty of allocation for all subjects (here: mod_g1_9$uncertainty). This gives us a list -->
-<!-- of uncertainties for all subjects. Next, we can evaluate the extent of uncertainty by looking -->
-<!-- at the maximum uncertainty or, for instance, the averaged uncertainty across subjects: -->
-<!-- max(mod_g1_9$uncertainty) -->
-
-<!-- ## [1] 0.003852604 -->
-<!-- mean(mod_g1_9$uncertainty) -->
-
-<!-- ## [1] 2.607299e-05 -->
-<!-- Here, the uncertainty is atypically low because of the used idealized data. In many cases, -->
-<!-- uncertainty is higher and it may be of interest to investigate it further, for instance, by -->
-<!-- looking at (e.g., plotting or tabulating) the uncertainty per class: -->
-<!-- cprob <- cbind(mod_g1_9$z, mod_g1_9$classification) -->
-<!-- cprob <- as.data.frame(cprob) -->
-<!-- colnames(cprob) <- c("prob (class 1)", "prob (class 2)", "prob (class 3)", "class") -->
-<!-- aggregate(cprob[, 1:3], list(cprob$class), mean) -->
-
-<!-- ##   Group.1 prob (class 1) prob (class 2) prob (class 3) -->
-<!-- ## 1       1   9.999824e-01   1.761611e-05   3.350070e-10 -->
-<!-- ## 2       2   5.486300e-05   9.999451e-01   7.192849e-08 -->
-<!-- ## 3       3   2.591358e-08   1.675400e-07   9.999998e-01 -->
-<!-- Here we can see that for each class, the mean probability of membership of that class was -->
-
-<!--                                                  9 -->
-<!-- very high and probabilities for the other classes were very low. Again, these values reflect the -->
-<!-- idealized data; in many cases, more differentiated class-probability patterns may be observed. -->
-
-<!-- 5.1.2 ‘tidyLPA’ -->
-<!-- When running ‘mclust’ models from the ‘tidyLPA’ package, the estimations are approached -->
-<!-- a little differently. By default, the number of models that can be estimated is restricted to -->
-<!-- six relatively common variants (see help(tidyLPA::estimate_profiles) for details). Of -->
-<!-- these models, four can be estimated with ‘mclust’. These models differ in terms of how the -->
-<!-- variances and covariances are allowed to vary or constrained to be equal across classes and -->
-<!-- whether covariances are or are not fixed to zero within classes. These models have each been -->
-<!-- allocated a number (1-6) in ‘tidyLPA’. See the table below how these numbers correspond to -->
-<!-- the above mentioned model configurations. -->
-<!--  Model variant    model number -->
-<!--  EEI              1 -->
-<!--  EEE              3 -->
-<!--  VVI              2 -->
-<!--  VVV              6 -->
-<!-- Models four and five are different and can only be fit when R is interfacing with Mplus using -->
-<!-- the ‘MplusAutomation’ package. This is outside the scope of this tutorial. -->
-<!-- To fit an LPA model in ‘tidyLPA’, we use the estimate_profiles() function. Here, we need -->
-<!-- to enter the data.frame to be used (df) and the number of classes to estimate (n_profiles). -->
-<!-- To determine what kind of model configuration to estimate, the authors have provided two -->
-<!-- different ways in the estimate_profiles() command. In the first approach, we simply tell what -->
-<!-- model number (see Table 2) we want to estimate using the models= argument For instance, if -->
-<!-- we want to estimate LPA models with 1 to 9 classes, with an EEI configuration, we can use: -->
-<!-- <!-- suppressMessages(library(tidyLPA)) --> -->
-<!-- <!-- suppressMessages(mod_1c_v1 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, --> -->
-<!-- <!--     models = 1)) --> -->
-
-<!-- The package issues a standard message by default notifying us that we use the models -->
-<!-- argument and that the variances and covariances arguments are ignored. Here, we suppress -->
-<!-- the message. The output shows us the model-fit information for the 1 to 9-class EEI models: -->
-<!-- <!-- mod_1c_v1 --> -->
-
-<!-- ## tidyLPA analysis using mclust: -->
-<!-- ## -->
-<!-- ## Model Classes AIC       BIC      Entropy            prob_min   prob_max   n_min   n_max   BLRT_p -->
-<!-- ## 1      1       14587.53 14661.61 1.00               1.00       1.00       1.00    1.00 -->
-<!-- ## 1      2       13160.84 13275.65 1.00               1.00       1.00       0.30    0.70    0.01 -->
-<!-- ## 1      3       11986.55 12142.11 1.00               1.00       1.00       0.30    0.37    0.01 -->
-<!-- ## 1      4       11982.14 12178.44 0.91               0.79       1.00       0.14    0.33    0.05 -->
-<!-- ## 1      5       11986.09 12223.14 0.89               0.75       1.00       0.06    0.33    0.30 -->
-
-<!--                                               10 -->
-<!-- ##   1      6         11986.25    12264.04   0.84      0.78       1.00        0.08   0.30    0.16 -->
-<!-- ##   1      7         11950.86    12269.39   0.81      0.77       0.91        0.08   0.17    0.01 -->
-<!-- ##   1      8         11939.04    12298.30   0.82      0.79       0.92        0.08   0.17    0.02 -->
-<!-- ##   1      9         11947.10    12347.11   0.82      0.76       0.92        0.06   0.16    0.57 -->
-<!-- Here, we can see that the BIC takes on different values compared to ‘mclust’, with lower -->
-<!-- rather than higher values indicating best fit (see explanation above). Instead of drawing the -->
-<!-- fit indices (e.g., BIC, AIC) directly from the ‘mclust’ package, ‘tidyLPA’ only draws the ‘raw’ -->
-<!-- log likelihood values and posterior probabilities from the ‘mclust’ output and (re)calculates -->
-<!-- the BIC, AIC, entropy etc. to mirror as well as possible those provided in Mplus. In addition, -->
-<!-- p-values of the the bootstrapped likelihood ratio test (BLRT) are given by default. These -->
-<!-- p-values indicate for each k-class model whether adding the kth class adds significantly to -->
-<!-- model fit. For instance, the BLRT_p of 0.04 for the 4-class model indicates that adding -->
-<!-- a fourth class led to an improvement of model fit compared to the 3-class model that was -->
-<!-- just significant when using an alpha of 0.05. When using ’mclust’ as a stand-alone package, -->
-<!-- the BLRT is not calculated by default, but can be obtained with the mclustBootstrapLRT -->
-<!-- function. -->
-<!-- Next, we can run similar commands, with different values for the models argument. It is -->
-<!-- also possible to estimate more than one model variant in a single run by providing a vector -->
-<!-- of model numbers (e.g., models=c(1,6)). -->
-<!-- The second approach to determine the model variant to be estimated is by using the variances -->
-<!-- and covariances arguments in the estimate_profiles command. The variances argument -->
-<!-- can have the values "equal" (i.e. variable variances constrained to be equal across classes) or -->
-<!-- "varying" (i.e. variances allowed to vary across classes). The covariances argument can -->
-<!-- have the values: "zero" (i.e. all variable covariances fixed to zero), "equal" (i.e. covariances -->
-<!-- constrained to be equal across classes) or "varying" (i.e. covariances allowed to vary across -->
-<!-- classes). Now, if we again want to estimate LPA models with 1 to 9 classes, with an EEI -->
-<!-- configuration, we can use: -->
-<!-- <!-- mod_1c_v2 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, variances = "equal", --> -->
-<!-- <!--     covariances = "zero") --> -->
-<!-- <!-- mod_1c_v2$model_1_class_2$fit --> -->
-
-<!-- <!-- ##         Model           Classes        LogLik                   AIC              AWE --> -->
-<!-- <!-- ## 1.000000e+00       2.000000e+00 -6.549418e+03          1.316084e+04     1.354347e+04 --> -->
-<!-- <!-- ##           BIC              CAIC           CLC                   KIC            SABIC --> -->
-<!-- <!-- ## 1.327565e+04       1.330665e+04 1.310084e+04           1.319484e+04     1.317734e+04 --> -->
-<!-- <!-- ##           ICL           Entropy      prob_min              prob_max            n_min --> -->
-<!-- <!-- ## -1.327568e+04      9.996750e-01 9.999484e-01           9.999990e-01     3.000000e-01 --> -->
-<!-- <!-- ##         n_max          BLRT_val        BLRT_p --> -->
-<!-- <!-- ## 7.000000e-01       1.448699e+03 9.900990e-03 --> -->
-<!-- As expected, the output is exactly the same as the one we got when using the models -->
-<!-- argument. Now, if we want to, we can continue fitting other model variants using either -->
-<!-- approach. Next, we can manually compare the models in terms of their BIC and AIC values -->
-
-<!--                                               11 -->
-<!-- or we can use compare_solutions to do this in an automated fashion. Interestingly, the -->
-<!-- best model can be selected based on the integrated information about several fit indices -->
-<!-- (analytic hierarchy process; see help(tidyLPA::AHP)) for more details). We obtain the model -->
-<!-- comparison with the following code. -->
-<!-- <!-- comp <- suppressWarnings(compare_solutions(mod_1c_v1)) --> -->
-<!-- <!-- comp$fits --> -->
-
-<!-- <!-- ##   # A tibble: 9 x 18 --> -->
-<!-- <!-- ##     Model Classes LogLik    AIC    AWE    BIC   CAIC    CLC    KIC SABIC      ICL --> -->
-<!-- <!-- ##     <dbl>   <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>           <dbl> --> -->
-<!-- <!-- ##   1     1       1 -7274. 14588. 14834. 14662. 14682. 14550. 14611. 14598. -14662. --> -->
-<!-- <!-- ##   2     1       2 -6549. 13161. 13543. 13276. 13307. 13101. 13195. 13177. -13276. --> -->
-<!-- <!-- ##   3     1       3 -5951. 11987. 12506. 12142. 12184. 11905. 12032. 12009. -12142. --> -->
-<!-- <!-- ##   4     1       4 -5938. 11982. 12638. 12178. 12231. 11878. 12038. 12010. -12218. --> -->
-<!-- <!-- ##   5     1       5 -5929. 11986. 12778. 12223. 12287. 11860. 12053. 12020. -12273. --> -->
-<!-- <!-- ##   6     1       6 -5918. 11986. 12915. 12264. 12339. 11838. 12064. 12026. -12353. --> -->
-<!-- <!-- ##   7     1       7 -5889. 11951. 13016. 12269. 12355. 11780. 12040. 11997. -12376. --> -->
-<!-- <!-- ##   8     1       8 -5873. 11939. 13141. 12298. 12395. 11747. 12039. 11991. -12409. --> -->
-<!-- <!-- ##   9     1       9 -5866. 11947. 13285. 12347. 12455. 11733. 12058. 12005. -12458. --> -->
-<!-- <!-- ##   # ... with 7 more variables: Entropy <dbl>, prob_min <dbl>, prob_max <dbl>, --> -->
-<!-- <!-- ##   #   n_min <dbl>, n_max <dbl>, BLRT_val <dbl>, BLRT_p <dbl> --> -->
-<!-- <!-- comp$best --> -->
-
-<!-- <!-- ## LogLik       AIC      AWE      BIC     CAIC        CLC    KIC   SABIC       ICL --> -->
-<!-- <!-- ##      9         8        3        3        3          9      3       8         3 --> -->
-<!-- As shown in the ‘mclust’ examples above, the EEI model (models=1) with 3 classes fit the -->
-<!-- data best, not only according to the BIC, but according to the integrated information from a -->
-<!-- broader range of fit indices. Note that we did not fit any competing EEE, VVV and VVI -->
-<!-- models here. If we would have wanted to compare fit to these model variants too, we should -->
-<!-- have actively told the package to fit these too, either by specifying a vector of different types -->
-<!-- of models to fit in the models argument or by repeating the analysis with different values for -->
-<!-- the variances and covariances arguments. -->
-
-<!-- 5.2 Comparison to Mplus -->
-<!-- For comparison, we fit the equivalent EEI model variants in Mplus. For this, we can mostly -->
-<!-- rely on the defaults of Mplus. We only need to specify that we want to estimate class-specific -->
-<!-- variable means. The variances are automatically constrained to be equal across classes and -->
-<!-- within class variable covariances are fixed to 0. Here we have the separate codes for fitting -->
-<!-- the first 4 models wit 1-4 class models (code for the 5-9 class models not shown). -->
-<!-- <!-- !! 1-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-
-<!-- <!--                                                  12 --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(1); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-
-<!-- <!-- !! 2-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(2); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#2% --> -->
-<!-- <!--   [v1-v10]; --> -->
-
-<!-- <!-- !! 3-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(3); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#2% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#3% --> -->
-<!-- <!--   [v1-v10]; --> -->
-
-<!-- <!--                             13 --> -->
-<!-- <!-- !! 3-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(4); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#2% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#3% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#4% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- Each of the models was fitted with multiple initial stage random starts and multiple final -->
-<!-- stage optimizations to reduce the risk of finding solutions on a local maximum. For the 1-6 -->
-<!-- class models, 250 iniitial and 50 final starts were used. For the 7-class model 1000 initial -->
-<!-- starts and 200 final optimizations were used and for the 8- and 9-class models, a replicable -->
-<!-- global solution could only be obtained with 5000 initial stage starts and 1000 final stage -->
-<!-- optimizations. The fit indices for each of the fitted models is displayed in the table below. -->
-<!-- For now, we only look at the estimated fit indices. -->
-<!--                 EEI models -->
-<!--  Classes   AIC       BIC -->
-<!--  1         14587.535 14587.535 -->
-<!--  2         13160.836 13275.653 -->
-<!--  3         11986.551 12142.11 -->
-<!--  4         11957.71  12154.01 -->
-<!--  5         11951.766 12188.808 -->
-<!--  6         11944.632 12222.415 -->
-<!--  7         11935.501 12254.027 -->
-<!--  8         11929.428 12288.695 -->
-<!--  9         11925.203 12325.212 -->
-<!-- If we compare the BIC values for the EEI models to those obtained with ‘mclust’ directly -->
-<!-- and via ‘tidyLPA’, we can see that the results are quite similar for the less complex models, -->
-<!-- with the BIC-values for the 2- and 3-class models being exactly the same in terms of their -->
-<!-- absolute values. When taking a closer look at the 3-class model that was previously found to -->
-<!-- be the best model, we can also see that that the Mplus-based classifications (n=100, n=110, -->
-<!-- and n=90) and entropy estimates (entropy=1.0) are similar to those obtained with ‘mclust’ -->
-
-<!--                                              14 -->
-<!-- and/or ‘tidyLPA’. -->
-<!--  If we look at the larger picture and evaluate the overlap between the BIC values obtained -->
-<!--  for all models with class numbers ranging from 1 to 9, we can see that the estimated BIC -->
-<!-- values show more differences between the R-packages and Mplus for the models with larger -->
-<!--  numbers of classes. This can be explained by the different methods used by Mplus and -->
-<!-- ‘mclust’/’tidyLPA’ to generate the start values for estimation (see below). These differences -->
-<!--  can cause the packages to yield different results for more complex models and/or models that -->
-<!--  are increasingly misspecified (as the models with more than 3 classes were in our example). -->
-
-<!-- 6. Final comments -->
-<!-- In this tutorial we have seen that it is relatively easy to run an LPA in R. As with all -->
-<!-- data-driven analyses, care should be taken not to over-interpret the results. In the end, LPA -->
-<!-- identifies classes in such a way as to optimally explain variance on a range of variables and -->
-<!-- not to optimize interpretability or usefulness. -->
-<!-- Another remark with regard to LPA results is that class-membership is probabilistic in nature: -->
-<!-- each subject in an analyzed sample has a probability of being in each of the model’s classes. -->
-<!-- Subjects can be allocated to a class based on their highest class-probability. Importantly, -->
-<!-- this can only be done with enough certainty if separation between classes is sufficient, which -->
-<!-- means that we see that each subject has a clear highest probability (e.g., p=0.9) for one of -->
-<!-- the classes. The entropy statistic is often used to quantify this separation, with values <0.8 -->
-<!-- being taken to indicate insufficient separation to allocate subjects to one class. In such cases, -->
-<!-- the class probabilities themselves can still be used in further investigations. -->
-<!-- Depending on one’s preferences, one can choose either the ‘mclust’ or the ‘tidyLPA’ package. -->
-<!-- The latter may be especially attractive for persons that already have experience with Mplus -->
-<!-- or comparable software. It is important to note, though, that differences exist between the -->
-<!-- estimation approaches taken by both R-packages and by Mplus. -->
-<!-- Mixture models such as LPA need to be estimated in an iterative fashion, using an EM -->
-<!-- algorithm that is very sensitive to the used start values, where poor start values can lead the -->
-<!-- estimation process to a poor solution. In addition, there is often a chance that the iterative -->
-<!-- EM process leads to a solution that is at a local, rather than a global maximum in the -->
-<!-- likelihood ‘landscape’. -->
-
-### Starting values
-
-<!-- Different methods have been developed to initialize the estimation (get starting values) in -->
-<!-- such a way as to optimize the chance of arriving at an accurate model solution (Biernacki et -->
-<!-- al., 2003; Shireman et al., 2016). Mplus and ‘mclust’ take two different approaches. Mplus -->
-<!-- uses a ‘brute force’ approach and reruns the model with multiple sets of random start values, -->
-<!-- each generated from uniform distributions of values with ranges that are based on the data -->
-<!-- (Muthén & Muthén, 1998-2015). The best solution is the model with the highest log likelihood -->
-<!-- that was arrived upon from at least two different starting points. In contrast, ‘mclust’ uses -->
-<!-- the data itself to generate a set of plausible start values using hierarchical clustering. This -->
-<!-- means that the start values are informed by the (hierarchical structure of) the data, which -->
-<!-- can work well. A downside is that only a single set of start values is used, so the risk of -->
-<!-- arriving at a local optimum is not addressed. These approaches show clear differences and -->
-<!-- have been shown previously to not always arrive at the same solutions, with some authors -->
-<!-- being especially critical of the ‘mclust’ approach (Shireman et al., 2017). In our example, -->
-<!-- we did not see much difference between the two approaches, but it should be noted that -->
-<!-- many real-world data sets do not have such a clear latent structure, making it much more -->
-<!-- challenging to arrive at an accurate solution, given the data. -->
-<!-- Based on the provided examples and these technical differences between ‘mclust’ and Mplus, -->
-<!-- it is probably better to see ‘mclust’ and ‘tidyLPA’ as useful alternatives for Mplus, rather -->
-<!-- than as tools to exactly mimic what can be done with the latter software. -->
-
-<!-- References -->
-<!-- Banfield, J., Raftery, A.E. (1993). Model-based Gaussian and non-Gaussian clustering. -->
-<!-- Biometrics. 49: 803–821. -->
-<!-- Biernacki,C.,Celeux,G., Govaert,G.(2003).Choosing starting values for the EM algorithm -->
-<!-- for getting the highest likelihood in multivariate Gaussian mixture models. Computational -->
-<!-- Studies and Data Analysis, 41, 561–575. -->
-<!-- Celeux, G., Govaert, G. (1995) Gaussian parsimonious clustering models. Pattern Recognition. -->
-<!-- 28:781–793 -->
-<!-- Fraley, C., Raftery, A.E. (1998) How many clusters? Which clustering method? Answers via -->
-<!-- model-based cluster analysis. The Computer Journal 41: 578-588. -->
-<!-- Gibson, W. A. (1959). Three multivariate models: Factor analysis, latent structure analysis, -->
-<!-- and latent profile analysis. Psychometrika, 24, 229–252. -->
-<!-- Linzer, D.A., Lewis J. (2011). “poLCA: an R Package for Polytomous Variable Latent Class -->
-<!-- Analysis.” Journal of Statistical Software. 42(10): 1-29 -->
-<!-- Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., Leisch, F. (2019). e1071: Misc -->
-<!-- Functions of the Department of Statistics, Probability Theory Group (Formerly: E1071), TU -->
-<!-- Wien. R package version 1.7-3. https://CRAN.R-project.org/package=e1071 -->
-<!-- Muthén, L., Muthén, B., 1998–2015. Mplus User’s Guide, 7th ed., Muthén & Muthén, Los -->
-<!-- Angeles. -->
-<!-- Neale, M.C., Hunter, M.D., Pritikin, J.N., Zahery, M., Brick, T.R., Kirkpatrick, R.M., -->
-<!-- Estabrook, R., Bates, T.C., Maes, H.H., Boker, S.M. (2016). “OpenMx 2.0: Extended -->
-<!-- structural equation and statistical modeling.” Psychometrika 81(2): 535-549. -->
-<!-- Nylund-Gibson, K., Choi, A. Y. (2018). Ten frequently asked questions about latent class -->
-<!-- analysis. Translational Issues in Psychological Sciences, 4, 440–461. -->
-<!-- Oberski, D. (2016). Mixture models: Latent profile and latent class analysis. In J. Robertson -->
-<!-- & M. Kaptein (Eds.), Modern statistical methods for HCI (pp. 275–287). Cham, Switzerland: -->
-<!-- Springer Inter-national Publishing. -->
-
-<!--                                              16 -->
-<!-- Pastor, D.A., Barron, K.E., Miller, B.J., Davis, S.L. (2007). A latent profile analysis of -->
-<!-- college students’ achievement goal orientation. Contemporary Educational Psychology 32(1): -->
-<!-- 8-47. -->
-<!-- Rosenberg, J.M., Beymer, P.N., Anderson, D.J., Van Lissa, C.J., & Schmidt, J.A. (2018). -->
-<!-- tidyLPA: An R Package to Easily Carry Out Latent Profile Analysis (LPA) Using Open-Source -->
-<!-- or Commercial Software. Journal of Open Source Software 3(30): 978. -->
-<!-- Scrucca, L., Fop M., Murphy, T.B., Raftery A.E. (2016) mclust 5: clustering, classification -->
-<!-- and density estimation using Gaussian finite mixture models The R Journal 8/1: 289-317. -->
-<!-- Shireman, E. M., Steinley, D., & Brusco, M. J. (2016). Local Optima in Mixture Modeling. -->
-<!-- Multivariate Behavioral Research 51(4): 466–481. -->
-<!-- Shireman, E., Steinley, D., Brusco, M.J. (2017). Examining the effect of initialization -->
-<!-- strategies on the performance of Gaussian mixture modeling. Behavioral Research Methods -->
-<!-- 49(1):282-293. -->
-<!-- Sterba, S.K. (2013). Understanding Linkages Among Mixture Models. Multivariate Behav -->
-<!-- Res. 2013 48(6): 775-815. -->
-<!-- Wolfe, J. (1970). Pattern clustering by multivariate mixture analysis. Multivariate Behavioral -->
-<!-- Research, 5: 329–350. -->
-
-<!-- Shireman 2016: -->
-<!-- https://www.tandfonline.com/doi/full/10.1080/00273171.2016.1160359?casa_token=YMvI52JWPq0AAAAA%3A20nlLRukoCkUmIil2lRWnsUS-33AlU6HmDRGsu07aJ0nTzS-skjvHnroJBtBVHM5HE0X8iStM1eyqw -->
-
-<!-- mixture models have two possible uses: either (a) classifying observations into groups through the estimation of distinct, underlying distributions or (b) using a set of well-defined parametric distributions (often times the Gaussian distribution) to approximate a messy data space that cannot be approximated by standard models that assume a single distribution (whether univariate or multivariate).
-While both are legitimate uses of mixture modeling, the only necessary and sufficient condition for fitting a mixture model is nonnormality in the data (see Bauer & Curran, 2003).
-It is certainly true that if there are distinct groups arising from distinct distributions, it is necessary to have a set of parameters to model each of those groups.
-Unfortunately, it may often be the case (in fact, it is probably likely) we are in the converse situation where the full data space is either nonlinear, asymmetric, or both, and there exist no distinct groups in the population.
-In this case, the mixture model is capturing the departure from “clean” structure, but the groups themselves are not meaningful in the sense that they are giving rise to distinct subpopulations.
-As such, their boundaries are going to be amorphous, and their parameters, regardless of their stability, lose meaning in terms of reflecting an underlying data-generative mechanism, making the interpretation of the means, covariances, or a specific individuals' membership an exercise in folly.
-Obviously, knowing which situation a data set falls in is paramount, and using the prevalence of local optima as a potential indicator is a crucial first step.
--->
-
-<!-- The optimal number of latent classes (i.e.
-identity configurations) was determined using the BIC (Fraley & Raftery, 1998); with lower values indicating a better statistical fit.
-Solutions were considered to be admissible if they consisted of at least 5% of the sample (Depaoli, 2013), if the entropy (a measure of class separability) exceeded .8 (Wang, Deng, Yang, 2017), and if the solution was theoretically interpretable and qualitatively distinct from the other class solutions (Howard & Hoffman, 2018).
--->
+\end{verbatim}
+
+As the open science movement is gaining momentum,
+researchers are becoming increasingly aware
+how important it is that analyses can be reproduced and audited.
+In line with open science principles, one of the suggested reporting standards relates to reproducible code.
+In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., \emph{Mplus} and \emph{Latent GOLD}).
+A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because the underlying source code is proprietary.
+To overcome these limitations, the present paper introduces new user-friendly functions in the \texttt{tidySEM} R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package \texttt{OpenMx}.
+The reporting guidelines described in this paper are adopted in \texttt{tidySEM} by default.
+The \texttt{tidySEM} R-package thus makes advanced mixture modeling based on best practices widely accessible,
+and facilitates the adoption of the estimation and reporting guidelines described in this paper.
+
+\hypertarget{best-practices-in-visualization}{%
+\subsection{Best practices in visualization}\label{best-practices-in-visualization}}
+
+\hypertarget{tutorial}{%
+\section{Tutorial}\label{tutorial}}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+--\textgreater{}
+
+\hypertarget{starting-values-1}{%
+\subsubsection{Starting values}\label{starting-values-1}}
+
+\end{document}

### 2022-08-01 manuscript.Rmd
update 2022-08-01 10:01:15

@@
 
-Latent class analysis has become quite popular across scientific fields,  and under a number of different names.
+Latent class analysis is an umbrella term that refers to a number of techniques for estimating unobserved group membership
+based on a parametric model of one or more observed indicators of group membership.
+This method has become quite popular across scientific fields, and under a number of different names; most notably (finite Gaussian) mixture modeling and latent profile analysis.
 <!-- Vermunt and Magidson (2004) defined it more generally as virtually any statistical model where “some of the parameters … differ across unobserved subgroups” (Vermunt, J.K.
@@
 -->
-The purpose of this technique is to estimate unobserved group membership
-based on a parametric model of one or more observed indicators of group membership.
-Despite the popularity of the method, there is a noted absence of guidelines for estimating and reporting latent class analyses [].
+Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
 <!-- Prior work resulted in the "GRoLTS-Checklist"; reporting guidelines for a specific subcategory of latent class analysis known as latent growth models. vermuntj.k.LatentClassAnalysis2004-->
@@
 and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on best practice.
-
-Throughout this paper, examples make use of free, open source software for latent class analysis in R.
-
-<!-- In practice, this technique is often used to  -->
+The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
+Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, this paper exclusively relies on free, open source software for latent class analysis in R.
+Our goal is to make best-practices latent class analysis widely accessable.
 
@@
 
-When a latent class analysis relies on categorical indicators,
-typically of a binary or ordinal measurement level,
-it has a different parameterization than a mixture model.
-One common assumption is that each categorical variable reflects an underlying standard normal distribution.
-The parameters in such a model are then "thresholds", corresponding to the quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
+The term latent class analysis originally referred to cases where the observed indicators were categorical.
+Nowadays, it is more commonly used as an umbrella term.
+To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as latent class analysis with ordinal indicators.
+Latent class models with ordinal indicators are parameterized differently from mixture models.
+One common parameterization assumes that each categorical variable reflects an underlying standard normal distribution.
+The parameters are "thresholds" that correspond to quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
 These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
@@
 For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses.
-Mixture modeling can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
+Latent class analysis can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
 If results indicate that the data are better described by a different number of classes,
@@
 
+<!-- CJ: Include these use cases too: -->
+Descriptive analysis; identify few prototypes based on many variables
 <!-- Inductively identify number of latent classes -->
@@
 <!-- Dimension reduction -->
-<!-- Violation of assumptions -->
+<!-- Deal with violation of assumptions, e.g. when data are clearly not normally distributed -->
 
@@
 The best practices in estimation, as outlined in Table \@ref(tab:checkest),
-are inspired by prior recommendations for best practices for estimating
+are rooted in existing recommendations for best practices for estimating
 specific sub-types of latent class analyses,
@@
 ```
+
 ### Examine observed data
@@
 Pay special attention to level of measurement of the indicators.
-Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions [see @vermuntKmeansMayPerform2011].
+Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
+Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermuntKmeansMayPerform2011].
 Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
-suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is perhaps the most common source of user error.
+suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
 Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @normanLikertScalesLevels2010],
@@
 It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
-The problem becomes eggregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
+The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
 In sum, Likert-type scales are rarely suitable for mixture modeling;
-a latent class analysis with ordinal indicators may be more appropriate.
+latent class analysis with ordinal indicators is more appropriate.
+
+Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [REF SPURK].
+This is incorrect.
+When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
+<!-- CJ: Expand on this here -->
 
@@
 Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
-For this tutorial, I have re-implemented it in the `mice` package as `mice::MCAR()`,
+For this tutorial, I have re-implemented it in the `mice` package as `mice::mcar()`,
 with a fast backend in C++ and new printing and plotting methods.
@@
 To conclude; our recommendation is to inspect missingness (e.g., using `mice::MCAR()`)
-and report the number of missings per variable (e.g., using `tidySEM::descriptives()`),
+and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
 before proceeding with FIML.
@@
 One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer the most parsimoneous model that has best fit according to any of the available fit indices.
+Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
 Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogulComparisonInformationCriteria2016].
@@
 Many software packages are available for the estimation of latent class analyses.
-Some of these packages have limited functionality, or implement particular innovations.
+Some of these packages have limited functionality, or implement specific innovations.
 Other packages implement latent class analyses in the context of a more flexible structural equation modeling framework.
@@
 and implement sensible defaults for complex analyses, including latent class analysis.
-This somewhat lowers the threshold for applied researchers to adopt such methods.
+This lowers the threshold for applied researchers to adopt such methods.
 Commercial software also has several downsides, however.
@@
 If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
-This latter property allows it to escape local optima.
+This latter property allows it to escape local optima, and find the global optimum.
 
@@
 
-In the case of Latent Profile Analysis, it has been suggested that when tests of multivariate normality indicate a violation, 
-maximum likelihood with robust standard errors should be used (Spurk et al., 2020). 
-This statement is problematic for two reasons. 
-Firstly, multivariate normality will always be violated in mixture models by definition. 
-We assume that the population consists of several groups each characterized by its own normal distribution. 
-When two or more normal distributions with different means and standard deviations comprise a mixed population distribution, 
-this combined distribution is no longer normal. 
-For example, you can imagine a bimodal distribution when dealing with two latent classes. 
-If this were not the case and our mixed population distribution were normal, 
-there would be no latent classes to extract as the population would consists of a single class only. 
-For this reason, multivariate normality will always be violated when the population is comprised of several latent groups. 
+One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators of a latent class analysis are not normally distributed (Spurk et al., 2020). 
+This statement is incorrect, and may lead readers to believe that they must use commercial software to estimate latent class analyses, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
+
+As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
+<!-- We assume that the population consists of several groups each characterized by its own normal distribution.  -->
+<!-- When two or more normal distributions with different means and standard deviations comprise a mixed population distribution,  -->
+<!-- this combined distribution is no longer normal.  -->
+<!-- For example, you can imagine a bimodal distribution when dealing with two latent classes.  -->
+<!-- If this were not the case and our mixed population distribution were normal,  -->
+<!-- there would be no latent classes to extract as the population would consists of a single class only.  -->
+<!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups.  -->
 Secondly, the claim that maximum likelihood with robust standard errors should be preferred 
@@
 
-### Starting values
-
-The EM algorithm needs starting values for each parameter it wants to estimate. 
-If it were provided a single starting value for each parameter, 
-the EM optimizer would be at risk of finding a local optimum instead of the global optimum. 
-In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates. 
-To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter
-in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004). 
-Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values.
-<!-- add more on this -->
-Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
-
-Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence
-or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004).
-The key is to use a large number of starting value sets. 
-Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models,
-and a greater number of sets for more complex models. 
-A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately.
+<!-- ### Starting values -->
+
+<!-- The EM algorithm needs starting values for each parameter it wants to estimate.  -->
+<!-- If it were provided a single starting value for each parameter,  -->
+<!-- the EM optimizer would be at risk of finding a local optimum instead of the global optimum.  -->
+<!-- In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates.  -->
+<!-- To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter -->
+<!-- in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004).  -->
+<!-- Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values. -->
+<!-- <!-- add more on this --> -->
+<!-- Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).  -->
+
+<!-- Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence -->
+<!-- or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004). -->
+<!-- The key is to use a large number of starting value sets.  -->
+<!-- Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, -->
+<!-- and a greater number of sets for more complex models.  -->
+<!-- A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately. -->
 

### 2022-08-01 manuscript.Rmd
update 2022-08-01 11:18:11

@@
 <!-- CJ: Include these use cases too: -->
-Descriptive analysis; identify few prototypes based on many variables
-<!-- Inductively identify number of latent classes -->
-<!-- Classify individuals -->
+<!-- Descriptive analysis; identify few prototypes based on many variables, e.g. DJA science funding -->
+<!-- Inductively identify number of latent classes: I know it's a group, but not how many groups -->
+<!-- Classify individuals, e.g. welzijnsmeter -->
 <!-- Identify items (observed characteristics) that -->
@@
 and use these as starting values for a mixture model.
+<!-- In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
+Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
+[@biernackiChoosingStartingValues2003].
+-->
 One remaining concern is that this approach may result in starting values close to a local optimum,
@@
 
-One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators of a latent class analysis are not normally distributed (Spurk et al., 2020). 
-This statement is incorrect, and may lead readers to believe that they must use commercial software to estimate latent class analyses, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
-
+One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed (Spurk et al., 2020). 
+This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
 As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
+<!-- CJ: Is there even any literature that says robust standard erorrs are better with latent class analysis? -->
+
 <!-- We assume that the population consists of several groups each characterized by its own normal distribution.  -->
@@
 <!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups.  -->
-Secondly, the claim that maximum likelihood with robust standard errors should be preferred 
-implicitly implies that only software packages with this estimation technique can be used.
+<!-- Secondly, the claim that maximum likelihood with robust standard errors should be preferred  -->
+<!-- implicitly implies that only software packages with this estimation technique can be used.  -->
 
@@
 
-[@nylund-gibsonTenFrequentlyAsked2018; ]
-
-    In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
-Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
-[@biernackiChoosingStartingValues2003]
-
-### Assessing Estimation Results
-<!-- maybe change to class enumeration -->
+### Class enumeration
 
@@
 with each additional model estimating one more class than the previous model.
+<!-- CJ: Not always; you may have a theory about a fixed number. -->
+<!-- CJ: Argue that, even in this case, it is useful to know if the theoretical model is markedly better than other numbers of classes -->
 The final model called the final class solution is chosen based on both theoretical and statistical criteria.
@@
 and the Bayes factor (Weller, Bowen & Faubert, 2020).
-
-Absolute model fit can be assessed using the likelihood ratio (LR) $\chi^2$ goodness-of-fit test. 
-It can compare two nested models when they have the same number of classes (Lanza et al., 2003).
-Typically, we compare our final class solution (a constrained model) 
-to an unconstrained one (a model characterized by the perfect fit).
-The former model is a special, less complex version of the latter.
-In this case, we compare our model-based response pattern frequencies to those in the dataset.
-Therefore, we make a statement about how well our model fits the data.
-The perfect fit would be indicated by the $\chi^2\ =\ 0$ and $p\ =\ 1$.
-Ideally, the LR $\chi^2$ test will be non-significant as this implies
-that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model.
-The underlying assumption of this procedure is that the test statistic is asymptotically distributed
-as a $\chi^2$ meaning that it can be approximated using this distribution.
-This is the case when the sample size is adequate (Lanza et al., 2003). 
-The dependence of the LR $\chi^2$ test statistic on sample size has its downsides (Masyn, 2013). 
-Very large samples will often effortlessly reject the null even when the final class solution is a good fit,
-and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic.
-In fact, small samples might not be well approximated using the $\chi^2$ distribution,
-which is a problem shared with large yet sparse datasets.
-In such cases, special caution is advised when interpreting the resulting p-values (Masyn, 2013).
 
@@
 The likelihood ratio test can only compare two models at a time (Lanza et al., 2003).
+
+### Model fit indices
+
+<!-- Absolute model fit can be assessed using the likelihood ratio (LR) $\chi^2$ goodness-of-fit test.  -->
+<!-- It can compare two nested models when they have the same number of classes (Lanza et al., 2003). -->
+<!-- Typically, we compare our final class solution (a constrained model)  -->
+<!-- to an unconstrained one (a model characterized by the perfect fit). -->
+<!-- The former model is a special, less complex version of the latter. -->
+<!-- In this case, we compare our model-based response pattern frequencies to those in the dataset. -->
+<!-- Therefore, we make a statement about how well our model fits the data. -->
+<!-- The perfect fit would be indicated by the $\chi^2\ =\ 0$ and $p\ =\ 1$. -->
+<!-- Ideally, the LR $\chi^2$ test will be non-significant as this implies -->
+<!-- that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model. -->
+<!-- The underlying assumption of this procedure is that the test statistic is asymptotically distributed -->
+<!-- as a $\chi^2$ meaning that it can be approximated using this distribution. -->
+<!-- This is the case when the sample size is adequate (Lanza et al., 2003).  -->
+<!-- The dependence of the LR $\chi^2$ test statistic on sample size has its downsides (Masyn, 2013).  -->
+<!-- Very large samples will often effortlessly reject the null even when the final class solution is a good fit, -->
+<!-- and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic. -->
+<!-- In fact, small samples might not be well approximated using the $\chi^2$ distribution, -->
+<!-- which is a problem shared with large yet sparse datasets. -->
+<!-- In such cases, special caution is advised when interpreting the resulting p-values (Masyn, 2013). -->
 
@@
 <!-- decide whether to keep the following paragraph, if so, where to place it -->
-When estimators which require sets of starting values are used,
-each fitted model will have a corresponding log likelihood value. 
-Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times 
-with each log likelihood value generated by a different starting value set. 
-Successful replications of the largest log likelihood value across different starting values
-instill confidence that the estimation was successful and we indeed found the global maximum (Masyn, 2013). 
-The model with the highest replicated likelihood will be chosen as our best-fit. 
-If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012).
-Such a model does not have a unique solution. For a discussion on model identification for LCA, see Masyn (2013).
-Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated 
-in at least 3-10% of the models initialized with different starting values in the first step of the optimization. 
-This consideration does not apply in case of models which use simulated annealing since these do not have starting values.
+<!-- When estimators which require sets of starting values are used, -->
+<!-- each fitted model will have a corresponding log likelihood value.  -->
+<!-- Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times  -->
+<!-- with each log likelihood value generated by a different starting value set.  -->
+<!-- Successful replications of the largest log likelihood value across different starting values -->
+<!-- instill confidence that the estimation was successful and we indeed found the global maximum (Masyn, 2013).  -->
+<!-- The model with the highest replicated likelihood will be chosen as our best-fit.  -->
+<!-- If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012). -->
+<!-- Such a model does not have a unique solution. For a discussion on model identification for LCA, see Masyn (2013). -->
+<!-- Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated  -->
+<!-- in at least 3-10% of the models initialized with different starting values in the first step of the optimization.  -->
+<!-- This consideration does not apply in case of models which use simulated annealing since these do not have starting values. -->
 <!-- Masyn 2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
@@
 
-Conditional item probabilities indicate the probability of a positive score on an item 
-given that the person belongs to a particular latent class. 
+<!-- CJ: This is about model interpretation; conditional item probabilities are the analogue of mean and standard deviation, but when the observed data are biary/ordinal  -->
+<!-- Conditional item probabilities indicate the probability of a positive score on an item  -->
+<!-- given that the person belongs to a particular latent class.  -->
+
 <!-- If a particular item is endorsed by two or more classes at markedly different rates, -->
@@
 the classes have high homogeneity with respect to this item (Masyn, 2013). MY OWN WRITING BUT NOT FLOWY ENOUGH-->
-However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1), 
-we are observing boundary parameter estimates. This can be a sign of an invalid solution, 
-warn us that too many classes were extracted, or indicate a local optimum. 
-Boundary parameter estimates are more likely to happen when working with sparse data 
-and should be interpreted with caution (Geiser, 2012).
-
-An important outcome of LCA are unconditional class probabilities. 
-These communicate the probability that an individual belongs to each class. 
-A limitation is that our final class solution need not reflect true class membership. 
-Once we do accept a particular class solution, special care also needs to be taken when naming the classes. 
+<!-- However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1),  -->
+<!-- we are observing boundary parameter estimates. This can be a sign of an invalid solution,  -->
+<!-- warn us that too many classes were extracted, or indicate a local optimum.  -->
+<!-- Boundary parameter estimates are more likely to happen when working with sparse data  -->
+<!-- and should be interpreted with caution (Geiser, 2012).  -->
+
+<!-- CJ: THis talks about the problem of inadmissible solultions.
+With ordinal items, that can happen because of sparse cross-tables (e.g., some combinations of responses never occur).
+WIth continuous items, that could happen e.g.
+because you have a zero-inflated, otherwise normally distributed variable; one class would have a mean at zero and its SD would be hard to determine.
+Or it could happen if you model a five-category Likert scale with a five-component latent profile analysis; each response category will have its own mean, but there's no variance around that (everybody in class 3 has exactly the response value 3).
+These are all the same problems.
+-->
+
+<!-- CJ: Merge with beginning of the paper, where we explain that LCA simultaneously estimates class membership and the descriptives of those classes. -->
+<!-- An important outcome of LCA are unconditional class probabilities.  -->
+<!-- These communicate the probability that an individual belongs to each class.  -->
+<!-- A limitation is that our final class solution need not reflect true class membership.  -->
+<!-- Once we do accept a particular class solution, special care also needs to be taken when naming the classes.  -->
+
+### Interpreting the results
+
 Class names should be chosen to accurately reflect group membership. 

### 2022-08-01 manuscript.tex
update 2022-08-01 11:27:43

@@
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 \ifLuaTeX
@@
 
-Latent class analysis has become quite popular across scientific fields, and under a number of different names.
-The purpose of this technique is to estimate unobserved group membership
+Latent class analysis is an umbrella term that refers to a number of techniques for estimating unobserved group membership
 based on a parametric model of one or more observed indicators of group membership.
-Despite the popularity of the method, there is a noted absence of guidelines for estimating and reporting latent class analyses {[}{]}.
+This method has become quite popular across scientific fields, and under a number of different names; most notably (finite Gaussian) mixture modeling and latent profile analysis.
+Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
 This complicates manuscript review and assessment of the quality of published studies,
 and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on best practice.
-
-Throughout this paper, examples make use of free, open source software for latent class analysis in R.
+The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
+Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, this paper exclusively relies on free, open source software for latent class analysis in R.
+Our goal is to make best-practices latent class analysis widely accessable.
 
@@
 One difference between the two techniques is that factor analysis can be considered as a way to group observed \emph{variables} into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
-By contrast, mixture modeling groups \emph{individuals} into classes (see \textbf{nylund-gibsonTenFrequentlyAsked2018?}).
+By contrast, mixture modeling groups \emph{individuals} into classes (see Nylund-Gibson \& Choi, 2018).
 In line with this distinction, latent class analysis is sometimes referred to as a ``person-centered'' technique, and factor analysis as a ``variable-centered'' technique.
@@
 Finally, in the context of machine learning, latent class analysis can be considered as an
-\emph{unsupervised classification} problem (\textbf{figueiredoUnsupervisedLearningFinite2002?}).
+\emph{unsupervised classification} problem (Figueiredo \& Jain, 2002).
 The term \emph{unsupervised} refers to the fact that the outcome variable, true class membership,
@@
 In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
-From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators (\textbf{vermuntj.k.LatentClassAnalysis2004?}).
+From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators (Vermunt, J.K. et al., 2004).
 Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
@@
 
-When a latent class analysis relies on categorical indicators,
-typically of a binary or ordinal measurement level,
-it has a different parameterization than a mixture model.
-One common assumption is that each categorical variable reflects an underlying standard normal distribution.
-The parameters in such a model are then ``thresholds'', corresponding to the quantiles of a standard normal distribution (with \(N(\mu = 0, \sigma = 1)\)).
+The term latent class analysis originally referred to cases where the observed indicators were categorical.
+Nowadays, it is more commonly used as an umbrella term.
+To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as latent class analysis with ordinal indicators.
+Latent class models with ordinal indicators are parameterized differently from mixture models.
+One common parameterization assumes that each categorical variable reflects an underlying standard normal distribution.
+The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution (with \(N(\mu = 0, \sigma = 1)\)).
 These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
@@
 For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses.
-Mixture modeling can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
+Latent class analysis can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
 If results indicate that the data are better described by a different number of classes,
@@
 or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical (\textbf{baughmanMixtureModelAnalysis2006?}) or mental (\textbf{wuAbuseDependencePrescription2011?}) health problems.
+For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
 The example of shoe size is a rudimentary illustration of this type of application.
@@
 The best practices in estimation, as outlined in Table \ref{tab:checkest},
-are inspired by prior recommendations for best practices for estimating
+are rooted in existing recommendations for best practices for estimating
 specific sub-types of latent class analyses,
 including latent class growth analysis (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?})
-and latent class analysis with ordinal indicators (e.g., \textbf{nylund-gibsonTenFrequentlyAsked2018?}).
+and latent class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
 These were generalized to be more relevant to all types of latent class analyses,
@@
 Pay special attention to level of measurement of the indicators.
-Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions (see \textbf{vermuntKmeansMayPerform2011?}).
+Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
+Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models (see Vermunt, 2011).
 Personal experience consulting on latent class analyses and moderating the \texttt{tidyLPA} Google group
-suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is perhaps the most common source of user error.
-Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous (e.g., \textbf{normanLikertScalesLevels2010?}),
+suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
+Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous (e.g., Norman, 2010),
 this certainly does not imply that all methods are.
 It is certainly unlikely that such ordinal variables can be treated as a \emph{mixture} of multiple normal distributions.
-The problem becomes eggregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
+The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
 In sum, Likert-type scales are rarely suitable for mixture modeling;
-a latent class analysis with ordinal indicators may be more appropriate.
+latent class analysis with ordinal indicators is more appropriate.
+
+Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed {[}REF SPURK{]}.
+This is incorrect.
+When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
 
@@
 Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible (consider using a GitHub repository as a comprehensive public research archive, as explained in \textbf{vanlissaWORCSWorkflowOpen2020?}).
+Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible (consider using a GitHub repository as a comprehensive public research archive, as explained in Van Lissa et al., 2020).
 
@@
 Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?}).
-Three types of missingness have been distinguished in the literature (\textbf{rubinInferenceMissingData1976?}):
+Three types of missingness have been distinguished in the literature (Rubin, 1976):
 Missing completely at random (MCAR), which means that missingness is random;
@@
 It is possible to conduct a so-called ``MCAR'' test,
-for example the non-parametric MCAR test (\textbf{jamshidianTestsHomoscedasticityNormality2010?}).
+for example the non-parametric MCAR test (Jamshidian \& Jalal, 2010).
 But note that the name ``MCAR test'' is somewhat misleading,
@@
 A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (\textbf{littleTestMissingCompletely1988?}).
+As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (Little, 1988).
 This assumption is tenuous in the context of latent class analysis.
-A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable (\textbf{jamshidianTestsHomoscedasticityNormality2010?}).
+A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable (2010).
 Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
-For this tutorial, I have re-implemented it in the \texttt{mice} package as \texttt{mice::MCAR()},
+For this tutorial, I have re-implemented it in the \texttt{mice} package as \texttt{mice::mcar()},
 with a fast backend in C++ and new printing and plotting methods.
@@
 which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data; on par with multiple imputation (\textbf{leeComparisonFullInformation2021?}).
+FIML is a best-practice solution for handling missing data; on par with multiple imputation (Lee \& Shi, 2021).
 FIML estimation assumes that missingness is either MCAR or MAR.
@@
 To conclude; our recommendation is to inspect missingness (e.g., using \texttt{mice::MCAR()})
-and report the number of missings per variable (e.g., using \texttt{tidySEM::descriptives()}),
+and report the proportion of missingness per variable (e.g., using \texttt{tidySEM::descriptives()}),
 before proceeding with FIML.
@@
 
-Prior literature on latent class analysis has emphasized exploratory applications of the method (see \textbf{nylundDecidingNumberClasses2007?}).
+Prior literature on latent class analysis has emphasized exploratory applications of the method (see Nylund, Asparouhov, \& Muthén, 2007).
 In exploratory analyses, a large number of models are typically estimated in batch,
@@
 One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer the most parsimoneous model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (\textbf{akogulComparisonInformationCriteria2016?}).
+Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
+Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016).
 
@@
 
-Van de Schoot and colleagues (\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}) argue that there are many choices to be made in the specification of latent trajectory models,
+Van de Schoot and colleagues (2017) argue that there are many choices to be made in the specification of latent trajectory models,
 and that alternative
@@
 If the destination likelihood is \emph{lower}, the estimator still moves there occasionally, based on probability.
-This latter property allows it to escape local optima.
+This latter property allows it to escape local optima, and find the global optimum.
 
@@
 
-In the case of Latent Profile Analysis, it has been suggested that when tests of multivariate normality indicate a violation,
-maximum likelihood with robust standard errors should be used (Spurk et al., 2020).
-This statement is problematic for two reasons.
-Firstly, multivariate normality will always be violated in mixture models by definition.
-We assume that the population consists of several groups each characterized by its own normal distribution.
-When two or more normal distributions with different means and standard deviations comprise a mixed population distribution,
-this combined distribution is no longer normal.
-For example, you can imagine a bimodal distribution when dealing with two latent classes.
-If this were not the case and our mixed population distribution were normal,
-there would be no latent classes to extract as the population would consists of a single class only.
-For this reason, multivariate normality will always be violated when the population is comprised of several latent groups.
-Secondly, the claim that maximum likelihood with robust standard errors should be preferred
-implicitly implies that only software packages with this estimation technique can be used.
+One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed (Spurk et al., 2020).
+This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
+As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
 
-\hypertarget{starting-values}{%
-\subsubsection{Starting values}\label{starting-values}}
+--\textgreater{}
 
-The EM algorithm needs starting values for each parameter it wants to estimate.
-If it were provided a single starting value for each parameter,
-the EM optimizer would be at risk of finding a local optimum instead of the global optimum.
-In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates.
-To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter
-in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan \& Peel, 2004).
-Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values.
-Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).
-
-Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence
-or in case of having poor initial values the algorithm might diverge altogether (McLachlan \& Peel, 2004).
-The key is to use a large number of starting value sets.
-Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models,
-and a greater number of sets for more complex models.
-A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately.
-
-{[}(\textbf{nylund-gibsonTenFrequentlyAsked2018?}); {]}
-
-\begin{verbatim}
-In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
-Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
-[@biernackiChoosingStartingValues2003]
-\end{verbatim}
-
-\hypertarget{assessing-estimation-results}{%
-\subsubsection{Assessing Estimation Results}\label{assessing-estimation-results}}
+\hypertarget{class-enumeration}{%
+\subsubsection{Class enumeration}\label{class-enumeration}}
 
@@
 
-Absolute model fit can be assessed using the likelihood ratio (LR) \(\chi^2\) goodness-of-fit test.
-It can compare two nested models when they have the same number of classes (Lanza et al., 2003).
-Typically, we compare our final class solution (a constrained model)
-to an unconstrained one (a model characterized by the perfect fit).
-The former model is a special, less complex version of the latter.
-In this case, we compare our model-based response pattern frequencies to those in the dataset.
-Therefore, we make a statement about how well our model fits the data.
-The perfect fit would be indicated by the \(\chi^2\ =\ 0\) and \(p\ =\ 1\).
-Ideally, the LR \(\chi^2\) test will be non-significant as this implies
-that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model.
-The underlying assumption of this procedure is that the test statistic is asymptotically distributed
-as a \(\chi^2\) meaning that it can be approximated using this distribution.
-This is the case when the sample size is adequate (Lanza et al., 2003).
-The dependence of the LR \(\chi^2\) test statistic on sample size has its downsides (Masyn, 2013).
-Very large samples will often effortlessly reject the null even when the final class solution is a good fit,
-and small samples will often be underpowered resulting in non-significant \(\chi^2\) statistic.
-In fact, small samples might not be well approximated using the \(\chi^2\) distribution,
-which is a problem shared with large yet sparse datasets.
-In such cases, special caution is advised when interpreting the resulting p-values (Masyn, 2013).
-
 Relative model fit can be examined using the likelihood ratio test.
@@
 The likelihood ratio test can only compare two models at a time (Lanza et al., 2003).
+
+\hypertarget{model-fit-indices}{%
+\subsubsection{Model fit indices}\label{model-fit-indices}}
 
@@
 and these models need not be nested (Masyn, 2013).
-
-When estimators which require sets of starting values are used,
-each fitted model will have a corresponding log likelihood value.
-Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times
-with each log likelihood value generated by a different starting value set.
-Successful replications of the largest log likelihood value across different starting values
-instill confidence that the estimation was successful and we indeed found the global maximum (Masyn, 2013).
-The model with the highest replicated likelihood will be chosen as our best-fit.
-If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012).
-Such a model does not have a unique solution. For a discussion on model identification for LCA, see Masyn (2013).
-Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated
-in at least 3-10\% of the models initialized with different starting values in the first step of the optimization.
-This consideration does not apply in case of models which use simulated annealing since these do not have starting values.
 
@@
 
-Conditional item probabilities indicate the probability of a positive score on an item
-given that the person belongs to a particular latent class.
-However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1),
-we are observing boundary parameter estimates. This can be a sign of an invalid solution,
-warn us that too many classes were extracted, or indicate a local optimum.
-Boundary parameter estimates are more likely to happen when working with sparse data
-and should be interpreted with caution (Geiser, 2012).
+\hypertarget{interpreting-the-results}{%
+\subsubsection{Interpreting the results}\label{interpreting-the-results}}
 
-An important outcome of LCA are unconditional class probabilities.
-These communicate the probability that an individual belongs to each class.
-A limitation is that our final class solution need not reflect true class membership.
-Once we do accept a particular class solution, special care also needs to be taken when naming the classes.
 Class names should be chosen to accurately reflect group membership.
@@
 
-\hypertarget{starting-values-1}{%
-\subsubsection{Starting values}\label{starting-values-1}}
+\hypertarget{starting-values}{%
+\subsubsection{Starting values}\label{starting-values}}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-akogulComparisonInformationCriteria2016}{}}%
+Akogul, S., \& Erisoglu, M.
+(2016).
+A {Comparison} of {Information Criteria} in {Clustering Based} on {Mixture} of {Multivariate Normal Distributions}.
+\emph{Mathematical and Computational Applications}, \emph{21}(3), 34.
+\url{https://doi.org/10.3390/mca21030034}
+
+\leavevmode\vadjust pre{\hypertarget{ref-baughmanMixtureModelAnalysis2006}{}}%
+Baughman, A.
+L., Bisgard, K.
+M., Lynn, F., \& Meade, B.
+D.
+(2006).
+Mixture model analysis for establishing a diagnostic cut-off point for pertussis antibody levels.
+\emph{Statistics in Medicine}, \emph{25}(17), 2994--3010.
+\url{https://doi.org/10.1002/sim.2442}
+
+\leavevmode\vadjust pre{\hypertarget{ref-figueiredoUnsupervisedLearningFinite2002}{}}%
+Figueiredo, M.
+A.
+T., \& Jain, A.
+K.
+(2002).
+Unsupervised learning of finite mixture models.
+\emph{IEEE Transactions on Pattern Analysis and Machine Intelligence}, \emph{24}(3), 381--396.
+\url{https://doi.org/10.1109/34.990138}
+
+\leavevmode\vadjust pre{\hypertarget{ref-jamshidianTestsHomoscedasticityNormality2010}{}}%
+Jamshidian, M., \& Jalal, S.
+(2010).
+Tests of {Homoscedasticity}, {Normality}, and {Missing Completely} at {Random} for {Incomplete Multivariate Data}.
+\emph{Psychometrika}, \emph{75}(4), 649--674.
+\url{https://doi.org/10.1007/s11336-010-9175-3}
+
+\leavevmode\vadjust pre{\hypertarget{ref-leeComparisonFullInformation2021}{}}%
+Lee, T., \& Shi, D.
+(2021).
+A comparison of full information maximum likelihood and multiple imputation in structural equation modeling with missing data.
+\emph{Psychological Methods}, No Pagination Specified--No Pagination Specified.
+\url{https://doi.org/10.1037/met0000381}
+
+\leavevmode\vadjust pre{\hypertarget{ref-littleTestMissingCompletely1988}{}}%
+Little, R.
+J.
+A.
+(1988).
+A {Test} of {Missing Completely} at {Random} for {Multivariate Data} with {Missing Values}.
+\emph{Journal of the American Statistical Association}, \emph{83}(404), pp.
+1198--1202.
+\url{https://doi.org/10.2307/2290157}
+
+\leavevmode\vadjust pre{\hypertarget{ref-normanLikertScalesLevels2010}{}}%
+Norman, G.
+(2010).
+Likert scales, levels of measurement and the {``laws''} of statistics.
+\emph{Advances in Health Sciences Education}, \emph{15}(5), 625--632.
+\url{https://doi.org/10.1007/s10459-010-9222-y}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nylundDecidingNumberClasses2007}{}}%
+Nylund, K.
+L., Asparouhov, T., \& Muthén, B.
+O.
+(2007).
+Deciding on the {Number} of {Classes} in {Latent Class Analysis} and {Growth Mixture Modeling}: {A Monte Carlo Simulation Study}.
+\emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{14}(4), 535--569.
+\url{https://doi.org/10.1080/10705510701575396}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nylund-gibsonTenFrequentlyAsked2018}{}}%
+Nylund-Gibson, K., \& Choi, A.
+Y.
+(2018).
+Ten frequently asked questions about latent class analysis.
+\emph{Translational Issues in Psychological Science}, \emph{4}(4), 440--461.
+\url{https://doi.org/10.1037/tps0000176}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rubinInferenceMissingData1976}{}}%
+Rubin, D. B. (1976). Inference and {Missing Data}. \emph{Biometrika}, \emph{63}(3), 581--592. \url{https://doi.org/10.2307/2335739}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vandeschootGRoLTSChecklistGuidelinesReporting2017}{}}%
+Van De Schoot, R., Sijbrandij, M., Winter, S.
+D., Depaoli, S., \& Vermunt, J.
+K.
+(2017).
+The {GRoLTS-Checklist}: {Guidelines} for {Reporting} on {Latent Trajectory Studies}.
+\emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{24}(3), 451--467.
+\url{https://doi.org/10.1080/10705511.2016.1247646}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vanlissaWORCSWorkflowOpen2020}{}}%
+Van Lissa, C.
+J., Brandmaier, A.
+M., Brinkman, L., Lamprecht, A.-L., Peikert, A., Struiksma, M.
+E., \& Vreede, B.
+(2020).
+\emph{{WORCS}: {A Workflow} for {Open Reproducible Code} in {Science}}.
+\url{https://doi.org/10.17605/OSF.IO/ZCVBS}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vermuntKmeansMayPerform2011}{}}%
+Vermunt, J.
+K.
+(2011).
+K-means may perform as well as mixture model clustering but may also be much worse: {Comment} on {Steinley} and {Brusco} (2011).
+\emph{Psychological Methods}, \emph{16}(1), 82--88.
+\url{https://doi.org/10.1037/a0020144}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vermuntj.k.LatentClassAnalysis2004}{}}%
+Vermunt, J.K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.F., \& Department of Methodology and Statistics.
+(2004).
+Latent class analysis.
+In \emph{The {Sage} encyclopedia of social sciences research methods} (pp.
+549--553).
+{Sage}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wuAbuseDependencePrescription2011}{}}%
+Wu, L.-T., Woody, G.
+E., Yang, C., Pan, J.-J., \& Blazer, D.
+G.
+(2011).
+Abuse and dependence on prescription opioids in adults: A mixture categorical and dimensional approach to diagnostic classification.
+\emph{Psychological Medicine}, \emph{41}(3), 653--664.
+\url{https://doi.org/10.1017/S0033291710000954}
+
+\end{CSLReferences}
 

### 2022-08-01 manuscript.Rmd

@@
 wordcount         : "X"
-
+bibliography      : "mixture.bib"
 figsintext        : no
@@
 specific sub-types of latent class analyses,
-including latent class growth analysis [@schootGRoLTSChecklistGuidelinesReporting2017]
+including latent class growth analysis [@vandeschootGRoLTSChecklistGuidelinesReporting2017]
 and latent class analysis with ordinal indicators [e.g., @nylund-gibsonTenFrequentlyAsked2018].
@@
 
-One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed (Spurk et al., 2020). 
+One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurkLatentProfileAnalysis2020]. 
 This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.

### 2022-08-15 example_analysis/Daniel_LCA.Rmd
LCA Example Analysis

---
title: "LCA in R"
output: html_notebook
---

This is an example of a Latent Class Analysis (LCA).

In Part 1, I use the Lazarsfeld dataset and OpenMX.
In Part 2, I use the Drinking dataset and OpenMx.
In Part 3, I redo the analysis from Part 2 using tidySEM.

```{r}
library(tidySEM)
library(OpenMx)
```

Part 1: LCA using OpenMx and Lazarsfeld Dataset

Analysis based on: https://rdrr.io/cran/OpenMx/src/inst/models/passing/LCAlazarsfeld.R

```{r}
data("lazarsfeld", package ="OpenMx", verbose= TRUE)
data <- lazarsfeld

freq <- data[,5]
data[,1] <- as.ordered(data[,1])
data[,2] <- as.ordered(data[,2])
data[,3] <- as.ordered(data[,3])
data[,4] <- as.ordered(data[,4])
vars <- data[,1:4]

nclass <- 2
nvar <- 4
nthresh <- 1

nameList <- names(vars)

class1 <- mxModel("Class1", 
            mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
            mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
            mxMatrix("Full", name = "ThresholdsClass1", nrow = 1, ncol = nvar, 
            	dimnames = list("Threshold",nameList), free=TRUE),
		  mxExpectationNormal(covariance="R", means="M", dimnames=nameList, 
				      thresholds="ThresholdsClass1"),
		  mxFitFunctionML(vector=TRUE))

class2 <- mxModel("Class2", 
            mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
            mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
            mxMatrix("Full", name = "ThresholdsClass2", nrow = 1, ncol = nvar, 
            	dimnames = list("Threshold",nameList), free=TRUE),
		  mxExpectationNormal(covariance="R", means="M", dimnames=nameList,
				      thresholds="ThresholdsClass2"),
		  mxFitFunctionML(vector=TRUE))

# Define the model

lcamodel <- mxModel("lcamodel", class1, class2, mxData(vars, type="raw"),

# Create class membership probabilities, constrain them to be in the range 0-1, and make their sum equal 1.0
            mxMatrix("Full", name = "ClassMembershipProbabilities", nrow = nclass, ncol = 1, free=TRUE, 
			        labels = c(paste("pclass", 1:nclass, sep=""))),
            mxBounds(c(paste("pclass", 1:nclass, sep="")),0,1),
		    mxMatrix("Iden", nrow = 1, name = "constraintLHS"),
		    mxAlgebra(sum(ClassMembershipProbabilities), name = "constraintRHS"),
            mxConstraint(constraintLHS == constraintRHS),

# Define the objective function
            mxAlgebra(-2*sum(freq * log(pclass1%x%Class1.objective + pclass2%x%Class2.objective)), 
            	name="lca"),
            mxFitFunctionAlgebra("lca")
)

# Run the job
model <- mxRun(lcamodel, suppressWarnings=TRUE)
summary(model)
model$matrices

# Check results against those hard-coded from old Mx:
negativeThresholds <- as.vector(mxEval(Class1.ThresholdsClass1, model))
positiveThresholds <- as.vector(mxEval(Class2.ThresholdsClass2, model))
if (negativeThresholds[[1]] > 0) {
	temp <- negativeThresholds
	negativeThresholds <- positiveThresholds
	positiveThresholds <- temp
}

omxCheckCloseEnough(negativeThresholds, c(-1.3247, -0.2909, -0.1466, -0.0044),.01)
omxCheckCloseEnough(positiveThresholds, c(0.1805, 0.9071, 1.3169, 1.5869),.01)
omxCheckCloseEnough(sort(as.vector(mxEval(ClassMembershipProbabilities, model))), c(0.4440, 0.5560),.01)
omxCheckCloseEnough(model$output$Minus2LogLikelihood, 4696.444, 0.01)
```

Part 2: LCA Using OpenMx and Drinking Dataset,

Dataset from: https://stats.oarc.ucla.edu/sas/dae/latent-class-analysis/

```{r}
data <- read.csv('C:/Users/danie/OneDrive/Desktop/LCA/drinking_data.csv', header = T, sep = ',')

#check if two rows have the same response pattern
library(DAKS)
pattern(data)

dat1 <- c(1,0,0,0,0,0,0,0,0, 126)
dat2 <- c(0,0,0,0,0,0,0,0,0, 102)
dat3 <- c(1,0,0,0,0,0,0,0,1, 47)
dat4 <- c(1,1,0,0,0,0,0,0,0, 47)
dat5 <- c(1,0,0,0,0,1,0,0,0, 43)
dat6 <- c(126, 102, 47, 47, 43)

dat <- data.frame(dat1,dat2,dat3,dat4,dat5)
dat <- t(dat)
dat <- as.data.frame(dat)
names(dat) <- c("Q1", "Q2", "Q3", "Q4", "Q5", "Q6", "Q7", "Q8", "Q9", "freq")
rm("data", "dat1", "dat2", "dat3", "dat4", "dat5", "dat6")

dat[,1] <- as.ordered(dat[,1])
dat[,2] <- as.ordered(dat[,2])
dat[,3] <- as.ordered(dat[,3])
dat[,4] <- as.ordered(dat[,4])
dat[,5] <- as.ordered(dat[,5])
dat[,6] <- as.ordered(dat[,6])
dat[,7] <- as.ordered(dat[,7])
dat[,8] <- as.ordered(dat[,8])
dat[,9] <- as.ordered(dat[,9])

freq <- dat[,10]
vars <- dat[,1:9]

nclass <- 2
nvar <- 9
nthresh <- 1
nameList <- names(vars)

class1 <- mxModel("Class1", 
                  mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
                  mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
                  mxMatrix("Full", name = "ThresholdsClass1", nrow = 1, ncol = nvar, 
                           dimnames = list("Threshold",nameList), free=TRUE),
                  mxExpectationNormal(covariance="R", means="M", dimnames=nameList, 
                                      thresholds="ThresholdsClass1"),
                  mxFitFunctionML(vector=TRUE))

class2 <- mxModel("Class2", 
                  mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
                  mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
                  mxMatrix("Full", name = "ThresholdsClass2", nrow = 1, ncol = nvar, 
                           dimnames = list("Threshold",nameList), free=TRUE),
                  mxExpectationNormal(covariance="R", means="M", dimnames=nameList,
                                      thresholds="ThresholdsClass2"),
                  mxFitFunctionML(vector=TRUE))

# Define the model

lcamodel <- mxModel(model = "lcamodel", class1, class2, mxData(vars, type="raw"))

# Create class membership probabilities, constrain them to be in the range 0-1, and make their sum equal 1.0
mxMatrix("Full", name = "ClassMembershipProbabilities", nrow = nclass, ncol = 1, free=TRUE, labels = c(paste("pclass", 1:nclass, sep=""))) 
mxBounds(c(paste("pclass", 1:nclass, sep="")),0,1) 
mxMatrix("Iden", nrow = 1, name = "constraintLHS") 
mxAlgebra(sum(ClassMembershipProbabilities), name = "constraintRHS") 
mxConstraint(constraintLHS == constraintRHS)

# Define the objective function
mxAlgebra(-2*sum(freq * log(pclass1%x%Class1.objective + pclass2%x%Class2.objective)), name="lca")

mxFitFunctionAlgebra("lca")

# Run the job
model <- mxRun(lcamodel, suppressWarnings=TRUE)
summary(model)
model$matrices

# Check results against those hard-coded from old Mx:
negativeThresholds <- as.vector(mxEval(Class1.ThresholdsClass1, model))
positiveThresholds <- as.vector(mxEval(Class2.ThresholdsClass2, model))
if (negativeThresholds[[1]] > 0) {
  temp <- negativeThresholds
  negativeThresholds <- positiveThresholds
  positiveThresholds <- temp
}

```

Part 3: LCA using tidySEM

# Part 1: Using mclust [incomplete]

```{r}
library(mclust)
```

```{r}
# Fit 1-9 class model
model <- Mclust(data)
```

### 2022-08-17 manuscript.Rmd
LCA Example Analysis
Use Cases Draft
Created paragraphs based on Caspar's comment about use cases.
Needs examples.

@@
 
-Latent class analysis is an umbrella term that refers to a number of techniques for estimating unobserved group membership
+Latent class analysis (LCA) is an umbrella term that refers to a number of techniques for estimating unobserved group membership
 based on a parametric model of one or more observed indicators of group membership.
@@
 Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, this paper exclusively relies on free, open source software for latent class analysis in R.
-Our goal is to make best-practices latent class analysis widely accessable.
+Our goal is to make best-practices latent class analysis widely accessible.
 
@@
 the mean and the variance.
-Commonly, the same model is estimated accross all classes, but with different parameters for each class (i.e., class-specific means and variances).
+Commonly, the same model is estimated across all classes, but with different parameters for each class (i.e., class-specific means and variances).
 Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.
@@
 When estimating a two-class mixture model, the detective observes that the model estimates
-the mean shoesize of the two groups are equal to `r report(estparam[1], equals = FALSE)` and  `r report(estparam[2], equals = FALSE)`,
+the mean shoe size of the two groups are equal to `r report(estparam[1], equals = FALSE)` and  `r report(estparam[2], equals = FALSE)`,
 which is close to the true means of the two groups, namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
@@
 
+LCA can be used as a descriptive analysis where a researcher wishes to describe their sample 
+and identify a few prototypes based on many variables.
+<!-- add an example and elaborate -->
+
+With LCA, our goal could be to inductively identify the number of classes. 
+For instance, if we believe that a variable represents a group, 
+but don't know how many groups there are, LCA may be an appropriate technique to answer this question.
+<!-- add an example -->
+
+Another application of LCA is to classify individuals.
+<!-- add an example and elaborate -->
+
+LCA is also appropriate when we wish to identify indicators that capture classes well.
+<!-- add an example and elaborate -->
+
+An extension of LCA is that containing covariates which can be used to predict class membership.
+<!-- add an example and elaborate -->
+
+When our interest is the prediction of one or more outcomes,
+LCA can be used to construct classes as categorical predictors.
+<!-- add an example and elaborate -->
+
+In addition to these applications, LCA can be used for dimensionality reduction
+as the resulting groups summarize response patterns on a large number of indicators.
+<!-- add an example -->
+
+Finally, LCA can be used to deal with data which violate the normality assumption.
+<!-- add an example and elaborate, check with Caspar if there are other assumptions -->
+
 <!-- CJ: Include these use cases too: -->
@@
 
-In LCA, a sequence of models is fitted to the data 
+LCA can be done in an exploratory or in a confirmatory fashion.
+In exploratory LCA, a sequence of models is fitted to the data 
 with each additional model estimating one more class than the previous model.
-<!-- CJ: Not always; you may have a theory about a fixed number. -->
-<!-- CJ: Argue that, even in this case, it is useful to know if the theoretical model is markedly better than other numbers of classes -->
-The final model called the final class solution is chosen based on both theoretical and statistical criteria.
+These models are then compared and the best solution is selected as the final class solution.
+In some cases, prior theory can inform the researcher about the number of classes to expect.
+Even in such confirmatory LCA cases, it is nonetheless useful to know 
+if the theoretical model is markedly better than those with differing numbers of classes.
+
+From a sequence of models, the final class solution is chosen based on both theoretical and statistical criteria.
 Theory should drive the selection of indicator variables, inform the expectations and reflect on the findings.

### 2022-08-31 example_analysis/Daniel_LCA.Rmd
Updates 23 August
Updated the manuscript.

All the references are now in the [@key] format.  New .bib file  with references has been made.

Reworked the model fit section to reflect statistics tidySEM reports.
Stated that LRT cannot be used.
Referred to the bootstrapped LRT being available albeit extremely slow; we are currently developing a faster version.

Adjusted classification diagnostics to reflect what tidySEM provides. Entropy is finished. Needs more information on the range of the posterior classification probabilities by most likely class membership and the range of the sample size.

@@
 ---
+
+USING TIDYSEM:
+
+see functions starting in mx
+
+```{r}
+library(tidySEM)
+df <- data_mix_ordinal
+df[1:4] <- lapply(df, ordered)
+mx_lca(data = df,
+       classes = 2) -> res
+```
+
+```{r}
+res$class1
+res$latentVars
+```
+
+USING DIFFERENT PACKAGES
 
@@
 
-# Part 1: Using mclust [incomplete]
-
 ```{r}
-library(mclust)
+#adjust this code
+# The argument `classes` indicates which class solutions should be estimated (e.g., 1 through 6).
+# The argument `variances` specifies whether variances should be `"equal"` or `"varying"` across classes.
+# The argument `covariances` specifies whether covariances should be constrained to `"zero"`, `"equal"` or `"varying"` across classes.
+tidySEM::mx_profiles(classes, variances, covariances)
 ```
 
-```{r}
-# Fit 1-9 class model
-model <- Mclust(data)
-```
+After all models have been estimated, the function `tidySEM::table_fit()` can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
+Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
+because this test is very computationally expensive.
+It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.

### 2022-08-31 manuscript.Rmd
Updates 23 August
Updated the manuscript.

All the references are now in the [@key] format.  New .bib file  with references has been made.

Reworked the model fit section to reflect statistics tidySEM reports.
Stated that LRT cannot be used.
Referred to the bootstrapped LRT being available albeit extremely slow; we are currently developing a faster version.

Adjusted classification diagnostics to reflect what tidySEM provides. Entropy is finished. Needs more information on the range of the posterior classification probabilities by most likely class membership and the range of the sample size.

@@
   using free and open source software in R. To this end, this tutorial
-  introduces new functionality for estimating and reporting mixture mixture
+  introduces new functionality for estimating and reporting mixture
   models in the `tidySEM` R-package. These functions rely on estimation using
@@
 One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
-By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibsonTenFrequentlyAsked2018].
+By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibson_ten_2018].
 In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.
@@
 In line with this perspective, mixture modeling is sometimes described as "model-based
-clustering" (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
-Raftery, 2017).
+clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
 Many clustering algorithms apply some recursive splitting algorithm to the data.
@@
 Finally, in the context of machine learning, latent class analysis can be considered as an
-*unsupervised classification* problem [@figueiredoUnsupervisedLearningFinite2002].
+*unsupervised classification* problem [@figueiredo_unsupervised_2002].
 The term *unsupervised* refers to the fact that the outcome variable, true class membership,
@@
 In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
-From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermuntj.k.LatentClassAnalysis2004].
+From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermunt_jk_latent_2004].
 Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
@@
 
-<!-- f a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis.
+<!-- If a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis.
 Modified to handle discrete data, this constrained analysis is known as LCA.
@@
 
-<!-- ecause LPA is model-based, a number of different model parameterizations can be -->
+<!-- Because LPA is model-based, a number of different model parameterizations can be -->
 <!-- estimated. These models differ in terms of whether--and how--parameters are -->
@@
 or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughmanMixtureModelAnalysis2006] or mental [@wuAbuseDependencePrescription2011] health problems.
+For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
 The example of shoe size is a rudimentary illustration of this type of application.
@@
 <!--                                                1 -->
-<!-- A -->
-<!-- downside is that these models are more complex (much more parameters to be estimated). -->
+<!-- A downside is that these models are more complex (much more parameters to be estimated). -->
 <!-- Using criteria, such as the Bayesian Information criterion (BIC) helps to find a model that -->
@@
 specific sub-types of latent class analyses,
-including latent class growth analysis [@vandeschootGRoLTSChecklistGuidelinesReporting2017]
-and latent class analysis with ordinal indicators [e.g., @nylund-gibsonTenFrequentlyAsked2018].
+including latent class growth analysis [@van_de_schoot_grolts-checklist_2017]
+and latent class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
 These were generalized to be more relevant to all types of latent class analyses,
@@
 Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
-Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermuntKmeansMayPerform2011].
+Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
 Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
-suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
-Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @normanLikertScalesLevels2010],
+suggest that the misapplication of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
+Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @norman_likert_2010],
 this certainly does not imply that all methods are.
@@
 
-Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [REF SPURK].
+Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [@spurk_latent_2020].
 This is incorrect.
@@
 Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @vanlissaWORCSWorkflowOpen2020].
+Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @van_lissa_worcs_2021.
+
+<!--  Adjust the following part, it belongs here:  -->
+
+<!-- CJ: This is about model interpretation; conditional item probabilities are the analogue of mean and standard deviation, but when the observed data are binary/ordinal  -->
+<!-- Conditional item probabilities indicate the probability of a positive score on an item  -->
+<!-- given that the person belongs to a particular latent class.  -->
+
+<!-- If a particular item is endorsed by two or more classes at markedly different rates, -->
+<!-- it is said to discriminate well between classes and is consequently considered a good indicator. -->
+<!-- The classes are considered highly homogeneous with respect to an item when for a particular item -->
+<!-- there are a distinct difference in conditional item probabilities of two or more classes. -->
+<!-- If an item is endorsed below 30% for one class and above 70% for another class, 
+the classes have high homogeneity with respect to this item [@masyn_latent_2013]. -->
+<!-- However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1),  -->
+<!-- we are observing boundary parameter estimates. This can be a sign of an invalid solution,  -->
+<!-- warn us that too many classes were extracted, or indicate a local optimum.  -->
+<!-- Boundary parameter estimates are more likely to happen when working with sparse data  -->
+<!-- and should be interpreted with caution (Geiser, 2012).  -->
+
+<!-- CJ: This talks about the problem of inadmissible solutions.
+With ordinal items, that can happen because of sparse cross-tables (e.g., some combinations of responses never occur).
+WIth continuous items, that could happen e.g.
+because you have a zero-inflated, otherwise normally distributed variable; one class would have a mean at zero and its SD would be hard to determine.
+Or it could happen if you model a five-category Likert scale with a five-component latent profile analysis; each response category will have its own mean, but there's no variance around that (everybody in class 3 has exactly the response value 3).
+These are all the same problems.
+-->
+
+<!-- An important outcome of LCA are unconditional class probabilities.  -->
+<!-- These communicate the probability that an individual belongs to each class.  -->
+<!-- A limitation is that our final class solution need not reflect true class membership.  -->
+<!-- Once we do accept a particular class solution, special care also needs to be taken when naming the classes.  -->
+
+A problem which can occur is that of inadmissible solutions. 
+With binary indicators, LCA is modelling a cross-table with all the predictors.
+The problem with such cross-tables is that they will often contain empty cells,
+i.e. combinations of responses that never occur together. 
+This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
+Such boundary parameter estimates could indicate that the solution is invalid.
+Boundary parameter estimates can also happen with continuous indicators.
+For instance, if we have a zero-inflated normal distribution and a two class solution,
+one class might have the mean of zero and its standard deviation cannot be determined 
+since there is little variance. 
+This too could be a sign of an invalid solution,
+warn us that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].
 
@@
 
-Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled [@schootGRoLTSChecklistGuidelinesReporting2017].
-Three types of missingness have been distinguished in the literature [@rubinInferenceMissingData1976]:
+Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled [@van_de_schoot_grolts-checklist_2017].
+Three types of missingness have been distinguished in the literature [@rubin_inference_1976]:
 Missing completely at random (MCAR), which means that missingness is random;
@@
 It is possible to conduct a so-called "MCAR" test, 
-for example the non-parametric MCAR test [@jamshidianTestsHomoscedasticityNormality2010].
+for example the non-parametric MCAR test [@jamshidian_tests_2010].
 But note that the name "MCAR test" is somewhat misleading,
@@
 A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@littleTestMissingCompletely1988].
+As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@little_test_1988].
 This assumption is tenuous in the context of latent class analysis.
-A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable [-@jamshidianTestsHomoscedasticityNormality2010].
+A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable [@jamshidian_tests_2010].
 Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
@@
 which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data; on par with multiple imputation [@leeComparisonFullInformation2021].
+FIML is a best-practice solution for handling missing data; on par with multiple imputation [@lee_comparison_2021].
 FIML estimation assumes that missingness is either MCAR or MAR.
@@
 Latent class analyses with ordinal indicators have thresholds.
-All of these parameters can be freely estimated across classes, or constrained,
-or fixed (e.g., to zero).
+All of these parameters can be freely estimated across classes, constrained to be equal across classes,
+or fixed to a certain value (e.g., to zero).
 The total number of parameters thus scales with the number of estimated classes.
 Consequently, latent class analyses have a potentially very high number of parameters.
-As any of these parameters could be mis-specified,
+As any of these parameters could be misspecified,
 it is important to consider alternative model specifications.
@@
 
-Prior literature on latent class analysis has emphasized exploratory applications of the method [see @nylundDecidingNumberClasses2007].
+Prior literature on latent class analysis has emphasized exploratory applications of the method [see @nylund_deciding_2007].
 In exploratory analyses, a large number of models are typically estimated in batch,
@@
 It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2 log likelihood (which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance fit and model complexity.
-The BIC usually applies a stronger penalty for complexity that scales logarithmically with the sample size.
-
-Fit indices may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogulComparisonInformationCriteria2016].
+More on fit indices can be found in the Model fit indices subsection of this paper.
 
@@
 For example, in the context of preregistered analyses, the main models of interest may have been specified a priori.
-
-Van de Schoot and colleagues [-@vandeschootGRoLTSChecklistGuidelinesReporting2017] argue that there are many choices to be made in the specification of latent trajectory models,
+Therefore, it may not be necessary to test different model specifications when using the confirmatory approach.
+
+<!-- see the following article and expand on the paragraph -->
+@van_de_schoot_grolts-checklist_2017 argue that there are many choices to be made in the specification of latent trajectory models,
 and that alternative
@@
 Conversely, the free open source program OpenMx is very flexible,
-but not very user-friendly.
-
-New functionality in the R-package `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
+but not very user-friendly. We directly address this limitation using the `tidySEM` R-package.
+New functionality in `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
 It adheres to best practices in estimation and reporting, as described in this paper.
@@
 
-Because of the limitations in ex tools, we set out to develop a tool that a)
-provided sensible defaults and were easy to use, but provided the option to
-access and modify all of the inputs to the model (i.e., low barrier, high
-ceiling), b) interfaced to existing tools, and are able to translate between
+Because of the limitations of the aforementioned tools, we set out to develop a free tool that
+provides sensible defaults and is easy to use, but provides the option to
+access and modify all of the model inputs (i.e., low barrier, high
+ceiling). `tidySEM` interfaces with existing tools, and is able to translate between
 what existing tools are capable of and what researchers and analysts
-carrying-out person-oriented analyses would like to specify,
-
-fully-reproducible analyses and
+carrying-out person-oriented analyses would like to specify. 
+Furthermore,`tidySEM` facilitates fully-reproducible analyses and contributes to open science.
 
@@
 <!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
-<!-- analytic functionality. However, the majority of packages is focused on analyses with discrete -->
+<!-- analytic functionality. However, the majority of packages are focused on analyses with discrete -->
 <!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
@@
 The choice of the estimator depends on the presence of missing values,
-sample size, number of indicators, and available computational resources (Weller, Bowen & Faubert, 2020).
+sample size, number of indicators, and available computational resources [@weller_latent_2020].
 A commonly used technique is maximum likelihood (ML) estimation
@@
 
-One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurkLatentProfileAnalysis2020]. 
+One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
 This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
 As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
-<!-- CJ: Is there even any literature that says robust standard erorrs are better with latent class analysis? -->
+<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->
 
@@
 if the theoretical model is markedly better than those with differing numbers of classes.
+Therefore, it may always be useful to compare different class solutions.
 
@@
 These include but are not limited to likelihood ratio tests, information criteria, 
-and the Bayes factor (Weller, Bowen & Faubert, 2020).
+and the Bayes factor [@weller_latent_2020].
 
@@
 in order to give support to the simpler model. 
-The likelihood ratio test can only compare two models at a time (Lanza et al., 2003).
+The likelihood ratio test can only compare two nested models at a time [@lanza_introduction_2003].
+
+<!-- DA: This is just a goodness-of-fit test, and if it is significant the model fits the data poorly.  -->
+<!-- We don't get this test with `tidySEM`, so do not focus on this too much. We only provide information criteria-->
+
+<!-- DA: the bootstrap likelihood ratio test is the only one that is valid one according to Caspar, but it is not easily implemented in `tidySEM` (too computationally intensive), which is a problem, takes a lot of time to compute, maybe write about this,
+see @masyn_latent_2013 'parametric bootstrapping' it is not widely implemented. 
+Do not write about bootstrapping, too computationally intensive, we do not have a good solution for model fit tests, downplay that, MPLUS has MLR adjusted maximum likelihood test but unsure if to compare two different class solutions or just misfit, however not something we have, it is a specialized solution available only in MPLUS-->
 
 ### Model fit indices
+
+Fit indices typically used for determining the optimal number of classes include
+the Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC).
+Both information criteria are based on the -2*log likelihood (which is lower for better fitting models),
+and add a penalty for the number of parameters (thus incentivizing simpler models).
+This helps balance fit and model complexity.
+The lower the value of an information criterion, the better the overall fit of the model.
+The BIC applies a stronger penalty for model complexity 
+that scales logarithmically with the sample size.
+The literature suggests the BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
+Both the AIC and the BIC are available in the `tidySEM` output.
+
+Information criteria may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
+One option is to select a specific fit index before analyzing the data.
+Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
+Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016].
+Finally, one might make an elbow plot and compare multiple information criteria
+[for an example see @nylund-gibson_ten_2018].
+
+Another common test of model fit is the likelihood ratio $\chi^2$ goodness-of-fit test.
+However, this test is not implemented in `tidySEM`.
+
+An alternative is using the bootstrapped likelihood ratio test
+which can be run using `tidySEM::BLRT()`.
+Currently, this test is computationally expensive and can be slow on most computers.
+The development of a faster version of this test is well under way.
+
+<!-- DA: NOTES ABOUT LRT test: It can compare two nested models when they have the same number of classes
+<!-- If we wish to simultaneously compare multiple models based on their relative fit, -->
+<!-- this can be done through a comparison of multiple information criteria. -->
+<!-- Examples include the Akaike information criterion (AIC), -->
+<!-- and the Bayesian information criterion (BIC). -->
+<!-- Information criteria are a sum of a measure of fit (usually a form of the converged maximum log likelihood value)  -->
+<!-- and a penalty for model complexity (a combination of sample size and the number of modeled parameters). -->
+<!-- As a general rule, the lower the value of an information criterion, the better the model fits the data. -->
+<!-- When examining several competing models, each with differing number of classes, -->
+<!-- we expect the information criteria values to drop with each successive model until the final class solution is reached. -->
+<!-- Further models with additional classes should show worse fit as information criteria reach an optimum  -->
+<!-- before their values rise again with an increasing number of classes. -->
+<!-- Multiple information criteria should be compared when choosing the final class solution.  -->
+<!-- This can be done by means of an elbow plot [for an example see @nylund-gibson_ten_2018]. -->
+<!-- Advantages of using information criteria are that we can compare multiple models at a time, -->
+<!-- and these models need not be nested @masyn_latent_2013. -->
+
+<!-- Add a paragraph about using the Bayes factor see @weller_latent_2020-->
+
+<!-- decide whether to keep the following paragraph, if so, where to place it -->
+<!-- When estimators which require sets of starting values are used, -->
+<!-- each fitted model will have a corresponding log likelihood value.  -->
+<!-- Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times  -->
+<!-- with each log likelihood value generated by a different starting value set.  -->
+<!-- Successful replications of the largest log likelihood value across different starting values -->
+<!-- instill confidence that the estimation was successful and we indeed found the global maximum [@masyn_latent_2013].  -->
+<!-- The model with the highest replicated likelihood will be chosen as our best-fit.  -->
+<!-- If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012). -->
+<!-- Such a model does not have a unique solution. For a discussion on model identification for LCA, see @masyn_latent_2013. -->
+<!-- Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated  -->
+<!-- in at least 3-10% of the models initialized with different starting values in the first step of the optimization.  -->
+<!-- This consideration does not apply in case of models which use simulated annealing since these do not have starting values. -->
+<!-- @masyn_latent_2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
+<!-- to ensure that a global (rather than a local) maximum has been reached.". For discussion see Hipp & Bauer, 2006 -->
 
@@
 <!-- This is the case when the sample size is adequate (Lanza et al., 2003).  -->
-<!-- The dependence of the LR $\chi^2$ test statistic on sample size has its downsides (Masyn, 2013).  -->
+<!-- The dependence of the LR $\chi^2$ test statistic on sample size has its downsides [@masyn_latent_2013].  -->
 <!-- Very large samples will often effortlessly reject the null even when the final class solution is a good fit, -->
@@
 <!-- which is a problem shared with large yet sparse datasets. -->
-<!-- In such cases, special caution is advised when interpreting the resulting p-values (Masyn, 2013). -->
-
-If we wish to simultaneously compare multiple models based on their relative fit,
-this can be done through a comparison of multiple information criteria.
-Examples include the Akaike information criterion (AIC), the Consistent Akaike Information Criterion (CAIC),
-the Bayesian information criterion (BIC), and the Sample-size Adjusted Bayesian Information Criterion (SABIC).
-Information criteria are a sum of a measure of fit (usually a form of the converged maximum log likelihood value) 
-and a penalty for model complexity (combination of sample size and number of modeled parameters).
-As a general rule, the lower the value of an information criterion, the better the model fits the data.
-When examining several competing models, each with differing number of classes,
-we expect the information criteria values to drop with each successive model until the final class solution is reached.
-Further models with additional classes should show worse fit as information criteria reach an optimum 
-before their values rise again with an increasing number of classes.
-Multiple information criteria should be compared when choosing the final class solution. 
-This can be done by means of an elbow plot (for an example see Nylund-Gibson and Choi, 2018).
-Advantages of using information criteria are that we can compare multiple models at a time,
-and these models need not be nested (Masyn, 2013).
-
-<!-- Add a paragraph about using the Bayes factor see Weller, Bowen & Faubert, 2020-->
-
-<!-- decide whether to keep the following paragraph, if so, where to place it -->
-<!-- When estimators which require sets of starting values are used, -->
-<!-- each fitted model will have a corresponding log likelihood value.  -->
-<!-- Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times  -->
-<!-- with each log likelihood value generated by a different starting value set.  -->
-<!-- Successful replications of the largest log likelihood value across different starting values -->
-<!-- instill confidence that the estimation was successful and we indeed found the global maximum (Masyn, 2013).  -->
-<!-- The model with the highest replicated likelihood will be chosen as our best-fit.  -->
-<!-- If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012). -->
-<!-- Such a model does not have a unique solution. For a discussion on model identification for LCA, see Masyn (2013). -->
-<!-- Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated  -->
-<!-- in at least 3-10% of the models initialized with different starting values in the first step of the optimization.  -->
-<!-- This consideration does not apply in case of models which use simulated annealing since these do not have starting values. -->
-<!-- Masyn 2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
-<!-- to ensure that a global (rather than a local) maximum has been reached.". For discussion see Hipp & Bauer, 2006 -->
+<!-- In such cases, special caution is advised when interpreting the resulting p-values [@masyn_latent_2013]. -->
 
@@
 Classification diagnostics give us a way to assess the degree to which this is the case.
-They are separate from the absolute and relative goodness-of-fit 
-as a model can fit the data well but show poor latent class separation (Masyn, 2013).
-A fundamental concept when examining classification precision and accuracy are the posterior class probabilities.
-A probability of belonging to each latent class is computed for each individual.
-The highest posterior class probability is then determined and the individual is assigned to the corresponding class.
-We want each individual's posterior class probabilities to be high for one and low for the remaining latent classes.
-This is considered a high classification accuracy and means that the classes are distinct.
-<!-- see if you find content on homogeneity of classes -->
-Three important classification diagnostic measures are entropy, the average posterior probability and the modal class assignment proportion.
+They are separate from model fit indices as a model can fit the data well 
+but show poor latent class separation [@masyn_latent_2013].
+Classification diagnostics should not be used for model selection,
+but they can be used to disqualify certain solutions because they are uninterpretable.
+Interpretability should always be a consideration when considering different class solutions [@nylund-gibson_ten_2018].
+
+Three important classification diagnostics provided by `tidySEM` are are entropy, 
+the range of the posterior classification probabilities by most likely class membership,
+and the range of the sample size.
 
@@
 
-<!-- Masyn 2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
+The range of the posterior classification probabilities by most likely class membership are
+
+Another classification diagnostic measure is the range of the sample size.
+
+<!-- A fundamental concept when examining classification accuracy is the posterior class probability. -->
+<!-- This concept represents the probability of belonging to each latent class computed for each individual. -->
+<!-- The highest posterior class probability is then determined and the individual is assigned to the corresponding class. -->
+<!-- We want each individual's posterior class probabilities to be high for one and low for the remaining latent classes. -->
+<!-- This is considered a high classification accuracy and means that the classes are distinct. -->
+
+<!-- @masyn_latent_2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
 <!-- add content about entropy -->
 
-The average posterior probability is a measure of classification uncertainty for each latent class.
-
-The modal class assignment proportion
-
-<!-- CJ: This is about model interpretation; conditional item probabilities are the analogue of mean and standard deviation, but when the observed data are biary/ordinal  -->
-<!-- Conditional item probabilities indicate the probability of a positive score on an item  -->
-<!-- given that the person belongs to a particular latent class.  -->
-
-<!-- If a particular item is endorsed by two or more classes at markedly different rates, -->
-<!-- it is said to discriminate well between classes and is consequently considered a good indicator. -->
-<!-- The classes are considered highly homogeneous with respect to an item when for a particular item -->
-<!-- there are a distinct difference in conditional item probabilities of two or more classes. -->
-<!-- If an item is endorsed below 30% for one class and above 70% for another class, 
-the classes have high homogeneity with respect to this item (Masyn, 2013). MY OWN WRITING BUT NOT FLOWY ENOUGH-->
-<!-- However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1),  -->
-<!-- we are observing boundary parameter estimates. This can be a sign of an invalid solution,  -->
-<!-- warn us that too many classes were extracted, or indicate a local optimum.  -->
-<!-- Boundary parameter estimates are more likely to happen when working with sparse data  -->
-<!-- and should be interpreted with caution (Geiser, 2012).  -->
-
-<!-- CJ: THis talks about the problem of inadmissible solultions.
-With ordinal items, that can happen because of sparse cross-tables (e.g., some combinations of responses never occur).
-WIth continuous items, that could happen e.g.
-because you have a zero-inflated, otherwise normally distributed variable; one class would have a mean at zero and its SD would be hard to determine.
-Or it could happen if you model a five-category Likert scale with a five-component latent profile analysis; each response category will have its own mean, but there's no variance around that (everybody in class 3 has exactly the response value 3).
-These are all the same problems.
--->
-
-<!-- CJ: Merge with beginning of the paper, where we explain that LCA simultaneously estimates class membership and the descriptives of those classes. -->
-<!-- An important outcome of LCA are unconditional class probabilities.  -->
-<!-- These communicate the probability that an individual belongs to each class.  -->
-<!-- A limitation is that our final class solution need not reflect true class membership.  -->
-<!-- Once we do accept a particular class solution, special care also needs to be taken when naming the classes.  -->
-
-### Interpreting the results
-
-Class names should be chosen to accurately reflect group membership. 
-Overly simplified and generalized class names may prove misleading to both audiences and researches alike
-leading to what is known as a naming fallacy (Weller, Bowen & Faubert, 2020).
+<!-- DA: Write about prob_min and prob_max (see name formatting), -->
+<!-- the posterior probability from the model, -->
+<!-- it calculates the entropy and the range of the most likely classification probability, -->
+<!-- the diagonal, lets say you have 3 classes, there is the class the person was assigned to, -->
+<!-- and the probability the person should belong to that class, -->
+<!-- what this is telling us is the range of the average classification probabilities, -->
+<!-- what you want to see is that people have a high probability of being in the class they were assigned to, -->
+<!-- and you want that to be high for all the classes,  -->
+<!-- for each class we want a high number -->
+
+<!-- n_min and n_max are the range of the proprotion of people assigned to those classes. -->
+<!-- The smallest class has XX percent of people, and the largest class has XXX percent of people. -->
+<!-- Sometimes the class is too small to be interpreted, e.g. 5% or less of people, that would be too few to estimate parameters.  -->
+
+<!-- The average posterior probability is a measure of classification uncertainty for each latent class. -->
+
+<!-- The modal class assignment proportion -->
+
+`tidySEM::table_prob()` returns thresholds for ordinal dependent variables in probability scale.
+These are conditional item probabilities
+which communicate item endorsement given a class
+for models with binary and ordinal variables.
 
 ### Label switching
+<!-- optional subsection: interpreting class solutions (with class naming) -->
 
@@
 even when the only difference between the models is in their starting values. 
-Label switching is something to be mindful of when comparing different LCA models (Masyn, 2013).
+Label switching is something to be mindful of when comparing different LCA models [@masyn_latent_2013].
+
+The order of clusters is nondeterministic when using K-means in `tidySEM`. 
+Therefore label switching is still a consideration.
+A simple solution to this is setting a random seed number one line prior to fitting the model.
+We advise `tidySEM` users to always do so in order to circumvent label switching.
 
@@
 
+Class names should be chosen to accurately reflect group membership. 
+Overly simplified and generalized class names may prove misleading to both audiences and researches alike
+leading to what is known as a naming fallacy [@weller_latent_2020].
+
 ## Best practices in reporting
 
-Among studies using LCA, reporting practices vary significantly (Weller, Bowen & Faubert, 2020).
-Various authors have tried to better and standardize ways of reporting LCA (e.g. Masyn, 2013; Weller, Bowen & Faubert, 2020). 
+Among studies using LCA, reporting practices vary significantly [@weller_latent_2020].
+Various authors have tried to better and standardize ways of reporting LCA [e.g. @masyn_latent_2013; @weller_latent_2020]. 
 Building on their work, we provide guidelines with a strong emphasis on scientific reproducibility and transparency
-better known as the open science framework (OSF). Open science framework promotes scientific openness, reproducibility and integrity 
-through an establishment of guidelines for the scientific process (Foster & Deardorff, 2017)
-To this end, van Lissa and colleagues (2020) developed WORCS, a workflow for open reproducible code in science.
+better known as the open science framework (OSF). The OSF promotes scientific openness, reproducibility and integrity 
+through an establishment of guidelines for researchers (Foster & Deardorff, 2017)
+To this end, @van_lissa_worcs_2020 developed WORCS, a workflow for open reproducible code in science.
 WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by Nosek and colleagues (2015).
 It can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
-version control, citation and formatting, among others (Van Lissa et al., 2020)
-
-	* Use comprehensive citation; this includes referencing the software used.
-	* Share code and (if possible) data. Van Lissa and colleagues suggest sharing synthetic data in case the original data cannot be shared, and provide functions to generate such synthetic data.
-	* Ideally, make the entire research project reproducible so that others may download it and reproduce it with just one click; for guidance, see Van Lissa and colleagues.
-
-<!-- CJ: I think this belongs here -->
+version control, citation and formatting, among others [@van_lissa_worcs_2020].
+
+The guidelines emphasise the use of comprehensive citation (including referencing the software used in the analysis),
+as well as code and data sharing wherever possible.
+@van_lissa_worcs_2020 suggest sharing synthetic data in case the original data cannot be shared, 
+and provide functions to generate such synthetic data.
+Ideally, the entire research project is made reproducible 
+so that others may download it and reproduce it with just one click; for guidance, see @van_lissa_worcs_2020.
+
 As the open science movement is gaining momentum,
@@
 
-### Starting values
-
+<!-- ### Starting values -->
 <!-- Different methods have been developed to initialize the estimation (get starting values) in -->

### 2022-09-02 manuscript.Rmd
Updates 02 September
Included relevant info from comments up to and including "Use cases for latent class analysis"

Wrote examples for "Use cases for latent class analysis".

@@
 based on a parametric model of one or more observed indicators of group membership.
-This method has become quite popular across scientific fields, and under a number of different names; most notably (finite Gaussian) mixture modeling and latent profile analysis.
-<!-- Vermunt and Magidson (2004) defined it more generally as virtually any statistical model where “some of the parameters … differ across unobserved subgroups” (Vermunt, J.K.
-and Magidson, J., 2004.
-Latent class analysis.
-In: M.S.
-Lewis-Beck, A.
-Bryman, and T.F.
-Liao (eds.), The Sage Encyclopedia of Social Sciences Research Methods, 549-553.
-Thousand Oaks, CA: Sage Publications).
--->
+This method has become quite popular across scientific fields, and under a number of different names;
+most notably (finite Gaussian) mixture modeling and latent profile analysis.
+@vermunt_jk_latent_2004 defined it more generally as virtually any statistical model where “some of the parameters […] differ across unobserved subgroups”.
+
 Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
-<!-- Prior work resulted in the "GRoLTS-Checklist"; reporting guidelines for a specific subcategory of latent class analysis known as latent growth models. vermuntj.k.LatentClassAnalysis2004-->
-<!-- General reporting guidelines for latent class analysis are still lacking, however. -->
+While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines for a specific subcategory of LCA known as latent growth models,
+general reporting guidelines for latent class analysis are still lacking.
 This complicates manuscript review and assessment of the quality of published studies,
@@
 The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
-Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, this paper exclusively relies on free, open source software for latent class analysis in R.
-Our goal is to make best-practices latent class analysis widely accessible.
+Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, 
+this paper exclusively relies on free, open source software for latent class analysis in R.
+Our goal is to make best-practices in latent class analysis widely accessible.
 
@@
 The concept of latent class analysis can be understood in different ways.
-Generally speaking, it can be said that a mixture model assumes that the study population
-is composed of $K$ subpopulations, or classes.
+Generally speaking, a mixture model assumes that the study population
+is composed of $K$ subpopulations or classes.
 It further assumes that the observed data are a mixture of data generated by class-specific models.
@@
 Commonly, the same model is estimated across all classes, but with different parameters for each class (i.e., class-specific means and variances).
-Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.
+Mixture modeling then estimates both the parameters for each class, 
+and the probability that an individual belongs to each class.
 
@@
 which is close to the true means of the two groups, namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
-When tabulating estimated group membership against observed (known) group membership, it can be seen that women are classified with a high degree of accuracy, but men are not:
+When tabulating estimated group membership against observed (known) group membership, 
+it can be seen that women are classified with a high degree of accuracy, but men are not:
 
@@
 
-One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
 A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
@@
 When the focus is on the model parameters in each group,
-then latent class analysis can be thought of as similar to a multi-group structural equation model.
+then LCA can be thought of as similar to a multi-group structural equation model.
 The main distinction is that group membership is not known a-priori,
 but is instead estimated - with measurement error - based on the data.
-Whereas in a multigroup model, the data are split by group and treated as independent samples,
+Whereas in a multi-group model, the data are split by group and treated as independent samples,
 in a mixture model, all cases contribute to the estimation of all parameters in all groups.
@@
 
-One common type of latent class analyses is the *finite Gaussian mixture model*;
+One common type of LCA is the *finite Gaussian mixture model*;
 a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
@@
 the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
-These parameters can be estimated freely, or constrained, across classes.
+These parameters can be estimated freely, or set to be constrained across classes.
 
@@
 which assumes conditional independence of the indicators.
-Conditional independence means that, 
-after class membership is accounted for,
+Conditional independence means that, after class membership is accounted for,
 the covariances/correlations between indicators are assumed to be zero.
@@
 Nowadays, it is more commonly used as an umbrella term.
-To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as latent class analysis with ordinal indicators.
+To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as *latent class analysis with ordinal indicators*.
 Latent class models with ordinal indicators are parameterized differently from mixture models.
@@
 If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.
+
 This paper will primarily focus on mixture models and special cases thereof,
@@
 
+<!-- DA: Do we still need the following comment? In my opinion the included text is sufficient. -->
 <!-- If a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis.
@@
 
-LCA can be used as a descriptive analysis where a researcher wishes to describe their sample 
-and identify a few prototypes based on many variables.
-<!-- add an example and elaborate -->
+LCA can be used as a descriptive analysis where a researcher wishes to describe their sample and identify a few prototypes based on many variables.
+As an example, if a survey among all Dutch academics was carried out
+with the goal of bringing about a funding reform, 
+LCA could be used to discover the types of publications that get funding.
 
@@
 but don't know how many groups there are, LCA may be an appropriate technique to answer this question.
-<!-- add an example -->
+For example, @hopfer_social_2014 studied the substance use, sexual behavior, and mental health status of urban population in Winnipeg, Canada. 
+The underlying assumption was that there were different risk profiles,
+but their number was not known.
+From a collection of indicators, the LCA provided evidence that there may be four distinct risk profiles in the Winnipeg area.
 
 Another application of LCA is to classify individuals.
-<!-- add an example and elaborate -->
+In a peer harassment study, @giang_using_2008 used latent class analysis to classify over 2,000 sixth grade students into aggressor and victim latent classes.
+The five-class solution comprised of classes of victims, aggressors, and socially adjusted students.
+For instance, it revealed that there were two types of victims: highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
 
 LCA is also appropriate when we wish to identify indicators that capture classes well.
-<!-- add an example and elaborate -->
+High quality indicators are strongly related to the latent variable
+and lead to good class separation.
+This relationship of high quality indicators to the latent variable
+is reflected in very high or very low conditional response probabilities.
+As a rule of thumb, high indicator quality can be indicated 
+by conditional response probabilities above .9 and below .1. 
+Moderate indicator quality can be indicated by probabilities above .8 and below .2,
+and low indicator quality would be above .7 and below .3.
+For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
+Therefore, one could use conditional response probabilities for each item to assess its quality
+with regards to how well it helps separate the latent classes.
+From this, a theory about the selection of indicators could be informed.
+
+<!-- "However, Galindo-Garre and Vermunt (2006) found that indicators with high population values (i.e., conditional response probabilities close to one) were more likely to produce boundary estimates in LCA applications, even though this effect decreased with larger N.
+Still, there is evidence that using high quality indicators is generally beneficial, at least in the context of structural equation models with continuous latent variables (Marsh et al., 1998).
+Having more high quality indicators can stabilize the estimation by increasing the information available to estimate latent variable parameters.
+In this vein, Collins and Wugalter (1992) speculated that having sufficiently high quality indicators may compensate for having few indicators and may aid parameter recovery in LTA models." direct quote from @geiser_is_2014 -->
 
 An extension of LCA is that containing covariates which can be used to predict class membership.
-<!-- add an example and elaborate -->
+In this approach, we not only model the latent class variable based on indicators,
+but we also relate the class membership to other explanatory variables (@vermunt_latent_2017).
+An example of using covariates comes from @nozadi_moderating_2016 
+who applied LCA to identify the probability of children's membership to an anxiety class.
+The authors tested several covariates including children's age, sex, and accuracy scores.
+Age and sex were not found to be related to the children's latent class membership,
+hence these covariates were excluded from the analysis.
+Accuracy scores were related to probabilities of being in anxiety and attention-
+anxiety classes and therefore this covariate was kept as a valuable predictor.
 
 When our interest is the prediction of one or more outcomes,
-LCA can be used to construct classes as categorical predictors.
-<!-- add an example and elaborate -->
+LCA can be used to construct latent classes as categorical predictors.
+@lanza_latent_2013 demonstrated how LCA can be used to classify adolescents into depression classes,
+and subsequently these classes can be used to predict smoking, grades, and delinquency.
+The study showed that the outcomes predicted by class membership can be binary (regular smoking), 
+continuous (grades) or count (delinquency).
 
@@
 as the resulting groups summarize response patterns on a large number of indicators.
-<!-- add an example -->
-
-Finally, LCA can be used to deal with data which violate the normality assumption.
-<!-- add an example and elaborate, check with Caspar if there are other assumptions -->
-
+For example, @macgregor_symptom_2021 investigated symptom profiles among injured U.S. military personnel. 
+They used fifteen dichotomous items from the Post-Deployment Health Assessment survey as LCA indicators.
+Combinatorics informs us that fifteen dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
+Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a method of dimensionality reduction.
+A five class solution was found to have the best fit according to both statistical criteria and clinical interpretability.
+
+<!-- from @macgregor_symptom_2021 "Incorporating LCA with the symptom checklist was a novel method of dimensionality -->
+<!-- reduction, and the identified symptom profiles provided more granularity than examining individual symptoms -->
+<!-- alone.
+There are also limitations that warrant mention.
+At times, LCA can be prone to classification errors, but in the present analysis, the model entropy value above 0.80 indicated good separation of the classes, and the MeanPP values showed a low rate of misclassification." -->
+<!-- [...] -->
+<!-- [listing the five classes] "(1) 50.4% were low -->
+<!-- morbidity (i.e., no or very few symptoms reported), (2) -->
+<!-- 14.0% musculoskeletal (i.e., muscle aches; swollen, stiff or -->
+<!-- painful joints; back pain), (3) 8.9% auditory (i.e., trouble -->
+<!-- hearing; ringing in the ears), (4) 11.1% psycho-cognitive -->
+<!-- (i.e., trouble concentrating, easily distracted; forgetful or -->
+<!-- trouble remembering things; increased irritability), and (5) -->
+<!-- 15.6% multimorbidity (i.e., multiple symptoms reported)." -->
+
+Finally, LCA can be used to deal with data which violate certain assumptions.
+For instance, nested data usually imply a violation of the independence assumption.
+In this case, an extension of LCA called multilevel LCA can be used (e.g. @henry_multilevel_2010).
+As discussed in the shoe size example, LCA can deal with violations of normality.
+In fact, LCA assumes the population distribution in a non-normal mixture of $K$ normal distributions,
+and LCA can discover the value of $K$, i.e. generate a $K$-class solution.
+
+<!-- check with Caspar if there are other assumptions and how to expand this paragraph -->
+
+<!-- OLD: -->
 <!-- CJ: Include these use cases too: -->

### 2022-09-04 example_analysis/Daniel_LCA.Rmd
Three LCA models
Fitted 3 LCA models using tidySEM functionality

@@
 ---
-title: "LCA in R"
+title: "LCA in R using tidySEM"
 output: html_notebook
@@
 
-USING TIDYSEM:
-
-see functions starting in mx
+# LOADING THE DATASET
 
@@
 library(tidySEM) 
-df <- data_mix_ordinal
+df <- data_mix_ordinal #ask Caspar to rename the dataset in documentation
 df[1:4] <- lapply(df, ordered)
-mx_lca(data = df,
-       classes = 2) -> res
+```
+
+# 2 class solution
+
+```{r}
+# fitting the model
+set.seed(123)
+mx_lca(data = df, classes = 2) -> res_2_class
 ```
@@
 ```{r}
-res$class1
-res$latentVars
+# printing model fit table
+print(table_2_class <- tidySEM::table_fit(res_2_class))
 ```
+
+```{r}
+# obtain latent class probabilities
+LC_Probabilities_2_class <- tidySEM::class_prob(res_2_class)
+print(LC_Probabilities_2_class[1:4])
+```
+```{r}
+# conditional item probabilities
+tidySEM::table_prob(res_2_class)
+```
+
+# 3 class solution
+
+```{r}
+# fitting the model
+set.seed(123)
+mx_lca(data = df, classes = 3) -> res_3_class
+```
+
+```{r}
+# printing model fit table
+print(table_3_class <- tidySEM::table_fit(res_3_class))
+```
+
+```{r}
+# obtain latent class probabilities
+LC_Probabilities_3_class <- tidySEM::class_prob(res_3_class)
+print(LC_Probabilities_3_class[1:4])
+```
+```{r}
+# conditional item probabilities
+tidySEM::table_prob(res_3_class)
+```
+
+# 4 class solution
+
+```{r}
+# fitting the model
+set.seed(123)
+mx_lca(data = df, classes = 4) -> res_4_class
+```
+
+```{r}
+# printing model fit table
+print(table_4_class <- tidySEM::table_fit(res_4_class))
+```
+
+```{r}
+# obtain latent class probabilities
+LC_Probabilities_4_class <- tidySEM::class_prob(res_4_class)
+print(LC_Probabilities_4_class[1:4])
+```
+```{r}
+# conditional item probabilities
+tidySEM::table_prob(res_4_class)
+```
+
+Comparing model fit
+
+```{r}
+print("BIC")
+print(c("2 class", table_2_class$BIC)) #lowest
+print(c("3 class", table_3_class$BIC))
+print(c("4 class", table_4_class$BIC))
+
+print("AIC")
+print(c("2 class", table_2_class$AIC))
+print(c("3 class", table_3_class$AIC))
+print(c("4 class", table_4_class$AIC)) #lowest
+
+print("Entropy")
+print(c("2 class", table_2_class$Entropy))
+print(c("3 class", table_3_class$Entropy)) #highest
+print(c("4 class", table_4_class$Entropy))
+```
+Using the Bootstrapped Likelihood Ratio Test to compare models
+
+```{r}
+tidySEM::BLRT(res_2_class)
+```
+
+# APPENDIX
 

### 2022-09-04 example_analysis/Daniel_LCA.Rmd
Update Daniel_LCA.Rmd
Added descriptives and one-class solution

@@
 df[1:4] <- lapply(df, ordered)
+```
+
+# Exploring the data
+
+```{r}
+tidySEM::descriptives(df)
+```
+
+# 1 class solution
+
+```{r}
+# fitting the model
+set.seed(123)
+mx_lca(data = df, classes = 1) -> res_1_class
+```
+
+```{r}
+# printing model fit table
+print(table_1_class <- tidySEM::table_fit(res_1_class))
+```
+
+```{r}
+# obtain latent class probabilities
+LC_Probabilities_1_class <- tidySEM::class_prob(res_1_class)
+print(LC_Probabilities_1_class[1:4])
+```
+```{r}
+# conditional item probabilities
+tidySEM::table_prob(res_1_class)
 ```
@@
 print("BIC")
+print(c("1 class", table_1_class$BIC))
 print(c("2 class", table_2_class$BIC)) #lowest
@@
 print(c("4 class", table_4_class$BIC))
-
+```
+```{r}
 print("AIC")
+print(c("1 class", table_1_class$AIC))
 print(c("2 class", table_2_class$AIC))
@@
 print(c("4 class", table_4_class$AIC)) #lowest
+```
 
+```{r}
 print("Entropy")
+print(c("1 class", table_1_class$Entropy)) # entropy = 1 
 print(c("2 class", table_2_class$Entropy))
-print(c("3 class", table_3_class$Entropy)) #highest
+print(c("3 class", table_3_class$Entropy)) # highest
 print(c("4 class", table_4_class$Entropy))
 ```
+
 Using the Bootstrapped Likelihood Ratio Test to compare models
@@
 ```{r}
-tidySEM::BLRT(res_2_class)
+tidySEM::BLRT()
 ```

### 2022-09-05 manuscript.Rmd
Updates 05 September 22
Expanded on Spurk criticism, wrote about conditional item probabilities, started writing on alternative parametrizations

@@
 
+<!-- DA: An assumption LCA has is that of conditional or local independence. -->
+<!-- LCA assumes conditional independence of indicators. -->
+<!-- This means class membership explains all the relationships between the indicators, -->
+<!-- and when the latent class is controlled for, the indicators become unrelated [@masyn_latent_2013] -->
+
 <!-- OLD: -->
@@
 
-Examining observed data is essential for any analysis,
+Examining observed data is essential for any analysis
 as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
 Pay special attention to level of measurement of the indicators.
-Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
+Finite Gaussian mixture models (including LPA) are strictly speaking only suitable for continuous variables.
 Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
@@
 It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
-The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
+The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; 
+in this case, each class-specific mean could describe a single response category, 
+and a class-specific variance component would be nonsensical.
 In sum, Likert-type scales are rarely suitable for mixture modeling;
@@
 When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
-<!-- CJ: Expand on this here -->
+In our shoe size example, it can be seen that the population distribution comprised of two normal distributions
+is bimodal and hence not normal. 
+Yet, this is a prototypical example of a mixed population distribution where LCA can discover latent groups.
+If the population distribution were instead normal, there would be no classes to extract as the whole population would belong to a single class.
 
@@
 Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @van_lissa_worcs_2021.
-
-<!--  Adjust the following part, it belongs here:  -->
-
-<!-- CJ: This is about model interpretation; conditional item probabilities are the analogue of mean and standard deviation, but when the observed data are binary/ordinal  -->
-<!-- Conditional item probabilities indicate the probability of a positive score on an item  -->
-<!-- given that the person belongs to a particular latent class.  -->
-
-<!-- If a particular item is endorsed by two or more classes at markedly different rates, -->
-<!-- it is said to discriminate well between classes and is consequently considered a good indicator. -->
-<!-- The classes are considered highly homogeneous with respect to an item when for a particular item -->
-<!-- there are a distinct difference in conditional item probabilities of two or more classes. -->
-<!-- If an item is endorsed below 30% for one class and above 70% for another class, 
-the classes have high homogeneity with respect to this item [@masyn_latent_2013]. -->
-<!-- However, if many conditional item probabilities in estimation are found to be extreme (as in exactly 0 or 1),  -->
-<!-- we are observing boundary parameter estimates. This can be a sign of an invalid solution,  -->
-<!-- warn us that too many classes were extracted, or indicate a local optimum.  -->
-<!-- Boundary parameter estimates are more likely to happen when working with sparse data  -->
-<!-- and should be interpreted with caution (Geiser, 2012).  -->
-
-<!-- CJ: This talks about the problem of inadmissible solutions.
-With ordinal items, that can happen because of sparse cross-tables (e.g., some combinations of responses never occur).
-WIth continuous items, that could happen e.g.
-because you have a zero-inflated, otherwise normally distributed variable; one class would have a mean at zero and its SD would be hard to determine.
-Or it could happen if you model a five-category Likert scale with a five-component latent profile analysis; each response category will have its own mean, but there's no variance around that (everybody in class 3 has exactly the response value 3).
-These are all the same problems.
--->
-
-<!-- An important outcome of LCA are unconditional class probabilities.  -->
-<!-- These communicate the probability that an individual belongs to each class.  -->
-<!-- A limitation is that our final class solution need not reflect true class membership.  -->
-<!-- Once we do accept a particular class solution, special care also needs to be taken when naming the classes.  -->
+Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @van_lissa_worcs_2021].
+
+<!-- DA: Does this following part about conditional item probabilities and inadmissible solutions belong here?  It's about model interpretation.-->
+
+An important outcome of LCA are conditional item probabilities,
+also known as class-specific item probabilities [@masyn_latent_2013], 
+conditional response or conditional solution probabilities [@geiser_data_2012].
+They indicate the probability of an item being endorsed
+given that the observation belongs to a particular latent class.
+Conditional item probabilities can be obtained using `tidySEM::table_prob()`
+If a particular item is endorsed by two or more classes at markedly different rates,
+it is said to discriminate well between the classes and is consequently considered a good indicator.
+Classes are considered highly homogeneous with respect to an item when for a particular item
+there is a distinct difference in conditional item probabilities for two or more classes.
+For instance, if an item is endorsed below 30% for one class and above 70% for another class,
+the classes have high homogeneity with respect to this item [@masyn_latent_2013].
+Conditional item probabilities are the analogue of mean and standard deviation
+when the indicators are binary/ordinal.
 
@@
 This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
-Such boundary parameter estimates could indicate that the solution is invalid.
+Such boundary parameter estimates could indicate that the solution is invalid [@geiser_data_2012].
 Boundary parameter estimates can also happen with continuous indicators.
@@
 
-### Missing data
+### Handling missing data
 
@@
 The "correct" model specification is then determined based on a combination of fit indices,
-significance tests,
-and interpretability.
+significance tests, and interpretability.
 For latent profile analysis, the function `tidySEM::mx_profiles(classes, variances, covariances)`
@@
 
-<!-- see the following article and expand on the paragraph -->
-@van_de_schoot_grolts-checklist_2017 argue that there are many choices to be made in the specification of latent trajectory models,
-and that alternative
-
-	* Consider alternative parameterizations; e.g., free variances/covariances 
-	* See Nylund
-
-## Software
-
-Many software packages are available for the estimation of latent class analyses.
-Some of these packages have limited functionality, or implement specific innovations.
-Other packages implement latent class analyses in the context of a more flexible structural equation modeling framework.
-The most notable examples of the latter are the commercial programs Mplus and Latent GOLD, 
-and the free open source R-package OpenMx.
-The commercial packages stand out because they offer relatively user-friendly interfaces
-and implement sensible defaults for complex analyses, including latent class analysis.
-This lowers the threshold for applied researchers to adopt such methods.
-Commercial software also has several downsides, however.
-One such downside is that use of the software is restricted to those individuals and institutions who can afford a license.
-A second downside is that the source code, being proprietary, cannot be audited, debugged, or enhanced by third parties.
-This incurs the risk that mistakes in the source code may go unnoticed, and curbs progress as software developers cannot add new functionality.
-
-Conversely, the free open source program OpenMx is very flexible,
-but not very user-friendly. We directly address this limitation using the `tidySEM` R-package.
-New functionality in `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
-It adheres to best practices in estimation and reporting, as described in this paper.
-The user interface is simple, making use of the model syntax of the widely used `lavaan` R-package.
-This syntax offers a human-readable way to specify latent variable models.
-Minor enhancements are made to simplify the specification of latent class analysis.
-
-Because of the limitations of the aforementioned tools, we set out to develop a free tool that
-provides sensible defaults and is easy to use, but provides the option to
-access and modify all of the model inputs (i.e., low barrier, high
-ceiling). `tidySEM` interfaces with existing tools, and is able to translate between
-what existing tools are capable of and what researchers and analysts
-carrying-out person-oriented analyses would like to specify. 
-Furthermore,`tidySEM` facilitates fully-reproducible analyses and contributes to open science.
-
-<!-- 3. Packages -->
-<!-- 3.1 R-packages -->
-<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
-<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
-<!-- analytic functionality. However, the majority of packages are focused on analyses with discrete -->
-<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
-<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
-<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
-<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
-<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
-<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
-<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
-<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
-<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
-<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
-<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->
-
-<!-- Although ‘tidyLPA’ is easy to use, this comes at the expense of restricting modeling options -->
-<!-- to a few, oft-used options. For completeness, both the ‘mclust’ and ‘tidyLPA’ approach will -->
-<!-- be illustrated. -->
-
-## Best practices in estimation
-
-### Algorithm
-
-Mixture model parameters and model fit statistics can be estimated in a variety of ways.
-The choice of the estimator depends on the presence of missing values,
-sample size, number of indicators, and available computational resources [@weller_latent_2020].
-A commonly used technique is maximum likelihood (ML) estimation
-with the expectation-maximization (EM) algorithm as a local optimizer. 
-Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator
-(ignoring the variance for now). The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes the likelihood ($LL$) of all observed data. 
-In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. 
-We can think of this optimization problem as a three-dimensional landscape: 
-The X and Y dimensions are determined by the class-specific means, 
-so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. 
-The optimizer must find the deepest "valley" in this landscape, 
-which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble, dropped in this landscape.
-It is dropped at some random point in space, and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
-(this is known as "convergence").
-It cannot climb out again.
-Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley 
-(a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
-One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
-choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
-This is the "random starts" approach.
-
-One problem with the random starts approach is that 
-it is computationally expensive to run this many replications.
-Moreover, because the algorithm begins with random starting values,
-many of the marbles are likely to be very far away from a "good enough" solution.
-Two innovations may improve the estimation procedure.
-The first is that, instead of picking random starting values, 
-a "reasonable solution" may be used for the starting values.
-For example, if we assume that the different classes are likely to have different mean values on the indicators, 
-then the K-means clustering algorithm can be used to determine these cluster centroids.
-We can compute the expected values of all model parameters by treating the K-means solution as a known class solution,
-and use these as starting values for a mixture model.
-<!-- In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
-Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
-[@biernackiChoosingStartingValues2003].
--->
-One remaining concern is that this approach may result in starting values close to a local optimum,
-and that the EM algorithm will thus never find the global optimum.
-A second innovation addresses this concern.
-
-Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some "destination" in the landscape, 
-and compares its likelihood to the current one. If the destination likelihood is higher, 
-the estimator moves there.
-If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
-This latter property allows it to escape local optima, and find the global optimum.
-
-By default, `tidySEM` employs this solution of deriving starting values using K-means clustering,
-and identifying the global optimum solution using simulated annealing.
-Once a solution has been found,
-simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for the parameters that can be used to compute standard errors.
-Note that these defaults can be manually overridden.
-
-One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
-This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
-As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
-<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->
-
-<!-- We assume that the population consists of several groups each characterized by its own normal distribution.  -->
-<!-- When two or more normal distributions with different means and standard deviations comprise a mixed population distribution,  -->
-<!-- this combined distribution is no longer normal.  -->
-<!-- For example, you can imagine a bimodal distribution when dealing with two latent classes.  -->
-<!-- If this were not the case and our mixed population distribution were normal,  -->
-<!-- there would be no latent classes to extract as the population would consists of a single class only.  -->
-<!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups.  -->
-<!-- Secondly, the claim that maximum likelihood with robust standard errors should be preferred  -->
-<!-- implicitly implies that only software packages with this estimation technique can be used.  -->
-
-<!-- ### Starting values -->
-
-<!-- The EM algorithm needs starting values for each parameter it wants to estimate.  -->
-<!-- If it were provided a single starting value for each parameter,  -->
-<!-- the EM optimizer would be at risk of finding a local optimum instead of the global optimum.  -->
-<!-- In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates.  -->
-<!-- To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter -->
-<!-- in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004).  -->
-<!-- Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values. -->
-<!-- <!-- add more on this --> -->
-<!-- Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).  -->
-
-<!-- Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence -->
-<!-- or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004). -->
-<!-- The key is to use a large number of starting value sets.  -->
-<!-- Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, -->
-<!-- and a greater number of sets for more complex models.  -->
-<!-- A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately. -->
-
-### Class enumeration
-
-LCA can be done in an exploratory or in a confirmatory fashion.
-In exploratory LCA, a sequence of models is fitted to the data 
-with each additional model estimating one more class than the previous model.
-These models are then compared and the best solution is selected as the final class solution.
-In some cases, prior theory can inform the researcher about the number of classes to expect.
-Even in such confirmatory LCA cases, it is nonetheless useful to know 
-if the theoretical model is markedly better than those with differing numbers of classes.
-Therefore, it may always be useful to compare different class solutions.
-
-From a sequence of models, the final class solution is chosen based on both theoretical and statistical criteria.
-Theory should drive the selection of indicator variables, inform the expectations and reflect on the findings.
-In addition to this, there are several statistical criteria to consider in model selection.
-These include but are not limited to likelihood ratio tests, information criteria, 
-and the Bayes factor [@weller_latent_2020].
-
-Relative model fit can be examined using the likelihood ratio test. 
-This is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as the difference in maximum log likelihoods of the two models,
-with the new degrees of freedom being the difference in their degrees of freedom.
-This statistic also follows the $\chi^2$ distribution.
-Similar to the LR $\chi^2$ goodness-of-fit test, we want the test statistic to be non-significant
-in order to give support to the simpler model. 
-The likelihood ratio test can only compare two nested models at a time [@lanza_introduction_2003].
-
-<!-- DA: This is just a goodness-of-fit test, and if it is significant the model fits the data poorly.  -->
-<!-- We don't get this test with `tidySEM`, so do not focus on this too much. We only provide information criteria-->
-
-<!-- DA: the bootstrap likelihood ratio test is the only one that is valid one according to Caspar, but it is not easily implemented in `tidySEM` (too computationally intensive), which is a problem, takes a lot of time to compute, maybe write about this,
-see @masyn_latent_2013 'parametric bootstrapping' it is not widely implemented. 
-Do not write about bootstrapping, too computationally intensive, we do not have a good solution for model fit tests, downplay that, MPLUS has MLR adjusted maximum likelihood test but unsure if to compare two different class solutions or just misfit, however not something we have, it is a specialized solution available only in MPLUS-->
-
-### Model fit indices
-
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2*log likelihood (which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance fit and model complexity.
-The lower the value of an information criterion, the better the overall fit of the model.
-The BIC applies a stronger penalty for model complexity 
-that scales logarithmically with the sample size.
-The literature suggests the BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
-Both the AIC and the BIC are available in the `tidySEM` output.
-
-Information criteria may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016].
-Finally, one might make an elbow plot and compare multiple information criteria
-[for an example see @nylund-gibson_ten_2018].
-
-Another common test of model fit is the likelihood ratio $\chi^2$ goodness-of-fit test.
-However, this test is not implemented in `tidySEM`.
-
-An alternative is using the bootstrapped likelihood ratio test
-which can be run using `tidySEM::BLRT()`.
-Currently, this test is computationally expensive and can be slow on most computers.
-The development of a faster version of this test is well under way.
-
-<!-- DA: NOTES ABOUT LRT test: It can compare two nested models when they have the same number of classes
-<!-- If we wish to simultaneously compare multiple models based on their relative fit, -->
-<!-- this can be done through a comparison of multiple information criteria. -->
-<!-- Examples include the Akaike information criterion (AIC), -->
-<!-- and the Bayesian information criterion (BIC). -->
-<!-- Information criteria are a sum of a measure of fit (usually a form of the converged maximum log likelihood value)  -->
-<!-- and a penalty for model complexity (a combination of sample size and the number of modeled parameters). -->
-<!-- As a general rule, the lower the value of an information criterion, the better the model fits the data. -->
-<!-- When examining several competing models, each with differing number of classes, -->
-<!-- we expect the information criteria values to drop with each successive model until the final class solution is reached. -->
-<!-- Further models with additional classes should show worse fit as information criteria reach an optimum  -->
-<!-- before their values rise again with an increasing number of classes. -->
-<!-- Multiple information criteria should be compared when choosing the final class solution.  -->
-<!-- This can be done by means of an elbow plot [for an example see @nylund-gibson_ten_2018]. -->
-<!-- Advantages of using information criteria are that we can compare multiple models at a time, -->
-<!-- and these models need not be nested @masyn_latent_2013. -->
-
-<!-- Add a paragraph about using the Bayes factor see @weller_latent_2020-->
-
-<!-- decide whether to keep the following paragraph, if so, where to place it -->
-<!-- When estimators which require sets of starting values are used, -->
-<!-- each fitted model will have a corresponding log likelihood value.  -->
-<!-- Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times  -->
-<!-- with each log likelihood value generated by a different starting value set.  -->
-<!-- Successful replications of the largest log likelihood value across different starting values -->
-<!-- instill confidence that the estimation was successful and we indeed found the global maximum [@masyn_latent_2013].  -->
-<!-- The model with the highest replicated likelihood will be chosen as our best-fit.  -->
-<!-- If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012). -->
-<!-- Such a model does not have a unique solution. For a discussion on model identification for LCA, see @masyn_latent_2013. -->
-<!-- Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated  -->
-<!-- in at least 3-10% of the models initialized with different starting values in the first step of the optimization.  -->
-<!-- This consideration does not apply in case of models which use simulated annealing since these do not have starting values. -->
-<!-- @masyn_latent_2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
-<!-- to ensure that a global (rather than a local) maximum has been reached.". For discussion see Hipp & Bauer, 2006 -->
-
-<!-- Absolute model fit can be assessed using the likelihood ratio (LR) $\chi^2$ goodness-of-fit test.  -->
-<!-- It can compare two nested models when they have the same number of classes (Lanza et al., 2003). -->
-<!-- Typically, we compare our final class solution (a constrained model)  -->
-<!-- to an unconstrained one (a model characterized by the perfect fit). -->
-<!-- The former model is a special, less complex version of the latter. -->
-<!-- In this case, we compare our model-based response pattern frequencies to those in the dataset. -->
-<!-- Therefore, we make a statement about how well our model fits the data. -->
-<!-- The perfect fit would be indicated by the $\chi^2\ =\ 0$ and $p\ =\ 1$. -->
-<!-- Ideally, the LR $\chi^2$ test will be non-significant as this implies -->
-<!-- that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model. -->
-<!-- The underlying assumption of this procedure is that the test statistic is asymptotically distributed -->
-<!-- as a $\chi^2$ meaning that it can be approximated using this distribution. -->
-<!-- This is the case when the sample size is adequate (Lanza et al., 2003).  -->
-<!-- The dependence of the LR $\chi^2$ test statistic on sample size has its downsides [@masyn_latent_2013].  -->
-<!-- Very large samples will often effortlessly reject the null even when the final class solution is a good fit, -->
-<!-- and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic. -->
-<!-- In fact, small samples might not be well approximated using the $\chi^2$ distribution, -->
-<!-- which is a problem shared with large yet sparse datasets. -->
-<!-- In such cases, special caution is advised when interpreting the resulting p-values [@masyn_latent_2013]. -->
-
-### Classification Diagnostics
-
-Best models will divide the sample into subgroups which are internally homogeneous and externally distinct.
-Classification diagnostics give us a way to assess the degree to which this is the case.
-They are separate from model fit indices as a model can fit the data well 
-but show poor latent class separation [@masyn_latent_2013].
-Classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions because they are uninterpretable.
-Interpretability should always be a consideration when considering different class solutions [@nylund-gibson_ten_2018].
-
-Three important classification diagnostics provided by `tidySEM` are are entropy, 
-the range of the posterior classification probabilities by most likely class membership,
-and the range of the sample size.
-
-Entropy is a summary measure of posterior class probabilities across classes and individuals.
-It ranges from 0 (model classification no better than random chance) to 1 (perfect classification).
-As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
-Entropy should not be used to select a particular class solution. 
-An appropriate use of entropy is that it can disqualify certain solutions if class separability is too low 
-or if one of the latent classes is too small to be meaningful or to calculate descriptive statistics.
-
-The range of the posterior classification probabilities by most likely class membership are
-
-Another classification diagnostic measure is the range of the sample size.
-
-<!-- A fundamental concept when examining classification accuracy is the posterior class probability. -->
-<!-- This concept represents the probability of belonging to each latent class computed for each individual. -->
-<!-- The highest posterior class probability is then determined and the individual is assigned to the corresponding class. -->
-<!-- We want each individual's posterior class probabilities to be high for one and low for the remaining latent classes. -->
-<!-- This is considered a high classification accuracy and means that the classes are distinct. -->
-
-<!-- @masyn_latent_2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
-<!-- add content about entropy -->
-
-<!-- DA: Write about prob_min and prob_max (see name formatting), -->
-<!-- the posterior probability from the model, -->
-<!-- it calculates the entropy and the range of the most likely classification probability, -->
-<!-- the diagonal, lets say you have 3 classes, there is the class the person was assigned to, -->
-<!-- and the probability the person should belong to that class, -->
-<!-- what this is telling us is the range of the average classification probabilities, -->
-<!-- what you want to see is that people have a high probability of being in the class they were assigned to, -->
-<!-- and you want that to be high for all the classes,  -->
-<!-- for each class we want a high number -->
-
-<!-- n_min and n_max are the range of the proprotion of people assigned to those classes. -->
-<!-- The smallest class has XX percent of people, and the largest class has XXX percent of people. -->
-<!-- Sometimes the class is too small to be interpreted, e.g. 5% or less of people, that would be too few to estimate parameters.  -->
-
-<!-- The average posterior probability is a measure of classification uncertainty for each latent class. -->
-
-<!-- The modal class assignment proportion -->
-
-`tidySEM::table_prob()` returns thresholds for ordinal dependent variables in probability scale.
-These are conditional item probabilities
-which communicate item endorsement given a class
-for models with binary and ordinal variables.
-
-### Label switching
-<!-- optional subsection: interpreting class solutions (with class naming) -->
-
-The final class solution will usually discover and enumerate several classes. 
-The class ordering however is completely arbitrary.
-The class labeled as Class 1 in one solution may become Class 2 or Class 3 in another model,
-even when the only difference between the models is in their starting values. 
-Label switching is something to be mindful of when comparing different LCA models [@masyn_latent_2013].
-
-The order of clusters is nondeterministic when using K-means in `tidySEM`. 
-Therefore label switching is still a consideration.
-A simple solution to this is setting a random seed number one line prior to fitting the model.
-We advise `tidySEM` users to always do so in order to circumvent label switching.
-
-<!-- Article discussing label switching: see Chung, Loken & Schafer -->
-
-<!-- Initialization around a Single Mode -->
-<!-- Another common approach is to run a single chain or to initialize the parameters near realistic values.36 -->
-
-<!-- This can work better than the hard constraint approach if reasonable initial values can be found and the labels do not switch within a Markov chain.
-The result is that all chains are glued to a neighborhood of a particular mode in the posterior.
--->
-<!-- https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/1467-9868.00265?casa_token=Tjk8gR4w8AEAAAAA:RygJL5OgdBaVWY6oaB8XlnyMOnzQgOVDBd2EHDyQD6TH-DzxRxpmU1J4AG8N-HpL5dHHNINRybiaMl6w -->
-
-Class names should be chosen to accurately reflect group membership. 
-Overly simplified and generalized class names may prove misleading to both audiences and researches alike
-leading to what is known as a naming fallacy [@weller_latent_2020].
-
-## Best practices in reporting
-
-Among studies using LCA, reporting practices vary significantly [@weller_latent_2020].
-Various authors have tried to better and standardize ways of reporting LCA [e.g. @masyn_latent_2013; @weller_latent_2020]. 
-Building on their work, we provide guidelines with a strong emphasis on scientific reproducibility and transparency
-better known as the open science framework (OSF). The OSF promotes scientific openness, reproducibility and integrity 
-through an establishment of guidelines for researchers (Foster & Deardorff, 2017)
-To this end, @van_lissa_worcs_2020 developed WORCS, a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by Nosek and colleagues (2015).
-It can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
-version control, citation and formatting, among others [@van_lissa_worcs_2020].
-
-The guidelines emphasise the use of comprehensive citation (including referencing the software used in the analysis),
-as well as code and data sharing wherever possible.
-@van_lissa_worcs_2020 suggest sharing synthetic data in case the original data cannot be shared, 
-and provide functions to generate such synthetic data.
-Ideally, the entire research project is made reproducible 
-so that others may download it and reproduce it with just one click; for guidance, see @van_lissa_worcs_2020.
-
-As the open science movement is gaining momentum,
-researchers are becoming increasingly aware
-how important it is that analyses can be reproduced and audited.
-In line with open science principles, one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., *Mplus* and *Latent GOLD*).
-A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because the underlying source code is proprietary.
-To overcome these limitations, the present paper introduces new user-friendly functions in the `tidySEM` R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package `OpenMx`.
-The reporting guidelines described in this paper are adopted in `tidySEM` by default.
-The `tidySEM` R-package thus makes advanced mixture modeling based on best practices widely accessible,
-and facilitates the adoption of the estimation and reporting guidelines described in this paper.
+In order to aid researchers working with latent trajectory models,
+@van_de_schoot_grolts-checklist_2017 developed a protocol called Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
+Throughout this paper, we adapt those GRoLTS protocols that are applicable to latent class analysis.
+One of the guidelines (namely, 6a) refers to considering alternative model specifications.
+In LCA, means, variances, and covariances between the indicators 
+can either be constrained across classes, or estimated freely.
+Researchers using LCA should transparently report their chosen parametrization 
+as well as discuss different parametrizations that were tested in the process.
+
+<!-- DA: - would researchers ideally anticipate the parameterization of the within-class var/cov structure? How does preregistration connect to this?  -->
 
@@
 ```
+
+## Software
+
+Many software packages are available for the estimation of latent class analyses.
+Some of these packages have limited functionality, or implement specific innovations.
+Other packages implement latent class analyses in the context of a more flexible structural equation modeling framework.
+The most notable examples of the latter are the commercial programs Mplus and Latent GOLD, 
+and the free open source R-package OpenMx.
+The commercial packages stand out because they offer relatively user-friendly interfaces
+and implement sensible defaults for complex analyses, including latent class analysis.
+This lowers the threshold for applied researchers to adopt such methods.
+Commercial software also has several downsides, however.
+One such downside is that use of the software is restricted to those individuals and institutions who can afford a license.
+A second downside is that the source code, being proprietary, cannot be audited, debugged, or enhanced by third parties.
+This incurs the risk that mistakes in the source code may go unnoticed, and curbs progress as software developers cannot add new functionality.
+
+Conversely, the free open source program OpenMx is very flexible,
+but not very user-friendly. We directly address this limitation using the `tidySEM` R-package.
+New functionality in `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
+It adheres to best practices in estimation and reporting, as described in this paper.
+The user interface is simple, making use of the model syntax of the widely used `lavaan` R-package.
+This syntax offers a human-readable way to specify latent variable models.
+Minor enhancements are made to simplify the specification of latent class analysis.
+
+Because of the limitations of the aforementioned tools, we set out to develop a free tool that
+provides sensible defaults and is easy to use, but provides the option to
+access and modify all of the model inputs (i.e., low barrier, high
+ceiling). `tidySEM` interfaces with existing tools, and is able to translate between
+what existing tools are capable of and what researchers and analysts
+carrying-out person-oriented analyses would like to specify. 
+Furthermore,`tidySEM` facilitates fully-reproducible analyses and contributes to open science.
+
+<!-- 3. Packages -->
+<!-- 3.1 R-packages -->
+<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
+<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
+<!-- analytic functionality. However, the majority of packages are focused on analyses with discrete -->
+<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
+<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
+<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
+<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
+<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
+<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
+<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
+<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
+<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
+<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
+<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->
+
+<!-- Although ‘tidyLPA’ is easy to use, this comes at the expense of restricting modeling options -->
+<!-- to a few, oft-used options. For completeness, both the ‘mclust’ and ‘tidyLPA’ approach will -->
+<!-- be illustrated. -->
+
+## Best practices in estimation
+
+### Algorithm
+
+Mixture model parameters and model fit statistics can be estimated in a variety of ways.
+The choice of the estimator depends on the presence of missing values,
+sample size, number of indicators, and available computational resources [@weller_latent_2020].
+A commonly used technique is maximum likelihood (ML) estimation
+with the expectation-maximization (EM) algorithm as a local optimizer. 
+Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator
+(ignoring the variance for now). The EM algorithm will attempt to find a combination of values
+for these two parameters that maximizes the likelihood ($LL$) of all observed data. 
+In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. 
+We can think of this optimization problem as a three-dimensional landscape: 
+The X and Y dimensions are determined by the class-specific means, 
+so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. 
+The optimizer must find the deepest "valley" in this landscape, 
+which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
+The EM optimizer behaves somewhat like a marble, dropped in this landscape.
+It is dropped at some random point in space, and will roll into the nearest valley.
+The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
+(this is known as "convergence").
+It cannot climb out again.
+Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley 
+(a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
+One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
+choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
+This is the "random starts" approach.
+
+One problem with the random starts approach is that 
+it is computationally expensive to run this many replications.
+Moreover, because the algorithm begins with random starting values,
+many of the marbles are likely to be very far away from a "good enough" solution.
+Two innovations may improve the estimation procedure.
+The first is that, instead of picking random starting values, 
+a "reasonable solution" may be used for the starting values.
+For example, if we assume that the different classes are likely to have different mean values on the indicators, 
+then the K-means clustering algorithm can be used to determine these cluster centroids.
+We can compute the expected values of all model parameters by treating the K-means solution as a known class solution,
+and use these as starting values for a mixture model.
+<!-- In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
+Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
+[@biernackiChoosingStartingValues2003].
+-->
+One remaining concern is that this approach may result in starting values close to a local optimum,
+and that the EM algorithm will thus never find the global optimum.
+A second innovation addresses this concern.
+
+Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
+Simulated annealing iteratively considers some "destination" in the landscape, 
+and compares its likelihood to the current one. If the destination likelihood is higher, 
+the estimator moves there.
+If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
+This latter property allows it to escape local optima, and find the global optimum.
+
+By default, `tidySEM` employs this solution of deriving starting values using K-means clustering,
+and identifying the global optimum solution using simulated annealing.
+Once a solution has been found,
+simulated annealing is followed up with a short run of the EM algorithm,
+as EM inherently produces an asymptotic covariance matrix for the parameters that can be used to compute standard errors.
+Note that these defaults can be manually overridden.
+
+One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
+This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
+As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
+<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->
+
+<!-- We assume that the population consists of several groups each characterized by its own normal distribution.  -->
+<!-- When two or more normal distributions with different means and standard deviations comprise a mixed population distribution,  -->
+<!-- this combined distribution is no longer normal.  -->
+<!-- For example, you can imagine a bimodal distribution when dealing with two latent classes.  -->
+<!-- If this were not the case and our mixed population distribution were normal,  -->
+<!-- there would be no latent classes to extract as the population would consists of a single class only.  -->
+<!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups.  -->
+<!-- Secondly, the claim that maximum likelihood with robust standard errors should be preferred  -->
+<!-- implicitly implies that only software packages with this estimation technique can be used.  -->
+
+<!-- ### Starting values -->
+
+<!-- The EM algorithm needs starting values for each parameter it wants to estimate.  -->
+<!-- If it were provided a single starting value for each parameter,  -->
+<!-- the EM optimizer would be at risk of finding a local optimum instead of the global optimum.  -->
+<!-- In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates.  -->
+<!-- To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter -->
+<!-- in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004).  -->
+<!-- Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values. -->
+<!-- <!-- add more on this --> -->
+<!-- Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).  -->
+
+<!-- Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence -->
+<!-- or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004). -->
+<!-- The key is to use a large number of starting value sets.  -->
+<!-- Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, -->
+<!-- and a greater number of sets for more complex models.  -->
+<!-- A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately. -->
+
+### Class enumeration
+
+LCA can be done in an exploratory or in a confirmatory fashion.
+In exploratory LCA, a sequence of models is fitted to the data 
+with each additional model estimating one more class than the previous model.
+These models are then compared and the best solution is selected as the final class solution.
+In some cases, prior theory can inform the researcher about the number of classes to expect.
+Even in such confirmatory LCA cases, it is nonetheless useful to know 
+if the theoretical model is markedly better than those with differing numbers of classes.
+Therefore, it may always be useful to compare different class solutions.
+
+From a sequence of models, the final class solution is chosen based on both theoretical and statistical criteria.
+Theory should drive the selection of indicator variables, inform the expectations and reflect on the findings.
+In addition to this, there are several statistical criteria to consider in model selection.
+These include but are not limited to likelihood ratio tests, information criteria, 
+and the Bayes factor [@weller_latent_2020].
+
+Relative model fit can be examined using the likelihood ratio test. 
+This is only appropriate when the two models we wish to compare are nested.
+The likelihood ratio test statistic is computed as the difference in maximum log likelihoods of the two models,
+with the new degrees of freedom being the difference in their degrees of freedom.
+This statistic also follows the $\chi^2$ distribution.
+Similar to the LR $\chi^2$ goodness-of-fit test, we want the test statistic to be non-significant
+in order to give support to the simpler model. 
+The likelihood ratio test can only compare two nested models at a time [@lanza_introduction_2003].
+
+<!-- DA: This is just a goodness-of-fit test, and if it is significant the model fits the data poorly.  -->
+<!-- We don't get this test with `tidySEM`, so do not focus on this too much. We only provide information criteria-->
+
+<!-- DA: the bootstrap likelihood ratio test is the only one that is valid one according to Caspar, but it is not easily implemented in `tidySEM` (too computationally intensive), which is a problem, takes a lot of time to compute, maybe write about this,
+see @masyn_latent_2013 'parametric bootstrapping' it is not widely implemented. 
+Do not write about bootstrapping, too computationally intensive, we do not have a good solution for model fit tests, downplay that, MPLUS has MLR adjusted maximum likelihood test but unsure if to compare two different class solutions or just misfit, however not something we have, it is a specialized solution available only in MPLUS-->
+
+### Model fit indices
+
+Fit indices typically used for determining the optimal number of classes include
+the Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC).
+Both information criteria are based on the -2*log likelihood (which is lower for better fitting models),
+and add a penalty for the number of parameters (thus incentivizing simpler models).
+This helps balance fit and model complexity.
+The lower the value of an information criterion, the better the overall fit of the model.
+The BIC applies a stronger penalty for model complexity 
+that scales logarithmically with the sample size.
+The literature suggests the BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
+Both the AIC and the BIC are available in the `tidySEM` output.
+
+Information criteria may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
+One option is to select a specific fit index before analyzing the data.
+Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
+Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016].
+Finally, one might make an elbow plot and compare multiple information criteria
+[for an example see @nylund-gibson_ten_2018].
+
+Another common test of model fit is the likelihood ratio $\chi^2$ goodness-of-fit test.
+However, this test is not implemented in `tidySEM`.
+
+An alternative is using the bootstrapped likelihood ratio test
+which can be run using `tidySEM::BLRT()`.
+Currently, this test is computationally expensive and can be slow on most computers.
+The development of a faster version of this test is well under way.
+
+<!-- DA: NOTES ABOUT LRT test: It can compare two nested models when they have the same number of classes
+<!-- If we wish to simultaneously compare multiple models based on their relative fit, -->
+<!-- this can be done through a comparison of multiple information criteria. -->
+<!-- Examples include the Akaike information criterion (AIC), -->
+<!-- and the Bayesian information criterion (BIC). -->
+<!-- Information criteria are a sum of a measure of fit (usually a form of the converged maximum log likelihood value)  -->
+<!-- and a penalty for model complexity (a combination of sample size and the number of modeled parameters). -->
+<!-- As a general rule, the lower the value of an information criterion, the better the model fits the data. -->
+<!-- When examining several competing models, each with differing number of classes, -->
+<!-- we expect the information criteria values to drop with each successive model until the final class solution is reached. -->
+<!-- Further models with additional classes should show worse fit as information criteria reach an optimum  -->
+<!-- before their values rise again with an increasing number of classes. -->
+<!-- Multiple information criteria should be compared when choosing the final class solution.  -->
+<!-- This can be done by means of an elbow plot [for an example see @nylund-gibson_ten_2018]. -->
+<!-- Advantages of using information criteria are that we can compare multiple models at a time, -->
+<!-- and these models need not be nested @masyn_latent_2013. -->
+
+<!-- Add a paragraph about using the Bayes factor see @weller_latent_2020-->
+
+<!-- decide whether to keep the following paragraph, if so, where to place it -->
+<!-- When estimators which require sets of starting values are used, -->
+<!-- each fitted model will have a corresponding log likelihood value.  -->
+<!-- Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times  -->
+<!-- with each log likelihood value generated by a different starting value set.  -->
+<!-- Successful replications of the largest log likelihood value across different starting values -->
+<!-- instill confidence that the estimation was successful and we indeed found the global maximum [@masyn_latent_2013].  -->
+<!-- The model with the highest replicated likelihood will be chosen as our best-fit.  -->
+<!-- If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012). -->
+<!-- Such a model does not have a unique solution. For a discussion on model identification for LCA, see @masyn_latent_2013. -->
+<!-- Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated  -->
+<!-- in at least 3-10% of the models initialized with different starting values in the first step of the optimization.  -->
+<!-- This consideration does not apply in case of models which use simulated annealing since these do not have starting values. -->
+<!-- @masyn_latent_2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
+<!-- to ensure that a global (rather than a local) maximum has been reached.". For discussion see Hipp & Bauer, 2006 -->
+
+<!-- Absolute model fit can be assessed using the likelihood ratio (LR) $\chi^2$ goodness-of-fit test.  -->
+<!-- It can compare two nested models when they have the same number of classes (Lanza et al., 2003). -->
+<!-- Typically, we compare our final class solution (a constrained model)  -->
+<!-- to an unconstrained one (a model characterized by the perfect fit). -->
+<!-- The former model is a special, less complex version of the latter. -->
+<!-- In this case, we compare our model-based response pattern frequencies to those in the dataset. -->
+<!-- Therefore, we make a statement about how well our model fits the data. -->
+<!-- The perfect fit would be indicated by the $\chi^2\ =\ 0$ and $p\ =\ 1$. -->
+<!-- Ideally, the LR $\chi^2$ test will be non-significant as this implies -->
+<!-- that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model. -->
+<!-- The underlying assumption of this procedure is that the test statistic is asymptotically distributed -->
+<!-- as a $\chi^2$ meaning that it can be approximated using this distribution. -->
+<!-- This is the case when the sample size is adequate (Lanza et al., 2003).  -->
+<!-- The dependence of the LR $\chi^2$ test statistic on sample size has its downsides [@masyn_latent_2013].  -->
+<!-- Very large samples will often effortlessly reject the null even when the final class solution is a good fit, -->
+<!-- and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic. -->
+<!-- In fact, small samples might not be well approximated using the $\chi^2$ distribution, -->
+<!-- which is a problem shared with large yet sparse datasets. -->
+<!-- In such cases, special caution is advised when interpreting the resulting p-values [@masyn_latent_2013]. -->
+
+### Classification Diagnostics
+
+Best models will divide the sample into subgroups which are internally homogeneous and externally distinct.
+Classification diagnostics give us a way to assess the degree to which this is the case.
+They are separate from model fit indices as a model can fit the data well 
+but show poor latent class separation [@masyn_latent_2013].
+Classification diagnostics should not be used for model selection,
+but they can be used to disqualify certain solutions because they are uninterpretable.
+Interpretability should always be a consideration when considering different class solutions [@nylund-gibson_ten_2018].
+
+Three important classification diagnostics provided by `tidySEM` are are entropy, 
+the range of the posterior classification probabilities by most likely class membership,
+and the range of the sample size.
+
+Entropy is a summary measure of posterior class probabilities across classes and individuals.
+It ranges from 0 (model classification no better than random chance) to 1 (perfect classification).
+As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
+Entropy should not be used to select a particular class solution. 
+An appropriate use of entropy is that it can disqualify certain solutions if class separability is too low 
+or if one of the latent classes is too small to be meaningful or to calculate descriptive statistics.
+
+The range of the posterior classification probabilities by most likely class membership are
+
+Another classification diagnostic measure is the range of the sample size.
+
+<!-- A fundamental concept when examining classification accuracy is the posterior class probability. -->
+<!-- This concept represents the probability of belonging to each latent class computed for each individual. -->
+<!-- The highest posterior class probability is then determined and the individual is assigned to the corresponding class. -->
+<!-- We want each individual's posterior class probabilities to be high for one and low for the remaining latent classes. -->
+<!-- This is considered a high classification accuracy and means that the classes are distinct. -->
+
+<!-- @masyn_latent_2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
+<!-- add content about entropy -->
+
+<!-- DA: Write about prob_min and prob_max (see name formatting), -->
+<!-- the posterior probability from the model, -->
+<!-- it calculates the entropy and the range of the most likely classification probability, -->
+<!-- the diagonal, lets say you have 3 classes, there is the class the person was assigned to, -->
+<!-- and the probability the person should belong to that class, -->
+<!-- what this is telling us is the range of the average classification probabilities, -->
+<!-- what you want to see is that people have a high probability of being in the class they were assigned to, -->
+<!-- and you want that to be high for all the classes,  -->
+<!-- for each class we want a high number -->
+
+<!-- n_min and n_max are the range of the proprotion of people assigned to those classes. -->
+<!-- The smallest class has XX percent of people, and the largest class has XXX percent of people. -->
+<!-- Sometimes the class is too small to be interpreted, e.g. 5% or less of people, that would be too few to estimate parameters.  -->
+
+<!-- The average posterior probability is a measure of classification uncertainty for each latent class. -->
+
+<!-- The modal class assignment proportion -->
+
+### Label switching
+<!-- optional subsection: interpreting class solutions (with class naming) -->
+
+The final class solution will usually discover and enumerate several classes. 
+The class ordering however is completely arbitrary.
+The class labeled as Class 1 in one solution may become Class 2 or Class 3 in another model,
+even when the only difference between the models is in their starting values. 
+Label switching is something to be mindful of when comparing different LCA models [@masyn_latent_2013].
+
+The order of clusters is nondeterministic when using K-means in `tidySEM`. 
+Therefore label switching is still a consideration.
+A simple solution to this is setting a random seed number one line prior to fitting the model.
+We advise `tidySEM` users to always do so in order to circumvent label switching.
+
+<!-- Article discussing label switching: see Chung, Loken & Schafer -->
+
+<!-- Initialization around a Single Mode -->
+<!-- Another common approach is to run a single chain or to initialize the parameters near realistic values.36 -->
+
+<!-- This can work better than the hard constraint approach if reasonable initial values can be found and the labels do not switch within a Markov chain.
+The result is that all chains are glued to a neighborhood of a particular mode in the posterior.
+-->
+<!-- https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/1467-9868.00265?casa_token=Tjk8gR4w8AEAAAAA:RygJL5OgdBaVWY6oaB8XlnyMOnzQgOVDBd2EHDyQD6TH-DzxRxpmU1J4AG8N-HpL5dHHNINRybiaMl6w -->
+
+Class names should be chosen to accurately reflect group membership. 
+Overly simplified and generalized class names may prove misleading to both audiences and researches alike
+leading to what is known as a naming fallacy [@weller_latent_2020].
+
+## Best practices in reporting
+
+Among studies using LCA, reporting practices vary significantly [@weller_latent_2020].
+Various authors have tried to better and standardize ways of reporting LCA [e.g. @masyn_latent_2013; @weller_latent_2020]. 
+Building on their work, we provide guidelines with a strong emphasis on scientific reproducibility and transparency
+better known as the open science framework (OSF). The OSF promotes scientific openness, reproducibility and integrity 
+through an establishment of guidelines for researchers (Foster & Deardorff, 2017)
+To this end, @van_lissa_worcs_2020 developed WORCS, a workflow for open reproducible code in science.
+WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by Nosek and colleagues (2015).
+It can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
+version control, citation and formatting, among others [@van_lissa_worcs_2020].
+
+The guidelines emphasise the use of comprehensive citation (including referencing the software used in the analysis),
+as well as code and data sharing wherever possible.
+@van_lissa_worcs_2020 suggest sharing synthetic data in case the original data cannot be shared, 
+and provide functions to generate such synthetic data.
+Ideally, the entire research project is made reproducible 
+so that others may download it and reproduce it with just one click; for guidance, see @van_lissa_worcs_2020.
+
+As the open science movement is gaining momentum,
+researchers are becoming increasingly aware
+how important it is that analyses can be reproduced and audited.
+In line with open science principles, one of the suggested reporting standards relates to reproducible code.
+In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., *Mplus* and *Latent GOLD*).
+A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because the underlying source code is proprietary.
+To overcome these limitations, the present paper introduces new user-friendly functions in the `tidySEM` R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package `OpenMx`.
+The reporting guidelines described in this paper are adopted in `tidySEM` by default.
+The `tidySEM` R-package thus makes advanced mixture modeling based on best practices widely accessible,
+and facilitates the adoption of the estimation and reporting guidelines described in this paper.
 

### 2022-09-06 example_analysis/Daniel_LCA.Rmd
Update Daniel_LCA.Rmd

--- 2022-08-31 example_analysis/Daniel_LCA.Rmd
@@
 library(tidySEM) 
-df <- data_mix_ordinal
+df <- data_mix_ordinal #ask Caspar to rename the dataset in documentation
 df[1:4] <- lapply(df, ordered)
-mx_lca(data = df,
-       classes = 2) -> res
 ```
 
+# Exploring the data
+
 ```{r}
-res$class1
-res$latentVars
+tidySEM::descriptives(df)
 ```
+
+# Conduct latent class analysis
+
+```{r}
+# fitting the model
+set.seed(123)
+res <- mx_lca(data = df, classes = 1:4)
+```
+# Class enumeration
+
+```{r}
+# printing model fit table
+fit_table <- table_fit(res)
+fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")]
+```
+
+# Interpreting class solution
+
+```{r}
+library(ggplot2)
+plot_prob(res[[3]]) + scale_fill_manual(name = "bla")
+table_prob(res[[3]])
+```
+
+# Extracting posterior class probabilities
+
+```{r}
+probs <- class_prob(res[[3]])
+probs$mostlikely.class
+probs$individual
+```
+
+# APPENDIX
 

### 2022-09-14 example_analysis/Daniel_LCA.Rmd
Update 14 Septemeber
In the manuscript.Rmd:

Added Shapiro-Wilks test when discussing Spurk.
Removed some obsolete comments (we still have backup just in case).
Wrote about alternative model parametrizations in LCA, referring back to GRoLTS.
Under Model Fit Indices, I wrote about -2*LL, mentioned we are working on an implementation of the lazy bootstrap.
Finished the Classification Diagnostics and Interpreting Class Solution sections.

In Daniel_LCA:

Finished annotating the code. We can include this section in the manuscript in the next commit.

@@
 ---
+This is an example of an exploratory LCA using `tidySEM`.
 
-# LOADING THE DATASET
+# Loading the Dataset
+
+In this example, we use *data_mix_ordinal*, 
+a simulated data for mixture modelling with ordinal indicators. 
+This dataset is built-in to `tidySEM`.
+For more information about the dataset, 
+type `?tidySEM::data_mix_ordinal` into the R console
+after loading the `tidySEM` using library().
+
+We load the dataset and convert the indicators into ordered factors.
+
+```{r, include = TRUE}
+library(tidySEM) # loading tidySEM
+df <- data_mix_ordinal  # dataset
+df[1:4] <- lapply(df, ordered) # indicators as ordered factors
+```
+
+# Exploring the Data
+
+An important step to preceding any statistical analysis is data exploration.
+Here we use `tidySEM::descriptives()` 
+to describe the dataset we are using.
 
 ```{r}
-library(tidySEM) 
-df <- data_mix_ordinal #ask Caspar to rename the dataset in documentation
-df[1:4] <- lapply(df, ordered)
+tidySEM::descriptives(df) # descriptives
 ```
+As we can see, the output includes various descriptives of our dataset.
+Special attention should be paid to examining the pattern of missingness,
+as discussed in the *Handling missing data* section of this paper.
+In our example, we see that there are no missing values,
+hence we proceed with our analyses.
 
-# Exploring the data
+# Conducting Latent Class Analysis
+
+Before we fit a series of LCA models, we set a random seed
+using set.seed().
+The seed number can be any digit.
+This is an important step as 
+there is some inherent randomness in the LCA computations,
+and having the same seed number ensures that
+two separate researcher obtain exactly the same results
+when fitting LCA models.
+
+Finally, we reach the step of fitting LCA models.
+To do so, we use `tidySEM::mx_lca()`
+which takes data and number of classes as inputs.
+In our example, we want to fit 1 to 4 class solutions
+and compare their output.
+Depending on your computer's computational power,
+this might take a while.
 
 ```{r}
-tidySEM::descriptives(df)
-```
-
-# Conduct latent class analysis
-
-```{r}
-# fitting the model
-set.seed(123)
-res <- mx_lca(data = df, classes = 1:4)
+set.seed(123) # setting seed 
+res <- mx_lca(data = df, classes = 1:4) # fitting LCA 
 ```
@@
 
+In class enumeration, 
+we want to compare a sequence of LCA models fitted to the data.
+To do so, we use `tidySEM::table_fit()` 
+with the results object as the input.
+This function creates the model fit table formatted for publication.
+However, it contains a lot of information on each of the four models.
+It may be more helpful to focus on a subset of model fit indices
+and classification diagnostics which we do in the subsequent step.
+
 ```{r}
-# printing model fit table
-fit_table <- table_fit(res)
-fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")]
+fit_table <- table_fit(res) # model fit table
+fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # a selection from the model fit table
 ```
+The model fit table is important for deciding on the final class solution.
+Our subset of fit indices and classification diagnostics includes:
 
-# Interpreting class solution
+**Name** indicates the $K$-class solution.
+In our case, we chose to fit 1 - 4 class solutions,
+each is represented by its row in the table.
+**LL** is the -2*log-likelihood of each model.
+**AIC** and **BIC** are the Akaike 
+and the Bayesian Information Criterion, respectively.
+**prob_min** and **prob_max** are 
+the lowest and the highest posterior class probability
+by most likely class membership.
+**n_min** and **n_max** are the lowest and the highest
+class proportion based on the posterior class probabilities.
+
+In this paper, we discussed several possible strategies 
+to select the final class solution.
+
+Here, we apply our own. 
+First, we examine the -2*log likelihood 
+which falls successively with each added class.
+
+Then, we look at the BIC value
+which is fairly close for the one, two, and three-class solutions,
+but high for the four-class solution.
+
+As previously stated, classification diagnostics should not be used for model selection,
+but they can be used to disqualify certain solutions because they are uninterpretable.
+We see that prob_min for the four-class solution is low,
+therefore we disqualify this solution.
+
+Out of the remaining three solutions,
+we notice that entropy is the highest for the three-class solution,
+and it has a satisfactory prob_min and n_min.
+Based on this, we can retain the three class solution in model selection.
+Note that entropy for the one-class solution will always equal to one,
+as it is 100% true that every case is in that class.
+Based on the low entropy of the two-class solution,
+we eliminate this model.
+
+Finally, when comparing the one and three-class solutions,
+we consult the BIC which tells us that the added complexity
+of having three classes still explains the data better
+than a one-class solution.
+Therefore, we select the three-class solution as our final-class solution.
+
+# Interpreting Final Class Solution
+
+To aid our understanding of the final class solution,
+we use `ggplot2::plot_prob()` 
+with the results of the three-class model as the input.
+The resulting graph shows response patterns 
+on all the indicators for each group.
+
+If we want to know the probability of each response option's endoresement 
+for each class, we can use `tidySEM::table_prob`.
+These are thresholds for ordinal dependent variables in the probability scale.
 
@@
 library(ggplot2)
-plot_prob(res[[3]]) + scale_fill_manual(name = "bla")
-table_prob(res[[3]])
+plot_prob(res[[3]]) # plotting the response patterns for the final class solution
+table_prob(res[[3]]) # tabulating the response patterns for the final class solution
 ```
+In the plot, we can see the distributions each of the response probabilities on the indicators
+for each of the three classes.
+For instance, we see in Class 1 the most common response to u2 is 2,
+while in Class 2 and Class 3 this is 0.
+We can also see that response 1 is a rare response not forming the majority in 
+any class.
+Class 2 distinguishes itself because the majority scores the response 0 category of u3 and u4,
+while in Class 1 and 2 this is not the case.
+Class 3 distinguishes itself because the most common response to u3 and u4 is 3.
+
+We can also interpret the response patterns numerically.
 
@@
 
+Another step is to extract posterior class probabilities.
+This is done by the use of `TidySEM::class_prob`
+with the results of the final class solution as the input.
+
 ```{r}
-probs <- class_prob(res[[3]])
-probs$mostlikely.class
-probs$individual
+probs <- class_prob(res[[3]]) # extracting the posterior class probabilities of the final class solution
+probs$mostlikely.class # posterior class probabilities by most likely class membership
+probs$individual # individual posterior class probabilities
 ```
 
-# APPENDIX
+BACKUP TUTORIAL (OLD)
 
-USING DIFFERENT PACKAGES
+This used to be in the tutorial section of the paper.
+I store it here for backup, but I don't think it is longer relevant.
 
-This is an example of a Latent Class Analysis (LCA).
+<!-- 5. Tutorial -->
+<!-- 5.1 ‘mclust’ -->
+<!-- To run LPA using the ‘mclust’ package, the Mclust() function can be used. This function -->
+<!-- requires data as its minimum input. In addition, the number of clusters to fit (G) and model -->
+<!-- variants to fit and select from (modelNames) can be entered (defaults are: G=1-9 and all -->
+<!-- fourteen model variants). In many cases, we may want to consider a more limited range of -->
+<!-- model variants to with different numbers of classes. For multivariate data, a total of fourteen -->
+<!-- options are available (see help("mclustModelNames") for an overview). Each model variant -->
+<!-- is estimated for each value of the specified range of G using Expectation Maximization (EM), -->
+<!-- using the Bayesian Information Criterion (BIC) to compare different model variants and -->
+<!-- select the optimal one. -->
+<!-- Conducting a latent variable model or mixture analysis with Mplus or comparable software -->
+<!-- usually entails fitting models with different numbers of classes (that are otherwise configured -->
+<!-- similarly) and to compare the fit between these models to select the best model. The -->
+<!-- default approach taken in the ‘mclust’ package is slightly different in that the main modeling -->
+<!-- command Mclust() fits multiple models for all possible combinations of the specified number -->
+<!-- of classes and model configurations . The eventual output of ‘mclust’ is the BIC-selected best -->
+<!-- model variant with the optimal combination of number of classes and model configuration. -->
+<!-- An advantage of the ‘mclust’ approach is that it is efficient and flexible: given a number of -->
+<!-- classes, the best model configuration is selected straight away and we do not have to run all -->
+<!-- model combinations one by one. This makes the approach especially suitable for exploratory -->
+<!-- analyses. However, like stated above, this approach is somewhat less usual in some fields, -->
+<!-- where estimation of LPA is approached in a more confirmatory fashion. In addition, it does -->
+<!-- take some control away from the researcher, who may want to primarily focus on selecting -->
+<!-- the number of classes, given a particular , theory-based model configuration that is kept -->
+<!-- constant throughout analyses, which is often the chosen strategy in software like Mplus and -->
+<!-- made easily available through the ‘tidyLPA’ package. -->
+<!-- Here, both approaches are shown: (1) the ‘mclust’ aproach and (2) the ‘tidyLPA’ approach. -->
 
-In Part 1, I use the Lazarsfeld dataset and OpenMX.
-In Part 2, I use the Drinking dataset and OpenMx.
-In Part 3, I redo the analysis from Part 2 using tidySEM.
+<!--                                               5 -->
+<!-- 5.1.1 The ‘mclust’ approach -->
+<!-- In this approach, we run consecutive models with increasing numbers of classes (G=1 to -->
+<!-- G=9), for each model we let the package fit the four above described model variants (“EEI”, -->
+<!-- “EEE”, “VVI” and “VVV”) and select the best-fitting one. We start by loading the package -->
+<!-- (library(mclust)) and creating an object mnames that contains the names of the four models -->
+<!-- we want to be fitted. Next, we use the Mclust() function to fit the models: -->
+<!-- library(mclust) -->
 
-```{r}
-library(tidySEM)
-library(OpenMx)
-```
+<!-- ## Package 'mclust' version 5.4.7 -->
+<!-- ## Type 'citation("mclust")' for citing this R package in publications. -->
+<!-- mnames <- c("EEI", "EEE", "VVI", "VVV") -->
 
-Part 1: LCA using OpenMx and Lazarsfeld Dataset
+<!-- # Fit 1-5 class model -->
+<!-- mod_g1_9 <- Mclust(dat1[, 1:10], modelNames = mnames) -->
 
-Analysis based on: https://rdrr.io/cran/OpenMx/src/inst/models/passing/LCAlazarsfeld.R
+<!-- We can first look at the optimal number of classes and the optimal model variant that were -->
+<!-- selected, using the following code: -->
+<!-- # Opimal number of classes -->
+<!-- mod_g1_9$G -->
 
-```{r}
-data("lazarsfeld", package ="OpenMx", verbose= TRUE)
-data <- lazarsfeld
+<!-- ## [1] 3 -->
+<!-- # Optimal model variant -->
+<!-- mod_g1_9$modelName -->
 
-freq <- data[,5]
-data[,1] <- as.ordered(data[,1])
-data[,2] <- as.ordered(data[,2])
-data[,3] <- as.ordered(data[,3])
-data[,4] <- as.ordered(data[,4])
-vars <- data[,1:4]
+<!-- ## [1] "EEI" -->
+<!-- This shows us that a 3-class EEI model was selected as the best model based on the BIC. We -->
+<!-- can get a better perspective of this model’s performance if we compare it to the other fitted -->
+<!-- models. We can do this by taking a closer look at the other models’ BIC values: -->
+<!-- mod_g1_9$BIC -->
 
-nclass <- 2
-nvar <- 4
-nthresh <- 1
+<!-- ##   Bayesian Information Criterion (BIC): -->
+<!-- ##           EEI       EEE       VVI       VVV -->
+<!-- ##   1 -14661.61 -12986.57 -14661.61 -12986.57 -->
+<!-- ##   2 -13275.65 -12611.60 -13095.82 -12678.12 -->
+<!-- ##   3 -12142.11 -12290.68 -12163.08 -12731.18 -->
+<!-- ##   4 -12178.44 -12330.04 -12241.21 -12970.57 -->
+<!-- ##   5 -12223.14 -12379.96 -12320.47 -13231.35 -->
+<!-- ##   6 -12264.04 -12423.72 -12406.59 -13522.40 -->
+<!-- ##   7 -12269.39 -12451.25 -12497.03 -13748.91 -->
+<!-- ##   8 -12298.30 -12483.35 -12568.62 -13980.82 -->
+<!-- ##   9 -12347.11 -12529.40 -12645.50 -14269.34 -->
+<!-- ## -->
 
-nameList <- names(vars)
+<!--                                              6 -->
+<!-- ## Top 3 models based on the BIC criterion: -->
+<!-- ##     EEI,3     VVI,3     EEI,4 -->
+<!-- ## -12142.11 -12163.08 -12178.44 -->
+<!-- mod_g1_9$loglik -->
 
-class1 <- mxModel("Class1", 
-            mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
-            mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
-            mxMatrix("Full", name = "ThresholdsClass1", nrow = 1, ncol = nvar, 
-            	dimnames = list("Threshold",nameList), free=TRUE),
-		  mxExpectationNormal(covariance="R", means="M", dimnames=nameList, 
-				      thresholds="ThresholdsClass1"),
-		  mxFitFunctionML(vector=TRUE))
+<!-- ## [1] -5951.275 -->
+<!-- mod_g1_9$df -->
 
-class2 <- mxModel("Class2", 
-            mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
-            mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
-            mxMatrix("Full", name = "ThresholdsClass2", nrow = 1, ncol = nvar, 
-            	dimnames = list("Threshold",nameList), free=TRUE),
-		  mxExpectationNormal(covariance="R", means="M", dimnames=nameList,
-				      thresholds="ThresholdsClass2"),
-		  mxFitFunctionML(vector=TRUE))
+<!-- ## [1] 42 -->
+<!-- Here, we can see the BICs for all fitted models (in this case 1-9 classes and 4 model variants: -->
+<!-- 36 models in total). We can see that the closest contenders were a 3-class model with a VVI -->
+<!-- configuration (with variances allowed to vary both within and between classes) and a 4-class -->
+<!-- EEI model. -->
+<!-- Note that the BIC values are negative, and that values that are more negative are considered -->
+<!-- to fit poorer than values closer to 0. This may strike some users as odd, given that in -->
+<!-- many modeling applications, we are used to consider lower BIC values to indicate better -->
+<!-- fit. However, the ‘mclust’ approach of maximizing the BIC is correct for the used modeling -->
+<!-- approach (see e.g., Fraley & Raftery, 2003; Banfield & Raftery, 1993). For theoretical reasons, -->
+<!-- the BIC is calculated in’mclust’ as: -->
 
-# Define the model
+<!--                            BIC = 2(Loglikelihood) - df (log(n)) -->
 
-lcamodel <- mxModel("lcamodel", class1, class2, mxData(vars, type="raw"),
+<!-- where df is the number of parameters (degrees of freedom) and n is the sample size. We can -->
+<!-- see, that here, higher log likelihood values lead to a higher BIC. For the best fitting model in -->
+<!-- our example, the log likelihood is -5951.275, the number of parameters is 42 and the sample -->
+<!-- size is 300. If we plug these into the formula, we get the following BIC: -->
 
-# Create class membership probabilities, constrain them to be in the range 0-1, and make their sum equal 1.0
-            mxMatrix("Full", name = "ClassMembershipProbabilities", nrow = nclass, ncol = 1, free=TRUE, 
-			        labels = c(paste("pclass", 1:nclass, sep=""))),
-            mxBounds(c(paste("pclass", 1:nclass, sep="")),0,1),
-		    mxMatrix("Iden", nrow = 1, name = "constraintLHS"),
-		    mxAlgebra(sum(ClassMembershipProbabilities), name = "constraintRHS"),
-            mxConstraint(constraintLHS == constraintRHS),
+<!--                      BIC = 2(-5951.275) - 42(log(300)) = -12142.11 -->
 
-# Define the objective function
-            mxAlgebra(-2*sum(freq * log(pclass1%x%Class1.objective + pclass2%x%Class2.objective)), 
-            	name="lca"),
-            mxFitFunctionAlgebra("lca")
-)
+<!-- As expected, the obtained value indeed corresponds to the BIC we got in the model output -->
+<!-- (see above). Other software packages, such as Mplus and ‘tidyLPA’ (see below) use a slightly -->
+<!-- different formula to calculate the BIC, where higher log likelihood values lead to lower BIC -->
+<!-- values: -->
 
-# Run the job
-model <- mxRun(lcamodel, suppressWarnings=TRUE)
-summary(model)
-model$matrices
+<!--                            BIC = df (log(n)) - 2(Loglikelihood) -->
 
-# Check results against those hard-coded from old Mx:
-negativeThresholds <- as.vector(mxEval(Class1.ThresholdsClass1, model))
-positiveThresholds <- as.vector(mxEval(Class2.ThresholdsClass2, model))
-if (negativeThresholds[[1]] > 0) {
-	temp <- negativeThresholds
-	negativeThresholds <- positiveThresholds
-	positiveThresholds <- temp
-}
+<!-- When we plug the best-model values into this formula, we get: -->
 
-omxCheckCloseEnough(negativeThresholds, c(-1.3247, -0.2909, -0.1466, -0.0044),.01)
-omxCheckCloseEnough(positiveThresholds, c(0.1805, 0.9071, 1.3169, 1.5869),.01)
-omxCheckCloseEnough(sort(as.vector(mxEval(ClassMembershipProbabilities, model))), c(0.4440, 0.5560),.01)
-omxCheckCloseEnough(model$output$Minus2LogLikelihood, 4696.444, 0.01)
-```
+<!--                       BIC = 42(log(300)) - 2(-5951.275) = 12142.11 -->
 
-Part 2: LCA Using OpenMx and Drinking Dataset,
+<!-- We can see that the absolute BIC value is the same, irrespective of the used formula, with -->
+<!-- only the sign differing between the two approaches. -->
 
-Dataset from: https://stats.oarc.ucla.edu/sas/dae/latent-class-analysis/
+<!--                                               7 -->
+<!-- Now, let us continue by taking a closer look at the class sizes and the mean values and -->
+<!-- variances of the ten variables for the three identified classes. Because the best model has and -->
+<!-- EEI configuration, printing out a single variance matrix from one class will suffice, because -->
+<!-- the variances are the same across classes (this is different in cases where the optimal model -->
+<!-- allows for between-class differences in (co)variances: -->
+<!-- # tabulate class-membership numbers -->
+<!-- table(summary(mod_g1_9)$classification) -->
 
-```{r}
-data <- read.csv('C:/Users/danie/OneDrive/Desktop/LCA/drinking_data.csv', header = T, sep = ',')
+<!-- ## -->
+<!-- ##   1   2      3 -->
+<!-- ## 100 110     90 -->
+<!-- # display the means per class -->
+<!-- mod_g1_9$parameters$mean -->
 
-#check if two rows have the same response pattern
-library(DAKS)
-pattern(data)
+<!-- ##             [,1]         [,2]        [,3] -->
+<!-- ##   var1 2.023572      6.083019   2.0304980 -->
+<!-- ##   var2 3.056837      6.937785   4.8997993 -->
+<!-- ##   var3 1.718166      1.107873   7.9385872 -->
+<!-- ##   var4 6.086727      2.854402   6.9657536 -->
+<!-- ##   var5 7.101109      2.225675   4.7416266 -->
+<!-- ##   var6 3.691936      8.081570   2.2469977 -->
+<!-- ##   var7 5.009819      2.817889   0.8862725 -->
+<!-- ##   var8 8.316894      8.080891   3.9350536 -->
+<!-- ##   var9 1.978136      1.896294   9.1537131 -->
+<!-- ##   var10 1.191249     5.142590   8.0332628 -->
+<!-- # select the first of 3 10-by-10 variance matrices -->
+<!-- mod_g1_9$parameters$variance$sigma[1:10, 1:10, 1] -->
 
-dat1 <- c(1,0,0,0,0,0,0,0,0, 126)
-dat2 <- c(0,0,0,0,0,0,0,0,0, 102)
-dat3 <- c(1,0,0,0,0,0,0,0,1, 47)
-dat4 <- c(1,1,0,0,0,0,0,0,0, 47)
-dat5 <- c(1,0,0,0,0,1,0,0,0, 43)
-dat6 <- c(126, 102, 47, 47, 43)
+<!-- ##               var1       var2       var3       var4       var5       var6       var7       var8 -->
+<!-- ##   var1    2.539537   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
+<!-- ##   var2    0.000000   2.117941   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
+<!-- ##   var3    0.000000   0.000000   2.509035   0.000000   0.000000   0.000000   0.000000   0.000000 -->
+<!-- ##   var4    0.000000   0.000000   0.000000   2.812796   0.000000   0.000000   0.000000   0.000000 -->
+<!-- ##   var5    0.000000   0.000000   0.000000   0.000000   2.647399   0.000000   0.000000   0.000000 -->
+<!-- ##   var6    0.000000   0.000000   0.000000   0.000000   0.000000   2.694156   0.000000   0.000000 -->
+<!-- ##   var7    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.368379   0.000000 -->
+<!-- ##   var8    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.540704 -->
+<!-- ##   var9    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
+<!-- ##   var10   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
+<!-- ##               var9      var10 -->
+<!-- ##   var1    0.000000   0.000000 -->
+<!-- ##   var2    0.000000   0.000000 -->
+<!-- ##   var3    0.000000   0.000000 -->
+<!-- ##   var4    0.000000   0.000000 -->
 
-dat <- data.frame(dat1,dat2,dat3,dat4,dat5)
-dat <- t(dat)
-dat <- as.data.frame(dat)
-names(dat) <- c("Q1", "Q2", "Q3", "Q4", "Q5", "Q6", "Q7", "Q8", "Q9", "freq")
-rm("data", "dat1", "dat2", "dat3", "dat4", "dat5", "dat6")
+<!--                                                8 -->
+<!-- ##   var5    0.000000   0.000000 -->
+<!-- ##   var6    0.000000   0.000000 -->
+<!-- ##   var7    0.000000   0.000000 -->
+<!-- ##   var8    0.000000   0.000000 -->
+<!-- ##   var9    2.509146   0.000000 -->
+<!-- ##   var10   0.000000   2.204457 -->
+<!-- From this output, we can see that the three classes have class-sizes of 100, 110 and 90, -->
+<!-- respectively. These sizes along with the patterns of class-specific mean variable scores -->
+<!-- correspond with those that served as the input for the simulation. As specified, the variances -->
+<!-- vary slightly across variables. -->
+<!-- The class allocation in LPA is probabilistic in nature: each subject in the data is assigned a -->
+<!-- probability for each of the estimated classes, based on their pattern of scores on the input -->
+<!-- variables. These probabilities can be inspected in the z-matrix (here: mod_g1_9$z). Subjects -->
+<!-- can be allocated to one of the classes based on their highest class-probability. These posterior -->
+<!-- allocations can be found in the classification matrix (here: mod_g1_9$classification). -->
+<!-- These class-allocations were tabulated above to evaluate class sizes. -->
+<!-- Note that class-allocation in this way only yields useful classifications if the patterns of -->
+<!-- class-probabilities allow for allocation of each subject to a single class with sufficient certainty. -->
+<!-- In case of too much uncertainty, using the classification is not advised. We can inspect the -->
+<!-- uncertainty of allocation for all subjects (here: mod_g1_9$uncertainty). This gives us a list -->
+<!-- of uncertainties for all subjects. Next, we can evaluate the extent of uncertainty by looking -->
+<!-- at the maximum uncertainty or, for instance, the averaged uncertainty across subjects: -->
+<!-- max(mod_g1_9$uncertainty) -->
 
-dat[,1] <- as.ordered(dat[,1])
-dat[,2] <- as.ordered(dat[,2])
-dat[,3] <- as.ordered(dat[,3])
-dat[,4] <- as.ordered(dat[,4])
-dat[,5] <- as.ordered(dat[,5])
-dat[,6] <- as.ordered(dat[,6])
-dat[,7] <- as.ordered(dat[,7])
-dat[,8] <- as.ordered(dat[,8])
-dat[,9] <- as.ordered(dat[,9])
+<!-- ## [1] 0.003852604 -->
+<!-- mean(mod_g1_9$uncertainty) -->
 
-freq <- dat[,10]
-vars <- dat[,1:9]
+<!-- ## [1] 2.607299e-05 -->
+<!-- Here, the uncertainty is atypically low because of the used idealized data. In many cases, -->
+<!-- uncertainty is higher and it may be of interest to investigate it further, for instance, by -->
+<!-- looking at (e.g., plotting or tabulating) the uncertainty per class: -->
+<!-- cprob <- cbind(mod_g1_9$z, mod_g1_9$classification) -->
+<!-- cprob <- as.data.frame(cprob) -->
+<!-- colnames(cprob) <- c("prob (class 1)", "prob (class 2)", "prob (class 3)", "class") -->
+<!-- aggregate(cprob[, 1:3], list(cprob$class), mean) -->
 
-nclass <- 2
-nvar <- 9
-nthresh <- 1
-nameList <- names(vars)
+<!-- ##   Group.1 prob (class 1) prob (class 2) prob (class 3) -->
+<!-- ## 1       1   9.999824e-01   1.761611e-05   3.350070e-10 -->
+<!-- ## 2       2   5.486300e-05   9.999451e-01   7.192849e-08 -->
+<!-- ## 3       3   2.591358e-08   1.675400e-07   9.999998e-01 -->
+<!-- Here we can see that for each class, the mean probability of membership of that class was -->
 
-class1 <- mxModel("Class1", 
-                  mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
-                  mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
-                  mxMatrix("Full", name = "ThresholdsClass1", nrow = 1, ncol = nvar, 
-                           dimnames = list("Threshold",nameList), free=TRUE),
-                  mxExpectationNormal(covariance="R", means="M", dimnames=nameList, 
-                                      thresholds="ThresholdsClass1"),
-                  mxFitFunctionML(vector=TRUE))
+<!--                                                  9 -->
+<!-- very high and probabilities for the other classes were very low. Again, these values reflect the -->
+<!-- idealized data; in many cases, more differentiated class-probability patterns may be observed. -->
 
-class2 <- mxModel("Class2", 
-                  mxMatrix("Iden", name = "R", nrow = nvar, ncol = nvar, free=FALSE),
-                  mxMatrix("Full", name = "M", nrow = 1, ncol = nvar, free=FALSE),
-                  mxMatrix("Full", name = "ThresholdsClass2", nrow = 1, ncol = nvar, 
-                           dimnames = list("Threshold",nameList), free=TRUE),
-                  mxExpectationNormal(covariance="R", means="M", dimnames=nameList,
-                                      thresholds="ThresholdsClass2"),
-                  mxFitFunctionML(vector=TRUE))
+<!-- 5.1.2 ‘tidyLPA’ -->
+<!-- When running ‘mclust’ models from the ‘tidyLPA’ package, the estimations are approached -->
+<!-- a little differently. By default, the number of models that can be estimated is restricted to -->
+<!-- six relatively common variants (see help(tidyLPA::estimate_profiles) for details). Of -->
+<!-- these models, four can be estimated with ‘mclust’. These models differ in terms of how the -->
+<!-- variances and covariances are allowed to vary or constrained to be equal across classes and -->
+<!-- whether covariances are or are not fixed to zero within classes. These models have each been -->
+<!-- allocated a number (1-6) in ‘tidyLPA’. See the table below how these numbers correspond to -->
+<!-- the above mentioned model configurations. -->
+<!--  Model variant    model number -->
+<!--  EEI              1 -->
+<!--  EEE              3 -->
+<!--  VVI              2 -->
+<!--  VVV              6 -->
+<!-- Models four and five are different and can only be fit when R is interfacing with Mplus using -->
+<!-- the ‘MplusAutomation’ package. This is outside the scope of this tutorial. -->
+<!-- To fit an LPA model in ‘tidyLPA’, we use the estimate_profiles() function. Here, we need -->
+<!-- to enter the data.frame to be used (df) and the number of classes to estimate (n_profiles). -->
+<!-- To determine what kind of model configuration to estimate, the authors have provided two -->
+<!-- different ways in the estimate_profiles() command. In the first approach, we simply tell what -->
+<!-- model number (see Table 2) we want to estimate using the models= argument For instance, if -->
+<!-- we want to estimate LPA models with 1 to 9 classes, with an EEI configuration, we can use: -->
+<!-- <!-- suppressMessages(library(tidyLPA)) --> -->
+<!-- <!-- suppressMessages(mod_1c_v1 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, --> -->
+<!-- <!--     models = 1)) --> -->
 
-# Define the model
+<!-- The package issues a standard message by default notifying us that we use the models -->
+<!-- argument and that the variances and covariances arguments are ignored. Here, we suppress -->
+<!-- the message. The output shows us the model-fit information for the 1 to 9-class EEI models: -->
+<!-- <!-- mod_1c_v1 --> -->
 
-lcamodel <- mxModel(model = "lcamodel", class1, class2, mxData(vars, type="raw"))
+<!-- ## tidyLPA analysis using mclust: -->
+<!-- ## -->
+<!-- ## Model Classes AIC       BIC      Entropy            prob_min   prob_max   n_min   n_max   BLRT_p -->
+<!-- ## 1      1       14587.53 14661.61 1.00               1.00       1.00       1.00    1.00 -->
+<!-- ## 1      2       13160.84 13275.65 1.00               1.00       1.00       0.30    0.70    0.01 -->
+<!-- ## 1      3       11986.55 12142.11 1.00               1.00       1.00       0.30    0.37    0.01 -->
+<!-- ## 1      4       11982.14 12178.44 0.91               0.79       1.00       0.14    0.33    0.05 -->
+<!-- ## 1      5       11986.09 12223.14 0.89               0.75       1.00       0.06    0.33    0.30 -->
 
-# Create class membership probabilities, constrain them to be in the range 0-1, and make their sum equal 1.0
-mxMatrix("Full", name = "ClassMembershipProbabilities", nrow = nclass, ncol = 1, free=TRUE, labels = c(paste("pclass", 1:nclass, sep=""))) 
-mxBounds(c(paste("pclass", 1:nclass, sep="")),0,1) 
-mxMatrix("Iden", nrow = 1, name = "constraintLHS") 
-mxAlgebra(sum(ClassMembershipProbabilities), name = "constraintRHS") 
-mxConstraint(constraintLHS == constraintRHS)
+<!--                                               10 -->
+<!-- ##   1      6         11986.25    12264.04   0.84      0.78       1.00        0.08   0.30    0.16 -->
+<!-- ##   1      7         11950.86    12269.39   0.81      0.77       0.91        0.08   0.17    0.01 -->
+<!-- ##   1      8         11939.04    12298.30   0.82      0.79       0.92        0.08   0.17    0.02 -->
+<!-- ##   1      9         11947.10    12347.11   0.82      0.76       0.92        0.06   0.16    0.57 -->
+<!-- Here, we can see that the BIC takes on different values compared to ‘mclust’, with lower -->
+<!-- rather than higher values indicating best fit (see explanation above). Instead of drawing the -->
+<!-- fit indices (e.g., BIC, AIC) directly from the ‘mclust’ package, ‘tidyLPA’ only draws the ‘raw’ -->
+<!-- log likelihood values and posterior probabilities from the ‘mclust’ output and (re)calculates -->
+<!-- the BIC, AIC, entropy etc. to mirror as well as possible those provided in Mplus. In addition, -->
+<!-- p-values of the the bootstrapped likelihood ratio test (BLRT) are given by default. These -->
+<!-- p-values indicate for each k-class model whether adding the kth class adds significantly to -->
+<!-- model fit. For instance, the BLRT_p of 0.04 for the 4-class model indicates that adding -->
+<!-- a fourth class led to an improvement of model fit compared to the 3-class model that was -->
+<!-- just significant when using an alpha of 0.05. When using ’mclust’ as a stand-alone package, -->
+<!-- the BLRT is not calculated by default, but can be obtained with the mclustBootstrapLRT -->
+<!-- function. -->
+<!-- Next, we can run similar commands, with different values for the models argument. It is -->
+<!-- also possible to estimate more than one model variant in a single run by providing a vector -->
+<!-- of model numbers (e.g., models=c(1,6)). -->
+<!-- The second approach to determine the model variant to be estimated is by using the variances -->
+<!-- and covariances arguments in the estimate_profiles command. The variances argument -->
+<!-- can have the values "equal" (i.e. variable variances constrained to be equal across classes) or -->
+<!-- "varying" (i.e. variances allowed to vary across classes). The covariances argument can -->
+<!-- have the values: "zero" (i.e. all variable covariances fixed to zero), "equal" (i.e. covariances -->
+<!-- constrained to be equal across classes) or "varying" (i.e. covariances allowed to vary across -->
+<!-- classes). Now, if we again want to estimate LPA models with 1 to 9 classes, with an EEI -->
+<!-- configuration, we can use: -->
+<!-- <!-- mod_1c_v2 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, variances = "equal", --> -->
+<!-- <!--     covariances = "zero") --> -->
+<!-- <!-- mod_1c_v2$model_1_class_2$fit --> -->
 
-# Define the objective function
-mxAlgebra(-2*sum(freq * log(pclass1%x%Class1.objective + pclass2%x%Class2.objective)), name="lca")
+<!-- <!-- ##         Model           Classes        LogLik                   AIC              AWE --> -->
+<!-- <!-- ## 1.000000e+00       2.000000e+00 -6.549418e+03          1.316084e+04     1.354347e+04 --> -->
+<!-- <!-- ##           BIC              CAIC           CLC                   KIC            SABIC --> -->
+<!-- <!-- ## 1.327565e+04       1.330665e+04 1.310084e+04           1.319484e+04     1.317734e+04 --> -->
+<!-- <!-- ##           ICL           Entropy      prob_min              prob_max            n_min --> -->
+<!-- <!-- ## -1.327568e+04      9.996750e-01 9.999484e-01           9.999990e-01     3.000000e-01 --> -->
+<!-- <!-- ##         n_max          BLRT_val        BLRT_p --> -->
+<!-- <!-- ## 7.000000e-01       1.448699e+03 9.900990e-03 --> -->
+<!-- As expected, the output is exactly the same as the one we got when using the models -->
+<!-- argument. Now, if we want to, we can continue fitting other model variants using either -->
+<!-- approach. Next, we can manually compare the models in terms of their BIC and AIC values -->
 
-mxFitFunctionAlgebra("lca")
+<!--                                               11 -->
+<!-- or we can use compare_solutions to do this in an automated fashion. Interestingly, the -->
+<!-- best model can be selected based on the integrated information about several fit indices -->
+<!-- (analytic hierarchy process; see help(tidyLPA::AHP)) for more details). We obtain the model -->
+<!-- comparison with the following code. -->
+<!-- <!-- comp <- suppressWarnings(compare_solutions(mod_1c_v1)) --> -->
+<!-- <!-- comp$fits --> -->
 
-# Run the job
-model <- mxRun(lcamodel, suppressWarnings=TRUE)
-summary(model)
-model$matrices
+<!-- <!-- ##   # A tibble: 9 x 18 --> -->
+<!-- <!-- ##     Model Classes LogLik    AIC    AWE    BIC   CAIC    CLC    KIC SABIC      ICL --> -->
+<!-- <!-- ##     <dbl>   <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>           <dbl> --> -->
+<!-- <!-- ##   1     1       1 -7274. 14588. 14834. 14662. 14682. 14550. 14611. 14598. -14662. --> -->
+<!-- <!-- ##   2     1       2 -6549. 13161. 13543. 13276. 13307. 13101. 13195. 13177. -13276. --> -->
+<!-- <!-- ##   3     1       3 -5951. 11987. 12506. 12142. 12184. 11905. 12032. 12009. -12142. --> -->
+<!-- <!-- ##   4     1       4 -5938. 11982. 12638. 12178. 12231. 11878. 12038. 12010. -12218. --> -->
+<!-- <!-- ##   5     1       5 -5929. 11986. 12778. 12223. 12287. 11860. 12053. 12020. -12273. --> -->
+<!-- <!-- ##   6     1       6 -5918. 11986. 12915. 12264. 12339. 11838. 12064. 12026. -12353. --> -->
+<!-- <!-- ##   7     1       7 -5889. 11951. 13016. 12269. 12355. 11780. 12040. 11997. -12376. --> -->
+<!-- <!-- ##   8     1       8 -5873. 11939. 13141. 12298. 12395. 11747. 12039. 11991. -12409. --> -->
+<!-- <!-- ##   9     1       9 -5866. 11947. 13285. 12347. 12455. 11733. 12058. 12005. -12458. --> -->
+<!-- <!-- ##   # ... with 7 more variables: Entropy <dbl>, prob_min <dbl>, prob_max <dbl>, --> -->
+<!-- <!-- ##   #   n_min <dbl>, n_max <dbl>, BLRT_val <dbl>, BLRT_p <dbl> --> -->
+<!-- <!-- comp$best --> -->
 
-# Check results against those hard-coded from old Mx:
-negativeThresholds <- as.vector(mxEval(Class1.ThresholdsClass1, model))
-positiveThresholds <- as.vector(mxEval(Class2.ThresholdsClass2, model))
-if (negativeThresholds[[1]] > 0) {
-  temp <- negativeThresholds
-  negativeThresholds <- positiveThresholds
-  positiveThresholds <- temp
-}
+<!-- <!-- ## LogLik       AIC      AWE      BIC     CAIC        CLC    KIC   SABIC       ICL --> -->
+<!-- <!-- ##      9         8        3        3        3          9      3       8         3 --> -->
+<!-- As shown in the ‘mclust’ examples above, the EEI model (models=1) with 3 classes fit the -->
+<!-- data best, not only according to the BIC, but according to the integrated information from a -->
+<!-- broader range of fit indices. Note that we did not fit any competing EEE, VVV and VVI -->
+<!-- models here. If we would have wanted to compare fit to these model variants too, we should -->
+<!-- have actively told the package to fit these too, either by specifying a vector of different types -->
+<!-- of models to fit in the models argument or by repeating the analysis with different values for -->
+<!-- the variances and covariances arguments. -->
 
-```
+<!-- 5.2 Comparison to Mplus -->
+<!-- For comparison, we fit the equivalent EEI model variants in Mplus. For this, we can mostly -->
+<!-- rely on the defaults of Mplus. We only need to specify that we want to estimate class-specific -->
+<!-- variable means. The variances are automatically constrained to be equal across classes and -->
+<!-- within class variable covariances are fixed to 0. Here we have the separate codes for fitting -->
+<!-- the first 4 models wit 1-4 class models (code for the 5-9 class models not shown). -->
+<!-- <!-- !! 1-class model --> -->
+<!-- <!--   DATA: file='dat1.dat'; --> -->
+<!-- <!--   VARIABLE: --> -->
 
-Part 3: LCA using tidySEM
+<!-- <!--                                                  12 --> -->
+<!-- <!--     names=id v1-v10; --> -->
+<!-- <!--     usevariables= v1-v10; --> -->
+<!-- <!--     classes = c(1); --> -->
+<!-- <!--   ANALYSIS: --> -->
+<!-- <!--     type = mixture; --> -->
+<!-- <!--     starts= 250 50; --> -->
+<!-- <!--   MODEL: --> -->
+<!-- <!--   %overall% --> -->
+<!-- <!--   %C#1% --> -->
+<!-- <!--   [v1-v10]; --> -->
 
-```{r}
-#adjust this code
-# The argument `classes` indicates which class solutions should be estimated (e.g., 1 through 6).
-# The argument `variances` specifies whether variances should be `"equal"` or `"varying"` across classes.
-# The argument `covariances` specifies whether covariances should be constrained to `"zero"`, `"equal"` or `"varying"` across classes.
-tidySEM::mx_profiles(classes, variances, covariances)
-```
+<!-- <!-- !! 2-class model --> -->
+<!-- <!--   DATA: file='dat1.dat'; --> -->
+<!-- <!--   VARIABLE: --> -->
+<!-- <!--     names=id v1-v10; --> -->
+<!-- <!--     usevariables= v1-v10; --> -->
+<!-- <!--     classes = c(2); --> -->
+<!-- <!--   ANALYSIS: --> -->
+<!-- <!--     type = mixture; --> -->
+<!-- <!--     starts= 250 50; --> -->
+<!-- <!--   MODEL: --> -->
+<!-- <!--   %overall% --> -->
+<!-- <!--   %C#1% --> -->
+<!-- <!--   [v1-v10]; --> -->
+<!-- <!--   %C#2% --> -->
+<!-- <!--   [v1-v10]; --> -->
 
-After all models have been estimated, the function `tidySEM::table_fit()` can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
-Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
-because this test is very computationally expensive.
-It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
+<!-- <!-- !! 3-class model --> -->
+<!-- <!--   DATA: file='dat1.dat'; --> -->
+<!-- <!--   VARIABLE: --> -->
+<!-- <!--     names=id v1-v10; --> -->
+<!-- <!--     usevariables= v1-v10; --> -->
+<!-- <!--     classes = c(3); --> -->
+<!-- <!--   ANALYSIS: --> -->
+<!-- <!--     type = mixture; --> -->
+<!-- <!--     starts= 250 50; --> -->
+<!-- <!--   MODEL: --> -->
+<!-- <!--   %overall% --> -->
+<!-- <!--   %C#1% --> -->
+<!-- <!--   [v1-v10]; --> -->
+<!-- <!--   %C#2% --> -->
+<!-- <!--   [v1-v10]; --> -->
+<!-- <!--   %C#3% --> -->
+<!-- <!--   [v1-v10]; --> -->
+
+<!-- <!--                             13 --> -->
+<!-- <!-- !! 3-class model --> -->
+<!-- <!--   DATA: file='dat1.dat'; --> -->
+<!-- <!--   VARIABLE: --> -->
+<!-- <!--     names=id v1-v10; --> -->
+<!-- <!--     usevariables= v1-v10; --> -->
+<!-- <!--     classes = c(4); --> -->
+<!-- <!--   ANALYSIS: --> -->
+<!-- <!--     type = mixture; --> -->
+<!-- <!--     starts= 250 50; --> -->
+<!-- <!--   MODEL: --> -->
+<!-- <!--   %overall% --> -->
+<!-- <!--   %C#1% --> -->
+<!-- <!--   [v1-v10]; --> -->
+<!-- <!--   %C#2% --> -->
+<!-- <!--   [v1-v10]; --> -->
+<!-- <!--   %C#3% --> -->
+<!-- <!--   [v1-v10]; --> -->
+<!-- <!--   %C#4% --> -->
+<!-- <!--   [v1-v10]; --> -->
+<!-- Each of the models was fitted with multiple initial stage random starts and multiple final -->
+<!-- stage optimizations to reduce the risk of finding solutions on a local maximum. For the 1-6 -->
+<!-- class models, 250 iniitial and 50 final starts were used. For the 7-class model 1000 initial -->
+<!-- starts and 200 final optimizations were used and for the 8- and 9-class models, a replicable -->
+<!-- global solution could only be obtained with 5000 initial stage starts and 1000 final stage -->
+<!-- optimizations. The fit indices for each of the fitted models is displayed in the table below. -->
+<!-- For now, we only look at the estimated fit indices. -->
+<!--                 EEI models -->
+<!--  Classes   AIC       BIC -->
+<!--  1         14587.535 14587.535 -->
+<!--  2         13160.836 13275.653 -->
+<!--  3         11986.551 12142.11 -->
+<!--  4         11957.71  12154.01 -->
+<!--  5         11951.766 12188.808 -->
+<!--  6         11944.632 12222.415 -->
+<!--  7         11935.501 12254.027 -->
+<!--  8         11929.428 12288.695 -->
+<!--  9         11925.203 12325.212 -->
+<!-- If we compare the BIC values for the EEI models to those obtained with ‘mclust’ directly -->
+<!-- and via ‘tidyLPA’, we can see that the results are quite similar for the less complex models, -->
+<!-- with the BIC-values for the 2- and 3-class models being exactly the same in terms of their -->
+<!-- absolute values. When taking a closer look at the 3-class model that was previously found to -->
+<!-- be the best model, we can also see that that the Mplus-based classifications (n=100, n=110, -->
+<!-- and n=90) and entropy estimates (entropy=1.0) are similar to those obtained with ‘mclust’ -->
+
+<!--                                              14 -->
+<!-- and/or ‘tidyLPA’. -->
+<!--  If we look at the larger picture and evaluate the overlap between the BIC values obtained -->
+<!--  for all models with class numbers ranging from 1 to 9, we can see that the estimated BIC -->
+<!-- values show more differences between the R-packages and Mplus for the models with larger -->
+<!--  numbers of classes. This can be explained by the different methods used by Mplus and -->
+<!-- ‘mclust’/’tidyLPA’ to generate the start values for estimation (see below). These differences -->
+<!--  can cause the packages to yield different results for more complex models and/or models that -->
+<!--  are increasingly misspecified (as the models with more than 3 classes were in our example). -->
+
+<!-- 6. Final comments -->
+<!-- In this tutorial we have seen that it is relatively easy to run an LPA in R. As with all -->
+<!-- data-driven analyses, care should be taken not to over-interpret the results. In the end, LPA -->
+<!-- identifies classes in such a way as to optimally explain variance on a range of variables and -->
+<!-- not to optimize interpretability or usefulness. -->
+<!-- Another remark with regard to LPA results is that class-membership is probabilistic in nature: -->
+<!-- each subject in an analyzed sample has a probability of being in each of the model’s classes. -->
+<!-- Subjects can be allocated to a class based on their highest class-probability. Importantly, -->
+<!-- this can only be done with enough certainty if separation between classes is sufficient, which -->
+<!-- means that we see that each subject has a clear highest probability (e.g., p=0.9) for one of -->
+<!-- the classes. The entropy statistic is often used to quantify this separation, with values <0.8 -->
+<!-- being taken to indicate insufficient separation to allocate subjects to one class. In such cases, -->
+<!-- the class probabilities themselves can still be used in further investigations. -->
+<!-- Depending on one’s preferences, one can choose either the ‘mclust’ or the ‘tidyLPA’ package. -->
+<!-- The latter may be especially attractive for persons that already have experience with Mplus -->
+<!-- or comparable software. It is important to note, though, that differences exist between the -->
+<!-- estimation approaches taken by both R-packages and by Mplus. -->
+<!-- Mixture models such as LPA need to be estimated in an iterative fashion, using an EM -->
+<!-- algorithm that is very sensitive to the used start values, where poor start values can lead the -->
+<!-- estimation process to a poor solution. In addition, there is often a chance that the iterative -->
+<!-- EM process leads to a solution that is at a local, rather than a global maximum in the -->
+<!-- likelihood ‘landscape’. -->
+
+<!-- ### Starting values -->
+<!-- Different methods have been developed to initialize the estimation (get starting values) in -->
+<!-- such a way as to optimize the chance of arriving at an accurate model solution (Biernacki et -->
+<!-- al., 2003; Shireman et al., 2016). Mplus and ‘mclust’ take two different approaches. Mplus -->
+<!-- uses a ‘brute force’ approach and reruns the model with multiple sets of random start values, -->
+<!-- each generated from uniform distributions of values with ranges that are based on the data -->
+<!-- (Muthén & Muthén, 1998-2015). The best solution is the model with the highest log likelihood -->
+<!-- that was arrived upon from at least two different starting points. In contrast, ‘mclust’ uses -->
+<!-- the data itself to generate a set of plausible start values using hierarchical clustering. This -->
+<!-- means that the start values are informed by the (hierarchical structure of) the data, which -->
+<!-- can work well. A downside is that only a single set of start values is used, so the risk of -->
+<!-- arriving at a local optimum is not addressed. These approaches show clear differences and -->
+<!-- have been shown previously to not always arrive at the same solutions, with some authors -->
+<!-- being especially critical of the ‘mclust’ approach (Shireman et al., 2017). In our example, -->
+<!-- we did not see much difference between the two approaches, but it should be noted that -->
+<!-- many real-world data sets do not have such a clear latent structure, making it much more -->
+<!-- challenging to arrive at an accurate solution, given the data. -->
+<!-- Based on the provided examples and these technical differences between ‘mclust’ and Mplus, -->
+<!-- it is probably better to see ‘mclust’ and ‘tidyLPA’ as useful alternatives for Mplus, rather -->
+<!-- than as tools to exactly mimic what can be done with the latter software. -->
+
+<!-- References -->
+<!-- Banfield, J., Raftery, A.E. (1993). Model-based Gaussian and non-Gaussian clustering. -->
+<!-- Biometrics. 49: 803–821. -->
+<!-- Biernacki,C.,Celeux,G., Govaert,G.(2003).Choosing starting values for the EM algorithm -->
+<!-- for getting the highest likelihood in multivariate Gaussian mixture models. Computational -->
+<!-- Studies and Data Analysis, 41, 561–575. -->
+<!-- Celeux, G., Govaert, G. (1995) Gaussian parsimonious clustering models. Pattern Recognition. -->
+<!-- 28:781–793 -->
+<!-- Fraley, C., Raftery, A.E. (1998) How many clusters? Which clustering method? Answers via -->
+<!-- model-based cluster analysis. The Computer Journal 41: 578-588. -->
+<!-- Gibson, W. A. (1959). Three multivariate models: Factor analysis, latent structure analysis, -->
+<!-- and latent profile analysis. Psychometrika, 24, 229–252. -->
+<!-- Linzer, D.A., Lewis J. (2011). “poLCA: an R Package for Polytomous Variable Latent Class -->
+<!-- Analysis.” Journal of Statistical Software. 42(10): 1-29 -->
+<!-- Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., Leisch, F. (2019). e1071: Misc -->
+<!-- Functions of the Department of Statistics, Probability Theory Group (Formerly: E1071), TU -->
+<!-- Wien. R package version 1.7-3. https://CRAN.R-project.org/package=e1071 -->
+<!-- Muthén, L., Muthén, B., 1998–2015. Mplus User’s Guide, 7th ed., Muthén & Muthén, Los -->
+<!-- Angeles. -->
+<!-- Neale, M.C., Hunter, M.D., Pritikin, J.N., Zahery, M., Brick, T.R., Kirkpatrick, R.M., -->
+<!-- Estabrook, R., Bates, T.C., Maes, H.H., Boker, S.M. (2016). “OpenMx 2.0: Extended -->
+<!-- structural equation and statistical modeling.” Psychometrika 81(2): 535-549. -->
+<!-- Nylund-Gibson, K., Choi, A. Y. (2018). Ten frequently asked questions about latent class -->
+<!-- analysis. Translational Issues in Psychological Sciences, 4, 440–461. -->
+<!-- Oberski, D. (2016). Mixture models: Latent profile and latent class analysis. In J. Robertson -->
+<!-- & M. Kaptein (Eds.), Modern statistical methods for HCI (pp. 275–287). Cham, Switzerland: -->
+<!-- Springer Inter-national Publishing. -->
+
+<!--                                              16 -->
+<!-- Pastor, D.A., Barron, K.E., Miller, B.J., Davis, S.L. (2007). A latent profile analysis of -->
+<!-- college students’ achievement goal orientation. Contemporary Educational Psychology 32(1): -->
+<!-- 8-47. -->
+<!-- Rosenberg, J.M., Beymer, P.N., Anderson, D.J., Van Lissa, C.J., & Schmidt, J.A. (2018). -->
+<!-- tidyLPA: An R Package to Easily Carry Out Latent Profile Analysis (LPA) Using Open-Source -->
+<!-- or Commercial Software. Journal of Open Source Software 3(30): 978. -->
+<!-- Scrucca, L., Fop M., Murphy, T.B., Raftery A.E. (2016) mclust 5: clustering, classification -->
+<!-- and density estimation using Gaussian finite mixture models The R Journal 8/1: 289-317. -->
+<!-- Shireman, E. M., Steinley, D., & Brusco, M. J. (2016). Local Optima in Mixture Modeling. -->
+<!-- Multivariate Behavioral Research 51(4): 466–481. -->
+<!-- Shireman, E., Steinley, D., Brusco, M.J. (2017). Examining the effect of initialization -->
+<!-- strategies on the performance of Gaussian mixture modeling. Behavioral Research Methods -->
+<!-- 49(1):282-293. -->
+<!-- Sterba, S.K. (2013). Understanding Linkages Among Mixture Models. Multivariate Behav -->
+<!-- Res. 2013 48(6): 775-815. -->
+<!-- Wolfe, J. (1970). Pattern clustering by multivariate mixture analysis. Multivariate Behavioral -->
+<!-- Research, 5: 329–350. -->
+
+<!-- Shireman 2016: -->
+<!-- https://www.tandfonline.com/doi/full/10.1080/00273171.2016.1160359?casa_token=YMvI52JWPq0AAAAA%3A20nlLRukoCkUmIil2lRWnsUS-33AlU6HmDRGsu07aJ0nTzS-skjvHnroJBtBVHM5HE0X8iStM1eyqw -->
+
+<!-- mixture models have two possible uses: either (a) classifying observations into groups through the estimation of distinct, underlying distributions or (b) using a set of well-defined parametric distributions (often times the Gaussian distribution) to approximate a messy data space that cannot be approximated by standard models that assume a single distribution (whether univariate or multivariate).
+While both are legitimate uses of mixture modeling, the only necessary and sufficient condition for fitting a mixture model is nonnormality in the data (see Bauer & Curran, 2003).
+It is certainly true that if there are distinct groups arising from distinct distributions, it is necessary to have a set of parameters to model each of those groups.
+Unfortunately, it may often be the case (in fact, it is probably likely) we are in the converse situation where the full data space is either nonlinear, asymmetric, or both, and there exist no distinct groups in the population.
+In this case, the mixture model is capturing the departure from “clean” structure, but the groups themselves are not meaningful in the sense that they are giving rise to distinct subpopulations.
+As such, their boundaries are going to be amorphous, and their parameters, regardless of their stability, lose meaning in terms of reflecting an underlying data-generative mechanism, making the interpretation of the means, covariances, or a specific individuals' membership an exercise in folly.
+Obviously, knowing which situation a data set falls in is paramount, and using the prevalence of local optima as a potential indicator is a crucial first step.
+-->
+
+<!-- The optimal number of latent classes (i.e.
+identity configurations) was determined using the BIC (Fraley & Raftery, 1998); with lower values indicating a better statistical fit.
+Solutions were considered to be admissible if they consisted of at least 5% of the sample (Depaoli, 2013), if the entropy (a measure of class separability) exceeded .8 (Wang, Deng, Yang, 2017), and if the solution was theoretically interpretable and qualitatively distinct from the other class solutions (Howard & Hoffman, 2018).
+-->

### 2022-09-14 manuscript.Rmd
Update 14 Septemeber
In the manuscript.Rmd:

Added Shapiro-Wilks test when discussing Spurk.
Removed some obsolete comments (we still have backup just in case).
Wrote about alternative model parametrizations in LCA, referring back to GRoLTS.
Under Model Fit Indices, I wrote about -2*LL, mentioned we are working on an implementation of the lazy bootstrap.
Finished the Classification Diagnostics and Interpreting Class Solution sections.

In Daniel_LCA:

Finished annotating the code. We can include this section in the manuscript in the next commit.

@@
   models in the `tidySEM` R-package. These functions rely on estimation using
-  the OpenMx R-package.
+  the `OpenMx` R-package.
 
@@
 obsparam <- tapply(X, C, mean)
+shoesize_shapiro <- shapiro.test(X) # Shapiro-Wilk test of normality
 ```
@@
 Many clustering algorithms apply some recursive splitting algorithm to the data.
-By contrast "model-based" clustering refers to the fact that latent class analysis 
+By contrast "model-based" clustering refers to the fact that LCA 
 estimates cluster membership based on a parametric model.
@@
 
-Finally, in the context of machine learning, latent class analysis can be considered as an
+Finally, in the context of machine learning, LCA can be considered as an
 *unsupervised classification* problem [@figueiredo_unsupervised_2002].
 The term *unsupervised* refers to the fact that the outcome variable, true class membership,
-is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome (class membership).
+is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome -- class membership.
 
@@
 As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
-A class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
+a class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
 From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.
+<!-- DA: what is the source for plastic particles? -->
 
@@
 This paper will primarily focus on mixture models and special cases thereof,
-although most of the suggested guidelines are applicable to latent class analyses.
-
-<!-- DA: Do we still need the following comment? In my opinion the included text is sufficient. -->
-<!-- If a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis.
-Modified to handle discrete data, this constrained analysis is known as LCA.
-Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit.
--->
-
-<!-- Because LPA is model-based, a number of different model parameterizations can be -->
-<!-- estimated. These models differ in terms of whether--and how--parameters are -->
-<!-- estimated across the profiles. These parameters are the means for the different -->
-<!-- profiles, which, in this approach, always are estimated freely across the -->
-<!-- profiles; the variances for the variables used to create the profiles, which can -->
-<!-- be estimated freely or can be estimated to be the same, or equal, across -->
-<!-- profiles; and the covariances of the variables used to create the profiles, -->
-<!-- which can be freely-estimated, estimated to be equal, or fixed to be zero. -->
-
-<!-- In general, as more parameters are estimated (i.e., those that are fixed to zero -->
-<!-- are estimated as being equal across profiles; or those estimated as being equal -->
-<!-- across profiles are freely-estimated across them), the model becomes more -->
-<!-- complex; the model may fit better, but also be overfit, meaning that the -->
-<!-- profiles identified may be challenging to replicate with another, separate data -->
-<!-- set. Even still, flexibility in terms of which models can be estimated also has -->
-<!-- affordances. For example, the varying means, equal variances, and covariances -->
-<!-- fixed to 0. A researcher might choose this model specification if she wants to -->
-<!-- model the variables to be used to create profiles that are independent. This -->
-<!-- model is very simple, as no covariances are estimated and the variances are -->
-<!-- estimated to be the same across profiles. As we estimate more parameters (and -->
-<!-- decrease the degrees of freedom), we are more likely to fit the data, but less -->
-<!-- likely to be able to replicate the model with a second set of data. As we -->
-<!-- progress toward more complex models (with increasingly complex -->
-<!-- parameterization), then we are more likely to fit the data better. -->
-<!-- In all, this flexibility associated with LPA also has a more pragmatic cost. It -->
-<!-- can be very difficult to efficiently---and reproducibly---estimate and compare a -->
-<!-- number of models. This cost means that analysts may focus on a subset of models. -->
-<!-- It may also mean that unintentional errors are introduced through copying and -->
-<!-- pasting the output of model estimations. Part of why we developed tidyLPA was to -->
-<!-- make the model estimation process more efficient and more reproducible. As we -->
-<!-- describe later, we do this using R, which has a number of benefits, including -->
-<!-- being open-source, powerful (in terms of pre-processing and using the output of -->
-<!-- models), and increasingly widely-used in the psychological sciences. We also -->
-<!-- provide an interface to perhaps the most widely-used software for LPA, MPlus. In -->
-<!-- this way, tidyLPA can both allow researchers to estimate models they already -->
-<!-- estimate using MPlus more efficiently and reproducibly, and, for those without -->
-<!-- access to this software, to do so entirely using R and the open-source R package -->
-<!-- mclust. -->
+although most of the suggested guidelines are applicable to all latent class analyses.
 
 ## Use cases for latent class analysis
+
+<!-- CJ: It's good but for practical purposes it's preferable -->
+<!-- to use examples from Caspar's published papers. -->
+<!-- Don't spend too much time on this, -->
+<!-- Caspar will check if I can replace them with any of my papers. -->
 
@@
 is reflected in very high or very low conditional response probabilities.
-As a rule of thumb, high indicator quality can be indicated 
-by conditional response probabilities above .9 and below .1. 
-Moderate indicator quality can be indicated by probabilities above .8 and below .2,
-and low indicator quality would be above .7 and below .3.
+<!-- As a rule of thumb, high indicator quality can be indicated  -->
+<!-- by conditional response probabilities above .9 and below .1.  -->
+<!-- Moderate indicator quality can be indicated by probabilities above .8 and below .2, -->
+<!-- and low indicator quality would be above .7 and below .3. <- DA's writing -->
+<!-- This quality rule of thumb is from @geiser_is_2014 where they state that -->
+<!-- "Indicators in high quality conditions were generated to either have CRPs (of endorsing the second category) of 0.9 or 0.1, while moderate quality conditions had CRPs of 0.8 and 0.2, and low quality indicators were generated at 0.7 or 0.3, following Collins and Wugalter’s (1992) strong and weak measurement strength conditions." direct quote from @geiser_is_2014 -->
 For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
@@
 
+<!-- CJ: (on the next paragraph)  -->
+<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
+<!-- That is not what this paragraph is about.  -->
+<!-- This paragraph is about motivations for using LCA. -->
+<!-- One motivation is if your data violates multivariate normality -->
+<!-- because violating multivariate normality means that you have multiple groups -->
+<!-- and you can use LCA to extract those. That is the only assumption  -->
+<!-- we have to discuss here, we just need to explain it. -->
 Finally, LCA can be used to deal with data which violate certain assumptions.
-For instance, nested data usually imply a violation of the independence assumption.
-In this case, an extension of LCA called multilevel LCA can be used (e.g. @henry_multilevel_2010).
 As discussed in the shoe size example, LCA can deal with violations of normality.
 In fact, LCA assumes the population distribution in a non-normal mixture of $K$ normal distributions,
-and LCA can discover the value of $K$, i.e. generate a $K$-class solution.
-
-<!-- check with Caspar if there are other assumptions and how to expand this paragraph -->
-
-<!-- DA: An assumption LCA has is that of conditional or local independence. -->
-<!-- LCA assumes conditional independence of indicators. -->
-<!-- This means class membership explains all the relationships between the indicators, -->
-<!-- and when the latent class is controlled for, the indicators become unrelated [@masyn_latent_2013] -->
-
-<!-- OLD: -->
+and it can discover the value of $K$, i.e. generate a $K$-class solution.
+
 <!-- CJ: Include these use cases too: -->
@@
 
-```{r checkest}
+```{r checkest, include=FALSE}
 tab_check <- data.frame(Item = c(
@@
   "Alternative model specifications",
-  "Starting values",
-  "Algorithm"
+  "Software",
+  "Best practices in estimation",
+  "Algorithm",
+  "Class enumeration",
+  "Model Fit Indices",
+  "Classification Diagnostics",
+  "Interpreting Class Solution",
+  "Label switching",
+  "Best Practices in Reporting",
+  "Tutorial"
+
 ))
@@
 as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
-Pay special attention to level of measurement of the indicators.
-Finite Gaussian mixture models (including LPA) are strictly speaking only suitable for continuous variables.
+Special attention should be paid to level of measurement of the indicators.
+Finite Gaussian mixture models (including LPA) are only suitable for continuous variables.
 Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
@@
 When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
-In our shoe size example, it can be seen that the population distribution comprised of two normal distributions
-is bimodal and hence not normal. 
+In our shoe size example, it can be seen that the population distribution comprised of two normal distributions.
+When examined visually, the population distribution is evidently bimodal. 
+The Shapiro-Wilk normality test ($W = 0.971$, $p < 0.05$ ) 
+rejects the null hypothesis that 
+the sample comes from a normally distributed population.
 Yet, this is a prototypical example of a mixed population distribution where LCA can discover latent groups.
-If the population distribution were instead normal, there would be no classes to extract as the whole population would belong to a single class.
+If the population distribution were instead normal, 
+there would be no classes to extract 
+as the whole population would belong to a single class.
 
@@
 Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @van_lissa_worcs_2021].
-
-<!-- DA: Does this following part about conditional item probabilities and inadmissible solutions belong here?  It's about model interpretation.-->
-
-An important outcome of LCA are conditional item probabilities,
-also known as class-specific item probabilities [@masyn_latent_2013], 
-conditional response or conditional solution probabilities [@geiser_data_2012].
-They indicate the probability of an item being endorsed
-given that the observation belongs to a particular latent class.
-Conditional item probabilities can be obtained using `tidySEM::table_prob()`
-If a particular item is endorsed by two or more classes at markedly different rates,
-it is said to discriminate well between the classes and is consequently considered a good indicator.
-Classes are considered highly homogeneous with respect to an item when for a particular item
-there is a distinct difference in conditional item probabilities for two or more classes.
-For instance, if an item is endorsed below 30% for one class and above 70% for another class,
-the classes have high homogeneity with respect to this item [@masyn_latent_2013].
-Conditional item probabilities are the analogue of mean and standard deviation
-when the indicators are binary/ordinal.
-
-A problem which can occur is that of inadmissible solutions. 
-With binary indicators, LCA is modelling a cross-table with all the predictors.
-The problem with such cross-tables is that they will often contain empty cells,
-i.e. combinations of responses that never occur together. 
-This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
-Such boundary parameter estimates could indicate that the solution is invalid [@geiser_data_2012].
-Boundary parameter estimates can also happen with continuous indicators.
-For instance, if we have a zero-inflated normal distribution and a two class solution,
-one class might have the mean of zero and its standard deviation cannot be determined 
-since there is little variance. 
-This too could be a sign of an invalid solution,
-warn us that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].
 
@@
 While we concur that investigating missingness is due dilligence,
-it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as `OpenMx` (and e.g., Mplus).
+it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, 
+such as Mplus, and `OpenMx` which is the backend of `tidySEM`.
 These packages use Full Information Maximum Likelihood (FIML) estimation,
@@
 
-The different types of latent class models have different parameters.
-For example, mixture models and latent profile analyses typically have class-specific means, variances, and covariances.
+In order to aid researchers working with latent trajectory models,
+@van_de_schoot_grolts-checklist_2017 developed a protocol called Guidelines 
+for Reporting on Latent Trajectory Studies (GRoLTS).
+Two of the GRoLTS guidelines (namely, 6a and 6b) refer to considering alternative model specifications.
+Both are discussing specific cases in latent trajectory model specification. 
+The first is about whether the variance of the growth parameter is estimated freely or fixed,
+and the second is about whether conditional independence is assumed
+as the researcher might also want to free-up the variance-covariance structure.
+In LCA, there are also many different ways to specify the model.
+Means, variances, and covariances between the indicators 
+can either be constrained across classes, or estimated freely.
+Researchers using LCA should transparently report their chosen parametrization 
+as well as discuss different parametrizations that were tested in the process.
+
+Different types of latent class models have different parameters.
+For example, mixture models and latent profile analyses 
+typically have class-specific means, variances, and covariances.
 Latent growth analyses have the same parameters, but with respect to the latent growth variables.
@@
 Confirmatory analyses typically require less comprehensive alternative model specifications.
-For example, in the context of preregistered analyses, the main models of interest may have been specified a priori.
-Therefore, it may not be necessary to test different model specifications when using the confirmatory approach.
-
-In order to aid researchers working with latent trajectory models,
-@van_de_schoot_grolts-checklist_2017 developed a protocol called Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
-Throughout this paper, we adapt those GRoLTS protocols that are applicable to latent class analysis.
-One of the guidelines (namely, 6a) refers to considering alternative model specifications.
-In LCA, means, variances, and covariances between the indicators 
-can either be constrained across classes, or estimated freely.
-Researchers using LCA should transparently report their chosen parametrization 
-as well as discuss different parametrizations that were tested in the process.
-
-<!-- DA: - would researchers ideally anticipate the parameterization of the within-class var/cov structure? How does preregistration connect to this?  -->
-
-```{r}
+For example, in the context of preregistered analyses, 
+the main models of interest may have been specified a priori. 
+Even in case of confirmatory LCA, the theoretical model could be compared to a few others to contextualize it.
+
+```{r, include = FALSE}
 grolts <- structure(list(Number = c("1.", "2.", "3a.", "3b.", "3c.", "4.", 
@@
 <!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
-<!-- analytic functionality. However, the majority of packages are focused on analyses with discrete -->
+<!-- analytic functionality. However, the majority of packages is focused on analyses with discrete -->
 <!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
@@
 <!-- If this were not the case and our mixed population distribution were normal, -->
-<!-- there would be no latent classes to extract as the population would consists of a single class only.  -->
+<!-- there would be no latent classes to extract  -->
+<!-- as the population would consists of a single class only. -->
 <!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups. -->
@@
 <!-- the EM optimizer would be at risk of finding a local optimum instead of the global optimum. -->
-<!-- In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates.  -->
+<!-- In other words, it would converge on a suboptimal solution -->
+<!-- resulting in biased parameter and model fit estimates. -->
 <!-- To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter -->
@@
 <!-- Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values. -->
-<!-- <!-- add more on this --> -->
+
 <!-- Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012). -->
 
-<!-- Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence -->
+<!-- Furthermore, the choice of the starting values  -->
+<!-- can impact the speed of the algorithm's convergence -->
 <!-- or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004). -->
@@
 
-### Model fit indices
+### Model Fit Indices
 
@@
 
+LCA studies commonly report -2*log likelihood of the final class solution.
+This is a basic fit measure used to compute most information criteria.
+However, since log likelihood is not penalized for model complexity,
+it will continuously fall with the addition of more classes.
+
 An alternative is using the bootstrapped likelihood ratio test
@@
 Currently, this test is computationally expensive and can be slow on most computers.
-The development of a faster version of this test is well under way.
+A faster version of this test, 
+namely an implementation of *the lazy bootstrap* [@van_kollenburg_lazy_2018]
+to `tidySEM` is being developed.
 
@@
 
-Three important classification diagnostics provided by `tidySEM` are are entropy, 
-the range of the posterior classification probabilities by most likely class membership,
-and the range of the sample size.
+Three important classification diagnostics provided by `tidySEM` are are 
+*the minimum* and *maximum percentage of the sample assigned to a particular class*,
+*the range of the posterior class probabilities by most likely class membership*, 
+and *entropy*. All three are based on posterior class probabilities.
+
+The posterior class probability is a measure of classification uncertainty
+which can be computed for each individual,
+or averaged for each latent class.
+When the posterior class probability is computed for each individual in the dataset,
+it represents each person's probability of belonging to each latent class.
+For each person, the highest posterior class probability is then determined 
+and the individual is assigned to the corresponding class. 
+We want each individual's posterior class probabilities to be high for one 
+and low for the remaining latent classes. 
+This is considered a high classification accuracy and means that the classes are distinct.
+To obtain posterior class probabilities, run `tidySEM::class_prob()`.
+This function produces an output comprised of several elements.
+Namely:
+
+`$sum.posterior` is a summary table of the posterior class probabilities 
+indicating what proportion of the data contributes to each class.
+
+`$sum.mostlikely` is a summary table of the most likely class membership
+based on the highest posterior class probability.
+From this table, we compute the minimum 
+and maximum percentage of the sample assigned to a particular class,
+, i.e. **n_min** 
+(the smallest class proportion based on the posterior class probabilities)
+and **n_max** (the largest class proportion based on the posterior class probabilities).
+We are especially interested in **n_min** as
+if it is very small and comprised of few observations,
+the model for that group might not be locally identified.
+Estimating LCA parameters on small subsamples
+might lead to bias in the results.
+Therefore, we advise caution when dealing with small classes.
+
+`$mostlikely.class` is a table with rows representing
+the class the person was assigned to,
+and the columns indicating the average posterior probability.
+The diagonal represents the probability that 
+observations in each class will be correctly classified.
+If any of the values on the diagonal of this table is low,
+we might consider not to interpret that solution.
+In `tidySEM` we use the diagonal to compute 
+the range of the posterior class probabilities by most likely class membership
+which consists of
+the lowest class posterior probability (**prob_min**),
+and the highest posterior probability (**prob_max**).
+Both **prob_min** and **prob_max** can be used to (dis)qualify certain class solutions,
+and are a convenient way to summarize class separation in LCA.
+We want both **prob_min** and **prob_max** to be high as 
+that means that for all classes
+the people who were assigned to that class have a high probability of being there.
+**prob_min** is especially important as it can diagnose if there is a class
+with low posterior probabilities 
+which could make one reconsider that class solution.
+<!-- DA: I would like to rephrase the first sentence of mostlikely.class
+to be in the same format as other post. class probs. -->
+
+`$avg.mostlikely` contains the average posterior probabilities for each class, 
+for the subset of observations with most likely class of 1:k, 
+where k is the number of classes.
+
+`$individual` is the individual posterior probability matrix, 
+with dimensions n (number of cases in the data) x k (number of classes).
+Individual class probabilities are often useful for researchers
+who wish to do follow up analyses.
 
@@
 As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
-Entropy should not be used to select a particular class solution. 
 An appropriate use of entropy is that it can disqualify certain solutions if class separability is too low 
 or if one of the latent classes is too small to be meaningful or to calculate descriptive statistics. 
-
-The range of the posterior classification probabilities by most likely class membership are
-
-Another classification diagnostic measure is the range of the sample size.
-
-<!-- A fundamental concept when examining classification accuracy is the posterior class probability. -->
-<!-- This concept represents the probability of belonging to each latent class computed for each individual. -->
-<!-- The highest posterior class probability is then determined and the individual is assigned to the corresponding class. -->
-<!-- We want each individual's posterior class probabilities to be high for one and low for the remaining latent classes. -->
-<!-- This is considered a high classification accuracy and means that the classes are distinct. -->
-
-<!-- @masyn_latent_2013 "entropy was never intended for, nor should it be used for, model selection during the class enumeration process" p. 570 -->
-<!-- add content about entropy -->
-
-<!-- DA: Write about prob_min and prob_max (see name formatting), -->
-<!-- the posterior probability from the model, -->
-<!-- it calculates the entropy and the range of the most likely classification probability, -->
-<!-- the diagonal, lets say you have 3 classes, there is the class the person was assigned to, -->
-<!-- and the probability the person should belong to that class, -->
-<!-- what this is telling us is the range of the average classification probabilities, -->
-<!-- what you want to see is that people have a high probability of being in the class they were assigned to, -->
-<!-- and you want that to be high for all the classes,  -->
-<!-- for each class we want a high number -->
-
-<!-- n_min and n_max are the range of the proprotion of people assigned to those classes. -->
-<!-- The smallest class has XX percent of people, and the largest class has XXX percent of people. -->
-<!-- Sometimes the class is too small to be interpreted, e.g. 5% or less of people, that would be too few to estimate parameters.  -->
-
-<!-- The average posterior probability is a measure of classification uncertainty for each latent class. -->
-
-<!-- The modal class assignment proportion -->
+Entropy was not built for nor should it be used for model selection during class enumeration [@masyn_latent_2013].
+
+**n_min**, **n_max**, **prob_min**, **prob_max**, and **entropy** and can be obtained using `tidySEM::table_fit()`.
+
+## Interpreting Class Solution
+
+An important outcome of LCA are conditional item probabilities,
+also known as class-specific item probabilities [@masyn_latent_2013], 
+conditional response or conditional solution probabilities [@geiser_data_2012].
+They indicate the probability of an item being endorsed
+given that the observation belongs to a particular latent class.
+Conditional item probabilities can be obtained using `tidySEM::table_prob()`
+If a particular item is endorsed by two or more classes at markedly different rates,
+it is said to discriminate well between the classes and is consequently considered a good indicator.
+Classes are considered highly homogeneous with respect to an item when for a particular item
+there is a distinct difference in conditional item probabilities for two or more classes.
+For instance, if an item is endorsed below 30% for one class and above 70% for another class,
+the classes have high homogeneity with respect to this item [@masyn_latent_2013].
+Conditional item probabilities are the analogue of mean and standard deviation
+when the indicators are binary or ordinal.
+
+A problem which can occur is that of inadmissible solutions. 
+With binary indicators, LCA is modelling a cross-table with all the predictors.
+The problem with such cross-tables is that they will often contain empty cells,
+i.e. combinations of responses that never occur together. 
+This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
+Such boundary parameter estimates could indicate that the solution is invalid [@geiser_data_2012].
+Boundary parameter estimates can also happen with continuous indicators.
+For instance, if we have a zero-inflated normal distribution and a two class solution,
+one class might have the mean of zero and its standard deviation cannot be determined 
+since there is little variance. 
+This too could be a sign of an invalid solution,
+warn us that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].
 
 ### Label switching
-<!-- optional subsection: interpreting class solutions (with class naming) -->
 
@@
 
-## Best practices in reporting
+## Best Practices in Reporting
 
 Among studies using LCA, reporting practices vary significantly [@weller_latent_2020].
-Various authors have tried to better and standardize ways of reporting LCA [e.g. @masyn_latent_2013; @weller_latent_2020]. 
-Building on their work, we provide guidelines with a strong emphasis on scientific reproducibility and transparency
-better known as the open science framework (OSF). The OSF promotes scientific openness, reproducibility and integrity 
-through an establishment of guidelines for researchers (Foster & Deardorff, 2017)
-To this end, @van_lissa_worcs_2020 developed WORCS, a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by Nosek and colleagues (2015).
-It can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
+Various authors have tried to better and standardize ways of reporting LCA [e.g. @masyn_latent_2013; @weller_latent_2020], but more work is needed.
+@van_lissa_worcs_2020 developed WORCS, a workflow for open reproducible code in science.
+WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by @nosek_promoting_2015.
+WORCS workflow can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
 version control, citation and formatting, among others [@van_lissa_worcs_2020].
 
-The guidelines emphasise the use of comprehensive citation (including referencing the software used in the analysis),
-as well as code and data sharing wherever possible.
+TOP-guidelines emphasise the use of comprehensive citation (including referencing the software used in the analysis),
+as well as code and data sharing wherever possible [@nosek_promoting_2015].
 @van_lissa_worcs_2020 suggest sharing synthetic data in case the original data cannot be shared, 
@@
 
-## Best practices in visualization
+## Best Practices in Visualization
 
@@
 
-<!-- 5. Tutorial -->
-<!-- 5.1 ‘mclust’ -->
-<!-- To run LPA using the ‘mclust’ package, the Mclust() function can be used. This function -->
-<!-- requires data as its minimum input. In addition, the number of clusters to fit (G) and model -->
-<!-- variants to fit and select from (modelNames) can be entered (defaults are: G=1-9 and all -->
-<!-- fourteen model variants). In many cases, we may want to consider a more limited range of -->
-<!-- model variants to with different numbers of classes. For multivariate data, a total of fourteen -->
-<!-- options are available (see help("mclustModelNames") for an overview). Each model variant -->
-<!-- is estimated for each value of the specified range of G using Expectation Maximization (EM), -->
-<!-- using the Bayesian Information Criterion (BIC) to compare different model variants and -->
-<!-- select the optimal one. -->
-<!-- Conducting a latent variable model or mixture analysis with Mplus or comparable software -->
-<!-- usually entails fitting models with different numbers of classes (that are otherwise configured -->
-<!-- similarly) and to compare the fit between these models to select the best model. The -->
-<!-- default approach taken in the ‘mclust’ package is slightly different in that the main modeling -->
-<!-- command Mclust() fits multiple models for all possible combinations of the specified number -->
-<!-- of classes and model configurations . The eventual output of ‘mclust’ is the BIC-selected best -->
-<!-- model variant with the optimal combination of number of classes and model configuration. -->
-<!-- An advantage of the ‘mclust’ approach is that it is efficient and flexible: given a number of -->
-<!-- classes, the best model configuration is selected straight away and we do not have to run all -->
-<!-- model combinations one by one. This makes the approach especially suitable for exploratory -->
-<!-- analyses. However, like stated above, this approach is somewhat less usual in some fields, -->
-<!-- where estimation of LPA is approached in a more confirmatory fashion. In addition, it does -->
-<!-- take some control away from the researcher, who may want to primarily focus on selecting -->
-<!-- the number of classes, given a particular , theory-based model configuration that is kept -->
-<!-- constant throughout analyses, which is often the chosen strategy in software like Mplus and -->
-<!-- made easily available through the ‘tidyLPA’ package. -->
-<!-- Here, both approaches are shown: (1) the ‘mclust’ aproach and (2) the ‘tidyLPA’ approach. -->
-
-<!--                                               5 -->
-<!-- 5.1.1 The ‘mclust’ approach -->
-<!-- In this approach, we run consecutive models with increasing numbers of classes (G=1 to -->
-<!-- G=9), for each model we let the package fit the four above described model variants (“EEI”, -->
-<!-- “EEE”, “VVI” and “VVV”) and select the best-fitting one. We start by loading the package -->
-<!-- (library(mclust)) and creating an object mnames that contains the names of the four models -->
-<!-- we want to be fitted. Next, we use the Mclust() function to fit the models: -->
-<!-- library(mclust) -->
-
-<!-- ## Package 'mclust' version 5.4.7 -->
-<!-- ## Type 'citation("mclust")' for citing this R package in publications. -->
-<!-- mnames <- c("EEI", "EEE", "VVI", "VVV") -->
-
-<!-- # Fit 1-5 class model -->
-<!-- mod_g1_9 <- Mclust(dat1[, 1:10], modelNames = mnames) -->
-
-<!-- We can first look at the optimal number of classes and the optimal model variant that were -->
-<!-- selected, using the following code: -->
-<!-- # Opimal number of classes -->
-<!-- mod_g1_9$G -->
-
-<!-- ## [1] 3 -->
-<!-- # Optimal model variant -->
-<!-- mod_g1_9$modelName -->
-
-<!-- ## [1] "EEI" -->
-<!-- This shows us that a 3-class EEI model was selected as the best model based on the BIC. We -->
-<!-- can get a better perspective of this model’s performance if we compare it to the other fitted -->
-<!-- models. We can do this by taking a closer look at the other models’ BIC values: -->
-<!-- mod_g1_9$BIC -->
-
-<!-- ##   Bayesian Information Criterion (BIC): -->
-<!-- ##           EEI       EEE       VVI       VVV -->
-<!-- ##   1 -14661.61 -12986.57 -14661.61 -12986.57 -->
-<!-- ##   2 -13275.65 -12611.60 -13095.82 -12678.12 -->
-<!-- ##   3 -12142.11 -12290.68 -12163.08 -12731.18 -->
-<!-- ##   4 -12178.44 -12330.04 -12241.21 -12970.57 -->
-<!-- ##   5 -12223.14 -12379.96 -12320.47 -13231.35 -->
-<!-- ##   6 -12264.04 -12423.72 -12406.59 -13522.40 -->
-<!-- ##   7 -12269.39 -12451.25 -12497.03 -13748.91 -->
-<!-- ##   8 -12298.30 -12483.35 -12568.62 -13980.82 -->
-<!-- ##   9 -12347.11 -12529.40 -12645.50 -14269.34 -->
-<!-- ## -->
-
-<!--                                              6 -->
-<!-- ## Top 3 models based on the BIC criterion: -->
-<!-- ##     EEI,3     VVI,3     EEI,4 -->
-<!-- ## -12142.11 -12163.08 -12178.44 -->
-<!-- mod_g1_9$loglik -->
-
-<!-- ## [1] -5951.275 -->
-<!-- mod_g1_9$df -->
-
-<!-- ## [1] 42 -->
-<!-- Here, we can see the BICs for all fitted models (in this case 1-9 classes and 4 model variants: -->
-<!-- 36 models in total). We can see that the closest contenders were a 3-class model with a VVI -->
-<!-- configuration (with variances allowed to vary both within and between classes) and a 4-class -->
-<!-- EEI model. -->
-<!-- Note that the BIC values are negative, and that values that are more negative are considered -->
-<!-- to fit poorer than values closer to 0. This may strike some users as odd, given that in -->
-<!-- many modeling applications, we are used to consider lower BIC values to indicate better -->
-<!-- fit. However, the ‘mclust’ approach of maximizing the BIC is correct for the used modeling -->
-<!-- approach (see e.g., Fraley & Raftery, 2003; Banfield & Raftery, 1993). For theoretical reasons, -->
-<!-- the BIC is calculated in’mclust’ as: -->
-
-<!--                            BIC = 2(Loglikelihood) - df (log(n)) -->
-
-<!-- where df is the number of parameters (degrees of freedom) and n is the sample size. We can -->
-<!-- see, that here, higher log likelihood values lead to a higher BIC. For the best fitting model in -->
-<!-- our example, the log likelihood is -5951.275, the number of parameters is 42 and the sample -->
-<!-- size is 300. If we plug these into the formula, we get the following BIC: -->
-
-<!--                      BIC = 2(-5951.275) - 42(log(300)) = -12142.11 -->
-
-<!-- As expected, the obtained value indeed corresponds to the BIC we got in the model output -->
-<!-- (see above). Other software packages, such as Mplus and ‘tidyLPA’ (see below) use a slightly -->
-<!-- different formula to calculate the BIC, where higher log likelihood values lead to lower BIC -->
-<!-- values: -->
-
-<!--                            BIC = df (log(n)) - 2(Loglikelihood) -->
-
-<!-- When we plug the best-model values into this formula, we get: -->
-
-<!--                       BIC = 42(log(300)) - 2(-5951.275) = 12142.11 -->
-
-<!-- We can see that the absolute BIC value is the same, irrespective of the used formula, with -->
-<!-- only the sign differing between the two approaches. -->
-
-<!--                                               7 -->
-<!-- Now, let us continue by taking a closer look at the class sizes and the mean values and -->
-<!-- variances of the ten variables for the three identified classes. Because the best model has and -->
-<!-- EEI configuration, printing out a single variance matrix from one class will suffice, because -->
-<!-- the variances are the same across classes (this is different in cases where the optimal model -->
-<!-- allows for between-class differences in (co)variances: -->
-<!-- # tabulate class-membership numbers -->
-<!-- table(summary(mod_g1_9)$classification) -->
-
-<!-- ## -->
-<!-- ##   1   2      3 -->
-<!-- ## 100 110     90 -->
-<!-- # display the means per class -->
-<!-- mod_g1_9$parameters$mean -->
-
-<!-- ##             [,1]         [,2]        [,3] -->
-<!-- ##   var1 2.023572      6.083019   2.0304980 -->
-<!-- ##   var2 3.056837      6.937785   4.8997993 -->
-<!-- ##   var3 1.718166      1.107873   7.9385872 -->
-<!-- ##   var4 6.086727      2.854402   6.9657536 -->
-<!-- ##   var5 7.101109      2.225675   4.7416266 -->
-<!-- ##   var6 3.691936      8.081570   2.2469977 -->
-<!-- ##   var7 5.009819      2.817889   0.8862725 -->
-<!-- ##   var8 8.316894      8.080891   3.9350536 -->
-<!-- ##   var9 1.978136      1.896294   9.1537131 -->
-<!-- ##   var10 1.191249     5.142590   8.0332628 -->
-<!-- # select the first of 3 10-by-10 variance matrices -->
-<!-- mod_g1_9$parameters$variance$sigma[1:10, 1:10, 1] -->
-
-<!-- ##               var1       var2       var3       var4       var5       var6       var7       var8 -->
-<!-- ##   var1    2.539537   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var2    0.000000   2.117941   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var3    0.000000   0.000000   2.509035   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var4    0.000000   0.000000   0.000000   2.812796   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var5    0.000000   0.000000   0.000000   0.000000   2.647399   0.000000   0.000000   0.000000 -->
-<!-- ##   var6    0.000000   0.000000   0.000000   0.000000   0.000000   2.694156   0.000000   0.000000 -->
-<!-- ##   var7    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.368379   0.000000 -->
-<!-- ##   var8    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   2.540704 -->
-<!-- ##   var9    0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##   var10   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000   0.000000 -->
-<!-- ##               var9      var10 -->
-<!-- ##   var1    0.000000   0.000000 -->
-<!-- ##   var2    0.000000   0.000000 -->
-<!-- ##   var3    0.000000   0.000000 -->
-<!-- ##   var4    0.000000   0.000000 -->
-
-<!--                                                8 -->
-<!-- ##   var5    0.000000   0.000000 -->
-<!-- ##   var6    0.000000   0.000000 -->
-<!-- ##   var7    0.000000   0.000000 -->
-<!-- ##   var8    0.000000   0.000000 -->
-<!-- ##   var9    2.509146   0.000000 -->
-<!-- ##   var10   0.000000   2.204457 -->
-<!-- From this output, we can see that the three classes have class-sizes of 100, 110 and 90, -->
-<!-- respectively. These sizes along with the patterns of class-specific mean variable scores -->
-<!-- correspond with those that served as the input for the simulation. As specified, the variances -->
-<!-- vary slightly across variables. -->
-<!-- The class allocation in LPA is probabilistic in nature: each subject in the data is assigned a -->
-<!-- probability for each of the estimated classes, based on their pattern of scores on the input -->
-<!-- variables. These probabilities can be inspected in the z-matrix (here: mod_g1_9$z). Subjects -->
-<!-- can be allocated to one of the classes based on their highest class-probability. These posterior -->
-<!-- allocations can be found in the classification matrix (here: mod_g1_9$classification). -->
-<!-- These class-allocations were tabulated above to evaluate class sizes. -->
-<!-- Note that class-allocation in this way only yields useful classifications if the patterns of -->
-<!-- class-probabilities allow for allocation of each subject to a single class with sufficient certainty. -->
-<!-- In case of too much uncertainty, using the classification is not advised. We can inspect the -->
-<!-- uncertainty of allocation for all subjects (here: mod_g1_9$uncertainty). This gives us a list -->
-<!-- of uncertainties for all subjects. Next, we can evaluate the extent of uncertainty by looking -->
-<!-- at the maximum uncertainty or, for instance, the averaged uncertainty across subjects: -->
-<!-- max(mod_g1_9$uncertainty) -->
-
-<!-- ## [1] 0.003852604 -->
-<!-- mean(mod_g1_9$uncertainty) -->
-
-<!-- ## [1] 2.607299e-05 -->
-<!-- Here, the uncertainty is atypically low because of the used idealized data. In many cases, -->
-<!-- uncertainty is higher and it may be of interest to investigate it further, for instance, by -->
-<!-- looking at (e.g., plotting or tabulating) the uncertainty per class: -->
-<!-- cprob <- cbind(mod_g1_9$z, mod_g1_9$classification) -->
-<!-- cprob <- as.data.frame(cprob) -->
-<!-- colnames(cprob) <- c("prob (class 1)", "prob (class 2)", "prob (class 3)", "class") -->
-<!-- aggregate(cprob[, 1:3], list(cprob$class), mean) -->
-
-<!-- ##   Group.1 prob (class 1) prob (class 2) prob (class 3) -->
-<!-- ## 1       1   9.999824e-01   1.761611e-05   3.350070e-10 -->
-<!-- ## 2       2   5.486300e-05   9.999451e-01   7.192849e-08 -->
-<!-- ## 3       3   2.591358e-08   1.675400e-07   9.999998e-01 -->
-<!-- Here we can see that for each class, the mean probability of membership of that class was -->
-
-<!--                                                  9 -->
-<!-- very high and probabilities for the other classes were very low. Again, these values reflect the -->
-<!-- idealized data; in many cases, more differentiated class-probability patterns may be observed. -->
-
-<!-- 5.1.2 ‘tidyLPA’ -->
-<!-- When running ‘mclust’ models from the ‘tidyLPA’ package, the estimations are approached -->
-<!-- a little differently. By default, the number of models that can be estimated is restricted to -->
-<!-- six relatively common variants (see help(tidyLPA::estimate_profiles) for details). Of -->
-<!-- these models, four can be estimated with ‘mclust’. These models differ in terms of how the -->
-<!-- variances and covariances are allowed to vary or constrained to be equal across classes and -->
-<!-- whether covariances are or are not fixed to zero within classes. These models have each been -->
-<!-- allocated a number (1-6) in ‘tidyLPA’. See the table below how these numbers correspond to -->
-<!-- the above mentioned model configurations. -->
-<!--  Model variant    model number -->
-<!--  EEI              1 -->
-<!--  EEE              3 -->
-<!--  VVI              2 -->
-<!--  VVV              6 -->
-<!-- Models four and five are different and can only be fit when R is interfacing with Mplus using -->
-<!-- the ‘MplusAutomation’ package. This is outside the scope of this tutorial. -->
-<!-- To fit an LPA model in ‘tidyLPA’, we use the estimate_profiles() function. Here, we need -->
-<!-- to enter the data.frame to be used (df) and the number of classes to estimate (n_profiles). -->
-<!-- To determine what kind of model configuration to estimate, the authors have provided two -->
-<!-- different ways in the estimate_profiles() command. In the first approach, we simply tell what -->
-<!-- model number (see Table 2) we want to estimate using the models= argument For instance, if -->
-<!-- we want to estimate LPA models with 1 to 9 classes, with an EEI configuration, we can use: -->
-<!-- <!-- suppressMessages(library(tidyLPA)) --> -->
-<!-- <!-- suppressMessages(mod_1c_v1 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, --> -->
-<!-- <!--     models = 1)) --> -->
-
-<!-- The package issues a standard message by default notifying us that we use the models -->
-<!-- argument and that the variances and covariances arguments are ignored. Here, we suppress -->
-<!-- the message. The output shows us the model-fit information for the 1 to 9-class EEI models: -->
-<!-- <!-- mod_1c_v1 --> -->
-
-<!-- ## tidyLPA analysis using mclust: -->
-<!-- ## -->
-<!-- ## Model Classes AIC       BIC      Entropy            prob_min   prob_max   n_min   n_max   BLRT_p -->
-<!-- ## 1      1       14587.53 14661.61 1.00               1.00       1.00       1.00    1.00 -->
-<!-- ## 1      2       13160.84 13275.65 1.00               1.00       1.00       0.30    0.70    0.01 -->
-<!-- ## 1      3       11986.55 12142.11 1.00               1.00       1.00       0.30    0.37    0.01 -->
-<!-- ## 1      4       11982.14 12178.44 0.91               0.79       1.00       0.14    0.33    0.05 -->
-<!-- ## 1      5       11986.09 12223.14 0.89               0.75       1.00       0.06    0.33    0.30 -->
-
-<!--                                               10 -->
-<!-- ##   1      6         11986.25    12264.04   0.84      0.78       1.00        0.08   0.30    0.16 -->
-<!-- ##   1      7         11950.86    12269.39   0.81      0.77       0.91        0.08   0.17    0.01 -->
-<!-- ##   1      8         11939.04    12298.30   0.82      0.79       0.92        0.08   0.17    0.02 -->
-<!-- ##   1      9         11947.10    12347.11   0.82      0.76       0.92        0.06   0.16    0.57 -->
-<!-- Here, we can see that the BIC takes on different values compared to ‘mclust’, with lower -->
-<!-- rather than higher values indicating best fit (see explanation above). Instead of drawing the -->
-<!-- fit indices (e.g., BIC, AIC) directly from the ‘mclust’ package, ‘tidyLPA’ only draws the ‘raw’ -->
-<!-- log likelihood values and posterior probabilities from the ‘mclust’ output and (re)calculates -->
-<!-- the BIC, AIC, entropy etc. to mirror as well as possible those provided in Mplus. In addition, -->
-<!-- p-values of the the bootstrapped likelihood ratio test (BLRT) are given by default. These -->
-<!-- p-values indicate for each k-class model whether adding the kth class adds significantly to -->
-<!-- model fit. For instance, the BLRT_p of 0.04 for the 4-class model indicates that adding -->
-<!-- a fourth class led to an improvement of model fit compared to the 3-class model that was -->
-<!-- just significant when using an alpha of 0.05. When using ’mclust’ as a stand-alone package, -->
-<!-- the BLRT is not calculated by default, but can be obtained with the mclustBootstrapLRT -->
-<!-- function. -->
-<!-- Next, we can run similar commands, with different values for the models argument. It is -->
-<!-- also possible to estimate more than one model variant in a single run by providing a vector -->
-<!-- of model numbers (e.g., models=c(1,6)). -->
-<!-- The second approach to determine the model variant to be estimated is by using the variances -->
-<!-- and covariances arguments in the estimate_profiles command. The variances argument -->
-<!-- can have the values "equal" (i.e. variable variances constrained to be equal across classes) or -->
-<!-- "varying" (i.e. variances allowed to vary across classes). The covariances argument can -->
-<!-- have the values: "zero" (i.e. all variable covariances fixed to zero), "equal" (i.e. covariances -->
-<!-- constrained to be equal across classes) or "varying" (i.e. covariances allowed to vary across -->
-<!-- classes). Now, if we again want to estimate LPA models with 1 to 9 classes, with an EEI -->
-<!-- configuration, we can use: -->
-<!-- <!-- mod_1c_v2 <- estimate_profiles(df = dat1[1:10], n_profiles = 1:9, variances = "equal", --> -->
-<!-- <!--     covariances = "zero") --> -->
-<!-- <!-- mod_1c_v2$model_1_class_2$fit --> -->
-
-<!-- <!-- ##         Model           Classes        LogLik                   AIC              AWE --> -->
-<!-- <!-- ## 1.000000e+00       2.000000e+00 -6.549418e+03          1.316084e+04     1.354347e+04 --> -->
-<!-- <!-- ##           BIC              CAIC           CLC                   KIC            SABIC --> -->
-<!-- <!-- ## 1.327565e+04       1.330665e+04 1.310084e+04           1.319484e+04     1.317734e+04 --> -->
-<!-- <!-- ##           ICL           Entropy      prob_min              prob_max            n_min --> -->
-<!-- <!-- ## -1.327568e+04      9.996750e-01 9.999484e-01           9.999990e-01     3.000000e-01 --> -->
-<!-- <!-- ##         n_max          BLRT_val        BLRT_p --> -->
-<!-- <!-- ## 7.000000e-01       1.448699e+03 9.900990e-03 --> -->
-<!-- As expected, the output is exactly the same as the one we got when using the models -->
-<!-- argument. Now, if we want to, we can continue fitting other model variants using either -->
-<!-- approach. Next, we can manually compare the models in terms of their BIC and AIC values -->
-
-<!--                                               11 -->
-<!-- or we can use compare_solutions to do this in an automated fashion. Interestingly, the -->
-<!-- best model can be selected based on the integrated information about several fit indices -->
-<!-- (analytic hierarchy process; see help(tidyLPA::AHP)) for more details). We obtain the model -->
-<!-- comparison with the following code. -->
-<!-- <!-- comp <- suppressWarnings(compare_solutions(mod_1c_v1)) --> -->
-<!-- <!-- comp$fits --> -->
-
-<!-- <!-- ##   # A tibble: 9 x 18 --> -->
-<!-- <!-- ##     Model Classes LogLik    AIC    AWE    BIC   CAIC    CLC    KIC SABIC      ICL --> -->
-<!-- <!-- ##     <dbl>   <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>           <dbl> --> -->
-<!-- <!-- ##   1     1       1 -7274. 14588. 14834. 14662. 14682. 14550. 14611. 14598. -14662. --> -->
-<!-- <!-- ##   2     1       2 -6549. 13161. 13543. 13276. 13307. 13101. 13195. 13177. -13276. --> -->
-<!-- <!-- ##   3     1       3 -5951. 11987. 12506. 12142. 12184. 11905. 12032. 12009. -12142. --> -->
-<!-- <!-- ##   4     1       4 -5938. 11982. 12638. 12178. 12231. 11878. 12038. 12010. -12218. --> -->
-<!-- <!-- ##   5     1       5 -5929. 11986. 12778. 12223. 12287. 11860. 12053. 12020. -12273. --> -->
-<!-- <!-- ##   6     1       6 -5918. 11986. 12915. 12264. 12339. 11838. 12064. 12026. -12353. --> -->
-<!-- <!-- ##   7     1       7 -5889. 11951. 13016. 12269. 12355. 11780. 12040. 11997. -12376. --> -->
-<!-- <!-- ##   8     1       8 -5873. 11939. 13141. 12298. 12395. 11747. 12039. 11991. -12409. --> -->
-<!-- <!-- ##   9     1       9 -5866. 11947. 13285. 12347. 12455. 11733. 12058. 12005. -12458. --> -->
-<!-- <!-- ##   # ... with 7 more variables: Entropy <dbl>, prob_min <dbl>, prob_max <dbl>, --> -->
-<!-- <!-- ##   #   n_min <dbl>, n_max <dbl>, BLRT_val <dbl>, BLRT_p <dbl> --> -->
-<!-- <!-- comp$best --> -->
-
-<!-- <!-- ## LogLik       AIC      AWE      BIC     CAIC        CLC    KIC   SABIC       ICL --> -->
-<!-- <!-- ##      9         8        3        3        3          9      3       8         3 --> -->
-<!-- As shown in the ‘mclust’ examples above, the EEI model (models=1) with 3 classes fit the -->
-<!-- data best, not only according to the BIC, but according to the integrated information from a -->
-<!-- broader range of fit indices. Note that we did not fit any competing EEE, VVV and VVI -->
-<!-- models here. If we would have wanted to compare fit to these model variants too, we should -->
-<!-- have actively told the package to fit these too, either by specifying a vector of different types -->
-<!-- of models to fit in the models argument or by repeating the analysis with different values for -->
-<!-- the variances and covariances arguments. -->
-
-<!-- 5.2 Comparison to Mplus -->
-<!-- For comparison, we fit the equivalent EEI model variants in Mplus. For this, we can mostly -->
-<!-- rely on the defaults of Mplus. We only need to specify that we want to estimate class-specific -->
-<!-- variable means. The variances are automatically constrained to be equal across classes and -->
-<!-- within class variable covariances are fixed to 0. Here we have the separate codes for fitting -->
-<!-- the first 4 models wit 1-4 class models (code for the 5-9 class models not shown). -->
-<!-- <!-- !! 1-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-
-<!-- <!--                                                  12 --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(1); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-
-<!-- <!-- !! 2-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(2); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#2% --> -->
-<!-- <!--   [v1-v10]; --> -->
-
-<!-- <!-- !! 3-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(3); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#2% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#3% --> -->
-<!-- <!--   [v1-v10]; --> -->
-
-<!-- <!--                             13 --> -->
-<!-- <!-- !! 3-class model --> -->
-<!-- <!--   DATA: file='dat1.dat'; --> -->
-<!-- <!--   VARIABLE: --> -->
-<!-- <!--     names=id v1-v10; --> -->
-<!-- <!--     usevariables= v1-v10; --> -->
-<!-- <!--     classes = c(4); --> -->
-<!-- <!--   ANALYSIS: --> -->
-<!-- <!--     type = mixture; --> -->
-<!-- <!--     starts= 250 50; --> -->
-<!-- <!--   MODEL: --> -->
-<!-- <!--   %overall% --> -->
-<!-- <!--   %C#1% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#2% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#3% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- <!--   %C#4% --> -->
-<!-- <!--   [v1-v10]; --> -->
-<!-- Each of the models was fitted with multiple initial stage random starts and multiple final -->
-<!-- stage optimizations to reduce the risk of finding solutions on a local maximum. For the 1-6 -->
-<!-- class models, 250 iniitial and 50 final starts were used. For the 7-class model 1000 initial -->
-<!-- starts and 200 final optimizations were used and for the 8- and 9-class models, a replicable -->
-<!-- global solution could only be obtained with 5000 initial stage starts and 1000 final stage -->
-<!-- optimizations. The fit indices for each of the fitted models is displayed in the table below. -->
-<!-- For now, we only look at the estimated fit indices. -->
-<!--                 EEI models -->
-<!--  Classes   AIC       BIC -->
-<!--  1         14587.535 14587.535 -->
-<!--  2         13160.836 13275.653 -->
-<!--  3         11986.551 12142.11 -->
-<!--  4         11957.71  12154.01 -->
-<!--  5         11951.766 12188.808 -->
-<!--  6         11944.632 12222.415 -->
-<!--  7         11935.501 12254.027 -->
-<!--  8         11929.428 12288.695 -->
-<!--  9         11925.203 12325.212 -->
-<!-- If we compare the BIC values for the EEI models to those obtained with ‘mclust’ directly -->
-<!-- and via ‘tidyLPA’, we can see that the results are quite similar for the less complex models, -->
-<!-- with the BIC-values for the 2- and 3-class models being exactly the same in terms of their -->
-<!-- absolute values. When taking a closer look at the 3-class model that was previously found to -->
-<!-- be the best model, we can also see that that the Mplus-based classifications (n=100, n=110, -->
-<!-- and n=90) and entropy estimates (entropy=1.0) are similar to those obtained with ‘mclust’ -->
-
-<!--                                              14 -->
-<!-- and/or ‘tidyLPA’. -->
-<!--  If we look at the larger picture and evaluate the overlap between the BIC values obtained -->
-<!--  for all models with class numbers ranging from 1 to 9, we can see that the estimated BIC -->
-<!-- values show more differences between the R-packages and Mplus for the models with larger -->
-<!--  numbers of classes. This can be explained by the different methods used by Mplus and -->
-<!-- ‘mclust’/’tidyLPA’ to generate the start values for estimation (see below). These differences -->
-<!--  can cause the packages to yield different results for more complex models and/or models that -->
-<!--  are increasingly misspecified (as the models with more than 3 classes were in our example). -->
-
-<!-- 6. Final comments -->
-<!-- In this tutorial we have seen that it is relatively easy to run an LPA in R. As with all -->
-<!-- data-driven analyses, care should be taken not to over-interpret the results. In the end, LPA -->
-<!-- identifies classes in such a way as to optimally explain variance on a range of variables and -->
-<!-- not to optimize interpretability or usefulness. -->
-<!-- Another remark with regard to LPA results is that class-membership is probabilistic in nature: -->
-<!-- each subject in an analyzed sample has a probability of being in each of the model’s classes. -->
-<!-- Subjects can be allocated to a class based on their highest class-probability. Importantly, -->
-<!-- this can only be done with enough certainty if separation between classes is sufficient, which -->
-<!-- means that we see that each subject has a clear highest probability (e.g., p=0.9) for one of -->
-<!-- the classes. The entropy statistic is often used to quantify this separation, with values <0.8 -->
-<!-- being taken to indicate insufficient separation to allocate subjects to one class. In such cases, -->
-<!-- the class probabilities themselves can still be used in further investigations. -->
-<!-- Depending on one’s preferences, one can choose either the ‘mclust’ or the ‘tidyLPA’ package. -->
-<!-- The latter may be especially attractive for persons that already have experience with Mplus -->
-<!-- or comparable software. It is important to note, though, that differences exist between the -->
-<!-- estimation approaches taken by both R-packages and by Mplus. -->
-<!-- Mixture models such as LPA need to be estimated in an iterative fashion, using an EM -->
-<!-- algorithm that is very sensitive to the used start values, where poor start values can lead the -->
-<!-- estimation process to a poor solution. In addition, there is often a chance that the iterative -->
-<!-- EM process leads to a solution that is at a local, rather than a global maximum in the -->
-<!-- likelihood ‘landscape’. -->
-
-<!-- ### Starting values -->
-<!-- Different methods have been developed to initialize the estimation (get starting values) in -->
-<!-- such a way as to optimize the chance of arriving at an accurate model solution (Biernacki et -->
-<!-- al., 2003; Shireman et al., 2016). Mplus and ‘mclust’ take two different approaches. Mplus -->
-<!-- uses a ‘brute force’ approach and reruns the model with multiple sets of random start values, -->
-<!-- each generated from uniform distributions of values with ranges that are based on the data -->
-<!-- (Muthén & Muthén, 1998-2015). The best solution is the model with the highest log likelihood -->
-<!-- that was arrived upon from at least two different starting points. In contrast, ‘mclust’ uses -->
-<!-- the data itself to generate a set of plausible start values using hierarchical clustering. This -->
-<!-- means that the start values are informed by the (hierarchical structure of) the data, which -->
-<!-- can work well. A downside is that only a single set of start values is used, so the risk of -->
-<!-- arriving at a local optimum is not addressed. These approaches show clear differences and -->
-<!-- have been shown previously to not always arrive at the same solutions, with some authors -->
-<!-- being especially critical of the ‘mclust’ approach (Shireman et al., 2017). In our example, -->
-<!-- we did not see much difference between the two approaches, but it should be noted that -->
-<!-- many real-world data sets do not have such a clear latent structure, making it much more -->
-<!-- challenging to arrive at an accurate solution, given the data. -->
-<!-- Based on the provided examples and these technical differences between ‘mclust’ and Mplus, -->
-<!-- it is probably better to see ‘mclust’ and ‘tidyLPA’ as useful alternatives for Mplus, rather -->
-<!-- than as tools to exactly mimic what can be done with the latter software. -->
-
-<!-- References -->
-<!-- Banfield, J., Raftery, A.E. (1993). Model-based Gaussian and non-Gaussian clustering. -->
-<!-- Biometrics. 49: 803–821. -->
-<!-- Biernacki,C.,Celeux,G., Govaert,G.(2003).Choosing starting values for the EM algorithm -->
-<!-- for getting the highest likelihood in multivariate Gaussian mixture models. Computational -->
-<!-- Studies and Data Analysis, 41, 561–575. -->
-<!-- Celeux, G., Govaert, G. (1995) Gaussian parsimonious clustering models. Pattern Recognition. -->
-<!-- 28:781–793 -->
-<!-- Fraley, C., Raftery, A.E. (1998) How many clusters? Which clustering method? Answers via -->
-<!-- model-based cluster analysis. The Computer Journal 41: 578-588. -->
-<!-- Gibson, W. A. (1959). Three multivariate models: Factor analysis, latent structure analysis, -->
-<!-- and latent profile analysis. Psychometrika, 24, 229–252. -->
-<!-- Linzer, D.A., Lewis J. (2011). “poLCA: an R Package for Polytomous Variable Latent Class -->
-<!-- Analysis.” Journal of Statistical Software. 42(10): 1-29 -->
-<!-- Meyer, D., Dimitriadou, E., Hornik, K., Weingessel, A., Leisch, F. (2019). e1071: Misc -->
-<!-- Functions of the Department of Statistics, Probability Theory Group (Formerly: E1071), TU -->
-<!-- Wien. R package version 1.7-3. https://CRAN.R-project.org/package=e1071 -->
-<!-- Muthén, L., Muthén, B., 1998–2015. Mplus User’s Guide, 7th ed., Muthén & Muthén, Los -->
-<!-- Angeles. -->
-<!-- Neale, M.C., Hunter, M.D., Pritikin, J.N., Zahery, M., Brick, T.R., Kirkpatrick, R.M., -->
-<!-- Estabrook, R., Bates, T.C., Maes, H.H., Boker, S.M. (2016). “OpenMx 2.0: Extended -->
-<!-- structural equation and statistical modeling.” Psychometrika 81(2): 535-549. -->
-<!-- Nylund-Gibson, K., Choi, A. Y. (2018). Ten frequently asked questions about latent class -->
-<!-- analysis. Translational Issues in Psychological Sciences, 4, 440–461. -->
-<!-- Oberski, D. (2016). Mixture models: Latent profile and latent class analysis. In J. Robertson -->
-<!-- & M. Kaptein (Eds.), Modern statistical methods for HCI (pp. 275–287). Cham, Switzerland: -->
-<!-- Springer Inter-national Publishing. -->
-
-<!--                                              16 -->
-<!-- Pastor, D.A., Barron, K.E., Miller, B.J., Davis, S.L. (2007). A latent profile analysis of -->
-<!-- college students’ achievement goal orientation. Contemporary Educational Psychology 32(1): -->
-<!-- 8-47. -->
-<!-- Rosenberg, J.M., Beymer, P.N., Anderson, D.J., Van Lissa, C.J., & Schmidt, J.A. (2018). -->
-<!-- tidyLPA: An R Package to Easily Carry Out Latent Profile Analysis (LPA) Using Open-Source -->
-<!-- or Commercial Software. Journal of Open Source Software 3(30): 978. -->
-<!-- Scrucca, L., Fop M., Murphy, T.B., Raftery A.E. (2016) mclust 5: clustering, classification -->
-<!-- and density estimation using Gaussian finite mixture models The R Journal 8/1: 289-317. -->
-<!-- Shireman, E. M., Steinley, D., & Brusco, M. J. (2016). Local Optima in Mixture Modeling. -->
-<!-- Multivariate Behavioral Research 51(4): 466–481. -->
-<!-- Shireman, E., Steinley, D., Brusco, M.J. (2017). Examining the effect of initialization -->
-<!-- strategies on the performance of Gaussian mixture modeling. Behavioral Research Methods -->
-<!-- 49(1):282-293. -->
-<!-- Sterba, S.K. (2013). Understanding Linkages Among Mixture Models. Multivariate Behav -->
-<!-- Res. 2013 48(6): 775-815. -->
-<!-- Wolfe, J. (1970). Pattern clustering by multivariate mixture analysis. Multivariate Behavioral -->
-<!-- Research, 5: 329–350. -->
-
-<!-- Shireman 2016: -->
-<!-- https://www.tandfonline.com/doi/full/10.1080/00273171.2016.1160359?casa_token=YMvI52JWPq0AAAAA%3A20nlLRukoCkUmIil2lRWnsUS-33AlU6HmDRGsu07aJ0nTzS-skjvHnroJBtBVHM5HE0X8iStM1eyqw -->
-
-<!-- mixture models have two possible uses: either (a) classifying observations into groups through the estimation of distinct, underlying distributions or (b) using a set of well-defined parametric distributions (often times the Gaussian distribution) to approximate a messy data space that cannot be approximated by standard models that assume a single distribution (whether univariate or multivariate).
-While both are legitimate uses of mixture modeling, the only necessary and sufficient condition for fitting a mixture model is nonnormality in the data (see Bauer & Curran, 2003).
-It is certainly true that if there are distinct groups arising from distinct distributions, it is necessary to have a set of parameters to model each of those groups.
-Unfortunately, it may often be the case (in fact, it is probably likely) we are in the converse situation where the full data space is either nonlinear, asymmetric, or both, and there exist no distinct groups in the population.
-In this case, the mixture model is capturing the departure from “clean” structure, but the groups themselves are not meaningful in the sense that they are giving rise to distinct subpopulations.
-As such, their boundaries are going to be amorphous, and their parameters, regardless of their stability, lose meaning in terms of reflecting an underlying data-generative mechanism, making the interpretation of the means, covariances, or a specific individuals' membership an exercise in folly.
-Obviously, knowing which situation a data set falls in is paramount, and using the prevalence of local optima as a potential indicator is a crucial first step.
--->
-
-<!-- The optimal number of latent classes (i.e.
-identity configurations) was determined using the BIC (Fraley & Raftery, 1998); with lower values indicating a better statistical fit.
-Solutions were considered to be admissible if they consisted of at least 5% of the sample (Depaoli, 2013), if the entropy (a measure of class separability) exceeded .8 (Wang, Deng, Yang, 2017), and if the solution was theoretically interpretable and qualitatively distinct from the other class solutions (Howard & Hoffman, 2018).
--->
+<!-- insert the Daniel_LCA file from the example_analysis folder here -->
+
+\newpage
+
+# References

### 2022-10-16 example_analysis/Daniel_LCA.Rmd
Update 16 October
Changed title and short title. Adjusted abstract, Added keywords. Made the use of "LCA" and "mixture models" more consistent based on whether we refer to the whole group of LCA techniques, or a mixture subset.  Minor adjustment to wording throughout the paper. Shortened lines throughout. Capitalized titles and made them uniform. Adjusted section-subsection-subsubsection depth (# vs ## vs ###).  Removed irrelevant comments. Wrote best practices in visualization (##Visualization).  Added the tutorial (now set to eval = F for knitting, however code chunks do not show when using

@@
 Before we fit a series of LCA models, we set a random seed
-using set.seed().
+using `set.seed()`.
 The seed number can be any digit.
@@
 ```
-# Class enumeration
+# Class Enumeration
 
@@
 we want to compare a sequence of LCA models fitted to the data.
-To do so, we use `tidySEM::table_fit()` 
+To aid the process, we create a model fit table using `tidySEM::table_fit()` 
 with the results object as the input.
-This function creates the model fit table formatted for publication.
-However, it contains a lot of information on each of the four models.
-It may be more helpful to focus on a subset of model fit indices
-and classification diagnostics which we do in the subsequent step.
+As the output contains a lot of information on each of the four fitted models,
+we select a subset of helpful model fit indices and classification diagnostics.
 
@@
 fit_table <- table_fit(res) # model fit table
-fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # a selection from the model fit table
+fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # our selection
 ```
-The model fit table is important for deciding on the final class solution.
-Our subset of fit indices and classification diagnostics includes:
+Our selection of fit indices and classification diagnostics includes:
 
-**Name** indicates the $K$-class solution.
-In our case, we chose to fit 1 - 4 class solutions,
-each is represented by its row in the table.
-**LL** is the -2*log-likelihood of each model.
-**AIC** and **BIC** are the Akaike 
-and the Bayesian Information Criterion, respectively.
-**prob_min** and **prob_max** are 
-the lowest and the highest posterior class probability
-by most likely class membership.
-**n_min** and **n_max** are the lowest and the highest
-class proportion based on the posterior class probabilities.
+- **Name** = the $K$-class solution
+- **LL** = the -2*log-likelihood of each model
+- **AIC** = the Akaike Information Criterion value
+- **BIC** = the Bayesian Information Criterion value
+- **prob_min** = the lowest posterior class probability
+                 by most likely class membership
+- **prob_max** = the highest posterior class probability
+                 by most likely class membership
+- **n_min**  = the lowest class proportion based on the posterior class probabilities
+- **n_max** = the highest class proportion based on the posterior class probabilities
 
-In this paper, we discussed several possible strategies 
-to select the final class solution.
+We discussed several possible strategies to select the final class solution. 
+Here, we apply our own.
 
-Here, we apply our own. 
-First, we examine the -2*log likelihood 
-which falls successively with each added class.
+To aid our interpretation of the results, 
+we create an elbow plot showing the trends in information criteria across four models.
 
-Then, we look at the BIC value
-which is fairly close for the one, two, and three-class solutions,
-but high for the four-class solution.
+```{r elbowplot, message=F, warning=F}
+library(tidyverse) # for data-wrangling
+library(ggplot2) # for plots
 
+elbow_plot <- fit_table[ , c("Name", "AIC", "BIC")] # extract ICs
+elbow_plot %<>% pivot_longer(cols = c("AIC", "BIC"), names_to = "IC", values_to = "Value") # to long format
+
+ggplot(elbow_plot, aes(x = Name, y = Value, group = IC))+
+  geom_point(aes(color = IC))+
+  geom_line(aes(color = IC))+
+  labs(title="Elbow Plot of Information Criteria per Class Solution", x = "Class Solution", y = " Value")+
+  scale_color_manual(name = "Information Criterion", values = c(AIC = 'blue', BIC = 'red'))+
+  theme_minimal()
+
+```
+From the elbow plot, we see that AIC has a lower penalty for model complexity.
+However, we are more interested in the BIC values, which are similar
+for the one, two and three-class solutions, but the four-class solution fits significantly worse.
+For this reason, we eliminate the four-class solution from the selection process.
+
+Then we examine the model fit table.
+As expected, the -2*log likelihood falls successively with each added class.
 As previously stated, classification diagnostics should not be used for model selection,
@@
 We see that prob_min for the four-class solution is low,
-therefore we disqualify this solution.
+knowing that this solution also has a high BIC, we disqualify this solution.
 
@@
 and it has a satisfactory prob_min and n_min.
-Based on this, we can retain the three class solution in model selection.
+Based on this, we retain the three class solution in model selection.
 Note that entropy for the one-class solution will always equal to one,
@@
 Finally, when comparing the one and three-class solutions,
-we consult the BIC which tells us that the added complexity
+we inspect the information criteria.
+For BIC, the one-class solution fits better, but the difference is marginal.
+AIC tells us that the added complexity
 of having three classes still explains the data better
@@
 
-If we want to know the probability of each response option's endoresement 
+If we want to know the probability of each response option's endorsement 
 for each class, we can use `tidySEM::table_prob`.
@@
 ```{r}
-library(ggplot2)
 plot_prob(res[[3]]) # plotting the response patterns for the final class solution
@@
 ```
-In the plot, we can see the distributions each of the response probabilities on the indicators
+In the plot, we can see the distributions of the response probabilities on the indicators
 for each of the three classes.
-For instance, we see in Class 1 the most common response to u2 is 2,
+For instance, we see that in Class 1 the most common response to u2 is 2,
 while in Class 2 and Class 3 this is 0.
@@
 
-# Extracting posterior class probabilities
+# Extracting Posterior Class Probabilities
 

### 2022-10-16 manuscript.Rmd
Update 16 October
Changed title and short title. Adjusted abstract, Added keywords. Made the use of "LCA" and "mixture models" more consistent based on whether we refer to the whole group of LCA techniques, or a mixture subset.  Minor adjustment to wording throughout the paper. Shortened lines throughout. Capitalized titles and made them uniform. Adjusted section-subsection-subsubsection depth (# vs ## vs ###).  Removed irrelevant comments. Wrote best practices in visualization (##Visualization).  Added the tutorial (now set to eval = F for knitting, however code chunks do not show when using

@@
 ---
-title             : "Best Practices in Mixture Modeling using Free Open Source Software"
-shorttitle        : "BEST PRACTICES MIXTURE MODELING"
+title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
+shorttitle        : "BEST PRACTICES LCA"
 
@@
 abstract: |
-  Latent class analysis is a popular technique for identifying groups in data based
-  on a parametric model. Examples of this technique are known as mixture models, latent profile
-  analysis, latent class analysis, growth mixture modeling, and latent class
+  Latent class analysis is a family of techniques for identifying groups in data based
+  on a parametric model. Its examples include mixture models, latent profile
+  analysis, latent class analysis with ordinal indicators, and latent class
   growth analysis. Despite the popularity of this technique, there is limited
@@
   introduces new functionality for estimating and reporting mixture
-  models in the `tidySEM` R-package. These functions rely on estimation using
+  models in the `tidySEM` R-package whose backend relies on
   the `OpenMx` R-package.
 
-keywords          : "keywords"
-wordcount         : "X"
+keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
+wordcount         : "8206"
 bibliography      : "mixture.bib"
-figsintext        : no
+figsintext        : yes
 figurelist        : no
@@
 footnotelist      : no
-lineno            : yes
+lineno            : no
 mask              : no
@@
 output            : papaja::apa6_pdf
+
 ---
@@
 
-Latent class analysis (LCA) is an umbrella term that refers to a number of techniques for estimating unobserved group membership
+Latent class analysis (LCA) is an umbrella term 
+that refers to a number of techniques for estimating unobserved group membership
 based on a parametric model of one or more observed indicators of group membership.
-This method has become quite popular across scientific fields, and under a number of different names;
-most notably (finite Gaussian) mixture modeling and latent profile analysis.
-@vermunt_jk_latent_2004 defined it more generally as virtually any statistical model where “some of the parameters […] differ across unobserved subgroups”.
-
-Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
-While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines for a specific subcategory of LCA known as latent growth models,
-general reporting guidelines for latent class analysis are still lacking.
+The types of LCA have become quite popular across scientific fields,
+most notably finite Gaussian mixture modeling and latent profile analysis.
+@vermunt_jk_latent_2004 defined LCA more generally as virtually any statistical model 
+where “some of the parameters […] differ across unobserved subgroups”.
+
+Despite its popularity, there is a lack of standards for estimating and reporting LCA.
+While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines 
+for a specific type of LCA known as latent growth models,
+general reporting guidelines for all LCA are still lacking.
 This complicates manuscript review and assessment of the quality of published studies,
 and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
+The present paper seeks to address this gap in the literature 
+by suggesting updated guidelines for estimation and reporting of LCA, 
+based on current best practices.
 Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, 
-this paper exclusively relies on free, open source software for latent class analysis in R.
-Our goal is to make best-practices in latent class analysis widely accessible.
-
-## Defining latent class analysis
-
-Latent class analysis can be understood as a method for estimating unobserved groups
+this paper exclusively relies on free, open source software in R.
+Our goal is to make best-practices in LCA widely accessible.
+
+## Defining Latent Class Analysis
+
+Latent class analysis is a group of methods for estimating unobserved groups
 based on a parametric model of observed indicators of group membership.
-The concept of latent class analysis can be understood in different ways.
-Generally speaking, a mixture model assumes that the study population
+The concept of LCA can be understood in different ways.
+A mixture model assumes that the study population
 is composed of $K$ subpopulations or classes.
 It further assumes that the observed data are a mixture of data generated by class-specific models.
-The simplest univariate "model" is a normal distribution, which can be described with two parameters:
-the mean and the variance.
-Commonly, the same model is estimated across all classes, but with different parameters for each class (i.e., class-specific means and variances).
+The simplest univariate model is a normal distribution, 
+which can be described with two parameters: the mean and the variance.
+Commonly, the same model is estimated across all classes, 
+but with different parameters for each class (i.e., class-specific means and variances).
 Mixture modeling then estimates both the parameters for each class, 
@@
 
-As an illustrative example, imagine that a detective wants to know if it would be
-possible to use mixture modeling to identify the sex of a suspect,
+As an illustrative example, imagine that a detective wants to know 
+if it would be possible to use mixture modeling to identify the sex of a suspect,
 based on footprints found at the crime scene.
@@
 When estimating a two-class mixture model, the detective observes that the model estimates
-the mean shoe size of the two groups are equal to `r report(estparam[1], equals = FALSE)` and  `r report(estparam[2], equals = FALSE)`,
-which is close to the true means of the two groups, namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
+the mean shoe size of the two groups are equal to `r report(estparam[1], equals = FALSE)` 
+and  `r report(estparam[2], equals = FALSE)`,
+which is close to the true means of the two groups, 
+namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
 When tabulating estimated group membership against observed (known) group membership, 
@@
 
-A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
-One difference between the two techniques is that factor analysis can be considered as a way to group observed *variables* into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
-By contrast, mixture modeling groups *individuals* into classes [see @nylund-gibson_ten_2018].
-In line with this distinction, latent class analysis is sometimes referred to as a "person-centered" technique, and factor analysis as a "variable-centered" technique.
+A LCA is like confirmatory factor analysis, 
+except that the continuous latent variable is substituted with a categorical latent variable.
+One difference between the two techniques is that 
+factor analysis can be considered as a way to group observed *variables* into latent constructs, 
+and LCA groups *individuals* into classes.
+In line with this distinction, LCA is sometimes referred to as a "person-centered" technique, 
+and factor analysis as a "variable-centered" technique [@nylund-gibson_ten_2018].
 
 When the focus is on the model parameters in each group,
-then LCA can be thought of as similar to a multi-group structural equation model.
+LCA can be thought of as similar to a multi-group structural equation model.
 The main distinction is that group membership is not known a-priori,
-but is instead estimated - with measurement error - based on the data.
+but is instead estimated -- with measurement error -- based on the data.
 Whereas in a multi-group model, the data are split by group and treated as independent samples,
-in a mixture model, all cases contribute to the estimation of all parameters in all groups.
-The relative contribution of each case to the parameters of each group is determined by that case's posterior probability of belonging to that group.
+in LCA, all cases contribute to the estimation of all parameters in all groups.
+The relative contribution of each case to the parameters of each group
+is determined by that case's posterior probability of belonging to that group.
 
 When the focus is on each individual's estimated class membership,
-latent class analysis can be thought of as a type of clustering algorithm.
-In line with this perspective, mixture modeling is sometimes described as "model-based
+LCA can be thought of as a type of clustering algorithm.
+In line with this perspective, LCA is sometimes described as "model-based
 clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
@@
 *unsupervised classification* problem [@figueiredo_unsupervised_2002].
-The term *unsupervised* refers to the fact that the outcome variable, true class membership,
-is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome -- class membership.
-
-## A taxonomy of latent class analyses
-
-In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
-From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators [@vermunt_jk_latent_2004].
-Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
-The term "latent class analysis" seems most appropriate as an umbrella term for this broader class of models,
-as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
-Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.
-
-One common type of LCA is the *finite Gaussian mixture model*;
-a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
-The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
+The term *unsupervised* refers to the fact that the outcome variable 
+-- true class membership -- is not known, 
+and the term *classification* refers to the fact that 
+the algorithm is predicting a categorical outcome -- class membership.
+
+## A Taxonomy of Latent Class Analysis Methods
+
+In this paper, we use the term latent class analysis to refer to 
+a family of techniques that estimate latent class membership 
+based on a parametric model of observed indicators.
+From a historical perspective, the term was initially conceived to 
+refer to analyses with categorical, usually binary indicators [@vermunt_jk_latent_2004].
+Nowadays, there are a number of related techniques, 
+known by distinct names, that serve a similar purpose.
+The term "latent class analysis" seems most appropriate 
+as an umbrella term for this broader class of models,
+as it only refers to the purpose of the analysis, 
+and does not imply restrictions to the model used, 
+or the level of measurement of the indicators.
+Given the abundance of terms in use for closely related classes of models, 
+we will provide a rudimentary taxonomy of LCA methods.
+
+A common type of LCA is the *finite Gaussian mixture model*; 
+a univariate analysis where the observed distribution of a single variable 
+is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
+The parameters of a finite Gaussian mixture model are 
+the means and variances of these underlying normal distributions.
 The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
-In the multivatiate case, with more than one indicator variable,
-the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
+In the multivariate case, with more than one indicator variable,
+the parameters of a mixture model are the means, variances, and covariances 
+between the indicators (which can be standardized to obtain correlations).
 These parameters can be estimated freely, or set to be constrained across classes.
@@
 In some cases, such constraints will be inappropriate;
-for example, when the cohesion between indicators is expected to differ between classes.
-As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
-a class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
-From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.
+for instance when the cohesion between indicators is expected to differ between classes.
+As an example, a LPA of ocean plastic particles 
+found two classes of particles based on length and width:
+a class of smaller particles with a high correlation between length and width, 
+meaning that these particles were approximately round or square in shape, 
+and a class of larger particles with a low correlation between length and width, 
+meaning that these particles were heterogenous in shape.
+From a theoretical perspective, this makes sense, 
+because the smaller particles have been ground down to a more uniform shape by the elements.
 <!-- DA: what is the source for plastic particles? -->
@@
 It is also possible to estimate a mixture model based on latent indicators.
-This means that, within each class, one or more continuous latent variables are estimated based on the observed indicators.
-Categorical latent variable membership is then estimated based on these continuous latent variables.
-A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
-Examples of this approach include *growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
-These techniques estimate a latent growth model to describe individual trajectories over time.
-The growth mixture model is a latent class model where the parameters that indicate class membership 
-are the intercepts and variances (and typically covariances) of the latent growth variables, e.g., a latent intercept and slope.
+This means that, within each class, 
+one or more continuous latent variables are estimated based on the observed indicators.
+Categorical latent variable membership is then estimated 
+based on these continuous latent variables.
+A common application of this approach is in longitudinal research, 
+where the indicators reflect one construct assessed at different time points.
+Examples of this approach include *growth mixture models* (GMM) 
+and *latent class growth analyses* (LCGA).
+These techniques estimate a latent growth model
+to describe individual trajectories over time.
+The growth mixture model is a latent class model 
+where the parameters that indicate class membership 
+are the intercepts and variances, 
+and typically covariances of the latent growth variables, 
+e.g., a latent intercept and slope.
 This technique assumes that individuals within a class can have heterogenous trajectories.
-If the variance of the growth parameters is fixed to zero, it is known as a latent class growth analysis.
-This latter approach assumes that all individuals within a class share the same identical trajectory,
-and that any variance in the indicators not explained by the class-specific latent trajectories is due to residual error variance.
-
-The term latent class analysis originally referred to cases where the observed indicators were categorical.
+If the variance of the growth parameters is fixed to zero, 
+it is known as a latent class growth analysis.
+This latter approach assumes that all individuals within a class 
+share the same identical trajectory,
+and that any variance in the indicators not explained by the class-specific latent trajectories 
+is due to residual error variance.
+
+The term latent class analysis originally referred to cases 
+where the observed indicators were categorical.
 Nowadays, it is more commonly used as an umbrella term.
-To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as *latent class analysis with ordinal indicators*.
+To prevent ambiguity, the special case 
+where indicators are of binary or ordinal measurement level 
+might be described as *latent class analysis with ordinal indicators*.
 Latent class models with ordinal indicators are parameterized differently from mixture models.
-One common parameterization assumes that each categorical variable reflects an underlying standard normal distribution.
-The parameters are "thresholds" that correspond to quantiles of a standard normal distribution (with $N(\mu = 0, \sigma = 1)$).
-These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
+One common parameterization assumes that 
+each categorical variable reflects an underlying standard normal distribution.
+The parameters are "thresholds" that correspond to quantiles of a standard normal distribution 
+(with $N(\mu = 0, \sigma = 1)$).
+These thresholds are estimated based on 
+the proportion of individuals in each of the response categories of the indicator variable.
 For example, a binary indicator has a single threshold that distinguishes the two response categories.
@@
 This paper will primarily focus on mixture models and special cases thereof,
-although most of the suggested guidelines are applicable to all latent class analyses.
-
-## Use cases for latent class analysis
-
-<!-- CJ: It's good but for practical purposes it's preferable -->
-<!-- to use examples from Caspar's published papers. -->
-<!-- Don't spend too much time on this, -->
-<!-- Caspar will check if I can replace them with any of my papers. -->
-
-There are several use cases for which latent class analyses are suitable.
+although most of the suggested guidelines are applicable to all LCA methods.
+
+## Use Cases for Latent Class Analysis
+
+<!-- Caspar will check if I can replace them with examples from Caspar's published papers. -->
+
+There are several use cases for which LCA methods are suitable.
 One example is to test a theory that postulates the existence of a categorical latent variable.
-For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses.
-Latent class analysis can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
+For example, *identity status theory* posits that, at any given point in time, 
+adolescents reside in one of four identity statuses.
+Latent class analysis can be used to identify these four statuses 
+based on observed indicators (e.g., self-reported identity exploration and commitment).
 If results indicate that the data are better described by a different number of classes,
-or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, then the theory may be called into question.
-
-Another use case is unsupervised learning;
+or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, 
+then the theory may be called into question.
+
+Another use case is unsupervised learning:
 when the goal is to restore unobserved class membership based on observed indicators,
 or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
+For example, a mixture model can be used as a diagnostic aid 
+when several clinical indicators can be used to distinguish between 
+a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
 The example of shoe size is a rudimentary illustration of this type of application.
 
-LCA can be used as a descriptive analysis where a researcher wishes to describe their sample and identify a few prototypes based on many variables.
+LCA can be used as a descriptive analysis 
+where a researcher wishes to describe their sample 
+and identify a few prototypes based on many variables.
 As an example, if a survey among all Dutch academics was carried out
@@
 For instance, if we believe that a variable represents a group, 
-but don't know how many groups there are, LCA may be an appropriate technique to answer this question.
-For example, @hopfer_social_2014 studied the substance use, sexual behavior, and mental health status of urban population in Winnipeg, Canada. 
+but don't know how many groups there are, 
+LCA may be an appropriate technique to answer this question.
+For example, @hopfer_social_2014 studied substance use, sexual behavior, 
+and mental health status of urban population in Winnipeg, Canada. 
 The underlying assumption was that there were different risk profiles,
 but their number was not known.
-From a collection of indicators, the LCA provided evidence that there may be four distinct risk profiles in the Winnipeg area.
+From a collection of indicators, LCA provided evidence 
+that there are four distinct risk profiles in the Winnipeg area.
 
 Another application of LCA is to classify individuals.
-In a peer harassment study, @giang_using_2008 used latent class analysis to classify over 2,000 sixth grade students into aggressor and victim latent classes.
-The five-class solution comprised of classes of victims, aggressors, and socially adjusted students.
-For instance, it revealed that there were two types of victims: highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
+In a peer harassment study, 
+@giang_using_2008 used LCA to classify over 2,000 sixth grade students 
+into aggressor and victim latent classes.
+The five-class solution comprised of classes of 
+victims, aggressors, and socially adjusted students. 
+For instance, it revealed that there were two types of victims: 
+highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
 
 LCA is also appropriate when we wish to identify indicators that capture classes well.
-High quality indicators are strongly related to the latent variable
-and lead to good class separation.
+High quality indicators will be strongly related to the latent variable
+and will lead to good class separation.
 This relationship of high quality indicators to the latent variable
 is reflected in very high or very low conditional response probabilities.
-<!-- As a rule of thumb, high indicator quality can be indicated  -->
-<!-- by conditional response probabilities above .9 and below .1.  -->
-<!-- Moderate indicator quality can be indicated by probabilities above .8 and below .2, -->
-<!-- and low indicator quality would be above .7 and below .3. <- DA's writing -->
-<!-- This quality rule of thumb is from @geiser_is_2014 where they state that -->
-<!-- "Indicators in high quality conditions were generated to either have CRPs (of endorsing the second category) of 0.9 or 0.1, while moderate quality conditions had CRPs of 0.8 and 0.2, and low quality indicators were generated at 0.7 or 0.3, following Collins and Wugalter’s (1992) strong and weak measurement strength conditions." direct quote from @geiser_is_2014 -->
 For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
@@
 From this, a theory about the selection of indicators could be informed.
-
-<!-- "However, Galindo-Garre and Vermunt (2006) found that indicators with high population values (i.e., conditional response probabilities close to one) were more likely to produce boundary estimates in LCA applications, even though this effect decreased with larger N.
-Still, there is evidence that using high quality indicators is generally beneficial, at least in the context of structural equation models with continuous latent variables (Marsh et al., 1998).
-Having more high quality indicators can stabilize the estimation by increasing the information available to estimate latent variable parameters.
-In this vein, Collins and Wugalter (1992) speculated that having sufficiently high quality indicators may compensate for having few indicators and may aid parameter recovery in LTA models." direct quote from @geiser_is_2014 -->
 
@@
 hence these covariates were excluded from the analysis.
-Accuracy scores were related to probabilities of being in anxiety and attention-
-anxiety classes and therefore this covariate was kept as a valuable predictor.
+Accuracy scores, however, were related to probabilities of 
+being in anxiety and attention-anxiety classes. 
+Therefore this covariate was kept as a valuable predictor.
 
@@
 LCA can be used to construct latent classes as categorical predictors.
-@lanza_latent_2013 demonstrated how LCA can be used to classify adolescents into depression classes,
+@lanza_latent_2013 demonstrated how LCA can be used 
+to classify adolescents into depression classes,
 and subsequently these classes can be used to predict smoking, grades, and delinquency.
-The study showed that the outcomes predicted by class membership can be binary (regular smoking), 
-continuous (grades) or count (delinquency).
+The study showed that the outcomes predicted by class membership 
+can be binary (regular smoking), continuous (grades) or count (delinquency).
 
@@
 as the resulting groups summarize response patterns on a large number of indicators.
-For example, @macgregor_symptom_2021 investigated symptom profiles among injured U.S. military personnel. 
-They used fifteen dichotomous items from the Post-Deployment Health Assessment survey as LCA indicators.
-Combinatorics informs us that fifteen dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
-Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a method of dimensionality reduction.
-A five class solution was found to have the best fit according to both statistical criteria and clinical interpretability.
-
-<!-- from @macgregor_symptom_2021 "Incorporating LCA with the symptom checklist was a novel method of dimensionality -->
-<!-- reduction, and the identified symptom profiles provided more granularity than examining individual symptoms -->
-<!-- alone.
-There are also limitations that warrant mention.
-At times, LCA can be prone to classification errors, but in the present analysis, the model entropy value above 0.80 indicated good separation of the classes, and the MeanPP values showed a low rate of misclassification." -->
-<!-- [...] -->
-<!-- [listing the five classes] "(1) 50.4% were low -->
-<!-- morbidity (i.e., no or very few symptoms reported), (2) -->
-<!-- 14.0% musculoskeletal (i.e., muscle aches; swollen, stiff or -->
-<!-- painful joints; back pain), (3) 8.9% auditory (i.e., trouble -->
-<!-- hearing; ringing in the ears), (4) 11.1% psycho-cognitive -->
-<!-- (i.e., trouble concentrating, easily distracted; forgetful or -->
-<!-- trouble remembering things; increased irritability), and (5) -->
-<!-- 15.6% multimorbidity (i.e., multiple symptoms reported)." -->
+For example, @macgregor_symptom_2021 investigated symptom profiles 
+among injured U.S. military personnel. 
+They used fifteen dichotomous items 
+from the Post-Deployment Health Assessment survey as LCA indicators.
+Combinatorics informs us that fifteen dichotomous items have 
+$2^{15}$ or $32,768$ unique symptom combinations.
+Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA 
+as a method of dimensionality reduction.
+A five class solution was found to have the best fit 
+according to both statistical criteria and clinical interpretability.
 
@@
 As discussed in the shoe size example, LCA can deal with violations of normality.
-In fact, LCA assumes the population distribution in a non-normal mixture of $K$ normal distributions,
+In fact, LCA assumes the population distribution is a non-normal mixture 
+of $K$ normal distributions,
 and it can discover the value of $K$, i.e. generate a $K$-class solution.
 
-<!-- CJ: Include these use cases too: -->
-<!-- Descriptive analysis; identify few prototypes based on many variables, e.g. DJA science funding -->
-<!-- Inductively identify number of latent classes: I know it's a group, but not how many groups -->
-<!-- Classify individuals, e.g. welzijnsmeter -->
-<!-- Identify items (observed characteristics) that -->
-<!-- indicate these classes well -->
-<!-- Predict class membership from covariates -->
-<!-- Predict outcomes from class membership -->
-<!-- Dimension reduction -->
-<!-- Deal with violation of assumptions, e.g. when data are clearly not normally distributed -->
-
-<!-- In medical and psychological science, LPA can be useful when -->
-<!-- considerable between-subject heterogeneity exists in scores on a range of variables and when -->
-<!-- this variation cannot be explained by known, manifest variables (e.g., Wolfe 1970; Sterba -->
-<!-- 2013). Here, LPA can help to identify or approximate possibly meaningful sub-grouping of -->
-<!-- subjects that may help to better understand sample heterogeneity (Sterba 2013). -->
-<!-- Generally, LPA works under the assumption that sample (residual) variance can be reduced -->
-<!-- by assuming a categorical latent variable that effectively subdivides the sample into >=2 -->
-<!-- subgroups that are more homogeneous in terms of their patterns of variable means and -->
-<!-- (co)variances. In case an LPA model is found to fit a dataset well, it is often the case that -->
-<!-- subjects in each class resemble each other relatively well in terms of their scores on the input -->
-<!-- variables. Depending on the model configuration, the identified classes can show different -->
-<!-- class-specific patterns of means and class-specific or class-varying variances. -->
-
-<!--                                                1 -->
-<!-- A downside is that these models are more complex (much more parameters to be estimated). -->
-<!-- Using criteria, such as the Bayesian Information criterion (BIC) helps to find a model that -->
-<!-- strikes a good balance between model-fit and model-complexity. When doing an LPA, most -->
-<!-- applied researchers will usually be most interested in the differences and/or overlap between -->
-<!-- the classes’ specific patterns of parameter estimates. These can be used to characterize the -->
-<!-- classes and, possibly, provide clues about underlying mechanisms (Sterba 2013). -->
-
-<!-- This tutorial -->
-<!-- LPA is less widely used than other latent variable models and, possibly due to that, has -->
-<!-- long been only available in specialized software packages such as Mplus. Luckily, ongoing -->
-<!-- developments in many different scientific fields (e.g., ecology, econometrics) have yielded a -->
-<!-- number of packages that also allow users to conduct LPA in the open-source R-platform. -->
-<!-- However, the use of R does require experience and documentation of packages can be rather -->
-<!-- limited or technical, making it less easily accessible for applied researchers. Therefore, this -->
-<!-- tutorial is aimed to help applied researchers to get going with LPA in R, illustrating the use -->
-<!-- of several packages and, for reference, providing a comparison of the results obtained in R -->
-<!-- with results obtained with Mplus. -->
-
-# Best practices
-
-## In estimation
+# Best Practices
+
+## Best Practices in Estimation
 
 The best practices in estimation, as outlined in Table \@ref(tab:checkest),
-are rooted in existing recommendations for best practices for estimating
-specific sub-types of latent class analyses,
+are rooted in existing recommendations for best practices 
+for estimating specific subtypes of LCA,
 including latent class growth analysis [@van_de_schoot_grolts-checklist_2017]
 and latent class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
-These were generalized to be more relevant to all types of latent class analyses,
+These were generalized to be relevant to all types of LCA,
 and updated to current best practices, as explained below.
 
-```{r checkest, include=FALSE}
+```{r checkest, include=TRUE}
 tab_check <- data.frame(Item = c(
-  "Examine observed data",
-  "Handling missing data",
-  "Alternative model specifications",
+  "Examining Observed Data",
+  "Handling Missing Data",
+  "Alternative Model Specifications",
   "Software",
-  "Best practices in estimation",
   "Algorithm",
-  "Class enumeration",
+  "Class Enumeration",
   "Model Fit Indices",
   "Classification Diagnostics",
-  "Interpreting Class Solution",
-  "Label switching",
-  "Best Practices in Reporting",
-  "Tutorial"
+  "Interpreting Class Solutions",
+  "Label switching"
 
@@
 
-### Examine observed data
+### Examining Observed Data
 
 Examining observed data is essential for any analysis
-as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
-Special attention should be paid to level of measurement of the indicators.
-Finite Gaussian mixture models (including LPA) are only suitable for continuous variables.
-Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
-Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
-suggest that the misapplication of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
-Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @norman_likert_2010],
+as it may reveal patterns and violations of assumptions 
+that had not been considered prior to data collection.
+Special attention should be paid 
+to level of measurement of the indicators.
+Finite Gaussian mixture models (including LPA) 
+are only suitable for continuous variables.
+Indicators with an ordinal level of measurement 
+are likely to violate the assumption of 
+within-class normal distributions of mixture models
+[see @vermunt_k-means_2011].
+Personal experience consulting on LCA methods and 
+moderating the `tidyLPA` Google group
+suggest that the misapplication of mixture models 
+to ordinal (e.g., Likert-type) indicators 
+is the most common source of user error.
+Whereas it has been argued that some parametric methods are robust 
+when scales with 7+ indicators are treated as continuous 
+[e.g., @norman_likert_2010],
 this certainly does not imply that all methods are.
-It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
-The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; 
+It is certainly unlikely that such ordinal variables 
+can be treated as a *mixture* of multiple normal distributions.
+The problem becomes egregious when 
+the number of classes estimated equals or exceeds the number of categories; 
 in this case, each class-specific mean could describe a single response category, 
@@
 
-Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [@spurk_latent_2020].
+Relatedly, a recent publication claimed that 
+an assumption of mixture models is that 
+observed indicators are normally distributed [@spurk_latent_2020].
 This is incorrect.
-When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
-In our shoe size example, it can be seen that the population distribution comprised of two normal distributions.
+When the number of classes is greater than one, 
+mixture models assume that the observed indicators 
+are a mixture of multiple (multivariate) normal distributions.
+In our shoe size example, it can be seen that the population distribution
+is comprised of two normal distributions.
 When examined visually, the population distribution is evidently bimodal. 
@@
 the sample comes from a normally distributed population.
-Yet, this is a prototypical example of a mixed population distribution where LCA can discover latent groups.
+Yet, this is a prototypical example of a mixed population distribution 
+where LCA can discover latent groups.
 If the population distribution were instead normal, 
@@
 
-Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function `tidySEM::descriptives(data)`.
-Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
-Plots (density plots for continuous variables, and bar charts for categorical ones) may be more diagnostic.
-Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible [consider using a GitHub repository as a comprehensive public research archive, as explained in @van_lissa_worcs_2021].
-
-### Handling missing data
-
-Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled [@van_de_schoot_grolts-checklist_2017].
-Three types of missingness have been distinguished in the literature [@rubin_inference_1976]:
-Missing completely at random (MCAR), which means that missingness is random;
-missing at random (MAR), which means that missingness is contingent on the *observed* data (and can thus be accounted for);
-and finally missing not at random (MNAR), which means that missingness is related to unobserved factors.
+Extensive descriptive statistics 
+(including the number of unique values, 
+variance of categorical variables, and missingness; see next subsection)
+can be obtained using the function `tidySEM::descriptives(data)`.
+Note, however, that sample-level descriptive statistics 
+are of limited value when the goal of a study is 
+to identify subsamples using latent class analysis.
+Plots (density plots for continuous variables, 
+and bar charts for categorical ones) may be more diagnostic.
+Note that density plots can also aid in the choice of the number of classes, 
+as further explained in the section on visualization.
+Descriptive statistics and plots can be relegated to online supplements, 
+provided that these are readily accessible 
+[consider using a GitHub repository as a comprehensive public research archive, 
+as explained in @van_lissa_worcs_2021].
+
+### Handling Missing Data
+
+Previous work has emphasized the importance of examining 
+the pattern of missing data and 
+reporting how missingness was handled 
+[@van_de_schoot_grolts-checklist_2017].
+Three types of missingness have been distinguished
+in the literature [@rubin_inference_1976]:
+Missing completely at random (MCAR), 
+which means that missingness is random;
+missing at random (MAR), 
+which means that missingness is contingent on the *observed* data
+(and can thus be accounted for);
+and finally missing not at random (MNAR), 
+which means that missingness is related to unobserved factors.
 It is possible to conduct a so-called "MCAR" test, 
@@
 as the null-hypothesis of this test is that the data are not MAR,
-and a significant test statistic indicates that missingness is related to the observed data (MAR).
+and a significant test statistic indicates that 
+missingness is related to the observed data (MAR).
 A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@little_test_1988].
-This assumption is tenuous in the context of latent class analysis.
-A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable [@jamshidian_tests_2010].
-Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
-For this tutorial, I have re-implemented it in the `mice` package as `mice::mcar()`,
+As Little's classic MCAR test relies on 
+the comparison of variances across groups 
+with different patterns of missing data, 
+it assumes normality [@little_test_1988].
+This assumption is tenuous in the context of LCA.
+A non-parametric MCAR test, 
+as provided by Jamshidian and Jalal, 
+may be more suitable [@jamshidian_tests_2010].
+Unfortunately, this test was removed from the central R-repository CRAN 
+due to lack of maintenance.
+For this tutorial, I have re-implemented it in the `mice` package 
+as `mice::mcar()`,
 with a fast backend in C++ and new printing and plotting methods.
 
-While we concur that investigating missingness is due dilligence,
-it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, 
+While we concur that investigating missingness is due diligence,
+it is important to emphasize that 
+missingness is adequately handled by default in many software packages for LCA, 
 such as Mplus, and `OpenMx` which is the backend of `tidySEM`.
@@
 which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data; on par with multiple imputation [@lee_comparison_2021].
+FIML is a best-practice solution for handling missing data; 
+on par with multiple imputation [@lee_comparison_2021].
 FIML estimation assumes that missingness is either MCAR or MAR.
-Thus, one would typically proceed with FIML regardless of the outcome of an MCAR test.
-Although FIML does not, by default, handle missingness in exogenous variables - all indicator variables in latent class analysis are endogenous, so this is not a concern.
-
-Multiple imputation is less suitable to latent class analyses for two reasons.
-First, because latent class analyses are often computationally expensive,
+Thus, one would typically proceed with FIML, 
+regardless of the outcome of an MCAR test.
+Although FIML does not, by default, handle missingness in exogenous variables
+-- all indicator variables in LCA are endogenous, 
+so this is not a concern.
+
+Multiple imputation is less suitable to LCA for two reasons.
+First, because LCA methods are often computationally expensive,
 and conducting them on multiple imputed datasets may be unfeasible.
-Second, because there is no straightforward way to integrate latent class analysis results across multiple datasets.
-To conclude; our recommendation is to inspect missingness (e.g., using `mice::MCAR()`)
-and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
+Second, because there is no straightforward way 
+to integrate LCA results across multiple datasets.
+To conclude; our recommendation is to inspect missingness 
+(e.g., using `mice::MCAR()`)
+and report the proportion of missingness per variable 
+(e.g., using `tidySEM::descriptives()`),
 before proceeding with FIML.
@@
 
-### Alternative model specifications
+### Alternative Model Specifications
 
 In order to aid researchers working with latent trajectory models,
-@van_de_schoot_grolts-checklist_2017 developed a protocol called Guidelines 
-for Reporting on Latent Trajectory Studies (GRoLTS).
-Two of the GRoLTS guidelines (namely, 6a and 6b) refer to considering alternative model specifications.
+@van_de_schoot_grolts-checklist_2017 developed a protocol called 
+Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
+Two of the GRoLTS guidelines (namely, 6a and 6b) refer to 
+considering alternative model specifications.
 Both are discussing specific cases in latent trajectory model specification. 
-The first is about whether the variance of the growth parameter is estimated freely or fixed,
-and the second is about whether conditional independence is assumed
+The first is about whether the variance of the growth parameter 
+is estimated freely or fixed,
+and the second is about whether conditional independence is assumed,
 as the researcher might also want to free-up the variance-covariance structure.
@@
 typically have class-specific means, variances, and covariances.
-Latent growth analyses have the same parameters, but with respect to the latent growth variables.
+Latent growth analyses have the same parameters, 
+but with respect to the latent growth variables.
 Latent class analyses with ordinal indicators have thresholds.
-All of these parameters can be freely estimated across classes, constrained to be equal across classes,
+All of these parameters can be freely estimated, 
+constrained to be equal across classes,
 or fixed to a certain value (e.g., to zero).
-The total number of parameters thus scales with the number of estimated classes.
-Consequently, latent class analyses have a potentially very high number of parameters.
+The total number of parameters thus 
+scales with the number of estimated classes.
+Consequently, LCA methods have a potentially very high number of parameters.
 As any of these parameters could be misspecified,
 it is important to consider alternative model specifications.
-However, alternative model specifications may be approached differently depending on whether an analysis is data driven (exploratory), or theoretically driven (confirmatory).
+However, alternative model specifications may be approached differently 
+depending on whether an analysis is data driven (exploratory), 
+or theory driven (confirmatory).
 This distinction has remained underemphasized in prior writing.
 
-Prior literature on latent class analysis has emphasized exploratory applications of the method [see @nylund_deciding_2007].
-In exploratory analyses, a large number of models are typically estimated in batch,
+Prior literature on LCA has emphasized its exploratory applications 
+[@nylund_deciding_2007].
+In exploratory LCA, 
+a large number of models are typically estimated in batch,
 with varying numbers of classes and model specifications.
-The "correct" model specification is then determined based on a combination of fit indices,
+The "correct" model specification is then determined 
+based on a combination of fit indices,
 significance tests, and interpretability.
-For latent profile analysis, the function `tidySEM::mx_profiles(classes, variances, covariances)`
+For latent profile analysis, 
+the function `tidySEM::mx_profiles(classes, variances, covariances)`
 largely automates this process.
-The argument `classes` indicates which class solutions should be estimated (e.g., 1 through 6).
-The argument `variances` specifies whether variances should be `"equal"` or `"varying"` across classes.
-The argument `covariances` specifies whether covariances should be constrained to `"zero"`, `"equal"` or `"varying"` across classes.
-The means are free to vary across classes by default, although the more general function `tidySEM::mx_mixture()` could be used to circumvent this.
-After all models have been estimated, the function `tidySEM::table_fit()` can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
-Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
+The argument `classes` indicates which 
+class solutions should be estimated (e.g., 1 through 6).
+The argument `variances` specifies whether 
+variances should be `"equal"` or `"varying"` across classes.
+The argument `covariances` specifies whether 
+covariances should be constrained to 
+`"zero"`, `"equal"` or `"varying"` across classes.
+The means are free to vary across classes by default, 
+although the more general function `tidySEM::mx_mixture()` 
+could be used to circumvent this.
+After all models have been estimated, 
+the function `tidySEM::table_fit()` can be used 
+to obtain a model fit table suitable for 
+determining the optimal model according to best practices.
+Note, however, that this table does not include 
+the bootstrapped likelihood ratio test (BLRT) by default,
 because this test is very computationally expensive.
-It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
-More on fit indices can be found in the Model fit indices subsection of this paper.
-
-Confirmatory analyses typically require less comprehensive alternative model specifications.
-For example, in the context of preregistered analyses, 
-the main models of interest may have been specified a priori. 
-Even in case of confirmatory LCA, the theoretical model could be compared to a few others to contextualize it.
+It is recommended to use the function `tidySEM::BLRT()` 
+to compare a shortlist of likely candidate models 
+based on other fit indices. 
+More on fit indices can be found in 
+the Model Fit Indices subsection of this paper.
+
+Confirmatory LCA typically requires 
+less comprehensive alternative model specifications.
+For example, in the context of a preregistered analysis, 
+the main models of interest would have been specified a priori. 
+Even in this case, 
+the theoretical model could be compared to a few others to contextualize it.
 
@@
 
-## Software
-
-Many software packages are available for the estimation of latent class analyses.
-Some of these packages have limited functionality, or implement specific innovations.
-Other packages implement latent class analyses in the context of a more flexible structural equation modeling framework.
-The most notable examples of the latter are the commercial programs Mplus and Latent GOLD, 
+### Software
+
+Many software packages are available for the estimation of LCA models.
+Some of these packages have limited functionality, 
+or implement specific innovations.
+Other packages implement LCA in the context of 
+a more flexible structural equation modeling framework.
+The most notable examples of the latter are 
+the commercial programs Mplus and Latent GOLD, 
 and the free open source R-package OpenMx.
-The commercial packages stand out because they offer relatively user-friendly interfaces
-and implement sensible defaults for complex analyses, including latent class analysis.
+The commercial packages stand out 
+because they offer relatively user-friendly interfaces
+and implement sensible defaults for complex analyses, 
+including LCA methods.
 This lowers the threshold for applied researchers to adopt such methods.
 Commercial software also has several downsides, however.
-One such downside is that use of the software is restricted to those individuals and institutions who can afford a license.
-A second downside is that the source code, being proprietary, cannot be audited, debugged, or enhanced by third parties.
-This incurs the risk that mistakes in the source code may go unnoticed, and curbs progress as software developers cannot add new functionality.
+One such downside is that use of the software 
+is restricted to those individuals 
+and institutions who can afford a license.
+Another downside is that the source code, 
+being proprietary, cannot be audited, debugged, 
+or enhanced by third parties.
+This incurs the risk that mistakes in the source code 
+may go unnoticed, and curbs progress as 
+software developers cannot add new functionality.
 
 Conversely, the free open source program OpenMx is very flexible,
-but not very user-friendly. We directly address this limitation using the `tidySEM` R-package.
-New functionality in `tidySEM` seeks to lower the threshold for latent class analysis using `OpenMx`.
-It adheres to best practices in estimation and reporting, as described in this paper.
-The user interface is simple, making use of the model syntax of the widely used `lavaan` R-package.
+but not very user-friendly. 
+We directly address this limitation using the `tidySEM` R-package.
+New functionality in `tidySEM` seeks to lower the threshold 
+for latent class analysis using `OpenMx`.
+It adheres to best practices in estimation and reporting, 
+as described in this paper.
+The user interface is simple, 
+making use of the model syntax of the widely used `lavaan` R-package.
 This syntax offers a human-readable way to specify latent variable models.
-Minor enhancements are made to simplify the specification of latent class analysis.
-
-Because of the limitations of the aforementioned tools, we set out to develop a free tool that
-provides sensible defaults and is easy to use, but provides the option to
-access and modify all of the model inputs (i.e., low barrier, high
-ceiling). `tidySEM` interfaces with existing tools, and is able to translate between
-what existing tools are capable of and what researchers and analysts
+Minor enhancements are made to simplify the specification of LCA.
+
+Because of the limitations of the aforementioned tools, 
+we set out to develop a free tool that
+provides sensible defaults and is easy to use, 
+but provides the option to access and modify 
+all of the model inputs (i.e., low barrier, high ceiling). 
+`tidySEM` interfaces with existing tools, 
+and is able to translate between what existing tools are capable of 
+and what researchers and analysts
 carrying-out person-oriented analyses would like to specify. 
-Furthermore,`tidySEM` facilitates fully-reproducible analyses and contributes to open science.
-
-<!-- 3. Packages -->
-<!-- 3.1 R-packages -->
-<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
-<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
-<!-- analytic functionality. However, the majority of packages is focused on analyses with discrete -->
-<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
-<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
-<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
-<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
-<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
-<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
-<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
-<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
-<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
-<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
-<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->
-
-<!-- Although ‘tidyLPA’ is easy to use, this comes at the expense of restricting modeling options -->
-<!-- to a few, oft-used options. For completeness, both the ‘mclust’ and ‘tidyLPA’ approach will -->
-<!-- be illustrated. -->
-
-## Best practices in estimation
+Furthermore,`tidySEM` facilitates fully-reproducible analyses 
+and contributes to open science.
 
@@
 
-Mixture model parameters and model fit statistics can be estimated in a variety of ways.
-The choice of the estimator depends on the presence of missing values,
-sample size, number of indicators, and available computational resources [@weller_latent_2020].
+LCA parameters and model fit statistics 
+can be estimated in a variety of ways.
+The choice of the estimator 
+depends on the presence of missing values,
+sample size, number of indicators, 
+and available computational resources [@weller_latent_2020].
 A commonly used technique is maximum likelihood (ML) estimation
 with the expectation-maximization (EM) algorithm as a local optimizer. 
-Imagine we are estimating two parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator
-(ignoring the variance for now). The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes the likelihood ($LL$) of all observed data. 
-In practice, instead of maximizing $LL$, often $-2*LL$ is minimized, as this offers computational advantages. 
-We can think of this optimization problem as a three-dimensional landscape: 
+Imagine we are estimating two parameters, 
+e.g. the class-specific means $\mu_c$ on a continuous indicator
+(ignoring the variance for now). 
+The EM algorithm will attempt to find a combination of values
+for these two parameters that maximizes
+the likelihood ($LL$) of all observed data. 
+In practice, instead of maximizing $LL$, 
+often $-2*LL$ is minimized, 
+as this offers computational advantages. 
+We can think of this optimization problem 
+as a three-dimensional landscape: 
 The X and Y dimensions are determined by the class-specific means, 
-so $X = \mu_1$ and $Y = \mu_2$ - and the Z-dimension is determined by $Z = -2*LL$. 
+so $X = \mu_1$ and $Y = \mu_2$ - 
+and the Z-dimension is determined by $Z = -2*LL$. 
 The optimizer must find the deepest "valley" in this landscape, 
-which reflects the combination of $\mu_1$ and $\mu_2$ that maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble, dropped in this landscape.
-It is dropped at some random point in space, and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
+which reflects the combination of $\mu_1$ and $\mu_2$ that 
+maximizes the likelihood of the data.
+The EM optimizer behaves somewhat like a marble dropped in this landscape.
+It is dropped at some random point in space, 
+and will roll into the nearest valley.
+The problem is that, once EM rolls into a valley, 
+it will settle on the bottom of that valley
 (this is known as "convergence").
 It cannot climb out again.
-Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley 
-(a "local optimum"), and never discovers the deepest valley (the "global optimum", or best solution).
-One solution to this problem is to drop many marbles at random places, compare their final $-2*LL$ values,
-choose the solution with the lowest $-2*LL$, and make sure that several marbles replicated this solution.
+Thus, if their are multiple valleys, 
+the risk is that the optimizer gets stuck in a shallower valley (a "local optimum"), 
+and never discovers the deepest valley (the "global optimum", or best solution).
+One solution to this problem is to drop many marbles at random places, 
+compare their final $-2*LL$ values,
+choose the solution with the lowest $-2*LL$, 
+and make sure that several marbles replicated this solution.
 This is the "random starts" approach.
@@
 a "reasonable solution" may be used for the starting values.
-For example, if we assume that the different classes are likely to have different mean values on the indicators, 
-then the K-means clustering algorithm can be used to determine these cluster centroids.
-We can compute the expected values of all model parameters by treating the K-means solution as a known class solution,
+For example, if we assume that the different classes are likely to have 
+different mean values on the indicators, 
+then the K-means clustering algorithm can be used to 
+determine these cluster centroids.
+We can compute the expected values of all model parameters 
+by treating the K-means solution as a known class solution,
 and use these as starting values for a mixture model.
-<!-- In many circumstances and especially in a cluster analysis setting, the mixture component means are expected to be different.
-Thus, a reasonable and largely employed way of initiating EM consists of starting from the solution of a K-means type algorithm.
-[@biernackiChoosingStartingValues2003].
--->
-One remaining concern is that this approach may result in starting values close to a local optimum,
+One remaining concern is that this approach may result 
+in starting values close to a local optimum,
 and that the EM algorithm will thus never find the global optimum.
 A second innovation addresses this concern.
-
-Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
+Instead of using EM, 
+it is possible to use an optimizer that can climb out of a valley.
 Simulated annealing iteratively considers some "destination" in the landscape, 
-and compares its likelihood to the current one. If the destination likelihood is higher, 
+and compares its likelihood to the current one. 
+If the destination likelihood is higher, 
 the estimator moves there.
-If the destination likelihood is *lower*, the estimator still moves there occasionally, based on probability.
-This latter property allows it to escape local optima, and find the global optimum.
-
-By default, `tidySEM` employs this solution of deriving starting values using K-means clustering,
-and identifying the global optimum solution using simulated annealing.
+If the destination likelihood is *lower*, 
+the estimator still moves there occasionally, 
+based on probability.
+This latter property allows it to escape local optima,
+and find the global optimum.
+
+By default, `tidySEM` derives starting values using K-means clustering,
+and identifies the global optimum solution using simulated annealing.
 Once a solution has been found,
 simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for the parameters that can be used to compute standard errors.
+as EM inherently produces an asymptotic covariance matrix for 
+the parameters that can be used to compute standard errors.
 Note that these defaults can be manually overridden.
 
-One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
-This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
-As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
+A recent paper suggested maximum likelihood with robust standard errors 
+should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
+This statement is incorrect, and may lead readers to believe that 
+they must use commercial software, 
+as robust maximum likelihood is currently only implemented 
+in Mplus and latentGOLD.
+As explained before, 
+mixture modeling assumes that observed data are 
+a mixture of (multivariate) normal distributions; 
+thus, the observed indicators will likely not be normally distributed.
 <!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->
 
-<!-- We assume that the population consists of several groups each characterized by its own normal distribution. -->
-<!-- When two or more normal distributions with different means and standard deviations comprise a mixed population distribution, -->
-<!-- this combined distribution is no longer normal. -->
-<!-- For example, you can imagine a bimodal distribution when dealing with two latent classes. -->
-<!-- If this were not the case and our mixed population distribution were normal, -->
-<!-- there would be no latent classes to extract  -->
-<!-- as the population would consists of a single class only. -->
-<!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups. -->
-<!-- Secondly, the claim that maximum likelihood with robust standard errors should be preferred -->
-<!-- implicitly implies that only software packages with this estimation technique can be used. -->
-
-<!-- ### Starting values -->
-
-<!-- The EM algorithm needs starting values for each parameter it wants to estimate. -->
-<!-- If it were provided a single starting value for each parameter, -->
-<!-- the EM optimizer would be at risk of finding a local optimum instead of the global optimum. -->
-<!-- In other words, it would converge on a suboptimal solution -->
-<!-- resulting in biased parameter and model fit estimates. -->
-<!-- To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter -->
-<!-- in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004). -->
-<!-- Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values. -->
-
-<!-- Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012). -->
-
-<!-- Furthermore, the choice of the starting values  -->
-<!-- can impact the speed of the algorithm's convergence -->
-<!-- or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004). -->
-<!-- The key is to use a large number of starting value sets. -->
-<!-- Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, -->
-<!-- and a greater number of sets for more complex models. -->
-<!-- A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately. -->
-
-### Class enumeration
-
-LCA can be done in an exploratory or in a confirmatory fashion.
+### Class Enumeration
+
+As explained, LCA can be done in an exploratory or in a confirmatory fashion.
 In exploratory LCA, a sequence of models is fitted to the data 
 with each additional model estimating one more class than the previous model.
-These models are then compared and the best solution is selected as the final class solution.
-In some cases, prior theory can inform the researcher about the number of classes to expect.
-Even in such confirmatory LCA cases, it is nonetheless useful to know 
-if the theoretical model is markedly better than those with differing numbers of classes.
+These models are then compared 
+and the best solution is selected as the final class solution.
+In some cases, prior theory can inform the researcher 
+about the number of classes to expect.
+Even in such confirmatory LCA cases, 
+it is nonetheless useful to know if the theoretical model 
+is markedly better than those with differing numbers of classes.
 Therefore, it may always be useful to compare different class solutions.
 
-From a sequence of models, the final class solution is chosen based on both theoretical and statistical criteria.
-Theory should drive the selection of indicator variables, inform the expectations and reflect on the findings.
-In addition to this, there are several statistical criteria to consider in model selection.
-These include but are not limited to likelihood ratio tests, information criteria, 
+From a sequence of models, the final class solution is chosen 
+based on both theoretical and statistical criteria.
+Theory should drive the selection of indicator variables, 
+inform the expectations and reflect on the findings.
+In addition to this, there are several statistical criteria 
+to consider in model selection.
+These include but are not limited to likelihood ratio tests, 
+information criteria, 
 and the Bayes factor [@weller_latent_2020].
@@
 This is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as the difference in maximum log likelihoods of the two models,
-with the new degrees of freedom being the difference in their degrees of freedom.
-This statistic also follows the $\chi^2$ distribution.
-Similar to the LR $\chi^2$ goodness-of-fit test, we want the test statistic to be non-significant
+The likelihood ratio test statistic is computed as 
+the difference in maximum log likelihoods of the two models,
+with the test degrees of freedom being 
+the difference in the degrees of freedom of the two compared models.
+The test statistic follows the $\chi^2$ distribution,
+and we want it to be non-significant
 in order to give support to the simpler model. 
@@
 
-<!-- DA: This is just a goodness-of-fit test, and if it is significant the model fits the data poorly.  -->
-<!-- We don't get this test with `tidySEM`, so do not focus on this too much. We only provide information criteria-->
-
-<!-- DA: the bootstrap likelihood ratio test is the only one that is valid one according to Caspar, but it is not easily implemented in `tidySEM` (too computationally intensive), which is a problem, takes a lot of time to compute, maybe write about this,
-see @masyn_latent_2013 'parametric bootstrapping' it is not widely implemented. 
-Do not write about bootstrapping, too computationally intensive, we do not have a good solution for model fit tests, downplay that, MPLUS has MLR adjusted maximum likelihood test but unsure if to compare two different class solutions or just misfit, however not something we have, it is a specialized solution available only in MPLUS-->
-
 ### Model Fit Indices
@@
 Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2*log likelihood (which is lower for better fitting models),
+the Akaike Information Criterion (AIC) 
+and the Bayesian Information Criterion (BIC).
+Both information criteria are based on the -2*log likelihood 
+(which is lower for better fitting models),
 and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance fit and model complexity.
-The lower the value of an information criterion, the better the overall fit of the model.
+This helps balance model fit and model complexity.
+The lower the value of an information criterion, 
+the better the overall fit of the model.
 The BIC applies a stronger penalty for model complexity 
 that scales logarithmically with the sample size.
-The literature suggests the BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
+The literature suggests the BIC may be 
+the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
 Both the AIC and the BIC are available in the `tidySEM` output.
 
-Information criteria may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
+Information criteria may occasionally contradict each other, 
+so it is important to identify a suitable strategy to reconcile them.
 One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016].
-Finally, one might make an elbow plot and compare multiple information criteria
-[for an example see @nylund-gibson_ten_2018].
-
-Another common test of model fit is the likelihood ratio $\chi^2$ goodness-of-fit test.
+Another option is to always prefer 
+the most parsimonious model that has best fit 
+according to any of the available fit indices.
+Yet another option is to incorporate information from 
+multiple fit indices using the analytic hierarchy process 
+[@akogul_comparison_2016].
+Finally, one might make an elbow plot and 
+compare multiple information criteria
+[@nylund-gibson_ten_2018].
+
+Another common test of model fit is 
+the likelihood ratio $\chi^2$ goodness-of-fit test.
 However, this test is not implemented in `tidySEM`.
@@
 which can be run using `tidySEM::BLRT()`.
-Currently, this test is computationally expensive and can be slow on most computers.
+Currently, this test is computationally expensive and 
+can be slow on most computers.
 A faster version of this test, 
@@
 
-<!-- DA: NOTES ABOUT LRT test: It can compare two nested models when they have the same number of classes
-<!-- If we wish to simultaneously compare multiple models based on their relative fit, -->
-<!-- this can be done through a comparison of multiple information criteria. -->
-<!-- Examples include the Akaike information criterion (AIC), -->
-<!-- and the Bayesian information criterion (BIC). -->
-<!-- Information criteria are a sum of a measure of fit (usually a form of the converged maximum log likelihood value)  -->
-<!-- and a penalty for model complexity (a combination of sample size and the number of modeled parameters). -->
-<!-- As a general rule, the lower the value of an information criterion, the better the model fits the data. -->
-<!-- When examining several competing models, each with differing number of classes, -->
-<!-- we expect the information criteria values to drop with each successive model until the final class solution is reached. -->
-<!-- Further models with additional classes should show worse fit as information criteria reach an optimum  -->
-<!-- before their values rise again with an increasing number of classes. -->
-<!-- Multiple information criteria should be compared when choosing the final class solution.  -->
-<!-- This can be done by means of an elbow plot [for an example see @nylund-gibson_ten_2018]. -->
-<!-- Advantages of using information criteria are that we can compare multiple models at a time, -->
-<!-- and these models need not be nested @masyn_latent_2013. -->
-
-<!-- Add a paragraph about using the Bayes factor see @weller_latent_2020-->
-
-<!-- decide whether to keep the following paragraph, if so, where to place it -->
-<!-- When estimators which require sets of starting values are used, -->
-<!-- each fitted model will have a corresponding log likelihood value.  -->
-<!-- Ideally, the largest log likelihood value of the final class solution will be replicated a large number of times  -->
-<!-- with each log likelihood value generated by a different starting value set.  -->
-<!-- Successful replications of the largest log likelihood value across different starting values -->
-<!-- instill confidence that the estimation was successful and we indeed found the global maximum [@masyn_latent_2013].  -->
-<!-- The model with the highest replicated likelihood will be chosen as our best-fit.  -->
-<!-- If the largest log likelihood value is not replicated many times, the model may be unidentified (Geiser, 2012). -->
-<!-- Such a model does not have a unique solution. For a discussion on model identification for LCA, see @masyn_latent_2013. -->
-<!-- Nylund-Gibson and Choi (2018) recommend that the highest log likelihood should be replicated  -->
-<!-- in at least 3-10% of the models initialized with different starting values in the first step of the optimization.  -->
-<!-- This consideration does not apply in case of models which use simulated annealing since these do not have starting values. -->
-<!-- @masyn_latent_2013 claims that "replication of the likelihood value is neither a necessary nor a sufficient requirement -->
-<!-- to ensure that a global (rather than a local) maximum has been reached.". For discussion see Hipp & Bauer, 2006 -->
-
-<!-- Absolute model fit can be assessed using the likelihood ratio (LR) $\chi^2$ goodness-of-fit test.  -->
-<!-- It can compare two nested models when they have the same number of classes (Lanza et al., 2003). -->
-<!-- Typically, we compare our final class solution (a constrained model)  -->
-<!-- to an unconstrained one (a model characterized by the perfect fit). -->
-<!-- The former model is a special, less complex version of the latter. -->
-<!-- In this case, we compare our model-based response pattern frequencies to those in the dataset. -->
-<!-- Therefore, we make a statement about how well our model fits the data. -->
-<!-- The perfect fit would be indicated by the $\chi^2\ =\ 0$ and $p\ =\ 1$. -->
-<!-- Ideally, the LR $\chi^2$ test will be non-significant as this implies -->
-<!-- that while our constrained model is more parsimonious, it fits the data just as well as the unconstrained model. -->
-<!-- The underlying assumption of this procedure is that the test statistic is asymptotically distributed -->
-<!-- as a $\chi^2$ meaning that it can be approximated using this distribution. -->
-<!-- This is the case when the sample size is adequate (Lanza et al., 2003).  -->
-<!-- The dependence of the LR $\chi^2$ test statistic on sample size has its downsides [@masyn_latent_2013].  -->
-<!-- Very large samples will often effortlessly reject the null even when the final class solution is a good fit, -->
-<!-- and small samples will often be underpowered resulting in non-significant $\chi^2$ statistic. -->
-<!-- In fact, small samples might not be well approximated using the $\chi^2$ distribution, -->
-<!-- which is a problem shared with large yet sparse datasets. -->
-<!-- In such cases, special caution is advised when interpreting the resulting p-values [@masyn_latent_2013]. -->
-
 ### Classification Diagnostics
 
-Best models will divide the sample into subgroups which are internally homogeneous and externally distinct.
-Classification diagnostics give us a way to assess the degree to which this is the case.
-They are separate from model fit indices as a model can fit the data well 
+Best models will divide the sample into subgroups 
+which are internally homogeneous and externally distinct.
+Classification diagnostics give us a way to assess 
+the degree to which this is the case.
+They are separate from model fit indices as 
+a model can fit the data well 
 but show poor latent class separation [@masyn_latent_2013].
 Classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions because they are uninterpretable.
-Interpretability should always be a consideration when considering different class solutions [@nylund-gibson_ten_2018].
-
-Three important classification diagnostics provided by `tidySEM` are are 
-*the minimum* and *maximum percentage of the sample assigned to a particular class*,
-*the range of the posterior class probabilities by most likely class membership*, 
-and *entropy*. All three are based on posterior class probabilities.
+but they can be used to disqualify certain solutions 
+because they are uninterpretable.
+Interpretability should always be a consideration 
+when considering different class solutions [@nylund-gibson_ten_2018].
+
+Three important classification diagnostics provided by `tidySEM` are: 
+(1) *the minimum* and *maximum percentage of the sample assigned to a particular class*,
+(2) *the range of the posterior class probabilities by most likely class membership*, 
+and (3) *entropy*. 
+All three are based on posterior class probabilities.
 
@@
 and the individual is assigned to the corresponding class. 
-We want each individual's posterior class probabilities to be high for one 
-and low for the remaining latent classes. 
-This is considered a high classification accuracy and means that the classes are distinct.
+We want each individual's posterior class probabilities to be 
+high for one and low for the remaining latent classes. 
+This is considered a high classification accuracy 
+and means that the classes are distinct.
 To obtain posterior class probabilities, run `tidySEM::class_prob()`.
-This function produces an output comprised of several elements.
-Namely:
+This function produces output comprised of several elements:
 
@@
 based on the highest posterior class probability.
-From this table, we compute the minimum 
-and maximum percentage of the sample assigned to a particular class,
+From this table, we compute the minimum and maximum 
+percentage of the sample assigned to a particular class,
 , i.e. **n_min** 
 (the smallest class proportion based on the posterior class probabilities)
-and **n_max** (the largest class proportion based on the posterior class probabilities).
+and **n_max** 
+(the largest class proportion based on the posterior class probabilities).
 We are especially interested in **n_min** as
@@
 the model for that group might not be locally identified.
+It may be impossible to calculate descriptive statistics for such a small class.
 Estimating LCA parameters on small subsamples
@@
 we might consider not to interpret that solution.
-In `tidySEM` we use the diagonal to compute 
-the range of the posterior class probabilities by most likely class membership
+In `tidySEM` we use the diagonal to compute the range of 
+the posterior class probabilities by most likely class membership
 which consists of
@@
 and the highest posterior probability (**prob_max**).
-Both **prob_min** and **prob_max** can be used to (dis)qualify certain class solutions,
+Both **prob_min** and **prob_max** can be used to 
+disqualify certain class solutions,
 and are a convenient way to summarize class separation in LCA.
@@
 that means that for all classes
-the people who were assigned to that class have a high probability of being there.
-**prob_min** is especially important as it can diagnose if there is a class
-with low posterior probabilities 
+the people who were assigned to that class 
+have a high probability of being there.
+**prob_min** is especially important as it can diagnose 
+if there is a class with low posterior probabilities 
 which could make one reconsider that class solution.
-<!-- DA: I would like to rephrase the first sentence of mostlikely.class
-to be in the same format as other post. class probs. -->
 
@@
 
-Entropy is a summary measure of posterior class probabilities across classes and individuals.
-It ranges from 0 (model classification no better than random chance) to 1 (perfect classification).
-As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
-An appropriate use of entropy is that it can disqualify certain solutions if class separability is too low 
-or if one of the latent classes is too small to be meaningful or to calculate descriptive statistics. 
-Entropy was not built for nor should it be used for model selection during class enumeration [@masyn_latent_2013].
+Entropy is a summary measure of posterior class probabilities 
+across classes and individuals.
+It ranges from 0 (model classification no better than random chance) 
+to 1 (perfect classification).
+As a rule of thumb, values above .80 are deemed acceptable 
+and those approaching 1 are considered ideal.
+An appropriate use of entropy is that it can disqualify certain solutions 
+if class separability is too low. 
+Entropy was not built for nor should it be used for 
+model selection during class enumeration [@masyn_latent_2013].
 
@@
 
-## Interpreting Class Solution
+### Interpreting Class Solutions
 
@@
 also known as class-specific item probabilities [@masyn_latent_2013], 
-conditional response or conditional solution probabilities [@geiser_data_2012].
+conditional response, or conditional solution probabilities [@geiser_data_2012].
 They indicate the probability of an item being endorsed
 given that the observation belongs to a particular latent class.
-Conditional item probabilities can be obtained using `tidySEM::table_prob()`
+Conditional item probabilities can be obtained using `tidySEM::table_prob()`.
 If a particular item is endorsed by two or more classes at markedly different rates,
-it is said to discriminate well between the classes and is consequently considered a good indicator.
-Classes are considered highly homogeneous with respect to an item when for a particular item
-there is a distinct difference in conditional item probabilities for two or more classes.
-For instance, if an item is endorsed below 30% for one class and above 70% for another class,
+it is said to discriminate well between the classes 
+and is consequently considered a good indicator.
+Classes are considered highly homogeneous with respect to an item 
+when for a particular item 
+there is a distinct difference in conditional item probabilities
+for two or more classes.
+For instance, if an item is endorsed below 30% for one class and 
+above 70% for another class,
 the classes have high homogeneity with respect to this item [@masyn_latent_2013].
@@
 This too could be a sign of an invalid solution,
-warn us that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].
-
-### Label switching
+warn that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].
+
+### Label Switching
 
 The final class solution will usually discover and enumerate several classes. 
-The class ordering however is completely arbitrary.
-The class labeled as Class 1 in one solution may become Class 2 or Class 3 in another model,
+The class ordering, however, is completely arbitrary.
+The class labeled as Class 1 in one solution 
+may become Class 2 or Class 3 in another model,
 even when the only difference between the models is in their starting values. 
@@
 
-<!-- Article discussing label switching: see Chung, Loken & Schafer -->
-
-<!-- Initialization around a Single Mode -->
-<!-- Another common approach is to run a single chain or to initialize the parameters near realistic values.36 -->
-
-<!-- This can work better than the hard constraint approach if reasonable initial values can be found and the labels do not switch within a Markov chain.
-The result is that all chains are glued to a neighborhood of a particular mode in the posterior.
--->
-<!-- https://rss.onlinelibrary.wiley.com/doi/abs/10.1111/1467-9868.00265?casa_token=Tjk8gR4w8AEAAAAA:RygJL5OgdBaVWY6oaB8XlnyMOnzQgOVDBd2EHDyQD6TH-DzxRxpmU1J4AG8N-HpL5dHHNINRybiaMl6w -->
-
 Class names should be chosen to accurately reflect group membership. 
-Overly simplified and generalized class names may prove misleading to both audiences and researches alike
+Overly simplified and generalized class names may prove misleading 
+to both audiences and researches alike
 leading to what is known as a naming fallacy [@weller_latent_2020].
@@
 Among studies using LCA, reporting practices vary significantly [@weller_latent_2020].
-Various authors have tried to better and standardize ways of reporting LCA [e.g. @masyn_latent_2013; @weller_latent_2020], but more work is needed.
-@van_lissa_worcs_2020 developed WORCS, a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by @nosek_promoting_2015.
-WORCS workflow can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
+Various authors have tried to improve and standardize ways of reporting LCA 
+[e.g. @masyn_latent_2013; @weller_latent_2020], 
+but more work is needed.
+@van_lissa_worcs_2020 developed WORCS,
+a workflow for open reproducible code in science.
+WORCS consists of step-by-step guidelines for research projects 
+based on the TOP-guidelines developed by @nosek_promoting_2015.
+WORCS workflow can be easily implemented in R in form of 
+an R package which facilitates preregistration, article drafting,
 version control, citation and formatting, among others [@van_lissa_worcs_2020].
 
-TOP-guidelines emphasise the use of comprehensive citation (including referencing the software used in the analysis),
+TOP-guidelines emphasise the use of comprehensive citation 
+(including referencing the software used in the analysis),
 as well as code and data sharing wherever possible [@nosek_promoting_2015].
-@van_lissa_worcs_2020 suggest sharing synthetic data in case the original data cannot be shared, 
+@van_lissa_worcs_2020 suggest sharing synthetic data in case 
+the original data cannot be shared, 
 and provide functions to generate such synthetic data.
 Ideally, the entire research project is made reproducible 
-so that others may download it and reproduce it with just one click; for guidance, see @van_lissa_worcs_2020.
+so that others may download it and reproduce it with just one click; 
+for guidance, see @van_lissa_worcs_2020.
 
@@
 how important it is that analyses can be reproduced and audited.
-In line with open science principles, one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., *Mplus* and *Latent GOLD*).
-A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because the underlying source code is proprietary.
-To overcome these limitations, the present paper introduces new user-friendly functions in the `tidySEM` R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package `OpenMx`.
-The reporting guidelines described in this paper are adopted in `tidySEM` by default.
-The `tidySEM` R-package thus makes advanced mixture modeling based on best practices widely accessible,
-and facilitates the adoption of the estimation and reporting guidelines described in this paper.
-
-## Best Practices in Visualization
+In line with open science principles, 
+one of the suggested reporting standards relates to reproducible code.
+In this context, it is important to note that 
+user-friendly methods for estimating LCA models 
+have predominantly been available in commercial software packages 
+(e.g., *Mplus* and *Latent GOLD*).
+A potential downside of commercial software is that 
+it restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because 
+the underlying source code is proprietary.
+To overcome these limitations, 
+the present paper introduces new user-friendly functions in 
+the `tidySEM` R-package that can be used to 
+estimate a wide range of LCA models 
+using the free, open-source R-package `OpenMx`.
+The reporting guidelines described in this paper 
+are adopted in `tidySEM` by default.
+The `tidySEM` R-package thus makes advanced mixture modeling 
+based on best practices widely accessible,
+and facilitates the adoption of the estimation and 
+reporting guidelines described in this paper.
+
+### Visualization
+
+Plots can greatly improve the interpretability of LCA models.
+There are several stages where this is an important consideration.
+
+First, in class enumeration, 
+we want to compare several competing class solutions.
+This can be done by means of the AIC and BIC.
+Here, we suggest using an elbow plot
+with the $K-class\ solution$ on the X-axis,
+and information criteria value on the Y-axis.
+We show an implementation of such an elbow plot 
+in our Tutorial subsection.
+<!-- For continuous variables, density plots can aid class enumeration. -->
+<!-- DA: Do we have density plots implemented in tidySEM? -->
+<!-- If so, we should add this example to the tutorial. -->
+
+Second, once we have decided on the final class solution,
+we want to interpret the response patterns
+on the indicators for each latent class.
+While `tidySEM` can create a table showing
+the probability of each item's response endorsement 
+(using `tidySEM::table_prob`),
+it may be easier to inspect these probabilities visually.
+For this reason, we created `ggplot2::plot_prob()`.
+The resulting shows response patterns 
+on all indicators for each group. 
+We give an example of this plot in the Tutorial subsection.
 
@@
 
-<!-- insert the Daniel_LCA file from the example_analysis folder here -->
+This is an example of an exploratory LCA using `tidySEM`.
+
+## Loading the Dataset
+
+In this example, we use *data_mix_ordinal*, 
+a simulated data for mixture modelling with ordinal indicators. 
+This dataset is built-in to `tidySEM`.
+For more information about the dataset, 
+type `?tidySEM::data_mix_ordinal` into the R console
+after loading the `tidySEM` using library().
+
+We load the dataset and convert the indicators into ordered factors.
+
+```{r, include = TRUE, eval=F}
+library(tidySEM) # loading tidySEM
+df <- data_mix_ordinal  # dataset
+df[1:4] <- lapply(df, ordered) # indicators as ordered factors
+```
+
+## Exploring the Data
+
+An important step to preceding any statistical analysis is data exploration.
+Here we use `tidySEM::descriptives()` 
+to describe the dataset we are using.
+
+```{r, include = TRUE, eval=F}
+tidySEM::descriptives(df) # descriptives
+```
+As we can see, the output includes various descriptives of our dataset.
+Special attention should be paid to examining the pattern of missingness,
+as discussed in the *Handling missing data* section of this paper.
+In our example, we see that there are no missing values,
+hence we proceed with our analysis.
+
+## Conducting Latent Class Analysis
+
+Before we fit a series of LCA models, we set a random seed
+using `set.seed()`.
+The seed number can be any digit.
+This is an important step as 
+there is some inherent randomness in the LCA computations,
+and having the same seed number ensures that
+two separate researcher obtain exactly the same results
+when fitting LCA models.
+
+Finally, we reach the step of fitting LCA models.
+To do so, we use `tidySEM::mx_lca()`
+which takes data and number of classes as inputs.
+In our example, we want to fit 1 to 4 class solutions
+and compare their output.
+Depending on your computer's computational power,
+this might take a while.
+
+```{r fitting_LCA, include = TRUE, eval=F}
+set.seed(123) # setting seed 
+res <- mx_lca(data = df, classes = 1:4) # fitting LCA 
+```
+
+## Class Enumeration
+
+In class enumeration, 
+we want to compare a sequence of LCA models fitted to the data.
+To aid the process, we create a model fit table using `tidySEM::table_fit()` 
+with the results object as the input.
+As the output contains a lot of information on each of the four fitted models,
+we select a subset of helpful model fit indices and classification diagnostics.
+
+```{r fit_table, include = TRUE, eval=F}
+fit_table <- table_fit(res) # model fit table
+fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # our selection
+```
+Our selection of fit indices and classification diagnostics includes:
+
+```{r, echo=F}
+Selection <- c("Name", "LL", "AIC", "BIC", "prob_min", "prob_max",
+          "n_min", "n_max")
+Description <- c("the $K$-class solution",
+                 "the -2*log-likelihood of each model",
+                 "the Akaike Information Criterion value",
+                 "the Bayesian Information Criterion value",
+                 "the lowest posterior class probability by most likely class membership",
+                 "the highest posterior class probability by most likely class membership",
+                 "the lowest class proportion based on the posterior class probabilities",
+                 "the highest class proportion based on the posterior class probabilities")
+selection <- cbind(Selection, Description)
+knitr::kable(selection, format = 'pipe', caption = "Selection of Fit Indices and Classification Diagnostics")
+```
+We discussed several possible strategies to select the final class solution. 
+Here, we apply our own.
+
+To aid our interpretation of the results, 
+we create an elbow plot showing the trends in information criteria across four models.
+
+```{r elbow_plot, include = TRUE, message=F, warning=F, eval=F}
+library(tidyverse) # for data-wrangling
+library(ggplot2) # for plots
+
+elbow_plot <- fit_table[ , c("Name", "AIC", "BIC")] # extract ICs
+elbow_plot %<>% pivot_longer(cols = c("AIC", "BIC"), names_to = "IC", values_to = "Value") # to long format
+
+ggplot(elbow_plot, aes(x = Name, y = Value, group = IC))+
+  geom_point(aes(color = IC))+
+  geom_line(aes(color = IC))+
+  labs(title="Elbow Plot of Information Criteria per Class Solution", x = "Class Solution", y = " Value")+
+  scale_color_manual(name = "Information Criterion", values = c(AIC = 'blue', BIC = 'red'))+
+  theme_minimal()
+
+```
+From the elbow plot, we see that AIC has a lower penalty for model complexity.
+However, we are more interested in the BIC values, which are similar
+for the one, two and three-class solutions, but the four-class solution fits significantly worse.
+For this reason, we eliminate the four-class solution from the selection process.
+
+Then we examine the model fit table.
+As expected, the -2*log likelihood falls successively with each added class.
+As previously stated, classification diagnostics should not be used for model selection,
+but they can be used to disqualify certain solutions because they are uninterpretable.
+We see that prob_min for the four-class solution is low,
+knowing that this solution also has a high BIC, we disqualify this solution.
+
+Out of the remaining three solutions,
+we notice that entropy is the highest for the three-class solution,
+and it has a satisfactory prob_min and n_min.
+Based on this, we retain the three class solution in model selection.
+Note that entropy for the one-class solution will always equal to one,
+as it is 100% true that every case is in that class.
+Based on the low entropy of the two-class solution,
+we eliminate this model.
+
+Finally, when comparing the one and three-class solutions,
+we inspect the information criteria.
+For BIC, the one-class solution fits better, but the difference is marginal.
+AIC tells us that the added complexity
+of having three classes still explains the data better
+than a one-class solution. 
+Therefore, we select the three-class solution as our final-class solution.
+
+## Interpreting the Final Class Solution
+
+To aid our understanding of the final class solution,
+we use `ggplot2::plot_prob()` 
+with the results of the three-class model as the input.
+The resulting graph shows response patterns 
+on all the indicators for each group.
+
+If we want to know the probability of each response option's endorsement 
+for each class, we can use `tidySEM::table_prob`.
+These are thresholds for ordinal dependent variables in the probability scale.
+
+```{r plot_prob, include = TRUE, eval=F}
+plot_prob(res[[3]]) # visualizing the response patterns for the final class solution
+table_prob(res[[3]]) # tabulating the response patterns for the final class solution
+```
+In the plot, we can see the distributions of the response probabilities on the indicators
+for each of the three classes.
+For instance, we see that in Class 1 the most common response to u2 is 2,
+while in Class 2 and Class 3 this is 0.
+We can also see that response 1 is a rare response not forming the majority in 
+any class.
+Class 2 distinguishes itself because the majority scores the response 0 category of u3 and u4,
+while in Class 1 and 2 this is not the case.
+Class 3 distinguishes itself because the most common response to u3 and u4 is 3.
+
+We can also interpret the response patterns numerically.
+
+## Extracting Posterior Class Probabilities
+
+Another step is to extract posterior class probabilities.
+This is done by the use of `TidySEM::class_prob`
+with the results of the final class solution as the input.
+
+```{r extract_Post_Class_Prob, include = TRUE, eval=F}
+probs <- class_prob(res[[3]]) # extracting the posterior class probabilities of the final class solution
+probs$mostlikely.class # posterior class probabilities by most likely class membership
+probs$individual # individual posterior class probabilities
+```
 

### 2022-10-16 manuscript.tex
Update 16 October
Changed title and short title. Adjusted abstract, Added keywords. Made the use of "LCA" and "mixture models" more consistent based on whether we refer to the whole group of LCA techniques, or a mixture subset.  Minor adjustment to wording throughout the paper. Shortened lines throughout. Capitalized titles and made them uniform. Adjusted section-subsection-subsubsection depth (# vs ## vs ###).  Removed irrelevant comments. Wrote best practices in visualization (##Visualization).  Added the tutorial (now set to eval = F for knitting, however code chunks do not show when using

@@
 \documentclass[
-  ,man]{apa6}
+  ,man,floatsintext]{apa6}
 \usepackage{amsmath,amssymb}
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Best Practices in Mixture Modeling using Free Open Source Software},
-  pdfauthor={Caspar J. van Lissa1,2},
-  pdflang={en-EN},
-  pdfkeywords={keywords},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 {}{\PatchFailed}
-\keywords{keywords\newline\indent Word count: X}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
+\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 8206}
 \usepackage{csquotes}
@@
 \fi
-
-\title{Best Practices in Mixture Modeling using Free Open Source Software}
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
+  pdfauthor={Caspar J. van Lissa1,2},
+  pdflang={en-EN},
+  pdfkeywords={latent class analysis, mixture models, best practices, free open source software, tidySEM},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
+
+\title{Best Practices in Latent Class Analysis using Free Open Source Software}
 \author{Caspar J. van Lissa\textsuperscript{1,2}}
@@
 
-\shorttitle{BEST PRACTICES MIXTURE MODELING}
+\shorttitle{BEST PRACTICES LCA}
 
@@
 \abstract{%
-Latent class analysis is a popular technique for identifying groups in data based
-on a parametric model. Examples of this technique are known as mixture models, latent profile
-analysis, latent class analysis, growth mixture modeling, and latent class
+Latent class analysis is a family of techniques for identifying groups in data based
+on a parametric model. Its examples include mixture models, latent profile
+analysis, latent class analysis with ordinal indicators, and latent class
 growth analysis. Despite the popularity of this technique, there is limited
@@
 using free and open source software in R. To this end, this tutorial
-introduces new functionality for estimating and reporting mixture mixture
-models in the \texttt{tidySEM} R-package. These functions rely on estimation using
-the OpenMx R-package.
+introduces new functionality for estimating and reporting mixture
+models in the \texttt{tidySEM} R-package whose backend relies on
+the \texttt{OpenMx} R-package.
 }
@@
 
-Latent class analysis is an umbrella term that refers to a number of techniques for estimating unobserved group membership
+Latent class analysis (LCA) is an umbrella term
+that refers to a number of techniques for estimating unobserved group membership
 based on a parametric model of one or more observed indicators of group membership.
-This method has become quite popular across scientific fields, and under a number of different names; most notably (finite Gaussian) mixture modeling and latent profile analysis.
-Despite the popularity of the method, there is a lack of standards for estimating and reporting latent class analyses.
+The types of LCA have become quite popular across scientific fields,
+most notably finite Gaussian mixture modeling and latent profile analysis.
+Vermunt, J.K. et al. (2004) defined LCA more generally as virtually any statistical model
+where ``some of the parameters {[}\ldots{]} differ across unobserved subgroups''.
+
+Despite its popularity, there is a lack of standards for estimating and reporting LCA.
+While Van De Schoot, Sijbrandij, Winter, Depaoli, and Vermunt (2017) developed reporting guidelines
+for a specific type of LCA known as latent growth models,
+general reporting guidelines for all LCA are still lacking.
 This complicates manuscript review and assessment of the quality of published studies,
 and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature by suggesting updated guidelines for estimation and reporting on latent class analysis, based on current best practice.
-Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, this paper exclusively relies on free, open source software for latent class analysis in R.
-Our goal is to make best-practices latent class analysis widely accessable.
+The present paper seeks to address this gap in the literature
+by suggesting updated guidelines for estimation and reporting of LCA,
+based on current best practices.
+Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples,
+this paper exclusively relies on free, open source software in R.
+Our goal is to make best-practices in LCA widely accessible.
 
 \hypertarget{defining-latent-class-analysis}{%
-\subsection{Defining latent class analysis}\label{defining-latent-class-analysis}}
-
-Latent class analysis can be understood as a method for estimating unobserved groups
+\subsection{Defining Latent Class Analysis}\label{defining-latent-class-analysis}}
+
+Latent class analysis is a group of methods for estimating unobserved groups
 based on a parametric model of observed indicators of group membership.
-The concept of latent class analysis can be understood in different ways.
-Generally speaking, it can be said that a mixture model assumes that the study population
-is composed of \(K\) subpopulations, or classes.
+The concept of LCA can be understood in different ways.
+A mixture model assumes that the study population
+is composed of \(K\) subpopulations or classes.
 It further assumes that the observed data are a mixture of data generated by class-specific models.
-The simplest univariate ``model'' is a normal distribution, which can be described with two parameters:
-the mean and the variance.
-Commonly, the same model is estimated accross all classes, but with different parameters for each class (i.e., class-specific means and variances).
-Mixture modeling then estimates both the parameters for each class, and the probability for each that an individual belongs to each class.
-
-As an illustrative example, imagine that a detective wants to know if it would be
-possible to use mixture modeling to identify the sex of a suspect,
+The simplest univariate model is a normal distribution,
+which can be described with two parameters: the mean and the variance.
+Commonly, the same model is estimated across all classes,
+but with different parameters for each class (i.e., class-specific means and variances).
+Mixture modeling then estimates both the parameters for each class,
+and the probability that an individual belongs to each class.
+
+As an illustrative example, imagine that a detective wants to know
+if it would be possible to use mixture modeling to identify the sex of a suspect,
 based on footprints found at the crime scene.
@@
 When estimating a two-class mixture model, the detective observes that the model estimates
-the mean shoesize of the two groups are equal to 7.25 and 9.22,
-which is close to the true means of the two groups, namely 9.04 and 6.93.
-When tabulating estimated group membership against observed (known) group membership, it can be seen that women are classified with a high degree of accuracy, but men are not:
+the mean shoe size of the two groups are equal to 7.25
+and 9.22,
+which is close to the true means of the two groups,
+namely 9.04 and 6.93.
+When tabulating estimated group membership against observed (known) group membership,
+it can be seen that women are classified with a high degree of accuracy, but men are not:
 
@@
 
-One way to is to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
-A mixture model is like confirmatory factor analysis, except that the continuous latent variable is substituted with a categorical latent variable.
-One difference between the two techniques is that factor analysis can be considered as a way to group observed \emph{variables} into latent constructs, with factor loadings indicating which items belong are most indicative of a construct.
-By contrast, mixture modeling groups \emph{individuals} into classes (see Nylund-Gibson \& Choi, 2018).
-In line with this distinction, latent class analysis is sometimes referred to as a ``person-centered'' technique, and factor analysis as a ``variable-centered'' technique.
+A LCA is like confirmatory factor analysis,
+except that the continuous latent variable is substituted with a categorical latent variable.
+One difference between the two techniques is that
+factor analysis can be considered as a way to group observed \emph{variables} into latent constructs,
+and LCA groups \emph{individuals} into classes.
+In line with this distinction, LCA is sometimes referred to as a ``person-centered'' technique,
+and factor analysis as a ``variable-centered'' technique (Nylund-Gibson \& Choi, 2018).
 
 When the focus is on the model parameters in each group,
-then latent class analysis can be thought of as similar to a multi-group structural equation model.
+LCA can be thought of as similar to a multi-group structural equation model.
 The main distinction is that group membership is not known a-priori,
-but is instead estimated - with measurement error - based on the data.
-Whereas in a multigroup model, the data are split by group and treated as independent samples,
-in a mixture model, all cases contribute to the estimation of all parameters in all groups.
-The relative contribution of each case to the parameters of each group is determined by that case's posterior probability of belonging to that group.
+but is instead estimated -- with measurement error -- based on the data.
+Whereas in a multi-group model, the data are split by group and treated as independent samples,
+in LCA, all cases contribute to the estimation of all parameters in all groups.
+The relative contribution of each case to the parameters of each group
+is determined by that case's posterior probability of belonging to that group.
 
 When the focus is on each individual's estimated class membership,
-latent class analysis can be thought of as a type of clustering algorithm.
-In line with this perspective, mixture modeling is sometimes described as ``model-based
-clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \&
-Raftery, 2017).
+LCA can be thought of as a type of clustering algorithm.
+In line with this perspective, LCA is sometimes described as ``model-based
+clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
 Many clustering algorithms apply some recursive splitting algorithm to the data.
-By contrast ``model-based'' clustering refers to the fact that latent class analysis
+By contrast ``model-based'' clustering refers to the fact that LCA
 estimates cluster membership based on a parametric model.
@@
 
-Finally, in the context of machine learning, latent class analysis can be considered as an
+Finally, in the context of machine learning, LCA can be considered as an
 \emph{unsupervised classification} problem (Figueiredo \& Jain, 2002).
-The term \emph{unsupervised} refers to the fact that the outcome variable, true class membership,
-is not known, and the term \emph{classification} refers to the fact that the algorithm is predicting a categorical outcome (class membership).
-
-\hypertarget{a-taxonomy-of-latent-class-analyses}{%
-\subsection{A taxonomy of latent class analyses}\label{a-taxonomy-of-latent-class-analyses}}
-
-In this paper, we use the term latent class analysis to refer to techniques that estimate latent class membership based on a parametric model of observed indicators.
-From a historical perspective, the term latent class analysis was initially conceived to refer to analyses with categorical (binary) indicators (Vermunt, J.K. et al., 2004).
-Nowadays, there are a number of related techniques, known by distinct names, that serve a similar purpose.
-The term ``latent class analysis'' seems most appropriate as an umbrella term for this broader class of models,
-as it only refers to the purpose of the analysis, and does not imply restrictions to the model used, or the level of measurement of the indicators.
-Given the abundance of terms in use for closely related classes of models, we will provide a rudimentary taxonomy of latent class analyses.
-
-One common type of latent class analyses is the \emph{finite Gaussian mixture model};
-a univariate analysis where the observed distribution of a single variable is assumed to result from a mixture of a known number of Gaussian (normal) distributions.
-The parameters of a finite Gaussian mixture model are the means and variances of these underlying normal distributions.
+The term \emph{unsupervised} refers to the fact that the outcome variable
+-- true class membership -- is not known,
+and the term \emph{classification} refers to the fact that
+the algorithm is predicting a categorical outcome -- class membership.
+
+\hypertarget{a-taxonomy-of-latent-class-analysis-methods}{%
+\subsection{A Taxonomy of Latent Class Analysis Methods}\label{a-taxonomy-of-latent-class-analysis-methods}}
+
+In this paper, we use the term latent class analysis to refer to
+a family of techniques that estimate latent class membership
+based on a parametric model of observed indicators.
+From a historical perspective, the term was initially conceived to
+refer to analyses with categorical, usually binary indicators (Vermunt, J.K. et al., 2004).
+Nowadays, there are a number of related techniques,
+known by distinct names, that serve a similar purpose.
+The term ``latent class analysis'' seems most appropriate
+as an umbrella term for this broader class of models,
+as it only refers to the purpose of the analysis,
+and does not imply restrictions to the model used,
+or the level of measurement of the indicators.
+Given the abundance of terms in use for closely related classes of models,
+we will provide a rudimentary taxonomy of LCA methods.
+
+A common type of LCA is the \emph{finite Gaussian mixture model};
+a univariate analysis where the observed distribution of a single variable
+is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
+The parameters of a finite Gaussian mixture model are
+the means and variances of these underlying normal distributions.
 The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
-In the multivatiate case, with more than one indicator variable,
-the parameters of a mixture model are the means, variances, and covariances between the indicators (which can be standardized to obtain correlations).
-These parameters can be estimated freely, or constrained, across classes.
+In the multivariate case, with more than one indicator variable,
+the parameters of a mixture model are the means, variances, and covariances
+between the indicators (which can be standardized to obtain correlations).
+These parameters can be estimated freely, or set to be constrained across classes.
 
@@
 which assumes conditional independence of the indicators.
-Conditional independence means that,
-after class membership is accounted for,
+Conditional independence means that, after class membership is accounted for,
 the covariances/correlations between indicators are assumed to be zero.
@@
 In some cases, such constraints will be inappropriate;
-for example, when the cohesion between indicators is expected to differ between classes.
-As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
-A class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
-From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.
+for instance when the cohesion between indicators is expected to differ between classes.
+As an example, a LPA of ocean plastic particles
+found two classes of particles based on length and width:
+a class of smaller particles with a high correlation between length and width,
+meaning that these particles were approximately round or square in shape,
+and a class of larger particles with a low correlation between length and width,
+meaning that these particles were heterogenous in shape.
+From a theoretical perspective, this makes sense,
+because the smaller particles have been ground down to a more uniform shape by the elements.
 
 It is also possible to estimate a mixture model based on latent indicators.
-This means that, within each class, one or more continuous latent variables are estimated based on the observed indicators.
-Categorical latent variable membership is then estimated based on these continuous latent variables.
-A common application of this approach is in longitudinal research, where the indicators reflect one construct assessed at different time points.
-Examples of this approach include \emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA).
-These techniques estimate a latent growth model to describe individual trajectories over time.
-The growth mixture model is a latent class model where the parameters that indicate class membership
-are the intercepts and variances (and typically covariances) of the latent growth variables, e.g., a latent intercept and slope.
+This means that, within each class,
+one or more continuous latent variables are estimated based on the observed indicators.
+Categorical latent variable membership is then estimated
+based on these continuous latent variables.
+A common application of this approach is in longitudinal research,
+where the indicators reflect one construct assessed at different time points.
+Examples of this approach include \emph{growth mixture models} (GMM)
+and \emph{latent class growth analyses} (LCGA).
+These techniques estimate a latent growth model
+to describe individual trajectories over time.
+The growth mixture model is a latent class model
+where the parameters that indicate class membership
+are the intercepts and variances,
+and typically covariances of the latent growth variables,
+e.g., a latent intercept and slope.
 This technique assumes that individuals within a class can have heterogenous trajectories.
-If the variance of the growth parameters is fixed to zero, it is known as a latent class growth analysis.
-This latter approach assumes that all individuals within a class share the same identical trajectory,
-and that any variance in the indicators not explained by the class-specific latent trajectories is due to residual error variance.
-
-The term latent class analysis originally referred to cases where the observed indicators were categorical.
+If the variance of the growth parameters is fixed to zero,
+it is known as a latent class growth analysis.
+This latter approach assumes that all individuals within a class
+share the same identical trajectory,
+and that any variance in the indicators not explained by the class-specific latent trajectories
+is due to residual error variance.
+
+The term latent class analysis originally referred to cases
+where the observed indicators were categorical.
 Nowadays, it is more commonly used as an umbrella term.
-To prevent ambiguity, the special case where indicators are of binary or ordinal measurement level might be described as latent class analysis with ordinal indicators.
+To prevent ambiguity, the special case
+where indicators are of binary or ordinal measurement level
+might be described as \emph{latent class analysis with ordinal indicators}.
 Latent class models with ordinal indicators are parameterized differently from mixture models.
-One common parameterization assumes that each categorical variable reflects an underlying standard normal distribution.
-The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution (with \(N(\mu = 0, \sigma = 1)\)).
-These thresholds are estimated based on the proportion of individuals in each of the response categories of the indicator variable.
+One common parameterization assumes that
+each categorical variable reflects an underlying standard normal distribution.
+The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution
+(with \(N(\mu = 0, \sigma = 1)\)).
+These thresholds are estimated based on
+the proportion of individuals in each of the response categories of the indicator variable.
 For example, a binary indicator has a single threshold that distinguishes the two response categories.
@@
 If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).
+
 This paper will primarily focus on mixture models and special cases thereof,
-although most of the suggested guidelines are applicable to latent class analyses.
+although most of the suggested guidelines are applicable to all LCA methods.
 
 \hypertarget{use-cases-for-latent-class-analysis}{%
-\subsection{Use cases for latent class analysis}\label{use-cases-for-latent-class-analysis}}
-
-There are several use cases for which latent class analyses are suitable.
+\subsection{Use Cases for Latent Class Analysis}\label{use-cases-for-latent-class-analysis}}
+
+There are several use cases for which LCA methods are suitable.
 One example is to test a theory that postulates the existence of a categorical latent variable.
-For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses.
-Latent class analysis can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment).
+For example, \emph{identity status theory} posits that, at any given point in time,
+adolescents reside in one of four identity statuses.
+Latent class analysis can be used to identify these four statuses
+based on observed indicators (e.g., self-reported identity exploration and commitment).
 If results indicate that the data are better described by a different number of classes,
-or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, then the theory may be called into question.
-
-Another use case is unsupervised learning;
+or that the four-class solution does not correspond to the predicted pattern of responses on the indicators,
+then the theory may be called into question.
+
+Another use case is unsupervised learning:
 when the goal is to restore unobserved class membership based on observed indicators,
 or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid when several clinical indicators can be used to distinguish between a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
+For example, a mixture model can be used as a diagnostic aid
+when several clinical indicators can be used to distinguish between
+a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
 The example of shoe size is a rudimentary illustration of this type of application.
 
+LCA can be used as a descriptive analysis
+where a researcher wishes to describe their sample
+and identify a few prototypes based on many variables.
+As an example, if a survey among all Dutch academics was carried out
+with the goal of bringing about a funding reform,
+LCA could be used to discover the types of publications that get funding.
+
+With LCA, our goal could be to inductively identify the number of classes.
+For instance, if we believe that a variable represents a group,
+but don't know how many groups there are,
+LCA may be an appropriate technique to answer this question.
+For example, Hopfer, Tan, and Wylie (2014) studied substance use, sexual behavior,
+and mental health status of urban population in Winnipeg, Canada.
+The underlying assumption was that there were different risk profiles,
+but their number was not known.
+From a collection of indicators, LCA provided evidence
+that there are four distinct risk profiles in the Winnipeg area.
+
+Another application of LCA is to classify individuals.
+In a peer harassment study,
+Giang and Graham (2008) used LCA to classify over 2,000 sixth grade students
+into aggressor and victim latent classes.
+The five-class solution comprised of classes of
+victims, aggressors, and socially adjusted students.
+For instance, it revealed that there were two types of victims:
+highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
+
+LCA is also appropriate when we wish to identify indicators that capture classes well.
+High quality indicators will be strongly related to the latent variable
+and will lead to good class separation.
+This relationship of high quality indicators to the latent variable
+is reflected in very high or very low conditional response probabilities.
+For a simulation study exploring the effects of indicator quality on LCA, see Geiser and Wurpts (2014).
+Therefore, one could use conditional response probabilities for each item to assess its quality
+with regards to how well it helps separate the latent classes.
+From this, a theory about the selection of indicators could be informed.
+
+An extension of LCA is that containing covariates which can be used to predict class membership.
+In this approach, we not only model the latent class variable based on indicators,
+but we also relate the class membership to other explanatory variables (Vermunt (2017)).
+An example of using covariates comes from Nozadi et al. (2016)
+who applied LCA to identify the probability of children's membership to an anxiety class.
+The authors tested several covariates including children's age, sex, and accuracy scores.
+Age and sex were not found to be related to the children's latent class membership,
+hence these covariates were excluded from the analysis.
+Accuracy scores, however, were related to probabilities of
+being in anxiety and attention-anxiety classes.
+Therefore this covariate was kept as a valuable predictor.
+
+When our interest is the prediction of one or more outcomes,
+LCA can be used to construct latent classes as categorical predictors.
+Lanza, Tan, and Bray (2013) demonstrated how LCA can be used
+to classify adolescents into depression classes,
+and subsequently these classes can be used to predict smoking, grades, and delinquency.
+The study showed that the outcomes predicted by class membership
+can be binary (regular smoking), continuous (grades) or count (delinquency).
+
+In addition to these applications, LCA can be used for dimensionality reduction
+as the resulting groups summarize response patterns on a large number of indicators.
+For example, MacGregor et al. (2021) investigated symptom profiles
+among injured U.S. military personnel.
+They used fifteen dichotomous items
+from the Post-Deployment Health Assessment survey as LCA indicators.
+Combinatorics informs us that fifteen dichotomous items have
+\(2^{15}\) or \(32,768\) unique symptom combinations.
+Perhaps for this reason, MacGregor et al. (2021) incorporated LCA
+as a method of dimensionality reduction.
+A five class solution was found to have the best fit
+according to both statistical criteria and clinical interpretability.
+
+Finally, LCA can be used to deal with data which violate certain assumptions.
+As discussed in the shoe size example, LCA can deal with violations of normality.
+In fact, LCA assumes the population distribution is a non-normal mixture
+of \(K\) normal distributions,
+and it can discover the value of \(K\), i.e.~generate a \(K\)-class solution.
+
 \hypertarget{best-practices}{%
-\section{Best practices}\label{best-practices}}
-
-\hypertarget{in-estimation}{%
-\subsection{In estimation}\label{in-estimation}}
+\section{Best Practices}\label{best-practices}}
+
+\hypertarget{best-practices-in-estimation}{%
+\subsection{Best Practices in Estimation}\label{best-practices-in-estimation}}
 
 The best practices in estimation, as outlined in Table \ref{tab:checkest},
-are rooted in existing recommendations for best practices for estimating
-specific sub-types of latent class analyses,
-including latent class growth analysis (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?})
+are rooted in existing recommendations for best practices
+for estimating specific subtypes of LCA,
+including latent class growth analysis (Van De Schoot et al., 2017)
 and latent class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
-These were generalized to be more relevant to all types of latent class analyses,
+These were generalized to be relevant to all types of LCA,
 and updated to current best practices, as explained below.
@@
 \hline
-1. & Examine observed data\\
+1. & Examining Observed Data\\
 \hline
-2. & Handling missing data\\
+2. & Handling Missing Data\\
 \hline
-3. & Alternative model specifications\\
+3. & Alternative Model Specifications\\
 \hline
-4. & Starting values\\
+4. & Software\\
 \hline
@@
 \hline
+6. & Class Enumeration\\
+\hline
+7. & Model Fit Indices\\
+\hline
+8. & Classification Diagnostics\\
+\hline
+9. & Interpreting Class Solutions\\
+\hline
+10. & Label switching\\
+\hline
 \end{tabular}
 
-\hypertarget{examine-observed-data}{%
-\subsubsection{Examine observed data}\label{examine-observed-data}}
-
-Examining observed data is essential for any analysis,
-as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
-Pay special attention to level of measurement of the indicators.
-Mixture modeling (including LPA) is strictly speaking only suitable for continuous variables.
-Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models (see Vermunt, 2011).
-Personal experience consulting on latent class analyses and moderating the \texttt{tidyLPA} Google group
-suggest that the mis-application of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
-Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous (e.g., Norman, 2010),
+\hypertarget{examining-observed-data}{%
+\subsubsection{Examining Observed Data}\label{examining-observed-data}}
+
+Examining observed data is essential for any analysis
+as it may reveal patterns and violations of assumptions
+that had not been considered prior to data collection.
+Special attention should be paid
+to level of measurement of the indicators.
+Finite Gaussian mixture models (including LPA)
+are only suitable for continuous variables.
+Indicators with an ordinal level of measurement
+are likely to violate the assumption of
+within-class normal distributions of mixture models
+(see Vermunt, 2011).
+Personal experience consulting on LCA methods and
+moderating the \texttt{tidyLPA} Google group
+suggest that the misapplication of mixture models
+to ordinal (e.g., Likert-type) indicators
+is the most common source of user error.
+Whereas it has been argued that some parametric methods are robust
+when scales with 7+ indicators are treated as continuous
+(e.g., Norman, 2010),
 this certainly does not imply that all methods are.
-It is certainly unlikely that such ordinal variables can be treated as a \emph{mixture} of multiple normal distributions.
-The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; in this case, each class-specific mean could describe a single response category, and a class-specific variance component would be nonsensical.
+It is certainly unlikely that such ordinal variables
+can be treated as a \emph{mixture} of multiple normal distributions.
+The problem becomes egregious when
+the number of classes estimated equals or exceeds the number of categories;
+in this case, each class-specific mean could describe a single response category,
+and a class-specific variance component would be nonsensical.
 In sum, Likert-type scales are rarely suitable for mixture modeling;
@@
 
-Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed {[}REF SPURK{]}.
+Relatedly, a recent publication claimed that
+an assumption of mixture models is that
+observed indicators are normally distributed (Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020).
 This is incorrect.
-When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
-
-Extensive descriptive statistics (including the number of unique values, variance of categorical variables, and missingness; see next paragraph) can be obtained using the function \texttt{tidySEM::descriptives(data)}.
-Note, however, that sample-level descriptive statistics are of limited value when the goal of a study is to identify sub-samples using latent class analysis.
-Plots (density plots for continuous variables, and bar charts for categorical ones) may be more diagnostic.
-Note that density plots can also aid in the choice of the number of classes, as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, provided that these are readily accessible (consider using a GitHub repository as a comprehensive public research archive, as explained in Van Lissa et al., 2020).
-
-\hypertarget{missing-data}{%
-\subsubsection{Missing data}\label{missing-data}}
-
-Previous work has emphasized the importance of examining the pattern of missing data and reporting how missingness was handled (\textbf{schootGRoLTSChecklistGuidelinesReporting2017?}).
-Three types of missingness have been distinguished in the literature (Rubin, 1976):
-Missing completely at random (MCAR), which means that missingness is random;
-missing at random (MAR), which means that missingness is contingent on the \emph{observed} data (and can thus be accounted for);
-and finally missing not at random (MNAR), which means that missingness is related to unobserved factors.
+When the number of classes is greater than one,
+mixture models assume that the observed indicators
+are a mixture of multiple (multivariate) normal distributions.
+In our shoe size example, it can be seen that the population distribution
+is comprised of two normal distributions.
+When examined visually, the population distribution is evidently bimodal.
+The Shapiro-Wilk normality test (\(W = 0.971\), \(p < 0.05\) )
+rejects the null hypothesis that
+the sample comes from a normally distributed population.
+Yet, this is a prototypical example of a mixed population distribution
+where LCA can discover latent groups.
+If the population distribution were instead normal,
+there would be no classes to extract
+as the whole population would belong to a single class.
+
+Extensive descriptive statistics
+(including the number of unique values,
+variance of categorical variables, and missingness; see next subsection)
+can be obtained using the function \texttt{tidySEM::descriptives(data)}.
+Note, however, that sample-level descriptive statistics
+are of limited value when the goal of a study is
+to identify subsamples using latent class analysis.
+Plots (density plots for continuous variables,
+and bar charts for categorical ones) may be more diagnostic.
+Note that density plots can also aid in the choice of the number of classes,
+as further explained in the section on visualization.
+Descriptive statistics and plots can be relegated to online supplements,
+provided that these are readily accessible
+(consider using a GitHub repository as a comprehensive public research archive,
+as explained in Van Lissa et al., 2021).
+
+\hypertarget{handling-missing-data}{%
+\subsubsection{Handling Missing Data}\label{handling-missing-data}}
+
+Previous work has emphasized the importance of examining
+the pattern of missing data and
+reporting how missingness was handled
+(Van De Schoot et al., 2017).
+Three types of missingness have been distinguished
+in the literature (Rubin, 1976):
+Missing completely at random (MCAR),
+which means that missingness is random;
+missing at random (MAR),
+which means that missingness is contingent on the \emph{observed} data
+(and can thus be accounted for);
+and finally missing not at random (MNAR),
+which means that missingness is related to unobserved factors.
 It is possible to conduct a so-called ``MCAR'' test,
@@
 as the null-hypothesis of this test is that the data are not MAR,
-and a significant test statistic indicates that missingness is related to the observed data (MAR).
+and a significant test statistic indicates that
+missingness is related to the observed data (MAR).
 A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (Little, 1988).
-This assumption is tenuous in the context of latent class analysis.
-A non-parametric MCAR test, as provided by Jamshidian and Jalal, may be more suitable (2010).
-Unfortunately, this test was removed from the central R-repository CRAN due to lack of maintenance.
-For this tutorial, I have re-implemented it in the \texttt{mice} package as \texttt{mice::mcar()},
+As Little's classic MCAR test relies on
+the comparison of variances across groups
+with different patterns of missing data,
+it assumes normality (Little, 1988).
+This assumption is tenuous in the context of LCA.
+A non-parametric MCAR test,
+as provided by Jamshidian and Jalal,
+may be more suitable (Jamshidian \& Jalal, 2010).
+Unfortunately, this test was removed from the central R-repository CRAN
+due to lack of maintenance.
+For this tutorial, I have re-implemented it in the \texttt{mice} package
+as \texttt{mice::mcar()},
 with a fast backend in C++ and new printing and plotting methods.
 
-While we concur that investigating missingness is due dilligence,
-it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as \texttt{OpenMx} (and e.g., Mplus).
+While we concur that investigating missingness is due diligence,
+it is important to emphasize that
+missingness is adequately handled by default in many software packages for LCA,
+such as Mplus, and \texttt{OpenMx} which is the backend of \texttt{tidySEM}.
 These packages use Full Information Maximum Likelihood (FIML) estimation,
 which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data; on par with multiple imputation (Lee \& Shi, 2021).
+FIML is a best-practice solution for handling missing data;
+on par with multiple imputation (Lee \& Shi, 2021).
 FIML estimation assumes that missingness is either MCAR or MAR.
-Thus, one would typically proceed with FIML regardless of the outcome of an MCAR test.
-Although FIML does not, by default, handle missingness in exogenous variables - all indicator variables in latent class analysis are endogenous, so this is not a concern.
-
-Multiple imputation is less suitable to latent class analyses for two reasons.
-First, because latent class analyses are often computationally expensive,
+Thus, one would typically proceed with FIML,
+regardless of the outcome of an MCAR test.
+Although FIML does not, by default, handle missingness in exogenous variables
+-- all indicator variables in LCA are endogenous,
+so this is not a concern.
+
+Multiple imputation is less suitable to LCA for two reasons.
+First, because LCA methods are often computationally expensive,
 and conducting them on multiple imputed datasets may be unfeasible.
-Second, because there is no straightforward way to integrate latent class analysis results across multiple datasets.
-To conclude; our recommendation is to inspect missingness (e.g., using \texttt{mice::MCAR()})
-and report the proportion of missingness per variable (e.g., using \texttt{tidySEM::descriptives()}),
+Second, because there is no straightforward way
+to integrate LCA results across multiple datasets.
+To conclude; our recommendation is to inspect missingness
+(e.g., using \texttt{mice::MCAR()})
+and report the proportion of missingness per variable
+(e.g., using \texttt{tidySEM::descriptives()}),
 before proceeding with FIML.
@@
 \hypertarget{alternative-model-specifications}{%
-\subsubsection{Alternative model specifications}\label{alternative-model-specifications}}
-
-The different types of latent class models have different parameters.
-For example, mixture models and latent profile analyses typically have class-specific means, variances, and covariances.
-Latent growth analyses have the same parameters, but with respect to the latent growth variables.
+\subsubsection{Alternative Model Specifications}\label{alternative-model-specifications}}
+
+In order to aid researchers working with latent trajectory models,
+Van De Schoot et al. (2017) developed a protocol called
+Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
+Two of the GRoLTS guidelines (namely, 6a and 6b) refer to
+considering alternative model specifications.
+Both are discussing specific cases in latent trajectory model specification.
+The first is about whether the variance of the growth parameter
+is estimated freely or fixed,
+and the second is about whether conditional independence is assumed,
+as the researcher might also want to free-up the variance-covariance structure.
+In LCA, there are also many different ways to specify the model.
+Means, variances, and covariances between the indicators
+can either be constrained across classes, or estimated freely.
+Researchers using LCA should transparently report their chosen parametrization
+as well as discuss different parametrizations that were tested in the process.
+
+Different types of latent class models have different parameters.
+For example, mixture models and latent profile analyses
+typically have class-specific means, variances, and covariances.
+Latent growth analyses have the same parameters,
+but with respect to the latent growth variables.
 Latent class analyses with ordinal indicators have thresholds.
-All of these parameters can be freely estimated across classes, or constrained,
-or fixed (e.g., to zero).
-The total number of parameters thus scales with the number of estimated classes.
-Consequently, latent class analyses have a potentially very high number of parameters.
-As any of these parameters could be mis-specified,
+All of these parameters can be freely estimated,
+constrained to be equal across classes,
+or fixed to a certain value (e.g., to zero).
+The total number of parameters thus
+scales with the number of estimated classes.
+Consequently, LCA methods have a potentially very high number of parameters.
+As any of these parameters could be misspecified,
 it is important to consider alternative model specifications.
-However, alternative model specifications may be approached differently depending on whether an analysis is data driven (exploratory), or theoretically driven (confirmatory).
+However, alternative model specifications may be approached differently
+depending on whether an analysis is data driven (exploratory),
+or theory driven (confirmatory).
 This distinction has remained underemphasized in prior writing.
 
-Prior literature on latent class analysis has emphasized exploratory applications of the method (see Nylund, Asparouhov, \& Muthén, 2007).
-In exploratory analyses, a large number of models are typically estimated in batch,
+Prior literature on LCA has emphasized its exploratory applications
+(Nylund, Asparouhov, \& Muthén, 2007).
+In exploratory LCA,
+a large number of models are typically estimated in batch,
 with varying numbers of classes and model specifications.
-The ``correct'' model specification is then determined based on a combination of fit indices,
-significance tests,
-and interpretability.
-For latent profile analysis, the function \texttt{tidySEM::mx\_profiles(classes,\ variances,\ covariances)}
+The ``correct'' model specification is then determined
+based on a combination of fit indices,
+significance tests, and interpretability.
+For latent profile analysis,
+the function \texttt{tidySEM::mx\_profiles(classes,\ variances,\ covariances)}
 largely automates this process.
-The argument \texttt{classes} indicates which class solutions should be estimated (e.g., 1 through 6).
-The argument \texttt{variances} specifies whether variances should be \texttt{"equal"} or \texttt{"varying"} across classes.
-The argument \texttt{covariances} specifies whether covariances should be constrained to \texttt{"zero"}, \texttt{"equal"} or \texttt{"varying"} across classes.
-The means are free to vary across classes by default, although the more general function \texttt{tidySEM::mx\_mixture()} could be used to circumvent this.
-After all models have been estimated, the function \texttt{tidySEM::table\_fit()} can be used to obtain a model fit table suitable for determining the optimal model according to best practices.
-Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
+The argument \texttt{classes} indicates which
+class solutions should be estimated (e.g., 1 through 6).
+The argument \texttt{variances} specifies whether
+variances should be \texttt{"equal"} or \texttt{"varying"} across classes.
+The argument \texttt{covariances} specifies whether
+covariances should be constrained to
+\texttt{"zero"}, \texttt{"equal"} or \texttt{"varying"} across classes.
+The means are free to vary across classes by default,
+although the more general function \texttt{tidySEM::mx\_mixture()}
+could be used to circumvent this.
+After all models have been estimated,
+the function \texttt{tidySEM::table\_fit()} can be used
+to obtain a model fit table suitable for
+determining the optimal model according to best practices.
+Note, however, that this table does not include
+the bootstrapped likelihood ratio test (BLRT) by default,
 because this test is very computationally expensive.
-It is recommended to use the function \texttt{tidySEM::BLRT()} to compare a shortlist of likely candidate models based on other fit indices.
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2 log likelihood (which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance fit and model complexity.
-The BIC usually applies a stronger penalty for complexity that scales logarithmically with the sample size.
-
-Fit indices may occasionally contradict each other, so it is important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016).
-
-Confirmatory analyses typically require less comprehensive alternative model specifications.
-For example, in the context of preregistered analyses, the main models of interest may have been specified a priori.
-
-Van de Schoot and colleagues (2017) argue that there are many choices to be made in the specification of latent trajectory models,
-and that alternative
-
-\begin{verbatim}
-* Consider alternative parameterizations; e.g., free variances/covariances 
-* See Nylund
-\end{verbatim}
+It is recommended to use the function \texttt{tidySEM::BLRT()}
+to compare a shortlist of likely candidate models
+based on other fit indices.
+More on fit indices can be found in
+the Model Fit Indices subsection of this paper.
+
+Confirmatory LCA typically requires
+less comprehensive alternative model specifications.
+For example, in the context of a preregistered analysis,
+the main models of interest would have been specified a priori.
+Even in this case,
+the theoretical model could be compared to a few others to contextualize it.
 
 \hypertarget{software}{%
-\subsection{Software}\label{software}}
-
-Many software packages are available for the estimation of latent class analyses.
-Some of these packages have limited functionality, or implement specific innovations.
-Other packages implement latent class analyses in the context of a more flexible structural equation modeling framework.
-The most notable examples of the latter are the commercial programs Mplus and Latent GOLD,
+\subsubsection{Software}\label{software}}
+
+Many software packages are available for the estimation of LCA models.
+Some of these packages have limited functionality,
+or implement specific innovations.
+Other packages implement LCA in the context of
+a more flexible structural equation modeling framework.
+The most notable examples of the latter are
+the commercial programs Mplus and Latent GOLD,
 and the free open source R-package OpenMx.
-The commercial packages stand out because they offer relatively user-friendly interfaces
-and implement sensible defaults for complex analyses, including latent class analysis.
+The commercial packages stand out
+because they offer relatively user-friendly interfaces
+and implement sensible defaults for complex analyses,
+including LCA methods.
 This lowers the threshold for applied researchers to adopt such methods.
 Commercial software also has several downsides, however.
-One such downside is that use of the software is restricted to those individuals and institutions who can afford a license.
-A second downside is that the source code, being proprietary, cannot be audited, debugged, or enhanced by third parties.
-This incurs the risk that mistakes in the source code may go unnoticed, and curbs progress as software developers cannot add new functionality.
+One such downside is that use of the software
+is restricted to those individuals
+and institutions who can afford a license.
+Another downside is that the source code,
+being proprietary, cannot be audited, debugged,
+or enhanced by third parties.
+This incurs the risk that mistakes in the source code
+may go unnoticed, and curbs progress as
+software developers cannot add new functionality.
 
@@
 but not very user-friendly.
-
-New functionality in the R-package \texttt{tidySEM} seeks to lower the threshold for latent class analysis using \texttt{OpenMx}.
-It adheres to best practices in estimation and reporting, as described in this paper.
-The user interface is simple, making use of the model syntax of the widely used \texttt{lavaan} R-package.
+We directly address this limitation using the \texttt{tidySEM} R-package.
+New functionality in \texttt{tidySEM} seeks to lower the threshold
+for latent class analysis using \texttt{OpenMx}.
+It adheres to best practices in estimation and reporting,
+as described in this paper.
+The user interface is simple,
+making use of the model syntax of the widely used \texttt{lavaan} R-package.
 This syntax offers a human-readable way to specify latent variable models.
-Minor enhancements are made to simplify the specification of latent class analysis.
-
-Because of the limitations in ex tools, we set out to develop a tool that a)
-provided sensible defaults and were easy to use, but provided the option to
-access and modify all of the inputs to the model (i.e., low barrier, high
-ceiling), b) interfaced to existing tools, and are able to translate between
-what existing tools are capable of and what researchers and analysts
-carrying-out person-oriented analyses would like to specify,
-
-fully-reproducible analyses and
-
-\hypertarget{best-practices-in-estimation}{%
-\subsection{Best practices in estimation}\label{best-practices-in-estimation}}
+Minor enhancements are made to simplify the specification of LCA.
+
+Because of the limitations of the aforementioned tools,
+we set out to develop a free tool that
+provides sensible defaults and is easy to use,
+but provides the option to access and modify
+all of the model inputs (i.e., low barrier, high ceiling).
+\texttt{tidySEM} interfaces with existing tools,
+and is able to translate between what existing tools are capable of
+and what researchers and analysts
+carrying-out person-oriented analyses would like to specify.
+Furthermore,\texttt{tidySEM} facilitates fully-reproducible analyses
+and contributes to open science.
 
@@
 
-Mixture model parameters and model fit statistics can be estimated in a variety of ways.
-The choice of the estimator depends on the presence of missing values,
-sample size, number of indicators, and available computational resources (Weller, Bowen \& Faubert, 2020).
+LCA parameters and model fit statistics
+can be estimated in a variety of ways.
+The choice of the estimator
+depends on the presence of missing values,
+sample size, number of indicators,
+and available computational resources (Weller, Bowen, \& Faubert, 2020).
 A commonly used technique is maximum likelihood (ML) estimation
 with the expectation-maximization (EM) algorithm as a local optimizer.
-Imagine we are estimating two parameters, e.g.~the class-specific means \(\mu_c\) on a continuous indicator
-(ignoring the variance for now). The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes the likelihood (\(LL\)) of all observed data.
-In practice, instead of maximizing \(LL\), often \(-2*LL\) is minimized, as this offers computational advantages.
-We can think of this optimization problem as a three-dimensional landscape:
+Imagine we are estimating two parameters,
+e.g.~the class-specific means \(\mu_c\) on a continuous indicator
+(ignoring the variance for now).
+The EM algorithm will attempt to find a combination of values
+for these two parameters that maximizes
+the likelihood (\(LL\)) of all observed data.
+In practice, instead of maximizing \(LL\),
+often \(-2*LL\) is minimized,
+as this offers computational advantages.
+We can think of this optimization problem
+as a three-dimensional landscape:
 The X and Y dimensions are determined by the class-specific means,
-so \(X = \mu_1\) and \(Y = \mu_2\) - and the Z-dimension is determined by \(Z = -2*LL\).
+so \(X = \mu_1\) and \(Y = \mu_2\) -
+and the Z-dimension is determined by \(Z = -2*LL\).
 The optimizer must find the deepest ``valley'' in this landscape,
-which reflects the combination of \(\mu_1\) and \(\mu_2\) that maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble, dropped in this landscape.
-It is dropped at some random point in space, and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley, it will settle on the bottom of that valley
+which reflects the combination of \(\mu_1\) and \(\mu_2\) that
+maximizes the likelihood of the data.
+The EM optimizer behaves somewhat like a marble dropped in this landscape.
+It is dropped at some random point in space,
+and will roll into the nearest valley.
+The problem is that, once EM rolls into a valley,
+it will settle on the bottom of that valley
 (this is known as ``convergence'').
 It cannot climb out again.
-Thus, if their are multiple valleys, the risk is that the optimizer gets stuck in a shallower valley
-(a ``local optimum''), and never discovers the deepest valley (the ``global optimum'', or best solution).
-One solution to this problem is to drop many marbles at random places, compare their final \(-2*LL\) values,
-choose the solution with the lowest \(-2*LL\), and make sure that several marbles replicated this solution.
+Thus, if their are multiple valleys,
+the risk is that the optimizer gets stuck in a shallower valley (a ``local optimum''),
+and never discovers the deepest valley (the ``global optimum'', or best solution).
+One solution to this problem is to drop many marbles at random places,
+compare their final \(-2*LL\) values,
+choose the solution with the lowest \(-2*LL\),
+and make sure that several marbles replicated this solution.
 This is the ``random starts'' approach.
@@
 a ``reasonable solution'' may be used for the starting values.
-For example, if we assume that the different classes are likely to have different mean values on the indicators,
-then the K-means clustering algorithm can be used to determine these cluster centroids.
-We can compute the expected values of all model parameters by treating the K-means solution as a known class solution,
+For example, if we assume that the different classes are likely to have
+different mean values on the indicators,
+then the K-means clustering algorithm can be used to
+determine these cluster centroids.
+We can compute the expected values of all model parameters
+by treating the K-means solution as a known class solution,
 and use these as starting values for a mixture model.
-One remaining concern is that this approach may result in starting values close to a local optimum,
+One remaining concern is that this approach may result
+in starting values close to a local optimum,
 and that the EM algorithm will thus never find the global optimum.
 A second innovation addresses this concern.
-
-Instead of using EM, it is possible to use an optimizer that can climb out of a valley.
+Instead of using EM,
+it is possible to use an optimizer that can climb out of a valley.
 Simulated annealing iteratively considers some ``destination'' in the landscape,
-and compares its likelihood to the current one. If the destination likelihood is higher,
+and compares its likelihood to the current one.
+If the destination likelihood is higher,
 the estimator moves there.
-If the destination likelihood is \emph{lower}, the estimator still moves there occasionally, based on probability.
-This latter property allows it to escape local optima, and find the global optimum.
-
-By default, \texttt{tidySEM} employs this solution of deriving starting values using K-means clustering,
-and identifying the global optimum solution using simulated annealing.
+If the destination likelihood is \emph{lower},
+the estimator still moves there occasionally,
+based on probability.
+This latter property allows it to escape local optima,
+and find the global optimum.
+
+By default, \texttt{tidySEM} derives starting values using K-means clustering,
+and identifies the global optimum solution using simulated annealing.
 Once a solution has been found,
 simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for the parameters that can be used to compute standard errors.
+as EM inherently produces an asymptotic covariance matrix for
+the parameters that can be used to compute standard errors.
 Note that these defaults can be manually overridden.
 
-One recent paper suggested maximum likelihood with robust standard errors should be used when the observed indicators are not normally distributed (Spurk et al., 2020).
-This statement is incorrect, and may lead readers to believe that they must use commercial software, as robust maximum likelihood is currently only implemented in Mplus and latentGOLD.
-As explained before, mixture modeling assumes that observed data are a mixture of (multivariate) normal distributions; thus, the observed indicators will likely not be normally distributed.
-
---\textgreater{}
+A recent paper suggested maximum likelihood with robust standard errors
+should be used when the observed indicators are not normally distributed (Spurk et al., 2020).
+This statement is incorrect, and may lead readers to believe that
+they must use commercial software,
+as robust maximum likelihood is currently only implemented
+in Mplus and latentGOLD.
+As explained before,
+mixture modeling assumes that observed data are
+a mixture of (multivariate) normal distributions;
+thus, the observed indicators will likely not be normally distributed.
 
 \hypertarget{class-enumeration}{%
-\subsubsection{Class enumeration}\label{class-enumeration}}
-
-In LCA, a sequence of models is fitted to the data
+\subsubsection{Class Enumeration}\label{class-enumeration}}
+
+As explained, LCA can be done in an exploratory or in a confirmatory fashion.
+In exploratory LCA, a sequence of models is fitted to the data
 with each additional model estimating one more class than the previous model.
-The final model called the final class solution is chosen based on both theoretical and statistical criteria.
-Theory should drive the selection of indicator variables, inform the expectations and reflect on the findings.
-In addition to this, there are several statistical criteria to consider in model selection.
-These include but are not limited to likelihood ratio tests, information criteria,
-and the Bayes factor (Weller, Bowen \& Faubert, 2020).
+These models are then compared
+and the best solution is selected as the final class solution.
+In some cases, prior theory can inform the researcher
+about the number of classes to expect.
+Even in such confirmatory LCA cases,
+it is nonetheless useful to know if the theoretical model
+is markedly better than those with differing numbers of classes.
+Therefore, it may always be useful to compare different class solutions.
+
+From a sequence of models, the final class solution is chosen
+based on both theoretical and statistical criteria.
+Theory should drive the selection of indicator variables,
+inform the expectations and reflect on the findings.
+In addition to this, there are several statistical criteria
+to consider in model selection.
+These include but are not limited to likelihood ratio tests,
+information criteria,
+and the Bayes factor (Weller et al., 2020).
 
@@
 This is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as the difference in maximum log likelihoods of the two models,
-with the new degrees of freedom being the difference in their degrees of freedom.
-This statistic also follows the \(\chi^2\) distribution.
-Similar to the LR \(\chi^2\) goodness-of-fit test, we want the test statistic to be non-significant
+The likelihood ratio test statistic is computed as
+the difference in maximum log likelihoods of the two models,
+with the test degrees of freedom being
+the difference in the degrees of freedom of the two compared models.
+The test statistic follows the \(\chi^2\) distribution,
+and we want it to be non-significant
 in order to give support to the simpler model.
-The likelihood ratio test can only compare two models at a time (Lanza et al., 2003).
+The likelihood ratio test can only compare two nested models at a time (Lanza, Bray, \& Collins, 2003).
 
 \hypertarget{model-fit-indices}{%
-\subsubsection{Model fit indices}\label{model-fit-indices}}
-
-If we wish to simultaneously compare multiple models based on their relative fit,
-this can be done through a comparison of multiple information criteria.
-Examples include the Akaike information criterion (AIC), the Consistent Akaike Information Criterion (CAIC),
-the Bayesian information criterion (BIC), and the Sample-size Adjusted Bayesian Information Criterion (SABIC).
-Information criteria are a sum of a measure of fit (usually a form of the converged maximum log likelihood value)
-and a penalty for model complexity (combination of sample size and number of modeled parameters).
-As a general rule, the lower the value of an information criterion, the better the model fits the data.
-When examining several competing models, each with differing number of classes,
-we expect the information criteria values to drop with each successive model until the final class solution is reached.
-Further models with additional classes should show worse fit as information criteria reach an optimum
-before their values rise again with an increasing number of classes.
-Multiple information criteria should be compared when choosing the final class solution.
-This can be done by means of an elbow plot (for an example see Nylund-Gibson and Choi, 2018).
-Advantages of using information criteria are that we can compare multiple models at a time,
-and these models need not be nested (Masyn, 2013).
+\subsubsection{Model Fit Indices}\label{model-fit-indices}}
+
+Fit indices typically used for determining the optimal number of classes include
+the Akaike Information Criterion (AIC)
+and the Bayesian Information Criterion (BIC).
+Both information criteria are based on the -2*log likelihood
+(which is lower for better fitting models),
+and add a penalty for the number of parameters (thus incentivizing simpler models).
+This helps balance model fit and model complexity.
+The lower the value of an information criterion,
+the better the overall fit of the model.
+The BIC applies a stronger penalty for model complexity
+that scales logarithmically with the sample size.
+The literature suggests the BIC may be
+the most appropriate information criterion to use for model comparison (Nylund-Gibson \& Choi, 2018).
+Both the AIC and the BIC are available in the \texttt{tidySEM} output.
+
+Information criteria may occasionally contradict each other,
+so it is important to identify a suitable strategy to reconcile them.
+One option is to select a specific fit index before analyzing the data.
+Another option is to always prefer
+the most parsimonious model that has best fit
+according to any of the available fit indices.
+Yet another option is to incorporate information from
+multiple fit indices using the analytic hierarchy process
+(Akogul \& Erisoglu, 2016).
+Finally, one might make an elbow plot and
+compare multiple information criteria
+(Nylund-Gibson \& Choi, 2018).
+
+Another common test of model fit is
+the likelihood ratio \(\chi^2\) goodness-of-fit test.
+However, this test is not implemented in \texttt{tidySEM}.
+
+LCA studies commonly report -2*log likelihood of the final class solution.
+This is a basic fit measure used to compute most information criteria.
+However, since log likelihood is not penalized for model complexity,
+it will continuously fall with the addition of more classes.
+
+An alternative is using the bootstrapped likelihood ratio test
+which can be run using \texttt{tidySEM::BLRT()}.
+Currently, this test is computationally expensive and
+can be slow on most computers.
+A faster version of this test,
+namely an implementation of \emph{the lazy bootstrap} (Kollenburg, Mulder, \& Vermunt, 2018)
+to \texttt{tidySEM} is being developed.
 
@@
 
-Best models will divide the sample into subgroups which are internally homogeneous and externally distinct.
-Classification diagnostics give us a way to assess the degree to which this is the case.
-They are separate from the absolute and relative goodness-of-fit
-as a model can fit the data well but show poor latent class separation (Masyn, 2013).
-A fundamental concept when examining classification precision and accuracy are the posterior class probabilities.
-A probability of belonging to each latent class is computed for each individual.
-The highest posterior class probability is then determined and the individual is assigned to the corresponding class.
-We want each individual's posterior class probabilities to be high for one and low for the remaining latent classes.
-This is considered a high classification accuracy and means that the classes are distinct.
-Three important classification diagnostic measures are entropy, the average posterior probability and the modal class assignment proportion.
-
-Entropy is a summary measure of posterior class probabilities across classes and individuals.
-It ranges from 0 (model classification no better than random chance) to 1 (perfect classification).
-As a rule of thumb, values above .80 are deemed acceptable and those approaching 1 are considered ideal.
-Entropy should not be used to select a particular class solution.
-An appropriate use of entropy is that it can disqualify certain solutions if class separability is too low
-or if one of the latent classes is too small to be meaningful or to calculate descriptive statistics.
-
-The average posterior probability is a measure of classification uncertainty for each latent class.
-
-The modal class assignment proportion
-
-\hypertarget{interpreting-the-results}{%
-\subsubsection{Interpreting the results}\label{interpreting-the-results}}
-
-Class names should be chosen to accurately reflect group membership.
-Overly simplified and generalized class names may prove misleading to both audiences and researches alike
-leading to what is known as a naming fallacy (Weller, Bowen \& Faubert, 2020).
+Best models will divide the sample into subgroups
+which are internally homogeneous and externally distinct.
+Classification diagnostics give us a way to assess
+the degree to which this is the case.
+They are separate from model fit indices as
+a model can fit the data well
+but show poor latent class separation (Masyn, 2013).
+Classification diagnostics should not be used for model selection,
+but they can be used to disqualify certain solutions
+because they are uninterpretable.
+Interpretability should always be a consideration
+when considering different class solutions (Nylund-Gibson \& Choi, 2018).
+
+Three important classification diagnostics provided by \texttt{tidySEM} are:
+(1) \emph{the minimum} and \emph{maximum percentage of the sample assigned to a particular class},
+(2) \emph{the range of the posterior class probabilities by most likely class membership},
+and (3) \emph{entropy}.
+All three are based on posterior class probabilities.
+
+The posterior class probability is a measure of classification uncertainty
+which can be computed for each individual,
+or averaged for each latent class.
+When the posterior class probability is computed for each individual in the dataset,
+it represents each person's probability of belonging to each latent class.
+For each person, the highest posterior class probability is then determined
+and the individual is assigned to the corresponding class.
+We want each individual's posterior class probabilities to be
+high for one and low for the remaining latent classes.
+This is considered a high classification accuracy
+and means that the classes are distinct.
+To obtain posterior class probabilities, run \texttt{tidySEM::class\_prob()}.
+This function produces output comprised of several elements:
+
+\texttt{\$sum.posterior} is a summary table of the posterior class probabilities
+indicating what proportion of the data contributes to each class.
+
+\texttt{\$sum.mostlikely} is a summary table of the most likely class membership
+based on the highest posterior class probability.
+From this table, we compute the minimum and maximum
+percentage of the sample assigned to a particular class,
+, i.e.~\textbf{n\_min}
+(the smallest class proportion based on the posterior class probabilities)
+and \textbf{n\_max}
+(the largest class proportion based on the posterior class probabilities).
+We are especially interested in \textbf{n\_min} as
+if it is very small and comprised of few observations,
+the model for that group might not be locally identified.
+It may be impossible to calculate descriptive statistics for such a small class.
+Estimating LCA parameters on small subsamples
+might lead to bias in the results.
+Therefore, we advise caution when dealing with small classes.
+
+\texttt{\$mostlikely.class} is a table with rows representing
+the class the person was assigned to,
+and the columns indicating the average posterior probability.
+The diagonal represents the probability that
+observations in each class will be correctly classified.
+If any of the values on the diagonal of this table is low,
+we might consider not to interpret that solution.
+In \texttt{tidySEM} we use the diagonal to compute the range of
+the posterior class probabilities by most likely class membership
+which consists of
+the lowest class posterior probability (\textbf{prob\_min}),
+and the highest posterior probability (\textbf{prob\_max}).
+Both \textbf{prob\_min} and \textbf{prob\_max} can be used to
+disqualify certain class solutions,
+and are a convenient way to summarize class separation in LCA.
+We want both \textbf{prob\_min} and \textbf{prob\_max} to be high as
+that means that for all classes
+the people who were assigned to that class
+have a high probability of being there.
+\textbf{prob\_min} is especially important as it can diagnose
+if there is a class with low posterior probabilities
+which could make one reconsider that class solution.
+
+\texttt{\$avg.mostlikely} contains the average posterior probabilities for each class,
+for the subset of observations with most likely class of 1:k,
+where k is the number of classes.
+
+\texttt{\$individual} is the individual posterior probability matrix,
+with dimensions n (number of cases in the data) x k (number of classes).
+Individual class probabilities are often useful for researchers
+who wish to do follow up analyses.
+
+Entropy is a summary measure of posterior class probabilities
+across classes and individuals.
+It ranges from 0 (model classification no better than random chance)
+to 1 (perfect classification).
+As a rule of thumb, values above .80 are deemed acceptable
+and those approaching 1 are considered ideal.
+An appropriate use of entropy is that it can disqualify certain solutions
+if class separability is too low.
+Entropy was not built for nor should it be used for
+model selection during class enumeration (Masyn, 2013).
+
+\textbf{n\_min}, \textbf{n\_max}, \textbf{prob\_min}, \textbf{prob\_max}, and \textbf{entropy} and can be obtained using \texttt{tidySEM::table\_fit()}.
+
+\hypertarget{interpreting-class-solutions}{%
+\subsubsection{Interpreting Class Solutions}\label{interpreting-class-solutions}}
+
+An important outcome of LCA are conditional item probabilities,
+also known as class-specific item probabilities (Masyn, 2013),
+conditional response, or conditional solution probabilities (Geiser, 2012).
+They indicate the probability of an item being endorsed
+given that the observation belongs to a particular latent class.
+Conditional item probabilities can be obtained using \texttt{tidySEM::table\_prob()}.
+If a particular item is endorsed by two or more classes at markedly different rates,
+it is said to discriminate well between the classes
+and is consequently considered a good indicator.
+Classes are considered highly homogeneous with respect to an item
+when for a particular item
+there is a distinct difference in conditional item probabilities
+for two or more classes.
+For instance, if an item is endorsed below 30\% for one class and
+above 70\% for another class,
+the classes have high homogeneity with respect to this item (Masyn, 2013).
+Conditional item probabilities are the analogue of mean and standard deviation
+when the indicators are binary or ordinal.
+
+A problem which can occur is that of inadmissible solutions.
+With binary indicators, LCA is modelling a cross-table with all the predictors.
+The problem with such cross-tables is that they will often contain empty cells,
+i.e.~combinations of responses that never occur together.
+This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
+Such boundary parameter estimates could indicate that the solution is invalid (Geiser, 2012).
+Boundary parameter estimates can also happen with continuous indicators.
+For instance, if we have a zero-inflated normal distribution and a two class solution,
+one class might have the mean of zero and its standard deviation cannot be determined
+since there is little variance.
+This too could be a sign of an invalid solution,
+warn that too many classes were extracted, or indicate a local optimum (Geiser, 2012).
 
 \hypertarget{label-switching}{%
-\subsubsection{Label switching}\label{label-switching}}
+\subsubsection{Label Switching}\label{label-switching}}
 
 The final class solution will usually discover and enumerate several classes.
-The class ordering however is completely arbitrary.
-The class labeled as Class 1 in one solution may become Class 2 or Class 3 in another model,
+The class ordering, however, is completely arbitrary.
+The class labeled as Class 1 in one solution
+may become Class 2 or Class 3 in another model,
 even when the only difference between the models is in their starting values.
@@
 
+The order of clusters is nondeterministic when using K-means in \texttt{tidySEM}.
+Therefore label switching is still a consideration.
+A simple solution to this is setting a random seed number one line prior to fitting the model.
+We advise \texttt{tidySEM} users to always do so in order to circumvent label switching.
+
+Class names should be chosen to accurately reflect group membership.
+Overly simplified and generalized class names may prove misleading
+to both audiences and researches alike
+leading to what is known as a naming fallacy (Weller et al., 2020).
+
 \hypertarget{best-practices-in-reporting}{%
-\subsection{Best practices in reporting}\label{best-practices-in-reporting}}
-
-Among studies using LCA, reporting practices vary significantly (Weller, Bowen \& Faubert, 2020).
-Various authors have tried to better and standardize ways of reporting LCA (e.g.~Masyn, 2013; Weller, Bowen \& Faubert, 2020).
-Building on their work, we provide guidelines with a strong emphasis on scientific reproducibility and transparency
-better known as the open science framework (OSF). Open science framework promotes scientific openness, reproducibility and integrity
-through an establishment of guidelines for the scientific process (Foster \& Deardorff, 2017)
-To this end, van Lissa and colleagues (2020) developed WORCS, a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects based on the TOP-guidelines developed by Nosek and colleagues (2015).
-It can be easily implemented in R in form of an R package which facilitates preregistration, article drafting,
-version control, citation and formatting, among others (Van Lissa et al., 2020)
-
-\begin{verbatim}
-* Use comprehensive citation; this includes referencing the software used.
-* Share code and (if possible) data. Van Lissa and colleagues suggest sharing synthetic data in case the original data cannot be shared, and provide functions to generate such synthetic data.
-* Ideally, make the entire research project reproducible so that others may download it and reproduce it with just one click; for guidance, see Van Lissa and colleagues.
-\end{verbatim}
+\subsection{Best Practices in Reporting}\label{best-practices-in-reporting}}
+
+Among studies using LCA, reporting practices vary significantly (Weller et al., 2020).
+Various authors have tried to improve and standardize ways of reporting LCA
+(e.g. Masyn, 2013; Weller et al., 2020),
+but more work is needed.
+Van Lissa et al. (2020) developed WORCS,
+a workflow for open reproducible code in science.
+WORCS consists of step-by-step guidelines for research projects
+based on the TOP-guidelines developed by Nosek et al. (2015).
+WORCS workflow can be easily implemented in R in form of
+an R package which facilitates preregistration, article drafting,
+version control, citation and formatting, among others (Van Lissa et al., 2020).
+
+TOP-guidelines emphasise the use of comprehensive citation
+(including referencing the software used in the analysis),
+as well as code and data sharing wherever possible (Nosek et al., 2015).
+Van Lissa et al. (2020) suggest sharing synthetic data in case
+the original data cannot be shared,
+and provide functions to generate such synthetic data.
+Ideally, the entire research project is made reproducible
+so that others may download it and reproduce it with just one click;
+for guidance, see Van Lissa et al. (2020).
 
@@
 how important it is that analyses can be reproduced and audited.
-In line with open science principles, one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that user-friendly methods for estimating latent class analyses have predominantly been available in commercial software packages (e.g., \emph{Mplus} and \emph{Latent GOLD}).
-A potential downside of commercial software is that it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because the underlying source code is proprietary.
-To overcome these limitations, the present paper introduces new user-friendly functions in the \texttt{tidySEM} R-package that can be used to estimate a wide range of latent class analysis models using the free, open-source R-package \texttt{OpenMx}.
-The reporting guidelines described in this paper are adopted in \texttt{tidySEM} by default.
-The \texttt{tidySEM} R-package thus makes advanced mixture modeling based on best practices widely accessible,
-and facilitates the adoption of the estimation and reporting guidelines described in this paper.
-
-\hypertarget{best-practices-in-visualization}{%
-\subsection{Best practices in visualization}\label{best-practices-in-visualization}}
+In line with open science principles,
+one of the suggested reporting standards relates to reproducible code.
+In this context, it is important to note that
+user-friendly methods for estimating LCA models
+have predominantly been available in commercial software packages
+(e.g., \emph{Mplus} and \emph{Latent GOLD}).
+A potential downside of commercial software is that
+it restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because
+the underlying source code is proprietary.
+To overcome these limitations,
+the present paper introduces new user-friendly functions in
+the \texttt{tidySEM} R-package that can be used to
+estimate a wide range of LCA models
+using the free, open-source R-package \texttt{OpenMx}.
+The reporting guidelines described in this paper
+are adopted in \texttt{tidySEM} by default.
+The \texttt{tidySEM} R-package thus makes advanced mixture modeling
+based on best practices widely accessible,
+and facilitates the adoption of the estimation and
+reporting guidelines described in this paper.
+
+\hypertarget{visualization}{%
+\subsubsection{Visualization}\label{visualization}}
+
+Plots can greatly improve the interpretability of LCA models.
+There are several stages where this is an important consideration.
+
+First, in class enumeration,
+we want to compare several competing class solutions.
+This can be done by means of the AIC and BIC.
+Here, we suggest using an elbow plot
+with the \(K-class\ solution\) on the X-axis,
+and information criteria value on the Y-axis.
+We show an implementation of such an elbow plot
+in our Tutorial subsection.
+
+Second, once we have decided on the final class solution,
+we want to interpret the response patterns
+on the indicators for each latent class.
+While \texttt{tidySEM} can create a table showing
+the probability of each item's response endorsement
+(using \texttt{tidySEM::table\_prob}),
+it may be easier to inspect these probabilities visually.
+For this reason, we created \texttt{ggplot2::plot\_prob()}.
+The resulting shows response patterns
+on all indicators for each group.
+We give an example of this plot in the Tutorial subsection.
 
@@
 
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
---\textgreater{}
-
-\hypertarget{starting-values}{%
-\subsubsection{Starting values}\label{starting-values}}
+This is an example of an exploratory LCA using \texttt{tidySEM}.
+
+\hypertarget{loading-the-dataset}{%
+\subsection{Loading the Dataset}\label{loading-the-dataset}}
+
+In this example, we use \emph{data\_mix\_ordinal},
+a simulated data for mixture modelling with ordinal indicators.
+This dataset is built-in to \texttt{tidySEM}.
+For more information about the dataset,
+type \texttt{?tidySEM::data\_mix\_ordinal} into the R console
+after loading the \texttt{tidySEM} using library().
+
+We load the dataset and convert the indicators into ordered factors.
+
+\hypertarget{exploring-the-data}{%
+\subsection{Exploring the Data}\label{exploring-the-data}}
+
+An important step to preceding any statistical analysis is data exploration.
+Here we use \texttt{tidySEM::descriptives()}
+to describe the dataset we are using.
+
+As we can see, the output includes various descriptives of our dataset.
+Special attention should be paid to examining the pattern of missingness,
+as discussed in the \emph{Handling missing data} section of this paper.
+In our example, we see that there are no missing values,
+hence we proceed with our analysis.
+
+\hypertarget{conducting-latent-class-analysis}{%
+\subsection{Conducting Latent Class Analysis}\label{conducting-latent-class-analysis}}
+
+Before we fit a series of LCA models, we set a random seed
+using \texttt{set.seed()}.
+The seed number can be any digit.
+This is an important step as
+there is some inherent randomness in the LCA computations,
+and having the same seed number ensures that
+two separate researcher obtain exactly the same results
+when fitting LCA models.
+
+Finally, we reach the step of fitting LCA models.
+To do so, we use \texttt{tidySEM::mx\_lca()}
+which takes data and number of classes as inputs.
+In our example, we want to fit 1 to 4 class solutions
+and compare their output.
+Depending on your computer's computational power,
+this might take a while.
+
+\hypertarget{class-enumeration-1}{%
+\subsection{Class Enumeration}\label{class-enumeration-1}}
+
+In class enumeration,
+we want to compare a sequence of LCA models fitted to the data.
+To aid the process, we create a model fit table using \texttt{tidySEM::table\_fit()}
+with the results object as the input.
+As the output contains a lot of information on each of the four fitted models,
+we select a subset of helpful model fit indices and classification diagnostics.
+
+Our selection of fit indices and classification diagnostics includes:
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1220}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8780}}@{}}
+\caption{\label{tab:unnamed-chunk-6}Selection of Fit Indices and Classification Diagnostics}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Selection
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Description
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Selection
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Description
+\end{minipage} \\
+\midrule()
+\endhead
+Name & the \(K\)-class solution \\
+LL & the -2*log-likelihood of each model \\
+AIC & the Akaike Information Criterion value \\
+BIC & the Bayesian Information Criterion value \\
+prob\_min & the lowest posterior class probability by most likely class membership \\
+prob\_max & the highest posterior class probability by most likely class membership \\
+n\_min & the lowest class proportion based on the posterior class probabilities \\
+n\_max & the highest class proportion based on the posterior class probabilities \\
+\bottomrule()
+\end{longtable}
+
+We discussed several possible strategies to select the final class solution.
+Here, we apply our own.
+
+To aid our interpretation of the results,
+we create an elbow plot showing the trends in information criteria across four models.
+
+From the elbow plot, we see that AIC has a lower penalty for model complexity.
+However, we are more interested in the BIC values, which are similar
+for the one, two and three-class solutions, but the four-class solution fits significantly worse.
+For this reason, we eliminate the four-class solution from the selection process.
+
+Then we examine the model fit table.
+As expected, the -2*log likelihood falls successively with each added class.
+As previously stated, classification diagnostics should not be used for model selection,
+but they can be used to disqualify certain solutions because they are uninterpretable.
+We see that prob\_min for the four-class solution is low,
+knowing that this solution also has a high BIC, we disqualify this solution.
+
+Out of the remaining three solutions,
+we notice that entropy is the highest for the three-class solution,
+and it has a satisfactory prob\_min and n\_min.
+Based on this, we retain the three class solution in model selection.
+Note that entropy for the one-class solution will always equal to one,
+as it is 100\% true that every case is in that class.
+Based on the low entropy of the two-class solution,
+we eliminate this model.
+
+Finally, when comparing the one and three-class solutions,
+we inspect the information criteria.
+For BIC, the one-class solution fits better, but the difference is marginal.
+AIC tells us that the added complexity
+of having three classes still explains the data better
+than a one-class solution.
+Therefore, we select the three-class solution as our final-class solution.
+
+\hypertarget{interpreting-the-final-class-solution}{%
+\subsection{Interpreting the Final Class Solution}\label{interpreting-the-final-class-solution}}
+
+To aid our understanding of the final class solution,
+we use \texttt{ggplot2::plot\_prob()}
+with the results of the three-class model as the input.
+The resulting graph shows response patterns
+on all the indicators for each group.
+
+If we want to know the probability of each response option's endorsement
+for each class, we can use \texttt{tidySEM::table\_prob}.
+These are thresholds for ordinal dependent variables in the probability scale.
+
+In the plot, we can see the distributions of the response probabilities on the indicators
+for each of the three classes.
+For instance, we see that in Class 1 the most common response to u2 is 2,
+while in Class 2 and Class 3 this is 0.
+We can also see that response 1 is a rare response not forming the majority in
+any class.
+Class 2 distinguishes itself because the majority scores the response 0 category of u3 and u4,
+while in Class 1 and 2 this is not the case.
+Class 3 distinguishes itself because the most common response to u3 and u4 is 3.
+
+We can also interpret the response patterns numerically.
+
+\hypertarget{extracting-posterior-class-probabilities}{%
+\subsection{Extracting Posterior Class Probabilities}\label{extracting-posterior-class-probabilities}}
+
+Another step is to extract posterior class probabilities.
+This is done by the use of \texttt{TidySEM::class\_prob}
+with the results of the final class solution as the input.
+
+\newpage
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-akogulComparisonInformationCriteria2016}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-akogul_comparison_2016}{}}%
 Akogul, S., \& Erisoglu, M.
 (2016).
-A {Comparison} of {Information Criteria} in {Clustering Based} on {Mixture} of {Multivariate Normal Distributions}.
+A {Comparison} of {Information} {Criteria} in {Clustering} {Based} on {Mixture} of {Multivariate} {Normal} {Distributions}.
 \emph{Mathematical and Computational Applications}, \emph{21}(3), 34.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-baughmanMixtureModelAnalysis2006}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-baughman_mixture_2006}{}}%
 Baughman, A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-figueiredoUnsupervisedLearningFinite2002}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-figueiredo_unsupervised_2002}{}}%
 Figueiredo, M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-jamshidianTestsHomoscedasticityNormality2010}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-geiser_data_2012}{}}%
+Geiser, C. (2012). \emph{Data {Analysis} with {Mplus}}. Guilford Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-geiser_is_2014}{}}%
+Geiser, C., \& Wurpts, I.
+(2014).
+Is adding more indicators to a latent class analysis beneficial or detrimental?
+{Results} of a {Monte} {Carlo} study.
+\emph{Frontiers in Psychology: Quantitative Psychology and Measurement}, \emph{5}.
+https://doi.org/\url{https://doi.org/10.3389/fpsyg.2014.00920}
+
+\leavevmode\vadjust pre{\hypertarget{ref-giang_using_2008}{}}%
+Giang, M.
+T., \& Graham, S.
+(2008).
+Using latent class analysis to identify aggressors and victims of peer harassment.
+\emph{Aggressive Behavior}, \emph{34}(2), 203--213.
+\url{https://doi.org/10.1002/ab.20233}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hennig_handbook_2015}{}}%
+Hennig, C., Meila, M., Murtagh, F., \& Rocci, R. (2015). \emph{Handbook of {Cluster} {Analysis}}. 28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hopfer_social_2014}{}}%
+Hopfer, S., Tan, X., \& Wylie, J.
+L.
+(2014).
+A {Social} {Network}--{Informed} {Latent} {Class} {Analysis} of {Patterns} of {Substance} {Use}, {Sexual} {Behavior}, and {Mental} {Health}: {Social} {Network} {Study} {III}, {Winnipeg}, {Manitoba}, {Canada}.
+\emph{American Journal of Public Health}, \emph{104}(5), 834--839.
+\url{https://doi.org/10.2105/AJPH.2013.301833}
+
+\leavevmode\vadjust pre{\hypertarget{ref-jamshidian_tests_2010}{}}%
 Jamshidian, M., \& Jalal, S.
 (2010).
-Tests of {Homoscedasticity}, {Normality}, and {Missing Completely} at {Random} for {Incomplete Multivariate Data}.
+Tests of {Homoscedasticity}, {Normality}, and {Missing} {Completely} at {Random} for {Incomplete} {Multivariate} {Data}.
 \emph{Psychometrika}, \emph{75}(4), 649--674.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-leeComparisonFullInformation2021}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-van_kollenburg_lazy_2018}{}}%
+Kollenburg, G. H. van, Mulder, J., \& Vermunt, J. K. (2018). \emph{The {Lazy} {Bootstrap}. {A} {Fast} {Resampling} {Method} for {Evaluating} {Latent} {Class} {Model} {Fit}}. 23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lanza_introduction_2003}{}}%
+Lanza, S.
+T., Bray, B.
+C., \& Collins, L.
+M.
+(2003).
+An {Introduction} to {Latent} {Class} and {Latent} {Transition} {Analysis}.
+In \emph{Handbook of {Psychology}: {Research} {Methods} in {Psychology}} (2nd ed., Vol.
+2, pp.
+690--712).
+John Wiley \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lanza_latent_2013}{}}%
+Lanza, S.
+T., Tan, X., \& Bray, B.
+C.
+(2013).
+Latent {Class} {Analysis} {With} {Distal} {Outcomes}: {A} {Flexible} {Model}-{Based} {Approach}.
+\emph{Structural Equation Modeling : A Multidisciplinary Journal}, \emph{20}(1), 1--26.
+\url{https://doi.org/10.1080/10705511.2013.742377}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lee_comparison_2021}{}}%
 Lee, T., \& Shi, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-littleTestMissingCompletely1988}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-little_test_1988}{}}%
 Little, R.
@@
 (1988).
-A {Test} of {Missing Completely} at {Random} for {Multivariate Data} with {Missing Values}.
+A {Test} of {Missing} {Completely} at {Random} for {Multivariate} {Data} with {Missing} {Values}.
 \emph{Journal of the American Statistical Association}, \emph{83}(404), pp.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-normanLikertScalesLevels2010}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-macgregor_symptom_2021}{}}%
+MacGregor, A.
+J., Dougherty, A.
+L., D'Souza, E.
+W., McCabe, C.
+T., Crouch, D.
+J., Zouris, J.
+M., \ldots{} Fraser, J.
+J.
+(2021).
+Symptom profiles following combat injury and long-term quality of life: A latent class analysis.
+\emph{Quality of Life Research}, \emph{30}(9), 2531--2540.
+\url{https://doi.org/10.1007/s11136-021-02836-y}
+
+\leavevmode\vadjust pre{\hypertarget{ref-masyn_latent_2013}{}}%
+Masyn, K.
+E.
+(2013).
+Latent {Class} {Analysis} and {Finite} {Mixture} {Modeling}.
+In \emph{The {Oxford} {Handbook} of {Quantitative} {Methods}}: \emph{Vol.} \emph{2: Statistical Analysis} (p.
+551).
+Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-norman_likert_2010}{}}%
 Norman, G.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nylundDecidingNumberClasses2007}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-nosek_promoting_2015}{}}%
+Nosek, B.
+A., Alter, G., Banks, G.
+C., Borsboom, D., Bowman, S.
+D., Breckler, S.
+J., \ldots{} Yarkoni, T.
+(2015).
+Promoting an open research culture.
+\emph{Science}, \emph{348}(6242), 1422--1425.
+\url{https://doi.org/10.1126/science.aab2374}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nozadi_moderating_2016}{}}%
+Nozadi, S.
+S., Troller-Renfree, S., White, L.
+K., Frenkel, T., Degnan, K.
+A., Bar-Haim, Y., \ldots{} Fox, N.
+A.
+(2016).
+The {Moderating} {Role} of {Attention} {Biases} in understanding the link between {Behavioral} {Inhibition} and {Anxiety}.
+\emph{Journal of Experimental Psychopathology}, \emph{7}(3), 451--465.
+\url{https://doi.org/10.5127/jep.052515}
+
+\leavevmode\vadjust pre{\hypertarget{ref-nylund_deciding_2007}{}}%
 Nylund, K.
@@
 (2007).
-Deciding on the {Number} of {Classes} in {Latent Class Analysis} and {Growth Mixture Modeling}: {A Monte Carlo Simulation Study}.
+Deciding on the {Number} of {Classes} in {Latent} {Class} {Analysis} and {Growth} {Mixture} {Modeling}: {A} {Monte} {Carlo} {Simulation} {Study}.
 \emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{14}(4), 535--569.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nylund-gibsonTenFrequentlyAsked2018}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-nylund-gibson_ten_2018}{}}%
 Nylund-Gibson, K., \& Choi, A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rubinInferenceMissingData1976}{}}%
-Rubin, D. B. (1976). Inference and {Missing Data}. \emph{Biometrika}, \emph{63}(3), 581--592. \url{https://doi.org/10.2307/2335739}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vandeschootGRoLTSChecklistGuidelinesReporting2017}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-rubin_inference_1976}{}}%
+Rubin, D. B. (1976). Inference and {Missing} {Data}. \emph{Biometrika}, \emph{63}(3), 581--592. \url{https://doi.org/10.2307/2335739}
+
+\leavevmode\vadjust pre{\hypertarget{ref-scrucca_mclust_2016}{}}%
+Scrucca, L., Fop, M., Murphy, T.
+B., \& Raftery, A.
+E.
+(2016).
+\href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5096736}{Mclust 5: {Clustering}, {Classification} and {Density} {Estimation} {Using} {Gaussian} {Finite} {Mixture} {Models}}.
+\emph{The R Journal}, \emph{8}(1), 289--317.
+
+\leavevmode\vadjust pre{\hypertarget{ref-spurk_latent_2020}{}}%
+Spurk, D., Hirschi, A., Wang, M., Valero, D., \& Kauffeld, S.
+(2020).
+Latent profile analysis: {A} review and {``how to''} guide of its application within vocational behavior research.
+\emph{Journal of Vocational Behavior}, \emph{120}, 103445.
+\url{https://doi.org/10.1016/j.jvb.2020.103445}
+
+\leavevmode\vadjust pre{\hypertarget{ref-van_de_schoot_grolts-checklist_2017}{}}%
 Van De Schoot, R., Sijbrandij, M., Winter, S.
@@
 (2017).
-The {GRoLTS-Checklist}: {Guidelines} for {Reporting} on {Latent Trajectory Studies}.
+The {GRoLTS}-{Checklist}: {Guidelines} for {Reporting} on {Latent} {Trajectory} {Studies}.
 \emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{24}(3), 451--467.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vanlissaWORCSWorkflowOpen2020}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_worcs_2020}{}}%
 Van Lissa, C.
@@
 (2020).
-\emph{{WORCS}: {A Workflow} for {Open Reproducible Code} in {Science}}.
+\emph{{WORCS}: {A} {Workflow} for {Open} {Reproducible} {Code} in {Science}}.
 \url{https://doi.org/10.17605/OSF.IO/ZCVBS}
 
-\leavevmode\vadjust pre{\hypertarget{ref-vermuntKmeansMayPerform2011}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_worcs_2021}{}}%
+Van Lissa, C.
+J., Brandmaier, A.
+M., Brinkman, L., Lamprecht, A.-L., Peikert, A., Struiksma, M.
+E., \& Vreede, B.
+M.
+I.
+(2021).
+{WORCS}: {A} workflow for open reproducible code in science.
+\emph{Data Science}, \emph{4}(1), 29--49.
+\url{https://doi.org/10.3233/DS-210031}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_k-means_2011}{}}%
 Vermunt, J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vermuntj.k.LatentClassAnalysis2004}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_latent_2017}{}}%
+Vermunt, J.
+K.
+(2017).
+Latent {Class} {Modeling} with {Covariates}: {Two} {Improved} {Three}-{Step} {Approaches}.
+\emph{Political Analysis}, \emph{18}(4), 450--469.
+\url{https://doi.org/10.1093/pan/mpq025}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_jk_latent_2004}{}}%
 Vermunt, J.K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.F., \& Department of Methodology and Statistics.
@@
 549--553).
-{Sage}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-wuAbuseDependencePrescription2011}{}}%
+Sage.
+Retrieved from \url{https://research.tilburguniversity.edu/en/publications/0caedd00-27c1-42bd-bb4e-d1dcb0864956}
+
+\leavevmode\vadjust pre{\hypertarget{ref-weller_latent_2020}{}}%
+Weller, B.
+E., Bowen, N.
+K., \& Faubert, S.
+J.
+(2020).
+Latent {Class} {Analysis}: {A} {Guide} to {Best} {Practice}.
+\emph{Journal of Black Psychology}, \emph{46}(4), 287--311.
+\url{https://doi.org/10.1177/0095798420930932}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wu_abuse_2011}{}}%
 Wu, L.-T., Woody, G.

### 2022-10-28 manuscript.Rmd
edits from Mauricio

@@
 shorttitle        : "BEST PRACTICES LCA"
-
 author: 
@@
     email         : "c.j.vanlissa@gmail.com"
-
+  - name          : "Mauricio Garnier-Villarreal"
+    affiliation   : "3"
+    corresponding : no   
+    email         : "m.garniervillarreal@vu.nl"
 affiliation:
@@
     institution   : "Open Science Community Utrecht"
-
+  - id            : "3"
+    institution   : "Vrije Universiteit Amsterdam, Sociology"
 author_note: |
@@
   3582CH Utrecht, The Netherlands
-
 abstract: |
   Latent class analysis is a family of techniques for identifying groups in data based
-  on a parametric model. Its examples include mixture models, latent profile
-  analysis, latent class analysis with ordinal indicators, and latent class
+  on a parametric model. Its examples include mixture models, latent class
+  analysis (with continuous and/or ordinal indicators), and latent class (mixture)
   growth analysis. Despite the popularity of this technique, there is limited
@@
   the `OpenMx` R-package.
-
 keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
-wordcount         : "8206"
+wordcount         : "9001"
 bibliography      : "mixture.bib"
@@
 mask              : no
-
 class             : "man"
 output            : papaja::apa6_pdf
-
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-Latent class analysis (LCA) is an umbrella term 
-that refers to a number of techniques for estimating unobserved group membership
-based on a parametric model of one or more observed indicators of group membership.
+Latent class analysis (LCA) is an umbrella term that refers to a number
+of techniques for estimating unobserved group membership based on a
+parametric model of one or more observed indicators of group membership.
 The types of LCA have become quite popular across scientific fields,
-most notably finite Gaussian mixture modeling and latent profile analysis.
-@vermunt_jk_latent_2004 defined LCA more generally as virtually any statistical model 
-where “some of the parameters […] differ across unobserved subgroups”.
-
-Despite its popularity, there is a lack of standards for estimating and reporting LCA.
-While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines 
-for a specific type of LCA known as latent growth models,
-general reporting guidelines for all LCA are still lacking.
-This complicates manuscript review and assessment of the quality of published studies,
-and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature 
-by suggesting updated guidelines for estimation and reporting of LCA, 
-based on current best practices.
-Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, 
-this paper exclusively relies on free, open source software in R.
-Our goal is to make best-practices in LCA widely accessible.
+most notably finite Gaussian mixture modeling and latent profile
+analysis. @vermunt_jk_latent_2004 defined LCA more generally as
+virtually any statistical model where "some of the parameters [...]
+differ across unobserved subgroups".
+
+Despite its popularity, there is a lack of standards for estimating and
+reporting LCA. While @van_de_schoot_grolts-checklist_2017 developed
+reporting guidelines for a specific type of LCA known as latent growth
+models, general reporting guidelines for all LCA are still lacking. This
+complicates manuscript review and assessment of the quality of published
+studies, and introduces a risk of misapplications of the technique. The
+present paper seeks to address this gap in the literature by suggesting
+updated guidelines for estimation and reporting of LCA, based on current
+best practices. Importantly, in order to lower the barrier of entry and
+ensure reproducibility of all examples, this paper exclusively relies on
+free, open source software in R. Our goal is to make best-practices in
+LCA widely accessible.
 
@@
 
-Latent class analysis is a group of methods for estimating unobserved groups
-based on a parametric model of observed indicators of group membership.
-The concept of LCA can be understood in different ways.
-A mixture model assumes that the study population
-is composed of $K$ subpopulations or classes.
-It further assumes that the observed data are a mixture of data generated by class-specific models.
-The simplest univariate model is a normal distribution, 
-which can be described with two parameters: the mean and the variance.
-Commonly, the same model is estimated across all classes, 
-but with different parameters for each class (i.e., class-specific means and variances).
-Mixture modeling then estimates both the parameters for each class, 
-and the probability that an individual belongs to each class.
-
-As an illustrative example, imagine that a detective wants to know 
-if it would be possible to use mixture modeling to identify the sex of a suspect,
-based on footprints found at the crime scene.
-To test the feasibility of this approach,
-the detective records the shoe sizes and sex of 100 volunteers.
-The resulting observed data look like this:
+Latent class analysis is a group of methods for estimating unobserved
+groups based on a parametric model of observed indicators of group
+membership. The concept of LCA can be understood in different ways. A
+mixture model assumes that the study population is composed of $K$
+subpopulations or classes. It further assumes that the observed data are
+a mixture of data generated by class-specific models. The simplest
+univariate model is a normal distribution, which can be described with
+two parameters: the mean and the variance. Commonly, the same model is
+estimated across all classes, but with different parameters for each
+class (i.e., class-specific means and variances). Mixture modeling then
+estimates both the parameters for each class, and the probability that
+an individual belongs to each class. This way, describing the whole
+sample with a single mean and variance is a special case where we assume
+the sample is homogeneous in function of these parameters.
+
+As an illustrative example, imagine that a detective wants to know if it
+would be possible to use mixture modeling to identify the sex of a
+suspect, based on footprints found at the crime scene. To test the
+feasibility of this approach, the detective records the shoe sizes and
+sex of 100 volunteers. The resulting observed data look like this:
 
@@
 
-The distribution is evidently bimodal, which bodes well for the intended mixture model.
-In this case, the number of classes is known a-priori.
-When estimating a two-class mixture model, the detective observes that the model estimates
-the mean shoe size of the two groups are equal to `r report(estparam[1], equals = FALSE)` 
-and  `r report(estparam[2], equals = FALSE)`,
-which is close to the true means of the two groups, 
-namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
-When tabulating estimated group membership against observed (known) group membership, 
-it can be seen that women are classified with a high degree of accuracy, but men are not:
+The distribution is evidently bimodal, which bodes well for the intended
+mixture model. In this case, the number of classes is known a-priori.
+When estimating a two-class mixture model, the detective observes that
+the model estimates the mean shoe size of the two groups are equal to
+`r report(estparam[1], equals = FALSE)` and
+`r report(estparam[2], equals = FALSE)`, which is close to the true
+means of the two groups, namely `r report(obsparam[1], equals = FALSE)`
+and `r report(obsparam[2], equals = FALSE)`. When tabulating estimated
+group membership against observed (known) group membership, it can be
+seen that women are classified with a high degree of accuracy, but men
+are not:
 
@@
 
-A LCA is like confirmatory factor analysis, 
-except that the continuous latent variable is substituted with a categorical latent variable.
-One difference between the two techniques is that 
-factor analysis can be considered as a way to group observed *variables* into latent constructs, 
-and LCA groups *individuals* into classes.
-In line with this distinction, LCA is sometimes referred to as a "person-centered" technique, 
-and factor analysis as a "variable-centered" technique [@nylund-gibson_ten_2018].
-
-When the focus is on the model parameters in each group,
-LCA can be thought of as similar to a multi-group structural equation model.
-The main distinction is that group membership is not known a-priori,
-but is instead estimated -- with measurement error -- based on the data.
-Whereas in a multi-group model, the data are split by group and treated as independent samples,
-in LCA, all cases contribute to the estimation of all parameters in all groups.
-The relative contribution of each case to the parameters of each group
-is determined by that case's posterior probability of belonging to that group.
-
-When the focus is on each individual's estimated class membership,
-LCA can be thought of as a type of clustering algorithm.
-In line with this perspective, LCA is sometimes described as "model-based
-clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
-Many clustering algorithms apply some recursive splitting algorithm to the data.
-By contrast "model-based" clustering refers to the fact that LCA 
-estimates cluster membership based on a parametric model.
-Specifically, the posterior class probability that an individual belongs to a latent class
-can be computed from the likelihood of that individual's observed data under given the class-specific model.
+LCA is like confirmatory factor analysis, except that the continuous
+latent variable is substituted with a categorical latent variable. Both
+assume that the indicators are related, and these relations can be
+represented by the underlying latent variable. One difference between
+the two techniques is that factor analysis can be considered as a way to
+group observed *variables* into latent constructs, and LCA groups
+*individuals* into classes.While factor analysis seeks to describe the
+relations between variables, LCA seeks to describe each unobserved group
+and the differences between them. In line with this distinction, LCA
+(mixture models) is referred to as a "person-centered" technique, and
+factor analysis as a "variable-centered" technique
+[@nylund-gibson_ten_2018; @masyn_latent_2013].
+
+When the focus is on the model parameters in each group, LCA can be
+thought of as similar to a multi-group structural equation model. The
+main distinction is that group membership is not known a-priori, but is
+instead estimated -- with measurement error -- based on the data.
+Whereas in a multi-group model, the data are split by group and treated
+as independent samples, in LCA, all cases contribute to the estimation
+of all parameters in all groups. The relative contribution of each case
+to the parameters of each group is determined by that case's posterior
+probability of belonging to that group.
+
+In factor analysis, the latent variable is continuous, so the relation
+between indicators is define by the communalities between them, and the
+latent variable represents the error corrected approximation of the
+sample. While in LCA, the relation between indicators is define by how
+well they separate the classes, as a multinomial regression between
+indicators and latent variable [@masyn_latent_2013]
+
+When the focus is on each individual's estimated class membership, LCA
+can be thought of as a type of clustering algorithm. In line with this
+perspective, LCA is sometimes described as "model-based clustering"
+[@hennig_handbook_2015; @scrucca_mclust_2016]. Many clustering
+algorithms apply some recursive splitting algorithm to the data. By
+contrast "model-based" clustering refers to the fact that LCA estimates
+cluster membership based on a parametric model. Specifically, the
+posterior class probability that an individual belongs to a latent class
+can be computed from the likelihood of that individual's observed data
+under given the class-specific model.
+For these differences, LCA is a clustering method, that corrects for (random) measurement error, while clustering algorithms do not correct for.
+This is something LCA has in common with factor analysis, as both model an underlying latent variable, which accounts for the communalities between indicators and correcting for measurement error.
+0
 
@@
 *unsupervised classification* problem [@figueiredo_unsupervised_2002].
-The term *unsupervised* refers to the fact that the outcome variable 
--- true class membership -- is not known, 
-and the term *classification* refers to the fact that 
-the algorithm is predicting a categorical outcome -- class membership.
+The term *unsupervised* refers to the fact that the outcome variable --
+true class membership -- is not known, and the term *classification*
+refers to the fact that the algorithm is predicting a categorical
+outcome -- class membership.
 
@@
 
-In this paper, we use the term latent class analysis to refer to 
-a family of techniques that estimate latent class membership 
-based on a parametric model of observed indicators.
-From a historical perspective, the term was initially conceived to 
-refer to analyses with categorical, usually binary indicators [@vermunt_jk_latent_2004].
-Nowadays, there are a number of related techniques, 
-known by distinct names, that serve a similar purpose.
-The term "latent class analysis" seems most appropriate 
-as an umbrella term for this broader class of models,
-as it only refers to the purpose of the analysis, 
-and does not imply restrictions to the model used, 
-or the level of measurement of the indicators.
-Given the abundance of terms in use for closely related classes of models, 
-we will provide a rudimentary taxonomy of LCA methods.
-
-A common type of LCA is the *finite Gaussian mixture model*; 
-a univariate analysis where the observed distribution of a single variable 
-is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
-The parameters of a finite Gaussian mixture model are 
-the means and variances of these underlying normal distributions.
-The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
-In the multivariate case, with more than one indicator variable,
-the parameters of a mixture model are the means, variances, and covariances 
-between the indicators (which can be standardized to obtain correlations).
-These parameters can be estimated freely, or set to be constrained across classes.
-
-The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
-which assumes conditional independence of the indicators.
-Conditional independence means that, after class membership is accounted for,
-the covariances/correlations between indicators are assumed to be zero.
-This can be conceived of as a restricted mixture model with covariances fixed to zero.
-In some cases, such constraints will be inappropriate;
-for instance when the cohesion between indicators is expected to differ between classes.
-As an example, a LPA of ocean plastic particles 
-found two classes of particles based on length and width:
-a class of smaller particles with a high correlation between length and width, 
-meaning that these particles were approximately round or square in shape, 
-and a class of larger particles with a low correlation between length and width, 
-meaning that these particles were heterogenous in shape.
-From a theoretical perspective, this makes sense, 
-because the smaller particles have been ground down to a more uniform shape by the elements.
+In this paper, we use the term latent class analysis to refer to a
+family of techniques that estimate latent class membership based on a
+parametric model of observed indicators. From a historical perspective,
+the term was initially conceived to refer to analyses with categorical,
+usually binary indicators [@vermunt_jk_latent_2004]. Nowadays, there are
+a number of related techniques, known by distinct names, that serve a
+similar purpose. The term "latent class analysis" seems most appropriate
+as an umbrella term for this broader class of models, as it only refers
+to the purpose of the analysis, and does not imply restrictions to the
+model used, or the level of measurement of the indicators. Given the
+abundance of terms in use for closely related classes of models, we will
+provide a rudimentary taxonomy of LCA methods.
+
+A common type of LCA is the *finite Gaussian mixture model*; a
+univariate analysis where the observed distribution of a single variable
+is assumed to result from a mixture of a known number of normal
+(Gaussian) distributions. The parameters of a finite Gaussian mixture
+model are the means and variances of these underlying normal
+distributions. The analysis of shoe sizes presented earlier is a
+canonical example of this type of analysis. In the multivariate case,
+with more than one indicator variable, the parameters of a mixture model
+are the means, variances, and covariances between the indicators (which
+can be standardized to obtain correlations). These parameters can be
+estimated freely, or set to be constrained across classes [@collins_latent_2010].
+
+The technique known as *latent profile analysis (LPA)* is a special case
+of the mixture model, which assumes conditional independence of the
+indicators. Conditional independence means that, after class membership
+is accounted for, the covariances/correlations between indicators are
+assumed to be zero. This can be conceived of as a restricted mixture
+model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
+will be inappropriate; for instance when the cohesion between indicators
+is expected to differ between classes [@collins_latent_2010]. As an example, a LPA of ocean
+plastic particles found two classes of particles based on length and
+width: a class of smaller particles with a high correlation between
+length and width, meaning that these particles were approximately round
+or square in shape, and a class of larger particles with a low
+correlation between length and width, meaning that these particles were
+heterogenous in shape. From a theoretical perspective, this makes sense,
+because the smaller particles have been ground down to a more uniform
+shape by the elements.
 <!-- DA: what is the source for plastic particles? -->
 
-It is also possible to estimate a mixture model based on latent indicators.
-This means that, within each class, 
-one or more continuous latent variables are estimated based on the observed indicators.
-Categorical latent variable membership is then estimated 
-based on these continuous latent variables.
-A common application of this approach is in longitudinal research, 
-where the indicators reflect one construct assessed at different time points.
-Examples of this approach include *growth mixture models* (GMM) 
-and *latent class growth analyses* (LCGA).
-These techniques estimate a latent growth model
-to describe individual trajectories over time.
-The growth mixture model is a latent class model 
-where the parameters that indicate class membership 
-are the intercepts and variances, 
-and typically covariances of the latent growth variables, 
-e.g., a latent intercept and slope.
-This technique assumes that individuals within a class can have heterogenous trajectories.
-If the variance of the growth parameters is fixed to zero, 
-it is known as a latent class growth analysis.
-This latter approach assumes that all individuals within a class 
-share the same identical trajectory,
-and that any variance in the indicators not explained by the class-specific latent trajectories 
-is due to residual error variance.
-
-The term latent class analysis originally referred to cases 
-where the observed indicators were categorical.
-Nowadays, it is more commonly used as an umbrella term.
-To prevent ambiguity, the special case 
-where indicators are of binary or ordinal measurement level 
-might be described as *latent class analysis with ordinal indicators*.
-Latent class models with ordinal indicators are parameterized differently from mixture models.
-One common parameterization assumes that 
-each categorical variable reflects an underlying standard normal distribution.
-The parameters are "thresholds" that correspond to quantiles of a standard normal distribution 
-(with $N(\mu = 0, \sigma = 1)$).
-These thresholds are estimated based on 
-the proportion of individuals in each of the response categories of the indicator variable.
-For example, a binary indicator has a single threshold that distinguishes the two response categories.
-If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
-If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.
-
-This paper will primarily focus on mixture models and special cases thereof,
-although most of the suggested guidelines are applicable to all LCA methods.
+It is also possible to estimate a mixture model based on latent
+indicators. This means that, within each class, one or more continuous
+latent variables are estimated based on the observed indicators.
+Categorical latent variable membership is then estimated based on these
+continuous latent variables. A common application of this approach is in
+longitudinal research, where the indicators reflect one construct
+assessed at different time points. Examples of this approach include
+*growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
+These techniques estimate a latent growth model to describe grouped
+trajectories over time. The growth mixture model is a latent class model
+where the parameters that indicate class membership are the intercepts
+and variances, and typically covariances of the latent growth variables,
+e.g., a latent intercept and slope. This technique assumes that
+individuals within a class can have heterogenous trajectories. If the
+variance of the growth parameters is fixed to zero, it is known as a
+latent class growth analysis. This latter approach assumes that all
+individuals within a class share the same identical trajectory, and that
+any variance in the indicators not explained by the class-specific
+latent trajectories is due to residual error variance.
+<!-- MGV: some reference for this description -->
+
+The term latent class analysis originally referred to cases where the
+observed indicators were categorical [@collins_latent_2010]. 
+Nowadays, it is more commonly used
+as an umbrella term. To prevent ambiguity, the special case where
+indicators are of binary or ordinal measurement level might be described
+as *latent class analysis with ordinal indicators*. Latent class models
+with ordinal indicators are parameterized differently from mixture
+models. One common parameterization assumes that each categorical
+variable reflects an underlying standard normal distribution. The
+parameters are "thresholds" that correspond to quantiles of a standard
+normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
+are estimated based on the proportion of individuals in each of the
+response categories of the indicator variable. For example, a binary
+indicator has a single threshold that distinguishes the two response
+categories. If responses are distributed 50/50, then the corresponding
+threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
+then the resulting threshold would be $t_1 = 0.25$.
+
+This paper will primarily focus on mixture models and special cases
+thereof, although most of the suggested guidelines are applicable to all
+LCA methods.
 
@@
 
-There are several use cases for which LCA methods are suitable.
-One example is to test a theory that postulates the existence of a categorical latent variable.
-For example, *identity status theory* posits that, at any given point in time, 
-adolescents reside in one of four identity statuses.
-Latent class analysis can be used to identify these four statuses 
-based on observed indicators (e.g., self-reported identity exploration and commitment).
-If results indicate that the data are better described by a different number of classes,
-or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, 
-then the theory may be called into question.
-
-Another use case is unsupervised learning:
-when the goal is to restore unobserved class membership based on observed indicators,
-or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid 
-when several clinical indicators can be used to distinguish between 
-a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
-The example of shoe size is a rudimentary illustration of this type of application.
-
-LCA can be used as a descriptive analysis 
-where a researcher wishes to describe their sample 
-and identify a few prototypes based on many variables.
-As an example, if a survey among all Dutch academics was carried out
-with the goal of bringing about a funding reform, 
-LCA could be used to discover the types of publications that get funding.
-
-With LCA, our goal could be to inductively identify the number of classes. 
-For instance, if we believe that a variable represents a group, 
-but don't know how many groups there are, 
-LCA may be an appropriate technique to answer this question.
-For example, @hopfer_social_2014 studied substance use, sexual behavior, 
-and mental health status of urban population in Winnipeg, Canada. 
-The underlying assumption was that there were different risk profiles,
-but their number was not known.
-From a collection of indicators, LCA provided evidence 
-that there are four distinct risk profiles in the Winnipeg area.
-
-Another application of LCA is to classify individuals.
-In a peer harassment study, 
-@giang_using_2008 used LCA to classify over 2,000 sixth grade students 
-into aggressor and victim latent classes.
-The five-class solution comprised of classes of 
-victims, aggressors, and socially adjusted students. 
-For instance, it revealed that there were two types of victims: 
-highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
-
-LCA is also appropriate when we wish to identify indicators that capture classes well.
-High quality indicators will be strongly related to the latent variable
-and will lead to good class separation.
-This relationship of high quality indicators to the latent variable
-is reflected in very high or very low conditional response probabilities.
-For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
-Therefore, one could use conditional response probabilities for each item to assess its quality
-with regards to how well it helps separate the latent classes.
-From this, a theory about the selection of indicators could be informed.
-
-An extension of LCA is that containing covariates which can be used to predict class membership.
-In this approach, we not only model the latent class variable based on indicators,
-but we also relate the class membership to other explanatory variables (@vermunt_latent_2017).
-An example of using covariates comes from @nozadi_moderating_2016 
-who applied LCA to identify the probability of children's membership to an anxiety class.
-The authors tested several covariates including children's age, sex, and accuracy scores.
-Age and sex were not found to be related to the children's latent class membership,
-hence these covariates were excluded from the analysis.
-Accuracy scores, however, were related to probabilities of 
-being in anxiety and attention-anxiety classes. 
-Therefore this covariate was kept as a valuable predictor.
-
-When our interest is the prediction of one or more outcomes,
-LCA can be used to construct latent classes as categorical predictors.
-@lanza_latent_2013 demonstrated how LCA can be used 
-to classify adolescents into depression classes,
-and subsequently these classes can be used to predict smoking, grades, and delinquency.
-The study showed that the outcomes predicted by class membership 
-can be binary (regular smoking), continuous (grades) or count (delinquency).
-
-In addition to these applications, LCA can be used for dimensionality reduction
-as the resulting groups summarize response patterns on a large number of indicators.
-For example, @macgregor_symptom_2021 investigated symptom profiles 
-among injured U.S. military personnel. 
-They used fifteen dichotomous items 
-from the Post-Deployment Health Assessment survey as LCA indicators.
-Combinatorics informs us that fifteen dichotomous items have 
-$2^{15}$ or $32,768$ unique symptom combinations.
-Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA 
-as a method of dimensionality reduction.
-A five class solution was found to have the best fit 
-according to both statistical criteria and clinical interpretability.
+There are several use cases for which LCA methods are suitable. One
+example is to test a theory that postulates the existence of a
+categorical latent variable. For example, *identity status theory*
+posits that, at any given point in time, adolescents reside in one of
+four identity statuses. Latent class analysis can be used to identify
+these four statuses based on observed indicators (e.g., self-reported
+identity exploration and commitment). If results indicate that the data
+are better described by a different number of classes, or that the
+four-class solution does not correspond to the predicted pattern of
+responses on the indicators, then the theory may be called into
+question.
+
+Another use case is unsupervised learning: when the goal is to restore
+unobserved class membership based on observed indicators, or to classify
+individuals. For example, a mixture model can be used as a diagnostic
+aid when several clinical indicators can be used to distinguish between
+a fixed number of physical [@baughman_mixture_2006], mental
+[@wu_abuse_2011] health problems, or health outcomes after recovering from cancer [@hardy_exploring_2018]. The example of shoe size is a
+rudimentary illustration of this type of application.
+
+LCA can be used as a descriptive analysis where a researcher wishes to
+describe their sample and identify a few prototypes based on many
+variables. As an example, if a survey among all Dutch academics was
+carried out with the goal of bringing about a funding reform, LCA could
+be used to discover the types of publications that get funding.
+
+With LCA, our goal could be to inductively identify the number of
+classes. For instance, if we believe that a variable represents a group,
+but don't know how many groups there are, LCA may be an appropriate
+technique to answer this question. For example, @hopfer_social_2014
+studied substance use, sexual behavior, and mental health status of
+urban population in Winnipeg, Canada. The underlying assumption was that
+there were different risk profiles, but their number was not known. From
+a collection of indicators, LCA provided evidence that there are four
+distinct risk profiles in the Winnipeg area.
+
+Another application of LCA is to classify individuals. In a peer
+harassment study, @giang_using_2008 used LCA to classify over 2,000
+sixth grade students into aggressor and victim latent classes. The
+five-class solution comprised of classes of victims, aggressors, and
+socially adjusted students. For instance, it revealed that there were
+two types of victims: highly-victimized aggressive-victims and
+highly-aggressive aggressive-victims.
+
+LCA is also appropriate when we wish to identify indicators that capture
+classes well. High quality indicators will be strongly related to the
+latent variable and will lead to good class separation. This
+relationship of high quality indicators to the latent variable is
+reflected in very high or very low conditional response probabilities.
+For a simulation study exploring the effects of indicator quality on
+LCA, see @geiser_is_2014. Therefore, one could use conditional response
+probabilities for each item to assess its quality with regards to how
+well it helps separate the latent classes. From this, a theory about the
+selection of indicators could be informed.
+
+An extension of LCA is that containing covariates which can be used to
+predict class membership. In this approach, we not only model the latent
+class variable based on indicators, but we also relate the class
+membership to other explanatory variables (@vermunt_latent_2017). An
+example of using covariates comes from @nozadi_moderating_2016 who
+applied LCA to identify the probability of children's membership to an
+anxiety class. The authors tested several covariates including
+children's age, sex, and accuracy scores. Age and sex were not found to
+be related to the children's latent class membership, hence these
+covariates were excluded from the analysis. Accuracy scores, however,
+were related to probabilities of being in anxiety and attention-anxiety
+classes. Therefore this covariate was kept as a valuable predictor.
+
+When our interest is the prediction of one or more outcomes, LCA can be
+used to construct latent classes as categorical predictors.
+@lanza_latent_2013 demonstrated how LCA can be used to classify
+adolescents into depression classes, and subsequently these classes can
+be used to predict smoking, grades, and delinquency. The study showed
+that the outcomes predicted by class membership can be binary (regular
+smoking), continuous (grades) or count (delinquency).
+
+These mixture models can also be useful to evaluate quality of measurement across instrument.
+For example, employment status was collected across register and survey data, and there is disagreement between them.
+You could think that register data is without error, but it has been shown that is victim of measurement error as well.
+In this case mixture models was used to evaluate the underlying factor of "error corrected employment status", by using two indicators of the same question and identify random and systamtic measurement error in either indicator [@pavlopoulos_measuring_2015].
+
+In addition to these applications, LCA can be used for dimensionality
+reduction as the resulting groups summarize response patterns on a large
+number of indicators. For example, @macgregor_symptom_2021 investigated
+symptom profiles among injured U.S. military personnel. They used
+fifteen dichotomous items from the Post-Deployment Health Assessment
+survey as LCA indicators. Combinatorics informs us that fifteen
+dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
+Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a
+method of dimensionality reduction. A five class solution was found to
+have the best fit according to both statistical criteria and clinical
+interpretability.
 
 <!-- CJ: (on the next paragraph)  -->
+
 <!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
+
 <!-- That is not what this paragraph is about.  -->
+
 <!-- This paragraph is about motivations for using LCA. -->
+
 <!-- One motivation is if your data violates multivariate normality -->
+
 <!-- because violating multivariate normality means that you have multiple groups -->
+
 <!-- and you can use LCA to extract those. That is the only assumption  -->
+
 <!-- we have to discuss here, we just need to explain it. -->
-Finally, LCA can be used to deal with data which violate certain assumptions.
-As discussed in the shoe size example, LCA can deal with violations of normality.
-In fact, LCA assumes the population distribution is a non-normal mixture 
-of $K$ normal distributions,
-and it can discover the value of $K$, i.e. generate a $K$-class solution.
+
+Finally, LCA can be used to deal with data which violate certain
+assumptions. As discussed in the shoe size example, LCA can deal with
+violations of normality. In fact, LCA assumes the population
+distribution is a non-normal mixture of $K$ normal distributions, and it
+can discover the value of $K$, i.e. generate a $K$-class solution.
+
+LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous of categorical.
+We encourage researchers to be agnostic about the **true** nature of the latent variable.
+And focus on the approach that fits the research question.
 
@@
 
-The best practices in estimation, as outlined in Table \@ref(tab:checkest),
-are rooted in existing recommendations for best practices 
-for estimating specific subtypes of LCA,
-including latent class growth analysis [@van_de_schoot_grolts-checklist_2017]
-and latent class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
-These were generalized to be relevant to all types of LCA,
-and updated to current best practices, as explained below.
+The best practices in estimation, as outlined in Table
+\@ref(tab:checkest), are rooted in existing recommendations for best
+practices for estimating specific subtypes of LCA, including latent
+class growth analysis [@van_de_schoot_grolts-checklist_2017] and latent
+class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
+These were generalized to be relevant to all types of LCA, and updated
+to current best practices, as explained below.
 
@@
 
-Examining observed data is essential for any analysis
-as it may reveal patterns and violations of assumptions 
-that had not been considered prior to data collection.
-Special attention should be paid 
-to level of measurement of the indicators.
-Finite Gaussian mixture models (including LPA) 
-are only suitable for continuous variables.
-Indicators with an ordinal level of measurement 
-are likely to violate the assumption of 
-within-class normal distributions of mixture models
-[see @vermunt_k-means_2011].
-Personal experience consulting on LCA methods and 
-moderating the `tidyLPA` Google group
-suggest that the misapplication of mixture models 
-to ordinal (e.g., Likert-type) indicators 
-is the most common source of user error.
-Whereas it has been argued that some parametric methods are robust 
-when scales with 7+ indicators are treated as continuous 
-[e.g., @norman_likert_2010],
-this certainly does not imply that all methods are.
-It is certainly unlikely that such ordinal variables 
-can be treated as a *mixture* of multiple normal distributions.
-The problem becomes egregious when 
-the number of classes estimated equals or exceeds the number of categories; 
-in this case, each class-specific mean could describe a single response category, 
-and a class-specific variance component would be nonsensical.
-In sum, Likert-type scales are rarely suitable for mixture modeling;
+Examining observed data is essential for any analysis as it may reveal
+patterns and violations of assumptions that had not been considered
+prior to data collection. Special attention should be paid to level of
+measurement of the indicators. Finite Gaussian mixture models (including
+LPA) are only suitable for continuous variables. Indicators with an
+ordinal level of measurement are likely to violate the assumption of
+within-class normal distributions of mixture models [see
+@vermunt_k-means_2011]. Personal experience consulting on LCA methods
+and moderating the `tidyLPA` Google group suggest that the
+misapplication of mixture models to ordinal (e.g., Likert-type)
+indicators is the most common source of user error. Whereas it has been
+argued that some parametric methods are robust when scales with 7+
+indicators are treated as continuous [e.g., @norman_likert_2010] or 
+with 5+ categories [@rhemtulla_when_2012], this
+certainly does not imply that all methods are. It is certainly unlikely
+that such ordinal variables can be treated as a *mixture* of multiple
+normal distributions. The problem becomes egregious when the number of
+classes estimated equals or exceeds the number of categories; in this
+case, each class-specific mean could describe a single response
+category, and a class-specific variance component would be nonsensical.
+In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
 latent class analysis with ordinal indicators is more appropriate.
 
-Relatedly, a recent publication claimed that 
-an assumption of mixture models is that 
-observed indicators are normally distributed [@spurk_latent_2020].
-This is incorrect.
-When the number of classes is greater than one, 
-mixture models assume that the observed indicators 
-are a mixture of multiple (multivariate) normal distributions.
-In our shoe size example, it can be seen that the population distribution
-is comprised of two normal distributions.
-When examined visually, the population distribution is evidently bimodal. 
-The Shapiro-Wilk normality test ($W = 0.971$, $p < 0.05$ ) 
-rejects the null hypothesis that 
-the sample comes from a normally distributed population.
-Yet, this is a prototypical example of a mixed population distribution 
-where LCA can discover latent groups.
-If the population distribution were instead normal, 
-there would be no classes to extract 
-as the whole population would belong to a single class.
-
-Extensive descriptive statistics 
-(including the number of unique values, 
+Relatedly, a recent publication claimed that an assumption of mixture
+models is that observed indicators are normally distributed
+[@spurk_latent_2020]. This is incorrect. When the number of classes is
+greater than one, Gaussian mixture models assume that the observed indicators are
+a mixture of multiple (multivariate) normal distributions. In our shoe
+size example, it can be seen that the population distribution is
+comprised of two normal distributions. When examined visually, the
+population distribution is evidently bimodal. The Shapiro-Wilk normality
+test ($W = 0.971$, $p < 0.05$ ) rejects the null hypothesis that the
+sample comes from a normally distributed population. Yet, this is a
+prototypical example of a mixed population distribution where LCA can
+discover latent groups. If the population distribution were instead
+normal, there would be no classes to extract as the whole population
+would belong to a single (homogeneous) class.
+
+Extensive descriptive statistics (including the number of unique values,
 variance of categorical variables, and missingness; see next subsection)
-can be obtained using the function `tidySEM::descriptives(data)`.
-Note, however, that sample-level descriptive statistics 
-are of limited value when the goal of a study is 
-to identify subsamples using latent class analysis.
-Plots (density plots for continuous variables, 
-and bar charts for categorical ones) may be more diagnostic.
-Note that density plots can also aid in the choice of the number of classes, 
-as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, 
-provided that these are readily accessible 
-[consider using a GitHub repository as a comprehensive public research archive, 
-as explained in @van_lissa_worcs_2021].
+can be obtained using the function `tidySEM::descriptives(data)`. Note,
+however, that sample-level descriptive statistics are of limited value
+when the goal of a study is to identify subsamples using latent class
+analysis. Plots (density plots for continuous variables, and bar charts
+for categorical ones) may be more diagnostic. Note that density plots
+can also aid in the choice of the number of classes, as further
+explained in the section on visualization. Descriptive statistics and
+plots can be relegated to online supplements, provided that these are
+readily accessible [consider using a GitHub repository as a
+comprehensive public research archive, as explained in
+@van_lissa_worcs_2021].
 
@@
 
-Previous work has emphasized the importance of examining 
-the pattern of missing data and 
-reporting how missingness was handled 
-[@van_de_schoot_grolts-checklist_2017].
-Three types of missingness have been distinguished
-in the literature [@rubin_inference_1976]:
-Missing completely at random (MCAR), 
-which means that missingness is random;
-missing at random (MAR), 
-which means that missingness is contingent on the *observed* data
-(and can thus be accounted for);
-and finally missing not at random (MNAR), 
-which means that missingness is related to unobserved factors.
-It is possible to conduct a so-called "MCAR" test, 
-for example the non-parametric MCAR test [@jamshidian_tests_2010].
-But note that the name "MCAR test" is somewhat misleading,
-as the null-hypothesis of this test is that the data are not MAR,
-and a significant test statistic indicates that 
-missingness is related to the observed data (MAR).
-A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on 
-the comparison of variances across groups 
-with different patterns of missing data, 
-it assumes normality [@little_test_1988].
-This assumption is tenuous in the context of LCA.
-A non-parametric MCAR test, 
-as provided by Jamshidian and Jalal, 
-may be more suitable [@jamshidian_tests_2010].
-Unfortunately, this test was removed from the central R-repository CRAN 
-due to lack of maintenance.
-For this tutorial, I have re-implemented it in the `mice` package 
-as `mice::mcar()`,
-with a fast backend in C++ and new printing and plotting methods.
-
-While we concur that investigating missingness is due diligence,
-it is important to emphasize that 
-missingness is adequately handled by default in many software packages for LCA, 
-such as Mplus, and `OpenMx` which is the backend of `tidySEM`.
-These packages use Full Information Maximum Likelihood (FIML) estimation,
-which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data; 
-on par with multiple imputation [@lee_comparison_2021].
-FIML estimation assumes that missingness is either MCAR or MAR.
-Thus, one would typically proceed with FIML, 
-regardless of the outcome of an MCAR test.
-Although FIML does not, by default, handle missingness in exogenous variables
--- all indicator variables in LCA are endogenous, 
-so this is not a concern.
-
-Multiple imputation is less suitable to LCA for two reasons.
-First, because LCA methods are often computationally expensive,
-and conducting them on multiple imputed datasets may be unfeasible.
-Second, because there is no straightforward way 
-to integrate LCA results across multiple datasets.
-To conclude; our recommendation is to inspect missingness 
-(e.g., using `mice::MCAR()`)
-and report the proportion of missingness per variable 
-(e.g., using `tidySEM::descriptives()`),
-before proceeding with FIML.
-One minor concern is that the K-means algorithm,
-which `tidySEM` uses for determining starting values,
-is *not* robust to missing values.
-When it fails, `tidySEM` automatically switches to hierarchical clustering,
-unless the user specifies a different
+Previous work has emphasized the importance of examining the pattern of
+missing data and reporting how missingness was handled
+[@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
+been distinguished in the literature [@rubin_inference_1976; @enders_applied_2010]: Missing
+completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
+missing at random (MAR), which means that missingness is contingent on
+the *observed* data (and can thus be accounted for); and finally missing
+not at random (MNAR), which means that missingness is related to
+unobserved factors. It is possible to conduct a so-called "MCAR" test,
+for example the non-parametric MCAR test [@jamshidian_tests_2010]. But
+note that the name "MCAR test" is somewhat misleading, as the
+null-hypothesis of this test is that the data are not MAR, and a
+significant test statistic indicates that missingness is related to the
+observed data (MAR). A non-significant test statistic does not
+distinguish between MCAR or MNAR. As Little's classic MCAR test relies
+on the comparison of variances across groups with different patterns of
+missing data, it assumes normality [@little_test_1988]. This assumption
+is tenuous in the context of LCA. A non-parametric MCAR test, as
+provided by Jamshidian and Jalal, may be more suitable
+[@jamshidian_tests_2010]. Unfortunately, this test was removed from the
+central R-repository CRAN due to lack of maintenance. For this tutorial,
+I have re-implemented it in the `mice` package as `mice::mcar()`, with a
+fast backend in C++ and new printing and plotting methods.
+
+While we concur that investigating missingness is due diligence, it is
+important to emphasize that missingness is adequately handled by default
+in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` which is
+the backend of `tidySEM`. These packages use Full Information Maximum
+Likelihood (FIML) estimation, which makes use of all available
+information without imputing missing values. FIML is a best-practice
+solution for handling missing data; on par with multiple imputation when the estimation model includes the same variables as the imputation model
+[@lee_comparison_2021]. FIML estimation assumes that missingness is
+either MCAR or MAR. Thus, one would typically proceed with FIML,
+regardless of the outcome of an MCAR test. Although FIML does not, by
+default, handle missingness in exogenous variables -- all indicator
+variables in LCA are endogenous, so this is not a concern.
+<!-- are you sure Mplus accounts for missing data automatically when the indicators are categorical? I know that is not the case with CFA/SEM models -->
+
+Multiple imputation is less suitable to LCA for two reasons. First,
+because LCA methods are often computationally expensive, and conducting
+them on multiple imputed datasets may be unfeasible. Second, because
+there is no straightforward way to integrate LCA results across multiple
+datasets. To conclude; our recommendation is to inspect missingness
+(e.g., using `mice::MCAR()`) and report the proportion of missingness
+per variable (e.g., using `tidySEM::descriptives()`), before proceeding
+with FIML. One minor concern is that the K-means algorithm, which
+`tidySEM` uses for determining starting values, is *not* robust to
+missing values. When it fails, `tidySEM` automatically switches to
+hierarchical clustering, unless the user specifies a different
 clustering algorithm or uses manual starting values.
@@
 @van_de_schoot_grolts-checklist_2017 developed a protocol called
-Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
-Two of the GRoLTS guidelines (namely, 6a and 6b) refer to 
-considering alternative model specifications.
-Both are discussing specific cases in latent trajectory model specification. 
-The first is about whether the variance of the growth parameter 
-is estimated freely or fixed,
-and the second is about whether conditional independence is assumed,
-as the researcher might also want to free-up the variance-covariance structure.
-In LCA, there are also many different ways to specify the model.
-Means, variances, and covariances between the indicators 
-can either be constrained across classes, or estimated freely.
-Researchers using LCA should transparently report their chosen parametrization 
-as well as discuss different parametrizations that were tested in the process.
-
-Different types of latent class models have different parameters.
-For example, mixture models and latent profile analyses 
-typically have class-specific means, variances, and covariances.
-Latent growth analyses have the same parameters, 
-but with respect to the latent growth variables.
-Latent class analyses with ordinal indicators have thresholds.
-All of these parameters can be freely estimated, 
-constrained to be equal across classes,
-or fixed to a certain value (e.g., to zero).
-The total number of parameters thus 
-scales with the number of estimated classes.
-Consequently, LCA methods have a potentially very high number of parameters.
-As any of these parameters could be misspecified,
-it is important to consider alternative model specifications.
-However, alternative model specifications may be approached differently 
-depending on whether an analysis is data driven (exploratory), 
-or theory driven (confirmatory).
-This distinction has remained underemphasized in prior writing.
+Guidelines for Reporting on Latent Trajectory Studies (GRoLTS). Two of
+the GRoLTS guidelines (namely, 6a and 6b) refer to considering
+alternative model specifications. Both are discussing specific cases in
+latent trajectory model specification. The first is about whether the
+variance of the growth parameter is estimated freely or fixed, and the
+second is about whether conditional independence is assumed, as the
+researcher might also want to free-up the variance-covariance structure.
+In LCA, there are also many different ways to specify the model. Means,
+variances, and covariances between the indicators can either be
+constrained across classes, or estimated freely. Researchers using LCA
+should transparently report their chosen parametrization as well as
+discuss different parametrizations that were tested in the process.
+
+Different types of latent class models have different parameters. For
+example, mixture models and latent profile analyses typically have
+class-specific means, variances, and covariances. Latent growth analyses
+have the same parameters, but with respect to the latent growth
+variables. Latent class analyses with ordinal indicators have
+thresholds. All of these parameters can be freely estimated, constrained
+to be equal across classes, or fixed to a certain value (e.g., to zero).
+The total number of parameters thus scales with the number of estimated
+classes. Consequently, LCA methods have a potentially very high number
+of parameters. As any of these parameters could be misspecified, it is
+important to consider alternative model specifications. However,
+alternative model specifications may be approached differently depending
+on whether an analysis is data driven (exploratory), or theory driven
+(confirmatory). This distinction has remained underemphasized in prior
+writing.
 
 Prior literature on LCA has emphasized its exploratory applications
-[@nylund_deciding_2007].
-In exploratory LCA, 
-a large number of models are typically estimated in batch,
-with varying numbers of classes and model specifications.
-The "correct" model specification is then determined 
-based on a combination of fit indices,
-significance tests, and interpretability.
-For latent profile analysis, 
-the function `tidySEM::mx_profiles(classes, variances, covariances)`
-largely automates this process.
-The argument `classes` indicates which 
-class solutions should be estimated (e.g., 1 through 6).
-The argument `variances` specifies whether 
-variances should be `"equal"` or `"varying"` across classes.
-The argument `covariances` specifies whether 
-covariances should be constrained to 
-`"zero"`, `"equal"` or `"varying"` across classes.
-The means are free to vary across classes by default, 
-although the more general function `tidySEM::mx_mixture()` 
-could be used to circumvent this.
-After all models have been estimated, 
-the function `tidySEM::table_fit()` can be used 
-to obtain a model fit table suitable for 
-determining the optimal model according to best practices.
-Note, however, that this table does not include 
-the bootstrapped likelihood ratio test (BLRT) by default,
-because this test is very computationally expensive.
-It is recommended to use the function `tidySEM::BLRT()` 
-to compare a shortlist of likely candidate models 
-based on other fit indices. 
-More on fit indices can be found in 
-the Model Fit Indices subsection of this paper.
-
-Confirmatory LCA typically requires 
-less comprehensive alternative model specifications.
-For example, in the context of a preregistered analysis, 
-the main models of interest would have been specified a priori. 
-Even in this case, 
-the theoretical model could be compared to a few others to contextualize it.
+[@nylund_deciding_2007]. In exploratory LCA, a large number of models
+are typically estimated in batch, with varying numbers of classes and
+model specifications. The "correct" model specification is then
+determined based on a combination of fit indices, significance tests,
+and interpretability. For latent profile analysis, the function
+`tidySEM::mx_profiles(classes, variances, covariances)` largely
+automates this process. The argument `classes` indicates which class
+solutions should be estimated (e.g., 1 through 6). The argument
+`variances` specifies whether variances should be `"equal"` or
+`"varying"` across classes. The argument `covariances` specifies whether
+covariances should be constrained to `"zero"`, `"equal"` or `"varying"`
+across classes. The means are free to vary across classes by default,
+although the more general function `tidySEM::mx_mixture()` could be used
+to circumvent this. After all models have been estimated, the function
+`tidySEM::table_fit()` can be used to obtain a model fit table suitable
+for determining the optimal model according to best practices. Note,
+however, that this table does not include the bootstrapped likelihood
+ratio test (BLRT) by default, because this test is very computationally
+expensive. It is recommended to use the function `tidySEM::BLRT()` to
+compare a shortlist of likely candidate models based on other fit
+indices. More on fit indices can be found in the Model Fit Indices
+subsection of this paper.
+
+Confirmatory LCA typically requires less comprehensive alternative model
+specifications. For example, in the context of a preregistered analysis,
+the main models of interest would have been specified a priori. Even in
+this case, the theoretical model could be compared to a few others to
+contextualize it.
 
@@
 Many software packages are available for the estimation of LCA models.
-Some of these packages have limited functionality, 
-or implement specific innovations.
-Other packages implement LCA in the context of 
-a more flexible structural equation modeling framework.
-The most notable examples of the latter are 
-the commercial programs Mplus and Latent GOLD, 
-and the free open source R-package OpenMx.
-The commercial packages stand out 
-because they offer relatively user-friendly interfaces
-and implement sensible defaults for complex analyses, 
-including LCA methods.
+Some of these packages have limited functionality, or implement specific
+innovations. Other packages implement LCA in the context of a more
+flexible structural equation modeling framework. The most notable
+examples of the latter are the commercial programs Mplus and Latent
+GOLD, and the free open source R-package OpenMx. The commercial packages
+stand out because they offer relatively user-friendly interfaces and
+implement sensible defaults for complex analyses, including LCA methods.
 This lowers the threshold for applied researchers to adopt such methods.
-Commercial software also has several downsides, however.
-One such downside is that use of the software 
-is restricted to those individuals 
-and institutions who can afford a license.
-Another downside is that the source code, 
-being proprietary, cannot be audited, debugged, 
-or enhanced by third parties.
-This incurs the risk that mistakes in the source code 
-may go unnoticed, and curbs progress as 
-software developers cannot add new functionality.
-
-Conversely, the free open source program OpenMx is very flexible,
-but not very user-friendly. 
-We directly address this limitation using the `tidySEM` R-package.
-New functionality in `tidySEM` seeks to lower the threshold 
-for latent class analysis using `OpenMx`.
-It adheres to best practices in estimation and reporting, 
-as described in this paper.
-The user interface is simple, 
-making use of the model syntax of the widely used `lavaan` R-package.
-This syntax offers a human-readable way to specify latent variable models.
-Minor enhancements are made to simplify the specification of LCA.
-
-Because of the limitations of the aforementioned tools, 
-we set out to develop a free tool that
-provides sensible defaults and is easy to use, 
-but provides the option to access and modify 
-all of the model inputs (i.e., low barrier, high ceiling). 
-`tidySEM` interfaces with existing tools, 
-and is able to translate between what existing tools are capable of 
-and what researchers and analysts
-carrying-out person-oriented analyses would like to specify. 
-Furthermore,`tidySEM` facilitates fully-reproducible analyses 
-and contributes to open science.
+Commercial software also has several downsides, however. One such
+downside is that use of the software is restricted to those individuals
+and institutions who can afford a license. Another downside is that the
+source code, being proprietary, cannot be audited, debugged, or enhanced
+by third parties. This incurs the risk that mistakes in the source code
+may go unnoticed, and curbs progress as software developers cannot add
+new functionality.
+
+Conversely, the free open source program OpenMx [@openmx] is very flexible, but
+not very user-friendly. We directly address this limitation using the
+`tidySEM` R-package. New functionality in `tidySEM` seeks to lower the
+threshold for latent class analysis using `OpenMx`. It adheres to best
+practices in estimation and reporting, as described in this paper. The
+user interface is simple, making use of the model syntax of the widely
+used `lavaan` R-package [@lavaan]. This syntax offers a human-readable way to
+specify latent variable models. Minor enhancements are made to simplify
+the specification of LCA.
+
+Because of the limitations of the aforementioned tools, we set out to
+develop a free tool that provides sensible defaults and is easy to use,
+but provides the option to access and modify all of the model inputs
+(i.e., low barrier, high ceiling). `tidySEM` interfaces with existing
+tools, and is able to translate between what existing tools are capable
+of and what researchers and analysts carrying-out person-oriented
+analyses would like to specify. Furthermore,`tidySEM` facilitates
+fully-reproducible analyses and contributes to open science.
 
@@
 
-LCA parameters and model fit statistics 
-can be estimated in a variety of ways.
-The choice of the estimator 
-depends on the presence of missing values,
-sample size, number of indicators, 
-and available computational resources [@weller_latent_2020].
-A commonly used technique is maximum likelihood (ML) estimation
-with the expectation-maximization (EM) algorithm as a local optimizer. 
-Imagine we are estimating two parameters, 
-e.g. the class-specific means $\mu_c$ on a continuous indicator
-(ignoring the variance for now). 
-The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes
-the likelihood ($LL$) of all observed data. 
-In practice, instead of maximizing $LL$, 
-often $-2*LL$ is minimized, 
-as this offers computational advantages. 
-We can think of this optimization problem 
-as a three-dimensional landscape: 
-The X and Y dimensions are determined by the class-specific means, 
-so $X = \mu_1$ and $Y = \mu_2$ - 
-and the Z-dimension is determined by $Z = -2*LL$. 
-The optimizer must find the deepest "valley" in this landscape, 
-which reflects the combination of $\mu_1$ and $\mu_2$ that 
-maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble dropped in this landscape.
-It is dropped at some random point in space, 
-and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley, 
-it will settle on the bottom of that valley
-(this is known as "convergence").
-It cannot climb out again.
-Thus, if their are multiple valleys, 
-the risk is that the optimizer gets stuck in a shallower valley (a "local optimum"), 
-and never discovers the deepest valley (the "global optimum", or best solution).
-One solution to this problem is to drop many marbles at random places, 
-compare their final $-2*LL$ values,
-choose the solution with the lowest $-2*LL$, 
-and make sure that several marbles replicated this solution.
-This is the "random starts" approach.
-
-One problem with the random starts approach is that 
-it is computationally expensive to run this many replications.
-Moreover, because the algorithm begins with random starting values,
-many of the marbles are likely to be very far away from a "good enough" solution.
-Two innovations may improve the estimation procedure.
-The first is that, instead of picking random starting values, 
-a "reasonable solution" may be used for the starting values.
-For example, if we assume that the different classes are likely to have 
-different mean values on the indicators, 
-then the K-means clustering algorithm can be used to 
-determine these cluster centroids.
-We can compute the expected values of all model parameters 
-by treating the K-means solution as a known class solution,
-and use these as starting values for a mixture model.
-One remaining concern is that this approach may result 
-in starting values close to a local optimum,
-and that the EM algorithm will thus never find the global optimum.
-A second innovation addresses this concern.
-Instead of using EM, 
-it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some "destination" in the landscape, 
-and compares its likelihood to the current one. 
-If the destination likelihood is higher, 
-the estimator moves there.
-If the destination likelihood is *lower*, 
-the estimator still moves there occasionally, 
-based on probability.
-This latter property allows it to escape local optima,
-and find the global optimum.
+LCA parameters and model fit statistics can be estimated in a variety of
+ways. The choice of the estimator depends on the presence of missing
+values, sample size, number of indicators, and available computational
+resources [@weller_latent_2020]. A commonly used technique is maximum
+likelihood (ML) estimation with the expectation-maximization (EM)
+algorithm as a local optimizer. Imagine we are estimating two
+parameters, e.g. the class-specific means $\mu_c$ on a continuous
+indicator (ignoring the variance for now). The EM algorithm will attempt
+to find a combination of values for these two parameters that maximizes
+the likelihood ($LL$) of all observed data. In practice, instead of
+maximizing $LL$, often $-2*LL$ is minimized, as this offers
+computational advantages. We can think of this optimization problem as a
+three-dimensional landscape: The X and Y dimensions are determined by
+the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the
+Z-dimension is determined by $Z = -2*LL$. The optimizer must find the
+deepest "valley" in this landscape, which reflects the combination of
+$\mu_1$ and $\mu_2$ that maximizes the likelihood of the data. The EM
+optimizer behaves somewhat like a marble dropped in this landscape. It
+is dropped at some random point in space, and will roll into the nearest
+valley. The problem is that, once EM rolls into a valley, it will settle
+on the bottom of that valley (this is known as "convergence"). It cannot
+climb out again. Thus, if their are multiple valleys, the risk is that
+the optimizer gets stuck in a shallower valley (a "local optimum"), and
+never discovers the deepest valley (the "global optimum", or best
+solution). One solution to this problem is to drop many marbles at
+random places, compare their final $-2*LL$ values, choose the solution
+with the lowest $-2*LL$, and make sure that several marbles replicated
+this solution. This is the "random starts" approach.
+
+One problem with the random starts approach is that it is
+computationally expensive to run this many replications. Moreover,
+because the algorithm begins with random starting values, many of the
+marbles are likely to be very far away from a "good enough" solution.
+Two innovations may improve the estimation procedure. The first is that,
+instead of picking random starting values, a "reasonable solution" may
+be used for the starting values. For example, if we assume that the
+different classes are likely to have different mean values on the
+indicators, then the K-means clustering algorithm can be used to
+determine these cluster centroids. We can compute the expected values of
+all model parameters by treating the K-means solution as a known class
+solution, and use these as starting values for a mixture model. One
+remaining concern is that this approach may result in starting values
+close to a local optimum, and that the EM algorithm will thus never find
+the global optimum. A second innovation addresses this concern. Instead
+of using EM, it is possible to use an optimizer that can climb out of a
+valley. Simulated annealing iteratively considers some "destination" in
+the landscape, and compares its likelihood to the current one. If the
+destination likelihood is higher, the estimator moves there. If the
+destination likelihood is *lower*, the estimator still moves there
+occasionally, based on probability. This latter property allows it to
+escape local optima, and find the global optimum.
 
@@
 and identifies the global optimum solution using simulated annealing.
-Once a solution has been found,
-simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for 
-the parameters that can be used to compute standard errors.
-Note that these defaults can be manually overridden.
+Once a solution has been found, simulated annealing is followed up with
+a short run of the EM algorithm, as EM inherently produces an asymptotic
+covariance matrix for the parameters that can be used to compute
+standard errors. Note that these defaults can be manually overridden.
 
 A recent paper suggested maximum likelihood with robust standard errors
-should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
-This statement is incorrect, and may lead readers to believe that 
-they must use commercial software, 
-as robust maximum likelihood is currently only implemented 
-in Mplus and latentGOLD.
-As explained before, 
-mixture modeling assumes that observed data are 
-a mixture of (multivariate) normal distributions; 
-thus, the observed indicators will likely not be normally distributed.
+should be used when the observed indicators are not normally distributed
+[@spurk_latent_2020]. This statement is incorrect, and may lead readers
+to believe that they must use commercial software, as robust maximum
+likelihood is currently only implemented in Mplus and LatentGOLD. As
+explained before, mixture modeling assumes that observed data are a
+mixture of (multivariate) normal distributions; thus, the observed
+indicators will likely not be normally distributed.
 <!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->
@@
 
-As explained, LCA can be done in an exploratory or in a confirmatory fashion.
-In exploratory LCA, a sequence of models is fitted to the data 
-with each additional model estimating one more class than the previous model.
-These models are then compared 
-and the best solution is selected as the final class solution.
-In some cases, prior theory can inform the researcher 
-about the number of classes to expect.
-Even in such confirmatory LCA cases, 
-it is nonetheless useful to know if the theoretical model 
-is markedly better than those with differing numbers of classes.
-Therefore, it may always be useful to compare different class solutions.
-
-From a sequence of models, the final class solution is chosen 
-based on both theoretical and statistical criteria.
-Theory should drive the selection of indicator variables, 
-inform the expectations and reflect on the findings.
-In addition to this, there are several statistical criteria 
-to consider in model selection.
-These include but are not limited to likelihood ratio tests, 
-information criteria, 
-and the Bayes factor [@weller_latent_2020].
-
-Relative model fit can be examined using the likelihood ratio test. 
-This is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as 
-the difference in maximum log likelihoods of the two models,
-with the test degrees of freedom being 
-the difference in the degrees of freedom of the two compared models.
-The test statistic follows the $\chi^2$ distribution,
-and we want it to be non-significant
-in order to give support to the simpler model. 
-The likelihood ratio test can only compare two nested models at a time [@lanza_introduction_2003].
+As explained, LCA can be done in an exploratory or in a confirmatory
+fashion. In exploratory LCA, a sequence of models is fitted to the data
+with each additional model estimating one more class than the previous
+model. These models are then compared and the best solution is selected
+as the final class solution. In some cases, prior theory can inform the
+researcher about the number of classes to expect. Even in such
+confirmatory LCA cases, it is nonetheless useful to know if the
+theoretical model is markedly better than those with differing numbers
+of classes. Therefore, it may always be useful to compare different
+class solutions.
+
+From a sequence of models, the final class solution is chosen based on
+both theoretical and statistical criteria. Theory should drive the
+selection of indicator variables, inform the expectations and reflect on
+the findings. In addition to this, there are several statistical
+criteria to consider in model selection. These include but are not
+limited to likelihood ratio tests, information criteria, and the Bayes
+factor [@weller_latent_2020; @masyn_latent_2013].
+
+Relative model fit can be examined using the likelihood ratio test. This
+is only appropriate when the two models we wish to compare are nested.
+The likelihood ratio test statistic is computed as the difference in
+maximum log-likelihoods of the two models, with the test degrees of
+freedom being the difference in the degrees of freedom of the two
+compared models. The test statistic follows the $\chi^2$ distribution,
+and we want it to be non-significant in order to give support to the
+simpler model. The likelihood ratio test can only compare two nested
+models at a time [@lanza_introduction_2003].
 
@@
 
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC) 
-and the Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2*log likelihood 
-(which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance model fit and model complexity.
-The lower the value of an information criterion, 
-the better the overall fit of the model.
-The BIC applies a stronger penalty for model complexity 
-that scales logarithmically with the sample size.
-The literature suggests the BIC may be 
-the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
-Both the AIC and the BIC are available in the `tidySEM` output.
-
-Information criteria may occasionally contradict each other, 
-so it is important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer 
-the most parsimonious model that has best fit 
-according to any of the available fit indices.
-Yet another option is to incorporate information from 
-multiple fit indices using the analytic hierarchy process 
-[@akogul_comparison_2016].
-Finally, one might make an elbow plot and 
-compare multiple information criteria
+Fit indices typically used for determining the optimal number of classes
+include the Akaike Information Criterion (AIC) and the Bayesian
+Information Criterion (BIC). Both information criteria are based on the
+-2\*log-likelihood (which is lower for better fitting models), and add a
+penalty for the number of parameters (thus incentivizing simpler
+models). This helps balance model fit and model complexity. The lower
+the value of an information criterion, the better the overall fit of the
+model.
+
+The general objective of the Information Criteria (IC) is to evaluate the model's out-of-sample predictive accuracy, so adjusting for over fitting.
+Fit measures like $R^2$ evaluate the in-sample predictive accuracy, meaning that they evaluate the models ability to predict the observed outcomes based on the same data that was use to build up the model.
+These metrics are positively bias, meaning that will present better model fit than it has in reality.
+While IC corrects for this positive bias by evaluating the models accuracy approximating the out-of-sample predictive accuracy, meaning that it is the ability to predict the outcome for observations that are not part of the training model.
+Ideally we would estimate the model a lot of time with less observations and predict their scores, but IC's approximate this by their different penalty metrics.
+For example $R^2$ will increase even if an added predictors are unnecessary, while IC's will show worst fit when a predictor (complexity) is unnecessary [@mcelreath_statistical_2020]
+
+The BIC applies a stronger penalty for model complexity that
+scales logarithmically with the sample size. The literature suggests the
+BIC may be the most appropriate information criterion to use for model
+comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. Both the AIC and the BIC are
+available in the `tidySEM` output.
+
+Information criteria may occasionally contradict each other, so it is
+important to identify a suitable strategy to reconcile them. One option
+is to select a specific fit index before analyzing the data. Another
+option is to always prefer the most parsimonious model that has best fit
+according to any of the available fit indices. Yet another option is to
+incorporate information from multiple fit indices using the analytic
+hierarchy process [@akogul_comparison_2016]. Finally, one might make an
+elbow plot and compare multiple information criteria
 [@nylund-gibson_ten_2018].
 
-Another common test of model fit is 
-the likelihood ratio $\chi^2$ goodness-of-fit test.
-However, this test is not implemented in `tidySEM`.
-
-LCA studies commonly report -2*log likelihood of the final class solution.
-This is a basic fit measure used to compute most information criteria.
-However, since log likelihood is not penalized for model complexity,
-it will continuously fall with the addition of more classes.
-
-An alternative is using the bootstrapped likelihood ratio test
-which can be run using `tidySEM::BLRT()`.
-Currently, this test is computationally expensive and 
-can be slow on most computers.
-A faster version of this test, 
-namely an implementation of *the lazy bootstrap* [@van_kollenburg_lazy_2018]
-to `tidySEM` is being developed.
+Another common test of model fit is the likelihood ratio $\chi^2$
+goodness-of-fit test. However, this test is not implemented in
+`tidySEM`.
+
+A disadvantage of the IC's is that we do not have standard errors for them, so we only have the absolute values without a measure of their variability.
+So, the difference between models IC can be very small.
+Still indicating that the model with the lower value is "better", but if this difference is  very small can considered them "functionally equal", and you should take into consideration the interpretability of the model.
+
+LCA studies commonly report -2\*log likelihood of the final class
+solution. This is a basic fit measure used to compute most information
+criteria. However, since log likelihood is not penalized for model
+complexity, it will continuously fall with the addition of more classes.
+
+An alternative is using the bootstrapped likelihood ratio test which can
+be run using `tidySEM::BLRT()`. Currently, this test is computationally
+expensive and can be slow on most computers. A faster version of this
+test, namely an implementation of *the lazy bootstrap*
+[@van_kollenburg_lazy_2018] to `tidySEM` is being developed.
 
@@
 
-Best models will divide the sample into subgroups 
-which are internally homogeneous and externally distinct.
-Classification diagnostics give us a way to assess 
-the degree to which this is the case.
-They are separate from model fit indices as 
-a model can fit the data well 
-but show poor latent class separation [@masyn_latent_2013].
-Classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions 
-because they are uninterpretable.
-Interpretability should always be a consideration 
-when considering different class solutions [@nylund-gibson_ten_2018].
+Best models will divide the sample into subgroups which are internally
+homogeneous and externally distinct. Classification diagnostics give us
+a way to assess the degree to which this is the case. They are separate
+from model fit indices as a model can fit the data well but show poor
+latent class separation [@masyn_latent_2013]. Classification diagnostics
+should not be used for model selection, but they can be used to
+disqualify certain solutions because they are uninterpretable.
+Interpretability should always be a consideration when considering
+different class solutions [@nylund-gibson_ten_2018].
 
 Three important classification diagnostics provided by `tidySEM` are:
-(1) *the minimum* and *maximum percentage of the sample assigned to a particular class*,
-(2) *the range of the posterior class probabilities by most likely class membership*, 
-and (3) *entropy*. 
-All three are based on posterior class probabilities.
-
-The posterior class probability is a measure of classification uncertainty
-which can be computed for each individual,
-or averaged for each latent class.
-When the posterior class probability is computed for each individual in the dataset,
-it represents each person's probability of belonging to each latent class.
-For each person, the highest posterior class probability is then determined 
-and the individual is assigned to the corresponding class. 
-We want each individual's posterior class probabilities to be 
-high for one and low for the remaining latent classes. 
-This is considered a high classification accuracy 
-and means that the classes are distinct.
-To obtain posterior class probabilities, run `tidySEM::class_prob()`.
-This function produces output comprised of several elements:
+(1) the *minimum* and *maximum* percentage of the sample assigned to a
+particular *class*, (2) the *range of the posterior class probabilities* by most likely class membership, and (3) *entropy*. All three are based
+on posterior class probabilities.
+<!-- MGV: I will work on adding average posterior class probability (AvePP) as well-->
+
+The posterior class probability is a measure of classification
+uncertainty which can be computed for each individual, or averaged for
+each latent class. When the posterior class probability is computed for
+each individual in the dataset, it represents each person's probability
+of belonging to each latent class. For each person, the highest
+posterior class probability is then determined and the individual is
+assigned to the corresponding class. We want each individual's posterior
+class probabilities to be high for one and low for the remaining latent
+classes. This is considered a high classification accuracy and means
+that the classes are distinct. To obtain posterior class probabilities,
+run `tidySEM::class_prob()`. This function produces output comprised of
+several elements:
 
@@
 `$sum.mostlikely` is a summary table of the most likely class membership
-based on the highest posterior class probability.
-From this table, we compute the minimum and maximum 
-percentage of the sample assigned to a particular class,
-, i.e. **n_min** 
-(the smallest class proportion based on the posterior class probabilities)
-and **n_max** 
-(the largest class proportion based on the posterior class probabilities).
-We are especially interested in **n_min** as
-if it is very small and comprised of few observations,
-the model for that group might not be locally identified.
-It may be impossible to calculate descriptive statistics for such a small class.
-Estimating LCA parameters on small subsamples
-might lead to bias in the results.
-Therefore, we advise caution when dealing with small classes.
-
-`$mostlikely.class` is a table with rows representing
-the class the person was assigned to,
-and the columns indicating the average posterior probability.
-The diagonal represents the probability that 
-observations in each class will be correctly classified.
-If any of the values on the diagonal of this table is low,
-we might consider not to interpret that solution.
-In `tidySEM` we use the diagonal to compute the range of 
-the posterior class probabilities by most likely class membership
-which consists of
-the lowest class posterior probability (**prob_min**),
-and the highest posterior probability (**prob_max**).
-Both **prob_min** and **prob_max** can be used to 
-disqualify certain class solutions,
-and are a convenient way to summarize class separation in LCA.
-We want both **prob_min** and **prob_max** to be high as 
-that means that for all classes
-the people who were assigned to that class 
-have a high probability of being there.
-**prob_min** is especially important as it can diagnose 
-if there is a class with low posterior probabilities 
-which could make one reconsider that class solution.
-
-`$avg.mostlikely` contains the average posterior probabilities for each class, 
-for the subset of observations with most likely class of 1:k, 
+based on the highest posterior class probability. From this table, we
+compute the minimum and maximum percentage of the sample assigned to a
+particular class, , i.e. **n_min** (the smallest class proportion based
+on the posterior class probabilities) and **n_max** (the largest class
+proportion based on the posterior class probabilities). We are
+especially interested in **n_min** as if it is very small and comprised
+of few observations, the model for that group might not be locally
+identified. It may be impossible to calculate descriptive statistics for
+such a small class. Estimating LCA parameters on small subsamples might
+lead to bias in the results. Therefore, we advise caution when dealing
+with small classes.
+
+`$mostlikely.class` is a table with rows representing the class the
+person was assigned to, and the columns indicating the average posterior
+probability. The diagonal represents the probability that observations
+in each class will be correctly classified. If any of the values on the
+diagonal of this table is low, we might consider not to interpret that
+solution. In `tidySEM` we use the diagonal to compute the range of the
+posterior class probabilities by most likely class membership which
+consists of the lowest class posterior probability (**prob_min**), and
+the highest posterior probability (**prob_max**). Both **prob_min** and
+**prob_max** can be used to disqualify certain class solutions, and are
+a convenient way to summarize class separation in LCA. We want both
+**prob_min** and **prob_max** to be high as that means that for all
+classes the people who were assigned to that class have a high
+probability of being there. **prob_min** is especially important as it
+can diagnose if there is a class with low posterior probabilities which
+could make one reconsider that class solution.
+
+`$avg.mostlikely` contains the average posterior probabilities for each
+class, for the subset of observations with most likely class of 1:k,
 where k is the number of classes.
 
-`$individual` is the individual posterior probability matrix, 
-with dimensions n (number of cases in the data) x k (number of classes).
-Individual class probabilities are often useful for researchers
-who wish to do follow up analyses.
-
-Entropy is a summary measure of posterior class probabilities 
-across classes and individuals.
-It ranges from 0 (model classification no better than random chance) 
-to 1 (perfect classification).
-As a rule of thumb, values above .80 are deemed acceptable 
-and those approaching 1 are considered ideal.
-An appropriate use of entropy is that it can disqualify certain solutions 
-if class separability is too low. 
-Entropy was not built for nor should it be used for 
-model selection during class enumeration [@masyn_latent_2013].
-
-**n_min**, **n_max**, **prob_min**, **prob_max**, and **entropy** and can be obtained using `tidySEM::table_fit()`.
+`$individual` is the individual posterior probability matrix, with
+dimensions n (number of cases in the data) x k (number of classes). Additionally it includes the `predicted` class in function of the highest predicted probability.
+Individual class probabilities and/or predicted class are often useful for researchers who wish to do follow up analyses.
+
+Entropy is a summary measure of posterior class probabilities across
+classes and individuals. It ranges from 0 (model classification no
+better than random chance) to 1 (perfect classification). As a rule of
+thumb, values above .80 are deemed acceptable and those approaching 1
+are considered ideal. An appropriate use of entropy is that it can
+disqualify certain solutions if class separability is too low. Entropy
+was not built for nor should it be used for model selection during class
+enumeration [@masyn_latent_2013].
+
+**n_min**, **n_max**, **prob_min**, **prob_max**, and **entropy** and
+can be obtained using `tidySEM::table_fit()`.
 
@@
 
-An important outcome of LCA are conditional item probabilities,
-also known as class-specific item probabilities [@masyn_latent_2013], 
-conditional response, or conditional solution probabilities [@geiser_data_2012].
-They indicate the probability of an item being endorsed
-given that the observation belongs to a particular latent class.
-Conditional item probabilities can be obtained using `tidySEM::table_prob()`.
-If a particular item is endorsed by two or more classes at markedly different rates,
-it is said to discriminate well between the classes 
-and is consequently considered a good indicator.
-Classes are considered highly homogeneous with respect to an item 
-when for a particular item 
-there is a distinct difference in conditional item probabilities
-for two or more classes.
-For instance, if an item is endorsed below 30% for one class and 
-above 70% for another class,
-the classes have high homogeneity with respect to this item [@masyn_latent_2013].
-Conditional item probabilities are the analogue of mean and standard deviation
-when the indicators are binary or ordinal.
-
-A problem which can occur is that of inadmissible solutions. 
-With binary indicators, LCA is modelling a cross-table with all the predictors.
-The problem with such cross-tables is that they will often contain empty cells,
-i.e. combinations of responses that never occur together. 
-This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
-Such boundary parameter estimates could indicate that the solution is invalid [@geiser_data_2012].
-Boundary parameter estimates can also happen with continuous indicators.
-For instance, if we have a zero-inflated normal distribution and a two class solution,
-one class might have the mean of zero and its standard deviation cannot be determined 
-since there is little variance. 
-This too could be a sign of an invalid solution,
-warn that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].
+An important outcome of LCA are conditional item probabilities, also
+known as class-specific item probabilities [@masyn_latent_2013],
+conditional response, or conditional solution probabilities
+[@geiser_data_2012]. They indicate the probability of an item being
+endorsed given that the observation belongs to a particular latent
+class. Conditional item probabilities can be obtained using
+`tidySEM::table_prob()`. If a particular item is endorsed by two or more
+classes at markedly different rates, it is said to discriminate well
+between the classes and is consequently considered a good indicator.
+Classes are considered highly homogeneous with respect to an item when
+for a particular item there is a distinct difference in conditional item
+probabilities for two or more classes. For instance, if an item is
+endorsed below 30% for one class and above 70% for another class, the
+classes have high homogeneity with respect to this item
+[@masyn_latent_2013]. Conditional item probabilities are the analogue of
+mean and standard deviation when the indicators are binary or ordinal.
+
+A problem which can occur is that of inadmissible solutions. With binary
+indicators, LCA is modelling a cross-table with all the predictors. The
+problem with such cross-tables is that they will often contain empty
+cells, i.e. combinations of responses that never occur together. This
+problem is reflected by extreme conditional item probabilities (as in
+exactly 0 or 1). Such boundary parameter estimates could indicate that
+the solution is invalid [@geiser_data_2012]. Boundary parameter
+estimates can also happen with continuous indicators. For instance, if
+we have a zero-inflated normal distribution and a two class solution,
+one class might have the mean of zero and its standard deviation cannot
+be determined since there is little variance. This too could be a sign
+of an invalid solution, warn that too many classes were extracted, or
+indicate a local optimum [@geiser_data_2012].
 
@@
 
-The final class solution will usually discover and enumerate several classes. 
-The class ordering, however, is completely arbitrary.
-The class labeled as Class 1 in one solution 
-may become Class 2 or Class 3 in another model,
-even when the only difference between the models is in their starting values. 
-Label switching is something to be mindful of when comparing different LCA models [@masyn_latent_2013].
-
-The order of clusters is nondeterministic when using K-means in `tidySEM`. 
-Therefore label switching is still a consideration.
-A simple solution to this is setting a random seed number one line prior to fitting the model.
-We advise `tidySEM` users to always do so in order to circumvent label switching.
-
-Class names should be chosen to accurately reflect group membership. 
-Overly simplified and generalized class names may prove misleading 
-to both audiences and researches alike
-leading to what is known as a naming fallacy [@weller_latent_2020].
+The final class solution will usually discover and enumerate several
+classes. The class ordering, however, is completely arbitrary. The class
+labeled as Class 1 in one solution may become Class 2 or Class 3 in
+another model, even when the only difference between the models is in
+their starting values. Label switching is something to be mindful of
+when comparing different LCA models [@masyn_latent_2013].
+
+The order of clusters is nondeterministic when using K-means in
+`tidySEM`. Therefore label switching is still a consideration. A simple
+solution to this is setting a random seed number one line prior to
+fitting the model. We advise `tidySEM` users to always do so in order to
+circumvent label switching.
+
+Class names should be chosen to accurately reflect group membership, as theoretically relevant names according to the interpretation of the class solutions.
+Overly simplified and generalized class names may prove misleading to
+both audiences and researches alike leading to what is known as a naming
+fallacy [@weller_latent_2020].
 
@@
 
-Among studies using LCA, reporting practices vary significantly [@weller_latent_2020].
-Various authors have tried to improve and standardize ways of reporting LCA 
-[e.g. @masyn_latent_2013; @weller_latent_2020], 
-but more work is needed.
-@van_lissa_worcs_2020 developed WORCS,
-a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects 
-based on the TOP-guidelines developed by @nosek_promoting_2015.
-WORCS workflow can be easily implemented in R in form of 
-an R package which facilitates preregistration, article drafting,
-version control, citation and formatting, among others [@van_lissa_worcs_2020].
-
-TOP-guidelines emphasise the use of comprehensive citation 
-(including referencing the software used in the analysis),
-as well as code and data sharing wherever possible [@nosek_promoting_2015].
-@van_lissa_worcs_2020 suggest sharing synthetic data in case 
-the original data cannot be shared, 
-and provide functions to generate such synthetic data.
-Ideally, the entire research project is made reproducible 
-so that others may download it and reproduce it with just one click; 
-for guidance, see @van_lissa_worcs_2020.
-
-As the open science movement is gaining momentum,
-researchers are becoming increasingly aware
-how important it is that analyses can be reproduced and audited.
-In line with open science principles, 
-one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that 
-user-friendly methods for estimating LCA models 
-have predominantly been available in commercial software packages 
-(e.g., *Mplus* and *Latent GOLD*).
-A potential downside of commercial software is that 
-it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because 
-the underlying source code is proprietary.
-To overcome these limitations, 
-the present paper introduces new user-friendly functions in 
-the `tidySEM` R-package that can be used to 
-estimate a wide range of LCA models 
-using the free, open-source R-package `OpenMx`.
-The reporting guidelines described in this paper 
-are adopted in `tidySEM` by default.
-The `tidySEM` R-package thus makes advanced mixture modeling 
-based on best practices widely accessible,
-and facilitates the adoption of the estimation and 
-reporting guidelines described in this paper.
+Among studies using LCA, reporting practices vary significantly
+[@weller_latent_2020]. Various authors have tried to improve and
+standardize ways of reporting LCA [e.g. @masyn_latent_2013;
+@weller_latent_2020], but more work is needed. @van_lissa_worcs_2020
+developed WORCS, a workflow for open reproducible code in science. WORCS
+consists of step-by-step guidelines for research projects based on the
+TOP-guidelines developed by @nosek_promoting_2015. WORCS workflow can be
+easily implemented in R in form of an R package which facilitates
+preregistration, article drafting, version control, citation and
+formatting, among others [@van_lissa_worcs_2020].
+
+TOP-guidelines emphasize the use of comprehensive citation (including
+referencing the software used in the analysis), as well as code and data
+sharing wherever possible [@nosek_promoting_2015]. @van_lissa_worcs_2020
+suggest sharing synthetic data in case the original data cannot be
+shared, and provide functions to generate such synthetic data. Ideally,
+the entire research project is made reproducible so that others may
+download it and reproduce it with just one click; for guidance, see
+@van_lissa_worcs_2020.
+
+As the open science movement is gaining momentum, researchers are
+becoming increasingly aware how important it is that analyses can be
+reproduced and audited. In line with open science principles, one of the
+suggested reporting standards relates to reproducible code. In this
+context, it is important to note that user-friendly methods for
+estimating LCA models have predominantly been available in commercial
+software packages (e.g., *Mplus* and *Latent GOLD*). A potential
+downside of commercial software is that it restricts the ability to
+reproduce analyses to license holders, and prevents auditing research
+because the underlying source code is proprietary. To overcome these
+limitations, the present paper introduces new user-friendly functions in
+the `tidySEM` R-package that can be used to estimate a wide range of LCA
+models using the free, open-source R-package `OpenMx` as the backend. The reporting
+guidelines described in this paper are adopted in `tidySEM` by default.
+The `tidySEM` R-package thus makes advanced mixture modeling based on
+best practices widely accessible, and facilitates the adoption of the
+estimation and reporting guidelines described in this paper.
 
@@
 
-Plots can greatly improve the interpretability of LCA models.
-There are several stages where this is an important consideration.
-
-First, in class enumeration, 
-we want to compare several competing class solutions.
-This can be done by means of the AIC and BIC.
-Here, we suggest using an elbow plot
-with the $K-class\ solution$ on the X-axis,
-and information criteria value on the Y-axis.
-We show an implementation of such an elbow plot 
-in our Tutorial subsection.
+Plots can greatly improve the interpretability of LCA models. There are
+several stages where this is an important consideration.
+
+First, in class enumeration, we want to compare several competing class
+solutions. This can be done by means of the AIC and BIC. Here, we
+suggest using an elbow plot with the $K-class\ solution$ on the X-axis,
+and information criteria value on the Y-axis. We show an implementation
+of such an elbow plot in our Tutorial subsection.
 <!-- For continuous variables, density plots can aid class enumeration. -->
-<!-- DA: Do we have density plots implemented in tidySEM? -->
+<!-- DA: Do we have density plots implemented in tidySEM?, yes, it is plot_density  -->
 <!-- If so, we should add this example to the tutorial. -->
 
-Second, once we have decided on the final class solution,
-we want to interpret the response patterns
-on the indicators for each latent class.
-While `tidySEM` can create a table showing
-the probability of each item's response endorsement 
-(using `tidySEM::table_prob`),
-it may be easier to inspect these probabilities visually.
-For this reason, we created `ggplot2::plot_prob()`.
-The resulting shows response patterns 
-on all indicators for each group. 
-We give an example of this plot in the Tutorial subsection.
+Second, once we have decided on the final class solution, we want to
+interpret the response patterns on the indicators for each latent class.
+While `tidySEM` can create a table showing the probability of each
+item's response endorsement (using `tidySEM::table_prob`), it may be
+easier to inspect these probabilities visually. For this reason, we
+created `ggplot2::plot_prob()`. The resulting shows response patterns on
+all indicators for each group. We give an example of this plot in the
+Tutorial subsection.
 
@@
 
-In this example, we use *data_mix_ordinal*, 
-a simulated data for mixture modelling with ordinal indicators. 
-This dataset is built-in to `tidySEM`.
-For more information about the dataset, 
-type `?tidySEM::data_mix_ordinal` into the R console
-after loading the `tidySEM` using library().
+In this example, we use *data_mix_ordinal*, a simulated data for mixture
+modelling with ordinal indicators. This dataset is built-in to
+`tidySEM`. For more information about the dataset, type
+`?tidySEM::data_mix_ordinal` into the R console after loading the
+`tidySEM` using library().
 
@@
 
-An important step to preceding any statistical analysis is data exploration.
-Here we use `tidySEM::descriptives()` 
-to describe the dataset we are using.
+An important step to preceding any statistical analysis is data
+exploration. Here we use `tidySEM::descriptives()` to describe the
+dataset we are using.
 
@@
 ```
+
 As we can see, the output includes various descriptives of our dataset.
-Special attention should be paid to examining the pattern of missingness,
-as discussed in the *Handling missing data* section of this paper.
-In our example, we see that there are no missing values,
-hence we proceed with our analysis.
+Special attention should be paid to examining the pattern of
+missingness, as discussed in the *Handling missing data* section of this
+paper. In our example, we see that there are no missing values, hence we
+proceed with our analysis.
 
@@
 
-Before we fit a series of LCA models, we set a random seed
-using `set.seed()`.
-The seed number can be any digit.
-This is an important step as 
-there is some inherent randomness in the LCA computations,
-and having the same seed number ensures that
-two separate researcher obtain exactly the same results
-when fitting LCA models.
-
-Finally, we reach the step of fitting LCA models.
-To do so, we use `tidySEM::mx_lca()`
-which takes data and number of classes as inputs.
-In our example, we want to fit 1 to 4 class solutions
-and compare their output.
-Depending on your computer's computational power,
-this might take a while.
+Before we fit a series of LCA models, we set a random seed using
+`set.seed()`. The seed number can be any digit. This is an important
+step as there is some inherent randomness in the LCA computations, and
+having the same seed number ensures that two separate researcher obtain
+exactly the same results when fitting LCA models.
+
+Finally, we reach the step of fitting LCA models. To do so, we use
+`tidySEM::mx_lca()` which takes data and number of classes as inputs.
+Notice that the analysis will be done on all the variable in the data set, so you have to make sure it only includes the relevant variables.
+In
+our example, we want to fit 1 to 4 class solutions and compare their
+output. Depending on your computer's computational power, this might
+take a while.
 
@@
 set.seed(123) # setting seed 
-res <- mx_lca(data = df, classes = 1:4) # fitting LCA 
-```
+res <- mx_lca(data = df, classes = 1:4) # fitting LCA - 1 to 4 classes
+```
+
+Notice that the model was run 10 times for each $K$ class solution.
+This because the functions runs a total of 10 **extra** times the model.
+This way we can choose the best fitting model across all the solutions.
+In this examples we see that all models converged without issues.
+
+```{r, include = TRUE, eval=F}
+is(res)
+is(res[[1]])
+```
+
+As `tidySEM::mx_lca()` was run for several models at the same time, the resulting object `res` is a `mixture_list` class where each $K$ class solution is an object in the list, and each of them are `OpenMx` models (`MxModel`).
 
@@
 
-In class enumeration, 
-we want to compare a sequence of LCA models fitted to the data.
-To aid the process, we create a model fit table using `tidySEM::table_fit()` 
-with the results object as the input.
-As the output contains a lot of information on each of the four fitted models,
-we select a subset of helpful model fit indices and classification diagnostics.
+In class enumeration, we want to compare a sequence of LCA models fitted
+to the data. To aid the process, we create a model fit table using
+`tidySEM::table_fit()` with the results object as the input. As the
+output contains a lot of information on each of the four fitted models,
+we select a subset of helpful model fit indices and classification
+diagnostics.
 
@@
 fit_table <- table_fit(res) # model fit table
-fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # our selection
-```
+fit_table[ , c("Name", "LL", "Parameters", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # our selection
+```
+
 Our selection of fit indices and classification diagnostics includes:
@@
 ```{r, echo=F}
-Selection <- c("Name", "LL", "AIC", "BIC", "prob_min", "prob_max",
+Selection <- c("Name", "LL", "Parameters", "AIC", "BIC", "prob_min", "prob_max",
           "n_min", "n_max")
@@
                  "the -2*log-likelihood of each model",
+                 "Number of estimated parameters in each model",
                  "the Akaike Information Criterion value",
@@
 ```
-We discussed several possible strategies to select the final class solution. 
-Here, we apply our own.
-
-To aid our interpretation of the results, 
-we create an elbow plot showing the trends in information criteria across four models.
+
+We discussed several possible strategies to select the final class
+solution. Here, we apply our own.
+
+To aid our interpretation of the results, we create an elbow plot
+showing the trends in information criteria across four models.
 
@@
 elbow_plot <- fit_table[ , c("Name", "AIC", "BIC")] # extract ICs
-elbow_plot %<>% pivot_longer(cols = c("AIC", "BIC"), names_to = "IC", values_to = "Value") # to long format
+elbow_plot <- pivot_longer(elbow_plot, cols = c("AIC", "BIC"), names_to = "IC", values_to = "Value") # to long format
 
@@
 ```
-From the elbow plot, we see that AIC has a lower penalty for model complexity.
-However, we are more interested in the BIC values, which are similar
-for the one, two and three-class solutions, but the four-class solution fits significantly worse.
-For this reason, we eliminate the four-class solution from the selection process.
-
-Then we examine the model fit table.
-As expected, the -2*log likelihood falls successively with each added class.
-As previously stated, classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions because they are uninterpretable.
-We see that prob_min for the four-class solution is low,
-knowing that this solution also has a high BIC, we disqualify this solution.
-
-Out of the remaining three solutions,
-we notice that entropy is the highest for the three-class solution,
-and it has a satisfactory prob_min and n_min.
-Based on this, we retain the three class solution in model selection.
-Note that entropy for the one-class solution will always equal to one,
-as it is 100% true that every case is in that class.
-Based on the low entropy of the two-class solution,
-we eliminate this model.
-
-Finally, when comparing the one and three-class solutions,
-we inspect the information criteria.
-For BIC, the one-class solution fits better, but the difference is marginal.
-AIC tells us that the added complexity
-of having three classes still explains the data better
-than a one-class solution. 
-Therefore, we select the three-class solution as our final-class solution.
+
+From the elbow plot, we see that AIC has a lower penalty for model
+complexity. However, we are more interested in the BIC values, which are
+similar for the one, two and three-class solutions, but the four-class
+solution fits significantly worse. For this reason, we eliminate the
+four-class solution from the selection process.
+
+Then we examine the model fit table. As expected, the -2\*log likelihood
+falls successively with each added class. As previously stated,
+classification diagnostics should not be used for model selection, but
+they can be used to disqualify certain solutions because they are
+uninterpretable. We see that prob_min and n_min for the four-class solution is
+low, knowing that this solution also has a high BIC, we disqualify this
+solution.
+
+Out of the remaining three solutions, we notice that entropy is the
+highest for the three-class solution, and it has a satisfactory prob_min
+and n_min. Based on this, we retain the three class solution in model
+selection. Note that entropy for the one-class solution will always
+equal to one, as it is 100% true that every case is in that class. Based
+on the low entropy of the two-class solution, we eliminate this model.
+
+Finally, when comparing the one and three-class solutions, we inspect
+the information criteria. For BIC, the one-class solution fits better,
+but the difference is marginal. AIC tells us that the added complexity
+of having three classes still explains the data better than a one-class
+solution. Therefore, we select the three-class solution as our
+final-class solution.
 
@@
 
-To aid our understanding of the final class solution,
-we use `ggplot2::plot_prob()` 
-with the results of the three-class model as the input.
-The resulting graph shows response patterns 
-on all the indicators for each group.
+To aid our understanding of the final class solution, we use
+`ggplot2::plot_prob()` with the results of the three-class model as the
+input. The resulting graph shows response patterns on all the indicators
+for each group.
 
 If we want to know the probability of each response option's endorsement
-for each class, we can use `tidySEM::table_prob`.
-These are thresholds for ordinal dependent variables in the probability scale.
+for each class, we can use `tidySEM::table_prob`. These are thresholds
+for ordinal dependent variables in the probability scale.
 
@@
 ```
-In the plot, we can see the distributions of the response probabilities on the indicators
-for each of the three classes.
-For instance, we see that in Class 1 the most common response to u2 is 2,
-while in Class 2 and Class 3 this is 0.
-We can also see that response 1 is a rare response not forming the majority in 
-any class.
-Class 2 distinguishes itself because the majority scores the response 0 category of u3 and u4,
-while in Class 1 and 2 this is not the case.
-Class 3 distinguishes itself because the most common response to u3 and u4 is 3.
+
+In the plot, we can see the distributions of the response probabilities
+on the indicators for each of the three classes. For instance, we see
+that in Class 1 the most common response to u2 is 2, while in Class 2
+and Class 3 this is 0. We can also see that response 1 is a rare
+response not forming the majority in any class. Class 2 distinguishes
+itself because the majority scores the response 0 category of u3 and u4,
+while in Class 1 and 2 this is not the case. Class 3 distinguishes
+itself because the most common response to u3 and u4 is 3.
 
 We can also interpret the response patterns numerically.
+It is a matter of preferences on how to interpret these probabilities.
+Here is where you would **name** each class, such that each response pattern is theoretically meaningful.
 
@@
 
-Another step is to extract posterior class probabilities.
-This is done by the use of `TidySEM::class_prob`
-with the results of the final class solution as the input.
+Another step is to extract posterior class probabilities. This is done
+by the use of `tidySEM::class_prob` with the results of the final class
+solution as the input.
 

### 2022-10-28 manuscript.tex
edits from Mauricio

@@
   pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
-  pdfauthor={Caspar J. van Lissa1,2},
+  pdfauthor={Caspar J. van Lissa1,2 \& Mauricio Garnier-Villarreal3},
   pdflang={en-EN},
@@
 \title{Best Practices in Latent Class Analysis using Free Open Source Software}
-\author{Caspar J. van Lissa\textsuperscript{1,2}}
+\author{Caspar J. van Lissa\textsuperscript{1,2} \& Mauricio Garnier-Villarreal\textsuperscript{3}}
 \date{}
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht\\\textsuperscript{3} Vrije Universiteit Amsterdam, Sociology}
 
@@
 Latent class analysis is a family of techniques for identifying groups in data based
-on a parametric model. Its examples include mixture models, latent profile
-analysis, latent class analysis with ordinal indicators, and latent class
+on a parametric model. Its examples include mixture models, latent class
+analysis (with continuous and/or ordinal indicators), and latent class (mixture)
 growth analysis. Despite the popularity of this technique, there is limited
@@
 
-Latent class analysis (LCA) is an umbrella term
-that refers to a number of techniques for estimating unobserved group membership
-based on a parametric model of one or more observed indicators of group membership.
+Latent class analysis (LCA) is an umbrella term that refers to a number
+of techniques for estimating unobserved group membership based on a
+parametric model of one or more observed indicators of group membership.
 The types of LCA have become quite popular across scientific fields,
-most notably finite Gaussian mixture modeling and latent profile analysis.
-Vermunt, J.K. et al. (2004) defined LCA more generally as virtually any statistical model
-where ``some of the parameters {[}\ldots{]} differ across unobserved subgroups''.
-
-Despite its popularity, there is a lack of standards for estimating and reporting LCA.
-While Van De Schoot, Sijbrandij, Winter, Depaoli, and Vermunt (2017) developed reporting guidelines
-for a specific type of LCA known as latent growth models,
-general reporting guidelines for all LCA are still lacking.
-This complicates manuscript review and assessment of the quality of published studies,
-and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature
-by suggesting updated guidelines for estimation and reporting of LCA,
-based on current best practices.
-Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples,
-this paper exclusively relies on free, open source software in R.
-Our goal is to make best-practices in LCA widely accessible.
+most notably finite Gaussian mixture modeling and latent profile
+analysis. Vermunt, J.K. et al. (2004) defined LCA more generally as
+virtually any statistical model where ``some of the parameters {[}\ldots{]}
+differ across unobserved subgroups''.
+
+Despite its popularity, there is a lack of standards for estimating and
+reporting LCA. While Van De Schoot, Sijbrandij, Winter, Depaoli, and Vermunt (2017) developed
+reporting guidelines for a specific type of LCA known as latent growth
+models, general reporting guidelines for all LCA are still lacking. This
+complicates manuscript review and assessment of the quality of published
+studies, and introduces a risk of misapplications of the technique. The
+present paper seeks to address this gap in the literature by suggesting
+updated guidelines for estimation and reporting of LCA, based on current
+best practices. Importantly, in order to lower the barrier of entry and
+ensure reproducibility of all examples, this paper exclusively relies on
+free, open source software in R. Our goal is to make best-practices in
+LCA widely accessible.
 
@@
 
-Latent class analysis is a group of methods for estimating unobserved groups
-based on a parametric model of observed indicators of group membership.
-The concept of LCA can be understood in different ways.
-A mixture model assumes that the study population
-is composed of \(K\) subpopulations or classes.
-It further assumes that the observed data are a mixture of data generated by class-specific models.
-The simplest univariate model is a normal distribution,
-which can be described with two parameters: the mean and the variance.
-Commonly, the same model is estimated across all classes,
-but with different parameters for each class (i.e., class-specific means and variances).
-Mixture modeling then estimates both the parameters for each class,
-and the probability that an individual belongs to each class.
-
-As an illustrative example, imagine that a detective wants to know
-if it would be possible to use mixture modeling to identify the sex of a suspect,
-based on footprints found at the crime scene.
-To test the feasibility of this approach,
-the detective records the shoe sizes and sex of 100 volunteers.
-The resulting observed data look like this:
+Latent class analysis is a group of methods for estimating unobserved
+groups based on a parametric model of observed indicators of group
+membership. The concept of LCA can be understood in different ways. A
+mixture model assumes that the study population is composed of \(K\)
+subpopulations or classes. It further assumes that the observed data are
+a mixture of data generated by class-specific models. The simplest
+univariate model is a normal distribution, which can be described with
+two parameters: the mean and the variance. Commonly, the same model is
+estimated across all classes, but with different parameters for each
+class (i.e., class-specific means and variances). Mixture modeling then
+estimates both the parameters for each class, and the probability that
+an individual belongs to each class. This way, describing the whole
+sample with a single mean and variance is a special case where we assume
+the sample is homogeneous in function of these parameters.
+
+As an illustrative example, imagine that a detective wants to know if it
+would be possible to use mixture modeling to identify the sex of a
+suspect, based on footprints found at the crime scene. To test the
+feasibility of this approach, the detective records the shoe sizes and
+sex of 100 volunteers. The resulting observed data look like this:
 
@@
 
-The distribution is evidently bimodal, which bodes well for the intended mixture model.
-In this case, the number of classes is known a-priori.
-When estimating a two-class mixture model, the detective observes that the model estimates
-the mean shoe size of the two groups are equal to 7.25
-and 9.22,
-which is close to the true means of the two groups,
-namely 9.04 and 6.93.
-When tabulating estimated group membership against observed (known) group membership,
-it can be seen that women are classified with a high degree of accuracy, but men are not:
+The distribution is evidently bimodal, which bodes well for the intended
+mixture model. In this case, the number of classes is known a-priori.
+When estimating a two-class mixture model, the detective observes that
+the model estimates the mean shoe size of the two groups are equal to
+9.22 and
+7.25, which is close to the true
+means of the two groups, namely 9.04
+and 6.93. When tabulating estimated
+group membership against observed (known) group membership, it can be
+seen that women are classified with a high degree of accuracy, but men
+are not:
 
@@
 \hline
-Man & 21 & 28\\
+Man & 28 & 21\\
 \hline
-Woman & 51 & 0\\
+Woman & 0 & 51\\
 \hline
@@
 
-A LCA is like confirmatory factor analysis,
-except that the continuous latent variable is substituted with a categorical latent variable.
-One difference between the two techniques is that
-factor analysis can be considered as a way to group observed \emph{variables} into latent constructs,
-and LCA groups \emph{individuals} into classes.
-In line with this distinction, LCA is sometimes referred to as a ``person-centered'' technique,
-and factor analysis as a ``variable-centered'' technique (Nylund-Gibson \& Choi, 2018).
-
-When the focus is on the model parameters in each group,
-LCA can be thought of as similar to a multi-group structural equation model.
-The main distinction is that group membership is not known a-priori,
-but is instead estimated -- with measurement error -- based on the data.
-Whereas in a multi-group model, the data are split by group and treated as independent samples,
-in LCA, all cases contribute to the estimation of all parameters in all groups.
-The relative contribution of each case to the parameters of each group
-is determined by that case's posterior probability of belonging to that group.
-
-When the focus is on each individual's estimated class membership,
-LCA can be thought of as a type of clustering algorithm.
-In line with this perspective, LCA is sometimes described as ``model-based
-clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
-Many clustering algorithms apply some recursive splitting algorithm to the data.
-By contrast ``model-based'' clustering refers to the fact that LCA
-estimates cluster membership based on a parametric model.
-Specifically, the posterior class probability that an individual belongs to a latent class
-can be computed from the likelihood of that individual's observed data under given the class-specific model.
+LCA is like confirmatory factor analysis, except that the continuous
+latent variable is substituted with a categorical latent variable. Both
+assume that the indicators are related, and these relations can be
+represented by the underlying latent variable. One difference between
+the two techniques is that factor analysis can be considered as a way to
+group observed \emph{variables} into latent constructs, and LCA groups
+\emph{individuals} into classes.While factor analysis seeks to describe the
+relations between variables, LCA seeks to describe each unobserved group
+and the differences between them. In line with this distinction, LCA
+(mixture models) is referred to as a ``person-centered'' technique, and
+factor analysis as a ``variable-centered'' technique
+(Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+
+When the focus is on the model parameters in each group, LCA can be
+thought of as similar to a multi-group structural equation model. The
+main distinction is that group membership is not known a-priori, but is
+instead estimated -- with measurement error -- based on the data.
+Whereas in a multi-group model, the data are split by group and treated
+as independent samples, in LCA, all cases contribute to the estimation
+of all parameters in all groups. The relative contribution of each case
+to the parameters of each group is determined by that case's posterior
+probability of belonging to that group.
+
+In factor analysis, the latent variable is continuous, so the relation
+between indicators is define by the communalities between them, and the
+latent variable represents the error corrected approximation of the
+sample. While in LCA, the relation between indicators is define by how
+well they separate the classes, as a multinomial regression between
+indicators and latent variable (Masyn, 2013)
+
+When the focus is on each individual's estimated class membership, LCA
+can be thought of as a type of clustering algorithm. In line with this
+perspective, LCA is sometimes described as ``model-based clustering''
+(Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016). Many clustering
+algorithms apply some recursive splitting algorithm to the data. By
+contrast ``model-based'' clustering refers to the fact that LCA estimates
+cluster membership based on a parametric model. Specifically, the
+posterior class probability that an individual belongs to a latent class
+can be computed from the likelihood of that individual's observed data
+under given the class-specific model.
+For these differences, LCA is a clustering method, that corrects for (random) measurement error, while clustering algorithms do not correct for.
+This is something LCA has in common with factor analysis, as both model an underlying latent variable, which accounts for the communalities between indicators and correcting for measurement error.
+0
 
@@
 \emph{unsupervised classification} problem (Figueiredo \& Jain, 2002).
-The term \emph{unsupervised} refers to the fact that the outcome variable
--- true class membership -- is not known,
-and the term \emph{classification} refers to the fact that
-the algorithm is predicting a categorical outcome -- class membership.
+The term \emph{unsupervised} refers to the fact that the outcome variable --
+true class membership -- is not known, and the term \emph{classification}
+refers to the fact that the algorithm is predicting a categorical
+outcome -- class membership.
 
@@
 
-In this paper, we use the term latent class analysis to refer to
-a family of techniques that estimate latent class membership
-based on a parametric model of observed indicators.
-From a historical perspective, the term was initially conceived to
-refer to analyses with categorical, usually binary indicators (Vermunt, J.K. et al., 2004).
-Nowadays, there are a number of related techniques,
-known by distinct names, that serve a similar purpose.
-The term ``latent class analysis'' seems most appropriate
-as an umbrella term for this broader class of models,
-as it only refers to the purpose of the analysis,
-and does not imply restrictions to the model used,
-or the level of measurement of the indicators.
-Given the abundance of terms in use for closely related classes of models,
-we will provide a rudimentary taxonomy of LCA methods.
-
-A common type of LCA is the \emph{finite Gaussian mixture model};
-a univariate analysis where the observed distribution of a single variable
-is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
-The parameters of a finite Gaussian mixture model are
-the means and variances of these underlying normal distributions.
-The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
-In the multivariate case, with more than one indicator variable,
-the parameters of a mixture model are the means, variances, and covariances
-between the indicators (which can be standardized to obtain correlations).
-These parameters can be estimated freely, or set to be constrained across classes.
-
-The technique known as \emph{latent profile analysis (LPA)} is a special case of the mixture model,
-which assumes conditional independence of the indicators.
-Conditional independence means that, after class membership is accounted for,
-the covariances/correlations between indicators are assumed to be zero.
-This can be conceived of as a restricted mixture model with covariances fixed to zero.
-In some cases, such constraints will be inappropriate;
-for instance when the cohesion between indicators is expected to differ between classes.
-As an example, a LPA of ocean plastic particles
-found two classes of particles based on length and width:
-a class of smaller particles with a high correlation between length and width,
-meaning that these particles were approximately round or square in shape,
-and a class of larger particles with a low correlation between length and width,
-meaning that these particles were heterogenous in shape.
-From a theoretical perspective, this makes sense,
-because the smaller particles have been ground down to a more uniform shape by the elements.
-
-It is also possible to estimate a mixture model based on latent indicators.
-This means that, within each class,
-one or more continuous latent variables are estimated based on the observed indicators.
-Categorical latent variable membership is then estimated
-based on these continuous latent variables.
-A common application of this approach is in longitudinal research,
-where the indicators reflect one construct assessed at different time points.
-Examples of this approach include \emph{growth mixture models} (GMM)
-and \emph{latent class growth analyses} (LCGA).
-These techniques estimate a latent growth model
-to describe individual trajectories over time.
-The growth mixture model is a latent class model
-where the parameters that indicate class membership
-are the intercepts and variances,
-and typically covariances of the latent growth variables,
-e.g., a latent intercept and slope.
-This technique assumes that individuals within a class can have heterogenous trajectories.
-If the variance of the growth parameters is fixed to zero,
-it is known as a latent class growth analysis.
-This latter approach assumes that all individuals within a class
-share the same identical trajectory,
-and that any variance in the indicators not explained by the class-specific latent trajectories
-is due to residual error variance.
-
-The term latent class analysis originally referred to cases
-where the observed indicators were categorical.
-Nowadays, it is more commonly used as an umbrella term.
-To prevent ambiguity, the special case
-where indicators are of binary or ordinal measurement level
-might be described as \emph{latent class analysis with ordinal indicators}.
-Latent class models with ordinal indicators are parameterized differently from mixture models.
-One common parameterization assumes that
-each categorical variable reflects an underlying standard normal distribution.
-The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution
-(with \(N(\mu = 0, \sigma = 1)\)).
-These thresholds are estimated based on
-the proportion of individuals in each of the response categories of the indicator variable.
-For example, a binary indicator has a single threshold that distinguishes the two response categories.
-If responses are distributed 50/50, then the corresponding threshold would be \(t_1 = 0.00\).
-If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).
-
-This paper will primarily focus on mixture models and special cases thereof,
-although most of the suggested guidelines are applicable to all LCA methods.
+In this paper, we use the term latent class analysis to refer to a
+family of techniques that estimate latent class membership based on a
+parametric model of observed indicators. From a historical perspective,
+the term was initially conceived to refer to analyses with categorical,
+usually binary indicators (Vermunt, J.K. et al., 2004). Nowadays, there are
+a number of related techniques, known by distinct names, that serve a
+similar purpose. The term ``latent class analysis'' seems most appropriate
+as an umbrella term for this broader class of models, as it only refers
+to the purpose of the analysis, and does not imply restrictions to the
+model used, or the level of measurement of the indicators. Given the
+abundance of terms in use for closely related classes of models, we will
+provide a rudimentary taxonomy of LCA methods.
+
+A common type of LCA is the \emph{finite Gaussian mixture model}; a
+univariate analysis where the observed distribution of a single variable
+is assumed to result from a mixture of a known number of normal
+(Gaussian) distributions. The parameters of a finite Gaussian mixture
+model are the means and variances of these underlying normal
+distributions. The analysis of shoe sizes presented earlier is a
+canonical example of this type of analysis. In the multivariate case,
+with more than one indicator variable, the parameters of a mixture model
+are the means, variances, and covariances between the indicators (which
+can be standardized to obtain correlations). These parameters can be
+estimated freely, or set to be constrained across classes (Collins \& Lanza, 2010).
+
+The technique known as \emph{latent profile analysis (LPA)} is a special case
+of the mixture model, which assumes conditional independence of the
+indicators. Conditional independence means that, after class membership
+is accounted for, the covariances/correlations between indicators are
+assumed to be zero. This can be conceived of as a restricted mixture
+model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
+will be inappropriate; for instance when the cohesion between indicators
+is expected to differ between classes (Collins \& Lanza, 2010). As an example, a LPA of ocean
+plastic particles found two classes of particles based on length and
+width: a class of smaller particles with a high correlation between
+length and width, meaning that these particles were approximately round
+or square in shape, and a class of larger particles with a low
+correlation between length and width, meaning that these particles were
+heterogenous in shape. From a theoretical perspective, this makes sense,
+because the smaller particles have been ground down to a more uniform
+shape by the elements.
+
+It is also possible to estimate a mixture model based on latent
+indicators. This means that, within each class, one or more continuous
+latent variables are estimated based on the observed indicators.
+Categorical latent variable membership is then estimated based on these
+continuous latent variables. A common application of this approach is in
+longitudinal research, where the indicators reflect one construct
+assessed at different time points. Examples of this approach include
+\emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA).
+These techniques estimate a latent growth model to describe grouped
+trajectories over time. The growth mixture model is a latent class model
+where the parameters that indicate class membership are the intercepts
+and variances, and typically covariances of the latent growth variables,
+e.g., a latent intercept and slope. This technique assumes that
+individuals within a class can have heterogenous trajectories. If the
+variance of the growth parameters is fixed to zero, it is known as a
+latent class growth analysis. This latter approach assumes that all
+individuals within a class share the same identical trajectory, and that
+any variance in the indicators not explained by the class-specific
+latent trajectories is due to residual error variance.
+
+The term latent class analysis originally referred to cases where the
+observed indicators were categorical (Collins \& Lanza, 2010).
+Nowadays, it is more commonly used
+as an umbrella term. To prevent ambiguity, the special case where
+indicators are of binary or ordinal measurement level might be described
+as \emph{latent class analysis with ordinal indicators}. Latent class models
+with ordinal indicators are parameterized differently from mixture
+models. One common parameterization assumes that each categorical
+variable reflects an underlying standard normal distribution. The
+parameters are ``thresholds'' that correspond to quantiles of a standard
+normal distribution (with \(N(\mu = 0, \sigma = 1)\)). These thresholds
+are estimated based on the proportion of individuals in each of the
+response categories of the indicator variable. For example, a binary
+indicator has a single threshold that distinguishes the two response
+categories. If responses are distributed 50/50, then the corresponding
+threshold would be \(t_1 = 0.00\). If the responses are distributed 60/40,
+then the resulting threshold would be \(t_1 = 0.25\).
+
+This paper will primarily focus on mixture models and special cases
+thereof, although most of the suggested guidelines are applicable to all
+LCA methods.
 
@@
 
-There are several use cases for which LCA methods are suitable.
-One example is to test a theory that postulates the existence of a categorical latent variable.
-For example, \emph{identity status theory} posits that, at any given point in time,
-adolescents reside in one of four identity statuses.
-Latent class analysis can be used to identify these four statuses
-based on observed indicators (e.g., self-reported identity exploration and commitment).
-If results indicate that the data are better described by a different number of classes,
-or that the four-class solution does not correspond to the predicted pattern of responses on the indicators,
-then the theory may be called into question.
-
-Another use case is unsupervised learning:
-when the goal is to restore unobserved class membership based on observed indicators,
-or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid
-when several clinical indicators can be used to distinguish between
-a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
-The example of shoe size is a rudimentary illustration of this type of application.
-
-LCA can be used as a descriptive analysis
-where a researcher wishes to describe their sample
-and identify a few prototypes based on many variables.
-As an example, if a survey among all Dutch academics was carried out
-with the goal of bringing about a funding reform,
-LCA could be used to discover the types of publications that get funding.
-
-With LCA, our goal could be to inductively identify the number of classes.
-For instance, if we believe that a variable represents a group,
-but don't know how many groups there are,
-LCA may be an appropriate technique to answer this question.
-For example, Hopfer, Tan, and Wylie (2014) studied substance use, sexual behavior,
-and mental health status of urban population in Winnipeg, Canada.
-The underlying assumption was that there were different risk profiles,
-but their number was not known.
-From a collection of indicators, LCA provided evidence
-that there are four distinct risk profiles in the Winnipeg area.
-
-Another application of LCA is to classify individuals.
-In a peer harassment study,
-Giang and Graham (2008) used LCA to classify over 2,000 sixth grade students
-into aggressor and victim latent classes.
-The five-class solution comprised of classes of
-victims, aggressors, and socially adjusted students.
-For instance, it revealed that there were two types of victims:
-highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
-
-LCA is also appropriate when we wish to identify indicators that capture classes well.
-High quality indicators will be strongly related to the latent variable
-and will lead to good class separation.
-This relationship of high quality indicators to the latent variable
-is reflected in very high or very low conditional response probabilities.
-For a simulation study exploring the effects of indicator quality on LCA, see Geiser and Wurpts (2014).
-Therefore, one could use conditional response probabilities for each item to assess its quality
-with regards to how well it helps separate the latent classes.
-From this, a theory about the selection of indicators could be informed.
-
-An extension of LCA is that containing covariates which can be used to predict class membership.
-In this approach, we not only model the latent class variable based on indicators,
-but we also relate the class membership to other explanatory variables (Vermunt (2017)).
-An example of using covariates comes from Nozadi et al. (2016)
-who applied LCA to identify the probability of children's membership to an anxiety class.
-The authors tested several covariates including children's age, sex, and accuracy scores.
-Age and sex were not found to be related to the children's latent class membership,
-hence these covariates were excluded from the analysis.
-Accuracy scores, however, were related to probabilities of
-being in anxiety and attention-anxiety classes.
-Therefore this covariate was kept as a valuable predictor.
-
-When our interest is the prediction of one or more outcomes,
-LCA can be used to construct latent classes as categorical predictors.
-Lanza, Tan, and Bray (2013) demonstrated how LCA can be used
-to classify adolescents into depression classes,
-and subsequently these classes can be used to predict smoking, grades, and delinquency.
-The study showed that the outcomes predicted by class membership
-can be binary (regular smoking), continuous (grades) or count (delinquency).
-
-In addition to these applications, LCA can be used for dimensionality reduction
-as the resulting groups summarize response patterns on a large number of indicators.
-For example, MacGregor et al. (2021) investigated symptom profiles
-among injured U.S. military personnel.
-They used fifteen dichotomous items
-from the Post-Deployment Health Assessment survey as LCA indicators.
-Combinatorics informs us that fifteen dichotomous items have
-\(2^{15}\) or \(32,768\) unique symptom combinations.
-Perhaps for this reason, MacGregor et al. (2021) incorporated LCA
-as a method of dimensionality reduction.
-A five class solution was found to have the best fit
-according to both statistical criteria and clinical interpretability.
-
-Finally, LCA can be used to deal with data which violate certain assumptions.
-As discussed in the shoe size example, LCA can deal with violations of normality.
-In fact, LCA assumes the population distribution is a non-normal mixture
-of \(K\) normal distributions,
-and it can discover the value of \(K\), i.e.~generate a \(K\)-class solution.
+There are several use cases for which LCA methods are suitable. One
+example is to test a theory that postulates the existence of a
+categorical latent variable. For example, \emph{identity status theory}
+posits that, at any given point in time, adolescents reside in one of
+four identity statuses. Latent class analysis can be used to identify
+these four statuses based on observed indicators (e.g., self-reported
+identity exploration and commitment). If results indicate that the data
+are better described by a different number of classes, or that the
+four-class solution does not correspond to the predicted pattern of
+responses on the indicators, then the theory may be called into
+question.
+
+Another use case is unsupervised learning: when the goal is to restore
+unobserved class membership based on observed indicators, or to classify
+individuals. For example, a mixture model can be used as a diagnostic
+aid when several clinical indicators can be used to distinguish between
+a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006), mental
+(Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems, or health outcomes after recovering from cancer (Hardy, Garnier-Villarreal, McCarthy, Anderson, \& Reynolds, 2018).
+The example of shoe size is a
+rudimentary illustration of this type of application.
+
+LCA can be used as a descriptive analysis where a researcher wishes to
+describe their sample and identify a few prototypes based on many
+variables. As an example, if a survey among all Dutch academics was
+carried out with the goal of bringing about a funding reform, LCA could
+be used to discover the types of publications that get funding.
+
+With LCA, our goal could be to inductively identify the number of
+classes. For instance, if we believe that a variable represents a group,
+but don't know how many groups there are, LCA may be an appropriate
+technique to answer this question. For example, Hopfer, Tan, and Wylie (2014)
+studied substance use, sexual behavior, and mental health status of
+urban population in Winnipeg, Canada. The underlying assumption was that
+there were different risk profiles, but their number was not known. From
+a collection of indicators, LCA provided evidence that there are four
+distinct risk profiles in the Winnipeg area.
+
+Another application of LCA is to classify individuals. In a peer
+harassment study, Giang and Graham (2008) used LCA to classify over 2,000
+sixth grade students into aggressor and victim latent classes. The
+five-class solution comprised of classes of victims, aggressors, and
+socially adjusted students. For instance, it revealed that there were
+two types of victims: highly-victimized aggressive-victims and
+highly-aggressive aggressive-victims.
+
+LCA is also appropriate when we wish to identify indicators that capture
+classes well. High quality indicators will be strongly related to the
+latent variable and will lead to good class separation. This
+relationship of high quality indicators to the latent variable is
+reflected in very high or very low conditional response probabilities.
+For a simulation study exploring the effects of indicator quality on
+LCA, see Geiser and Wurpts (2014). Therefore, one could use conditional response
+probabilities for each item to assess its quality with regards to how
+well it helps separate the latent classes. From this, a theory about the
+selection of indicators could be informed.
+
+An extension of LCA is that containing covariates which can be used to
+predict class membership. In this approach, we not only model the latent
+class variable based on indicators, but we also relate the class
+membership to other explanatory variables (Vermunt (2017)). An
+example of using covariates comes from Nozadi et al. (2016) who
+applied LCA to identify the probability of children's membership to an
+anxiety class. The authors tested several covariates including
+children's age, sex, and accuracy scores. Age and sex were not found to
+be related to the children's latent class membership, hence these
+covariates were excluded from the analysis. Accuracy scores, however,
+were related to probabilities of being in anxiety and attention-anxiety
+classes. Therefore this covariate was kept as a valuable predictor.
+
+When our interest is the prediction of one or more outcomes, LCA can be
+used to construct latent classes as categorical predictors.
+Lanza, Tan, and Bray (2013) demonstrated how LCA can be used to classify
+adolescents into depression classes, and subsequently these classes can
+be used to predict smoking, grades, and delinquency. The study showed
+that the outcomes predicted by class membership can be binary (regular
+smoking), continuous (grades) or count (delinquency).
+
+These mixture models can also be useful to evaluate quality of measurement across instrument.
+For example, employment status was collected across register and survey data, and there is disagreement between them.
+You could think that register data is without error, but it has been shown that is victim of measurement error as well.
+In this case mixture models was used to evaluate the underlying factor of ``error corrected employment status'', by using two indicators of the same question and identify random and systamtic measurement error in either indicator (Pavlopoulos \& Vermunt, 2015).
+
+In addition to these applications, LCA can be used for dimensionality
+reduction as the resulting groups summarize response patterns on a large
+number of indicators. For example, MacGregor et al. (2021) investigated
+symptom profiles among injured U.S. military personnel. They used
+fifteen dichotomous items from the Post-Deployment Health Assessment
+survey as LCA indicators. Combinatorics informs us that fifteen
+dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
+Perhaps for this reason, MacGregor et al. (2021) incorporated LCA as a
+method of dimensionality reduction. A five class solution was found to
+have the best fit according to both statistical criteria and clinical
+interpretability.
+
+Finally, LCA can be used to deal with data which violate certain
+assumptions. As discussed in the shoe size example, LCA can deal with
+violations of normality. In fact, LCA assumes the population
+distribution is a non-normal mixture of \(K\) normal distributions, and it
+can discover the value of \(K\), i.e.~generate a \(K\)-class solution.
+
+LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous of categorical.
+We encourage researchers to be agnostic about the \textbf{true} nature of the latent variable.
+And focus on the approach that fits the research question.
 
@@
 
-The best practices in estimation, as outlined in Table \ref{tab:checkest},
-are rooted in existing recommendations for best practices
-for estimating specific subtypes of LCA,
-including latent class growth analysis (Van De Schoot et al., 2017)
-and latent class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
-These were generalized to be relevant to all types of LCA,
-and updated to current best practices, as explained below.
+The best practices in estimation, as outlined in Table
+\ref{tab:checkest}, are rooted in existing recommendations for best
+practices for estimating specific subtypes of LCA, including latent
+class growth analysis (Van De Schoot et al., 2017) and latent
+class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
+These were generalized to be relevant to all types of LCA, and updated
+to current best practices, as explained below.
 
@@
 
-Examining observed data is essential for any analysis
-as it may reveal patterns and violations of assumptions
-that had not been considered prior to data collection.
-Special attention should be paid
-to level of measurement of the indicators.
-Finite Gaussian mixture models (including LPA)
-are only suitable for continuous variables.
-Indicators with an ordinal level of measurement
-are likely to violate the assumption of
-within-class normal distributions of mixture models
-(see Vermunt, 2011).
-Personal experience consulting on LCA methods and
-moderating the \texttt{tidyLPA} Google group
-suggest that the misapplication of mixture models
-to ordinal (e.g., Likert-type) indicators
-is the most common source of user error.
-Whereas it has been argued that some parametric methods are robust
-when scales with 7+ indicators are treated as continuous
-(e.g., Norman, 2010),
-this certainly does not imply that all methods are.
-It is certainly unlikely that such ordinal variables
-can be treated as a \emph{mixture} of multiple normal distributions.
-The problem becomes egregious when
-the number of classes estimated equals or exceeds the number of categories;
-in this case, each class-specific mean could describe a single response category,
-and a class-specific variance component would be nonsensical.
-In sum, Likert-type scales are rarely suitable for mixture modeling;
+Examining observed data is essential for any analysis as it may reveal
+patterns and violations of assumptions that had not been considered
+prior to data collection. Special attention should be paid to level of
+measurement of the indicators. Finite Gaussian mixture models (including
+LPA) are only suitable for continuous variables. Indicators with an
+ordinal level of measurement are likely to violate the assumption of
+within-class normal distributions of mixture models (see Vermunt, 2011). Personal experience consulting on LCA methods
+and moderating the \texttt{tidyLPA} Google group suggest that the
+misapplication of mixture models to ordinal (e.g., Likert-type)
+indicators is the most common source of user error. Whereas it has been
+argued that some parametric methods are robust when scales with 7+
+indicators are treated as continuous (e.g., Norman, 2010) or
+with 5+ categories (Rhemtulla, Brosseau-Liard, \& Savalei, 2012), this
+certainly does not imply that all methods are. It is certainly unlikely
+that such ordinal variables can be treated as a \emph{mixture} of multiple
+normal distributions. The problem becomes egregious when the number of
+classes estimated equals or exceeds the number of categories; in this
+case, each class-specific mean could describe a single response
+category, and a class-specific variance component would be nonsensical.
+In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
 latent class analysis with ordinal indicators is more appropriate.
 
-Relatedly, a recent publication claimed that
-an assumption of mixture models is that
-observed indicators are normally distributed (Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020).
-This is incorrect.
-When the number of classes is greater than one,
-mixture models assume that the observed indicators
-are a mixture of multiple (multivariate) normal distributions.
-In our shoe size example, it can be seen that the population distribution
-is comprised of two normal distributions.
-When examined visually, the population distribution is evidently bimodal.
-The Shapiro-Wilk normality test (\(W = 0.971\), \(p < 0.05\) )
-rejects the null hypothesis that
-the sample comes from a normally distributed population.
-Yet, this is a prototypical example of a mixed population distribution
-where LCA can discover latent groups.
-If the population distribution were instead normal,
-there would be no classes to extract
-as the whole population would belong to a single class.
-
-Extensive descriptive statistics
-(including the number of unique values,
+Relatedly, a recent publication claimed that an assumption of mixture
+models is that observed indicators are normally distributed
+(Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020). This is incorrect. When the number of classes is
+greater than one, Gaussian mixture models assume that the observed indicators are
+a mixture of multiple (multivariate) normal distributions. In our shoe
+size example, it can be seen that the population distribution is
+comprised of two normal distributions. When examined visually, the
+population distribution is evidently bimodal. The Shapiro-Wilk normality
+test (\(W = 0.971\), \(p < 0.05\) ) rejects the null hypothesis that the
+sample comes from a normally distributed population. Yet, this is a
+prototypical example of a mixed population distribution where LCA can
+discover latent groups. If the population distribution were instead
+normal, there would be no classes to extract as the whole population
+would belong to a single (homogeneous) class.
+
+Extensive descriptive statistics (including the number of unique values,
 variance of categorical variables, and missingness; see next subsection)
-can be obtained using the function \texttt{tidySEM::descriptives(data)}.
-Note, however, that sample-level descriptive statistics
-are of limited value when the goal of a study is
-to identify subsamples using latent class analysis.
-Plots (density plots for continuous variables,
-and bar charts for categorical ones) may be more diagnostic.
-Note that density plots can also aid in the choice of the number of classes,
-as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements,
-provided that these are readily accessible
-(consider using a GitHub repository as a comprehensive public research archive,
-as explained in Van Lissa et al., 2021).
+can be obtained using the function \texttt{tidySEM::descriptives(data)}. Note,
+however, that sample-level descriptive statistics are of limited value
+when the goal of a study is to identify subsamples using latent class
+analysis. Plots (density plots for continuous variables, and bar charts
+for categorical ones) may be more diagnostic. Note that density plots
+can also aid in the choice of the number of classes, as further
+explained in the section on visualization. Descriptive statistics and
+plots can be relegated to online supplements, provided that these are
+readily accessible (consider using a GitHub repository as a
+comprehensive public research archive, as explained in Van Lissa et al., 2021).
 
@@
 
-Previous work has emphasized the importance of examining
-the pattern of missing data and
-reporting how missingness was handled
-(Van De Schoot et al., 2017).
-Three types of missingness have been distinguished
-in the literature (Rubin, 1976):
-Missing completely at random (MCAR),
-which means that missingness is random;
-missing at random (MAR),
-which means that missingness is contingent on the \emph{observed} data
-(and can thus be accounted for);
-and finally missing not at random (MNAR),
-which means that missingness is related to unobserved factors.
-It is possible to conduct a so-called ``MCAR'' test,
-for example the non-parametric MCAR test (Jamshidian \& Jalal, 2010).
-But note that the name ``MCAR test'' is somewhat misleading,
-as the null-hypothesis of this test is that the data are not MAR,
-and a significant test statistic indicates that
-missingness is related to the observed data (MAR).
-A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on
-the comparison of variances across groups
-with different patterns of missing data,
-it assumes normality (Little, 1988).
-This assumption is tenuous in the context of LCA.
-A non-parametric MCAR test,
-as provided by Jamshidian and Jalal,
-may be more suitable (Jamshidian \& Jalal, 2010).
-Unfortunately, this test was removed from the central R-repository CRAN
-due to lack of maintenance.
-For this tutorial, I have re-implemented it in the \texttt{mice} package
-as \texttt{mice::mcar()},
-with a fast backend in C++ and new printing and plotting methods.
-
-While we concur that investigating missingness is due diligence,
-it is important to emphasize that
-missingness is adequately handled by default in many software packages for LCA,
-such as Mplus, and \texttt{OpenMx} which is the backend of \texttt{tidySEM}.
-These packages use Full Information Maximum Likelihood (FIML) estimation,
-which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data;
-on par with multiple imputation (Lee \& Shi, 2021).
-FIML estimation assumes that missingness is either MCAR or MAR.
-Thus, one would typically proceed with FIML,
-regardless of the outcome of an MCAR test.
-Although FIML does not, by default, handle missingness in exogenous variables
--- all indicator variables in LCA are endogenous,
-so this is not a concern.
-
-Multiple imputation is less suitable to LCA for two reasons.
-First, because LCA methods are often computationally expensive,
-and conducting them on multiple imputed datasets may be unfeasible.
-Second, because there is no straightforward way
-to integrate LCA results across multiple datasets.
-To conclude; our recommendation is to inspect missingness
-(e.g., using \texttt{mice::MCAR()})
-and report the proportion of missingness per variable
-(e.g., using \texttt{tidySEM::descriptives()}),
-before proceeding with FIML.
-One minor concern is that the K-means algorithm,
-which \texttt{tidySEM} uses for determining starting values,
-is \emph{not} robust to missing values.
-When it fails, \texttt{tidySEM} automatically switches to hierarchical clustering,
-unless the user specifies a different
+Previous work has emphasized the importance of examining the pattern of
+missing data and reporting how missingness was handled
+(Van De Schoot et al., 2017). Three types of missingness have
+been distinguished in the literature (Enders, 2010; Rubin, 1976): Missing
+completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
+missing at random (MAR), which means that missingness is contingent on
+the \emph{observed} data (and can thus be accounted for); and finally missing
+not at random (MNAR), which means that missingness is related to
+unobserved factors. It is possible to conduct a so-called ``MCAR'' test,
+for example the non-parametric MCAR test (Jamshidian \& Jalal, 2010). But
+note that the name ``MCAR test'' is somewhat misleading, as the
+null-hypothesis of this test is that the data are not MAR, and a
+significant test statistic indicates that missingness is related to the
+observed data (MAR). A non-significant test statistic does not
+distinguish between MCAR or MNAR. As Little's classic MCAR test relies
+on the comparison of variances across groups with different patterns of
+missing data, it assumes normality (Little, 1988). This assumption
+is tenuous in the context of LCA. A non-parametric MCAR test, as
+provided by Jamshidian and Jalal, may be more suitable
+(Jamshidian \& Jalal, 2010). Unfortunately, this test was removed from the
+central R-repository CRAN due to lack of maintenance. For this tutorial,
+I have re-implemented it in the \texttt{mice} package as \texttt{mice::mcar()}, with a
+fast backend in C++ and new printing and plotting methods.
+
+While we concur that investigating missingness is due diligence, it is
+important to emphasize that missingness is adequately handled by default
+in many software packages for LCA, such as Mplus, Latet GOLD, and \texttt{OpenMx} which is
+the backend of \texttt{tidySEM}. These packages use Full Information Maximum
+Likelihood (FIML) estimation, which makes use of all available
+information without imputing missing values. FIML is a best-practice
+solution for handling missing data; on par with multiple imputation when the estimation model includes the same variables as the imputation model
+(Lee \& Shi, 2021). FIML estimation assumes that missingness is
+either MCAR or MAR. Thus, one would typically proceed with FIML,
+regardless of the outcome of an MCAR test. Although FIML does not, by
+default, handle missingness in exogenous variables -- all indicator
+variables in LCA are endogenous, so this is not a concern.
+
+Multiple imputation is less suitable to LCA for two reasons. First,
+because LCA methods are often computationally expensive, and conducting
+them on multiple imputed datasets may be unfeasible. Second, because
+there is no straightforward way to integrate LCA results across multiple
+datasets. To conclude; our recommendation is to inspect missingness
+(e.g., using \texttt{mice::MCAR()}) and report the proportion of missingness
+per variable (e.g., using \texttt{tidySEM::descriptives()}), before proceeding
+with FIML. One minor concern is that the K-means algorithm, which
+\texttt{tidySEM} uses for determining starting values, is \emph{not} robust to
+missing values. When it fails, \texttt{tidySEM} automatically switches to
+hierarchical clustering, unless the user specifies a different
 clustering algorithm or uses manual starting values.
@@
 Van De Schoot et al. (2017) developed a protocol called
-Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
-Two of the GRoLTS guidelines (namely, 6a and 6b) refer to
-considering alternative model specifications.
-Both are discussing specific cases in latent trajectory model specification.
-The first is about whether the variance of the growth parameter
-is estimated freely or fixed,
-and the second is about whether conditional independence is assumed,
-as the researcher might also want to free-up the variance-covariance structure.
-In LCA, there are also many different ways to specify the model.
-Means, variances, and covariances between the indicators
-can either be constrained across classes, or estimated freely.
-Researchers using LCA should transparently report their chosen parametrization
-as well as discuss different parametrizations that were tested in the process.
-
-Different types of latent class models have different parameters.
-For example, mixture models and latent profile analyses
-typically have class-specific means, variances, and covariances.
-Latent growth analyses have the same parameters,
-but with respect to the latent growth variables.
-Latent class analyses with ordinal indicators have thresholds.
-All of these parameters can be freely estimated,
-constrained to be equal across classes,
-or fixed to a certain value (e.g., to zero).
-The total number of parameters thus
-scales with the number of estimated classes.
-Consequently, LCA methods have a potentially very high number of parameters.
-As any of these parameters could be misspecified,
-it is important to consider alternative model specifications.
-However, alternative model specifications may be approached differently
-depending on whether an analysis is data driven (exploratory),
-or theory driven (confirmatory).
-This distinction has remained underemphasized in prior writing.
+Guidelines for Reporting on Latent Trajectory Studies (GRoLTS). Two of
+the GRoLTS guidelines (namely, 6a and 6b) refer to considering
+alternative model specifications. Both are discussing specific cases in
+latent trajectory model specification. The first is about whether the
+variance of the growth parameter is estimated freely or fixed, and the
+second is about whether conditional independence is assumed, as the
+researcher might also want to free-up the variance-covariance structure.
+In LCA, there are also many different ways to specify the model. Means,
+variances, and covariances between the indicators can either be
+constrained across classes, or estimated freely. Researchers using LCA
+should transparently report their chosen parametrization as well as
+discuss different parametrizations that were tested in the process.
+
+Different types of latent class models have different parameters. For
+example, mixture models and latent profile analyses typically have
+class-specific means, variances, and covariances. Latent growth analyses
+have the same parameters, but with respect to the latent growth
+variables. Latent class analyses with ordinal indicators have
+thresholds. All of these parameters can be freely estimated, constrained
+to be equal across classes, or fixed to a certain value (e.g., to zero).
+The total number of parameters thus scales with the number of estimated
+classes. Consequently, LCA methods have a potentially very high number
+of parameters. As any of these parameters could be misspecified, it is
+important to consider alternative model specifications. However,
+alternative model specifications may be approached differently depending
+on whether an analysis is data driven (exploratory), or theory driven
+(confirmatory). This distinction has remained underemphasized in prior
+writing.
 
 Prior literature on LCA has emphasized its exploratory applications
-(Nylund, Asparouhov, \& Muthén, 2007).
-In exploratory LCA,
-a large number of models are typically estimated in batch,
-with varying numbers of classes and model specifications.
-The ``correct'' model specification is then determined
-based on a combination of fit indices,
-significance tests, and interpretability.
-For latent profile analysis,
-the function \texttt{tidySEM::mx\_profiles(classes,\ variances,\ covariances)}
-largely automates this process.
-The argument \texttt{classes} indicates which
-class solutions should be estimated (e.g., 1 through 6).
-The argument \texttt{variances} specifies whether
-variances should be \texttt{"equal"} or \texttt{"varying"} across classes.
-The argument \texttt{covariances} specifies whether
-covariances should be constrained to
-\texttt{"zero"}, \texttt{"equal"} or \texttt{"varying"} across classes.
-The means are free to vary across classes by default,
-although the more general function \texttt{tidySEM::mx\_mixture()}
-could be used to circumvent this.
-After all models have been estimated,
-the function \texttt{tidySEM::table\_fit()} can be used
-to obtain a model fit table suitable for
-determining the optimal model according to best practices.
-Note, however, that this table does not include
-the bootstrapped likelihood ratio test (BLRT) by default,
-because this test is very computationally expensive.
-It is recommended to use the function \texttt{tidySEM::BLRT()}
-to compare a shortlist of likely candidate models
-based on other fit indices.
-More on fit indices can be found in
-the Model Fit Indices subsection of this paper.
-
-Confirmatory LCA typically requires
-less comprehensive alternative model specifications.
-For example, in the context of a preregistered analysis,
-the main models of interest would have been specified a priori.
-Even in this case,
-the theoretical model could be compared to a few others to contextualize it.
+(Nylund, Asparouhov, \& Muthén, 2007). In exploratory LCA, a large number of models
+are typically estimated in batch, with varying numbers of classes and
+model specifications. The ``correct'' model specification is then
+determined based on a combination of fit indices, significance tests,
+and interpretability. For latent profile analysis, the function
+\texttt{tidySEM::mx\_profiles(classes,\ variances,\ covariances)} largely
+automates this process. The argument \texttt{classes} indicates which class
+solutions should be estimated (e.g., 1 through 6). The argument
+\texttt{variances} specifies whether variances should be \texttt{"equal"} or
+\texttt{"varying"} across classes. The argument \texttt{covariances} specifies whether
+covariances should be constrained to \texttt{"zero"}, \texttt{"equal"} or \texttt{"varying"}
+across classes. The means are free to vary across classes by default,
+although the more general function \texttt{tidySEM::mx\_mixture()} could be used
+to circumvent this. After all models have been estimated, the function
+\texttt{tidySEM::table\_fit()} can be used to obtain a model fit table suitable
+for determining the optimal model according to best practices. Note,
+however, that this table does not include the bootstrapped likelihood
+ratio test (BLRT) by default, because this test is very computationally
+expensive. It is recommended to use the function \texttt{tidySEM::BLRT()} to
+compare a shortlist of likely candidate models based on other fit
+indices. More on fit indices can be found in the Model Fit Indices
+subsection of this paper.
+
+Confirmatory LCA typically requires less comprehensive alternative model
+specifications. For example, in the context of a preregistered analysis,
+the main models of interest would have been specified a priori. Even in
+this case, the theoretical model could be compared to a few others to
+contextualize it.
 
@@
 Many software packages are available for the estimation of LCA models.
-Some of these packages have limited functionality,
-or implement specific innovations.
-Other packages implement LCA in the context of
-a more flexible structural equation modeling framework.
-The most notable examples of the latter are
-the commercial programs Mplus and Latent GOLD,
-and the free open source R-package OpenMx.
-The commercial packages stand out
-because they offer relatively user-friendly interfaces
-and implement sensible defaults for complex analyses,
-including LCA methods.
+Some of these packages have limited functionality, or implement specific
+innovations. Other packages implement LCA in the context of a more
+flexible structural equation modeling framework. The most notable
+examples of the latter are the commercial programs Mplus and Latent
+GOLD, and the free open source R-package OpenMx. The commercial packages
+stand out because they offer relatively user-friendly interfaces and
+implement sensible defaults for complex analyses, including LCA methods.
 This lowers the threshold for applied researchers to adopt such methods.
-Commercial software also has several downsides, however.
-One such downside is that use of the software
-is restricted to those individuals
-and institutions who can afford a license.
-Another downside is that the source code,
-being proprietary, cannot be audited, debugged,
-or enhanced by third parties.
-This incurs the risk that mistakes in the source code
-may go unnoticed, and curbs progress as
-software developers cannot add new functionality.
-
-Conversely, the free open source program OpenMx is very flexible,
-but not very user-friendly.
-We directly address this limitation using the \texttt{tidySEM} R-package.
-New functionality in \texttt{tidySEM} seeks to lower the threshold
-for latent class analysis using \texttt{OpenMx}.
-It adheres to best practices in estimation and reporting,
-as described in this paper.
-The user interface is simple,
-making use of the model syntax of the widely used \texttt{lavaan} R-package.
-This syntax offers a human-readable way to specify latent variable models.
-Minor enhancements are made to simplify the specification of LCA.
-
-Because of the limitations of the aforementioned tools,
-we set out to develop a free tool that
-provides sensible defaults and is easy to use,
-but provides the option to access and modify
-all of the model inputs (i.e., low barrier, high ceiling).
-\texttt{tidySEM} interfaces with existing tools,
-and is able to translate between what existing tools are capable of
-and what researchers and analysts
-carrying-out person-oriented analyses would like to specify.
-Furthermore,\texttt{tidySEM} facilitates fully-reproducible analyses
-and contributes to open science.
+Commercial software also has several downsides, however. One such
+downside is that use of the software is restricted to those individuals
+and institutions who can afford a license. Another downside is that the
+source code, being proprietary, cannot be audited, debugged, or enhanced
+by third parties. This incurs the risk that mistakes in the source code
+may go unnoticed, and curbs progress as software developers cannot add
+new functionality.
+
+Conversely, the free open source program OpenMx (Neale et al., 2016) is very flexible, but
+not very user-friendly. We directly address this limitation using the
+\texttt{tidySEM} R-package. New functionality in \texttt{tidySEM} seeks to lower the
+threshold for latent class analysis using \texttt{OpenMx}. It adheres to best
+practices in estimation and reporting, as described in this paper. The
+user interface is simple, making use of the model syntax of the widely
+used \texttt{lavaan} R-package (Rosseel, 2012). This syntax offers a human-readable way to
+specify latent variable models. Minor enhancements are made to simplify
+the specification of LCA.
+
+Because of the limitations of the aforementioned tools, we set out to
+develop a free tool that provides sensible defaults and is easy to use,
+but provides the option to access and modify all of the model inputs
+(i.e., low barrier, high ceiling). \texttt{tidySEM} interfaces with existing
+tools, and is able to translate between what existing tools are capable
+of and what researchers and analysts carrying-out person-oriented
+analyses would like to specify. Furthermore,\texttt{tidySEM} facilitates
+fully-reproducible analyses and contributes to open science.
 
@@
 
-LCA parameters and model fit statistics
-can be estimated in a variety of ways.
-The choice of the estimator
-depends on the presence of missing values,
-sample size, number of indicators,
-and available computational resources (Weller, Bowen, \& Faubert, 2020).
-A commonly used technique is maximum likelihood (ML) estimation
-with the expectation-maximization (EM) algorithm as a local optimizer.
-Imagine we are estimating two parameters,
-e.g.~the class-specific means \(\mu_c\) on a continuous indicator
-(ignoring the variance for now).
-The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes
-the likelihood (\(LL\)) of all observed data.
-In practice, instead of maximizing \(LL\),
-often \(-2*LL\) is minimized,
-as this offers computational advantages.
-We can think of this optimization problem
-as a three-dimensional landscape:
-The X and Y dimensions are determined by the class-specific means,
-so \(X = \mu_1\) and \(Y = \mu_2\) -
-and the Z-dimension is determined by \(Z = -2*LL\).
-The optimizer must find the deepest ``valley'' in this landscape,
-which reflects the combination of \(\mu_1\) and \(\mu_2\) that
-maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble dropped in this landscape.
-It is dropped at some random point in space,
-and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley,
-it will settle on the bottom of that valley
-(this is known as ``convergence'').
-It cannot climb out again.
-Thus, if their are multiple valleys,
-the risk is that the optimizer gets stuck in a shallower valley (a ``local optimum''),
-and never discovers the deepest valley (the ``global optimum'', or best solution).
-One solution to this problem is to drop many marbles at random places,
-compare their final \(-2*LL\) values,
-choose the solution with the lowest \(-2*LL\),
-and make sure that several marbles replicated this solution.
-This is the ``random starts'' approach.
-
-One problem with the random starts approach is that
-it is computationally expensive to run this many replications.
-Moreover, because the algorithm begins with random starting values,
-many of the marbles are likely to be very far away from a ``good enough'' solution.
-Two innovations may improve the estimation procedure.
-The first is that, instead of picking random starting values,
-a ``reasonable solution'' may be used for the starting values.
-For example, if we assume that the different classes are likely to have
-different mean values on the indicators,
-then the K-means clustering algorithm can be used to
-determine these cluster centroids.
-We can compute the expected values of all model parameters
-by treating the K-means solution as a known class solution,
-and use these as starting values for a mixture model.
-One remaining concern is that this approach may result
-in starting values close to a local optimum,
-and that the EM algorithm will thus never find the global optimum.
-A second innovation addresses this concern.
-Instead of using EM,
-it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some ``destination'' in the landscape,
-and compares its likelihood to the current one.
-If the destination likelihood is higher,
-the estimator moves there.
-If the destination likelihood is \emph{lower},
-the estimator still moves there occasionally,
-based on probability.
-This latter property allows it to escape local optima,
-and find the global optimum.
+LCA parameters and model fit statistics can be estimated in a variety of
+ways. The choice of the estimator depends on the presence of missing
+values, sample size, number of indicators, and available computational
+resources (Weller, Bowen, \& Faubert, 2020). A commonly used technique is maximum
+likelihood (ML) estimation with the expectation-maximization (EM)
+algorithm as a local optimizer. Imagine we are estimating two
+parameters, e.g.~the class-specific means \(\mu_c\) on a continuous
+indicator (ignoring the variance for now). The EM algorithm will attempt
+to find a combination of values for these two parameters that maximizes
+the likelihood (\(LL\)) of all observed data. In practice, instead of
+maximizing \(LL\), often \(-2*LL\) is minimized, as this offers
+computational advantages. We can think of this optimization problem as a
+three-dimensional landscape: The X and Y dimensions are determined by
+the class-specific means, so \(X = \mu_1\) and \(Y = \mu_2\) - and the
+Z-dimension is determined by \(Z = -2*LL\). The optimizer must find the
+deepest ``valley'' in this landscape, which reflects the combination of
+\(\mu_1\) and \(\mu_2\) that maximizes the likelihood of the data. The EM
+optimizer behaves somewhat like a marble dropped in this landscape. It
+is dropped at some random point in space, and will roll into the nearest
+valley. The problem is that, once EM rolls into a valley, it will settle
+on the bottom of that valley (this is known as ``convergence''). It cannot
+climb out again. Thus, if their are multiple valleys, the risk is that
+the optimizer gets stuck in a shallower valley (a ``local optimum''), and
+never discovers the deepest valley (the ``global optimum'', or best
+solution). One solution to this problem is to drop many marbles at
+random places, compare their final \(-2*LL\) values, choose the solution
+with the lowest \(-2*LL\), and make sure that several marbles replicated
+this solution. This is the ``random starts'' approach.
+
+One problem with the random starts approach is that it is
+computationally expensive to run this many replications. Moreover,
+because the algorithm begins with random starting values, many of the
+marbles are likely to be very far away from a ``good enough'' solution.
+Two innovations may improve the estimation procedure. The first is that,
+instead of picking random starting values, a ``reasonable solution'' may
+be used for the starting values. For example, if we assume that the
+different classes are likely to have different mean values on the
+indicators, then the K-means clustering algorithm can be used to
+determine these cluster centroids. We can compute the expected values of
+all model parameters by treating the K-means solution as a known class
+solution, and use these as starting values for a mixture model. One
+remaining concern is that this approach may result in starting values
+close to a local optimum, and that the EM algorithm will thus never find
+the global optimum. A second innovation addresses this concern. Instead
+of using EM, it is possible to use an optimizer that can climb out of a
+valley. Simulated annealing iteratively considers some ``destination'' in
+the landscape, and compares its likelihood to the current one. If the
+destination likelihood is higher, the estimator moves there. If the
+destination likelihood is \emph{lower}, the estimator still moves there
+occasionally, based on probability. This latter property allows it to
+escape local optima, and find the global optimum.
 
@@
 and identifies the global optimum solution using simulated annealing.
-Once a solution has been found,
-simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for
-the parameters that can be used to compute standard errors.
-Note that these defaults can be manually overridden.
+Once a solution has been found, simulated annealing is followed up with
+a short run of the EM algorithm, as EM inherently produces an asymptotic
+covariance matrix for the parameters that can be used to compute
+standard errors. Note that these defaults can be manually overridden.
 
 A recent paper suggested maximum likelihood with robust standard errors
-should be used when the observed indicators are not normally distributed (Spurk et al., 2020).
-This statement is incorrect, and may lead readers to believe that
-they must use commercial software,
-as robust maximum likelihood is currently only implemented
-in Mplus and latentGOLD.
-As explained before,
-mixture modeling assumes that observed data are
-a mixture of (multivariate) normal distributions;
-thus, the observed indicators will likely not be normally distributed.
+should be used when the observed indicators are not normally distributed
+(Spurk et al., 2020). This statement is incorrect, and may lead readers
+to believe that they must use commercial software, as robust maximum
+likelihood is currently only implemented in Mplus and LatentGOLD. As
+explained before, mixture modeling assumes that observed data are a
+mixture of (multivariate) normal distributions; thus, the observed
+indicators will likely not be normally distributed.
 
@@
 
-As explained, LCA can be done in an exploratory or in a confirmatory fashion.
-In exploratory LCA, a sequence of models is fitted to the data
-with each additional model estimating one more class than the previous model.
-These models are then compared
-and the best solution is selected as the final class solution.
-In some cases, prior theory can inform the researcher
-about the number of classes to expect.
-Even in such confirmatory LCA cases,
-it is nonetheless useful to know if the theoretical model
-is markedly better than those with differing numbers of classes.
-Therefore, it may always be useful to compare different class solutions.
-
-From a sequence of models, the final class solution is chosen
-based on both theoretical and statistical criteria.
-Theory should drive the selection of indicator variables,
-inform the expectations and reflect on the findings.
-In addition to this, there are several statistical criteria
-to consider in model selection.
-These include but are not limited to likelihood ratio tests,
-information criteria,
-and the Bayes factor (Weller et al., 2020).
-
-Relative model fit can be examined using the likelihood ratio test.
-This is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as
-the difference in maximum log likelihoods of the two models,
-with the test degrees of freedom being
-the difference in the degrees of freedom of the two compared models.
-The test statistic follows the \(\chi^2\) distribution,
-and we want it to be non-significant
-in order to give support to the simpler model.
-The likelihood ratio test can only compare two nested models at a time (Lanza, Bray, \& Collins, 2003).
+As explained, LCA can be done in an exploratory or in a confirmatory
+fashion. In exploratory LCA, a sequence of models is fitted to the data
+with each additional model estimating one more class than the previous
+model. These models are then compared and the best solution is selected
+as the final class solution. In some cases, prior theory can inform the
+researcher about the number of classes to expect. Even in such
+confirmatory LCA cases, it is nonetheless useful to know if the
+theoretical model is markedly better than those with differing numbers
+of classes. Therefore, it may always be useful to compare different
+class solutions.
+
+From a sequence of models, the final class solution is chosen based on
+both theoretical and statistical criteria. Theory should drive the
+selection of indicator variables, inform the expectations and reflect on
+the findings. In addition to this, there are several statistical
+criteria to consider in model selection. These include but are not
+limited to likelihood ratio tests, information criteria, and the Bayes
+factor (Masyn, 2013; Weller et al., 2020).
+
+Relative model fit can be examined using the likelihood ratio test. This
+is only appropriate when the two models we wish to compare are nested.
+The likelihood ratio test statistic is computed as the difference in
+maximum log-likelihoods of the two models, with the test degrees of
+freedom being the difference in the degrees of freedom of the two
+compared models. The test statistic follows the \(\chi^2\) distribution,
+and we want it to be non-significant in order to give support to the
+simpler model. The likelihood ratio test can only compare two nested
+models at a time (Lanza, Bray, \& Collins, 2003).
 
@@
 
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC)
-and the Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2*log likelihood
-(which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance model fit and model complexity.
-The lower the value of an information criterion,
-the better the overall fit of the model.
-The BIC applies a stronger penalty for model complexity
-that scales logarithmically with the sample size.
-The literature suggests the BIC may be
-the most appropriate information criterion to use for model comparison (Nylund-Gibson \& Choi, 2018).
-Both the AIC and the BIC are available in the \texttt{tidySEM} output.
-
-Information criteria may occasionally contradict each other,
-so it is important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer
-the most parsimonious model that has best fit
-according to any of the available fit indices.
-Yet another option is to incorporate information from
-multiple fit indices using the analytic hierarchy process
-(Akogul \& Erisoglu, 2016).
-Finally, one might make an elbow plot and
-compare multiple information criteria
+Fit indices typically used for determining the optimal number of classes
+include the Akaike Information Criterion (AIC) and the Bayesian
+Information Criterion (BIC). Both information criteria are based on the
+-2*log-likelihood (which is lower for better fitting models), and add a
+penalty for the number of parameters (thus incentivizing simpler
+models). This helps balance model fit and model complexity. The lower
+the value of an information criterion, the better the overall fit of the
+model.
+
+The general objective of the Information Criteria (IC) is to evaluate the model's out-of-sample predictive accuracy, so adjusting for over fitting.
+Fit measures like \(R^2\) evaluate the in-sample predictive accuracy, meaning that they evaluate the models ability to predict the observed outcomes based on the same data that was use to build up the model.
+These metrics are positively bias, meaning that will present better model fit than it has in reality.
+While IC corrects for this positive bias by evaluating the models accuracy approximating the out-of-sample predictive accuracy, meaning that it is the ability to predict the outcome for observations that are not part of the training model.
+Ideally we would estimate the model a lot of time with less observations and predict their scores, but IC's approximate this by their different penalty metrics.
+For example \(R^2\) will increase even if an added predictors are unnecessary, while IC's will show worst fit when a predictor (complexity) is unnecessary (McElreath, 2020)
+
+The BIC applies a stronger penalty for model complexity that
+scales logarithmically with the sample size. The literature suggests the
+BIC may be the most appropriate information criterion to use for model
+comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018). Both the AIC and the BIC are
+available in the \texttt{tidySEM} output.
+
+Information criteria may occasionally contradict each other, so it is
+important to identify a suitable strategy to reconcile them. One option
+is to select a specific fit index before analyzing the data. Another
+option is to always prefer the most parsimonious model that has best fit
+according to any of the available fit indices. Yet another option is to
+incorporate information from multiple fit indices using the analytic
+hierarchy process (Akogul \& Erisoglu, 2016). Finally, one might make an
+elbow plot and compare multiple information criteria
 (Nylund-Gibson \& Choi, 2018).
 
-Another common test of model fit is
-the likelihood ratio \(\chi^2\) goodness-of-fit test.
-However, this test is not implemented in \texttt{tidySEM}.
-
-LCA studies commonly report -2*log likelihood of the final class solution.
-This is a basic fit measure used to compute most information criteria.
-However, since log likelihood is not penalized for model complexity,
-it will continuously fall with the addition of more classes.
-
-An alternative is using the bootstrapped likelihood ratio test
-which can be run using \texttt{tidySEM::BLRT()}.
-Currently, this test is computationally expensive and
-can be slow on most computers.
-A faster version of this test,
-namely an implementation of \emph{the lazy bootstrap} (Kollenburg, Mulder, \& Vermunt, 2018)
-to \texttt{tidySEM} is being developed.
+Another common test of model fit is the likelihood ratio \(\chi^2\)
+goodness-of-fit test. However, this test is not implemented in
+\texttt{tidySEM}.
+
+A disadvantage of the IC's is that we do not have standard errors for them, so we only have the absolute values without a measure of their variability.
+So, the difference between models IC can be very small.
+Still indicating that the model with the lower value is ``better'', but if this difference is very small can considered them ``functionally equal'', and you should take into consideration the interpretability of the model.
+
+LCA studies commonly report -2*log likelihood of the final class
+solution. This is a basic fit measure used to compute most information
+criteria. However, since log likelihood is not penalized for model
+complexity, it will continuously fall with the addition of more classes.
+
+An alternative is using the bootstrapped likelihood ratio test which can
+be run using \texttt{tidySEM::BLRT()}. Currently, this test is computationally
+expensive and can be slow on most computers. A faster version of this
+test, namely an implementation of \emph{the lazy bootstrap}
+(Kollenburg, Mulder, \& Vermunt, 2018) to \texttt{tidySEM} is being developed.
 
@@
 
-Best models will divide the sample into subgroups
-which are internally homogeneous and externally distinct.
-Classification diagnostics give us a way to assess
-the degree to which this is the case.
-They are separate from model fit indices as
-a model can fit the data well
-but show poor latent class separation (Masyn, 2013).
-Classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions
-because they are uninterpretable.
-Interpretability should always be a consideration
-when considering different class solutions (Nylund-Gibson \& Choi, 2018).
+Best models will divide the sample into subgroups which are internally
+homogeneous and externally distinct. Classification diagnostics give us
+a way to assess the degree to which this is the case. They are separate
+from model fit indices as a model can fit the data well but show poor
+latent class separation (Masyn, 2013). Classification diagnostics
+should not be used for model selection, but they can be used to
+disqualify certain solutions because they are uninterpretable.
+Interpretability should always be a consideration when considering
+different class solutions (Nylund-Gibson \& Choi, 2018).
 
 Three important classification diagnostics provided by \texttt{tidySEM} are:
-(1) \emph{the minimum} and \emph{maximum percentage of the sample assigned to a particular class},
-(2) \emph{the range of the posterior class probabilities by most likely class membership},
-and (3) \emph{entropy}.
-All three are based on posterior class probabilities.
-
-The posterior class probability is a measure of classification uncertainty
-which can be computed for each individual,
-or averaged for each latent class.
-When the posterior class probability is computed for each individual in the dataset,
-it represents each person's probability of belonging to each latent class.
-For each person, the highest posterior class probability is then determined
-and the individual is assigned to the corresponding class.
-We want each individual's posterior class probabilities to be
-high for one and low for the remaining latent classes.
-This is considered a high classification accuracy
-and means that the classes are distinct.
-To obtain posterior class probabilities, run \texttt{tidySEM::class\_prob()}.
-This function produces output comprised of several elements:
+(1) the \emph{minimum} and \emph{maximum} percentage of the sample assigned to a
+particular \emph{class}, (2) the \emph{range of the posterior class probabilities} by most likely class membership, and (3) \emph{entropy}. All three are based
+on posterior class probabilities.
+
+The posterior class probability is a measure of classification
+uncertainty which can be computed for each individual, or averaged for
+each latent class. When the posterior class probability is computed for
+each individual in the dataset, it represents each person's probability
+of belonging to each latent class. For each person, the highest
+posterior class probability is then determined and the individual is
+assigned to the corresponding class. We want each individual's posterior
+class probabilities to be high for one and low for the remaining latent
+classes. This is considered a high classification accuracy and means
+that the classes are distinct. To obtain posterior class probabilities,
+run \texttt{tidySEM::class\_prob()}. This function produces output comprised of
+several elements:
 
@@
 \texttt{\$sum.mostlikely} is a summary table of the most likely class membership
-based on the highest posterior class probability.
-From this table, we compute the minimum and maximum
-percentage of the sample assigned to a particular class,
-, i.e.~\textbf{n\_min}
-(the smallest class proportion based on the posterior class probabilities)
-and \textbf{n\_max}
-(the largest class proportion based on the posterior class probabilities).
-We are especially interested in \textbf{n\_min} as
-if it is very small and comprised of few observations,
-the model for that group might not be locally identified.
-It may be impossible to calculate descriptive statistics for such a small class.
-Estimating LCA parameters on small subsamples
-might lead to bias in the results.
-Therefore, we advise caution when dealing with small classes.
-
-\texttt{\$mostlikely.class} is a table with rows representing
-the class the person was assigned to,
-and the columns indicating the average posterior probability.
-The diagonal represents the probability that
-observations in each class will be correctly classified.
-If any of the values on the diagonal of this table is low,
-we might consider not to interpret that solution.
-In \texttt{tidySEM} we use the diagonal to compute the range of
-the posterior class probabilities by most likely class membership
-which consists of
-the lowest class posterior probability (\textbf{prob\_min}),
-and the highest posterior probability (\textbf{prob\_max}).
-Both \textbf{prob\_min} and \textbf{prob\_max} can be used to
-disqualify certain class solutions,
-and are a convenient way to summarize class separation in LCA.
-We want both \textbf{prob\_min} and \textbf{prob\_max} to be high as
-that means that for all classes
-the people who were assigned to that class
-have a high probability of being there.
-\textbf{prob\_min} is especially important as it can diagnose
-if there is a class with low posterior probabilities
-which could make one reconsider that class solution.
-
-\texttt{\$avg.mostlikely} contains the average posterior probabilities for each class,
-for the subset of observations with most likely class of 1:k,
+based on the highest posterior class probability. From this table, we
+compute the minimum and maximum percentage of the sample assigned to a
+particular class, , i.e.~\textbf{n\_min} (the smallest class proportion based
+on the posterior class probabilities) and \textbf{n\_max} (the largest class
+proportion based on the posterior class probabilities). We are
+especially interested in \textbf{n\_min} as if it is very small and comprised
+of few observations, the model for that group might not be locally
+identified. It may be impossible to calculate descriptive statistics for
+such a small class. Estimating LCA parameters on small subsamples might
+lead to bias in the results. Therefore, we advise caution when dealing
+with small classes.
+
+\texttt{\$mostlikely.class} is a table with rows representing the class the
+person was assigned to, and the columns indicating the average posterior
+probability. The diagonal represents the probability that observations
+in each class will be correctly classified. If any of the values on the
+diagonal of this table is low, we might consider not to interpret that
+solution. In \texttt{tidySEM} we use the diagonal to compute the range of the
+posterior class probabilities by most likely class membership which
+consists of the lowest class posterior probability (\textbf{prob\_min}), and
+the highest posterior probability (\textbf{prob\_max}). Both \textbf{prob\_min} and
+\textbf{prob\_max} can be used to disqualify certain class solutions, and are
+a convenient way to summarize class separation in LCA. We want both
+\textbf{prob\_min} and \textbf{prob\_max} to be high as that means that for all
+classes the people who were assigned to that class have a high
+probability of being there. \textbf{prob\_min} is especially important as it
+can diagnose if there is a class with low posterior probabilities which
+could make one reconsider that class solution.
+
+\texttt{\$avg.mostlikely} contains the average posterior probabilities for each
+class, for the subset of observations with most likely class of 1:k,
 where k is the number of classes.
 
-\texttt{\$individual} is the individual posterior probability matrix,
-with dimensions n (number of cases in the data) x k (number of classes).
-Individual class probabilities are often useful for researchers
-who wish to do follow up analyses.
-
-Entropy is a summary measure of posterior class probabilities
-across classes and individuals.
-It ranges from 0 (model classification no better than random chance)
-to 1 (perfect classification).
-As a rule of thumb, values above .80 are deemed acceptable
-and those approaching 1 are considered ideal.
-An appropriate use of entropy is that it can disqualify certain solutions
-if class separability is too low.
-Entropy was not built for nor should it be used for
-model selection during class enumeration (Masyn, 2013).
-
-\textbf{n\_min}, \textbf{n\_max}, \textbf{prob\_min}, \textbf{prob\_max}, and \textbf{entropy} and can be obtained using \texttt{tidySEM::table\_fit()}.
+\texttt{\$individual} is the individual posterior probability matrix, with
+dimensions n (number of cases in the data) x k (number of classes). Additionally it includes the \texttt{predicted} class in function of the highest predicted probability.
+Individual class probabilities and/or predicted class are often useful for researchers who wish to do follow up analyses.
+
+Entropy is a summary measure of posterior class probabilities across
+classes and individuals. It ranges from 0 (model classification no
+better than random chance) to 1 (perfect classification). As a rule of
+thumb, values above .80 are deemed acceptable and those approaching 1
+are considered ideal. An appropriate use of entropy is that it can
+disqualify certain solutions if class separability is too low. Entropy
+was not built for nor should it be used for model selection during class
+enumeration (Masyn, 2013).
+
+\textbf{n\_min}, \textbf{n\_max}, \textbf{prob\_min}, \textbf{prob\_max}, and \textbf{entropy} and
+can be obtained using \texttt{tidySEM::table\_fit()}.
 
@@
 
-An important outcome of LCA are conditional item probabilities,
-also known as class-specific item probabilities (Masyn, 2013),
-conditional response, or conditional solution probabilities (Geiser, 2012).
-They indicate the probability of an item being endorsed
-given that the observation belongs to a particular latent class.
-Conditional item probabilities can be obtained using \texttt{tidySEM::table\_prob()}.
-If a particular item is endorsed by two or more classes at markedly different rates,
-it is said to discriminate well between the classes
-and is consequently considered a good indicator.
-Classes are considered highly homogeneous with respect to an item
-when for a particular item
-there is a distinct difference in conditional item probabilities
-for two or more classes.
-For instance, if an item is endorsed below 30\% for one class and
-above 70\% for another class,
-the classes have high homogeneity with respect to this item (Masyn, 2013).
-Conditional item probabilities are the analogue of mean and standard deviation
-when the indicators are binary or ordinal.
-
-A problem which can occur is that of inadmissible solutions.
-With binary indicators, LCA is modelling a cross-table with all the predictors.
-The problem with such cross-tables is that they will often contain empty cells,
-i.e.~combinations of responses that never occur together.
-This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
-Such boundary parameter estimates could indicate that the solution is invalid (Geiser, 2012).
-Boundary parameter estimates can also happen with continuous indicators.
-For instance, if we have a zero-inflated normal distribution and a two class solution,
-one class might have the mean of zero and its standard deviation cannot be determined
-since there is little variance.
-This too could be a sign of an invalid solution,
-warn that too many classes were extracted, or indicate a local optimum (Geiser, 2012).
+An important outcome of LCA are conditional item probabilities, also
+known as class-specific item probabilities (Masyn, 2013),
+conditional response, or conditional solution probabilities
+(Geiser, 2012). They indicate the probability of an item being
+endorsed given that the observation belongs to a particular latent
+class. Conditional item probabilities can be obtained using
+\texttt{tidySEM::table\_prob()}. If a particular item is endorsed by two or more
+classes at markedly different rates, it is said to discriminate well
+between the classes and is consequently considered a good indicator.
+Classes are considered highly homogeneous with respect to an item when
+for a particular item there is a distinct difference in conditional item
+probabilities for two or more classes. For instance, if an item is
+endorsed below 30\% for one class and above 70\% for another class, the
+classes have high homogeneity with respect to this item
+(Masyn, 2013). Conditional item probabilities are the analogue of
+mean and standard deviation when the indicators are binary or ordinal.
+
+A problem which can occur is that of inadmissible solutions. With binary
+indicators, LCA is modelling a cross-table with all the predictors. The
+problem with such cross-tables is that they will often contain empty
+cells, i.e.~combinations of responses that never occur together. This
+problem is reflected by extreme conditional item probabilities (as in
+exactly 0 or 1). Such boundary parameter estimates could indicate that
+the solution is invalid (Geiser, 2012). Boundary parameter
+estimates can also happen with continuous indicators. For instance, if
+we have a zero-inflated normal distribution and a two class solution,
+one class might have the mean of zero and its standard deviation cannot
+be determined since there is little variance. This too could be a sign
+of an invalid solution, warn that too many classes were extracted, or
+indicate a local optimum (Geiser, 2012).
 
@@
 
-The final class solution will usually discover and enumerate several classes.
-The class ordering, however, is completely arbitrary.
-The class labeled as Class 1 in one solution
-may become Class 2 or Class 3 in another model,
-even when the only difference between the models is in their starting values.
-Label switching is something to be mindful of when comparing different LCA models (Masyn, 2013).
-
-The order of clusters is nondeterministic when using K-means in \texttt{tidySEM}.
-Therefore label switching is still a consideration.
-A simple solution to this is setting a random seed number one line prior to fitting the model.
-We advise \texttt{tidySEM} users to always do so in order to circumvent label switching.
-
-Class names should be chosen to accurately reflect group membership.
-Overly simplified and generalized class names may prove misleading
-to both audiences and researches alike
-leading to what is known as a naming fallacy (Weller et al., 2020).
+The final class solution will usually discover and enumerate several
+classes. The class ordering, however, is completely arbitrary. The class
+labeled as Class 1 in one solution may become Class 2 or Class 3 in
+another model, even when the only difference between the models is in
+their starting values. Label switching is something to be mindful of
+when comparing different LCA models (Masyn, 2013).
+
+The order of clusters is nondeterministic when using K-means in
+\texttt{tidySEM}. Therefore label switching is still a consideration. A simple
+solution to this is setting a random seed number one line prior to
+fitting the model. We advise \texttt{tidySEM} users to always do so in order to
+circumvent label switching.
+
+Class names should be chosen to accurately reflect group membership, as theoretically relevant names according to the interpretation of the class solutions.
+Overly simplified and generalized class names may prove misleading to
+both audiences and researches alike leading to what is known as a naming
+fallacy (Weller et al., 2020).
 
@@
 
-Among studies using LCA, reporting practices vary significantly (Weller et al., 2020).
-Various authors have tried to improve and standardize ways of reporting LCA
-(e.g. Masyn, 2013; Weller et al., 2020),
-but more work is needed.
-Van Lissa et al. (2020) developed WORCS,
-a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects
-based on the TOP-guidelines developed by Nosek et al. (2015).
-WORCS workflow can be easily implemented in R in form of
-an R package which facilitates preregistration, article drafting,
-version control, citation and formatting, among others (Van Lissa et al., 2020).
-
-TOP-guidelines emphasise the use of comprehensive citation
-(including referencing the software used in the analysis),
-as well as code and data sharing wherever possible (Nosek et al., 2015).
-Van Lissa et al. (2020) suggest sharing synthetic data in case
-the original data cannot be shared,
-and provide functions to generate such synthetic data.
-Ideally, the entire research project is made reproducible
-so that others may download it and reproduce it with just one click;
-for guidance, see Van Lissa et al. (2020).
-
-As the open science movement is gaining momentum,
-researchers are becoming increasingly aware
-how important it is that analyses can be reproduced and audited.
-In line with open science principles,
-one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that
-user-friendly methods for estimating LCA models
-have predominantly been available in commercial software packages
-(e.g., \emph{Mplus} and \emph{Latent GOLD}).
-A potential downside of commercial software is that
-it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because
-the underlying source code is proprietary.
-To overcome these limitations,
-the present paper introduces new user-friendly functions in
-the \texttt{tidySEM} R-package that can be used to
-estimate a wide range of LCA models
-using the free, open-source R-package \texttt{OpenMx}.
-The reporting guidelines described in this paper
-are adopted in \texttt{tidySEM} by default.
-The \texttt{tidySEM} R-package thus makes advanced mixture modeling
-based on best practices widely accessible,
-and facilitates the adoption of the estimation and
-reporting guidelines described in this paper.
+Among studies using LCA, reporting practices vary significantly
+(Weller et al., 2020). Various authors have tried to improve and
+standardize ways of reporting LCA (e.g. Masyn, 2013; Weller et al., 2020), but more work is needed. Van Lissa et al. (2020)
+developed WORCS, a workflow for open reproducible code in science. WORCS
+consists of step-by-step guidelines for research projects based on the
+TOP-guidelines developed by Nosek et al. (2015). WORCS workflow can be
+easily implemented in R in form of an R package which facilitates
+preregistration, article drafting, version control, citation and
+formatting, among others (Van Lissa et al., 2020).
+
+TOP-guidelines emphasize the use of comprehensive citation (including
+referencing the software used in the analysis), as well as code and data
+sharing wherever possible (Nosek et al., 2015). Van Lissa et al. (2020)
+suggest sharing synthetic data in case the original data cannot be
+shared, and provide functions to generate such synthetic data. Ideally,
+the entire research project is made reproducible so that others may
+download it and reproduce it with just one click; for guidance, see
+Van Lissa et al. (2020).
+
+As the open science movement is gaining momentum, researchers are
+becoming increasingly aware how important it is that analyses can be
+reproduced and audited. In line with open science principles, one of the
+suggested reporting standards relates to reproducible code. In this
+context, it is important to note that user-friendly methods for
+estimating LCA models have predominantly been available in commercial
+software packages (e.g., \emph{Mplus} and \emph{Latent GOLD}). A potential
+downside of commercial software is that it restricts the ability to
+reproduce analyses to license holders, and prevents auditing research
+because the underlying source code is proprietary. To overcome these
+limitations, the present paper introduces new user-friendly functions in
+the \texttt{tidySEM} R-package that can be used to estimate a wide range of LCA
+models using the free, open-source R-package \texttt{OpenMx} as the backend. The reporting
+guidelines described in this paper are adopted in \texttt{tidySEM} by default.
+The \texttt{tidySEM} R-package thus makes advanced mixture modeling based on
+best practices widely accessible, and facilitates the adoption of the
+estimation and reporting guidelines described in this paper.
 
@@
 
-Plots can greatly improve the interpretability of LCA models.
-There are several stages where this is an important consideration.
-
-First, in class enumeration,
-we want to compare several competing class solutions.
-This can be done by means of the AIC and BIC.
-Here, we suggest using an elbow plot
-with the \(K-class\ solution\) on the X-axis,
-and information criteria value on the Y-axis.
-We show an implementation of such an elbow plot
-in our Tutorial subsection.
-
-Second, once we have decided on the final class solution,
-we want to interpret the response patterns
-on the indicators for each latent class.
-While \texttt{tidySEM} can create a table showing
-the probability of each item's response endorsement
-(using \texttt{tidySEM::table\_prob}),
-it may be easier to inspect these probabilities visually.
-For this reason, we created \texttt{ggplot2::plot\_prob()}.
-The resulting shows response patterns
-on all indicators for each group.
-We give an example of this plot in the Tutorial subsection.
+Plots can greatly improve the interpretability of LCA models. There are
+several stages where this is an important consideration.
+
+First, in class enumeration, we want to compare several competing class
+solutions. This can be done by means of the AIC and BIC. Here, we
+suggest using an elbow plot with the \(K-class\ solution\) on the X-axis,
+and information criteria value on the Y-axis. We show an implementation
+of such an elbow plot in our Tutorial subsection.
+
+Second, once we have decided on the final class solution, we want to
+interpret the response patterns on the indicators for each latent class.
+While \texttt{tidySEM} can create a table showing the probability of each
+item's response endorsement (using \texttt{tidySEM::table\_prob}), it may be
+easier to inspect these probabilities visually. For this reason, we
+created \texttt{ggplot2::plot\_prob()}. The resulting shows response patterns on
+all indicators for each group. We give an example of this plot in the
+Tutorial subsection.
 
@@
 
-In this example, we use \emph{data\_mix\_ordinal},
-a simulated data for mixture modelling with ordinal indicators.
-This dataset is built-in to \texttt{tidySEM}.
-For more information about the dataset,
-type \texttt{?tidySEM::data\_mix\_ordinal} into the R console
-after loading the \texttt{tidySEM} using library().
+In this example, we use \emph{data\_mix\_ordinal}, a simulated data for mixture
+modelling with ordinal indicators. This dataset is built-in to
+\texttt{tidySEM}. For more information about the dataset, type
+\texttt{?tidySEM::data\_mix\_ordinal} into the R console after loading the
+\texttt{tidySEM} using library().
 
@@
 
-An important step to preceding any statistical analysis is data exploration.
-Here we use \texttt{tidySEM::descriptives()}
-to describe the dataset we are using.
+An important step to preceding any statistical analysis is data
+exploration. Here we use \texttt{tidySEM::descriptives()} to describe the
+dataset we are using.
 
 As we can see, the output includes various descriptives of our dataset.
-Special attention should be paid to examining the pattern of missingness,
-as discussed in the \emph{Handling missing data} section of this paper.
-In our example, we see that there are no missing values,
-hence we proceed with our analysis.
+Special attention should be paid to examining the pattern of
+missingness, as discussed in the \emph{Handling missing data} section of this
+paper. In our example, we see that there are no missing values, hence we
+proceed with our analysis.
 
@@
 
-Before we fit a series of LCA models, we set a random seed
-using \texttt{set.seed()}.
-The seed number can be any digit.
-This is an important step as
-there is some inherent randomness in the LCA computations,
-and having the same seed number ensures that
-two separate researcher obtain exactly the same results
-when fitting LCA models.
-
-Finally, we reach the step of fitting LCA models.
-To do so, we use \texttt{tidySEM::mx\_lca()}
-which takes data and number of classes as inputs.
-In our example, we want to fit 1 to 4 class solutions
-and compare their output.
-Depending on your computer's computational power,
-this might take a while.
+Before we fit a series of LCA models, we set a random seed using
+\texttt{set.seed()}. The seed number can be any digit. This is an important
+step as there is some inherent randomness in the LCA computations, and
+having the same seed number ensures that two separate researcher obtain
+exactly the same results when fitting LCA models.
+
+Finally, we reach the step of fitting LCA models. To do so, we use
+\texttt{tidySEM::mx\_lca()} which takes data and number of classes as inputs.
+Notice that the analysis will be done on all the variable in the data set, so you have to make sure it only includes the relevant variables.
+In
+our example, we want to fit 1 to 4 class solutions and compare their
+output. Depending on your computer's computational power, this might
+take a while.
+
+Notice that the model was run 10 times for each \(K\) class solution.
+This because the functions runs a total of 10 \textbf{extra} times the model.
+This way we can choose the best fitting model across all the solutions.
+In this examples we see that all models converged without issues.
+
+As \texttt{tidySEM::mx\_lca()} was run for several models at the same time, the resulting object \texttt{res} is a \texttt{mixture\_list} class where each \(K\) class solution is an object in the list, and each of them are \texttt{OpenMx} models (\texttt{MxModel}).
 
@@
 
-In class enumeration,
-we want to compare a sequence of LCA models fitted to the data.
-To aid the process, we create a model fit table using \texttt{tidySEM::table\_fit()}
-with the results object as the input.
-As the output contains a lot of information on each of the four fitted models,
-we select a subset of helpful model fit indices and classification diagnostics.
+In class enumeration, we want to compare a sequence of LCA models fitted
+to the data. To aid the process, we create a model fit table using
+\texttt{tidySEM::table\_fit()} with the results object as the input. As the
+output contains a lot of information on each of the four fitted models,
+we select a subset of helpful model fit indices and classification
+diagnostics.
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1220}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8780}}@{}}
-\caption{\label{tab:unnamed-chunk-6}Selection of Fit Indices and Classification Diagnostics}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1325}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8675}}@{}}
+\caption{\label{tab:unnamed-chunk-7}Selection of Fit Indices and Classification Diagnostics}\tabularnewline
 \toprule()
@@
 LL & the -2*log-likelihood of each model \\
+Parameters & Number of estimated parameters in each model \\
 AIC & the Akaike Information Criterion value \\
@@
 
-We discussed several possible strategies to select the final class solution.
-Here, we apply our own.
-
-To aid our interpretation of the results,
-we create an elbow plot showing the trends in information criteria across four models.
-
-From the elbow plot, we see that AIC has a lower penalty for model complexity.
-However, we are more interested in the BIC values, which are similar
-for the one, two and three-class solutions, but the four-class solution fits significantly worse.
-For this reason, we eliminate the four-class solution from the selection process.
-
-Then we examine the model fit table.
-As expected, the -2*log likelihood falls successively with each added class.
-As previously stated, classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions because they are uninterpretable.
-We see that prob\_min for the four-class solution is low,
-knowing that this solution also has a high BIC, we disqualify this solution.
-
-Out of the remaining three solutions,
-we notice that entropy is the highest for the three-class solution,
-and it has a satisfactory prob\_min and n\_min.
-Based on this, we retain the three class solution in model selection.
-Note that entropy for the one-class solution will always equal to one,
-as it is 100\% true that every case is in that class.
-Based on the low entropy of the two-class solution,
-we eliminate this model.
-
-Finally, when comparing the one and three-class solutions,
-we inspect the information criteria.
-For BIC, the one-class solution fits better, but the difference is marginal.
-AIC tells us that the added complexity
-of having three classes still explains the data better
-than a one-class solution.
-Therefore, we select the three-class solution as our final-class solution.
+We discussed several possible strategies to select the final class
+solution. Here, we apply our own.
+
+To aid our interpretation of the results, we create an elbow plot
+showing the trends in information criteria across four models.
+
+From the elbow plot, we see that AIC has a lower penalty for model
+complexity. However, we are more interested in the BIC values, which are
+similar for the one, two and three-class solutions, but the four-class
+solution fits significantly worse. For this reason, we eliminate the
+four-class solution from the selection process.
+
+Then we examine the model fit table. As expected, the -2*log likelihood
+falls successively with each added class. As previously stated,
+classification diagnostics should not be used for model selection, but
+they can be used to disqualify certain solutions because they are
+uninterpretable. We see that prob\_min and n\_min for the four-class solution is
+low, knowing that this solution also has a high BIC, we disqualify this
+solution.
+
+Out of the remaining three solutions, we notice that entropy is the
+highest for the three-class solution, and it has a satisfactory prob\_min
+and n\_min. Based on this, we retain the three class solution in model
+selection. Note that entropy for the one-class solution will always
+equal to one, as it is 100\% true that every case is in that class. Based
+on the low entropy of the two-class solution, we eliminate this model.
+
+Finally, when comparing the one and three-class solutions, we inspect
+the information criteria. For BIC, the one-class solution fits better,
+but the difference is marginal. AIC tells us that the added complexity
+of having three classes still explains the data better than a one-class
+solution. Therefore, we select the three-class solution as our
+final-class solution.
 
@@
 
-To aid our understanding of the final class solution,
-we use \texttt{ggplot2::plot\_prob()}
-with the results of the three-class model as the input.
-The resulting graph shows response patterns
-on all the indicators for each group.
+To aid our understanding of the final class solution, we use
+\texttt{ggplot2::plot\_prob()} with the results of the three-class model as the
+input. The resulting graph shows response patterns on all the indicators
+for each group.
 
 If we want to know the probability of each response option's endorsement
-for each class, we can use \texttt{tidySEM::table\_prob}.
-These are thresholds for ordinal dependent variables in the probability scale.
-
-In the plot, we can see the distributions of the response probabilities on the indicators
-for each of the three classes.
-For instance, we see that in Class 1 the most common response to u2 is 2,
-while in Class 2 and Class 3 this is 0.
-We can also see that response 1 is a rare response not forming the majority in
-any class.
-Class 2 distinguishes itself because the majority scores the response 0 category of u3 and u4,
-while in Class 1 and 2 this is not the case.
-Class 3 distinguishes itself because the most common response to u3 and u4 is 3.
+for each class, we can use \texttt{tidySEM::table\_prob}. These are thresholds
+for ordinal dependent variables in the probability scale.
+
+In the plot, we can see the distributions of the response probabilities
+on the indicators for each of the three classes. For instance, we see
+that in Class 1 the most common response to u2 is 2, while in Class 2
+and Class 3 this is 0. We can also see that response 1 is a rare
+response not forming the majority in any class. Class 2 distinguishes
+itself because the majority scores the response 0 category of u3 and u4,
+while in Class 1 and 2 this is not the case. Class 3 distinguishes
+itself because the most common response to u3 and u4 is 3.
 
 We can also interpret the response patterns numerically.
+It is a matter of preferences on how to interpret these probabilities.
+Here is where you would \textbf{name} each class, such that each response pattern is theoretically meaningful.
 
@@
 
-Another step is to extract posterior class probabilities.
-This is done by the use of \texttt{TidySEM::class\_prob}
-with the results of the final class solution as the input.
+Another step is to extract posterior class probabilities. This is done
+by the use of \texttt{tidySEM::class\_prob} with the results of the final class
+solution as the input.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-collins_latent_2010}{}}%
+Collins, L.
+M., \& Lanza, S.
+T.
+(2010).
+\emph{Latent {Class} and {Latent} {Transition} {Analysis}: {With} {Applications} in the {Social}, {Behavioral}, and {Health} {Science}}.
+New Jersey, USA: John Wiley \& Sons, Inc.
+
+\leavevmode\vadjust pre{\hypertarget{ref-enders_applied_2010}{}}%
+Enders, C. K. (2010). \emph{Applied missing data analysis}. New York: Guilford Press.
+
 \leavevmode\vadjust pre{\hypertarget{ref-figueiredo_unsupervised_2002}{}}%
@@
 \url{https://doi.org/10.1002/ab.20233}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hardy_exploring_2018}{}}%
+Hardy, T.
+M., Garnier-Villarreal, M., McCarthy, D.
+O., Anderson, R.
+A., \& Reynolds, R.
+M.
+(2018).
+Exploring the {Ovarian} {Reserve} {Within} {Health} {Parameters}: {A} {Latent} {Class} {Analysis}.
+\emph{Western Journal of Nursing Research}, 1--16.
+\url{https://doi.org/10.1177/0193945918792303}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-mcelreath_statistical_2020}{}}%
+McElreath, R. (2020). \emph{Statistical rethinking: A {Bayesian} course with examples in {R} and {Stan}} (2nd ed.). Boca Raton: Taylor; Francis, CRC Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-openmx}{}}%
+Neale, M.
+C., Hunter, M.
+D., Pritikin, J.
+N., Zahery, M., Brick, T.
+R., Kirkpatrick, R.
+M., \ldots{} Boker, S.
+M.
+(2016).
+Open{M}x 2.0: {E}xtended structural equation and statistical modeling.
+\emph{Psychometrika}, \emph{81}(2), 535--549.
+\url{https://doi.org/10.1007/s11336-014-9435-8}
+
 \leavevmode\vadjust pre{\hypertarget{ref-norman_likert_2010}{}}%
@@
 \url{https://doi.org/10.1037/tps0000176}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pavlopoulos_measuring_2015}{}}%
+Pavlopoulos, D., \& Vermunt, J.
+K.
+(2015).
+Measuring {Temporary} {Employment}.
+{Do} {Survey} or {Register} {Data} {Tell} the {Truth}?
+\emph{Survey Methodology}, \emph{41}(1), 197--214.
+Retrieved from \url{https://www150.statcan.gc.ca/n1/pub/12-001-x/2015001/article/14151-eng.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rhemtulla_when_2012}{}}%
+Rhemtulla, M., Brosseau-Liard, P.
+É., \& Savalei, V.
+(2012).
+When can categorical variables be treated as continuous?
+{A} comparison of robust continuous and categorical {SEM} estimation methods under suboptimal conditions.
+\emph{Psychological Methods}, \emph{17}(3), 354--373.
+\url{https://doi.org/10.1037/a0029315}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lavaan}{}}%
+Rosseel, Y. (2012). {lavaan}: An {R} package for structural equation modeling. \emph{Journal of Statistical Software}, \emph{48}(2), 1--36. \url{https://doi.org/10.18637/jss.v048.i02}
 

### 2022-11-13 manuscript.Rmd
Update citations & Minor Adjustments

@@
     email         : "m.garniervillarreal@vu.nl"
+  - name          : "Daniel Anadria"
+    affiliation   : "1"
+    corresponding : no
+    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
+    email         : "d.anadria@uu.nl"
 affiliation:
@@
 most notably finite Gaussian mixture modeling and latent profile
-analysis. @vermunt_jk_latent_2004 defined LCA more generally as
+analysis. @vermunt_latent_2004 defined LCA more generally as
 virtually any statistical model where "some of the parameters [...]
@@
 In factor analysis, the latent variable is continuous, so the relation
-between indicators is define by the communalities between them, and the
+between indicators is defined by the communalities between them, and the
 latent variable represents the error corrected approximation of the
@@
 the term was initially conceived to refer to analyses with categorical,
-usually binary indicators [@vermunt_jk_latent_2004]. Nowadays, there are
+usually binary indicators [@vermunt_latent_2004]. Nowadays, there are
 a number of related techniques, known by distinct names, that serve a
@@
 can be standardized to obtain correlations). These parameters can be
-estimated freely, or set to be constrained across classes [@collins_latent_2010].
+estimated freely, or set to be constrained across classes [@collins_latent_2009].
 
@@
 will be inappropriate; for instance when the cohesion between indicators
-is expected to differ between classes [@collins_latent_2010]. As an example, a LPA of ocean
+is expected to differ between classes [@collins_latent_2009]. As an example, a LPA of ocean
 plastic particles found two classes of particles based on length and
@@
 The term latent class analysis originally referred to cases where the
-observed indicators were categorical [@collins_latent_2010]. 
+observed indicators were categorical [@collins_latent_2009]. 
 Nowadays, it is more commonly used
@@
 [@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
-been distinguished in the literature [@rubin_inference_1976; @enders_applied_2010]: Missing
+been distinguished in the literature [@rubin_inference_1976; @enders_applied_2022]: Missing
 completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
@@
 
-Conversely, the free open source program OpenMx [@openmx] is very flexible, but
+Conversely, the free open source program OpenMx [@neale_openmx_2016] is very flexible, but
 not very user-friendly. We directly address this limitation using the
@@
 user interface is simple, making use of the model syntax of the widely
-used `lavaan` R-package [@lavaan]. This syntax offers a human-readable way to
+used `lavaan` R-package [@rosseel_lavaan_2012]. This syntax offers a human-readable way to
 specify latent variable models. Minor enhancements are made to simplify

### 2022-11-22 example_analysis/LCA_Tutorial.Rmd
Update November 23
- updated the literature list
- added use case examples
- corrected spelling/grammar mistakes
- adjusted spacing, also in the tutorial

*** things to resolve:
1) Caspar should add Well-being compass study to Zotero and update the in-text citation for that use case
2) the plot_density function doesn't work, also the example in the documentation;
I would like to show how to use it in class enumeration section of the tutorial
3) final proof-reading, anything to add?

--- 2022-10-16 manuscript.Rmd
@@
 ---
-title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
-shorttitle        : "BEST PRACTICES LCA"
-
-author: 
-  - name          : "Caspar J. van Lissa"
-    affiliation   : "1,2"
-    corresponding : yes    # Define only one corresponding author
-    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
-    email         : "c.j.vanlissa@gmail.com"
-
-affiliation:
-  - id            : "1"
-    institution   : "Utrecht University, Methodology & Statistics"
-  - id            : "2"
-    institution   : "Open Science Community Utrecht"
-
-author_note: |
-  Methodology & Statistics
-  Padualaan 14
-  3582CH Utrecht, The Netherlands
-
-abstract: |
-  Latent class analysis is a family of techniques for identifying groups in data based
-  on a parametric model. Its examples include mixture models, latent profile
-  analysis, latent class analysis with ordinal indicators, and latent class
-  growth analysis. Despite the popularity of this technique, there is limited
-  guidance with respect to best practices in estimating and reporting mixture
-  models. Moreover, although user-friendly interfaces for advanced mixture
-  modeling have long been available in commercial software packages, open
-  source alternatives have remained somewhat inaccessible. This tutorial
-  describes best practices for the estimation and reporting of latent class analysis,
-  using free and open source software in R. To this end, this tutorial
-  introduces new functionality for estimating and reporting mixture
-  models in the `tidySEM` R-package whose backend relies on
-  the `OpenMx` R-package.
-
-keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
-wordcount         : "8206"
-bibliography      : "mixture.bib"
-figsintext        : yes
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-lineno            : no
-mask              : no
-
-class             : "man"
-output            : papaja::apa6_pdf
-
+title: "LCA in R using tidySEM"
+output: html_notebook
 ---
-
-```{r load_packages, include = FALSE}
-library("papaja")
-#bibliography      : ["r-references.bib"]
-library(worcs)
-run_everything = FALSE
-```
-
-Latent class analysis (LCA) is an umbrella term 
-that refers to a number of techniques for estimating unobserved group membership
-based on a parametric model of one or more observed indicators of group membership.
-The types of LCA have become quite popular across scientific fields,
-most notably finite Gaussian mixture modeling and latent profile analysis.
-@vermunt_jk_latent_2004 defined LCA more generally as virtually any statistical model 
-where “some of the parameters […] differ across unobserved subgroups”.
-
-Despite its popularity, there is a lack of standards for estimating and reporting LCA.
-While @van_de_schoot_grolts-checklist_2017 developed reporting guidelines 
-for a specific type of LCA known as latent growth models,
-general reporting guidelines for all LCA are still lacking.
-This complicates manuscript review and assessment of the quality of published studies,
-and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature 
-by suggesting updated guidelines for estimation and reporting of LCA, 
-based on current best practices.
-Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples, 
-this paper exclusively relies on free, open source software in R.
-Our goal is to make best-practices in LCA widely accessible.
-
-## Defining Latent Class Analysis
-
-Latent class analysis is a group of methods for estimating unobserved groups
-based on a parametric model of observed indicators of group membership.
-The concept of LCA can be understood in different ways.
-A mixture model assumes that the study population
-is composed of $K$ subpopulations or classes.
-It further assumes that the observed data are a mixture of data generated by class-specific models.
-The simplest univariate model is a normal distribution, 
-which can be described with two parameters: the mean and the variance.
-Commonly, the same model is estimated across all classes, 
-but with different parameters for each class (i.e., class-specific means and variances).
-Mixture modeling then estimates both the parameters for each class, 
-and the probability that an individual belongs to each class.
-
-As an illustrative example, imagine that a detective wants to know 
-if it would be possible to use mixture modeling to identify the sex of a suspect,
-based on footprints found at the crime scene.
-To test the feasibility of this approach,
-the detective records the shoe sizes and sex of 100 volunteers.
-The resulting observed data look like this:
-
-```{r echo = FALSE}
-library(tidySEM)
-library(ggplot2)
-set.seed(1)
-n = 100
-C <- sample(c("Man", "Woman"), n, replace = TRUE)
-means <- c(Man = 9, Woman = 7)
-X <- rnorm(n, mean = means[C], sd = 1)
-X <- round(X*2)/2
-obsparam <- tapply(X, C, mean)
-shoesize_shapiro <- shapiro.test(X) # Shapiro-Wilk test of normality
-```
-
-```{r echo = FALSE, eval = run_everything}
-est <- mx_profiles(data.frame(X), 2)
-estparam <- table_results(est, c("label", "est", "matrix"))
-estparam <- as.numeric(estparam$est[estparam$matrix == "M"])
-props <- class_prob(est, "individual")$individual[ , "predicted"]
-dput(props, "props.txt")
-dput(estparam, "estparam.txt")
-
-p <- ggplot(data.frame(Shoesize = X), aes(x = Shoesize)) + geom_density() + theme_bw()
-ggsave("shoedens.png", p, device = "png")
-```
-
-```{r shoedens, echo = FALSE, fig.cap="Kernel density plot of shoe sizes."}
-props <- eval(parse("props.txt"))
-estparam <- eval(parse("estparam.txt"))
-knitr::include_graphics("shoedens.png")
-```
-
-The distribution is evidently bimodal, which bodes well for the intended mixture model.
-In this case, the number of classes is known a-priori.
-When estimating a two-class mixture model, the detective observes that the model estimates
-the mean shoe size of the two groups are equal to `r report(estparam[1], equals = FALSE)` 
-and  `r report(estparam[2], equals = FALSE)`,
-which is close to the true means of the two groups, 
-namely `r report(obsparam[1], equals = FALSE)` and  `r report(obsparam[2], equals = FALSE)`.
-When tabulating estimated group membership against observed (known) group membership, 
-it can be seen that women are classified with a high degree of accuracy, but men are not:
-
-```{r tabshoe, results="asis"}
-library(kableExtra)
-tab <- table(C, props)
-tab <- data.frame(Observed = rownames(tab), as.data.frame.array(tab))
-names(tab)[2:3] <- c("Class 1", "Class 2")
-rownames(tab) <- NULL
-kbl(tab, caption = "Observed group membership by estimated class membership.")
-```
-
-A LCA is like confirmatory factor analysis, 
-except that the continuous latent variable is substituted with a categorical latent variable.
-One difference between the two techniques is that 
-factor analysis can be considered as a way to group observed *variables* into latent constructs, 
-and LCA groups *individuals* into classes.
-In line with this distinction, LCA is sometimes referred to as a "person-centered" technique, 
-and factor analysis as a "variable-centered" technique [@nylund-gibson_ten_2018].
-
-When the focus is on the model parameters in each group,
-LCA can be thought of as similar to a multi-group structural equation model.
-The main distinction is that group membership is not known a-priori,
-but is instead estimated -- with measurement error -- based on the data.
-Whereas in a multi-group model, the data are split by group and treated as independent samples,
-in LCA, all cases contribute to the estimation of all parameters in all groups.
-The relative contribution of each case to the parameters of each group
-is determined by that case's posterior probability of belonging to that group.
-
-When the focus is on each individual's estimated class membership,
-LCA can be thought of as a type of clustering algorithm.
-In line with this perspective, LCA is sometimes described as "model-based
-clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
-Many clustering algorithms apply some recursive splitting algorithm to the data.
-By contrast "model-based" clustering refers to the fact that LCA 
-estimates cluster membership based on a parametric model.
-Specifically, the posterior class probability that an individual belongs to a latent class
-can be computed from the likelihood of that individual's observed data under given the class-specific model.
-
-Finally, in the context of machine learning, LCA can be considered as an
-*unsupervised classification* problem [@figueiredo_unsupervised_2002].
-The term *unsupervised* refers to the fact that the outcome variable 
--- true class membership -- is not known, 
-and the term *classification* refers to the fact that 
-the algorithm is predicting a categorical outcome -- class membership.
-
-## A Taxonomy of Latent Class Analysis Methods
-
-In this paper, we use the term latent class analysis to refer to 
-a family of techniques that estimate latent class membership 
-based on a parametric model of observed indicators.
-From a historical perspective, the term was initially conceived to 
-refer to analyses with categorical, usually binary indicators [@vermunt_jk_latent_2004].
-Nowadays, there are a number of related techniques, 
-known by distinct names, that serve a similar purpose.
-The term "latent class analysis" seems most appropriate 
-as an umbrella term for this broader class of models,
-as it only refers to the purpose of the analysis, 
-and does not imply restrictions to the model used, 
-or the level of measurement of the indicators.
-Given the abundance of terms in use for closely related classes of models, 
-we will provide a rudimentary taxonomy of LCA methods.
-
-A common type of LCA is the *finite Gaussian mixture model*; 
-a univariate analysis where the observed distribution of a single variable 
-is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
-The parameters of a finite Gaussian mixture model are 
-the means and variances of these underlying normal distributions.
-The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
-In the multivariate case, with more than one indicator variable,
-the parameters of a mixture model are the means, variances, and covariances 
-between the indicators (which can be standardized to obtain correlations).
-These parameters can be estimated freely, or set to be constrained across classes.
-
-The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
-which assumes conditional independence of the indicators.
-Conditional independence means that, after class membership is accounted for,
-the covariances/correlations between indicators are assumed to be zero.
-This can be conceived of as a restricted mixture model with covariances fixed to zero.
-In some cases, such constraints will be inappropriate;
-for instance when the cohesion between indicators is expected to differ between classes.
-As an example, a LPA of ocean plastic particles 
-found two classes of particles based on length and width:
-a class of smaller particles with a high correlation between length and width, 
-meaning that these particles were approximately round or square in shape, 
-and a class of larger particles with a low correlation between length and width, 
-meaning that these particles were heterogenous in shape.
-From a theoretical perspective, this makes sense, 
-because the smaller particles have been ground down to a more uniform shape by the elements.
-<!-- DA: what is the source for plastic particles? -->
-
-It is also possible to estimate a mixture model based on latent indicators.
-This means that, within each class, 
-one or more continuous latent variables are estimated based on the observed indicators.
-Categorical latent variable membership is then estimated 
-based on these continuous latent variables.
-A common application of this approach is in longitudinal research, 
-where the indicators reflect one construct assessed at different time points.
-Examples of this approach include *growth mixture models* (GMM) 
-and *latent class growth analyses* (LCGA).
-These techniques estimate a latent growth model
-to describe individual trajectories over time.
-The growth mixture model is a latent class model 
-where the parameters that indicate class membership 
-are the intercepts and variances, 
-and typically covariances of the latent growth variables, 
-e.g., a latent intercept and slope.
-This technique assumes that individuals within a class can have heterogenous trajectories.
-If the variance of the growth parameters is fixed to zero, 
-it is known as a latent class growth analysis.
-This latter approach assumes that all individuals within a class 
-share the same identical trajectory,
-and that any variance in the indicators not explained by the class-specific latent trajectories 
-is due to residual error variance.
-
-The term latent class analysis originally referred to cases 
-where the observed indicators were categorical.
-Nowadays, it is more commonly used as an umbrella term.
-To prevent ambiguity, the special case 
-where indicators are of binary or ordinal measurement level 
-might be described as *latent class analysis with ordinal indicators*.
-Latent class models with ordinal indicators are parameterized differently from mixture models.
-One common parameterization assumes that 
-each categorical variable reflects an underlying standard normal distribution.
-The parameters are "thresholds" that correspond to quantiles of a standard normal distribution 
-(with $N(\mu = 0, \sigma = 1)$).
-These thresholds are estimated based on 
-the proportion of individuals in each of the response categories of the indicator variable.
-For example, a binary indicator has a single threshold that distinguishes the two response categories.
-If responses are distributed 50/50, then the corresponding threshold would be  $t_1 = 0.00$.
-If the responses are distributed 60/40, then the resulting threshold would be $t_1 = 0.25$.
-
-This paper will primarily focus on mixture models and special cases thereof,
-although most of the suggested guidelines are applicable to all LCA methods.
-
-## Use Cases for Latent Class Analysis
-
-<!-- Caspar will check if I can replace them with examples from Caspar's published papers. -->
-
-There are several use cases for which LCA methods are suitable.
-One example is to test a theory that postulates the existence of a categorical latent variable.
-For example, *identity status theory* posits that, at any given point in time, 
-adolescents reside in one of four identity statuses.
-Latent class analysis can be used to identify these four statuses 
-based on observed indicators (e.g., self-reported identity exploration and commitment).
-If results indicate that the data are better described by a different number of classes,
-or that the four-class solution does not correspond to the predicted pattern of responses on the indicators, 
-then the theory may be called into question.
-
-Another use case is unsupervised learning:
-when the goal is to restore unobserved class membership based on observed indicators,
-or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid 
-when several clinical indicators can be used to distinguish between 
-a fixed number of physical [@baughman_mixture_2006] or mental [@wu_abuse_2011] health problems.
-The example of shoe size is a rudimentary illustration of this type of application.
-
-LCA can be used as a descriptive analysis 
-where a researcher wishes to describe their sample 
-and identify a few prototypes based on many variables.
-As an example, if a survey among all Dutch academics was carried out
-with the goal of bringing about a funding reform, 
-LCA could be used to discover the types of publications that get funding.
-
-With LCA, our goal could be to inductively identify the number of classes. 
-For instance, if we believe that a variable represents a group, 
-but don't know how many groups there are, 
-LCA may be an appropriate technique to answer this question.
-For example, @hopfer_social_2014 studied substance use, sexual behavior, 
-and mental health status of urban population in Winnipeg, Canada. 
-The underlying assumption was that there were different risk profiles,
-but their number was not known.
-From a collection of indicators, LCA provided evidence 
-that there are four distinct risk profiles in the Winnipeg area.
-
-Another application of LCA is to classify individuals.
-In a peer harassment study, 
-@giang_using_2008 used LCA to classify over 2,000 sixth grade students 
-into aggressor and victim latent classes.
-The five-class solution comprised of classes of 
-victims, aggressors, and socially adjusted students. 
-For instance, it revealed that there were two types of victims: 
-highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
-
-LCA is also appropriate when we wish to identify indicators that capture classes well.
-High quality indicators will be strongly related to the latent variable
-and will lead to good class separation.
-This relationship of high quality indicators to the latent variable
-is reflected in very high or very low conditional response probabilities.
-For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
-Therefore, one could use conditional response probabilities for each item to assess its quality
-with regards to how well it helps separate the latent classes.
-From this, a theory about the selection of indicators could be informed.
-
-An extension of LCA is that containing covariates which can be used to predict class membership.
-In this approach, we not only model the latent class variable based on indicators,
-but we also relate the class membership to other explanatory variables (@vermunt_latent_2017).
-An example of using covariates comes from @nozadi_moderating_2016 
-who applied LCA to identify the probability of children's membership to an anxiety class.
-The authors tested several covariates including children's age, sex, and accuracy scores.
-Age and sex were not found to be related to the children's latent class membership,
-hence these covariates were excluded from the analysis.
-Accuracy scores, however, were related to probabilities of 
-being in anxiety and attention-anxiety classes. 
-Therefore this covariate was kept as a valuable predictor.
-
-When our interest is the prediction of one or more outcomes,
-LCA can be used to construct latent classes as categorical predictors.
-@lanza_latent_2013 demonstrated how LCA can be used 
-to classify adolescents into depression classes,
-and subsequently these classes can be used to predict smoking, grades, and delinquency.
-The study showed that the outcomes predicted by class membership 
-can be binary (regular smoking), continuous (grades) or count (delinquency).
-
-In addition to these applications, LCA can be used for dimensionality reduction
-as the resulting groups summarize response patterns on a large number of indicators.
-For example, @macgregor_symptom_2021 investigated symptom profiles 
-among injured U.S. military personnel. 
-They used fifteen dichotomous items 
-from the Post-Deployment Health Assessment survey as LCA indicators.
-Combinatorics informs us that fifteen dichotomous items have 
-$2^{15}$ or $32,768$ unique symptom combinations.
-Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA 
-as a method of dimensionality reduction.
-A five class solution was found to have the best fit 
-according to both statistical criteria and clinical interpretability.
-
-<!-- CJ: (on the next paragraph)  -->
-<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
-<!-- That is not what this paragraph is about.  -->
-<!-- This paragraph is about motivations for using LCA. -->
-<!-- One motivation is if your data violates multivariate normality -->
-<!-- because violating multivariate normality means that you have multiple groups -->
-<!-- and you can use LCA to extract those. That is the only assumption  -->
-<!-- we have to discuss here, we just need to explain it. -->
-Finally, LCA can be used to deal with data which violate certain assumptions.
-As discussed in the shoe size example, LCA can deal with violations of normality.
-In fact, LCA assumes the population distribution is a non-normal mixture 
-of $K$ normal distributions,
-and it can discover the value of $K$, i.e. generate a $K$-class solution.
-
-# Best Practices
-
-## Best Practices in Estimation
-
-The best practices in estimation, as outlined in Table \@ref(tab:checkest),
-are rooted in existing recommendations for best practices 
-for estimating specific subtypes of LCA,
-including latent class growth analysis [@van_de_schoot_grolts-checklist_2017]
-and latent class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
-These were generalized to be relevant to all types of LCA,
-and updated to current best practices, as explained below.
-
-```{r checkest, include=TRUE}
-tab_check <- data.frame(Item = c(
-  "Examining Observed Data",
-  "Handling Missing Data",
-  "Alternative Model Specifications",
-  "Software",
-  "Algorithm",
-  "Class Enumeration",
-  "Model Fit Indices",
-  "Classification Diagnostics",
-  "Interpreting Class Solutions",
-  "Label switching"
-
-))
-tab_check$`#` = paste0(1:length(tab_check$Item), ".")
-kbl(tab_check[, c("#", "Item")])
-```
-
-### Examining Observed Data
-
-Examining observed data is essential for any analysis
-as it may reveal patterns and violations of assumptions 
-that had not been considered prior to data collection.
-Special attention should be paid 
-to level of measurement of the indicators.
-Finite Gaussian mixture models (including LPA) 
-are only suitable for continuous variables.
-Indicators with an ordinal level of measurement 
-are likely to violate the assumption of 
-within-class normal distributions of mixture models
-[see @vermunt_k-means_2011].
-Personal experience consulting on LCA methods and 
-moderating the `tidyLPA` Google group
-suggest that the misapplication of mixture models 
-to ordinal (e.g., Likert-type) indicators 
-is the most common source of user error.
-Whereas it has been argued that some parametric methods are robust 
-when scales with 7+ indicators are treated as continuous 
-[e.g., @norman_likert_2010],
-this certainly does not imply that all methods are.
-It is certainly unlikely that such ordinal variables 
-can be treated as a *mixture* of multiple normal distributions.
-The problem becomes egregious when 
-the number of classes estimated equals or exceeds the number of categories; 
-in this case, each class-specific mean could describe a single response category, 
-and a class-specific variance component would be nonsensical.
-In sum, Likert-type scales are rarely suitable for mixture modeling;
-latent class analysis with ordinal indicators is more appropriate.
-
-Relatedly, a recent publication claimed that 
-an assumption of mixture models is that 
-observed indicators are normally distributed [@spurk_latent_2020].
-This is incorrect.
-When the number of classes is greater than one, 
-mixture models assume that the observed indicators 
-are a mixture of multiple (multivariate) normal distributions.
-In our shoe size example, it can be seen that the population distribution
-is comprised of two normal distributions.
-When examined visually, the population distribution is evidently bimodal. 
-The Shapiro-Wilk normality test ($W = 0.971$, $p < 0.05$ ) 
-rejects the null hypothesis that 
-the sample comes from a normally distributed population.
-Yet, this is a prototypical example of a mixed population distribution 
-where LCA can discover latent groups.
-If the population distribution were instead normal, 
-there would be no classes to extract 
-as the whole population would belong to a single class.
-
-Extensive descriptive statistics 
-(including the number of unique values, 
-variance of categorical variables, and missingness; see next subsection)
-can be obtained using the function `tidySEM::descriptives(data)`.
-Note, however, that sample-level descriptive statistics 
-are of limited value when the goal of a study is 
-to identify subsamples using latent class analysis.
-Plots (density plots for continuous variables, 
-and bar charts for categorical ones) may be more diagnostic.
-Note that density plots can also aid in the choice of the number of classes, 
-as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements, 
-provided that these are readily accessible 
-[consider using a GitHub repository as a comprehensive public research archive, 
-as explained in @van_lissa_worcs_2021].
-
-### Handling Missing Data
-
-Previous work has emphasized the importance of examining 
-the pattern of missing data and 
-reporting how missingness was handled 
-[@van_de_schoot_grolts-checklist_2017].
-Three types of missingness have been distinguished
-in the literature [@rubin_inference_1976]:
-Missing completely at random (MCAR), 
-which means that missingness is random;
-missing at random (MAR), 
-which means that missingness is contingent on the *observed* data
-(and can thus be accounted for);
-and finally missing not at random (MNAR), 
-which means that missingness is related to unobserved factors.
-It is possible to conduct a so-called "MCAR" test, 
-for example the non-parametric MCAR test [@jamshidian_tests_2010].
-But note that the name "MCAR test" is somewhat misleading,
-as the null-hypothesis of this test is that the data are not MAR,
-and a significant test statistic indicates that 
-missingness is related to the observed data (MAR).
-A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on 
-the comparison of variances across groups 
-with different patterns of missing data, 
-it assumes normality [@little_test_1988].
-This assumption is tenuous in the context of LCA.
-A non-parametric MCAR test, 
-as provided by Jamshidian and Jalal, 
-may be more suitable [@jamshidian_tests_2010].
-Unfortunately, this test was removed from the central R-repository CRAN 
-due to lack of maintenance.
-For this tutorial, I have re-implemented it in the `mice` package 
-as `mice::mcar()`,
-with a fast backend in C++ and new printing and plotting methods.
-
-While we concur that investigating missingness is due diligence,
-it is important to emphasize that 
-missingness is adequately handled by default in many software packages for LCA, 
-such as Mplus, and `OpenMx` which is the backend of `tidySEM`.
-These packages use Full Information Maximum Likelihood (FIML) estimation,
-which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data; 
-on par with multiple imputation [@lee_comparison_2021].
-FIML estimation assumes that missingness is either MCAR or MAR.
-Thus, one would typically proceed with FIML, 
-regardless of the outcome of an MCAR test.
-Although FIML does not, by default, handle missingness in exogenous variables
--- all indicator variables in LCA are endogenous, 
-so this is not a concern.
-
-Multiple imputation is less suitable to LCA for two reasons.
-First, because LCA methods are often computationally expensive,
-and conducting them on multiple imputed datasets may be unfeasible.
-Second, because there is no straightforward way 
-to integrate LCA results across multiple datasets.
-To conclude; our recommendation is to inspect missingness 
-(e.g., using `mice::MCAR()`)
-and report the proportion of missingness per variable 
-(e.g., using `tidySEM::descriptives()`),
-before proceeding with FIML.
-One minor concern is that the K-means algorithm,
-which `tidySEM` uses for determining starting values,
-is *not* robust to missing values.
-When it fails, `tidySEM` automatically switches to hierarchical clustering,
-unless the user specifies a different
-clustering algorithm or uses manual starting values.
-
-### Alternative Model Specifications
-
-In order to aid researchers working with latent trajectory models,
-@van_de_schoot_grolts-checklist_2017 developed a protocol called 
-Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
-Two of the GRoLTS guidelines (namely, 6a and 6b) refer to 
-considering alternative model specifications.
-Both are discussing specific cases in latent trajectory model specification. 
-The first is about whether the variance of the growth parameter 
-is estimated freely or fixed,
-and the second is about whether conditional independence is assumed,
-as the researcher might also want to free-up the variance-covariance structure.
-In LCA, there are also many different ways to specify the model.
-Means, variances, and covariances between the indicators 
-can either be constrained across classes, or estimated freely.
-Researchers using LCA should transparently report their chosen parametrization 
-as well as discuss different parametrizations that were tested in the process.
-
-Different types of latent class models have different parameters.
-For example, mixture models and latent profile analyses 
-typically have class-specific means, variances, and covariances.
-Latent growth analyses have the same parameters, 
-but with respect to the latent growth variables.
-Latent class analyses with ordinal indicators have thresholds.
-All of these parameters can be freely estimated, 
-constrained to be equal across classes,
-or fixed to a certain value (e.g., to zero).
-The total number of parameters thus 
-scales with the number of estimated classes.
-Consequently, LCA methods have a potentially very high number of parameters.
-As any of these parameters could be misspecified,
-it is important to consider alternative model specifications.
-However, alternative model specifications may be approached differently 
-depending on whether an analysis is data driven (exploratory), 
-or theory driven (confirmatory).
-This distinction has remained underemphasized in prior writing.
-
-Prior literature on LCA has emphasized its exploratory applications 
-[@nylund_deciding_2007].
-In exploratory LCA, 
-a large number of models are typically estimated in batch,
-with varying numbers of classes and model specifications.
-The "correct" model specification is then determined 
-based on a combination of fit indices,
-significance tests, and interpretability.
-For latent profile analysis, 
-the function `tidySEM::mx_profiles(classes, variances, covariances)`
-largely automates this process.
-The argument `classes` indicates which 
-class solutions should be estimated (e.g., 1 through 6).
-The argument `variances` specifies whether 
-variances should be `"equal"` or `"varying"` across classes.
-The argument `covariances` specifies whether 
-covariances should be constrained to 
-`"zero"`, `"equal"` or `"varying"` across classes.
-The means are free to vary across classes by default, 
-although the more general function `tidySEM::mx_mixture()` 
-could be used to circumvent this.
-After all models have been estimated, 
-the function `tidySEM::table_fit()` can be used 
-to obtain a model fit table suitable for 
-determining the optimal model according to best practices.
-Note, however, that this table does not include 
-the bootstrapped likelihood ratio test (BLRT) by default,
-because this test is very computationally expensive.
-It is recommended to use the function `tidySEM::BLRT()` 
-to compare a shortlist of likely candidate models 
-based on other fit indices. 
-More on fit indices can be found in 
-the Model Fit Indices subsection of this paper.
-
-Confirmatory LCA typically requires 
-less comprehensive alternative model specifications.
-For example, in the context of a preregistered analysis, 
-the main models of interest would have been specified a priori. 
-Even in this case, 
-the theoretical model could be compared to a few others to contextualize it.
-
-```{r, include = FALSE}
-grolts <- structure(list(Number = c("1.", "2.", "3a.", "3b.", "3c.", "4.", 
-"5.", "6a.", "6b.", "7.", "8.", "9.", "10.", "11.", "12.", "13.", 
-"14a.", "14b.", "14c.", "15.", "16."), Item = c("Is the metric of time used in the statistical model reported?", 
-"Is information presented about the mean and variance of time within a wave?", 
-"Is the missing data mechanism reported?", "Is a description provided of what variables are related to attrition/missing data?", 
-"Is a description provided of how missing data in the analyses were dealt with?", 
-"Is information about the distribution of the observed variables included?", 
-"Is the software mentioned?", "Are alternative specifications of within-class heterogeneity considered (e.g., LGCA vs.
-LGMM) and clearly documented?
-If not, was sufficient justification provided as to eliminate certain specifications from consideration?",
-"Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented?
-If not, was sufficient justification provided as to eliminate certain specifications from consideration?",
-"Are alternative shape/functional forms of the trajectories described?", 
-"If covariates have been used, can analyses still be replicated?", 
-"Is information reported about the number of random start values and final iterations included?", 
-"Are the model comparison (and selection) tools described from a statistical perspective?", 
-"Are the total number of fitted models reported, including a one-class solution?", 
-"Are the number of cases per class reported for each model (absolute sample size, or proportion)?", 
-"If classification of cases in a trajectory is the goal, is entropy reported?", 
-"Is a plot included with the estimated mean trajectories of the final solution?", 
-"Are plots included with the estimated mean trajectories for each model?", 
-"Is a plot included of the combination of estimated means of the final model and the observed individual trajectories split out for each latent class?", 
-"Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?", 
-"Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?"
-), relevant = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
-TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
-TRUE, TRUE, TRUE)), row.names = c(NA, -21L), class = "data.frame")
-```
-
-### Software
-
-Many software packages are available for the estimation of LCA models.
-Some of these packages have limited functionality, 
-or implement specific innovations.
-Other packages implement LCA in the context of 
-a more flexible structural equation modeling framework.
-The most notable examples of the latter are 
-the commercial programs Mplus and Latent GOLD, 
-and the free open source R-package OpenMx.
-The commercial packages stand out 
-because they offer relatively user-friendly interfaces
-and implement sensible defaults for complex analyses, 
-including LCA methods.
-This lowers the threshold for applied researchers to adopt such methods.
-Commercial software also has several downsides, however.
-One such downside is that use of the software 
-is restricted to those individuals 
-and institutions who can afford a license.
-Another downside is that the source code, 
-being proprietary, cannot be audited, debugged, 
-or enhanced by third parties.
-This incurs the risk that mistakes in the source code 
-may go unnoticed, and curbs progress as 
-software developers cannot add new functionality.
-
-Conversely, the free open source program OpenMx is very flexible,
-but not very user-friendly. 
-We directly address this limitation using the `tidySEM` R-package.
-New functionality in `tidySEM` seeks to lower the threshold 
-for latent class analysis using `OpenMx`.
-It adheres to best practices in estimation and reporting, 
-as described in this paper.
-The user interface is simple, 
-making use of the model syntax of the widely used `lavaan` R-package.
-This syntax offers a human-readable way to specify latent variable models.
-Minor enhancements are made to simplify the specification of LCA.
-
-Because of the limitations of the aforementioned tools, 
-we set out to develop a free tool that
-provides sensible defaults and is easy to use, 
-but provides the option to access and modify 
-all of the model inputs (i.e., low barrier, high ceiling). 
-`tidySEM` interfaces with existing tools, 
-and is able to translate between what existing tools are capable of 
-and what researchers and analysts
-carrying-out person-oriented analyses would like to specify. 
-Furthermore,`tidySEM` facilitates fully-reproducible analyses 
-and contributes to open science.
-
-### Algorithm
-
-LCA parameters and model fit statistics 
-can be estimated in a variety of ways.
-The choice of the estimator 
-depends on the presence of missing values,
-sample size, number of indicators, 
-and available computational resources [@weller_latent_2020].
-A commonly used technique is maximum likelihood (ML) estimation
-with the expectation-maximization (EM) algorithm as a local optimizer. 
-Imagine we are estimating two parameters, 
-e.g. the class-specific means $\mu_c$ on a continuous indicator
-(ignoring the variance for now). 
-The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes
-the likelihood ($LL$) of all observed data. 
-In practice, instead of maximizing $LL$, 
-often $-2*LL$ is minimized, 
-as this offers computational advantages. 
-We can think of this optimization problem 
-as a three-dimensional landscape: 
-The X and Y dimensions are determined by the class-specific means, 
-so $X = \mu_1$ and $Y = \mu_2$ - 
-and the Z-dimension is determined by $Z = -2*LL$. 
-The optimizer must find the deepest "valley" in this landscape, 
-which reflects the combination of $\mu_1$ and $\mu_2$ that 
-maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble dropped in this landscape.
-It is dropped at some random point in space, 
-and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley, 
-it will settle on the bottom of that valley
-(this is known as "convergence").
-It cannot climb out again.
-Thus, if their are multiple valleys, 
-the risk is that the optimizer gets stuck in a shallower valley (a "local optimum"), 
-and never discovers the deepest valley (the "global optimum", or best solution).
-One solution to this problem is to drop many marbles at random places, 
-compare their final $-2*LL$ values,
-choose the solution with the lowest $-2*LL$, 
-and make sure that several marbles replicated this solution.
-This is the "random starts" approach.
-
-One problem with the random starts approach is that 
-it is computationally expensive to run this many replications.
-Moreover, because the algorithm begins with random starting values,
-many of the marbles are likely to be very far away from a "good enough" solution.
-Two innovations may improve the estimation procedure.
-The first is that, instead of picking random starting values, 
-a "reasonable solution" may be used for the starting values.
-For example, if we assume that the different classes are likely to have 
-different mean values on the indicators, 
-then the K-means clustering algorithm can be used to 
-determine these cluster centroids.
-We can compute the expected values of all model parameters 
-by treating the K-means solution as a known class solution,
-and use these as starting values for a mixture model.
-One remaining concern is that this approach may result 
-in starting values close to a local optimum,
-and that the EM algorithm will thus never find the global optimum.
-A second innovation addresses this concern.
-Instead of using EM, 
-it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some "destination" in the landscape, 
-and compares its likelihood to the current one. 
-If the destination likelihood is higher, 
-the estimator moves there.
-If the destination likelihood is *lower*, 
-the estimator still moves there occasionally, 
-based on probability.
-This latter property allows it to escape local optima,
-and find the global optimum.
-
-By default, `tidySEM` derives starting values using K-means clustering,
-and identifies the global optimum solution using simulated annealing.
-Once a solution has been found,
-simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for 
-the parameters that can be used to compute standard errors.
-Note that these defaults can be manually overridden.
-
-A recent paper suggested maximum likelihood with robust standard errors 
-should be used when the observed indicators are not normally distributed [@spurk_latent_2020]. 
-This statement is incorrect, and may lead readers to believe that 
-they must use commercial software, 
-as robust maximum likelihood is currently only implemented 
-in Mplus and latentGOLD.
-As explained before, 
-mixture modeling assumes that observed data are 
-a mixture of (multivariate) normal distributions; 
-thus, the observed indicators will likely not be normally distributed.
-<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->
-
-### Class Enumeration
-
-As explained, LCA can be done in an exploratory or in a confirmatory fashion.
-In exploratory LCA, a sequence of models is fitted to the data 
-with each additional model estimating one more class than the previous model.
-These models are then compared 
-and the best solution is selected as the final class solution.
-In some cases, prior theory can inform the researcher 
-about the number of classes to expect.
-Even in such confirmatory LCA cases, 
-it is nonetheless useful to know if the theoretical model 
-is markedly better than those with differing numbers of classes.
-Therefore, it may always be useful to compare different class solutions.
-
-From a sequence of models, the final class solution is chosen 
-based on both theoretical and statistical criteria.
-Theory should drive the selection of indicator variables, 
-inform the expectations and reflect on the findings.
-In addition to this, there are several statistical criteria 
-to consider in model selection.
-These include but are not limited to likelihood ratio tests, 
-information criteria, 
-and the Bayes factor [@weller_latent_2020].
-
-Relative model fit can be examined using the likelihood ratio test. 
-This is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as 
-the difference in maximum log likelihoods of the two models,
-with the test degrees of freedom being 
-the difference in the degrees of freedom of the two compared models.
-The test statistic follows the $\chi^2$ distribution,
-and we want it to be non-significant
-in order to give support to the simpler model. 
-The likelihood ratio test can only compare two nested models at a time [@lanza_introduction_2003].
-
-### Model Fit Indices
-
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC) 
-and the Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2*log likelihood 
-(which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance model fit and model complexity.
-The lower the value of an information criterion, 
-the better the overall fit of the model.
-The BIC applies a stronger penalty for model complexity 
-that scales logarithmically with the sample size.
-The literature suggests the BIC may be 
-the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018].
-Both the AIC and the BIC are available in the `tidySEM` output.
-
-Information criteria may occasionally contradict each other, 
-so it is important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer 
-the most parsimonious model that has best fit 
-according to any of the available fit indices.
-Yet another option is to incorporate information from 
-multiple fit indices using the analytic hierarchy process 
-[@akogul_comparison_2016].
-Finally, one might make an elbow plot and 
-compare multiple information criteria
-[@nylund-gibson_ten_2018].
-
-Another common test of model fit is 
-the likelihood ratio $\chi^2$ goodness-of-fit test.
-However, this test is not implemented in `tidySEM`.
-
-LCA studies commonly report -2*log likelihood of the final class solution.
-This is a basic fit measure used to compute most information criteria.
-However, since log likelihood is not penalized for model complexity,
-it will continuously fall with the addition of more classes.
-
-An alternative is using the bootstrapped likelihood ratio test
-which can be run using `tidySEM::BLRT()`.
-Currently, this test is computationally expensive and 
-can be slow on most computers.
-A faster version of this test, 
-namely an implementation of *the lazy bootstrap* [@van_kollenburg_lazy_2018]
-to `tidySEM` is being developed.
-
-### Classification Diagnostics
-
-Best models will divide the sample into subgroups 
-which are internally homogeneous and externally distinct.
-Classification diagnostics give us a way to assess 
-the degree to which this is the case.
-They are separate from model fit indices as 
-a model can fit the data well 
-but show poor latent class separation [@masyn_latent_2013].
-Classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions 
-because they are uninterpretable.
-Interpretability should always be a consideration 
-when considering different class solutions [@nylund-gibson_ten_2018].
-
-Three important classification diagnostics provided by `tidySEM` are: 
-(1) *the minimum* and *maximum percentage of the sample assigned to a particular class*,
-(2) *the range of the posterior class probabilities by most likely class membership*, 
-and (3) *entropy*. 
-All three are based on posterior class probabilities.
-
-The posterior class probability is a measure of classification uncertainty
-which can be computed for each individual,
-or averaged for each latent class.
-When the posterior class probability is computed for each individual in the dataset,
-it represents each person's probability of belonging to each latent class.
-For each person, the highest posterior class probability is then determined 
-and the individual is assigned to the corresponding class. 
-We want each individual's posterior class probabilities to be 
-high for one and low for the remaining latent classes. 
-This is considered a high classification accuracy 
-and means that the classes are distinct.
-To obtain posterior class probabilities, run `tidySEM::class_prob()`.
-This function produces output comprised of several elements:
-
-`$sum.posterior` is a summary table of the posterior class probabilities 
-indicating what proportion of the data contributes to each class.
-
-`$sum.mostlikely` is a summary table of the most likely class membership
-based on the highest posterior class probability.
-From this table, we compute the minimum and maximum 
-percentage of the sample assigned to a particular class,
-, i.e. **n_min** 
-(the smallest class proportion based on the posterior class probabilities)
-and **n_max** 
-(the largest class proportion based on the posterior class probabilities).
-We are especially interested in **n_min** as
-if it is very small and comprised of few observations,
-the model for that group might not be locally identified.
-It may be impossible to calculate descriptive statistics for such a small class.
-Estimating LCA parameters on small subsamples
-might lead to bias in the results.
-Therefore, we advise caution when dealing with small classes.
-
-`$mostlikely.class` is a table with rows representing
-the class the person was assigned to,
-and the columns indicating the average posterior probability.
-The diagonal represents the probability that 
-observations in each class will be correctly classified.
-If any of the values on the diagonal of this table is low,
-we might consider not to interpret that solution.
-In `tidySEM` we use the diagonal to compute the range of 
-the posterior class probabilities by most likely class membership
-which consists of
-the lowest class posterior probability (**prob_min**),
-and the highest posterior probability (**prob_max**).
-Both **prob_min** and **prob_max** can be used to 
-disqualify certain class solutions,
-and are a convenient way to summarize class separation in LCA.
-We want both **prob_min** and **prob_max** to be high as 
-that means that for all classes
-the people who were assigned to that class 
-have a high probability of being there.
-**prob_min** is especially important as it can diagnose 
-if there is a class with low posterior probabilities 
-which could make one reconsider that class solution.
-
-`$avg.mostlikely` contains the average posterior probabilities for each class, 
-for the subset of observations with most likely class of 1:k, 
-where k is the number of classes.
-
-`$individual` is the individual posterior probability matrix, 
-with dimensions n (number of cases in the data) x k (number of classes).
-Individual class probabilities are often useful for researchers
-who wish to do follow up analyses.
-
-Entropy is a summary measure of posterior class probabilities 
-across classes and individuals.
-It ranges from 0 (model classification no better than random chance) 
-to 1 (perfect classification).
-As a rule of thumb, values above .80 are deemed acceptable 
-and those approaching 1 are considered ideal.
-An appropriate use of entropy is that it can disqualify certain solutions 
-if class separability is too low. 
-Entropy was not built for nor should it be used for 
-model selection during class enumeration [@masyn_latent_2013].
-
-**n_min**, **n_max**, **prob_min**, **prob_max**, and **entropy** and can be obtained using `tidySEM::table_fit()`.
-
-### Interpreting Class Solutions
-
-An important outcome of LCA are conditional item probabilities,
-also known as class-specific item probabilities [@masyn_latent_2013], 
-conditional response, or conditional solution probabilities [@geiser_data_2012].
-They indicate the probability of an item being endorsed
-given that the observation belongs to a particular latent class.
-Conditional item probabilities can be obtained using `tidySEM::table_prob()`.
-If a particular item is endorsed by two or more classes at markedly different rates,
-it is said to discriminate well between the classes 
-and is consequently considered a good indicator.
-Classes are considered highly homogeneous with respect to an item 
-when for a particular item 
-there is a distinct difference in conditional item probabilities
-for two or more classes.
-For instance, if an item is endorsed below 30% for one class and 
-above 70% for another class,
-the classes have high homogeneity with respect to this item [@masyn_latent_2013].
-Conditional item probabilities are the analogue of mean and standard deviation
-when the indicators are binary or ordinal.
-
-A problem which can occur is that of inadmissible solutions. 
-With binary indicators, LCA is modelling a cross-table with all the predictors.
-The problem with such cross-tables is that they will often contain empty cells,
-i.e. combinations of responses that never occur together. 
-This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
-Such boundary parameter estimates could indicate that the solution is invalid [@geiser_data_2012].
-Boundary parameter estimates can also happen with continuous indicators.
-For instance, if we have a zero-inflated normal distribution and a two class solution,
-one class might have the mean of zero and its standard deviation cannot be determined 
-since there is little variance. 
-This too could be a sign of an invalid solution,
-warn that too many classes were extracted, or indicate a local optimum [@geiser_data_2012].
-
-### Label Switching
-
-The final class solution will usually discover and enumerate several classes. 
-The class ordering, however, is completely arbitrary.
-The class labeled as Class 1 in one solution 
-may become Class 2 or Class 3 in another model,
-even when the only difference between the models is in their starting values. 
-Label switching is something to be mindful of when comparing different LCA models [@masyn_latent_2013].
-
-The order of clusters is nondeterministic when using K-means in `tidySEM`. 
-Therefore label switching is still a consideration.
-A simple solution to this is setting a random seed number one line prior to fitting the model.
-We advise `tidySEM` users to always do so in order to circumvent label switching.
-
-Class names should be chosen to accurately reflect group membership. 
-Overly simplified and generalized class names may prove misleading 
-to both audiences and researches alike
-leading to what is known as a naming fallacy [@weller_latent_2020].
-
-## Best Practices in Reporting
-
-Among studies using LCA, reporting practices vary significantly [@weller_latent_2020].
-Various authors have tried to improve and standardize ways of reporting LCA 
-[e.g. @masyn_latent_2013; @weller_latent_2020], 
-but more work is needed.
-@van_lissa_worcs_2020 developed WORCS,
-a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects 
-based on the TOP-guidelines developed by @nosek_promoting_2015.
-WORCS workflow can be easily implemented in R in form of 
-an R package which facilitates preregistration, article drafting,
-version control, citation and formatting, among others [@van_lissa_worcs_2020].
-
-TOP-guidelines emphasise the use of comprehensive citation 
-(including referencing the software used in the analysis),
-as well as code and data sharing wherever possible [@nosek_promoting_2015].
-@van_lissa_worcs_2020 suggest sharing synthetic data in case 
-the original data cannot be shared, 
-and provide functions to generate such synthetic data.
-Ideally, the entire research project is made reproducible 
-so that others may download it and reproduce it with just one click; 
-for guidance, see @van_lissa_worcs_2020.
-
-As the open science movement is gaining momentum,
-researchers are becoming increasingly aware
-how important it is that analyses can be reproduced and audited.
-In line with open science principles, 
-one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that 
-user-friendly methods for estimating LCA models 
-have predominantly been available in commercial software packages 
-(e.g., *Mplus* and *Latent GOLD*).
-A potential downside of commercial software is that 
-it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because 
-the underlying source code is proprietary.
-To overcome these limitations, 
-the present paper introduces new user-friendly functions in 
-the `tidySEM` R-package that can be used to 
-estimate a wide range of LCA models 
-using the free, open-source R-package `OpenMx`.
-The reporting guidelines described in this paper 
-are adopted in `tidySEM` by default.
-The `tidySEM` R-package thus makes advanced mixture modeling 
-based on best practices widely accessible,
-and facilitates the adoption of the estimation and 
-reporting guidelines described in this paper.
-
-### Visualization
-
-Plots can greatly improve the interpretability of LCA models.
-There are several stages where this is an important consideration.
-
-First, in class enumeration, 
-we want to compare several competing class solutions.
-This can be done by means of the AIC and BIC.
-Here, we suggest using an elbow plot
-with the $K-class\ solution$ on the X-axis,
-and information criteria value on the Y-axis.
-We show an implementation of such an elbow plot 
-in our Tutorial subsection.
-<!-- For continuous variables, density plots can aid class enumeration. -->
-<!-- DA: Do we have density plots implemented in tidySEM? -->
-<!-- If so, we should add this example to the tutorial. -->
-
-Second, once we have decided on the final class solution,
-we want to interpret the response patterns
-on the indicators for each latent class.
-While `tidySEM` can create a table showing
-the probability of each item's response endorsement 
-(using `tidySEM::table_prob`),
-it may be easier to inspect these probabilities visually.
-For this reason, we created `ggplot2::plot_prob()`.
-The resulting shows response patterns 
-on all indicators for each group. 
-We give an example of this plot in the Tutorial subsection.
-
 # Tutorial
@@
 
-In this example, we use *data_mix_ordinal*, 
-a simulated data for mixture modelling with ordinal indicators. 
-This dataset is built-in to `tidySEM`.
-For more information about the dataset, 
-type `?tidySEM::data_mix_ordinal` into the R console
-after loading the `tidySEM` using library().
+In this example, we use *data_mix_ordinal*, a simulated data for mixture
+modeling with ordinal indicators. This dataset is built-in to
+`tidySEM`. For more information about the dataset, type
+`?tidySEM::data_mix_ordinal` into the R console after loading the
+`tidySEM` using library().
 
@@
 
-Before we fit a series of LCA models, we set a random seed
-using `set.seed()`.
-The seed number can be any digit.
-This is an important step as 
-there is some inherent randomness in the LCA computations,
-and having the same seed number ensures that
-two separate researcher obtain exactly the same results
-when fitting LCA models.
+Before we fit a series of LCA models, we set a random seed using
+`set.seed()`. The seed number can be any digit. This is an important
+step as there is some inherent randomness in the LCA computations, and
+having the same seed number ensures that two separate researcher obtain
+exactly the same results when fitting LCA models.
 
-Finally, we reach the step of fitting LCA models.
-To do so, we use `tidySEM::mx_lca()`
-which takes data and number of classes as inputs.
-In our example, we want to fit 1 to 4 class solutions
-and compare their output.
-Depending on your computer's computational power,
+Finally, we reach the step of fitting LCA models. To do so, we use
+`tidySEM::mx_lca()` which takes data and number of classes as inputs. 
+Notice that the analysis will be done on all the variable in the data set, 
+so you have to make sure it only includes the relevant variables. 
+In our example, we want to fit 1 to 4 class solutions and compare 
+their output. Depending on your computer's computational power, 
 this might take a while.
@@
 set.seed(123) # setting seed 
-res <- mx_lca(data = df, classes = 1:4) # fitting LCA 
+res <- mx_lca(data = df, classes = 1:4) # fitting LCA 1 to 4 class solutions
 ```
+
+Notice that the model run 10 times for each $K$ class solution. 
+This way we can choose the best fitting model across all the solutions. 
+In this example, we see that all models converged without issues.
+
+We can inspect the class of resulting objects.
+
+```{r, include = TRUE, eval=F}
+is(res)
+is(res[[1]])
+```
+
+As `tidySEM::mx_lca()` run several models at a time, the resulting object `res` 
+is a `mixture_list` class where each $K$ class solution is a list element, or
+more specifically an `OpenMx` model (`MxModel`).
 
@@
 fit_table <- table_fit(res) # model fit table
-fit_table[ , c("Name", "LL", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # our selection
+fit_table[ , c("Name", "LL", "Parameters", 
+               "AIC", "BIC", "Entropy", 
+               "prob_min", "prob_max", 
+               "n_min", "n_max")] # our selection
 ```
+
 Our selection of fit indices and classification diagnostics includes:
@@
 ```{r, echo=F}
-Selection <- c("Name", "LL", "AIC", "BIC", "prob_min", "prob_max",
-          "n_min", "n_max")
-Description <- c("the $K$-class solution",
+Selection <- c(
+    "Name", "LL", "Parameters", 
+    "AIC", "BIC", "prob_min", 
+    "prob_max", "n_min", "n_max")
+
+Description <- c(
+    "the $K$-class solution",
     "the -2*log-likelihood of each model",
+    "Number of estimated parameters in each model",
     "the Akaike Information Criterion value",
@@
     "the highest class proportion based on the posterior class probabilities")
+
 selection <- cbind(Selection, Description)
-knitr::kable(selection, format = 'pipe', caption = "Selection of Fit Indices and Classification Diagnostics")
+
+knitr::kable(selection, format = 'pipe', 
+             caption = "Selection of Fit Indices and 
+                        Classification Diagnostics")
 ```
-We discussed several possible strategies to select the final class solution. 
-Here, we apply our own.
 
-To aid our interpretation of the results, 
-we create an elbow plot showing the trends in information criteria across four models.
+We discussed several possible strategies to select the final class
+solution. Here, we apply our own.
+
+To aid our interpretation of the results, we create an elbow plot
+showing the trends in information criteria across four models.
 
@@
 ```
-From the elbow plot, we see that AIC has a lower penalty for model complexity.
-However, we are more interested in the BIC values, which are similar
-for the one, two and three-class solutions, but the four-class solution fits significantly worse.
-For this reason, we eliminate the four-class solution from the selection process.
 
-Then we examine the model fit table.
-As expected, the -2*log likelihood falls successively with each added class.
-As previously stated, classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions because they are uninterpretable.
-We see that prob_min for the four-class solution is low,
-knowing that this solution also has a high BIC, we disqualify this solution.
+From the elbow plot, we see that AIC has a lower penalty for model
+complexity. However, we are more interested in the BIC values, which are
+similar for the one, two and three-class solutions, but the four-class
+solution fits significantly worse. For this reason, we eliminate the
+four-class solution from the selection process.
 
-Out of the remaining three solutions,
-we notice that entropy is the highest for the three-class solution,
-and it has a satisfactory prob_min and n_min.
-Based on this, we retain the three class solution in model selection.
-Note that entropy for the one-class solution will always equal to one,
-as it is 100% true that every case is in that class.
-Based on the low entropy of the two-class solution,
-we eliminate this model.
+Then we examine the model fit table. As expected, the -2\*log likelihood
+falls successively with each added class. As previously stated,
+classification diagnostics should not be used for model selection, but
+they can be used to disqualify certain solutions because they are
+uninterpretable. We see that prob_min and n_min for the four-class solution is
+low, knowing that this solution also has a high BIC, we disqualify this
+solution.
 
-Finally, when comparing the one and three-class solutions,
-we inspect the information criteria.
-For BIC, the one-class solution fits better, but the difference is marginal.
-AIC tells us that the added complexity
-of having three classes still explains the data better
-than a one-class solution. 
-Therefore, we select the three-class solution as our final-class solution.
+Out of the remaining three solutions, we notice that entropy is the
+highest for the three-class solution, and it has a satisfactory prob_min
+and n_min. Based on this, we retain the three class solution in model
+selection. Note that entropy for the one-class solution will always
+equal to one, as it is 100% true that every case is in that class. Based
+on the low entropy of the two-class solution, we eliminate this model.
+
+Finally, when comparing the one and three-class solutions, we inspect
+the information criteria. For BIC, the one-class solution fits better,
+but the difference is marginal. AIC tells us that the added complexity
+of having three classes still explains the data better than a one-class
+solution. Therefore, we select the three-class solution as our
+final-class solution.
 
@@
 ```
-
-\newpage
-
-# References

### 2022-11-22 manuscript.Rmd
Update November 23
- updated the literature list
- added use case examples
- corrected spelling/grammar mistakes
- adjusted spacing, also in the tutorial

*** things to resolve:
1) Caspar should add Well-being compass study to Zotero and update the in-text citation for that use case
2) the plot_density function doesn't work, also the example in the documentation;
I would like to show how to use it in class enumeration section of the tutorial
3) final proof-reading, anything to add?

@@
 group observed *variables* into latent constructs, and LCA groups
-*individuals* into classes.While factor analysis seeks to describe the
+*individuals* into classes. While factor analysis seeks to describe the
 relations between variables, LCA seeks to describe each unobserved group
@@
 
-In factor analysis, the latent variable is continuous, so the relation
-between indicators is defined by the communalities between them, and the
+In factor analysis, the latent variable is continuous, so the relationships
+between indicators are defined by the communalities between them. Here, the
 latent variable represents the error corrected approximation of the
-sample. While in LCA, the relation between indicators is define by how
-well they separate the classes, as a multinomial regression between
-indicators and latent variable [@masyn_latent_2013]
+sample. In LCA, by contrast, the relation between indicators is defined by how
+well they separate the classes, as a multinomial regression between the
+indicators and the latent variable [@masyn_latent_2013].
 
@@
 can be computed from the likelihood of that individual's observed data
-under given the class-specific model.
-For these differences, LCA is a clustering method, that corrects for (random) measurement error, while clustering algorithms do not correct for.
-This is something LCA has in common with factor analysis, as both model an underlying latent variable, which accounts for the communalities between indicators and correcting for measurement error.
-0
+under given the class-specific model. For these differences, 
+LCA is a clustering method, that corrects for (random) measurement error, 
+while clustering algorithms do not correct for. This is something LCA has 
+in common with factor analysis, as both model an underlying latent variable, 
+which accounts for the communalities between indicators 
+and correcting for measurement error.
 
@@
 are the means, variances, and covariances between the indicators (which
-can be standardized to obtain correlations). These parameters can be
-estimated freely, or set to be constrained across classes [@collins_latent_2009].
+can be standardized to obtain correlations). These parameters can be estimated 
+freely, or set to be constrained across classes [@collins_latent_2009].
 
@@
 assumed to be zero. This can be conceived of as a restricted mixture
-model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
+model with covariances fixed to zero. Stating that the covariances/correlations
+are accounted by the latent variable. In some cases, such constraints
 will be inappropriate; for instance when the cohesion between indicators
-is expected to differ between classes [@collins_latent_2009]. As an example, a LPA of ocean
-plastic particles found two classes of particles based on length and
-width: a class of smaller particles with a high correlation between
+is expected to differ between classes [@collins_latent_2009]. As an example,
+a LPA of ocean plastic particles found two classes of particles based on length
+and width: a class of smaller particles with a high correlation between
 length and width, meaning that these particles were approximately round
@@
 correlation between length and width, meaning that these particles were
-heterogenous in shape. From a theoretical perspective, this makes sense,
+heterogenous in shape [@alkema_maximizing_2022]. From a theoretical perspective, this makes sense,
 because the smaller particles have been ground down to a more uniform
 shape by the elements.
-<!-- DA: what is the source for plastic particles? -->
 
@@
 assessed at different time points. Examples of this approach include
-*growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
-These techniques estimate a latent growth model to describe grouped
-trajectories over time. The growth mixture model is a latent class model
-where the parameters that indicate class membership are the intercepts
-and variances, and typically covariances of the latent growth variables,
-e.g., a latent intercept and slope. This technique assumes that
+*growth mixture models* (GMM) and *latent class growth analyses* (LCGA) 
+[@jung_introduction_2008]. These techniques estimate a latent growth model 
+to describe grouped trajectories over time. The growth mixture model is a 
+latent class model where the parameters that indicate class membership are the 
+intercepts and variances, and typically covariances of the latent growth 
+variables, e.g., a latent intercept and slope. This technique assumes that
 individuals within a class can have heterogenous trajectories. If the
@@
 latent trajectories is due to residual error variance.
-<!-- MGV: some reference for this description -->
 
@@
 
-<!-- Caspar will check if I can replace them with examples from Caspar's published papers. -->
-
 There are several use cases for which LCA methods are suitable. One
-example is to test a theory that postulates the existence of a
-categorical latent variable. For example, *identity status theory*
+such application is to test a theory that postulates the existence of a
+categorical latent variable. For instance, *identity status theory*
 posits that, at any given point in time, adolescents reside in one of
-four identity statuses. Latent class analysis can be used to identify
-these four statuses based on observed indicators (e.g., self-reported
-identity exploration and commitment). If results indicate that the data
+four identity statuses [@marcia_development_1966; @marcia_identity_1980]. 
+LCA can be used to test whether these four statuses
+can be recovered based on observed indicators (e.g., self-reported
+identity exploration and commitment). If the results indicate that the data
 are better described by a different number of classes, or that the
@@
 responses on the indicators, then the theory may be called into
-question.
+question. A recent study by @maene_perceived_2022 applied this principle to 
+test whether identities of Belgian adolescents with migration background can be 
+summarized as those relating to their heritage, national and regional identity.
+A LCA revealed three identity profiles: full integration (high belonging to 
+heritage, national and regional identity), national integration (high heritage
+and national identity with low regional identity), and weak separation
+(high heritage identity with neutral attitudes towards the nation and region).
 
@@
 unobserved class membership based on observed indicators, or to classify
-individuals. For example, a mixture model can be used as a diagnostic
-aid when several clinical indicators can be used to distinguish between
-a fixed number of physical [@baughman_mixture_2006], mental
-[@wu_abuse_2011] health problems, or health outcomes after recovering from cancer [@hardy_exploring_2018]. The example of shoe size is a
-rudimentary illustration of this type of application.
+individuals. For example, a mixture model can be utilized as a diagnostic
+aid when using several clinical indicators to distinguish between
+a fixed number of physical [@baughman_mixture_2006], mental health
+[@wu_abuse_2011] problems, or health outcomes after recovering from
+cancer [@hardy_exploring_2018]. The example of shoe size is a rudimentary
+illustration of this type of application. As another example, a recent study 
+identified two suicide risk profiles in military service members and veterans
+[@gromatsky_characteristics_2022].  One class was defined by higher-risk
+history and another by lower-risk history. The higher-risk group comprised of
+individuals with greater intent to die using methods with higher lethality,
+and the lower-risk group was characterized by greater ambivalence towards death
+and usage of lower lethality methods.
+
+A related use case is where class-membership is known but imperfectly measured.
+Mixture models can also be useful to evaluate the quality of measurement
+across instruments. For example, a recent study had employment status 
+information collected across register and survey data. There was disagreement 
+between the two data sources. While it might be tempting to assume that the
+register data is free of error, this has been shown to be false.
+The authors used a mixture model to evaluate the underlying factor of 
+"error corrected employment status" by using two indicators of the same question
+and identifying random and systematic measurement error in both indicators 
+[@pavlopoulos_measuring_2015].
 
@@
 be used to discover the types of publications that get funding.
-
-With LCA, our goal could be to inductively identify the number of
-classes. For instance, if we believe that a variable represents a group,
-but don't know how many groups there are, LCA may be an appropriate
+Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean.
+Prior to this study, ways to categorize ocean microplastics in terms of length,
+width and polymer type were rather arbitrary, and assumed that these properties
+were independent. Also, there was a lack of information regarding variability 
+within these categories. The authors used finite Gaussian mixture modeling to 
+build a classification system for ocean microplastic particles based on a set 
+of measured characteristics. With regards to particle dimensions, two classes
+emerged: one with smaller fragments whose width and height were highly 
+correlated, and another consisting of larger fragments with low correlation 
+between particle width and height.
+
+With LCA, our goal could be to inductively identify the number of classes. 
+For instance, if we believe that a variable represents a group,
+but do not know how many groups there are, LCA may be an appropriate
 technique to answer this question. For example, @hopfer_social_2014
@@
 urban population in Winnipeg, Canada. The underlying assumption was that
-there were different risk profiles, but their number was not known. From
-a collection of indicators, LCA provided evidence that there are four
-distinct risk profiles in the Winnipeg area.
-
-Another application of LCA is to classify individuals. In a peer
-harassment study, @giang_using_2008 used LCA to classify over 2,000
-sixth grade students into aggressor and victim latent classes. The
-five-class solution comprised of classes of victims, aggressors, and
-socially adjusted students. For instance, it revealed that there were
-two types of victims: highly-victimized aggressive-victims and
-highly-aggressive aggressive-victims.
+there were different risk profiles, but their number was not known. 
+From a collection of indicators, LCA provided evidence that there are four
+distinct risk profiles in the Winnipeg area. Another study investigated 
+adolescent empathy development and its influence on parent-child conflict 
+[@van_lissa_divergence_2015]. The data was collected over six years. 
+Using LCGA, adolescents with similar empathy development trajectories 
+were grouped into: high-, average-, and low-empathy classes.
+
+Another application of LCA is to classify individuals. In a peer harassment 
+study, @giang_using_2008 used LCA to classify over 2,000 sixth grade students 
+into aggressor and victim classes. The five-class solution comprised of 
+classes of victims, aggressors, and socially adjusted students. 
+The study revealed that there were two types of victims: highly-victimized 
+aggressive-victims and highly-aggressive aggressive-victims.
+Another example is the "Well-being compass" study by van Lissa, currently 
+a preprint. The study was conducted in collaboration with a mental health care
+facility and focused on care providers. Latent class analysis has identified 
+four different risk profiles for care providers. Not only was class membership 
+estimated for the participants of the original study, but the class solution 
+was used as a risk assessment aid for new care providers, 
+so they could be more adequately supported by mental health care professionals.
+<!-- DA: this study is not in zotero yet, I lack information (title, authors, etc.) to enter it, @Caspar could you enter this to Zotero and adjust the in-text citation-->
 
@@
 selection of indicators could be informed.
+An example of such use of LCA is a study by @tsaousis_measurement_2020 who
+studied differential item functioning using LCA. The study focused 
+on university admission tests of Saudi Arabian students. LCA has revealed three
+latent classes: high-, average-, and low-scorers. Subsequently, the authors
+used the Multiple Indicator Multiple Causes model to identify both uniform
+and non-uniform differential item functioning. The main finding was that gender
+was a potential source of differential item functioning for latent class
+indicators. Therefore, gender should be included as a covariate when fitting 
+a LCA model in order to obtain unbiased estimates.
 
@@
 classes. Therefore this covariate was kept as a valuable predictor.
-
-When our interest is the prediction of one or more outcomes, LCA can be
-used to construct latent classes as categorical predictors.
-@lanza_latent_2013 demonstrated how LCA can be used to classify
-adolescents into depression classes, and subsequently these classes can
-be used to predict smoking, grades, and delinquency. The study showed
-that the outcomes predicted by class membership can be binary (regular
-smoking), continuous (grades) or count (delinquency).
-
-These mixture models can also be useful to evaluate quality of measurement across instrument.
-For example, employment status was collected across register and survey data, and there is disagreement between them.
-You could think that register data is without error, but it has been shown that is victim of measurement error as well.
-In this case mixture models was used to evaluate the underlying factor of "error corrected employment status", by using two indicators of the same question and identify random and systamtic measurement error in either indicator [@pavlopoulos_measuring_2015].
+Another example comes from @van_lissa_divergence_2015 who included sex 
+as a predictor of the developmental trajectories and class membership in 
+the previously discussed adolescent empathy development study. The rationale
+was that adding sex as a covariate would account for known sex differences 
+in empathy development, and lead to a model with lower bias.
+
+When our interest is the prediction of one or more outcomes, LCA can be used 
+to construct latent classes as categorical predictors. @lanza_latent_2013 
+demonstrated how LCA can be used to classify adolescents into depression 
+classes, and subsequently these classes can be used to predict smoking, grades,
+and delinquency. The study showed that the outcomes predicted by class 
+membership can be binary (regular smoking), continuous (grades) or 
+count (delinquency). In the previously discussed study of Belgian adolescent
+identities, @maene_perceived_2022 first constructed the latent variable
+representing different multiple identification strategies. Subsequently, 
+the most likely class membership based on posterior class probability was
+used as an "observed" categorical variable for prediction of depressive symptoms.
 
@@
 have the best fit according to both statistical criteria and clinical
-interpretability.
-
-<!-- CJ: (on the next paragraph)  -->
-
-<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
-
-<!-- That is not what this paragraph is about.  -->
-
-<!-- This paragraph is about motivations for using LCA. -->
-
-<!-- One motivation is if your data violates multivariate normality -->
-
-<!-- because violating multivariate normality means that you have multiple groups -->
-
-<!-- and you can use LCA to extract those. That is the only assumption  -->
-
-<!-- we have to discuss here, we just need to explain it. -->
+interpretability. In the longitudinal adolescent empathy development study, 
+@van_lissa_divergence_2015 first investigated heterogeneity in empathy 
+development. This was done through identification of classes of individuals 
+characterized by similar developmental trajectories. The authors used LCGA
+as a dimensionality reduction technique where 467 individual trajectories
+were summarized into three empathy classes. Such reduction can help researchers 
+gain a better perspective of the grand patterns in the data.
 
@@
 distribution is a non-normal mixture of $K$ normal distributions, and it
-can discover the value of $K$, i.e. generate a $K$-class solution.
-
-LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous of categorical.
+can recover the value of $K$, i.e. generate a $K$-class solution.
+For instance, @alkema_maximizing_2022 studied ocean plastic debris. 
+Plastic particle width and height are non-normally distributed. The authors 
+demonstrate that the non-normal distributions of particle dimensions are
+due to the measurements being a mixture composed of several particle types, 
+each with its own normal distribution.
+
+<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical.
 We encourage researchers to be agnostic about the **true** nature of the latent variable.
 And focus on the approach that fits the research question.
-
+-->
+<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->
 # Best Practices
@@
 [@spurk_latent_2020]. This is incorrect. When the number of classes is
-greater than one, Gaussian mixture models assume that the observed indicators are
-a mixture of multiple (multivariate) normal distributions. In our shoe
+greater than one, Gaussian mixture models assume that the observed indicators 
+are a mixture of multiple (multivariate) normal distributions. In our shoe
 size example, it can be seen that the population distribution is
@@
 [@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
-been distinguished in the literature [@rubin_inference_1976; @enders_applied_2022]: Missing
-completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
+been distinguished in the literature [@rubin_inference_1976; 
+@enders_applied_2022]: Missing completely at random (MCAR), which means that 
+missingness is random and unrelated to any observed or unobserved variables;
 missing at random (MAR), which means that missingness is contingent on
@@
 important to emphasize that missingness is adequately handled by default
-in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` which is
-the backend of `tidySEM`. These packages use Full Information Maximum
+in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` 
+which is the backend of `tidySEM`. These packages use Full Information Maximum
 Likelihood (FIML) estimation, which makes use of all available
 information without imputing missing values. FIML is a best-practice
-solution for handling missing data; on par with multiple imputation when the estimation model includes the same variables as the imputation model
+solution for handling missing data; on par with multiple imputation when the 
+estimation model includes the same variables as the imputation model
 [@lee_comparison_2021]. FIML estimation assumes that missingness is
@@
 <!-- are you sure Mplus accounts for missing data automatically when the indicators are categorical? I know that is not the case with CFA/SEM models -->
+<!-- CJ: yes, it does -->
 
@@
 Information Criterion (BIC). Both information criteria are based on the
--2\*log-likelihood (which is lower for better fitting models), and add a
-penalty for the number of parameters (thus incentivizing simpler
+-2\*log-likelihood (which is lower for better fitting models), and a
+penalty term for the number of parameters (thus incentivizing simpler
 models). This helps balance model fit and model complexity. The lower
@@
 model. 
-
-The general objective of the Information Criteria (IC) is to evaluate the model's out-of-sample predictive accuracy, so adjusting for over fitting.
-Fit measures like $R^2$ evaluate the in-sample predictive accuracy, meaning that they evaluate the models ability to predict the observed outcomes based on the same data that was use to build up the model.
-These metrics are positively bias, meaning that will present better model fit than it has in reality.
-While IC corrects for this positive bias by evaluating the models accuracy approximating the out-of-sample predictive accuracy, meaning that it is the ability to predict the outcome for observations that are not part of the training model.
-Ideally we would estimate the model a lot of time with less observations and predict their scores, but IC's approximate this by their different penalty metrics.
-For example $R^2$ will increase even if an added predictors are unnecessary, while IC's will show worst fit when a predictor (complexity) is unnecessary [@mcelreath_statistical_2020]
+XXXXXX
+The general objective of information criteria (ICs) is to evaluate the model's 
+out-of-sample predictive accuracy, thus adjusting it for overfitting. 
+Fit measures such as $R^2$ evaluate the in-sample predictive accuracy, 
+i.e. the model's ability to predict the observed outcomes based on the same 
+data that was used to fit the model. In-sample metrics are positively biased, 
+meaning that they overestimate the fit the model has in reality, i.e. when 
+presented with new data. While common ways to overcome this bias is through
+resampling methods such as cross-validation, or use of out-of-sample data such
+as a test set, ICs correct for this positive bias by evaluating the model's 
+accuracy in a way which approximates the out-of-sample predictive accuracy. 
+ICs achieve this by applying different penalty metrics. For example, $R^2$ will 
+increase even if an added predictor is fully redundant. ICs show preference for
+simpler models and indicate worse fit when a predictor, or its added complexity 
+is unnecessary [@mcelreath_statistical_2020].
 
@@
 BIC may be the most appropriate information criterion to use for model
-comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. Both the AIC and the BIC are
-available in the `tidySEM` output.
+comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
+Both the AIC and the BIC are available in the `tidySEM` output.
 
@@
 
-A disadvantage of the IC's is that we do not have standard errors for them, so we only have the absolute values without a measure of their variability.
-So, the difference between models IC can be very small.
-Still indicating that the model with the lower value is "better", but if this difference is  very small can considered them "functionally equal", and you should take into consideration the interpretability of the model.
+A disadvantage of ICs is that they are not accompanied by standard errors or 
+any measure of variability, only their absolute value. The difference between 
+two models' ICs can be rather small. The lower information criterion would 
+still indicate the better model, but if the ICs difference is very small, 
+two models can be considered "functionally equal". In such situations, the 
+interpretability of the model should take priority.
 
@@
 (1) the *minimum* and *maximum* percentage of the sample assigned to a 
-particular *class*, (2) the *range of the posterior class probabilities* by most likely class membership, and (3) *entropy*. All three are based
-on posterior class probabilities.
+particular *class*, (2) the *range of the posterior class probabilities* by most
+likely class membership, and (3) *entropy*. All three are based on posterior 
+class probabilities.
 <!-- MGV: I will work on adding average posterior class probability (AvePP) as well-->
@@
 `$individual` is the individual posterior probability matrix, with
-dimensions n (number of cases in the data) x k (number of classes). Additionally it includes the `predicted` class in function of the highest predicted probability.
-Individual class probabilities and/or predicted class are often useful for researchers who wish to do follow up analyses.
+dimensions n (number of cases in the data) x k (number of classes). 
+Additionally, it includes the `predicted` class as a function of the highest 
+predicted probability. Individual class probabilities and/or predicted class 
+are often useful to researchers who wish to carry out follow up analyses.
 
@@
 A problem which can occur is that of inadmissible solutions. With binary
-indicators, LCA is modelling a cross-table with all the predictors. The
+indicators, LCA is modeling a cross-table with all the predictors. The
 problem with such cross-tables is that they will often contain empty
@@
 
-Class names should be chosen to accurately reflect group membership, as theoretically relevant names according to the interpretation of the class solutions.
-Overly simplified and generalized class names may prove misleading to
-both audiences and researches alike leading to what is known as a naming
+Class names should be chosen to accurately reflect group membership,
+as theoretically relevant names according to the interpretation of the class 
+solutions. Overly simplified and generalized class names may prove misleading 
+to both audiences and researches alike leading to what is known as a naming
 fallacy [@weller_latent_2020].
@@
 the `tidySEM` R-package that can be used to estimate a wide range of LCA
-models using the free, open-source R-package `OpenMx` as the backend. The reporting
-guidelines described in this paper are adopted in `tidySEM` by default.
-The `tidySEM` R-package thus makes advanced mixture modeling based on
-best practices widely accessible, and facilitates the adoption of the
+models using the free, open-source R-package `OpenMx` as the backend. The 
+reporting guidelines described in this paper are adopted in `tidySEM` by 
+default. The `tidySEM` R-package thus makes advanced mixture modeling based 
+on best practices widely accessible, and facilitates the adoption of the
 estimation and reporting guidelines described in this paper.
@@
 suggest using an elbow plot with the $K-class\ solution$ on the X-axis,
-and information criteria value on the Y-axis. We show an implementation
-of such an elbow plot in our Tutorial subsection.
-<!-- For continuous variables, density plots can aid class enumeration. -->
-<!-- DA: Do we have density plots implemented in tidySEM?, yes, it is plot_density  -->
-<!-- If so, we should add this example to the tutorial. -->
+and information criteria value on the Y-axis. For continuous variables, 
+density plots can aid class enumeration. We demonstrate how to implement 
+both elbow and density plots in the Tutorial subsection of this paper.
 
@@
 In this example, we use *data_mix_ordinal*, a simulated data for mixture
-modelling with ordinal indicators. This dataset is built-in to
+modeling with ordinal indicators. This dataset is built-in to
 `tidySEM`. For more information about the dataset, type
@@
 `tidySEM::mx_lca()` which takes data and number of classes as inputs. 
-Notice that the analysis will be done on all the variable in the data set, so you have to make sure it only includes the relevant variables.
-In
-our example, we want to fit 1 to 4 class solutions and compare their
-output. Depending on your computer's computational power, this might
-take a while.
+Notice that the analysis will be done on all the variable in the data set, 
+so you have to make sure it only includes the relevant variables. 
+In our example, we want to fit 1 to 4 class solutions and compare 
+their output. Depending on your computer's computational power, 
+this might take a while.
 
@@
 set.seed(123) # setting seed 
-res <- mx_lca(data = df, classes = 1:4) # fitting LCA - 1 to 4 classes
-```
-
-Notice that the model was run 10 times for each $K$ class solution.
-This because the functions runs a total of 10 **extra** times the model.
+res <- mx_lca(data = df, classes = 1:4) # fitting LCA 1 to 4 class solutions
+```
+
+Notice that the model run 10 times for each $K$ class solution. 
 This way we can choose the best fitting model across all the solutions. 
-In this examples we see that all models converged without issues.
+In this example, we see that all models converged without issues.
+
+We can inspect the class of resulting objects.
 
@@
 
-As `tidySEM::mx_lca()` was run for several models at the same time, the resulting object `res` is a `mixture_list` class where each $K$ class solution is an object in the list, and each of them are `OpenMx` models (`MxModel`).
+As `tidySEM::mx_lca()` run several models at a time, the resulting object `res` 
+is a `mixture_list` class where each $K$ class solution is a list element, or
+more specifically an `OpenMx` model (`MxModel`).
 
@@
 fit_table <- table_fit(res) # model fit table
-fit_table[ , c("Name", "LL", "Parameters", "AIC", "BIC", "Entropy", "prob_min", "prob_max", "n_min", "n_max")] # our selection
+fit_table[ , c("Name", "LL", "Parameters", 
+               "AIC", "BIC", "Entropy", 
+               "prob_min", "prob_max", 
+               "n_min", "n_max")] # our selection
 ```
@@
 ```{r, echo=F}
-Selection <- c("Name", "LL", "Parameters", "AIC", "BIC", "prob_min", "prob_max",
-          "n_min", "n_max")
-Description <- c("the $K$-class solution",
+Selection <- c(
+    "Name", "LL", "Parameters", 
+    "AIC", "BIC", "prob_min", 
+    "prob_max", "n_min", "n_max")
+
+Description <- c(
+    "the $K$-class solution",
     "the -2*log-likelihood of each model",
@@
     "the highest class proportion based on the posterior class probabilities")
+
 selection <- cbind(Selection, Description)
-knitr::kable(selection, format = 'pipe', caption = "Selection of Fit Indices and Classification Diagnostics")
+
+knitr::kable(selection, format = 'pipe', 
+             caption = "Selection of Fit Indices and 
+                        Classification Diagnostics")
 ```
@@
 elbow_plot <- fit_table[ , c("Name", "AIC", "BIC")] # extract ICs
-elbow_plot <- pivot_longer(elbow_plot, cols = c("AIC", "BIC"), names_to = "IC", values_to = "Value") # to long format
+elbow_plot <- pivot_longer(elbow_plot, cols = c("AIC", "BIC"), 
+                           names_to = "IC", values_to = "Value") # to long format
 
@@
   geom_line(aes(color = IC))+
-  labs(title="Elbow Plot of Information Criteria per Class Solution", x = "Class Solution", y = " Value")+
-  scale_color_manual(name = "Information Criterion", values = c(AIC = 'blue', BIC = 'red'))+
+  labs(title="Elbow Plot of Information Criteria per Class Solution", 
+       x = "Class Solution", y = " Value")+
+  scale_color_manual(name = "Information Criterion", 
+                     values = c(AIC = 'blue', BIC = 'red'))+
   theme_minimal()
@@
 ```{r plot_prob, include = TRUE, eval=F}
-plot_prob(res[[3]]) # visualizing the response patterns for the final class solution
-table_prob(res[[3]]) # tabulating the response patterns for the final class solution
+plot_prob(res[[3]]) # visualizing the response patterns for the final model
+table_prob(res[[3]]) # tabulating the response patterns for the final model
 ```
@@
 
-We can also interpret the response patterns numerically.
-It is a matter of preferences on how to interpret these probabilities.
-Here is where you would **name** each class, such that each response pattern is theoretically meaningful.
+We can also interpret the response patterns numerically. It is a matter of 
+preferences on how to interpret these probabilities. Here is where you would
+**name** each class, such that each response pattern is theoretically meaningful.
 
@@
 ```{r extract_Post_Class_Prob, include = TRUE, eval=F}
-probs <- class_prob(res[[3]]) # extracting the posterior class probabilities of the final class solution
-probs$mostlikely.class # posterior class probabilities by most likely class membership
+probs <- class_prob(res[[3]]) # extracting the posterior class probabilities
+probs$mostlikely.class # posterior probabilities by most likely class membership
 probs$individual # individual posterior class probabilities

### 2022-11-30 manuscript.tex

@@
 {}{\PatchFailed}
-\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 8206}
+\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 9001}
 \usepackage{csquotes}
@@
   pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
-  pdfauthor={Caspar J. van Lissa1,2 \& Mauricio Garnier-Villarreal3},
+  pdfauthor={Caspar J. van Lissa1,2, Mauricio Garnier-Villarreal3, \& Daniel Anadria1},
   pdflang={en-EN},
@@
 \title{Best Practices in Latent Class Analysis using Free Open Source Software}
-\author{Caspar J. van Lissa\textsuperscript{1,2} \& Mauricio Garnier-Villarreal\textsuperscript{3}}
+\author{Caspar J. van Lissa\textsuperscript{1,2}, Mauricio Garnier-Villarreal\textsuperscript{3}, \& Daniel Anadria\textsuperscript{1}}
 \date{}
@@
 most notably finite Gaussian mixture modeling and latent profile
-analysis. Vermunt, J.K. et al. (2004) defined LCA more generally as
+analysis. Vermunt et al. (2004) defined LCA more generally as
 virtually any statistical model where ``some of the parameters {[}\ldots{]}
@@
 the term was initially conceived to refer to analyses with categorical,
-usually binary indicators (Vermunt, J.K. et al., 2004). Nowadays, there are
+usually binary indicators (Vermunt et al., 2004). Nowadays, there are
 a number of related techniques, known by distinct names, that serve a
@@
 can be standardized to obtain correlations). These parameters can be
-estimated freely, or set to be constrained across classes (Collins \& Lanza, 2010).
+estimated freely, or set to be constrained across classes (Collins \& Lanza, 2009).
 
@@
 will be inappropriate; for instance when the cohesion between indicators
-is expected to differ between classes (Collins \& Lanza, 2010). As an example, a LPA of ocean
+is expected to differ between classes (Collins \& Lanza, 2009). As an example, a LPA of ocean
 plastic particles found two classes of particles based on length and
@@
 The term latent class analysis originally referred to cases where the
-observed indicators were categorical (Collins \& Lanza, 2010).
+observed indicators were categorical (Collins \& Lanza, 2009).
 Nowadays, it is more commonly used
@@
 (Van De Schoot et al., 2017). Three types of missingness have
-been distinguished in the literature (Enders, 2010; Rubin, 1976): Missing
+been distinguished in the literature (Enders, 2022; Rubin, 1976): Missing
 completely at random (MCAR), which means that missingness is random and unrelated to any observed or unobserved variables;
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-collins_latent_2010}{}}%
-Collins, L.
-M., \& Lanza, S.
-T.
-(2010).
-\emph{Latent {Class} and {Latent} {Transition} {Analysis}: {With} {Applications} in the {Social}, {Behavioral}, and {Health} {Science}}.
-New Jersey, USA: John Wiley \& Sons, Inc.
-
-\leavevmode\vadjust pre{\hypertarget{ref-enders_applied_2010}{}}%
-Enders, C. K. (2010). \emph{Applied missing data analysis}. New York: Guilford Press.
+\leavevmode\vadjust pre{\hypertarget{ref-collins_latent_2009}{}}%
+Collins, L. M., \& Lanza, S. T. (2009). \emph{Latent {Class} and {Latent} {Transition} {Analysis}: {With} {Applications} in the {Social}, {Behavioral}, and {Health} {Sciences}}. John Wiley \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-enders_applied_2022}{}}%
+Enders, C. K. (2022). \emph{Applied {Missing} {Data} {Analysis}}. Guilford Publications.
 
@@
 Exploring the {Ovarian} {Reserve} {Within} {Health} {Parameters}: {A} {Latent} {Class} {Analysis}.
-\emph{Western Journal of Nursing Research}, 1--16.
+\emph{Western Journal of Nursing Research}, \emph{40}(12), 1903--1918.
 \url{https://doi.org/10.1177/0193945918792303}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mcelreath_statistical_2020}{}}%
-McElreath, R. (2020). \emph{Statistical rethinking: A {Bayesian} course with examples in {R} and {Stan}} (2nd ed.). Boca Raton: Taylor; Francis, CRC Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-openmx}{}}%
+McElreath, R.
+(2020).
+\emph{Statistical {Rethinking} {\textbar} {A} {Bayesian} {Course} with {Examples} in {R} and {STAN}} (2nd ed.).
+Chapman; Hall/CRC.
+Retrieved from \url{https://www.taylorfrancis.com/books/mono/10.1201/9780429029608/statistical-rethinking-richard-mcelreath}
+
+\leavevmode\vadjust pre{\hypertarget{ref-neale_openmx_2016}{}}%
 Neale, M.
@@
 (2016).
-Open{M}x 2.0: {E}xtended structural equation and statistical modeling.
+{OpenMx} 2.0: {Extended} {Structural} {Equation} and {Statistical} {Modeling}.
 \emph{Psychometrika}, \emph{81}(2), 535--549.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pavlopoulos_measuring_2015}{}}%
-Pavlopoulos, D., \& Vermunt, J.
-K.
-(2015).
-Measuring {Temporary} {Employment}.
-{Do} {Survey} or {Register} {Data} {Tell} the {Truth}?
-\emph{Survey Methodology}, \emph{41}(1), 197--214.
-Retrieved from \url{https://www150.statcan.gc.ca/n1/pub/12-001-x/2015001/article/14151-eng.pdf}
+Pavlopoulos, D., \& Vermunt, J. K. (2015). \emph{Measuring temporary employment. {Do} survey or register data tell the truth?} 37.
 
@@
 {A} comparison of robust continuous and categorical {SEM} estimation methods under suboptimal conditions.
-\emph{Psychological Methods}, \emph{17}(3), 354--373.
+\emph{Psychological Methods}, \emph{17}, 354--373.
 \url{https://doi.org/10.1037/a0029315}
 
-\leavevmode\vadjust pre{\hypertarget{ref-lavaan}{}}%
-Rosseel, Y. (2012). {lavaan}: An {R} package for structural equation modeling. \emph{Journal of Statistical Software}, \emph{48}(2), 1--36. \url{https://doi.org/10.18637/jss.v048.i02}
+\leavevmode\vadjust pre{\hypertarget{ref-rosseel_lavaan_2012}{}}%
+Rosseel, Y. (2012). Lavaan: {An} {R} {Package} for {Structural} {Equation} {Modeling}. \emph{Journal of Statistical Software}, \emph{48}, 1--36. \url{https://doi.org/10.18637/jss.v048.i02}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vermunt_jk_latent_2004}{}}%
-Vermunt, J.K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.F., \& Department of Methodology and Statistics.
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_latent_2004}{}}%
+Vermunt, J.
+K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.
+F., \& Department of Methodology and Statistics.
 (2004).

### 2022-11-30 manuscript.Rmd
update 2022-11-30 15:30:14

@@
 title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
-shorttitle        : "BEST PRACTICES LCA"
+shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
 author: 
@@
 abstract: |
-  Latent class analysis is a family of techniques for identifying groups in data based
-  on a parametric model. Its examples include mixture models, latent class
-  analysis (with continuous and/or ordinal indicators), and latent class (mixture)
-  growth analysis. Despite the popularity of this technique, there is limited
+  Latent class analysis refers to a family of techniques for identifying groups
+  in data based on a parametric model.
+  Its examples include mixture models, latent class analysis with ordinal
+  indicators, and latent class growth analysis.
+  Despite the popularity of this technique, there is limited
   guidance with respect to best practices in estimating and reporting mixture
@@
   introduces new functionality for estimating and reporting mixture
-  models in the `tidySEM` R-package whose backend relies on
-  the `OpenMx` R-package.
+  models in the `tidySEM` R-package whose backend relies on `OpenMx`.
 keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
@@
 reporting guidelines for a specific type of LCA known as latent growth
-models, general reporting guidelines for all LCA are still lacking. This
-complicates manuscript review and assessment of the quality of published
-studies, and introduces a risk of misapplications of the technique. The
-present paper seeks to address this gap in the literature by suggesting
-updated guidelines for estimation and reporting of LCA, based on current
-best practices. Importantly, in order to lower the barrier of entry and
-ensure reproducibility of all examples, this paper exclusively relies on
-free, open source software in R. Our goal is to make best-practices in
-LCA widely accessible.
+models, general reporting guidelines for best practices in LCA are still
+lacking.
+This introduces a risk of misapplications of the technique,
+and complicates manuscript review and assessment of the quality of published studies.
+The present paper seeks to address this gap in the literature by introducing updated guidelines for estimation and reporting on latent class analysis,
+based on current best practice. 
+Importantly, in order to lower the barrier of entry to latent class analysis
+and ensure reproducibility of all examples,
+this paper exclusively uses free, open source software for latent class analysis in R.
+To this end, new functionality was developed in the R-package `tidySEM` [REF tidysem].
 
@@
 mixture model assumes that the study population is composed of $K$
-subpopulations or classes. It further assumes that the observed data are
-a mixture of data generated by class-specific models. The simplest
-univariate model is a normal distribution, which can be described with
-two parameters: the mean and the variance. Commonly, the same model is
-estimated across all classes, but with different parameters for each
-class (i.e., class-specific means and variances). Mixture modeling then
-estimates both the parameters for each class, and the probability that
-an individual belongs to each class. This way, describing the whole
-sample with a single mean and variance is a special case where we assume
-the sample is homogeneous in function of these parameters.
+subpopulations or classes.
+It further assumes that the observed data are
+a mixture of data originating from those subpopulations;
+hence the name mixture model.
+Consider the simplest possible
+univariate "model", which is a normal distribution.
+This model has
+two parameters: the mean and the variance.
+Latent class analysis aims to estimate the values of those parameters across a known number of classes $k$,
+as well as the probability of belonging to classes $1 \ldots k$ for every individual $i$.
+Commonly, the same model is estimated across all classes,
+but with different parameters for each class (i.e., class-specific means and variances).
+From this perspective, reporting the mean and variance of a variable is a special case of a mixture model with $k = 1$ class.
 
@@
 
-LCA is like confirmatory factor analysis, except that the continuous
-latent variable is substituted with a categorical latent variable. Both
-assume that the indicators are related, and these relations can be
-represented by the underlying latent variable. One difference between
+Another way to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
+A mixture model is like confirmatory factor analysis,
+except that the continuous latent variable is substituted with a categorical latent variable.
+One difference between
 the two techniques is that factor analysis can be considered as a way to
@@
 
-When the focus is on each individual's estimated class membership, LCA
-can be thought of as a type of clustering algorithm. In line with this
-perspective, LCA is sometimes described as "model-based clustering"
-[@hennig_handbook_2015; @scrucca_mclust_2016]. Many clustering
-algorithms apply some recursive splitting algorithm to the data. By
+When the focus is on each individual's estimated class membership,
+LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
+Specifically, the posterior class probability that an individual belongs to a latent class
+can be computed from the likelihood of that individual's observed data under given the class-specific model.
+These posterior class probabilities can be used to weight follow-up analyses,
+or individuals can be assigned to classes based on their highest class probability.
+If the classes are clearly distinct, measurement error is low, and the latter approach is probably acceptable.
+This perspective on mixture modeling is sometimes referred to as "model-based
+clustering" [@hennig_handbook_2015; @scrucca_mclust_2016]. 
+Many clustering algorithms apply some recursive splitting algorithm to the data. By
 contrast "model-based" clustering refers to the fact that LCA estimates
-cluster membership based on a parametric model. Specifically, the
-posterior class probability that an individual belongs to a latent class
-can be computed from the likelihood of that individual's observed data
-under given the class-specific model. For these differences, 
-LCA is a clustering method, that corrects for (random) measurement error, 
-while clustering algorithms do not correct for. This is something LCA has 
-in common with factor analysis, as both model an underlying latent variable, 
-which accounts for the communalities between indicators 
-and correcting for measurement error.
+cluster membership based on a parametric model.
+Since a parametric model estimates only a fixed number of parameters,
+it can be a relatively parsimonious solution compared to non-parametric techniques.
 
 Finally, in the context of machine learning, LCA can be considered as an
-*unsupervised classification* problem [@figueiredo_unsupervised_2002].
+*unsupervised classification* method [@figueiredo_unsupervised_2002].
 The term *unsupervised* refers to the fact that the outcome variable --
@@
 indicators. Conditional independence means that, after class membership
-is accounted for, the covariances/correlations between indicators are
-assumed to be zero. This can be conceived of as a restricted mixture
-model with covariances fixed to zero. Stating that the covariances/correlations
-are accounted by the latent variable. In some cases, such constraints
+is accounted for,
+the residual covariances/correlations between indicators are
+assumed to be zero.
+In some cases, such constraints
 will be inappropriate; for instance when the cohesion between indicators
-is expected to differ between classes [@collins_latent_2009]. As an example,
-a LPA of ocean plastic particles found two classes of particles based on length
-and width: a class of smaller particles with a high correlation between
-length and width, meaning that these particles were approximately round
-or square in shape, and a class of larger particles with a low
-correlation between length and width, meaning that these particles were
-heterogenous in shape [@alkema_maximizing_2022]. From a theoretical perspective, this makes sense,
-because the smaller particles have been ground down to a more uniform
-shape by the elements.
+is expected to differ between classes [@collins_latent_2009].
+Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
+A class of larger particles with a low correlation between length and width,
+and a class of smaller particles with a high correlation between length and width.
+The reason for this difference in correlations makes theoretical sense:
+the large particles were heterogenous in shape,
+whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements [@alkema_maximizing_2022].
 
@@
 The term latent class analysis originally referred to cases where the
-observed indicators were categorical [@collins_latent_2009]. 
+observed indicators were ordinal [@collins_latent_2009]. 
 Nowadays, it is more commonly used
@@
 indicators are of binary or ordinal measurement level might be described
-as *latent class analysis with ordinal indicators*. Latent class models
-with ordinal indicators are parameterized differently from mixture
-models. One common parameterization assumes that each categorical
-variable reflects an underlying standard normal distribution. The
-parameters are "thresholds" that correspond to quantiles of a standard
+as *latent class analysis with ordinal indicators*.
+Latent class models with ordinal indicators are parametrized differently from mixture models.
+One common parametrization assumes that each categorical variable reflects an underlying standard normal distribution.
+The parameters are "thresholds" that correspond to quantiles of a standard
 normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
@@
 then the resulting threshold would be $t_1 = 0.25$.
-
-This paper will primarily focus on mixture models and special cases
-thereof, although most of the suggested guidelines are applicable to all
-LCA methods.
+Although this paper primarily focuses on latent class analysis with continuous indicators,
+most of the suggested guidelines are equally applicable to ordinal indicators.
 
@@
 
-There are several use cases for which LCA methods are suitable. One
-such application is to test a theory that postulates the existence of a
-categorical latent variable. For instance, *identity status theory*
-posits that, at any given point in time, adolescents reside in one of
-four identity statuses [@marcia_development_1966; @marcia_identity_1980]. 
-LCA can be used to test whether these four statuses
-can be recovered based on observed indicators (e.g., self-reported
-identity exploration and commitment). If the results indicate that the data
-are better described by a different number of classes, or that the
-four-class solution does not correspond to the predicted pattern of
-responses on the indicators, then the theory may be called into
-question. A recent study by @maene_perceived_2022 applied this principle to 
+There are several use cases for which LCA methods are suitable. 
+Although latent class analysis is often discussed as an exploratory analysis technique,
+it can also be used in a confirmatory manner.
+Given that latent class analysis is similar to confirmatory factor analysis,
+but with a categorical latent variable, it can be used to similar ends:
+To assess whether a theoretical measurement model holds.
+This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
+For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
+LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
+One challenge is that, unlike confirmatory factor analysis,
+LCA does not provide absolute fit indices that can be used to perform inference.
+Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
+Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
+either qualitatively or quantitatively, using significance tests for pre-specified values.
+A recent study by @maene_perceived_2022 applied this principle to 
 test whether identities of Belgian adolescents with migration background can be 
 summarized as those relating to their heritage, national and regional identity.
-A LCA revealed three identity profiles: full integration (high belonging to 
-heritage, national and regional identity), national integration (high heritage
-and national identity with low regional identity), and weak separation
-(high heritage identity with neutral attitudes towards the nation and region).
 

### 2022-12-01 manuscript.Rmd
update 2022-12-01 08:01:21

@@
 There are several use cases for which LCA methods are suitable.
+
+### Testing theory
+
 Although latent class analysis is often discussed as an exploratory analysis technique,
@@
 
+### Unsupervised learning
+
 Another use case is unsupervised learning: when the goal is to restore
@@
 a fixed number of physical [@baughman_mixture_2006], mental health
-[@wu_abuse_2011] problems, or health outcomes after recovering from
-cancer [@hardy_exploring_2018]. The example of shoe size is a rudimentary
-illustration of this type of application. As another example, a recent study 
-identified two suicide risk profiles in military service members and veterans
-[@gromatsky_characteristics_2022].  One class was defined by higher-risk
-history and another by lower-risk history. The higher-risk group comprised of
-individuals with greater intent to die using methods with higher lethality,
-and the lower-risk group was characterized by greater ambivalence towards death
-and usage of lower lethality methods.
-
-A related use case is where class-membership is known but imperfectly measured.
-Mixture models can also be useful to evaluate the quality of measurement
-across instruments. For example, a recent study had employment status 
+[@wu_abuse_2011] problems. The shoe size example in Figure \@ref(fig:shoedens)
+is a rudimentary illustration of this use case.
+As another example, a recent study identified latent profiles of high versus low suicide risk in military service members and veterans,
+based on self-reported suicidal ideation and information about prior attempted methods and severity
+[@gromatsky_characteristics_2022].
+Applications like these can aid the clinical decision making process,
+and help decide whether additional support is warranted,
+or what type of intervention is appropriate.
+
+### Measurement model
+
+Another use case relies on LCA's similarity to confirmatory factor analysis.
+This approach is useful when class membership is measured imperfectly by several indicators.
+It can be used to partial out measurement error and restore most likely class membership,
+and to evaluate the reliability of different indicators.
+For example, a recent study had employment status 
 information collected across register and survey data. There was disagreement 
@@
 register data is free of error, this has been shown to be false.
-The authors used a mixture model to evaluate the underlying factor of 
+The authors used a mixture model to estimate the
 "error corrected employment status" by using two indicators of the same question
@@
 
-LCA can be used as a descriptive analysis where a researcher wishes to
-describe their sample and identify a few prototypes based on many
-variables. As an example, if a survey among all Dutch academics was
-carried out with the goal of bringing about a funding reform, LCA could
-be used to discover the types of publications that get funding.
-Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean.
-Prior to this study, ways to categorize ocean microplastics in terms of length,
-width and polymer type were rather arbitrary, and assumed that these properties
-were independent. Also, there was a lack of information regarding variability 
-within these categories. The authors used finite Gaussian mixture modeling to 
-build a classification system for ocean microplastic particles based on a set 
-of measured characteristics. With regards to particle dimensions, two classes
-emerged: one with smaller fragments whose width and height were highly 
-correlated, and another consisting of larger fragments with low correlation 
-between particle width and height.
+### Dimension reduction
+
+LCA can also be used to reduce the dimensionality of data,
+by reducing many variables to a few prototypes.
+This is similar to the use of factor analysis for dimension reduction,
+with the key distinction that factor analysis merely accounts for linear covariance between indicators, whereas LCA can accommodate more complex patterns.
+As this is a pragmatic application of LCA,
+the existence of a true categorical latent variable is questionable (cf. the section on Testing theory).
+Consequently, the results of this use case should be treated as descriptive and not reified (i.e., treating the found classes as evidence of the existence of a categorical latent variable) or straightforwardly generalized.
+A common use of LCA for dimension reduction is in longitudinal research:
+Often, one developmental process is summarized using latent class growth analysis (LCGA), and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
+For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy [@van_lissa_divergence_2015].
+These groups were then used as moderator of a latent growth analysis of adolescent- and parent-reported conflict.
+Results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than the other groups,
+and that the low empathy group had more conflict with parents than the other groups.
+Note that both these key findings are non-linear.
+Modeling them in a single-group model would be difficult;
+using LCGA greatly simplifies model specification and interpretation,
+albeit at a cost of some loss of information and, potentially, generalizability.
+<!-- Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean. -->
+<!-- Prior to this study, ways to categorize ocean microplastics in terms of length, -->
+<!-- width and polymer type were rather arbitrary, and assumed that these properties -->
+<!-- were independent. Also, there was a lack of information regarding variability  -->
+<!-- within these categories. The authors used finite Gaussian mixture modeling to  -->
+<!-- build a classification system for ocean microplastic particles based on a set  -->
+<!-- of measured characteristics. With regards to particle dimensions, two classes -->
+<!-- emerged: one with smaller fragments whose width and height were highly  -->
+<!-- correlated, and another consisting of larger fragments with low correlation  -->
+<!-- between particle width and height. -->
+
+### Class enumeration
 
@@
 From a collection of indicators, LCA provided evidence that there are four
-distinct risk profiles in the Winnipeg area. Another study investigated 
-adolescent empathy development and its influence on parent-child conflict 
-[@van_lissa_divergence_2015]. The data was collected over six years. 
-Using LCGA, adolescents with similar empathy development trajectories 
-were grouped into: high-, average-, and low-empathy classes.
+distinct risk profiles in the Winnipeg area.
 
@@
 <!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->
+
 # Best Practices
@@
 argued that some parametric methods are robust when scales with 7+
-indicators are treated as continuous [e.g., @norman_likert_2010] or 
-with 5+ categories [@rhemtulla_when_2012], this
-certainly does not imply that all methods are. It is certainly unlikely
-that such ordinal variables can be treated as a *mixture* of multiple
+indicators are treated as continuous [e.g., @norman_likert_2010],
+this certainly does not imply that all methods are.
+It is very unlikely
+that an ordinal variable can be modeled as a *mixture* of multiple
 normal distributions. The problem becomes egregious when the number of
@@
 population distribution is evidently bimodal. The Shapiro-Wilk normality
-test ($W = 0.971$, $p < 0.05$ ) rejects the null hypothesis that the
+test ($W `r report(shoesize_shapiro[["statistic"]])`$, $p `r report(shoesize_shapiro[["p.value"]])`$ ) rejects the null hypothesis that the
 sample comes from a normally distributed population. Yet, this is a

### 2022-12-01 manuscript.Rmd
update 2022-12-01 12:53:17

@@
 
-### Unsupervised learning
-
-Another use case is unsupervised learning: when the goal is to restore
-unobserved class membership based on observed indicators, or to classify
-individuals. For example, a mixture model can be utilized as a diagnostic
-aid when using several clinical indicators to distinguish between
-a fixed number of physical [@baughman_mixture_2006], mental health
-[@wu_abuse_2011] problems. The shoe size example in Figure \@ref(fig:shoedens)
-is a rudimentary illustration of this use case.
-As another example, a recent study identified latent profiles of high versus low suicide risk in military service members and veterans,
-based on self-reported suicidal ideation and information about prior attempted methods and severity
-[@gromatsky_characteristics_2022].
-Applications like these can aid the clinical decision making process,
-and help decide whether additional support is warranted,
-or what type of intervention is appropriate.
-
 ### Measurement model
@@
 This approach is useful when class membership is measured imperfectly by several indicators.
-It can be used to partial out measurement error and restore most likely class membership,
-and to evaluate the reliability of different indicators.
-For example, a recent study had employment status 
-information collected across register and survey data. There was disagreement 
-between the two data sources. While it might be tempting to assume that the
+This approach can be used to partial out measurement error and restore most likely class membership.
+For example, a recent study found descrepancies between self-reported employment status and official register data. 
+While it might be tempting to assume that the
 register data is free of error, this has been shown to be false.
@@
 
+This example also illustrates that LCA allows us to evaluate the reliability of different indicators [@geiser_is_2014].
+High quality indicators will be strongly related to the
+latent variable and will lead to good class separation.
+A continuous indicator that discriminates well between classes will show large differences in class-specific means;
+an ordinal indicator will show very high or very low conditional response probabilities for some classes but not for others.
+This information can be used to select the indicators that are most diagnostic of class membership.
+For an example of research on differential item functioning,
+consider a study on university admission tests of Saudi Arabian students [@tsaousis_measurement_2020].
+LCA revealed three
+latent classes: high-, average-, and low-scorers. Subsequently, the authors
+used the Multiple Indicator Multiple Causes model to identify both uniform
+and non-uniform differential item functioning. The main finding was that gender
+was a potential source of differential item functioning for latent class
+indicators. Therefore, gender should be included as a covariate when fitting 
+a LCA model in order to obtain unbiased estimates.
+
 ### Dimension reduction
@@
 by reducing many variables to a few prototypes.
-This is similar to the use of factor analysis for dimension reduction,
-with the key distinction that factor analysis merely accounts for linear covariance between indicators, whereas LCA can accommodate more complex patterns.
+This is similar to the use of factor analysis,
+with the key distinction that factor analysis accounts for linear covariance between indicators,
+whereas LCA can accommodate complex - but discrete - patterns.
 As this is a pragmatic application of LCA,
 the existence of a true categorical latent variable is questionable (cf. the section on Testing theory).
-Consequently, the results of this use case should be treated as descriptive and not reified (i.e., treating the found classes as evidence of the existence of a categorical latent variable) or straightforwardly generalized.
-A common use of LCA for dimension reduction is in longitudinal research:
-Often, one developmental process is summarized using latent class growth analysis (LCGA), and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
+Consequently, the results of this use case should be treated as descriptive and not "reified": i.e., the found classes should not be treated as evidence of the existence of a categorical latent variable.
+The use of LCA for dimension reduction is common in longitudinal research,
+where one developmental process is summarized using latent class growth analysis (LCGA),
+and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
 For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy [@van_lissa_divergence_2015].
@@
 and that the low empathy group had more conflict with parents than the other groups.
-Note that both these key findings are non-linear.
-Modeling them in a single-group model would be difficult;
-using LCGA greatly simplifies model specification and interpretation,
+Note that both these key findings are non-linear;
+modeling them in a single-group model would be difficult.
+LCGA greatly simplifies model specification and interpretation,
 albeit at a cost of some loss of information and, potentially, generalizability.
@@
 
+When using ordinal indicators,
+the need for dimension reduction can be especially pertinent.
+For example, one study measured
+fifteen dichotomous health indicators among injured military personnel [@macgregor_symptom_2021]. 
+Combinatorics informs us that fifteen
+dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
+However, LCA was able to reduce response patterns to five clinically meaningful latent classes.
+
 ### Class enumeration
 
-With LCA, our goal could be to inductively identify the number of classes. 
-For instance, if we believe that a variable represents a group,
-but do not know how many groups there are, LCA may be an appropriate
-technique to answer this question. For example, @hopfer_social_2014
-studied substance use, sexual behavior, and mental health status of
-urban population in Winnipeg, Canada. The underlying assumption was that
-there were different risk profiles, but their number was not known. 
-From a collection of indicators, LCA provided evidence that there are four
-distinct risk profiles in the Winnipeg area.
-
-Another application of LCA is to classify individuals. In a peer harassment 
-study, @giang_using_2008 used LCA to classify over 2,000 sixth grade students 
-into aggressor and victim classes. The five-class solution comprised of 
-classes of victims, aggressors, and socially adjusted students. 
-The study revealed that there were two types of victims: highly-victimized 
-aggressive-victims and highly-aggressive aggressive-victims.
-Another example is the "Well-being compass" study by van Lissa, currently 
-a preprint. The study was conducted in collaboration with a mental health care
-facility and focused on care providers. Latent class analysis has identified 
-four different risk profiles for care providers. Not only was class membership 
-estimated for the participants of the original study, but the class solution 
-was used as a risk assessment aid for new care providers, 
-so they could be more adequately supported by mental health care professionals.
+Another use case is to explore potential heterogeneity in a population,
+and to determine whether it is comprised of latent subpopulations.
+For example, researchers suspected the existence of different risk profiles of substance use, sexual behavior, and mental health outcomes, and used LCA to determine how many such risk profiles could be distinguished [@hopfer_social_2014].
+Another explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
+and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
+[@gromatsky_characteristics_2022].
+
+### Classification
+
+Another use case is to estimate individuals' unobserved class membership based on observed indicators.
+The shoe size example in Figure \@ref(fig:shoedens)
+is a rudimentary illustration of this application.
+In applied research, LCA is often used for classification in clinical contexts.
+For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough [@baughman_mixture_2006].
+Another study used LCA to not only classify the sample used to estimate the model, but also future cases [Zegwaard et al, in prep].
+The study identified four types of care providers among those who supported close kin with mental health problems.
+Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers, to provide more targeted support.
+To obtain predicted class membership based on an existing LCA model,
+the authors computed the likelihood of data from a new participant under the multivariate normal distributions estimated within all latent classes.
 <!-- DA: this study is not in zotero yet, I lack information (title, authors, etc.) to enter it, @Caspar could you enter this to Zotero and adjust the in-text citation-->
-
-LCA is also appropriate when we wish to identify indicators that capture
-classes well. High quality indicators will be strongly related to the
-latent variable and will lead to good class separation. This
-relationship of high quality indicators to the latent variable is
-reflected in very high or very low conditional response probabilities.
-For a simulation study exploring the effects of indicator quality on
-LCA, see @geiser_is_2014. Therefore, one could use conditional response
-probabilities for each item to assess its quality with regards to how
-well it helps separate the latent classes. From this, a theory about the
-selection of indicators could be informed.
-An example of such use of LCA is a study by @tsaousis_measurement_2020 who
-studied differential item functioning using LCA. The study focused 
-on university admission tests of Saudi Arabian students. LCA has revealed three
-latent classes: high-, average-, and low-scorers. Subsequently, the authors
-used the Multiple Indicator Multiple Causes model to identify both uniform
-and non-uniform differential item functioning. The main finding was that gender
-was a potential source of differential item functioning for latent class
-indicators. Therefore, gender should be included as a covariate when fitting 
-a LCA model in order to obtain unbiased estimates.
-
+These applications of LCA aid the clinical decision making process,
+and help decide whether additional support is warranted,
+or what type of intervention is appropriate.
+
+### 
 An extension of LCA is that containing covariates which can be used to
@@
 used as an "observed" categorical variable for prediction of depressive symptoms.
-
-In addition to these applications, LCA can be used for dimensionality
-reduction as the resulting groups summarize response patterns on a large
-number of indicators. For example, @macgregor_symptom_2021 investigated
-symptom profiles among injured U.S. military personnel. They used
-fifteen dichotomous items from the Post-Deployment Health Assessment
-survey as LCA indicators. Combinatorics informs us that fifteen
-dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
-Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a
-method of dimensionality reduction. A five class solution was found to
-have the best fit according to both statistical criteria and clinical
-interpretability. In the longitudinal adolescent empathy development study, 
-@van_lissa_divergence_2015 first investigated heterogeneity in empathy 
-development. This was done through identification of classes of individuals 
-characterized by similar developmental trajectories. The authors used LCGA
-as a dimensionality reduction technique where 467 individual trajectories
-were summarized into three empathy classes. Such reduction can help researchers 
-gain a better perspective of the grand patterns in the data.
 

### 2022-12-01 manuscript.Rmd
update 2022-12-01 15:26:44

@@
 
-### 
+### Violations of normality
+
+Finally, LCA can be used to deal with data which violate certain
+assumptions. As discussed in the shoe size example, LCA can deal with
+violations of normality. In fact, LCA assumes the population
+distribution is a non-normal mixture of $K$ normal distributions, and it
+can recover the value of $K$, i.e. generate a $K$-class solution.
+For instance, @alkema_maximizing_2022 studied ocean plastic debris. 
+Plastic particle width and height are non-normally distributed. The authors 
+demonstrate that the non-normal distributions of particle dimensions are
+due to the measurements being a mixture composed of several particle types, 
+each with its own normal distribution.
+
+<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical.
+We encourage researchers to be agnostic about the **true** nature of the latent variable.
+And focus on the approach that fits the research question.
+-->
+<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->
+
+### Controlling for covariates
+
 An extension of LCA is that containing covariates which can be used to
@@
 
+### Distal outcomes
+
 When our interest is the prediction of one or more outcomes, LCA can be used 
@@
 used as an "observed" categorical variable for prediction of depressive symptoms.
-
-Finally, LCA can be used to deal with data which violate certain
-assumptions. As discussed in the shoe size example, LCA can deal with
-violations of normality. In fact, LCA assumes the population
-distribution is a non-normal mixture of $K$ normal distributions, and it
-can recover the value of $K$, i.e. generate a $K$-class solution.
-For instance, @alkema_maximizing_2022 studied ocean plastic debris. 
-Plastic particle width and height are non-normally distributed. The authors 
-demonstrate that the non-normal distributions of particle dimensions are
-due to the measurements being a mixture composed of several particle types, 
-each with its own normal distribution.
-
-<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical.
-We encourage researchers to be agnostic about the **true** nature of the latent variable.
-And focus on the approach that fits the research question.
--->
-<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->
 
@@
 latent class analysis with ordinal indicators is more appropriate.
-
-Relatedly, a recent publication claimed that an assumption of mixture
-models is that observed indicators are normally distributed
-[@spurk_latent_2020]. This is incorrect. When the number of classes is
-greater than one, Gaussian mixture models assume that the observed indicators 
-are a mixture of multiple (multivariate) normal distributions. In our shoe
-size example, it can be seen that the population distribution is
-comprised of two normal distributions. When examined visually, the
-population distribution is evidently bimodal. The Shapiro-Wilk normality
-test ($W `r report(shoesize_shapiro[["statistic"]])`$, $p `r report(shoesize_shapiro[["p.value"]])`$ ) rejects the null hypothesis that the
-sample comes from a normally distributed population. Yet, this is a
-prototypical example of a mixed population distribution where LCA can
-discover latent groups. If the population distribution were instead
-normal, there would be no classes to extract as the whole population
-would belong to a single (homogeneous) class.
 
@@
 
+## Myths and misunderstandings
+
+Relatedly, a recent publication claimed that an assumption of mixture
+models is that observed indicators are normally distributed
+[@spurk_latent_2020]. This is incorrect. When the number of classes is
+greater than one, Gaussian mixture models assume that the observed indicators 
+are a mixture of multiple (multivariate) normal distributions. In our shoe
+size example, it can be seen that the population distribution is
+comprised of two normal distributions. When examined visually, the
+population distribution is evidently bimodal. The Shapiro-Wilk normality
+test ($W `r report(shoesize_shapiro[["statistic"]])`$, $p `r report(shoesize_shapiro[["p.value"]])`$ ) rejects the null hypothesis that the
+sample comes from a normally distributed population. Yet, this is a
+prototypical example of a mixed population distribution where LCA can
+discover latent groups. If the population distribution were instead
+normal, there would be no classes to extract as the whole population
+would belong to a single (homogeneous) class.
+
+<!-- Treating ordinal data as continuous -->
+<!-- People omitting the 1-class solution in class enumeration -->
+
 # Tutorial

### 2022-12-04 manuscript.Rmd
update 2022-12-04 10:03:24

@@
 groups based on a parametric model of observed indicators of group
-membership. The concept of LCA can be understood in different ways. A
-mixture model assumes that the study population is composed of $K$
+membership. The concept of LCA can be understood in different ways.
+A mixture model assumes that the study population is composed of $K$
 subpopulations or classes.
@@
 
-Another way to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling.
+Another way to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling [@molenaarArbitraryNatureLatent1994].
 A mixture model is like confirmatory factor analysis,
@@
 group observed *variables* into latent constructs, and LCA groups
-*individuals* into classes. While factor analysis seeks to describe the
-relations between variables, LCA seeks to describe each unobserved group
-and the differences between them. In line with this distinction, LCA
+*individuals* into classes.
+While factor analysis seeks to describe the relations between variables,
+LCA seeks to describe unobserved groups and classify individual observations to those groups.
+In line with this distinction, LCA
 (mixture models) is referred to as a "person-centered" technique, and
@@
 Another use case is to explore potential heterogeneity in a population,
-and to determine whether it is comprised of latent subpopulations.
+and to determine whether the population is comprised of latent subpopulations, and if so, how many.
+The primary focus in this use case is on class enumeration.
 For example, researchers suspected the existence of different risk profiles of substance use, sexual behavior, and mental health outcomes, and used LCA to determine how many such risk profiles could be distinguished [@hopfer_social_2014].
@@
 
-### Controlling for covariates
-
-An extension of LCA is that containing covariates which can be used to
-predict class membership. In this approach, we not only model the latent
-class variable based on indicators, but we also relate the class
-membership to other explanatory variables (@vermunt_latent_2017). An
-example of using covariates comes from @nozadi_moderating_2016 who
-applied LCA to identify the probability of children's membership to an
-anxiety class. The authors tested several covariates including
-children's age, sex, and accuracy scores. Age and sex were not found to
-be related to the children's latent class membership, hence these
-covariates were excluded from the analysis. Accuracy scores, however,
-were related to probabilities of being in anxiety and attention-anxiety
-classes. Therefore this covariate was kept as a valuable predictor.
-Another example comes from @van_lissa_divergence_2015 who included sex 
-as a predictor of the developmental trajectories and class membership in 
-the previously discussed adolescent empathy development study. The rationale
-was that adding sex as a covariate would account for known sex differences 
-in empathy development, and lead to a model with lower bias.
-
-### Distal outcomes
-
-When our interest is the prediction of one or more outcomes, LCA can be used 
-to construct latent classes as categorical predictors. @lanza_latent_2013 
-demonstrated how LCA can be used to classify adolescents into depression 
-classes, and subsequently these classes can be used to predict smoking, grades,
-and delinquency. The study showed that the outcomes predicted by class 
-membership can be binary (regular smoking), continuous (grades) or 
-count (delinquency). In the previously discussed study of Belgian adolescent
-identities, @maene_perceived_2022 first constructed the latent variable
-representing different multiple identification strategies. Subsequently, 
-the most likely class membership based on posterior class probability was
-used as an "observed" categorical variable for prediction of depressive symptoms.
+<!-- ### Controlling for covariates -->
+<!-- CJ: I'm not sure we should call this a use case... in principle, -->
+<!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->
+<!-- An extension of LCA is that containing covariates which can be used to -->
+<!-- predict class membership. In this approach, we not only model the latent -->
+<!-- class variable based on indicators, but we also relate the class -->
+<!-- membership to other explanatory variables (@vermunt_latent_2017). An -->
+<!-- example of using covariates comes from @nozadi_moderating_2016 who -->
+<!-- applied LCA to identify the probability of children's membership to an -->
+<!-- anxiety class. The authors tested several covariates including -->
+<!-- children's age, sex, and accuracy scores. Age and sex were not found to -->
+<!-- be related to the children's latent class membership, hence these -->
+<!-- covariates were excluded from the analysis. Accuracy scores, however, -->
+<!-- were related to probabilities of being in anxiety and attention-anxiety -->
+<!-- classes. Therefore this covariate was kept as a valuable predictor. -->
+<!-- Another example comes from @van_lissa_divergence_2015 who included sex  -->
+<!-- as a predictor of the developmental trajectories and class membership in  -->
+<!-- the previously discussed adolescent empathy development study. The rationale -->
+<!-- was that adding sex as a covariate would account for known sex differences  -->
+<!-- in empathy development, and lead to a model with lower bias. -->
+
+<!-- ### Distal outcomes -->
+<!-- CJ: I'm not sure we should call this a use case... in principle, -->
+<!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->
+<!-- When our interest is the prediction of one or more outcomes, LCA can be used  -->
+<!-- to construct latent classes as categorical predictors. @lanza_latent_2013  -->
+<!-- demonstrated how LCA can be used to classify adolescents into depression  -->
+<!-- classes, and subsequently these classes can be used to predict smoking, grades, -->
+<!-- and delinquency. The study showed that the outcomes predicted by class  -->
+<!-- membership can be binary (regular smoking), continuous (grades) or  -->
+<!-- count (delinquency). In the previously discussed study of Belgian adolescent -->
+<!-- identities, @maene_perceived_2022 first constructed the latent variable -->
+<!-- representing different multiple identification strategies. Subsequently,  -->
+<!-- the most likely class membership based on posterior class probability was -->
+<!-- used as an "observed" categorical variable for prediction of depressive symptoms. -->
 
 # Best Practices
-
-## Best Practices in Estimation
 
@@
 \@ref(tab:checkest), are rooted in existing recommendations for best
-practices for estimating specific subtypes of LCA, including latent
+practices for estimating specific subtypes of LCA,
+including latent
 class growth analysis [@van_de_schoot_grolts-checklist_2017] and latent
@@
 to current best practices, as explained below.
+In order to aid researchers working with latent trajectory models,
+@van_de_schoot_grolts-checklist_2017 developed a protocol called
+Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
+
+## Best Practices in Estimation
 
@@
 prior to data collection. Special attention should be paid to level of
-measurement of the indicators. Finite Gaussian mixture models (including
-LPA) are only suitable for continuous variables. Indicators with an
+measurement of the indicators.
+Finite Gaussian mixture models (including
+LPA) are only suitable for continuous variables.
+Indicators with an
 ordinal level of measurement are likely to violate the assumption of
 within-class normal distributions of mixture models [see
-@vermunt_k-means_2011]. Personal experience consulting on LCA methods
+@vermunt_k-means_2011].
+Personal experience consulting on LCA methods
 and moderating the `tidyLPA` Google group suggest that the
@@
 argued that some parametric methods are robust when scales with 7+
-indicators are treated as continuous [e.g., @norman_likert_2010],
-this certainly does not imply that all methods are.
-It is very unlikely
+response categories are treated as continuous [e.g., @norman_likert_2010],
+it is very unlikely
 that an ordinal variable can be modeled as a *mixture* of multiple
-normal distributions. The problem becomes egregious when the number of
-classes estimated equals or exceeds the number of categories; in this
-case, each class-specific mean could describe a single response
-category, and a class-specific variance component would be nonsensical.
+normal distributions.
+To illustrate the problem, consider what happens when estimating a 5-class solution on a 5-point Likert scale.
+Each class-specific mean could describe a single response
+category,
+and a class-specific variance component would be nonsensical.
 In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
@@
 missing data and reporting how missingness was handled
-[@van_de_schoot_grolts-checklist_2017]. Three types of missingness have
+[@van_de_schoot_grolts-checklist_2017].
+While we concur that investigating missingness is due diligence,
+it is important to emphasize that missingness is adequately handled by default in `OpenMx`, which is the computational backend of `tidySEM`.
+Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
+information without imputing missing values [@endersRelativePerformanceFull2001].
+FIML is a best-practice solution for handling missing data;
+on par with multiple imputation when the
+estimation model includes the same variables as the imputation model
+[@lee_comparison_2021].
+Although FIML does not, by
+default, handle missingness in exogenous variables - all indicator
+variables in LCA are endogenous, so this is not a concern for LCA.
+Multiple imputation, a common alternative to FIML,
+is less suitable to LCA for two reasons. First,
+because LCA methods are often computationally expensive,
+and conducting them on multiple imputed datasets may be unfeasible.
+Second, because many multiple imputation methods have some assumption of normality, which is likely to be violated when it is assumed that the data come from a heterogenous population.
+
+Three types of missingness have
 been distinguished in the literature [@rubin_inference_1976; 
-@enders_applied_2022]: Missing completely at random (MCAR), which means that 
-missingness is random and unrelated to any observed or unobserved variables;
-missing at random (MAR), which means that missingness is contingent on
-the *observed* data (and can thus be accounted for); and finally missing
-not at random (MNAR), which means that missingness is related to
-unobserved factors. It is possible to conduct a so-called "MCAR" test,
-for example the non-parametric MCAR test [@jamshidian_tests_2010]. But
-note that the name "MCAR test" is somewhat misleading, as the
-null-hypothesis of this test is that the data are not MAR, and a
-significant test statistic indicates that missingness is related to the
-observed data (MAR). A non-significant test statistic does not
-distinguish between MCAR or MNAR. As Little's classic MCAR test relies
-on the comparison of variances across groups with different patterns of
-missing data, it assumes normality [@little_test_1988]. This assumption
-is tenuous in the context of LCA. A non-parametric MCAR test, as
-provided by Jamshidian and Jalal, may be more suitable
-[@jamshidian_tests_2010]. Unfortunately, this test was removed from the
-central R-repository CRAN due to lack of maintenance. For this tutorial,
-I have re-implemented it in the `mice` package as `mice::mcar()`, with a
+@enders_applied_2022]: random missingness (MCAR);
+missingness contingent on *observed* variables that can thus be known and controlled for (MAR);
+and missingness related to unobserved variables, which can neither be known nor controlled for (MNAR).
+A so-called "MCAR" test is often reported in relation to missingness
+[@jamshidian_tests_2010].
+But note that this name is somewhat misleading,
+as a significant test indicates that missingness is MAR,
+but a non-significant test statistic does not distinguish between MCAR or MNAR.
+As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@little_test_1988].
+This assumption is tenuous in the context of LCA.
+A non-parametric MCAR test may be more suitable [@jamshidian_tests_2010].
+For this tutorial,
+the lead author has implemented this test in the `mice` package as `mice::mcar()`, with a
 fast backend in C++ and new printing and plotting methods.
-
-While we concur that investigating missingness is due diligence, it is
-important to emphasize that missingness is adequately handled by default
-in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` 
-which is the backend of `tidySEM`. These packages use Full Information Maximum
-Likelihood (FIML) estimation, which makes use of all available
-information without imputing missing values. FIML is a best-practice
-solution for handling missing data; on par with multiple imputation when the 
-estimation model includes the same variables as the imputation model
-[@lee_comparison_2021]. FIML estimation assumes that missingness is
-either MCAR or MAR. Thus, one would typically proceed with FIML,
-regardless of the outcome of an MCAR test. Although FIML does not, by
-default, handle missingness in exogenous variables -- all indicator
-variables in LCA are endogenous, so this is not a concern.
-<!-- are you sure Mplus accounts for missing data automatically when the indicators are categorical? I know that is not the case with CFA/SEM models -->
-<!-- CJ: yes, it does -->
-
-Multiple imputation is less suitable to LCA for two reasons. First,
-because LCA methods are often computationally expensive, and conducting
-them on multiple imputed datasets may be unfeasible. Second, because
-there is no straightforward way to integrate LCA results across multiple
-datasets. To conclude; our recommendation is to inspect missingness
-(e.g., using `mice::MCAR()`) and report the proportion of missingness
-per variable (e.g., using `tidySEM::descriptives()`), before proceeding
-with FIML. One minor concern is that the K-means algorithm, which
-`tidySEM` uses for determining starting values, is *not* robust to
-missing values. When it fails, `tidySEM` automatically switches to
-hierarchical clustering, unless the user specifies a different
-clustering algorithm or uses manual starting values.
-
-### Alternative Model Specifications
-
-In order to aid researchers working with latent trajectory models,
-@van_de_schoot_grolts-checklist_2017 developed a protocol called
-Guidelines for Reporting on Latent Trajectory Studies (GRoLTS). Two of
-the GRoLTS guidelines (namely, 6a and 6b) refer to considering
-alternative model specifications. Both are discussing specific cases in
-latent trajectory model specification. The first is about whether the
-variance of the growth parameter is estimated freely or fixed, and the
-second is about whether conditional independence is assumed, as the
-researcher might also want to free-up the variance-covariance structure.
-In LCA, there are also many different ways to specify the model. Means,
-variances, and covariances between the indicators can either be
-constrained across classes, or estimated freely. Researchers using LCA
-should transparently report their chosen parametrization as well as
-discuss different parametrizations that were tested in the process.
-
-Different types of latent class models have different parameters. For
-example, mixture models and latent profile analyses typically have
-class-specific means, variances, and covariances. Latent growth analyses
-have the same parameters, but with respect to the latent growth
-variables. Latent class analyses with ordinal indicators have
-thresholds. All of these parameters can be freely estimated, constrained
-to be equal across classes, or fixed to a certain value (e.g., to zero).
-The total number of parameters thus scales with the number of estimated
-classes. Consequently, LCA methods have a potentially very high number
-of parameters. As any of these parameters could be misspecified, it is
-important to consider alternative model specifications. However,
-alternative model specifications may be approached differently depending
-on whether an analysis is data driven (exploratory), or theory driven
-(confirmatory). This distinction has remained underemphasized in prior
-writing.
-
-Prior literature on LCA has emphasized its exploratory applications
-[@nylund_deciding_2007]. In exploratory LCA, a large number of models
-are typically estimated in batch, with varying numbers of classes and
-model specifications. The "correct" model specification is then
-determined based on a combination of fit indices, significance tests,
-and interpretability. For latent profile analysis, the function
-`tidySEM::mx_profiles(classes, variances, covariances)` largely
-automates this process. The argument `classes` indicates which class
-solutions should be estimated (e.g., 1 through 6). The argument
-`variances` specifies whether variances should be `"equal"` or
-`"varying"` across classes. The argument `covariances` specifies whether
-covariances should be constrained to `"zero"`, `"equal"` or `"varying"`
-across classes. The means are free to vary across classes by default,
-although the more general function `tidySEM::mx_mixture()` could be used
-to circumvent this. After all models have been estimated, the function
-`tidySEM::table_fit()` can be used to obtain a model fit table suitable
-for determining the optimal model according to best practices. Note,
-however, that this table does not include the bootstrapped likelihood
-ratio test (BLRT) by default, because this test is very computationally
-expensive. It is recommended to use the function `tidySEM::BLRT()` to
-compare a shortlist of likely candidate models based on other fit
-indices. More on fit indices can be found in the Model Fit Indices
-subsection of this paper.
-
-Confirmatory LCA typically requires less comprehensive alternative model
-specifications. For example, in the context of a preregistered analysis,
-the main models of interest would have been specified a priori. Even in
-this case, the theoretical model could be compared to a few others to
-contextualize it.
+Our recommendation is to inspect missingness and report the proportion of missingness per variable using `tidySEM::descriptives()`,
+and testing for MAR using `mice::mcar()`,
+before proceeding with FIML.
+
+One minor concern is that the K-means algorithm, which
+`tidySEM` uses to determine starting values, is *not* robust to
+partially missing data [@hartiganAlgorithm136KMeans1979].
+When missing data causes it to fail,
+`tidySEM` automatically switches to hierarchical clustering [@murtaghWardHierarchicalAgglomerative2014].
+The user can also specify a different
+clustering algorithm or specify starting values manually.
+Cases with complete missing data should be removed,
+or their data should be imputed another way.
+
+### Model Specification
+
+Before estimating a LCA model, researchers should carefully specify the model,
+based on theoretical and empirical considerations.
+This will ensure that the model accurately represents the data and provides reliable results.
+Two factors play a role in LCA model specification:
+the number of latent classes and the parametrization within each class.
+The total number of parameters of a latent class model consists of the number of parameters within each class times the number of classes,
+plus a classification vector with $K-1$ free elements.
+Different types of latent class models have different parameters.
+For example, latent profile analyses typically assume class-specific means,
+equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`.
+LCA with ordinal indicators typically assume free thresholds across classes,
+where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`.
+Like latent profile analyses, latent class growth analyses typically have free means across classes, but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`.
+In the general sense, any structural equation model can be specified,
+either using `lavaan` model syntax or using `OpenMx` RAM specification,
+and supplied to the general function `mx_mixture()`.
+The resulting model may have means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables.
+All of these parameters can be freely estimated,
+constrained to be equal across classes using equality constraints (e.g., `x ~~ a*x` fixes the variance of `x` to be equal across classes),
+or fixed to a certain value (e.g., `x ~.5*1` fixes the mean of `x` to `.5`).
+
+As any of these parameters could be misspecified,
+prior literature has emphasized the importance of considering alternative model specifications [@van_de_schoot_grolts-checklist_2017].
+While we agree with this recommendation,
+two considerations have remained underemphasized in prior literature.
+One consideration is that alternative model specifications may be approached differently depending on whether the LCA is data driven (exploratory), or theory driven
+(confirmatory).
+The section on Use Cases illustrates these different applications.
+
+Prior literature has focused on exploratory LCA.
+In this context, model specification typically consists of an exhaustive grid search along varying numbers of classes and model specifications.
+The "correct" model specification is then determined based on a combination of fit indices, significance tests, and interpretability [@nylund_deciding_2007].
+Confirmatory LCA does not require such an extensive search over the model space.
+Theoretical models can be compared to a smaller set.
+If hypothesized models have been specified a priori,
+and possibly even preregistered,
+it makes sense to focus on the estimation and interpretation of these models, rather than proceed in a purely data-driven manner.
+One key challenge is that LCA lacks objective fit indices,
+and the fit of a theoretical model can thus only be determined relative to other likely candidates.
+For this reason, the theoretical model should still be compared to a few others.
+A one-class model is a sensible benchmark, to test whether LCA is even appropriate.
+A theoretical model would be supported by a substantial (and significant) difference in fit with a one-class model.
+Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
+Here again, a substantial difference in fit supports the theoretical model - but note that small differences in fit do not give cause to reject a theoretical model.
+The theoretical model may still be preferred, for example, to ensure comparability with prior work.
+The fact that the theoretical model received little differential support should of course be discussed,
+along with potential implications for future work.
+The BIC is the preferred fit index for comparing
+
+<!-- HIER -->
+<!-- For latent profile analysis, the function -->
+<!-- `tidySEM::mx_profiles(classes, variances, covariances)` largely -->
+<!-- automates this process. The argument `classes` indicates which class -->
+<!-- solutions should be estimated (e.g., 1 through 6). The argument -->
+<!-- `variances` specifies whether variances should be `"equal"` or -->
+<!-- `"varying"` across classes. The argument `covariances` specifies whether -->
+<!-- covariances should be constrained to `"zero"`, `"equal"` or `"varying"` -->
+<!-- across classes. The means are free to vary across classes by default, -->
+<!-- although the more general function `tidySEM::mx_mixture()` could be used -->
+<!-- to circumvent this. After all models have been estimated, the function -->
+<!-- `tidySEM::table_fit()` can be used to obtain a model fit table suitable -->
+<!-- for determining the optimal model according to best practices. Note, -->
+<!-- however, that this table does not include the bootstrapped likelihood -->
+<!-- ratio test (BLRT) by default, because this test is very computationally -->
+<!-- expensive. It is recommended to use the function `tidySEM::BLRT()` to -->
+<!-- compare a shortlist of likely candidate models based on other fit -->
+<!-- indices. More on fit indices can be found in the Model Fit Indices -->
+<!-- subsection of this paper. -->
+
+The second consideration is the risk of overfitting.
+We address each in turn.
+
+As explained before, the total number of parameters scales with the number of estimated classes.
+Consequently, LCA methods have a potentially very high number of parameters.
 
@@
 standardize ways of reporting LCA [e.g. @masyn_latent_2013;
-@weller_latent_2020], but more work is needed. @van_lissa_worcs_2020
-developed WORCS, a workflow for open reproducible code in science. WORCS
-consists of step-by-step guidelines for research projects based on the
-TOP-guidelines developed by @nosek_promoting_2015. WORCS workflow can be
-easily implemented in R in form of an R package which facilitates
-preregistration, article drafting, version control, citation and
-formatting, among others [@van_lissa_worcs_2020].
+@weller_latent_2020], but more work is needed. 
+<!-- This does not belong here: -->
+<!-- @van_lissa_worcs_2020 -->
+<!-- developed WORCS, a workflow for open reproducible code in science. WORCS -->
+<!-- consists of step-by-step guidelines for research projects based on the -->
+<!-- TOP-guidelines developed by @nosek_promoting_2015. WORCS workflow can be -->
+<!-- easily implemented in R in form of an R package which facilitates -->
+<!-- preregistration, article drafting, version control, citation and -->
+<!-- formatting, among others [@van_lissa_worcs_2020]. -->
 
@@
 sharing wherever possible [@nosek_promoting_2015]. @van_lissa_worcs_2020
-suggest sharing synthetic data in case the original data cannot be
-shared, and provide functions to generate such synthetic data. Ideally,
+If the original data cannot be shared, sharing synthetic data is a good alternative.
+The function `worcs::synthetic()` provides a generic, non-parametric method to create a synthetic copy of data.
+When synthesizing data for LCA, it is also possible to synthesize data directly from the model.
+LCA models constructed in `tidySEM` are estimated using `OpenMx`,
+and as such, can benefit from all existing functions in that package.
+To synthesize data from a mixture model `mix`, use the function `mxGenerateData(mix)`.
+
 the entire research project is made reproducible so that others may
@@
 ```{r, include = TRUE, eval=F}
-library(tidySEM) # loading tidySEM
-df <- data_mix_ordinal  # dataset
-df[1:4] <- lapply(df, ordered) # indicators as ordered factors
+# Load required packages
+library(tidySEM) 
+# Load data
+df <- data_mix_ordinal
 ```
@@
 ```{r, include = TRUE, eval=F}
-tidySEM::descriptives(df) # descriptives
+tidySEM::descriptives(df)
 ```
@@
 Before we fit a series of LCA models, we set a random seed using
-`set.seed()`. The seed number can be any digit. This is an important
+`set.seed()`. This is an important
 step as there is some inherent randomness in the LCA computations, and
@@
 
-Finally, we reach the step of fitting LCA models. To do so, we use
-`tidySEM::mx_lca()` which takes data and number of classes as inputs. 
-Notice that the analysis will be done on all the variable in the data set, 
-so you have to make sure it only includes the relevant variables. 
-In our example, we want to fit 1 to 4 class solutions and compare 
-their output. Depending on your computer's computational power, 
-this might take a while.
+Finally, we reach the step of fitting LCA models.
+As all variables are ordinal, we can use the convenience function
+`tidySEM::mx_lca()`, which is a wrapper for the generic function `mx_mixture()`,
+optimized for latent class analysis with ordinal data.
+Any mixture model can be specified through `mx_mixture()`,
+but at the time of writing, there are two more wrapper functions for specialized models:
+`mx_profiles()`, for latent profile analysis, and `mx_growth_mixture()`, for latent growth analysis and growth mixture models.
+All of these functions have arguments `data` and number of `classes`.
+All variables in `data` are included in the analysis,
+so relevant variables must be selected first.
+Below, we fit 1 to 4 class solutions.
+<!-- Depending on your computer's computational power, -->
+<!-- this might take a while. -->
+<!-- As `tidySEM::mx_lca()` run several models at a time, the resulting object `res`  -->
+<!-- is a `mixture_list` class where each $K$ class solution is a list element, or -->
+<!-- more specifically an `OpenMx` model (`MxModel`). -->
 
@@
 
-Notice that the model run 10 times for each $K$ class solution. 
-This way we can choose the best fitting model across all the solutions. 
-In this example, we see that all models converged without issues.
-
-We can inspect the class of resulting objects.
-
-```{r, include = TRUE, eval=F}
+<!-- Notice that the model run 10 times for each $K$ class solution. -->
+<!-- This way we can choose the best fitting model across all the solutions. -->
+<!-- In this example, we see that all models converged without issues.  -->
+<!-- We can inspect the class of resulting objects. -->
+
+```{r, include = TRUE, eval=F, echo = F}
 is(res)
@@
 ```
-
-As `tidySEM::mx_lca()` run several models at a time, the resulting object `res` 
-is a `mixture_list` class where each $K$ class solution is a list element, or
-more specifically an `OpenMx` model (`MxModel`).
 
@@
 Description <- c(
-    "the $K$-class solution",
-    "the -2*log-likelihood of each model",
-    "Number of estimated parameters in each model",
-    "the Akaike Information Criterion value",
-    "the Bayesian Information Criterion value",
-    "the lowest posterior class probability by most likely class membership",
-    "the highest posterior class probability by most likely class membership",
-    "the lowest class proportion based on the posterior class probabilities",
-    "the highest class proportion based on the posterior class probabilities")
+    "Number of classes $K$",
+    "Basic fit measure of the natural log of the likelihood of the data",
+    "Total number of model parameters (in all classes)",
+    "Relative fit measure: Akaike Information Criterion",
+    "Relative fit measure: Bayesian Information Criterion",
+    "Lowest average posterior class probability for class cases were assigned to",
+    "Highest average posterior class probability for class cases were assigned to",
+    "Proportion of cases in the smallest class based on posterior class probability",
+    "Proportion of cases in the largest class based on posterior class probability")
 

### 2022-12-16 manuscript.Rmd
Changes

@@
   in data based on a parametric model.
-  Its examples include mixture models, latent class analysis with ordinal
+  Examples include mixture models, latent class analysis with ordinal
   indicators, and latent class growth analysis.
@@
 this paper exclusively uses free, open source software for latent class analysis in R.
-To this end, new functionality was developed in the R-package `tidySEM` [REF tidysem].
+To this end, new functionality was developed in the R-package `tidySEM` [@vanlissaTidySEMTidyStructural2022].
 
@@
 groups based on a parametric model of observed indicators of group
-membership. The concept of LCA can be understood in different ways.
-A mixture model assumes that the study population is composed of $K$
-subpopulations or classes.
+membership [@vermuntLatentClassAnalysis2004].
+This method has become quite popular across scientific fields,
+under a number of different names;
+most notably (finite Gaussian) mixture modeling and latent profile analysis.
+The concept of LCA can be understood in different ways.
+Generally speaking, LCA assumes that the study population is composed of $K$
+subpopulations.
 It further assumes that the observed data are
@@
 two parameters: the mean and the variance.
-Latent class analysis aims to estimate the values of those parameters across a known number of classes $k$,
-as well as the probability of belonging to classes $1 \ldots k$ for every individual $i$.
+LCA aims to estimate the values of those parameters across a known number of classes $K$,
+as well as the probability of every individual $i$ belonging to classes $1 \ldots k$.
 Commonly, the same model is estimated across all classes,
 but with different parameters for each class (i.e., class-specific means and variances).
-From this perspective, reporting the mean and variance of a variable is a special case of a mixture model with $k = 1$ class.
 
@@
 
-Another way to conceptualize latent class analysis is by analogy to a measurement model in structural equation modeling [@molenaarArbitraryNatureLatent1994].
+Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling [@molenaarArbitraryNatureLatent1994].
 A mixture model is like confirmatory factor analysis,
-except that the continuous latent variable is substituted with a categorical latent variable.
+except with a categorical instead of a continuous latent variable.
 One difference between
@@
 or individuals can be assigned to classes based on their highest class probability.
-If the classes are clearly distinct, measurement error is low, and the latter approach is probably acceptable.
+If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
 This perspective on mixture modeling is sometimes referred to as "model-based
@@
 The term latent class analysis originally referred to cases where the
-observed indicators were ordinal [@collins_latent_2009]. 
+observed indicators were ordinal (ordered categorical) [@collins_latent_2009]. 
 Nowadays, it is more commonly used
@@
 as *latent class analysis with ordinal indicators*.
-Latent class models with ordinal indicators are parametrized differently from mixture models.
-One common parametrization assumes that each categorical variable reflects an underlying standard normal distribution.
-The parameters are "thresholds" that correspond to quantiles of a standard
-normal distribution (with $N(\mu = 0, \sigma = 1)$). These thresholds
-are estimated based on the proportion of individuals in each of the
-response categories of the indicator variable. For example, a binary
-indicator has a single threshold that distinguishes the two response
-categories. If responses are distributed 50/50, then the corresponding
-threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
-then the resulting threshold would be $t_1 = 0.25$.
-Although this paper primarily focuses on latent class analysis with continuous indicators,
-most of the suggested guidelines are equally applicable to ordinal indicators.
+
+<!-- Although this paper primarily focuses on latent class analysis with continuous indicators, -->
+<!-- most of the suggested guidelines are equally applicable to ordinal indicators. -->
 
@@
 LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
-One challenge is that, unlike confirmatory factor analysis,
-LCA does not provide absolute fit indices that can be used to perform inference.
+Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA [@donnellanResilientOvercontrolledUndercontrolled2010].
+One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices.
 Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
@@
 variance of categorical variables, and missingness; see next subsection)
-can be obtained using the function `tidySEM::descriptives(data)`. Note,
+can be obtained using `tidySEM::descriptives(data)`. Note,
 however, that sample-level descriptive statistics are of limited value
@@
 
+### Data preprocessing
+
+LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
+Numeric variables should have an interval measurement level,
+such that it is sensible to estimate means and (co)variances.
+Note that this assumption is likely violated for Likert-type scales.
+For the remaining three measurement levels,
+some data preprocessing is required.
+The exact data types accepted by `tidySEM` at the time of writing are `numeric` and `ordered`.
+To correctly specify ordered variables,
+such as Likert scales
+use the function `OpenMx::mxFactor()`.
+Binary variables,
+such as presence or absence of some symptom, are also treated as ordinal.
+Finally, nominal variables are converted to binary indicators of group membership via dummy coding, for example, using `mx_dummies()`.
+
+A second consideration in data preprocessing is that models with indicators on different scales may present with convergence issues.
+The reason for this is that the parameter space is high-dimensional with many potential local optima.
+If one indicator is on a much larger scale than another, the parameter space may be more extensively explored for that parameter.
+The problem can be resolved by transforming the indicators to place them on (roughly) similar scales.
+The most common transformation is standardization, which involves subtracting the mean of each variable from each value and dividing by the standard deviation.
+In R, this transformation is applied using the `scale()` function.
+Standardization retains all available information,
+and model parameters can be transformed back to the original scale of the indicators by reversing the transformation.
+The transformed variable has a mean of zero and a standard deviation of one.
+Note that standardization is typically not necessary if indicators are already on the same, or similar, scales.
+
 ### Handling Missing Data
@@
 FIML is a best-practice solution for handling missing data;
-on par with multiple imputation when the
-estimation model includes the same variables as the imputation model
-[@lee_comparison_2021].
-Although FIML does not, by
-default, handle missingness in exogenous variables - all indicator
-variables in LCA are endogenous, so this is not a concern for LCA.
-Multiple imputation, a common alternative to FIML,
-is less suitable to LCA for two reasons. First,
+at least on par with multiple imputation [@lee_comparison_2021].
+Multiple imputation is less suitable to LCA for two reasons. First,
 because LCA methods are often computationally expensive,
@@
 Second, because many multiple imputation methods have some assumption of normality, which is likely to be violated when it is assumed that the data come from a heterogenous population.
+FIML is not suitable for cases with complete missing data.
 
@@
 For this tutorial,
-the lead author has implemented this test in the `mice` package as `mice::mcar()`, with a
-fast backend in C++ and new printing and plotting methods.
-Our recommendation is to inspect missingness and report the proportion of missingness per variable using `tidySEM::descriptives()`,
-and testing for MAR using `mice::mcar()`,
-before proceeding with FIML.
-
-One minor concern is that the K-means algorithm, which
-`tidySEM` uses to determine starting values, is *not* robust to
-partially missing data [@hartiganAlgorithm136KMeans1979].
-When missing data causes it to fail,
-`tidySEM` automatically switches to hierarchical clustering [@murtaghWardHierarchicalAgglomerative2014].
-The user can also specify a different
-clustering algorithm or specify starting values manually.
-Cases with complete missing data should be removed,
-or their data should be imputed another way.
+the lead author has implemented this test in the `mice` package as `mice::mcar()`.
+
+To conclude; our recommendation is to inspect and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
+and to test for MAR using `mice::mcar()`.
+As FIML estimation assumes that missingness is either MCAR or MAR,
+one would proceed with FIML regardless of the outcome of the test.
+One minor concern is that the approach `tidySEM` uses to determine starting values is *not* robust to
+partially missing data.
+This should not be a problem, however, as uses a global optimizer to estimate LCA models,
+which should be independent of starting values [@coranaMinimizingMultimodalFunctions1987].
 
@@
 
-Before estimating a LCA model, researchers should carefully specify the model,
-based on theoretical and empirical considerations.
-This will ensure that the model accurately represents the data and provides reliable results.
-Two factors play a role in LCA model specification:
-the number of latent classes and the parametrization within each class.
-The total number of parameters of a latent class model consists of the number of parameters within each class times the number of classes,
-plus a classification vector with $K-1$ free elements.
-Different types of latent class models have different parameters.
-For example, latent profile analyses typically assume class-specific means,
-equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`.
-LCA with ordinal indicators typically assume free thresholds across classes,
-where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`.
-Like latent profile analyses, latent class growth analyses typically have free means across classes, but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`.
-In the general sense, any structural equation model can be specified,
-either using `lavaan` model syntax or using `OpenMx` RAM specification,
-and supplied to the general function `mx_mixture()`.
-The resulting model may have means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables.
-All of these parameters can be freely estimated,
-constrained to be equal across classes using equality constraints (e.g., `x ~~ a*x` fixes the variance of `x` to be equal across classes),
-or fixed to a certain value (e.g., `x ~.5*1` fixes the mean of `x` to `.5`).
-
-As any of these parameters could be misspecified,
-prior literature has emphasized the importance of considering alternative model specifications [@van_de_schoot_grolts-checklist_2017].
-While we agree with this recommendation,
-two considerations have remained underemphasized in prior literature.
-One consideration is that alternative model specifications may be approached differently depending on whether the LCA is data driven (exploratory), or theory driven
-(confirmatory).
-The section on Use Cases illustrates these different applications.
-
-Prior literature has focused on exploratory LCA.
-In this context, model specification typically consists of an exhaustive grid search along varying numbers of classes and model specifications.
-The "correct" model specification is then determined based on a combination of fit indices, significance tests, and interpretability [@nylund_deciding_2007].
+Any LCA model should be carefully specified based on theoretical and empirical considerations.
+The parameters of LCA depend on the measurement level of the indicators.
+LCA with continuous indicators can be parametrized in terms of means, variances, and covariances.
+Latent class growth analyses have the same parameters,
+but with respect to the latent growth variables.
+LCA with ordinal indicators can be parametrized differently.
+The approach implemented in `tidySEM` assumes that each ordinal variable reflects an underlying standard normal distribution.
+The parameters are "thresholds" that correspond to quantiles of a standard
+normal distribution (with $N(\mu = 0, \sigma = 1)$).
+These thresholds
+are estimated based on the proportion of individuals in each of the
+response categories of the indicator variable. For example, a binary
+indicator has a single threshold that distinguishes the two response
+categories. If responses are distributed 50/50, then the corresponding
+threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
+then the resulting threshold would be $t_1 = 0.25$.
+All of the aforementioned parameters can be freely estimated,
+or constrained across classes, or fixed (e.g., to zero).
+
+Prior literature emphasized the importance of considering alternative model specifications [@vandeschootGRoLTSChecklistGuidelinesReporting2017].
+However, what has remained underemphasized is that alternative model specifications may be approached differently depending on whether an analysis is exploratory or confirmatory (see Use Cases).
+Most prior literature has emphasized exploratory applications of LCA [e.g., @nylund_deciding_2007]
+In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
+A "final" model is then selected based on a combination of fit indices,
+significance tests,
+and interpretability (see Class Enumeration).
+
 Confirmatory LCA does not require such an extensive search over the model space.
-Theoretical models can be compared to a smaller set.
 If hypothesized models have been specified a priori,
 and possibly even preregistered,
-it makes sense to focus on the estimation and interpretation of these models, rather than proceed in a purely data-driven manner.
-One key challenge is that LCA lacks objective fit indices,
-and the fit of a theoretical model can thus only be determined relative to other likely candidates.
+it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
+Nonetheless, an important reason to consider alternative model specifications is that LCA lacks objective fit indices.
+The fit of a theoretical model can thus only be assessed by comparison to other candidates.
 For this reason, the theoretical model should still be compared to a few others.
 A one-class model is a sensible benchmark, to test whether LCA is even appropriate.
-A theoretical model would be supported by a substantial (and significant) difference in fit with a one-class model.
 Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
-Here again, a substantial difference in fit supports the theoretical model - but note that small differences in fit do not give cause to reject a theoretical model.
+If the theoretical model has much better fit than other models,
+this provides evidence for the theory.
+If differences in fit are small, however, this may not  do not give cause to reject a theoretical model.
 The theoretical model may still be preferred, for example, to ensure comparability with prior work.
@@
 along with potential implications for future work.
-The BIC is the preferred fit index for comparing
+
+<!-- This will ensure that the model accurately represents the data and provides reliable results. -->
+<!-- Two factors play a role in LCA model specification: -->
+<!-- the number of latent classes and the model estimated within each class. -->
+<!-- Assuming the same model is estimated across classes, -->
+<!-- the total number of parameters in LCA is equal to the number of parameters of the class-specific model times the number of classes, -->
+<!-- plus a vector of class proportions with $K-1$ free elements. -->
+
+<!-- In the general sense, the class-specific model can be any structural equation model. -->
+<!-- Its parameters may include means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables. -->
+<!-- All of these parameters can be freely estimated, -->
+<!-- constrained to be equal across classes, -->
+<!-- or fixed to a known value. -->
+<!-- The specific types LCA introduced in the Taxonomy all have different parameters. -->
+<!-- For example, LPA typically assumes class-specific means, -->
+<!-- equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`. -->
+<!-- By contrast, LCA with ordinal indicators assumes free thresholds across classes, -->
+<!-- where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`. -->
+<!-- Like LPA, LCGA and GMM typically have free means across classes, -->
+<!-- but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`. -->
+<!-- The distinction between LCGA and GMM is whether variances of the latent growth variables are constrained to zero, or freely estimated (either across or within classes). -->
+<!-- either using `lavaan` model syntax or using `OpenMx` RAM specification, -->
+<!-- and supplied to the general function `mx_mixture()`. -->
+<!-- The resulting model may have -->
 
@@
 
-The second consideration is the risk of overfitting.
-We address each in turn.
-
-As explained before, the total number of parameters scales with the number of estimated classes.
-Consequently, LCA methods have a potentially very high number of parameters.
+A second consideration in model specification is the risk of overfitting [@hastieElementsStatisticalLearning2009].
+Overfitting means that a model captures idiosyncratic noise in a sample,
+and thus generalizes poorly to new samples.
+It occurs when a model has many parameters relative to the number of observations
+As explained before, the number of parameters in LCA scales with the number of estimated classes.
+Consequently, LCA with many classes have a potentially very high number of parameters.
+
+In Bayesian mixture modeling,
+it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased [@rousseauAsymptoticBehaviourPosterior2011].
+Although it is not known whether frequentist LCA exhibits the same behavior,
+this suggests that low class counts might indicate overfitting.
 
@@
 
-Many software packages are available for the estimation of LCA models.
+LCA models can be estimated in a variety of ways.
 Some of these packages have limited functionality, or implement specific
@@
 parameters, e.g. the class-specific means $\mu_c$ on a continuous
-indicator (ignoring the variance for now). The EM algorithm will attempt
-to find a combination of values for these two parameters that maximizes
-the likelihood ($LL$) of all observed data. In practice, instead of
-maximizing $LL$, often $-2*LL$ is minimized, as this offers
-computational advantages. We can think of this optimization problem as a
-three-dimensional landscape: The X and Y dimensions are determined by
-the class-specific means, so $X = \mu_1$ and $Y = \mu_2$ - and the
-Z-dimension is determined by $Z = -2*LL$. The optimizer must find the
-deepest "valley" in this landscape, which reflects the combination of
-$\mu_1$ and $\mu_2$ that maximizes the likelihood of the data. The EM
+(ignoring the variance for now).
+EM attempts to find a combination of values for these two parameters that maximizes the log likelihood ($LL$) of the data.
+In practice, instead of maximizing $LL$,
+$-2*LL$ is minimized.
+Transforming the likelihood this way does not change the resulting parameter estimates,
+but it is mathematically convenient.
+Moreover, the $-2*LL$ is asymptotically $\chi^2$ distributed if the null-hypothesis is true, and thus this quantity can be used for model fit tests [@wilksLargeSampleDistributionLikelihood1938].
+To understand the optimization problem, imagine a three-dimensional landscape: 
+The X and Y dimensions represent potential values of the class-specific means, $X = \mu_1$ and $Y = \mu_2$, and the Z-dimension is determined by $Z = -2*LL$. 
+The optimizer must find the
+deepest "valley" in this landscape,
+which reflects the combination of $\mu_1$ and $\mu_2$ that minimizes the $-2*LL$.
+The EM
 optimizer behaves somewhat like a marble dropped in this landscape. It
 is dropped at some random point in space, and will roll into the nearest
-valley. The problem is that, once EM rolls into a valley, it will settle
-on the bottom of that valley (this is known as "convergence"). It cannot
-climb out again. Thus, if their are multiple valleys, the risk is that
-the optimizer gets stuck in a shallower valley (a "local optimum"), and
-never discovers the deepest valley (the "global optimum", or best
-solution). One solution to this problem is to drop many marbles at
-random places, compare their final $-2*LL$ values, choose the solution
+valley.
+The challenge is that LCA models are often highly parametrized.
+This means that the landscape is not three-dimensional, but (very) high-dimensional,
+with potentially many valleys and sparse data to provide information about which valley is deepest.
+If the marble rolls into one of these valleys,
+it will settle there and not climb out again.
+The estimator has "converged".
+The risk is that the optimizer gets stuck in a shallower valley (a "local optimum"),
+and never discovers the deepest valley (the "global optimum", or best solution).
+One solution to this problem is to drop many marbles at random places,
+compare their final $-2*LL$ values,
+choose the solution
 with the lowest $-2*LL$, and make sure that several marbles replicated
-this solution. This is the "random starts" approach.
+this solution.
+This is the "random starts" approach implemented in, e.g., Mplus.
 
@@
 computationally expensive to run this many replications. Moreover,
-because the algorithm begins with random starting values, many of the
-marbles are likely to be very far away from a "good enough" solution.
-Two innovations may improve the estimation procedure. The first is that,
-instead of picking random starting values, a "reasonable solution" may
-be used for the starting values. For example, if we assume that the
-different classes are likely to have different mean values on the
-indicators, then the K-means clustering algorithm can be used to
-determine these cluster centroids. We can compute the expected values of
-all model parameters by treating the K-means solution as a known class
-solution, and use these as starting values for a mixture model. One
-remaining concern is that this approach may result in starting values
-close to a local optimum, and that the EM algorithm will thus never find
-the global optimum. A second innovation addresses this concern. Instead
-of using EM, it is possible to use an optimizer that can climb out of a
-valley. Simulated annealing iteratively considers some "destination" in
-the landscape, and compares its likelihood to the current one. If the
-destination likelihood is higher, the estimator moves there. If the
-destination likelihood is *lower*, the estimator still moves there
-occasionally, based on probability. This latter property allows it to
-escape local optima, and find the global optimum.
-
+because the algorithm begins with random starting values,
+it is inefficient because many marbles are likely to start very far away from any valley.
+Finally, if all sets of starting values are close to a local optimum,
+there is a risk that the global optimum is not discovered.
+An alternative solution that overcomes these challenges is to use a global optimizer, like the simulated annealing algorithm [SA, @coranaMinimizingMultimodalFunctions1987].
+SA does not behave quite like the marbles of the random starts approach;
+instead, it iteratively considers some "destination" in the landscape,
+and compares its $-2*LL$ to the current one.
+If the destination $-2*LL$ is lower, 
+the estimator moves there.
+However, if the destination  $-2*LL$ is *higher*,
+the estimator still moves there occasionally, with a certain probability.
+Think of this as occasionally flicking the marble back out of valleys,
+to see if it rolls back into the current valley, or finds another,
+even deeper one.
+This property allows SA it to escape local optima and find the global optimum.
+
+Since SA is a global optimizer, it should be independent of starting values.
+With this in mind, it is efficient to start estimation from a "reasonable solution", instead of a random starting point.
+For example, if we assume that the different classes are likely to have different mean values on the indicators, 
+then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernackiChoosingStartingValues2003].
 By default, `tidySEM` derives starting values using K-means clustering,
-and identifies the global optimum solution using simulated annealing.
-Once a solution has been found, simulated annealing is followed up with
-a short run of the EM algorithm, as EM inherently produces an asymptotic
+computing the mixture model starting values by treating the K-means solution as multi-group model.
+Next, SA is used to find the global optimum solution.
+Finally, SA is followed up with a short run of the ML algorithm,
+as EM inherently produces an asymptotic
 covariance matrix for the parameters that can be used to compute
-standard errors. Note that these defaults can be manually overridden.
-
-A recent paper suggested maximum likelihood with robust standard errors
-should be used when the observed indicators are not normally distributed
-[@spurk_latent_2020]. This statement is incorrect, and may lead readers
-to believe that they must use commercial software, as robust maximum
-likelihood is currently only implemented in Mplus and LatentGOLD. As
-explained before, mixture modeling assumes that observed data are a
-mixture of (multivariate) normal distributions; thus, the observed
-indicators will likely not be normally distributed.
-<!-- CJ: Is there even any literature that says robust standard errors are better with latent class analysis? -->
+standard errors.
+Note that these defaults can be manually overridden.
+
+In case of any (suspected) convergence problems,
+note that `tidySEM` LCA models are `OpenMx` models.
+Thus, existing `OpenMx` functions to aid model convergence can be used, like `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
 
@@
 
-As explained, LCA can be done in an exploratory or in a confirmatory
-fashion. In exploratory LCA, a sequence of models is fitted to the data
-with each additional model estimating one more class than the previous
-model. These models are then compared and the best solution is selected
-as the final class solution. In some cases, prior theory can inform the
-researcher about the number of classes to expect. Even in such
-confirmatory LCA cases, it is nonetheless useful to know if the
-theoretical model is markedly better than those with differing numbers
-of classes. Therefore, it may always be useful to compare different
-class solutions.
+Class enumeration refers to the common practice of comparing models with different numbers of classes in LCA.
+
+<!-- The total number of parameters scales with the number of estimated classes. -->
+<!-- Consequently, latent class analyses have a potentially very high number of parameters. -->
+<!-- As any of these parameters could be mis-specified, -->
+<!-- it is important to consider alternative model specifications. -->
+<!-- One way to obtain relative evidence in favor of different models is by using BIC weights, obtained by the function `tidySEM::ic_weights()` [@wagenmakersAICModelSelection2004]. -->
+
+<!-- From sinhaPractitionerGuideLatent2021: -->
+<!-- Testing for number of classes.
+In addition to indexing model fit, there are tests comparing a model with k classes to one with k-1 classes.
+Developed by Lo, Mendell and Rubin (36) based on work by Vuong (54) the VLMR test assumes multivariate normality, and it is not clear how sensitive the p-value is to violation of that assumption.
+It is also possible to use a bootstrapped p-value, although its validity outside of normally distributed data remains unknown.(55) In a Monte Carlo simulation of LCA using only categorical indicators, Nylund and colleagues found that the bootstrapped test (BLMR) consistently outperformed the simple Lo-Mendell-Rubin test (LMR).(34) Again in simulation studies with latent profile analysis (continuous indicators), Tein and colleagues found that the BLMR had higher statistical power than LMR, although both tests performed well at detecting the true class.(56) In our practice, across multiple analyses with real-life data consisting of a mixture of categorical and continuous indicators, we have found that the BLMR consistently favours k classes over k - 1 class to the point of being of limited value.
+-->
+Class enumeration refers to the process of determining the appropriate number of classes.
+This topic is closely related to model specification.
+As explained before,
+LCA can be done in an exploratory or in a confirmatory fashion.
+In the confirmatory case, the set of models to be compared will be smaller.
+In the exploratory case, the set of models typically consists of a sequence from 1 to some higher number of classes $K$.
+The maximum number of classes $K$ may be chosen a-priori, or on theoretical accounts.
+In practice, the limit is often dictated by the data;
+after a certain number,
+convergence problems may arise, 
+or the proportion of cases assigned to the smallest classes may be so low that such a class would not be locally identified.
+It is sensible to limit $K$ to the maximum number of classes that result in valid solutions.
+<!-- In exploratory LCA, a sequence of models is fitted to the data -->
+<!-- with each additional model estimating one more class than the previous -->
+<!-- model. These models are then compared and the best solution is selected -->
+<!-- as the final class solution. In some cases, prior theory can inform the -->
+<!-- researcher about the number of classes to expect. Even in such -->
+<!-- confirmatory LCA cases, it is nonetheless useful to know if the -->
+<!-- theoretical model is markedly better than those with differing numbers -->
+<!-- of classes. Therefore, it may always be useful to compare different -->
+<!-- class solutions. -->
 
@@
 
-Fit indices typically used for determining the optimal number of classes
-include the Akaike Information Criterion (AIC) and the Bayesian
-Information Criterion (BIC). Both information criteria are based on the
--2\*log-likelihood (which is lower for better fitting models), and a
-penalty term for the number of parameters (thus incentivizing simpler
-models). This helps balance model fit and model complexity. The lower
+As there are no absolute fit indices for LCA,
+information criteria are typically used to assess relative fit.
+The original information criterion is the Akaike Information Criterion (AIC);
+it was designed to balance model fit and model complexity.
+This is achieved by adding a penalty for the number of parameters to the -2\*log-likelihood (which is the most fundamental measure of model fit).
+The lower
 the value of an information criterion, the better the overall fit of the
 model.
-XXXXXX
-The general objective of information criteria (ICs) is to evaluate the model's 
-out-of-sample predictive accuracy, thus adjusting it for overfitting. 
-Fit measures such as $R^2$ evaluate the in-sample predictive accuracy, 
-i.e. the model's ability to predict the observed outcomes based on the same 
-data that was used to fit the model. In-sample metrics are positively biased, 
-meaning that they overestimate the fit the model has in reality, i.e. when 
-presented with new data. While common ways to overcome this bias is through
-resampling methods such as cross-validation, or use of out-of-sample data such
-as a test set, ICs correct for this positive bias by evaluating the model's 
-accuracy in a way which approximates the out-of-sample predictive accuracy. 
-ICs achieve this by applying different penalty metrics. For example, $R^2$ will 
-increase even if an added predictor is fully redundant. ICs show preference for
-simpler models and indicate worse fit when a predictor, or its added complexity 
-is unnecessary [@mcelreath_statistical_2020].
-
-The BIC applies a stronger penalty for model complexity that
-scales logarithmically with the sample size. The literature suggests the
-BIC may be the most appropriate information criterion to use for model
-comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
-Both the AIC and the BIC are available in the `tidySEM` output.
+The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size,
+such that complex models are penalized more in smaller samples.
+Finally,
+the so-called *sample size adjusted* BIC (saBIC) implements a different penalty for sample size.
+
+Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
+because this test is very computationally expensive.
+It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
+
+Fit indices typically used for determining the optimal number of classes include
+the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC).
+Both information criteria are based on the -2 log likelihood (which is lower for better fitting models),
+and add a penalty for the number of parameters (thus incentivizing simpler models).
+This helps balance fit and model complexity.
+The BIC usually applies a stronger penalty for complexity that scales logarithmically with the sample size.
 
@@
 [@nylund-gibson_ten_2018].
+
+The literature suggests that the saBIC performs especially well for class enumeration.
+However, unlike the AIC and BIC, its penalty for sample size is not grounded in theory.
+BIC also performs well and is grounded in theory; as such, it may be the most appropriate information criterion to use for model
+comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
+All three ICs are available in `tidySEM` output.
+
+<!-- The general objective of information criteria is to evaluate the model's  -->
+<!-- out-of-sample predictive accuracy, -->
+<!-- thus curtailing overfitting. -->
+<!-- Fit measures such as $R^2$ evaluate the in-sample predictive accuracy,  -->
+<!-- i.e. the model's ability to predict the observed outcomes based on the same  -->
+<!-- data that was used to fit the model. In-sample metrics are positively biased,  -->
+<!-- meaning that they overestimate the fit the model has in reality, i.e. when  -->
+<!-- presented with new data. While common ways to overcome this bias is through -->
+<!-- resampling methods such as cross-validation, or use of out-of-sample data such -->
+<!-- as a test set, ICs correct for this positive bias by evaluating the model's  -->
+<!-- accuracy in a way which approximates the out-of-sample predictive accuracy.  -->
+<!-- ICs achieve this by applying different penalty metrics. For example, $R^2$ will  -->
+<!-- increase even if an added predictor is fully redundant. ICs show preference for -->
+<!-- simpler models and indicate worse fit when a predictor, or its added complexity  -->
+<!-- is unnecessary [@mcelreath_statistical_2020].  -->
+
+### Nested model tests
 
@@
 
+A recent paper suggested maximum likelihood with robust standard errors
+should be used when the observed indicators are not normally distributed
+[@spurk_latent_2020].
+This statement is incorrect;
+LCA does not have an assumption of marginal normality, and there is - to our knowledge - no literature on the merits of robust standard errors for LCA.
+
+<!-- and may lead readers -->
+<!-- to believe that they must use commercial software, as robust maximum -->
+<!-- likelihood is currently only implemented in Mplus and LatentGOLD. As -->
+<!-- explained before, mixture modeling assumes that observed data are a -->
+<!-- mixture of (multivariate) normal distributions; thus, the observed -->
+<!-- indicators will likely not be normally distributed. -->
+<!-- CJ: there is no literature that says robust standard errors are better with latent class analysis? -->
+
 <!-- Treating ordinal data as continuous -->

### 2022-12-22 manuscript.Rmd
update 2022-12-22 14:51:39

@@
 
-Despite its popularity, there is a lack of standards for estimating and
-reporting LCA. While @van_de_schoot_grolts-checklist_2017 developed
-reporting guidelines for a specific type of LCA known as latent growth
-models, general reporting guidelines for best practices in LCA are still
-lacking.
+Despite the popularity of LCA,
+two challenges impede broader adoption and correct application of the method.
+The first challenge is that most existing software that implements these models is commercial and closed-source.
+The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
 This introduces a risk of misapplications of the technique,
 and complicates manuscript review and assessment of the quality of published studies.
-The present paper seeks to address this gap in the literature by introducing updated guidelines for estimation and reporting on latent class analysis,
+The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
 based on current best practice. 
-Importantly, in order to lower the barrier of entry to latent class analysis
-and ensure reproducibility of all examples,
-this paper exclusively uses free, open source software for latent class analysis in R.
-To this end, new functionality was developed in the R-package `tidySEM` [@vanlissaTidySEMTidyStructural2022].
+Furthermore, meeting the increasing demand for open source research software,
+this paper introduces new functionality in the R-package `tidySEM` [@vanlissaTidySEMTidyStructural2022] to perform LCA using free open source software (FOSS).
+
+## The Argument for Free Open Source Software
+
+Despite the popularity of latent class analysis (LCA),
+most existing software is commercial and closed-source,
+for example, Mplus [@muthenMplusUserGuide1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquistMplusAutomationPackageFacilitating2018], and LatentGOLD [REF].
+<!-- software that implements these models. -->
+<!-- One important limitation is that the most fully-features -->
+A downside of commercial software is that its use is restricted to individuals and institutions who can afford a license.
+Another downside is that the proprietary source code cannot be audited, debugged, or enhanced by third parties.
+This goes against open science principles [@lamprechtFAIRPrinciplesResearch2019],
+incurs a risk that mistakes in the source code go unnoticed,
+and curbs progress as independent developers cannot fix bugs or add functionality.
+
+There are several compelling reasons why researchers should prefer free open source software (FOSS) instead.
+FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
+This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
+Furthermore, FOSS promotes collaboration within the scientific community,
+as its users can contribute to the development, maintenance, and support of the software.
+Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
+researchers can review and verify the methods and algorithms used,
+which helps to ensure the integrity and reliability of research findings.
+Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.
+
+Notable FOSS for LCA includes Mclust [REF], depmixS4 [REF], and OpenMx [@neale_openmx_2016].
+A crucial limitation of existing FOSS is that some packages have limited functionality (e.g., Mclust),
+or implement specific innovations (e.g., depmixS4).
+OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
+Remaining obstacles to the use of `OpenMx` is that it is not very user-friendly and its functionality for LCA is poorly documented.
+
+These limitations are addressed by new functionality in the `tidySEM` package,
+which provide a user-friendly interface for LCA in `OpenMx`.
+These functions adhere to best practices in estimation and reporting as described in this paper.
+The `tidySEM` package aims to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
+The user interface is simple, as models can be specified using the widely used `lavaan` syntax [@rosseel_lavaan_2012].
+The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
+but the resulting models can still be fully customized (low barrier, high ceiling).
+Specifically, LCA models constructed in `tidySEM` inherit from `OpenMx`' format `MxModel`, and remain compatible with downstream functions that process such models.
+Being FOSS, `tidySEM` facilitates fully-reproducible analyses and contributes to open science.
+It was developed with open science in mind,
+and adheres to the FAIR software principles [Findable, Accessible, Interoperable, and Reusable, @lamprechtFAIRPrinciplesResearch2019],
+and the [OpenSSF Best Practices](https://bestpractices.coreinfrastructure.org/en/criteria) for FOSS software.
 
@@
 
+<!-- A third consideration in preprocessing relates to the univariate distributions of indicators. -->
+<!-- For example, the `plas_depression` data included with `tidySEM` show extreme right skew; -->
+<!-- these data assess depression, and as such,  -->
+
 ### Handling Missing Data
@@
 ```
-
-### Software
-
-LCA models can be estimated in a variety of ways.
-Some of these packages have limited functionality, or implement specific
-innovations. Other packages implement LCA in the context of a more
-flexible structural equation modeling framework. The most notable
-examples of the latter are the commercial programs Mplus and Latent
-GOLD, and the free open source R-package OpenMx. The commercial packages
-stand out because they offer relatively user-friendly interfaces and
-implement sensible defaults for complex analyses, including LCA methods.
-This lowers the threshold for applied researchers to adopt such methods.
-Commercial software also has several downsides, however. One such
-downside is that use of the software is restricted to those individuals
-and institutions who can afford a license. Another downside is that the
-source code, being proprietary, cannot be audited, debugged, or enhanced
-by third parties. This incurs the risk that mistakes in the source code
-may go unnoticed, and curbs progress as software developers cannot add
-new functionality.
-
-Conversely, the free open source program OpenMx [@neale_openmx_2016] is very flexible, but
-not very user-friendly. We directly address this limitation using the
-`tidySEM` R-package. New functionality in `tidySEM` seeks to lower the
-threshold for latent class analysis using `OpenMx`. It adheres to best
-practices in estimation and reporting, as described in this paper. The
-user interface is simple, making use of the model syntax of the widely
-used `lavaan` R-package [@rosseel_lavaan_2012]. This syntax offers a human-readable way to
-specify latent variable models. Minor enhancements are made to simplify
-the specification of LCA.
-
-Because of the limitations of the aforementioned tools, we set out to
-develop a free tool that provides sensible defaults and is easy to use,
-but provides the option to access and modify all of the model inputs
-(i.e., low barrier, high ceiling). `tidySEM` interfaces with existing
-tools, and is able to translate between what existing tools are capable
-of and what researchers and analysts carrying-out person-oriented
-analyses would like to specify. Furthermore,`tidySEM` facilitates
-fully-reproducible analyses and contributes to open science.
 
@@
 note that `tidySEM` LCA models are `OpenMx` models.
-Thus, existing `OpenMx` functions to aid model convergence can be used, like `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
+Thus, existing `OpenMx` functions to aid model convergence can be used, like `mxTryHard()` or `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
+Often, these functions return an acceptable solution for models with convergence issues.
 
 ### Class Enumeration
-
-Class enumeration refers to the common practice of comparing models with different numbers of classes in LCA.
 
@@
 Class enumeration refers to the process of determining the appropriate number of classes.
-This topic is closely related to model specification.
+This topic is closely related to model specification,
+because the number of latent classes also influences the number of model parameters.
 As explained before,
 LCA can be done in an exploratory or in a confirmatory fashion.
-In the confirmatory case, the set of models to be compared will be smaller.
 In the exploratory case, the set of models typically consists of a sequence from 1 to some higher number of classes $K$.
-The maximum number of classes $K$ may be chosen a-priori, or on theoretical accounts.
-In practice, the limit is often dictated by the data;
-after a certain number,
-convergence problems may arise, 
+The maximum number of classes $K$ may be chosen a-priori, or on theoretical grounds, or dictated by the data:
+after a certain number of classes,
+convergence problems often arise, 
 or the proportion of cases assigned to the smallest classes may be so low that such a class would not be locally identified.
-It is sensible to limit $K$ to the maximum number of classes that result in valid solutions.
+In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
+the proportion of cases assigned to the excessive classes converges to zero as the sample size increases [@rousseauAsymptoticBehaviourPosterior2011].
+It is not known whether this phenomenon also occurs in the frequentist case;
+nonetheless, near-zero class memberships should be taken into account when determining the maximum $K$.
+It is sensible to limit $K$ to the maximum number of classes that result in valid and interpretable solutions.
+In the confirmatory case,
+the set of models with different numbers of classes to be compared is typically smaller;
+its purpose is to provide context for the relative fit of the theoretical model.
+
+During class enumeration, the set of models is compared on several statistical and substantive criteria.
+In the exploratory case, these criteria may be decisive in selecting a final model;
+in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
+Statistical criteria to consider in class enumeration include model fit indices and
+likelihood ratio tests [@weller_latent_2020; @masyn_latent_2013].
+Substantive criteria include convergence and model identification,
+class separability,
+and interpretability.
+We address each of these criteria below.
+
 <!-- In exploratory LCA, a sequence of models is fitted to the data -->
@@
 
-From a sequence of models, the final class solution is chosen based on
-both theoretical and statistical criteria. Theory should drive the
-selection of indicator variables, inform the expectations and reflect on
-the findings. In addition to this, there are several statistical
-criteria to consider in model selection. These include but are not
-limited to likelihood ratio tests, information criteria, and the Bayes
-factor [@weller_latent_2020; @masyn_latent_2013].
-
-Relative model fit can be examined using the likelihood ratio test. This
-is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as the difference in
-maximum log-likelihoods of the two models, with the test degrees of
-freedom being the difference in the degrees of freedom of the two
-compared models. The test statistic follows the $\chi^2$ distribution,
-and we want it to be non-significant in order to give support to the
-simpler model. The likelihood ratio test can only compare two nested
-models at a time [@lanza_introduction_2003].
-
 ### Model Fit Indices
@@
 information criteria are typically used to assess relative fit.
-The original information criterion is the Akaike Information Criterion (AIC);
-it was designed to balance model fit and model complexity.
-This is achieved by adding a penalty for the number of parameters to the -2\*log-likelihood (which is the most fundamental measure of model fit).
+Information criteria are computed from the -2\*log-likelihood ($-2LL$),
+which is the most fundamental measure of model fit.
+A penalty for model complexity is added to the $-2LL$ to penalize more complex models.
+Information criteria thus balance fit and complexity.
 The lower
@@
 model.
-The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size,
+The original information criterion is the Akaike Information Criterion (AIC),
+computed as $-2LL+2p$, where $p$ is the number of parameters.
+The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
 such that complex models are penalized more in smaller samples.
 Finally,
-the so-called *sample size adjusted* BIC (saBIC) implements a different penalty for sample size.
-
-Note however that this table does not include the bootstrapped likelihood ratio test (BLRT) by default,
-because this test is very computationally expensive.
-It is recommended to use the function `tidySEM::BLRT()` to compare a shortlist of likely candidate models based on other fit indices.
-
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2 log likelihood (which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance fit and model complexity.
-The BIC usually applies a stronger penalty for complexity that scales logarithmically with the sample size.
+the so-called *sample size adjusted* BIC (saBIC) implements a different ad-hoc penalty for sample size,
+$\ln(\frac{n + 2}{24})$.
+Although a simulation study showed that the saBIC performed well for class enumeration,
+its penalty is not based on theory.
+BIC also performs well and is theoretically substantiated;
+as such, it may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
+All three ICs are available in the output of `table_fit()`.
 
 Information criteria may occasionally contradict each other, so it is
-important to identify a suitable strategy to reconcile them. One option
-is to select a specific fit index before analyzing the data. Another
-option is to always prefer the most parsimonious model that has best fit
-according to any of the available fit indices. Yet another option is to
-incorporate information from multiple fit indices using the analytic
-hierarchy process [@akogul_comparison_2016]. Finally, one might make an
-elbow plot and compare multiple information criteria
-[@nylund-gibson_ten_2018].
-
-The literature suggests that the saBIC performs especially well for class enumeration.
-However, unlike the AIC and BIC, its penalty for sample size is not grounded in theory.
-BIC also performs well and is grounded in theory; as such, it may be the most appropriate information criterion to use for model
-comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
-All three ICs are available in `tidySEM` output.
+important to identify a suitable strategy to reconcile them.
+One option is to select a specific fit index, such as the BIC,
+before analyzing the data.
+Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
+Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016],
+which is implemented in `tidyLPA` [@rosenbergTidyLPAPackageEasily2018].
+Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs [@nylund-gibson_ten_2018].
+The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
+Adding this many classes results in substantial decreases of the IC,
+and adding further classes results in relatively small decreases.
+
+Given a specific choice of information criterion,
+it is also possible to calculate so-called "IC-weights",
+which indicate the relative support the data provide for each model in the set, expressed as a proportion of total support [@wagenmakersAICModelSelection2004].
+In `tidySEM`, IC-weights can be calculated by calling `ic_weights(x = table_fit(), ic = "BIC")` on a model fit table.
+Note, however, that IC-weights will (strongly) prefer the model with the lowest IC value;
+the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
 
@@
 
-Another common test of model fit is the likelihood ratio $\chi^2$
-goodness-of-fit test. However, this test is not implemented in
-`tidySEM`.
-
-A disadvantage of ICs is that they are not accompanied by standard errors or 
-any measure of variability, only their absolute value. The difference between 
-two models' ICs can be rather small. The lower information criterion would 
-still indicate the better model, but if the ICs difference is very small, 
-two models can be considered "functionally equal". In such situations, the 
-interpretability of the model should take priority.
-
-LCA studies commonly report -2\*log likelihood of the final class
-solution. This is a basic fit measure used to compute most information
-criteria. However, since log likelihood is not penalized for model
-complexity, it will continuously fall with the addition of more classes.
-
-An alternative is using the bootstrapped likelihood ratio test which can
-be run using `tidySEM::BLRT()`. Currently, this test is computationally
-expensive and can be slow on most computers. A faster version of this
-test, namely an implementation of *the lazy bootstrap*
-[@van_kollenburg_lazy_2018] to `tidySEM` is being developed.
+When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
+The test statistic is computed as the -2 times the log of the ratio of likelihoods,
+and its degrees of freedom are equal to the difference in the degrees of freedom of the two compared models.
+Since a $k-1$-class model is nested in a $k$-class model,
+it is possible to test the difference in model fit using a LR test.
+One challenge is that the LR test statistic is not asymptotically $\chi^2$ distributed in LCA.
+The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR [REF lo 2001].
+This test is implemented in `tidySEM`, and is part of the default output of `table_fit()`.
+
+Another solution is to derive an empirical sampling distribution for the LR difference.
+The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the $k-1$-class model $B$ times (e.g., 1000),
+and fitting the $k$- and $k-1$-class models to each simulated dataset [REF Nylund et al., 2007].
+The p-value of this test is defined as the proportion of $B$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
+A crucial limitation of the BLRT is that it is very computationally expensive.
+With this in mind, `table_fit()` does not provide the BLRT by default.
+It is possible to compute it for a shortlist of likely candidate models,
+using the function `tidySEM::BLRT()`.
+However, we have noticed that this function often runs out of memory or takes prohbitively long to compute.
+
+Simulation studies comparing the LMR and BLMR suggest that the latter has greater statistical,
+although both performed comparably well at detecting the true number of classes [REF Nylund 2007, REF Tein 2013].
+However, anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
+returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
+Other authors have reported similar experiences [@sinhaPractitionerGuideLatent2021].
+This suggests that, like existing ICs,
+LR tests do not always adequately balance fit and complexity in LCA.
+Better solutions for class enumeration are thus an important target for future research.
+
+One promising alternative solution is the so-called *lazy bootstrap* (LB),
+which is much more efficient because it is based on sample statistics of model-implied data generated from the $k-1$ and $k$-class models,
+rather than on fit statistics of LCA models fit to those data [@van_kollenburg_lazy_2018].
+It is similar to the posterior predictive p-value in Bayesian statistics,
+and describes how well each model is able to reproduce the observed data [see @mcelreath_statistical_2020].
+One version of the lazy bootstrap is already implemented in the commercial program latentGOLD.
+At the time of writing, we are developing extentions of the LB and investigating their utility for class enumeration.
+Future versions of `tidySEM` will likely incorporate this test.
 
@@
 
-Best models will divide the sample into subgroups which are internally
-homogeneous and externally distinct. Classification diagnostics give us
-a way to assess the degree to which this is the case. They are separate
-from model fit indices as a model can fit the data well but show poor
-latent class separation [@masyn_latent_2013]. Classification diagnostics
-should not be used for model selection, but they can be used to
-disqualify certain solutions because they are uninterpretable.
+<!-- 1. We do not consider classes with, on average, fewer than 5 participants per parameter in a class due to potential local underidentification -->
+<!-- 1. We do not consider solutions with entropy < .90 because poor class separability compromises interpretability of the results -->
+<!-- 1. We do not consider solutions with minimum posterior classification probability < .90 because poor class separability compromises interpretability of the results -->
+
+<!-- Nevertheless, researchers can set an a-priori threshold for the minimum proportion of cases per class, -->
+<!-- for example by considering only solutions with a number of participants at least five times as large as the number of parameters per class. -->
+
+In most use cases of LCA, the goal is to identify 
+subgroups that are internally homogeneous and externally distinct.
+Classification diagnostics assess whether this is the case.
+These diagnostics are distinct from model fit indices,
+because a model can fit the data well but show poor latent class separation [@masyn_latent_2013].
+Classification diagnostics should therefore not be used for model selection.
+Nonetheless, they can help restrict the search space of acceptable models.
 Interpretability should always be a consideration when considering
 different class solutions [@nylund-gibson_ten_2018].
-
-Three important classification diagnostics provided by `tidySEM` are:
-(1) the *minimum* and *maximum* percentage of the sample assigned to a 
-particular *class*, (2) the *range of the posterior class probabilities* by most
-likely class membership, and (3) *entropy*. All three are based on posterior 
-class probabilities.
-<!-- MGV: I will work on adding average posterior class probability (AvePP) as well-->
-
-The posterior class probability is a measure of classification
-uncertainty which can be computed for each individual, or averaged for
-each latent class. When the posterior class probability is computed for
-each individual in the dataset, it represents each person's probability
-of belonging to each latent class. For each person, the highest
-posterior class probability is then determined and the individual is
-assigned to the corresponding class. We want each individual's posterior
-class probabilities to be high for one and low for the remaining latent
-classes. This is considered a high classification accuracy and means
-that the classes are distinct. To obtain posterior class probabilities,
-run `tidySEM::class_prob()`. This function produces output comprised of
-several elements:
-
-`$sum.posterior` is a summary table of the posterior class probabilities
-indicating what proportion of the data contributes to each class.
-
-`$sum.mostlikely` is a summary table of the most likely class membership
-based on the highest posterior class probability. From this table, we
-compute the minimum and maximum percentage of the sample assigned to a
-particular class, , i.e. **n_min** (the smallest class proportion based
-on the posterior class probabilities) and **n_max** (the largest class
-proportion based on the posterior class probabilities). We are
-especially interested in **n_min** as if it is very small and comprised
-of few observations, the model for that group might not be locally
-identified. It may be impossible to calculate descriptive statistics for
-such a small class. Estimating LCA parameters on small subsamples might
-lead to bias in the results. Therefore, we advise caution when dealing
-with small classes.
-
-`$mostlikely.class` is a table with rows representing the class the
-person was assigned to, and the columns indicating the average posterior
-probability. The diagonal represents the probability that observations
-in each class will be correctly classified. If any of the values on the
-diagonal of this table is low, we might consider not to interpret that
-solution. In `tidySEM` we use the diagonal to compute the range of the
-posterior class probabilities by most likely class membership which
-consists of the lowest class posterior probability (**prob_min**), and
-the highest posterior probability (**prob_max**). Both **prob_min** and
-**prob_max** can be used to disqualify certain class solutions, and are
-a convenient way to summarize class separation in LCA. We want both
-**prob_min** and **prob_max** to be high as that means that for all
-classes the people who were assigned to that class have a high
-probability of being there. **prob_min** is especially important as it
-can diagnose if there is a class with low posterior probabilities which
-could make one reconsider that class solution.
-
-`$avg.mostlikely` contains the average posterior probabilities for each
-class, for the subset of observations with most likely class of 1:k,
-where k is the number of classes.
-
-`$individual` is the individual posterior probability matrix, with
-dimensions n (number of cases in the data) x k (number of classes). 
-Additionally, it includes the `predicted` class as a function of the highest 
-predicted probability. Individual class probabilities and/or predicted class 
-are often useful to researchers who wish to carry out follow up analyses.
-
-Entropy is a summary measure of posterior class probabilities across
-classes and individuals. It ranges from 0 (model classification no
-better than random chance) to 1 (perfect classification). As a rule of
-thumb, values above .80 are deemed acceptable and those approaching 1
-are considered ideal. An appropriate use of entropy is that it can
-disqualify certain solutions if class separability is too low. Entropy
-was not built for nor should it be used for model selection during class
-enumeration [@masyn_latent_2013].
-
-**n_min**, **n_max**, **prob_min**, **prob_max**, and **entropy** and
-can be obtained using `tidySEM::table_fit()`.
+If the goal is to interpret properties of (members of) classes,
+it makes sense to consider only models that meet certain minimum thresholds for classification diagnostics.
+
+All classification diagnostics are derived from the posterior classification probabilities: an $n \times k$ matrix with the probability that every individual $n$ belongs to latent class $k$.
+When classification accuracy is high, and classes are distinct,
+each individual's posterior class probabilities are high for one class, and low for the remaining classes.
+This matrix of individual classification probabilities $P$ is obtained by running `class_prob(res)` and extracting the element labeled `"individual"`.
+Note that one column is added to the matrix $P$:
+the final column represents most likely class membership $M$,
+which is obtained by assigning each observation to the class corresponding to their maximum posterior class probability.
+Thus, if an individual has posterior probabilities `c(0.2, 0.5, 0.3)` for classes 1-3,
+they would be assigned to class 2 ($M = 2$).
+The predicted class membership is often of interest to researchers who wish to carry out follow up analyses.
+Note, however, that such follow-up analyses should take classification error into account. The function `BCH()` does this automatically.
+
+The matrix $P$ can be summarized in different ways, and the following summaries are provided by default:
+First, a frequency table of the latent categorical variable,
+based on posterior probabilities, referred to as $F_P$ (`"sum.posterior"`).
+It is obtained by calculating the column sums of $P$.
+These numbers are fractional because individuals can contribute partially to multiple classes.
+This table thus takes classification error into account.
+
+Second, most likely class membership based on the highest posterior class probability, $F_M$ (`"sum.mostlikely"`).
+This is the frequency table of the variable $M$, and it ignores classification error.
+Note that this table can differ from $F_P$,
+where some observations may partly contribute to multiple classes.
+If every observation is classified with perfect accuracy,
+these tables will be identical.
+Part of the output of `table_fit()` for LCA analyses
+are the minimum and maximum percentage of the sample assigned to a
+particular class, `n_min` and `n_max`, which are derived from $F_M$.
+Of particular interest is the smallest class proportion, `n_min`.
+When the smallest class (`n_min`) is less than 10% of the total sample size,
+proceed with caution:
+The model might not be locally identified in that class.
+Check whether the number of participants in this class is larger than the number of parameters per class (e.g., five times as large),
+and check whether this smallest class has a meaningful substantive interpretation.
+
+To assess the classification accuracy, we turn to the third diagnostic table:
+The average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$ (`"avg.mostlikely"`):
+for participants assigned to one particular class (rows) according to $M$,
+this table gives the mean probability of being assigned to any class (columns).
+This table is obtained by calculating the column means of $P$ for the 
+$k$ subsets of observations with most likely class of $M = 1 \ldots k$.
+The diagonal represents the probability that observations in each class will be correctly classified.
+
+Finally, the classification probabilities for the most likely latent class $M$ by latent class, $P_{p_{(M = m | C = c)}}$ is similar to the aforementioned table $P_{p_{(M = m)}}$,
+except that it accounts for classification uncertainty.
+If $C$ is the true class of an observation, which is imperfectly measured,
+and $M$ is the most likely class,
+then this table shows the probability of an observation being assigned to class $m$ in $M$, given that its true class is $c$ in $C$, or $p(M=m|C=c)$.
+Values on the diagonal can thus be conceptualized as the reliability with which each level of the categorical latent variable is measured by $M$,
+and off-diagonal elements can be conceptualized as measurement error.
+The minimum and maximum of the diagonal of this table are given by default as part of the output of `table_fit()` for LCA analyses (`prob_min` and `prob_max`).
+These probabilities are a convenient way to summarize classification accuracy.
+The minimum probability is especially important however,
+as it can diagnose if there is at least one class with poor classification accuracy.
+
+Finally, class separability can be summarized in a single statistic:
+the entropy criterion.
+In physics, the concept of entropy is associated with randomness or uncertainty.
+High entropy corresponds to maximum uncertainty,
+and low entropy corresponds to a strict arrangement of the units of study.
+Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
+Low entropy, by contrast, means that every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
+Crucially, in `tidySEM`, the entropy criterion is defined as 1-entropy (a convention originating in Mplus).
+The interpretation of the entropy criterion is reversed:
+0 means the model classification is no better than chance,
+and 1 means perfect classification. 
+<!-- Celeux, G., & Soromenho, G. (1996). An entropy criterion for assessing the number of clusters -->
+<!-- in a mixture model. Journal of Classification, 13, 195-212. -->
+As a rule of
+thumb, values above .80 are deemed acceptable and above .90 are considered good.
+Values close to 1 indicate that classes are very well separated,
+and treating the most likely class membership $M$ as observed in follow-up analyses is probably defensible.
+Like other classification diagnostics,
+entropy is not a model fit measure and should not be used for model selection [@masyn_latent_2013].
+However, it can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
 
@@
 The order of clusters is nondeterministic when using K-means in
-`tidySEM`. Therefore label switching is still a consideration. A simple
-solution to this is setting a random seed number one line prior to
-fitting the model. We advise `tidySEM` users to always do so in order to
+`tidySEM`.
+Therefore label switching is still a consideration.
+A simple solution to this is setting a random seed number one line prior to
+fitting the model, using `set.seed()`.
+We advise `tidySEM` users to always do so in order to
 circumvent label switching.
+Another solution is to specify starting values manually,
+but this is more challenging for two reasons:
+First, because this requires modifying the underlying `MxModel` objects, which is not straightforward.
+Second, if the starting values are not very good, label switching may still occur, and bad starting values could even lead to non-convergence.
 
 Class names should be chosen to accurately reflect group membership,
-as theoretically relevant names according to the interpretation of the class 
-solutions. Overly simplified and generalized class names may prove misleading 
+as theoretically relevant names according to the interpretation of the class solutions.
+Overly simplified and generalized class names may prove misleading 
 to both audiences and researches alike leading to what is known as a naming
@@
 
+There is substantial merit in pre-specifying, and ideally preregistering, the criteria used for class enumeration.
+This is especially pertinent for LCA because there are no objective fit indices for LCA,
+and because there is a risk that information criteria and likelihood ratio tests will suggest a larger number of classes, even if there are concerns about overfitting, classification accuracy, or model interpretability.
+
 Among studies using LCA, reporting practices vary significantly
-[@weller_latent_2020]. Various authors have tried to improve and
+[@weller_latent_2020].
+Various authors have tried to improve and
 standardize ways of reporting LCA [e.g. @masyn_latent_2013;
@@
 
-This is an example of an exploratory LCA using `tidySEM`.
-
-## Loading the Dataset
-
-In this example, we use *data_mix_ordinal*, a simulated data for mixture
-modeling with ordinal indicators. This dataset is built-in to
-`tidySEM`. For more information about the dataset, type
-`?tidySEM::data_mix_ordinal` into the R console after loading the
-`tidySEM` using library().
-
-We load the dataset and convert the indicators into ordered factors.
+This is an example of exploratory LCA with ordinal indicators using `tidySEM`.
+At the time of writing, the `tidySEM` package includes two more vignettes to demonstrate latent profile analysis (LCA with continuous indicators),
+and Latent Class Growth Analysis.
+The present example uses a simulated dataset with ordinal indicators included in `tidySEM`.
+To view its documentation,
+run the command `?tidySEM::data_mix_ordinal` in the R console.
+
+## Loading the Data
+
+To load the data, one simply needs to attach the `tidySEM` package.
+For convenience, we assign the data to an object called `df`:
 
@@
 
-## Exploring the Data
-
-An important step to preceding any statistical analysis is data
-exploration. Here we use `tidySEM::descriptives()` to describe the
-dataset we are using.
+## Examining the Data
+
+As per the best practices,
+the first step in LCA is examining the observed data.
+We use `tidySEM::descriptives()` to describe the data.
 
@@
 As we can see, the output includes various descriptives of our dataset.
-Special attention should be paid to examining the pattern of
-missingness, as discussed in the *Handling missing data* section of this
-paper. In our example, we see that there are no missing values, hence we
+For example, it provides invaluable information about the measurement level of the indicators,
+which is used to specify the model correctly.
+If these data had not been coded as ordinal factors,
+these descriptive statistics would have revealed that each variable has only 3 unique values.
+Attention should alo be paid to the pattern of
+missingness, as discussed in the *Handling Missing Data* section.
+The proportion of missing values is reported in the `"missing"` column.
+If any variables had missing values, we would report an MCAR test with `mice::mcar()`,
+and explain that missing data are accounted for using FIML.
+In our example, we see that there are no missing values, hence we
 proceed with our analysis.
 
+In this case, no preprocessing is necessary:
+all variables are on the same 3-point scale.
+For an example on LCA with extremely skewed variables, see the LCGA vignette.
+
 ## Conducting Latent Class Analysis
@@
 Before we fit a series of LCA models, we set a random seed using
-`set.seed()`. This is an important
-step as there is some inherent randomness in the LCA computations, and
-having the same seed number ensures that two separate researcher obtain
-exactly the same results when fitting LCA models.
-
-Finally, we reach the step of fitting LCA models.
+`set.seed()`.
+This is important because there is some inherent randomness in the estimation procedure,
+and using a seed ensures that we (and others) can exactly reproduce the results.
+
+Next, we fit the LCA models.
 As all variables are ordinal, we can use the convenience function
-`tidySEM::mx_lca()`, which is a wrapper for the generic function `mx_mixture()`,
-optimized for latent class analysis with ordinal data.
-Any mixture model can be specified through `mx_mixture()`,
-but at the time of writing, there are two more wrapper functions for specialized models:
+`tidySEM::mx_lca()`, which is a wrapper for the generic function `mx_mixture()` optimized for latent class analysis with ordinal data.
+Any mixture model can be specified through `mx_mixture()`.
+At the time of writing, there are two other wrapper functions for special cases:
 `mx_profiles()`, for latent profile analysis, and `mx_growth_mixture()`, for latent growth analysis and growth mixture models.
@@
 so relevant variables must be selected first.
-Below, we fit 1 to 4 class solutions.
+Below, we fit 1 to 4 class solutions:
+
+```{r fitting_LCA, include = TRUE, eval=F}
+set.seed(123) # setting seed 
+res <- mx_lca(data = df, classes = 1:4) # fitting LCA 1 to 4 class solutions
+```
+
+In this example, all models converge without issues.
+If this is not the case, however, we can try to aid convergence using `mxTryHard()`, which expands the search for optimal parameter values,
+and related functions, `mxTryHardWideSearch()`, which considers a wider range of potential parameter values,
+and `mxTryHardOrdinal()` for models with ordinal indicators.
+We can inspect the class of resulting objects.
+
+```{r, include = TRUE, eval=F, echo = F}
+is(res)
+is(res[[1]])
+```
+
 <!-- Depending on your computer's computational power, -->
@@
 <!-- more specifically an `OpenMx` model (`MxModel`). -->
-
-```{r fitting_LCA, include = TRUE, eval=F}
-set.seed(123) # setting seed 
-res <- mx_lca(data = df, classes = 1:4) # fitting LCA 1 to 4 class solutions
-```
-
-<!-- Notice that the model run 10 times for each $K$ class solution. -->
-<!-- This way we can choose the best fitting model across all the solutions. -->
-<!-- In this example, we see that all models converged without issues.  -->
-<!-- We can inspect the class of resulting objects. -->
-
-```{r, include = TRUE, eval=F, echo = F}
-is(res)
-is(res[[1]])
-```
 

### 2023-01-16 manuscript.Rmd
draft for feedback

@@
 abstract: |
-  Latent class analysis refers to a family of techniques for identifying groups
+  Latent class analysis (LCA) refers to a family of techniques for identifying groups
   in data based on a parametric model.
-  Examples include mixture models, latent class analysis with ordinal
+  Examples include mixture models, LCA with ordinal
   indicators, and latent class growth analysis.
@@
   source alternatives have remained somewhat inaccessible. This tutorial
-  describes best practices for the estimation and reporting of latent class analysis,
+  describes best practices for the estimation and reporting of LCA,
   using free and open source software in R. To this end, this tutorial
@@
 
-Despite the popularity of latent class analysis (LCA),
+Despite the popularity of LCA,
 most existing software is commercial and closed-source,
@@
 
-There are several compelling reasons why researchers should prefer free open source software (FOSS) instead.
+There are several compelling reasons to prefer free open source software (FOSS) instead.
 FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
@@
 Notable FOSS for LCA includes Mclust [REF], depmixS4 [REF], and OpenMx [@neale_openmx_2016].
-A crucial limitation of existing FOSS is that some packages have limited functionality (e.g., Mclust),
+A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
 or implement specific innovations (e.g., depmixS4).
@@
 
-These limitations are addressed by new functionality in the `tidySEM` package,
+These limitations are addressed by the `tidySEM` package,
 which provide a user-friendly interface for LCA in `OpenMx`.
-These functions adhere to best practices in estimation and reporting as described in this paper.
-The `tidySEM` package aims to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
+The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting LCA, as described in this paper.
+The primary purpose of the `tidySEM` package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
 The user interface is simple, as models can be specified using the widely used `lavaan` syntax [@rosseel_lavaan_2012].
@@
 
-Latent class analysis is a group of methods for estimating unobserved
+LCA refers to methods for estimating unobserved
 groups based on a parametric model of observed indicators of group
 membership [@vermuntLatentClassAnalysis2004].
-This method has become quite popular across scientific fields,
+It has become popular across scientific fields,
 under a number of different names;
@@
 LCA seeks to describe unobserved groups and classify individual observations to those groups.
+<!-- In factor analysis, the latent variable is continuous, so the relationships -->
+<!-- between indicators are defined by their communalities. -->
+<!-- Here, the -->
+<!-- latent variable represents the error corrected approximation of the -->
+<!-- sample. In LCA, by contrast, the relation between indicators is defined by how -->
+<!-- well they separate the classes, as a multinomial regression between the -->
+<!-- indicators and the latent variable [@masyn_latent_2013]. -->
 In line with this distinction, LCA
@@
 thought of as similar to a multi-group structural equation model. The
-main distinction is that group membership is not known a-priori, but is
-instead estimated -- with measurement error -- based on the data.
+main distinction is that group membership is not known a-priori, but estimated from the data.
 Whereas in a multi-group model, the data are split by group and treated
@@
 probability of belonging to that group.
-
-In factor analysis, the latent variable is continuous, so the relationships
-between indicators are defined by the communalities between them. Here, the
-latent variable represents the error corrected approximation of the
-sample. In LCA, by contrast, the relation between indicators is defined by how
-well they separate the classes, as a multinomial regression between the
-indicators and the latent variable [@masyn_latent_2013].
 
@@
 Finally, in the context of machine learning, LCA can be considered as an
-*unsupervised classification* method [@figueiredo_unsupervised_2002].
+*unsupervised* learning method for *classification* [@figueiredo_unsupervised_2002].
 The term *unsupervised* refers to the fact that the outcome variable --
@@
 refers to the fact that the algorithm is predicting a categorical
-outcome -- class membership.
+outcome: class membership.
 
@@
 
-In this paper, we use the term latent class analysis to refer to a
-family of techniques that estimate latent class membership based on a
-parametric model of observed indicators. From a historical perspective,
-the term was initially conceived to refer to analyses with categorical,
-usually binary indicators [@vermunt_latent_2004]. Nowadays, there are
-a number of related techniques, known by distinct names, that serve a
-similar purpose. The term "latent class analysis" seems most appropriate
-as an umbrella term for this broader class of models, as it only refers
+The term LCA was initially conceived to refer to analyses with categorical,
+usually binary indicators [@collins_latent_2009].
+It was later generalized to refer to any technique that estimates a categorical latent variable based on a
+parametric model of observed indicators, as it only refers
 to the purpose of the analysis, and does not imply restrictions to the
-model used, or the level of measurement of the indicators. Given the
-abundance of terms in use for closely related classes of models, we will
-provide a rudimentary taxonomy of LCA methods.
+model used, or the level of measurement of the indicators [@vermunt_latent_2004].
+To prevent ambiguity, we refer to LCA with binary or ordinal indicators as *LCA with ordinal indicators*.
+Given the abundance of redundant terms in use for closely related types of LCA models,
+we here provide a rudimentary taxonomy.
 
@@
 
-It is also possible to estimate a mixture model based on latent
-indicators. This means that, within each class, one or more continuous
-latent variables are estimated based on the observed indicators.
-Categorical latent variable membership is then estimated based on these
-continuous latent variables. A common application of this approach is in
-longitudinal research, where the indicators reflect one construct
-assessed at different time points. Examples of this approach include
+Finally, some LCA models use latent variables as indicators.
+A common application of this approach is in
+longitudinal research with repeated measures of a construct.
+Examples of this approach include
 *growth mixture models* (GMM) and *latent class growth analyses* (LCGA) 
-[@jung_introduction_2008]. These techniques estimate a latent growth model 
-to describe grouped trajectories over time. The growth mixture model is a 
-latent class model where the parameters that indicate class membership are the 
-intercepts and variances, and typically covariances of the latent growth 
-variables, e.g., a latent intercept and slope. This technique assumes that
-individuals within a class can have heterogenous trajectories. If the
-variance of the growth parameters is fixed to zero, it is known as a
-latent class growth analysis. This latter approach assumes that all
-individuals within a class share the same identical trajectory, and that
-any variance in the indicators not explained by the class-specific
-latent trajectories is due to residual error variance.
-
-The term latent class analysis originally referred to cases where the
-observed indicators were ordinal (ordered categorical) [@collins_latent_2009]. 
-Nowadays, it is more commonly used
-as an umbrella term. To prevent ambiguity, the special case where
-indicators are of binary or ordinal measurement level might be described
-as *latent class analysis with ordinal indicators*.
-
+[@jung_introduction_2008].
+These techniques estimate a latent growth model 
+to describe grouped trajectories over time.
+The indicators of class membership in a GMM are the 
+intercepts, and variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
+For linear trajectories, these latent variables are an intercept and slope.
+The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes homogeneity of trajectories within classes.
+This is achieved by restricting the (co)variances of the growth parameters to zero.
+Any variance in the indicators not explained by the class-specific
+latent trajectories is assumed to be error variance.
 <!-- Although this paper primarily focuses on latent class analysis with continuous indicators, -->
@@
 
-Although latent class analysis is often discussed as an exploratory analysis technique,
+Although LCA is often discussed as an exploratory analysis technique,
 it can also be used in a confirmatory manner.
-Given that latent class analysis is similar to confirmatory factor analysis,
-but with a categorical latent variable, it can be used to similar ends:
+Given that LCA is similar to confirmatory factor analysis,
+it can be used to similar ends:
 To assess whether a theoretical measurement model holds.
@@
 Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA [@donnellanResilientOvercontrolledUndercontrolled2010].
-One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices.
+
+One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices (see Model Fit Indices).
 Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
 Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
-either qualitatively or quantitatively, using significance tests for pre-specified values.
-A recent study by @maene_perceived_2022 applied this principle to 
-test whether identities of Belgian adolescents with migration background can be 
-summarized as those relating to their heritage, national and regional identity.
+either qualitatively or quantitatively, using significance tests for pre-specified values (see Inference).
+
+Appendix A is a tutorial to confirmatory LCA in `tidySEM`,
+based on a recent study that tested the theory that identities of Belgian adolescents with migration backgrounds could be summarized as those relating to their heritage, national and regional identity [@maene_perceived_2022].
+
+### Exploring Heterogeneity
+
+Another use case is to explore whether a population is comprised of latent subpopulations, and if so, how many.
+The primary focus in this use case is on class enumeration.
+For example, 
+<!-- researchers suspected the existence of different risk profiles of substance use, sexual behavior, and mental health outcomes, and used LCA to determine how many such risk profiles could be distinguished [@hopfer_social_2014].
+-->
+<!-- Another  -->
+one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
+and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
+[@gromatsky_characteristics_2022].
+Knowing that the population of suicidal service members is heterogeneous can inform future research, policy, and care provided.
+Appendix B offers a tutorial for exploratory LCA,
+based on the aforementioned study of the distribution of ocean microplastics [@alkema_maximizing_2022].
 
@@
 
-Another use case relies on LCA's similarity to confirmatory factor analysis.
-This approach is useful when class membership is measured imperfectly by several indicators.
-This approach can be used to partial out measurement error and restore most likely class membership.
-For example, a recent study found descrepancies between self-reported employment status and official register data. 
-While it might be tempting to assume that the
-register data is free of error, this has been shown to be false.
-The authors used a mixture model to estimate the
-"error corrected employment status" by using two indicators of the same question
-and identifying random and systematic measurement error in both indicators 
-[@pavlopoulos_measuring_2015].
+LCA can also be used as a measurement model, similar to confirmatory factor analysis.
+This is useful when class membership is measured imperfectly by several indicators:
+LCA can partial out measurement error and restore most likely class membership.
+For example, one study found discrepancies between self-reported employment status and official register data [@pavlopoulos_measuring_2015]. 
+While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
+A mixture model was used to estimate employment status, correcting for measurement error, from both  self-report and registry data.
+The study investigated random and systematic measurement error in both indicators,
+which reveal strengths and weaknesses of both types of assessment.
 
 This example also illustrates that LCA allows us to evaluate the reliability of different indicators [@geiser_is_2014].
-High quality indicators will be strongly related to the
-latent variable and will lead to good class separation.
+<!-- High quality indicators will be strongly related to the -->
+<!-- latent variable and will lead to good class separation. -->
 A continuous indicator that discriminates well between classes will show large differences in class-specific means;
@@
 This information can be used to select the indicators that are most diagnostic of class membership.
-For an example of research on differential item functioning,
+For an example of LCA with a focus on differential item functioning,
 consider a study on university admission tests of Saudi Arabian students [@tsaousis_measurement_2020].
-LCA revealed three
-latent classes: high-, average-, and low-scorers. Subsequently, the authors
-used the Multiple Indicator Multiple Causes model to identify both uniform
-and non-uniform differential item functioning. The main finding was that gender
-was a potential source of differential item functioning for latent class
-indicators. Therefore, gender should be included as a covariate when fitting 
-a LCA model in order to obtain unbiased estimates.
+<!-- LCA revealed three -->
+<!-- latent classes: high-, average-, and low-scorers. Subsequently, the authors -->
+<!-- used the Multiple Indicator Multiple Causes model to identify both uniform -->
+<!-- and non-uniform differential item functioning. The main finding was that gender -->
+<!-- was a potential source of differential item functioning for latent class -->
+<!-- indicators. Therefore, gender should be included as a covariate when fitting  -->
+<!-- a LCA model in order to obtain unbiased estimates. -->
 
@@
 
-LCA can also be used to reduce the dimensionality of data,
-by reducing many variables to a few prototypes.
-This is similar to the use of factor analysis,
-with the key distinction that factor analysis accounts for linear covariance between indicators,
+LCA can also be used to reduce high-dimensional data
+to just a few prototypes.
+Factor analysis is also sometimes used for dimension reduction,
+but only accounts for linear covariance between indicators,
 whereas LCA can accommodate complex - but discrete - patterns.
-As this is a pragmatic application of LCA,
-the existence of a true categorical latent variable is questionable (cf. the section on Testing theory).
-Consequently, the results of this use case should be treated as descriptive and not "reified": i.e., the found classes should not be treated as evidence of the existence of a categorical latent variable.
+For example, imagine that there exist two indicators - $x$ and $y$, and some individuals score high on both, some score low on both, and some score high on $x$ but low on $y$.
+Factor analysis would only be able to capture the differences between the first two groups well; LCA captures differences between all three.
+Note that this is a pragmatic use of LCA;
+there is no theory about the existence of a categorical latent variable (cf. Testing theory).
+Consequently, the class solution should be treated as descriptive of the sample,
+and not be "reified" - treated as evidence of the existence of a categorical latent variable.
+
 The use of LCA for dimension reduction is common in longitudinal research,
@@
 albeit at a cost of some loss of information and, potentially, generalizability.
+Appendix C is a tutorial on LCGA,
+based on ongoing research on trajectories of depressive symptoms in military service members after deployment.
 <!-- Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean. -->
@@
 
-When using ordinal indicators,
-the need for dimension reduction can be especially pertinent.
+The need for dimension reduction can be especially pertinent when using ordinal indicators.
 For example, one study measured
@@
 
-### Class enumeration
-
-Another use case is to explore potential heterogeneity in a population,
-and to determine whether the population is comprised of latent subpopulations, and if so, how many.
-The primary focus in this use case is on class enumeration.
-For example, researchers suspected the existence of different risk profiles of substance use, sexual behavior, and mental health outcomes, and used LCA to determine how many such risk profiles could be distinguished [@hopfer_social_2014].
-Another explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
-and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
-[@gromatsky_characteristics_2022].
-
 ### Classification
@@
 For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough [@baughman_mixture_2006].
-Another study used LCA to not only classify the sample used to estimate the model, but also future cases [Zegwaard et al, in prep].
+Another study used LCA to not only classify the sample used to estimate the model, but also future cases [REF Zegwaard et al, in prep].
 The study identified four types of care providers among those who supported close kin with mental health problems.
@@
 
-Finally, LCA can be used to deal with data which violate certain
-assumptions. As discussed in the shoe size example, LCA can deal with
-violations of normality. In fact, LCA assumes the population
-distribution is a non-normal mixture of $K$ normal distributions, and it
-can recover the value of $K$, i.e. generate a $K$-class solution.
-For instance, @alkema_maximizing_2022 studied ocean plastic debris. 
-Plastic particle width and height are non-normally distributed. The authors 
-demonstrate that the non-normal distributions of particle dimensions are
-due to the measurements being a mixture composed of several particle types, 
-each with its own normal distribution.
-
+Finally, LCA can be used to deal with data which violate certain assumptions.
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughmanMixtureModelFramework2007].
+These models estimate one class to explain the "true zeros",
+and another class with a different distribution to explain all oter values, which can include zero.
+Note that this LCA has different models for both classes,
+and thus falls beyond the scope of this paper.
+<!-- CJ: This has been said elsewhere: -->
+<!-- In fact, LCA assumes the population -->
+<!-- distribution is a non-normal mixture of $K$ normal distributions, and it -->
+<!-- can recover the value of $K$, i.e. generate a $K$-class solution. -->
+<!-- CJ: This example is not perfect, we need another example still.. -->
+For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
+A mixture of two normal distributions with distinct variances described the observed distribution better.
 <!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical.
@@
 class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
-These were generalized to be relevant to all types of LCA, and updated
-to current best practices, as explained below.
-In order to aid researchers working with latent trajectory models,
-@van_de_schoot_grolts-checklist_2017 developed a protocol called
-Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
+These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.
+<!-- In order to aid researchers working with latent trajectory models, -->
+<!-- @van_de_schoot_grolts-checklist_2017 developed a protocol called -->
+<!-- Guidelines for Reporting on Latent Trajectory Studies (GRoLTS). -->
 
@@
 patterns and violations of assumptions that had not been considered
-prior to data collection. Special attention should be paid to level of
-measurement of the indicators.
-Finite Gaussian mixture models (including
-LPA) are only suitable for continuous variables.
-Indicators with an
-ordinal level of measurement are likely to violate the assumption of
-within-class normal distributions of mixture models [see
-@vermunt_k-means_2011].
-Personal experience consulting on LCA methods
+prior to data collection.
+The function `descriptives(data)` provides extensive descriptive statistics,
+including the number of unique values,
+variance of continuous and categorical variables,
+missingness, skew and kurtosis (peakedness of the distribution). 
+However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
+Plots convey additional information; specifically, density plots for continuous variables, and bar charts
+for categorical ones.
+Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.
+
+Particular points of attention include measurement level of the indicators.
+Recall that `tidySEM` can estimate mixture models with either numeric (type `numeric` and `integer`) or `ordered` indicators.
+The column `type` should thus only contain these types.
+Also check that the number of `unique` values match the intended measurement level.
+Numeric variables with few unique values are typically better modeled as ordinal.
+Our experience consulting on LCA methods
 and moderating the `tidyLPA` Google group suggest that the
@@
 normal distributions.
-To illustrate the problem, consider what happens when estimating a 5-class solution on a 5-point Likert scale.
-Each class-specific mean could describe a single response
+To illustrate the problem, consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
+Each class-specific mean would describe a single response
 category,
-and a class-specific variance component would be nonsensical.
-In sum, Likert-type scales are rarely suitable for Gaussian mixture modeling;
-latent class analysis with ordinal indicators is more appropriate.
-
-Extensive descriptive statistics (including the number of unique values,
-variance of categorical variables, and missingness; see next subsection)
-can be obtained using `tidySEM::descriptives(data)`. Note,
-however, that sample-level descriptive statistics are of limited value
-when the goal of a study is to identify subsamples using latent class
-analysis. Plots (density plots for continuous variables, and bar charts
-for categorical ones) may be more diagnostic. Note that density plots
-can also aid in the choice of the number of classes, as further
-explained in the section on visualization. Descriptive statistics and
-plots can be relegated to online supplements, provided that these are
-readily accessible [consider using a GitHub repository as a
-comprehensive public research archive, as explained in
-@van_lissa_worcs_2021].
+and a class-specific variance component would be empirically underidentified.
+In sum, variables with few unique values might not be suitable for Gaussian mixture modeling;
+LCA with ordinal indicators is more appropriate.
+<!-- Descriptive statistics and -->
+<!-- plots can be relegated to online supplements, provided that these are -->
+<!-- readily accessible [consider using a GitHub repository as a -->
+<!-- comprehensive public research archive, as explained in -->
+<!-- @van_lissa_worcs_2021]. -->
 
@@
 
+<!-- Any variables that of class `character` or `factor` must be recoded to ordinal. -->
+<!-- Nominal variables must be recoded to multiple ordinal dummy variables, e.g., using the function `mx_dummies()`.  -->
 LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
@@
 such that it is sensible to estimate means and (co)variances.
-Note that this assumption is likely violated for Likert-type scales.
 For the remaining three measurement levels,
-some data preprocessing is required.
+some preprocessing is required.
 The exact data types accepted by `tidySEM` at the time of writing are `numeric` and `ordered`.
-To correctly specify ordered variables,
-such as Likert scales
+To correctly specify ordered variables
+such as Likert scales,
 use the function `OpenMx::mxFactor()`.
 Binary variables,
-such as presence or absence of some symptom, are also treated as ordinal.
-Finally, nominal variables are converted to binary indicators of group membership via dummy coding, for example, using `mx_dummies()`.
+such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
+Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using `mx_dummies()`.
 
@@
 
-<!-- A third consideration in preprocessing relates to the univariate distributions of indicators. -->
-<!-- For example, the `plas_depression` data included with `tidySEM` show extreme right skew; -->
-<!-- these data assess depression, and as such,  -->
+A third consideration in preprocessing relates to the univariate distributions of indicators.
+For example, the `plas_depression` data show extreme right skew in depressive symptoms:
+most respondents report few symptoms,
+and few respondents report many symptoms.
+Similarly, the `alkema_microplastics` data are extremely right-skewed: Very small plastic fragments abound in the oceans, and large pieces are rare.
+These distributional characteristics must be taken into account either in preprocessing, or model specification.
+In preprocessing, the data can be transformed to reduce skew.
+After reducing skew, a simpler LCA model may fit the data well.
+Alternatively,
+one can specify a model that accommodates the distributional idiosyncrasies.
+For example, allowing free variances across classes might result in one class with a near-zero mean and small variance to account for the fact that most respondents have few depressive symptoms,
+and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.
 
@@
 Multiple imputation is less suitable to LCA for two reasons. First,
-because LCA methods are often computationally expensive,
-and conducting them on multiple imputed datasets may be unfeasible.
-Second, because many multiple imputation methods have some assumption of normality, which is likely to be violated when it is assumed that the data come from a heterogenous population.
-FIML is not suitable for cases with complete missing data.
+because LCA methods are often computationally intensive,
+and conducting them on several imputed datasets multiplies the load.
+Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population.
 
@@
 @enders_applied_2022]: random missingness (MCAR);
-missingness contingent on *observed* variables that can thus be known and controlled for (MAR);
-and missingness related to unobserved variables, which can neither be known nor controlled for (MNAR).
-A so-called "MCAR" test is often reported in relation to missingness
-[@jamshidian_tests_2010].
-But note that this name is somewhat misleading,
+missingness contingent on *observed* variables that can be known and controlled for (MAR);
+and missingness related to unobserved variables,
+which can be neither known nor controlled for (MNAR).
+A so-called "MCAR" test is often reported [@jamshidian_tests_2010],
+but note that this name is somewhat misleading,
 as a significant test indicates that missingness is MAR,
@@
 the lead author has implemented this test in the `mice` package as `mice::mcar()`.
-
-To conclude; our recommendation is to inspect and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
-and to test for MAR using `mice::mcar()`.
+Our recommendation is to inspect and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
+and report the `mice::mcar()` test.
 As FIML estimation assumes that missingness is either MCAR or MAR,
-one would proceed with FIML regardless of the outcome of the test.
-One minor concern is that the approach `tidySEM` uses to determine starting values is *not* robust to
-partially missing data.
-This should not be a problem, however, as uses a global optimizer to estimate LCA models,
-which should be independent of starting values [@coranaMinimizingMultimodalFunctions1987].
+one would proceed with FIML regardless of the outcome.
+
+One minor concern is that `tidySEM` uses a different algorithm to determine starting values for complete data (K-means) and for missing data (hierarchical clustering).
+This should not be a problem, however, as the global optimizer used to estimate the model should be independent of starting values [@coranaMinimizingMultimodalFunctions1987].
+A second concern is that FIML is not suitable for cases with complete missing data.
+If such cases must be included, their values must be imputed first.
 
@@
 
-Any LCA model should be carefully specified based on theoretical and empirical considerations.
-The parameters of LCA depend on the measurement level of the indicators.
-LCA with continuous indicators can be parametrized in terms of means, variances, and covariances.
-Latent class growth analyses have the same parameters,
+In principle, LCA can be based on any parametric model.
+At present, `tidySEM` only facilitates estimation of models that can be specified in `lavaan` syntax.
+The available model parameters depend on the measurement level of the indicators.
+Continuous indicators can be parametrized in terms of means, variances, and covariances.
+LCGA and GMM have these same parameters,
 but with respect to the latent growth variables.
-LCA with ordinal indicators can be parametrized differently.
-The approach implemented in `tidySEM` assumes that each ordinal variable reflects an underlying standard normal distribution.
-The parameters are "thresholds" that correspond to quantiles of a standard
-normal distribution (with $N(\mu = 0, \sigma = 1)$).
-These thresholds
-are estimated based on the proportion of individuals in each of the
-response categories of the indicator variable. For example, a binary
+Ordinal indicators can be parametrized differently;
+the default approach in `tidySEM` assumes that each ordinal indicator reflects an underlying standard normal distribution, $N(\mu = 0, \sigma = 1)$.
+The parameters are "thresholds" that correspond to quantiles of this distribution, estimated based on the proportion of individuals in each of the
+response categories of the indicator variable.
+For example, a binary
 indicator has a single threshold that distinguishes the two response
@@
 Prior literature emphasized the importance of considering alternative model specifications [@vandeschootGRoLTSChecklistGuidelinesReporting2017].
-However, what has remained underemphasized is that alternative model specifications may be approached differently depending on whether an analysis is exploratory or confirmatory (see Use Cases).
-Most prior literature has emphasized exploratory applications of LCA [e.g., @nylund_deciding_2007]
+What has remained underemphasized is that alternative model specifications should be approached differently for exploratory versus confirmatory analyses (see Use Cases).
+Most prior literature has emphasized exploratory LCA [e.g., @nylund_deciding_2007].
 In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
@@
 it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
-Nonetheless, an important reason to consider alternative model specifications is that LCA lacks objective fit indices.
+Nonetheless, the theoretical model specification should still be compared to a few alternatives because LCA lacks objective fit indices.
 The fit of a theoretical model can thus only be assessed by comparison to other candidates.
-For this reason, the theoretical model should still be compared to a few others.
-A one-class model is a sensible benchmark, to test whether LCA is even appropriate.
+A one-class model is a sensible benchmark
+to test whether the theoretical model fits better than a model that assumes a homogeneous population.
 Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
 If the theoretical model has much better fit than other models,
+for example, based on IC weights,
 this provides evidence for the theory.
-If differences in fit are small, however, this may not  do not give cause to reject a theoretical model.
-The theoretical model may still be preferred, for example, to ensure comparability with prior work.
-The fact that the theoretical model received little differential support should of course be discussed,
-along with potential implications for future work.
-
+If another model fits substantially better,
+this provides evidence against the theory.
+If fit differences are small, the theoretical model may still be preferred, as theory is a relevant consideration in model selection [@jungIntroductionLatentClass2008].
+The fact that the theoretical model received little differential support should of course be discussed.
 <!-- This will ensure that the model accurately represents the data and provides reliable results. -->
@@
 <!-- The resulting model may have -->
-
 <!-- HIER -->
@@
 A second consideration in model specification is the risk of overfitting [@hastieElementsStatisticalLearning2009].
-Overfitting means that a model captures idiosyncratic noise in a sample,
-and thus generalizes poorly to new samples.
-It occurs when a model has many parameters relative to the number of observations
+Overfitting occurs when a model has many parameters relative to the number of observations,
+causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
 As explained before, the number of parameters in LCA scales with the number of estimated classes.
 Consequently, LCA with many classes have a potentially very high number of parameters.
-
 In Bayesian mixture modeling,
@@
 Although it is not known whether frequentist LCA exhibits the same behavior,
-this suggests that low class counts might indicate overfitting.
+low class counts are still an indicator of potential overfitting.
+The section on Class Enumeration addresses ways to guard against overfitting.
+
+Models can be specified in `tidySEM` in different ways.
+The wrapper functions `mx_profiles()` and `mx_lca()` construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
+The function `mx_profiles()` has two arguments to free elements of the covariance matrix: `variances` and `covariances`.
+By default, `mx_profiles()` fixes covariances to zero, and thus performs latent profile analysis.
+Finally, function `mx_mixture()` accepts a more flexible `model` argument, which can be a character string specifying a model in `lavaan` syntax,
+which is translated to `OpenMx` models via the function `as_ram()`,
+or a list of `OpenMx` models.
 
@@
 
-### Algorithm
-
-LCA parameters and model fit statistics can be estimated in a variety of
-ways. The choice of the estimator depends on the presence of missing
+### Estimation and Convergence
+
+LCA parameters and model fit statistics can be estimated in different
+ways.
+The choice of estimator depends on the presence of missing
 values, sample size, number of indicators, and available computational
-resources [@weller_latent_2020]. A commonly used technique is maximum
-likelihood (ML) estimation with the expectation-maximization (EM)
-algorithm as a local optimizer. Imagine we are estimating two
-parameters, e.g. the class-specific means $\mu_c$ on a continuous
+resources [@weller_latent_2020].
+LCA in `tidySEM` uses maximum likelihood estimation (ML), which can be conceptually understood as follows:
+Imagine we are estimating two
+parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator for two classes
 (ignoring the variance for now).
-EM attempts to find a combination of values for these two parameters that maximizes the log likelihood ($LL$) of the data.
-In practice, instead of maximizing $LL$,
-$-2*LL$ is minimized.
-Transforming the likelihood this way does not change the resulting parameter estimates,
-but it is mathematically convenient.
-Moreover, the $-2*LL$ is asymptotically $\chi^2$ distributed if the null-hypothesis is true, and thus this quantity can be used for model fit tests [@wilksLargeSampleDistributionLikelihood1938].
-To understand the optimization problem, imagine a three-dimensional landscape: 
-The X and Y dimensions represent potential values of the class-specific means, $X = \mu_1$ and $Y = \mu_2$, and the Z-dimension is determined by $Z = -2*LL$. 
+ML attempts to find a combination of values for these two parameters that minimizes minus two times the log likelihood ($-2LL$) of the data.
+<!-- The $-2LL$ is asymptotically $\chi^2$ distributed if the null-hypothesis is true, and can thus be used for model fit tests [@wilksLargeSampleDistributionLikelihood1938]. -->
+To understand the ML optimization procedure,
+imagine a three-dimensional landscape: 
+The X and Y dimensions represent potential values of the class-specific means, $X = \mu_1$ and $Y = \mu_2$, and the Z-dimension is determined by $Z = -2LL$. 
 The optimizer must find the
 deepest "valley" in this landscape,
-which reflects the combination of $\mu_1$ and $\mu_2$ that minimizes the $-2*LL$.
-The EM
+which reflects the combination of $\mu_1$ and $\mu_2$ that minimizes the $-2LL$.
+The ML
 optimizer behaves somewhat like a marble dropped in this landscape. It
-is dropped at some random point in space, and will roll into the nearest
-valley.
+is dropped at some random point in space, and rolls into the nearest valley (a combination of values for $\mu_1$ and $\mu_2$ that has a low $-2LL$).
 The challenge is that LCA models are often highly parametrized.
 This means that the landscape is not three-dimensional, but (very) high-dimensional,
-with potentially many valleys and sparse data to provide information about which valley is deepest.
+with many valleys and sparse data to provide information about which valley is deepest.
 If the marble rolls into one of these valleys,
@@
 The estimator has "converged".
-The risk is that the optimizer gets stuck in a shallower valley (a "local optimum"),
+
+A unique challenge in LCA is that not only model parameters,
+but also class membership must be estimated.
+Parameters in all classes are estimated from the same dataset,
+but for each class,
+observations are weighted by their probability of being in that class - which is in turn computed based on the class-specific parameters.
+The resulting chicken-and-egg problem is solved by the expectation-maximization (EM) algorithm:
+It alternates between estimating model parameters,
+calculating posterior class probabilities from those parameters,
+re-estimating model parameters using those new class probabilities as weights, and so on, until both have converged.
+
+A risk with this approach is that the optimizer might get stuck in a shallower valley (a "local optimum"),
 and never discovers the deepest valley (the "global optimum", or best solution).
 One solution to this problem is to drop many marbles at random places,
-compare their final $-2*LL$ values,
+compare their final $-2LL$ values,
 choose the solution
-with the lowest $-2*LL$, and make sure that several marbles replicated
+with the lowest $-2LL$, and make sure that several marbles replicated
 this solution.
@@
 because the algorithm begins with random starting values,
-it is inefficient because many marbles are likely to start very far away from any valley.
-Finally, if all sets of starting values are close to a local optimum,
-there is a risk that the global optimum is not discovered.
-An alternative solution that overcomes these challenges is to use a global optimizer, like the simulated annealing algorithm [SA, @coranaMinimizingMultimodalFunctions1987].
+it is inefficient because many marbles are likely to start far away from any valley.
+Additionally, if all sets of starting values are close to a specific local optimum,
+there is a risk that the global optimum is never found.
+An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing [SA, @coranaMinimizingMultimodalFunctions1987].
 SA does not behave quite like the marbles of the random starts approach;
 instead, it iteratively considers some "destination" in the landscape,
-and compares its $-2*LL$ to the current one.
-If the destination $-2*LL$ is lower, 
+and compares its $-2LL$ to the current one.
+If the destination $-2LL$ is lower, 
 the estimator moves there.
-However, if the destination  $-2*LL$ is *higher*,
-the estimator still moves there occasionally, with a certain probability.
-Think of this as occasionally flicking the marble back out of valleys,
+However, if the destination  $-2LL$ is *higher*,
+it still moves there occasionally, with a certain probability.
+Think of this as occasionally flicking the marble,
 to see if it rolls back into the current valley, or finds another,
@@
 then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernackiChoosingStartingValues2003].
-By default, `tidySEM` derives starting values using K-means clustering,
+By default, `tidySEM` derives starting values using K-means clustering [REF],
 computing the mixture model starting values by treating the K-means solution as multi-group model.
@@
 Finally, SA is followed up with a short run of the ML algorithm,
-as EM inherently produces an asymptotic
+as ML inherently produces an asymptotic
 covariance matrix for the parameters that can be used to compute
@@
 
-In case of any (suspected) convergence problems,
-note that `tidySEM` LCA models are `OpenMx` models.
-Thus, existing `OpenMx` functions to aid model convergence can be used, like `mxTryHard()` or `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
-Often, these functions return an acceptable solution for models with convergence issues.
+In practice, models do not always converge.
+Reasons may include local optima,
+bad starting values,
+incorrect model specification (the model may be too complex for the data, or conversely, too simple to capture important patterns),
+or poor data quality (small data sets, or distributions that are difficult to model).
+Convergence problems may be indicated by errors and warnings from the estimator,
+or they may be suspected, for example,
+when results change if the analyses are reproduced.
+
+Nonconvergence may be due to bad model specification.
+Consider, for example, an LCA with multiple ordinal indicators.
+If certain combinations of responses never occur together,
+in general or within one of the classes,
+this can result in extreme conditional item probabilities (exactly 0 or 1).
+Such extreme values impede model convergence.
+This problem could be addressed by trying alternative model specifications (e.g., fewer classes),
+or by merging categories of the ordinal indicators with very few responses.
+Similarly problems occur with continuous indicators.
+For instance, consider a zero-inflated indicator and a two class LPA.
+To account for the excess zeroes, one class will have a mean of zero.
+It is difficult to estimate the standard deviation for this class, however, because all of its members have the same value.
+If standard deviations are constrained to be equal across classes,
+this model will likely not converge.
+This problem can be addressed by considering more appropriate model specifications (e.g., free class variances).
+
+Nonconvergence due to bad starting values
+can be overcome by manually specifying starting values:
+Specify the argument `run = FALSE` in the call to `mx_mixture()` and related functions.
+This returns the model with default starting values, which can be updated.
+Nonconvergence and local optima may also be overcome by re-estimating the model while permuting the starting values.
+The `OpenMx` package contains a family of functions for this purpose: `mxTryHard()` and `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
+When poor data quality is to blame, the only solution is to collect more, or higher quality, data.
+
+<!-- A problem which can occur is that of inadmissible solutions. With binary -->
+<!-- indicators, LCA is modeling a cross-table with all the predictors. The -->
+<!-- problem with such cross-tables is that they will often contain empty -->
+<!-- cells, i.e. combinations of responses that never occur together. This -->
+<!-- problem is reflected by extreme conditional item probabilities (as in -->
+<!-- exactly 0 or 1). Such boundary parameter estimates could indicate that -->
+<!-- the solution is invalid [@geiser_data_2012]. Boundary parameter -->
+<!-- estimates can also happen with continuous indicators. For instance, if -->
+<!-- we have a zero-inflated normal distribution and a two class solution, -->
+<!-- one class might have the mean of zero and its standard deviation cannot -->
+<!-- be determined since there is little variance. This too could be a sign -->
+<!-- of an invalid solution, warn that too many classes were extracted, or -->
+<!-- indicate a local optimum [@geiser_data_2012]. -->
 
@@
 Class enumeration refers to the process of determining the appropriate number of classes.
-This topic is closely related to model specification,
-because the number of latent classes also influences the number of model parameters.
-As explained before,
-LCA can be done in an exploratory or in a confirmatory fashion.
-In the exploratory case, the set of models typically consists of a sequence from 1 to some higher number of classes $K$.
-The maximum number of classes $K$ may be chosen a-priori, or on theoretical grounds, or dictated by the data:
-after a certain number of classes,
-convergence problems often arise, 
-or the proportion of cases assigned to the smallest classes may be so low that such a class would not be locally identified.
+It is approached differently for exploratory versus confirmatory LCA.
+In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes $K$. 
+The number $K$ may be chosen a-priori or dictated by the data.
+It is sensible to limit $K$ to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
+Above a certain number of classes,
+convergence problems typically arise, 
+or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
 In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
 the proportion of cases assigned to the excessive classes converges to zero as the sample size increases [@rousseauAsymptoticBehaviourPosterior2011].
-It is not known whether this phenomenon also occurs in the frequentist case;
+It is not known whether this also happens in frequentist LCA;
 nonetheless, near-zero class memberships should be taken into account when determining the maximum $K$.
-It is sensible to limit $K$ to the maximum number of classes that result in valid and interpretable solutions.
-In the confirmatory case,
-the set of models with different numbers of classes to be compared is typically smaller;
-its purpose is to provide context for the relative fit of the theoretical model.
-
-During class enumeration, the set of models is compared on several statistical and substantive criteria.
-In the exploratory case, these criteria may be decisive in selecting a final model;
+
+In confirmatory LCA,
+the set of models with different numbers of classes to be compared is typically smaller.
+Its purpose is to provide context for the relative fit of the theoretical model, or to compare multiple theoretical models.
+
+In class enumeration, the set of models is compared on several statistical and substantive criteria.
+In exploratory LCA, these criteria may be decisive in selecting a final model;
 in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
@@
 information criteria are typically used to assess relative fit.
-Information criteria are computed from the -2\*log-likelihood ($-2LL$),
-which is the most fundamental measure of model fit.
-A penalty for model complexity is added to the $-2LL$ to penalize more complex models.
-Information criteria thus balance fit and complexity.
+Information criteria are computed from the $-2LL$,
+which is the most basic fit measure.
+Information criteria add a penalty for model complexity to the $-2LL$, and thus balance fit and complexity.
 The lower
@@
 Although a simulation study showed that the saBIC performed well for class enumeration,
-its penalty is not based on theory.
+its penalty is not based on theory [REF].
 BIC also performs well and is theoretically substantiated;
@@
 One option is to select a specific fit index, such as the BIC,
-before analyzing the data.
+before seeing the results.
+This choice can be preregistered to avoid concerns about cherry picking.
 Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
@@
 
-Given a specific choice of information criterion,
+Given a specific IC,
 it is also possible to calculate so-called "IC-weights",
-which indicate the relative support the data provide for each model in the set, expressed as a proportion of total support [@wagenmakersAICModelSelection2004].
-In `tidySEM`, IC-weights can be calculated by calling `ic_weights(x = table_fit(), ic = "BIC")` on a model fit table.
-Note, however, that IC-weights will (strongly) prefer the model with the lowest IC value;
+which indicate the relative support the data provide for each model in the set [@wagenmakersAICModelSelection2004].
+This is particularly useful for confirmatory LCA,
+as it allows one to quantify the support for the theoretical model.
+IC-weights are calculated by calling `ic_weights()` on the model fit table.
+Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
 the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
-
 <!-- The general objective of information criteria is to evaluate the model's  -->
@@
 
-### Nested model tests
+### Nested Model Tests
 
@@
 it is possible to test the difference in model fit using a LR test.
-One challenge is that the LR test statistic is not asymptotically $\chi^2$ distributed in LCA.
+One challenge is that the LR test statistic is not asymptotically $\chi^2$ distributed for LCA.
 The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR [REF lo 2001].
-This test is implemented in `tidySEM`, and is part of the default output of `table_fit()`.
+This test is implemented in `tidySEM`, and is part of the default output of `table_fit()`, comparing subsequent rows of the table.
 
@@
 and fitting the $k$- and $k-1$-class models to each simulated dataset [REF Nylund et al., 2007].
-The p-value of this test is defined as the proportion of $B$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
+The p-value of this test is the proportion of $B$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
+Simulation studies suggest that the BLRT has greater statistical power than the LRT,
+although both performed comparably well at detecting the true number of classes [REF Nylund 2007, REF Tein 2013].
 A crucial limitation of the BLRT is that it is very computationally expensive.
 With this in mind, `table_fit()` does not provide the BLRT by default.
-It is possible to compute it for a shortlist of likely candidate models,
-using the function `tidySEM::BLRT()`.
-However, we have noticed that this function often runs out of memory or takes prohbitively long to compute.
-
-Simulation studies comparing the LMR and BLMR suggest that the latter has greater statistical,
-although both performed comparably well at detecting the true number of classes [REF Nylund 2007, REF Tein 2013].
-However, anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
+Specific models can be compared using the function `BLRT()`.
+
+Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
 returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
@@
 This suggests that, like existing ICs,
-LR tests do not always adequately balance fit and complexity in LCA.
+LR tests might not adequately balance fit and complexity in LCA.
 Better solutions for class enumeration are thus an important target for future research.
 
-One promising alternative solution is the so-called *lazy bootstrap* (LB),
-which is much more efficient because it is based on sample statistics of model-implied data generated from the $k-1$ and $k$-class models,
+A promising novel solution is the so-called *lazy bootstrap*,
+which is much more efficient than the BLRT because it is based on sample statistics of model-implied data generated from the $k-1$ and $k$-class models,
 rather than on fit statistics of LCA models fit to those data [@van_kollenburg_lazy_2018].
-It is similar to the posterior predictive p-value in Bayesian statistics,
-and describes how well each model is able to reproduce the observed data [see @mcelreath_statistical_2020].
-One version of the lazy bootstrap is already implemented in the commercial program latentGOLD.
-At the time of writing, we are developing extentions of the LB and investigating their utility for class enumeration.
+This simulation-based approach to inference is similar to the posterior predictive p-value in Bayesian statistics,
+which describes how well each model is able to reproduce the observed correlation matrix [see @mcelreath_statistical_2020].
+One version of the lazy bootstrap is already implemented in the commercial program latentGOLD [REF].
+At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
 Future versions of `tidySEM` will likely incorporate this test.
@@
 <!-- 1. We do not consider solutions with minimum posterior classification probability < .90 because poor class separability compromises interpretability of the results -->
-
 <!-- Nevertheless, researchers can set an a-priori threshold for the minimum proportion of cases per class, -->
 <!-- for example by considering only solutions with a number of participants at least five times as large as the number of parameters per class. -->
-
-In most use cases of LCA, the goal is to identify 
-subgroups that are internally homogeneous and externally distinct.
-Classification diagnostics assess whether this is the case.
-These diagnostics are distinct from model fit indices,
-because a model can fit the data well but show poor latent class separation [@masyn_latent_2013].
-Classification diagnostics should therefore not be used for model selection.
-Nonetheless, they can help restrict the search space of acceptable models.
-Interpretability should always be a consideration when considering
-different class solutions [@nylund-gibson_ten_2018].
-If the goal is to interpret properties of (members of) classes,
-it makes sense to consider only models that meet certain minimum thresholds for classification diagnostics.
-
-All classification diagnostics are derived from the posterior classification probabilities: an $n \times k$ matrix with the probability that every individual $n$ belongs to latent class $k$.
+Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
+These diagnostics are not model fit indices and should not be used for class enumeration,
+because a model can fit the data well and still show poor latent class separation [@masyn_latent_2013].
+However, classification diagnostics can be used to eliminate models:
+If one analysis goal is to interpret properties of (members of) classes,
+it makes sense to only consider models that classify well [@nylund-gibson_ten_2018].
+Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.
+
+All classification diagnostics are derived from the posterior classification probabilities: an $n \times k$ matrix called $P$, with the probability that every individual $n$ belongs to latent class $k$.
 When classification accuracy is high, and classes are distinct,
 each individual's posterior class probabilities are high for one class, and low for the remaining classes.
-This matrix of individual classification probabilities $P$ is obtained by running `class_prob(res)` and extracting the element labeled `"individual"`.
-Note that one column is added to the matrix $P$:
-the final column represents most likely class membership $M$,
-which is obtained by assigning each observation to the class corresponding to their maximum posterior class probability.
-Thus, if an individual has posterior probabilities `c(0.2, 0.5, 0.3)` for classes 1-3,
-they would be assigned to class 2 ($M = 2$).
-The predicted class membership is often of interest to researchers who wish to carry out follow up analyses.
-Note, however, that such follow-up analyses should take classification error into account. The function `BCH()` does this automatically.
-
-The matrix $P$ can be summarized in different ways, and the following summaries are provided by default:
-First, a frequency table of the latent categorical variable,
-based on posterior probabilities, referred to as $F_P$ (`"sum.posterior"`).
-It is obtained by calculating the column sums of $P$.
-These numbers are fractional because individuals can contribute partially to multiple classes.
-This table thus takes classification error into account.
-
-Second, most likely class membership based on the highest posterior class probability, $F_M$ (`"sum.mostlikely"`).
-This is the frequency table of the variable $M$, and it ignores classification error.
-Note that this table can differ from $F_P$,
-where some observations may partly contribute to multiple classes.
+The individual classification probabilities $P$ are obtained by running `class_prob(res, type = "individual")`.
+Note that one column is added to the matrix, with each individual's most likely class membership $M$.
+$M$ is obtained by assigning each individual to the class corresponding to their maximum posterior class probability.
+<!-- Thus, if an individual has posterior probabilities `c(0.2, 0.5, 0.3)` for classes 1-3, -->
+<!-- they would be assigned to class 2 ($M = 2$). -->
+Predicted class membership $M$ is often of interest to researchers who wish to carry out follow up analyses.
+Note, however, that follow-up analyses should take classification error into account.
+The function `BCH()` does this automatically.
+
+The matrix $P$ can be summarized in different way.
+For example, the estimated frequency table of the latent categorical variable can be obtained by summing the posterior probabilities by class (i.e., the column sums of $P$).
+This frequency table $F_P$ is obtained by calling `class_prob(res, type = "sum.posterior"`.
+These numbers are fractional because this table takes classification error into account.
+Individuals can contribute partially to multiple classes.
+
+Second, most likely class membership based on the highest posterior class probability, $F_M$, is obtained by running `class_prob(res, "sum.mostlikely")`.
+This is the frequency table of the variable $M$.
+Unlike $F_P$, $F_M$ ignores classification error.
 If every observation is classified with perfect accuracy,
-these tables will be identical.
+$F_M$ and $F_P$ are identical.
 Part of the output of `table_fit()` for LCA analyses
-are the minimum and maximum percentage of the sample assigned to a
-particular class, `n_min` and `n_max`, which are derived from $F_M$.
-Of particular interest is the smallest class proportion, `n_min`.
-When the smallest class (`n_min`) is less than 10% of the total sample size,
+are the minimum and maximum percentage of the sample assigned to a particular class, `n_min` and `n_max`, derived from $F_M$.
+This is reported by default because the smallest class proportion, `n_min`, is of particular interest.
+When the size of the smallest class is not much larger than the number of parameters per class,
 proceed with caution:
-The model might not be locally identified in that class.
-Check whether the number of participants in this class is larger than the number of parameters per class (e.g., five times as large),
-and check whether this smallest class has a meaningful substantive interpretation.
-
-To assess the classification accuracy, we turn to the third diagnostic table:
-The average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$ (`"avg.mostlikely"`):
+The model might not be locally identified in that class [@depaoliMixtureClassRecovery2013].
+Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
+This is not particularly meaningful,
+however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
+Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
+Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
+If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.
+
+To assess classification accuracy, we turn to the third diagnostic table:
+The average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$, `class_prob(res, type = "avg.mostlikely")`:
 for participants assigned to one particular class (rows) according to $M$,
@@
 The diagonal represents the probability that observations in each class will be correctly classified.
-
-Finally, the classification probabilities for the most likely latent class $M$ by latent class, $P_{p_{(M = m | C = c)}}$ is similar to the aforementioned table $P_{p_{(M = m)}}$,
+<!-- Mauricio: Can you fact-check this?!? I really struglle here -->
+
+The table of classification probabilities for the most likely latent class $M$ by latent class, $P_{p_{(M = m | C = c)}}$ is similar to the aforementioned table $P_{p_{(M = m)}}$,
 except that it accounts for classification uncertainty.
@@
 and off-diagonal elements can be conceptualized as measurement error.
+<!-- Mauricio: Can you fact-check this?!? I really struglle here -->
 The minimum and maximum of the diagonal of this table are given by default as part of the output of `table_fit()` for LCA analyses (`prob_min` and `prob_max`).
-These probabilities are a convenient way to summarize classification accuracy.
-The minimum probability is especially important however,
-as it can diagnose if there is at least one class with poor classification accuracy.
+These probabilities are of particular interest because they conveniently summarize classification accuracy per class.
+Both probabilities should be high, as this means that all classes are reliably classified.
+If the minimum probability is low,
+there is at least one class with poor classification accuracy.
 
 Finally, class separability can be summarized in a single statistic:
-the entropy criterion.
-In physics, the concept of entropy is associated with randomness or uncertainty.
-High entropy corresponds to maximum uncertainty,
+the entropy criterion [REF].
+<!-- Celeux, G., & Soromenho, G. (1996). An entropy criterion for assessing the number of clusters -->
+<!-- in a mixture model. Journal of Classification, 13, 195-212. -->
+In physics, high entropy refers to maximum randomness or uncertainty,
 and low entropy corresponds to a strict arrangement of the units of study.
 Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
-Low entropy, by contrast, means that every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
+Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
 Crucially, in `tidySEM`, the entropy criterion is defined as 1-entropy (a convention originating in Mplus).
-The interpretation of the entropy criterion is reversed:
+Its interpretation is reversed;
 0 means the model classification is no better than chance,
 and 1 means perfect classification. 
-<!-- Celeux, G., & Soromenho, G. (1996). An entropy criterion for assessing the number of clusters -->
-<!-- in a mixture model. Journal of Classification, 13, 195-212. -->
-As a rule of
-thumb, values above .80 are deemed acceptable and above .90 are considered good.
-Values close to 1 indicate that classes are very well separated,
-and treating the most likely class membership $M$ as observed in follow-up analyses is probably defensible.
+There are no rules of thumb for entropy,
+because entropy is also sensitive to the sample size (entropy tends to be lower in larger samples),
+number of latent classes (entropy tends to be lower for higher number of classes), 
+and number of indicators.
 Like other classification diagnostics,
@@
 
-### Interpreting Class Solutions
-
-An important outcome of LCA are conditional item probabilities, also
-known as class-specific item probabilities [@masyn_latent_2013],
-conditional response, or conditional solution probabilities
-[@geiser_data_2012]. They indicate the probability of an item being
-endorsed given that the observation belongs to a particular latent
-class. Conditional item probabilities can be obtained using
-`tidySEM::table_prob()`. If a particular item is endorsed by two or more
-classes at markedly different rates, it is said to discriminate well
-between the classes and is consequently considered a good indicator.
-Classes are considered highly homogeneous with respect to an item when
-for a particular item there is a distinct difference in conditional item
-probabilities for two or more classes. For instance, if an item is
-endorsed below 30% for one class and above 70% for another class, the
-classes have high homogeneity with respect to this item
-[@masyn_latent_2013]. Conditional item probabilities are the analogue of
-mean and standard deviation when the indicators are binary or ordinal.
-
-A problem which can occur is that of inadmissible solutions. With binary
-indicators, LCA is modeling a cross-table with all the predictors. The
-problem with such cross-tables is that they will often contain empty
-cells, i.e. combinations of responses that never occur together. This
-problem is reflected by extreme conditional item probabilities (as in
-exactly 0 or 1). Such boundary parameter estimates could indicate that
-the solution is invalid [@geiser_data_2012]. Boundary parameter
-estimates can also happen with continuous indicators. For instance, if
-we have a zero-inflated normal distribution and a two class solution,
-one class might have the mean of zero and its standard deviation cannot
-be determined since there is little variance. This too could be a sign
-of an invalid solution, warn that too many classes were extracted, or
-indicate a local optimum [@geiser_data_2012].
-
-### Label Switching
-
-The final class solution will usually discover and enumerate several
-classes. The class ordering, however, is completely arbitrary. The class
-labeled as Class 1 in one solution may become Class 2 or Class 3 in
-another model, even when the only difference between the models is in
-their starting values. Label switching is something to be mindful of
-when comparing different LCA models [@masyn_latent_2013].
-
-The order of clusters is nondeterministic when using K-means in
-`tidySEM`.
-Therefore label switching is still a consideration.
-A simple solution to this is setting a random seed number one line prior to
-fitting the model, using `set.seed()`.
-We advise `tidySEM` users to always do so in order to
-circumvent label switching.
-Another solution is to specify starting values manually,
-but this is more challenging for two reasons:
-First, because this requires modifying the underlying `MxModel` objects, which is not straightforward.
-Second, if the starting values are not very good, label switching may still occur, and bad starting values could even lead to non-convergence.
-
-Class names should be chosen to accurately reflect group membership,
-as theoretically relevant names according to the interpretation of the class solutions.
-Overly simplified and generalized class names may prove misleading 
-to both audiences and researches alike leading to what is known as a naming
-fallacy [@weller_latent_2020].
+### Labelling Classes
+
+Class names should be chosen to accurately reflect theoretically relevant characteristics of group membership.
+These class names should aid the interpretation of the class solutions and serve as a shorthand throughout the manuscript.
+For example, one study used a dual-trajectory LCGA of cognitive and affective empathy development with gender as a covariate.
+Although this model has dozens of parameters,
+the resulting classes were labeled low, average, and high empathy because level differences were the most distinguishing feature.
+However, overly simplistic and general class names can be misleading, a phenomenon known as the *naming fallacy* [@weller_latent_2020].
+For example, it would be misleading to call one group "low empathy" and another group "medium empathy" if the empathy differences were non-significant,
+or if other differences, like the effect of the covariate gender, were more substantial than level differences.
+One way to avoid the naming fallacy is comprehensively reporting model parameters,
+and explicating that class names are just a shorthand.
+
+When labeling classes, it is important to note that LCA models are identified up to the class order.
+<!-- This means that, if an LCA model has converged on the global optimum solution, -->
+<!-- its results are stable if the analysis is replicated. -->
+This means that the order of classes is arbitrary and may change if the analysis is replicated.
+A simple solution to ensure replicable results is setting a random seed before fitting the model, using `set.seed()`.
+This will ensure that the classes stay in the same order upon replication,
+but that may not be the preferred order for, e.g., interpreting the results.
+To address this problem, the function `mx_switch_labels()` can be used.
+By default, the function orders classes from largest to smallest class count - but classes can also be ordered by values of a specific parameter (e.g., the mean intercept in a latent class growth model).
+Note that label switching is a much bigger problem when using Bayesian estimation,
+where it can occur during model estimation.
+This is beyond the scope of the present paper.
+
+### Follow-up Analyses
+
+LCA is rarely performed in isolation;
+often, researchers wish to relate class membership to auxiliary variables, or even auxiliary models.
+It is important to keep in mind that model-implied class membership is subject to classification error.
+Treating class membership as an observed variable in follow-up analyses disregards this error.
+The results will be biased,
+except when class separation is very high - as indicated by high (near 1) entropy and minimum posterior classification probability.
+It is therefore important to account for classification error.
+
+There are several different ways to do follow-up analyses in latent class analysis (LCA) while accounting for classification error.
+At present, the so-called BCH method is considered one of the best approaches [after its creators, ].
+It has been implemented in `tidySEM`.
+The advantage of the BCH method is that it not only allows for the comparison of individual variables, but also entire auxiliary models across latent classes.
+
+Using posterior class probabilities: In LCA, individuals are assigned to classes based on their maximum posterior class probabilities.
+However, posterior class probabilities can be used to assess the degree of uncertainty in class assignment.
+By using posterior class probabilities rather than discrete class assignments, follow-up analyses can account for classification error.
+
+Using sensitivity and specificity: In LCA, one can calculate sensitivity and specificity statistics to assess the accuracy of class assignment.
+These statistics can be used to adjust follow-up analyses, such as prevalence estimates or regression coefficients, to account for classification error.
+
+Using bootstrap methods: Bootstrap methods can be used to estimate the distribution of parameter estimates and classification error.
+This can be used to construct confidence intervals or to assess the stability of results across different samples.
+
+Using Bayesian model averaging: In Bayesian LCA, one can use Bayesian model averaging to account for model uncertainty.
+This can be used to account for the uncertainty in class assignments, resulting from different models fitting the data.
+
+In summary these methods are based on accounting the probability of the individual being classified in a specific group rather than a hard classification and adjusting the anlysis based on that, also taking uncertainty into account by bootstrapping, or Bayesian model averaging.
+
+c.j.vanlissa@tilburguniversity.edu
+What about the three-step method?
+The three-step method is another way to perform follow-up analyses in latent class analysis while accounting for classification error. The method consists of three steps:
+
+Estimate the latent class model: The first step is to estimate the latent class model using data on observed variables.
+This step produces estimates of the class-specific probabilities for each observed variable and the class membership probabilities for each individual.
+
+Use the estimated latent class model to classify individuals: The second step is to use the estimated model to classify each individual into the class that maximizes their posterior class probability.
+This results in a set of discrete class assignments.
+
+Adjust the analyses for classification error: The third step is to adjust the follow-up analyses for classification error.
+This can be done by using techniques such as weighting, matching, or regression adjustment, which take into account the posterior class probabilities and the uncertainty of class assignment.
+
+This method is based on a three-step process that allows the researcher to classify individuals, consider the uncertainty in the classification, and adjust the analysis accordingly.
+This can produce more accurate results and increase the robustness of the analysis.
 
 ## Best Practices in Reporting
-
-There is substantial merit in pre-specifying, and ideally preregistering, the criteria used for class enumeration.
-This is especially pertinent for LCA because there are no objective fit indices for LCA,
-and because there is a risk that information criteria and likelihood ratio tests will suggest a larger number of classes, even if there are concerns about overfitting, classification accuracy, or model interpretability.
 
@@
 @weller_latent_2020], but more work is needed.
-<!-- This does not belong here: -->
-<!-- @van_lissa_worcs_2020 -->
-<!-- developed WORCS, a workflow for open reproducible code in science. WORCS -->
-<!-- consists of step-by-step guidelines for research projects based on the -->
-<!-- TOP-guidelines developed by @nosek_promoting_2015. WORCS workflow can be -->
-<!-- easily implemented in R in form of an R package which facilitates -->
-<!-- preregistration, article drafting, version control, citation and -->
-<!-- formatting, among others [@van_lissa_worcs_2020]. -->
-
-TOP-guidelines emphasize the use of comprehensive citation (including
-referencing the software used in the analysis), as well as code and data
-sharing wherever possible [@nosek_promoting_2015]. @van_lissa_worcs_2020
-If the original data cannot be shared, sharing synthetic data is a good alternative.
-The function `worcs::synthetic()` provides a generic, non-parametric method to create a synthetic copy of data.
-When synthesizing data for LCA, it is also possible to synthesize data directly from the model.
-LCA models constructed in `tidySEM` are estimated using `OpenMx`,
-and as such, can benefit from all existing functions in that package.
-To synthesize data from a mixture model `mix`, use the function `mxGenerateData(mix)`.
-
-the entire research project is made reproducible so that others may
-download it and reproduce it with just one click; for guidance, see
-@van_lissa_worcs_2020.
-
-As the open science movement is gaining momentum, researchers are
-becoming increasingly aware how important it is that analyses can be
-reproduced and audited. In line with open science principles, one of the
-suggested reporting standards relates to reproducible code. In this
-context, it is important to note that user-friendly methods for
-estimating LCA models have predominantly been available in commercial
-software packages (e.g., *Mplus* and *Latent GOLD*). A potential
-downside of commercial software is that it restricts the ability to
-reproduce analyses to license holders, and prevents auditing research
-because the underlying source code is proprietary. To overcome these
-limitations, the present paper introduces new user-friendly functions in
-the `tidySEM` R-package that can be used to estimate a wide range of LCA
-models using the free, open-source R-package `OpenMx` as the backend. The 
-reporting guidelines described in this paper are adopted in `tidySEM` by 
-default. The `tidySEM` R-package thus makes advanced mixture modeling based 
-on best practices widely accessible, and facilitates the adoption of the
-estimation and reporting guidelines described in this paper.
+
+### Open and Reproducible Science
+
+<!-- TOP-guidelines emphasize the use of comprehensive citation (including -->
+<!-- referencing the software used in the analysis), as well as code and data -->
+<!-- sharing wherever possible [@nosek_promoting_2015]. @van_lissa_worcs_2020 -->
+<!-- If the original data cannot be shared, sharing synthetic data is a good alternative. -->
+<!-- The function `worcs::synthetic()` provides a generic, non-parametric method to create a synthetic copy of data. -->
+<!-- When synthesizing data for LCA, it is also possible to synthesize data directly from the model. -->
+<!-- LCA models constructed in `tidySEM` are estimated using `OpenMx`, -->
+<!-- and as such, can benefit from all existing functions in that package. -->
+<!-- To synthesize data from a mixture model `mix`, use the function `mxGenerateData(mix)`. -->
+
+<!-- the entire research project is made reproducible so that others may -->
+<!-- download it and reproduce it with just one click; for guidance, see -->
+<!-- @van_lissa_worcs_2020. -->
+
+<!-- As the open science movement is gaining momentum, researchers are -->
+<!-- becoming increasingly aware how important it is that analyses can be -->
+<!-- reproduced and audited. In line with open science principles, one of the -->
+<!-- suggested reporting standards relates to reproducible code. In this -->
+<!-- context, it is important to note that user-friendly methods for -->
+<!-- estimating LCA models have predominantly been available in commercial -->
+<!-- software packages (e.g., *Mplus* and *Latent GOLD*). A potential -->
+<!-- downside of commercial software is that it restricts the ability to -->
+<!-- reproduce analyses to license holders, and prevents auditing research -->
+<!-- because the underlying source code is proprietary. To overcome these -->
+<!-- limitations, the present paper introduces new user-friendly functions in -->
+<!-- the `tidySEM` R-package that can be used to estimate a wide range of LCA -->
+<!-- models using the free, open-source R-package `OpenMx` as the backend. The  -->
+<!-- reporting guidelines described in this paper are adopted in `tidySEM` by  -->
+<!-- default. The `tidySEM` R-package thus makes advanced mixture modeling based  -->
+<!-- on best practices widely accessible, and facilitates the adoption of the -->
+<!-- estimation and reporting guidelines described in this paper. -->
+
+Journals and funding bodies increasingly require research to be open and reproducible.
+For instance, many journals (and the APA) have adopted the TOP guidelines, which recommend comprehensive citation of literature, software, and data sources;
+sharing data and analysis code required to exactly reproduce analyses,
+and pre-registering study plans prior to data collection or analysis [@nosekTransparencyOpennessPromotion2016a].
+A key advantage of using open source software for LCA is that, compared to commercial software, it enables much greater adherence to open science principles.
+Sharing code has limited utility if it can only be evaluated in commercial software,
+as this restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because the underlying source code is proprietary.
+The `tidySEM` package enables performing LCA in free open source software,
+thus increasing the utility of shared code.
+
+Adopting FOSS makes it possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
+The Workflow for Open Reproducible Code in Science describes how to create a fully reproducible research archive,
+and the `worcs` R-package automates many of the steps involved [@vanlissaWORCSWorkflowOpen2021a].
+<!-- brings down the learning curve by implementing best practices for creating a reproducible digital research repository. 
+-->
+This workflow is based on three principles: First,
+the paper (or analysis section) is written as a *dynamic document* that contains both prose and analysis code.
+Results are automatically rendered as text, tables, and plots.
+Second,
+the research archive is *version controlled*, which means that the entire historical record of the project is retained.
+This version controlled archive can be made public.
+Third, *dependency management* ensures that all software used is documented, including its version number and provenance.
+Without dependency management, code might not work (for example, because it is not clear where certain functions come from),
+or its results may change when some dependencies are updated.
+By combining `tidySEM` and `worcs`,
+it is possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
+
+An additional argument for sharing data in the context of LCA is that reproducing LCA analyses requires individual participant data,
+not just sufficient statistics like many other methods (i.e., a correlation or covariance matrix).
+If the original data cannot be shared,
+sharing synthetic data at least enables reproducing and reusing analysis code.
+The `worcs` package provides a generic, non-parametric method to create a synthetic copy of data when data cannot be shared.
+It is also possible to synthesize data directly from the LCA model using `mxGenerateData()`,
+which often provides a better basis for reproducing the LCA.
+<!-- In sum, open science practices are particularly relevant for LCA, -->
+<!-- and the use of FOSS enables making analyses fully reproducible. -->
+
+<!-- Furthermore, the FAIR principles focus on *how* research output is shared, -->
+<!-- including data, code, and preprints. -->
+<!-- These outputs should be Findable in standardized search-indexed repositories, like [psyarxiv.org](https://psyarxiv.org) or [osf.io](https://osf.io), -->
+<!-- made Accessible by using open source software and file formats, -->
+<!-- made Interoperable so that others can build upon prior work, -->
+<!-- and licensed with permission for Reuse, thus enabling others to use and cite the work. -->
+
+### Preregistration
+
+LCA analyses inherently involve some subjectivity,
+because of the potential for disagreement between criteria for class enumeration and the lack of objective fit indices.
+In our experience, reviewers (and authors) often disagree about which criteria should be used.
+Preregistering the class enumeration criteria as part of the analysis plan reduces disagreement and eliminates concerns about cherry picking fit measures after the results are known.
+It is particularly important to pre-specify the criteria that will be used to determine the maximum number of classes,
+because there is a risk that information criteria and likelihood ratio tests will suggest a larger number of classes, even if there are concerns about overfitting, classification accuracy, or model interpretability.
+
+The aforementioned Workflow for Open Reproducible Code in Science facilitates creating a preregistration.
+More interestingly, it also enables creating a *Preregistration As Code* (PAC):
+A draft paper with reproducible analysis section based on fake (synthetic or simulated) data [@peikertReproducibleResearchTutorial2021].
+When the real data are accessed or collected,
+the analysis is simply reproduced with this new data file,
+and the results in the paper are automatically updated.
+PACs are quite unambiguous; the selection of the final model can even be automated by hard-coding the criteria into the analysis section.
+For an example of PAC, see @vanlissaComplementingPreregisteredConfirmatory2022.
+
+### Reporting Model Fit
+
+The function `table_fit()` reports an overview of fit indices, customized for LCA.
+In reporting model fit, it is important, firstly, to report the effective sample size, which should be the same for all models in order to compare relative fit measures.
+Secondly, it should be clearly reported,
+for example, in the strategy of analyses,
+which parameters are estimated across and within classes.
+In case of doubt, a full overview of the parameters is given by the `coef()` and `table_results()` functions.
+The number of parameters for each model should be included in the fit table.
+Thirdly, it is important that readers can assess the relative fit of different solutions.
+The most fundamental fit measure is the $-2LL$,
+as this is the quantity optimized during estimation.
+From these statistics, information criteria can be computed.
+By default, `table_fit()` includes the AIC, BIC, and saBIC.
+We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.
+
+Classification diagnostics should be reported,
+as they may determine whether models are excluded from consideration.
+Of particular interest is the diagonal of either the average posterior probabilities by most likely class membership,
+or of the most likely class membership by true class membership.
+The former indicates the average probability that individuals belong to the class they were assigned to,
+and the latter indicates the probability that individuals are assigned to a class if they actually belong to that class.
+By default, `table_fit()` reports the latter as `prob_min` and `prob_max`.
+The entropy is also of interest as an overall measure of class separation.
+Finally, the range of proportions of cases assigned to each class is important;
+the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant [see @depaoliMixtureClassRecovery2013].
+By default, `table_fit()` reports the minimum and maximum proportion of cases,
+based on most likely class membership.
+Fifth, the LMR or BLRT can be reported if it is part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
+By default, `table_fit()` reports the VLM for subsequent models in the table.
+
+### Reporting results
+
+Reporting LCA results involves reporting the class proportions,
+obtained via `class_prob(res, "sum.posterior")`,
+and model parameters, obtained via `table_results(res)`.
+The class proportions are obtained by standardizing the class weights:
+A $k$-class model has $k-1$ free class weights,
+and one weight constrained to 1 for model identification.
+The free weights are included in the model results,
+but are redundant with the aforementioned class proportions.
+Other model parameters may include means, (co)variances, regression coefficients (rarely), and thresholds for ordinal indicators.
+It is good practice to report the point estimate and confidence interval of these parameters.
+Although it is conventional to report p-values as well,
+note that the default p-values tests the hypothesis that the parameter is equal to zero.
+This hypothesis may be irrelevant in LCA when the focus is on parameter differences between classes.
+Reporting standard errors allows readers to compute p-values for other more informative hypotheses.
+Thresholds are more readily interpretable when converted to conditional item probabilities, using the function `table_prob()`.
+These indicate the probability of an item being
+endorsed given that the observation belongs to a particular latent [@geiser_data_2012].
+
+### Inference
+
+As mentioned before, the results table included null-hypothesis tests for all parameters by default.
+Whether these tests are informative depends on the study goals and scale of the data.
+For example, if LCA indicators are mean-centered,
+then the p-values of class-specific means indicate whether the mean in that class differs significantly from the sample average.
+It is also possible to test more informative hypotheses,
+using the function `wald_test()`.
+Equality-constrained informative hypotheses are specified using the `bain` syntax,
+and evaluated on the model object.
+This enables testing theoretical hypotheses - for example, if specific values for class-specific means had been hypothesized.
+The same function allows testing parameters for equality, either within or across classes.
 
@@
 
-Plots can greatly improve the interpretability of LCA models. There are
-several stages where this is an important consideration.
-
-First, in class enumeration, we want to compare several competing class
-solutions. This can be done by means of the AIC and BIC. Here, we
-suggest using an elbow plot with the $K-class\ solution$ on the X-axis,
-and information criteria value on the Y-axis. For continuous variables, 
-density plots can aid class enumeration. We demonstrate how to implement 
-both elbow and density plots in the Tutorial subsection of this paper.
-
-Second, once we have decided on the final class solution, we want to
-interpret the response patterns on the indicators for each latent class.
-While `tidySEM` can create a table showing the probability of each
-item's response endorsement (using `tidySEM::table_prob`), it may be
-easier to inspect these probabilities visually. For this reason, we
-created `ggplot2::plot_prob()`. The resulting shows response patterns on
-all indicators for each group. We give an example of this plot in the
-Tutorial subsection.
-
-## Myths and misunderstandings
-
-Relatedly, a recent publication claimed that an assumption of mixture
+In an informal review of published graphics for LCA,
+we found that most visualizations focused on the estimated model parameters,
+displaying class means as points connected by lines, with point shape and linetype indicating class membership.
+Note that the lines introduce spurious information.
+Furthermore, they omit uncertainty about class means and do not illustrate how well the model describes the data.
+
+The Grammar of Graphics addresses best practices in visualization, which we summarize here with specific emphasis on the application to LCA.
+Most importantly, good graphics start with the *data* in mind.
+Visualizing raw data allows one to assess class separability,
+and which aspects of the data are captured well or poorly by the model.
+A second consideration are which *aesthetics* to use.
+In a color plot, class membership can be indicated by color.
+In a black and white plot, it can be indicated by linetype and point shape.
+A third consideration is which geometric objects (geoms) to use.
+As mentioned before, common practice is to plot parameter estimates as points, and connect these points with lines.
+Points are suitable geoms to visualize point estimates and raw data.
+For continuous indicators, additional relevant geoms include error bars for standard errors and standard deviations,
+and density plots to visualize the distribution of raw data (smoothed with a kernel density function).
+In two dimensions, (co)variances can be visualized as ellipses.
+Finally, when plotting multiple class solutions,
+it makes sense to *facet* a plot into multiple panels by class solution.
+The plots in `tidySEM` are based on best practices as outlined in the grammar of graphics,
+and are implemented in `ggplot2`.
+As such, these plots can be further modified just like any other `ggplot2` object.
+One remaining limitation is that these best practices have only been implemented for LCA with continuous indicators, in `plot_profiles()`, `plot_density()` and `plot_bivariate()`, and for LCGA in `plot_growth()`.
+Plots for LCA with ordinal indicators are still rudimentary, more work remains to be done in this area.
+<!-- Plots can greatly aid the interpretability of LCA models. There are several stages where this is an important consideration. -->
+<!-- Visualization can aid class enumeration by means of a scree plot for the chosen information criterion. -->
+<!-- Calling `plot()` on the LCA object renders a scree plot by default, with the $1:K$-class solutions on the X-axis, -->
+<!-- and an IC on the Y-axis. -->
+<!-- The preferred model is typically either the one with the minimum IC value, -->
+<!-- or the inflection point at which the line graph flattens out. -->
+<!-- For continuous variables, plots can also aid in class enumeration and interpretation of the chosen solution. -->
+<!-- A good plot has the following properties:  -->
+
+<!-- Second, once we have decided on the final class solution, we want to -->
+<!-- interpret the response patterns on the indicators for each latent class. -->
+<!-- While `tidySEM` can create a table showing the probability of each -->
+<!-- item's response endorsement (using `tidySEM::table_prob`), it may be -->
+<!-- easier to inspect these probabilities visually. For this reason, we -->
+<!-- created `ggplot2::plot_prob()`. The resulting shows response patterns on -->
+<!-- all indicators for each group. We give an example of this plot in the -->
+<!-- Tutorial subsection. -->
+
+## Potential Pitfalls
+
+Based on our experience in statistical consultations,
+there are a few common pitfalls and misunderstandings about LCA.
+First, perhaps the most common problem we see in consultations is treating ordinal data as continuous.
+As explained previously, LCA for ordinal indicators should be used for Likert-type scales and other data with few unique values.
+A second common pitfall is omitting the 1-class solution from class enumeration.
+If a one-class solution has better or similar fit to models with a higher number of classes, LCA may not be appropriate.
+A one-class solution should thus always be included as a benchmark.
+Third, although it is widely understood that classification diagnostics (e.g., entropy, posterior probability, class counts) are not class enumeration criteria,
+they still have a role in class enumeration.
+Specifically, they can be used to exclude solutions that do not meet predefined criteria from consideration.
+Fourth, previous reporting standards called for reporting of sample descriptive statistics [see @schreiberLatentClassAnalysis2017].
+While we agree that this is due dilligence,
+sample descriptives have two limitations in the context of LCA:
+they are not sufficient to check or reproduce the analysis,
+and many use cases of LCA assume that the population is heterogeneous, which limits the utility of whole-sample descriptives.
+These two limitations can be addressed by data sharing and reporting descriptive statistics for each latent class separately.
+Fifth, a recent publication mistakenly stated that an assumption of mixture
 models is that observed indicators are normally distributed
-[@spurk_latent_2020]. This is incorrect. When the number of classes is
-greater than one, Gaussian mixture models assume that the observed indicators 
-are a mixture of multiple (multivariate) normal distributions. In our shoe
-size example, it can be seen that the population distribution is
-comprised of two normal distributions. When examined visually, the
-population distribution is evidently bimodal. The Shapiro-Wilk normality
-test ($W `r report(shoesize_shapiro[["statistic"]])`$, $p `r report(shoesize_shapiro[["p.value"]])`$ ) rejects the null hypothesis that the
-sample comes from a normally distributed population. Yet, this is a
-prototypical example of a mixed population distribution where LCA can
-discover latent groups. If the population distribution were instead
-normal, there would be no classes to extract as the whole population
+[@spurk_latent_2020].
+This is incorrect, as evident from the shoe size example:
+When the number of classes is greater than one,
+Gaussian mixture models assume that the distribution of observed indicators is a *mixture of multiple normal distributions*.
+<!-- In our shoe -->
+<!-- size example, it can be seen that the population distribution is -->
+<!-- comprised of two normal distributions. When examined visually, the -->
+<!-- population distribution is evidently bimodal. The Shapiro-Wilk normality -->
+<!-- test ($W `r report(shoesize_shapiro[["statistic"]])`$, $p `r report(shoesize_shapiro[["p.value"]])`$ ) rejects the null hypothesis that the -->
+<!-- sample comes from a normally distributed population. Yet, this is a -->
+<!-- prototypical example of a mixed population distribution where LCA can -->
+<!-- discover latent groups. -->
+If the indicators were normally distributed,
+there would be no classes to extract as the whole population
 would belong to a single (homogeneous) class.
-
-A recent paper suggested maximum likelihood with robust standard errors
-should be used when the observed indicators are not normally distributed
+Finally, the same paper suggested that robust standard errors
+should be used when indicators are not normally distributed
 [@spurk_latent_2020].
-This statement is incorrect;
-LCA does not have an assumption of marginal normality, and there is - to our knowledge - no literature on the merits of robust standard errors for LCA.
-
-<!-- and may lead readers -->
-<!-- to believe that they must use commercial software, as robust maximum -->
-<!-- likelihood is currently only implemented in Mplus and LatentGOLD. As -->
-<!-- explained before, mixture modeling assumes that observed data are a -->
-<!-- mixture of (multivariate) normal distributions; thus, the observed -->
-<!-- indicators will likely not be normally distributed. -->
-<!-- CJ: there is no literature that says robust standard errors are better with latent class analysis? -->
-
-<!-- Treating ordinal data as continuous -->
-<!-- People omitting the 1-class solution in class enumeration -->
-
-# Tutorial
-
-This is an example of exploratory LCA with ordinal indicators using `tidySEM`.
-At the time of writing, the `tidySEM` package includes two more vignettes to demonstrate latent profile analysis (LCA with continuous indicators),
-and Latent Class Growth Analysis.
-The present example uses a simulated dataset with ordinal indicators included in `tidySEM`.
-To view its documentation,
-run the command `?tidySEM::data_mix_ordinal` in the R console.
-
-## Loading the Data
-
-To load the data, one simply needs to attach the `tidySEM` package.
-For convenience, we assign the data to an object called `df`:
-
-```{r, include = TRUE, eval=F}
-# Load required packages
-library(tidySEM) 
-# Load data
-df <- data_mix_ordinal
-```
-
-## Examining the Data
-
-As per the best practices,
-the first step in LCA is examining the observed data.
-We use `tidySEM::descriptives()` to describe the data.
-
-```{r, include = TRUE, eval=F}
-tidySEM::descriptives(df)
-```
-
-As we can see, the output includes various descriptives of our dataset.
-For example, it provides invaluable information about the measurement level of the indicators,
-which is used to specify the model correctly.
-If these data had not been coded as ordinal factors,
-these descriptive statistics would have revealed that each variable has only 3 unique values.
-Attention should alo be paid to the pattern of
-missingness, as discussed in the *Handling Missing Data* section.
-The proportion of missing values is reported in the `"missing"` column.
-If any variables had missing values, we would report an MCAR test with `mice::mcar()`,
-and explain that missing data are accounted for using FIML.
-In our example, we see that there are no missing values, hence we
-proceed with our analysis.
-
-In this case, no preprocessing is necessary:
-all variables are on the same 3-point scale.
-For an example on LCA with extremely skewed variables, see the LCGA vignette.
-
-## Conducting Latent Class Analysis
-
-Before we fit a series of LCA models, we set a random seed using
-`set.seed()`.
-This is important because there is some inherent randomness in the estimation procedure,
-and using a seed ensures that we (and others) can exactly reproduce the results.
-
-Next, we fit the LCA models.
-As all variables are ordinal, we can use the convenience function
-`tidySEM::mx_lca()`, which is a wrapper for the generic function `mx_mixture()` optimized for latent class analysis with ordinal data.
-Any mixture model can be specified through `mx_mixture()`.
-At the time of writing, there are two other wrapper functions for special cases:
-`mx_profiles()`, for latent profile analysis, and `mx_growth_mixture()`, for latent growth analysis and growth mixture models.
-All of these functions have arguments `data` and number of `classes`.
-All variables in `data` are included in the analysis,
-so relevant variables must be selected first.
-Below, we fit 1 to 4 class solutions:
-
-```{r fitting_LCA, include = TRUE, eval=F}
-set.seed(123) # setting seed 
-res <- mx_lca(data = df, classes = 1:4) # fitting LCA 1 to 4 class solutions
-```
-
-In this example, all models converge without issues.
-If this is not the case, however, we can try to aid convergence using `mxTryHard()`, which expands the search for optimal parameter values,
-and related functions, `mxTryHardWideSearch()`, which considers a wider range of potential parameter values,
-and `mxTryHardOrdinal()` for models with ordinal indicators.
-We can inspect the class of resulting objects.
-
-```{r, include = TRUE, eval=F, echo = F}
-is(res)
-is(res[[1]])
-```
-
-<!-- Depending on your computer's computational power, -->
-<!-- this might take a while. -->
-<!-- As `tidySEM::mx_lca()` run several models at a time, the resulting object `res`  -->
-<!-- is a `mixture_list` class where each $K$ class solution is a list element, or -->
-<!-- more specifically an `OpenMx` model (`MxModel`). -->
-
-## Class Enumeration
-
-In class enumeration, we want to compare a sequence of LCA models fitted
-to the data. To aid the process, we create a model fit table using
-`tidySEM::table_fit()` with the results object as the input. As the
-output contains a lot of information on each of the four fitted models,
-we select a subset of helpful model fit indices and classification
-diagnostics.
-
-```{r fit_table, include = TRUE, eval=F}
-fit_table <- table_fit(res) # model fit table
-fit_table[ , c("Name", "LL", "Parameters", 
-               "AIC", "BIC", "Entropy", 
-               "prob_min", "prob_max", 
-               "n_min", "n_max")] # our selection
-```
-
-Our selection of fit indices and classification diagnostics includes:
-
-```{r, echo=F}
-Selection <- c(
-    "Name", "LL", "Parameters", 
-    "AIC", "BIC", "prob_min", 
-    "prob_max", "n_min", "n_max")
-
-Description <- c(
-    "Number of classes $K$",
-    "Basic fit measure of the natural log of the likelihood of the data",
-    "Total number of model parameters (in all classes)",
-    "Relative fit measure: Akaike Information Criterion",
-    "Relative fit measure: Bayesian Information Criterion",
-    "Lowest average posterior class probability for class cases were assigned to",
-    "Highest average posterior class probability for class cases were assigned to",
-    "Proportion of cases in the smallest class based on posterior class probability",
-    "Proportion of cases in the largest class based on posterior class probability")
-
-selection <- cbind(Selection, Description)
-
-knitr::kable(selection, format = 'pipe', 
-             caption = "Selection of Fit Indices and 
-                        Classification Diagnostics")
-```
-
-We discussed several possible strategies to select the final class
-solution. Here, we apply our own.
-
-To aid our interpretation of the results, we create an elbow plot
-showing the trends in information criteria across four models.
-
-```{r elbow_plot, include = TRUE, message=F, warning=F, eval=F}
-library(tidyverse) # for data-wrangling
-library(ggplot2) # for plots
-
-elbow_plot <- fit_table[ , c("Name", "AIC", "BIC")] # extract ICs
-elbow_plot <- pivot_longer(elbow_plot, cols = c("AIC", "BIC"), 
-                           names_to = "IC", values_to = "Value") # to long format
-
-ggplot(elbow_plot, aes(x = Name, y = Value, group = IC))+
-  geom_point(aes(color = IC))+
-  geom_line(aes(color = IC))+
-  labs(title="Elbow Plot of Information Criteria per Class Solution", 
-       x = "Class Solution", y = " Value")+
-  scale_color_manual(name = "Information Criterion", 
-                     values = c(AIC = 'blue', BIC = 'red'))+
-  theme_minimal()
-
-```
-
-From the elbow plot, we see that AIC has a lower penalty for model
-complexity. However, we are more interested in the BIC values, which are
-similar for the one, two and three-class solutions, but the four-class
-solution fits significantly worse. For this reason, we eliminate the
-four-class solution from the selection process.
-
-Then we examine the model fit table. As expected, the -2\*log likelihood
-falls successively with each added class. As previously stated,
-classification diagnostics should not be used for model selection, but
-they can be used to disqualify certain solutions because they are
-uninterpretable. We see that prob_min and n_min for the four-class solution is
-low, knowing that this solution also has a high BIC, we disqualify this
-solution.
-
-Out of the remaining three solutions, we notice that entropy is the
-highest for the three-class solution, and it has a satisfactory prob_min
-and n_min. Based on this, we retain the three class solution in model
-selection. Note that entropy for the one-class solution will always
-equal to one, as it is 100% true that every case is in that class. Based
-on the low entropy of the two-class solution, we eliminate this model.
-
-Finally, when comparing the one and three-class solutions, we inspect
-the information criteria. For BIC, the one-class solution fits better,
-but the difference is marginal. AIC tells us that the added complexity
-of having three classes still explains the data better than a one-class
-solution. Therefore, we select the three-class solution as our
-final-class solution.
-
-## Interpreting the Final Class Solution
-
-To aid our understanding of the final class solution, we use
-`ggplot2::plot_prob()` with the results of the three-class model as the
-input. The resulting graph shows response patterns on all the indicators
-for each group.
-
-If we want to know the probability of each response option's endorsement
-for each class, we can use `tidySEM::table_prob`. These are thresholds
-for ordinal dependent variables in the probability scale.
-
-```{r plot_prob, include = TRUE, eval=F}
-plot_prob(res[[3]]) # visualizing the response patterns for the final model
-table_prob(res[[3]]) # tabulating the response patterns for the final model
-```
-
-In the plot, we can see the distributions of the response probabilities
-on the indicators for each of the three classes. For instance, we see
-that in Class 1 the most common response to u2 is 2, while in Class 2
-and Class 3 this is 0. We can also see that response 1 is a rare
-response not forming the majority in any class. Class 2 distinguishes
-itself because the majority scores the response 0 category of u3 and u4,
-while in Class 1 and 2 this is not the case. Class 3 distinguishes
-itself because the most common response to u3 and u4 is 3.
-
-We can also interpret the response patterns numerically. It is a matter of 
-preferences on how to interpret these probabilities. Here is where you would
-**name** each class, such that each response pattern is theoretically meaningful.
-
-## Extracting Posterior Class Probabilities
-
-Another step is to extract posterior class probabilities. This is done
-by the use of `tidySEM::class_prob` with the results of the final class
-solution as the input.
-
-```{r extract_Post_Class_Prob, include = TRUE, eval=F}
-probs <- class_prob(res[[3]]) # extracting the posterior class probabilities
-probs$mostlikely.class # posterior probabilities by most likely class membership
-probs$individual # individual posterior class probabilities
-```
+This is also incorrect;
+to our knowledge, there is no literature on the merits of robust standard errors for LCA.
+Finally, we caution against step-wise class enumeration,
+whereby additional models are estimated until the selected class enumeration criteria no longer show improvement of model fit [e.g., see @spurk_latent_2020].
+This practice is problematic,
+because it is possible that an even higher number of classes would lead to a (significant) improvement in fit.
+Consider this thought experiment: there exist two groups of three relatively similar classes each (six in total).
+Step-wise class enumeration shows improvement in model fit going from a 1-class to 2-class model: the two classes capture the difference between the two groups of three classes.
+Then, no noteworthy improvement in model fit is found for adding more classes, until the 6-class solution is reached.
+In the step-wise approach, the analysis would be terminated after the 2-class model - even though the correct number of classes is 6.
+
+## SMART-LCA Checklist
+
+The best practices discussed in this paper are summarized in the SMART-LCA Checklist: Standards for More Accuracy in Reporting of different Types of Latent Class Analysis.
+
+1. Report item descriptives
+    + Are all items on similar scales?
+    + Is the measurement level correctly coded?
+    + Any other reasons why the observed (multivariate) distribution could not be described by the LCA (e.g., extreme skew)?
+2. Report missingness
+    + Proportion of missingness per variable
+    + MAR test
+    + Missing data method for handling missingness (typically FIML)
+3. Report model specification
+    + In exploratory LCA, compare all sensible alternative specifications
+    + In confirmatory LCA, justify alternative models on theoretical grounds
+    + Check that the observed (multivariate) distribution could theoretically be captured by the specified model (see #1).
+4. Class enumeration
+    + Justify the maximum number of classes $K$
+    + In exploratory LCA, estimate 1:$K$ classes
+    + In confirmatory LCA, estimate the theoretical number of classes, a 1-class model, and optionally other numbers of classes for comparison
+5. Justify the (preregistered) criteria used for class enumeration, which can include:
+    + Information criteria
+    + LR tests
+6. Justify the (preregistered) criteria used to eliminate models from consideration, including:
+    + Classification diagnostics
+    + Local identifiability (acceptable number of observations per parameter in each class)
+    + Model convergence
+    + Theoretical interpretability of the solution
+7. Report the following:
+    + Fit of all models under consideration
+    + Classification diagnostics for all models under consideration
+    + All parameters of the chosen model (means, (co)variances, thresholds)
+    + Proportion of cases per class
+8. Provide an informative plot of the chosen solution
+    + Show raw data
+    + Show parameters
+    + Show confidence bounds if possible
+    + Show multivariate distributions for multivariate models, if possible
+9. Account for classification inaccuracy in follow-up analyses
+    + If entropy > .95 and posterior classification probability for each class > .95, this is less crucial
+
+# Conclusions
+
+Write me
 

### 2023-01-22 manuscript.Rmd

@@
 There are several different ways to do follow-up analyses in latent class analysis (LCA) while accounting for classification error.
-At present, the so-called BCH method is considered one of the best approaches [after its creators, ].
+At present, the so-called BCH method is considered one of the best approaches [after its creators, REF].
 It has been implemented in `tidySEM`.
@@
 
-Using posterior class probabilities: In LCA, individuals are assigned to classes based on their maximum posterior class probabilities.
+<!-- Using posterior class probabilities: In LCA, individuals are assigned to classes based on their maximum posterior class probabilities.
 However, posterior class probabilities can be used to assess the degree of uncertainty in class assignment.
 By using posterior class probabilities rather than discrete class assignments, follow-up analyses can account for classification error.
-
-Using sensitivity and specificity: In LCA, one can calculate sensitivity and specificity statistics to assess the accuracy of class assignment.
-These statistics can be used to adjust follow-up analyses, such as prevalence estimates or regression coefficients, to account for classification error.
-
-Using bootstrap methods: Bootstrap methods can be used to estimate the distribution of parameter estimates and classification error.
-This can be used to construct confidence intervals or to assess the stability of results across different samples.
-
-Using Bayesian model averaging: In Bayesian LCA, one can use Bayesian model averaging to account for model uncertainty.
-This can be used to account for the uncertainty in class assignments, resulting from different models fitting the data.
-
-In summary these methods are based on accounting the probability of the individual being classified in a specific group rather than a hard classification and adjusting the anlysis based on that, also taking uncertainty into account by bootstrapping, or Bayesian model averaging.
-
-c.j.vanlissa@tilburguniversity.edu
-What about the three-step method?
-The three-step method is another way to perform follow-up analyses in latent class analysis while accounting for classification error. The method consists of three steps:
-
-Estimate the latent class model: The first step is to estimate the latent class model using data on observed variables.
+-->
+
+<!-- One way to account for classification error in follow-up analyses is using the three-step method.
+The first step is to estimate the latent class model using data on observed variables.
 This step produces estimates of the class-specific probabilities for each observed variable and the class membership probabilities for each individual.
-
-Use the estimated latent class model to classify individuals: The second step is to use the estimated model to classify each individual into the class that maximizes their posterior class probability.
+The second step is to use the estimated model to classify each individual into the class that maximizes their posterior class probability.
 This results in a set of discrete class assignments.
-
-Adjust the analyses for classification error: The third step is to adjust the follow-up analyses for classification error.
+The third step is to adjust the follow-up analyses for classification error.
 This can be done by using techniques such as weighting, matching, or regression adjustment, which take into account the posterior class probabilities and the uncertainty of class assignment.
-
-This method is based on a three-step process that allows the researcher to classify individuals, consider the uncertainty in the classification, and adjust the analysis accordingly.
-This can produce more accurate results and increase the robustness of the analysis.
+-->
 

### 2023-01-22 manuscript.tex

--- 2022-10-16 manuscript.tex
@@
 \usepackage{xcolor}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
   pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
-  pdfauthor={Caspar J. van Lissa1,2},
+  pdfauthor={Caspar J. van Lissa1,2, Mauricio Garnier-Villarreal3, \& Daniel Anadria1},
   pdflang={en-EN},
@@
 \title{Best Practices in Latent Class Analysis using Free Open Source Software}
-\author{Caspar J. van Lissa\textsuperscript{1,2}}
+\author{Caspar J. van Lissa\textsuperscript{1,2}, Mauricio Garnier-Villarreal\textsuperscript{3}, \& Daniel Anadria\textsuperscript{1}}
 \date{}
 
-\shorttitle{BEST PRACTICES LCA}
+\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}
 
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht\\\textsuperscript{3} Vrije Universiteit Amsterdam, Sociology}
 
 \abstract{%
-Latent class analysis is a family of techniques for identifying groups in data based
-on a parametric model. Its examples include mixture models, latent profile
-analysis, latent class analysis with ordinal indicators, and latent class
-growth analysis. Despite the popularity of this technique, there is limited
+Latent class analysis (LCA) refers to a family of techniques for identifying groups
+in data based on a parametric model.
+Examples include mixture models, LCA with ordinal
+indicators, and latent class growth analysis.
+Despite the popularity of this technique, there is limited
 guidance with respect to best practices in estimating and reporting mixture
@@
 source alternatives have remained somewhat inaccessible. This tutorial
-describes best practices for the estimation and reporting of latent class analysis,
+describes best practices for the estimation and reporting of LCA,
 using free and open source software in R. To this end, this tutorial
 introduces new functionality for estimating and reporting mixture
-models in the \texttt{tidySEM} R-package whose backend relies on
-the \texttt{OpenMx} R-package.
+models in the \texttt{tidySEM} R-package whose backend relies on \texttt{OpenMx}.
 }
@@
 
-Latent class analysis (LCA) is an umbrella term
-that refers to a number of techniques for estimating unobserved group membership
-based on a parametric model of one or more observed indicators of group membership.
+Latent class analysis (LCA) is an umbrella term that refers to a number
+of techniques for estimating unobserved group membership based on a
+parametric model of one or more observed indicators of group membership.
 The types of LCA have become quite popular across scientific fields,
-most notably finite Gaussian mixture modeling and latent profile analysis.
-Vermunt, J.K. et al. (2004) defined LCA more generally as virtually any statistical model
-where ``some of the parameters {[}\ldots{]} differ across unobserved subgroups''.
-
-Despite its popularity, there is a lack of standards for estimating and reporting LCA.
-While Van De Schoot, Sijbrandij, Winter, Depaoli, and Vermunt (2017) developed reporting guidelines
-for a specific type of LCA known as latent growth models,
-general reporting guidelines for all LCA are still lacking.
-This complicates manuscript review and assessment of the quality of published studies,
-and introduces a risk of misapplications of the technique.
-The present paper seeks to address this gap in the literature
-by suggesting updated guidelines for estimation and reporting of LCA,
-based on current best practices.
-Importantly, in order to lower the barrier of entry and ensure reproducibility of all examples,
-this paper exclusively relies on free, open source software in R.
-Our goal is to make best-practices in LCA widely accessible.
+most notably finite Gaussian mixture modeling and latent profile
+analysis. Vermunt et al. (2004) defined LCA more generally as
+virtually any statistical model where ``some of the parameters {[}\ldots{]}
+differ across unobserved subgroups''.
+
+Despite the popularity of LCA,
+two challenges impede broader adoption and correct application of the method.
+The first challenge is that most existing software that implements these models is commercial and closed-source.
+The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
+This introduces a risk of misapplications of the technique,
+and complicates manuscript review and assessment of the quality of published studies.
+The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
+based on current best practice.
+Furthermore, meeting the increasing demand for open source research software,
+this paper introduces new functionality in the R-package \texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free open source software (FOSS).
+
+\hypertarget{the-argument-for-free-open-source-software}{%
+\subsection{The Argument for Free Open Source Software}\label{the-argument-for-free-open-source-software}}
+
+Despite the popularity of LCA,
+most existing software is commercial and closed-source,
+for example, Mplus (\textbf{muthenMplusUserGuide1998?}), for which the R-package \texttt{MplusAutomation} offers augmented LCA functionality (\textbf{hallquistMplusAutomationPackageFacilitating2018?}), and LatentGOLD {[}REF{]}.
+A downside of commercial software is that its use is restricted to individuals and institutions who can afford a license.
+Another downside is that the proprietary source code cannot be audited, debugged, or enhanced by third parties.
+This goes against open science principles (\textbf{lamprechtFAIRPrinciplesResearch2019?}),
+incurs a risk that mistakes in the source code go unnoticed,
+and curbs progress as independent developers cannot fix bugs or add functionality.
+
+There are several compelling reasons why researchers should prefer free open source software (FOSS) instead.
+FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
+This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
+Furthermore, FOSS promotes collaboration within the scientific community,
+as its users can contribute to the development, maintenance, and support of the software.
+Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
+researchers can review and verify the methods and algorithms used,
+which helps to ensure the integrity and reliability of research findings.
+Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.
+
+Notable FOSS for LCA includes Mclust {[}REF{]}, depmixS4 {[}REF{]}, and OpenMx (Neale et al., 2016).
+A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
+or implement specific innovations (e.g., depmixS4).
+OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
+Remaining obstacles to the use of \texttt{OpenMx} is that it is not very user-friendly and its functionality for LCA is poorly documented.
+
+These limitations are addressed by the \texttt{tidySEM} package,
+which provide a user-friendly interface for LCA in \texttt{OpenMx}.
+The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting LCA, as described in this paper.
+The primary purpose of the \texttt{tidySEM} package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
+The user interface is simple, as models can be specified using the widely used \texttt{lavaan} syntax (Rosseel, 2012).
+The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
+but the resulting models can still be fully customized (low barrier, high ceiling).
+Specifically, LCA models constructed in \texttt{tidySEM} inherit from \texttt{OpenMx}' format \texttt{MxModel}, and remain compatible with downstream functions that process such models.
+Being FOSS, \texttt{tidySEM} facilitates fully-reproducible analyses and contributes to open science.
+It was developed with open science in mind,
+and adheres to the FAIR software principles (Findable, Accessible, Interoperable, and Reusable, \textbf{lamprechtFAIRPrinciplesResearch2019?}),
+and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{OpenSSF Best Practices} for FOSS software.
 
@@
 
-Latent class analysis is a group of methods for estimating unobserved groups
-based on a parametric model of observed indicators of group membership.
+LCA refers to methods for estimating unobserved
+groups based on a parametric model of observed indicators of group
+membership (\textbf{vermuntLatentClassAnalysis2004?}).
+It has become popular across scientific fields,
+under a number of different names;
+most notably (finite Gaussian) mixture modeling and latent profile analysis.
 The concept of LCA can be understood in different ways.
-A mixture model assumes that the study population
-is composed of \(K\) subpopulations or classes.
-It further assumes that the observed data are a mixture of data generated by class-specific models.
-The simplest univariate model is a normal distribution,
-which can be described with two parameters: the mean and the variance.
+Generally speaking, LCA assumes that the study population is composed of \(K\)
+subpopulations.
+It further assumes that the observed data are
+a mixture of data originating from those subpopulations;
+hence the name mixture model.
+Consider the simplest possible
+univariate ``model'', which is a normal distribution.
+This model has
+two parameters: the mean and the variance.
+LCA aims to estimate the values of those parameters across a known number of classes \(K\),
+as well as the probability of every individual \(i\) belonging to classes \(1 \ldots k\).
 Commonly, the same model is estimated across all classes,
 but with different parameters for each class (i.e., class-specific means and variances).
-Mixture modeling then estimates both the parameters for each class,
-and the probability that an individual belongs to each class.
-
-As an illustrative example, imagine that a detective wants to know
-if it would be possible to use mixture modeling to identify the sex of a suspect,
-based on footprints found at the crime scene.
-To test the feasibility of this approach,
-the detective records the shoe sizes and sex of 100 volunteers.
-The resulting observed data look like this:
+
+As an illustrative example, imagine that a detective wants to know if it
+would be possible to use mixture modeling to identify the sex of a
+suspect, based on footprints found at the crime scene. To test the
+feasibility of this approach, the detective records the shoe sizes and
+sex of 100 volunteers. The resulting observed data look like this:
 
@@
 
-A LCA is like confirmatory factor analysis,
-except that the continuous latent variable is substituted with a categorical latent variable.
-One difference between the two techniques is that
-factor analysis can be considered as a way to group observed \emph{variables} into latent constructs,
-and LCA groups \emph{individuals} into classes.
-In line with this distinction, LCA is sometimes referred to as a ``person-centered'' technique,
-and factor analysis as a ``variable-centered'' technique (Nylund-Gibson \& Choi, 2018).
-
-When the focus is on the model parameters in each group,
-LCA can be thought of as similar to a multi-group structural equation model.
-The main distinction is that group membership is not known a-priori,
-but is instead estimated -- with measurement error -- based on the data.
-Whereas in a multi-group model, the data are split by group and treated as independent samples,
-in LCA, all cases contribute to the estimation of all parameters in all groups.
-The relative contribution of each case to the parameters of each group
-is determined by that case's posterior probability of belonging to that group.
+Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling (\textbf{molenaarArbitraryNatureLatent1994?}).
+A mixture model is like confirmatory factor analysis,
+except with a categorical instead of a continuous latent variable.
+One difference between
+the two techniques is that factor analysis can be considered as a way to
+group observed \emph{variables} into latent constructs, and LCA groups
+\emph{individuals} into classes.
+While factor analysis seeks to describe the relations between variables,
+LCA seeks to describe unobserved groups and classify individual observations to those groups.
+In line with this distinction, LCA
+(mixture models) is referred to as a ``person-centered'' technique, and
+factor analysis as a ``variable-centered'' technique
+(Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+
+When the focus is on the model parameters in each group, LCA can be
+thought of as similar to a multi-group structural equation model. The
+main distinction is that group membership is not known a-priori, but is
+instead estimated -- with measurement error -- based on the data.
+Whereas in a multi-group model, the data are split by group and treated
+as independent samples, in LCA, all cases contribute to the estimation
+of all parameters in all groups. The relative contribution of each case
+to the parameters of each group is determined by that case's posterior
+probability of belonging to that group.
 
 When the focus is on each individual's estimated class membership,
-LCA can be thought of as a type of clustering algorithm.
-In line with this perspective, LCA is sometimes described as ``model-based
-clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
-Many clustering algorithms apply some recursive splitting algorithm to the data.
-By contrast ``model-based'' clustering refers to the fact that LCA
-estimates cluster membership based on a parametric model.
+LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
 Specifically, the posterior class probability that an individual belongs to a latent class
 can be computed from the likelihood of that individual's observed data under given the class-specific model.
+These posterior class probabilities can be used to weight follow-up analyses,
+or individuals can be assigned to classes based on their highest class probability.
+If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
+This perspective on mixture modeling is sometimes referred to as ``model-based
+clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
+Many clustering algorithms apply some recursive splitting algorithm to the data. By
+contrast ``model-based'' clustering refers to the fact that LCA estimates
+cluster membership based on a parametric model.
+Since a parametric model estimates only a fixed number of parameters,
+it can be a relatively parsimonious solution compared to non-parametric techniques.
 
 Finally, in the context of machine learning, LCA can be considered as an
-\emph{unsupervised classification} problem (Figueiredo \& Jain, 2002).
-The term \emph{unsupervised} refers to the fact that the outcome variable
--- true class membership -- is not known,
-and the term \emph{classification} refers to the fact that
-the algorithm is predicting a categorical outcome -- class membership.
+\emph{unsupervised classification} method (Figueiredo \& Jain, 2002).
+The term \emph{unsupervised} refers to the fact that the outcome variable --
+true class membership -- is not known, and the term \emph{classification}
+refers to the fact that the algorithm is predicting a categorical
+outcome: class membership.
 
@@
 
-In this paper, we use the term latent class analysis to refer to
-a family of techniques that estimate latent class membership
-based on a parametric model of observed indicators.
-From a historical perspective, the term was initially conceived to
-refer to analyses with categorical, usually binary indicators (Vermunt, J.K. et al., 2004).
-Nowadays, there are a number of related techniques,
-known by distinct names, that serve a similar purpose.
-The term ``latent class analysis'' seems most appropriate
-as an umbrella term for this broader class of models,
-as it only refers to the purpose of the analysis,
-and does not imply restrictions to the model used,
-or the level of measurement of the indicators.
-Given the abundance of terms in use for closely related classes of models,
-we will provide a rudimentary taxonomy of LCA methods.
-
-A common type of LCA is the \emph{finite Gaussian mixture model};
-a univariate analysis where the observed distribution of a single variable
-is assumed to result from a mixture of a known number of normal (Gaussian) distributions.
-The parameters of a finite Gaussian mixture model are
-the means and variances of these underlying normal distributions.
-The analysis of shoe sizes presented earlier is a canonical example of this type of analysis.
-In the multivariate case, with more than one indicator variable,
-the parameters of a mixture model are the means, variances, and covariances
-between the indicators (which can be standardized to obtain correlations).
-These parameters can be estimated freely, or set to be constrained across classes.
-
-The technique known as \emph{latent profile analysis (LPA)} is a special case of the mixture model,
-which assumes conditional independence of the indicators.
-Conditional independence means that, after class membership is accounted for,
-the covariances/correlations between indicators are assumed to be zero.
-This can be conceived of as a restricted mixture model with covariances fixed to zero.
-In some cases, such constraints will be inappropriate;
-for instance when the cohesion between indicators is expected to differ between classes.
-As an example, a LPA of ocean plastic particles
-found two classes of particles based on length and width:
-a class of smaller particles with a high correlation between length and width,
-meaning that these particles were approximately round or square in shape,
-and a class of larger particles with a low correlation between length and width,
-meaning that these particles were heterogenous in shape.
-From a theoretical perspective, this makes sense,
-because the smaller particles have been ground down to a more uniform shape by the elements.
-
-It is also possible to estimate a mixture model based on latent indicators.
-This means that, within each class,
-one or more continuous latent variables are estimated based on the observed indicators.
-Categorical latent variable membership is then estimated
-based on these continuous latent variables.
-A common application of this approach is in longitudinal research,
-where the indicators reflect one construct assessed at different time points.
-Examples of this approach include \emph{growth mixture models} (GMM)
-and \emph{latent class growth analyses} (LCGA).
+The term LCA was initially conceived to refer to analyses with categorical,
+usually binary indicators (Collins \& Lanza, 2009).
+It was later generalized to refer to any technique that estimates a categorical latent variable based on a
+parametric model of observed indicators, as it only refers
+to the purpose of the analysis, and does not imply restrictions to the
+model used, or the level of measurement of the indicators (Vermunt et al., 2004).
+To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
+Given the abundance of redundant terms in use for closely related types of LCA models,
+we here provide a rudimentary taxonomy.
+
+A common type of LCA is the \emph{finite Gaussian mixture model}; a
+univariate analysis where the observed distribution of a single variable
+is assumed to result from a mixture of a known number of normal
+(Gaussian) distributions. The parameters of a finite Gaussian mixture
+model are the means and variances of these underlying normal
+distributions. The analysis of shoe sizes presented earlier is a
+canonical example of this type of analysis. In the multivariate case,
+with more than one indicator variable, the parameters of a mixture model
+are the means, variances, and covariances between the indicators (which
+can be standardized to obtain correlations). These parameters can be estimated
+freely, or set to be constrained across classes (Collins \& Lanza, 2009).
+
+The technique known as \emph{latent profile analysis (LPA)} is a special case
+of the mixture model, which assumes conditional independence of the
+indicators. Conditional independence means that, after class membership
+is accounted for,
+the residual covariances/correlations between indicators are
+assumed to be zero.
+In some cases, such constraints
+will be inappropriate; for instance when the cohesion between indicators
+is expected to differ between classes (Collins \& Lanza, 2009).
+Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
+A class of larger particles with a low correlation between length and width,
+and a class of smaller particles with a high correlation between length and width.
+The reason for this difference in correlations makes theoretical sense:
+the large particles were heterogenous in shape,
+whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).
+
+Finally, some LCA models use latent variables as indicators.
+A common application of this approach is in
+longitudinal research with repeated measures of a construct.
+Examples of this approach include
+\emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
+(Jung \& Wickrama, 2008).
 These techniques estimate a latent growth model
-to describe individual trajectories over time.
-The growth mixture model is a latent class model
-where the parameters that indicate class membership
-are the intercepts and variances,
-and typically covariances of the latent growth variables,
-e.g., a latent intercept and slope.
-This technique assumes that individuals within a class can have heterogenous trajectories.
-If the variance of the growth parameters is fixed to zero,
-it is known as a latent class growth analysis.
-This latter approach assumes that all individuals within a class
-share the same identical trajectory,
-and that any variance in the indicators not explained by the class-specific latent trajectories
-is due to residual error variance.
-
-The term latent class analysis originally referred to cases
-where the observed indicators were categorical.
-Nowadays, it is more commonly used as an umbrella term.
-To prevent ambiguity, the special case
-where indicators are of binary or ordinal measurement level
-might be described as \emph{latent class analysis with ordinal indicators}.
-Latent class models with ordinal indicators are parameterized differently from mixture models.
-One common parameterization assumes that
-each categorical variable reflects an underlying standard normal distribution.
-The parameters are ``thresholds'' that correspond to quantiles of a standard normal distribution
-(with \(N(\mu = 0, \sigma = 1)\)).
-These thresholds are estimated based on
-the proportion of individuals in each of the response categories of the indicator variable.
-For example, a binary indicator has a single threshold that distinguishes the two response categories.
-If responses are distributed 50/50, then the corresponding threshold would be \(t_1 = 0.00\).
-If the responses are distributed 60/40, then the resulting threshold would be \(t_1 = 0.25\).
-
-This paper will primarily focus on mixture models and special cases thereof,
-although most of the suggested guidelines are applicable to all LCA methods.
+to describe grouped trajectories over time.
+The indicators of class membership in a GMM are the
+intercepts, and variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
+For linear trajectories, these latent variables are an intercept and slope.
+The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes homogeneity of trajectories within classes.
+This is achieved by restricting the (co)variances of the growth parameters to zero.
+Any variance in the indicators not explained by the class-specific
+latent trajectories is assumed to be error variance.
 
@@
 There are several use cases for which LCA methods are suitable.
-One example is to test a theory that postulates the existence of a categorical latent variable.
-For example, \emph{identity status theory} posits that, at any given point in time,
-adolescents reside in one of four identity statuses.
-Latent class analysis can be used to identify these four statuses
-based on observed indicators (e.g., self-reported identity exploration and commitment).
-If results indicate that the data are better described by a different number of classes,
-or that the four-class solution does not correspond to the predicted pattern of responses on the indicators,
-then the theory may be called into question.
-
-Another use case is unsupervised learning:
-when the goal is to restore unobserved class membership based on observed indicators,
-or to classify individuals.
-For example, a mixture model can be used as a diagnostic aid
-when several clinical indicators can be used to distinguish between
-a fixed number of physical (Baughman, Bisgard, Lynn, \& Meade, 2006) or mental (Wu, Woody, Yang, Pan, \& Blazer, 2011) health problems.
-The example of shoe size is a rudimentary illustration of this type of application.
-
-LCA can be used as a descriptive analysis
-where a researcher wishes to describe their sample
-and identify a few prototypes based on many variables.
-As an example, if a survey among all Dutch academics was carried out
-with the goal of bringing about a funding reform,
-LCA could be used to discover the types of publications that get funding.
-
-With LCA, our goal could be to inductively identify the number of classes.
-For instance, if we believe that a variable represents a group,
-but don't know how many groups there are,
-LCA may be an appropriate technique to answer this question.
-For example, Hopfer, Tan, and Wylie (2014) studied substance use, sexual behavior,
-and mental health status of urban population in Winnipeg, Canada.
-The underlying assumption was that there were different risk profiles,
-but their number was not known.
-From a collection of indicators, LCA provided evidence
-that there are four distinct risk profiles in the Winnipeg area.
-
-Another application of LCA is to classify individuals.
-In a peer harassment study,
-Giang and Graham (2008) used LCA to classify over 2,000 sixth grade students
-into aggressor and victim latent classes.
-The five-class solution comprised of classes of
-victims, aggressors, and socially adjusted students.
-For instance, it revealed that there were two types of victims:
-highly-victimized aggressive-victims and highly-aggressive aggressive-victims.
-
-LCA is also appropriate when we wish to identify indicators that capture classes well.
-High quality indicators will be strongly related to the latent variable
-and will lead to good class separation.
-This relationship of high quality indicators to the latent variable
-is reflected in very high or very low conditional response probabilities.
-For a simulation study exploring the effects of indicator quality on LCA, see Geiser and Wurpts (2014).
-Therefore, one could use conditional response probabilities for each item to assess its quality
-with regards to how well it helps separate the latent classes.
-From this, a theory about the selection of indicators could be informed.
-
-An extension of LCA is that containing covariates which can be used to predict class membership.
-In this approach, we not only model the latent class variable based on indicators,
-but we also relate the class membership to other explanatory variables (Vermunt (2017)).
-An example of using covariates comes from Nozadi et al. (2016)
-who applied LCA to identify the probability of children's membership to an anxiety class.
-The authors tested several covariates including children's age, sex, and accuracy scores.
-Age and sex were not found to be related to the children's latent class membership,
-hence these covariates were excluded from the analysis.
-Accuracy scores, however, were related to probabilities of
-being in anxiety and attention-anxiety classes.
-Therefore this covariate was kept as a valuable predictor.
-
-When our interest is the prediction of one or more outcomes,
-LCA can be used to construct latent classes as categorical predictors.
-Lanza, Tan, and Bray (2013) demonstrated how LCA can be used
-to classify adolescents into depression classes,
-and subsequently these classes can be used to predict smoking, grades, and delinquency.
-The study showed that the outcomes predicted by class membership
-can be binary (regular smoking), continuous (grades) or count (delinquency).
-
-In addition to these applications, LCA can be used for dimensionality reduction
-as the resulting groups summarize response patterns on a large number of indicators.
-For example, MacGregor et al. (2021) investigated symptom profiles
-among injured U.S. military personnel.
-They used fifteen dichotomous items
-from the Post-Deployment Health Assessment survey as LCA indicators.
-Combinatorics informs us that fifteen dichotomous items have
-\(2^{15}\) or \(32,768\) unique symptom combinations.
-Perhaps for this reason, MacGregor et al. (2021) incorporated LCA
-as a method of dimensionality reduction.
-A five class solution was found to have the best fit
-according to both statistical criteria and clinical interpretability.
+
+\hypertarget{testing-theory}{%
+\subsubsection{Testing theory}\label{testing-theory}}
+
+Although LCA is often discussed as an exploratory analysis technique,
+it can also be used in a confirmatory manner.
+Given that LCA is similar to confirmatory factor analysis,
+it can be used to similar ends:
+To assess whether a theoretical measurement model holds.
+This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
+For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses (Marcia, 1966).
+LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) (\textbf{luyckxDevelopmentalTypologiesIdentity2008?}).
+Similarly, \emph{personality type theory} states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA (\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).
+One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices (see Model Fit Indices).
+Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
+Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
+either qualitatively or quantitatively, using significance tests for pre-specified values (see Inference).
+A recent study by Maene, D'hondt, Van Lissa, Thijs, and Stevens (2022) applied this principle to
+test the theory that identities of Belgian adolescents with migration background could be summarized as those relating to their heritage, national and regional identity.
+
+\hypertarget{measurement-model}{%
+\subsubsection{Measurement model}\label{measurement-model}}
+
+LCA can also be used as a measurement model, similar to confirmatory factor analysis.
+This is useful when class membership is measured imperfectly by several indicators:
+LCA can partial out measurement error and restore most likely class membership.
+For example, a recent study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
+While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
+A mixture model was used to estimate ``error corrected employment status'', using self-report and registry data as indicators.
+Among the outcomes of interest were random and systematic measurement error in both indicators,
+which reveal strengths and weaknesses of both types of assessment.
+
+This example also illustrates that LCA allows us to evaluate the reliability of different indicators (Geiser \& Wurpts, 2014).
+A continuous indicator that discriminates well between classes will show large differences in class-specific means;
+an ordinal indicator will show very high or very low conditional response probabilities for some classes but not for others.
+This information can be used to select the indicators that are most diagnostic of class membership.
+For an example of LCA with a focus on differential item functioning,
+consider a study on university admission tests of Saudi Arabian students (Tsaousis, Sideridis, \& AlGhamdi, 2020).
+
+\hypertarget{dimension-reduction}{%
+\subsubsection{Dimension reduction}\label{dimension-reduction}}
+
+LCA can also be used to reduce the dimensionality of data
+by reducing many variables to a few prototypes.
+Factor analysis is also sometimes used for dimension reduction,
+but only accounts for linear covariance between indicators,
+whereas LCA can accommodate complex - but discrete - patterns.
+For example, imagine that there exist two indicators - \(x\) and \(y\), and some individuals score high on both, some score low on both, and some score high on \(x\) but low on \(y\).
+Factor analysis would only be able to capture the differences between the first two groups well; LCA captures differences between all three.
+Note that this is a pragmatic use of LCA;
+there is no theory about the existence of a categorical latent variable (cf.~Testing theory).
+Consequently, the class solution should be treated as descriptive of the sample,
+and not be ``reified'' - treated as evidence of the existence of a categorical latent variable.
+The use of LCA for dimension reduction is common in longitudinal research,
+where one developmental process is summarized using latent class growth analysis (LCGA),
+and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
+For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy (Van Lissa et al., 2015).
+These groups were then used as moderator of a latent growth analysis of adolescent- and parent-reported conflict.
+Results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than the other groups,
+and that the low empathy group had more conflict with parents than the other groups.
+Note that both these key findings are non-linear;
+modeling them in a single-group model would be difficult.
+LCGA greatly simplifies model specification and interpretation,
+albeit at a cost of some loss of information and, potentially, generalizability.
+
+The need for dimension reduction can be especially pertinent when using ordinal indicators.
+For example, one study measured
+fifteen dichotomous health indicators among injured military personnel (MacGregor et al., 2021).
+Combinatorics informs us that fifteen
+dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
+However, LCA was able to reduce response patterns to five clinically meaningful latent classes.
+
+\hypertarget{class-enumeration}{%
+\subsubsection{Class enumeration}\label{class-enumeration}}
+
+Another use case is to explore heterogeneity in a population,
+and determine whether the population is comprised of latent subpopulations, and if so, how many.
+The primary focus in this use case is on class enumeration.
+For example,
+one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
+and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
+(Gromatsky et al., 2022).
+
+\hypertarget{classification}{%
+\subsubsection{Classification}\label{classification}}
+
+Another use case is to estimate individuals' unobserved class membership based on observed indicators.
+The shoe size example in Figure \ref{fig:shoedens}
+is a rudimentary illustration of this application.
+In applied research, LCA is often used for classification in clinical contexts.
+For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough (Baughman, Bisgard, Lynn, \& Meade, 2006).
+Another study used LCA to not only classify the sample used to estimate the model, but also future cases {[}REF Zegwaard et al, in prep{]}.
+The study identified four types of care providers among those who supported close kin with mental health problems.
+Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers, to provide more targeted support.
+To obtain predicted class membership based on an existing LCA model,
+the authors computed the likelihood of data from a new participant under the multivariate normal distributions estimated within all latent classes.
+These applications of LCA aid the clinical decision making process,
+and help decide whether additional support is warranted,
+or what type of intervention is appropriate.
+
+\hypertarget{violations-of-normality}{%
+\subsubsection{Violations of normality}\label{violations-of-normality}}
 
 Finally, LCA can be used to deal with data which violate certain assumptions.
-As discussed in the shoe size example, LCA can deal with violations of normality.
-In fact, LCA assumes the population distribution is a non-normal mixture
-of \(K\) normal distributions,
-and it can discover the value of \(K\), i.e.~generate a \(K\)-class solution.
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (\textbf{baughmanMixtureModelFramework2007?}).
+These models estimate one class to explain the ``true zeros'',
+and another class with a different distribution to explain all oter values, which can include zero.
+Note that this LCA has different models for both classes,
+and thus falls beyond the scope of this paper.
+For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
+A mixture of two normal distributions with distinct variances described the observed distribution better.
 
@@
 
+The best practices in estimation, as outlined in Table
+\ref{tab:checkest}, are rooted in existing recommendations for best
+practices for estimating specific subtypes of LCA,
+including latent
+class growth analysis (Van De Schoot, Sijbrandij, Winter, Depaoli, \& Vermunt, 2017) and latent
+class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
+These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.
+
 \hypertarget{best-practices-in-estimation}{%
 \subsection{Best Practices in Estimation}\label{best-practices-in-estimation}}
-
-The best practices in estimation, as outlined in Table \ref{tab:checkest},
-are rooted in existing recommendations for best practices
-for estimating specific subtypes of LCA,
-including latent class growth analysis (Van De Schoot et al., 2017)
-and latent class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
-These were generalized to be relevant to all types of LCA,
-and updated to current best practices, as explained below.
 
@@
 
-Examining observed data is essential for any analysis
-as it may reveal patterns and violations of assumptions
-that had not been considered prior to data collection.
-Special attention should be paid
-to level of measurement of the indicators.
-Finite Gaussian mixture models (including LPA)
-are only suitable for continuous variables.
-Indicators with an ordinal level of measurement
-are likely to violate the assumption of
-within-class normal distributions of mixture models
-(see Vermunt, 2011).
-Personal experience consulting on LCA methods and
-moderating the \texttt{tidyLPA} Google group
-suggest that the misapplication of mixture models
-to ordinal (e.g., Likert-type) indicators
-is the most common source of user error.
-Whereas it has been argued that some parametric methods are robust
-when scales with 7+ indicators are treated as continuous
-(e.g., Norman, 2010),
-this certainly does not imply that all methods are.
-It is certainly unlikely that such ordinal variables
-can be treated as a \emph{mixture} of multiple normal distributions.
-The problem becomes egregious when
-the number of classes estimated equals or exceeds the number of categories;
-in this case, each class-specific mean could describe a single response category,
-and a class-specific variance component would be nonsensical.
-In sum, Likert-type scales are rarely suitable for mixture modeling;
-latent class analysis with ordinal indicators is more appropriate.
-
-Relatedly, a recent publication claimed that
-an assumption of mixture models is that
-observed indicators are normally distributed (Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020).
-This is incorrect.
-When the number of classes is greater than one,
-mixture models assume that the observed indicators
-are a mixture of multiple (multivariate) normal distributions.
-In our shoe size example, it can be seen that the population distribution
-is comprised of two normal distributions.
-When examined visually, the population distribution is evidently bimodal.
-The Shapiro-Wilk normality test (\(W = 0.971\), \(p < 0.05\) )
-rejects the null hypothesis that
-the sample comes from a normally distributed population.
-Yet, this is a prototypical example of a mixed population distribution
-where LCA can discover latent groups.
-If the population distribution were instead normal,
-there would be no classes to extract
-as the whole population would belong to a single class.
-
-Extensive descriptive statistics
-(including the number of unique values,
-variance of categorical variables, and missingness; see next subsection)
-can be obtained using the function \texttt{tidySEM::descriptives(data)}.
-Note, however, that sample-level descriptive statistics
-are of limited value when the goal of a study is
-to identify subsamples using latent class analysis.
-Plots (density plots for continuous variables,
-and bar charts for categorical ones) may be more diagnostic.
-Note that density plots can also aid in the choice of the number of classes,
-as further explained in the section on visualization.
-Descriptive statistics and plots can be relegated to online supplements,
-provided that these are readily accessible
-(consider using a GitHub repository as a comprehensive public research archive,
-as explained in Van Lissa et al., 2021).
+Examining observed data is essential for any analysis as it may reveal
+patterns and violations of assumptions that had not been considered
+prior to data collection.
+The function \texttt{descriptives(data)} provides extensive descriptive statistics,
+including the number of unique values,
+variance of continuous and categorical variables,
+missingness, skew and kurtosis (peakedness of the distribution).
+However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
+Plots convey additional information; specifically, density plots for continuous variables, and bar charts
+for categorical ones.
+Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.
+
+Particular points of attention include measurement level of the indicators.
+Recall that \texttt{tidySEM} can estimate mixture models with either numeric (type \texttt{numeric} and \texttt{integer}) or \texttt{ordered} indicators.
+The column \texttt{type} should thus only contain these types.
+Also check that the number of \texttt{unique} values match the intended measurement level.
+Numeric variables with few unique values are typically better modeled as ordinal.
+Our experience consulting on LCA methods
+and moderating the \texttt{tidyLPA} Google group suggest that the
+misapplication of mixture models to ordinal (e.g., Likert-type)
+indicators is the most common source of user error. Whereas it has been
+argued that some parametric methods are robust when scales with 7+
+response categories are treated as continuous (e.g., Norman, 2010),
+it is very unlikely
+that an ordinal variable can be modeled as a \emph{mixture} of multiple
+normal distributions.
+To illustrate the problem, consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
+Each class-specific mean would describe a single response
+category,
+and a class-specific variance component would be empirically underidentified.
+In sum, variables with few unique values might not be suitable for Gaussian mixture modeling;
+LCA with ordinal indicators is more appropriate.
+
+\hypertarget{data-preprocessing}{%
+\subsubsection{Data preprocessing}\label{data-preprocessing}}
+
+LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
+Numeric variables should have an interval measurement level,
+such that it is sensible to estimate means and (co)variances.
+For the remaining three measurement levels,
+some preprocessing is required.
+The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
+To correctly specify ordered variables,
+such as Likert scales
+use the function \texttt{OpenMx::mxFactor()}.
+Binary variables,
+such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
+Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.
+
+A second consideration in data preprocessing is that models with indicators on different scales may present with convergence issues.
+The reason for this is that the parameter space is high-dimensional with many potential local optima.
+If one indicator is on a much larger scale than another, the parameter space may be more extensively explored for that parameter.
+The problem can be resolved by transforming the indicators to place them on (roughly) similar scales.
+The most common transformation is standardization, which involves subtracting the mean of each variable from each value and dividing by the standard deviation.
+In R, this transformation is applied using the \texttt{scale()} function.
+Standardization retains all available information,
+and model parameters can be transformed back to the original scale of the indicators by reversing the transformation.
+The transformed variable has a mean of zero and a standard deviation of one.
+Note that standardization is typically not necessary if indicators are already on the same, or similar, scales.
+
+A third consideration in preprocessing relates to the univariate distributions of indicators.
+For example, the \texttt{plas\_depression} data included with \texttt{tidySEM} show extreme right skew.
+These data assess depressive symptoms;
+many respondents report (near) zero scores,
+and some respondents report elevated scores.
+To model skewed variables like this,
+it makes sense to preprocess the data by transforming the raw scores to reduce skew.
+Alternatively (or additionally), one can specify a model with free variances across classes:
+this might result in one class with a near-zero mean and small variance to account for the fact that many respondents have no depressive symptoms,
+and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.
 
@@
 
-Previous work has emphasized the importance of examining
-the pattern of missing data and
-reporting how missingness was handled
+Previous work has emphasized the importance of examining the pattern of
+missing data and reporting how missingness was handled
 (Van De Schoot et al., 2017).
-Three types of missingness have been distinguished
-in the literature (Rubin, 1976):
-Missing completely at random (MCAR),
-which means that missingness is random;
-missing at random (MAR),
-which means that missingness is contingent on the \emph{observed} data
-(and can thus be accounted for);
-and finally missing not at random (MNAR),
-which means that missingness is related to unobserved factors.
-It is possible to conduct a so-called ``MCAR'' test,
-for example the non-parametric MCAR test (Jamshidian \& Jalal, 2010).
-But note that the name ``MCAR test'' is somewhat misleading,
-as the null-hypothesis of this test is that the data are not MAR,
-and a significant test statistic indicates that
-missingness is related to the observed data (MAR).
-A non-significant test statistic does not distinguish between MCAR or MNAR.
-As Little's classic MCAR test relies on
-the comparison of variances across groups
-with different patterns of missing data,
-it assumes normality (Little, 1988).
+While we concur that investigating missingness is due diligence,
+it is important to emphasize that missingness is adequately handled by default in \texttt{OpenMx}, which is the computational backend of \texttt{tidySEM}.
+Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
+information without imputing missing values (\textbf{endersRelativePerformanceFull2001?}).
+FIML is a best-practice solution for handling missing data;
+at least on par with multiple imputation (Lee \& Shi, 2021).
+Multiple imputation is less suitable to LCA for two reasons. First,
+because LCA methods are often computationally expensive,
+and conducting them on multiple imputed datasets may be unfeasible.
+Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population.
+
+Three types of missingness have
+been distinguished in the literature (Enders, 2022; Rubin, 1976): random missingness (MCAR);
+missingness contingent on \emph{observed} variables that can be known and controlled for (MAR);
+and missingness related to unobserved variables,
+which can be neither known nor controlled for (MNAR).
+A so-called ``MCAR'' test is often reported (Jamshidian \& Jalal, 2010),
+but note that this name is somewhat misleading,
+as a significant test indicates that missingness is MAR,
+but a non-significant test statistic does not distinguish between MCAR or MNAR.
+As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (Little, 1988).
 This assumption is tenuous in the context of LCA.
-A non-parametric MCAR test,
-as provided by Jamshidian and Jalal,
-may be more suitable (Jamshidian \& Jalal, 2010).
-Unfortunately, this test was removed from the central R-repository CRAN
-due to lack of maintenance.
-For this tutorial, I have re-implemented it in the \texttt{mice} package
-as \texttt{mice::mcar()},
-with a fast backend in C++ and new printing and plotting methods.
-
-While we concur that investigating missingness is due diligence,
-it is important to emphasize that
-missingness is adequately handled by default in many software packages for LCA,
-such as Mplus, and \texttt{OpenMx} which is the backend of \texttt{tidySEM}.
-These packages use Full Information Maximum Likelihood (FIML) estimation,
-which makes use of all available information without imputing missing values.
-FIML is a best-practice solution for handling missing data;
-on par with multiple imputation (Lee \& Shi, 2021).
-FIML estimation assumes that missingness is either MCAR or MAR.
-Thus, one would typically proceed with FIML,
-regardless of the outcome of an MCAR test.
-Although FIML does not, by default, handle missingness in exogenous variables
--- all indicator variables in LCA are endogenous,
-so this is not a concern.
-
-Multiple imputation is less suitable to LCA for two reasons.
-First, because LCA methods are often computationally expensive,
-and conducting them on multiple imputed datasets may be unfeasible.
-Second, because there is no straightforward way
-to integrate LCA results across multiple datasets.
-To conclude; our recommendation is to inspect missingness
-(e.g., using \texttt{mice::MCAR()})
-and report the proportion of missingness per variable
-(e.g., using \texttt{tidySEM::descriptives()}),
-before proceeding with FIML.
-One minor concern is that the K-means algorithm,
-which \texttt{tidySEM} uses for determining starting values,
-is \emph{not} robust to missing values.
-When it fails, \texttt{tidySEM} automatically switches to hierarchical clustering,
-unless the user specifies a different
-clustering algorithm or uses manual starting values.
-
-\hypertarget{alternative-model-specifications}{%
-\subsubsection{Alternative Model Specifications}\label{alternative-model-specifications}}
-
-In order to aid researchers working with latent trajectory models,
-Van De Schoot et al. (2017) developed a protocol called
-Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
-Two of the GRoLTS guidelines (namely, 6a and 6b) refer to
-considering alternative model specifications.
-Both are discussing specific cases in latent trajectory model specification.
-The first is about whether the variance of the growth parameter
-is estimated freely or fixed,
-and the second is about whether conditional independence is assumed,
-as the researcher might also want to free-up the variance-covariance structure.
-In LCA, there are also many different ways to specify the model.
-Means, variances, and covariances between the indicators
-can either be constrained across classes, or estimated freely.
-Researchers using LCA should transparently report their chosen parametrization
-as well as discuss different parametrizations that were tested in the process.
-
-Different types of latent class models have different parameters.
-For example, mixture models and latent profile analyses
-typically have class-specific means, variances, and covariances.
-Latent growth analyses have the same parameters,
+A non-parametric MCAR test may be more suitable (Jamshidian \& Jalal, 2010).
+For this tutorial,
+the lead author has implemented this test in the \texttt{mice} package as \texttt{mice::mcar()}.
+
+To conclude; our recommendation is to inspect and report the proportion of missingness per variable (e.g., using \texttt{tidySEM::descriptives()}),
+and to test for MAR using \texttt{mice::mcar()}.
+As FIML estimation assumes that missingness is either MCAR or MAR,
+one would proceed with FIML regardless of the outcome of the test.
+
+One minor concern is that \texttt{tidySEM} uses a different algorithm to determine starting values for complete data (K-means) and for missing data (hierarchical clustering).
+This should not be a problem, however, as the global optimizer used to estimate the model should be independent of starting values (\textbf{coranaMinimizingMultimodalFunctions1987?}).
+A second concern is that FIML is not suitable for cases with complete missing data.
+If such cases must be included, their values must be imputed first.
+
+\hypertarget{model-specification}{%
+\subsubsection{Model Specification}\label{model-specification}}
+
+Any LCA model should be specified based on theoretical and empirical considerations.
+The model parameters depend on the measurement level of the indicators.
+Continuous indicators can be parametrized in terms of means, variances, and covariances.
+LCGA and GMM have these same parameters,
 but with respect to the latent growth variables.
-Latent class analyses with ordinal indicators have thresholds.
-All of these parameters can be freely estimated,
-constrained to be equal across classes,
-or fixed to a certain value (e.g., to zero).
-The total number of parameters thus
-scales with the number of estimated classes.
-Consequently, LCA methods have a potentially very high number of parameters.
-As any of these parameters could be misspecified,
-it is important to consider alternative model specifications.
-However, alternative model specifications may be approached differently
-depending on whether an analysis is data driven (exploratory),
-or theory driven (confirmatory).
-This distinction has remained underemphasized in prior writing.
-
-Prior literature on LCA has emphasized its exploratory applications
-(Nylund, Asparouhov, \& Muthén, 2007).
-In exploratory LCA,
-a large number of models are typically estimated in batch,
-with varying numbers of classes and model specifications.
-The ``correct'' model specification is then determined
-based on a combination of fit indices,
-significance tests, and interpretability.
-For latent profile analysis,
-the function \texttt{tidySEM::mx\_profiles(classes,\ variances,\ covariances)}
-largely automates this process.
-The argument \texttt{classes} indicates which
-class solutions should be estimated (e.g., 1 through 6).
-The argument \texttt{variances} specifies whether
-variances should be \texttt{"equal"} or \texttt{"varying"} across classes.
-The argument \texttt{covariances} specifies whether
-covariances should be constrained to
-\texttt{"zero"}, \texttt{"equal"} or \texttt{"varying"} across classes.
-The means are free to vary across classes by default,
-although the more general function \texttt{tidySEM::mx\_mixture()}
-could be used to circumvent this.
-After all models have been estimated,
-the function \texttt{tidySEM::table\_fit()} can be used
-to obtain a model fit table suitable for
-determining the optimal model according to best practices.
-Note, however, that this table does not include
-the bootstrapped likelihood ratio test (BLRT) by default,
-because this test is very computationally expensive.
-It is recommended to use the function \texttt{tidySEM::BLRT()}
-to compare a shortlist of likely candidate models
-based on other fit indices.
-More on fit indices can be found in
-the Model Fit Indices subsection of this paper.
-
-Confirmatory LCA typically requires
-less comprehensive alternative model specifications.
-For example, in the context of a preregistered analysis,
-the main models of interest would have been specified a priori.
-Even in this case,
-the theoretical model could be compared to a few others to contextualize it.
-
-\hypertarget{software}{%
-\subsubsection{Software}\label{software}}
-
-Many software packages are available for the estimation of LCA models.
-Some of these packages have limited functionality,
-or implement specific innovations.
-Other packages implement LCA in the context of
-a more flexible structural equation modeling framework.
-The most notable examples of the latter are
-the commercial programs Mplus and Latent GOLD,
-and the free open source R-package OpenMx.
-The commercial packages stand out
-because they offer relatively user-friendly interfaces
-and implement sensible defaults for complex analyses,
-including LCA methods.
-This lowers the threshold for applied researchers to adopt such methods.
-Commercial software also has several downsides, however.
-One such downside is that use of the software
-is restricted to those individuals
-and institutions who can afford a license.
-Another downside is that the source code,
-being proprietary, cannot be audited, debugged,
-or enhanced by third parties.
-This incurs the risk that mistakes in the source code
-may go unnoticed, and curbs progress as
-software developers cannot add new functionality.
-
-Conversely, the free open source program OpenMx is very flexible,
-but not very user-friendly.
-We directly address this limitation using the \texttt{tidySEM} R-package.
-New functionality in \texttt{tidySEM} seeks to lower the threshold
-for latent class analysis using \texttt{OpenMx}.
-It adheres to best practices in estimation and reporting,
-as described in this paper.
-The user interface is simple,
-making use of the model syntax of the widely used \texttt{lavaan} R-package.
-This syntax offers a human-readable way to specify latent variable models.
-Minor enhancements are made to simplify the specification of LCA.
-
-Because of the limitations of the aforementioned tools,
-we set out to develop a free tool that
-provides sensible defaults and is easy to use,
-but provides the option to access and modify
-all of the model inputs (i.e., low barrier, high ceiling).
-\texttt{tidySEM} interfaces with existing tools,
-and is able to translate between what existing tools are capable of
-and what researchers and analysts
-carrying-out person-oriented analyses would like to specify.
-Furthermore,\texttt{tidySEM} facilitates fully-reproducible analyses
-and contributes to open science.
-
-\hypertarget{algorithm}{%
-\subsubsection{Algorithm}\label{algorithm}}
-
-LCA parameters and model fit statistics
-can be estimated in a variety of ways.
-The choice of the estimator
-depends on the presence of missing values,
-sample size, number of indicators,
-and available computational resources (Weller, Bowen, \& Faubert, 2020).
-A commonly used technique is maximum likelihood (ML) estimation
-with the expectation-maximization (EM) algorithm as a local optimizer.
-Imagine we are estimating two parameters,
-e.g.~the class-specific means \(\mu_c\) on a continuous indicator
+Ordinal indicators can be parametrized differently;
+the default approach in \texttt{tidySEM} assumes that each ordinal indicator reflects an underlying standard normal distribution, \(N(\mu = 0, \sigma = 1)\).
+The parameters are ``thresholds'' that correspond to quantiles of this distribution, estimated based on the proportion of individuals in each of the
+response categories of the indicator variable.
+For example, a binary
+indicator has a single threshold that distinguishes the two response
+categories. If responses are distributed 50/50, then the corresponding
+threshold would be \(t_1 = 0.00\). If the responses are distributed 60/40,
+then the resulting threshold would be \(t_1 = 0.25\).
+All of the aforementioned parameters can be freely estimated,
+or constrained across classes, or fixed (e.g., to zero).
+
+Prior literature emphasized the importance of considering alternative model specifications (\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}).
+However, what has remained underemphasized is that alternative model specifications should be approached differently for exploratory versus confirmatory analyses (see Use Cases).
+Most prior literature has emphasized exploratory LCA (e.g., Nylund, Asparouhov, \& Muthén, 2007).
+In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
+A ``final'' model is then selected based on a combination of fit indices,
+significance tests,
+and interpretability (see Class Enumeration).
+
+Confirmatory LCA does not require such an extensive search over the model space.
+If hypothesized models have been specified a priori,
+and possibly even preregistered,
+it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
+Nonetheless, an important reason to consider alternative model specifications is that LCA lacks objective fit indices.
+The fit of a theoretical model can thus only be assessed by comparison to other candidates.
+For this reason, the theoretical model should still be compared to a few others.
+A one-class model is a sensible benchmark,
+to test whether theoretical model fits better than a model without latent classes.
+Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
+If the theoretical model has much better fit than other models,
+this provides evidence for the theory.
+If differences in fit are small, however, this may not do not give cause to reject a theoretical model.
+The theoretical model may still be preferred, for example, to ensure comparability with prior work.
+The fact that the theoretical model received little differential support should of course be discussed,
+along with potential implications for future work.
+
+A second consideration in model specification is the risk of overfitting (\textbf{hastieElementsStatisticalLearning2009?}).
+Overfitting occurs when a model has many parameters relative to the number of observations.
+Such a overparametrized model captures much idiosyncratic noise in a sample,
+limiting its generalizability to new samples.
+As explained before, the number of parameters in LCA scales with the number of estimated classes.
+Consequently, LCA with many classes have a potentially very high number of parameters.
+In Bayesian mixture modeling,
+it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (\textbf{rousseauAsymptoticBehaviourPosterior2011?}).
+Although it is not known whether frequentist LCA exhibits the same behavior,
+this suggests that low class counts might indicate overfitting.
+The section on Class Enumeration addresses further ways to guard against overfitting.
+
+Models can be specified in \texttt{tidySEM} in different ways.
+The wrapper functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
+The function \texttt{mx\_profiles()} has two arguments to free elements of the covariance matrix: \texttt{variances} and \texttt{covariances}.
+By default, \texttt{mx\_profiles()} fixes covariances to zero, and thus performs latent profile analysis.
+Finally, function \texttt{mx\_mixture()} accepts a more flexible \texttt{model} argument, which can be a character string specifying a model in \texttt{lavaan} syntax,
+which is translated to \texttt{OpenMx} models via the function \texttt{as\_ram()},
+or a list of \texttt{OpenMx} models.
+
+\hypertarget{estimation-and-convergence}{%
+\subsubsection{Estimation and Convergence}\label{estimation-and-convergence}}
+
+LCA parameters and model fit statistics can be estimated in different
+ways.
+The choice of estimator depends on the presence of missing
+values, sample size, number of indicators, and available computational
+resources (Weller, Bowen, \& Faubert, 2020).
+The \texttt{tidySEM} package uses maximum likelihood estimation (ML), which can be conceptually understood as follows:
+Imagine we are estimating two
+parameters, e.g.~the class-specific means \(\mu_c\) on a continuous indicator for two classes
 (ignoring the variance for now).
-The EM algorithm will attempt to find a combination of values
-for these two parameters that maximizes
-the likelihood (\(LL\)) of all observed data.
-In practice, instead of maximizing \(LL\),
-often \(-2*LL\) is minimized,
-as this offers computational advantages.
-We can think of this optimization problem
-as a three-dimensional landscape:
-The X and Y dimensions are determined by the class-specific means,
-so \(X = \mu_1\) and \(Y = \mu_2\) -
-and the Z-dimension is determined by \(Z = -2*LL\).
-The optimizer must find the deepest ``valley'' in this landscape,
-which reflects the combination of \(\mu_1\) and \(\mu_2\) that
-maximizes the likelihood of the data.
-The EM optimizer behaves somewhat like a marble dropped in this landscape.
-It is dropped at some random point in space,
-and will roll into the nearest valley.
-The problem is that, once EM rolls into a valley,
-it will settle on the bottom of that valley
-(this is known as ``convergence'').
-It cannot climb out again.
-Thus, if their are multiple valleys,
-the risk is that the optimizer gets stuck in a shallower valley (a ``local optimum''),
+ML attempts to find a combination of values for these two parameters that minimizes minus two times the log likelihood (\(-2LL\)) of the data.
+To understand the ML optimization procedure,
+imagine a three-dimensional landscape:
+The X and Y dimensions represent potential values of the class-specific means, \(X = \mu_1\) and \(Y = \mu_2\), and the Z-dimension is determined by \(Z = -2LL\).
+The optimizer must find the
+deepest ``valley'' in this landscape,
+which reflects the combination of \(\mu_1\) and \(\mu_2\) that minimizes the \(-2LL\).
+The ML
+optimizer behaves somewhat like a marble dropped in this landscape. It
+is dropped at some random point in space, and rolls into the nearest valley (a combination of values for \(\mu_1\) and \(\mu_2\) that has a low \(-2LL\)).
+The challenge is that LCA models are often highly parametrized.
+This means that the landscape is not three-dimensional, but (very) high-dimensional,
+with many valleys and sparse data to provide information about which valley is deepest.
+If the marble rolls into one of these valleys,
+it will settle there and not climb out again.
+The estimator has ``converged''.
+
+A unique challenge in LCA is that not only model parameters,
+but also the proportion of observations in each class must be estimated.
+Parameters in each class are estimated from the same dataset,
+but for each class,
+observations are weighted by their probability of being in that class - which is itself computed based on the class-specific parameters.
+The resulting chicken-and-egg problem is solved by the expectation-maximization (EM) algorithm:
+It alternates between estimating model parameters,
+calculating posterior class probabilities from those parameters,
+re-estimating model parameters using those new class probabilities as weights, and so on, until both have converged.
+
+A risk with this approach is that the optimizer might get stuck in a shallower valley (a ``local optimum''),
 and never discovers the deepest valley (the ``global optimum'', or best solution).
 One solution to this problem is to drop many marbles at random places,
-compare their final \(-2*LL\) values,
-choose the solution with the lowest \(-2*LL\),
-and make sure that several marbles replicated this solution.
-This is the ``random starts'' approach.
-
-One problem with the random starts approach is that
-it is computationally expensive to run this many replications.
-Moreover, because the algorithm begins with random starting values,
-many of the marbles are likely to be very far away from a ``good enough'' solution.
-Two innovations may improve the estimation procedure.
-The first is that, instead of picking random starting values,
-a ``reasonable solution'' may be used for the starting values.
-For example, if we assume that the different classes are likely to have
-different mean values on the indicators,
-then the K-means clustering algorithm can be used to
-determine these cluster centroids.
-We can compute the expected values of all model parameters
-by treating the K-means solution as a known class solution,
-and use these as starting values for a mixture model.
-One remaining concern is that this approach may result
-in starting values close to a local optimum,
-and that the EM algorithm will thus never find the global optimum.
-A second innovation addresses this concern.
-Instead of using EM,
-it is possible to use an optimizer that can climb out of a valley.
-Simulated annealing iteratively considers some ``destination'' in the landscape,
-and compares its likelihood to the current one.
-If the destination likelihood is higher,
+compare their final \(-2LL\) values,
+choose the solution
+with the lowest \(-2LL\), and make sure that several marbles replicated
+this solution.
+This is the ``random starts'' approach implemented in, e.g., Mplus.
+
+One problem with the random starts approach is that it is
+computationally expensive to run this many replications. Moreover,
+because the algorithm begins with random starting values,
+it is inefficient because many marbles are likely to start far away from any valley.
+Additionally, if all sets of starting values are close to a specific local optimum,
+there is a risk that the global optimum is never found.
+An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing (SA, \textbf{coranaMinimizingMultimodalFunctions1987?}).
+SA does not behave quite like the marbles of the random starts approach;
+instead, it iteratively considers some ``destination'' in the landscape,
+and compares its \(-2LL\) to the current one.
+If the destination \(-2LL\) is lower,
 the estimator moves there.
-If the destination likelihood is \emph{lower},
-the estimator still moves there occasionally,
-based on probability.
-This latter property allows it to escape local optima,
-and find the global optimum.
-
-By default, \texttt{tidySEM} derives starting values using K-means clustering,
-and identifies the global optimum solution using simulated annealing.
-Once a solution has been found,
-simulated annealing is followed up with a short run of the EM algorithm,
-as EM inherently produces an asymptotic covariance matrix for
-the parameters that can be used to compute standard errors.
+However, if the destination \(-2LL\) is \emph{higher},
+it still moves there occasionally, with a certain probability.
+Think of this as occasionally flicking the marble back out of valleys,
+to see if it rolls back into the current valley, or finds another,
+even deeper one.
+This property allows SA it to escape local optima and find the global optimum.
+
+Since SA is a global optimizer, it should be independent of starting values.
+With this in mind, it is efficient to start estimation from a ``reasonable solution'', instead of a random starting point.
+For example, if we assume that the different classes are likely to have different mean values on the indicators,
+then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids (see also \textbf{biernackiChoosingStartingValues2003?}).
+By default, \texttt{tidySEM} derives starting values using K-means clustering {[}REF{]},
+computing the mixture model starting values by treating the K-means solution as multi-group model.
+Next, SA is used to find the global optimum solution.
+Finally, SA is followed up with a short run of the ML algorithm,
+as ML inherently produces an asymptotic
+covariance matrix for the parameters that can be used to compute
+standard errors.
 Note that these defaults can be manually overridden.
 
-A recent paper suggested maximum likelihood with robust standard errors
-should be used when the observed indicators are not normally distributed (Spurk et al., 2020).
-This statement is incorrect, and may lead readers to believe that
-they must use commercial software,
-as robust maximum likelihood is currently only implemented
-in Mplus and latentGOLD.
-As explained before,
-mixture modeling assumes that observed data are
-a mixture of (multivariate) normal distributions;
-thus, the observed indicators will likely not be normally distributed.
-
-\hypertarget{class-enumeration}{%
-\subsubsection{Class Enumeration}\label{class-enumeration}}
-
-As explained, LCA can be done in an exploratory or in a confirmatory fashion.
-In exploratory LCA, a sequence of models is fitted to the data
-with each additional model estimating one more class than the previous model.
-These models are then compared
-and the best solution is selected as the final class solution.
-In some cases, prior theory can inform the researcher
-about the number of classes to expect.
-Even in such confirmatory LCA cases,
-it is nonetheless useful to know if the theoretical model
-is markedly better than those with differing numbers of classes.
-Therefore, it may always be useful to compare different class solutions.
-
-From a sequence of models, the final class solution is chosen
-based on both theoretical and statistical criteria.
-Theory should drive the selection of indicator variables,
-inform the expectations and reflect on the findings.
-In addition to this, there are several statistical criteria
-to consider in model selection.
-These include but are not limited to likelihood ratio tests,
-information criteria,
-and the Bayes factor (Weller et al., 2020).
-
-Relative model fit can be examined using the likelihood ratio test.
-This is only appropriate when the two models we wish to compare are nested.
-The likelihood ratio test statistic is computed as
-the difference in maximum log likelihoods of the two models,
-with the test degrees of freedom being
-the difference in the degrees of freedom of the two compared models.
-The test statistic follows the \(\chi^2\) distribution,
-and we want it to be non-significant
-in order to give support to the simpler model.
-The likelihood ratio test can only compare two nested models at a time (Lanza, Bray, \& Collins, 2003).
+In practice, models do not always converge.
+Reasons may include local optima,
+bad starting values,
+incorrect model specification (the model may be too complex for the data, or conversely, too simple to capture important patterns),
+or poor data quality (small data sets, or distributions that are difficult to model).
+Convergence problems may be indicated by errors and warning messages from the estimator,
+or they may be suspected, for example,
+when results change if the analyses are reproduced.
+
+Nonconvergence may be due to bad model specification.
+Consider, for example, an LCA with multiple ordinal indicators.
+If certain combinations of responses never occur together,
+in general or within one of the classes,
+this can result in extreme conditional item probabilities (exactly 0 or 1).
+Such extreme values impede model convergence.
+This problem could be addressed by trying alternative model specifications (e.g., fewer classes),
+or by merging categories of the ordinal indicators with very few responses.
+Similarly problems occur with continuous indicators.
+For instance, consider a zero-inflated indicator and a two class LPA.
+To account for the excess zeroes, one class will have a mean of zero.
+It is difficult to estimate the standard deviation for this class, however, because all of its members have the same value.
+If standard deviations are constrained to be equal across classes,
+this model will likely not converge.
+This problem can be addressed by considering more appropriate model specifications (e.g., free class variances).
+
+When poor data quality is to blame, the only solution is to collect more, or higher quality, data.
+Nonconvergence due to bad starting values and local optima may be overcome by re-estimating the model, while permuting the starting values.
+A non-converged model can be updated an re-run using \texttt{OpenMx} functions to aid model convergence,
+like \texttt{mxTryHard()} or \texttt{mxTryHardWideSearch()} for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal indicators.
+Often, these functions return an acceptable solution for models with convergence issues.
+
+\hypertarget{class-enumeration-1}{%
+\subsubsection{Class Enumeration}\label{class-enumeration-1}}
+
+Class enumeration refers to the process of determining the appropriate number of classes.
+It is approached differently for exploratory versus confirmatory LCA.
+In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes \(K\).
+The number \(K\) may be chosen a-priori or dictated by the data:
+It is sensible to limit \(K\) to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
+Above a certain number of classes,
+convergence problems typically arise,
+or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
+In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
+the proportion of cases assigned to the excessive classes converges to zero as the sample size increases (\textbf{rousseauAsymptoticBehaviourPosterior2011?}).
+It is not known whether this also happens in frequentist LCA;
+nonetheless, near-zero class memberships should be taken into account when determining the maximum \(K\).
+
+In confirmatory LCA,
+the set of models with different numbers of classes to be compared is typically smaller.
+Its purpose is to provide context for the relative fit of the theoretical model, or to compare multiple theoretical models.
+
+In class enumeration, the set of models is compared on several statistical and substantive criteria.
+In exploratory LCA, these criteria may be decisive in selecting a final model;
+in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
+Statistical criteria to consider in class enumeration include model fit indices and
+likelihood ratio tests (Masyn, 2013; Weller et al., 2020).
+Substantive criteria include convergence and model identification,
+class separability,
+and interpretability.
+We address each of these criteria below.
 
@@
 
-Fit indices typically used for determining the optimal number of classes include
-the Akaike Information Criterion (AIC)
-and the Bayesian Information Criterion (BIC).
-Both information criteria are based on the -2*log likelihood
-(which is lower for better fitting models),
-and add a penalty for the number of parameters (thus incentivizing simpler models).
-This helps balance model fit and model complexity.
-The lower the value of an information criterion,
-the better the overall fit of the model.
-The BIC applies a stronger penalty for model complexity
-that scales logarithmically with the sample size.
-The literature suggests the BIC may be
-the most appropriate information criterion to use for model comparison (Nylund-Gibson \& Choi, 2018).
-Both the AIC and the BIC are available in the \texttt{tidySEM} output.
-
-Information criteria may occasionally contradict each other,
-so it is important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index before analyzing the data.
-Another option is to always prefer
-the most parsimonious model that has best fit
-according to any of the available fit indices.
-Yet another option is to incorporate information from
-multiple fit indices using the analytic hierarchy process
-(Akogul \& Erisoglu, 2016).
-Finally, one might make an elbow plot and
-compare multiple information criteria
-(Nylund-Gibson \& Choi, 2018).
-
-Another common test of model fit is
-the likelihood ratio \(\chi^2\) goodness-of-fit test.
-However, this test is not implemented in \texttt{tidySEM}.
-
-LCA studies commonly report -2*log likelihood of the final class solution.
-This is a basic fit measure used to compute most information criteria.
-However, since log likelihood is not penalized for model complexity,
-it will continuously fall with the addition of more classes.
-
-An alternative is using the bootstrapped likelihood ratio test
-which can be run using \texttt{tidySEM::BLRT()}.
-Currently, this test is computationally expensive and
-can be slow on most computers.
-A faster version of this test,
-namely an implementation of \emph{the lazy bootstrap} (Kollenburg, Mulder, \& Vermunt, 2018)
-to \texttt{tidySEM} is being developed.
+As there are no absolute fit indices for LCA,
+information criteria are typically used to assess relative fit.
+Information criteria are computed from the \(-2LL\),
+which is the most basic fit measure.
+Information criteria add a penalty for model complexity to the \(-2LL\), and thus balance fit and complexity.
+The lower
+the value of an information criterion, the better the overall fit of the
+model.
+The original information criterion is the Akaike Information Criterion (AIC),
+computed as \(-2LL+2p\), where \(p\) is the number of parameters.
+The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
+such that complex models are penalized more in smaller samples.
+Finally,
+the so-called \emph{sample size adjusted} BIC (saBIC) implements a different ad-hoc penalty for sample size,
+\(\ln(\frac{n + 2}{24})\).
+Although a simulation study showed that the saBIC performed well for class enumeration,
+its penalty is not based on theory {[}REF{]}.
+BIC also performs well and is theoretically substantiated;
+as such, it may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+All three ICs are available in the output of \texttt{table\_fit()}.
+
+Information criteria may occasionally contradict each other, so it is
+important to identify a suitable strategy to reconcile them.
+One option is to select a specific fit index, such as the BIC,
+before analyzing the data.
+Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
+Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016),
+which is implemented in \texttt{tidyLPA} (\textbf{rosenbergTidyLPAPackageEasily2018?}).
+Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs (Nylund-Gibson \& Choi, 2018).
+The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
+Adding this many classes results in substantial decreases of the IC,
+and adding further classes results in relatively small decreases.
+
+Given a specific IC,
+it is also possible to calculate so-called ``IC-weights'',
+which indicate the relative support the data provide for each model in the set (\textbf{wagenmakersAICModelSelection2004?}).
+In \texttt{tidySEM}, IC-weights are calculated by calling \texttt{ic\_weights()} on a model fit table.
+Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
+the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
+
+\hypertarget{nested-model-tests}{%
+\subsubsection{Nested Model Tests}\label{nested-model-tests}}
+
+When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
+The test statistic is computed as the -2 times the log of the ratio of likelihoods,
+and its degrees of freedom are equal to the difference in the degrees of freedom of the two compared models.
+Since a \(k-1\)-class model is nested in a \(k\)-class model,
+it is possible to test the difference in model fit using a LR test.
+One challenge is that the LR test statistic is not asymptotically \(\chi^2\) distributed for LCA.
+The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR {[}REF lo 2001{]}.
+This test is implemented in \texttt{tidySEM}, and is part of the default output of \texttt{table\_fit()}, comparing subsequent rows of the table.
+
+Another solution is to derive an empirical sampling distribution for the LR difference.
+The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the \(k-1\)-class model \(B\) times (e.g., 1000),
+and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset {[}REF Nylund et al., 2007{]}.
+The p-value of this test is defined as the proportion of \(B\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
+Simulation studies suggest that the BLMT has greater statistical power than the LRT,
+although both performed comparably well at detecting the true number of classes {[}REF Nylund 2007, REF Tein 2013{]}.
+A crucial limitation of the BLRT is that it is very computationally expensive.
+With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
+Specific models can be compared using the function \texttt{BLRT()}.
+However, this function often takes prohbitively long to compute or runs out of memory.
+
+Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
+returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
+Other authors have reported similar experiences (\textbf{sinhaPractitionerGuideLatent2021?}).
+This suggests that, like existing ICs,
+LR tests might not adequately balance fit and complexity in LCA.
+Better solutions for class enumeration are thus an important target for future research.
+
+A promising novel solution is the so-called \emph{lazy bootstrap},
+which is much more efficient than the BLRT because it is based on sample statistics of model-implied data generated from the \(k-1\) and \(k\)-class models,
+rather than on fit statistics of LCA models fit to those data (Kollenburg, Mulder, \& Vermunt, 2018).
+This simulation-based approach to inference is similar to the posterior predictive p-value in Bayesian statistics,
+which describes how well each model is able to reproduce the observed correlation matrix (see McElreath, 2020).
+One version of the lazy bootstrap is already implemented in the commercial program latentGOLD {[}REF{]}.
+At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
+Future versions of \texttt{tidySEM} will likely incorporate this test.
 
@@
 
-Best models will divide the sample into subgroups
-which are internally homogeneous and externally distinct.
-Classification diagnostics give us a way to assess
-the degree to which this is the case.
-They are separate from model fit indices as
-a model can fit the data well
-but show poor latent class separation (Masyn, 2013).
-Classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions
-because they are uninterpretable.
-Interpretability should always be a consideration
-when considering different class solutions (Nylund-Gibson \& Choi, 2018).
-
-Three important classification diagnostics provided by \texttt{tidySEM} are:
-(1) \emph{the minimum} and \emph{maximum percentage of the sample assigned to a particular class},
-(2) \emph{the range of the posterior class probabilities by most likely class membership},
-and (3) \emph{entropy}.
-All three are based on posterior class probabilities.
-
-The posterior class probability is a measure of classification uncertainty
-which can be computed for each individual,
-or averaged for each latent class.
-When the posterior class probability is computed for each individual in the dataset,
-it represents each person's probability of belonging to each latent class.
-For each person, the highest posterior class probability is then determined
-and the individual is assigned to the corresponding class.
-We want each individual's posterior class probabilities to be
-high for one and low for the remaining latent classes.
-This is considered a high classification accuracy
-and means that the classes are distinct.
-To obtain posterior class probabilities, run \texttt{tidySEM::class\_prob()}.
-This function produces output comprised of several elements:
-
-\texttt{\$sum.posterior} is a summary table of the posterior class probabilities
-indicating what proportion of the data contributes to each class.
-
-\texttt{\$sum.mostlikely} is a summary table of the most likely class membership
-based on the highest posterior class probability.
-From this table, we compute the minimum and maximum
-percentage of the sample assigned to a particular class,
-, i.e.~\textbf{n\_min}
-(the smallest class proportion based on the posterior class probabilities)
-and \textbf{n\_max}
-(the largest class proportion based on the posterior class probabilities).
-We are especially interested in \textbf{n\_min} as
-if it is very small and comprised of few observations,
-the model for that group might not be locally identified.
-It may be impossible to calculate descriptive statistics for such a small class.
-Estimating LCA parameters on small subsamples
-might lead to bias in the results.
-Therefore, we advise caution when dealing with small classes.
-
-\texttt{\$mostlikely.class} is a table with rows representing
-the class the person was assigned to,
-and the columns indicating the average posterior probability.
-The diagonal represents the probability that
-observations in each class will be correctly classified.
-If any of the values on the diagonal of this table is low,
-we might consider not to interpret that solution.
-In \texttt{tidySEM} we use the diagonal to compute the range of
-the posterior class probabilities by most likely class membership
-which consists of
-the lowest class posterior probability (\textbf{prob\_min}),
-and the highest posterior probability (\textbf{prob\_max}).
-Both \textbf{prob\_min} and \textbf{prob\_max} can be used to
-disqualify certain class solutions,
-and are a convenient way to summarize class separation in LCA.
-We want both \textbf{prob\_min} and \textbf{prob\_max} to be high as
-that means that for all classes
-the people who were assigned to that class
-have a high probability of being there.
-\textbf{prob\_min} is especially important as it can diagnose
-if there is a class with low posterior probabilities
-which could make one reconsider that class solution.
-
-\texttt{\$avg.mostlikely} contains the average posterior probabilities for each class,
-for the subset of observations with most likely class of 1:k,
-where k is the number of classes.
-
-\texttt{\$individual} is the individual posterior probability matrix,
-with dimensions n (number of cases in the data) x k (number of classes).
-Individual class probabilities are often useful for researchers
-who wish to do follow up analyses.
-
-Entropy is a summary measure of posterior class probabilities
-across classes and individuals.
-It ranges from 0 (model classification no better than random chance)
-to 1 (perfect classification).
-As a rule of thumb, values above .80 are deemed acceptable
-and those approaching 1 are considered ideal.
-An appropriate use of entropy is that it can disqualify certain solutions
-if class separability is too low.
-Entropy was not built for nor should it be used for
-model selection during class enumeration (Masyn, 2013).
-
-\textbf{n\_min}, \textbf{n\_max}, \textbf{prob\_min}, \textbf{prob\_max}, and \textbf{entropy} and can be obtained using \texttt{tidySEM::table\_fit()}.
-
-\hypertarget{interpreting-class-solutions}{%
-\subsubsection{Interpreting Class Solutions}\label{interpreting-class-solutions}}
-
-An important outcome of LCA are conditional item probabilities,
-also known as class-specific item probabilities (Masyn, 2013),
-conditional response, or conditional solution probabilities (Geiser, 2012).
-They indicate the probability of an item being endorsed
-given that the observation belongs to a particular latent class.
-Conditional item probabilities can be obtained using \texttt{tidySEM::table\_prob()}.
-If a particular item is endorsed by two or more classes at markedly different rates,
-it is said to discriminate well between the classes
-and is consequently considered a good indicator.
-Classes are considered highly homogeneous with respect to an item
-when for a particular item
-there is a distinct difference in conditional item probabilities
-for two or more classes.
-For instance, if an item is endorsed below 30\% for one class and
-above 70\% for another class,
-the classes have high homogeneity with respect to this item (Masyn, 2013).
-Conditional item probabilities are the analogue of mean and standard deviation
-when the indicators are binary or ordinal.
-
-A problem which can occur is that of inadmissible solutions.
-With binary indicators, LCA is modelling a cross-table with all the predictors.
-The problem with such cross-tables is that they will often contain empty cells,
-i.e.~combinations of responses that never occur together.
-This problem is reflected by extreme conditional item probabilities (as in exactly 0 or 1).
-Such boundary parameter estimates could indicate that the solution is invalid (Geiser, 2012).
-Boundary parameter estimates can also happen with continuous indicators.
-For instance, if we have a zero-inflated normal distribution and a two class solution,
-one class might have the mean of zero and its standard deviation cannot be determined
-since there is little variance.
-This too could be a sign of an invalid solution,
-warn that too many classes were extracted, or indicate a local optimum (Geiser, 2012).
-
-\hypertarget{label-switching}{%
-\subsubsection{Label Switching}\label{label-switching}}
-
-The final class solution will usually discover and enumerate several classes.
-The class ordering, however, is completely arbitrary.
-The class labeled as Class 1 in one solution
-may become Class 2 or Class 3 in another model,
-even when the only difference between the models is in their starting values.
-Label switching is something to be mindful of when comparing different LCA models (Masyn, 2013).
-
-The order of clusters is nondeterministic when using K-means in \texttt{tidySEM}.
-Therefore label switching is still a consideration.
-A simple solution to this is setting a random seed number one line prior to fitting the model.
-We advise \texttt{tidySEM} users to always do so in order to circumvent label switching.
-
-Class names should be chosen to accurately reflect group membership.
-Overly simplified and generalized class names may prove misleading
-to both audiences and researches alike
-leading to what is known as a naming fallacy (Weller et al., 2020).
+Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
+These diagnostics are not model fit indices,
+because a model can fit the data well but show poor latent class separation (Masyn, 2013).
+Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.
+Interpretability should always be a consideration when considering
+different class solutions (Nylund-Gibson \& Choi, 2018).
+If the goal is to interpret properties of (members of) classes,
+it makes sense to only consider models that classify well.
+
+All classification diagnostics are derived from the posterior classification probabilities: an \(n \times k\) matrix called \(P\), with the probability that every individual \(n\) belongs to latent class \(k\).
+When classification accuracy is high, and classes are distinct,
+each individual's posterior class probabilities are high for one class, and low for the remaining classes.
+The individual classification probabilities \(P\) are obtained by running \texttt{class\_prob(res,\ type\ =\ "individual")}.
+Note that one column is added to the matrix, with each individual's most likely class membership \(M\).
+\(M\) is obtained by assigning each individual to the class corresponding to their maximum posterior class probability.
+Predicted class membership \(M\) is often of interest to researchers who wish to carry out follow up analyses.
+Note, however, that follow-up analyses should take classification error into account.
+The function \texttt{BCH()} does this automatically.
+
+The matrix \(P\) can be summarized in different way.
+For example, the estimated frequency table of the latent categorical variable can be obtained by summing the posterior probabilities by class (i.e., the column sums of \(P\)).
+This frequency table \(F_P\) is obtained by calling \texttt{class\_prob(res,\ type\ =\ "sum.posterior"}.
+These numbers are fractional because this table takes classification error into account.
+Individuals can contribute partially to multiple classes.
+
+Second, most likely class membership based on the highest posterior class probability, \(F_M\), is obtained by running \texttt{class\_prob(res,\ "sum.mostlikely")}.
+This is the frequency table of the variable \(M\).
+Unlike \(F_P\), \(F_M\) ignores classification error.
+If every observation is classified with perfect accuracy,
+\(F_M\) and \(F_P\) are identical.
+Part of the output of \texttt{table\_fit()} for LCA analyses
+are the minimum and maximum percentage of the sample assigned to a particular class, \texttt{n\_min} and \texttt{n\_max}, derived from \(F_M\).
+This is reported by default because the smallest class proportion, \texttt{n\_min}, is of particular interest.
+When the smallest class (\texttt{n\_min}) is less than 10\% of the total sample size,
+proceed with caution:
+The model might not be locally identified in that class.
+Check whether the number of participants in this class is larger than the number of parameters per class (e.g., five times as large),
+and check whether this smallest class has a meaningful substantive interpretation.
+When the smallest class is less than 1\%, the class is near-empty and might be redundant.
+
+To assess classification accuracy, we turn to the third diagnostic table:
+The average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\), \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}:
+for participants assigned to one particular class (rows) according to \(M\),
+this table gives the mean probability of being assigned to any class (columns).
+This table is obtained by calculating the column means of \(P\) for the
+\(k\) subsets of observations with most likely class of \(M = 1 \ldots k\).
+The diagonal represents the probability that observations in each class will be correctly classified.
+
+The table of classification probabilities for the most likely latent class \(M\) by latent class, \(P_{p_{(M = m | C = c)}}\) is similar to the aforementioned table \(P_{p_{(M = m)}}\),
+except that it accounts for classification uncertainty.
+If \(C\) is the true class of an observation, which is imperfectly measured,
+and \(M\) is the most likely class,
+then this table shows the probability of an observation being assigned to class \(m\) in \(M\), given that its true class is \(c\) in \(C\), or \(p(M=m|C=c)\).
+Values on the diagonal can thus be conceptualized as the reliability with which each level of the categorical latent variable is measured by \(M\),
+and off-diagonal elements can be conceptualized as measurement error.
+The minimum and maximum of the diagonal of this table are given by default as part of the output of \texttt{table\_fit()} for LCA analyses (\texttt{prob\_min} and \texttt{prob\_max}).
+These probabilities are of particular interest because they conveniently summarize classification accuracy per class.
+Both probabilities should be high, as this means that all classes are reliably classified.
+If the minimum probability is low,
+there is at least one class with poor classification accuracy.
+
+Finally, class separability can be summarized in a single statistic:
+the entropy criterion {[}REF{]}.
+In physics, entropy refers to randomness or uncertainty.
+High entropy corresponds to maximum uncertainty,
+and low entropy corresponds to a strict arrangement of the units of study.
+Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
+Low entropy, by contrast, means that every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
+Crucially, in \texttt{tidySEM}, the entropy criterion is defined as 1-entropy (a convention originating in Mplus).
+The interpretation of the entropy criterion is reversed:
+0 means the model classification is no better than chance,
+and 1 means perfect classification.
+As a rule of
+thumb, values above .80 are deemed acceptable and above .90 are considered good.
+Values close to 1 indicate that classes are very well separated,
+and treating the most likely class membership \(M\) as observed in follow-up analyses is probably defensible.
+Like other classification diagnostics,
+entropy is not a model fit measure and should not be used for model selection (Masyn, 2013).
+However, it can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
+
+\hypertarget{labeling-classes}{%
+\subsubsection{Labeling Classes}\label{labeling-classes}}
+
+\hypertarget{labelling-classes}{%
+\subsubsection{Labelling Classes}\label{labelling-classes}}
+
+Class names should be chosen to accurately reflect theoretically relevant characteristics of group membership.
+These class names should aid the interpretation of the class solutions and serve as a shorthand throughout the manuscript.
+For example, one study used a dual-trajectory LCGA of cognitive and affective empathy development with gender as a covariate.
+Although this model has dozens of parameters,
+the resulting classes were labeled low, average, and high empathy because level differences were the most distinguishing feature.
+However, overly simplistic and general class names can also be misleading, a phenomenon known as the naming
+fallacy (Weller et al., 2020).
+For example, it would be misleading to call one group ``low empathy'' and another group ``medium empathy'' if the difference in empathy were non-significant,
+or if other differences, like the effect of the covariate gender, were more substantial than level differences.
+One way to avoid the naming fallacy is comprehensively reporting model parameters,
+and explicating that class names are just a shorthand.
+
+When labeling classes, it is important to note that LCA models are identified up to the class order.
+This means that the order of classes is arbitrary and may change if the analysis is replicated.
+A simple solution to ensure replicable results is setting a random seed before fitting the model, using \texttt{set.seed()}.
+This will ensure that the classes stay in the same order upon replication,
+but that may not be the preferred order for, e.g., interpreting the results.
+To address this problem, the function \texttt{mx\_switch\_labels()} can be used.
+By default, the function orders classes from largest to smallest class count - but classes can also be ordered by values of a specific parameter (e.g., the mean intercept in a latent class growth model).
+Note that label switching is a much bigger problem when using Bayesian estimation,
+where it can occur during model estimation.
+This is beyond the scope of the present paper.
 
@@
 
-Among studies using LCA, reporting practices vary significantly (Weller et al., 2020).
-Various authors have tried to improve and standardize ways of reporting LCA
-(e.g. Masyn, 2013; Weller et al., 2020),
-but more work is needed.
-Van Lissa et al. (2020) developed WORCS,
-a workflow for open reproducible code in science.
-WORCS consists of step-by-step guidelines for research projects
-based on the TOP-guidelines developed by Nosek et al. (2015).
-WORCS workflow can be easily implemented in R in form of
-an R package which facilitates preregistration, article drafting,
-version control, citation and formatting, among others (Van Lissa et al., 2020).
-
-TOP-guidelines emphasise the use of comprehensive citation
-(including referencing the software used in the analysis),
-as well as code and data sharing wherever possible (Nosek et al., 2015).
-Van Lissa et al. (2020) suggest sharing synthetic data in case
-the original data cannot be shared,
-and provide functions to generate such synthetic data.
-Ideally, the entire research project is made reproducible
-so that others may download it and reproduce it with just one click;
-for guidance, see Van Lissa et al. (2020).
-
-As the open science movement is gaining momentum,
-researchers are becoming increasingly aware
-how important it is that analyses can be reproduced and audited.
-In line with open science principles,
-one of the suggested reporting standards relates to reproducible code.
-In this context, it is important to note that
-user-friendly methods for estimating LCA models
-have predominantly been available in commercial software packages
-(e.g., \emph{Mplus} and \emph{Latent GOLD}).
-A potential downside of commercial software is that
-it restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because
-the underlying source code is proprietary.
-To overcome these limitations,
-the present paper introduces new user-friendly functions in
-the \texttt{tidySEM} R-package that can be used to
-estimate a wide range of LCA models
-using the free, open-source R-package \texttt{OpenMx}.
-The reporting guidelines described in this paper
-are adopted in \texttt{tidySEM} by default.
-The \texttt{tidySEM} R-package thus makes advanced mixture modeling
-based on best practices widely accessible,
-and facilitates the adoption of the estimation and
-reporting guidelines described in this paper.
+Among studies using LCA, reporting practices vary significantly
+(Weller et al., 2020).
+Various authors have tried to improve and
+standardize ways of reporting LCA (e.g. Masyn, 2013; Weller et al., 2020), but more work is needed.
+
+\hypertarget{open-and-reproducible-science}{%
+\subsubsection{Open and Reproducible Science}\label{open-and-reproducible-science}}
+
+Journals and funding bodies increasingly require research to be open and reproducible.
+For instance, many journals (and the APA) have adopted the TOP guidelines, which recommend comprehensive citation of literature, software, and data sources;
+sharing data and analysis code required to exactly reproduce analyses,
+and pre-registering study plans prior to data collection or analysis (\textbf{nosekTransparencyOpennessPromotion2016a?}).
+A key advantage of using open source software for LCA is that, compared to commercial software, it enables much greater adherence to open science principles.
+Sharing code has limited utility if it can only be evaluated in commercial software,
+as this restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because the underlying source code is proprietary.
+The \texttt{tidySEM} package enables performing LCA in free open source software,
+thus increasing the utility of shared code.
+
+Adopting FOSS makes it possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
+The Workflow for Open Reproducible Code in Science describes how to create a fully reproducible research archive,
+and the \texttt{worcs} R-package automates many of the steps involved (\textbf{vanlissaWORCSWorkflowOpen2021a?}).
+This workflow is based on three principles: First,
+the paper (or analysis section) is written as a \emph{dynamic document} that contains both prose and analysis code.
+Results are automatically rendered as text, tables, and plots.
+Second,
+the research archive is \emph{version controlled}, which means that the entire historical record of the project is retained.
+This version controlled archive can be made public.
+Third, \emph{dependency management} ensures that all software used is documented, including its version number and provenance.
+Without dependency management, code might not work (for example, because it is not clear where certain functions come from),
+or its results may change when some dependencies are updated.
+By combining \texttt{tidySEM} and \texttt{worcs},
+it is possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
+
+An additional argument for sharing data in the context of LCA is that reproducing LCA analyses requires individual participant data,
+not just sufficient statistics like many other methods (i.e., a correlation or covariance matrix).
+If the original data cannot be shared,
+sharing synthetic data at least enables reproducing and reusing analysis code.
+The \texttt{worcs} package provides a generic, non-parametric method to create a synthetic copy of data when data cannot be shared.
+It is also possible to synthesize data directly from the LCA model using \texttt{mxGenerateData()},
+which often provides a better basis for reproducing the LCA.
+
+\hypertarget{preregistration}{%
+\subsubsection{Preregistration}\label{preregistration}}
+
+LCA analyses inherently involve some subjectivity,
+because of the potential for disagreement between criteria for class enumeration and the lack of objective fit indices.
+In our experience, reviewers (and authors) often disagree about which criteria should be used.
+Preregistering the class enumeration criteria as part of the analysis plan reduces disagreement and eliminates concerns about cherry picking fit measures after the results are known.
+It is particularly important to pre-specify the criteria that will be used to determine the maximum number of classes,
+because there is a risk that information criteria and likelihood ratio tests will suggest a larger number of classes, even if there are concerns about overfitting, classification accuracy, or model interpretability.
+
+The aforementioned Workflow for Open Reproducible Code in Science facilitates creating a preregistration.
+More interestingly, it also enables creating a \emph{Preregistration As Code} (PAC):
+A draft paper with reproducible analysis section based on fake (synthetic or simulated) data (\textbf{peikertReproducibleResearchTutorial2021?}).
+When the real data are accessed or collected,
+the analysis is simply reproduced with this new data file,
+and the results in the paper are automatically updated.
+PACs are quite unambiguous; the selection of the final model can even be automated by hard-coding the criteria into the analysis section.
+For an example of PAC, see (\textbf{vanlissaComplementingPreregisteredConfirmatory2022?}).
+
+\hypertarget{reporting-model-fit}{%
+\subsubsection{Reporting Model Fit}\label{reporting-model-fit}}
+
+The function \texttt{table\_fit()} reports an overview of fit indices, customized for LCA.
+In reporting model fit, it is important, firstly, to report the effective sample size, which should be the same for all models in order to compare relative fit measures.
+Secondly, it should be clearly reported,
+for example, in the strategy of analyses,
+which parameters are estimated across and within classes.
+In case of doubt, a full overview of the parameters is given by the \texttt{coef()} and \texttt{table\_results()} functions.
+The number of parameters for each model should be included in the fit table.
+Thirdly, it is important that readers can assess the relative fit of different solutions.
+The most fundamental fit measure is the \(-2LL\),
+as this is the quantity optimized during estimation.
+From these statistics, information criteria can be computed.
+By default, \texttt{table\_fit()} includes the AIC, BIC, and saBIC.
+We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.
+
+Classification diagnostics should be reported,
+as they may determine whether models are excluded from consideration.
+Of particular interest is the diagonal of either the average posterior probabilities by most likely class membership,
+or of the most likely class membership by true class membership.
+The former indicates the average probability that individuals belong to the class they were assigned to,
+and the latter indicates the probability that individuals are assigned to a class if they actually belong to that class.
+By default, \texttt{table\_fit()} reports the latter as \texttt{prob\_min} and \texttt{prob\_max}.
+The entropy is also of interest as an overall measure of class separation.
+Finally, the range of proportions of cases assigned to each class is important;
+the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant.
+By default, \texttt{table\_fit()} reports the minimum and maximum proportion of cases,
+based on most likely class membership.
+Fifth, the LMR or BLRT can be reported if it is part of the class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}.
+By default, \texttt{table\_fit()} reports the VLM for subsequent models in the table.
+
+\hypertarget{reporting-results}{%
+\subsubsection{Reporting results}\label{reporting-results}}
+
+Reporting LCA results involves reporting the class proportions,
+obtained via \texttt{class\_prob(res,\ "sum.posterior")},
+and model parameters, obtained via \texttt{table\_results(res)}.
+The class proportions are obtained by standardizing the class weights:
+A \(k\)-class model has \(k-1\) free class weights,
+and one weight constrained to 1 for model identification.
+The free weights are included in the model results,
+but are redundant with the aforementioned class proportions.
+Other model parameters may include means, (co)variances, regression coefficients (rarely), and thresholds for ordinal indicators.
+It is good practice to report the point estimate and confidence interval of these parameters.
+Although it is conventional to report p-values as well,
+note that the default p-values tests the hypothesis that the parameter is equal to zero.
+This hypothesis may be irrelevant in LCA when the focus is on parameter differences between classes.
+Reporting standard errors allows readers to compute p-values for other more informative hypotheses.
+Thresholds are more readily interpretable when converted to conditional item probabilities, using the function \texttt{table\_prob()}.
+These indicate the probability of an item being
+endorsed given that the observation belongs to a particular latent (Geiser, 2012).
+
+\hypertarget{inference}{%
+\subsubsection{Inference}\label{inference}}
+
+As mentioned before, the results table included null-hypothesis tests for all parameters by default.
+Whether these tests are informative depends on the study goals and scale of the data.
+For example, if LCA indicators are mean-centered,
+then the p-values of class-specific means indicate whether the mean in that class differs significantly from the sample average.
+It is also possible to test more informative hypotheses,
+using the function \texttt{wald\_test()}.
+Equality-constrained informative hypotheses are specified using the \texttt{bain} syntax,
+and evaluated on the model object.
+This enables testing theoretical hypotheses - for example, if specific values for class-specific means had been hypothesized.
+The same function allows testing parameters for equality, either within or across classes.
 
@@
 
-Plots can greatly improve the interpretability of LCA models.
-There are several stages where this is an important consideration.
-
-First, in class enumeration,
-we want to compare several competing class solutions.
-This can be done by means of the AIC and BIC.
-Here, we suggest using an elbow plot
-with the \(K-class\ solution\) on the X-axis,
-and information criteria value on the Y-axis.
-We show an implementation of such an elbow plot
-in our Tutorial subsection.
-
-Second, once we have decided on the final class solution,
-we want to interpret the response patterns
-on the indicators for each latent class.
-While \texttt{tidySEM} can create a table showing
-the probability of each item's response endorsement
-(using \texttt{tidySEM::table\_prob}),
-it may be easier to inspect these probabilities visually.
-For this reason, we created \texttt{ggplot2::plot\_prob()}.
-The resulting shows response patterns
-on all indicators for each group.
-We give an example of this plot in the Tutorial subsection.
-
-\hypertarget{tutorial}{%
-\section{Tutorial}\label{tutorial}}
-
-This is an example of an exploratory LCA using \texttt{tidySEM}.
-
-\hypertarget{loading-the-dataset}{%
-\subsection{Loading the Dataset}\label{loading-the-dataset}}
-
-In this example, we use \emph{data\_mix\_ordinal},
-a simulated data for mixture modelling with ordinal indicators.
-This dataset is built-in to \texttt{tidySEM}.
-For more information about the dataset,
-type \texttt{?tidySEM::data\_mix\_ordinal} into the R console
-after loading the \texttt{tidySEM} using library().
-
-We load the dataset and convert the indicators into ordered factors.
-
-\hypertarget{exploring-the-data}{%
-\subsection{Exploring the Data}\label{exploring-the-data}}
-
-An important step to preceding any statistical analysis is data exploration.
-Here we use \texttt{tidySEM::descriptives()}
-to describe the dataset we are using.
-
-As we can see, the output includes various descriptives of our dataset.
-Special attention should be paid to examining the pattern of missingness,
-as discussed in the \emph{Handling missing data} section of this paper.
-In our example, we see that there are no missing values,
-hence we proceed with our analysis.
-
-\hypertarget{conducting-latent-class-analysis}{%
-\subsection{Conducting Latent Class Analysis}\label{conducting-latent-class-analysis}}
-
-Before we fit a series of LCA models, we set a random seed
-using \texttt{set.seed()}.
-The seed number can be any digit.
-This is an important step as
-there is some inherent randomness in the LCA computations,
-and having the same seed number ensures that
-two separate researcher obtain exactly the same results
-when fitting LCA models.
-
-Finally, we reach the step of fitting LCA models.
-To do so, we use \texttt{tidySEM::mx\_lca()}
-which takes data and number of classes as inputs.
-In our example, we want to fit 1 to 4 class solutions
-and compare their output.
-Depending on your computer's computational power,
-this might take a while.
-
-\hypertarget{class-enumeration-1}{%
-\subsection{Class Enumeration}\label{class-enumeration-1}}
-
-In class enumeration,
-we want to compare a sequence of LCA models fitted to the data.
-To aid the process, we create a model fit table using \texttt{tidySEM::table\_fit()}
-with the results object as the input.
-As the output contains a lot of information on each of the four fitted models,
-we select a subset of helpful model fit indices and classification diagnostics.
-
-Our selection of fit indices and classification diagnostics includes:
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1220}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8780}}@{}}
-\caption{\label{tab:unnamed-chunk-6}Selection of Fit Indices and Classification Diagnostics}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Selection
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Description
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Selection
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Description
-\end{minipage} \\
-\midrule()
-\endhead
-Name & the \(K\)-class solution \\
-LL & the -2*log-likelihood of each model \\
-AIC & the Akaike Information Criterion value \\
-BIC & the Bayesian Information Criterion value \\
-prob\_min & the lowest posterior class probability by most likely class membership \\
-prob\_max & the highest posterior class probability by most likely class membership \\
-n\_min & the lowest class proportion based on the posterior class probabilities \\
-n\_max & the highest class proportion based on the posterior class probabilities \\
-\bottomrule()
-\end{longtable}
-
-We discussed several possible strategies to select the final class solution.
-Here, we apply our own.
-
-To aid our interpretation of the results,
-we create an elbow plot showing the trends in information criteria across four models.
-
-From the elbow plot, we see that AIC has a lower penalty for model complexity.
-However, we are more interested in the BIC values, which are similar
-for the one, two and three-class solutions, but the four-class solution fits significantly worse.
-For this reason, we eliminate the four-class solution from the selection process.
-
-Then we examine the model fit table.
-As expected, the -2*log likelihood falls successively with each added class.
-As previously stated, classification diagnostics should not be used for model selection,
-but they can be used to disqualify certain solutions because they are uninterpretable.
-We see that prob\_min for the four-class solution is low,
-knowing that this solution also has a high BIC, we disqualify this solution.
-
-Out of the remaining three solutions,
-we notice that entropy is the highest for the three-class solution,
-and it has a satisfactory prob\_min and n\_min.
-Based on this, we retain the three class solution in model selection.
-Note that entropy for the one-class solution will always equal to one,
-as it is 100\% true that every case is in that class.
-Based on the low entropy of the two-class solution,
-we eliminate this model.
-
-Finally, when comparing the one and three-class solutions,
-we inspect the information criteria.
-For BIC, the one-class solution fits better, but the difference is marginal.
-AIC tells us that the added complexity
-of having three classes still explains the data better
-than a one-class solution.
-Therefore, we select the three-class solution as our final-class solution.
-
-\hypertarget{interpreting-the-final-class-solution}{%
-\subsection{Interpreting the Final Class Solution}\label{interpreting-the-final-class-solution}}
-
-To aid our understanding of the final class solution,
-we use \texttt{ggplot2::plot\_prob()}
-with the results of the three-class model as the input.
-The resulting graph shows response patterns
-on all the indicators for each group.
-
-If we want to know the probability of each response option's endorsement
-for each class, we can use \texttt{tidySEM::table\_prob}.
-These are thresholds for ordinal dependent variables in the probability scale.
-
-In the plot, we can see the distributions of the response probabilities on the indicators
-for each of the three classes.
-For instance, we see that in Class 1 the most common response to u2 is 2,
-while in Class 2 and Class 3 this is 0.
-We can also see that response 1 is a rare response not forming the majority in
-any class.
-Class 2 distinguishes itself because the majority scores the response 0 category of u3 and u4,
-while in Class 1 and 2 this is not the case.
-Class 3 distinguishes itself because the most common response to u3 and u4 is 3.
-
-We can also interpret the response patterns numerically.
-
-\hypertarget{extracting-posterior-class-probabilities}{%
-\subsection{Extracting Posterior Class Probabilities}\label{extracting-posterior-class-probabilities}}
-
-Another step is to extract posterior class probabilities.
-This is done by the use of \texttt{TidySEM::class\_prob}
-with the results of the final class solution as the input.
+In an informal review of published graphics for LCA,
+we found that most visualizations focused on the estimated model parameters,
+displaying class means as points connected by lines, with point shape and linetype indicating class membership.
+Note that the lines introduce spurious information.
+Furthermore, they omit uncertainty about class means and do not illustrate how well the model describes the data.
+
+The Grammar of Graphics addresses best practices in visualization, which we summarize here with specific emphasis on the application to LCA.
+Most importantly, good graphics start with the \emph{data} in mind.
+Visualizing raw data allows one to assess class separability,
+and which aspects of the data are captured well or poorly by the model.
+A second consideration are which \emph{aesthetics} to use.
+In a color plot, class membership can be indicated by color.
+In a black and white plot, it can be indicated by linetype and point shape.
+A third consideration is which geometric objects (geoms) to use.
+As mentioned before, common practice is to plot parameter estimates as points, and connect these points with lines.
+Points are suitable geoms to visualize point estimates and raw data.
+For continuous indicators, additional relevant geoms include error bars for standard errors and standard deviations,
+and density plots to visualize the distribution of raw data (smoothed with a kernel density function).
+In two dimensions, (co)variances can be visualized as ellipses.
+Finally, when plotting multiple class solutions,
+it makes sense to \emph{facet} a plot into multiple panels by class solution.
+The plots in \texttt{tidySEM} are based on best practices as outlined in the grammar of graphics,
+and are implemented in \texttt{ggplot2}.
+As such, these plots can be further modified just like any other \texttt{ggplot2} object.
+One remaining limitation is that these best practices have only been implemented for LCA with continuous indicators, in \texttt{plot\_profiles()}, \texttt{plot\_density()} and \texttt{plot\_bivariate()}, and for LCGA in \texttt{plot\_growth()}.
+Plots for LCA with ordinal indicators are still rudimentary, more work remains to be done in this area.
+
+\hypertarget{potential-pitfalls}{%
+\subsection{Potential Pitfalls}\label{potential-pitfalls}}
+
+Based on our experience in statistical consultations,
+there are a few common pitfalls and misunderstandings about LCA.
+First, perhaps the most common problem we see in consultations is treating ordinal data as continuous.
+As explained previously, LCA for ordinal indicators should be used for Likert-type scales and other data with few unique values.
+A second common pitfall is omitting the 1-class solution from class enumeration.
+If a one-class solution has better or similar fit to models with a higher number of classes, LCA may not be appropriate.
+A one-class solution should thus always be included as a benchmark.
+Third, although it is widely understood that classification diagnostics (e.g., entropy, posterior probability, class counts) are not class enumeration criteria,
+they still have a role in class enumeration.
+Specifically, they can be used to exclude solutions that do not meet predefined criteria from consideration.
+Fourth, previous reporting standards called for reporting of sample descriptive statistics (see \textbf{schreiberLatentClassAnalysis2017?}).
+While we agree that this is due dilligence,
+sample descriptives have two limitations in the context of LCA:
+they are not sufficient to check or reproduce the analysis,
+and many use cases of LCA assume that the population is heterogeneous, which limits the utility of whole-sample descriptives.
+These two limitations can be addressed by data sharing and reporting descriptive statistics for each latent class separately.
+Fifth, a recent publication mistakenly stated that an assumption of mixture
+models is that observed indicators are normally distributed
+(Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020).
+This is incorrect, as evident from the shoe size example:
+When the number of classes is greater than one,
+Gaussian mixture models assume that the distribution of observed indicators is a \emph{mixture of multiple normal distributions}.
+If the indicators were normally distributed,
+there would be no classes to extract as the whole population
+would belong to a single (homogeneous) class.
+Finally, the same paper suggested that robust standard errors
+should be used when indicators are not normally distributed
+(Spurk et al., 2020).
+This is also incorrect;
+to our knowledge, there is no literature on the merits of robust standard errors for LCA.
+
+\hypertarget{smart-lca-checklist}{%
+\subsection{SMART-LCA Checklist}\label{smart-lca-checklist}}
+
+The best practices discussed in this paper are summarized in the SMART-LCA Checklist: Standards for More Accuracy in Reporting of different Types of Latent Class Analysis.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Report item descriptives
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Are all items on similar scales?
+  \item
+    Is the measurement level correctly coded?
+  \item
+    Any other reasons why the observed (multivariate) distribution could not be described by the LCA (e.g., extreme skew)?
+  \end{itemize}
+\item
+  Report missingness
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Proportion of missingness per variable
+  \item
+    MAR test
+  \item
+    Missing data method for handling missingness (typically FIML)
+  \end{itemize}
+\item
+  Report model specification
+
+  \begin{itemize}
+  \tightlist
+  \item
+    In exploratory LCA, compare all sensible alternative specifications
+  \item
+    In confirmatory LCA, justify alternative models on theoretical grounds
+  \item
+    Check that the observed (multivariate) distribution could theoretically be captured by the specified model (see \#1).
+  \end{itemize}
+\item
+  Class enumeration
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Justify the maximum number of classes \(K\)
+  \item
+    In exploratory LCA, estimate 1:\(K\) classes
+  \item
+    In confirmatory LCA, estimate the theoretical number of classes, a 1-class model, and optionally other numbers of classes for comparison
+  \end{itemize}
+\item
+  Justify the (preregistered) criteria used for class enumeration, which can include:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Information criteria
+  \item
+    LR tests
+  \end{itemize}
+\item
+  Justify the (preregistered) criteria used to eliminate models from consideration, including:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Classification diagnostics
+  \item
+    Local identifiability (acceptable number of observations per parameter in each class)
+  \item
+    Model convergence
+  \item
+    Theoretical interpretability of the solution
+  \end{itemize}
+\item
+  Report the following:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Fit of all models under consideration
+  \item
+    Classification diagnostics for all models under consideration
+  \item
+    All parameters of the chosen model (means, (co)variances, thresholds)
+  \item
+    Proportion of cases per class
+  \end{itemize}
+\item
+  Provide an informative plot of the chosen solution
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Show raw data
+  \item
+    Show parameters
+  \item
+    Show confidence bounds if possible
+  \item
+    Show multivariate distributions for multivariate models, if possible
+  \end{itemize}
+\item
+  Account for classification inaccuracy in follow-up analyses
+
+  \begin{itemize}
+  \tightlist
+  \item
+    If entropy \textgreater{} .95 and posterior classification probability for each class \textgreater{} .95, this is less crucial
+  \end{itemize}
+\end{enumerate}
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+Write me
 
@@
 \url{https://doi.org/10.3390/mca21030034}
+
+\leavevmode\vadjust pre{\hypertarget{ref-alkema_maximizing_2022}{}}%
+Alkema, L.
+M., Van Lissa, C.
+J., Kooi, M., \& Koelmans, A.
+A.
+(2022).
+Maximizing {Realism}: {Mapping} {Plastic} {Particles} at the {Ocean} {Surface} {Using} {Mixtures} of {Normal} {Distributions}.
+\emph{Environmental Science \& Technology}, \emph{56}(22), 15552--15562.
+\url{https://doi.org/10.1021/acs.est.2c03559}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-collins_latent_2009}{}}%
+Collins, L. M., \& Lanza, S. T. (2009). \emph{Latent {Class} and {Latent} {Transition} {Analysis}: {With} {Applications} in the {Social}, {Behavioral}, and {Health} {Sciences}}. John Wiley \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-enders_applied_2022}{}}%
+Enders, C. K. (2022). \emph{Applied {Missing} {Data} {Analysis}}. Guilford Publications.
+
 \leavevmode\vadjust pre{\hypertarget{ref-figueiredo_unsupervised_2002}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-giang_using_2008}{}}%
-Giang, M.
-T., \& Graham, S.
-(2008).
-Using latent class analysis to identify aggressors and victims of peer harassment.
-\emph{Aggressive Behavior}, \emph{34}(2), 203--213.
-\url{https://doi.org/10.1002/ab.20233}
+\leavevmode\vadjust pre{\hypertarget{ref-gromatsky_characteristics_2022}{}}%
+Gromatsky, M., Edwards, E.
+R., Sullivan, S.
+R., Van Lissa, C.
+J., Lane, R., Spears, A.
+P., \ldots{} Goodman, M.
+(2022).
+Characteristics of suicide attempts associated with lethality and method: {A} latent class analysis of the {Military} {Suicide} {Research} {Consortium}.
+\emph{Journal of Psychiatric Research}, \emph{149}, 54--61.
 
@@
 Hennig, C., Meila, M., Murtagh, F., \& Rocci, R. (2015). \emph{Handbook of {Cluster} {Analysis}}. 28.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hopfer_social_2014}{}}%
-Hopfer, S., Tan, X., \& Wylie, J.
-L.
-(2014).
-A {Social} {Network}--{Informed} {Latent} {Class} {Analysis} of {Patterns} of {Substance} {Use}, {Sexual} {Behavior}, and {Mental} {Health}: {Social} {Network} {Study} {III}, {Winnipeg}, {Manitoba}, {Canada}.
-\emph{American Journal of Public Health}, \emph{104}(5), 834--839.
-\url{https://doi.org/10.2105/AJPH.2013.301833}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-jung_introduction_2008}{}}%
+Jung, T., \& Wickrama, K.
+a.
+S.
+(2008).
+An {Introduction} to {Latent} {Class} {Growth} {Analysis} and {Growth} {Mixture} {Modeling}.
+\emph{Social and Personality Psychology Compass}, \emph{2}(1), 302--317.
+\url{https://doi.org/10.1111/j.1751-9004.2007.00054.x}
+
 \leavevmode\vadjust pre{\hypertarget{ref-van_kollenburg_lazy_2018}{}}%
 Kollenburg, G. H. van, Mulder, J., \& Vermunt, J. K. (2018). \emph{The {Lazy} {Bootstrap}. {A} {Fast} {Resampling} {Method} for {Evaluating} {Latent} {Class} {Model} {Fit}}. 23.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lanza_introduction_2003}{}}%
-Lanza, S.
-T., Bray, B.
-C., \& Collins, L.
-M.
-(2003).
-An {Introduction} to {Latent} {Class} and {Latent} {Transition} {Analysis}.
-In \emph{Handbook of {Psychology}: {Research} {Methods} in {Psychology}} (2nd ed., Vol.
-2, pp.
-690--712).
-John Wiley \& Sons.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lanza_latent_2013}{}}%
-Lanza, S.
-T., Tan, X., \& Bray, B.
-C.
-(2013).
-Latent {Class} {Analysis} {With} {Distal} {Outcomes}: {A} {Flexible} {Model}-{Based} {Approach}.
-\emph{Structural Equation Modeling : A Multidisciplinary Journal}, \emph{20}(1), 1--26.
-\url{https://doi.org/10.1080/10705511.2013.742377}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-maene_perceived_2022}{}}%
+Maene, C., D'hondt, F., Van Lissa, C.
+J., Thijs, J., \& Stevens, P.
+A.
+J.
+(2022).
+Perceived {Teacher} {Discrimination} and {Depressive} {Feelings} in {Adolescents}: {The} {Role} of {National}, {Regional}, and {Heritage} {Identities} in {Flemish} {Schools}.
+\emph{Journal of Youth and Adolescence}, \emph{51}(12), 2281--2293.
+\url{https://doi.org/10.1007/s10964-022-01665-7}
+
+\leavevmode\vadjust pre{\hypertarget{ref-marcia_development_1966}{}}%
+Marcia, J. E. (1966). Development and validation of ego-identity status. \emph{Journal of Personality and Social Psychology}, \emph{3}, 551--558. \url{https://doi.org/10.1037/h0023281}
+
 \leavevmode\vadjust pre{\hypertarget{ref-masyn_latent_2013}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-mcelreath_statistical_2020}{}}%
+McElreath, R.
+(2020).
+\emph{Statistical {Rethinking} {\textbar} {A} {Bayesian} {Course} with {Examples} in {R} and {STAN}} (2nd ed.).
+Chapman; Hall/CRC.
+Retrieved from \url{https://www.taylorfrancis.com/books/mono/10.1201/9780429029608/statistical-rethinking-richard-mcelreath}
+
+\leavevmode\vadjust pre{\hypertarget{ref-neale_openmx_2016}{}}%
+Neale, M.
+C., Hunter, M.
+D., Pritikin, J.
+N., Zahery, M., Brick, T.
+R., Kirkpatrick, R.
+M., \ldots{} Boker, S.
+M.
+(2016).
+{OpenMx} 2.0: {Extended} {Structural} {Equation} and {Statistical} {Modeling}.
+\emph{Psychometrika}, \emph{81}(2), 535--549.
+\url{https://doi.org/10.1007/s11336-014-9435-8}
+
 \leavevmode\vadjust pre{\hypertarget{ref-norman_likert_2010}{}}%
@@
 \url{https://doi.org/10.1007/s10459-010-9222-y}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nosek_promoting_2015}{}}%
-Nosek, B.
-A., Alter, G., Banks, G.
-C., Borsboom, D., Bowman, S.
-D., Breckler, S.
-J., \ldots{} Yarkoni, T.
-(2015).
-Promoting an open research culture.
-\emph{Science}, \emph{348}(6242), 1422--1425.
-\url{https://doi.org/10.1126/science.aab2374}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nozadi_moderating_2016}{}}%
-Nozadi, S.
-S., Troller-Renfree, S., White, L.
-K., Frenkel, T., Degnan, K.
-A., Bar-Haim, Y., \ldots{} Fox, N.
-A.
-(2016).
-The {Moderating} {Role} of {Attention} {Biases} in understanding the link between {Behavioral} {Inhibition} and {Anxiety}.
-\emph{Journal of Experimental Psychopathology}, \emph{7}(3), 451--465.
-\url{https://doi.org/10.5127/jep.052515}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-pavlopoulos_measuring_2015}{}}%
+Pavlopoulos, D., \& Vermunt, J. K. (2015). \emph{Measuring temporary employment. {Do} survey or register data tell the truth?} 37.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rosseel_lavaan_2012}{}}%
+Rosseel, Y. (2012). Lavaan: {An} {R} {Package} for {Structural} {Equation} {Modeling}. \emph{Journal of Statistical Software}, \emph{48}, 1--36. \url{https://doi.org/10.18637/jss.v048.i02}
+
 \leavevmode\vadjust pre{\hypertarget{ref-rubin_inference_1976}{}}%
@@
 \url{https://doi.org/10.1016/j.jvb.2020.103445}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tsaousis_measurement_2020}{}}%
+Tsaousis, I., Sideridis, G.
+D., \& AlGhamdi, H.
+M.
+(2020).
+Measurement {Invariance} and {Differential} {Item} {Functioning} {Across} {Gender} {Within} a {Latent} {Class} {Analysis} {Framework}: {Evidence} {From} a {High}-{Stakes} {Test} for {University} {Admission} in {Saudi} {Arabia}.
+\emph{Frontiers in Psychology}, \emph{11}.
+Retrieved from \url{https://www.frontiersin.org/articles/10.3389/fpsyg.2020.00622}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_worcs_2020}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_divergence_2015}{}}%
 Van Lissa, C.
-J., Brandmaier, A.
-M., Brinkman, L., Lamprecht, A.-L., Peikert, A., Struiksma, M.
-E., \& Vreede, B.
-(2020).
-\emph{{WORCS}: {A} {Workflow} for {Open} {Reproducible} {Code} in {Science}}.
-\url{https://doi.org/10.17605/OSF.IO/ZCVBS}
-
-\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_worcs_2021}{}}%
-Van Lissa, C.
-J., Brandmaier, A.
-M., Brinkman, L., Lamprecht, A.-L., Peikert, A., Struiksma, M.
-E., \& Vreede, B.
-M.
-I.
-(2021).
-{WORCS}: {A} workflow for open reproducible code in science.
-\emph{Data Science}, \emph{4}(1), 29--49.
-\url{https://doi.org/10.3233/DS-210031}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vermunt_k-means_2011}{}}%
+J., Hawk, S.
+T., Branje, S.
+J.
+T., Koot, H.
+M., Van Lier, P.
+A.
+C., \& Meeus, W.
+H.
+J.
+(2015).
+Divergence {Between} {Adolescent} and {Parental} {Perceptions} of {Conflict} in {Relationship} to {Adolescent} {Empathy} {Development}.
+\emph{Journal of Youth and Adolescence}, \emph{44}(1), 48--61.
+\url{https://doi.org/10.1007/s10964-014-0152-5}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_latent_2004}{}}%
 Vermunt, J.
-K.
-(2011).
-K-means may perform as well as mixture model clustering but may also be much worse: {Comment} on {Steinley} and {Brusco} (2011).
-\emph{Psychological Methods}, \emph{16}(1), 82--88.
-\url{https://doi.org/10.1037/a0020144}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vermunt_latent_2017}{}}%
-Vermunt, J.
-K.
-(2017).
-Latent {Class} {Modeling} with {Covariates}: {Two} {Improved} {Three}-{Step} {Approaches}.
-\emph{Political Analysis}, \emph{18}(4), 450--469.
-\url{https://doi.org/10.1093/pan/mpq025}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vermunt_jk_latent_2004}{}}%
-Vermunt, J.K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.F., \& Department of Methodology and Statistics.
+K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.
+F., \& Department of Methodology and Statistics.
 (2004).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wu_abuse_2011}{}}%
-Wu, L.-T., Woody, G.
-E., Yang, C., Pan, J.-J., \& Blazer, D.
-G.
-(2011).
-Abuse and dependence on prescription opioids in adults: A mixture categorical and dimensional approach to diagnostic classification.
-\emph{Psychological Medicine}, \emph{41}(3), 653--664.
-\url{https://doi.org/10.1017/S0033291710000954}
-
 \end{CSLReferences}

### 2023-01-23 appendix_a.Rmd
edits by Mauricio

---
title             : "Appendix A: Confirmatory LPA for the Caregiver Compass"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
bibliography      : "r-references.bib"
floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
library(yaml)
library(scales)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(width.cutoff = 60), tidy = TRUE
)
options(scipen = 1, digits = 2)

run_everything = FALSE

pkgdown <- TRUE
```

This is an example of confirmatory LPA using `tidySEM`.
The simulated data are based on work by Zegwaard and colleagues, who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
Qualitative research among experts resulted in a theory postulating the existence of four types of caregivers (translated from the original Dutch):

**Balanced**

> The balanced caregiver experiences relative balance between the costs and benefits of caring for a close other.

**Imbalanced**

> The imbalanced caregiver experiences a precarious balance between the costs and benefits of caring for a close other.

**Lonely**

> The lonely caregiver experiences a strong sense of isolation.

**Entrapped**

> The entrapped caregiver strongly feels a sense of being entangled in responsibilities which are difficult to fulfill.

To view the data documentation, run the command `?tidySEM::zegwaard_carecompass` in the R console.

## Loading the Data

To load the data, simply attach the `tidySEM` package.
For convenience, we assign the variables used for analysis to an object called `df`.
We first only use the four scales: `c("burdened", "trapped", "negaffect", "loneliness")`.
\`.

```{r, echo = TRUE, eval=TRUE}
# Load required packages
library(tidySEM) 
library(ggplot2)
# Load data
df <- zegwaard_carecompass[, c("burdened", "trapped", "negaffect", "loneliness")]
```

## Examining the Data

As per the best practices, the first step in LCA is examining the observed data.
We use `tidySEM::descriptives()` to describe the data numerically.
Because all scales are continuous, we select only columns for continuous data to de-clutter the table:

```{r echo = TRUE, eval=FALSE, results='asis'}
desc <- tidySEM::descriptives(df)
desc <- desc[, c("name", "n", "missing", "unique",
                 "mean", "median", "sd", "min", "max",
                 "skew_2se", "kurt_2se")]
desc
```

```{r tabdesc, echo = FALSE, eval=TRUE, results='asis'}
desc <- tidySEM::descriptives(df)
desc <- desc[, c("name", "n", "missing", "unique",
                 "mean", "median", "sd", "min", "max",
                 "skew_2se", "kurt_2se")]
papaja::apa_table(desc, caption = "Descriptive statistics")
```

The table indicates two potential causes for concern: there is a small percentage of missingness, and all variables have relatively high kurtosis.
Since there are some missing values, we can conduct an MCAR test using `mice::mcar(df)`.
According to Hawkins' test, there is no evidence to reject the assumptions of multivariate normality and MCAR, $\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71$.
Missing data will be accounted for using FIML.

Additionally, we can plot the data.
The `ggplot2` function `geom_density()` is useful for continuous data.
Visual inspection confirms the conclusions from the `descriptives()` table: the data are kurtotic (peaked).

```{r, echo = TRUE, eval = FALSE}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
```

```{r, echo = FALSE, eval = run_everything}
# mcartest <- mice::mcar(df)
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
p <- ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
ggsave("appendices/plot_lpa_desc.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```

```{r figdesc, echo = FALSE, eval = pkgdown}
knitr::include_graphics("appendices/plot_lpa_desc.pdf")
```

## Conducting Latent Profile Analysis

As all variables are continuous, we can use the convenience function `tidySEM::mx_profiles()`, which is a wrapper for the generic function `mx_mixture()` optimized for continuous indicators.
Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
Its arguments are `data` and number of `classes`.
All variables in `data` are included in the analysis, which is why we first selected the indicator variables.
As this is a confirmatory LCA, we do not follow a strictly data-driven class enumeration procedure.
We will set the maximum number of classes $K$ to one more than the theoretically expected number.
We set a seed to ensure replicable results.

```{r fitlca, eval = run_everything, echo = FALSE}
set.seed(123) # setting seed 
res <- mx_profiles(data = df,
                   classes = 1:5)
saveRDS(res, "res_lpa.RData")
fit <- table_fit(res)
write.csv(fit, "lpatabfit.csv", row.names = FALSE)
```

```{r eval = FALSE, echo = TRUE}
set.seed(123)
res <- mx_profiles(data = df,
                   classes = 1:5)
```

```{r eval = TRUE, echo = FALSE}
fit <- read.csv("appendices/lpatabfit.csv", stringsAsFactors = FALSE)
class(fit) <- c("tidy_fit", "data.frame")
```

This analysis should produce some warnings about cluster initialization.
These relate to the selection of starting values, which relies on the K-means algorithm and is not robust to missing data.
The algorithm automatically switches to hierarchical clustering, no further action is required.

## Class Enumeration

To compare the fit of the theoretical model against other models, we create a model fit table using `table_fit()` and retain relevant columns.
We also determine whether any models can be disqualified.

In this example, all models converge without issues.
If, for example, the two-class solution had not converged, we could use the function `res[[2]] <- mxTryHard(res[[2]])` to aid convergence.

Next, we check for local identifiability.
The sample size is consistently reported as 513, which means that partially missing cases were indeed included via FIML.
The smallest class size occurs in the 5-class model, where the smallest class is assigned 7% of cases, or 38 cases.
This model has 28 parameters, approximately 6 per class.
We thus have at least five observations per parameter in every class, and do not disqualify the 5-class model.

There are concerns about theoretical interpretability of all solutions, as the entropies and minimum classification probabilities are all low.
However, in this confirmatory use case, we address this when interpreting the results.

```{r fit_table, include = TRUE, eval=F}
fit <- table_fit(res) # model fit table
fit[ , c("Name", "LL", "Parameters", "n",
         "BIC", "Entropy",
         "prob_min", "prob_max", 
         "n_min", "n_max",
         "lmr_p")]
```

```{r tabfit, echo = FALSE, eval = TRUE}
tbl <- fit[ , c("Name", "LL", "Parameters", "n",
         "BIC", "Entropy",
         "prob_min", "prob_max", 
         "n_min", "n_max",
         "lmr_p")]
names(tbl) <- c("Name", "LL", "p", "n",
         "BIC", "Entropy",
         "p_min", "p_max", 
         "n_min", "n_max",
         "lmr_p")
papaja::apa_table(tbl, caption = "Model fit table")
```

Class enumeration unambiguously indicates a preference for the 4-class solution: the 4-class solution has the lowest BIC, which means it is preferred over all other solutions including a 1-class solution and a solution with more classes.
The p-values of the LMR LR test are significant for all pairwise model comparisons, except for the 5-class model.
Note that a scree plot for the BIC can be plotted by calling `plot(fit)`.
Following the elbow criterion, a three-class solution would also be defensible.
The function `ic_weights(tab_compare)` allows us to compute IC weights; it indicates that, conditional on the set of models, the 4-class model has a posterior model probability of nearly 100%.

## Optional: Alternative Model Specifications

In the case of confirmatory LCA, the theory would be refuted by strong evidence against the hypothesized model and number of classes.
In the preceding, we only compared the theoretical model against models with different number of classes.
Imagine, however, that a Reviewer argues that variance ought to be freely estimated across classes.
We could compare our theoretical model against their competing model as follows.
Note that we can put two models into a list to compare them.

```{r eval = run_everything, echo = FALSE}
res_alt <- mx_profiles(df, classes = 4, variances = "varying")
compare <- list(res[[4]], res_alt)
fit_compare <- table_fit(compare)
write.csv(fit_compare, "lpa_fit_compare.csv", row.names = FALSE)
```

```{r, echo = TRUE, eval = FALSE}
res_alt <- mx_profiles(df, classes = 4, variances = "varying")
compare <- list(res[[4]], res_alt)
table_fit(compare)
```

```{r tabfitcomp, echo = FALSE, eval = TRUE}
fit_compare <- read.csv("appendices/lpa_fit_compare.csv", stringsAsFactors = FALSE)
class(fit_compare) <- c("tidy_fit", "data.frame")
papaja::apa_table(fit_compare[ , c("Name", "LL", "Parameters",
         "BIC", "Entropy",
         "prob_min", "prob_max", 
         "n_min", "n_max")], caption = "Comparing competing theoretical models")
```

The alternative model incurs 12 additional parameters for the free variances.
Yet, it has a higher BIC, which indicates that this additional complexity does not outweigh the increase in fit.

## Interpreting the Final Class Solution

To interpret the final class solution, we first reorder the 4-class model by class size.
This helps prevent label switching.

```{r, echo = TRUE, eval =FALSE}
res_final <- mx_switch_labels(res[[4]])
```

```{r, echo = FALSE, eval = run_everything}
res_final <- mx_switch_labels(res[[4]])
cp <- class_prob(res_final)
out <- list(counts = cp$sum.posterior$proportion)
```

The 4-class model yielded classes of reasonable size; the largest class comprised 33%, and the smallest comprised 16% of cases.
However, the entropy was low, $S = .75$, indicating poor class separability.
Furthermore, the posterior classification probability ranged from $[.81, .92]$, which means that at least some classes had a high classification error.
We produce a table of the results below.

```{r echo = TRUE, eval = FALSE}
table_results(res_final, columns = c("label", "est", "se", "confint", "class"))
```

```{r echo = FALSE, eval = run_everything}
tab <- table_results(res_final, columns = c("label", "est", "se", "confint", "class"))
write.csv(tab, "lpa_tab_res.csv", row.names = FALSE)
```

```{r, eval = TRUE, echo=FALSE}
tab <- read.csv("appendices/lpa_tab_res.csv", stringsAsFactors = FALSE)
papaja::apa_table(tab, caption = "Four-class model results")
```

The results are best interpreted by examining a plot of the model and data, however.
Relevant plot functions are `plot_bivariate()`, `plot_density()`, and `plot_profiles()`.
However, we omit the density plots, because `plot_bivariate()` also includes them.

```{r echo = TRUE, eval = FALSE}
plot_bivariate(res_final)
```

```{r echo = FALSE, eval = run_everything}
p <- plot_bivariate(res_final, return_list = TRUE)
p[[1]] <- p[[1]] + scale_y_continuous(breaks= c(0, .1, .2, .3), labels = c(".0", ".1", ".2", ".3"))
p <- tidySEM:::merge_corplots(p)
ggsave("appendices/lpa_bivariate.pdf", p, device = "pdf", width = 210, height = 100, units = "mm", scale = 1.5)
```

```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/lpa_bivariate.pdf")
```

On the diagonal of the bivariate plot are weighted density plots: normal approximations of the density function of observed data, weighed by class probability.
On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.
As this model has zero covariances, all circles are round (albeit warped by the different scales of the X and Y axes)

The marginal density plots show that trappedness distinguishes classes rather well.
For all other indicators, groups are not always clearly separated in terms of marginal density: class 2 and 3 coalesce on negative affect, 1 and 2 coalesce on loneliness, and 1 and 2 coalesce on burden.
Nevertheless, the off-diagonal scatterplots show reasonable bivariate separation for all classes.

We can obtain a more classic profile plot using `plot_profiles(res_final)`.
This plot conveys less information than the bivariate plot, but is readily interpretable.
Below is a comparison between the most common type of visualization for LPA, and the best-practices visualization provided by `tidySEM`.
Note that the best practices plot includes class means and error bars, standard deviations, and a ribbon plot of raw data weighted by class probability to indicate how well the classes describe the observed distribution.
The overlap between the classes is clearly visible in this figure; this is why the entropy and classification probabilities are relatively low.

Based on the bivariate plot, we can label class 1 as the *balanced* type (33%), class 2 as the *imbalanced* type (29%), class 3 as the *entrapped* type (22%), and class 4 as the *lonely* type (16%).

```{r echo = TRUE, eval = FALSE}
plot_profiles(res_final)
```

```{r echo = FALSE, eval = run_everything}
p1 <- plot_profiles(res_final, ci = NULL, sd= FALSE, add_line = TRUE, rawdata = FALSE)
p1 <- p1 + theme(legend.position = c(.85, .2))
p2 <- plot_profiles(res_final)
p2 <- p2 + theme(legend.position = "none")
p <- ggpubr::ggarrange(p1, p2)
ggsave("appendices/lpa_profiles.pdf", p, device = "pdf", width = 210, height = 100, units = "mm", scale = 1)
```

```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/lpa_profiles.pdf")
```

## Auxiliary Analyses

Finally, we may want to compare the different classes on auxiliary variables or models.
The `BCH()` function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
We consider two examples: a single variable, and an auxiliary model.

### Comparing Means or Proportions Across Classes

For a single (continuous or ordinal) variable, we can call the BCH function and simply supply the auxiliary variable to the `data` argument, omitting the `model` argument.
Below, we estimate an auxiliary model to compare the sex of patients between classes:

```{r echo = TRUE, eval=FALSE}
aux_sex <- BCH(res_final, data = zegwaard_carecompass$sexpatient)
```

```{r echo = FALSE, eval=run_everything}
aux_sex <- BCH(res_final, data = zegwaard_carecompass$sexpatient)
saveRDS(aux_sex, "lpa_aux_sex.RData")
```

```{r echo = FALSE, eval=FALSE}
aux_sex <- readRDS("appendices/lpa_aux_sex.RData")
```

To obtain an omnibus likelihood ratio test of the significance of these sex differences across classes, as well as pairwise comparisons between classes, use `lr_test(aux_sex)`.
The results indicate that there are significant sex differences across classes, $\Delta LL(1) = 8.7, p = .003$.
Pairwise comparisons indicate that class 3 differs significantly from classes 1 and 2.
The results can be reported in probability scale using `table_prob(aux_sex)`.
It appears that the entrapped class disproportionately cares for female patients.

### Comparing Auxiliary Models Across Classes

Finally, we compare a simple model between classes.
Specifically, we will examine whether the distance predicts the frequency of visits differently across classes (treated as continuous).

```{r echo = TRUE, eval=FALSE}
df_aux <- zegwaard_carecompass[, c("freqvisit", "distance")]
df_aux$freqvisit <- as.numeric(df_aux$freqvisit)
aux_model <- BCH(res_final, model = "freqvisit ~ distance",
                 data = df_aux)
```

```{r echo = FALSE, eval=run_everything}
df_aux <- zegwaard_carecompass[, c("freqvisit", "distance")]
df_aux$freqvisit <- as.numeric(df_aux$freqvisit)
aux_model <- BCH(res_final, model = "freqvisit ~ distance",
               data = df_aux)
saveRDS(aux_model, "lpa_aux_model.RData")
```

```{r echo = FALSE, eval=FALSE}
aux_model <- readRDS("appendices/lpa_aux_model.RData")
```

To obtain an omnibus likelihood ratio test of the difference in regression coefficients across classes and pairwise comparisons between classes, use `lr_test(aux_model, compare = "A")`.
The results indicate that there are no significant sex differences across classes, $\Delta LL(3) = 0.98, p = .81$.
The results can be reported using `table_results(aux_model)`.

<!-- # References -->

<!-- ::: {#refs custom-style="Bibliography"} -->

<!-- ::: -->

## R session

```{r echo = TRUE, eval=TRUE}
sessionInfo()
```

### 2023-01-23 appendix_a.tex
edits by Mauricio

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Appendix A: Confirmatory LPA for the Caregiver Compass},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Appendix A: Confirmatory LPA for the Caregiver Compass}
\author{\phantom{0}}
\date{}

\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}

\affiliation{\phantom{0}}

\begin{document}
\maketitle

This is an example of confirmatory LPA using \texttt{tidySEM}.
The simulated data are based on work by Zegwaard and colleagues, who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
Qualitative research among experts resulted in a theory postulating the existence of four types of caregivers (translated from the original Dutch):

\textbf{Balanced}

\begin{quote}
The balanced caregiver experiences relative balance between the costs and benefits of caring for a close other.
\end{quote}

\textbf{Imbalanced}

\begin{quote}
The imbalanced caregiver experiences a precarious balance between the costs and benefits of caring for a close other.
\end{quote}

\textbf{Lonely}

\begin{quote}
The lonely caregiver experiences a strong sense of isolation.
\end{quote}

\textbf{Entrapped}

\begin{quote}
The entrapped caregiver strongly feels a sense of being entangled in responsibilities which are difficult to fulfill.
\end{quote}

To view the data documentation, run the command \texttt{?tidySEM::zegwaard\_carecompass} in the R console.

\hypertarget{loading-the-data}{%
\subsection{Loading the Data}\label{loading-the-data}}

To load the data, simply attach the \texttt{tidySEM} package.
For convenience, we assign the variables used for analysis to an object called \texttt{df}.
We first only use the four scales: \texttt{c("burdened",\ "trapped",\ "negaffect",\ "loneliness")}.
`.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Load required packages}
\FunctionTok{library}\NormalTok{(tidySEM) }
\FunctionTok{library}\NormalTok{(ggplot2)}
\CommentTok{\# Load data}
\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ zegwaard\_carecompass[, }\FunctionTok{c}\NormalTok{(}\StringTok{"burdened"}\NormalTok{, }\StringTok{"trapped"}\NormalTok{, }\StringTok{"negaffect"}\NormalTok{, }\StringTok{"loneliness"}\NormalTok{)]}
\end{Highlighting}
\end{Shaded}

\hypertarget{examining-the-data}{%
\subsection{Examining the Data}\label{examining-the-data}}

As per the best practices, the first step in LCA is examining the observed data.
We use \texttt{tidySEM::descriptives()} to describe the data numerically.
Because all scales are continuous, we select only columns for continuous data to de-clutter the table:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ tidySEM}\SpecialCharTok{::}\FunctionTok{descriptives}\NormalTok{(df)}
\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{,}
                 \StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{,}
                 \StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
\NormalTok{desc}
\end{Highlighting}
\end{Shaded}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:tabdesc}Descriptive statistics}

\begin{tabular}{lllllllllll}
\toprule
name & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{missing} & \multicolumn{1}{c}{unique} & \multicolumn{1}{c}{mean} & \multicolumn{1}{c}{median} & \multicolumn{1}{c}{sd} & \multicolumn{1}{c}{min} & \multicolumn{1}{c}{max} & \multicolumn{1}{c}{skew\_2se} & \multicolumn{1}{c}{kurt\_2se}\\
\midrule
burdened & 509 & 0.01 & 509 & 3.38 & 3.39 & 0.75 & 1.20 & 5.28 & 0.17 & 6.51\\
trapped & 505 & 0.02 & 505 & 1.73 & 1.79 & 0.90 & -0.86 & 3.81 & -1.03 & 5.39\\
negaffect & 506 & 0.01 & 506 & 2.50 & 2.52 & 0.69 & 0.71 & 4.98 & 0.08 & 6.52\\
loneliness & 510 & 0.01 & 510 & 2.66 & 2.69 & 0.62 & 0.98 & 4.22 & -0.33 & 6.31\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

The table indicates two potential causes for concern: there is a small percentage of missingness, and all variables have relatively high kurtosis.
Since there are some missing values, we can conduct an MCAR test using \texttt{mice::mcar(df)}.
According to Hawkins' test, there is no evidence to reject the assumptions of multivariate normality and MCAR, \(\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71\).
Missing data will be accounted for using FIML.

Additionally, we can plot the data.
The \texttt{ggplot2} function \texttt{geom\_density()} is useful for continuous data.
Visual inspection confirms the conclusions from the \texttt{descriptives()} table: the data are kurtotic (peaked).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df\_plot }\OtherTok{\textless{}{-}}\NormalTok{ df}
\FunctionTok{names}\NormalTok{(df\_plot) }\OtherTok{\textless{}{-}} \FunctionTok{paste0}\NormalTok{(}\StringTok{"Value."}\NormalTok{, }\FunctionTok{names}\NormalTok{(df\_plot))}
\NormalTok{df\_plot }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(df\_plot, }\AttributeTok{varying =} \FunctionTok{names}\NormalTok{(df\_plot), }\AttributeTok{direction =} \StringTok{"long"}\NormalTok{,}
                   \AttributeTok{timevar =} \StringTok{"Variable"}\NormalTok{)}
\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value)) }\SpecialCharTok{+}
  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{Variable)}\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\includegraphics{appendices/plot_lpa_desc.pdf}

\hypertarget{conducting-latent-profile-analysis}{%
\subsection{Conducting Latent Profile Analysis}\label{conducting-latent-profile-analysis}}

As all variables are continuous, we can use the convenience function \texttt{tidySEM::mx\_profiles()}, which is a wrapper for the generic function \texttt{mx\_mixture()} optimized for continuous indicators.
Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
Its arguments are \texttt{data} and number of \texttt{classes}.
All variables in \texttt{data} are included in the analysis, which is why we first selected the indicator variables.
As this is a confirmatory LCA, we do not follow a strictly data-driven class enumeration procedure.
We will set the maximum number of classes \(K\) to one more than the theoretically expected number.
We set a seed to ensure replicable results.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df,}
                   \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

This analysis should produce some warnings about cluster initialization.
These relate to the selection of starting values, which relies on the K-means algorithm and is not robust to missing data.
The algorithm automatically switches to hierarchical clustering, no further action is required.

\hypertarget{class-enumeration}{%
\subsection{Class Enumeration}\label{class-enumeration}}

To compare the fit of the theoretical model against other models, we create a model fit table using \texttt{table\_fit()} and retain relevant columns.
We also determine whether any models can be disqualified.

In this example, all models converge without issues.
If, for example, the two-class solution had not converged, we could use the function \texttt{res{[}{[}2{]}{]}\ \textless{}-\ mxTryHard(res{[}{[}2{]}{]})} to aid convergence.

Next, we check for local identifiability.
The sample size is consistently reported as 513, which means that partially missing cases were indeed included via FIML.
The smallest class size occurs in the 5-class model, where the smallest class is assigned 7\% of cases, or 38 cases.
This model has 28 parameters, approximately 6 per class.
We thus have at least five observations per parameter in every class, and do not disqualify the 5-class model.

There are concerns about theoretical interpretability of all solutions, as the entropies and minimum classification probabilities are all low.
However, in this confirmatory use case, we address this when interpreting the results.

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:tabfit}Model fit table}

\begin{tabular}{lllllllllll}
\toprule
Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{p} & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Entropy} & \multicolumn{1}{c}{p\_min} & \multicolumn{1}{c}{p\_max} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{n\_max} & \multicolumn{1}{c}{lmr\_p}\\
\midrule
equal var 1 & -2,241.98 & 8 & 513 & 4,533.89 & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 & NA\\
equal var 2 & -2,031.37 & 13 & 513 & 4,143.85 & 0.74 & 0.91 & 0.93 & 0.42 & 0.58 & 0.00\\
equal var 3 & -1,951.35 & 18 & 513 & 4,015.02 & 0.78 & 0.89 & 0.91 & 0.19 & 0.54 & 0.00\\
equal var 4 & -1,916.25 & 23 & 513 & 3,976.03 & 0.75 & 0.81 & 0.92 & 0.16 & 0.34 & 0.00\\
equal var 5 & -1,912.93 & 28 & 513 & 4,000.59 & 0.70 & 0.39 & 0.92 & 0.07 & 0.31 & 0.28\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

Class enumeration unambiguously indicates a preference for the 4-class solution: the 4-class solution has the lowest BIC, which means it is preferred over all other solutions including a 1-class solution and a solution with more classes.
The p-values of the LMR LR test are significant for all pairwise model comparisons, except for the 5-class model.
Note that a scree plot for the BIC can be plotted by calling \texttt{plot(fit)}.
Following the elbow criterion, a three-class solution would also be defensible.
The function \texttt{ic\_weights(tab\_compare)} allows us to compute IC weights; it indicates that, conditional on the set of models, the 4-class model has a posterior model probability of nearly 100\%.

\hypertarget{optional-alternative-model-specifications}{%
\subsection{Optional: Alternative Model Specifications}\label{optional-alternative-model-specifications}}

In the case of confirmatory LCA, the theory would be refuted by strong evidence against the hypothesized model and number of classes.
In the preceding, we only compared the theoretical model against models with different number of classes.
Imagine, however, that a Reviewer argues that variance ought to be freely estimated across classes.
We could compare our theoretical model against their competing model as follows.
Note that we can put two models into a list to compare them.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res\_alt }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(df, }\AttributeTok{classes =} \DecValTok{4}\NormalTok{, }\AttributeTok{variances =} \StringTok{"varying"}\NormalTok{)}
\NormalTok{compare }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(res[[}\DecValTok{4}\NormalTok{]], res\_alt)}
\FunctionTok{table\_fit}\NormalTok{(compare)}
\end{Highlighting}
\end{Shaded}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:tabfitcomp}Comparing competing theoretical models}

\begin{tabular}{lllllllll}
\toprule
Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{Parameters} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Entropy} & \multicolumn{1}{c}{prob\_min} & \multicolumn{1}{c}{prob\_max} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{n\_max}\\
\midrule
1 & -1,916.25 & 23 & 3,976.03 & 0.75 & 0.81 & 0.92 & 0.16 & 0.34\\
2 & -1,909.23 & 35 & 4,036.87 & 0.78 & 0.84 & 0.92 & 0.16 & 0.32\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

The alternative model incurs 12 additional parameters for the free variances.
Yet, it has a higher BIC, which indicates that this additional complexity does not outweigh the increase in fit.

\hypertarget{interpreting-the-final-class-solution}{%
\subsection{Interpreting the Final Class Solution}\label{interpreting-the-final-class-solution}}

To interpret the final class solution, we first reorder the 4-class model by class size.
This helps prevent label switching.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res\_final }\OtherTok{\textless{}{-}} \FunctionTok{mx\_switch\_labels}\NormalTok{(res[[}\DecValTok{4}\NormalTok{]])}
\end{Highlighting}
\end{Shaded}

The 4-class model yielded classes of reasonable size; the largest class comprised 33\%, and the smallest comprised 16\% of cases.
However, the entropy was low, \(S = .75\), indicating poor class separability.
Furthermore, the posterior classification probability ranged from \([.81, .92]\), which means that at least some classes had a high classification error.
We produce a table of the results below.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{table\_results}\NormalTok{(res\_final, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"confint"}\NormalTok{, }\StringTok{"class"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:unnamed-chunk-13}Four-class model results}

\begin{tabular}{lllll}
\toprule
label & \multicolumn{1}{c}{est} & \multicolumn{1}{c}{se} & \multicolumn{1}{c}{confint} & \multicolumn{1}{c}{class}\\
\midrule
mix4.weights[1,2] & 0.86 & 0.15 & {}[0.56, 1.15] & NA\\
mix4.weights[1,3] & 0.66 & 0.11 & {}[0.44, 0.88] & NA\\
mix4.weights[1,4] & 0.47 & 0.08 & {}[0.32, 0.63] & NA\\
Variances.burdened & 0.23 & 0.02 & {}[0.19, 0.27] & class1\\
Variances.trapped & 0.17 & 0.02 & {}[0.14, 0.20] & class1\\
Variances.negaffect & 0.31 & 0.02 & {}[0.27, 0.36] & class1\\
Variances.loneliness & 0.24 & 0.02 & {}[0.20, 0.28] & class1\\
Means.burdened & 3.27 & 0.04 & {}[3.18, 3.36] & class1\\
Means.trapped & 1.28 & 0.05 & {}[1.18, 1.38] & class1\\
Means.negaffect & 2.31 & 0.06 & {}[2.20, 2.42] & class1\\
Means.loneliness & 2.73 & 0.04 & {}[2.64, 2.82] & class1\\
Means.burdened & 3.40 & 0.06 & {}[3.28, 3.52] & class2\\
Means.trapped & 2.27 & 0.06 & {}[2.15, 2.38] & class2\\
Means.negaffect & 2.81 & 0.06 & {}[2.70, 2.93] & class2\\
Means.loneliness & 2.79 & 0.06 & {}[2.66, 2.91] & class2\\
Means.burdened & 4.25 & 0.07 & {}[4.12, 4.38] & class3\\
Means.trapped & 2.67 & 0.05 & {}[2.58, 2.77] & class3\\
Means.negaffect & 2.92 & 0.06 & {}[2.80, 3.03] & class3\\
Means.loneliness & 2.01 & 0.06 & {}[1.89, 2.14] & class3\\
Means.burdened & 2.38 & 0.06 & {}[2.26, 2.50] & class4\\
Means.trapped & 0.38 & 0.05 & {}[0.28, 0.49] & class4\\
Means.negaffect & 1.78 & 0.07 & {}[1.65, 1.91] & class4\\
Means.loneliness & 3.18 & 0.06 & {}[3.07, 3.30] & class4\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

The results are best interpreted by examining a plot of the model and data, however.
Relevant plot functions are \texttt{plot\_bivariate()}, \texttt{plot\_density()}, and \texttt{plot\_profiles()}.
However, we omit the density plots, because \texttt{plot\_bivariate()} also includes them.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot\_bivariate}\NormalTok{(res\_final)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=1\linewidth]{appendices/lpa_bivariate} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-16}
\end{figure}

On the diagonal of the bivariate plot are weighted density plots: normal approximations of the density function of observed data, weighed by class probability.
On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.
As this model has zero covariances, all circles are round (albeit warped by the different scales of the X and Y axes)

The marginal density plots show that trappedness distinguishes classes rather well.
For all other indicators, groups are not always clearly separated in terms of marginal density: class 2 and 3 coalesce on negative affect, 1 and 2 coalesce on loneliness, and 1 and 2 coalesce on burden.
Nevertheless, the off-diagonal scatterplots show reasonable bivariate separation for all classes.

We can obtain a more classic profile plot using \texttt{plot\_profiles(res\_final)}.
This plot conveys less information than the bivariate plot, but is readily interpretable.
Below is a comparison between the most common type of visualization for LPA, and the best-practices visualization provided by \texttt{tidySEM}.
Note that the best practices plot includes class means and error bars, standard deviations, and a ribbon plot of raw data weighted by class probability to indicate how well the classes describe the observed distribution.
The overlap between the classes is clearly visible in this figure; this is why the entropy and classification probabilities are relatively low.

Based on the bivariate plot, we can label class 1 as the \emph{balanced} type (33\%), class 2 as the \emph{imbalanced} type (29\%), class 3 as the \emph{entrapped} type (22\%), and class 4 as the \emph{lonely} type (16\%).

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot\_profiles}\NormalTok{(res\_final)}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=1\linewidth]{appendices/lpa_profiles} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-19}
\end{figure}

\hypertarget{auxiliary-analyses}{%
\subsection{Auxiliary Analyses}\label{auxiliary-analyses}}

Finally, we may want to compare the different classes on auxiliary variables or models.
The \texttt{BCH()} function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
We consider two examples: a single variable, and an auxiliary model.

\hypertarget{comparing-means-or-proportions-across-classes}{%
\subsubsection{Comparing Means or Proportions Across Classes}\label{comparing-means-or-proportions-across-classes}}

For a single (continuous or ordinal) variable, we can call the BCH function and simply supply the auxiliary variable to the \texttt{data} argument, omitting the \texttt{model} argument.
Below, we estimate an auxiliary model to compare the sex of patients between classes:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{aux\_sex }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_final, }\AttributeTok{data =}\NormalTok{ zegwaard\_carecompass}\SpecialCharTok{$}\NormalTok{sexpatient)}
\end{Highlighting}
\end{Shaded}

To obtain an omnibus likelihood ratio test of the significance of these sex differences across classes, as well as pairwise comparisons between classes, use \texttt{lr\_test(aux\_sex)}.
The results indicate that there are significant sex differences across classes, \(\Delta LL(1) = 8.7, p = .003\).
Pairwise comparisons indicate that class 3 differs significantly from classes 1 and 2.
The results can be reported in probability scale using \texttt{table\_prob(aux\_sex)}.
It appears that the entrapped class disproportionately cares for female patients.

\hypertarget{comparing-auxiliary-models-across-classes}{%
\subsubsection{Comparing Auxiliary Models Across Classes}\label{comparing-auxiliary-models-across-classes}}

Finally, we compare a simple model between classes.
Specifically, we will examine whether the distance predicts the frequency of visits differently across classes (treated as continuous).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df\_aux }\OtherTok{\textless{}{-}}\NormalTok{ zegwaard\_carecompass[, }\FunctionTok{c}\NormalTok{(}\StringTok{"freqvisit"}\NormalTok{, }\StringTok{"distance"}\NormalTok{)]}
\NormalTok{df\_aux}\SpecialCharTok{$}\NormalTok{freqvisit }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(df\_aux}\SpecialCharTok{$}\NormalTok{freqvisit)}
\NormalTok{aux\_model }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_final, }\AttributeTok{model =} \StringTok{"freqvisit \textasciitilde{} distance"}\NormalTok{,}
                 \AttributeTok{data =}\NormalTok{ df\_aux)}
\end{Highlighting}
\end{Shaded}

To obtain an omnibus likelihood ratio test of the difference in regression coefficients across classes and pairwise comparisons between classes, use \texttt{lr\_test(aux\_model,\ compare\ =\ "A")}.
The results indicate that there are no significant sex differences across classes, \(\Delta LL(3) = 0.98, p = .81\).
The results can be reported using \texttt{table\_results(aux\_model)}.

\hypertarget{r-session}{%
\subsection{R session}\label{r-session}}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{sessionInfo}\NormalTok{()}
\CommentTok{\#\textgreater{} R version 4.1.3 (2022{-}03{-}10)}
\CommentTok{\#\textgreater{} Platform: x86\_64{-}w64{-}mingw32/x64 (64{-}bit)}
\CommentTok{\#\textgreater{} Running under: Windows 10 x64 (build 19045)}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Matrix products: default}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} locale:}
\CommentTok{\#\textgreater{} [1] LC\_COLLATE=English\_United Kingdom.1252 }
\CommentTok{\#\textgreater{} [2] LC\_CTYPE=English\_United Kingdom.1252   }
\CommentTok{\#\textgreater{} [3] LC\_MONETARY=English\_United Kingdom.1252}
\CommentTok{\#\textgreater{} [4] LC\_NUMERIC=C                           }
\CommentTok{\#\textgreater{} [5] LC\_TIME=English\_United Kingdom.1252    }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} attached base packages:}
\CommentTok{\#\textgreater{} [1] stats     graphics  grDevices utils     datasets  methods   base     }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} other attached packages:}
\CommentTok{\#\textgreater{} [1] ggplot2\_3.4.0    tidySEM\_0.2.4.6  OpenMx\_2.21.1    scales\_1.2.1    }
\CommentTok{\#\textgreater{} [5] yaml\_2.3.6       papaja\_0.1.1     tinylabels\_0.2.3}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} loaded via a namespace (and not attached):}
\CommentTok{\#\textgreater{}   [1] TH.data\_1.1{-}1         colorspace\_2.1{-}0      estimability\_1.4.1   }
\CommentTok{\#\textgreater{}   [4] parameters\_0.20.1     rstudioapi\_0.14       listenv\_0.9.0        }
\CommentTok{\#\textgreater{}   [7] lavaan\_0.6{-}13         rstan\_2.26.13         fansi\_1.0.4          }
\CommentTok{\#\textgreater{}  [10] mvtnorm\_1.1{-}3         codetools\_0.2{-}18      splines\_4.1.3        }
\CommentTok{\#\textgreater{}  [13] mnormt\_2.1.1          knitr\_1.41            texreg\_1.38.6        }
\CommentTok{\#\textgreater{}  [16] bayesplot\_1.10.0      jsonlite\_1.8.4        effectsize\_0.8.2     }
\CommentTok{\#\textgreater{}  [19] compiler\_4.1.3        httr\_1.4.4            emmeans\_1.8.4{-}1      }
\CommentTok{\#\textgreater{}  [22] backports\_1.4.1       assertthat\_0.2.1      Matrix\_1.5{-}3         }
\CommentTok{\#\textgreater{}  [25] fastmap\_1.1.0         cli\_3.6.0             htmltools\_0.5.4      }
\CommentTok{\#\textgreater{}  [28] prettyunits\_1.1.1     tools\_4.1.3           igraph\_1.3.5         }
\CommentTok{\#\textgreater{}  [31] coda\_0.19{-}4           gtable\_0.3.1          glue\_1.6.2           }
\CommentTok{\#\textgreater{}  [34] RANN\_2.6.1            dplyr\_1.0.10          V8\_4.2.2             }
\CommentTok{\#\textgreater{}  [37] Rcpp\_1.0.10           carData\_3.0{-}5         vctrs\_0.5.2          }
\CommentTok{\#\textgreater{}  [40] nlme\_3.1{-}161          psych\_2.2.9           insight\_0.18.8       }
\CommentTok{\#\textgreater{}  [43] xfun\_0.36             stringr\_1.5.0         globals\_0.16.2       }
\CommentTok{\#\textgreater{}  [46] ps\_1.7.2              proto\_1.0.0           CompQuadForm\_1.4.3   }
\CommentTok{\#\textgreater{}  [49] lifecycle\_1.0.3       future\_1.30.0         MASS\_7.3{-}58.2        }
\CommentTok{\#\textgreater{}  [52] zoo\_1.8{-}11            parallel\_4.1.3        sandwich\_3.0{-}2       }
\CommentTok{\#\textgreater{}  [55] inline\_0.3.19         curl\_5.0.0            gridExtra\_2.3        }
\CommentTok{\#\textgreater{}  [58] pander\_0.6.5          blavaan\_0.4{-}3         loo\_2.5.1            }
\CommentTok{\#\textgreater{}  [61] StanHeaders\_2.26.13   stringi\_1.7.12        highr\_0.10           }
\CommentTok{\#\textgreater{}  [64] bayestestR\_0.13.0     fastDummies\_1.6.3     checkmate\_2.1.0      }
\CommentTok{\#\textgreater{}  [67] boot\_1.3{-}28           pkgbuild\_1.4.0        rlang\_1.0.6          }
\CommentTok{\#\textgreater{}  [70] pkgconfig\_2.0.3       matrixStats\_0.63.0    evaluate\_0.20        }
\CommentTok{\#\textgreater{}  [73] lattice\_0.20{-}45       rstantools\_2.2.0      processx\_3.8.0       }
\CommentTok{\#\textgreater{}  [76] tidyselect\_1.2.0      parallelly\_1.34.0     plyr\_1.8.8           }
\CommentTok{\#\textgreater{}  [79] magrittr\_2.0.3        bookdown\_0.32         R6\_2.5.1             }
\CommentTok{\#\textgreater{}  [82] generics\_0.1.3        multcomp\_1.4{-}20       DBI\_1.1.3            }
\CommentTok{\#\textgreater{}  [85] withr\_2.5.0           gsubfn\_0.7            pillar\_1.8.1         }
\CommentTok{\#\textgreater{}  [88] survival\_3.5{-}0        datawizard\_0.6.5      abind\_1.4{-}5          }
\CommentTok{\#\textgreater{}  [91] tibble\_3.1.8          future.apply\_1.10.0   crayon\_1.5.2         }
\CommentTok{\#\textgreater{}  [94] car\_3.1{-}1             nonnest2\_0.5{-}5        utf8\_1.2.2           }
\CommentTok{\#\textgreater{}  [97] tmvnsim\_1.0{-}2         MplusAutomation\_1.1.0 rmarkdown\_2.20       }
\CommentTok{\#\textgreater{} [100] grid\_4.1.3            data.table\_1.14.6     pbivnorm\_0.6.0       }
\CommentTok{\#\textgreater{} [103] bain\_0.2.8            callr\_3.7.3           digest\_0.6.31        }
\CommentTok{\#\textgreater{} [106] xtable\_1.8{-}4          dbscan\_1.1{-}11         RcppParallel\_5.1.6   }
\CommentTok{\#\textgreater{} [109] stats4\_4.1.3          munsell\_0.5.0}
\end{Highlighting}
\end{Shaded}

\end{document}

### 2023-01-23 appendix_b.Rmd
edits by Mauricio

---
title             : "Appendix B: Exploratory LPA for Ocean Microplastics"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
bibliography      : "r-references.bib"
floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
library(yaml)
library(scales)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(width.cutoff = 60), tidy = TRUE
)
options(scipen = 1, digits = 2)

run_everything = FALSE

pkgdown <- TRUE
```

This is an example of exploratory latent class analysis (LCA) with continuous indicators, otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using `tidySEM`.
The present example uses data collected by Alkema as part of a study on ocean microplastics.
The purpose of this study was to provide a more nuanced model for the distribution of different sizes of ocean microplastics than the commonly used normal distribution.
To this end, a mixture of normals was used.
Since there is no theoretical reason to expect a certain number of classes, this is an exploratory LCA.
To view its documentation, run the command `?tidySEM::alkema_microplastics` in the R console.
The original analyses are available at <https://github.com/cjvanlissa/lise_microplastics>; in this vignette, we take a different approach to the analysis to showcase other possibilities.

## Loading the Data

To load the data, simply attach the `tidySEM` package.
For convenience, we assign the variables used for analysis to an object called `df`.
As explained in the paper, the classes are quite different for lines, films, and fragments.
For this reason, we here only use data from fragments.
The indicators are fragments' length and width.

```{r, echo = TRUE, eval=TRUE}
# Load required packages
library(tidySEM) 
library(ggplot2)
# Load data
df_analyze <- alkema_microplastics[alkema_microplastics$category == "Fragment", ]
df <- df_analyze[ ,c("length", "width")]
```

## Examining the Data

As per the best practices, the first step in LCA is examining the observed data.
We use `tidySEM::descriptives()` to describe the data numerically.
Because all items are categorical, we remove columns for continuous data to de-clutter the table:

```{r tabdesc, echo = TRUE, eval=TRUE, results='asis'}
desc <- tidySEM::descriptives(df)
desc <- desc[, c("name", "type", "n", "missing", "unique", 
"mean", "median", "sd", "min", "max", "skew_2se", "kurt_2se")]
papaja::apa_table(desc, caption = "Descriptive statistics")
```

Additionally, we can plot the data.
The `ggplot2` function `geom_density()` is useful to visualize continuous data:

```{r, echo = TRUE, eval = FALSE}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
```

```{r, echo = FALSE, eval = run_everything}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
p <- ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
ggsave("appendices/plot_gmm_desc.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```

```{r figdesc, echo = FALSE, eval = pkgdown}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
knitr::include_graphics("appendices/plot_gmm_desc.pdf")
```

The data are correctly coded as `numeric`.
There are no missing values; if any variables had missing values, we would report an MCAR test with `mice::mcar()`, and explain that missing data are accounted for using FIML.
Both the table above and the density plot indicate that the data are extremely right-skewed and kurtotic.
With this in mind, it can be useful to transform and rescale the data.
We will use a log transformation.

```{r, echo = TRUE, eval = FALSE}
df_plot$Value <- log(df_plot$Value)
ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
```

```{r, echo = FALSE, eval = run_everything}
df_plot$Value <- log(df_plot$Value)
p <- ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
ggsave("appendices/plot_gmm_desc_log.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```

```{r figdesclog, echo = FALSE, eval = pkgdown}
knitr::include_graphics("appendices/plot_gmm_desc_log.pdf")
```

The log transformation addresses the aforementioned concerns regarding skew and kurtosis.
To confirm this, reshape the data to wide format and examine a scatterplot:

```{r, eval = FALSE, echo = TRUE}
df <- reshape(df_plot, direction = "wide", v.names = "Value")[, -1]
names(df) <- gsub("Value.", "", names(df), fixed = TRUE)
ggplot(df, aes(x = length, y = width)) +
  geom_point(alpha = .1) +
  theme_bw()
```

```{r, eval = run_everything, echo = FALSE}
df <- reshape(df_plot, direction = "wide", v.names = "Value")[, -1]
names(df) <- gsub("Value.", "", names(df), fixed = TRUE)
p <- ggplot(df, aes(x = length, y = width)) +
  geom_point(alpha = .1) +
  theme_bw()
ggsave("appendices/plot_gmm_scatter.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```

```{r figscatter, eval = pkgdown, echo = FALSE}
knitr::include_graphics("appendices/plot_gmm_scatter.pdf")
```

<!-- As evident from the scatterplot, there is a very strong linear tend in the data. -->

<!-- This indicates that fragments tend to be coextensive (when length goes up, width goes up). -->

<!-- We can analyze these data in their original dimensions (length and width). -->

<!-- Alternatively, we can use PCA to rotate the data such that the first dimension can be interpreted as "size", -->

<!-- and the second dimension can be interpreted as "shape": -->

```{r echo = FALSE, eval = FALSE}
pca <- princomp(df)
df <- data.frame(pca$scores)
names(df) <- c("pc1", "pc2")
```

## Conducting Latent Profile Analysis

As all variables are continuous, we can use the convenience function `tidySEM::mx_profiles()`, which is a wrapper for the generic function `mx_mixture()` optimized for continuous indicators.
Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
Its arguments are `data` and number of `classes`.
All variables in `data` are included in the analysis, which is why we first selected the indicator variables.

As this is an exploratory LCA, we will conduct a rather extensive search across model specifications and number of classes.
We will set the maximum number of classes $K$ to four; depending on the results, we can always choose to increase it later.
We set a seed to ensure replicable results.

As the analysis takes a long time to compute, it is prudent to save the results to disk immediately, so as not to lose them.
For this, we use the function `saveRDS()`.
We can later use `res <- readRDS("res_gmm.RData")` to load the analysis from the file.

```{r fitlca, eval = FALSE, echo = TRUE}
set.seed(123)
res <- mx_profiles(data = df,
                   classes = 1:4,
                   variances = c("equal", "varying"),
                   covariances = c("zero", "equal",
                                   "varying"),
                   expand_grid = TRUE)
saveRDS(res, "res_gmm.RData")
```

```{r eval = run_everything, echo = FALSE}
set.seed(123)
res <- mx_profiles(data = df,
                   classes = 1:4,
                   variances = c("equal", "varying"),
                   covariances = c("zero", "equal",
                                   "varying"),
                   expand_grid = TRUE)
saveRDS(res, "appendices/res_gmm.RData")
fit <- table_fit(res)
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
saveRDS(fit, "appendices/gmm_tabfit.RData")
#"Warning: In model 'mix4' Optimizer returned a non-zero status code 6. The model does not satisfy the first-order optimality conditions to the required accuracy, and no improved point for the merit function could be found during the final linesearch (Mx status RED)"
```

```{r eval = TRUE, echo = FALSE}
fit <- readRDS("appendices/gmm_tabfit.RData")
```

<!-- Note that one model (the most complex one) returns the following warning: -->

<!-- > Warning: In model 'mix4' Optimizer returned a non-zero status code 6. The model does not satisfy the first-order optimality conditions to the required accuracy, and no improved point for the merit function could be found during the final linesearch (Mx status RED) -->

<!-- This suggests a potential problem with with model convergence. -->

<!-- To try and get this model to converge, it is possible to use the function `mxTryHard()`, which permutes the starting values and tries to find a better solution. -->

<!-- To do this, run `res[[24]] <- mxTryHard(res[[24]])`. -->

<!-- However, at present, we will skip this step, -->

<!-- because there are also other indications that this 4-class model is overfitted (see below). -->

## Class Enumeration

To compare the fit of the estimated models, we create a model fit table using `table_fit()`.

```{r echo = TRUE, eval=F}
fit <- table_fit(res)
```

First, we determine whether any models can be disqualified.
There were no indications of convergence problems during estimation, so this is not a reason to disqualify solutions.
Next, we check for local identifiability.
The sample size is `5605`.
We can calculate the ratio of observations to parameters and append it to the fit table:

```{r echo = TRUE, eval = FALSE}
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
fit[ , c("Name", "LL", "Parameters", "par_ratio",
         "BIC", "Entropy",
         "prob_min", "prob_max", 
         "n_min", "n_max",
         "lmr_p")]
```

```{r tabfit, echo = FALSE, eval = TRUE}
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
tbl <- fit[ , c("Name", "LL", "Parameters", "par_ratio",
         "BIC", "Entropy",
         "prob_min",
         "n_min",
         "lmr_p")]
names(tbl) <- c("Name", "LL", "p", "par_ratio",
         "BIC", "Ent.",
         "p_min",
         "n_min",
         "lmr_p")
papaja::apa_table(tbl, caption = "Model fit table. Some columns removed to fit page.")
```

As can be seen from the fit table, the lowest ratio of observations to parameters is 18, which is no cause for concern.

However, note that we have a very large sample, and for many models, the smallest class comprises only a very small percentage of the total sample.
Since the purpose of this analysis is to better represent the distribution of ocean microplastics, we can wonder whether it makes sense to allow for classes that only describe a small percentage of the cases.
We therefore set a lower bound for class size and only consider solutions that capture at least 10% of the sample.

An interesting characteristic of this data is that the BIC and the entropy are strongly correlated.
The raw correlation between these two metrics is .66, `cor(fit$BIC, fit$Entropy)`.
If we omit the 1-class models, for which entropy is technically not defined, the correlation is even as high as .85, `cor(fit$BIC[!fit$Classes == 1], fit$Entropy[!fit$Classes == 1])`.

This strong correlation indicates that an increase in fit comes with a decrease in class separability.
This illustrates why entropy should not be treated as a model fit criterion.
It also illustrates that criteria for class enumeration should be explicit, because we will likely come to a different decision depending on which criteria are used.

As mentioned before, we drop models with \< 10% of cases in the smallest class:

```{r echo = TRUE, eval = run_everything}
fit <- fit[!fit$n_min < .1, ]
```

If our strategy is to optimize fit, we can examine the fit table above, or plot a scree plot for the BIC by calling `plot(fit)`.
Note that, due to the large sample size, all ICs give identical conclusions.

```{r echo = TRUE, eval = FALSE}
plot(fit) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r echo = FALSE, eval = run_everything}
p <- plot(fit) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("appendices/gmm_plotfit.pdf", p, device = "pdf", width = 100, height = 100, units = "mm", scale = 1)
```

```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/gmm_plotfit.pdf")
```

Looking at the blocks of 1-4 class models for each model specification, it appears that the BIC keeps decreasing with the addition of more classes.
Across the blocks, the BIC keeps decreasing with increasingly complex model specifications.
Similarly, all LMR tests are significant, indicating a preference for more complex models.

Out of the 16 models that remain after removing those with \< 10% of cases in the smallest class, one model stands out: The 2-class model with free (co)variances.
<!-- The function `ic_weights(fit)` allows us to compute IC weights for all models in the set; --> <!-- it prefers the most complex model out of the 16 models with a posterior model probability of nearly 100%. --> We thus select this as our final model.

## Interpreting the Final Class Solution

and extract the classification probabilities and model results.
We here request the estimates (`est`) and standardized estimates `std_est`, because the latter allows us to interpret the correlations between length and width.
Note that there is little sense in requesting the standard errors and p-values: With a sample size of 5606, every parameter is significantly different from zero.

```{r echo = TRUE, eval = FALSE}
res_bic <- res[["free var, free cov 2"]]
cp <- class_prob(res_bic)
results <- table_results(res_bic, columns = c("label", "est", "std_est"))
results
```

```{r echo = FALSE, eval = run_everything}
res_bic <- res[["free var, free cov 2"]]
cp <- class_prob(res_bic)
results <- table_results(res_bic, columns = c("label", "est", "est_std"))
saveRDS(results, "appendices/gmm_results.RData")
```

```{r eval = TRUE, echo = FALSE}
results <- readRDS("appendices/gmm_results.RData")
papaja::apa_table(results, digits = 2, caption = "Results of a 2-class model with free (co)variances")
```

Interpreting the results is facilitated by examining a plot of the model and data.
Relevant plot functions are `plot_bivariate()`, `plot_density()`, and `plot_profiles()`.
However, we omit the density plots, because `plot_bivariate()` also includes them.

```{r echo = TRUE, eval = FALSE}
plot_bivariate(res_bic)
```

```{r echo = FALSE, eval = run_everything}
p <- plot_bivariate(res_bic)
ggsave("appendices/gmm_bivariate_bic.pdf", p, device = "pdf", width = 210, height = 100, units = "mm", scale = 1.5)
```

```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/gmm_bivariate_bic.pdf")
```

On the diagonal of the bivariate plot are weighted density plots: normal approximations of the density function of observed data, weighed by class probability.
On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.

The bivariate and marginal plots show that the classes are not clearly separable, as also evident from the low entropy.
At the same time however, it is clear that the distributions are non-normal, and the second class accounts for some of this non-normality (there is a smaller 'bump' to the right of the mode, which could be the mean of a second normal distribution).
The first class (57%) accounts for smaller fragments, and the second class (43%) accounts for some of the right-skew in fragments' length and width.
We can simply label class 1 as *small fragments*, and class 2 as *larger fragments*.

It also appears that the correlation between length and width is stronger for small fragments than for large fragments.
To test the difference, use `wald_test(res_bic, hypothesis = "c11 = c21")`.
Results indicate that the correlation is indeed significantly larger for small fragments ($r = .92$) than for larger fragments ($r = .85$), $\chi^2(1) = 11.56, p < .001$.
Thus, small fragments are more coextensive than large fragments.

There are, however, concerns about the interpretability of this solutions: the entropy is `.56` and the minimum classification probability is`.81`.
This is because of substantial overlap in the distributions of the two classes.

## Auxiliary Analyses

Finally, we may want to compare the different classes on auxiliary variables or models.
The `BCH()` function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
For example, we can test whether polymer type differs between the two classes:

```{r echo = TRUE, eval=FALSE}
df_pt <- mx_dummies(df_analyze$poly_type)
aux_pt <- BCH(res_bic, model = "poly_typeOther | t1
                                poly_typePE | t1
                                poly_typePP | t1", data = df_pt)
aux_pt <- mxTryHardOrdinal(aux_pt)
```

```{r echo = FALSE, eval=run_everything}
df_pt <- mx_dummies(df_analyze$poly_type)
aux_pt <- BCH(res_final, model = "poly_typeOther | t1
                                  poly_typePE | t1
                                  poly_typePP | t1", data = df_pt)
aux_pt <- mxTryHardOrdinal(aux_pt)
saveRDS(aux_pt, "appendices/gmm_aux_pt.RData")
```

```{r echo = FALSE, eval=FALSE}
aux_pt <- readRDS("appendices/gmm_aux_pt.RData")
```

To obtain an omnibus likelihood ratio test of the significance of the differences in polymer type across classes, use `lr_test(aux_pt)`.
The results indicate that there are significant differences in polymer types across classes, $\Delta LL(3) = 17.14, p < .001$.
The results can be reported in probability scale using `table_prob(aux_pt)`.
To test differences for specific polymer types, we can use Wald tests:

```{r echo = TRUE, eval = FALSE}
wald_test(aux_pt, "class1.Thresholds[1,1] = class2.Thresholds[1,1];
          class1.Thresholds[1,2] = class2.Thresholds[1,2];
          class1.Thresholds[1,3] = class2.Thresholds[1,3]")
```

The results indicate that there is no significant difference in the prevalence of "Other" polymer types across classes.
However, PE is significantly more prevalent in class 1, and PP is significantly more prevalent in class 2.

<!-- \newpage -->

<!-- # References -->

<!-- ::: {#refs custom-style="Bibliography"} -->

<!-- ::: -->

## R session

```{r echo = TRUE, eval=TRUE}
sessionInfo()
```

### 2023-01-23 appendix_b.tex
edits by Mauricio

--- 2023-01-23 appendix_a.tex
@@
 \hypersetup{
-  pdftitle={Appendix A: Confirmatory LPA for the Caregiver Compass},
+  pdftitle={Appendix B: Exploratory LPA for Ocean Microplastics},
   pdflang={en-EN},
@@
 
-\title{Appendix A: Confirmatory LPA for the Caregiver Compass}
+\title{Appendix B: Exploratory LPA for Ocean Microplastics}
 \author{\phantom{0}}
@@
 
-This is an example of confirmatory LPA using \texttt{tidySEM}.
-The simulated data are based on work by Zegwaard and colleagues, who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
-Qualitative research among experts resulted in a theory postulating the existence of four types of caregivers (translated from the original Dutch):
-
-\textbf{Balanced}
-
-\begin{quote}
-The balanced caregiver experiences relative balance between the costs and benefits of caring for a close other.
-\end{quote}
-
-\textbf{Imbalanced}
-
-\begin{quote}
-The imbalanced caregiver experiences a precarious balance between the costs and benefits of caring for a close other.
-\end{quote}
-
-\textbf{Lonely}
-
-\begin{quote}
-The lonely caregiver experiences a strong sense of isolation.
-\end{quote}
-
-\textbf{Entrapped}
-
-\begin{quote}
-The entrapped caregiver strongly feels a sense of being entangled in responsibilities which are difficult to fulfill.
-\end{quote}
-
-To view the data documentation, run the command \texttt{?tidySEM::zegwaard\_carecompass} in the R console.
+This is an example of exploratory latent class analysis (LCA) with continuous indicators, otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using \texttt{tidySEM}.
+The present example uses data collected by Alkema as part of a study on ocean microplastics.
+The purpose of this study was to provide a more nuanced model for the distribution of different sizes of ocean microplastics than the commonly used normal distribution.
+To this end, a mixture of normals was used.
+Since there is no theoretical reason to expect a certain number of classes, this is an exploratory LCA.
+To view its documentation, run the command \texttt{?tidySEM::alkema\_microplastics} in the R console.
+The original analyses are available at \url{https://github.com/cjvanlissa/lise_microplastics}; in this vignette, we take a different approach to the analysis to showcase other possibilities.
 
@@
 For convenience, we assign the variables used for analysis to an object called \texttt{df}.
-We first only use the four scales: \texttt{c("burdened",\ "trapped",\ "negaffect",\ "loneliness")}.
-`.
+As explained in the paper, the classes are quite different for lines, films, and fragments.
+For this reason, we here only use data from fragments.
+The indicators are fragments' length and width.
 
@@
 \CommentTok{\# Load data}
-\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ zegwaard\_carecompass[, }\FunctionTok{c}\NormalTok{(}\StringTok{"burdened"}\NormalTok{, }\StringTok{"trapped"}\NormalTok{, }\StringTok{"negaffect"}\NormalTok{, }\StringTok{"loneliness"}\NormalTok{)]}
+\NormalTok{df\_analyze }\OtherTok{\textless{}{-}}\NormalTok{ alkema\_microplastics[alkema\_microplastics}\SpecialCharTok{$}\NormalTok{category }\SpecialCharTok{==} \StringTok{"Fragment"}\NormalTok{, ]}
+\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ df\_analyze[ ,}\FunctionTok{c}\NormalTok{(}\StringTok{"length"}\NormalTok{, }\StringTok{"width"}\NormalTok{)]}
 \end{Highlighting}
@@
 We use \texttt{tidySEM::descriptives()} to describe the data numerically.
-Because all scales are continuous, we select only columns for continuous data to de-clutter the table:
+Because all items are categorical, we remove columns for continuous data to de-clutter the table:
 
@@
 \NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ tidySEM}\SpecialCharTok{::}\FunctionTok{descriptives}\NormalTok{(df)}
-\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{,}
-                 \StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{,}
-                 \StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
-\NormalTok{desc}
+\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"type"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{, }
+\StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{, }\StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
+\NormalTok{papaja}\SpecialCharTok{::}\FunctionTok{apa\_table}\NormalTok{(desc, }\AttributeTok{caption =} \StringTok{"Descriptive statistics"}\NormalTok{)}
 \end{Highlighting}
@@
 
-\begin{tabular}{lllllllllll}
+\begin{tabular}{llllllllllll}
 \toprule
-name & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{missing} & \multicolumn{1}{c}{unique} & \multicolumn{1}{c}{mean} & \multicolumn{1}{c}{median} & \multicolumn{1}{c}{sd} & \multicolumn{1}{c}{min} & \multicolumn{1}{c}{max} & \multicolumn{1}{c}{skew\_2se} & \multicolumn{1}{c}{kurt\_2se}\\
+name & \multicolumn{1}{c}{type} & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{missing} & \multicolumn{1}{c}{unique} & \multicolumn{1}{c}{mean} & \multicolumn{1}{c}{median} & \multicolumn{1}{c}{sd} & \multicolumn{1}{c}{min} & \multicolumn{1}{c}{max} & \multicolumn{1}{c}{skew\_2se} & \multicolumn{1}{c}{kurt\_2se}\\
 \midrule
-burdened & 509 & 0.01 & 509 & 3.38 & 3.39 & 0.75 & 1.20 & 5.28 & 0.17 & 6.51\\
-trapped & 505 & 0.02 & 505 & 1.73 & 1.79 & 0.90 & -0.86 & 3.81 & -1.03 & 5.39\\
-negaffect & 506 & 0.01 & 506 & 2.50 & 2.52 & 0.69 & 0.71 & 4.98 & 0.08 & 6.52\\
-loneliness & 510 & 0.01 & 510 & 2.66 & 2.69 & 0.62 & 0.98 & 4.22 & -0.33 & 6.31\\
+length & numeric & 5605 & 0.00 & 2086 & 2.94 & 2.37 & 1.89 & 1.00 & 69.16 & 137.38 & 2,116.43\\
+width & numeric & 5605 & 0.00 & 2079 & 2.00 & 1.62 & 1.11 & 0.20 & 6.76 & 22.23 & 37.08\\
 \bottomrule
@@
 
-The table indicates two potential causes for concern: there is a small percentage of missingness, and all variables have relatively high kurtosis.
-Since there are some missing values, we can conduct an MCAR test using \texttt{mice::mcar(df)}.
-According to Hawkins' test, there is no evidence to reject the assumptions of multivariate normality and MCAR, \(\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71\).
-Missing data will be accounted for using FIML.
-
 Additionally, we can plot the data.
-The \texttt{ggplot2} function \texttt{geom\_density()} is useful for continuous data.
-Visual inspection confirms the conclusions from the \texttt{descriptives()} table: the data are kurtotic (peaked).
+The \texttt{ggplot2} function \texttt{geom\_density()} is useful to visualize continuous data:
 
@@
 
-\includegraphics{appendices/plot_lpa_desc.pdf}
+\includegraphics{appendices/plot_gmm_desc.pdf}
+
+The data are correctly coded as \texttt{numeric}.
+There are no missing values; if any variables had missing values, we would report an MCAR test with \texttt{mice::mcar()}, and explain that missing data are accounted for using FIML.
+Both the table above and the density plot indicate that the data are extremely right-skewed and kurtotic.
+With this in mind, it can be useful to transform and rescale the data.
+We will use a log transformation.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{df\_plot}\SpecialCharTok{$}\NormalTok{Value }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(df\_plot}\SpecialCharTok{$}\NormalTok{Value)}
+\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value)) }\SpecialCharTok{+}
+  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
+  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{Variable)}\SpecialCharTok{+}
+  \FunctionTok{theme\_bw}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{appendices/plot_gmm_desc_log.pdf}
+
+The log transformation addresses the aforementioned concerns regarding skew and kurtosis.
+To confirm this, reshape the data to wide format and examine a scatterplot:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{df }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(df\_plot, }\AttributeTok{direction =} \StringTok{"wide"}\NormalTok{, }\AttributeTok{v.names =} \StringTok{"Value"}\NormalTok{)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\FunctionTok{names}\NormalTok{(df) }\OtherTok{\textless{}{-}} \FunctionTok{gsub}\NormalTok{(}\StringTok{"Value."}\NormalTok{, }\StringTok{""}\NormalTok{, }\FunctionTok{names}\NormalTok{(df), }\AttributeTok{fixed =} \ConstantTok{TRUE}\NormalTok{)}
+\FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ length, }\AttributeTok{y =}\NormalTok{ width)) }\SpecialCharTok{+}
+  \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{alpha =}\NormalTok{ .}\DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
+  \FunctionTok{theme\_bw}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\includegraphics{appendices/plot_gmm_scatter.pdf}
 
@@
 All variables in \texttt{data} are included in the analysis, which is why we first selected the indicator variables.
-As this is a confirmatory LCA, we do not follow a strictly data-driven class enumeration procedure.
-We will set the maximum number of classes \(K\) to one more than the theoretically expected number.
+
+As this is an exploratory LCA, we will conduct a rather extensive search across model specifications and number of classes.
+We will set the maximum number of classes \(K\) to four; depending on the results, we can always choose to increase it later.
 We set a seed to ensure replicable results.
+
+As the analysis takes a long time to compute, it is prudent to save the results to disk immediately, so as not to lose them.
+For this, we use the function \texttt{saveRDS()}.
+We can later use \texttt{res\ \textless{}-\ readRDS("res\_gmm.RData")} to load the analysis from the file.
 
@@
 \NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df,}
-                   \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{)}
+                   \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{,}
+                   \AttributeTok{variances =} \FunctionTok{c}\NormalTok{(}\StringTok{"equal"}\NormalTok{, }\StringTok{"varying"}\NormalTok{),}
+                   \AttributeTok{covariances =} \FunctionTok{c}\NormalTok{(}\StringTok{"zero"}\NormalTok{, }\StringTok{"equal"}\NormalTok{,}
+                                   \StringTok{"varying"}\NormalTok{),}
+                   \AttributeTok{expand\_grid =} \ConstantTok{TRUE}\NormalTok{)}
+\FunctionTok{saveRDS}\NormalTok{(res, }\StringTok{"res\_gmm.RData"}\NormalTok{)}
 \end{Highlighting}
 \end{Shaded}
-
-This analysis should produce some warnings about cluster initialization.
-These relate to the selection of starting values, which relies on the K-means algorithm and is not robust to missing data.
-The algorithm automatically switches to hierarchical clustering, no further action is required.
 
@@
 
-To compare the fit of the theoretical model against other models, we create a model fit table using \texttt{table\_fit()} and retain relevant columns.
-We also determine whether any models can be disqualified.
+To compare the fit of the estimated models, we create a model fit table using \texttt{table\_fit()}.
 
-In this example, all models converge without issues.
-If, for example, the two-class solution had not converged, we could use the function \texttt{res{[}{[}2{]}{]}\ \textless{}-\ mxTryHard(res{[}{[}2{]}{]})} to aid convergence.
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{table\_fit}\NormalTok{(res)}
+\end{Highlighting}
+\end{Shaded}
 
+First, we determine whether any models can be disqualified.
+There were no indications of convergence problems during estimation, so this is not a reason to disqualify solutions.
 Next, we check for local identifiability.
-The sample size is consistently reported as 513, which means that partially missing cases were indeed included via FIML.
-The smallest class size occurs in the 5-class model, where the smallest class is assigned 7\% of cases, or 38 cases.
-This model has 28 parameters, approximately 6 per class.
-We thus have at least five observations per parameter in every class, and do not disqualify the 5-class model.
+The sample size is \texttt{5605}.
+We can calculate the ratio of observations to parameters and append it to the fit table:
 
-There are concerns about theoretical interpretability of all solutions, as the entropies and minimum classification probabilities are all low.
-However, in this confirmatory use case, we address this when interpreting the results.
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{fit}\SpecialCharTok{$}\NormalTok{par\_ratio }\OtherTok{\textless{}{-}}\NormalTok{ (}\DecValTok{5605}\SpecialCharTok{*}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{n\_min) }\SpecialCharTok{/}\NormalTok{ (fit}\SpecialCharTok{$}\NormalTok{Parameters}\SpecialCharTok{/}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{Classes)}
+\NormalTok{fit[ , }\FunctionTok{c}\NormalTok{(}\StringTok{"Name"}\NormalTok{, }\StringTok{"LL"}\NormalTok{, }\StringTok{"Parameters"}\NormalTok{, }\StringTok{"par\_ratio"}\NormalTok{,}
+         \StringTok{"BIC"}\NormalTok{, }\StringTok{"Entropy"}\NormalTok{,}
+         \StringTok{"prob\_min"}\NormalTok{, }\StringTok{"prob\_max"}\NormalTok{, }
+         \StringTok{"n\_min"}\NormalTok{, }\StringTok{"n\_max"}\NormalTok{,}
+         \StringTok{"lmr\_p"}\NormalTok{)]}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\caption{\label{tab:tabfit}Model fit table}
+\caption{\label{tab:tabfit}Model fit table. Some columns removed to fit page.}
 
-\begin{tabular}{lllllllllll}
+\begin{tabular}{lllllllll}
 \toprule
-Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{p} & \multicolumn{1}{c}{n} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Entropy} & \multicolumn{1}{c}{p\_min} & \multicolumn{1}{c}{p\_max} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{n\_max} & \multicolumn{1}{c}{lmr\_p}\\
+Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{p} & \multicolumn{1}{c}{par\_ratio} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Ent.} & \multicolumn{1}{c}{p\_min} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{lmr\_p}\\
 \midrule
-equal var 1 & -2,241.98 & 8 & 513 & 4,533.89 & 1.00 & 1.00 & 1.00 & 1.00 & 1.00 & NA\\
-equal var 2 & -2,031.37 & 13 & 513 & 4,143.85 & 0.74 & 0.91 & 0.93 & 0.42 & 0.58 & 0.00\\
-equal var 3 & -1,951.35 & 18 & 513 & 4,015.02 & 0.78 & 0.89 & 0.91 & 0.19 & 0.54 & 0.00\\
-equal var 4 & -1,916.25 & 23 & 513 & 3,976.03 & 0.75 & 0.81 & 0.92 & 0.16 & 0.34 & 0.00\\
-equal var 5 & -1,912.93 & 28 & 513 & 4,000.59 & 0.70 & 0.39 & 0.92 & 0.07 & 0.31 & 0.28\\
+equal var 1 & -8,107.31 & 4 & 1,401.25 & 16,249.14 & 1.00 & 1.00 & 1.00 & NA\\
+equal var 2 & -5,211.25 & 7 & 564.00 & 10,482.91 & 0.87 & 0.94 & 0.35 & 0.00\\
+equal var 3 & -4,138.27 & 10 & 341.70 & 8,362.85 & 0.83 & 0.88 & 0.20 & 0.00\\
+equal var 4 & -3,500.42 & 13 & 246.15 & 7,113.04 & 0.83 & 0.89 & 0.14 & 0.00\\
+free var 1 & -8,107.31 & 4 & 1,401.25 & 16,249.14 & 1.00 & 1.00 & 1.00 & NA\\
+free var 2 & -5,137.90 & 9 & 501.33 & 10,353.49 & 0.85 & 0.94 & 0.40 & 0.00\\
+free var 3 & -4,005.10 & 14 & 374.36 & 8,131.04 & 0.83 & 0.89 & 0.31 & 0.00\\
+free var 4 & -3,330.87 & 19 & 245.05 & 6,825.74 & 0.84 & 0.88 & 0.21 & 0.00\\
+equal var, equal cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
+equal var, equal cov 2 & -3,082.31 & 8 & 432.25 & 6,233.66 & 0.72 & 0.86 & 0.31 & 0.00\\
+equal var, equal cov 3 & -3,030.10 & 11 & 256.36 & 6,155.14 & 0.67 & 0.71 & 0.17 & 0.00\\
+equal var, equal cov 4 & -3,020.67 & 14 & 220.29 & 6,162.19 & 0.61 & 0.68 & 0.14 & 0.00\\
+free var, equal cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
+free var, equal cov 2 & -2,544.74 & 10 & 97.00 & 5,175.79 & 0.64 & 0.51 & 0.09 & 0.00\\
+free var, equal cov 3 & -2,256.95 & 15 & 71.80 & 4,643.36 & 0.68 & 0.54 & 0.06 & 0.00\\
+free var, equal cov 4 & -2,068.70 & 20 & 26.00 & 4,310.02 & 0.63 & 0.52 & 0.02 & 0.00\\
+equal var, free cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
+equal var, free cov 2 & -2,551.86 & 9 & 107.78 & 5,181.40 & 0.65 & 0.52 & 0.09 & 0.00\\
+equal var, free cov 3 & -2,359.39 & 13 & 84.23 & 4,830.99 & 0.68 & 0.56 & 0.07 & 0.00\\
+equal var, free cov 4 & -2,173.85 & 17 & 27.53 & 4,494.44 & 0.63 & 0.60 & 0.02 & 0.00\\
+free var, free cov 1 & -3,388.56 & 5 & 1,121.00 & 6,820.28 & 1.00 & 1.00 & 1.00 & NA\\
+free var, free cov 2 & -2,574.87 & 11 & 406.91 & 5,244.69 & 0.56 & 0.81 & 0.40 & 0.00\\
+free var, free cov 3 & -2,111.38 & 17 & 36.00 & 4,369.50 & 0.65 & 0.51 & 0.04 & 0.00\\
+free var, free cov 4 & -2,024.10 & 23 & 15.48 & 4,246.73 & 0.62 & 0.50 & 0.02 & 0.00\\
 \bottomrule
@@
 
-Class enumeration unambiguously indicates a preference for the 4-class solution: the 4-class solution has the lowest BIC, which means it is preferred over all other solutions including a 1-class solution and a solution with more classes.
-The p-values of the LMR LR test are significant for all pairwise model comparisons, except for the 5-class model.
-Note that a scree plot for the BIC can be plotted by calling \texttt{plot(fit)}.
-Following the elbow criterion, a three-class solution would also be defensible.
-The function \texttt{ic\_weights(tab\_compare)} allows us to compute IC weights; it indicates that, conditional on the set of models, the 4-class model has a posterior model probability of nearly 100\%.
+As can be seen from the fit table, the lowest ratio of observations to parameters is 18, which is no cause for concern.
 
-\hypertarget{optional-alternative-model-specifications}{%
-\subsection{Optional: Alternative Model Specifications}\label{optional-alternative-model-specifications}}
+However, note that we have a very large sample, and for many models, the smallest class comprises only a very small percentage of the total sample.
+Since the purpose of this analysis is to better represent the distribution of ocean microplastics, we can wonder whether it makes sense to allow for classes that only describe a small percentage of the cases.
+We therefore set a lower bound for class size and only consider solutions that capture at least 10\% of the sample.
 
-In the case of confirmatory LCA, the theory would be refuted by strong evidence against the hypothesized model and number of classes.
-In the preceding, we only compared the theoretical model against models with different number of classes.
-Imagine, however, that a Reviewer argues that variance ought to be freely estimated across classes.
-We could compare our theoretical model against their competing model as follows.
-Note that we can put two models into a list to compare them.
+An interesting characteristic of this data is that the BIC and the entropy are strongly correlated.
+The raw correlation between these two metrics is .66, \texttt{cor(fit\$BIC,\ fit\$Entropy)}.
+If we omit the 1-class models, for which entropy is technically not defined, the correlation is even as high as .85, \texttt{cor(fit\$BIC{[}!fit\$Classes\ ==\ 1{]},\ fit\$Entropy{[}!fit\$Classes\ ==\ 1{]})}.
+
+This strong correlation indicates that an increase in fit comes with a decrease in class separability.
+This illustrates why entropy should not be treated as a model fit criterion.
+It also illustrates that criteria for class enumeration should be explicit, because we will likely come to a different decision depending on which criteria are used.
+
+As mentioned before, we drop models with \textless{} 10\% of cases in the smallest class:
 
@@
 \begin{Highlighting}[]
-\NormalTok{res\_alt }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(df, }\AttributeTok{classes =} \DecValTok{4}\NormalTok{, }\AttributeTok{variances =} \StringTok{"varying"}\NormalTok{)}
-\NormalTok{compare }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(res[[}\DecValTok{4}\NormalTok{]], res\_alt)}
-\FunctionTok{table\_fit}\NormalTok{(compare)}
+\NormalTok{fit }\OtherTok{\textless{}{-}}\NormalTok{ fit[}\SpecialCharTok{!}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{n\_min }\SpecialCharTok{\textless{}}\NormalTok{ .}\DecValTok{1}\NormalTok{, ]}
+\end{Highlighting}
+\end{Shaded}
+
+If our strategy is to optimize fit, we can examine the fit table above, or plot a scree plot for the BIC by calling \texttt{plot(fit)}.
+Note that, due to the large sample size, all ICs give identical conclusions.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{plot}\NormalTok{(fit) }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \DecValTok{90}\NormalTok{, }\AttributeTok{vjust =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{hjust=}\DecValTok{1}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{appendices/gmm_plotfit} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-16}
+\end{figure}
+
+Looking at the blocks of 1-4 class models for each model specification, it appears that the BIC keeps decreasing with the addition of more classes.
+Across the blocks, the BIC keeps decreasing with increasingly complex model specifications.
+Similarly, all LMR tests are significant, indicating a preference for more complex models.
+
+Out of the 16 models that remain after removing those with \textless{} 10\% of cases in the smallest class, one model stands out: The 2-class model with free (co)variances.
+We thus select this as our final model.
+
+\hypertarget{interpreting-the-final-class-solution}{%
+\subsection{Interpreting the Final Class Solution}\label{interpreting-the-final-class-solution}}
+
+and extract the classification probabilities and model results.
+We here request the estimates (\texttt{est}) and standardized estimates \texttt{std\_est}, because the latter allows us to interpret the correlations between length and width.
+Note that there is little sense in requesting the standard errors and p-values: With a sample size of 5606, every parameter is significantly different from zero.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{res\_bic }\OtherTok{\textless{}{-}}\NormalTok{ res[[}\StringTok{"free var, free cov 2"}\NormalTok{]]}
+\NormalTok{cp }\OtherTok{\textless{}{-}} \FunctionTok{class\_prob}\NormalTok{(res\_bic)}
+\NormalTok{results }\OtherTok{\textless{}{-}} \FunctionTok{table\_results}\NormalTok{(res\_bic, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"std\_est"}\NormalTok{))}
+\NormalTok{results}
 \end{Highlighting}
@@
 
-\caption{\label{tab:tabfitcomp}Comparing competing theoretical models}
+\caption{\label{tab:unnamed-chunk-19}Results of a 2-class model with free (co)variances}
 
-\begin{tabular}{lllllllll}
+\begin{tabular}{lll}
 \toprule
-Name & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{Parameters} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Entropy} & \multicolumn{1}{c}{prob\_min} & \multicolumn{1}{c}{prob\_max} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{n\_max}\\
+label & \multicolumn{1}{c}{est} & \multicolumn{1}{c}{std\_est}\\
 \midrule
-1 & -1,916.25 & 23 & 3,976.03 & 0.75 & 0.81 & 0.92 & 0.16 & 0.34\\
-2 & -1,909.23 & 35 & 4,036.87 & 0.78 & 0.84 & 0.92 & 0.16 & 0.32\\
+Means.length & 0.65 & 2.04\\
+Means.width & 0.34 & 1.07\\
+Variances.length & 0.10 & 1.00\\
+Covariances.length.WITH.width & 0.09 & 0.92\\
+Variances.width & 0.10 & 1.00\\
+Means.length & 1.33 & 3.03\\
+Means.width & 0.86 & 1.64\\
+Variances.length & 0.19 & 1.00\\
+Covariances.length.WITH.width & 0.20 & 0.85\\
+Variances.width & 0.28 & 1.00\\
+mix2.weights[1,2] & 0.75 & NA\\
 \bottomrule
@@
 
-The alternative model incurs 12 additional parameters for the free variances.
-Yet, it has a higher BIC, which indicates that this additional complexity does not outweigh the increase in fit.
-
-\hypertarget{interpreting-the-final-class-solution}{%
-\subsection{Interpreting the Final Class Solution}\label{interpreting-the-final-class-solution}}
-
-To interpret the final class solution, we first reorder the 4-class model by class size.
-This helps prevent label switching.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{res\_final }\OtherTok{\textless{}{-}} \FunctionTok{mx\_switch\_labels}\NormalTok{(res[[}\DecValTok{4}\NormalTok{]])}
-\end{Highlighting}
-\end{Shaded}
-
-The 4-class model yielded classes of reasonable size; the largest class comprised 33\%, and the smallest comprised 16\% of cases.
-However, the entropy was low, \(S = .75\), indicating poor class separability.
-Furthermore, the posterior classification probability ranged from \([.81, .92]\), which means that at least some classes had a high classification error.
-We produce a table of the results below.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{table\_results}\NormalTok{(res\_final, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"confint"}\NormalTok{, }\StringTok{"class"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{table}[tbp]
-
-\begin{center}
-\begin{threeparttable}
-
-\caption{\label{tab:unnamed-chunk-13}Four-class model results}
-
-\begin{tabular}{lllll}
-\toprule
-label & \multicolumn{1}{c}{est} & \multicolumn{1}{c}{se} & \multicolumn{1}{c}{confint} & \multicolumn{1}{c}{class}\\
-\midrule
-mix4.weights[1,2] & 0.86 & 0.15 & {}[0.56, 1.15] & NA\\
-mix4.weights[1,3] & 0.66 & 0.11 & {}[0.44, 0.88] & NA\\
-mix4.weights[1,4] & 0.47 & 0.08 & {}[0.32, 0.63] & NA\\
-Variances.burdened & 0.23 & 0.02 & {}[0.19, 0.27] & class1\\
-Variances.trapped & 0.17 & 0.02 & {}[0.14, 0.20] & class1\\
-Variances.negaffect & 0.31 & 0.02 & {}[0.27, 0.36] & class1\\
-Variances.loneliness & 0.24 & 0.02 & {}[0.20, 0.28] & class1\\
-Means.burdened & 3.27 & 0.04 & {}[3.18, 3.36] & class1\\
-Means.trapped & 1.28 & 0.05 & {}[1.18, 1.38] & class1\\
-Means.negaffect & 2.31 & 0.06 & {}[2.20, 2.42] & class1\\
-Means.loneliness & 2.73 & 0.04 & {}[2.64, 2.82] & class1\\
-Means.burdened & 3.40 & 0.06 & {}[3.28, 3.52] & class2\\
-Means.trapped & 2.27 & 0.06 & {}[2.15, 2.38] & class2\\
-Means.negaffect & 2.81 & 0.06 & {}[2.70, 2.93] & class2\\
-Means.loneliness & 2.79 & 0.06 & {}[2.66, 2.91] & class2\\
-Means.burdened & 4.25 & 0.07 & {}[4.12, 4.38] & class3\\
-Means.trapped & 2.67 & 0.05 & {}[2.58, 2.77] & class3\\
-Means.negaffect & 2.92 & 0.06 & {}[2.80, 3.03] & class3\\
-Means.loneliness & 2.01 & 0.06 & {}[1.89, 2.14] & class3\\
-Means.burdened & 2.38 & 0.06 & {}[2.26, 2.50] & class4\\
-Means.trapped & 0.38 & 0.05 & {}[0.28, 0.49] & class4\\
-Means.negaffect & 1.78 & 0.07 & {}[1.65, 1.91] & class4\\
-Means.loneliness & 3.18 & 0.06 & {}[3.07, 3.30] & class4\\
-\bottomrule
-\end{tabular}
-
-\end{threeparttable}
-\end{center}
-
-\end{table}
-
-The results are best interpreted by examining a plot of the model and data, however.
+Interpreting the results is facilitated by examining a plot of the model and data.
 Relevant plot functions are \texttt{plot\_bivariate()}, \texttt{plot\_density()}, and \texttt{plot\_profiles()}.
@@
 \begin{Highlighting}[]
-\FunctionTok{plot\_bivariate}\NormalTok{(res\_final)}
+\FunctionTok{plot\_bivariate}\NormalTok{(res\_bic)}
 \end{Highlighting}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{appendices/lpa_bivariate} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-16}
+\includegraphics[width=1\linewidth]{appendices/gmm_bivariate_bic} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-22}
 \end{figure}
@@
 On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.
-As this model has zero covariances, all circles are round (albeit warped by the different scales of the X and Y axes)
 
-The marginal density plots show that trappedness distinguishes classes rather well.
-For all other indicators, groups are not always clearly separated in terms of marginal density: class 2 and 3 coalesce on negative affect, 1 and 2 coalesce on loneliness, and 1 and 2 coalesce on burden.
-Nevertheless, the off-diagonal scatterplots show reasonable bivariate separation for all classes.
+The bivariate and marginal plots show that the classes are not clearly separable, as also evident from the low entropy.
+At the same time however, it is clear that the distributions are non-normal, and the second class accounts for some of this non-normality (there is a smaller `bump' to the right of the mode, which could be the mean of a second normal distribution).
+The first class (57\%) accounts for smaller fragments, and the second class (43\%) accounts for some of the right-skew in fragments' length and width.
+We can simply label class 1 as \emph{small fragments}, and class 2 as \emph{larger fragments}.
 
-We can obtain a more classic profile plot using \texttt{plot\_profiles(res\_final)}.
-This plot conveys less information than the bivariate plot, but is readily interpretable.
-Below is a comparison between the most common type of visualization for LPA, and the best-practices visualization provided by \texttt{tidySEM}.
-Note that the best practices plot includes class means and error bars, standard deviations, and a ribbon plot of raw data weighted by class probability to indicate how well the classes describe the observed distribution.
-The overlap between the classes is clearly visible in this figure; this is why the entropy and classification probabilities are relatively low.
+It also appears that the correlation between length and width is stronger for small fragments than for large fragments.
+To test the difference, use \texttt{wald\_test(res\_bic,\ hypothesis\ =\ "c11\ =\ c21")}.
+Results indicate that the correlation is indeed significantly larger for small fragments (\(r = .92\)) than for larger fragments (\(r = .85\)), \(\chi^2(1) = 11.56, p < .001\).
+Thus, small fragments are more coextensive than large fragments.
 
-Based on the bivariate plot, we can label class 1 as the \emph{balanced} type (33\%), class 2 as the \emph{imbalanced} type (29\%), class 3 as the \emph{entrapped} type (22\%), and class 4 as the \emph{lonely} type (16\%).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{plot\_profiles}\NormalTok{(res\_final)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{appendices/lpa_profiles} \caption{Bivariate profile plot}\label{fig:unnamed-chunk-19}
-\end{figure}
+There are, however, concerns about the interpretability of this solutions: the entropy is \texttt{.56} and the minimum classification probability is\texttt{.81}.
+This is because of substantial overlap in the distributions of the two classes.
 
@@
 The \texttt{BCH()} function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
-We consider two examples: a single variable, and an auxiliary model.
-
-\hypertarget{comparing-means-or-proportions-across-classes}{%
-\subsubsection{Comparing Means or Proportions Across Classes}\label{comparing-means-or-proportions-across-classes}}
-
-For a single (continuous or ordinal) variable, we can call the BCH function and simply supply the auxiliary variable to the \texttt{data} argument, omitting the \texttt{model} argument.
-Below, we estimate an auxiliary model to compare the sex of patients between classes:
+For example, we can test whether polymer type differs between the two classes:
 
@@
 \begin{Highlighting}[]
-\NormalTok{aux\_sex }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_final, }\AttributeTok{data =}\NormalTok{ zegwaard\_carecompass}\SpecialCharTok{$}\NormalTok{sexpatient)}
+\NormalTok{df\_pt }\OtherTok{\textless{}{-}} \FunctionTok{mx\_dummies}\NormalTok{(df\_analyze}\SpecialCharTok{$}\NormalTok{poly\_type)}
+\NormalTok{aux\_pt }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_bic, }\AttributeTok{model =} \StringTok{"poly\_typeOther | t1}
+\StringTok{                                poly\_typePE | t1}
+\StringTok{                                poly\_typePP | t1"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ df\_pt)}
+\NormalTok{aux\_pt }\OtherTok{\textless{}{-}} \FunctionTok{mxTryHardOrdinal}\NormalTok{(aux\_pt)}
 \end{Highlighting}
@@
 
-To obtain an omnibus likelihood ratio test of the significance of these sex differences across classes, as well as pairwise comparisons between classes, use \texttt{lr\_test(aux\_sex)}.
-The results indicate that there are significant sex differences across classes, \(\Delta LL(1) = 8.7, p = .003\).
-Pairwise comparisons indicate that class 3 differs significantly from classes 1 and 2.
-The results can be reported in probability scale using \texttt{table\_prob(aux\_sex)}.
-It appears that the entrapped class disproportionately cares for female patients.
-
-\hypertarget{comparing-auxiliary-models-across-classes}{%
-\subsubsection{Comparing Auxiliary Models Across Classes}\label{comparing-auxiliary-models-across-classes}}
-
-Finally, we compare a simple model between classes.
-Specifically, we will examine whether the distance predicts the frequency of visits differently across classes (treated as continuous).
+To obtain an omnibus likelihood ratio test of the significance of the differences in polymer type across classes, use \texttt{lr\_test(aux\_pt)}.
+The results indicate that there are significant differences in polymer types across classes, \(\Delta LL(3) = 17.14, p < .001\).
+The results can be reported in probability scale using \texttt{table\_prob(aux\_pt)}.
+To test differences for specific polymer types, we can use Wald tests:
 
@@
 \begin{Highlighting}[]
-\NormalTok{df\_aux }\OtherTok{\textless{}{-}}\NormalTok{ zegwaard\_carecompass[, }\FunctionTok{c}\NormalTok{(}\StringTok{"freqvisit"}\NormalTok{, }\StringTok{"distance"}\NormalTok{)]}
-\NormalTok{df\_aux}\SpecialCharTok{$}\NormalTok{freqvisit }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(df\_aux}\SpecialCharTok{$}\NormalTok{freqvisit)}
-\NormalTok{aux\_model }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_final, }\AttributeTok{model =} \StringTok{"freqvisit \textasciitilde{} distance"}\NormalTok{,}
-                 \AttributeTok{data =}\NormalTok{ df\_aux)}
+\FunctionTok{wald\_test}\NormalTok{(aux\_pt, }\StringTok{"class1.Thresholds[1,1] = class2.Thresholds[1,1];}
+\StringTok{          class1.Thresholds[1,2] = class2.Thresholds[1,2];}
+\StringTok{          class1.Thresholds[1,3] = class2.Thresholds[1,3]"}\NormalTok{)}
 \end{Highlighting}
@@
 
-To obtain an omnibus likelihood ratio test of the difference in regression coefficients across classes and pairwise comparisons between classes, use \texttt{lr\_test(aux\_model,\ compare\ =\ "A")}.
-The results indicate that there are no significant sex differences across classes, \(\Delta LL(3) = 0.98, p = .81\).
-The results can be reported using \texttt{table\_results(aux\_model)}.
+The results indicate that there is no significant difference in the prevalence of ``Other'' polymer types across classes.
+However, PE is significantly more prevalent in class 1, and PP is significantly more prevalent in class 2.
 

### 2023-01-23 appendix_c.Rmd
edits by Mauricio

---
title             : "Appendix C: Latent Class Growth Analysis"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
bibliography      : "r-references.bib"
floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
library(yaml)
library(scales)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  tidy.opts = list(width.cutoff = 60), tidy = TRUE
)
options(scipen = 1, digits = 2)

run_everything = FALSE

pkgdown <- TRUE

```

This vignette illustrated `tidySEM`'s ability to perform latent class
growth analysis, or growth mixture modeling. The simulated data used for
this example are inspired by work in progress by Plas and colleagues, on
heterogeneity in depression trajectories among Dutch military personnel
who were deployed to Afghanistan. The original data were collected as
part of the *Prospection in Stress-related Military Research (PRISMO)*
study, which examined of psychological problems after deployment in more
than 1,000 Dutch military personnel from 2005-2019.

First, we load all required packages:

```{r echo = TRUE, eval = TRUE}
library(tidySEM)
library(ggplot2)
library(MASS)
```

## Data preprocessing

We first examined the descriptive statistics for the sum score scales:

```{r, results='asis'}
# Get descriptives
df <- plas_depression
desc <- descriptives(df)
desc <- desc[, c("name", "mean", "median", "sd", "min", "max", 
"skew_2se", "kurt_2se")
]
papaja::apa_table(desc, caption = "Item descriptives")
```

Note that all variables were extremely right-skewed due to censoring at
the lower end of the scale.

We can examine these distributions visually as well:

```{r echo = TRUE, eval = FALSE}
df_plot <- reshape(df, direction = "long", varying = names(df))
ggplot(df_plot, aes(x = scl)) +
  geom_density() +
  facet_wrap(~time) + theme_bw()
```

```{r, eval = run_everything, echo = FALSE}
df_plot <- reshape(df, direction = "long", varying = names(df))
p = ggplot(df_plot, aes(x = scl)) +
  geom_density() +
  facet_wrap(~time) + theme_bw()
ggsave("appendices/plot_dist.png", p, width = 150, height = 120, units = "mm")
```

```{r, echo = FALSE, out.width="50%", eval = pkgdown}
knitr::include_graphics("appendices/plot_dist.png")
```

As this type of skew can result in convergence problems in LCGA, we
compared several transformations to reduce skew: The square and cube
root, log, inverse, and Box-Cox transformations.

```{r echo = TRUE, eval = FALSE, results='hide'}
df_scores <- df_plot
# Store original range of SCL
rng_scl <- range(df_scores$scl)
# Log-transform
df_scores$log <- scales::rescale(log(df_scores$scl), to = c(0, 1))
# Square root transform
df_scores$sqrt <- scales::rescale(sqrt(df_scores$scl), to = c(0, 1))
# Cube root transform
df_scores$qrt <- scales::rescale(df_scores$scl^.33, to = c(0, 1))
# Reciprocal transform
df_scores$reciprocal <- scales::rescale(1/df_scores$scl, to = c(0, 1))
# Define function for Box-Cox transformation
bc <- function(x, lambda){
  (((x ^ lambda) - 1) / lambda)
}
# Inverse Box-Cox transformation
invbc <- function(x, lambda){
  ((x*lambda)+1)^(1/lambda)
}
# Box-Cox transform
b <- MASS::boxcox(lm(df_scores$scl ~ 1), plotit = FALSE)
lambda <- b$x[which.max(b$y)]
df_scores$boxcox <- bc(df_scores$scl, lambda)
# Store range of Box-Cox transformed data
rng_bc <- range(df_scores$boxcox)
df_scores$boxcox <- scales::rescale(df_scores$boxcox, to = c(0, 1))
# Rescale SCL
df_scores$scl <- scales::rescale(df_scores$scl, to = c(0, 1))
```

We can plot these transformations:

```{r, eval = FALSE, echo = TRUE}
# Make plot data
df_plot <- do.call(rbind, lapply(c("scl", "log", "sqrt", "qrt", "boxcox"), function(n){
  data.frame(df_scores[c("time", "id")],
             Value = df_scores[[n]],
             Transformation = n)
}))
# Plot
ggplot(df_plot, aes(x = Value, colour = Transformation)) +
  geom_density() +
  facet_wrap(~time) +
  scale_y_sqrt() +
  xlab("scl (rescaled to 0-1)") +
  theme_bw()
```

```{r, eval = run_everything, echo = FALSE}
df_plot <- reshape(df, direction = "long", varying = names(df))
df_scores <- df_plot
# Store original range of SCL
rng_scl <- range(df_scores$scl)
# Log-transform
df_scores$log <- scales::rescale(log(df_scores$scl), to = c(0, 1))
# Square root transform
df_scores$sqrt <- scales::rescale(sqrt(df_scores$scl), to = c(0, 1))
# Cube root transform
df_scores$qrt <- scales::rescale(df_scores$scl^.33, to = c(0, 1))
# Reciprocal transform
df_scores$reciprocal <- scales::rescale(1/df_scores$scl, to = c(0, 1))
# Define function for Box-Cox transformation
bc <- function(x, lambda){
  (((x ^ lambda) - 1) / lambda)
}
# Inverse Box-Cox transformation
invbc <- function(x, lambda){
  ((x*lambda)+1)^(1/lambda)
}
# Box-Cox transform
b <- MASS::boxcox(lm(df_scores$scl ~ 1), plotit = FALSE)
lambda <- b$x[which.max(b$y)]
df_scores$boxcox <- bc(df_scores$scl, lambda)
# Store range of Box-Cox transformed data
rng_bc <- range(df_scores$boxcox)
df_scores$boxcox <- scales::rescale(df_scores$boxcox, to = c(0, 1))
# Rescale SCL
df_scores$scl <- scales::rescale(df_scores$scl, to = c(0, 1))

df_plot <- do.call(rbind, lapply(c("scl", "log", "sqrt", "qrt", "boxcox"), function(n){
  data.frame(df_scores[c("time", "id")],
             Value = df_scores[[n]],
             Transformation = n)
}))
# Plot
p <- ggplot(df_plot, aes(x = Value, colour = Transformation)) +
  geom_density() +
  facet_wrap(~time) +
  scale_y_sqrt() +
  xlab("scl (rescaled to 0-1)") +
  theme_bw()
ggsave("appendices/plot_trans.png", p, width = 150, height = 120, units = "mm")
```

```{r, echo = FALSE, eval = pkgdown, out.width="50%"}
knitr::include_graphics("appendices/plot_trans.png")
```

Evidently, the Box-Cox transformation reduced skew the most.
Consequently, we proceeded with the Box-Cox transformed scores for
analysis.

```{r, echo = TRUE, eval = run_everything}
dat <- df_scores[, c("id", "time", "boxcox")]
dat <- reshape(dat, direction = "wide", v.names = "boxcox", timevar = "time", idvar = "id")
names(dat) <- gsub("boxcox.", "scl", names(dat))
```

# Latent Class Growth Analysis

Next, we estimated a latent class growth model for SCL. The model
included an overall intercept, centered at T1, `i`. To model the
potential effect of deployment on depresion, we also included a dummy
variable that was zero before deployment, and 1 after deployment,
`step`. Finally, to model potential change (or recovery) in depression
post-deployment, we included a linear slope from T2-T6, `s`. All
variances of growth parameters were fixed to zero due to the sparse
nature of the data. In this vignette, we do not consider more than 5
classes, because the analyses are computationally very intensive and the
data were simulated from a 3-class model.

It is important to not that in LCGA, the subgroups will be limited by
the specify growth structure, meaning that LCA will identify distinctive
growth patterns within the intercept, step, and slope growth. For
example, if there is a subgroup that follows a quadratic growth pattern
this models will not be able to identify it.

**NOTE: The time scales in this model are not correct; it currently
assumes that all measurements are equidistant. Feel free to experiment
with adjusting this.**

```{r lcga, echo = TRUE, eval = FALSE}
set.seed(27796)
dat[["id"]] <- NULL
res_step <- mx_growth_mixture(
  model =
  "
  i =~ 1*scl1 + 1*scl2 + 1*scl3 +1*scl4 +1*scl5 +1*scl6
  step =~ 0*scl1 + 1*scl2 + 1*scl3 +1*scl4 +1*scl5 +1*scl6
  s =~ 0*scl1 + 0*scl2 + 1*scl3 +2*scl4 +3*scl5 +4*scl6
  scl1 ~~ vscl1*scl1
  scl2 ~~ vscl2*scl2
  scl3 ~~ vscl3*scl3
  scl4 ~~ vscl4*scl4
  scl5 ~~ vscl5*scl5
  scl6 ~~ vscl6*scl6
  i ~~ 0*i
  step ~~ 0*step
  s ~~ 0*s
  i ~~ 0*s
  i ~~ 0*step
  s ~~ 0*step", classes = 1:5,
  data = dat)
# Additional iterations because of
# convergence problems for model 1:
res_step[[1]] <- mxTryHardWideSearch(res_step[[1]], extraTries = 50)
```

```{r echo = FALSE, eval = run_everything}
set.seed(27796)
dat[["id"]] <- NULL
res_step <- mx_growth_mixture(
  model =
"i =~ 1*scl1 + 1*scl2 + 1*scl3 +1*scl4 +1*scl5 +1*scl6
step =~ 0*scl1 + 1*scl2 + 1*scl3 +1*scl4 +1*scl5 +1*scl6
s =~ 0*scl1 + 0*scl2 + 1*scl3 +2*scl4 +3*scl5 +4*scl6
scl1 ~~ vscl1*scl1
scl2 ~~ vscl2*scl2
scl3 ~~ vscl3*scl3
scl4 ~~ vscl4*scl4
scl5 ~~ vscl5*scl5
scl6 ~~ vscl6*scl6
i ~~ 0*i
step ~~ 0*step
s ~~ 0*s
i ~~ 0*s
i ~~ 0*step
s ~~ 0*step", classes = 1:5,
  data = dat)
# Additional iterations because of
# convergence problems for model 1:
res_step[[1]] <- mxTryHardWideSearch(res_step[[1]], extraTries = 50)
saveRDS(res_step, "appendices/res_step.RData")
fit <- table_fit(res_step)
saveRDS(fit, "appendices/lcga_fit.RData")
```

```{r, echo = FALSE, eval = FALSE}
res_step <- readRDS("appendices/res_step.RData")
```

Note that the first model showed convergence problems, throwing the
error: *The model does not satisfy the first-order optimality conditions
to the required accuracy, and no improved point for the merit function
could be found during the final linesearch.* To address this problem, we
performed additional iterations to\
find a better solution, using `OpenMx::mxTryHardWideSearch()`. This also
illustrates that `tidySEM` mixture models inherit from `OpenMx`'s
`MxModel`, and thus, different `OpenMx` functions can be used to act on
models specified via `tidySEM`.

The fifth model also evidenced convergence problems, but this (as we
will see) is because the solution is overfitted.

## Class enumeration

To determine the correct number of classes, we considered the following
criteria:

1.  We do not consider classes with, on average, fewer than 5
    participants per parameter in a class due to potential local
    underidentification
2.  Lower values for information criteria (AIC, BIC, saBIC) indicate
    better fit
3.  Significant Lo-Mendell-Rubin LRT test indicates better fit for $k$
    vs $k-1$ classes
4.  We do not consider solutions with entropy \< .90 because poor class
    separability compromises interpretability of the results
5.  We do not consider solutions with minimum posterior classification
    probability \< .90 because poor class separability compromises
    interpretability of the results

```{r eval = FALSE, echo = TRUE}
# Get fit table fit
tab_fit <- table_fit(res_step)
# Select columns
tab_fit <- tab_fit[, c("Name", "Classes", "LL", "Parameters", "BIC", "Entropy", "prob_min", "n_min", "warning", "lmr_p")]
tab_fit
```

```{r tabfit, eval = TRUE, echo = FALSE, results = "asis"}
# Get fit table fit
tab_fit <- readRDS("appendices/lcga_fit.RData")
tab_fit <- tab_fit[, c("Name", "Classes", "LL", "Parameters", "BIC", "Entropy", "prob_min", "n_min", "warning", "lmr_p")]
names(tab_fit) <- c("Name", "Classes", "LL", "p", "BIC", "Ent.", "p_min", "n_min", "warn", "lmr_p")
papaja::apa_table(tab_fit, digits = 2, caption = "Fit of LCGA models")
```

According to the Table, increasing the number of classes keeps
increasing model fit according to all ICs except the BIC, which
increased after 3 classes.

The first two LMR tests are significant, indicating that a 2- and
3-class solution were a significant improvement over a 1- and 2-class
solution, respectively. However, solutions with \>3 classes had entropy
and minimum posterior classification probability below the pre-specified
thresholds. Models with \>3 solutions also had fewer than five
observations per parameter. This suggests that the preferred model
should be selected from 1-3 classes.

### Scree plot

A scree plot indicates that the largest decrease in ICs occurs from 1-2
classes, and the inflection point for all ICs is at 3 classes. Moreover,
the BIC increased after 3 classes. A three-class solution thus appears
to be the most parsimonious solution with good fit.

```{r plotscree, echo = TRUE, eval = FALSE}
plot(tab_fit, statistics = c("AIC", "BIC", "saBIC"))
```

```{r echo = FALSE, eval = run_everything}
p <- plot(tab_fit, statistics = c("AIC", "BIC", "saBIC"))
ggsave("appendices/lcga_plot_fit.png", p, width = 150, height = 120, units = "mm")
```

```{r echo = FALSE, eval = pkgdown, out.width="50%"}
knitr::include_graphics("appendices/lcga_plot_fit.png")
```

Based on the aforementioned criteria, we selected a 3-class model for
further analyses. First, to prevent label switching, we re-order these
classes by the value of the intercept `i`. Then, we report the estimated
parameters.

```{r echo = TRUE, eval = FALSE}
res_final <- mx_switch_labels(res_step[[3]], param = "M[1,7]", decreasing = FALSE)
tab_res <- table_results(res_final, columns = NULL)
# Select rows and columns
tab_res <- tab_res[tab_res$Category %in% c("Means", "Variances"), c("Category", "lhs", "est", "se", "pval", "confint", "name")]
tab_res
```

```{r echo = FALSE, eval = run_everything, results='asis'}
res_final <- mx_switch_labels(res_step[[3]], param = "M[1,7]", decreasing = FALSE)
tab_res <- table_results(res_final, columns = NULL)
write.csv(tab_res, "lcga_tab_res.csv", row.names = FALSE)
```

```{r tabres, echo=FALSE, eval = TRUE, results='asis'}
tab_res <- read.csv("appendices/lcga_tab_res.csv", stringsAsFactors = FALSE)
class(tab_res) <- c("tidy_results", "data.frame")
tab_res <- tab_res[tab_res$Category %in% c("Means", "Variances"), c("Category", "lhs", "est", "se", "pval", "confint", "name")]
papaja::apa_table(tab_res, digits = 2, caption = "Results from 3-class LCGA model", longtable = TRUE)
```

As evident from these results, Class 1 started at a relatively lower
level of depressive symptoms, experienced a decrease after deployment,
followed by increase over time. Class 2 started at a moderate level of
depressive symptoms, experienced an increase after deployment, followed
by significant increase over time from T2-T6. Class 3 started at a
relatively higher level, experienced an increase after deployment,
followed by stability.

## Wald tests

To test whether parameters are significantly different between classes,
we can use Wald tests. Wald tests can be specified for all parameters in
the model, using the hypothesis syntax from the `bain` package for
informative hypothesis testing.

To identify the names of parameters in the model, we can use the `name`
column of the results table above. Alternatively, to see all parameters
in the model, run:

```{r params, echo = TRUE, eval = FALSE}
names(coef(res_final))
```

```{r echo = FALSE, eval = TRUE}
c("mix3.weights[1,2]", "mix3.weights[1,3]", "vscl1", "vscl2", 
"vscl3", "vscl4", "vscl5", "vscl6", "class1.M[1,7]", "class1.M[1,8]", 
"class1.M[1,9]", "class2.M[1,7]", "class2.M[1,8]", "class2.M[1,9]", 
"class3.M[1,7]", "class3.M[1,8]", "class3.M[1,9]")
```

Next, specify equality constrained hypotheses. For example, a hypothesis
that states that the mean intercept is equal across groups is specified
as follows:

`"class1.M[1,7] = class2.M[1,7] & class1.M[1,7] = class3.M[1,7]`

It is also possible to consider comparisons between two classes at a
time. When conducting many significance tests, consider correcting for
multiple comparisons however.

```{r echo = TRUE, eval = FALSE}
wald_tests <- wald_test(res_final, 
                   "
                   class1.M[1,7] = class2.M[1,7]&
                   class1.M[1,7] = class3.M[1,7];
                   class1.M[1,8] = class2.M[1,8]&
                   class1.M[1,8] = class3.M[1,8];
                   class1.M[1,9] = class2.M[1,9]&
                   class1.M[1,9] = class3.M[1,9]")
# Rename the hypothesis
wald_tests$Hypothesis <- c("Mean i", "Mean step", "Mean slope")
papaja::apa_table(wald_tests, digits = 2, caption = "Wald tests")
```

```{r echo = FALSE, eval = run_everything}
wald_tests <- wald_test(res_final, 
                   "class1.M[1,7] = class2.M[1,7]&class1.M[1,7] = class3.M[1,7];class1.M[1,8] = class2.M[1,8]&class1.M[1,8] = class3.M[1,8];class1.M[1,9] = class2.M[1,9]&class1.M[1,9] = class3.M[1,9]")
# Rename the hypothesis
wald_tests$Hypothesis <- c("Mean i", "Mean step", "Mean slope")
write.csv(wald_tests, "lcga_wald_tests.csv", row.names = FALSE)
```

```{r waldtests, echo = FALSE, eval = TRUE, results='asis'}
wald_tests <- read.csv("appendices/lcga_wald_tests.csv", stringsAsFactors = FALSE)
class(wald_tests) <- c("wald_test", "data.frame")
papaja::apa_table(wald_tests, digits = 2, caption = "Wald tests")
```

All Wald tests are significant, indicating that there are significant
differences between the intercepts, step function, and slopes of the
three classes.

## Trajectory plot

Finally, we can plot the growth trajectories. This can help interpret
the results better, as well as the residual heterogeneity around class
trajectories.

```{r makelcgaplot, echo = TRUE, eval = FALSE}
p <- plot_growth(res_step[[3]], rawdata = TRUE, alpha_range = c(0, .05))
# Add Y-axis breaks in original scale
brks <- seq(0, 1, length.out = 5)
labs <- round(invbc(scales::rescale(brks, from = c(0, 1), to = rng_bc), lambda))
p <- p + scale_y_continuous(breaks = seq(0, 1, length.out = 5), labels = labs) + ylab("SCL (rescaled from Box-Cox)")
p
```

```{r echo = FALSE, eval = run_everything}
p <- plot_growth(res_step[[3]], rawdata = TRUE, alpha_range = c(0, .05))
# Add Y-axis breaks in original scale
brks <- seq(0, 1, length.out = 5)
labs <- round(invbc(scales::rescale(brks, from = c(0, 1), to = rng_bc), lambda))
p <- p + scale_y_continuous(breaks = seq(0, 1, length.out = 5), labels = labs) + ylab("SCL (rescaled from Box-Cox)")
ggsave("appendices/plot_traj.png", p, width = 150, height = 120, units = "mm")
```

```{r echo = FALSE, eval = pkgdown, out.width="50%"}
knitr::include_graphics("appendices/plot_traj.png")
```

Note that the observed individual trajectories show very high
variability within classes.

<!-- \newpage -->

<!-- # References -->

<!-- ::: {#refs custom-style="Bibliography"} -->

<!-- ::: -->

## R session

```{r echo = TRUE, eval=TRUE}
sessionInfo()
```

### 2023-01-23 appendix_c.tex
edits by Mauricio

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  man,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
% See https://github.com/crsh/papaja/issues/292
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-1em}%
  {\normalfont\normalsize\bfseries\itshape\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
  {-\z@\relax}%
  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Appendix C: Latent Class Growth Analysis},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Appendix C: Latent Class Growth Analysis}
\author{\phantom{0}}
\date{}

\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}

\affiliation{\phantom{0}}

\begin{document}
\maketitle

This vignette illustrated \texttt{tidySEM}'s ability to perform latent class
growth analysis, or growth mixture modeling. The simulated data used for
this example are inspired by work in progress by Plas and colleagues, on
heterogeneity in depression trajectories among Dutch military personnel
who were deployed to Afghanistan. The original data were collected as
part of the \emph{Prospection in Stress-related Military Research (PRISMO)}
study, which examined of psychological problems after deployment in more
than 1,000 Dutch military personnel from 2005-2019.

First, we load all required packages:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(tidySEM)}
\FunctionTok{library}\NormalTok{(ggplot2)}
\FunctionTok{library}\NormalTok{(MASS)}
\end{Highlighting}
\end{Shaded}

\hypertarget{data-preprocessing}{%
\subsection{Data preprocessing}\label{data-preprocessing}}

We first examined the descriptive statistics for the sum score scales:

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:unnamed-chunk-2}Item descriptives}

\begin{tabular}{llllllll}
\toprule
name & \multicolumn{1}{c}{mean} & \multicolumn{1}{c}{median} & \multicolumn{1}{c}{sd} & \multicolumn{1}{c}{min} & \multicolumn{1}{c}{max} & \multicolumn{1}{c}{skew\_2se} & \multicolumn{1}{c}{kurt\_2se}\\
\midrule
scl.1 & 20.20 & 20.00 & 2.36 & 17.00 & 38.00 & 14.52 & 40.39\\
scl.2 & 20.42 & 19.00 & 3.51 & 16.00 & 64.00 & 25.90 & 111.96\\
scl.3 & 20.49 & 20.00 & 3.41 & 17.00 & 59.00 & 26.41 & 107.36\\
scl.4 & 20.61 & 20.00 & 3.36 & 16.00 & 50.00 & 18.23 & 54.25\\
scl.5 & 20.93 & 20.00 & 4.06 & 16.00 & 64.00 & 24.54 & 93.40\\
scl.6 & 21.07 & 20.00 & 4.10 & 16.00 & 58.00 & 20.44 & 65.76\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

Note that all variables were extremely right-skewed due to censoring at
the lower end of the scale.

We can examine these distributions visually as well:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df\_plot }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(df, }\AttributeTok{direction =} \StringTok{"long"}\NormalTok{, }\AttributeTok{varying =} \FunctionTok{names}\NormalTok{(df))}
\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ scl)) }\SpecialCharTok{+}
  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{time) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=0.5\linewidth]{appendices/plot_dist} \caption{ }\label{fig:unnamed-chunk-5}
\end{figure}

As this type of skew can result in convergence problems in LCGA, we
compared several transformations to reduce skew: The square and cube
root, log, inverse, and Box-Cox transformations.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df\_scores }\OtherTok{\textless{}{-}}\NormalTok{ df\_plot}
\CommentTok{\# Store original range of SCL}
\NormalTok{rng\_scl }\OtherTok{\textless{}{-}} \FunctionTok{range}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl)}
\CommentTok{\# Log{-}transform}
\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{log }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\FunctionTok{log}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl), }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\CommentTok{\# Square root transform}
\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{sqrt }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\FunctionTok{sqrt}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl), }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\CommentTok{\# Cube root transform}
\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{qrt }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl}\SpecialCharTok{\^{}}\NormalTok{.}\DecValTok{33}\NormalTok{, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\CommentTok{\# Reciprocal transform}
\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{reciprocal }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{scl, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\CommentTok{\# Define function for Box{-}Cox transformation}
\NormalTok{bc }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, lambda)\{}
\NormalTok{  (((x }\SpecialCharTok{\^{}}\NormalTok{ lambda) }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{) }\SpecialCharTok{/}\NormalTok{ lambda)}
\NormalTok{\}}
\CommentTok{\# Inverse Box{-}Cox transformation}
\NormalTok{invbc }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, lambda)\{}
\NormalTok{  ((x}\SpecialCharTok{*}\NormalTok{lambda)}\SpecialCharTok{+}\DecValTok{1}\NormalTok{)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{lambda)}
\NormalTok{\}}
\CommentTok{\# Box{-}Cox transform}
\NormalTok{b }\OtherTok{\textless{}{-}}\NormalTok{ MASS}\SpecialCharTok{::}\FunctionTok{boxcox}\NormalTok{(}\FunctionTok{lm}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl }\SpecialCharTok{\textasciitilde{}} \DecValTok{1}\NormalTok{), }\AttributeTok{plotit =} \ConstantTok{FALSE}\NormalTok{)}
\NormalTok{lambda }\OtherTok{\textless{}{-}}\NormalTok{ b}\SpecialCharTok{$}\NormalTok{x[}\FunctionTok{which.max}\NormalTok{(b}\SpecialCharTok{$}\NormalTok{y)]}
\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{boxcox }\OtherTok{\textless{}{-}} \FunctionTok{bc}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl, lambda)}
\CommentTok{\# Store range of Box{-}Cox transformed data}
\NormalTok{rng\_bc }\OtherTok{\textless{}{-}} \FunctionTok{range}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{boxcox)}
\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{boxcox }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{boxcox, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\CommentTok{\# Rescale SCL}
\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{scl }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

We can plot these transformations:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Make plot data}
\NormalTok{df\_plot }\OtherTok{\textless{}{-}} \FunctionTok{do.call}\NormalTok{(rbind, }\FunctionTok{lapply}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"scl"}\NormalTok{, }\StringTok{"log"}\NormalTok{, }\StringTok{"sqrt"}\NormalTok{, }\StringTok{"qrt"}\NormalTok{, }\StringTok{"boxcox"}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(n)\{}
  \FunctionTok{data.frame}\NormalTok{(df\_scores[}\FunctionTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"id"}\NormalTok{)],}
             \AttributeTok{Value =}\NormalTok{ df\_scores[[n]],}
             \AttributeTok{Transformation =}\NormalTok{ n)}
\NormalTok{\}))}
\CommentTok{\# Plot}
\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value, }\AttributeTok{colour =}\NormalTok{ Transformation)) }\SpecialCharTok{+}
  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{time) }\SpecialCharTok{+}
  \FunctionTok{scale\_y\_sqrt}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{xlab}\NormalTok{(}\StringTok{"scl (rescaled to 0{-}1)"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_bw}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=0.5\linewidth]{appendices/plot_trans} \caption{ }\label{fig:unnamed-chunk-9}
\end{figure}

Evidently, the Box-Cox transformation reduced skew the most.
Consequently, we proceeded with the Box-Cox transformed scores for
analysis.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dat }\OtherTok{\textless{}{-}}\NormalTok{ df\_scores[, }\FunctionTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"time"}\NormalTok{, }\StringTok{"boxcox"}\NormalTok{)]}
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(dat, }\AttributeTok{direction =} \StringTok{"wide"}\NormalTok{, }\AttributeTok{v.names =} \StringTok{"boxcox"}\NormalTok{, }\AttributeTok{timevar =} \StringTok{"time"}\NormalTok{, }\AttributeTok{idvar =} \StringTok{"id"}\NormalTok{)}
\FunctionTok{names}\NormalTok{(dat) }\OtherTok{\textless{}{-}} \FunctionTok{gsub}\NormalTok{(}\StringTok{"boxcox."}\NormalTok{, }\StringTok{"scl"}\NormalTok{, }\FunctionTok{names}\NormalTok{(dat))}
\end{Highlighting}
\end{Shaded}

\hypertarget{latent-class-growth-analysis}{%
\section{Latent Class Growth Analysis}\label{latent-class-growth-analysis}}

Next, we estimated a latent class growth model for SCL. The model
included an overall intercept, centered at T1, \texttt{i}. To model the
potential effect of deployment on depresion, we also included a dummy
variable that was zero before deployment, and 1 after deployment,
\texttt{step}. Finally, to model potential change (or recovery) in depression
post-deployment, we included a linear slope from T2-T6, \texttt{s}. All
variances of growth parameters were fixed to zero due to the sparse
nature of the data. In this vignette, we do not consider more than 5
classes, because the analyses are computationally very intensive and the
data were simulated from a 3-class model.

It is important to not that in LCGA, the subgroups will be limited by
the specify growth structure, meaning that LCA will identify distinctive
growth patterns within the intercept, step, and slope growth. For
example, if there is a subgroup that follows a quadratic growth pattern
this models will not be able to identify it.

\textbf{NOTE: The time scales in this model are not correct; it currently
assumes that all measurements are equidistant. Feel free to experiment
with adjusting this.}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{27796}\NormalTok{)}
\NormalTok{dat[[}\StringTok{"id"}\NormalTok{]] }\OtherTok{\textless{}{-}} \ConstantTok{NULL}
\NormalTok{res\_step }\OtherTok{\textless{}{-}} \FunctionTok{mx\_growth\_mixture}\NormalTok{(}
  \AttributeTok{model =}
  \StringTok{"}
\StringTok{  i =\textasciitilde{} 1*scl1 + 1*scl2 + 1*scl3 +1*scl4 +1*scl5 +1*scl6}
\StringTok{  step =\textasciitilde{} 0*scl1 + 1*scl2 + 1*scl3 +1*scl4 +1*scl5 +1*scl6}
\StringTok{  s =\textasciitilde{} 0*scl1 + 0*scl2 + 1*scl3 +2*scl4 +3*scl5 +4*scl6}
\StringTok{  scl1 \textasciitilde{}\textasciitilde{} vscl1*scl1}
\StringTok{  scl2 \textasciitilde{}\textasciitilde{} vscl2*scl2}
\StringTok{  scl3 \textasciitilde{}\textasciitilde{} vscl3*scl3}
\StringTok{  scl4 \textasciitilde{}\textasciitilde{} vscl4*scl4}
\StringTok{  scl5 \textasciitilde{}\textasciitilde{} vscl5*scl5}
\StringTok{  scl6 \textasciitilde{}\textasciitilde{} vscl6*scl6}
\StringTok{  i \textasciitilde{}\textasciitilde{} 0*i}
\StringTok{  step \textasciitilde{}\textasciitilde{} 0*step}
\StringTok{  s \textasciitilde{}\textasciitilde{} 0*s}
\StringTok{  i \textasciitilde{}\textasciitilde{} 0*s}
\StringTok{  i \textasciitilde{}\textasciitilde{} 0*step}
\StringTok{  s \textasciitilde{}\textasciitilde{} 0*step"}\NormalTok{, }\AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{,}
  \AttributeTok{data =}\NormalTok{ dat)}
\CommentTok{\# Additional iterations because of}
\CommentTok{\# convergence problems for model 1:}
\NormalTok{res\_step[[}\DecValTok{1}\NormalTok{]] }\OtherTok{\textless{}{-}} \FunctionTok{mxTryHardWideSearch}\NormalTok{(res\_step[[}\DecValTok{1}\NormalTok{]], }\AttributeTok{extraTries =} \DecValTok{50}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Note that the first model showed convergence problems, throwing the
error: \emph{The model does not satisfy the first-order optimality conditions
to the required accuracy, and no improved point for the merit function
could be found during the final linesearch.} To address this problem, we
performed additional iterations to\\
find a better solution, using \texttt{OpenMx::mxTryHardWideSearch()}. This also
illustrates that \texttt{tidySEM} mixture models inherit from \texttt{OpenMx}'s
\texttt{MxModel}, and thus, different \texttt{OpenMx} functions can be used to act on
models specified via \texttt{tidySEM}.

The fifth model also evidenced convergence problems, but this (as we
will see) is because the solution is overfitted.

\hypertarget{class-enumeration}{%
\subsection{Class enumeration}\label{class-enumeration}}

To determine the correct number of classes, we considered the following
criteria:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We do not consider classes with, on average, fewer than 5
  participants per parameter in a class due to potential local
  underidentification
\item
  Lower values for information criteria (AIC, BIC, saBIC) indicate
  better fit
\item
  Significant Lo-Mendell-Rubin LRT test indicates better fit for \(k\)
  vs \(k-1\) classes
\item
  We do not consider solutions with entropy \textless{} .90 because poor class
  separability compromises interpretability of the results
\item
  We do not consider solutions with minimum posterior classification
  probability \textless{} .90 because poor class separability compromises
  interpretability of the results
\end{enumerate}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Get fit table fit}
\NormalTok{tab\_fit }\OtherTok{\textless{}{-}} \FunctionTok{table\_fit}\NormalTok{(res\_step)}
\CommentTok{\# Select columns}
\NormalTok{tab\_fit }\OtherTok{\textless{}{-}}\NormalTok{ tab\_fit[, }\FunctionTok{c}\NormalTok{(}\StringTok{"Name"}\NormalTok{, }\StringTok{"Classes"}\NormalTok{, }\StringTok{"LL"}\NormalTok{, }\StringTok{"Parameters"}\NormalTok{, }\StringTok{"BIC"}\NormalTok{, }\StringTok{"Entropy"}\NormalTok{, }\StringTok{"prob\_min"}\NormalTok{, }\StringTok{"n\_min"}\NormalTok{, }\StringTok{"warning"}\NormalTok{, }\StringTok{"lmr\_p"}\NormalTok{)]}
\NormalTok{tab\_fit}
\end{Highlighting}
\end{Shaded}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:tabfit}Fit of LCGA models}

\begin{tabular}{llllllllll}
\toprule
Name & \multicolumn{1}{c}{Classes} & \multicolumn{1}{c}{LL} & \multicolumn{1}{c}{p} & \multicolumn{1}{c}{BIC} & \multicolumn{1}{c}{Ent.} & \multicolumn{1}{c}{p\_min} & \multicolumn{1}{c}{n\_min} & \multicolumn{1}{c}{warn} & \multicolumn{1}{c}{lmr\_p}\\
\midrule
1 & 1.00 & 2,592.32 & 9 & -5,122.67 & 1.00 & 1.00 & 1.00 & NA & NA\\
2 & 2.00 & 3,797.78 & 13 & -7,506.04 & 0.92 & 0.97 & 0.31 & NA & 0.00\\
3 & 3.00 & 4,264.43 & 17 & -8,411.80 & 0.95 & 0.96 & 0.10 & NA & 0.00\\
4 & 4.00 & 4,264.45 & 21 & -8,384.30 & 0.72 & 0.00 & 0.00 & NA & 1.00\\
5 & 5.00 & 4,264.45 & 25 & -8,356.75 & 0.55 & 0.00 & 0.00 & TRUE & 1.00\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

According to the Table, increasing the number of classes keeps
increasing model fit according to all ICs except the BIC, which
increased after 3 classes.

The first two LMR tests are significant, indicating that a 2- and
3-class solution were a significant improvement over a 1- and 2-class
solution, respectively. However, solutions with \textgreater3 classes had entropy
and minimum posterior classification probability below the pre-specified
thresholds. Models with \textgreater3 solutions also had fewer than five
observations per parameter. This suggests that the preferred model
should be selected from 1-3 classes.

\hypertarget{scree-plot}{%
\subsubsection{Scree plot}\label{scree-plot}}

A scree plot indicates that the largest decrease in ICs occurs from 1-2
classes, and the inflection point for all ICs is at 3 classes. Moreover,
the BIC increased after 3 classes. A three-class solution thus appears
to be the most parsimonious solution with good fit.

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{plot}\NormalTok{(tab\_fit, }\AttributeTok{statistics =} \FunctionTok{c}\NormalTok{(}\StringTok{"AIC"}\NormalTok{, }\StringTok{"BIC"}\NormalTok{, }\StringTok{"saBIC"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=0.5\linewidth]{appendices/lcga_plot_fit} \caption{ }\label{fig:unnamed-chunk-15}
\end{figure}

Based on the aforementioned criteria, we selected a 3-class model for
further analyses. First, to prevent label switching, we re-order these
classes by the value of the intercept \texttt{i}. Then, we report the estimated
parameters.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res\_final }\OtherTok{\textless{}{-}} \FunctionTok{mx\_switch\_labels}\NormalTok{(res\_step[[}\DecValTok{3}\NormalTok{]], }\AttributeTok{param =} \StringTok{"M[1,7]"}\NormalTok{, }\AttributeTok{decreasing =} \ConstantTok{FALSE}\NormalTok{)}
\NormalTok{tab\_res }\OtherTok{\textless{}{-}} \FunctionTok{table\_results}\NormalTok{(res\_final, }\AttributeTok{columns =} \ConstantTok{NULL}\NormalTok{)}
\CommentTok{\# Select rows and columns}
\NormalTok{tab\_res }\OtherTok{\textless{}{-}}\NormalTok{ tab\_res[tab\_res}\SpecialCharTok{$}\NormalTok{Category }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{"Means"}\NormalTok{, }\StringTok{"Variances"}\NormalTok{), }\FunctionTok{c}\NormalTok{(}\StringTok{"Category"}\NormalTok{, }\StringTok{"lhs"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"pval"}\NormalTok{, }\StringTok{"confint"}\NormalTok{, }\StringTok{"name"}\NormalTok{)]}
\NormalTok{tab\_res}
\end{Highlighting}
\end{Shaded}

\begin{center}
\begin{ThreePartTable}

\begin{longtable}{llllllll}\noalign{\getlongtablewidth\global\LTcapwidth=\longtablewidth}
\caption{\label{tab:tabres}Results from 3-class LCGA model}\\
\toprule
 & \multicolumn{1}{c}{Category} & \multicolumn{1}{c}{lhs} & \multicolumn{1}{c}{est} & \multicolumn{1}{c}{se} & \multicolumn{1}{c}{pval} & \multicolumn{1}{c}{confint} & \multicolumn{1}{c}{name}\\
\midrule
\endfirsthead
\caption*{\normalfont{Table \ref{tab:tabres} continued}}\\
\toprule
 & \multicolumn{1}{c}{Category} & \multicolumn{1}{c}{lhs} & \multicolumn{1}{c}{est} & \multicolumn{1}{c}{se} & \multicolumn{1}{c}{pval} & \multicolumn{1}{c}{confint} & \multicolumn{1}{c}{name}\\
\midrule
\endhead
16 & Means & i & 0.35 & 0.00 & 0.00 & {}[0.35, 0.36] & class1.M[1,7]\\
17 & Means & step & -0.03 & 0.01 & 0.00 & {}[-0.04, -0.02] & class1.M[1,8]\\
18 & Means & s & 0.01 & 0.00 & 0.00 & {}[0.00, 0.01] & class1.M[1,9]\\
19 & Variances & scl1 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class1.S[1,1]\\
20 & Variances & scl2 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class1.S[2,2]\\
21 & Variances & scl3 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class1.S[3,3]\\
22 & Variances & scl4 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class1.S[4,4]\\
23 & Variances & scl5 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class1.S[5,5]\\
24 & Variances & scl6 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.02] & class1.S[6,6]\\
40 & Means & i & 0.46 & 0.01 & 0.00 & {}[0.44, 0.47] & class2.M[1,7]\\
41 & Means & step & 0.03 & 0.01 & 0.00 & {}[0.01, 0.04] & class2.M[1,8]\\
42 & Means & s & 0.01 & 0.00 & 0.00 & {}[0.01, 0.02] & class2.M[1,9]\\
43 & Variances & scl1 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class2.S[1,1]\\
44 & Variances & scl2 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class2.S[2,2]\\
45 & Variances & scl3 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class2.S[3,3]\\
46 & Variances & scl4 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class2.S[4,4]\\
47 & Variances & scl5 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class2.S[5,5]\\
48 & Variances & scl6 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.02] & class2.S[6,6]\\
64 & Means & i & 0.63 & 0.01 & 0.00 & {}[0.61, 0.65] & class3.M[1,7]\\
65 & Means & step & 0.07 & 0.01 & 0.00 & {}[0.05, 0.10] & class3.M[1,8]\\
66 & Means & s & 0.01 & 0.00 & 0.13 & {}[-0.00, 0.01] & class3.M[1,9]\\
67 & Variances & scl1 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class3.S[1,1]\\
68 & Variances & scl2 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class3.S[2,2]\\
69 & Variances & scl3 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class3.S[3,3]\\
70 & Variances & scl4 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class3.S[4,4]\\
71 & Variances & scl5 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.01] & class3.S[5,5]\\
72 & Variances & scl6 & 0.01 & 0.00 & 0.00 & {}[0.01, 0.02] & class3.S[6,6]\\
\bottomrule
\end{longtable}

\end{ThreePartTable}
\end{center}

As evident from these results, Class 1 started at a relatively lower
level of depressive symptoms, experienced a decrease after deployment,
followed by increase over time. Class 2 started at a moderate level of
depressive symptoms, experienced an increase after deployment, followed
by significant increase over time from T2-T6. Class 3 started at a
relatively higher level, experienced an increase after deployment,
followed by stability.

\hypertarget{wald-tests}{%
\subsection{Wald tests}\label{wald-tests}}

To test whether parameters are significantly different between classes,
we can use Wald tests. Wald tests can be specified for all parameters in
the model, using the hypothesis syntax from the \texttt{bain} package for
informative hypothesis testing.

To identify the names of parameters in the model, we can use the \texttt{name}
column of the results table above. Alternatively, to see all parameters
in the model, run:

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{names}\NormalTok{(}\FunctionTok{coef}\NormalTok{(res\_final))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#>  [1] "mix3.weights[1,2]" "mix3.weights[1,3]" "vscl1"            
#>  [4] "vscl2"             "vscl3"             "vscl4"            
#>  [7] "vscl5"             "vscl6"             "class1.M[1,7]"    
#> [10] "class1.M[1,8]"     "class1.M[1,9]"     "class2.M[1,7]"    
#> [13] "class2.M[1,8]"     "class2.M[1,9]"     "class3.M[1,7]"    
#> [16] "class3.M[1,8]"     "class3.M[1,9]"
\end{verbatim}

Next, specify equality constrained hypotheses. For example, a hypothesis
that states that the mean intercept is equal across groups is specified
as follows:

\texttt{"class1.M{[}1,7{]}\ =\ class2.M{[}1,7{]}\ \&\ class1.M{[}1,7{]}\ =\ class3.M{[}1,7{]}}

It is also possible to consider comparisons between two classes at a
time. When conducting many significance tests, consider correcting for
multiple comparisons however.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{wald\_tests }\OtherTok{\textless{}{-}} \FunctionTok{wald\_test}\NormalTok{(res\_final, }
                   \StringTok{"}
\StringTok{                   class1.M[1,7] = class2.M[1,7]\&}
\StringTok{                   class1.M[1,7] = class3.M[1,7];}
\StringTok{                   class1.M[1,8] = class2.M[1,8]\&}
\StringTok{                   class1.M[1,8] = class3.M[1,8];}
\StringTok{                   class1.M[1,9] = class2.M[1,9]\&}
\StringTok{                   class1.M[1,9] = class3.M[1,9]"}\NormalTok{)}
\CommentTok{\# Rename the hypothesis}
\NormalTok{wald\_tests}\SpecialCharTok{$}\NormalTok{Hypothesis }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\StringTok{"Mean i"}\NormalTok{, }\StringTok{"Mean step"}\NormalTok{, }\StringTok{"Mean slope"}\NormalTok{)}
\NormalTok{papaja}\SpecialCharTok{::}\FunctionTok{apa\_table}\NormalTok{(wald\_tests, }\AttributeTok{digits =} \DecValTok{2}\NormalTok{, }\AttributeTok{caption =} \StringTok{"Wald tests"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{table}[tbp]

\begin{center}
\begin{threeparttable}

\caption{\label{tab:waldtests}Wald tests}

\begin{tabular}{llll}
\toprule
Hypothesis & \multicolumn{1}{c}{df} & \multicolumn{1}{c}{chisq} & \multicolumn{1}{c}{p}\\
\midrule
Mean i & 2 & 628.38 & 0.00\\
Mean step & 2 & 65.32 & 0.00\\
Mean slope & 2 & 13.03 & 0.00\\
\bottomrule
\end{tabular}

\end{threeparttable}
\end{center}

\end{table}

All Wald tests are significant, indicating that there are significant
differences between the intercepts, step function, and slopes of the
three classes.

\hypertarget{trajectory-plot}{%
\subsection{Trajectory plot}\label{trajectory-plot}}

Finally, we can plot the growth trajectories. This can help interpret
the results better, as well as the residual heterogeneity around class
trajectories.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{plot\_growth}\NormalTok{(res\_step[[}\DecValTok{3}\NormalTok{]], }\AttributeTok{rawdata =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{alpha\_range =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{05}\NormalTok{))}
\CommentTok{\# Add Y{-}axis breaks in original scale}
\NormalTok{brks }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{length.out =} \DecValTok{5}\NormalTok{)}
\NormalTok{labs }\OtherTok{\textless{}{-}} \FunctionTok{round}\NormalTok{(}\FunctionTok{invbc}\NormalTok{(scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(brks, }\AttributeTok{from =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{to =}\NormalTok{ rng\_bc), lambda))}
\NormalTok{p }\OtherTok{\textless{}{-}}\NormalTok{ p }\SpecialCharTok{+} \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{length.out =} \DecValTok{5}\NormalTok{), }\AttributeTok{labels =}\NormalTok{ labs) }\SpecialCharTok{+} \FunctionTok{ylab}\NormalTok{(}\StringTok{"SCL (rescaled from Box{-}Cox)"}\NormalTok{)}
\NormalTok{p}
\end{Highlighting}
\end{Shaded}

\begin{figure}
\includegraphics[width=0.5\linewidth]{appendices/plot_traj} \caption{ }\label{fig:unnamed-chunk-22}
\end{figure}

Note that the observed individual trajectories show very high
variability within classes.

\hypertarget{r-session}{%
\subsection{R session}\label{r-session}}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{sessionInfo}\NormalTok{()}
\CommentTok{\#\textgreater{} R version 4.1.3 (2022{-}03{-}10)}
\CommentTok{\#\textgreater{} Platform: x86\_64{-}w64{-}mingw32/x64 (64{-}bit)}
\CommentTok{\#\textgreater{} Running under: Windows 10 x64 (build 19045)}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Matrix products: default}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} locale:}
\CommentTok{\#\textgreater{} [1] LC\_COLLATE=English\_United Kingdom.1252 }
\CommentTok{\#\textgreater{} [2] LC\_CTYPE=English\_United Kingdom.1252   }
\CommentTok{\#\textgreater{} [3] LC\_MONETARY=English\_United Kingdom.1252}
\CommentTok{\#\textgreater{} [4] LC\_NUMERIC=C                           }
\CommentTok{\#\textgreater{} [5] LC\_TIME=English\_United Kingdom.1252    }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} attached base packages:}
\CommentTok{\#\textgreater{} [1] stats     graphics  grDevices utils     datasets  methods   base     }
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} other attached packages:}
\CommentTok{\#\textgreater{} [1] MASS\_7.3{-}58.2    ggplot2\_3.4.0    tidySEM\_0.2.4.6  OpenMx\_2.21.1   }
\CommentTok{\#\textgreater{} [5] scales\_1.2.1     yaml\_2.3.6       papaja\_0.1.1     tinylabels\_0.2.3}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} loaded via a namespace (and not attached):}
\CommentTok{\#\textgreater{}   [1] TH.data\_1.1{-}1         colorspace\_2.1{-}0      estimability\_1.4.1   }
\CommentTok{\#\textgreater{}   [4] parameters\_0.20.1     rstudioapi\_0.14       listenv\_0.9.0        }
\CommentTok{\#\textgreater{}   [7] lavaan\_0.6{-}13         rstan\_2.26.13         fansi\_1.0.4          }
\CommentTok{\#\textgreater{}  [10] mvtnorm\_1.1{-}3         codetools\_0.2{-}18      splines\_4.1.3        }
\CommentTok{\#\textgreater{}  [13] mnormt\_2.1.1          knitr\_1.41            texreg\_1.38.6        }
\CommentTok{\#\textgreater{}  [16] bayesplot\_1.10.0      jsonlite\_1.8.4        effectsize\_0.8.2     }
\CommentTok{\#\textgreater{}  [19] compiler\_4.1.3        httr\_1.4.4            emmeans\_1.8.4{-}1      }
\CommentTok{\#\textgreater{}  [22] backports\_1.4.1       assertthat\_0.2.1      Matrix\_1.5{-}3         }
\CommentTok{\#\textgreater{}  [25] fastmap\_1.1.0         cli\_3.6.0             htmltools\_0.5.4      }
\CommentTok{\#\textgreater{}  [28] prettyunits\_1.1.1     tools\_4.1.3           igraph\_1.3.5         }
\CommentTok{\#\textgreater{}  [31] coda\_0.19{-}4           gtable\_0.3.1          glue\_1.6.2           }
\CommentTok{\#\textgreater{}  [34] RANN\_2.6.1            dplyr\_1.0.10          V8\_4.2.2             }
\CommentTok{\#\textgreater{}  [37] Rcpp\_1.0.10           carData\_3.0{-}5         vctrs\_0.5.2          }
\CommentTok{\#\textgreater{}  [40] nlme\_3.1{-}161          psych\_2.2.9           insight\_0.18.8       }
\CommentTok{\#\textgreater{}  [43] xfun\_0.36             stringr\_1.5.0         globals\_0.16.2       }
\CommentTok{\#\textgreater{}  [46] ps\_1.7.2              proto\_1.0.0           CompQuadForm\_1.4.3   }
\CommentTok{\#\textgreater{}  [49] lifecycle\_1.0.3       future\_1.30.0         zoo\_1.8{-}11           }
\CommentTok{\#\textgreater{}  [52] parallel\_4.1.3        sandwich\_3.0{-}2        inline\_0.3.19        }
\CommentTok{\#\textgreater{}  [55] curl\_5.0.0            gridExtra\_2.3         pander\_0.6.5         }
\CommentTok{\#\textgreater{}  [58] blavaan\_0.4{-}3         loo\_2.5.1             StanHeaders\_2.26.13  }
\CommentTok{\#\textgreater{}  [61] stringi\_1.7.12        bayestestR\_0.13.0     fastDummies\_1.6.3    }
\CommentTok{\#\textgreater{}  [64] checkmate\_2.1.0       boot\_1.3{-}28           pkgbuild\_1.4.0       }
\CommentTok{\#\textgreater{}  [67] rlang\_1.0.6           pkgconfig\_2.0.3       matrixStats\_0.63.0   }
\CommentTok{\#\textgreater{}  [70] evaluate\_0.20         lattice\_0.20{-}45       rstantools\_2.2.0     }
\CommentTok{\#\textgreater{}  [73] processx\_3.8.0        tidyselect\_1.2.0      parallelly\_1.34.0    }
\CommentTok{\#\textgreater{}  [76] plyr\_1.8.8            magrittr\_2.0.3        bookdown\_0.32        }
\CommentTok{\#\textgreater{}  [79] R6\_2.5.1              generics\_0.1.3        multcomp\_1.4{-}20      }
\CommentTok{\#\textgreater{}  [82] DBI\_1.1.3             withr\_2.5.0           gsubfn\_0.7           }
\CommentTok{\#\textgreater{}  [85] pillar\_1.8.1          survival\_3.5{-}0        datawizard\_0.6.5     }
\CommentTok{\#\textgreater{}  [88] abind\_1.4{-}5           tibble\_3.1.8          future.apply\_1.10.0  }
\CommentTok{\#\textgreater{}  [91] crayon\_1.5.2          car\_3.1{-}1             nonnest2\_0.5{-}5       }
\CommentTok{\#\textgreater{}  [94] utf8\_1.2.2            tmvnsim\_1.0{-}2         MplusAutomation\_1.1.0}
\CommentTok{\#\textgreater{}  [97] rmarkdown\_2.20        grid\_4.1.3            data.table\_1.14.6    }
\CommentTok{\#\textgreater{} [100] pbivnorm\_0.6.0        bain\_0.2.8            callr\_3.7.3          }
\CommentTok{\#\textgreater{} [103] digest\_0.6.31         xtable\_1.8{-}4          dbscan\_1.1{-}11        }
\CommentTok{\#\textgreater{} [106] RcppParallel\_5.1.6    stats4\_4.1.3          munsell\_0.5.0}
\end{Highlighting}
\end{Shaded}

\end{document}

### 2023-01-23 manuscript.Rmd
edits by Mauricio

@@
     wrap: 72
+#bibliography: references.bib
 ---
@@
 most notably finite Gaussian mixture modeling and latent profile
-analysis. @vermunt_latent_2004 defined LCA more generally as
-virtually any statistical model where "some of the parameters [...]
-differ across unobserved subgroups".
-
-Despite the popularity of LCA,
-two challenges impede broader adoption and correct application of the method.
-The first challenge is that most existing software that implements these models is commercial and closed-source.
-The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
-This introduces a risk of misapplications of the technique,
-and complicates manuscript review and assessment of the quality of published studies.
-The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
-based on current best practice. 
-Furthermore, meeting the increasing demand for open source research software,
-this paper introduces new functionality in the R-package `tidySEM` [@vanlissaTidySEMTidyStructural2022] to perform LCA using free open source software (FOSS).
+analysis. @vermunt_latent_2004 defined LCA more generally as virtually
+any statistical model where "some of the parameters [...] differ across
+unobserved subgroups".
+
+Despite the popularity of LCA, two challenges impede broader adoption
+and correct application of the method. The first challenge is that most
+existing software that implements these models is commercial and
+closed-source. The second challenge is that there is a lack of standards
+for estimating and reporting LCA, and some myths and misunderstandings
+have been perpetuated in prior literature. This introduces a risk of
+misapplications of the technique, and complicates manuscript review and
+assessment of the quality of published studies. The present paper seeks
+to address these challenges by introducing updated guidelines for
+estimating and reporting LCA, based on current best practice.
+Furthermore, meeting the increasing demand for open source research
+software, this paper introduces new functionality in the R-package
+`tidySEM` [@vanlissaTidySEMTidyStructural2022] to perform LCA using free
+open source software (FOSS).
 
@@
 
-Despite the popularity of LCA,
-most existing software is commercial and closed-source,
-for example, Mplus [@muthenMplusUserGuide1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquistMplusAutomationPackageFacilitating2018], and LatentGOLD [REF].
-<!-- software that implements these models. -->
-<!-- One important limitation is that the most fully-features -->
-A downside of commercial software is that its use is restricted to individuals and institutions who can afford a license.
-Another downside is that the proprietary source code cannot be audited, debugged, or enhanced by third parties.
-This goes against open science principles [@lamprechtFAIRPrinciplesResearch2019],
-incurs a risk that mistakes in the source code go unnoticed,
-and curbs progress as independent developers cannot fix bugs or add functionality.
-
-There are several compelling reasons to prefer free open source software (FOSS) instead.
-FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
-This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
-Furthermore, FOSS promotes collaboration within the scientific community,
-as its users can contribute to the development, maintenance, and support of the software.
-Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
-researchers can review and verify the methods and algorithms used,
-which helps to ensure the integrity and reliability of research findings.
-Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.
-
-Notable FOSS for LCA includes Mclust [REF], depmixS4 [REF], and OpenMx [@neale_openmx_2016].
-A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
-or implement specific innovations (e.g., depmixS4).
-OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
-Remaining obstacles to the use of `OpenMx` is that it is not very user-friendly and its functionality for LCA is poorly documented.
-
-These limitations are addressed by the `tidySEM` package,
-which provide a user-friendly interface for LCA in `OpenMx`.
-The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting LCA, as described in this paper.
-The primary purpose of the `tidySEM` package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
-The user interface is simple, as models can be specified using the widely used `lavaan` syntax [@rosseel_lavaan_2012].
-The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
-but the resulting models can still be fully customized (low barrier, high ceiling).
-Specifically, LCA models constructed in `tidySEM` inherit from `OpenMx`' format `MxModel`, and remain compatible with downstream functions that process such models.
-Being FOSS, `tidySEM` facilitates fully-reproducible analyses and contributes to open science.
-It was developed with open science in mind,
-and adheres to the FAIR software principles [Findable, Accessible, Interoperable, and Reusable, @lamprechtFAIRPrinciplesResearch2019],
-and the [OpenSSF Best Practices](https://bestpractices.coreinfrastructure.org/en/criteria) for FOSS software.
+Despite the popularity of LCA, most existing software is commercial and
+closed-source, for example, Mplus [@muthenMplusUserGuide1998], for which
+the R-package `MplusAutomation` offers augmented LCA functionality
+[@hallquistMplusAutomationPackageFacilitating2018], and LatentGOLD
+[REF]. <!-- software that implements these models. -->
+<!-- One important limitation is that the most fully-features --> A
+downside of commercial software is that its use is restricted to
+individuals and institutions who can afford a license. Another downside
+is that the proprietary source code cannot be audited, debugged, or
+enhanced by third parties. This goes against open science principles
+[@lamprechtFAIRPrinciplesResearch2019], incurs a risk that mistakes in
+the source code go unnoticed, and curbs progress as independent
+developers cannot fix bugs or add functionality.
+
+There are several compelling reasons to prefer free open source software
+(FOSS) instead. FOSS is freely available for anyone to use, modify, and
+distribute, which means that scientists can access and use a wide range
+of powerful tools without financial impediments. This is invaluable for
+reducing gatekeeping of researchers with limited budgets, such as those
+from developing countries. Furthermore, FOSS promotes collaboration
+within the scientific community, as its users can contribute to the
+development, maintenance, and support of the software. Finally, FOSS
+promotes transparency and reproducibility: Because the source code for
+FOSS is accessible, researchers can review and verify the methods and
+algorithms used, which helps to ensure the integrity and reliability of
+research findings. Overall, FOSS has the potential to significantly
+improve the efficiency and effectiveness of scientific research, while
+also promoting transparency, collaboration, and inclusion within the
+scientific community.
+
+Notable FOSS for LCA includes mclust [@scrucca_mclust_2016], depmixS4
+[@depmixS4], and OpenMx [@neale_openmx_2016]. A crucial limitation of
+these existing packages is that some have limited functionality (e.g.,
+Mclust), or implement specific innovations (e.g., depmixS4). OpenMx is
+unique in that, like Mplus, it implement LCA in the context of a fully
+featured structural equation modeling framework. Remaining obstacles to
+the use of `OpenMx` is that it is not very user-friendly and its
+functionality for LCA is poorly documented.
+
+These limitations are addressed by the `tidySEM` package, which provide
+a user-friendly interface for LCA in `OpenMx`. The package also includes
+convenience functions to obtain relevant statistics, tests and plots in
+accordance with best practices in estimation and reporting LCA, as
+described in this paper. The primary purpose of the `tidySEM` package is
+to provide a tidy workflow for generating, estimating, reporting, and
+plotting structural equation models in a software-agnostic way. The user
+interface is simple, as models can be specified using the widely used
+`lavaan` syntax [@rosseel_lavaan_2012]. The package contains
+user-friendly wrapper functions to conduct LCA with sensible defaults,
+but the resulting models can still be fully customized (low barrier,
+high ceiling). Specifically, LCA models constructed in `tidySEM` inherit
+from `OpenMx`' format `MxModel`, and remain compatible with downstream
+functions that process such models. Being FOSS, `tidySEM` facilitates
+fully-reproducible analyses and contributes to open science. It was
+developed with open science in mind, and adheres to the FAIR software
+principles [Findable, Accessible, Interoperable, and Reusable,
+@lamprechtFAIRPrinciplesResearch2019], and the [OpenSSF Best
+Practices](https://bestpractices.coreinfrastructure.org/en/criteria) for
+FOSS software.
 
@@
 
-LCA refers to methods for estimating unobserved
-groups based on a parametric model of observed indicators of group
-membership [@vermuntLatentClassAnalysis2004].
-It has become popular across scientific fields,
-under a number of different names;
-most notably (finite Gaussian) mixture modeling and latent profile analysis.
-The concept of LCA can be understood in different ways.
-Generally speaking, LCA assumes that the study population is composed of $K$
-subpopulations.
-It further assumes that the observed data are
-a mixture of data originating from those subpopulations;
-hence the name mixture model.
-Consider the simplest possible
-univariate "model", which is a normal distribution.
-This model has
-two parameters: the mean and the variance.
-LCA aims to estimate the values of those parameters across a known number of classes $K$,
-as well as the probability of every individual $i$ belonging to classes $1 \ldots k$.
-Commonly, the same model is estimated across all classes,
-but with different parameters for each class (i.e., class-specific means and variances).
+LCA refers to methods for estimating unobserved groups based on a
+parametric model of observed indicators of group membership
+[@vermuntLatentClassAnalysis2004]. It has become popular across
+scientific fields, under a number of different names; most notably
+(finite Gaussian) mixture modeling and latent profile analysis. The
+concept of LCA can be understood in different ways. Generally speaking,
+LCA assumes that the study population is composed of $K$ subpopulations.
+It further assumes that the observed data are a mixture of data
+originating from those subpopulations; hence the name mixture model.
+Consider the simplest possible univariate "model", which is a normal
+distribution. This model has two parameters: the mean and the variance.
+LCA aims to estimate the values of those parameters across a known
+number of classes $K$, as well as the probability of every individual
+$i$ belonging to classes $1 \ldots k$. Commonly, the same model is
+estimated across all classes, but with different parameters for each
+class (i.e., class-specific means and variances).
 
@@
 
-Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling [@molenaarArbitraryNatureLatent1994].
-A mixture model is like confirmatory factor analysis,
-except with a categorical instead of a continuous latent variable.
-One difference between
-the two techniques is that factor analysis can be considered as a way to
-group observed *variables* into latent constructs, and LCA groups
-*individuals* into classes.
-While factor analysis seeks to describe the relations between variables,
-LCA seeks to describe unobserved groups and classify individual observations to those groups.
+Another way to conceptualize LCA is by analogy to a measurement model in
+structural equation modeling [@molenaarArbitraryNatureLatent1994]. A
+mixture model is like confirmatory factor analysis, except with a
+categorical instead of a continuous latent variable. One difference
+between the two techniques is that factor analysis can be considered as
+a way to group observed *variables* into latent constructs, and LCA
+groups *individuals* into classes. While factor analysis seeks to
+describe the relations between variables, LCA seeks to describe
+unobserved groups and classify individual observations to those groups.
 <!-- In factor analysis, the latent variable is continuous, so the relationships -->
@@
 <!-- well they separate the classes, as a multinomial regression between the -->
-<!-- indicators and the latent variable [@masyn_latent_2013]. -->
-In line with this distinction, LCA
-(mixture models) is referred to as a "person-centered" technique, and
-factor analysis as a "variable-centered" technique
-[@nylund-gibson_ten_2018; @masyn_latent_2013].
+<!-- indicators and the latent variable [@masyn_latent_2013]. --> In
+line with this distinction, LCA (mixture models) is referred to as a
+"person-centered" technique, and factor analysis as a
+"variable-centered" technique [@nylund-gibson_ten_2018;
+@masyn_latent_2013].
 
@@
 thought of as similar to a multi-group structural equation model. The
-main distinction is that group membership is not known a-priori, but estimated from the data.
-Whereas in a multi-group model, the data are split by group and treated
-as independent samples, in LCA, all cases contribute to the estimation
-of all parameters in all groups. The relative contribution of each case
-to the parameters of each group is determined by that case's posterior
-probability of belonging to that group.
-
-When the focus is on each individual's estimated class membership,
-LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
-Specifically, the posterior class probability that an individual belongs to a latent class
-can be computed from the likelihood of that individual's observed data under given the class-specific model.
-These posterior class probabilities can be used to weight follow-up analyses,
-or individuals can be assigned to classes based on their highest class probability.
-If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
-This perspective on mixture modeling is sometimes referred to as "model-based
-clustering" [@hennig_handbook_2015; @scrucca_mclust_2016]. 
-Many clustering algorithms apply some recursive splitting algorithm to the data. By
-contrast "model-based" clustering refers to the fact that LCA estimates
-cluster membership based on a parametric model.
-Since a parametric model estimates only a fixed number of parameters,
-it can be a relatively parsimonious solution compared to non-parametric techniques.
+main distinction is that group membership is not known a-priori, but
+estimated from the data. Whereas in a multi-group model, the data are
+split by group and treated as independent samples, in LCA, all cases
+contribute to the estimation of all parameters in all groups. The
+relative contribution of each case to the parameters of each group is
+determined by that case's posterior probability of belonging to that
+group.
+
+When the focus is on each individual's estimated class membership, LCA
+can be thought of as a type of clustering algorithm that corrects for
+(random) measurement error. Specifically, the posterior class
+probability that an individual belongs to a latent class can be computed
+from the likelihood of that individual's observed data under given the
+class-specific model. These posterior class probabilities can be used to
+weight follow-up analyses, or individuals can be assigned to classes
+based on their highest class probability. If the classes are clearly
+distinct, measurement error is low, and the latter approach may be
+acceptable. This perspective on mixture modeling is sometimes referred
+to as "model-based clustering" [@hennig_handbook_2015;
+@scrucca_mclust_2016]. Many clustering algorithms apply some recursive
+splitting algorithm to the data. By contrast "model-based" clustering
+refers to the fact that LCA estimates cluster membership based on a
+parametric model. Since a parametric model estimates only a fixed number
+of parameters, it can be a relatively parsimonious solution compared to
+non-parametric techniques.
 
 Finally, in the context of machine learning, LCA can be considered as an
-*unsupervised* learning method for *classification* [@figueiredo_unsupervised_2002].
-The term *unsupervised* refers to the fact that the outcome variable --
-true class membership -- is not known, and the term *classification*
-refers to the fact that the algorithm is predicting a categorical
-outcome: class membership.
+*unsupervised* learning method for *classification*
+[@figueiredo_unsupervised_2002]. The term *unsupervised* refers to the
+fact that the outcome variable -- true class membership -- is not known,
+and the term *classification* refers to the fact that the algorithm is
+predicting a categorical outcome: class membership.
 
@@
 
-The term LCA was initially conceived to refer to analyses with categorical,
-usually binary indicators [@collins_latent_2009].
-It was later generalized to refer to any technique that estimates a categorical latent variable based on a
-parametric model of observed indicators, as it only refers
-to the purpose of the analysis, and does not imply restrictions to the
-model used, or the level of measurement of the indicators [@vermunt_latent_2004].
-To prevent ambiguity, we refer to LCA with binary or ordinal indicators as *LCA with ordinal indicators*.
-Given the abundance of redundant terms in use for closely related types of LCA models,
-we here provide a rudimentary taxonomy.
+The term LCA was initially conceived to refer to analyses with
+categorical, usually binary indicators [@collins_latent_2009]. It was
+later generalized to refer to any technique that estimates a categorical
+latent variable based on a parametric model of observed indicators, as
+it only refers to the purpose of the analysis, and does not imply
+restrictions to the model used, or the level of measurement of the
+indicators [@vermunt_latent_2004]. To prevent ambiguity, we refer to LCA
+with binary or ordinal indicators as *LCA with ordinal indicators*.
+Given the abundance of redundant terms in use for closely related types
+of LCA models, we here provide a rudimentary taxonomy.
 
 A common type of LCA is the *finite Gaussian mixture model*; a
-univariate analysis where the observed distribution of a single variable
-is assumed to result from a mixture of a known number of normal
-(Gaussian) distributions. The parameters of a finite Gaussian mixture
-model are the means and variances of these underlying normal
-distributions. The analysis of shoe sizes presented earlier is a
+univariate analysis where the observed distribution of a single
+continuous variable is assumed to result from a mixture of a known
+number of normal (Gaussian) distributions. The parameters of a finite
+Gaussian mixture model are the means and variances of these underlying
+normal distributions. The analysis of shoe sizes presented earlier is a
 canonical example of this type of analysis. In the multivariate case,
@@
 are the means, variances, and covariances between the indicators (which
-can be standardized to obtain correlations). These parameters can be estimated 
-freely, or set to be constrained across classes [@collins_latent_2009].
+can be standardized to obtain correlations). These parameters can be
+estimated freely, or set to be constrained across classes
+[@collins_latent_2009].
 
@@
 indicators. Conditional independence means that, after class membership
-is accounted for,
-the residual covariances/correlations between indicators are
-assumed to be zero.
-In some cases, such constraints
-will be inappropriate; for instance when the cohesion between indicators
-is expected to differ between classes [@collins_latent_2009].
-Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
-A class of larger particles with a low correlation between length and width,
-and a class of smaller particles with a high correlation between length and width.
+is accounted for, the residual covariances/correlations between
+indicators are assumed to be zero. In some cases, such constraints will
+be inappropriate; for instance when the cohesion between indicators is
+expected to differ between classes [@collins_latent_2009]. Consider, for
+example, a mixture model of ocean plastic particles, which found two
+classes of particles based on length and width: A class of larger
+particles with a low correlation between length and width, and a class
+of smaller particles with a high correlation between length and width.
 The reason for this difference in correlations makes theoretical sense:
-the large particles were heterogenous in shape,
-whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements [@alkema_maximizing_2022].
-
-Finally, some LCA models use latent variables as indicators.
-A common application of this approach is in
-longitudinal research with repeated measures of a construct.
-Examples of this approach include
-*growth mixture models* (GMM) and *latent class growth analyses* (LCGA) 
-[@jung_introduction_2008].
-These techniques estimate a latent growth model 
-to describe grouped trajectories over time.
-The indicators of class membership in a GMM are the 
-intercepts, and variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
-For linear trajectories, these latent variables are an intercept and slope.
-The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes homogeneity of trajectories within classes.
-This is achieved by restricting the (co)variances of the growth parameters to zero.
-Any variance in the indicators not explained by the class-specific
+the large particles were heterogenous in shape, whereas the smaller
+particles had been polished to a more uniform (rounded) shape by the
+elements [@alkema_maximizing_2022].
+
+A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
+called Latent Transition Analysis), which estimates the change in class
+membership over time with the markov assumption over time. The markov
+assumption states that the class membership at time $t$ is only affected
+by the class membership at time $t-1$. The structural part of the model
+is comprise of the initial states (proportion of the sample in each
+class at the first time point), and transition probabilities
+(probabilities to switch class membership across time). The transition
+probabilities are estimated as multinomial logistic regressions between
+class memberships [@vermunt_longitudinal_2010].
+
+Finally, some LCA models use latent variables as indicators. A common
+application of this approach is in longitudinal research with repeated
+measures of a construct. Examples of this approach include *growth
+mixture models* (GMM) and *latent class growth analyses* (LCGA)
+[@jung_introduction_2008]. These techniques estimate a latent growth
+model to describe grouped trajectories over time. The indicators of
+class membership in a GMM are the intercepts, and variances, and
+covariances of the latent growth variables that describe individual
+trajectories on the observed variables. For linear trajectories, these
+latent variables are an intercept and slope. The free variances allow
+trajectories to vary within classes. The LCGA differs from the GMM in
+that it assumes homogeneity of trajectories within classes. This is
+achieved by restricting the (co)variances of the growth parameters to
+zero. Any variance in the indicators not explained by the class-specific
 latent trajectories is assumed to be error variance.
@@
 
-Although LCA is often discussed as an exploratory analysis technique,
-it can also be used in a confirmatory manner.
-Given that LCA is similar to confirmatory factor analysis,
-it can be used to similar ends:
-To assess whether a theoretical measurement model holds.
-This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
-For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
-LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
-Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA [@donnellanResilientOvercontrolledUndercontrolled2010].
-
-One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices (see Model Fit Indices).
-Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
-Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
-either qualitatively or quantitatively, using significance tests for pre-specified values (see Inference).
-
-Appendix A is a tutorial to confirmatory LCA in `tidySEM`,
-based on a recent study that tested the theory that identities of Belgian adolescents with migration backgrounds could be summarized as those relating to their heritage, national and regional identity [@maene_perceived_2022].
+Although LCA is often discussed as an exploratory analysis technique, it
+can also be used in a confirmatory manner. Given that LCA is similar to
+confirmatory factor analysis, it can be used to similar ends: To assess
+whether a theoretical measurement model holds. This use case is relevant
+when testing a theory that postulates the existence of a categorical
+latent variable. For example, *identity status theory* posits that, at
+any given point in time, adolescents reside in one of four identity
+statuses [@marcia_development_1966]. LCA can be used to identify these
+four statuses based on observed indicators (e.g., self-reported identity
+exploration and commitment)
+[@luyckxDevelopmentalTypologiesIdentity2008]. Similarly, *personality
+type theory* states that dimensional differences in personality can
+largely be explained by an undercontrolled, overcontrolled, and
+resilient type, which could also be restored using LCA
+[@donnellanResilientOvercontrolledUndercontrolled2010].
+
+One challenge is that, unlike confirmatory factor analysis, LCA does not
+provide absolute fit indices (see Model Fit Indices). Thus, to test a
+theory, one can ascertain that the data are better described by the
+number of classes dictated by theory than by a different number of
+classes. Furthermore, it would be possible to test whether the observed
+pattern of class-specific means corresponds to a hypothesized pattern;
+either qualitatively or quantitatively, using significance tests for
+pre-specified values (see Inference).
+
+Appendix A is a tutorial to confirmatory LCA in `tidySEM`, based on a
+recent study that tested the theory that identities of Belgian
+adolescents with migration backgrounds could be summarized as those
+relating to their heritage, national and regional identity
+[@maene_perceived_2022].
 
@@
 
-Another use case is to explore whether a population is comprised of latent subpopulations, and if so, how many.
-The primary focus in this use case is on class enumeration.
-For example, 
+Another use case is to explore whether a population is comprised of
+latent subpopulations, and if so, how many. The primary focus in this
+use case is on class enumeration. For example,
 <!-- researchers suspected the existence of different risk profiles of substance use, sexual behavior, and mental health outcomes, and used LCA to determine how many such risk profiles could be distinguished [@hopfer_social_2014].
 -->
-<!-- Another  -->
-one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
-and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
-[@gromatsky_characteristics_2022].
-Knowing that the population of suicidal service members is heterogeneous can inform future research, policy, and care provided.
-Appendix B offers a tutorial for exploratory LCA,
-based on the aforementioned study of the distribution of ocean microplastics [@alkema_maximizing_2022].
+<!-- Another  --> one study explored heterogeneity in the population of
+military service members and veterans with prior suicide attempts, and
+identified two latent profiles characterized by high versus low suicide
+risk based on self-reported suicidal ideation and information about
+prior attempted methods and severity [@gromatsky_characteristics_2022].
+Knowing that the population of suicidal service members is heterogeneous
+can inform future research, policy, and care provided. Appendix B offers
+a tutorial for exploratory LCA, based on the aforementioned study of the
+distribution of ocean microplastics [@alkema_maximizing_2022].
 
@@
 
-LCA can also be used as a measurement model, similar to confirmatory factor analysis.
-This is useful when class membership is measured imperfectly by several indicators:
-LCA can partial out measurement error and restore most likely class membership.
-For example, one study found discrepancies between self-reported employment status and official register data [@pavlopoulos_measuring_2015]. 
-While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
-A mixture model was used to estimate employment status, correcting for measurement error, from both  self-report and registry data.
-The study investigated random and systematic measurement error in both indicators,
-which reveal strengths and weaknesses of both types of assessment.
-
-This example also illustrates that LCA allows us to evaluate the reliability of different indicators [@geiser_is_2014].
+LCA can also be used as a measurement model, similar to confirmatory
+factor analysis. This is useful when class membership is measured
+imperfectly by several indicators: LCA can partial out measurement error
+and restore most likely class membership. For example, one study found
+discrepancies between self-reported employment status and official
+register data [@pavlopoulos_measuring_2015]. While it might be tempting
+to assume that register data is free of error, this assumption was shown
+to be false. A mixture model was used to estimate employment status,
+correcting for measurement error, from both self-report and registry
+data. The study investigated random and systematic measurement error in
+both indicators, which reveal strengths and weaknesses of both types of
+assessment.
+
+This example also illustrates that LCA allows us to evaluate the
+reliability of different indicators [@geiser_is_2014].
 <!-- High quality indicators will be strongly related to the -->
-<!-- latent variable and will lead to good class separation. -->
-A continuous indicator that discriminates well between classes will show large differences in class-specific means;
-an ordinal indicator will show very high or very low conditional response probabilities for some classes but not for others.
-This information can be used to select the indicators that are most diagnostic of class membership.
-For an example of LCA with a focus on differential item functioning,
-consider a study on university admission tests of Saudi Arabian students [@tsaousis_measurement_2020].
-<!-- LCA revealed three -->
+<!-- latent variable and will lead to good class separation. --> A
+continuous indicator that discriminates well between classes will show
+large differences in class-specific means; an ordinal indicator will
+show very high or very low conditional response probabilities for some
+classes but not for others. This information can be used to select the
+indicators that are most diagnostic of class membership. For an example
+of LCA with a focus on differential item functioning, consider a study
+on university admission tests of Saudi Arabian students
+[@tsaousis_measurement_2020]. <!-- LCA revealed three -->
 <!-- latent classes: high-, average-, and low-scorers. Subsequently, the authors -->
@@
 
-LCA can also be used to reduce high-dimensional data
-to just a few prototypes.
-Factor analysis is also sometimes used for dimension reduction,
-but only accounts for linear covariance between indicators,
-whereas LCA can accommodate complex - but discrete - patterns.
-For example, imagine that there exist two indicators - $x$ and $y$, and some individuals score high on both, some score low on both, and some score high on $x$ but low on $y$.
-Factor analysis would only be able to capture the differences between the first two groups well; LCA captures differences between all three.
-Note that this is a pragmatic use of LCA;
-there is no theory about the existence of a categorical latent variable (cf. Testing theory).
-Consequently, the class solution should be treated as descriptive of the sample,
-and not be "reified" - treated as evidence of the existence of a categorical latent variable.
-
-The use of LCA for dimension reduction is common in longitudinal research,
-where one developmental process is summarized using latent class growth analysis (LCGA),
-and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
-For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy [@van_lissa_divergence_2015].
-These groups were then used as moderator of a latent growth analysis of adolescent- and parent-reported conflict.
-Results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than the other groups,
-and that the low empathy group had more conflict with parents than the other groups.
-Note that both these key findings are non-linear;
-modeling them in a single-group model would be difficult.
-LCGA greatly simplifies model specification and interpretation,
-albeit at a cost of some loss of information and, potentially, generalizability.
-Appendix C is a tutorial on LCGA,
-based on ongoing research on trajectories of depressive symptoms in military service members after deployment.
+LCA can also be used to reduce high-dimensional data to just a few
+prototypes. Factor analysis is also sometimes used for dimension
+reduction, but only accounts for linear covariance between indicators,
+whereas LCA can accommodate complex - but discrete - patterns. For
+example, imagine that there exist two indicators - $x$ and $y$, and some
+individuals score high on both, some score low on both, and some score
+high on $x$ but low on $y$. Factor analysis would only be able to
+capture the differences between the first two groups well; LCA captures
+differences between all three. Note that this is a pragmatic use of LCA;
+there is no theory about the existence of a categorical latent variable
+(cf. Testing theory). Consequently, the class solution should be treated
+as descriptive of the sample, and not be "reified" - treated as evidence
+of the existence of a categorical latent variable.
+
+The use of LCA for dimension reduction is common in longitudinal
+research, where one developmental process is summarized using latent
+class growth analysis (LCGA), and the resulting categories are then used
+as a moderator of another developmental process in a multi-group model.
+For example, one study conducted a LCGA of adolescent empathy
+development on two dimensions of empathy and found three groups: high,
+average, and low empathy [@van_lissa_divergence_2015]. These groups were
+then used as moderator of a latent growth analysis of adolescent- and
+parent-reported conflict. Results showed that the high-empathy group
+evidenced greater adolescent-parent agreement about the incidence of
+conflict than the other groups, and that the low empathy group had more
+conflict with parents than the other groups. Note that both these key
+findings are non-linear; modeling them in a single-group model would be
+difficult. LCGA greatly simplifies model specification and
+interpretation, albeit at a cost of some loss of information and,
+potentially, generalizability. Appendix C is a tutorial on LCGA, based
+on ongoing research on trajectories of depressive symptoms in military
+service members after deployment.
 <!-- Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean. -->
@@
 
-The need for dimension reduction can be especially pertinent when using ordinal indicators.
-For example, one study measured
-fifteen dichotomous health indicators among injured military personnel [@macgregor_symptom_2021]. 
-Combinatorics informs us that fifteen
+The need for dimension reduction can be especially pertinent when using
+ordinal indicators. For example, one study measured fifteen dichotomous
+health indicators among injured military personnel
+[@macgregor_symptom_2021]. Combinatorics informs us that fifteen
 dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
-However, LCA was able to reduce response patterns to five clinically meaningful latent classes.
+However, LCA was able to reduce response patterns to five clinically
+meaningful latent classes.
 
@@
 
-Another use case is to estimate individuals' unobserved class membership based on observed indicators.
-The shoe size example in Figure \@ref(fig:shoedens)
-is a rudimentary illustration of this application.
-In applied research, LCA is often used for classification in clinical contexts.
-For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough [@baughman_mixture_2006].
-Another study used LCA to not only classify the sample used to estimate the model, but also future cases [REF Zegwaard et al, in prep].
-The study identified four types of care providers among those who supported close kin with mental health problems.
-Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers, to provide more targeted support.
-To obtain predicted class membership based on an existing LCA model,
-the authors computed the likelihood of data from a new participant under the multivariate normal distributions estimated within all latent classes.
+Another use case is to estimate individuals' unobserved class membership
+based on observed indicators. The shoe size example in Figure
+\@ref(fig:shoedens) is a rudimentary illustration of this application.
+In applied research, LCA is often used for classification in clinical
+contexts. For example, LCA was used to determine clinical cut-off scores
+for diagnosing whooping cough [@baughman_mixture_2006]. Another study
+used LCA to not only classify the sample used to estimate the model, but
+also future cases [REF Zegwaard et al, in prep]. The study identified
+four types of care providers among those who supported close kin with
+mental health problems. Importantly, the LCA model was embedded in an
+interactive app that could be used by healthcare professionals to
+determine the most likely class membership for new care providers, to
+provide more targeted support. To obtain predicted class membership
+based on an existing LCA model, the authors computed the likelihood of
+data from a new participant under the multivariate normal distributions
+estimated within all latent classes.
 <!-- DA: this study is not in zotero yet, I lack information (title, authors, etc.) to enter it, @Caspar could you enter this to Zotero and adjust the in-text citation-->
-These applications of LCA aid the clinical decision making process,
-and help decide whether additional support is warranted,
-or what type of intervention is appropriate.
+These applications of LCA aid the clinical decision making process, and
+help decide whether additional support is warranted, or what type of
+intervention is appropriate.
 
@@
 
-Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughmanMixtureModelFramework2007].
-These models estimate one class to explain the "true zeros",
-and another class with a different distribution to explain all oter values, which can include zero.
-Note that this LCA has different models for both classes,
-and thus falls beyond the scope of this paper.
+Finally, LCA can be used to deal with data which violate certain
+assumptions. For example, zero-inflated and hurdle models are special
+cases of LCA for dealing data with a very high proportion of zeros
+(extreme right skew) [@baughmanMixtureModelFramework2007]. These models
+estimate one class to explain the "true zeros", and another class with a
+different distribution to explain all other values, which can include
+zero. Note that this LCA has different models for both classes, and thus
+falls beyond the scope of this paper.
 <!-- CJ: This has been said elsewhere: -->
@@
 <!-- CJ: This example is not perfect, we need another example still.. -->
-For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
-A mixture of two normal distributions with distinct variances described the observed distribution better.
+For instance, the aforementioned study on marine plastic noted that many
+prior studies had treated the distribution of particles as
+(multivariate) normal, despite clear deviations from normality
+[@alkema_maximizing_2022]. A mixture of two normal distributions with
+distinct variances described the observed distribution better.
 <!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical.
@@
 <!-- ### Controlling for covariates -->
+
 <!-- CJ: I'm not sure we should call this a use case... in principle, -->
+
 <!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->
+
 <!-- An extension of LCA is that containing covariates which can be used to -->
+
 <!-- predict class membership. In this approach, we not only model the latent -->
+
 <!-- class variable based on indicators, but we also relate the class -->
+
 <!-- membership to other explanatory variables (@vermunt_latent_2017). An -->
+
 <!-- example of using covariates comes from @nozadi_moderating_2016 who -->
+
 <!-- applied LCA to identify the probability of children's membership to an -->
+
 <!-- anxiety class. The authors tested several covariates including -->
+
 <!-- children's age, sex, and accuracy scores. Age and sex were not found to -->
+
 <!-- be related to the children's latent class membership, hence these -->
+
 <!-- covariates were excluded from the analysis. Accuracy scores, however, -->
+
 <!-- were related to probabilities of being in anxiety and attention-anxiety -->
+
 <!-- classes. Therefore this covariate was kept as a valuable predictor. -->
+
 <!-- Another example comes from @van_lissa_divergence_2015 who included sex  -->
+
 <!-- as a predictor of the developmental trajectories and class membership in  -->
+
 <!-- the previously discussed adolescent empathy development study. The rationale -->
+
 <!-- was that adding sex as a covariate would account for known sex differences  -->
+
 <!-- in empathy development, and lead to a model with lower bias. -->
@@
 <!-- ### Distal outcomes -->
+
 <!-- CJ: I'm not sure we should call this a use case... in principle, -->
+
 <!-- any time you're doing some kind of LCA you may want to predict group membership or predict something from group membership... -->
+
 <!-- When our interest is the prediction of one or more outcomes, LCA can be used  -->
+
 <!-- to construct latent classes as categorical predictors. @lanza_latent_2013  -->
+
 <!-- demonstrated how LCA can be used to classify adolescents into depression  -->
+
 <!-- classes, and subsequently these classes can be used to predict smoking, grades, -->
+
 <!-- and delinquency. The study showed that the outcomes predicted by class  -->
+
 <!-- membership can be binary (regular smoking), continuous (grades) or  -->
+
 <!-- count (delinquency). In the previously discussed study of Belgian adolescent -->
+
 <!-- identities, @maene_perceived_2022 first constructed the latent variable -->
+
 <!-- representing different multiple identification strategies. Subsequently,  -->
+
 <!-- the most likely class membership based on posterior class probability was -->
+
 <!-- used as an "observed" categorical variable for prediction of depressive symptoms. -->
@@
 \@ref(tab:checkest), are rooted in existing recommendations for best
-practices for estimating specific subtypes of LCA,
-including latent
+practices for estimating specific subtypes of LCA, including latent
 class growth analysis [@van_de_schoot_grolts-checklist_2017] and latent
 class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
-These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.
+These were generalized to be relevant to all types of LCA, and updated
+to current best practices, as explained below.
 <!-- In order to aid researchers working with latent trajectory models, -->
@@
 patterns and violations of assumptions that had not been considered
-prior to data collection.
-The function `descriptives(data)` provides extensive descriptive statistics,
-including the number of unique values,
-variance of continuous and categorical variables,
-missingness, skew and kurtosis (peakedness of the distribution). 
-However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
-Plots convey additional information; specifically, density plots for continuous variables, and bar charts
-for categorical ones.
-Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.
-
-Particular points of attention include measurement level of the indicators.
-Recall that `tidySEM` can estimate mixture models with either numeric (type `numeric` and `integer`) or `ordered` indicators.
-The column `type` should thus only contain these types.
-Also check that the number of `unique` values match the intended measurement level.
-Numeric variables with few unique values are typically better modeled as ordinal.
-Our experience consulting on LCA methods
-and moderating the `tidyLPA` Google group suggest that the
-misapplication of mixture models to ordinal (e.g., Likert-type)
-indicators is the most common source of user error. Whereas it has been
-argued that some parametric methods are robust when scales with 7+
-response categories are treated as continuous [e.g., @norman_likert_2010],
-it is very unlikely
-that an ordinal variable can be modeled as a *mixture* of multiple
-normal distributions.
-To illustrate the problem, consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
-Each class-specific mean would describe a single response
-category,
-and a class-specific variance component would be empirically underidentified.
-In sum, variables with few unique values might not be suitable for Gaussian mixture modeling;
-LCA with ordinal indicators is more appropriate.
-<!-- Descriptive statistics and -->
+prior to data collection. The function `descriptives(data)` provides
+extensive descriptive statistics, including the number of unique values,
+variance of continuous and categorical variables, missingness, skew and
+kurtosis (peakedness of the distribution). However, sample-level
+descriptive statistics are of limited value when the goal of a study is
+to identify subsamples using LCA. Plots convey additional information;
+specifically, density plots for continuous variables, and bar charts for
+categorical ones. Density plots can also aid in the choice of the number
+of classes, as explained in the section on visualization.
+
+Particular points of attention include measurement level of the
+indicators. Recall that `tidySEM` can estimate mixture models with
+either numeric (type `numeric` and `integer`) or `ordered` indicators.
+The column `type` should thus only contain these types. Also check that
+the number of `unique` values match the intended measurement level.
+Numeric variables with few unique values are typically better modeled as
+ordinal. Our experience consulting on LCA methods and moderating the
+`tidyLPA` Google group suggest that the misapplication of mixture models
+to ordinal (e.g., Likert-type) indicators is the most common source of
+user error. Whereas it has been argued that some parametric methods are
+robust when scales with 7+ response categories are treated as continuous
+[e.g., @norman_likert_2010], it is very unlikely that an ordinal
+variable can be modeled as a *mixture* of multiple normal distributions.
+To illustrate the problem, consider what happens when estimating a
+5-class solution on a single 5-point Likert scale indicator. Each
+class-specific mean would describe a single response category, and a
+class-specific variance component would be empirically underidentified.
+In sum, variables with few unique values might not be suitable for
+Gaussian mixture modeling; LCA with ordinal indicators is more
+appropriate. <!-- Descriptive statistics and -->
 <!-- plots can be relegated to online supplements, provided that these are -->
@@
 <!-- Any variables that of class `character` or `factor` must be recoded to ordinal. -->
+
 <!-- Nominal variables must be recoded to multiple ordinal dummy variables, e.g., using the function `mx_dummies()`.  -->
-LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
-Numeric variables should have an interval measurement level,
-such that it is sensible to estimate means and (co)variances.
-For the remaining three measurement levels,
-some preprocessing is required.
-The exact data types accepted by `tidySEM` at the time of writing are `numeric` and `ordered`.
-To correctly specify ordered variables
-such as Likert scales,
-use the function `OpenMx::mxFactor()`.
-Binary variables,
-such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
-Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using `mx_dummies()`.
-
-A second consideration in data preprocessing is that models with indicators on different scales may present with convergence issues.
-The reason for this is that the parameter space is high-dimensional with many potential local optima.
-If one indicator is on a much larger scale than another, the parameter space may be more extensively explored for that parameter.
-The problem can be resolved by transforming the indicators to place them on (roughly) similar scales.
-The most common transformation is standardization, which involves subtracting the mean of each variable from each value and dividing by the standard deviation.
+
+LCA can accommodate variables of different measurement levels:
+continuous, binary, nominal, and ordinal. Numeric variables should have
+an interval measurement level, such that it is sensible to estimate
+means and (co)variances. For the remaining three measurement levels,
+some preprocessing is required. The exact data types accepted by
+`tidySEM` at the time of writing are `numeric` and `ordered`. To
+correctly specify ordered variables such as Likert scales, use the
+function `OpenMx::mxFactor()`. Binary variables, such as presence or
+absence of some symptom, are also treated as ordinal, with one category
+coded as zero and the other as one. Nominal variables are converted to
+binary indicators of group membership via dummy coding, for example,
+using `mx_dummies()`.
+
+A second consideration in data preprocessing is that models with
+indicators on different scales may present with convergence issues. The
+reason for this is that the parameter space is high-dimensional with
+many potential local optima. If one indicator is on a much larger scale
+than another, the parameter space may be more extensively explored for
+that parameter. The problem can be resolved by transforming the
+indicators to place them on (roughly) similar scales. The most common
+transformation is standardization, which involves subtracting the mean
+of each variable from each value and dividing by the standard deviation.
 In R, this transformation is applied using the `scale()` function.
-Standardization retains all available information,
-and model parameters can be transformed back to the original scale of the indicators by reversing the transformation.
-The transformed variable has a mean of zero and a standard deviation of one.
-Note that standardization is typically not necessary if indicators are already on the same, or similar, scales.
-
-A third consideration in preprocessing relates to the univariate distributions of indicators.
-For example, the `plas_depression` data show extreme right skew in depressive symptoms:
-most respondents report few symptoms,
-and few respondents report many symptoms.
-Similarly, the `alkema_microplastics` data are extremely right-skewed: Very small plastic fragments abound in the oceans, and large pieces are rare.
-These distributional characteristics must be taken into account either in preprocessing, or model specification.
-In preprocessing, the data can be transformed to reduce skew.
-After reducing skew, a simpler LCA model may fit the data well.
-Alternatively,
-one can specify a model that accommodates the distributional idiosyncrasies.
-For example, allowing free variances across classes might result in one class with a near-zero mean and small variance to account for the fact that most respondents have few depressive symptoms,
-and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.
+Standardization retains all available information, and model parameters
+can be transformed back to the original scale of the indicators by
+reversing the transformation. The transformed variable has a mean of
+zero and a standard deviation of one. Note that standardization is
+typically not necessary if indicators are already on the same, or
+similar, scales.
+
+A third consideration in preprocessing relates to the univariate
+distributions of indicators. For example, the `plas_depression` data
+show extreme right skew in depressive symptoms: most respondents report
+few symptoms, and few respondents report many symptoms. Similarly, the
+`alkema_microplastics` data are extremely right-skewed: Very small
+plastic fragments abound in the oceans, and large pieces are rare. These
+distributional characteristics must be taken into account either in
+preprocessing, or model specification. In preprocessing, the data can be
+transformed to reduce skew. After reducing skew, a simpler LCA model may
+fit the data well. Alternatively, one can specify a model that
+accommodates the distributional idiosyncrasies. For example, allowing
+free variances across classes might result in one class with a near-zero
+mean and small variance to account for the fact that most respondents
+have few depressive symptoms, and one class with an elevated mean and
+large variance to account for all respondents with non-zero symptoms.
 
@@
 missing data and reporting how missingness was handled
-[@van_de_schoot_grolts-checklist_2017].
-While we concur that investigating missingness is due diligence,
-it is important to emphasize that missingness is adequately handled by default in `OpenMx`, which is the computational backend of `tidySEM`.
-Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
-information without imputing missing values [@endersRelativePerformanceFull2001].
-FIML is a best-practice solution for handling missing data;
-at least on par with multiple imputation [@lee_comparison_2021].
-Multiple imputation is less suitable to LCA for two reasons. First,
-because LCA methods are often computationally intensive,
-and conducting them on several imputed datasets multiplies the load.
-Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population.
-
-Three types of missingness have
-been distinguished in the literature [@rubin_inference_1976; 
-@enders_applied_2022]: random missingness (MCAR);
-missingness contingent on *observed* variables that can be known and controlled for (MAR);
-and missingness related to unobserved variables,
-which can be neither known nor controlled for (MNAR).
-A so-called "MCAR" test is often reported [@jamshidian_tests_2010],
-but note that this name is somewhat misleading,
-as a significant test indicates that missingness is MAR,
-but a non-significant test statistic does not distinguish between MCAR or MNAR.
-As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality [@little_test_1988].
-This assumption is tenuous in the context of LCA.
-A non-parametric MCAR test may be more suitable [@jamshidian_tests_2010].
-For this tutorial,
-the lead author has implemented this test in the `mice` package as `mice::mcar()`.
-Our recommendation is to inspect and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
-and report the `mice::mcar()` test.
-As FIML estimation assumes that missingness is either MCAR or MAR,
-one would proceed with FIML regardless of the outcome.
-
-One minor concern is that `tidySEM` uses a different algorithm to determine starting values for complete data (K-means) and for missing data (hierarchical clustering).
-This should not be a problem, however, as the global optimizer used to estimate the model should be independent of starting values [@coranaMinimizingMultimodalFunctions1987].
-A second concern is that FIML is not suitable for cases with complete missing data.
-If such cases must be included, their values must be imputed first.
+[@van_de_schoot_grolts-checklist_2017]. While we concur that
+investigating missingness is due diligence, it is important to emphasize
+that missingness is adequately handled by default in `OpenMx`, which is
+the computational backend of `tidySEM`. Its Full Information Maximum
+Likelihood (FIML) estimator makes use of all available information
+without imputing missing values, by estimating the model based on all
+available information for each subject
+[@endersRelativePerformanceFull2001; @enders_applied_2022]. FIML is a
+best-practice solution for handling missing data; at least on par with
+multiple imputation [@lee_comparison_2021]. Multiple imputation is less
+suitable to LCA for two reasons. First, because LCA methods are often
+computationally intensive, and conducting them on several imputed
+datasets multiplies the load. Second, because many multiple imputation
+methods assume normality; an assumption likely to be violated when data
+come from a heterogeneous population.
+
+Three types of missingness have been distinguished in the literature
+[@rubin_inference_1976; @enders_applied_2022]: random missingness
+(MCAR); missingness contingent on *observed* variables that can be known
+and controlled for (MAR); and missingness related to unobserved
+variables, which can be neither known nor controlled for (MNAR). A
+so-called "MCAR" test is often reported [@jamshidian_tests_2010], but
+note that this name is somewhat misleading, as a significant test
+indicates that missingness is MAR, but a non-significant test statistic
+does not distinguish between MCAR or MNAR. As the classic MCAR test
+relies on the comparison of variances across groups with different
+patterns of missing data, it assumes normality [@little_test_1988]. This
+assumption is tenuous in the context of LCA. A non-parametric MCAR test
+may be more suitable [@jamshidian_tests_2010]. For this tutorial, the
+lead author has implemented this test in the `mice` package as
+`mice::mcar()`. Our recommendation is to inspect and report the
+proportion of missingness per variable (e.g., using
+`tidySEM::descriptives()`), and report the `mice::mcar()` test. As FIML
+estimation assumes that missingness is either MCAR or MAR, one would
+proceed with FIML regardless of the outcome.
+
+One minor concern is that `tidySEM` uses a different algorithm to
+determine starting values for complete data (K-means) and for missing
+data (hierarchical clustering). This should not be a problem, however,
+as the global optimizer used to estimate the model should be independent
+of starting values [@coranaMinimizingMultimodalFunctions1987]. A second
+concern is that FIML is not suitable for cases with complete missing
+data. If such cases must be included, their values must be imputed
+first.
 
@@
 
-In principle, LCA can be based on any parametric model.
-At present, `tidySEM` only facilitates estimation of models that can be specified in `lavaan` syntax.
-The available model parameters depend on the measurement level of the indicators.
-Continuous indicators can be parametrized in terms of means, variances, and covariances.
-LCGA and GMM have these same parameters,
-but with respect to the latent growth variables.
-Ordinal indicators can be parametrized differently;
-the default approach in `tidySEM` assumes that each ordinal indicator reflects an underlying standard normal distribution, $N(\mu = 0, \sigma = 1)$.
-The parameters are "thresholds" that correspond to quantiles of this distribution, estimated based on the proportion of individuals in each of the
-response categories of the indicator variable.
-For example, a binary
-indicator has a single threshold that distinguishes the two response
-categories. If responses are distributed 50/50, then the corresponding
-threshold would be $t_1 = 0.00$. If the responses are distributed 60/40,
-then the resulting threshold would be $t_1 = 0.25$.
-All of the aforementioned parameters can be freely estimated,
-or constrained across classes, or fixed (e.g., to zero).
-
-Prior literature emphasized the importance of considering alternative model specifications [@vandeschootGRoLTSChecklistGuidelinesReporting2017].
-What has remained underemphasized is that alternative model specifications should be approached differently for exploratory versus confirmatory analyses (see Use Cases).
-Most prior literature has emphasized exploratory LCA [e.g., @nylund_deciding_2007].
-In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
-A "final" model is then selected based on a combination of fit indices,
-significance tests,
-and interpretability (see Class Enumeration).
-
-Confirmatory LCA does not require such an extensive search over the model space.
-If hypothesized models have been specified a priori,
-and possibly even preregistered,
-it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
-Nonetheless, the theoretical model specification should still be compared to a few alternatives because LCA lacks objective fit indices.
-The fit of a theoretical model can thus only be assessed by comparison to other candidates.
-A one-class model is a sensible benchmark
-to test whether the theoretical model fits better than a model that assumes a homogeneous population.
-Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
-If the theoretical model has much better fit than other models,
-for example, based on IC weights,
-this provides evidence for the theory.
-If another model fits substantially better,
-this provides evidence against the theory.
-If fit differences are small, the theoretical model may still be preferred, as theory is a relevant consideration in model selection [@jungIntroductionLatentClass2008].
-The fact that the theoretical model received little differential support should of course be discussed.
+In principle, LCA can be based on any parametric model. At present,
+`tidySEM` only facilitates estimation of models that can be specified in
+`lavaan` syntax. The available model parameters depend on the
+measurement level of the indicators. Continuous indicators can be
+parametrized in terms of means, variances, and covariances. LCGA and GMM
+have these same parameters, but with respect to the latent growth
+variables. Ordinal indicators can be parametrized differently; the
+default approach in `tidySEM` assumes that each ordinal indicator
+reflects an underlying standard normal distribution,
+$N(\mu = 0, \sigma = 1)$. The parameters are "thresholds" that
+correspond to quantiles of this distribution, estimated based on the
+proportion of individuals in each of the response categories of the
+indicator variable. For example, a binary indicator has a single
+threshold that distinguishes the two response categories. If responses
+are distributed 50/50, then the corresponding threshold would be
+$t_1 = 0.00$. If the responses are distributed 60/40, then the resulting
+threshold would be $t_1 = 0.25$. All of the aforementioned parameters
+can be freely estimated, or constrained across classes, or fixed (e.g.,
+to zero).
+
+Prior literature emphasized the importance of considering alternative
+model specifications
+[@vandeschootGRoLTSChecklistGuidelinesReporting2017]. What has remained
+underemphasized is that alternative model specifications should be
+approached differently for exploratory versus confirmatory analyses (see
+Use Cases). Most prior literature has emphasized exploratory LCA [e.g.,
+@nylund_deciding_2007]. In this context, model specification typically
+consists of an exhaustive grid search along several model specifications
+and varying numbers of classes. A "final" model is then selected based
+on a combination of fit indices, significance tests, and
+interpretability (see Class Enumeration).
+
+Confirmatory LCA does not require such an extensive search over the
+model space. If hypothesized models have been specified a priori, and
+possibly even preregistered, it makes sense to focus on these models,
+rather than proceed in a purely data-driven manner. Nonetheless, the
+theoretical model specification should still be compared to a few
+alternatives because LCA lacks objective fit indices. The fit of a
+theoretical model can thus only be assessed by comparison to other
+candidates. A one-class model is a sensible benchmark to test whether
+the theoretical model fits better than a model that assumes a
+homogeneous population. Other candidates for comparison include
+competing theoretical models and those with a few more or less classes
+than the hypothesized number. If the theoretical model has much better
+fit than other models, for example, based on IC weights, this provides
+evidence for the theory. If another model fits substantially better,
+this provides evidence against the theory. If fit differences are small,
+the theoretical model may still be preferred, as theory is a relevant
+consideration in model selection [@jungIntroductionLatentClass2008]. The
+fact that the theoretical model received little differential support
+should of course be discussed.
 <!-- This will ensure that the model accurately represents the data and provides reliable results. -->
@@
 <!-- In the general sense, the class-specific model can be any structural equation model. -->
+
 <!-- Its parameters may include means, (co)variances, regression coefficients, factor loadings, and thresholds for ordinal variables. -->
+
 <!-- All of these parameters can be freely estimated, -->
+
 <!-- constrained to be equal across classes, -->
+
 <!-- or fixed to a known value. -->
+
 <!-- The specific types LCA introduced in the Taxonomy all have different parameters. -->
+
 <!-- For example, LPA typically assumes class-specific means, -->
+
 <!-- equal variances across classes, and zero covariances (i.e., indicators are independent after accounting for latent class membership), see `mx_profiles()`. -->
+
 <!-- By contrast, LCA with ordinal indicators assumes free thresholds across classes, -->
+
 <!-- where each threshold reflects a proportion of responses within a particular response category, see `mx_lca()`. -->
+
 <!-- Like LPA, LCGA and GMM typically have free means across classes, -->
+
 <!-- but with respect to the latent growth variables instead of the observed variables, see `mx_growth_mixture()`. -->
+
 <!-- The distinction between LCGA and GMM is whether variances of the latent growth variables are constrained to zero, or freely estimated (either across or within classes). -->
+
 <!-- either using `lavaan` model syntax or using `OpenMx` RAM specification, -->
+
 <!-- and supplied to the general function `mx_mixture()`. -->
+
 <!-- The resulting model may have -->
+
 <!-- HIER -->
+
 <!-- For latent profile analysis, the function -->
+
 <!-- `tidySEM::mx_profiles(classes, variances, covariances)` largely -->
+
 <!-- automates this process. The argument `classes` indicates which class -->
+
 <!-- solutions should be estimated (e.g., 1 through 6). The argument -->
+
 <!-- `variances` specifies whether variances should be `"equal"` or -->
+
 <!-- `"varying"` across classes. The argument `covariances` specifies whether -->
+
 <!-- covariances should be constrained to `"zero"`, `"equal"` or `"varying"` -->
+
 <!-- across classes. The means are free to vary across classes by default, -->
+
 <!-- although the more general function `tidySEM::mx_mixture()` could be used -->
+
 <!-- to circumvent this. After all models have been estimated, the function -->
+
 <!-- `tidySEM::table_fit()` can be used to obtain a model fit table suitable -->
+
 <!-- for determining the optimal model according to best practices. Note, -->
+
 <!-- however, that this table does not include the bootstrapped likelihood -->
+
 <!-- ratio test (BLRT) by default, because this test is very computationally -->
+
 <!-- expensive. It is recommended to use the function `tidySEM::BLRT()` to -->
+
 <!-- compare a shortlist of likely candidate models based on other fit -->
+
 <!-- indices. More on fit indices can be found in the Model Fit Indices -->
+
 <!-- subsection of this paper. -->
 
-A second consideration in model specification is the risk of overfitting [@hastieElementsStatisticalLearning2009].
-Overfitting occurs when a model has many parameters relative to the number of observations,
-causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
-As explained before, the number of parameters in LCA scales with the number of estimated classes.
-Consequently, LCA with many classes have a potentially very high number of parameters.
-In Bayesian mixture modeling,
-it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased [@rousseauAsymptoticBehaviourPosterior2011].
-Although it is not known whether frequentist LCA exhibits the same behavior,
-low class counts are still an indicator of potential overfitting.
-The section on Class Enumeration addresses ways to guard against overfitting.
-
-Models can be specified in `tidySEM` in different ways.
-The wrapper functions `mx_profiles()` and `mx_lca()` construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
-The function `mx_profiles()` has two arguments to free elements of the covariance matrix: `variances` and `covariances`.
-By default, `mx_profiles()` fixes covariances to zero, and thus performs latent profile analysis.
-Finally, function `mx_mixture()` accepts a more flexible `model` argument, which can be a character string specifying a model in `lavaan` syntax,
-which is translated to `OpenMx` models via the function `as_ram()`,
-or a list of `OpenMx` models.
+A second consideration in model specification is the risk of overfitting
+[@hastieElementsStatisticalLearning2009]. Overfitting occurs when a
+model has many parameters relative to the number of observations,
+causing it to capture idiosyncratic noise in the sample and limiting its
+generalizability to new samples. As explained before, the number of
+parameters in LCA scales with the number of estimated classes.
+Consequently, LCA with many classes have a potentially very high number
+of parameters. In Bayesian mixture modeling, it was observed that if the
+estimated number of classes exceeded the true number, the number of
+observations assigned to these classes tended towards zero as sample
+size increased [@rousseauAsymptoticBehaviourPosterior2011]. Although it
+is not known whether frequentist LCA exhibits the same behavior, low
+class counts are still an indicator of potential overfitting. The
+section on Class Enumeration addresses ways to guard against
+overfitting.
+
+Models can be specified in `tidySEM` in different ways. The wrapper
+functions `mx_profiles()` and `mx_lca()` construct LCA models for
+continuous and ordinal indicators respectively, without needing to
+specify model syntax. The function `mx_profiles()` has two arguments to
+free elements of the covariance matrix: `variances` and `covariances`.
+By default, `mx_profiles()` fixes covariances to zero, and thus performs
+latent profile analysis. Finally, function `mx_mixture()` accepts a more
+flexible `model` argument, which can be a character string specifying a
+model in `lavaan` syntax, which is translated to `OpenMx` models via the
+function `as_ram()`, or a list of `OpenMx` models.
 
@@
 LCA parameters and model fit statistics can be estimated in different
-ways.
-The choice of estimator depends on the presence of missing
-values, sample size, number of indicators, and available computational
-resources [@weller_latent_2020].
-LCA in `tidySEM` uses maximum likelihood estimation (ML), which can be conceptually understood as follows:
-Imagine we are estimating two
-parameters, e.g. the class-specific means $\mu_c$ on a continuous indicator for two classes
-(ignoring the variance for now).
-ML attempts to find a combination of values for these two parameters that minimizes minus two times the log likelihood ($-2LL$) of the data.
+ways. The choice of estimator depends on the presence of missing values,
+sample size, number of indicators, and available computational resources
+[@weller_latent_2020]. LCA in `tidySEM` uses maximum likelihood
+estimation (ML), which can be conceptually understood as follows:
+Imagine we are estimating two parameters, e.g. the class-specific means
+$\mu_c$ on a continuous indicator for two classes (ignoring the variance
+for now). ML attempts to find a combination of values for these two
+parameters that minimizes minus two times the log likelihood ($-2LL$) of
+the data.
 <!-- The $-2LL$ is asymptotically $\chi^2$ distributed if the null-hypothesis is true, and can thus be used for model fit tests [@wilksLargeSampleDistributionLikelihood1938]. -->
-To understand the ML optimization procedure,
-imagine a three-dimensional landscape: 
-The X and Y dimensions represent potential values of the class-specific means, $X = \mu_1$ and $Y = \mu_2$, and the Z-dimension is determined by $Z = -2LL$. 
-The optimizer must find the
-deepest "valley" in this landscape,
-which reflects the combination of $\mu_1$ and $\mu_2$ that minimizes the $-2LL$.
-The ML
-optimizer behaves somewhat like a marble dropped in this landscape. It
-is dropped at some random point in space, and rolls into the nearest valley (a combination of values for $\mu_1$ and $\mu_2$ that has a low $-2LL$).
-The challenge is that LCA models are often highly parametrized.
-This means that the landscape is not three-dimensional, but (very) high-dimensional,
-with many valleys and sparse data to provide information about which valley is deepest.
-If the marble rolls into one of these valleys,
-it will settle there and not climb out again.
-The estimator has "converged".
-
-A unique challenge in LCA is that not only model parameters,
-but also class membership must be estimated.
-Parameters in all classes are estimated from the same dataset,
-but for each class,
-observations are weighted by their probability of being in that class - which is in turn computed based on the class-specific parameters.
-The resulting chicken-and-egg problem is solved by the expectation-maximization (EM) algorithm:
-It alternates between estimating model parameters,
+To understand the ML optimization procedure, imagine a three-dimensional
+landscape: The X and Y dimensions represent potential values of the
+class-specific means, $X = \mu_1$ and $Y = \mu_2$, and the Z-dimension
+is determined by $Z = -2LL$. The optimizer must find the deepest
+"valley" in this landscape, which reflects the combination of $\mu_1$
+and $\mu_2$ that minimizes the $-2LL$. The ML optimizer behaves somewhat
+like a marble dropped in this landscape. It is dropped at some random
+point in space, and rolls into the nearest valley (a combination of
+values for $\mu_1$ and $\mu_2$ that has a low $-2LL$). The challenge is
+that LCA models are often highly parametrized. This means that the
+landscape is not three-dimensional, but (very) high-dimensional, with
+many valleys and sparse data to provide information about which valley
+is deepest. If the marble rolls into one of these valleys, it will
+settle there and not climb out again. The estimator has "converged".
+
+A unique challenge in LCA is that not only model parameters, but also
+class membership must be estimated. Parameters in all classes are
+estimated from the same dataset, but for each class, observations are
+weighted by their probability of being in that class - which is in turn
+computed based on the class-specific parameters. The resulting
+chicken-and-egg problem is solved by the expectation-maximization (EM)
+algorithm: It alternates between estimating model parameters,
 calculating posterior class probabilities from those parameters,
-re-estimating model parameters using those new class probabilities as weights, and so on, until both have converged.
-
-A risk with this approach is that the optimizer might get stuck in a shallower valley (a "local optimum"),
-and never discovers the deepest valley (the "global optimum", or best solution).
-One solution to this problem is to drop many marbles at random places,
-compare their final $-2LL$ values,
-choose the solution
-with the lowest $-2LL$, and make sure that several marbles replicated
-this solution.
-This is the "random starts" approach implemented in, e.g., Mplus.
+re-estimating model parameters using those new class probabilities as
+weights, and so on, until both have converged.
+
+A risk with this approach is that the optimizer might get stuck in a
+shallower valley (a "local optimum"), and never discovers the deepest
+valley (the "global optimum", or best solution). One solution to this
+problem is to drop many marbles at random places, compare their final
+$-2LL$ values, choose the solution with the lowest $-2LL$, and make sure
+that several marbles replicated this solution. This is the "random
+starts" approach implemented in, e.g., Mplus.
 
@@
 computationally expensive to run this many replications. Moreover,
-because the algorithm begins with random starting values,
-it is inefficient because many marbles are likely to start far away from any valley.
-Additionally, if all sets of starting values are close to a specific local optimum,
-there is a risk that the global optimum is never found.
-An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing [SA, @coranaMinimizingMultimodalFunctions1987].
-SA does not behave quite like the marbles of the random starts approach;
-instead, it iteratively considers some "destination" in the landscape,
-and compares its $-2LL$ to the current one.
-If the destination $-2LL$ is lower, 
-the estimator moves there.
-However, if the destination  $-2LL$ is *higher*,
-it still moves there occasionally, with a certain probability.
-Think of this as occasionally flicking the marble,
-to see if it rolls back into the current valley, or finds another,
-even deeper one.
-This property allows SA it to escape local optima and find the global optimum.
-
-Since SA is a global optimizer, it should be independent of starting values.
-With this in mind, it is efficient to start estimation from a "reasonable solution", instead of a random starting point.
-For example, if we assume that the different classes are likely to have different mean values on the indicators, 
-then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernackiChoosingStartingValues2003].
-By default, `tidySEM` derives starting values using K-means clustering [REF],
-computing the mixture model starting values by treating the K-means solution as multi-group model.
-Next, SA is used to find the global optimum solution.
-Finally, SA is followed up with a short run of the ML algorithm,
-as ML inherently produces an asymptotic
-covariance matrix for the parameters that can be used to compute
-standard errors.
-Note that these defaults can be manually overridden.
-
-In practice, models do not always converge.
-Reasons may include local optima,
-bad starting values,
-incorrect model specification (the model may be too complex for the data, or conversely, too simple to capture important patterns),
-or poor data quality (small data sets, or distributions that are difficult to model).
-Convergence problems may be indicated by errors and warnings from the estimator,
-or they may be suspected, for example,
-when results change if the analyses are reproduced.
-
-Nonconvergence may be due to bad model specification.
-Consider, for example, an LCA with multiple ordinal indicators.
-If certain combinations of responses never occur together,
-in general or within one of the classes,
-this can result in extreme conditional item probabilities (exactly 0 or 1).
-Such extreme values impede model convergence.
-This problem could be addressed by trying alternative model specifications (e.g., fewer classes),
-or by merging categories of the ordinal indicators with very few responses.
-Similarly problems occur with continuous indicators.
-For instance, consider a zero-inflated indicator and a two class LPA.
-To account for the excess zeroes, one class will have a mean of zero.
-It is difficult to estimate the standard deviation for this class, however, because all of its members have the same value.
-If standard deviations are constrained to be equal across classes,
-this model will likely not converge.
-This problem can be addressed by considering more appropriate model specifications (e.g., free class variances).
-
-Nonconvergence due to bad starting values
-can be overcome by manually specifying starting values:
-Specify the argument `run = FALSE` in the call to `mx_mixture()` and related functions.
-This returns the model with default starting values, which can be updated.
-Nonconvergence and local optima may also be overcome by re-estimating the model while permuting the starting values.
-The `OpenMx` package contains a family of functions for this purpose: `mxTryHard()` and `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
-When poor data quality is to blame, the only solution is to collect more, or higher quality, data.
+because the algorithm begins with random starting values, it is
+inefficient because many marbles are likely to start far away from any
+valley. Additionally, if all sets of starting values are close to a
+specific local optimum, there is a risk that the global optimum is never
+found. An alternative solution that overcomes these challenges is to use
+a global optimizer, like simulated annealing [SA,
+@coranaMinimizingMultimodalFunctions1987]. SA does not behave quite like
+the marbles of the random starts approach; instead, it iteratively
+considers some "destination" in the landscape, and compares its $-2LL$
+to the current one. If the destination $-2LL$ is lower, the estimator
+moves there. However, if the destination $-2LL$ is *higher*, it still
+moves there occasionally, with a certain probability. Think of this as
+occasionally flicking the marble, to see if it rolls back into the
+current valley, or finds another, even deeper one. This property allows
+SA it to escape local optima and find the global optimum.
+
+Since SA is a global optimizer, it should be independent of starting
+values. With this in mind, it is efficient to start estimation from a
+"reasonable solution", instead of a random starting point. For example,
+if we assume that the different classes are likely to have different
+mean values on the indicators, then a nonparametric clustering
+algorithm - like K-means or hierarchical clustering - can be used to
+determine these cluster centroids [see also
+@biernackiChoosingStartingValues2003]. By default, `tidySEM` derives
+starting values using K-means clustering [REF], computing the mixture
+model starting values by treating the K-means solution as multi-group
+model. Next, SA is used to find the global optimum solution. Finally, SA
+is followed up with a short run of the ML algorithm, as ML inherently
+produces an asymptotic covariance matrix for the parameters that can be
+used to compute standard errors. Note that these defaults can be
+manually overridden.
+
+In practice, models do not always converge. Reasons may include local
+optima, bad starting values, incorrect model specification (the model
+may be too complex for the data, or conversely, too simple to capture
+important patterns), or poor data quality (small data sets, or
+distributions that are difficult to model). Convergence problems may be
+indicated by errors and warnings from the estimator, or they may be
+suspected, for example, when results change if the analyses are
+reproduced.
+
+Nonconvergence may be due to bad model specification. Consider, for
+example, an LCA with multiple ordinal indicators. If certain
+combinations of responses never occur together, in general or within one
+of the classes, this can result in extreme conditional item
+probabilities (exactly 0 or 1). Such extreme values impede model
+convergence. This problem could be addressed by trying alternative model
+specifications (e.g., fewer classes), or by merging categories of the
+ordinal indicators with very few responses. Similarly problems occur
+with continuous indicators. For instance, consider a zero-inflated
+indicator and a two class LPA. To account for the excess zeroes, one
+class will have a mean of zero. It is difficult to estimate the standard
+deviation for this class, however, because all of its members have the
+same value. If standard deviations are constrained to be equal across
+classes, this model will likely not converge. This problem can be
+addressed by considering more appropriate model specifications (e.g.,
+free class variances).
+
+Nonconvergence due to bad starting values can be overcome by manually
+specifying starting values: Specify the argument `run = FALSE` in the
+call to `mx_mixture()` and related functions. This returns the model
+with default starting values, which can be updated. Nonconvergence and
+local optima may also be overcome by re-estimating the model while
+permuting the starting values. The `OpenMx` package contains a family of
+functions for this purpose: `mxTryHard()` and `mxTryHardWideSearch()`
+for continuous indicators, and `mxTryHardOrdinal()` for ordinal
+indicators. When poor data quality is to blame, the only solution is to
+collect more, or higher quality, data.
 
 <!-- A problem which can occur is that of inadmissible solutions. With binary -->
+
 <!-- indicators, LCA is modeling a cross-table with all the predictors. The -->
+
 <!-- problem with such cross-tables is that they will often contain empty -->
+
 <!-- cells, i.e. combinations of responses that never occur together. This -->
+
 <!-- problem is reflected by extreme conditional item probabilities (as in -->
+
 <!-- exactly 0 or 1). Such boundary parameter estimates could indicate that -->
+
 <!-- the solution is invalid [@geiser_data_2012]. Boundary parameter -->
+
 <!-- estimates can also happen with continuous indicators. For instance, if -->
+
 <!-- we have a zero-inflated normal distribution and a two class solution, -->
+
 <!-- one class might have the mean of zero and its standard deviation cannot -->
+
 <!-- be determined since there is little variance. This too could be a sign -->
+
 <!-- of an invalid solution, warn that too many classes were extracted, or -->
+
 <!-- indicate a local optimum [@geiser_data_2012]. -->
@@
 <!-- The total number of parameters scales with the number of estimated classes. -->
+
 <!-- Consequently, latent class analyses have a potentially very high number of parameters. -->
+
 <!-- As any of these parameters could be mis-specified, -->
+
 <!-- it is important to consider alternative model specifications. -->
+
 <!-- One way to obtain relative evidence in favor of different models is by using BIC weights, obtained by the function `tidySEM::ic_weights()` [@wagenmakersAICModelSelection2004]. -->
@@
 <!-- From sinhaPractitionerGuideLatent2021: -->
+
 <!-- Testing for number of classes.
@@
 -->
-Class enumeration refers to the process of determining the appropriate number of classes.
-It is approached differently for exploratory versus confirmatory LCA.
-In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes $K$. 
-The number $K$ may be chosen a-priori or dictated by the data.
-It is sensible to limit $K$ to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
-Above a certain number of classes,
-convergence problems typically arise, 
-or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
-In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
-the proportion of cases assigned to the excessive classes converges to zero as the sample size increases [@rousseauAsymptoticBehaviourPosterior2011].
-It is not known whether this also happens in frequentist LCA;
-nonetheless, near-zero class memberships should be taken into account when determining the maximum $K$.
-
-In confirmatory LCA,
-the set of models with different numbers of classes to be compared is typically smaller.
-Its purpose is to provide context for the relative fit of the theoretical model, or to compare multiple theoretical models.
-
-In class enumeration, the set of models is compared on several statistical and substantive criteria.
-In exploratory LCA, these criteria may be decisive in selecting a final model;
-in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
-Statistical criteria to consider in class enumeration include model fit indices and
+
+Class enumeration refers to the process of determining the appropriate
+number of classes. It is approached differently for exploratory versus
+confirmatory LCA. In exploratory LCA, the set of models typically
+consists of a sequence from 1 to the maximum number of classes $K$. The
+number $K$ may be chosen a-priori or dictated by the data. It is
+sensible to limit $K$ to the maximum number of classes that result in
+valid and interpretable solutions (also see Classification Diagnostics).
+Above a certain number of classes, convergence problems typically arise,
+or the proportion of cases assigned to the smallest classes may be low
+relative to the number of class-specific parameters. In Bayesian LCA, it
+has been observed that, when the estimated number of classes exceeds the
+true number, the proportion of cases assigned to the excessive classes
+converges to zero as the sample size increases
+[@rousseauAsymptoticBehaviourPosterior2011]. It is not known whether
+this also happens in frequentist LCA; nonetheless, near-zero class
+memberships should be taken into account when determining the maximum
+$K$.
+
+In confirmatory LCA, the set of models with different numbers of classes
+to be compared is typically smaller. Its purpose is to provide context
+for the relative fit of the theoretical model, or to compare multiple
+theoretical models.
+
+In class enumeration, the set of models is compared on several
+statistical and substantive criteria. In exploratory LCA, these criteria
+may be decisive in selecting a final model; in the confirmatory case,
+the burden of proof is reversed: there must be convincing evidence
+against the theoretical model in order to reject it. Statistical
+criteria to consider in class enumeration include model fit indices and
 likelihood ratio tests [@weller_latent_2020; @masyn_latent_2013].
-Substantive criteria include convergence and model identification,
-class separability,
-and interpretability.
-We address each of these criteria below.
+Substantive criteria include convergence and model identification, class
+separability, and interpretability. We address each of these criteria
+below.
 
 <!-- In exploratory LCA, a sequence of models is fitted to the data -->
+
 <!-- with each additional model estimating one more class than the previous -->
+
 <!-- model. These models are then compared and the best solution is selected -->
+
 <!-- as the final class solution. In some cases, prior theory can inform the -->
+
 <!-- researcher about the number of classes to expect. Even in such -->
+
 <!-- confirmatory LCA cases, it is nonetheless useful to know if the -->
+
 <!-- theoretical model is markedly better than those with differing numbers -->
+
 <!-- of classes. Therefore, it may always be useful to compare different -->
+
 <!-- class solutions. -->
@@
 
-As there are no absolute fit indices for LCA,
-information criteria are typically used to assess relative fit.
-Information criteria are computed from the $-2LL$,
-which is the most basic fit measure.
-Information criteria add a penalty for model complexity to the $-2LL$, and thus balance fit and complexity.
-The lower
-the value of an information criterion, the better the overall fit of the
-model.
-The original information criterion is the Akaike Information Criterion (AIC),
-computed as $-2LL+2p$, where $p$ is the number of parameters.
-The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
-such that complex models are penalized more in smaller samples.
-Finally,
-the so-called *sample size adjusted* BIC (saBIC) implements a different ad-hoc penalty for sample size,
-$\ln(\frac{n + 2}{24})$.
-Although a simulation study showed that the saBIC performed well for class enumeration,
-its penalty is not based on theory [REF].
-BIC also performs well and is theoretically substantiated;
-as such, it may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
-All three ICs are available in the output of `table_fit()`.
+As there are no absolute fit indices for LCA, information criteria are
+typically used to assess relative fit. Information criteria are computed
+from the $-2LL$, which is the most basic fit measure, testing the
+overall model fit if the model properly represents the data. Information
+criteria add a penalty for model complexity to the $-2LL$, and thus
+balance fit and complexity. The lower the value of an information
+criterion, the better the overall fit of the model. The original
+information criterion is the Akaike Information Criterion (AIC),
+computed as $-2LL+2p$, where $p$ is the number of parameters. The
+Bayesian Information Criterion (BIC) multiplies the penalty for
+complexity with a function of the sample size, $p \ln(n)$, such that
+complex models are penalized more in smaller samples. Finally, the
+so-called *sample size adjusted* BIC (saBIC) implements a different
+ad-hoc penalty for sample size, $\ln(\frac{n + 2}{24})$. Although a
+simulation study showed that the saBIC performed well for class
+enumeration, its penalty is not based on theory [REF]. BIC also performs
+well and is theoretically substantiated; as such, it may be the most
+appropriate information criterion to use for model comparison
+[@nylund-gibson_ten_2018; @masyn_latent_2013]. All three ICs are
+available in the output of `table_fit()`.
 
 Information criteria may occasionally contradict each other, so it is
-important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index, such as the BIC,
-before seeing the results.
-This choice can be preregistered to avoid concerns about cherry picking.
-Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016],
-which is implemented in `tidyLPA` [@rosenbergTidyLPAPackageEasily2018].
-Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs [@nylund-gibson_ten_2018].
-The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
-Adding this many classes results in substantial decreases of the IC,
-and adding further classes results in relatively small decreases.
-
-Given a specific IC,
-it is also possible to calculate so-called "IC-weights",
-which indicate the relative support the data provide for each model in the set [@wagenmakersAICModelSelection2004].
-This is particularly useful for confirmatory LCA,
-as it allows one to quantify the support for the theoretical model.
-IC-weights are calculated by calling `ic_weights()` on the model fit table.
-Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
-the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
+important to identify a suitable strategy to reconcile them. One option
+is to select a specific fit index, such as the BIC, before seeing the
+results. This choice can be preregistered to avoid concerns about cherry
+picking. Another option is to always prefer the most parsimonious model
+that has best fit according to any of the available fit indices. Other
+approach is to select several competing model based on the ICs, and
+evaluate the theoretical intepretation. Yet another option is to
+incorporate information from multiple fit indices using the analytic
+hierarchy process [@akogul_comparison_2016], which is implemented in
+`tidyLPA` [@rosenbergTidyLPAPackageEasily2018]. Finally, one can use a
+scree plot to determine the inflection point at which additional classes
+contribute little to further decreases of the ICs
+[@nylund-gibson_ten_2018]. The preferred number of classes is at the
+inflection point where the slope of the curve is clearly leveling off.
+Adding this many classes results in substantial decreases of the IC, and
+adding further classes results in relatively small decreases.
+
+Given a specific IC, it is also possible to calculate so-called
+"IC-weights", which indicate the relative support the data provide for
+each model in the set [@wagenmakersAICModelSelection2004]. This is
+particularly useful for confirmatory LCA, as it allows one to quantify
+the support for the theoretical model. IC-weights are calculated by
+calling `ic_weights()` on the model fit table. Note, however, that
+IC-weights will (strongly) favor the model with the lowest IC value; the
+scree plot approach above may be more suitable to select the most
+parsimonious model with adequate fit.
 <!-- The general objective of information criteria is to evaluate the model's  -->
@@
 
-When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
-The test statistic is computed as the -2 times the log of the ratio of likelihoods,
-and its degrees of freedom are equal to the difference in the degrees of freedom of the two compared models.
-Since a $k-1$-class model is nested in a $k$-class model,
-it is possible to test the difference in model fit using a LR test.
-One challenge is that the LR test statistic is not asymptotically $\chi^2$ distributed for LCA.
-The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR [REF lo 2001].
-This test is implemented in `tidySEM`, and is part of the default output of `table_fit()`, comparing subsequent rows of the table.
-
-Another solution is to derive an empirical sampling distribution for the LR difference.
-The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the $k-1$-class model $B$ times (e.g., 1000),
-and fitting the $k$- and $k-1$-class models to each simulated dataset [REF Nylund et al., 2007].
-The p-value of this test is the proportion of $B$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
-Simulation studies suggest that the BLRT has greater statistical power than the LRT,
-although both performed comparably well at detecting the true number of classes [REF Nylund 2007, REF Tein 2013].
-A crucial limitation of the BLRT is that it is very computationally expensive.
-With this in mind, `table_fit()` does not provide the BLRT by default.
-Specific models can be compared using the function `BLRT()`.
-
-Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
-returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
-Other authors have reported similar experiences [@sinhaPractitionerGuideLatent2021].
-This suggests that, like existing ICs,
-LR tests might not adequately balance fit and complexity in LCA.
-Better solutions for class enumeration are thus an important target for future research.
-
-A promising novel solution is the so-called *lazy bootstrap*,
-which is much more efficient than the BLRT because it is based on sample statistics of model-implied data generated from the $k-1$ and $k$-class models,
-rather than on fit statistics of LCA models fit to those data [@van_kollenburg_lazy_2018].
-This simulation-based approach to inference is similar to the posterior predictive p-value in Bayesian statistics,
-which describes how well each model is able to reproduce the observed correlation matrix [see @mcelreath_statistical_2020].
-One version of the lazy bootstrap is already implemented in the commercial program latentGOLD [REF].
-At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
-Future versions of `tidySEM` will likely incorporate this test.
+When two models are nested, the significance of the difference in fit
+between these models can be tested using a Likelihood Ratio test (LR).
+The test statistic is computed as the -2 times the log of the ratio of
+likelihoods, and its degrees of freedom are equal to the difference in
+the degrees of freedom of the two compared models. Since a $k-1$-class
+model is nested in a $k$-class model, it is possible to test the
+difference in model fit using a LR test. One challenge is that the LR
+test statistic is not asymptotically $\chi^2$ distributed for LCA. The
+so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem
+by applying an ad-hoc correction to the LR [REF lo 2001]. This test is
+implemented in `tidySEM`, and is part of the default output of
+`table_fit()`, comparing subsequent rows of the table.
+
+Another solution is to derive an empirical sampling distribution for the
+LR difference. The so-called bootstrapped Likelihood Ratio Test (BLRT)
+does so by simulating data from the $k-1$-class model $B$ times (e.g.,
+1000), and fitting the $k$- and $k-1$-class models to each simulated
+dataset [REF Nylund et al., 2007]. The p-value of this test is the
+proportion of $B$ in which the LR of the models fit to simulated data
+exceeds the LR observed in the real data. Simulation studies suggest
+that the BLRT has greater statistical power than the LRT, although both
+performed comparably well at detecting the true number of classes [REF
+Nylund 2007, REF Tein 2013]. A crucial limitation of the BLRT is that it
+is very computationally expensive. With this in mind, `table_fit()` does
+not provide the BLRT by default. Specific models can be compared using
+the function `BLRT()`.
+
+Anecdotal experience as a statistical co-author on LCA studies suggests
+that both these tests tend to be liberal, returning significant p-values
+for models with more classes even when other indicators suggest that
+those models are overfitted (e.g., near-zero class counts). Other
+authors have reported similar experiences
+[@sinhaPractitionerGuideLatent2021], which matches the differences
+between LRT and ICs from information theory as LRT tests and compare the
+models based on in sample predictive accuracy, and ICs evaluates and
+compare the models out-of-sample predictive accuracy which helps to
+control for overfitting [@mcelreath_statistical_2020]. This suggests
+that, like existing ICs, LR tests might not adequately balance fit and
+complexity in LCA. Better solutions for class enumeration are thus an
+important target for future research.
+
+A promising novel solution is the so-called *lazy bootstrap*, which is
+much more efficient than the BLRT because it is based on sample
+statistics of model-implied data generated from the $k-1$ and $k$-class
+models, rather than on fit statistics of LCA models fit to those data
+[@van_kollenburg_lazy_2018]. This simulation-based approach to inference
+is similar to the posterior predictive p-value in Bayesian statistics,
+which describes how well each model is able to reproduce the observed
+correlation matrix [see @mcelreath_statistical_2020]. One version of the
+lazy bootstrap is already implemented in the commercial program
+latentGOLD [REF]. At the time of writing, we are developing a more
+general simulation-based likelihood ratio test (SBLRT) for class
+enumeration. Future versions of `tidySEM` will likely incorporate this
+test.
 
@@
 <!-- 1. We do not consider classes with, on average, fewer than 5 participants per parameter in a class due to potential local underidentification -->
+
 <!-- 1. We do not consider solutions with entropy < .90 because poor class separability compromises interpretability of the results -->
+
 <!-- 1. We do not consider solutions with minimum posterior classification probability < .90 because poor class separability compromises interpretability of the results -->
+
 <!-- Nevertheless, researchers can set an a-priori threshold for the minimum proportion of cases per class, -->
+
 <!-- for example by considering only solutions with a number of participants at least five times as large as the number of parameters per class. -->
-Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
-These diagnostics are not model fit indices and should not be used for class enumeration,
-because a model can fit the data well and still show poor latent class separation [@masyn_latent_2013].
-However, classification diagnostics can be used to eliminate models:
-If one analysis goal is to interpret properties of (members of) classes,
-it makes sense to only consider models that classify well [@nylund-gibson_ten_2018].
-Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.
-
-All classification diagnostics are derived from the posterior classification probabilities: an $n \times k$ matrix called $P$, with the probability that every individual $n$ belongs to latent class $k$.
-When classification accuracy is high, and classes are distinct,
-each individual's posterior class probabilities are high for one class, and low for the remaining classes.
-The individual classification probabilities $P$ are obtained by running `class_prob(res, type = "individual")`.
-Note that one column is added to the matrix, with each individual's most likely class membership $M$.
-$M$ is obtained by assigning each individual to the class corresponding to their maximum posterior class probability.
+
+Classification diagnostics assess whether the classes identified by LCA
+are distinct and clearly separable. These diagnostics are not model fit
+indices and should not be used for class enumeration, because a model
+can fit the data well and still show poor latent class separation
+[@masyn_latent_2013]. However, classification diagnostics can be used to
+eliminate models: If one analysis goal is to interpret properties of
+(members of) classes, it makes sense to only consider models that
+classify well [@nylund-gibson_ten_2018]. Classification diagnostics
+should thus not be used for model selection, but they can help restrict
+the search space of acceptable models.
+
+All classification diagnostics are derived from the posterior
+classification probabilities: an $n \times k$ matrix called $P$, with
+the probability that every individual $n$ belongs to latent class $k$.
+When classification accuracy is high, and classes are distinct, each
+individual's posterior class probabilities are high for one class, and
+low for the remaining classes. The individual classification
+probabilities $P$ are obtained by running
+`class_prob(res, type = "individual")`. Note that one column is added to
+the matrix, with each individual's most likely class membership $M$. $M$
+is obtained by assigning each individual to the class corresponding to
+their maximum posterior class probability.
 <!-- Thus, if an individual has posterior probabilities `c(0.2, 0.5, 0.3)` for classes 1-3, -->
-<!-- they would be assigned to class 2 ($M = 2$). -->
-Predicted class membership $M$ is often of interest to researchers who wish to carry out follow up analyses.
-Note, however, that follow-up analyses should take classification error into account.
-The function `BCH()` does this automatically.
-
-The matrix $P$ can be summarized in different way.
-For example, the estimated frequency table of the latent categorical variable can be obtained by summing the posterior probabilities by class (i.e., the column sums of $P$).
-This frequency table $F_P$ is obtained by calling `class_prob(res, type = "sum.posterior"`.
-These numbers are fractional because this table takes classification error into account.
-Individuals can contribute partially to multiple classes.
-
-Second, most likely class membership based on the highest posterior class probability, $F_M$, is obtained by running `class_prob(res, "sum.mostlikely")`.
-This is the frequency table of the variable $M$.
-Unlike $F_P$, $F_M$ ignores classification error.
-If every observation is classified with perfect accuracy,
-$F_M$ and $F_P$ are identical.
-Part of the output of `table_fit()` for LCA analyses
-are the minimum and maximum percentage of the sample assigned to a particular class, `n_min` and `n_max`, derived from $F_M$.
-This is reported by default because the smallest class proportion, `n_min`, is of particular interest.
-When the size of the smallest class is not much larger than the number of parameters per class,
-proceed with caution:
-The model might not be locally identified in that class [@depaoliMixtureClassRecovery2013].
-Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
-This is not particularly meaningful,
-however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
-Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
-Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
-If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.
-
-To assess classification accuracy, we turn to the third diagnostic table:
-The average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$, `class_prob(res, type = "avg.mostlikely")`:
-for participants assigned to one particular class (rows) according to $M$,
-this table gives the mean probability of being assigned to any class (columns).
-This table is obtained by calculating the column means of $P$ for the 
-$k$ subsets of observations with most likely class of $M = 1 \ldots k$.
-The diagonal represents the probability that observations in each class will be correctly classified.
-<!-- Mauricio: Can you fact-check this?!? I really struglle here -->
-
-The table of classification probabilities for the most likely latent class $M$ by latent class, $P_{p_{(M = m | C = c)}}$ is similar to the aforementioned table $P_{p_{(M = m)}}$,
-except that it accounts for classification uncertainty.
-If $C$ is the true class of an observation, which is imperfectly measured,
-and $M$ is the most likely class,
-then this table shows the probability of an observation being assigned to class $m$ in $M$, given that its true class is $c$ in $C$, or $p(M=m|C=c)$.
-Values on the diagonal can thus be conceptualized as the reliability with which each level of the categorical latent variable is measured by $M$,
-and off-diagonal elements can be conceptualized as measurement error.
-<!-- Mauricio: Can you fact-check this?!? I really struglle here -->
-The minimum and maximum of the diagonal of this table are given by default as part of the output of `table_fit()` for LCA analyses (`prob_min` and `prob_max`).
-These probabilities are of particular interest because they conveniently summarize classification accuracy per class.
-Both probabilities should be high, as this means that all classes are reliably classified.
-If the minimum probability is low,
-there is at least one class with poor classification accuracy.
-
-Finally, class separability can be summarized in a single statistic:
-the entropy criterion [REF].
+<!-- they would be assigned to class 2 ($M = 2$). --> Predicted class
+membership $M$ is often of interest to researchers who wish to carry out
+follow up analyses. Note, however, that follow-up analyses should take
+classification error into account. The function `BCH()` does this
+automatically.
+
+The matrix $P$ can be summarized in different way. For example, the
+estimated frequency table of the latent categorical variable can be
+obtained by summing the posterior probabilities by class (i.e., the
+column sums of $P$). This frequency table $F_P$ is obtained by calling
+`class_prob(res, type = "sum.posterior")`. These numbers are fractional
+because this table takes classification error into account. Individuals
+can contribute partially to multiple classes.
+
+Second, most likely class membership based on the highest posterior
+class probability, $F_M$, is obtained by running
+`class_prob(res, "sum.mostlikely")`. This is the frequency table of the
+variable $M$. Unlike $F_P$, $F_M$ ignores classification error. If every
+observation is classified with perfect accuracy, $F_M$ and $F_P$ are
+identical. Part of the output of `table_fit()` for LCA analyses are the
+minimum and maximum percentage of the sample assigned to a particular
+class, `n_min` and `n_max`, derived from $F_M$. This is reported by
+default because the smallest class proportion, `n_min`, is of particular
+interest. When the size of the smallest class is not much larger than
+the number of parameters per class, proceed with caution: The model
+might not be locally identified in that class
+[@depaoliMixtureClassRecovery2013]. Existing rules of thumb have focused
+on smallest class size in terms of percentage of the total sample size.
+This is not particularly meaningful, however, as the primary concern is
+that small classes may comprise too few observations to accurately
+estimate class parameters. Thus, a better rule of thumb might be to
+ascertain that the smallest class has many more observations than the
+average number of parameters per class. Another good practice is to
+ensure that the smallest class has a unique and meaningful substantive
+interpretation. If the smallest class does not comprise at least a few
+observations per parameter, or is redundant with another larger class, a
+simpler solution is preferable.
+
+To assess classification accuracy, we turn to the third diagnostic
+table: The average posterior probabilities by most likely class
+membership, $P_{p_{(M = m)}}$,
+`class_prob(res, type = "avg.mostlikely")`: for participants assigned to
+one particular class (rows) according to $M$, this table gives the mean
+probability of being assigned to any class (columns). This table is
+obtained by calculating the column means of $P$ for the $k$ subsets of
+observations with most likely class of $M = 1 \ldots k$. The diagonal
+represents the average predictive probability of a the subjects related
+to their most likely class, indicating the certainty of class assignment
+for the subjects. Some guidelines have being suggested for this
+diagonal, like for them to be higher than 0.7 [@masyn2013], but we
+suggest researchers to be cautions about these guidelines as they cannot
+be generalized and should not be use as strict cutoffs to defined "good"
+or "bad" classification.
+
+The table of classification probabilities for the most likely latent
+class $M$ by latent class, $P_{p_{(M = m | C = c)}}$ is similar to the
+aforementioned table $P_{p_{(M = m)}}$, except that it accounts for
+classification uncertainty. If $C$ is the true class of an observation,
+which is imperfectly measured, and $M$ is the most likely class, then
+this table shows the probability of an observation being assigned to
+class $m$ in $M$, given that its true class is $c$ in $C$, or
+$p(M=m|C=c)$. Values on the diagonal can thus be conceptualized as the
+reliability with which each level of the categorical latent variable is
+measured by $M$, and off-diagonal elements can be conceptualized as
+measurement error. The minimum and maximum of the diagonal of this table
+are given by default as part of the output of `table_fit()` for LCA
+analyses (`prob_min` and `prob_max`). These probabilities are of
+particular interest because they conveniently summarize classification
+accuracy per class. Both probabilities should be high, as this means
+that all classes are reliably classified. If the minimum probability is
+low, there is at least one class with poor classification accuracy.
+
+Finally, class separability can be summarized in a single statistic: the
+entropy criterion [REF].
 <!-- Celeux, G., & Soromenho, G. (1996). An entropy criterion for assessing the number of clusters -->
-<!-- in a mixture model. Journal of Classification, 13, 195-212. -->
-In physics, high entropy refers to maximum randomness or uncertainty,
-and low entropy corresponds to a strict arrangement of the units of study.
-Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
-Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
-Crucially, in `tidySEM`, the entropy criterion is defined as 1-entropy (a convention originating in Mplus).
-Its interpretation is reversed;
-0 means the model classification is no better than chance,
-and 1 means perfect classification. 
-There are no rules of thumb for entropy,
-because entropy is also sensitive to the sample size (entropy tends to be lower in larger samples),
-number of latent classes (entropy tends to be lower for higher number of classes), 
-and number of indicators.
-Like other classification diagnostics,
-entropy is not a model fit measure and should not be used for model selection [@masyn_latent_2013].
-However, it can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
+<!-- in a mixture model. Journal of Classification, 13, 195-212. --> In
+physics, high entropy refers to maximum randomness or uncertainty, and
+low entropy corresponds to a strict arrangement of the units of study.
+Applied to LCA, high entropy means that classes are completely
+indistinguishable; every individual has equal probability of being in
+any class. Low entropy, by contrast, means that classes are fully
+separated: every individual has a near-one probability of being in one
+class only, and near-zero probability of being in any other class.
+Crucially, in `tidySEM`, the entropy criterion is defined as 1-entropy
+(a convention originating in Mplus). Its interpretation is reversed; 0
+means the model classification is no better than chance, and 1 means
+perfect classification. There are no rules of thumb for entropy, because
+entropy is also sensitive to the sample size (entropy tends to be lower
+in larger samples), number of latent classes (entropy tends to be lower
+for higher number of classes), and number of indicators. Like other
+classification diagnostics, entropy is not a model fit measure and
+should not be used for model selection [@masyn_latent_2013]. However, it
+can be used to disqualify certain solutions if the goal is to identify
+clearly distinct classes.
 
@@
 
-Class names should be chosen to accurately reflect theoretically relevant characteristics of group membership.
-These class names should aid the interpretation of the class solutions and serve as a shorthand throughout the manuscript.
-For example, one study used a dual-trajectory LCGA of cognitive and affective empathy development with gender as a covariate.
-Although this model has dozens of parameters,
-the resulting classes were labeled low, average, and high empathy because level differences were the most distinguishing feature.
-However, overly simplistic and general class names can be misleading, a phenomenon known as the *naming fallacy* [@weller_latent_2020].
-For example, it would be misleading to call one group "low empathy" and another group "medium empathy" if the empathy differences were non-significant,
-or if other differences, like the effect of the covariate gender, were more substantial than level differences.
-One way to avoid the naming fallacy is comprehensively reporting model parameters,
-and explicating that class names are just a shorthand.
-
-When labeling classes, it is important to note that LCA models are identified up to the class order.
+Class names should be chosen to accurately reflect theoretically
+relevant characteristics of group membership. These class names should
+aid the interpretation of the class solutions and serve as a shorthand
+throughout the manuscript. For example, one study used a dual-trajectory
+LCGA of cognitive and affective empathy development with gender as a
+covariate. Although this model has dozens of parameters, the resulting
+classes were labeled low, average, and high empathy because level
+differences were the most distinguishing feature. However, overly
+simplistic and general class names can be misleading, a phenomenon known
+as the *naming fallacy* [@weller_latent_2020]. For example, it would be
+misleading to call one group "low empathy" and another group "medium
+empathy" if the empathy differences were non-significant, or if other
+differences, like the effect of the covariate gender, were more
+substantial than level differences. One way to avoid the naming fallacy
+is comprehensively reporting model parameters, and explicating that
+class names are just a shorthand.
+
+When labeling classes, it is important to note that LCA models are
+identified up to the class order.
 <!-- This means that, if an LCA model has converged on the global optimum solution, -->
-<!-- its results are stable if the analysis is replicated. -->
-This means that the order of classes is arbitrary and may change if the analysis is replicated.
-A simple solution to ensure replicable results is setting a random seed before fitting the model, using `set.seed()`.
-This will ensure that the classes stay in the same order upon replication,
-but that may not be the preferred order for, e.g., interpreting the results.
-To address this problem, the function `mx_switch_labels()` can be used.
-By default, the function orders classes from largest to smallest class count - but classes can also be ordered by values of a specific parameter (e.g., the mean intercept in a latent class growth model).
-Note that label switching is a much bigger problem when using Bayesian estimation,
-where it can occur during model estimation.
-This is beyond the scope of the present paper.
+<!-- its results are stable if the analysis is replicated. --> This
+means that the order of classes is arbitrary and may change if the
+analysis is replicated. A simple solution to ensure replicable results
+is setting a random seed before fitting the model, using `set.seed()`.
+This will ensure that the classes stay in the same order upon
+replication, but that may not be the preferred order for, e.g.,
+interpreting the results. To address this problem, the function
+`mx_switch_labels()` can be used. By default, the function orders
+classes from largest to smallest class count - but classes can also be
+ordered by values of a specific parameter (e.g., the mean intercept in a
+latent class growth model). Note that label switching is a much bigger
+problem when using Bayesian estimation, where it can occur during model
+estimation. This is beyond the scope of the present paper.
 
@@
 
-LCA is rarely performed in isolation;
-often, researchers wish to relate class membership to auxiliary variables, or even auxiliary models.
-It is important to keep in mind that model-implied class membership is subject to classification error.
-Treating class membership as an observed variable in follow-up analyses disregards this error.
-The results will be biased,
-except when class separation is very high - as indicated by high (near 1) entropy and minimum posterior classification probability.
+LCA is rarely performed in isolation; often, researchers wish to relate
+class membership to auxiliary variables, or even auxiliary models. It is
+important to keep in mind that model-implied class membership is subject
+to classification error. Treating class membership as an observed
+variable in follow-up analyses disregards this error. The results will
+be biased, except when class separation is very high - as indicated by
+high (near 1) entropy and minimum posterior classification probability.
 It is therefore important to account for classification error.
 
-There are several different ways to do follow-up analyses in latent class analysis (LCA) while accounting for classification error.
-At present, the so-called BCH method is considered one of the best approaches [after its creators, REF].
-It has been implemented in `tidySEM`.
-The advantage of the BCH method is that it not only allows for the comparison of individual variables, but also entire auxiliary models across latent classes.
+There are several different ways to do follow-up analyses in latent
+class analysis (LCA) while accounting for classification error. At
+present, the so-called BCH method is considered one of the best
+approaches [after its creators, REF]. It has been implemented in
+`tidySEM`. The advantage of the BCH method is that it not only allows
+for the comparison of individual variables, but also entire auxiliary
+models across latent classes.
 
@@
 Among studies using LCA, reporting practices vary significantly
-[@weller_latent_2020].
-Various authors have tried to improve and
+[@weller_latent_2020]. Various authors have tried to improve and
 standardize ways of reporting LCA [e.g. @masyn_latent_2013;
@@
 <!-- TOP-guidelines emphasize the use of comprehensive citation (including -->
+
 <!-- referencing the software used in the analysis), as well as code and data -->
+
 <!-- sharing wherever possible [@nosek_promoting_2015]. @van_lissa_worcs_2020 -->
+
 <!-- If the original data cannot be shared, sharing synthetic data is a good alternative. -->
+
 <!-- The function `worcs::synthetic()` provides a generic, non-parametric method to create a synthetic copy of data. -->
+
 <!-- When synthesizing data for LCA, it is also possible to synthesize data directly from the model. -->
+
 <!-- LCA models constructed in `tidySEM` are estimated using `OpenMx`, -->
+
 <!-- and as such, can benefit from all existing functions in that package. -->
+
 <!-- To synthesize data from a mixture model `mix`, use the function `mxGenerateData(mix)`. -->
@@
 <!-- the entire research project is made reproducible so that others may -->
+
 <!-- download it and reproduce it with just one click; for guidance, see -->
+
 <!-- @van_lissa_worcs_2020. -->
@@
 <!-- As the open science movement is gaining momentum, researchers are -->
+
 <!-- becoming increasingly aware how important it is that analyses can be -->
+
 <!-- reproduced and audited. In line with open science principles, one of the -->
+
 <!-- suggested reporting standards relates to reproducible code. In this -->
+
 <!-- context, it is important to note that user-friendly methods for -->
+
 <!-- estimating LCA models have predominantly been available in commercial -->
+
 <!-- software packages (e.g., *Mplus* and *Latent GOLD*). A potential -->
+
 <!-- downside of commercial software is that it restricts the ability to -->
+
 <!-- reproduce analyses to license holders, and prevents auditing research -->
+
 <!-- because the underlying source code is proprietary. To overcome these -->
+
 <!-- limitations, the present paper introduces new user-friendly functions in -->
+
 <!-- the `tidySEM` R-package that can be used to estimate a wide range of LCA -->
+
 <!-- models using the free, open-source R-package `OpenMx` as the backend. The  -->
+
 <!-- reporting guidelines described in this paper are adopted in `tidySEM` by  -->
+
 <!-- default. The `tidySEM` R-package thus makes advanced mixture modeling based  -->
+
 <!-- on best practices widely accessible, and facilitates the adoption of the -->
+
 <!-- estimation and reporting guidelines described in this paper. -->
 
-Journals and funding bodies increasingly require research to be open and reproducible.
-For instance, many journals (and the APA) have adopted the TOP guidelines, which recommend comprehensive citation of literature, software, and data sources;
-sharing data and analysis code required to exactly reproduce analyses,
-and pre-registering study plans prior to data collection or analysis [@nosekTransparencyOpennessPromotion2016a].
-A key advantage of using open source software for LCA is that, compared to commercial software, it enables much greater adherence to open science principles.
-Sharing code has limited utility if it can only be evaluated in commercial software,
-as this restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because the underlying source code is proprietary.
-The `tidySEM` package enables performing LCA in free open source software,
-thus increasing the utility of shared code.
-
-Adopting FOSS makes it possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
-The Workflow for Open Reproducible Code in Science describes how to create a fully reproducible research archive,
-and the `worcs` R-package automates many of the steps involved [@vanlissaWORCSWorkflowOpen2021a].
+Journals and funding bodies increasingly require research to be open and
+reproducible. For instance, many journals (and the APA) have adopted the
+TOP guidelines, which recommend comprehensive citation of literature,
+software, and data sources; sharing data and analysis code required to
+exactly reproduce analyses, and pre-registering study plans prior to
+data collection or analysis [@nosekTransparencyOpennessPromotion2016a].
+A key advantage of using open source software for LCA is that, compared
+to commercial software, it enables much greater adherence to open
+science principles. Sharing code has limited utility if it can only be
+evaluated in commercial software, as this restricts the ability to
+reproduce analyses to license holders, and prevents auditing research
+because the underlying source code is proprietary. The `tidySEM` package
+enables performing LCA in free open source software, thus increasing the
+utility of shared code.
+
+Adopting FOSS makes it possible to conduct a fully reproducible LCA
+analysis compliant with best practices in open science. The Workflow for
+Open Reproducible Code in Science describes how to create a fully
+reproducible research archive, and the `worcs` R-package automates many
+of the steps involved [@vanlissaWORCSWorkflowOpen2021a].
 <!-- brings down the learning curve by implementing best practices for creating a reproducible digital research repository. 
--->
-This workflow is based on three principles: First,
-the paper (or analysis section) is written as a *dynamic document* that contains both prose and analysis code.
-Results are automatically rendered as text, tables, and plots.
-Second,
-the research archive is *version controlled*, which means that the entire historical record of the project is retained.
-This version controlled archive can be made public.
-Third, *dependency management* ensures that all software used is documented, including its version number and provenance.
-Without dependency management, code might not work (for example, because it is not clear where certain functions come from),
-or its results may change when some dependencies are updated.
-By combining `tidySEM` and `worcs`,
-it is possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
-
-An additional argument for sharing data in the context of LCA is that reproducing LCA analyses requires individual participant data,
-not just sufficient statistics like many other methods (i.e., a correlation or covariance matrix).
-If the original data cannot be shared,
-sharing synthetic data at least enables reproducing and reusing analysis code.
-The `worcs` package provides a generic, non-parametric method to create a synthetic copy of data when data cannot be shared.
-It is also possible to synthesize data directly from the LCA model using `mxGenerateData()`,
+--> This workflow is based on three principles: First, the paper (or
+analysis section) is written as a *dynamic document* that contains both
+prose and analysis code. Results are automatically rendered as text,
+tables, and plots. Second, the research archive is *version controlled*,
+which means that the entire historical record of the project is
+retained. This version controlled archive can be made public. Third,
+*dependency management* ensures that all software used is documented,
+including its version number and provenance. Without dependency
+management, code might not work (for example, because it is not clear
+where certain functions come from), or its results may change when some
+dependencies are updated. By combining `tidySEM` and `worcs`, it is
+possible to conduct a fully reproducible LCA analysis compliant with
+best practices in open science.
+
+An additional argument for sharing data in the context of LCA is that
+reproducing LCA analyses requires individual participant data, not just
+sufficient statistics like many other methods (i.e., a correlation or
+covariance matrix). If the original data cannot be shared, sharing
+synthetic data at least enables reproducing and reusing analysis code.
+The `worcs` package provides a generic, non-parametric method to create
+a synthetic copy of data when data cannot be shared. It is also possible
+to synthesize data directly from the LCA model using `mxGenerateData()`,
 which often provides a better basis for reproducing the LCA.
@@
 <!-- Furthermore, the FAIR principles focus on *how* research output is shared, -->
+
 <!-- including data, code, and preprints. -->
+
 <!-- These outputs should be Findable in standardized search-indexed repositories, like [psyarxiv.org](https://psyarxiv.org) or [osf.io](https://osf.io), -->
+
 <!-- made Accessible by using open source software and file formats, -->
+
 <!-- made Interoperable so that others can build upon prior work, -->
+
 <!-- and licensed with permission for Reuse, thus enabling others to use and cite the work. -->
@@
 
-LCA analyses inherently involve some subjectivity,
-because of the potential for disagreement between criteria for class enumeration and the lack of objective fit indices.
-In our experience, reviewers (and authors) often disagree about which criteria should be used.
-Preregistering the class enumeration criteria as part of the analysis plan reduces disagreement and eliminates concerns about cherry picking fit measures after the results are known.
-It is particularly important to pre-specify the criteria that will be used to determine the maximum number of classes,
-because there is a risk that information criteria and likelihood ratio tests will suggest a larger number of classes, even if there are concerns about overfitting, classification accuracy, or model interpretability.
-
-The aforementioned Workflow for Open Reproducible Code in Science facilitates creating a preregistration.
-More interestingly, it also enables creating a *Preregistration As Code* (PAC):
-A draft paper with reproducible analysis section based on fake (synthetic or simulated) data [@peikertReproducibleResearchTutorial2021].
-When the real data are accessed or collected,
-the analysis is simply reproduced with this new data file,
-and the results in the paper are automatically updated.
-PACs are quite unambiguous; the selection of the final model can even be automated by hard-coding the criteria into the analysis section.
-For an example of PAC, see @vanlissaComplementingPreregisteredConfirmatory2022.
+LCA analyses inherently involve some subjectivity, because of the
+potential for disagreement between criteria for class enumeration and
+the lack of objective fit indices. In our experience, reviewers (and
+authors) often disagree about which criteria should be used.
+Preregistering the class enumeration criteria as part of the analysis
+plan reduces disagreement and eliminates concerns about cherry picking
+fit measures after the results are known. It is particularly important
+to pre-specify the criteria that will be used to determine the maximum
+number of classes, because there is a risk that information criteria and
+likelihood ratio tests will suggest a larger number of classes, even if
+there are concerns about overfitting, classification accuracy, or model
+interpretability.
+
+The aforementioned Workflow for Open Reproducible Code in Science
+facilitates creating a preregistration. More interestingly, it also
+enables creating a *Preregistration As Code* (PAC): A draft paper with
+reproducible analysis section based on fake (synthetic or simulated)
+data [@peikertReproducibleResearchTutorial2021]. When the real data are
+accessed or collected, the analysis is simply reproduced with this new
+data file, and the results in the paper are automatically updated. PACs
+are quite unambiguous; the selection of the final model can even be
+automated by hard-coding the criteria into the analysis section. For an
+example of PAC, see @vanlissaComplementingPreregisteredConfirmatory2022.
 
@@
 
-The function `table_fit()` reports an overview of fit indices, customized for LCA.
-In reporting model fit, it is important, firstly, to report the effective sample size, which should be the same for all models in order to compare relative fit measures.
-Secondly, it should be clearly reported,
-for example, in the strategy of analyses,
-which parameters are estimated across and within classes.
-In case of doubt, a full overview of the parameters is given by the `coef()` and `table_results()` functions.
-The number of parameters for each model should be included in the fit table.
-Thirdly, it is important that readers can assess the relative fit of different solutions.
-The most fundamental fit measure is the $-2LL$,
-as this is the quantity optimized during estimation.
-From these statistics, information criteria can be computed.
-By default, `table_fit()` includes the AIC, BIC, and saBIC.
-We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.
-
-Classification diagnostics should be reported,
-as they may determine whether models are excluded from consideration.
-Of particular interest is the diagonal of either the average posterior probabilities by most likely class membership,
-or of the most likely class membership by true class membership.
-The former indicates the average probability that individuals belong to the class they were assigned to,
-and the latter indicates the probability that individuals are assigned to a class if they actually belong to that class.
-By default, `table_fit()` reports the latter as `prob_min` and `prob_max`.
-The entropy is also of interest as an overall measure of class separation.
-Finally, the range of proportions of cases assigned to each class is important;
-the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant [see @depaoliMixtureClassRecovery2013].
-By default, `table_fit()` reports the minimum and maximum proportion of cases,
-based on most likely class membership.
-Fifth, the LMR or BLRT can be reported if it is part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
-By default, `table_fit()` reports the VLM for subsequent models in the table.
+The function `table_fit()` reports an overview of fit indices,
+customized for LCA. In reporting model fit, it is important, firstly, to
+report the effective sample size, which should be the same for all
+models in order to compare relative fit measures. Secondly, it should be
+clearly reported, for example, in the strategy of analyses, which
+parameters are estimated across and within classes. In case of doubt, a
+full overview of the parameters is given by the `coef()` and
+`table_results()` functions. The number of parameters for each model
+should be included in the fit table. Thirdly, it is important that
+readers can assess the relative fit of different solutions. The most
+fundamental fit measure is the $-2LL$, as this is the quantity optimized
+during estimation. From these statistics, information criteria can be
+computed. By default, `table_fit()` includes the AIC, BIC, and saBIC. We
+suggest to report at least the BIC, as it is ubiquitous, grounded in
+theory, and performs well in class enumeration.
+
+Classification diagnostics should be reported, as they may determine
+whether models are excluded from consideration. Of particular interest
+is the diagonal of either the average posterior probabilities by most
+likely class membership, or of the most likely class membership by true
+class membership. The former indicates the average probability that
+individuals belong to the class they were assigned to, and the latter
+indicates the probability that individuals are assigned to a class if
+they actually belong to that class. By default, `table_fit()` reports
+the latter as `prob_min` and `prob_max`. The entropy is also of interest
+as an overall measure of class separation. Finally, the range of
+proportions of cases assigned to each class is important; the minimum
+proportion of cases per class can be used to diagnose classes that are
+too small to be locally identified or practically relevant [see
+@depaoliMixtureClassRecovery2013]. By default, `table_fit()` reports the
+minimum and maximum proportion of cases, based on most likely class
+membership. Fifth, the LMR or BLRT can be reported if it is part of the
+class enumeration strategy, using `lr_lmr()` and `BLRT()`. By default,
+`table_fit()` reports the VLM for subsequent models in the table.
 
@@
 
-Reporting LCA results involves reporting the class proportions,
-obtained via `class_prob(res, "sum.posterior")`,
-and model parameters, obtained via `table_results(res)`.
-The class proportions are obtained by standardizing the class weights:
-A $k$-class model has $k-1$ free class weights,
-and one weight constrained to 1 for model identification.
-The free weights are included in the model results,
-but are redundant with the aforementioned class proportions.
-Other model parameters may include means, (co)variances, regression coefficients (rarely), and thresholds for ordinal indicators.
-It is good practice to report the point estimate and confidence interval of these parameters.
-Although it is conventional to report p-values as well,
-note that the default p-values tests the hypothesis that the parameter is equal to zero.
-This hypothesis may be irrelevant in LCA when the focus is on parameter differences between classes.
-Reporting standard errors allows readers to compute p-values for other more informative hypotheses.
-Thresholds are more readily interpretable when converted to conditional item probabilities, using the function `table_prob()`.
-These indicate the probability of an item being
-endorsed given that the observation belongs to a particular latent [@geiser_data_2012].
+Reporting LCA results involves reporting the class proportions, obtained
+via `class_prob(res, "sum.posterior")`, and model parameters, obtained
+via `table_results(res)`. The class proportions are obtained by
+standardizing the class weights: A $k$-class model has $k-1$ free class
+weights, and one weight constrained to 1 for model identification. The
+free weights are included in the model results, but are redundant with
+the aforementioned class proportions. Other model parameters may include
+means, (co)variances, regression coefficients (rarely), and thresholds
+for ordinal indicators. It is good practice to report the point estimate
+and confidence interval of these parameters. Although it is conventional
+to report p-values as well, note that the default p-values tests the
+hypothesis that the parameter is equal to zero. This hypothesis may be
+irrelevant in LCA when the focus is on parameter differences between
+classes. Reporting standard errors allows readers to compute p-values
+for other more informative hypotheses. Thresholds are more readily
+interpretable when converted to conditional item probabilities, using
+the function `table_prob()`. These indicate the probability of an item
+being endorsed given that the observation belongs to a particular latent
+[@geiser_data_2012].
 
@@
 
-As mentioned before, the results table included null-hypothesis tests for all parameters by default.
-Whether these tests are informative depends on the study goals and scale of the data.
-For example, if LCA indicators are mean-centered,
-then the p-values of class-specific means indicate whether the mean in that class differs significantly from the sample average.
-It is also possible to test more informative hypotheses,
-using the function `wald_test()`.
-Equality-constrained informative hypotheses are specified using the `bain` syntax,
-and evaluated on the model object.
-This enables testing theoretical hypotheses - for example, if specific values for class-specific means had been hypothesized.
-The same function allows testing parameters for equality, either within or across classes.
+As mentioned before, the results table included null-hypothesis tests
+for all parameters by default. Whether these tests are informative
+depends on the study goals and scale of the data. For example, if LCA
+indicators are mean-centered, then the p-values of class-specific means
+indicate whether the mean in that class differs significantly from the
+sample average. It is also possible to test more informative hypotheses,
+using the function `wald_test()`. Equality-constrained informative
+hypotheses are specified using the `bain` syntax, and evaluated on the
+model object. This enables testing theoretical hypotheses - for example,
+if specific values for class-specific means had been hypothesized. The
+same function allows testing parameters for equality, either within or
+across classes.
 
@@
 
-In an informal review of published graphics for LCA,
-we found that most visualizations focused on the estimated model parameters,
-displaying class means as points connected by lines, with point shape and linetype indicating class membership.
-Note that the lines introduce spurious information.
-Furthermore, they omit uncertainty about class means and do not illustrate how well the model describes the data.
-
-The Grammar of Graphics addresses best practices in visualization, which we summarize here with specific emphasis on the application to LCA.
-Most importantly, good graphics start with the *data* in mind.
-Visualizing raw data allows one to assess class separability,
-and which aspects of the data are captured well or poorly by the model.
-A second consideration are which *aesthetics* to use.
-In a color plot, class membership can be indicated by color.
-In a black and white plot, it can be indicated by linetype and point shape.
-A third consideration is which geometric objects (geoms) to use.
-As mentioned before, common practice is to plot parameter estimates as points, and connect these points with lines.
-Points are suitable geoms to visualize point estimates and raw data.
-For continuous indicators, additional relevant geoms include error bars for standard errors and standard deviations,
-and density plots to visualize the distribution of raw data (smoothed with a kernel density function).
-In two dimensions, (co)variances can be visualized as ellipses.
-Finally, when plotting multiple class solutions,
-it makes sense to *facet* a plot into multiple panels by class solution.
-The plots in `tidySEM` are based on best practices as outlined in the grammar of graphics,
-and are implemented in `ggplot2`.
-As such, these plots can be further modified just like any other `ggplot2` object.
-One remaining limitation is that these best practices have only been implemented for LCA with continuous indicators, in `plot_profiles()`, `plot_density()` and `plot_bivariate()`, and for LCGA in `plot_growth()`.
-Plots for LCA with ordinal indicators are still rudimentary, more work remains to be done in this area.
+In an informal review of published graphics for LCA, we found that most
+visualizations focused on the estimated model parameters, displaying
+class means as points connected by lines, with point shape and linetype
+indicating class membership. Note that the lines introduce spurious
+information. Furthermore, they omit uncertainty about class means and do
+not illustrate how well the model describes the data.
+
+The Grammar of Graphics addresses best practices in visualization, which
+we summarize here with specific emphasis on the application to LCA. Most
+importantly, good graphics start with the *data* in mind. Visualizing
+raw data allows one to assess class separability, and which aspects of
+the data are captured well or poorly by the model. A second
+consideration are which *aesthetics* to use. In a color plot, class
+membership can be indicated by color. In a black and white plot, it can
+be indicated by linetype and point shape. A third consideration is which
+geometric objects (geoms) to use. As mentioned before, common practice
+is to plot parameter estimates as points, and connect these points with
+lines. Points are suitable geoms to visualize point estimates and raw
+data. For continuous indicators, additional relevant geoms include error
+bars for standard errors and standard deviations, and density plots to
+visualize the distribution of raw data (smoothed with a kernel density
+function). In two dimensions, (co)variances can be visualized as
+ellipses. Finally, when plotting multiple class solutions, it makes
+sense to *facet* a plot into multiple panels by class solution. The
+plots in `tidySEM` are based on best practices as outlined in the
+grammar of graphics, and are implemented in `ggplot2`. As such, these
+plots can be further modified just like any other `ggplot2` object. One
+remaining limitation is that these best practices have only been
+implemented for LCA with continuous indicators, in `plot_profiles()`,
+`plot_density()` and `plot_bivariate()`, and for LCGA in
+`plot_growth()`. Plots for LCA with ordinal indicators are still
+rudimentary, more work remains to be done in this area.
 <!-- Plots can greatly aid the interpretability of LCA models. There are several stages where this is an important consideration. -->
@@
 <!-- Second, once we have decided on the final class solution, we want to -->
+
 <!-- interpret the response patterns on the indicators for each latent class. -->
+
 <!-- While `tidySEM` can create a table showing the probability of each -->
+
 <!-- item's response endorsement (using `tidySEM::table_prob`), it may be -->
+
 <!-- easier to inspect these probabilities visually. For this reason, we -->
+
 <!-- created `ggplot2::plot_prob()`. The resulting shows response patterns on -->
+
 <!-- all indicators for each group. We give an example of this plot in the -->
+
 <!-- Tutorial subsection. -->
@@
 
-Based on our experience in statistical consultations,
-there are a few common pitfalls and misunderstandings about LCA.
-First, perhaps the most common problem we see in consultations is treating ordinal data as continuous.
-As explained previously, LCA for ordinal indicators should be used for Likert-type scales and other data with few unique values.
-A second common pitfall is omitting the 1-class solution from class enumeration.
-If a one-class solution has better or similar fit to models with a higher number of classes, LCA may not be appropriate.
-A one-class solution should thus always be included as a benchmark.
-Third, although it is widely understood that classification diagnostics (e.g., entropy, posterior probability, class counts) are not class enumeration criteria,
-they still have a role in class enumeration.
-Specifically, they can be used to exclude solutions that do not meet predefined criteria from consideration.
-Fourth, previous reporting standards called for reporting of sample descriptive statistics [see @schreiberLatentClassAnalysis2017].
-While we agree that this is due dilligence,
-sample descriptives have two limitations in the context of LCA:
-they are not sufficient to check or reproduce the analysis,
-and many use cases of LCA assume that the population is heterogeneous, which limits the utility of whole-sample descriptives.
-These two limitations can be addressed by data sharing and reporting descriptive statistics for each latent class separately.
-Fifth, a recent publication mistakenly stated that an assumption of mixture
-models is that observed indicators are normally distributed
-[@spurk_latent_2020].
-This is incorrect, as evident from the shoe size example:
-When the number of classes is greater than one,
-Gaussian mixture models assume that the distribution of observed indicators is a *mixture of multiple normal distributions*.
+Based on our experience in statistical consultations, there are a few
+common pitfalls and misunderstandings about LCA. First, perhaps the most
+common problem we see in consultations is treating ordinal data as
+continuous. As explained previously, LCA for ordinal indicators should
+be used for Likert-type scales and other data with few unique values. A
+second common pitfall is omitting the 1-class solution from class
+enumeration. If a one-class solution has better or similar fit to models
+with a higher number of classes, LCA may not be appropriate. A one-class
+solution should thus always be included as a benchmark. Third, although
+it is widely understood that classification diagnostics (e.g., entropy,
+posterior probability, class counts) are not class enumeration criteria,
+they still have a role in class enumeration. Specifically, they can be
+used to exclude solutions that do not meet predefined criteria from
+consideration. Fourth, previous reporting standards called for reporting
+of sample descriptive statistics [see
+@schreiberLatentClassAnalysis2017]. While we agree that this is due
+dilligence, sample descriptives have two limitations in the context of
+LCA: they are not sufficient to check or reproduce the analysis, and
+many use cases of LCA assume that the population is heterogeneous, which
+limits the utility of whole-sample descriptives. These two limitations
+can be addressed by data sharing and reporting descriptive statistics
+for each latent class separately. Fifth, a recent publication mistakenly
+stated that an assumption of mixture models is that observed indicators
+are normally distributed [@spurk_latent_2020]. This is incorrect, as
+evident from the shoe size example: When the number of classes is
+greater than one, Gaussian mixture models assume that the distribution
+of observed indicators is a *mixture of multiple normal distributions*.
 <!-- In our shoe -->
@@
 <!-- prototypical example of a mixed population distribution where LCA can -->
-<!-- discover latent groups. -->
-If the indicators were normally distributed,
-there would be no classes to extract as the whole population
-would belong to a single (homogeneous) class.
-Finally, the same paper suggested that robust standard errors
-should be used when indicators are not normally distributed
-[@spurk_latent_2020].
-This is also incorrect;
-to our knowledge, there is no literature on the merits of robust standard errors for LCA.
-Finally, we caution against step-wise class enumeration,
-whereby additional models are estimated until the selected class enumeration criteria no longer show improvement of model fit [e.g., see @spurk_latent_2020].
-This practice is problematic,
-because it is possible that an even higher number of classes would lead to a (significant) improvement in fit.
-Consider this thought experiment: there exist two groups of three relatively similar classes each (six in total).
-Step-wise class enumeration shows improvement in model fit going from a 1-class to 2-class model: the two classes capture the difference between the two groups of three classes.
-Then, no noteworthy improvement in model fit is found for adding more classes, until the 6-class solution is reached.
-In the step-wise approach, the analysis would be terminated after the 2-class model - even though the correct number of classes is 6.
+<!-- discover latent groups. --> If the indicators were normally
+distributed, there would be no classes to extract as the whole
+population would belong to a single (homogeneous) class. Finally, the
+same paper suggested that robust standard errors should be used when
+indicators are not normally distributed [@spurk_latent_2020]. This is
+also incorrect; to our knowledge, there is no literature on the merits
+of robust standard errors for LCA. Finally, we caution against step-wise
+class enumeration, whereby additional models are estimated until the
+selected class enumeration criteria no longer show improvement of model
+fit [e.g., see @spurk_latent_2020]. This practice is problematic,
+because it is possible that an even higher number of classes would lead
+to a (significant) improvement in fit. Consider this thought experiment:
+there exist two groups of three relatively similar classes each (six in
+total). Step-wise class enumeration shows improvement in model fit going
+from a 1-class to 2-class model: the two classes capture the difference
+between the two groups of three classes. Then, no noteworthy improvement
+in model fit is found for adding more classes, until the 6-class
+solution is reached. In the step-wise approach, the analysis would be
+terminated after the 2-class model - even though the correct number of
+classes is 6.
 
@@
 
-The best practices discussed in this paper are summarized in the SMART-LCA Checklist: Standards for More Accuracy in Reporting of different Types of Latent Class Analysis.
+The best practices discussed in this paper are summarized in the
+SMART-LCA Checklist: Standards for More Accuracy in Reporting of
+different Types of Latent Class Analysis.
 
 1.  Report item descriptives
-    + Are all items on similar scales?
-    + Is the measurement level correctly coded?
-    + Any other reasons why the observed (multivariate) distribution could not be described by the LCA (e.g., extreme skew)?
+    -   Are all items on similar scales?
+    -   Is the measurement level correctly coded?
+    -   Any other reasons why the observed (multivariate) distribution
+        could not be described by the LCA (e.g., extreme skew)?
 2.  Report missingness
-    + Proportion of missingness per variable
-    + MAR test
-    + Missing data method for handling missingness (typically FIML)
+    -   Proportion of missingness per variable
+    -   MAR test
+    -   Missing data method for handling missingness (typically FIML)
 3.  Report model specification
-    + In exploratory LCA, compare all sensible alternative specifications
-    + In confirmatory LCA, justify alternative models on theoretical grounds
-    + Check that the observed (multivariate) distribution could theoretically be captured by the specified model (see #1).
+    -   In exploratory LCA, compare all sensible alternative
+        specifications
+    -   In confirmatory LCA, justify alternative models on theoretical
+        grounds
+    -   Check that the observed (multivariate) distribution could
+        theoretically be captured by the specified model (see #1).
 4.  Class enumeration
-    + Justify the maximum number of classes $K$
-    + In exploratory LCA, estimate 1:$K$ classes
-    + In confirmatory LCA, estimate the theoretical number of classes, a 1-class model, and optionally other numbers of classes for comparison
-5. Justify the (preregistered) criteria used for class enumeration, which can include:
-    + Information criteria
-    + LR tests
-6. Justify the (preregistered) criteria used to eliminate models from consideration, including:
-    + Classification diagnostics
-    + Local identifiability (acceptable number of observations per parameter in each class)
-    + Model convergence
-    + Theoretical interpretability of the solution
+    -   Justify the maximum number of classes $K$
+    -   In exploratory LCA, estimate 1:$K$ classes
+    -   In confirmatory LCA, estimate the theoretical number of classes,
+        a 1-class model, and optionally other numbers of classes for
+        comparison
+5.  Justify the (preregistered) criteria used for class enumeration,
+    which can include:
+    -   Information criteria
+    -   LR tests
+    -   Theoretical interpretability of the solution
+6.  Justify the (preregistered) criteria used to eliminate models from
+    consideration, including:
+    -   Classification diagnostics
+    -   Local identifiability (acceptable number of observations per
+        parameter in each class)
+    -   Model convergence
+    -   Theoretical interpretability of the solution
 7.  Report the following:
-    + Fit of all models under consideration
-    + Classification diagnostics for all models under consideration
-    + All parameters of the chosen model (means, (co)variances, thresholds)
-    + Proportion of cases per class
+    -   Fit of all models under consideration
+    -   Classification diagnostics for all models under consideration
+    -   All parameters of the chosen model (means, (co)variances,
+        thresholds)
+    -   Proportion of cases per class
 8.  Provide an informative plot of the chosen solution
-    + Show raw data
-    + Show parameters
-    + Show confidence bounds if possible
-    + Show multivariate distributions for multivariate models, if possible
+    -   Show raw data
+    -   Show parameters
+    -   Show confidence bounds if possible
+    -   Show multivariate distributions for multivariate models, if
+        possible
 9.  Account for classification inaccuracy in follow-up analyses
-    + If entropy > .95 and posterior classification probability for each class > .95, this is less crucial
+    -   If entropy \> .95 and posterior classification probability for
+        each class \> .95, this is less crucial
 

### 2023-01-23 manuscript.tex
edits by Mauricio

@@
 most notably finite Gaussian mixture modeling and latent profile
-analysis. Vermunt et al. (2004) defined LCA more generally as
-virtually any statistical model where ``some of the parameters {[}\ldots{]}
-differ across unobserved subgroups''.
-
-Despite the popularity of LCA,
-two challenges impede broader adoption and correct application of the method.
-The first challenge is that most existing software that implements these models is commercial and closed-source.
-The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
-This introduces a risk of misapplications of the technique,
-and complicates manuscript review and assessment of the quality of published studies.
-The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
-based on current best practice.
-Furthermore, meeting the increasing demand for open source research software,
-this paper introduces new functionality in the R-package \texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free open source software (FOSS).
+analysis. Jeroen K. Vermunt et al. (2004) defined LCA more generally as virtually
+any statistical model where ``some of the parameters {[}\ldots{]} differ across
+unobserved subgroups''.
+
+Despite the popularity of LCA, two challenges impede broader adoption
+and correct application of the method. The first challenge is that most
+existing software that implements these models is commercial and
+closed-source. The second challenge is that there is a lack of standards
+for estimating and reporting LCA, and some myths and misunderstandings
+have been perpetuated in prior literature. This introduces a risk of
+misapplications of the technique, and complicates manuscript review and
+assessment of the quality of published studies. The present paper seeks
+to address these challenges by introducing updated guidelines for
+estimating and reporting LCA, based on current best practice.
+Furthermore, meeting the increasing demand for open source research
+software, this paper introduces new functionality in the R-package
+\texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free
+open source software (FOSS).
 
@@
 
-Despite the popularity of LCA,
-most existing software is commercial and closed-source,
-for example, Mplus (\textbf{muthenMplusUserGuide1998?}), for which the R-package \texttt{MplusAutomation} offers augmented LCA functionality (\textbf{hallquistMplusAutomationPackageFacilitating2018?}), and LatentGOLD {[}REF{]}.
-A downside of commercial software is that its use is restricted to individuals and institutions who can afford a license.
-Another downside is that the proprietary source code cannot be audited, debugged, or enhanced by third parties.
-This goes against open science principles (\textbf{lamprechtFAIRPrinciplesResearch2019?}),
-incurs a risk that mistakes in the source code go unnoticed,
-and curbs progress as independent developers cannot fix bugs or add functionality.
-
-There are several compelling reasons why researchers should prefer free open source software (FOSS) instead.
-FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
-This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
-Furthermore, FOSS promotes collaboration within the scientific community,
-as its users can contribute to the development, maintenance, and support of the software.
-Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
-researchers can review and verify the methods and algorithms used,
-which helps to ensure the integrity and reliability of research findings.
-Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.
-
-Notable FOSS for LCA includes Mclust {[}REF{]}, depmixS4 {[}REF{]}, and OpenMx (Neale et al., 2016).
-A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
-or implement specific innovations (e.g., depmixS4).
-OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
-Remaining obstacles to the use of \texttt{OpenMx} is that it is not very user-friendly and its functionality for LCA is poorly documented.
-
-These limitations are addressed by the \texttt{tidySEM} package,
-which provide a user-friendly interface for LCA in \texttt{OpenMx}.
-The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting LCA, as described in this paper.
-The primary purpose of the \texttt{tidySEM} package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
-The user interface is simple, as models can be specified using the widely used \texttt{lavaan} syntax (Rosseel, 2012).
-The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
-but the resulting models can still be fully customized (low barrier, high ceiling).
-Specifically, LCA models constructed in \texttt{tidySEM} inherit from \texttt{OpenMx}' format \texttt{MxModel}, and remain compatible with downstream functions that process such models.
-Being FOSS, \texttt{tidySEM} facilitates fully-reproducible analyses and contributes to open science.
-It was developed with open science in mind,
-and adheres to the FAIR software principles (Findable, Accessible, Interoperable, and Reusable, \textbf{lamprechtFAIRPrinciplesResearch2019?}),
-and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{OpenSSF Best Practices} for FOSS software.
+Despite the popularity of LCA, most existing software is commercial and
+closed-source, for example, Mplus (\textbf{muthenMplusUserGuide1998?}), for which
+the R-package \texttt{MplusAutomation} offers augmented LCA functionality
+(\textbf{hallquistMplusAutomationPackageFacilitating2018?}), and LatentGOLD
+{[}REF{]}.
+A
+downside of commercial software is that its use is restricted to
+individuals and institutions who can afford a license. Another downside
+is that the proprietary source code cannot be audited, debugged, or
+enhanced by third parties. This goes against open science principles
+(\textbf{lamprechtFAIRPrinciplesResearch2019?}), incurs a risk that mistakes in
+the source code go unnoticed, and curbs progress as independent
+developers cannot fix bugs or add functionality.
+
+There are several compelling reasons to prefer free open source software
+(FOSS) instead. FOSS is freely available for anyone to use, modify, and
+distribute, which means that scientists can access and use a wide range
+of powerful tools without financial impediments. This is invaluable for
+reducing gatekeeping of researchers with limited budgets, such as those
+from developing countries. Furthermore, FOSS promotes collaboration
+within the scientific community, as its users can contribute to the
+development, maintenance, and support of the software. Finally, FOSS
+promotes transparency and reproducibility: Because the source code for
+FOSS is accessible, researchers can review and verify the methods and
+algorithms used, which helps to ensure the integrity and reliability of
+research findings. Overall, FOSS has the potential to significantly
+improve the efficiency and effectiveness of scientific research, while
+also promoting transparency, collaboration, and inclusion within the
+scientific community.
+
+Notable FOSS for LCA includes mclust (Scrucca, Fop, Murphy, \& Raftery, 2016), depmixS4
+(Visser \& Speekenbrink, 2010), and OpenMx (Neale et al., 2016). A crucial limitation of
+these existing packages is that some have limited functionality (e.g.,
+Mclust), or implement specific innovations (e.g., depmixS4). OpenMx is
+unique in that, like Mplus, it implement LCA in the context of a fully
+featured structural equation modeling framework. Remaining obstacles to
+the use of \texttt{OpenMx} is that it is not very user-friendly and its
+functionality for LCA is poorly documented.
+
+These limitations are addressed by the \texttt{tidySEM} package, which provide
+a user-friendly interface for LCA in \texttt{OpenMx}. The package also includes
+convenience functions to obtain relevant statistics, tests and plots in
+accordance with best practices in estimation and reporting LCA, as
+described in this paper. The primary purpose of the \texttt{tidySEM} package is
+to provide a tidy workflow for generating, estimating, reporting, and
+plotting structural equation models in a software-agnostic way. The user
+interface is simple, as models can be specified using the widely used
+\texttt{lavaan} syntax (Rosseel, 2012). The package contains
+user-friendly wrapper functions to conduct LCA with sensible defaults,
+but the resulting models can still be fully customized (low barrier,
+high ceiling). Specifically, LCA models constructed in \texttt{tidySEM} inherit
+from \texttt{OpenMx}' format \texttt{MxModel}, and remain compatible with downstream
+functions that process such models. Being FOSS, \texttt{tidySEM} facilitates
+fully-reproducible analyses and contributes to open science. It was
+developed with open science in mind, and adheres to the FAIR software
+principles (Findable, Accessible, Interoperable, and Reusable, \textbf{lamprechtFAIRPrinciplesResearch2019?}), and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{OpenSSF Best
+Practices} for
+FOSS software.
 
@@
 
-LCA refers to methods for estimating unobserved
-groups based on a parametric model of observed indicators of group
-membership (\textbf{vermuntLatentClassAnalysis2004?}).
-It has become popular across scientific fields,
-under a number of different names;
-most notably (finite Gaussian) mixture modeling and latent profile analysis.
-The concept of LCA can be understood in different ways.
-Generally speaking, LCA assumes that the study population is composed of \(K\)
-subpopulations.
-It further assumes that the observed data are
-a mixture of data originating from those subpopulations;
-hence the name mixture model.
-Consider the simplest possible
-univariate ``model'', which is a normal distribution.
-This model has
-two parameters: the mean and the variance.
-LCA aims to estimate the values of those parameters across a known number of classes \(K\),
-as well as the probability of every individual \(i\) belonging to classes \(1 \ldots k\).
-Commonly, the same model is estimated across all classes,
-but with different parameters for each class (i.e., class-specific means and variances).
+LCA refers to methods for estimating unobserved groups based on a
+parametric model of observed indicators of group membership
+(\textbf{vermuntLatentClassAnalysis2004?}). It has become popular across
+scientific fields, under a number of different names; most notably
+(finite Gaussian) mixture modeling and latent profile analysis. The
+concept of LCA can be understood in different ways. Generally speaking,
+LCA assumes that the study population is composed of \(K\) subpopulations.
+It further assumes that the observed data are a mixture of data
+originating from those subpopulations; hence the name mixture model.
+Consider the simplest possible univariate ``model'', which is a normal
+distribution. This model has two parameters: the mean and the variance.
+LCA aims to estimate the values of those parameters across a known
+number of classes \(K\), as well as the probability of every individual
+\(i\) belonging to classes \(1 \ldots k\). Commonly, the same model is
+estimated across all classes, but with different parameters for each
+class (i.e., class-specific means and variances).
 
@@
 
-Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling (\textbf{molenaarArbitraryNatureLatent1994?}).
-A mixture model is like confirmatory factor analysis,
-except with a categorical instead of a continuous latent variable.
-One difference between
-the two techniques is that factor analysis can be considered as a way to
-group observed \emph{variables} into latent constructs, and LCA groups
-\emph{individuals} into classes.
-While factor analysis seeks to describe the relations between variables,
-LCA seeks to describe unobserved groups and classify individual observations to those groups.
-In line with this distinction, LCA
-(mixture models) is referred to as a ``person-centered'' technique, and
-factor analysis as a ``variable-centered'' technique
-(Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+Another way to conceptualize LCA is by analogy to a measurement model in
+structural equation modeling (\textbf{molenaarArbitraryNatureLatent1994?}). A
+mixture model is like confirmatory factor analysis, except with a
+categorical instead of a continuous latent variable. One difference
+between the two techniques is that factor analysis can be considered as
+a way to group observed \emph{variables} into latent constructs, and LCA
+groups \emph{individuals} into classes. While factor analysis seeks to
+describe the relations between variables, LCA seeks to describe
+unobserved groups and classify individual observations to those groups.
+In
+line with this distinction, LCA (mixture models) is referred to as a
+``person-centered'' technique, and factor analysis as a
+``variable-centered'' technique (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
 
@@
 thought of as similar to a multi-group structural equation model. The
-main distinction is that group membership is not known a-priori, but is
-instead estimated -- with measurement error -- based on the data.
-Whereas in a multi-group model, the data are split by group and treated
-as independent samples, in LCA, all cases contribute to the estimation
-of all parameters in all groups. The relative contribution of each case
-to the parameters of each group is determined by that case's posterior
-probability of belonging to that group.
-
-When the focus is on each individual's estimated class membership,
-LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
-Specifically, the posterior class probability that an individual belongs to a latent class
-can be computed from the likelihood of that individual's observed data under given the class-specific model.
-These posterior class probabilities can be used to weight follow-up analyses,
-or individuals can be assigned to classes based on their highest class probability.
-If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
-This perspective on mixture modeling is sometimes referred to as ``model-based
-clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \& Raftery, 2016).
-Many clustering algorithms apply some recursive splitting algorithm to the data. By
-contrast ``model-based'' clustering refers to the fact that LCA estimates
-cluster membership based on a parametric model.
-Since a parametric model estimates only a fixed number of parameters,
-it can be a relatively parsimonious solution compared to non-parametric techniques.
+main distinction is that group membership is not known a-priori, but
+estimated from the data. Whereas in a multi-group model, the data are
+split by group and treated as independent samples, in LCA, all cases
+contribute to the estimation of all parameters in all groups. The
+relative contribution of each case to the parameters of each group is
+determined by that case's posterior probability of belonging to that
+group.
+
+When the focus is on each individual's estimated class membership, LCA
+can be thought of as a type of clustering algorithm that corrects for
+(random) measurement error. Specifically, the posterior class
+probability that an individual belongs to a latent class can be computed
+from the likelihood of that individual's observed data under given the
+class-specific model. These posterior class probabilities can be used to
+weight follow-up analyses, or individuals can be assigned to classes
+based on their highest class probability. If the classes are clearly
+distinct, measurement error is low, and the latter approach may be
+acceptable. This perspective on mixture modeling is sometimes referred
+to as ``model-based clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016). Many clustering algorithms apply some recursive
+splitting algorithm to the data. By contrast ``model-based'' clustering
+refers to the fact that LCA estimates cluster membership based on a
+parametric model. Since a parametric model estimates only a fixed number
+of parameters, it can be a relatively parsimonious solution compared to
+non-parametric techniques.
 
 Finally, in the context of machine learning, LCA can be considered as an
-\emph{unsupervised classification} method (Figueiredo \& Jain, 2002).
-The term \emph{unsupervised} refers to the fact that the outcome variable --
-true class membership -- is not known, and the term \emph{classification}
-refers to the fact that the algorithm is predicting a categorical
-outcome: class membership.
+\emph{unsupervised} learning method for \emph{classification}
+(Figueiredo \& Jain, 2002). The term \emph{unsupervised} refers to the
+fact that the outcome variable -- true class membership -- is not known,
+and the term \emph{classification} refers to the fact that the algorithm is
+predicting a categorical outcome: class membership.
 
@@
 
-The term LCA was initially conceived to refer to analyses with categorical,
-usually binary indicators (Collins \& Lanza, 2009).
-It was later generalized to refer to any technique that estimates a categorical latent variable based on a
-parametric model of observed indicators, as it only refers
-to the purpose of the analysis, and does not imply restrictions to the
-model used, or the level of measurement of the indicators (Vermunt et al., 2004).
-To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
-Given the abundance of redundant terms in use for closely related types of LCA models,
-we here provide a rudimentary taxonomy.
+The term LCA was initially conceived to refer to analyses with
+categorical, usually binary indicators (Collins \& Lanza, 2009). It was
+later generalized to refer to any technique that estimates a categorical
+latent variable based on a parametric model of observed indicators, as
+it only refers to the purpose of the analysis, and does not imply
+restrictions to the model used, or the level of measurement of the
+indicators (Jeroen K. Vermunt et al., 2004). To prevent ambiguity, we refer to LCA
+with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
+Given the abundance of redundant terms in use for closely related types
+of LCA models, we here provide a rudimentary taxonomy.
 
 A common type of LCA is the \emph{finite Gaussian mixture model}; a
-univariate analysis where the observed distribution of a single variable
-is assumed to result from a mixture of a known number of normal
-(Gaussian) distributions. The parameters of a finite Gaussian mixture
-model are the means and variances of these underlying normal
-distributions. The analysis of shoe sizes presented earlier is a
+univariate analysis where the observed distribution of a single
+continuous variable is assumed to result from a mixture of a known
+number of normal (Gaussian) distributions. The parameters of a finite
+Gaussian mixture model are the means and variances of these underlying
+normal distributions. The analysis of shoe sizes presented earlier is a
 canonical example of this type of analysis. In the multivariate case,
@@
 are the means, variances, and covariances between the indicators (which
-can be standardized to obtain correlations). These parameters can be estimated
-freely, or set to be constrained across classes (Collins \& Lanza, 2009).
+can be standardized to obtain correlations). These parameters can be
+estimated freely, or set to be constrained across classes
+(Collins \& Lanza, 2009).
 
@@
 indicators. Conditional independence means that, after class membership
-is accounted for,
-the residual covariances/correlations between indicators are
-assumed to be zero.
-In some cases, such constraints
-will be inappropriate; for instance when the cohesion between indicators
-is expected to differ between classes (Collins \& Lanza, 2009).
-Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
-A class of larger particles with a low correlation between length and width,
-and a class of smaller particles with a high correlation between length and width.
+is accounted for, the residual covariances/correlations between
+indicators are assumed to be zero. In some cases, such constraints will
+be inappropriate; for instance when the cohesion between indicators is
+expected to differ between classes (Collins \& Lanza, 2009). Consider, for
+example, a mixture model of ocean plastic particles, which found two
+classes of particles based on length and width: A class of larger
+particles with a low correlation between length and width, and a class
+of smaller particles with a high correlation between length and width.
 The reason for this difference in correlations makes theoretical sense:
-the large particles were heterogenous in shape,
-whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).
-
-Finally, some LCA models use latent variables as indicators.
-A common application of this approach is in
-longitudinal research with repeated measures of a construct.
-Examples of this approach include
-\emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
-(Jung \& Wickrama, 2008).
-These techniques estimate a latent growth model
-to describe grouped trajectories over time.
-The indicators of class membership in a GMM are the
-intercepts, and variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
-For linear trajectories, these latent variables are an intercept and slope.
-The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes homogeneity of trajectories within classes.
-This is achieved by restricting the (co)variances of the growth parameters to zero.
-Any variance in the indicators not explained by the class-specific
+the large particles were heterogenous in shape, whereas the smaller
+particles had been polished to a more uniform (rounded) shape by the
+elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).
+
+A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
+called Latent Transition Analysis), which estimates the change in class
+membership over time with the markov assumption over time. The markov
+assumption states that the class membership at time \(t\) is only affected
+by the class membership at time \(t-1\). The structural part of the model
+is comprise of the initial states (proportion of the sample in each
+class at the first time point), and transition probabilities
+(probabilities to switch class membership across time). The transition
+probabilities are estimated as multinomial logistic regressions between
+class memberships (Jeroen K. Vermunt, 2010).
+
+Finally, some LCA models use latent variables as indicators. A common
+application of this approach is in longitudinal research with repeated
+measures of a construct. Examples of this approach include \emph{growth
+mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
+(Jung \& Wickrama, 2008). These techniques estimate a latent growth
+model to describe grouped trajectories over time. The indicators of
+class membership in a GMM are the intercepts, and variances, and
+covariances of the latent growth variables that describe individual
+trajectories on the observed variables. For linear trajectories, these
+latent variables are an intercept and slope. The free variances allow
+trajectories to vary within classes. The LCGA differs from the GMM in
+that it assumes homogeneity of trajectories within classes. This is
+achieved by restricting the (co)variances of the growth parameters to
+zero. Any variance in the indicators not explained by the class-specific
 latent trajectories is assumed to be error variance.
@@
 
-Although LCA is often discussed as an exploratory analysis technique,
-it can also be used in a confirmatory manner.
-Given that LCA is similar to confirmatory factor analysis,
-it can be used to similar ends:
-To assess whether a theoretical measurement model holds.
-This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
-For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses (Marcia, 1966).
-LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) (\textbf{luyckxDevelopmentalTypologiesIdentity2008?}).
-Similarly, \emph{personality type theory} states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA (\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).
-One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices (see Model Fit Indices).
-Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
-Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
-either qualitatively or quantitatively, using significance tests for pre-specified values (see Inference).
-A recent study by Maene, D'hondt, Van Lissa, Thijs, and Stevens (2022) applied this principle to
-test the theory that identities of Belgian adolescents with migration background could be summarized as those relating to their heritage, national and regional identity.
+Although LCA is often discussed as an exploratory analysis technique, it
+can also be used in a confirmatory manner. Given that LCA is similar to
+confirmatory factor analysis, it can be used to similar ends: To assess
+whether a theoretical measurement model holds. This use case is relevant
+when testing a theory that postulates the existence of a categorical
+latent variable. For example, \emph{identity status theory} posits that, at
+any given point in time, adolescents reside in one of four identity
+statuses (Marcia, 1966). LCA can be used to identify these
+four statuses based on observed indicators (e.g., self-reported identity
+exploration and commitment)
+(\textbf{luyckxDevelopmentalTypologiesIdentity2008?}). Similarly, \emph{personality
+type theory} states that dimensional differences in personality can
+largely be explained by an undercontrolled, overcontrolled, and
+resilient type, which could also be restored using LCA
+(\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).
+
+One challenge is that, unlike confirmatory factor analysis, LCA does not
+provide absolute fit indices (see Model Fit Indices). Thus, to test a
+theory, one can ascertain that the data are better described by the
+number of classes dictated by theory than by a different number of
+classes. Furthermore, it would be possible to test whether the observed
+pattern of class-specific means corresponds to a hypothesized pattern;
+either qualitatively or quantitatively, using significance tests for
+pre-specified values (see Inference).
+
+Appendix A is a tutorial to confirmatory LCA in \texttt{tidySEM}, based on a
+recent study that tested the theory that identities of Belgian
+adolescents with migration backgrounds could be summarized as those
+relating to their heritage, national and regional identity
+(Maene, D'hondt, Van Lissa, Thijs, \& Stevens, 2022).
+
+\hypertarget{exploring-heterogeneity}{%
+\subsubsection{Exploring Heterogeneity}\label{exploring-heterogeneity}}
+
+Another use case is to explore whether a population is comprised of
+latent subpopulations, and if so, how many. The primary focus in this
+use case is on class enumeration. For example,
+one study explored heterogeneity in the population of
+military service members and veterans with prior suicide attempts, and
+identified two latent profiles characterized by high versus low suicide
+risk based on self-reported suicidal ideation and information about
+prior attempted methods and severity (Gromatsky et al., 2022).
+Knowing that the population of suicidal service members is heterogeneous
+can inform future research, policy, and care provided. Appendix B offers
+a tutorial for exploratory LCA, based on the aforementioned study of the
+distribution of ocean microplastics (Alkema et al., 2022).
 
@@
 
-LCA can also be used as a measurement model, similar to confirmatory factor analysis.
-This is useful when class membership is measured imperfectly by several indicators:
-LCA can partial out measurement error and restore most likely class membership.
-For example, a recent study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
-While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
-A mixture model was used to estimate ``error corrected employment status'', using self-report and registry data as indicators.
-Among the outcomes of interest were random and systematic measurement error in both indicators,
-which reveal strengths and weaknesses of both types of assessment.
-
-This example also illustrates that LCA allows us to evaluate the reliability of different indicators (Geiser \& Wurpts, 2014).
-A continuous indicator that discriminates well between classes will show large differences in class-specific means;
-an ordinal indicator will show very high or very low conditional response probabilities for some classes but not for others.
-This information can be used to select the indicators that are most diagnostic of class membership.
-For an example of LCA with a focus on differential item functioning,
-consider a study on university admission tests of Saudi Arabian students (Tsaousis, Sideridis, \& AlGhamdi, 2020).
+LCA can also be used as a measurement model, similar to confirmatory
+factor analysis. This is useful when class membership is measured
+imperfectly by several indicators: LCA can partial out measurement error
+and restore most likely class membership. For example, one study found
+discrepancies between self-reported employment status and official
+register data (Pavlopoulos \& Vermunt, 2015). While it might be tempting
+to assume that register data is free of error, this assumption was shown
+to be false. A mixture model was used to estimate employment status,
+correcting for measurement error, from both self-report and registry
+data. The study investigated random and systematic measurement error in
+both indicators, which reveal strengths and weaknesses of both types of
+assessment.
+
+This example also illustrates that LCA allows us to evaluate the
+reliability of different indicators (Geiser \& Wurpts, 2014).
+A
+continuous indicator that discriminates well between classes will show
+large differences in class-specific means; an ordinal indicator will
+show very high or very low conditional response probabilities for some
+classes but not for others. This information can be used to select the
+indicators that are most diagnostic of class membership. For an example
+of LCA with a focus on differential item functioning, consider a study
+on university admission tests of Saudi Arabian students
+(Tsaousis, Sideridis, \& AlGhamdi, 2020).
 
@@
 
-LCA can also be used to reduce the dimensionality of data
-by reducing many variables to a few prototypes.
-Factor analysis is also sometimes used for dimension reduction,
-but only accounts for linear covariance between indicators,
-whereas LCA can accommodate complex - but discrete - patterns.
-For example, imagine that there exist two indicators - \(x\) and \(y\), and some individuals score high on both, some score low on both, and some score high on \(x\) but low on \(y\).
-Factor analysis would only be able to capture the differences between the first two groups well; LCA captures differences between all three.
-Note that this is a pragmatic use of LCA;
-there is no theory about the existence of a categorical latent variable (cf.~Testing theory).
-Consequently, the class solution should be treated as descriptive of the sample,
-and not be ``reified'' - treated as evidence of the existence of a categorical latent variable.
-The use of LCA for dimension reduction is common in longitudinal research,
-where one developmental process is summarized using latent class growth analysis (LCGA),
-and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
-For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy (Van Lissa et al., 2015).
-These groups were then used as moderator of a latent growth analysis of adolescent- and parent-reported conflict.
-Results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than the other groups,
-and that the low empathy group had more conflict with parents than the other groups.
-Note that both these key findings are non-linear;
-modeling them in a single-group model would be difficult.
-LCGA greatly simplifies model specification and interpretation,
-albeit at a cost of some loss of information and, potentially, generalizability.
-
-The need for dimension reduction can be especially pertinent when using ordinal indicators.
-For example, one study measured
-fifteen dichotomous health indicators among injured military personnel (MacGregor et al., 2021).
-Combinatorics informs us that fifteen
+LCA can also be used to reduce high-dimensional data to just a few
+prototypes. Factor analysis is also sometimes used for dimension
+reduction, but only accounts for linear covariance between indicators,
+whereas LCA can accommodate complex - but discrete - patterns. For
+example, imagine that there exist two indicators - \(x\) and \(y\), and some
+individuals score high on both, some score low on both, and some score
+high on \(x\) but low on \(y\). Factor analysis would only be able to
+capture the differences between the first two groups well; LCA captures
+differences between all three. Note that this is a pragmatic use of LCA;
+there is no theory about the existence of a categorical latent variable
+(cf.~Testing theory). Consequently, the class solution should be treated
+as descriptive of the sample, and not be ``reified'' - treated as evidence
+of the existence of a categorical latent variable.
+
+The use of LCA for dimension reduction is common in longitudinal
+research, where one developmental process is summarized using latent
+class growth analysis (LCGA), and the resulting categories are then used
+as a moderator of another developmental process in a multi-group model.
+For example, one study conducted a LCGA of adolescent empathy
+development on two dimensions of empathy and found three groups: high,
+average, and low empathy (Van Lissa et al., 2015). These groups were
+then used as moderator of a latent growth analysis of adolescent- and
+parent-reported conflict. Results showed that the high-empathy group
+evidenced greater adolescent-parent agreement about the incidence of
+conflict than the other groups, and that the low empathy group had more
+conflict with parents than the other groups. Note that both these key
+findings are non-linear; modeling them in a single-group model would be
+difficult. LCGA greatly simplifies model specification and
+interpretation, albeit at a cost of some loss of information and,
+potentially, generalizability. Appendix C is a tutorial on LCGA, based
+on ongoing research on trajectories of depressive symptoms in military
+service members after deployment.
+
+The need for dimension reduction can be especially pertinent when using
+ordinal indicators. For example, one study measured fifteen dichotomous
+health indicators among injured military personnel
+(MacGregor et al., 2021). Combinatorics informs us that fifteen
 dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
-However, LCA was able to reduce response patterns to five clinically meaningful latent classes.
-
-\hypertarget{class-enumeration}{%
-\subsubsection{Class enumeration}\label{class-enumeration}}
-
-Another use case is to explore heterogeneity in a population,
-and determine whether the population is comprised of latent subpopulations, and if so, how many.
-The primary focus in this use case is on class enumeration.
-For example,
-one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
-and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
-(Gromatsky et al., 2022).
+However, LCA was able to reduce response patterns to five clinically
+meaningful latent classes.
 
@@
 
-Another use case is to estimate individuals' unobserved class membership based on observed indicators.
-The shoe size example in Figure \ref{fig:shoedens}
-is a rudimentary illustration of this application.
-In applied research, LCA is often used for classification in clinical contexts.
-For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough (Baughman, Bisgard, Lynn, \& Meade, 2006).
-Another study used LCA to not only classify the sample used to estimate the model, but also future cases {[}REF Zegwaard et al, in prep{]}.
-The study identified four types of care providers among those who supported close kin with mental health problems.
-Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers, to provide more targeted support.
-To obtain predicted class membership based on an existing LCA model,
-the authors computed the likelihood of data from a new participant under the multivariate normal distributions estimated within all latent classes.
-These applications of LCA aid the clinical decision making process,
-and help decide whether additional support is warranted,
-or what type of intervention is appropriate.
+Another use case is to estimate individuals' unobserved class membership
+based on observed indicators. The shoe size example in Figure
+\ref{fig:shoedens} is a rudimentary illustration of this application.
+In applied research, LCA is often used for classification in clinical
+contexts. For example, LCA was used to determine clinical cut-off scores
+for diagnosing whooping cough (Baughman, Bisgard, Lynn, \& Meade, 2006). Another study
+used LCA to not only classify the sample used to estimate the model, but
+also future cases {[}REF Zegwaard et al, in prep{]}. The study identified
+four types of care providers among those who supported close kin with
+mental health problems. Importantly, the LCA model was embedded in an
+interactive app that could be used by healthcare professionals to
+determine the most likely class membership for new care providers, to
+provide more targeted support. To obtain predicted class membership
+based on an existing LCA model, the authors computed the likelihood of
+data from a new participant under the multivariate normal distributions
+estimated within all latent classes.
+These applications of LCA aid the clinical decision making process, and
+help decide whether additional support is warranted, or what type of
+intervention is appropriate.
 
@@
 
-Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (\textbf{baughmanMixtureModelFramework2007?}).
-These models estimate one class to explain the ``true zeros'',
-and another class with a different distribution to explain all oter values, which can include zero.
-Note that this LCA has different models for both classes,
-and thus falls beyond the scope of this paper.
-For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
-A mixture of two normal distributions with distinct variances described the observed distribution better.
+Finally, LCA can be used to deal with data which violate certain
+assumptions. For example, zero-inflated and hurdle models are special
+cases of LCA for dealing data with a very high proportion of zeros
+(extreme right skew) (\textbf{baughmanMixtureModelFramework2007?}). These models
+estimate one class to explain the ``true zeros'', and another class with a
+different distribution to explain all other values, which can include
+zero. Note that this LCA has different models for both classes, and thus
+falls beyond the scope of this paper.
+For instance, the aforementioned study on marine plastic noted that many
+prior studies had treated the distribution of particles as
+(multivariate) normal, despite clear deviations from normality
+(Alkema et al., 2022). A mixture of two normal distributions with
+distinct variances described the observed distribution better.
 
@@
 \ref{tab:checkest}, are rooted in existing recommendations for best
-practices for estimating specific subtypes of LCA,
-including latent
+practices for estimating specific subtypes of LCA, including latent
 class growth analysis (Van De Schoot, Sijbrandij, Winter, Depaoli, \& Vermunt, 2017) and latent
 class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
-These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.
+These were generalized to be relevant to all types of LCA, and updated
+to current best practices, as explained below.
 
@@
 patterns and violations of assumptions that had not been considered
-prior to data collection.
-The function \texttt{descriptives(data)} provides extensive descriptive statistics,
-including the number of unique values,
-variance of continuous and categorical variables,
-missingness, skew and kurtosis (peakedness of the distribution).
-However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
-Plots convey additional information; specifically, density plots for continuous variables, and bar charts
-for categorical ones.
-Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.
-
-Particular points of attention include measurement level of the indicators.
-Recall that \texttt{tidySEM} can estimate mixture models with either numeric (type \texttt{numeric} and \texttt{integer}) or \texttt{ordered} indicators.
-The column \texttt{type} should thus only contain these types.
-Also check that the number of \texttt{unique} values match the intended measurement level.
-Numeric variables with few unique values are typically better modeled as ordinal.
-Our experience consulting on LCA methods
-and moderating the \texttt{tidyLPA} Google group suggest that the
-misapplication of mixture models to ordinal (e.g., Likert-type)
-indicators is the most common source of user error. Whereas it has been
-argued that some parametric methods are robust when scales with 7+
-response categories are treated as continuous (e.g., Norman, 2010),
-it is very unlikely
-that an ordinal variable can be modeled as a \emph{mixture} of multiple
-normal distributions.
-To illustrate the problem, consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
-Each class-specific mean would describe a single response
-category,
-and a class-specific variance component would be empirically underidentified.
-In sum, variables with few unique values might not be suitable for Gaussian mixture modeling;
-LCA with ordinal indicators is more appropriate.
+prior to data collection. The function \texttt{descriptives(data)} provides
+extensive descriptive statistics, including the number of unique values,
+variance of continuous and categorical variables, missingness, skew and
+kurtosis (peakedness of the distribution). However, sample-level
+descriptive statistics are of limited value when the goal of a study is
+to identify subsamples using LCA. Plots convey additional information;
+specifically, density plots for continuous variables, and bar charts for
+categorical ones. Density plots can also aid in the choice of the number
+of classes, as explained in the section on visualization.
+
+Particular points of attention include measurement level of the
+indicators. Recall that \texttt{tidySEM} can estimate mixture models with
+either numeric (type \texttt{numeric} and \texttt{integer}) or \texttt{ordered} indicators.
+The column \texttt{type} should thus only contain these types. Also check that
+the number of \texttt{unique} values match the intended measurement level.
+Numeric variables with few unique values are typically better modeled as
+ordinal. Our experience consulting on LCA methods and moderating the
+\texttt{tidyLPA} Google group suggest that the misapplication of mixture models
+to ordinal (e.g., Likert-type) indicators is the most common source of
+user error. Whereas it has been argued that some parametric methods are
+robust when scales with 7+ response categories are treated as continuous
+(e.g., Norman, 2010), it is very unlikely that an ordinal
+variable can be modeled as a \emph{mixture} of multiple normal distributions.
+To illustrate the problem, consider what happens when estimating a
+5-class solution on a single 5-point Likert scale indicator. Each
+class-specific mean would describe a single response category, and a
+class-specific variance component would be empirically underidentified.
+In sum, variables with few unique values might not be suitable for
+Gaussian mixture modeling; LCA with ordinal indicators is more
+appropriate.
 
@@
 
-LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
-Numeric variables should have an interval measurement level,
-such that it is sensible to estimate means and (co)variances.
-For the remaining three measurement levels,
-some preprocessing is required.
-The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
-To correctly specify ordered variables,
-such as Likert scales
-use the function \texttt{OpenMx::mxFactor()}.
-Binary variables,
-such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
-Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.
-
-A second consideration in data preprocessing is that models with indicators on different scales may present with convergence issues.
-The reason for this is that the parameter space is high-dimensional with many potential local optima.
-If one indicator is on a much larger scale than another, the parameter space may be more extensively explored for that parameter.
-The problem can be resolved by transforming the indicators to place them on (roughly) similar scales.
-The most common transformation is standardization, which involves subtracting the mean of each variable from each value and dividing by the standard deviation.
+LCA can accommodate variables of different measurement levels:
+continuous, binary, nominal, and ordinal. Numeric variables should have
+an interval measurement level, such that it is sensible to estimate
+means and (co)variances. For the remaining three measurement levels,
+some preprocessing is required. The exact data types accepted by
+\texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}. To
+correctly specify ordered variables such as Likert scales, use the
+function \texttt{OpenMx::mxFactor()}. Binary variables, such as presence or
+absence of some symptom, are also treated as ordinal, with one category
+coded as zero and the other as one. Nominal variables are converted to
+binary indicators of group membership via dummy coding, for example,
+using \texttt{mx\_dummies()}.
+
+A second consideration in data preprocessing is that models with
+indicators on different scales may present with convergence issues. The
+reason for this is that the parameter space is high-dimensional with
+many potential local optima. If one indicator is on a much larger scale
+than another, the parameter space may be more extensively explored for
+that parameter. The problem can be resolved by transforming the
+indicators to place them on (roughly) similar scales. The most common
+transformation is standardization, which involves subtracting the mean
+of each variable from each value and dividing by the standard deviation.
 In R, this transformation is applied using the \texttt{scale()} function.
-Standardization retains all available information,
-and model parameters can be transformed back to the original scale of the indicators by reversing the transformation.
-The transformed variable has a mean of zero and a standard deviation of one.
-Note that standardization is typically not necessary if indicators are already on the same, or similar, scales.
-
-A third consideration in preprocessing relates to the univariate distributions of indicators.
-For example, the \texttt{plas\_depression} data included with \texttt{tidySEM} show extreme right skew.
-These data assess depressive symptoms;
-many respondents report (near) zero scores,
-and some respondents report elevated scores.
-To model skewed variables like this,
-it makes sense to preprocess the data by transforming the raw scores to reduce skew.
-Alternatively (or additionally), one can specify a model with free variances across classes:
-this might result in one class with a near-zero mean and small variance to account for the fact that many respondents have no depressive symptoms,
-and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.
+Standardization retains all available information, and model parameters
+can be transformed back to the original scale of the indicators by
+reversing the transformation. The transformed variable has a mean of
+zero and a standard deviation of one. Note that standardization is
+typically not necessary if indicators are already on the same, or
+similar, scales.
+
+A third consideration in preprocessing relates to the univariate
+distributions of indicators. For example, the \texttt{plas\_depression} data
+show extreme right skew in depressive symptoms: most respondents report
+few symptoms, and few respondents report many symptoms. Similarly, the
+\texttt{alkema\_microplastics} data are extremely right-skewed: Very small
+plastic fragments abound in the oceans, and large pieces are rare. These
+distributional characteristics must be taken into account either in
+preprocessing, or model specification. In preprocessing, the data can be
+transformed to reduce skew. After reducing skew, a simpler LCA model may
+fit the data well. Alternatively, one can specify a model that
+accommodates the distributional idiosyncrasies. For example, allowing
+free variances across classes might result in one class with a near-zero
+mean and small variance to account for the fact that most respondents
+have few depressive symptoms, and one class with an elevated mean and
+large variance to account for all respondents with non-zero symptoms.
 
@@
 missing data and reporting how missingness was handled
-(Van De Schoot et al., 2017).
-While we concur that investigating missingness is due diligence,
-it is important to emphasize that missingness is adequately handled by default in \texttt{OpenMx}, which is the computational backend of \texttt{tidySEM}.
-Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
-information without imputing missing values (\textbf{endersRelativePerformanceFull2001?}).
-FIML is a best-practice solution for handling missing data;
-at least on par with multiple imputation (Lee \& Shi, 2021).
-Multiple imputation is less suitable to LCA for two reasons. First,
-because LCA methods are often computationally expensive,
-and conducting them on multiple imputed datasets may be unfeasible.
-Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population.
-
-Three types of missingness have
-been distinguished in the literature (Enders, 2022; Rubin, 1976): random missingness (MCAR);
-missingness contingent on \emph{observed} variables that can be known and controlled for (MAR);
-and missingness related to unobserved variables,
-which can be neither known nor controlled for (MNAR).
-A so-called ``MCAR'' test is often reported (Jamshidian \& Jalal, 2010),
-but note that this name is somewhat misleading,
-as a significant test indicates that missingness is MAR,
-but a non-significant test statistic does not distinguish between MCAR or MNAR.
-As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (Little, 1988).
-This assumption is tenuous in the context of LCA.
-A non-parametric MCAR test may be more suitable (Jamshidian \& Jalal, 2010).
-For this tutorial,
-the lead author has implemented this test in the \texttt{mice} package as \texttt{mice::mcar()}.
-
-To conclude; our recommendation is to inspect and report the proportion of missingness per variable (e.g., using \texttt{tidySEM::descriptives()}),
-and to test for MAR using \texttt{mice::mcar()}.
-As FIML estimation assumes that missingness is either MCAR or MAR,
-one would proceed with FIML regardless of the outcome of the test.
-
-One minor concern is that \texttt{tidySEM} uses a different algorithm to determine starting values for complete data (K-means) and for missing data (hierarchical clustering).
-This should not be a problem, however, as the global optimizer used to estimate the model should be independent of starting values (\textbf{coranaMinimizingMultimodalFunctions1987?}).
-A second concern is that FIML is not suitable for cases with complete missing data.
-If such cases must be included, their values must be imputed first.
+(Van De Schoot et al., 2017). While we concur that
+investigating missingness is due diligence, it is important to emphasize
+that missingness is adequately handled by default in \texttt{OpenMx}, which is
+the computational backend of \texttt{tidySEM}. Its Full Information Maximum
+Likelihood (FIML) estimator makes use of all available information
+without imputing missing values, by estimating the model based on all
+available information for each subject
+(Enders, 2022; \textbf{endersRelativePerformanceFull2001?}). FIML is a
+best-practice solution for handling missing data; at least on par with
+multiple imputation (Lee \& Shi, 2021). Multiple imputation is less
+suitable to LCA for two reasons. First, because LCA methods are often
+computationally intensive, and conducting them on several imputed
+datasets multiplies the load. Second, because many multiple imputation
+methods assume normality; an assumption likely to be violated when data
+come from a heterogeneous population.
+
+Three types of missingness have been distinguished in the literature
+(Enders, 2022; Rubin, 1976): random missingness
+(MCAR); missingness contingent on \emph{observed} variables that can be known
+and controlled for (MAR); and missingness related to unobserved
+variables, which can be neither known nor controlled for (MNAR). A
+so-called ``MCAR'' test is often reported (Jamshidian \& Jalal, 2010), but
+note that this name is somewhat misleading, as a significant test
+indicates that missingness is MAR, but a non-significant test statistic
+does not distinguish between MCAR or MNAR. As the classic MCAR test
+relies on the comparison of variances across groups with different
+patterns of missing data, it assumes normality (Little, 1988). This
+assumption is tenuous in the context of LCA. A non-parametric MCAR test
+may be more suitable (Jamshidian \& Jalal, 2010). For this tutorial, the
+lead author has implemented this test in the \texttt{mice} package as
+\texttt{mice::mcar()}. Our recommendation is to inspect and report the
+proportion of missingness per variable (e.g., using
+\texttt{tidySEM::descriptives()}), and report the \texttt{mice::mcar()} test. As FIML
+estimation assumes that missingness is either MCAR or MAR, one would
+proceed with FIML regardless of the outcome.
+
+One minor concern is that \texttt{tidySEM} uses a different algorithm to
+determine starting values for complete data (K-means) and for missing
+data (hierarchical clustering). This should not be a problem, however,
+as the global optimizer used to estimate the model should be independent
+of starting values (\textbf{coranaMinimizingMultimodalFunctions1987?}). A second
+concern is that FIML is not suitable for cases with complete missing
+data. If such cases must be included, their values must be imputed
+first.
 
@@
 
-Any LCA model should be specified based on theoretical and empirical considerations.
-The model parameters depend on the measurement level of the indicators.
-Continuous indicators can be parametrized in terms of means, variances, and covariances.
-LCGA and GMM have these same parameters,
-but with respect to the latent growth variables.
-Ordinal indicators can be parametrized differently;
-the default approach in \texttt{tidySEM} assumes that each ordinal indicator reflects an underlying standard normal distribution, \(N(\mu = 0, \sigma = 1)\).
-The parameters are ``thresholds'' that correspond to quantiles of this distribution, estimated based on the proportion of individuals in each of the
-response categories of the indicator variable.
-For example, a binary
-indicator has a single threshold that distinguishes the two response
-categories. If responses are distributed 50/50, then the corresponding
-threshold would be \(t_1 = 0.00\). If the responses are distributed 60/40,
-then the resulting threshold would be \(t_1 = 0.25\).
-All of the aforementioned parameters can be freely estimated,
-or constrained across classes, or fixed (e.g., to zero).
-
-Prior literature emphasized the importance of considering alternative model specifications (\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}).
-However, what has remained underemphasized is that alternative model specifications should be approached differently for exploratory versus confirmatory analyses (see Use Cases).
-Most prior literature has emphasized exploratory LCA (e.g., Nylund, Asparouhov, \& Muthén, 2007).
-In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
-A ``final'' model is then selected based on a combination of fit indices,
-significance tests,
-and interpretability (see Class Enumeration).
-
-Confirmatory LCA does not require such an extensive search over the model space.
-If hypothesized models have been specified a priori,
-and possibly even preregistered,
-it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
-Nonetheless, an important reason to consider alternative model specifications is that LCA lacks objective fit indices.
-The fit of a theoretical model can thus only be assessed by comparison to other candidates.
-For this reason, the theoretical model should still be compared to a few others.
-A one-class model is a sensible benchmark,
-to test whether theoretical model fits better than a model without latent classes.
-Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
-If the theoretical model has much better fit than other models,
-this provides evidence for the theory.
-If differences in fit are small, however, this may not do not give cause to reject a theoretical model.
-The theoretical model may still be preferred, for example, to ensure comparability with prior work.
-The fact that the theoretical model received little differential support should of course be discussed,
-along with potential implications for future work.
-
-A second consideration in model specification is the risk of overfitting (\textbf{hastieElementsStatisticalLearning2009?}).
-Overfitting occurs when a model has many parameters relative to the number of observations.
-Such a overparametrized model captures much idiosyncratic noise in a sample,
-limiting its generalizability to new samples.
-As explained before, the number of parameters in LCA scales with the number of estimated classes.
-Consequently, LCA with many classes have a potentially very high number of parameters.
-In Bayesian mixture modeling,
-it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (\textbf{rousseauAsymptoticBehaviourPosterior2011?}).
-Although it is not known whether frequentist LCA exhibits the same behavior,
-this suggests that low class counts might indicate overfitting.
-The section on Class Enumeration addresses further ways to guard against overfitting.
-
-Models can be specified in \texttt{tidySEM} in different ways.
-The wrapper functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
-The function \texttt{mx\_profiles()} has two arguments to free elements of the covariance matrix: \texttt{variances} and \texttt{covariances}.
-By default, \texttt{mx\_profiles()} fixes covariances to zero, and thus performs latent profile analysis.
-Finally, function \texttt{mx\_mixture()} accepts a more flexible \texttt{model} argument, which can be a character string specifying a model in \texttt{lavaan} syntax,
-which is translated to \texttt{OpenMx} models via the function \texttt{as\_ram()},
-or a list of \texttt{OpenMx} models.
+In principle, LCA can be based on any parametric model. At present,
+\texttt{tidySEM} only facilitates estimation of models that can be specified in
+\texttt{lavaan} syntax. The available model parameters depend on the
+measurement level of the indicators. Continuous indicators can be
+parametrized in terms of means, variances, and covariances. LCGA and GMM
+have these same parameters, but with respect to the latent growth
+variables. Ordinal indicators can be parametrized differently; the
+default approach in \texttt{tidySEM} assumes that each ordinal indicator
+reflects an underlying standard normal distribution,
+\(N(\mu = 0, \sigma = 1)\). The parameters are ``thresholds'' that
+correspond to quantiles of this distribution, estimated based on the
+proportion of individuals in each of the response categories of the
+indicator variable. For example, a binary indicator has a single
+threshold that distinguishes the two response categories. If responses
+are distributed 50/50, then the corresponding threshold would be
+\(t_1 = 0.00\). If the responses are distributed 60/40, then the resulting
+threshold would be \(t_1 = 0.25\). All of the aforementioned parameters
+can be freely estimated, or constrained across classes, or fixed (e.g.,
+to zero).
+
+Prior literature emphasized the importance of considering alternative
+model specifications
+(\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}). What has remained
+underemphasized is that alternative model specifications should be
+approached differently for exploratory versus confirmatory analyses (see
+Use Cases). Most prior literature has emphasized exploratory LCA (e.g., Nylund, Asparouhov, \& Muthén, 2007). In this context, model specification typically
+consists of an exhaustive grid search along several model specifications
+and varying numbers of classes. A ``final'' model is then selected based
+on a combination of fit indices, significance tests, and
+interpretability (see Class Enumeration).
+
+Confirmatory LCA does not require such an extensive search over the
+model space. If hypothesized models have been specified a priori, and
+possibly even preregistered, it makes sense to focus on these models,
+rather than proceed in a purely data-driven manner. Nonetheless, the
+theoretical model specification should still be compared to a few
+alternatives because LCA lacks objective fit indices. The fit of a
+theoretical model can thus only be assessed by comparison to other
+candidates. A one-class model is a sensible benchmark to test whether
+the theoretical model fits better than a model that assumes a
+homogeneous population. Other candidates for comparison include
+competing theoretical models and those with a few more or less classes
+than the hypothesized number. If the theoretical model has much better
+fit than other models, for example, based on IC weights, this provides
+evidence for the theory. If another model fits substantially better,
+this provides evidence against the theory. If fit differences are small,
+the theoretical model may still be preferred, as theory is a relevant
+consideration in model selection (\textbf{jungIntroductionLatentClass2008?}). The
+fact that the theoretical model received little differential support
+should of course be discussed.
+
+A second consideration in model specification is the risk of overfitting
+(\textbf{hastieElementsStatisticalLearning2009?}). Overfitting occurs when a
+model has many parameters relative to the number of observations,
+causing it to capture idiosyncratic noise in the sample and limiting its
+generalizability to new samples. As explained before, the number of
+parameters in LCA scales with the number of estimated classes.
+Consequently, LCA with many classes have a potentially very high number
+of parameters. In Bayesian mixture modeling, it was observed that if the
+estimated number of classes exceeded the true number, the number of
+observations assigned to these classes tended towards zero as sample
+size increased (\textbf{rousseauAsymptoticBehaviourPosterior2011?}). Although it
+is not known whether frequentist LCA exhibits the same behavior, low
+class counts are still an indicator of potential overfitting. The
+section on Class Enumeration addresses ways to guard against
+overfitting.
+
+Models can be specified in \texttt{tidySEM} in different ways. The wrapper
+functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for
+continuous and ordinal indicators respectively, without needing to
+specify model syntax. The function \texttt{mx\_profiles()} has two arguments to
+free elements of the covariance matrix: \texttt{variances} and \texttt{covariances}.
+By default, \texttt{mx\_profiles()} fixes covariances to zero, and thus performs
+latent profile analysis. Finally, function \texttt{mx\_mixture()} accepts a more
+flexible \texttt{model} argument, which can be a character string specifying a
+model in \texttt{lavaan} syntax, which is translated to \texttt{OpenMx} models via the
+function \texttt{as\_ram()}, or a list of \texttt{OpenMx} models.
 
@@
 LCA parameters and model fit statistics can be estimated in different
-ways.
-The choice of estimator depends on the presence of missing
-values, sample size, number of indicators, and available computational
-resources (Weller, Bowen, \& Faubert, 2020).
-The \texttt{tidySEM} package uses maximum likelihood estimation (ML), which can be conceptually understood as follows:
-Imagine we are estimating two
-parameters, e.g.~the class-specific means \(\mu_c\) on a continuous indicator for two classes
-(ignoring the variance for now).
-ML attempts to find a combination of values for these two parameters that minimizes minus two times the log likelihood (\(-2LL\)) of the data.
-To understand the ML optimization procedure,
-imagine a three-dimensional landscape:
-The X and Y dimensions represent potential values of the class-specific means, \(X = \mu_1\) and \(Y = \mu_2\), and the Z-dimension is determined by \(Z = -2LL\).
-The optimizer must find the
-deepest ``valley'' in this landscape,
-which reflects the combination of \(\mu_1\) and \(\mu_2\) that minimizes the \(-2LL\).
-The ML
-optimizer behaves somewhat like a marble dropped in this landscape. It
-is dropped at some random point in space, and rolls into the nearest valley (a combination of values for \(\mu_1\) and \(\mu_2\) that has a low \(-2LL\)).
-The challenge is that LCA models are often highly parametrized.
-This means that the landscape is not three-dimensional, but (very) high-dimensional,
-with many valleys and sparse data to provide information about which valley is deepest.
-If the marble rolls into one of these valleys,
-it will settle there and not climb out again.
-The estimator has ``converged''.
-
-A unique challenge in LCA is that not only model parameters,
-but also the proportion of observations in each class must be estimated.
-Parameters in each class are estimated from the same dataset,
-but for each class,
-observations are weighted by their probability of being in that class - which is itself computed based on the class-specific parameters.
-The resulting chicken-and-egg problem is solved by the expectation-maximization (EM) algorithm:
-It alternates between estimating model parameters,
+ways. The choice of estimator depends on the presence of missing values,
+sample size, number of indicators, and available computational resources
+(Weller, Bowen, \& Faubert, 2020). LCA in \texttt{tidySEM} uses maximum likelihood
+estimation (ML), which can be conceptually understood as follows:
+Imagine we are estimating two parameters, e.g.~the class-specific means
+\(\mu_c\) on a continuous indicator for two classes (ignoring the variance
+for now). ML attempts to find a combination of values for these two
+parameters that minimizes minus two times the log likelihood (\(-2LL\)) of
+the data.
+To understand the ML optimization procedure, imagine a three-dimensional
+landscape: The X and Y dimensions represent potential values of the
+class-specific means, \(X = \mu_1\) and \(Y = \mu_2\), and the Z-dimension
+is determined by \(Z = -2LL\). The optimizer must find the deepest
+``valley'' in this landscape, which reflects the combination of \(\mu_1\)
+and \(\mu_2\) that minimizes the \(-2LL\). The ML optimizer behaves somewhat
+like a marble dropped in this landscape. It is dropped at some random
+point in space, and rolls into the nearest valley (a combination of
+values for \(\mu_1\) and \(\mu_2\) that has a low \(-2LL\)). The challenge is
+that LCA models are often highly parametrized. This means that the
+landscape is not three-dimensional, but (very) high-dimensional, with
+many valleys and sparse data to provide information about which valley
+is deepest. If the marble rolls into one of these valleys, it will
+settle there and not climb out again. The estimator has ``converged''.
+
+A unique challenge in LCA is that not only model parameters, but also
+class membership must be estimated. Parameters in all classes are
+estimated from the same dataset, but for each class, observations are
+weighted by their probability of being in that class - which is in turn
+computed based on the class-specific parameters. The resulting
+chicken-and-egg problem is solved by the expectation-maximization (EM)
+algorithm: It alternates between estimating model parameters,
 calculating posterior class probabilities from those parameters,
-re-estimating model parameters using those new class probabilities as weights, and so on, until both have converged.
-
-A risk with this approach is that the optimizer might get stuck in a shallower valley (a ``local optimum''),
-and never discovers the deepest valley (the ``global optimum'', or best solution).
-One solution to this problem is to drop many marbles at random places,
-compare their final \(-2LL\) values,
-choose the solution
-with the lowest \(-2LL\), and make sure that several marbles replicated
-this solution.
-This is the ``random starts'' approach implemented in, e.g., Mplus.
+re-estimating model parameters using those new class probabilities as
+weights, and so on, until both have converged.
+
+A risk with this approach is that the optimizer might get stuck in a
+shallower valley (a ``local optimum''), and never discovers the deepest
+valley (the ``global optimum'', or best solution). One solution to this
+problem is to drop many marbles at random places, compare their final
+\(-2LL\) values, choose the solution with the lowest \(-2LL\), and make sure
+that several marbles replicated this solution. This is the ``random
+starts'' approach implemented in, e.g., Mplus.
 
@@
 computationally expensive to run this many replications. Moreover,
-because the algorithm begins with random starting values,
-it is inefficient because many marbles are likely to start far away from any valley.
-Additionally, if all sets of starting values are close to a specific local optimum,
-there is a risk that the global optimum is never found.
-An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing (SA, \textbf{coranaMinimizingMultimodalFunctions1987?}).
-SA does not behave quite like the marbles of the random starts approach;
-instead, it iteratively considers some ``destination'' in the landscape,
-and compares its \(-2LL\) to the current one.
-If the destination \(-2LL\) is lower,
-the estimator moves there.
-However, if the destination \(-2LL\) is \emph{higher},
-it still moves there occasionally, with a certain probability.
-Think of this as occasionally flicking the marble back out of valleys,
-to see if it rolls back into the current valley, or finds another,
-even deeper one.
-This property allows SA it to escape local optima and find the global optimum.
-
-Since SA is a global optimizer, it should be independent of starting values.
-With this in mind, it is efficient to start estimation from a ``reasonable solution'', instead of a random starting point.
-For example, if we assume that the different classes are likely to have different mean values on the indicators,
-then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids (see also \textbf{biernackiChoosingStartingValues2003?}).
-By default, \texttt{tidySEM} derives starting values using K-means clustering {[}REF{]},
-computing the mixture model starting values by treating the K-means solution as multi-group model.
-Next, SA is used to find the global optimum solution.
-Finally, SA is followed up with a short run of the ML algorithm,
-as ML inherently produces an asymptotic
-covariance matrix for the parameters that can be used to compute
-standard errors.
-Note that these defaults can be manually overridden.
-
-In practice, models do not always converge.
-Reasons may include local optima,
-bad starting values,
-incorrect model specification (the model may be too complex for the data, or conversely, too simple to capture important patterns),
-or poor data quality (small data sets, or distributions that are difficult to model).
-Convergence problems may be indicated by errors and warning messages from the estimator,
-or they may be suspected, for example,
-when results change if the analyses are reproduced.
-
-Nonconvergence may be due to bad model specification.
-Consider, for example, an LCA with multiple ordinal indicators.
-If certain combinations of responses never occur together,
-in general or within one of the classes,
-this can result in extreme conditional item probabilities (exactly 0 or 1).
-Such extreme values impede model convergence.
-This problem could be addressed by trying alternative model specifications (e.g., fewer classes),
-or by merging categories of the ordinal indicators with very few responses.
-Similarly problems occur with continuous indicators.
-For instance, consider a zero-inflated indicator and a two class LPA.
-To account for the excess zeroes, one class will have a mean of zero.
-It is difficult to estimate the standard deviation for this class, however, because all of its members have the same value.
-If standard deviations are constrained to be equal across classes,
-this model will likely not converge.
-This problem can be addressed by considering more appropriate model specifications (e.g., free class variances).
-
-When poor data quality is to blame, the only solution is to collect more, or higher quality, data.
-Nonconvergence due to bad starting values and local optima may be overcome by re-estimating the model, while permuting the starting values.
-A non-converged model can be updated an re-run using \texttt{OpenMx} functions to aid model convergence,
-like \texttt{mxTryHard()} or \texttt{mxTryHardWideSearch()} for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal indicators.
-Often, these functions return an acceptable solution for models with convergence issues.
-
-\hypertarget{class-enumeration-1}{%
-\subsubsection{Class Enumeration}\label{class-enumeration-1}}
-
-Class enumeration refers to the process of determining the appropriate number of classes.
-It is approached differently for exploratory versus confirmatory LCA.
-In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes \(K\).
-The number \(K\) may be chosen a-priori or dictated by the data:
-It is sensible to limit \(K\) to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
-Above a certain number of classes,
-convergence problems typically arise,
-or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
-In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
-the proportion of cases assigned to the excessive classes converges to zero as the sample size increases (\textbf{rousseauAsymptoticBehaviourPosterior2011?}).
-It is not known whether this also happens in frequentist LCA;
-nonetheless, near-zero class memberships should be taken into account when determining the maximum \(K\).
-
-In confirmatory LCA,
-the set of models with different numbers of classes to be compared is typically smaller.
-Its purpose is to provide context for the relative fit of the theoretical model, or to compare multiple theoretical models.
-
-In class enumeration, the set of models is compared on several statistical and substantive criteria.
-In exploratory LCA, these criteria may be decisive in selecting a final model;
-in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
-Statistical criteria to consider in class enumeration include model fit indices and
+because the algorithm begins with random starting values, it is
+inefficient because many marbles are likely to start far away from any
+valley. Additionally, if all sets of starting values are close to a
+specific local optimum, there is a risk that the global optimum is never
+found. An alternative solution that overcomes these challenges is to use
+a global optimizer, like simulated annealing (SA, \textbf{coranaMinimizingMultimodalFunctions1987?}). SA does not behave quite like
+the marbles of the random starts approach; instead, it iteratively
+considers some ``destination'' in the landscape, and compares its \(-2LL\)
+to the current one. If the destination \(-2LL\) is lower, the estimator
+moves there. However, if the destination \(-2LL\) is \emph{higher}, it still
+moves there occasionally, with a certain probability. Think of this as
+occasionally flicking the marble, to see if it rolls back into the
+current valley, or finds another, even deeper one. This property allows
+SA it to escape local optima and find the global optimum.
+
+Since SA is a global optimizer, it should be independent of starting
+values. With this in mind, it is efficient to start estimation from a
+``reasonable solution'', instead of a random starting point. For example,
+if we assume that the different classes are likely to have different
+mean values on the indicators, then a nonparametric clustering
+algorithm - like K-means or hierarchical clustering - can be used to
+determine these cluster centroids (see also \textbf{biernackiChoosingStartingValues2003?}). By default, \texttt{tidySEM} derives
+starting values using K-means clustering {[}REF{]}, computing the mixture
+model starting values by treating the K-means solution as multi-group
+model. Next, SA is used to find the global optimum solution. Finally, SA
+is followed up with a short run of the ML algorithm, as ML inherently
+produces an asymptotic covariance matrix for the parameters that can be
+used to compute standard errors. Note that these defaults can be
+manually overridden.
+
+In practice, models do not always converge. Reasons may include local
+optima, bad starting values, incorrect model specification (the model
+may be too complex for the data, or conversely, too simple to capture
+important patterns), or poor data quality (small data sets, or
+distributions that are difficult to model). Convergence problems may be
+indicated by errors and warnings from the estimator, or they may be
+suspected, for example, when results change if the analyses are
+reproduced.
+
+Nonconvergence may be due to bad model specification. Consider, for
+example, an LCA with multiple ordinal indicators. If certain
+combinations of responses never occur together, in general or within one
+of the classes, this can result in extreme conditional item
+probabilities (exactly 0 or 1). Such extreme values impede model
+convergence. This problem could be addressed by trying alternative model
+specifications (e.g., fewer classes), or by merging categories of the
+ordinal indicators with very few responses. Similarly problems occur
+with continuous indicators. For instance, consider a zero-inflated
+indicator and a two class LPA. To account for the excess zeroes, one
+class will have a mean of zero. It is difficult to estimate the standard
+deviation for this class, however, because all of its members have the
+same value. If standard deviations are constrained to be equal across
+classes, this model will likely not converge. This problem can be
+addressed by considering more appropriate model specifications (e.g.,
+free class variances).
+
+Nonconvergence due to bad starting values can be overcome by manually
+specifying starting values: Specify the argument \texttt{run\ =\ FALSE} in the
+call to \texttt{mx\_mixture()} and related functions. This returns the model
+with default starting values, which can be updated. Nonconvergence and
+local optima may also be overcome by re-estimating the model while
+permuting the starting values. The \texttt{OpenMx} package contains a family of
+functions for this purpose: \texttt{mxTryHard()} and \texttt{mxTryHardWideSearch()}
+for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal
+indicators. When poor data quality is to blame, the only solution is to
+collect more, or higher quality, data.
+
+\hypertarget{class-enumeration}{%
+\subsubsection{Class Enumeration}\label{class-enumeration}}
+
+Class enumeration refers to the process of determining the appropriate
+number of classes. It is approached differently for exploratory versus
+confirmatory LCA. In exploratory LCA, the set of models typically
+consists of a sequence from 1 to the maximum number of classes \(K\). The
+number \(K\) may be chosen a-priori or dictated by the data. It is
+sensible to limit \(K\) to the maximum number of classes that result in
+valid and interpretable solutions (also see Classification Diagnostics).
+Above a certain number of classes, convergence problems typically arise,
+or the proportion of cases assigned to the smallest classes may be low
+relative to the number of class-specific parameters. In Bayesian LCA, it
+has been observed that, when the estimated number of classes exceeds the
+true number, the proportion of cases assigned to the excessive classes
+converges to zero as the sample size increases
+(\textbf{rousseauAsymptoticBehaviourPosterior2011?}). It is not known whether
+this also happens in frequentist LCA; nonetheless, near-zero class
+memberships should be taken into account when determining the maximum
+\(K\).
+
+In confirmatory LCA, the set of models with different numbers of classes
+to be compared is typically smaller. Its purpose is to provide context
+for the relative fit of the theoretical model, or to compare multiple
+theoretical models.
+
+In class enumeration, the set of models is compared on several
+statistical and substantive criteria. In exploratory LCA, these criteria
+may be decisive in selecting a final model; in the confirmatory case,
+the burden of proof is reversed: there must be convincing evidence
+against the theoretical model in order to reject it. Statistical
+criteria to consider in class enumeration include model fit indices and
 likelihood ratio tests (Masyn, 2013; Weller et al., 2020).
-Substantive criteria include convergence and model identification,
-class separability,
-and interpretability.
-We address each of these criteria below.
+Substantive criteria include convergence and model identification, class
+separability, and interpretability. We address each of these criteria
+below.
 
@@
 
-As there are no absolute fit indices for LCA,
-information criteria are typically used to assess relative fit.
-Information criteria are computed from the \(-2LL\),
-which is the most basic fit measure.
-Information criteria add a penalty for model complexity to the \(-2LL\), and thus balance fit and complexity.
-The lower
-the value of an information criterion, the better the overall fit of the
-model.
-The original information criterion is the Akaike Information Criterion (AIC),
-computed as \(-2LL+2p\), where \(p\) is the number of parameters.
-The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
-such that complex models are penalized more in smaller samples.
-Finally,
-the so-called \emph{sample size adjusted} BIC (saBIC) implements a different ad-hoc penalty for sample size,
-\(\ln(\frac{n + 2}{24})\).
-Although a simulation study showed that the saBIC performed well for class enumeration,
-its penalty is not based on theory {[}REF{]}.
-BIC also performs well and is theoretically substantiated;
-as such, it may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
-All three ICs are available in the output of \texttt{table\_fit()}.
+As there are no absolute fit indices for LCA, information criteria are
+typically used to assess relative fit. Information criteria are computed
+from the \(-2LL\), which is the most basic fit measure, testing the
+overall model fit if the model properly represents the data. Information
+criteria add a penalty for model complexity to the \(-2LL\), and thus
+balance fit and complexity. The lower the value of an information
+criterion, the better the overall fit of the model. The original
+information criterion is the Akaike Information Criterion (AIC),
+computed as \(-2LL+2p\), where \(p\) is the number of parameters. The
+Bayesian Information Criterion (BIC) multiplies the penalty for
+complexity with a function of the sample size, \(p \ln(n)\), such that
+complex models are penalized more in smaller samples. Finally, the
+so-called \emph{sample size adjusted} BIC (saBIC) implements a different
+ad-hoc penalty for sample size, \(\ln(\frac{n + 2}{24})\). Although a
+simulation study showed that the saBIC performed well for class
+enumeration, its penalty is not based on theory {[}REF{]}. BIC also performs
+well and is theoretically substantiated; as such, it may be the most
+appropriate information criterion to use for model comparison
+(Masyn, 2013; Nylund-Gibson \& Choi, 2018). All three ICs are
+available in the output of \texttt{table\_fit()}.
 
 Information criteria may occasionally contradict each other, so it is
-important to identify a suitable strategy to reconcile them.
-One option is to select a specific fit index, such as the BIC,
-before analyzing the data.
-Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
-Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016),
-which is implemented in \texttt{tidyLPA} (\textbf{rosenbergTidyLPAPackageEasily2018?}).
-Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs (Nylund-Gibson \& Choi, 2018).
-The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
-Adding this many classes results in substantial decreases of the IC,
-and adding further classes results in relatively small decreases.
-
-Given a specific IC,
-it is also possible to calculate so-called ``IC-weights'',
-which indicate the relative support the data provide for each model in the set (\textbf{wagenmakersAICModelSelection2004?}).
-In \texttt{tidySEM}, IC-weights are calculated by calling \texttt{ic\_weights()} on a model fit table.
-Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
-the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
+important to identify a suitable strategy to reconcile them. One option
+is to select a specific fit index, such as the BIC, before seeing the
+results. This choice can be preregistered to avoid concerns about cherry
+picking. Another option is to always prefer the most parsimonious model
+that has best fit according to any of the available fit indices. Other
+approach is to select several competing model based on the ICs, and
+evaluate the theoretical intepretation. Yet another option is to
+incorporate information from multiple fit indices using the analytic
+hierarchy process (Akogul \& Erisoglu, 2016), which is implemented in
+\texttt{tidyLPA} (\textbf{rosenbergTidyLPAPackageEasily2018?}). Finally, one can use a
+scree plot to determine the inflection point at which additional classes
+contribute little to further decreases of the ICs
+(Nylund-Gibson \& Choi, 2018). The preferred number of classes is at the
+inflection point where the slope of the curve is clearly leveling off.
+Adding this many classes results in substantial decreases of the IC, and
+adding further classes results in relatively small decreases.
+
+Given a specific IC, it is also possible to calculate so-called
+``IC-weights'', which indicate the relative support the data provide for
+each model in the set (\textbf{wagenmakersAICModelSelection2004?}). This is
+particularly useful for confirmatory LCA, as it allows one to quantify
+the support for the theoretical model. IC-weights are calculated by
+calling \texttt{ic\_weights()} on the model fit table. Note, however, that
+IC-weights will (strongly) favor the model with the lowest IC value; the
+scree plot approach above may be more suitable to select the most
+parsimonious model with adequate fit.
 
@@
 
-When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
-The test statistic is computed as the -2 times the log of the ratio of likelihoods,
-and its degrees of freedom are equal to the difference in the degrees of freedom of the two compared models.
-Since a \(k-1\)-class model is nested in a \(k\)-class model,
-it is possible to test the difference in model fit using a LR test.
-One challenge is that the LR test statistic is not asymptotically \(\chi^2\) distributed for LCA.
-The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR {[}REF lo 2001{]}.
-This test is implemented in \texttt{tidySEM}, and is part of the default output of \texttt{table\_fit()}, comparing subsequent rows of the table.
-
-Another solution is to derive an empirical sampling distribution for the LR difference.
-The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the \(k-1\)-class model \(B\) times (e.g., 1000),
-and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset {[}REF Nylund et al., 2007{]}.
-The p-value of this test is defined as the proportion of \(B\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
-Simulation studies suggest that the BLMT has greater statistical power than the LRT,
-although both performed comparably well at detecting the true number of classes {[}REF Nylund 2007, REF Tein 2013{]}.
-A crucial limitation of the BLRT is that it is very computationally expensive.
-With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
-Specific models can be compared using the function \texttt{BLRT()}.
-However, this function often takes prohbitively long to compute or runs out of memory.
-
-Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
-returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
-Other authors have reported similar experiences (\textbf{sinhaPractitionerGuideLatent2021?}).
-This suggests that, like existing ICs,
-LR tests might not adequately balance fit and complexity in LCA.
-Better solutions for class enumeration are thus an important target for future research.
-
-A promising novel solution is the so-called \emph{lazy bootstrap},
-which is much more efficient than the BLRT because it is based on sample statistics of model-implied data generated from the \(k-1\) and \(k\)-class models,
-rather than on fit statistics of LCA models fit to those data (Kollenburg, Mulder, \& Vermunt, 2018).
-This simulation-based approach to inference is similar to the posterior predictive p-value in Bayesian statistics,
-which describes how well each model is able to reproduce the observed correlation matrix (see McElreath, 2020).
-One version of the lazy bootstrap is already implemented in the commercial program latentGOLD {[}REF{]}.
-At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
-Future versions of \texttt{tidySEM} will likely incorporate this test.
+When two models are nested, the significance of the difference in fit
+between these models can be tested using a Likelihood Ratio test (LR).
+The test statistic is computed as the -2 times the log of the ratio of
+likelihoods, and its degrees of freedom are equal to the difference in
+the degrees of freedom of the two compared models. Since a \(k-1\)-class
+model is nested in a \(k\)-class model, it is possible to test the
+difference in model fit using a LR test. One challenge is that the LR
+test statistic is not asymptotically \(\chi^2\) distributed for LCA. The
+so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem
+by applying an ad-hoc correction to the LR {[}REF lo 2001{]}. This test is
+implemented in \texttt{tidySEM}, and is part of the default output of
+\texttt{table\_fit()}, comparing subsequent rows of the table.
+
+Another solution is to derive an empirical sampling distribution for the
+LR difference. The so-called bootstrapped Likelihood Ratio Test (BLRT)
+does so by simulating data from the \(k-1\)-class model \(B\) times (e.g.,
+1000), and fitting the \(k\)- and \(k-1\)-class models to each simulated
+dataset {[}REF Nylund et al., 2007{]}. The p-value of this test is the
+proportion of \(B\) in which the LR of the models fit to simulated data
+exceeds the LR observed in the real data. Simulation studies suggest
+that the BLRT has greater statistical power than the LRT, although both
+performed comparably well at detecting the true number of classes {[}REF
+Nylund 2007, REF Tein 2013{]}. A crucial limitation of the BLRT is that it
+is very computationally expensive. With this in mind, \texttt{table\_fit()} does
+not provide the BLRT by default. Specific models can be compared using
+the function \texttt{BLRT()}.
+
+Anecdotal experience as a statistical co-author on LCA studies suggests
+that both these tests tend to be liberal, returning significant p-values
+for models with more classes even when other indicators suggest that
+those models are overfitted (e.g., near-zero class counts). Other
+authors have reported similar experiences
+(\textbf{sinhaPractitionerGuideLatent2021?}), which matches the differences
+between LRT and ICs from information theory as LRT tests and compare the
+models based on in sample predictive accuracy, and ICs evaluates and
+compare the models out-of-sample predictive accuracy which helps to
+control for overfitting (McElreath, 2020). This suggests
+that, like existing ICs, LR tests might not adequately balance fit and
+complexity in LCA. Better solutions for class enumeration are thus an
+important target for future research.
+
+A promising novel solution is the so-called \emph{lazy bootstrap}, which is
+much more efficient than the BLRT because it is based on sample
+statistics of model-implied data generated from the \(k-1\) and \(k\)-class
+models, rather than on fit statistics of LCA models fit to those data
+(Kollenburg, Mulder, \& Vermunt, 2018). This simulation-based approach to inference
+is similar to the posterior predictive p-value in Bayesian statistics,
+which describes how well each model is able to reproduce the observed
+correlation matrix (see McElreath, 2020). One version of the
+lazy bootstrap is already implemented in the commercial program
+latentGOLD {[}REF{]}. At the time of writing, we are developing a more
+general simulation-based likelihood ratio test (SBLRT) for class
+enumeration. Future versions of \texttt{tidySEM} will likely incorporate this
+test.
 
@@
 
-Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
-These diagnostics are not model fit indices,
-because a model can fit the data well but show poor latent class separation (Masyn, 2013).
-Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.
-Interpretability should always be a consideration when considering
-different class solutions (Nylund-Gibson \& Choi, 2018).
-If the goal is to interpret properties of (members of) classes,
-it makes sense to only consider models that classify well.
-
-All classification diagnostics are derived from the posterior classification probabilities: an \(n \times k\) matrix called \(P\), with the probability that every individual \(n\) belongs to latent class \(k\).
-When classification accuracy is high, and classes are distinct,
-each individual's posterior class probabilities are high for one class, and low for the remaining classes.
-The individual classification probabilities \(P\) are obtained by running \texttt{class\_prob(res,\ type\ =\ "individual")}.
-Note that one column is added to the matrix, with each individual's most likely class membership \(M\).
-\(M\) is obtained by assigning each individual to the class corresponding to their maximum posterior class probability.
-Predicted class membership \(M\) is often of interest to researchers who wish to carry out follow up analyses.
-Note, however, that follow-up analyses should take classification error into account.
-The function \texttt{BCH()} does this automatically.
-
-The matrix \(P\) can be summarized in different way.
-For example, the estimated frequency table of the latent categorical variable can be obtained by summing the posterior probabilities by class (i.e., the column sums of \(P\)).
-This frequency table \(F_P\) is obtained by calling \texttt{class\_prob(res,\ type\ =\ "sum.posterior"}.
-These numbers are fractional because this table takes classification error into account.
-Individuals can contribute partially to multiple classes.
-
-Second, most likely class membership based on the highest posterior class probability, \(F_M\), is obtained by running \texttt{class\_prob(res,\ "sum.mostlikely")}.
-This is the frequency table of the variable \(M\).
-Unlike \(F_P\), \(F_M\) ignores classification error.
-If every observation is classified with perfect accuracy,
-\(F_M\) and \(F_P\) are identical.
-Part of the output of \texttt{table\_fit()} for LCA analyses
-are the minimum and maximum percentage of the sample assigned to a particular class, \texttt{n\_min} and \texttt{n\_max}, derived from \(F_M\).
-This is reported by default because the smallest class proportion, \texttt{n\_min}, is of particular interest.
-When the smallest class (\texttt{n\_min}) is less than 10\% of the total sample size,
-proceed with caution:
-The model might not be locally identified in that class.
-Check whether the number of participants in this class is larger than the number of parameters per class (e.g., five times as large),
-and check whether this smallest class has a meaningful substantive interpretation.
-When the smallest class is less than 1\%, the class is near-empty and might be redundant.
-
-To assess classification accuracy, we turn to the third diagnostic table:
-The average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\), \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}:
-for participants assigned to one particular class (rows) according to \(M\),
-this table gives the mean probability of being assigned to any class (columns).
-This table is obtained by calculating the column means of \(P\) for the
-\(k\) subsets of observations with most likely class of \(M = 1 \ldots k\).
-The diagonal represents the probability that observations in each class will be correctly classified.
-
-The table of classification probabilities for the most likely latent class \(M\) by latent class, \(P_{p_{(M = m | C = c)}}\) is similar to the aforementioned table \(P_{p_{(M = m)}}\),
-except that it accounts for classification uncertainty.
-If \(C\) is the true class of an observation, which is imperfectly measured,
-and \(M\) is the most likely class,
-then this table shows the probability of an observation being assigned to class \(m\) in \(M\), given that its true class is \(c\) in \(C\), or \(p(M=m|C=c)\).
-Values on the diagonal can thus be conceptualized as the reliability with which each level of the categorical latent variable is measured by \(M\),
-and off-diagonal elements can be conceptualized as measurement error.
-The minimum and maximum of the diagonal of this table are given by default as part of the output of \texttt{table\_fit()} for LCA analyses (\texttt{prob\_min} and \texttt{prob\_max}).
-These probabilities are of particular interest because they conveniently summarize classification accuracy per class.
-Both probabilities should be high, as this means that all classes are reliably classified.
-If the minimum probability is low,
-there is at least one class with poor classification accuracy.
-
-Finally, class separability can be summarized in a single statistic:
-the entropy criterion {[}REF{]}.
-In physics, entropy refers to randomness or uncertainty.
-High entropy corresponds to maximum uncertainty,
-and low entropy corresponds to a strict arrangement of the units of study.
-Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
-Low entropy, by contrast, means that every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
-Crucially, in \texttt{tidySEM}, the entropy criterion is defined as 1-entropy (a convention originating in Mplus).
-The interpretation of the entropy criterion is reversed:
-0 means the model classification is no better than chance,
-and 1 means perfect classification.
-As a rule of
-thumb, values above .80 are deemed acceptable and above .90 are considered good.
-Values close to 1 indicate that classes are very well separated,
-and treating the most likely class membership \(M\) as observed in follow-up analyses is probably defensible.
-Like other classification diagnostics,
-entropy is not a model fit measure and should not be used for model selection (Masyn, 2013).
-However, it can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
-
-\hypertarget{labeling-classes}{%
-\subsubsection{Labeling Classes}\label{labeling-classes}}
+Classification diagnostics assess whether the classes identified by LCA
+are distinct and clearly separable. These diagnostics are not model fit
+indices and should not be used for class enumeration, because a model
+can fit the data well and still show poor latent class separation
+(Masyn, 2013). However, classification diagnostics can be used to
+eliminate models: If one analysis goal is to interpret properties of
+(members of) classes, it makes sense to only consider models that
+classify well (Nylund-Gibson \& Choi, 2018). Classification diagnostics
+should thus not be used for model selection, but they can help restrict
+the search space of acceptable models.
+
+All classification diagnostics are derived from the posterior
+classification probabilities: an \(n \times k\) matrix called \(P\), with
+the probability that every individual \(n\) belongs to latent class \(k\).
+When classification accuracy is high, and classes are distinct, each
+individual's posterior class probabilities are high for one class, and
+low for the remaining classes. The individual classification
+probabilities \(P\) are obtained by running
+\texttt{class\_prob(res,\ type\ =\ "individual")}. Note that one column is added to
+the matrix, with each individual's most likely class membership \(M\). \(M\)
+is obtained by assigning each individual to the class corresponding to
+their maximum posterior class probability.
+Predicted class
+membership \(M\) is often of interest to researchers who wish to carry out
+follow up analyses. Note, however, that follow-up analyses should take
+classification error into account. The function \texttt{BCH()} does this
+automatically.
+
+The matrix \(P\) can be summarized in different way. For example, the
+estimated frequency table of the latent categorical variable can be
+obtained by summing the posterior probabilities by class (i.e., the
+column sums of \(P\)). This frequency table \(F_P\) is obtained by calling
+\texttt{class\_prob(res,\ type\ =\ "sum.posterior")}. These numbers are fractional
+because this table takes classification error into account. Individuals
+can contribute partially to multiple classes.
+
+Second, most likely class membership based on the highest posterior
+class probability, \(F_M\), is obtained by running
+\texttt{class\_prob(res,\ "sum.mostlikely")}. This is the frequency table of the
+variable \(M\). Unlike \(F_P\), \(F_M\) ignores classification error. If every
+observation is classified with perfect accuracy, \(F_M\) and \(F_P\) are
+identical. Part of the output of \texttt{table\_fit()} for LCA analyses are the
+minimum and maximum percentage of the sample assigned to a particular
+class, \texttt{n\_min} and \texttt{n\_max}, derived from \(F_M\). This is reported by
+default because the smallest class proportion, \texttt{n\_min}, is of particular
+interest. When the size of the smallest class is not much larger than
+the number of parameters per class, proceed with caution: The model
+might not be locally identified in that class
+(\textbf{depaoliMixtureClassRecovery2013?}). Existing rules of thumb have focused
+on smallest class size in terms of percentage of the total sample size.
+This is not particularly meaningful, however, as the primary concern is
+that small classes may comprise too few observations to accurately
+estimate class parameters. Thus, a better rule of thumb might be to
+ascertain that the smallest class has many more observations than the
+average number of parameters per class. Another good practice is to
+ensure that the smallest class has a unique and meaningful substantive
+interpretation. If the smallest class does not comprise at least a few
+observations per parameter, or is redundant with another larger class, a
+simpler solution is preferable.
+
+To assess classification accuracy, we turn to the third diagnostic
+table: The average posterior probabilities by most likely class
+membership, \(P_{p_{(M = m)}}\),
+\texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}: for participants assigned to
+one particular class (rows) according to \(M\), this table gives the mean
+probability of being assigned to any class (columns). This table is
+obtained by calculating the column means of \(P\) for the \(k\) subsets of
+observations with most likely class of \(M = 1 \ldots k\). The diagonal
+represents the average predictive probability of a the subjects related
+to their most likely class, indicating the certainty of class assignment
+for the subjects. Some guidelines have being suggested for this
+diagonal, like for them to be higher than 0.7 (\textbf{masyn2013?}), but we
+suggest researchers to be cautions about these guidelines as they cannot
+be generalized and should not be use as strict cutoffs to defined ``good''
+or ``bad'' classification.
+
+The table of classification probabilities for the most likely latent
+class \(M\) by latent class, \(P_{p_{(M = m | C = c)}}\) is similar to the
+aforementioned table \(P_{p_{(M = m)}}\), except that it accounts for
+classification uncertainty. If \(C\) is the true class of an observation,
+which is imperfectly measured, and \(M\) is the most likely class, then
+this table shows the probability of an observation being assigned to
+class \(m\) in \(M\), given that its true class is \(c\) in \(C\), or
+\(p(M=m|C=c)\). Values on the diagonal can thus be conceptualized as the
+reliability with which each level of the categorical latent variable is
+measured by \(M\), and off-diagonal elements can be conceptualized as
+measurement error. The minimum and maximum of the diagonal of this table
+are given by default as part of the output of \texttt{table\_fit()} for LCA
+analyses (\texttt{prob\_min} and \texttt{prob\_max}). These probabilities are of
+particular interest because they conveniently summarize classification
+accuracy per class. Both probabilities should be high, as this means
+that all classes are reliably classified. If the minimum probability is
+low, there is at least one class with poor classification accuracy.
+
+Finally, class separability can be summarized in a single statistic: the
+entropy criterion {[}REF{]}.
+In
+physics, high entropy refers to maximum randomness or uncertainty, and
+low entropy corresponds to a strict arrangement of the units of study.
+Applied to LCA, high entropy means that classes are completely
+indistinguishable; every individual has equal probability of being in
+any class. Low entropy, by contrast, means that classes are fully
+separated: every individual has a near-one probability of being in one
+class only, and near-zero probability of being in any other class.
+Crucially, in \texttt{tidySEM}, the entropy criterion is defined as 1-entropy
+(a convention originating in Mplus). Its interpretation is reversed; 0
+means the model classification is no better than chance, and 1 means
+perfect classification. There are no rules of thumb for entropy, because
+entropy is also sensitive to the sample size (entropy tends to be lower
+in larger samples), number of latent classes (entropy tends to be lower
+for higher number of classes), and number of indicators. Like other
+classification diagnostics, entropy is not a model fit measure and
+should not be used for model selection (Masyn, 2013). However, it
+can be used to disqualify certain solutions if the goal is to identify
+clearly distinct classes.
 
@@
 
-Class names should be chosen to accurately reflect theoretically relevant characteristics of group membership.
-These class names should aid the interpretation of the class solutions and serve as a shorthand throughout the manuscript.
-For example, one study used a dual-trajectory LCGA of cognitive and affective empathy development with gender as a covariate.
-Although this model has dozens of parameters,
-the resulting classes were labeled low, average, and high empathy because level differences were the most distinguishing feature.
-However, overly simplistic and general class names can also be misleading, a phenomenon known as the naming
-fallacy (Weller et al., 2020).
-For example, it would be misleading to call one group ``low empathy'' and another group ``medium empathy'' if the difference in empathy were non-significant,
-or if other differences, like the effect of the covariate gender, were more substantial than level differences.
-One way to avoid the naming fallacy is comprehensively reporting model parameters,
-and explicating that class names are just a shorthand.
-
-When labeling classes, it is important to note that LCA models are identified up to the class order.
-This means that the order of classes is arbitrary and may change if the analysis is replicated.
-A simple solution to ensure replicable results is setting a random seed before fitting the model, using \texttt{set.seed()}.
-This will ensure that the classes stay in the same order upon replication,
-but that may not be the preferred order for, e.g., interpreting the results.
-To address this problem, the function \texttt{mx\_switch\_labels()} can be used.
-By default, the function orders classes from largest to smallest class count - but classes can also be ordered by values of a specific parameter (e.g., the mean intercept in a latent class growth model).
-Note that label switching is a much bigger problem when using Bayesian estimation,
-where it can occur during model estimation.
-This is beyond the scope of the present paper.
+Class names should be chosen to accurately reflect theoretically
+relevant characteristics of group membership. These class names should
+aid the interpretation of the class solutions and serve as a shorthand
+throughout the manuscript. For example, one study used a dual-trajectory
+LCGA of cognitive and affective empathy development with gender as a
+covariate. Although this model has dozens of parameters, the resulting
+classes were labeled low, average, and high empathy because level
+differences were the most distinguishing feature. However, overly
+simplistic and general class names can be misleading, a phenomenon known
+as the \emph{naming fallacy} (Weller et al., 2020). For example, it would be
+misleading to call one group ``low empathy'' and another group ``medium
+empathy'' if the empathy differences were non-significant, or if other
+differences, like the effect of the covariate gender, were more
+substantial than level differences. One way to avoid the naming fallacy
+is comprehensively reporting model parameters, and explicating that
+class names are just a shorthand.
+
+When labeling classes, it is important to note that LCA models are
+identified up to the class order.
+This
+means that the order of classes is arbitrary and may change if the
+analysis is replicated. A simple solution to ensure replicable results
+is setting a random seed before fitting the model, using \texttt{set.seed()}.
+This will ensure that the classes stay in the same order upon
+replication, but that may not be the preferred order for, e.g.,
+interpreting the results. To address this problem, the function
+\texttt{mx\_switch\_labels()} can be used. By default, the function orders
+classes from largest to smallest class count - but classes can also be
+ordered by values of a specific parameter (e.g., the mean intercept in a
+latent class growth model). Note that label switching is a much bigger
+problem when using Bayesian estimation, where it can occur during model
+estimation. This is beyond the scope of the present paper.
+
+\hypertarget{follow-up-analyses}{%
+\subsubsection{Follow-up Analyses}\label{follow-up-analyses}}
+
+LCA is rarely performed in isolation; often, researchers wish to relate
+class membership to auxiliary variables, or even auxiliary models. It is
+important to keep in mind that model-implied class membership is subject
+to classification error. Treating class membership as an observed
+variable in follow-up analyses disregards this error. The results will
+be biased, except when class separation is very high - as indicated by
+high (near 1) entropy and minimum posterior classification probability.
+It is therefore important to account for classification error.
+
+There are several different ways to do follow-up analyses in latent
+class analysis (LCA) while accounting for classification error. At
+present, the so-called BCH method is considered one of the best
+approaches {[}after its creators, REF{]}. It has been implemented in
+\texttt{tidySEM}. The advantage of the BCH method is that it not only allows
+for the comparison of individual variables, but also entire auxiliary
+models across latent classes.
 
@@
 Among studies using LCA, reporting practices vary significantly
-(Weller et al., 2020).
-Various authors have tried to improve and
+(Weller et al., 2020). Various authors have tried to improve and
 standardize ways of reporting LCA (e.g. Masyn, 2013; Weller et al., 2020), but more work is needed.
@@
 
-Journals and funding bodies increasingly require research to be open and reproducible.
-For instance, many journals (and the APA) have adopted the TOP guidelines, which recommend comprehensive citation of literature, software, and data sources;
-sharing data and analysis code required to exactly reproduce analyses,
-and pre-registering study plans prior to data collection or analysis (\textbf{nosekTransparencyOpennessPromotion2016a?}).
-A key advantage of using open source software for LCA is that, compared to commercial software, it enables much greater adherence to open science principles.
-Sharing code has limited utility if it can only be evaluated in commercial software,
-as this restricts the ability to reproduce analyses to license holders,
-and prevents auditing research because the underlying source code is proprietary.
-The \texttt{tidySEM} package enables performing LCA in free open source software,
-thus increasing the utility of shared code.
-
-Adopting FOSS makes it possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
-The Workflow for Open Reproducible Code in Science describes how to create a fully reproducible research archive,
-and the \texttt{worcs} R-package automates many of the steps involved (\textbf{vanlissaWORCSWorkflowOpen2021a?}).
-This workflow is based on three principles: First,
-the paper (or analysis section) is written as a \emph{dynamic document} that contains both prose and analysis code.
-Results are automatically rendered as text, tables, and plots.
-Second,
-the research archive is \emph{version controlled}, which means that the entire historical record of the project is retained.
-This version controlled archive can be made public.
-Third, \emph{dependency management} ensures that all software used is documented, including its version number and provenance.
-Without dependency management, code might not work (for example, because it is not clear where certain functions come from),
-or its results may change when some dependencies are updated.
-By combining \texttt{tidySEM} and \texttt{worcs},
-it is possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
-
-An additional argument for sharing data in the context of LCA is that reproducing LCA analyses requires individual participant data,
-not just sufficient statistics like many other methods (i.e., a correlation or covariance matrix).
-If the original data cannot be shared,
-sharing synthetic data at least enables reproducing and reusing analysis code.
-The \texttt{worcs} package provides a generic, non-parametric method to create a synthetic copy of data when data cannot be shared.
-It is also possible to synthesize data directly from the LCA model using \texttt{mxGenerateData()},
+Journals and funding bodies increasingly require research to be open and
+reproducible. For instance, many journals (and the APA) have adopted the
+TOP guidelines, which recommend comprehensive citation of literature,
+software, and data sources; sharing data and analysis code required to
+exactly reproduce analyses, and pre-registering study plans prior to
+data collection or analysis (\textbf{nosekTransparencyOpennessPromotion2016a?}).
+A key advantage of using open source software for LCA is that, compared
+to commercial software, it enables much greater adherence to open
+science principles. Sharing code has limited utility if it can only be
+evaluated in commercial software, as this restricts the ability to
+reproduce analyses to license holders, and prevents auditing research
+because the underlying source code is proprietary. The \texttt{tidySEM} package
+enables performing LCA in free open source software, thus increasing the
+utility of shared code.
+
+Adopting FOSS makes it possible to conduct a fully reproducible LCA
+analysis compliant with best practices in open science. The Workflow for
+Open Reproducible Code in Science describes how to create a fully
+reproducible research archive, and the \texttt{worcs} R-package automates many
+of the steps involved (\textbf{vanlissaWORCSWorkflowOpen2021a?}).
+This workflow is based on three principles: First, the paper (or
+analysis section) is written as a \emph{dynamic document} that contains both
+prose and analysis code. Results are automatically rendered as text,
+tables, and plots. Second, the research archive is \emph{version controlled},
+which means that the entire historical record of the project is
+retained. This version controlled archive can be made public. Third,
+\emph{dependency management} ensures that all software used is documented,
+including its version number and provenance. Without dependency
+management, code might not work (for example, because it is not clear
+where certain functions come from), or its results may change when some
+dependencies are updated. By combining \texttt{tidySEM} and \texttt{worcs}, it is
+possible to conduct a fully reproducible LCA analysis compliant with
+best practices in open science.
+
+An additional argument for sharing data in the context of LCA is that
+reproducing LCA analyses requires individual participant data, not just
+sufficient statistics like many other methods (i.e., a correlation or
+covariance matrix). If the original data cannot be shared, sharing
+synthetic data at least enables reproducing and reusing analysis code.
+The \texttt{worcs} package provides a generic, non-parametric method to create
+a synthetic copy of data when data cannot be shared. It is also possible
+to synthesize data directly from the LCA model using \texttt{mxGenerateData()},
 which often provides a better basis for reproducing the LCA.
@@
 
-LCA analyses inherently involve some subjectivity,
-because of the potential for disagreement between criteria for class enumeration and the lack of objective fit indices.
-In our experience, reviewers (and authors) often disagree about which criteria should be used.
-Preregistering the class enumeration criteria as part of the analysis plan reduces disagreement and eliminates concerns about cherry picking fit measures after the results are known.
-It is particularly important to pre-specify the criteria that will be used to determine the maximum number of classes,
-because there is a risk that information criteria and likelihood ratio tests will suggest a larger number of classes, even if there are concerns about overfitting, classification accuracy, or model interpretability.
-
-The aforementioned Workflow for Open Reproducible Code in Science facilitates creating a preregistration.
-More interestingly, it also enables creating a \emph{Preregistration As Code} (PAC):
-A draft paper with reproducible analysis section based on fake (synthetic or simulated) data (\textbf{peikertReproducibleResearchTutorial2021?}).
-When the real data are accessed or collected,
-the analysis is simply reproduced with this new data file,
-and the results in the paper are automatically updated.
-PACs are quite unambiguous; the selection of the final model can even be automated by hard-coding the criteria into the analysis section.
-For an example of PAC, see (\textbf{vanlissaComplementingPreregisteredConfirmatory2022?}).
+LCA analyses inherently involve some subjectivity, because of the
+potential for disagreement between criteria for class enumeration and
+the lack of objective fit indices. In our experience, reviewers (and
+authors) often disagree about which criteria should be used.
+Preregistering the class enumeration criteria as part of the analysis
+plan reduces disagreement and eliminates concerns about cherry picking
+fit measures after the results are known. It is particularly important
+to pre-specify the criteria that will be used to determine the maximum
+number of classes, because there is a risk that information criteria and
+likelihood ratio tests will suggest a larger number of classes, even if
+there are concerns about overfitting, classification accuracy, or model
+interpretability.
+
+The aforementioned Workflow for Open Reproducible Code in Science
+facilitates creating a preregistration. More interestingly, it also
+enables creating a \emph{Preregistration As Code} (PAC): A draft paper with
+reproducible analysis section based on fake (synthetic or simulated)
+data (\textbf{peikertReproducibleResearchTutorial2021?}). When the real data are
+accessed or collected, the analysis is simply reproduced with this new
+data file, and the results in the paper are automatically updated. PACs
+are quite unambiguous; the selection of the final model can even be
+automated by hard-coding the criteria into the analysis section. For an
+example of PAC, see (\textbf{vanlissaComplementingPreregisteredConfirmatory2022?}).
 
@@
 
-The function \texttt{table\_fit()} reports an overview of fit indices, customized for LCA.
-In reporting model fit, it is important, firstly, to report the effective sample size, which should be the same for all models in order to compare relative fit measures.
-Secondly, it should be clearly reported,
-for example, in the strategy of analyses,
-which parameters are estimated across and within classes.
-In case of doubt, a full overview of the parameters is given by the \texttt{coef()} and \texttt{table\_results()} functions.
-The number of parameters for each model should be included in the fit table.
-Thirdly, it is important that readers can assess the relative fit of different solutions.
-The most fundamental fit measure is the \(-2LL\),
-as this is the quantity optimized during estimation.
-From these statistics, information criteria can be computed.
-By default, \texttt{table\_fit()} includes the AIC, BIC, and saBIC.
-We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.
-
-Classification diagnostics should be reported,
-as they may determine whether models are excluded from consideration.
-Of particular interest is the diagonal of either the average posterior probabilities by most likely class membership,
-or of the most likely class membership by true class membership.
-The former indicates the average probability that individuals belong to the class they were assigned to,
-and the latter indicates the probability that individuals are assigned to a class if they actually belong to that class.
-By default, \texttt{table\_fit()} reports the latter as \texttt{prob\_min} and \texttt{prob\_max}.
-The entropy is also of interest as an overall measure of class separation.
-Finally, the range of proportions of cases assigned to each class is important;
-the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant.
-By default, \texttt{table\_fit()} reports the minimum and maximum proportion of cases,
-based on most likely class membership.
-Fifth, the LMR or BLRT can be reported if it is part of the class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}.
-By default, \texttt{table\_fit()} reports the VLM for subsequent models in the table.
+The function \texttt{table\_fit()} reports an overview of fit indices,
+customized for LCA. In reporting model fit, it is important, firstly, to
+report the effective sample size, which should be the same for all
+models in order to compare relative fit measures. Secondly, it should be
+clearly reported, for example, in the strategy of analyses, which
+parameters are estimated across and within classes. In case of doubt, a
+full overview of the parameters is given by the \texttt{coef()} and
+\texttt{table\_results()} functions. The number of parameters for each model
+should be included in the fit table. Thirdly, it is important that
+readers can assess the relative fit of different solutions. The most
+fundamental fit measure is the \(-2LL\), as this is the quantity optimized
+during estimation. From these statistics, information criteria can be
+computed. By default, \texttt{table\_fit()} includes the AIC, BIC, and saBIC. We
+suggest to report at least the BIC, as it is ubiquitous, grounded in
+theory, and performs well in class enumeration.
+
+Classification diagnostics should be reported, as they may determine
+whether models are excluded from consideration. Of particular interest
+is the diagonal of either the average posterior probabilities by most
+likely class membership, or of the most likely class membership by true
+class membership. The former indicates the average probability that
+individuals belong to the class they were assigned to, and the latter
+indicates the probability that individuals are assigned to a class if
+they actually belong to that class. By default, \texttt{table\_fit()} reports
+the latter as \texttt{prob\_min} and \texttt{prob\_max}. The entropy is also of interest
+as an overall measure of class separation. Finally, the range of
+proportions of cases assigned to each class is important; the minimum
+proportion of cases per class can be used to diagnose classes that are
+too small to be locally identified or practically relevant (see \textbf{depaoliMixtureClassRecovery2013?}). By default, \texttt{table\_fit()} reports the
+minimum and maximum proportion of cases, based on most likely class
+membership. Fifth, the LMR or BLRT can be reported if it is part of the
+class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}. By default,
+\texttt{table\_fit()} reports the VLM for subsequent models in the table.
 
@@
 
-Reporting LCA results involves reporting the class proportions,
-obtained via \texttt{class\_prob(res,\ "sum.posterior")},
-and model parameters, obtained via \texttt{table\_results(res)}.
-The class proportions are obtained by standardizing the class weights:
-A \(k\)-class model has \(k-1\) free class weights,
-and one weight constrained to 1 for model identification.
-The free weights are included in the model results,
-but are redundant with the aforementioned class proportions.
-Other model parameters may include means, (co)variances, regression coefficients (rarely), and thresholds for ordinal indicators.
-It is good practice to report the point estimate and confidence interval of these parameters.
-Although it is conventional to report p-values as well,
-note that the default p-values tests the hypothesis that the parameter is equal to zero.
-This hypothesis may be irrelevant in LCA when the focus is on parameter differences between classes.
-Reporting standard errors allows readers to compute p-values for other more informative hypotheses.
-Thresholds are more readily interpretable when converted to conditional item probabilities, using the function \texttt{table\_prob()}.
-These indicate the probability of an item being
-endorsed given that the observation belongs to a particular latent (Geiser, 2012).
+Reporting LCA results involves reporting the class proportions, obtained
+via \texttt{class\_prob(res,\ "sum.posterior")}, and model parameters, obtained
+via \texttt{table\_results(res)}. The class proportions are obtained by
+standardizing the class weights: A \(k\)-class model has \(k-1\) free class
+weights, and one weight constrained to 1 for model identification. The
+free weights are included in the model results, but are redundant with
+the aforementioned class proportions. Other model parameters may include
+means, (co)variances, regression coefficients (rarely), and thresholds
+for ordinal indicators. It is good practice to report the point estimate
+and confidence interval of these parameters. Although it is conventional
+to report p-values as well, note that the default p-values tests the
+hypothesis that the parameter is equal to zero. This hypothesis may be
+irrelevant in LCA when the focus is on parameter differences between
+classes. Reporting standard errors allows readers to compute p-values
+for other more informative hypotheses. Thresholds are more readily
+interpretable when converted to conditional item probabilities, using
+the function \texttt{table\_prob()}. These indicate the probability of an item
+being endorsed given that the observation belongs to a particular latent
+(Geiser, 2012).
 
@@
 
-As mentioned before, the results table included null-hypothesis tests for all parameters by default.
-Whether these tests are informative depends on the study goals and scale of the data.
-For example, if LCA indicators are mean-centered,
-then the p-values of class-specific means indicate whether the mean in that class differs significantly from the sample average.
-It is also possible to test more informative hypotheses,
-using the function \texttt{wald\_test()}.
-Equality-constrained informative hypotheses are specified using the \texttt{bain} syntax,
-and evaluated on the model object.
-This enables testing theoretical hypotheses - for example, if specific values for class-specific means had been hypothesized.
-The same function allows testing parameters for equality, either within or across classes.
+As mentioned before, the results table included null-hypothesis tests
+for all parameters by default. Whether these tests are informative
+depends on the study goals and scale of the data. For example, if LCA
+indicators are mean-centered, then the p-values of class-specific means
+indicate whether the mean in that class differs significantly from the
+sample average. It is also possible to test more informative hypotheses,
+using the function \texttt{wald\_test()}. Equality-constrained informative
+hypotheses are specified using the \texttt{bain} syntax, and evaluated on the
+model object. This enables testing theoretical hypotheses - for example,
+if specific values for class-specific means had been hypothesized. The
+same function allows testing parameters for equality, either within or
+across classes.
 
@@
 
-In an informal review of published graphics for LCA,
-we found that most visualizations focused on the estimated model parameters,
-displaying class means as points connected by lines, with point shape and linetype indicating class membership.
-Note that the lines introduce spurious information.
-Furthermore, they omit uncertainty about class means and do not illustrate how well the model describes the data.
-
-The Grammar of Graphics addresses best practices in visualization, which we summarize here with specific emphasis on the application to LCA.
-Most importantly, good graphics start with the \emph{data} in mind.
-Visualizing raw data allows one to assess class separability,
-and which aspects of the data are captured well or poorly by the model.
-A second consideration are which \emph{aesthetics} to use.
-In a color plot, class membership can be indicated by color.
-In a black and white plot, it can be indicated by linetype and point shape.
-A third consideration is which geometric objects (geoms) to use.
-As mentioned before, common practice is to plot parameter estimates as points, and connect these points with lines.
-Points are suitable geoms to visualize point estimates and raw data.
-For continuous indicators, additional relevant geoms include error bars for standard errors and standard deviations,
-and density plots to visualize the distribution of raw data (smoothed with a kernel density function).
-In two dimensions, (co)variances can be visualized as ellipses.
-Finally, when plotting multiple class solutions,
-it makes sense to \emph{facet} a plot into multiple panels by class solution.
-The plots in \texttt{tidySEM} are based on best practices as outlined in the grammar of graphics,
-and are implemented in \texttt{ggplot2}.
-As such, these plots can be further modified just like any other \texttt{ggplot2} object.
-One remaining limitation is that these best practices have only been implemented for LCA with continuous indicators, in \texttt{plot\_profiles()}, \texttt{plot\_density()} and \texttt{plot\_bivariate()}, and for LCGA in \texttt{plot\_growth()}.
-Plots for LCA with ordinal indicators are still rudimentary, more work remains to be done in this area.
+In an informal review of published graphics for LCA, we found that most
+visualizations focused on the estimated model parameters, displaying
+class means as points connected by lines, with point shape and linetype
+indicating class membership. Note that the lines introduce spurious
+information. Furthermore, they omit uncertainty about class means and do
+not illustrate how well the model describes the data.
+
+The Grammar of Graphics addresses best practices in visualization, which
+we summarize here with specific emphasis on the application to LCA. Most
+importantly, good graphics start with the \emph{data} in mind. Visualizing
+raw data allows one to assess class separability, and which aspects of
+the data are captured well or poorly by the model. A second
+consideration are which \emph{aesthetics} to use. In a color plot, class
+membership can be indicated by color. In a black and white plot, it can
+be indicated by linetype and point shape. A third consideration is which
+geometric objects (geoms) to use. As mentioned before, common practice
+is to plot parameter estimates as points, and connect these points with
+lines. Points are suitable geoms to visualize point estimates and raw
+data. For continuous indicators, additional relevant geoms include error
+bars for standard errors and standard deviations, and density plots to
+visualize the distribution of raw data (smoothed with a kernel density
+function). In two dimensions, (co)variances can be visualized as
+ellipses. Finally, when plotting multiple class solutions, it makes
+sense to \emph{facet} a plot into multiple panels by class solution. The
+plots in \texttt{tidySEM} are based on best practices as outlined in the
+grammar of graphics, and are implemented in \texttt{ggplot2}. As such, these
+plots can be further modified just like any other \texttt{ggplot2} object. One
+remaining limitation is that these best practices have only been
+implemented for LCA with continuous indicators, in \texttt{plot\_profiles()},
+\texttt{plot\_density()} and \texttt{plot\_bivariate()}, and for LCGA in
+\texttt{plot\_growth()}. Plots for LCA with ordinal indicators are still
+rudimentary, more work remains to be done in this area.
 
@@
 
-Based on our experience in statistical consultations,
-there are a few common pitfalls and misunderstandings about LCA.
-First, perhaps the most common problem we see in consultations is treating ordinal data as continuous.
-As explained previously, LCA for ordinal indicators should be used for Likert-type scales and other data with few unique values.
-A second common pitfall is omitting the 1-class solution from class enumeration.
-If a one-class solution has better or similar fit to models with a higher number of classes, LCA may not be appropriate.
-A one-class solution should thus always be included as a benchmark.
-Third, although it is widely understood that classification diagnostics (e.g., entropy, posterior probability, class counts) are not class enumeration criteria,
-they still have a role in class enumeration.
-Specifically, they can be used to exclude solutions that do not meet predefined criteria from consideration.
-Fourth, previous reporting standards called for reporting of sample descriptive statistics (see \textbf{schreiberLatentClassAnalysis2017?}).
-While we agree that this is due dilligence,
-sample descriptives have two limitations in the context of LCA:
-they are not sufficient to check or reproduce the analysis,
-and many use cases of LCA assume that the population is heterogeneous, which limits the utility of whole-sample descriptives.
-These two limitations can be addressed by data sharing and reporting descriptive statistics for each latent class separately.
-Fifth, a recent publication mistakenly stated that an assumption of mixture
-models is that observed indicators are normally distributed
-(Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020).
-This is incorrect, as evident from the shoe size example:
-When the number of classes is greater than one,
-Gaussian mixture models assume that the distribution of observed indicators is a \emph{mixture of multiple normal distributions}.
-If the indicators were normally distributed,
-there would be no classes to extract as the whole population
-would belong to a single (homogeneous) class.
-Finally, the same paper suggested that robust standard errors
-should be used when indicators are not normally distributed
-(Spurk et al., 2020).
-This is also incorrect;
-to our knowledge, there is no literature on the merits of robust standard errors for LCA.
+Based on our experience in statistical consultations, there are a few
+common pitfalls and misunderstandings about LCA. First, perhaps the most
+common problem we see in consultations is treating ordinal data as
+continuous. As explained previously, LCA for ordinal indicators should
+be used for Likert-type scales and other data with few unique values. A
+second common pitfall is omitting the 1-class solution from class
+enumeration. If a one-class solution has better or similar fit to models
+with a higher number of classes, LCA may not be appropriate. A one-class
+solution should thus always be included as a benchmark. Third, although
+it is widely understood that classification diagnostics (e.g., entropy,
+posterior probability, class counts) are not class enumeration criteria,
+they still have a role in class enumeration. Specifically, they can be
+used to exclude solutions that do not meet predefined criteria from
+consideration. Fourth, previous reporting standards called for reporting
+of sample descriptive statistics (see \textbf{schreiberLatentClassAnalysis2017?}). While we agree that this is due
+dilligence, sample descriptives have two limitations in the context of
+LCA: they are not sufficient to check or reproduce the analysis, and
+many use cases of LCA assume that the population is heterogeneous, which
+limits the utility of whole-sample descriptives. These two limitations
+can be addressed by data sharing and reporting descriptive statistics
+for each latent class separately. Fifth, a recent publication mistakenly
+stated that an assumption of mixture models is that observed indicators
+are normally distributed (Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020). This is incorrect, as
+evident from the shoe size example: When the number of classes is
+greater than one, Gaussian mixture models assume that the distribution
+of observed indicators is a \emph{mixture of multiple normal distributions}.
+If the indicators were normally
+distributed, there would be no classes to extract as the whole
+population would belong to a single (homogeneous) class. Finally, the
+same paper suggested that robust standard errors should be used when
+indicators are not normally distributed (Spurk et al., 2020). This is
+also incorrect; to our knowledge, there is no literature on the merits
+of robust standard errors for LCA. Finally, we caution against step-wise
+class enumeration, whereby additional models are estimated until the
+selected class enumeration criteria no longer show improvement of model
+fit (e.g., see Spurk et al., 2020). This practice is problematic,
+because it is possible that an even higher number of classes would lead
+to a (significant) improvement in fit. Consider this thought experiment:
+there exist two groups of three relatively similar classes each (six in
+total). Step-wise class enumeration shows improvement in model fit going
+from a 1-class to 2-class model: the two classes capture the difference
+between the two groups of three classes. Then, no noteworthy improvement
+in model fit is found for adding more classes, until the 6-class
+solution is reached. In the step-wise approach, the analysis would be
+terminated after the 2-class model - even though the correct number of
+classes is 6.
 
@@
 
-The best practices discussed in this paper are summarized in the SMART-LCA Checklist: Standards for More Accuracy in Reporting of different Types of Latent Class Analysis.
+The best practices discussed in this paper are summarized in the
+SMART-LCA Checklist: Standards for More Accuracy in Reporting of
+different Types of Latent Class Analysis.
 
@@
   \item
-    Any other reasons why the observed (multivariate) distribution could not be described by the LCA (e.g., extreme skew)?
+    Any other reasons why the observed (multivariate) distribution
+    could not be described by the LCA (e.g., extreme skew)?
   \end{itemize}
@@
   \item
-    In exploratory LCA, compare all sensible alternative specifications
-  \item
-    In confirmatory LCA, justify alternative models on theoretical grounds
-  \item
-    Check that the observed (multivariate) distribution could theoretically be captured by the specified model (see \#1).
+    In exploratory LCA, compare all sensible alternative
+    specifications
+  \item
+    In confirmatory LCA, justify alternative models on theoretical
+    grounds
+  \item
+    Check that the observed (multivariate) distribution could
+    theoretically be captured by the specified model (see \#1).
   \end{itemize}
@@
   \item
-    In confirmatory LCA, estimate the theoretical number of classes, a 1-class model, and optionally other numbers of classes for comparison
+    In confirmatory LCA, estimate the theoretical number of classes,
+    a 1-class model, and optionally other numbers of classes for
+    comparison
   \end{itemize}
 \item
-  Justify the (preregistered) criteria used for class enumeration, which can include:
+  Justify the (preregistered) criteria used for class enumeration,
+  which can include:
 
@@
     LR tests
+  \item
+    Theoretical interpretability of the solution
   \end{itemize}
 \item
-  Justify the (preregistered) criteria used to eliminate models from consideration, including:
+  Justify the (preregistered) criteria used to eliminate models from
+  consideration, including:
 
@@
   \item
-    Local identifiability (acceptable number of observations per parameter in each class)
+    Local identifiability (acceptable number of observations per
+    parameter in each class)
   \item
@@
   \item
-    All parameters of the chosen model (means, (co)variances, thresholds)
+    All parameters of the chosen model (means, (co)variances,
+    thresholds)
   \item
@@
   \item
-    Show multivariate distributions for multivariate models, if possible
+    Show multivariate distributions for multivariate models, if
+    possible
   \end{itemize}
@@
   \item
-    If entropy \textgreater{} .95 and posterior classification probability for each class \textgreater{} .95, this is less crucial
+    If entropy \textgreater{} .95 and posterior classification probability for
+    each class \textgreater{} .95, this is less crucial
   \end{itemize}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_longitudinal_2010}{}}%
+Vermunt, Jeroen K.
+(2010).
+Longitudinal {Research} {Using} {Mixture} {Models}.
+In K.
+Montfort, J.
+H.
+Oud, \& A.
+Satorra (Eds.), \emph{Longitudinal {Research} with {Latent} {Variables}} (pp.
+119--152).
+Berlin/Heidelberg: Springer.
+\url{https://doi.org/10.1007/978-3-642-11760-2_4}
+
 \leavevmode\vadjust pre{\hypertarget{ref-vermunt_latent_2004}{}}%
-Vermunt, J.
-K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.
+Vermunt, Jeroen K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.
 F., \& Department of Methodology and Statistics.
@@
 Retrieved from \url{https://research.tilburguniversity.edu/en/publications/0caedd00-27c1-42bd-bb4e-d1dcb0864956}
+
+\leavevmode\vadjust pre{\hypertarget{ref-depmixS4}{}}%
+Visser, I., \& Speekenbrink, M.
+(2010).
+{depmixS4}: An {R} package for hidden markov models.
+\emph{Journal of Statistical Software}, \emph{36}(7), 1--21.
+Retrieved from \url{https://www.jstatsoft.org/v36/i07/}
 

### 2023-01-24 appendix_a.Rmd
revert MGV previous commit

@@
 shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
+
 bibliography      : "r-references.bib"
+
 floatsintext      : yes
@@
 mask              : no
+
 figurelist        : no
@@
 footnotelist      : no
+
 classoption       : "man"
 output            : papaja::apa6_pdf
-editor_options: 
-  markdown: 
-    wrap: sentence
 ---
@@
 This is an example of confirmatory LPA using `tidySEM`.
-The simulated data are based on work by Zegwaard and colleagues, who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
+The simulated data are based on work by Zegwaard and colleagues,
+who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
 Qualitative research among experts resulted in a theory postulating the existence of four types of caregivers (translated from the original Dutch):
@@
 
-To view the data documentation, run the command `?tidySEM::zegwaard_carecompass` in the R console.
+To view the data documentation,
+run the command `?tidySEM::zegwaard_carecompass` in the R console.
 
@@
 We first only use the four scales: `c("burdened", "trapped", "negaffect", "loneliness")`.
-\`.
+`.
 
@@
 
-As per the best practices, the first step in LCA is examining the observed data.
+As per the best practices,
+the first step in LCA is examining the observed data.
 We use `tidySEM::descriptives()` to describe the data numerically.
-Because all scales are continuous, we select only columns for continuous data to de-clutter the table:
+Because all scales are continuous,
+we select only columns for continuous data to de-clutter the table:
 
@@
 ```
-
 ```{r tabdesc, echo = FALSE, eval=TRUE, results='asis'}
@@
 
-The table indicates two potential causes for concern: there is a small percentage of missingness, and all variables have relatively high kurtosis.
-Since there are some missing values, we can conduct an MCAR test using `mice::mcar(df)`.
-According to Hawkins' test, there is no evidence to reject the assumptions of multivariate normality and MCAR, $\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71$.
+The table indicates two potential causes for concern:
+there is a small percentage of missingness,
+and all variables have relatively high kurtosis.
+Since there are some missing values,
+we can conduct an MCAR test using `mice::mcar(df)`.
+According to Hawkins' test,
+there is no evidence to reject the assumptions of multivariate normality and MCAR, $\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71$.
 Missing data will be accounted for using FIML.
@@
 The `ggplot2` function `geom_density()` is useful for continuous data.
-Visual inspection confirms the conclusions from the `descriptives()` table: the data are kurtotic (peaked).
+Visual inspection confirms the conclusions from the `descriptives()` table:
+the data are kurtotic (peaked).
 
@@
 ```
-
 ```{r, echo = FALSE, eval = run_everything}
@@
 ```
-
 ```{r figdesc, echo = FALSE, eval = pkgdown}
@@
 
-As all variables are continuous, we can use the convenience function `tidySEM::mx_profiles()`, which is a wrapper for the generic function `mx_mixture()` optimized for continuous indicators.
+As all variables are continuous, we can use the convenience function
+`tidySEM::mx_profiles()`,
+which is a wrapper for the generic function `mx_mixture()` optimized for continuous indicators.
 Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
 Its arguments are `data` and number of `classes`.
-All variables in `data` are included in the analysis, which is why we first selected the indicator variables.
-As this is a confirmatory LCA, we do not follow a strictly data-driven class enumeration procedure.
+All variables in `data` are included in the analysis,
+which is why we first selected the indicator variables.
+As this is a confirmatory LCA,
+we do not follow a strictly data-driven class enumeration procedure.
 We will set the maximum number of classes $K$ to one more than the theoretically expected number.
@@
 ```
-
 ```{r eval = FALSE, echo = TRUE}
@@
 ```
-
 ```{r eval = TRUE, echo = FALSE}
@@
 This analysis should produce some warnings about cluster initialization.
-These relate to the selection of starting values, which relies on the K-means algorithm and is not robust to missing data.
+These relate to the selection of starting values,
+which relies on the K-means algorithm and is not robust to missing data.
 The algorithm automatically switches to hierarchical clustering, no further action is required.
@@
 
-To compare the fit of the theoretical model against other models, we create a model fit table using `table_fit()` and retain relevant columns.
+To compare the fit of the theoretical model against other models,
+we create a model fit table using
+`table_fit()` and retain relevant columns.
 We also determine whether any models can be disqualified.
@@
 Next, we check for local identifiability.
-The sample size is consistently reported as 513, which means that partially missing cases were indeed included via FIML.
-The smallest class size occurs in the 5-class model, where the smallest class is assigned 7% of cases, or 38 cases.
+The sample size is consistently reported as 513,
+which means that partially missing cases were indeed included via FIML.
+The smallest class size occurs in the 5-class model,
+where the smallest class is assigned 7% of cases, or 38 cases.
 This model has 28 parameters, approximately 6 per class.
-We thus have at least five observations per parameter in every class, and do not disqualify the 5-class model.
+We thus have at least five observations per parameter in every class,
+and do not disqualify the 5-class model.
 
-There are concerns about theoretical interpretability of all solutions, as the entropies and minimum classification probabilities are all low.
+There are concerns about theoretical interpretability of all solutions,
+as the entropies and minimum classification probabilities are all low.
 However, in this confirmatory use case, we address this when interpreting the results.
@@
 ```
-
 ```{r tabfit, echo = FALSE, eval = TRUE}
@@
 
-Class enumeration unambiguously indicates a preference for the 4-class solution: the 4-class solution has the lowest BIC, which means it is preferred over all other solutions including a 1-class solution and a solution with more classes.
+Class enumeration unambiguously indicates a preference for the 4-class solution:
+the 4-class solution has the lowest BIC,
+which means it is preferred over all other solutions including a 1-class solution and a solution with more classes.
 The p-values of the LMR LR test are significant for all pairwise model comparisons, except for the 5-class model.
@@
 Following the elbow criterion, a three-class solution would also be defensible.
-The function `ic_weights(tab_compare)` allows us to compute IC weights; it indicates that, conditional on the set of models, the 4-class model has a posterior model probability of nearly 100%.
+The function `ic_weights(tab_compare)` allows us to compute IC weights;
+it indicates that, conditional on the set of models,
+the 4-class model has a posterior model probability of nearly 100%.
 
@@
 ```
-
 ```{r, echo = TRUE, eval = FALSE}
@@
 ```
-
 ```{r tabfitcomp, echo = FALSE, eval = TRUE}
@@
 
-To interpret the final class solution, we first reorder the 4-class model by class size.
+To interpret the final class solution,
+we first reorder the 4-class model by class size.
 This helps prevent label switching.
@@
 
-The 4-class model yielded classes of reasonable size; the largest class comprised 33%, and the smallest comprised 16% of cases.
+The 4-class model yielded classes of reasonable size;
+the largest class comprised 33%,
+and the smallest comprised 16% of cases.
 However, the entropy was low, $S = .75$, indicating poor class separability.
@@
 ```
-
 ```{r echo = FALSE, eval = run_everything}
@@
 ```
-
 ```{r, eval = TRUE, echo=FALSE}
@@
 ```
-
 ```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
@@
 ```
-
-On the diagonal of the bivariate plot are weighted density plots: normal approximations of the density function of observed data, weighed by class probability.
-On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.
-As this model has zero covariances, all circles are round (albeit warped by the different scales of the X and Y axes)
+On the diagonal of the bivariate plot are weighted density plots:
+normal approximations of the density function of observed data,
+weighed by class probability.
+On the off-diagonal are plots for each pair of indicators,
+with the class means indicated by a point,
+class standard deviations indicated by lines,
+and covariances indicated by circles.
+As this model has zero covariances,
+all circles are round (albeit warped by the different scales of the X and Y axes)
 
@@
 We can obtain a more classic profile plot using `plot_profiles(res_final)`.
-This plot conveys less information than the bivariate plot, but is readily interpretable.
-Below is a comparison between the most common type of visualization for LPA, and the best-practices visualization provided by `tidySEM`.
-Note that the best practices plot includes class means and error bars, standard deviations, and a ribbon plot of raw data weighted by class probability to indicate how well the classes describe the observed distribution.
-The overlap between the classes is clearly visible in this figure; this is why the entropy and classification probabilities are relatively low.
+This plot conveys less information than the bivariate plot,
+but is readily interpretable.
+Below is a comparison between the most common type of visualization
+for LPA, and the best-practices visualization provided by `tidySEM`.
+Note that the best practices plot includes class means and error bars,
+standard deviations,
+and a ribbon plot of raw data weighted by class probability to indicate how well the classes describe the observed distribution.
+The overlap between the classes is clearly visible in this figure;
+this is why the entropy and classification probabilities are relatively low.
 
-Based on the bivariate plot, we can label class 1 as the *balanced* type (33%), class 2 as the *imbalanced* type (29%), class 3 as the *entrapped* type (22%), and class 4 as the *lonely* type (16%).
+Based on the bivariate plot, we can label class 1 as the *balanced* type (33%),
+class 2 as the *imbalanced* type (29%), class 3 as the *entrapped* type (22%),
+and class 4 as the *lonely* type (16%).
 
@@
 Finally, we may want to compare the different classes on auxiliary variables or models.
-The `BCH()` function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
+The `BCH()` function applies three-step analysis,
+which compares the classes using a multi-group model,
+controlling for classification error.
 We consider two examples: a single variable, and an auxiliary model.
@@
 
-For a single (continuous or ordinal) variable, we can call the BCH function and simply supply the auxiliary variable to the `data` argument, omitting the `model` argument.
+For a single (continuous or ordinal) variable,
+we can call the BCH function and simply supply the auxiliary variable to the `data` argument, omitting the `model` argument.
 Below, we estimate an auxiliary model to compare the sex of patients between classes:
@@
 ```
-
 ```{r echo = FALSE, eval=run_everything}
@@
 ```
-
 ```{r echo = FALSE, eval=FALSE}
@@
 
-To obtain an omnibus likelihood ratio test of the significance of these sex differences across classes, as well as pairwise comparisons between classes, use `lr_test(aux_sex)`.
+To obtain an omnibus likelihood ratio test of the significance of these sex differences across classes,
+as well as pairwise comparisons between classes,
+use `lr_test(aux_sex)`.
 The results indicate that there are significant sex differences across classes, $\Delta LL(1) = 8.7, p = .003$.
@@
 
-To obtain an omnibus likelihood ratio test of the difference in regression coefficients across classes and pairwise comparisons between classes, use `lr_test(aux_model, compare = "A")`.
+To obtain an omnibus likelihood ratio test of the difference in regression coefficients across classes
+and pairwise comparisons between classes,
+use `lr_test(aux_model, compare = "A")`.
 The results indicate that there are no significant sex differences across classes, $\Delta LL(3) = 0.98, p = .81$.
 The results can be reported using `table_results(aux_model)`.
+
+<!-- \newpage -->
 
@@
 <!-- ::: {#refs custom-style="Bibliography"} -->
-
 <!-- ::: -->
-
-## R session
-
-```{r echo = TRUE, eval=TRUE}
-sessionInfo()
-```

### 2023-01-24 appendix_a.tex
revert MGV previous commit

@@
 This is an example of confirmatory LPA using \texttt{tidySEM}.
-The simulated data are based on work by Zegwaard and colleagues, who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
+The simulated data are based on work by Zegwaard and colleagues,
+who sought to establish a typology of caregivers who support a close other receiving outpatient psychological care.
 Qualitative research among experts resulted in a theory postulating the existence of four types of caregivers (translated from the original Dutch):
@@
 
-To view the data documentation, run the command \texttt{?tidySEM::zegwaard\_carecompass} in the R console.
+To view the data documentation,
+run the command \texttt{?tidySEM::zegwaard\_carecompass} in the R console.
 
@@
 \CommentTok{\# Load required packages}
-\FunctionTok{library}\NormalTok{(tidySEM) }
+\FunctionTok{library}\NormalTok{(tidySEM)}
 \FunctionTok{library}\NormalTok{(ggplot2)}
 \CommentTok{\# Load data}
-\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ zegwaard\_carecompass[, }\FunctionTok{c}\NormalTok{(}\StringTok{"burdened"}\NormalTok{, }\StringTok{"trapped"}\NormalTok{, }\StringTok{"negaffect"}\NormalTok{, }\StringTok{"loneliness"}\NormalTok{)]}
+\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ zegwaard\_carecompass[, }\FunctionTok{c}\NormalTok{(}\StringTok{"burdened"}\NormalTok{, }\StringTok{"trapped"}\NormalTok{, }\StringTok{"negaffect"}\NormalTok{,}
+    \StringTok{"loneliness"}\NormalTok{)]}
 \end{Highlighting}
@@
 
-As per the best practices, the first step in LCA is examining the observed data.
+As per the best practices,
+the first step in LCA is examining the observed data.
 We use \texttt{tidySEM::descriptives()} to describe the data numerically.
-Because all scales are continuous, we select only columns for continuous data to de-clutter the table:
+Because all scales are continuous,
+we select only columns for continuous data to de-clutter the table:
 
@@
 \NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ tidySEM}\SpecialCharTok{::}\FunctionTok{descriptives}\NormalTok{(df)}
-\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{,}
-                 \StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{,}
-                 \StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
+\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{, }\StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{,}
+    \StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{, }\StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
 \NormalTok{desc}
@@
 
-The table indicates two potential causes for concern: there is a small percentage of missingness, and all variables have relatively high kurtosis.
-Since there are some missing values, we can conduct an MCAR test using \texttt{mice::mcar(df)}.
-According to Hawkins' test, there is no evidence to reject the assumptions of multivariate normality and MCAR, \(\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71\).
+The table indicates two potential causes for concern:
+there is a small percentage of missingness,
+and all variables have relatively high kurtosis.
+Since there are some missing values,
+we can conduct an MCAR test using \texttt{mice::mcar(df)}.
+According to Hawkins' test,
+there is no evidence to reject the assumptions of multivariate normality and MCAR, \(\tilde{\chi^2}(6) = 3.78, \tilde{p} = 0.71\).
 Missing data will be accounted for using FIML.
@@
 The \texttt{ggplot2} function \texttt{geom\_density()} is useful for continuous data.
-Visual inspection confirms the conclusions from the \texttt{descriptives()} table: the data are kurtotic (peaked).
+Visual inspection confirms the conclusions from the \texttt{descriptives()} table:
+the data are kurtotic (peaked).
 
@@
     \AttributeTok{timevar =} \StringTok{"Variable"}\NormalTok{)}
-\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{Variable)}\SpecialCharTok{+}
+\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value)) }\SpecialCharTok{+} \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+} \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{Variable) }\SpecialCharTok{+}
     \FunctionTok{theme\_bw}\NormalTok{()}
@@
 
-As all variables are continuous, we can use the convenience function \texttt{tidySEM::mx\_profiles()}, which is a wrapper for the generic function \texttt{mx\_mixture()} optimized for continuous indicators.
+As all variables are continuous, we can use the convenience function
+\texttt{tidySEM::mx\_profiles()},
+which is a wrapper for the generic function \texttt{mx\_mixture()} optimized for continuous indicators.
 Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
 Its arguments are \texttt{data} and number of \texttt{classes}.
-All variables in \texttt{data} are included in the analysis, which is why we first selected the indicator variables.
-As this is a confirmatory LCA, we do not follow a strictly data-driven class enumeration procedure.
+All variables in \texttt{data} are included in the analysis,
+which is why we first selected the indicator variables.
+As this is a confirmatory LCA,
+we do not follow a strictly data-driven class enumeration procedure.
 We will set the maximum number of classes \(K\) to one more than the theoretically expected number.
@@
 \FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df,}
-                   \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{)}
+\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df, }\AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{)}
 \end{Highlighting}
@@
 
-To compare the fit of the theoretical model against other models, we create a model fit table using \texttt{table\_fit()} and retain relevant columns.
+To compare the fit of the theoretical model against other models,
+we create a model fit table using
+\texttt{table\_fit()} and retain relevant columns.
 We also determine whether any models can be disqualified.
@@
 Next, we check for local identifiability.
-The sample size is consistently reported as 513, which means that partially missing cases were indeed included via FIML.
-The smallest class size occurs in the 5-class model, where the smallest class is assigned 7\% of cases, or 38 cases.
+The sample size is consistently reported as 513,
+which means that partially missing cases were indeed included via FIML.
+The smallest class size occurs in the 5-class model,
+where the smallest class is assigned 7\% of cases, or 38 cases.
 This model has 28 parameters, approximately 6 per class.
-We thus have at least five observations per parameter in every class, and do not disqualify the 5-class model.
+We thus have at least five observations per parameter in every class,
+and do not disqualify the 5-class model.
 
-There are concerns about theoretical interpretability of all solutions, as the entropies and minimum classification probabilities are all low.
+There are concerns about theoretical interpretability of all solutions,
+as the entropies and minimum classification probabilities are all low.
 However, in this confirmatory use case, we address this when interpreting the results.
@@
 Following the elbow criterion, a three-class solution would also be defensible.
-The function \texttt{ic\_weights(tab\_compare)} allows us to compute IC weights; it indicates that, conditional on the set of models, the 4-class model has a posterior model probability of nearly 100\%.
+The function \texttt{ic\_weights(tab\_compare)} allows us to compute IC weights;
+it indicates that, conditional on the set of models,
+the 4-class model has a posterior model probability of nearly 100\%.
 
@@
 
-The 4-class model yielded classes of reasonable size; the largest class comprised 33\%, and the smallest comprised 16\% of cases.
+The 4-class model yielded classes of reasonable size;
+the largest class comprised 33\%,
+and the smallest comprised 16\% of cases.
 However, the entropy was low, \(S = .75\), indicating poor class separability.
@@
 \begin{Highlighting}[]
-\FunctionTok{table\_results}\NormalTok{(res\_final, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"confint"}\NormalTok{, }\StringTok{"class"}\NormalTok{))}
+\FunctionTok{table\_results}\NormalTok{(res\_final, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"confint"}\NormalTok{,}
+    \StringTok{"class"}\NormalTok{))}
 \end{Highlighting}
@@
 
-On the diagonal of the bivariate plot are weighted density plots: normal approximations of the density function of observed data, weighed by class probability.
-On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.
-As this model has zero covariances, all circles are round (albeit warped by the different scales of the X and Y axes)
+On the diagonal of the bivariate plot are weighted density plots:
+normal approximations of the density function of observed data,
+weighed by class probability.
+On the off-diagonal are plots for each pair of indicators,
+with the class means indicated by a point,
+class standard deviations indicated by lines,
+and covariances indicated by circles.
+As this model has zero covariances,
+all circles are round (albeit warped by the different scales of the X and Y axes)
 
@@
 We can obtain a more classic profile plot using \texttt{plot\_profiles(res\_final)}.
-This plot conveys less information than the bivariate plot, but is readily interpretable.
-Below is a comparison between the most common type of visualization for LPA, and the best-practices visualization provided by \texttt{tidySEM}.
-Note that the best practices plot includes class means and error bars, standard deviations, and a ribbon plot of raw data weighted by class probability to indicate how well the classes describe the observed distribution.
-The overlap between the classes is clearly visible in this figure; this is why the entropy and classification probabilities are relatively low.
+This plot conveys less information than the bivariate plot,
+but is readily interpretable.
+Below is a comparison between the most common type of visualization
+for LPA, and the best-practices visualization provided by \texttt{tidySEM}.
+Note that the best practices plot includes class means and error bars,
+standard deviations,
+and a ribbon plot of raw data weighted by class probability to indicate how well the classes describe the observed distribution.
+The overlap between the classes is clearly visible in this figure;
+this is why the entropy and classification probabilities are relatively low.
 
-Based on the bivariate plot, we can label class 1 as the \emph{balanced} type (33\%), class 2 as the \emph{imbalanced} type (29\%), class 3 as the \emph{entrapped} type (22\%), and class 4 as the \emph{lonely} type (16\%).
+Based on the bivariate plot, we can label class 1 as the \emph{balanced} type (33\%),
+class 2 as the \emph{imbalanced} type (29\%), class 3 as the \emph{entrapped} type (22\%),
+and class 4 as the \emph{lonely} type (16\%).
 
@@
 Finally, we may want to compare the different classes on auxiliary variables or models.
-The \texttt{BCH()} function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
+The \texttt{BCH()} function applies three-step analysis,
+which compares the classes using a multi-group model,
+controlling for classification error.
 We consider two examples: a single variable, and an auxiliary model.
@@
 
-For a single (continuous or ordinal) variable, we can call the BCH function and simply supply the auxiliary variable to the \texttt{data} argument, omitting the \texttt{model} argument.
+For a single (continuous or ordinal) variable,
+we can call the BCH function and simply supply the auxiliary variable to the \texttt{data} argument, omitting the \texttt{model} argument.
 Below, we estimate an auxiliary model to compare the sex of patients between classes:
@@
 
-To obtain an omnibus likelihood ratio test of the significance of these sex differences across classes, as well as pairwise comparisons between classes, use \texttt{lr\_test(aux\_sex)}.
+To obtain an omnibus likelihood ratio test of the significance of these sex differences across classes,
+as well as pairwise comparisons between classes,
+use \texttt{lr\_test(aux\_sex)}.
 The results indicate that there are significant sex differences across classes, \(\Delta LL(1) = 8.7, p = .003\).
@@
 \NormalTok{df\_aux}\SpecialCharTok{$}\NormalTok{freqvisit }\OtherTok{\textless{}{-}} \FunctionTok{as.numeric}\NormalTok{(df\_aux}\SpecialCharTok{$}\NormalTok{freqvisit)}
-\NormalTok{aux\_model }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_final, }\AttributeTok{model =} \StringTok{"freqvisit \textasciitilde{} distance"}\NormalTok{,}
-                 \AttributeTok{data =}\NormalTok{ df\_aux)}
+\NormalTok{aux\_model }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_final, }\AttributeTok{model =} \StringTok{"freqvisit \textasciitilde{} distance"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ df\_aux)}
 \end{Highlighting}
@@
 
-To obtain an omnibus likelihood ratio test of the difference in regression coefficients across classes and pairwise comparisons between classes, use \texttt{lr\_test(aux\_model,\ compare\ =\ "A")}.
+To obtain an omnibus likelihood ratio test of the difference in regression coefficients across classes
+and pairwise comparisons between classes,
+use \texttt{lr\_test(aux\_model,\ compare\ =\ "A")}.
 The results indicate that there are no significant sex differences across classes, \(\Delta LL(3) = 0.98, p = .81\).
@@
 
-\hypertarget{r-session}{%
-\subsection{R session}\label{r-session}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{sessionInfo}\NormalTok{()}
-\CommentTok{\#\textgreater{} R version 4.1.3 (2022{-}03{-}10)}
-\CommentTok{\#\textgreater{} Platform: x86\_64{-}w64{-}mingw32/x64 (64{-}bit)}
-\CommentTok{\#\textgreater{} Running under: Windows 10 x64 (build 19045)}
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} Matrix products: default}
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} locale:}
-\CommentTok{\#\textgreater{} [1] LC\_COLLATE=English\_United Kingdom.1252 }
-\CommentTok{\#\textgreater{} [2] LC\_CTYPE=English\_United Kingdom.1252   }
-\CommentTok{\#\textgreater{} [3] LC\_MONETARY=English\_United Kingdom.1252}
-\CommentTok{\#\textgreater{} [4] LC\_NUMERIC=C                           }
-\CommentTok{\#\textgreater{} [5] LC\_TIME=English\_United Kingdom.1252    }
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} attached base packages:}
-\CommentTok{\#\textgreater{} [1] stats     graphics  grDevices utils     datasets  methods   base     }
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} other attached packages:}
-\CommentTok{\#\textgreater{} [1] ggplot2\_3.4.0    tidySEM\_0.2.4.6  OpenMx\_2.21.1    scales\_1.2.1    }
-\CommentTok{\#\textgreater{} [5] yaml\_2.3.6       papaja\_0.1.1     tinylabels\_0.2.3}
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} loaded via a namespace (and not attached):}
-\CommentTok{\#\textgreater{}   [1] TH.data\_1.1{-}1         colorspace\_2.1{-}0      estimability\_1.4.1   }
-\CommentTok{\#\textgreater{}   [4] parameters\_0.20.1     rstudioapi\_0.14       listenv\_0.9.0        }
-\CommentTok{\#\textgreater{}   [7] lavaan\_0.6{-}13         rstan\_2.26.13         fansi\_1.0.4          }
-\CommentTok{\#\textgreater{}  [10] mvtnorm\_1.1{-}3         codetools\_0.2{-}18      splines\_4.1.3        }
-\CommentTok{\#\textgreater{}  [13] mnormt\_2.1.1          knitr\_1.41            texreg\_1.38.6        }
-\CommentTok{\#\textgreater{}  [16] bayesplot\_1.10.0      jsonlite\_1.8.4        effectsize\_0.8.2     }
-\CommentTok{\#\textgreater{}  [19] compiler\_4.1.3        httr\_1.4.4            emmeans\_1.8.4{-}1      }
-\CommentTok{\#\textgreater{}  [22] backports\_1.4.1       assertthat\_0.2.1      Matrix\_1.5{-}3         }
-\CommentTok{\#\textgreater{}  [25] fastmap\_1.1.0         cli\_3.6.0             htmltools\_0.5.4      }
-\CommentTok{\#\textgreater{}  [28] prettyunits\_1.1.1     tools\_4.1.3           igraph\_1.3.5         }
-\CommentTok{\#\textgreater{}  [31] coda\_0.19{-}4           gtable\_0.3.1          glue\_1.6.2           }
-\CommentTok{\#\textgreater{}  [34] RANN\_2.6.1            dplyr\_1.0.10          V8\_4.2.2             }
-\CommentTok{\#\textgreater{}  [37] Rcpp\_1.0.10           carData\_3.0{-}5         vctrs\_0.5.2          }
-\CommentTok{\#\textgreater{}  [40] nlme\_3.1{-}161          psych\_2.2.9           insight\_0.18.8       }
-\CommentTok{\#\textgreater{}  [43] xfun\_0.36             stringr\_1.5.0         globals\_0.16.2       }
-\CommentTok{\#\textgreater{}  [46] ps\_1.7.2              proto\_1.0.0           CompQuadForm\_1.4.3   }
-\CommentTok{\#\textgreater{}  [49] lifecycle\_1.0.3       future\_1.30.0         MASS\_7.3{-}58.2        }
-\CommentTok{\#\textgreater{}  [52] zoo\_1.8{-}11            parallel\_4.1.3        sandwich\_3.0{-}2       }
-\CommentTok{\#\textgreater{}  [55] inline\_0.3.19         curl\_5.0.0            gridExtra\_2.3        }
-\CommentTok{\#\textgreater{}  [58] pander\_0.6.5          blavaan\_0.4{-}3         loo\_2.5.1            }
-\CommentTok{\#\textgreater{}  [61] StanHeaders\_2.26.13   stringi\_1.7.12        highr\_0.10           }
-\CommentTok{\#\textgreater{}  [64] bayestestR\_0.13.0     fastDummies\_1.6.3     checkmate\_2.1.0      }
-\CommentTok{\#\textgreater{}  [67] boot\_1.3{-}28           pkgbuild\_1.4.0        rlang\_1.0.6          }
-\CommentTok{\#\textgreater{}  [70] pkgconfig\_2.0.3       matrixStats\_0.63.0    evaluate\_0.20        }
-\CommentTok{\#\textgreater{}  [73] lattice\_0.20{-}45       rstantools\_2.2.0      processx\_3.8.0       }
-\CommentTok{\#\textgreater{}  [76] tidyselect\_1.2.0      parallelly\_1.34.0     plyr\_1.8.8           }
-\CommentTok{\#\textgreater{}  [79] magrittr\_2.0.3        bookdown\_0.32         R6\_2.5.1             }
-\CommentTok{\#\textgreater{}  [82] generics\_0.1.3        multcomp\_1.4{-}20       DBI\_1.1.3            }
-\CommentTok{\#\textgreater{}  [85] withr\_2.5.0           gsubfn\_0.7            pillar\_1.8.1         }
-\CommentTok{\#\textgreater{}  [88] survival\_3.5{-}0        datawizard\_0.6.5      abind\_1.4{-}5          }
-\CommentTok{\#\textgreater{}  [91] tibble\_3.1.8          future.apply\_1.10.0   crayon\_1.5.2         }
-\CommentTok{\#\textgreater{}  [94] car\_3.1{-}1             nonnest2\_0.5{-}5        utf8\_1.2.2           }
-\CommentTok{\#\textgreater{}  [97] tmvnsim\_1.0{-}2         MplusAutomation\_1.1.0 rmarkdown\_2.20       }
-\CommentTok{\#\textgreater{} [100] grid\_4.1.3            data.table\_1.14.6     pbivnorm\_0.6.0       }
-\CommentTok{\#\textgreater{} [103] bain\_0.2.8            callr\_3.7.3           digest\_0.6.31        }
-\CommentTok{\#\textgreater{} [106] xtable\_1.8{-}4          dbscan\_1.1{-}11         RcppParallel\_5.1.6   }
-\CommentTok{\#\textgreater{} [109] stats4\_4.1.3          munsell\_0.5.0}
-\end{Highlighting}
-\end{Shaded}
-
 \end{document}

### 2023-01-24 appendix_b.Rmd
revert MGV previous commit

@@
 
-This is an example of exploratory latent class analysis (LCA) with continuous indicators, otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using `tidySEM`.
+This is an example of exploratory latent class analysis (LCA) with continuous indicators,
+otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using `tidySEM`.
 The present example uses data collected by Alkema as part of a study on ocean microplastics.
@@
 Since there is no theoretical reason to expect a certain number of classes, this is an exploratory LCA.
-To view its documentation, run the command `?tidySEM::alkema_microplastics` in the R console.
-The original analyses are available at <https://github.com/cjvanlissa/lise_microplastics>; in this vignette, we take a different approach to the analysis to showcase other possibilities.
+To view its documentation,
+run the command `?tidySEM::alkema_microplastics` in the R console.
+The original analyses are available at https://github.com/cjvanlissa/lise_microplastics;
+in this vignette, we take a different approach to the analysis to showcase other possibilities.
 
@@
 
-As per the best practices, the first step in LCA is examining the observed data.
+As per the best practices,
+the first step in LCA is examining the observed data.
 We use `tidySEM::descriptives()` to describe the data numerically.
-Because all items are categorical, we remove columns for continuous data to de-clutter the table:
+Because all items are categorical,
+we remove columns for continuous data to de-clutter the table:
 
@@
 The data are correctly coded as `numeric`.
-There are no missing values; if any variables had missing values, we would report an MCAR test with `mice::mcar()`, and explain that missing data are accounted for using FIML.
+There are no missing values; if any variables had missing values, we would report an MCAR test with `mice::mcar()`,
+and explain that missing data are accounted for using FIML.
 Both the table above and the density plot indicate that the data are extremely right-skewed and kurtotic.
@@
 ```
-
 ```{r figdesclog, echo = FALSE, eval = pkgdown}
@@
 The log transformation addresses the aforementioned concerns regarding skew and kurtosis.
-To confirm this, reshape the data to wide format and examine a scatterplot:
+To confirm this, reshape the data to wide format
+and examine a scatterplot:
 
@@
 ```
-
 ```{r, eval = run_everything, echo = FALSE}
@@
 ```
-
 ```{r figscatter, eval = pkgdown, echo = FALSE}
@@
 <!-- As evident from the scatterplot, there is a very strong linear tend in the data. -->
-
 <!-- This indicates that fragments tend to be coextensive (when length goes up, width goes up). -->
-
 <!-- We can analyze these data in their original dimensions (length and width). -->
-
 <!-- Alternatively, we can use PCA to rotate the data such that the first dimension can be interpreted as "size", -->
-
 <!-- and the second dimension can be interpreted as "shape": -->
@@
 
-As all variables are continuous, we can use the convenience function `tidySEM::mx_profiles()`, which is a wrapper for the generic function `mx_mixture()` optimized for continuous indicators.
+As all variables are continuous, we can use the convenience function
+`tidySEM::mx_profiles()`,
+which is a wrapper for the generic function `mx_mixture()` optimized for continuous indicators.
 Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
 Its arguments are `data` and number of `classes`.
-All variables in `data` are included in the analysis, which is why we first selected the indicator variables.
+All variables in `data` are included in the analysis,
+which is why we first selected the indicator variables.
 
-As this is an exploratory LCA, we will conduct a rather extensive search across model specifications and number of classes.
+As this is an exploratory LCA,
+we will conduct a rather extensive search across model specifications and number of classes.
 We will set the maximum number of classes $K$ to four; depending on the results, we can always choose to increase it later.
@@
 
-As the analysis takes a long time to compute, it is prudent to save the results to disk immediately, so as not to lose them.
+As the analysis takes a long time to compute,
+it is prudent to save the results to disk immediately, so as not to lose them.
 For this, we use the function `saveRDS()`.
@@
 ```
-
 ```{r eval = run_everything, echo = FALSE}
@@
 <!-- This suggests a potential problem with with model convergence. -->
-
 <!-- To try and get this model to converge, it is possible to use the function `mxTryHard()`, which permutes the starting values and tries to find a better solution. -->
-
 <!-- To do this, run `res[[24]] <- mxTryHard(res[[24]])`. -->
-
 <!-- However, at present, we will skip this step, -->
-
 <!-- because there are also other indications that this 4-class model is overfitted (see below). -->
@@
 
-To compare the fit of the estimated models, we create a model fit table using `table_fit()`.
+To compare the fit of the estimated models,
+we create a model fit table using
+`table_fit()`.
 
@@
 First, we determine whether any models can be disqualified.
-There were no indications of convergence problems during estimation, so this is not a reason to disqualify solutions.
+There were no indications of convergence problems during estimation,
+so this is not a reason to disqualify solutions.
 Next, we check for local identifiability.
@@
 
-As can be seen from the fit table, the lowest ratio of observations to parameters is 18, which is no cause for concern.
+As can be seen from the fit table,
+the lowest ratio of observations to parameters is 18, which is no cause for concern.
 
@@
 
-An interesting characteristic of this data is that the BIC and the entropy are strongly correlated.
+An interesting characteristic of this data is that
+the BIC and the entropy are strongly correlated.
 The raw correlation between these two metrics is .66, `cor(fit$BIC, fit$Entropy)`.
-If we omit the 1-class models, for which entropy is technically not defined, the correlation is even as high as .85, `cor(fit$BIC[!fit$Classes == 1], fit$Entropy[!fit$Classes == 1])`.
+If we omit the 1-class models, for which entropy is technically not defined,
+the correlation is even as high as .85, `cor(fit$BIC[!fit$Classes == 1], fit$Entropy[!fit$Classes == 1])`.
 
@@
 This illustrates why entropy should not be treated as a model fit criterion.
-It also illustrates that criteria for class enumeration should be explicit, because we will likely come to a different decision depending on which criteria are used.
+It also illustrates that criteria for class enumeration should be explicit,
+because we will likely come to a different decision depending on which criteria are used.
 
-As mentioned before, we drop models with \< 10% of cases in the smallest class:
+As mentioned before, we drop models with < 10% of cases in the smallest class:
 
@@
 
-If our strategy is to optimize fit, we can examine the fit table above, or plot a scree plot for the BIC by calling `plot(fit)`.
+If our strategy is to optimize fit,
+we can examine the fit table above,
+or plot a scree plot for the BIC by calling `plot(fit)`.
 Note that, due to the large sample size, all ICs give identical conclusions.
@@
 
-Looking at the blocks of 1-4 class models for each model specification, it appears that the BIC keeps decreasing with the addition of more classes.
+Looking at the blocks of 1-4 class models for each model specification,
+it appears that the BIC keeps decreasing with the addition of more classes.
 Across the blocks, the BIC keeps decreasing with increasingly complex model specifications.
@@
 
-Out of the 16 models that remain after removing those with \< 10% of cases in the smallest class, one model stands out: The 2-class model with free (co)variances.
-<!-- The function `ic_weights(fit)` allows us to compute IC weights for all models in the set; --> <!-- it prefers the most complex model out of the 16 models with a posterior model probability of nearly 100%.
---> We thus select this as our final model.
+Out of the 16 models that remain after removing those with < 10% of cases in the smallest class,
+one model stands out:
+The 2-class model with free (co)variances.
+<!-- The function `ic_weights(fit)` allows us to compute IC weights for all models in the set; -->
+<!-- it prefers the most complex model out of the 16 models with a posterior model probability of nearly 100%. -->
+We thus select this as our final model.
 
@@
 
-On the diagonal of the bivariate plot are weighted density plots: normal approximations of the density function of observed data, weighed by class probability.
-On the off-diagonal are plots for each pair of indicators, with the class means indicated by a point, class standard deviations indicated by lines, and covariances indicated by circles.
+On the diagonal of the bivariate plot are weighted density plots:
+normal approximations of the density function of observed data,
+weighed by class probability.
+On the off-diagonal are plots for each pair of indicators,
+with the class means indicated by a point,
+class standard deviations indicated by lines,
+and covariances indicated by circles.
 
@@
 The first class (57%) accounts for smaller fragments, and the second class (43%) accounts for some of the right-skew in fragments' length and width.
-We can simply label class 1 as *small fragments*, and class 2 as *larger fragments*.
+We can simply label class 1 as *small fragments*,
+and class 2 as *larger fragments*.
 
@@
 
-There are, however, concerns about the interpretability of this solutions: the entropy is `.56` and the minimum classification probability is`.81`.
+There are, however,
+concerns about the interpretability of this solutions:
+the entropy is `.56` and the minimum classification probability is`.81`.
 This is because of substantial overlap in the distributions of the two classes.
@@
 Finally, we may want to compare the different classes on auxiliary variables or models.
-The `BCH()` function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
+The `BCH()` function applies three-step analysis,
+which compares the classes using a multi-group model,
+controlling for classification error.
 For example, we can test whether polymer type differs between the two classes:
@@
 
-To obtain an omnibus likelihood ratio test of the significance of the differences in polymer type across classes, use `lr_test(aux_pt)`.
+To obtain an omnibus likelihood ratio test of the significance of the differences in polymer type across classes,
+use `lr_test(aux_pt)`.
 The results indicate that there are significant differences in polymer types across classes, $\Delta LL(3) = 17.14, p < .001$.

### 2023-01-24 appendix_b.tex
revert MGV previous commit

@@
 
-This is an example of exploratory latent class analysis (LCA) with continuous indicators, otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using \texttt{tidySEM}.
+This is an example of exploratory latent class analysis (LCA) with continuous indicators,
+otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using \texttt{tidySEM}.
 The present example uses data collected by Alkema as part of a study on ocean microplastics.
@@
 Since there is no theoretical reason to expect a certain number of classes, this is an exploratory LCA.
-To view its documentation, run the command \texttt{?tidySEM::alkema\_microplastics} in the R console.
-The original analyses are available at \url{https://github.com/cjvanlissa/lise_microplastics}; in this vignette, we take a different approach to the analysis to showcase other possibilities.
+To view its documentation,
+run the command \texttt{?tidySEM::alkema\_microplastics} in the R console.
+The original analyses are available at \url{https://github.com/cjvanlissa/lise_microplastics};
+in this vignette, we take a different approach to the analysis to showcase other possibilities.
 
@@
 \CommentTok{\# Load required packages}
-\FunctionTok{library}\NormalTok{(tidySEM) }
+\FunctionTok{library}\NormalTok{(tidySEM)}
 \FunctionTok{library}\NormalTok{(ggplot2)}
 \CommentTok{\# Load data}
-\NormalTok{df\_analyze }\OtherTok{\textless{}{-}}\NormalTok{ alkema\_microplastics[alkema\_microplastics}\SpecialCharTok{$}\NormalTok{category }\SpecialCharTok{==} \StringTok{"Fragment"}\NormalTok{, ]}
-\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ df\_analyze[ ,}\FunctionTok{c}\NormalTok{(}\StringTok{"length"}\NormalTok{, }\StringTok{"width"}\NormalTok{)]}
+\NormalTok{df\_analyze }\OtherTok{\textless{}{-}}\NormalTok{ alkema\_microplastics[alkema\_microplastics}\SpecialCharTok{$}\NormalTok{category }\SpecialCharTok{==}
+    \StringTok{"Fragment"}\NormalTok{, ]}
+\NormalTok{df }\OtherTok{\textless{}{-}}\NormalTok{ df\_analyze[, }\FunctionTok{c}\NormalTok{(}\StringTok{"length"}\NormalTok{, }\StringTok{"width"}\NormalTok{)]}
 \end{Highlighting}
@@
 
-As per the best practices, the first step in LCA is examining the observed data.
+As per the best practices,
+the first step in LCA is examining the observed data.
 We use \texttt{tidySEM::descriptives()} to describe the data numerically.
-Because all items are categorical, we remove columns for continuous data to de-clutter the table:
+Because all items are categorical,
+we remove columns for continuous data to de-clutter the table:
 
@@
 \NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ tidySEM}\SpecialCharTok{::}\FunctionTok{descriptives}\NormalTok{(df)}
-\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"type"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{, }
-\StringTok{"mean"}\NormalTok{, }\StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{, }\StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
+\NormalTok{desc }\OtherTok{\textless{}{-}}\NormalTok{ desc[, }\FunctionTok{c}\NormalTok{(}\StringTok{"name"}\NormalTok{, }\StringTok{"type"}\NormalTok{, }\StringTok{"n"}\NormalTok{, }\StringTok{"missing"}\NormalTok{, }\StringTok{"unique"}\NormalTok{, }\StringTok{"mean"}\NormalTok{,}
+    \StringTok{"median"}\NormalTok{, }\StringTok{"sd"}\NormalTok{, }\StringTok{"min"}\NormalTok{, }\StringTok{"max"}\NormalTok{, }\StringTok{"skew\_2se"}\NormalTok{, }\StringTok{"kurt\_2se"}\NormalTok{)]}
 \NormalTok{papaja}\SpecialCharTok{::}\FunctionTok{apa\_table}\NormalTok{(desc, }\AttributeTok{caption =} \StringTok{"Descriptive statistics"}\NormalTok{)}
@@
 The data are correctly coded as \texttt{numeric}.
-There are no missing values; if any variables had missing values, we would report an MCAR test with \texttt{mice::mcar()}, and explain that missing data are accounted for using FIML.
+There are no missing values; if any variables had missing values, we would report an MCAR test with \texttt{mice::mcar()},
+and explain that missing data are accounted for using FIML.
 Both the table above and the density plot indicate that the data are extremely right-skewed and kurtotic.
@@
 \NormalTok{df\_plot}\SpecialCharTok{$}\NormalTok{Value }\OtherTok{\textless{}{-}} \FunctionTok{log}\NormalTok{(df\_plot}\SpecialCharTok{$}\NormalTok{Value)}
-\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{Variable)}\SpecialCharTok{+}
+\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value)) }\SpecialCharTok{+} \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+} \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{Variable) }\SpecialCharTok{+}
     \FunctionTok{theme\_bw}\NormalTok{()}
@@
 \begin{Highlighting}[]
-\NormalTok{df }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(df\_plot, }\AttributeTok{direction =} \StringTok{"wide"}\NormalTok{, }\AttributeTok{v.names =} \StringTok{"Value"}\NormalTok{)[, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
+\NormalTok{df }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(df\_plot, }\AttributeTok{direction =} \StringTok{"wide"}\NormalTok{, }\AttributeTok{v.names =} \StringTok{"Value"}\NormalTok{)[,}
+    \SpecialCharTok{{-}}\DecValTok{1}\NormalTok{]}
 \FunctionTok{names}\NormalTok{(df) }\OtherTok{\textless{}{-}} \FunctionTok{gsub}\NormalTok{(}\StringTok{"Value."}\NormalTok{, }\StringTok{""}\NormalTok{, }\FunctionTok{names}\NormalTok{(df), }\AttributeTok{fixed =} \ConstantTok{TRUE}\NormalTok{)}
-\FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ length, }\AttributeTok{y =}\NormalTok{ width)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{alpha =}\NormalTok{ .}\DecValTok{1}\NormalTok{) }\SpecialCharTok{+}
+\FunctionTok{ggplot}\NormalTok{(df, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ length, }\AttributeTok{y =}\NormalTok{ width)) }\SpecialCharTok{+} \FunctionTok{geom\_point}\NormalTok{(}\AttributeTok{alpha =} \FloatTok{0.1}\NormalTok{) }\SpecialCharTok{+}
     \FunctionTok{theme\_bw}\NormalTok{()}
@@
 
-As all variables are continuous, we can use the convenience function \texttt{tidySEM::mx\_profiles()}, which is a wrapper for the generic function \texttt{mx\_mixture()} optimized for continuous indicators.
+As all variables are continuous, we can use the convenience function
+\texttt{tidySEM::mx\_profiles()},
+which is a wrapper for the generic function \texttt{mx\_mixture()} optimized for continuous indicators.
 Its default settings are appropriate for LPA, assuming fixed variances across classes and zero covariances.
 Its arguments are \texttt{data} and number of \texttt{classes}.
-All variables in \texttt{data} are included in the analysis, which is why we first selected the indicator variables.
+All variables in \texttt{data} are included in the analysis,
+which is why we first selected the indicator variables.
 
-As this is an exploratory LCA, we will conduct a rather extensive search across model specifications and number of classes.
+As this is an exploratory LCA,
+we will conduct a rather extensive search across model specifications and number of classes.
 We will set the maximum number of classes \(K\) to four; depending on the results, we can always choose to increase it later.
@@
 
-As the analysis takes a long time to compute, it is prudent to save the results to disk immediately, so as not to lose them.
+As the analysis takes a long time to compute,
+it is prudent to save the results to disk immediately, so as not to lose them.
 For this, we use the function \texttt{saveRDS()}.
@@
 \FunctionTok{set.seed}\NormalTok{(}\DecValTok{123}\NormalTok{)}
-\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df,}
-                   \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{,}
-                   \AttributeTok{variances =} \FunctionTok{c}\NormalTok{(}\StringTok{"equal"}\NormalTok{, }\StringTok{"varying"}\NormalTok{),}
-                   \AttributeTok{covariances =} \FunctionTok{c}\NormalTok{(}\StringTok{"zero"}\NormalTok{, }\StringTok{"equal"}\NormalTok{,}
-                                   \StringTok{"varying"}\NormalTok{),}
+\NormalTok{res }\OtherTok{\textless{}{-}} \FunctionTok{mx\_profiles}\NormalTok{(}\AttributeTok{data =}\NormalTok{ df, }\AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{4}\NormalTok{, }\AttributeTok{variances =} \FunctionTok{c}\NormalTok{(}\StringTok{"equal"}\NormalTok{,}
+    \StringTok{"varying"}\NormalTok{), }\AttributeTok{covariances =} \FunctionTok{c}\NormalTok{(}\StringTok{"zero"}\NormalTok{, }\StringTok{"equal"}\NormalTok{, }\StringTok{"varying"}\NormalTok{),}
     \AttributeTok{expand\_grid =} \ConstantTok{TRUE}\NormalTok{)}
@@
 
-To compare the fit of the estimated models, we create a model fit table using \texttt{table\_fit()}.
+To compare the fit of the estimated models,
+we create a model fit table using
+\texttt{table\_fit()}.
 
@@
 \begin{Highlighting}[]
-\NormalTok{fit}\SpecialCharTok{$}\NormalTok{par\_ratio }\OtherTok{\textless{}{-}}\NormalTok{ (}\DecValTok{5605}\SpecialCharTok{*}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{n\_min) }\SpecialCharTok{/}\NormalTok{ (fit}\SpecialCharTok{$}\NormalTok{Parameters}\SpecialCharTok{/}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{Classes)}
-\NormalTok{fit[ , }\FunctionTok{c}\NormalTok{(}\StringTok{"Name"}\NormalTok{, }\StringTok{"LL"}\NormalTok{, }\StringTok{"Parameters"}\NormalTok{, }\StringTok{"par\_ratio"}\NormalTok{,}
-         \StringTok{"BIC"}\NormalTok{, }\StringTok{"Entropy"}\NormalTok{,}
-         \StringTok{"prob\_min"}\NormalTok{, }\StringTok{"prob\_max"}\NormalTok{, }
-         \StringTok{"n\_min"}\NormalTok{, }\StringTok{"n\_max"}\NormalTok{,}
-         \StringTok{"lmr\_p"}\NormalTok{)]}
+\NormalTok{fit}\SpecialCharTok{$}\NormalTok{par\_ratio }\OtherTok{\textless{}{-}}\NormalTok{ (}\DecValTok{5605} \SpecialCharTok{*}\NormalTok{ fit}\SpecialCharTok{$}\NormalTok{n\_min)}\SpecialCharTok{/}\NormalTok{(fit}\SpecialCharTok{$}\NormalTok{Parameters}\SpecialCharTok{/}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{Classes)}
+\NormalTok{fit[, }\FunctionTok{c}\NormalTok{(}\StringTok{"Name"}\NormalTok{, }\StringTok{"LL"}\NormalTok{, }\StringTok{"Parameters"}\NormalTok{, }\StringTok{"par\_ratio"}\NormalTok{, }\StringTok{"BIC"}\NormalTok{, }\StringTok{"Entropy"}\NormalTok{,}
+    \StringTok{"prob\_min"}\NormalTok{, }\StringTok{"prob\_max"}\NormalTok{, }\StringTok{"n\_min"}\NormalTok{, }\StringTok{"n\_max"}\NormalTok{, }\StringTok{"lmr\_p"}\NormalTok{)]}
 \end{Highlighting}
@@
 
-An interesting characteristic of this data is that the BIC and the entropy are strongly correlated.
+An interesting characteristic of this data is that
+the BIC and the entropy are strongly correlated.
 The raw correlation between these two metrics is .66, \texttt{cor(fit\$BIC,\ fit\$Entropy)}.
-If we omit the 1-class models, for which entropy is technically not defined, the correlation is even as high as .85, \texttt{cor(fit\$BIC{[}!fit\$Classes\ ==\ 1{]},\ fit\$Entropy{[}!fit\$Classes\ ==\ 1{]})}.
+If we omit the 1-class models, for which entropy is technically not defined,
+the correlation is even as high as .85, \texttt{cor(fit\$BIC{[}!fit\$Classes\ ==\ 1{]},\ fit\$Entropy{[}!fit\$Classes\ ==\ 1{]})}.
 
@@
 This illustrates why entropy should not be treated as a model fit criterion.
-It also illustrates that criteria for class enumeration should be explicit, because we will likely come to a different decision depending on which criteria are used.
+It also illustrates that criteria for class enumeration should be explicit,
+because we will likely come to a different decision depending on which criteria are used.
 
@@
 \begin{Highlighting}[]
-\NormalTok{fit }\OtherTok{\textless{}{-}}\NormalTok{ fit[}\SpecialCharTok{!}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{n\_min }\SpecialCharTok{\textless{}}\NormalTok{ .}\DecValTok{1}\NormalTok{, ]}
+\NormalTok{fit }\OtherTok{\textless{}{-}}\NormalTok{ fit[}\SpecialCharTok{!}\NormalTok{fit}\SpecialCharTok{$}\NormalTok{n\_min }\SpecialCharTok{\textless{}} \FloatTok{0.1}\NormalTok{, ]}
 \end{Highlighting}
@@
 
-If our strategy is to optimize fit, we can examine the fit table above, or plot a scree plot for the BIC by calling \texttt{plot(fit)}.
+If our strategy is to optimize fit,
+we can examine the fit table above,
+or plot a scree plot for the BIC by calling \texttt{plot(fit)}.
 Note that, due to the large sample size, all ICs give identical conclusions.
@@
 \begin{Highlighting}[]
-\FunctionTok{plot}\NormalTok{(fit) }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \DecValTok{90}\NormalTok{, }\AttributeTok{vjust =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{hjust=}\DecValTok{1}\NormalTok{))}
+\FunctionTok{plot}\NormalTok{(fit) }\SpecialCharTok{+} \FunctionTok{theme}\NormalTok{(}\AttributeTok{axis.text.x =} \FunctionTok{element\_text}\NormalTok{(}\AttributeTok{angle =} \DecValTok{90}\NormalTok{, }\AttributeTok{vjust =} \FloatTok{0.5}\NormalTok{,}
+    \AttributeTok{hjust =} \DecValTok{1}\NormalTok{))}
 \end{Highlighting}
@@
 
-Out of the 16 models that remain after removing those with \textless{} 10\% of cases in the smallest class, one model stands out: The 2-class model with free (co)variances.
+Out of the 16 models that remain after removing those with \textless{} 10\% of cases in the smallest class,
+one model stands out:
+The 2-class model with free (co)variances.
 We thus select this as our final model.
@@
 \NormalTok{cp }\OtherTok{\textless{}{-}} \FunctionTok{class\_prob}\NormalTok{(res\_bic)}
-\NormalTok{results }\OtherTok{\textless{}{-}} \FunctionTok{table\_results}\NormalTok{(res\_bic, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"std\_est"}\NormalTok{))}
+\NormalTok{results }\OtherTok{\textless{}{-}} \FunctionTok{table\_results}\NormalTok{(res\_bic, }\AttributeTok{columns =} \FunctionTok{c}\NormalTok{(}\StringTok{"label"}\NormalTok{, }\StringTok{"est"}\NormalTok{,}
+    \StringTok{"std\_est"}\NormalTok{))}
 \NormalTok{results}
@@
 The first class (57\%) accounts for smaller fragments, and the second class (43\%) accounts for some of the right-skew in fragments' length and width.
-We can simply label class 1 as \emph{small fragments}, and class 2 as \emph{larger fragments}.
+We can simply label class 1 as \emph{small fragments},
+and class 2 as \emph{larger fragments}.
 
@@
 
-There are, however, concerns about the interpretability of this solutions: the entropy is \texttt{.56} and the minimum classification probability is\texttt{.81}.
+There are, however,
+concerns about the interpretability of this solutions:
+the entropy is \texttt{.56} and the minimum classification probability is\texttt{.81}.
 This is because of substantial overlap in the distributions of the two classes.
@@
 Finally, we may want to compare the different classes on auxiliary variables or models.
-The \texttt{BCH()} function applies three-step analysis, which compares the classes using a multi-group model, controlling for classification error.
+The \texttt{BCH()} function applies three-step analysis,
+which compares the classes using a multi-group model,
+controlling for classification error.
 For example, we can test whether polymer type differs between the two classes:
@@
 \NormalTok{df\_pt }\OtherTok{\textless{}{-}} \FunctionTok{mx\_dummies}\NormalTok{(df\_analyze}\SpecialCharTok{$}\NormalTok{poly\_type)}
-\NormalTok{aux\_pt }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_bic, }\AttributeTok{model =} \StringTok{"poly\_typeOther | t1}
+\NormalTok{aux\_pt }\OtherTok{\textless{}{-}} \FunctionTok{BCH}\NormalTok{(res\_bic, mo}
+\AttributeTok{l =} \StringTok{"poly\_typeOther | t1}
 \StringTok{                                poly\_typePE | t1}
-\StringTok{                                poly\_typePP | t1"}\NormalTok{, }\AttributeTok{data =}\NormalTok{ df\_pt)}
+\StringTok{                                poly\_typePP | t1"}\NormalTok{,}
+    \AttributeTok{data =}\NormalTok{ df\_pt)}
 \NormalTok{aux\_pt }\OtherTok{\textless{}{-}} \FunctionTok{mxTryHardOrdinal}\NormalTok{(aux\_pt)}
@@
 
-To obtain an omnibus likelihood ratio test of the significance of the differences in polymer type across classes, use \texttt{lr\_test(aux\_pt)}.
+To obtain an omnibus likelihood ratio test of the significance of the differences in polymer type across classes,
+use \texttt{lr\_test(aux\_pt)}.
 The results indicate that there are significant differences in polymer types across classes, \(\Delta LL(3) = 17.14, p < .001\).

### 2023-01-24 appendix_c.Rmd
revert MGV previous commit

@@
 footnotelist      : no
+
 classoption       : "man"
 output            : papaja::apa6_pdf
-editor_options: 
-  markdown: 
-    wrap: 72
 ---
@@
 
-This vignette illustrated `tidySEM`'s ability to perform latent class
-growth analysis, or growth mixture modeling. The simulated data used for
-this example are inspired by work in progress by Plas and colleagues, on
-heterogeneity in depression trajectories among Dutch military personnel
-who were deployed to Afghanistan. The original data were collected as
-part of the *Prospection in Stress-related Military Research (PRISMO)*
-study, which examined of psychological problems after deployment in more
-than 1,000 Dutch military personnel from 2005-2019.
+This vignette illustrated `tidySEM`'s ability to perform latent class growth analysis, or growth mixture modeling.
+The simulated data used for this example are inspired by work in progress by Plas and colleagues,
+on heterogeneity in depression trajectories among Dutch military personnel who were deployed to Afghanistan.
+The original data were collected as part of the *Prospection in Stress-related Military Research (PRISMO)* study,
+which examined of psychological problems after deployment in more than 1,000 Dutch military personnel from 2005-2019.
 
@@
 
-Note that all variables were extremely right-skewed due to censoring at
-the lower end of the scale.
+Note that all variables were extremely right-skewed due to censoring at the lower end of the scale.
 
@@
 ```
-
 ```{r, echo = FALSE, out.width="50%", eval = pkgdown}
@@
 
-As this type of skew can result in convergence problems in LCGA, we
-compared several transformations to reduce skew: The square and cube
-root, log, inverse, and Box-Cox transformations.
+As this type of skew can result in convergence problems in LCGA,
+we compared several transformations to reduce skew:
+The square and cube root, log, inverse, and Box-Cox transformations.
 
@@
 ```
-
 ```{r, echo = FALSE, eval = pkgdown, out.width="50%"}
@@
 Evidently, the Box-Cox transformation reduced skew the most.
-Consequently, we proceeded with the Box-Cox transformed scores for
-analysis.
+Consequently, we proceeded with the Box-Cox transformed scores for analysis.
 
@@
 
-Next, we estimated a latent class growth model for SCL. The model
-included an overall intercept, centered at T1, `i`. To model the
-potential effect of deployment on depresion, we also included a dummy
-variable that was zero before deployment, and 1 after deployment,
-`step`. Finally, to model potential change (or recovery) in depression
-post-deployment, we included a linear slope from T2-T6, `s`. All
-variances of growth parameters were fixed to zero due to the sparse
-nature of the data. In this vignette, we do not consider more than 5
-classes, because the analyses are computationally very intensive and the
-data were simulated from a 3-class model.
+Next, we estimated a latent class growth model
+for SCL.
+The model included an overall intercept, centered at T1, `i`.
+To model the potential effect of deployment on
+depresion,
+we also included a dummy variable that was zero before
+deployment, and 1 after deployment, `step`.
+Finally, to model potential change (or recovery) in depression post-deployment,
+we included a linear slope from T2-T6, `s`.
+All variances of growth parameters were fixed to zero due to the sparse nature of the data.
+In this vignette,
+we do not consider more than 5 classes,
+because the analyses are computationally very intensive and the data were simulated from a 3-class model.
 
-It is important to not that in LCGA, the subgroups will be limited by
-the specify growth structure, meaning that LCA will identify distinctive
-growth patterns within the intercept, step, and slope growth. For
-example, if there is a subgroup that follows a quadratic growth pattern
-this models will not be able to identify it.
-
-**NOTE: The time scales in this model are not correct; it currently
-assumes that all measurements are equidistant. Feel free to experiment
-with adjusting this.**
+**NOTE: The time scales in this model are not correct; it currently assumes that all measurements are equidistant. Feel free to experiment with adjusting this.**
 
@@
 ```
-
 ```{r, echo = FALSE, eval = FALSE}
@@
 
-Note that the first model showed convergence problems, throwing the
-error: *The model does not satisfy the first-order optimality conditions
-to the required accuracy, and no improved point for the merit function
-could be found during the final linesearch.* To address this problem, we
-performed additional iterations to\
-find a better solution, using `OpenMx::mxTryHardWideSearch()`. This also
-illustrates that `tidySEM` mixture models inherit from `OpenMx`'s
-`MxModel`, and thus, different `OpenMx` functions can be used to act on
-models specified via `tidySEM`.
+Note that the first model showed convergence problems, throwing the error:
+*The model does not satisfy the 
+first-order optimality conditions to
+the required accuracy, and no improved
+point for the merit function could be
+found during the final linesearch.*
+To address this problem, we performed
+additional iterations to  
+find a better solution, using `OpenMx::mxTryHardWideSearch()`.
+This also illustrates that `tidySEM` mixture models inherit from `OpenMx`'s `MxModel`,
+and thus, different `OpenMx` functions can be used to act on models specified via `tidySEM`.
 
-The fifth model also evidenced convergence problems, but this (as we
-will see) is because the solution is overfitted.
+The fifth model also evidenced convergence problems, but this (as we will see) is because the solution is overfitted.
 
@@
 
-To determine the correct number of classes, we considered the following
-criteria:
+To determine the correct number of classes, we considered the following criteria:
 
-1.  We do not consider classes with, on average, fewer than 5
-    participants per parameter in a class due to potential local
-    underidentification
-2.  Lower values for information criteria (AIC, BIC, saBIC) indicate
-    better fit
-3.  Significant Lo-Mendell-Rubin LRT test indicates better fit for $k$
-    vs $k-1$ classes
-4.  We do not consider solutions with entropy \< .90 because poor class
-    separability compromises interpretability of the results
-5.  We do not consider solutions with minimum posterior classification
-    probability \< .90 because poor class separability compromises
-    interpretability of the results
+1. We do not consider classes with, on average, fewer than 5 participants per parameter in a class due to potential local underidentification
+1. Lower values for information criteria (AIC, BIC, saBIC) indicate better fit
+1. Significant Lo-Mendell-Rubin LRT test indicates better fit for $k$ vs $k-1$ classes
+1. We do not consider solutions with entropy < .90 because poor class separability compromises interpretability of the results
+1. We do not consider solutions with minimum posterior classification probability < .90 because poor class separability compromises interpretability of the results
 
@@
 ```
-
 ```{r tabfit, eval = TRUE, echo = FALSE, results = "asis"}
@@
 
-According to the Table, increasing the number of classes keeps
-increasing model fit according to all ICs except the BIC, which
-increased after 3 classes.
+According to the Table, increasing the number of classes keeps increasing model fit according to all ICs except the BIC, which increased after 3 classes.
 
-The first two LMR tests are significant, indicating that a 2- and
-3-class solution were a significant improvement over a 1- and 2-class
-solution, respectively. However, solutions with \>3 classes had entropy
-and minimum posterior classification probability below the pre-specified
-thresholds. Models with \>3 solutions also had fewer than five
-observations per parameter. This suggests that the preferred model
-should be selected from 1-3 classes.
+The first two LMR tests are significant,
+indicating that a 2- and 3-class solution were a significant improvement over a 1- and 2-class solution, respectively.
+However, solutions with >3 classes had entropy and minimum posterior classification probability below the pre-specified thresholds.
+Models with >3 solutions also had fewer than five observations per parameter.
+This suggests that the preferred model should be selected from 1-3 classes.
 
@@
 
-A scree plot indicates that the largest decrease in ICs occurs from 1-2
-classes, and the inflection point for all ICs is at 3 classes. Moreover,
-the BIC increased after 3 classes. A three-class solution thus appears
-to be the most parsimonious solution with good fit.
+A scree plot indicates that
+the largest decrease in ICs occurs from 1-2 classes,
+and the inflection point for all ICs is at 3 classes.
+Moreover, the BIC increased after 3 classes.
+A three-class solution thus appears to be the most parsimonious
+solution with good fit.
 
@@
 ```
-
 ```{r echo = FALSE, eval = pkgdown, out.width="50%"}
@@
 
-Based on the aforementioned criteria, we selected a 3-class model for
-further analyses. First, to prevent label switching, we re-order these
-classes by the value of the intercept `i`. Then, we report the estimated
-parameters.
-
+Based on the aforementioned criteria,
+we selected a 3-class model for further analyses.
+First, to prevent label switching,
+we re-order these classes by the value of the intercept `i`.
+Then, we report the estimated parameters.
 ```{r echo = TRUE, eval = FALSE}
@@
 ```
-
 ```{r echo = FALSE, eval = run_everything, results='asis'}
@@
 ```
-
 ```{r tabres, echo=FALSE, eval = TRUE, results='asis'}
@@
 
-As evident from these results, Class 1 started at a relatively lower
-level of depressive symptoms, experienced a decrease after deployment,
-followed by increase over time. Class 2 started at a moderate level of
-depressive symptoms, experienced an increase after deployment, followed
-by significant increase over time from T2-T6. Class 3 started at a
-relatively higher level, experienced an increase after deployment,
-followed by stability.
+As evident from these results, 
+Class 1 started at a relatively lower level of depressive symptoms,
+experienced a decrease after deployment,
+followed by increase over time.
+Class 2 started at a moderate level of depressive symptoms,
+experienced an increase after deployment,
+followed by significant increase over time from T2-T6.
+Class 3 started at a relatively higher level,
+experienced an increase after deployment, followed by stability.
 
@@
 To test whether parameters are significantly different between classes,
-we can use Wald tests. Wald tests can be specified for all parameters in
-the model, using the hypothesis syntax from the `bain` package for
-informative hypothesis testing.
+we can use Wald tests.
+Wald tests can be specified for all parameters in the model,
+using the hypothesis syntax from the `bain` package for informative hypothesis testing.
 
-To identify the names of parameters in the model, we can use the `name`
-column of the results table above. Alternatively, to see all parameters
-in the model, run:
+To identify the names of parameters in the model,
+we can use the `name` column of the results table above.
+Alternatively, to see all parameters in the model, run:
 
@@
 ```
-
 ```{r echo = FALSE, eval = TRUE}
@@
 
-Next, specify equality constrained hypotheses. For example, a hypothesis
-that states that the mean intercept is equal across groups is specified
-as follows:
+Next, specify equality constrained hypotheses.
+For example, a hypothesis that states that the mean intercept is equal across groups is specified as follows:
 
@@
 
-It is also possible to consider comparisons between two classes at a
-time. When conducting many significance tests, consider correcting for
-multiple comparisons however.
+It is also possible to consider comparisons between two classes at a time.
+When conducting many significance tests,
+consider correcting for multiple comparisons however.
 
@@
 ```
-
 ```{r waldtests, echo = FALSE, eval = TRUE, results='asis'}
@@
 
-All Wald tests are significant, indicating that there are significant
-differences between the intercepts, step function, and slopes of the
-three classes.
+All Wald tests are significant, indicating that there are significant differences between the intercepts, step function, and slopes of the three classes.
 
@@
 
-Finally, we can plot the growth trajectories. This can help interpret
-the results better, as well as the residual heterogeneity around class
-trajectories.
+Finally, we can plot the growth trajectories.
+This can help interpret the results better,
+as well as the residual heterogeneity around class trajectories.
 
@@
 
-Note that the observed individual trajectories show very high
-variability within classes.
+Note that the observed individual trajectories show very high variability within classes.
 

### 2023-01-24 appendix_c.tex
revert MGV previous commit

@@
 
-This vignette illustrated \texttt{tidySEM}'s ability to perform latent class
-growth analysis, or growth mixture modeling. The simulated data used for
-this example are inspired by work in progress by Plas and colleagues, on
-heterogeneity in depression trajectories among Dutch military personnel
-who were deployed to Afghanistan. The original data were collected as
-part of the \emph{Prospection in Stress-related Military Research (PRISMO)}
-study, which examined of psychological problems after deployment in more
-than 1,000 Dutch military personnel from 2005-2019.
+This vignette illustrated \texttt{tidySEM}'s ability to perform latent class growth analysis, or growth mixture modeling.
+The simulated data used for this example are inspired by work in progress by Plas and colleagues,
+on heterogeneity in depression trajectories among Dutch military personnel who were deployed to Afghanistan.
+The original data were collected as part of the \emph{Prospection in Stress-related Military Research (PRISMO)} study,
+which examined of psychological problems after deployment in more than 1,000 Dutch military personnel from 2005-2019.
 
@@
 \NormalTok{df\_plot }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(df, }\AttributeTok{direction =} \StringTok{"long"}\NormalTok{, }\AttributeTok{varying =} \FunctionTok{names}\NormalTok{(df))}
-\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ scl)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{time) }\SpecialCharTok{+} \FunctionTok{theme\_bw}\NormalTok{()}
+\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ scl)) }\SpecialCharTok{+} \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+} \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{time) }\SpecialCharTok{+}
+    \FunctionTok{theme\_bw}\NormalTok{()}
 \end{Highlighting}
@@
 \CommentTok{\# Log{-}transform}
-\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{log }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\FunctionTok{log}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl), }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{log }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\FunctionTok{log}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl), }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}
+    \DecValTok{1}\NormalTok{))}
 \CommentTok{\# Square root transform}
-\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{sqrt }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\FunctionTok{sqrt}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl), }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{sqrt }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\FunctionTok{sqrt}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl), }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}
+    \DecValTok{1}\NormalTok{))}
 \CommentTok{\# Cube root transform}
-\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{qrt }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl}\SpecialCharTok{\^{}}\NormalTok{.}\DecValTok{33}\NormalTok{, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{qrt }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{scl}\SpecialCharTok{\^{}}\FloatTok{0.33}\NormalTok{, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}
+    \DecValTok{1}\NormalTok{))}
 \CommentTok{\# Reciprocal transform}
-\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{reciprocal }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{scl, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{reciprocal }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{scl, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}
+    \DecValTok{1}\NormalTok{))}
 \CommentTok{\# Define function for Box{-}Cox transformation}
-\NormalTok{bc }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, lambda)\{}
-\NormalTok{  (((x }\SpecialCharTok{\^{}}\NormalTok{ lambda) }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{) }\SpecialCharTok{/}\NormalTok{ lambda)}
+\NormalTok{bc }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, lambda) \{}
+\NormalTok{    (((x}\SpecialCharTok{\^{}}\NormalTok{lambda) }\SpecialCharTok{{-}} \DecValTok{1}\NormalTok{)}\SpecialCharTok{/}\NormalTok{lambda)}
 \NormalTok{\}}
 \CommentTok{\# Inverse Box{-}Cox transformation}
-\NormalTok{invbc }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, lambda)\{}
-\NormalTok{  ((x}\SpecialCharTok{*}\NormalTok{lambda)}\SpecialCharTok{+}\DecValTok{1}\NormalTok{)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{lambda)}
+\NormalTok{invbc }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, lambda) \{}
+\NormalTok{    ((x }\SpecialCharTok{*}\NormalTok{ lambda) }\SpecialCharTok{+} \DecValTok{1}\NormalTok{)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\NormalTok{lambda)}
 \NormalTok{\}}
@@
 \NormalTok{rng\_bc }\OtherTok{\textless{}{-}} \FunctionTok{range}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{boxcox)}
-\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{boxcox }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{boxcox, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{))}
+\NormalTok{df\_scores}\SpecialCharTok{$}\NormalTok{boxcox }\OtherTok{\textless{}{-}}\NormalTok{ scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(df\_scores}\SpecialCharTok{$}\NormalTok{boxcox, }\AttributeTok{to =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}
+    \DecValTok{1}\NormalTok{))}
 \CommentTok{\# Rescale SCL}
@@
 \CommentTok{\# Make plot data}
-\NormalTok{df\_plot }\OtherTok{\textless{}{-}} \FunctionTok{do.call}\NormalTok{(rbind, }\FunctionTok{lapply}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"scl"}\NormalTok{, }\StringTok{"log"}\NormalTok{, }\StringTok{"sqrt"}\NormalTok{, }\StringTok{"qrt"}\NormalTok{, }\StringTok{"boxcox"}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(n)\{}
-  \FunctionTok{data.frame}\NormalTok{(df\_scores[}\FunctionTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"id"}\NormalTok{)],}
-             \AttributeTok{Value =}\NormalTok{ df\_scores[[n]],}
+\NormalTok{df\_plot }\OtherTok{\textless{}{-}} \FunctionTok{do.call}\NormalTok{(rbind, }\FunctionTok{lapply}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"scl"}\NormalTok{, }\StringTok{"log"}\NormalTok{, }\StringTok{"sqrt"}\NormalTok{, }\StringTok{"qrt"}\NormalTok{,}
+    \StringTok{"boxcox"}\NormalTok{), }\ControlFlowTok{function}\NormalTok{(n) \{}
+    \FunctionTok{data.frame}\NormalTok{(df\_scores[}\FunctionTok{c}\NormalTok{(}\StringTok{"time"}\NormalTok{, }\StringTok{"id"}\NormalTok{)], }\AttributeTok{Value =}\NormalTok{ df\_scores[[n]],}
         \AttributeTok{Transformation =}\NormalTok{ n)}
@@
 \CommentTok{\# Plot}
-\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value, }\AttributeTok{colour =}\NormalTok{ Transformation)) }\SpecialCharTok{+}
-  \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{time) }\SpecialCharTok{+}
-  \FunctionTok{scale\_y\_sqrt}\NormalTok{() }\SpecialCharTok{+}
-  \FunctionTok{xlab}\NormalTok{(}\StringTok{"scl (rescaled to 0{-}1)"}\NormalTok{) }\SpecialCharTok{+}
+\FunctionTok{ggplot}\NormalTok{(df\_plot, }\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ Value, }\AttributeTok{colour =}\NormalTok{ Transformation)) }\SpecialCharTok{+} \FunctionTok{geom\_density}\NormalTok{() }\SpecialCharTok{+}
+    \FunctionTok{facet\_wrap}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{time) }\SpecialCharTok{+} \FunctionTok{scale\_y\_sqrt}\NormalTok{() }\SpecialCharTok{+} \FunctionTok{xlab}\NormalTok{(}\StringTok{"scl (rescaled to 0{-}1)"}\NormalTok{) }\SpecialCharTok{+}
     \FunctionTok{theme\_bw}\NormalTok{()}
@@
 \NormalTok{dat }\OtherTok{\textless{}{-}}\NormalTok{ df\_scores[, }\FunctionTok{c}\NormalTok{(}\StringTok{"id"}\NormalTok{, }\StringTok{"time"}\NormalTok{, }\StringTok{"boxcox"}\NormalTok{)]}
-\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(dat, }\AttributeTok{direction =} \StringTok{"wide"}\NormalTok{, }\AttributeTok{v.names =} \StringTok{"boxcox"}\NormalTok{, }\AttributeTok{timevar =} \StringTok{"time"}\NormalTok{, }\AttributeTok{idvar =} \StringTok{"id"}\NormalTok{)}
+\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{reshape}\NormalTok{(dat, }\AttributeTok{direction =} \StringTok{"wide"}\NormalTok{, }\AttributeTok{v.names =} \StringTok{"boxcox"}\NormalTok{, }\AttributeTok{timevar =} \StringTok{"time"}\NormalTok{,}
+    \AttributeTok{idvar =} \StringTok{"id"}\NormalTok{)}
 \FunctionTok{names}\NormalTok{(dat) }\OtherTok{\textless{}{-}} \FunctionTok{gsub}\NormalTok{(}\StringTok{"boxcox."}\NormalTok{, }\StringTok{"scl"}\NormalTok{, }\FunctionTok{names}\NormalTok{(dat))}
@@
 
-Next, we estimated a latent class growth model for SCL. The model
-included an overall intercept, centered at T1, \texttt{i}. To model the
-potential effect of deployment on depresion, we also included a dummy
-variable that was zero before deployment, and 1 after deployment,
-\texttt{step}. Finally, to model potential change (or recovery) in depression
-post-deployment, we included a linear slope from T2-T6, \texttt{s}. All
-variances of growth parameters were fixed to zero due to the sparse
-nature of the data. In this vignette, we do not consider more than 5
-classes, because the analyses are computationally very intensive and the
-data were simulated from a 3-class model.
+Next, we estimated a latent class growth model
+for SCL.
+The model included an overall intercept, centered at T1, \texttt{i}.
+To model the potential effect of deployment on
+depresion,
+we also included a dummy variable that was zero before
+deployment, and 1 after deployment, \texttt{step}.
+Finally, to model potential change (or recovery) in depression post-deployment,
+we included a linear slope from T2-T6, \texttt{s}.
+All variances of growth parameters were fixed to zero due to the sparse nature of the data.
+In this vignette,
+we do not consider more than 5 classes,
+because the analyses are computationally very intensive and the data were simulated from a 3-class model.
 
-It is important to not that in LCGA, the subgroups will be limited by
-the specify growth structure, meaning that LCA will identify distinctive
-growth patterns within the intercept, step, and slope growth. For
-example, if there is a subgroup that follows a quadratic growth pattern
-this models will not be able to identify it.
-
-\textbf{NOTE: The time scales in this model are not correct; it currently
-assumes that all measurements are equidistant. Feel free to experiment
-with adjusting this.}
+\textbf{NOTE: The time scales in this model are not correct; it currently assumes that all measurements are equidistant. Feel free to experiment with adjusting this.}
 
@@
 \NormalTok{dat[[}\StringTok{"id"}\NormalTok{]] }\OtherTok{\textless{}{-}} \ConstantTok{NULL}
-\NormalTok{res\_step }\OtherTok{\textless{}{-}} \FunctionTok{mx\_growth\_mixture}\NormalTok{(}
-  \AttributeTok{model =}
-  \StringTok{"}
+\NormalTok{res\_step }\OtherTok{\textless{}{-}} \FunctionTok{mx\_growth\_mixture}\NormalTok{(}\AttributeTok{model =} \StringTok{"}
 \StringTok{  i =\textasciitilde{} 1*scl1 + 1*scl2 + 1*scl3 +1*scl4 +1*scl5 +1*scl6}
@@
 \StringTok{  i \textasciitilde{}\textasciitilde{} 0*step}
-\StringTok{  s \textasciitilde{}\textasciitilde{} 0*step"}\NormalTok{, }\AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{,}
-  \AttributeTok{data =}\NormalTok{ dat)}
-\CommentTok{\# Additional iterations because of}
-\CommentTok{\# convergence problems for model 1:}
+\StringTok{  s \textasciitilde{}\textasciitilde{} 0*step"}\NormalTok{,}
+    \AttributeTok{classes =} \DecValTok{1}\SpecialCharTok{:}\DecValTok{5}\NormalTok{, }\AttributeTok{data =}\NormalTok{ dat)}
+\CommentTok{\# Additional iterations because of convergence problems for}
+\CommentTok{\# model 1:}
 \NormalTok{res\_step[[}\DecValTok{1}\NormalTok{]] }\OtherTok{\textless{}{-}} \FunctionTok{mxTryHardWideSearch}\NormalTok{(res\_step[[}\DecValTok{1}\NormalTok{]], }\AttributeTok{extraTries =} \DecValTok{50}\NormalTok{)}
@@
 
-Note that the first model showed convergence problems, throwing the
-error: \emph{The model does not satisfy the first-order optimality conditions
-to the required accuracy, and no improved point for the merit function
-could be found during the final linesearch.} To address this problem, we
-performed additional iterations to\\
-find a better solution, using \texttt{OpenMx::mxTryHardWideSearch()}. This also
-illustrates that \texttt{tidySEM} mixture models inherit from \texttt{OpenMx}'s
-\texttt{MxModel}, and thus, different \texttt{OpenMx} functions can be used to act on
-models specified via \texttt{tidySEM}.
+Note that the first model showed convergence problems, throwing the error:
+\emph{The model does not satisfy the
+first-order optimality conditions to
+the required accuracy, and no improved
+point for the merit function could be
+found during the final linesearch.}
+To address this problem, we performed
+additional iterations to\\
+find a better solution, using \texttt{OpenMx::mxTryHardWideSearch()}.
+This also illustrates that \texttt{tidySEM} mixture models inherit from \texttt{OpenMx}'s \texttt{MxModel},
+and thus, different \texttt{OpenMx} functions can be used to act on models specified via \texttt{tidySEM}.
 
-The fifth model also evidenced convergence problems, but this (as we
-will see) is because the solution is overfitted.
+The fifth model also evidenced convergence problems, but this (as we will see) is because the solution is overfitted.
 
@@
 
-To determine the correct number of classes, we considered the following
-criteria:
+To determine the correct number of classes, we considered the following criteria:
 
@@
 \item
-  We do not consider classes with, on average, fewer than 5
-  participants per parameter in a class due to potential local
-  underidentification
+  We do not consider classes with, on average, fewer than 5 participants per parameter in a class due to potential local underidentification
 \item
-  Lower values for information criteria (AIC, BIC, saBIC) indicate
-  better fit
+  Lower values for information criteria (AIC, BIC, saBIC) indicate better fit
 \item
-  Significant Lo-Mendell-Rubin LRT test indicates better fit for \(k\)
-  vs \(k-1\) classes
+  Significant Lo-Mendell-Rubin LRT test indicates better fit for \(k\) vs \(k-1\) classes
 \item
-  We do not consider solutions with entropy \textless{} .90 because poor class
-  separability compromises interpretability of the results
+  We do not consider solutions with entropy \textless{} .90 because poor class separability compromises interpretability of the results
 \item
-  We do not consider solutions with minimum posterior classification
-  probability \textless{} .90 because poor class separability compromises
-  interpretability of the results
+  We do not consider solutions with minimum posterior classification probability \textless{} .90 because poor class separability compromises interpretability of the results
 \end{enumerate}
@@
 \CommentTok{\# Select columns}
-\NormalTok{tab\_fit }\OtherTok{\textless{}{-}}\NormalTok{ tab\_fit[, }\FunctionTok{c}\NormalTok{(}\StringTok{"Name"}\NormalTok{, }\StringTok{"Classes"}\NormalTok{, }\StringTok{"LL"}\NormalTok{, }\StringTok{"Parameters"}\NormalTok{, }\StringTok{"BIC"}\NormalTok{, }\StringTok{"Entropy"}\NormalTok{, }\StringTok{"prob\_min"}\NormalTok{, }\StringTok{"n\_min"}\NormalTok{, }\StringTok{"warning"}\NormalTok{, }\StringTok{"lmr\_p"}\NormalTok{)]}
+\NormalTok{tab\_fit }\OtherTok{\textless{}{-}}\NormalTok{ tab\_fit[, }\FunctionTok{c}\NormalTok{(}\StringTok{"Name"}\NormalTok{, }\StringTok{"Classes"}\NormalTok{, }\StringTok{"LL"}\NormalTok{, }\StringTok{"Parameters"}\NormalTok{,}
+    \StringTok{"BIC"}\NormalTok{, }\StringTok{"Entropy"}\NormalTok{, }\StringTok{"prob\_min"}\NormalTok{, }\StringTok{"n\_min"}\NormalTok{, }\StringTok{"warning"}\NormalTok{, }\StringTok{"lmr\_p"}\NormalTok{)]}
 \NormalTok{tab\_fit}
@@
 
-According to the Table, increasing the number of classes keeps
-increasing model fit according to all ICs except the BIC, which
-increased after 3 classes.
+According to the Table, increasing the number of classes keeps increasing model fit according to all ICs except the BIC, which increased after 3 classes.
 
-The first two LMR tests are significant, indicating that a 2- and
-3-class solution were a significant improvement over a 1- and 2-class
-solution, respectively. However, solutions with \textgreater3 classes had entropy
-and minimum posterior classification probability below the pre-specified
-thresholds. Models with \textgreater3 solutions also had fewer than five
-observations per parameter. This suggests that the preferred model
-should be selected from 1-3 classes.
+The first two LMR tests are significant,
+indicating that a 2- and 3-class solution were a significant improvement over a 1- and 2-class solution, respectively.
+However, solutions with \textgreater3 classes had entropy and minimum posterior classification probability below the pre-specified thresholds.
+Models with \textgreater3 solutions also had fewer than five observations per parameter.
+This suggests that the preferred model should be selected from 1-3 classes.
 
@@
 
-Based on the aforementioned criteria, we selected a 3-class model for
-further analyses. First, to prevent label switching, we re-order these
-classes by the value of the intercept \texttt{i}. Then, we report the estimated
-parameters.
+Based on the aforementioned criteria,
+we selected a 3-class model for further analyses.
+First, to prevent label switching,
+we re-order these classes by the value of the intercept \texttt{i}.
+Then, we report the estimated parameters.
 
@@
 \begin{Highlighting}[]
-\NormalTok{res\_final }\OtherTok{\textless{}{-}} \FunctionTok{mx\_switch\_labels}\NormalTok{(res\_step[[}\DecValTok{3}\NormalTok{]], }\AttributeTok{param =} \StringTok{"M[1,7]"}\NormalTok{, }\AttributeTok{decreasing =} \ConstantTok{FALSE}\NormalTok{)}
+\NormalTok{res\_final }\OtherTok{\textless{}{-}} \FunctionTok{mx\_switch\_labels}\NormalTok{(res\_step[[}\DecValTok{3}\NormalTok{]], }\AttributeTok{param =} \StringTok{"M[1,7]"}\NormalTok{,}
+    \AttributeTok{decreasing =} \ConstantTok{FALSE}\NormalTok{)}
 \NormalTok{tab\_res }\OtherTok{\textless{}{-}} \FunctionTok{table\_results}\NormalTok{(res\_final, }\AttributeTok{columns =} \ConstantTok{NULL}\NormalTok{)}
 \CommentTok{\# Select rows and columns}
-\NormalTok{tab\_res }\OtherTok{\textless{}{-}}\NormalTok{ tab\_res[tab\_res}\SpecialCharTok{$}\NormalTok{Category }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{"Means"}\NormalTok{, }\StringTok{"Variances"}\NormalTok{), }\FunctionTok{c}\NormalTok{(}\StringTok{"Category"}\NormalTok{, }\StringTok{"lhs"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"pval"}\NormalTok{, }\StringTok{"confint"}\NormalTok{, }\StringTok{"name"}\NormalTok{)]}
+\NormalTok{tab\_res }\OtherTok{\textless{}{-}}\NormalTok{ tab\_res[tab\_res}\SpecialCharTok{$}\NormalTok{Category }\SpecialCharTok{\%in\%} \FunctionTok{c}\NormalTok{(}\StringTok{"Means"}\NormalTok{, }\StringTok{"Variances"}\NormalTok{),}
+    \FunctionTok{c}\NormalTok{(}\StringTok{"Category"}\NormalTok{, }\StringTok{"lhs"}\NormalTok{, }\StringTok{"est"}\NormalTok{, }\StringTok{"se"}\NormalTok{, }\StringTok{"pval"}\NormalTok{, }\StringTok{"confint"}\NormalTok{, }\StringTok{"name"}\NormalTok{)]}
 \NormalTok{tab\_res}
@@
 
-As evident from these results, Class 1 started at a relatively lower
-level of depressive symptoms, experienced a decrease after deployment,
-followed by increase over time. Class 2 started at a moderate level of
-depressive symptoms, experienced an increase after deployment, followed
-by significant increase over time from T2-T6. Class 3 started at a
-relatively higher level, experienced an increase after deployment,
-followed by stability.
+As evident from these results,
+Class 1 started at a relatively lower level of depressive symptoms,
+experienced a decrease after deployment,
+followed by increase over time.
+Class 2 started at a moderate level of depressive symptoms,
+experienced an increase after deployment,
+followed by significant increase over time from T2-T6.
+Class 3 started at a relatively higher level,
+experienced an increase after deployment, followed by stability.
 
@@
 To test whether parameters are significantly different between classes,
-we can use Wald tests. Wald tests can be specified for all parameters in
-the model, using the hypothesis syntax from the \texttt{bain} package for
-informative hypothesis testing.
+we can use Wald tests.
+Wald tests can be specified for all parameters in the model,
+using the hypothesis syntax from the \texttt{bain} package for informative hypothesis testing.
 
-To identify the names of parameters in the model, we can use the \texttt{name}
-column of the results table above. Alternatively, to see all parameters
-in the model, run:
+To identify the names of parameters in the model,
+we can use the \texttt{name} column of the results table above.
+Alternatively, to see all parameters in the model, run:
 
@@
 \begin{Highlighting}[]
-\NormalTok{wald\_tests }\OtherTok{\textless{}{-}} \FunctionTok{wald\_test}\NormalTok{(res\_final, }
-                   \StringTok{"}
+\NormalTok{wald\_tests }\OtherTok{\textless{}{-}} \FunctionTok{wald\_test}\NormalTok{(res\_final, }\StringTok{"}
 \StringTok{                   class1.M[1,7] = class2.M[1,7]\&}
@@
 \begin{Highlighting}[]
-\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{plot\_growth}\NormalTok{(res\_step[[}\DecValTok{3}\NormalTok{]], }\AttributeTok{rawdata =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{alpha\_range =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, .}\DecValTok{05}\NormalTok{))}
+\NormalTok{p }\OtherTok{\textless{}{-}} \FunctionTok{plot\_growth}\NormalTok{(res\_step[[}\DecValTok{3}\NormalTok{]], }\AttributeTok{rawdata =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{alpha\_range =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}
+    \FloatTok{0.05}\NormalTok{))}
 \CommentTok{\# Add Y{-}axis breaks in original scale}
 \NormalTok{brks }\OtherTok{\textless{}{-}} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{length.out =} \DecValTok{5}\NormalTok{)}
-\NormalTok{labs }\OtherTok{\textless{}{-}} \FunctionTok{round}\NormalTok{(}\FunctionTok{invbc}\NormalTok{(scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(brks, }\AttributeTok{from =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{to =}\NormalTok{ rng\_bc), lambda))}
-\NormalTok{p }\OtherTok{\textless{}{-}}\NormalTok{ p }\SpecialCharTok{+} \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{length.out =} \DecValTok{5}\NormalTok{), }\AttributeTok{labels =}\NormalTok{ labs) }\SpecialCharTok{+} \FunctionTok{ylab}\NormalTok{(}\StringTok{"SCL (rescaled from Box{-}Cox)"}\NormalTok{)}
+\NormalTok{labs }\OtherTok{\textless{}{-}} \FunctionTok{round}\NormalTok{(}\FunctionTok{invbc}\NormalTok{(scales}\SpecialCharTok{::}\FunctionTok{rescale}\NormalTok{(brks, }\AttributeTok{from =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{to =}\NormalTok{ rng\_bc),}
+\NormalTok{    lambda))}
+\NormalTok{p }\OtherTok{\textless{}{-}}\NormalTok{ p }\SpecialCharTok{+} \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{breaks =} \FunctionTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{1}\NormalTok{, }\AttributeTok{length.out =} \DecValTok{5}\NormalTok{),}
+    \AttributeTok{labels =}\NormalTok{ labs) }\SpecialCharTok{+} \FunctionTok{ylab}\NormalTok{(}\StringTok{"SCL (rescaled from Box{-}Cox)"}\NormalTok{)}
 \NormalTok{p}
@@
 
-Note that the observed individual trajectories show very high
-variability within classes.
-
-\hypertarget{r-session}{%
-\subsection{R session}\label{r-session}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{sessionInfo}\NormalTok{()}
-\CommentTok{\#\textgreater{} R version 4.1.3 (2022{-}03{-}10)}
-\CommentTok{\#\textgreater{} Platform: x86\_64{-}w64{-}mingw32/x64 (64{-}bit)}
-\CommentTok{\#\textgreater{} Running under: Windows 10 x64 (build 19045)}
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} Matrix products: default}
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} locale:}
-\CommentTok{\#\textgreater{} [1] LC\_COLLATE=English\_United Kingdom.1252 }
-\CommentTok{\#\textgreater{} [2] LC\_CTYPE=English\_United Kingdom.1252   }
-\CommentTok{\#\textgreater{} [3] LC\_MONETARY=English\_United Kingdom.1252}
-\CommentTok{\#\textgreater{} [4] LC\_NUMERIC=C                           }
-\CommentTok{\#\textgreater{} [5] LC\_TIME=English\_United Kingdom.1252    }
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} attached base packages:}
-\CommentTok{\#\textgreater{} [1] stats     graphics  grDevices utils     datasets  methods   base     }
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} other attached packages:}
-\CommentTok{\#\textgreater{} [1] MASS\_7.3{-}58.2    ggplot2\_3.4.0    tidySEM\_0.2.4.6  OpenMx\_2.21.1   }
-\CommentTok{\#\textgreater{} [5] scales\_1.2.1     yaml\_2.3.6       papaja\_0.1.1     tinylabels\_0.2.3}
-\CommentTok{\#\textgreater{} }
-\CommentTok{\#\textgreater{} loaded via a namespace (and not attached):}
-\CommentTok{\#\textgreater{}   [1] TH.data\_1.1{-}1         colorspace\_2.1{-}0      estimability\_1.4.1   }
-\CommentTok{\#\textgreater{}   [4] parameters\_0.20.1     rstudioapi\_0.14       listenv\_0.9.0        }
-\CommentTok{\#\textgreater{}   [7] lavaan\_0.6{-}13         rstan\_2.26.13         fansi\_1.0.4          }
-\CommentTok{\#\textgreater{}  [10] mvtnorm\_1.1{-}3         codetools\_0.2{-}18      splines\_4.1.3        }
-\CommentTok{\#\textgreater{}  [13] mnormt\_2.1.1          knitr\_1.41            texreg\_1.38.6        }
-\CommentTok{\#\textgreater{}  [16] bayesplot\_1.10.0      jsonlite\_1.8.4        effectsize\_0.8.2     }
-\CommentTok{\#\textgreater{}  [19] compiler\_4.1.3        httr\_1.4.4            emmeans\_1.8.4{-}1      }
-\CommentTok{\#\textgreater{}  [22] backports\_1.4.1       assertthat\_0.2.1      Matrix\_1.5{-}3         }
-\CommentTok{\#\textgreater{}  [25] fastmap\_1.1.0         cli\_3.6.0             htmltools\_0.5.4      }
-\CommentTok{\#\textgreater{}  [28] prettyunits\_1.1.1     tools\_4.1.3           igraph\_1.3.5         }
-\CommentTok{\#\textgreater{}  [31] coda\_0.19{-}4           gtable\_0.3.1          glue\_1.6.2           }
-\CommentTok{\#\textgreater{}  [34] RANN\_2.6.1            dplyr\_1.0.10          V8\_4.2.2             }
-\CommentTok{\#\textgreater{}  [37] Rcpp\_1.0.10           carData\_3.0{-}5         vctrs\_0.5.2          }
-\CommentTok{\#\textgreater{}  [40] nlme\_3.1{-}161          psych\_2.2.9           insight\_0.18.8       }
-\CommentTok{\#\textgreater{}  [43] xfun\_0.36             stringr\_1.5.0         globals\_0.16.2       }
-\CommentTok{\#\textgreater{}  [46] ps\_1.7.2              proto\_1.0.0           CompQuadForm\_1.4.3   }
-\CommentTok{\#\textgreater{}  [49] lifecycle\_1.0.3       future\_1.30.0         zoo\_1.8{-}11           }
-\CommentTok{\#\textgreater{}  [52] parallel\_4.1.3        sandwich\_3.0{-}2        inline\_0.3.19        }
-\CommentTok{\#\textgreater{}  [55] curl\_5.0.0            gridExtra\_2.3         pander\_0.6.5         }
-\CommentTok{\#\textgreater{}  [58] blavaan\_0.4{-}3         loo\_2.5.1             StanHeaders\_2.26.13  }
-\CommentTok{\#\textgreater{}  [61] stringi\_1.7.12        bayestestR\_0.13.0     fastDummies\_1.6.3    }
-\CommentTok{\#\textgreater{}  [64] checkmate\_2.1.0       boot\_1.3{-}28           pkgbuild\_1.4.0       }
-\CommentTok{\#\textgreater{}  [67] rlang\_1.0.6           pkgconfig\_2.0.3       matrixStats\_0.63.0   }
-\CommentTok{\#\textgreater{}  [70] evaluate\_0.20         lattice\_0.20{-}45       rstantools\_2.2.0     }
-\CommentTok{\#\textgreater{}  [73] processx\_3.8.0        tidyselect\_1.2.0      parallelly\_1.34.0    }
-\CommentTok{\#\textgreater{}  [76] plyr\_1.8.8            magrittr\_2.0.3        bookdown\_0.32        }
-\CommentTok{\#\textgreater{}  [79] R6\_2.5.1              generics\_0.1.3        multcomp\_1.4{-}20      }
-\CommentTok{\#\textgreater{}  [82] DBI\_1.1.3             withr\_2.5.0           gsubfn\_0.7           }
-\CommentTok{\#\textgreater{}  [85] pillar\_1.8.1          survival\_3.5{-}0        datawizard\_0.6.5     }
-\CommentTok{\#\textgreater{}  [88] abind\_1.4{-}5           tibble\_3.1.8          future.apply\_1.10.0  }
-\CommentTok{\#\textgreater{}  [91] crayon\_1.5.2          car\_3.1{-}1             nonnest2\_0.5{-}5       }
-\CommentTok{\#\textgreater{}  [94] utf8\_1.2.2            tmvnsim\_1.0{-}2         MplusAutomation\_1.1.0}
-\CommentTok{\#\textgreater{}  [97] rmarkdown\_2.20        grid\_4.1.3            data.table\_1.14.6    }
-\CommentTok{\#\textgreater{} [100] pbivnorm\_0.6.0        bain\_0.2.8            callr\_3.7.3          }
-\CommentTok{\#\textgreater{} [103] digest\_0.6.31         xtable\_1.8{-}4          dbscan\_1.1{-}11        }
-\CommentTok{\#\textgreater{} [106] RcppParallel\_5.1.6    stats4\_4.1.3          munsell\_0.5.0}
-\end{Highlighting}
-\end{Shaded}
+Note that the observed individual trajectories show very high variability within classes.
 

### 2023-01-24 manuscript.Rmd

--- 2023-01-22 manuscript.Rmd
@@
 Furthermore, meeting the increasing demand for open source research software,
-this paper introduces new functionality in the R-package `tidySEM` [@vanlissaTidySEMTidyStructural2022] to perform LCA using free open source software (FOSS).
+this paper introduces new functionality in the R-package `tidySEM` [@van_lissa_tidysem_2022] to perform LCA using free open source software (FOSS).
 
@@
 most existing software is commercial and closed-source,
-for example, Mplus [@muthenMplusUserGuide1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquistMplusAutomationPackageFacilitating2018], and LatentGOLD [REF].
+for example, Mplus [@muthen_mplus_1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquist_mplusautomation_2018], and LatentGOLD [REF].
 <!-- software that implements these models. -->
@@
 Another downside is that the proprietary source code cannot be audited, debugged, or enhanced by third parties.
-This goes against open science principles [@lamprechtFAIRPrinciplesResearch2019],
+This goes against open science principles [@lamprecht_towards_2020],
 incurs a risk that mistakes in the source code go unnoticed,
@@
 It was developed with open science in mind,
-and adheres to the FAIR software principles [Findable, Accessible, Interoperable, and Reusable, @lamprechtFAIRPrinciplesResearch2019],
+and adheres to the FAIR software principles [Findable, Accessible, Interoperable, and Reusable, @lamprecht_towards_2020],
 and the [OpenSSF Best Practices](https://bestpractices.coreinfrastructure.org/en/criteria) for FOSS software.
@@
 groups based on a parametric model of observed indicators of group
-membership [@vermuntLatentClassAnalysis2004].
+membership [@vermunt_latent_2004].
 It has become popular across scientific fields,
@@
 
-Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling [@molenaarArbitraryNatureLatent1994].
+Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling [@molenaar_arbitrary_1994].
 A mixture model is like confirmatory factor analysis,
@@
 LCA seeks to describe unobserved groups and classify individual observations to those groups.
-<!-- In factor analysis, the latent variable is continuous, so the relationships -->
-<!-- between indicators are defined by their communalities. -->
-<!-- Here, the -->
-<!-- latent variable represents the error corrected approximation of the -->
-<!-- sample. In LCA, by contrast, the relation between indicators is defined by how -->
-<!-- well they separate the classes, as a multinomial regression between the -->
-<!-- indicators and the latent variable [@masyn_latent_2013]. -->
 In line with this distinction, LCA
@@
 For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
-LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckxDevelopmentalTypologiesIdentity2008].
-Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA [@donnellanResilientOvercontrolledUndercontrolled2010].
+LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckx_developmental_2008].
+Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA [@donnellan_resilient_2010].
 
@@
 Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughmanMixtureModelFramework2007].
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007].
 These models estimate one class to explain the "true zeros",
@@
 Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
-information without imputing missing values [@endersRelativePerformanceFull2001].
+information without imputing missing values [@enders_relative_2001].
 FIML is a best-practice solution for handling missing data;
@@
 One minor concern is that `tidySEM` uses a different algorithm to determine starting values for complete data (K-means) and for missing data (hierarchical clustering).
-This should not be a problem, however, as the global optimizer used to estimate the model should be independent of starting values [@coranaMinimizingMultimodalFunctions1987].
+This should not be a problem, however, as the global optimizer used to estimate the model should be independent of starting values [@corana_minimizing_1987].
 A second concern is that FIML is not suitable for cases with complete missing data.
@@
 
-Prior literature emphasized the importance of considering alternative model specifications [@vandeschootGRoLTSChecklistGuidelinesReporting2017].
+Prior literature emphasized the importance of considering alternative model specifications [@van_de_schoot_grolts-checklist_2017].
 What has remained underemphasized is that alternative model specifications should be approached differently for exploratory versus confirmatory analyses (see Use Cases).
@@
 this provides evidence against the theory.
-If fit differences are small, the theoretical model may still be preferred, as theory is a relevant consideration in model selection [@jungIntroductionLatentClass2008].
+If fit differences are small, the theoretical model may still be preferred, as theory is a relevant consideration in model selection [@jung_introduction_2008].
 The fact that the theoretical model received little differential support should of course be discussed.
@@
 
-A second consideration in model specification is the risk of overfitting [@hastieElementsStatisticalLearning2009].
+A second consideration in model specification is the risk of overfitting [@hastie_elements_2009].
 Overfitting occurs when a model has many parameters relative to the number of observations,
@@
 In Bayesian mixture modeling,
-it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased [@rousseauAsymptoticBehaviourPosterior2011].
+it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased [@rousseau_asymptotic_2011].
 Although it is not known whether frequentist LCA exhibits the same behavior,
@@
 there is a risk that the global optimum is never found.
-An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing [SA, @coranaMinimizingMultimodalFunctions1987].
+An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing [SA, @corana_minimizing_1987].
 SA does not behave quite like the marbles of the random starts approach;
@@
 For example, if we assume that the different classes are likely to have different mean values on the indicators, 
-then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernackiChoosingStartingValues2003].
+then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
 By default, `tidySEM` derives starting values using K-means clustering [REF],
@@
 <!-- it is important to consider alternative model specifications. -->
-<!-- One way to obtain relative evidence in favor of different models is by using BIC weights, obtained by the function `tidySEM::ic_weights()` [@wagenmakersAICModelSelection2004]. -->
-
-<!-- From sinhaPractitionerGuideLatent2021: -->
+<!-- One way to obtain relative evidence in favor of different models is by using BIC weights, obtained by the function `tidySEM::ic_weights()` [@wagenmakers_aic_2004]. -->
+
+<!-- From sinha_practitioners_2021: -->
 <!-- Testing for number of classes.
@@
 In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
-the proportion of cases assigned to the excessive classes converges to zero as the sample size increases [@rousseauAsymptoticBehaviourPosterior2011].
+the proportion of cases assigned to the excessive classes converges to zero as the sample size increases [@rousseau_asymptotic_2011].
 It is not known whether this also happens in frequentist LCA;
@@
 Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016],
-which is implemented in `tidyLPA` [@rosenbergTidyLPAPackageEasily2018].
+which is implemented in `tidyLPA` [@rosenberg_tidylpa_2018].
 Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs [@nylund-gibson_ten_2018].
@@
 it is also possible to calculate so-called "IC-weights",
-which indicate the relative support the data provide for each model in the set [@wagenmakersAICModelSelection2004].
+which indicate the relative support the data provide for each model in the set [@wagenmakers_aic_2004].
 This is particularly useful for confirmatory LCA,
@@
 returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
-Other authors have reported similar experiences [@sinhaPractitionerGuideLatent2021].
+Other authors have reported similar experiences [@sinha_practitioners_2021].
 This suggests that, like existing ICs,
@@
 proceed with caution:
-The model might not be locally identified in that class [@depaoliMixtureClassRecovery2013].
+The model might not be locally identified in that class [@depaoli_mixture_2013].
 Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
@@
 sharing data and analysis code required to exactly reproduce analyses,
-and pre-registering study plans prior to data collection or analysis [@nosekTransparencyOpennessPromotion2016a].
+and pre-registering study plans prior to data collection or analysis [@nosek_transparency_2016].
 A key advantage of using open source software for LCA is that, compared to commercial software, it enables much greater adherence to open science principles.
@@
 The Workflow for Open Reproducible Code in Science describes how to create a fully reproducible research archive,
-and the `worcs` R-package automates many of the steps involved [@vanlissaWORCSWorkflowOpen2021a].
+and the `worcs` R-package automates many of the steps involved [@van_lissa_worcs_2021].
 <!-- brings down the learning curve by implementing best practices for creating a reproducible digital research repository. 
@@
 More interestingly, it also enables creating a *Preregistration As Code* (PAC):
-A draft paper with reproducible analysis section based on fake (synthetic or simulated) data [@peikertReproducibleResearchTutorial2021].
+A draft paper with reproducible analysis section based on fake (synthetic or simulated) data [@peikert_reproducible_2021].
 When the real data are accessed or collected,
@@
 PACs are quite unambiguous; the selection of the final model can even be automated by hard-coding the criteria into the analysis section.
-For an example of PAC, see @vanlissaComplementingPreregisteredConfirmatory2022.
+For an example of PAC, see @van_lissa_complementing_2022.
 
@@
 Finally, the range of proportions of cases assigned to each class is important;
-the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant [see @depaoliMixtureClassRecovery2013].
+the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant [see @depaoli_mixture_2013].
 By default, `table_fit()` reports the minimum and maximum proportion of cases,
@@
 Specifically, they can be used to exclude solutions that do not meet predefined criteria from consideration.
-Fourth, previous reporting standards called for reporting of sample descriptive statistics [see @schreiberLatentClassAnalysis2017].
+Fourth, previous reporting standards called for reporting of sample descriptive statistics [see @schreiber_latent_2017].
 While we agree that this is due dilligence,

### 2023-01-25 manuscript.Rmd
edits from Mauricio (with source editor)

@@
 and complicates manuscript review and assessment of the quality of published studies.
-The present paper seeks to address these challenges by introducing updated guidelines for estimatin and reporting LCA,
+The present paper seeks to address these challenges by introducing updated guidelines for estimating and reporting LCA,
 based on current best practice. 
@@
 most existing software is commercial and closed-source,
-for example, Mplus [@muthen_mplus_1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquist_mplusautomation_2018], and LatentGOLD [REF].
+for example, Mplus [@muthen_mplus_1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquist_mplusautomation_2018], and LatentGOLD [@LG_manual].
 <!-- software that implements these models. -->
@@
 
-Notable FOSS for LCA includes Mclust [REF], depmixS4 [REF], and OpenMx [@neale_openmx_2016].
+Notable FOSS for LCA includes mclust [@scrucca_mclust_2016], depmixS4 [@depmixS4], and OpenMx [@neale_openmx_2016].
 A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
@@
 When the focus is on each individual's estimated class membership,
-LCA can be thought of as a type of clustering algorithm that corrects for (random) measurement error.
+LCA can be thought of as a type of clustering algorithm that corrects for random measurement error by default, and is able to correct for systematic measurement error if it is included in the model.
 Specifically, the posterior class probability that an individual belongs to a latent class
@@
 whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements [@alkema_maximizing_2022].
+
+A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
+called Latent Transition Analysis), which estimates the change in class
+membership over time with the markov assumption over time. The markov
+assumption states that the class membership at time $t$ is only affected
+by the class membership at time $t-1$. The structural part of the model
+is comprise of the initial states (proportion of the sample in each
+class at the first time point), and transition probabilities
+(probabilities to switch class membership across time). The transition
+probabilities are estimated as multinomial logistic regressions between
+class memberships [@vermunt_longitudinal_2010].
 
@@
 and not be "reified" - treated as evidence of the existence of a categorical latent variable.
+These classes can be theoretically justified, and use as part of the theory building process, but they are not sufficient.
 
@@
 Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007].
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007; magnus_multidimensional_2021].
 These models estimate one class to explain the "true zeros",
-and another class with a different distribution to explain all oter values, which can include zero.
+and another class with a different distribution to explain all other values, which can include zero.
 Note that this LCA has different models for both classes,
@@
 argued that some parametric methods are robust when scales with 7+
-response categories are treated as continuous [e.g., @norman_likert_2010],
+response categories are treated as continuous [e.g., @norman_likert_2010; rhemtulla_when_2012],
 it is very unlikely
@@
 Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
-information without imputing missing values [@enders_relative_2001].
+information without imputing missing values, by estimating the model based on all
+available information for each subject [@enders_relative_2001; enders_applied_2022].
 FIML is a best-practice solution for handling missing data;
@@
 and conducting them on several imputed datasets multiplies the load.
-Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population.
+Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population, so the imputation model could bias the results as it does not account for the expected heterogeneity.
 
@@
 Our recommendation is to inspect and report the proportion of missingness per variable (e.g., using `tidySEM::descriptives()`),
-and report the `mice::mcar()` test.
+and report the `mice::mcar()` test, as way to identify variables that can help the parameter recovery instead of evidence of MCAR or MNAR.
 As FIML estimation assumes that missingness is either MCAR or MAR,
@@
 but with respect to the latent growth variables.
-Ordinal indicators can be parametrized differently;
+Ordinal indicators are parametrized differently;
 the default approach in `tidySEM` assumes that each ordinal indicator reflects an underlying standard normal distribution, $N(\mu = 0, \sigma = 1)$.
@@
 Information criteria are computed from the $-2LL$,
-which is the most basic fit measure.
+which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
 Information criteria add a penalty for model complexity to the $-2LL$, and thus balance fit and complexity.
@@
 Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
+Other approach is to select several competing model based on the ICs, and evaluate the theoretical interpretation.
 Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016],
@@
 returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
-Other authors have reported similar experiences [@sinha_practitioners_2021].
+Other authors have reported similar experiences [@sinha_practitioners_2021], which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting [@mcelreath_statistical_2020].
 This suggests that, like existing ICs,
@@
 $k$ subsets of observations with most likely class of $M = 1 \ldots k$.
-The diagonal represents the probability that observations in each class will be correctly classified.
-<!-- Mauricio: Can you fact-check this?!? I really struglle here -->
+The diagonal represents the average predictive probability of a the subjects related to their most likely class, indicating the certainty of class assignment for the subjects.
+Some guidelines have being suggested for this diagonal, like for them to be higher than 0.7 [@masyn2013], but we suggest researchers to be cautions about these guidelines as they cannot be generalized and should not be use as strict cutoffs as evidence "good" or "bad" classification.
 
@@
 and off-diagonal elements can be conceptualized as measurement error.
-<!-- Mauricio: Can you fact-check this?!? I really struglle here -->
+<!-- Mauricio: Can you fact-check this?!? I really struglle here. MGV: this looks good -->
 The minimum and maximum of the diagonal of this table are given by default as part of the output of `table_fit()` for LCA analyses (`prob_min` and `prob_max`).
@@
     + LR tests
+    + Theoretical interpretability of the solution
 6. Justify the (preregistered) criteria used to eliminate models from consideration, including:

### 2023-01-31 appendix_c.Rmd

@@
 because the analyses are computationally very intensive and the data were simulated from a 3-class model.
+
+It is important to highlight that in LCGA, the subgroups will be limited by
+the specify growth structure, meaning that LCA will identify distinctive
+growth patterns within the intercept, step, and slope growth. For
+example, if there is a subgroup that follows a quadratic growth pattern
+this models will not be able to identify it.
 
@@
 <!-- ::: -->
+
+## R session
+
+```{r echo = TRUE, eval=TRUE}
+sessionInfo()
+```

### 2023-02-16 manuscript.Rmd
Update pt. 1
Updated manuscript and cleaned references

@@
 title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
-shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
+shorttitle        : "BEST PRACTICES IN LATENT CLASS ANALYSIS"
 author: 
@@
   describes best practices for the estimation and reporting of LCA,
-  using free and open source software in R. To this end, this tutorial
-  introduces new functionality for estimating and reporting mixture
-  models in the `tidySEM` R-package whose backend relies on `OpenMx`.
+  using free and open source software in R. To this end, 
+  we introduce new functionality for estimating and reporting mixture models
+  using the `tidySEM` R-package.
+
 keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
@@
 The present paper seeks to address these challenges by introducing updated guidelines for estimating and reporting LCA,
-based on current best practice. 
+based on current best practices. 
 Furthermore, meeting the increasing demand for open source research software,
@@
 most existing software is commercial and closed-source,
-for example, Mplus [@muthen_mplus_1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquist_mplusautomation_2018], and LatentGOLD [@LG_manual].
+for example, *Mplus* [@muthen_mplus_1998], for which the R-package `MplusAutomation` offers augmented LCA functionality [@hallquist_mplusautomation_2018], and *Latent GOLD* [@vermunt_latent_2000].
 <!-- software that implements these models. -->
 <!-- One important limitation is that the most fully-features -->
-A downside of commercial software is that its use is restricted to individuals and institutions who can afford a license.
-Another downside is that the proprietary source code cannot be audited, debugged, or enhanced by third parties.
+A downside of commercial software is that its use is restricted to individuals and institutions that can afford a license.
+Another downside is that proprietary source code cannot be audited, debugged, or enhanced by third parties.
 This goes against open science principles [@lamprecht_towards_2020],
-incurs a risk that mistakes in the source code go unnoticed,
+incurs a risk that mistakes in the source code may go unnoticed,
 and curbs progress as independent developers cannot fix bugs or add functionality.
 
-There are several compelling reasons to prefer free open source software (FOSS) instead.
+There are several compelling reasons to prefer free open source software instead.
 FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
@@
 
-Notable FOSS for LCA includes mclust [@scrucca_mclust_2016], depmixS4 [@depmixS4], and OpenMx [@neale_openmx_2016].
-A crucial limitation of these existing packages is that some have limited functionality (e.g., Mclust),
-or implement specific innovations (e.g., depmixS4).
-OpenMx is unique in that, like Mplus, it implement LCA in the context of a fully featured structural equation modeling framework.
-Remaining obstacles to the use of `OpenMx` is that it is not very user-friendly and its functionality for LCA is poorly documented.
+Notable FOSS for LCA includes `mclust` [@scrucca_mclust_2016], `depmixS4` [@visser_depmixs4_2010], and `OpenMx` [@neale_openmx_2016].
+A crucial limitation of these existing packages is that some have limited functionality (e.g., `mclust`),
+or implement specific innovations (e.g., `depmixS4`).
+`OpenMx` is unique in that, like proprietary *Mplus*, it implements LCA in the context of a fully featured structural equation modeling framework.
+A major downside of `OpenMx` is that it is not very user-friendly and its functionality is poorly documented.
 
 These limitations are addressed by the `tidySEM` package,
-which provide a user-friendly interface for LCA in `OpenMx`.
-The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting LCA, as described in this paper.
+which provides a user-friendly interface for LCA in `OpenMx`.
+The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting of LCA, as described in this paper.
 The primary purpose of the `tidySEM` package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
@@
 The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
-but the resulting models can still be fully customized (low barrier, high ceiling).
-Specifically, LCA models constructed in `tidySEM` inherit from `OpenMx`' format `MxModel`, and remain compatible with downstream functions that process such models.
-Being FOSS, `tidySEM` facilitates fully-reproducible analyses and contributes to open science.
-It was developed with open science in mind,
-and adheres to the FAIR software principles [Findable, Accessible, Interoperable, and Reusable, @lamprecht_towards_2020],
-and the [OpenSSF Best Practices](https://bestpractices.coreinfrastructure.org/en/criteria) for FOSS software.
+but the resulting models can still be fully customized making it a low threshold - high ceiling package.
+For instance, LCA models constructed in `tidySEM` inherit the `MxModel` format from `OpenMx` which renders them compatible with various legacy functions from `OpenMx` and related LCA ecosystem .
+Since `tidySEM` was developed with open science in mind, it facilitates fully-reproducible analyses and adheres to the *FAIR (Findable, Accessible, Interoperable, and Reusable) software principles* [@lamprecht_towards_2020],
+and the [*OpenSSF Best Practices*](https://bestpractices.coreinfrastructure.org/en/criteria) for FOSS software.
 
@@
 under a number of different names;
-most notably (finite Gaussian) mixture modeling and latent profile analysis.
+most notably (*finite Gaussian*) *mixture modeling* and *latent profile analysis*.
 The concept of LCA can be understood in different ways.
-Generally speaking, LCA assumes that the study population is composed of $K$
+Generally speaking, LCA assumes that the study population is composed of $k$
 subpopulations.
 It further assumes that the observed data are
-a mixture of data originating from those subpopulations;
-hence the name mixture model.
+a mixture of observations originating from those subpopulations;
+hence the name *mixture model*.
 Consider the simplest possible
-univariate "model", which is a normal distribution.
-This model has
-two parameters: the mean and the variance.
-LCA aims to estimate the values of those parameters across a known number of classes $K$,
-as well as the probability of every individual $i$ belonging to classes $1 \ldots k$.
-Commonly, the same model is estimated across all classes,
-but with different parameters for each class (i.e., class-specific means and variances).
+univariate model, which is a normal distribution.
+This model has two parameters: the mean and the variance.
+LCA aims to estimate the values of those parameters across $k$ classes,
+as well as estimate the probability of every individual observation $i$ belonging to each of the $1 \ldots k$ classes.
+Commonly, the same model is fitted to all $k$ classes,
+but each class is characterized by a unique set of parameters for that model (i.e., class-specific means and variances).
 
@@
 When estimating a two-class mixture model, the detective observes that
-the model estimates the mean shoe size of the two groups are equal to
+the model estimates the mean shoe size of the two groups as equal to
 `r report(estparam[1], equals = FALSE)` and
@@
 
-Another way to conceptualize LCA is by analogy to a measurement model in structural equation modeling [@molenaar_arbitrary_1994].
+Another way to conceptualize LCA is through an analogy to a measurement model in structural equation modeling [@molenaar_arbitrary_1994].
 A mixture model is like confirmatory factor analysis,
-except with a categorical instead of a continuous latent variable.
-One difference between
-the two techniques is that factor analysis can be considered as a way to
-group observed *variables* into latent constructs, and LCA groups
-*individuals* into classes.
+except for the fact that the latent variable is categorical instead of continuous.
+Another difference between the two techniques is that factor analysis can group observed *variables* into latent constructs, while LCA can group *observations* into classes.
 While factor analysis seeks to describe the relations between variables,
 LCA seeks to describe unobserved groups and classify individual observations to those groups.
-In line with this distinction, LCA
-(mixture models) is referred to as a "person-centered" technique, and
+In line with this distinction, LCA is referred to as a "person-centered" technique, and
 factor analysis as a "variable-centered" technique
@@
 of all parameters in all groups. The relative contribution of each case
-to the parameters of each group is determined by that case's posterior
-probability of belonging to that group.
+to the parameters of each group is determined by that case's *posterior
+probability* of belonging to that group.
 
@@
 Specifically, the posterior class probability that an individual belongs to a latent class
-can be computed from the likelihood of that individual's observed data under given the class-specific model.
+can be computed from the likelihood of that individual's observed data given the class-specific model.
 These posterior class probabilities can be used to weight follow-up analyses,
-or individuals can be assigned to classes based on their highest class probability.
+or to assign individuals to classes based on their highest class probability.
 If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
@@
 model used, or the level of measurement of the indicators [@vermunt_latent_2004].
-To prevent ambiguity, we refer to LCA with binary or ordinal indicators as *LCA with ordinal indicators*.
+To prevent ambiguity, we refer to LCA with binary or ordinal indicators as *LCA with ordinal indicators*,
+and the sole *LCA* expression as an umbrella term.
 Given the abundance of redundant terms in use for closely related types of LCA models,
@@
 Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
-A class of larger particles with a low correlation between length and width,
+a class of larger particles with a low correlation between length and width,
 and a class of smaller particles with a high correlation between length and width.
@@
 
-A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
-called Latent Transition Analysis), which estimates the change in class
-membership over time with the markov assumption over time. The markov
-assumption states that the class membership at time $t$ is only affected
+A longitudinal extension of LCA is the *Hidden Markov Model* (*HMM*, also
+called *Latent Transition Analysis*), which estimates the change in class
+membership over time under the Markov assumption. The Markov
+assumption states that class membership at time $t$ is only affected
 by the class membership at time $t-1$. The structural part of the model
-is comprise of the initial states (proportion of the sample in each
+is comprised of the initial states (proportion of the sample in each
 class at the first time point), and transition probabilities
-(probabilities to switch class membership across time). The transition
+(probabilities of class membership switching across time). The transition
 probabilities are estimated as multinomial logistic regressions between
@@
 The indicators of class membership in a GMM are the 
-intercepts, and variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
-For linear trajectories, these latent variables are an intercept and slope.
-The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes homogeneity of trajectories within classes.
+intercepts, variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
+For linear trajectories, these latent variables are the intercept and slope.
+The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes within-class homogeneity of trajectories.
 This is achieved by restricting the (co)variances of the growth parameters to zero.
@@
 
-There are several use cases for which LCA methods are suitable.
+There are multiple use cases for which LCA methods are suitable. We present their brief overview.
 
@@
 it can be used to similar ends:
-To assess whether a theoretical measurement model holds.
+to assess whether a theoretical measurement model holds.
 This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
 For example, *identity status theory* posits that, at any given point in time, adolescents reside in one of four identity statuses [@marcia_development_1966].
-LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment) [@luyckx_developmental_2008].
-Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient type, which could also be restored using LCA [@donnellan_resilient_2010].
+LCA can be used to identify these four statuses based on observed indicators [e.g., self-reported identity exploration and commitment, @luyckx_developmental_2008].
+Similarly, *personality type theory* states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient personality types, which could also be restored using LCA [@donnellan_resilient_2010].
 
@@
 
-Appendix A is a tutorial to confirmatory LCA in `tidySEM`,
-based on a recent study that tested the theory that identities of Belgian adolescents with migration backgrounds could be summarized as those relating to their heritage, national and regional identity [@maene_perceived_2022].
+Appendix A showcases a step-by-step tutorial to confirmatory LCA in `tidySEM`,
+based on a recent study that tested a theory that identities of Belgian adolescents with migration backgrounds could be summarized as those relating to their heritage, national and regional identities [@maene_perceived_2022].
 
@@
 
-Another use case is to explore whether a population is comprised of latent subpopulations, and if so, how many.
+Another use case of LCA is to explore whether a population is comprised of latent subpopulations, and if so, how many.
 The primary focus in this use case is on class enumeration.
@@
 one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
-and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and information about prior attempted methods and severity
+and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and prior attempt methods and severity
 [@gromatsky_characteristics_2022].
-Knowing that the population of suicidal service members is heterogeneous can inform future research, policy, and care provided.
+Knowing that the population of suicidal service members is heterogeneous can inform future research, policy, and care.
 Appendix B offers a tutorial for exploratory LCA,
@@
 A mixture model was used to estimate employment status, correcting for measurement error, from both self-report and registry data.
-The study investigated random and systematic measurement error in both indicators,
-which reveal strengths and weaknesses of both types of assessment.
-
-This example also illustrates that LCA allows us to evaluate the reliability of different indicators [@geiser_is_2014].
+The study investigated random and systematic measurement error in the indicatorss stemming from the two registers,
+which revealed strengths and weaknesses of both types of assessment.
+
+This example also illustrates that LCA allows us to evaluate the reliability of different types of indicators [@geiser_is_2014].
 <!-- High quality indicators will be strongly related to the -->
 <!-- latent variable and will lead to good class separation. -->
-A continuous indicator that discriminates well between classes will show large differences in class-specific means;
-an ordinal indicator will show very high or very low conditional response probabilities for some classes but not for others.
+As a rule, a continuous indicator that discriminates well between classes will show large differences in class-specific means.
+An ordinal indicator with the same property will show very high or very low conditional response probabilities for some classes but not for others.
 This information can be used to select the indicators that are most diagnostic of class membership.
 For an example of LCA with a focus on differential item functioning,
-consider a study on university admission tests of Saudi Arabian students [@tsaousis_measurement_2020].
+see a study on university admission tests of Saudi Arabian students [@tsaousis_measurement_2020].
 <!-- LCA revealed three -->
@@
 whereas LCA can accommodate complex - but discrete - patterns.
-For example, imagine that there exist two indicators - $x$ and $y$, and some individuals score high on both, some score low on both, and some score high on $x$ but low on $y$.
-Factor analysis would only be able to capture the differences between the first two groups well; LCA captures differences between all three.
-Note that this is a pragmatic use of LCA;
-there is no theory about the existence of a categorical latent variable (cf. Testing theory).
-Consequently, the class solution should be treated as descriptive of the sample,
-and not be "reified" - treated as evidence of the existence of a categorical latent variable.
-These classes can be theoretically justified, and use as part of the theory building process, but they are not sufficient.
+For example, imagine that there exist two indicators - $x$ and $y$. Some individuals score high on both indicators, some score low on both, and other score high on $x$ but low on $y$.
+Factor analysis would only be able to capture the differences between the first two groups well, but LCA will capture differences between all three classes.
+Note, however, that this is a pragmatic use of LCA.
+There is no theory about the existence of a categorical latent variable (cf.
+Testing theory).
+Consequently, the final class solution should be treated as descriptive of the sample.
+It should not be reified or treated as evidence of the existence of a categorical latent variable without accumulation of additional empirical evidence and deliberation on the findings.
+The discovered classes could indeed prove to be theoretically justified and thus inform the theory building process, but a single exploratory fining is never sufficient for reaching such conclusions.
 
@@
 and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
-For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy and found three groups: high, average, and low empathy [@van_lissa_divergence_2015].
-These groups were then used as moderator of a latent growth analysis of adolescent- and parent-reported conflict.
-Results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than the other groups,
-and that the low empathy group had more conflict with parents than the other groups.
-Note that both these key findings are non-linear;
-modeling them in a single-group model would be difficult.
+For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy identifying: high, average, and low empathy classes [@van_lissa_divergence_2015].
+The membership to these three groups was then used as a moderator in a latent growth analysis of adolescent- and parent-reported conflict.
+The results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than other groups,
+and the low empathy group had more conflict with parents than other groups.
+Note that as both of these effects were non-linear,
+modeling them in a single-group framework would introduce unnecessary bias.
 LCGA greatly simplifies model specification and interpretation,
-albeit at a cost of some loss of information and, potentially, generalizability.
+albeit at a cost of some loss of information, and potentially generalizability.
 Appendix C is a tutorial on LCGA,
@@
 dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
-However, LCA was able to reduce response patterns to five clinically meaningful latent classes.
+However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data.
 
@@
 
-Another use case is to estimate individuals' unobserved class membership based on observed indicators.
+Another use case of LCA is to estimate individuals' unobserved class membership based on observed indicators.
 The shoe size example in Figure \@ref(fig:shoedens)
-is a rudimentary illustration of this application.
+is a rudimentary illustration of such application.
 In applied research, LCA is often used for classification in clinical contexts.
 For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough [@baughman_mixture_2006].
-Another study used LCA to not only classify the sample used to estimate the model, but also future cases [REF Zegwaard et al, in prep].
+Another study used LCA not only to classify the sample used to estimate the model, but also predict future cases [REF Zegwaard et al, in prep].
 The study identified four types of care providers among those who supported close kin with mental health problems.
-Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers, to provide more targeted support.
+Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers and pave the way to providing them more targeted support.
 To obtain predicted class membership based on an existing LCA model,
 the authors computed the likelihood of data from a new participant under the multivariate normal distributions estimated within all latent classes.
-<!-- DA: this study is not in zotero yet, I lack information (title, authors, etc.) to enter it, @Caspar could you enter this to Zotero and adjust the in-text citation-->
+<!-- DA: this study is not in zotero yet, I lack information (title, authors, etc.) to enter it, @Caspar could you share more info (i.e.
+the working title, authors, goal journal) enter this to Zotero and adjust the in-text citation-->
 These applications of LCA aid the clinical decision making process,
@@
 <!-- In fact, LCA assumes the population -->
-<!-- distribution is a non-normal mixture of $K$ normal distributions, and it -->
-<!-- can recover the value of $K$, i.e. generate a $K$-class solution. -->
+<!-- distribution is a non-normal mixture of $k$ normal distributions, and it -->
+<!-- can recover the value of $k$, i.e. generate a $k$-class solution. -->
 <!-- CJ: This example is not perfect, we need another example still.. -->
@@
 then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
-By default, `tidySEM` derives starting values using K-means clustering [REF],
+By default, `tidySEM` derives starting values using K-means clustering [@likas_global_2003],
 computing the mixture model starting values by treating the K-means solution as multi-group model.
@@
 It is approached differently for exploratory versus confirmatory LCA.
-In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes $K$. 
-The number $K$ may be chosen a-priori or dictated by the data.
-It is sensible to limit $K$ to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
+In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes $k$. 
+The number $k$ may be chosen a-priori or dictated by the data.
+It is sensible to limit $k$ to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
 Above a certain number of classes,
@@
 It is not known whether this also happens in frequentist LCA;
-nonetheless, near-zero class memberships should be taken into account when determining the maximum $K$.
+nonetheless, near-zero class memberships should be taken into account when determining the maximum $k$.
 
@@
 $\ln(\frac{n + 2}{24})$.
-Although a simulation study showed that the saBIC performed well for class enumeration,
-its penalty is not based on theory [REF].
+Although simulation studies have showed that the saBIC performed well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
+its penalty is not based on theory.
 BIC also performs well and is theoretically substantiated;
@@
 One challenge is that the LR test statistic is not asymptotically $\chi^2$ distributed for LCA.
-The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR [REF lo 2001].
+The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR [@lo_testing_2001].
 This test is implemented in `tidySEM`, and is part of the default output of `table_fit()`, comparing subsequent rows of the table.
@@
 The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the $k-1$-class model $B$ times (e.g., 1000),
-and fitting the $k$- and $k-1$-class models to each simulated dataset [REF Nylund et al., 2007].
+and fitting the $k$- and $k-1$-class models to each simulated dataset [@nylund_deciding_2007].
 The p-value of this test is the proportion of $B$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
 Simulation studies suggest that the BLRT has greater statistical power than the LRT,
-although both performed comparably well at detecting the true number of classes [REF Nylund 2007, REF Tein 2013].
+although both performed comparably well at detecting the true number of classes [@nylund_deciding_2007; @tein_statistical_2013].
 A crucial limitation of the BLRT is that it is very computationally expensive.
@@
 which describes how well each model is able to reproduce the observed correlation matrix [see @mcelreath_statistical_2020].
-One version of the lazy bootstrap is already implemented in the commercial program latentGOLD [REF].
+One version of the lazy bootstrap is already implemented in the commercial program *Latent GOLD* [@vermunt_upgrade_2021].
 At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
@@
 The diagonal represents the average predictive probability of a the subjects related to their most likely class, indicating the certainty of class assignment for the subjects.
-Some guidelines have being suggested for this diagonal, like for them to be higher than 0.7 [@masyn2013], but we suggest researchers to be cautions about these guidelines as they cannot be generalized and should not be use as strict cutoffs as evidence "good" or "bad" classification.
+Some guidelines have being suggested for this diagonal, like for them to be higher than 0.7 [@masyn_latent_2013], but we suggest researchers to be cautions about these guidelines as they cannot be generalized and should not be use as strict cutoffs as evidence "good" or "bad" classification.
 
@@
 Finally, class separability can be summarized in a single statistic:
-the entropy criterion [REF].
-<!-- Celeux, G., & Soromenho, G. (1996). An entropy criterion for assessing the number of clusters -->
-<!-- in a mixture model. Journal of Classification, 13, 195-212. -->
+the entropy criterion [@celeux_entropy_1996].
+Celeux, G., & Soromenho, G. (1996). 
 In physics, high entropy refers to maximum randomness or uncertainty,
@@
 Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
-Crucially, in `tidySEM`, the entropy criterion is defined as 1-entropy (a convention originating in Mplus).
+Crucially, in `tidySEM`, the entropy criterion is defined as 1-entropy (a convention originating in *Mplus*).
 Its interpretation is reversed;
@@
 There are several different ways to do follow-up analyses in latent class analysis (LCA) while accounting for classification error.
-At present, the so-called BCH method is considered one of the best approaches [after its creators, REF].
+At present, the so-called BCH method is considered one of the best approaches [@bolck_estimating_2004].
 It has been implemented in `tidySEM`.
@@
 4. Class enumeration
-    + Justify the maximum number of classes $K$
-    + In exploratory LCA, estimate 1:$K$ classes
+    + Justify the maximum number of classes $k$
+    + In exploratory LCA, estimate 1:$k$ classes
     + In confirmatory LCA, estimate the theoretical number of classes, a 1-class model, and optionally other numbers of classes for comparison

### 2023-02-16 manuscript.tex
Update pt. 1
Updated manuscript and cleaned references

@@
 
-\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}
+\shorttitle{BEST PRACTICES IN LATENT CLASS ANALYSIS}
 
@@
 describes best practices for the estimation and reporting of LCA,
-using free and open source software in R. To this end, this tutorial
-introduces new functionality for estimating and reporting mixture
-models in the \texttt{tidySEM} R-package whose backend relies on \texttt{OpenMx}.
+using free and open source software in R. To this end,
+we introduce new functionality for estimating and reporting mixture models
+using the \texttt{tidySEM} R-package.
 }
@@
 most notably finite Gaussian mixture modeling and latent profile
-analysis. Jeroen K. Vermunt et al. (2004) defined LCA more generally as virtually
-any statistical model where ``some of the parameters {[}\ldots{]} differ across
-unobserved subgroups''.
-
-Despite the popularity of LCA, two challenges impede broader adoption
-and correct application of the method. The first challenge is that most
-existing software that implements these models is commercial and
-closed-source. The second challenge is that there is a lack of standards
-for estimating and reporting LCA, and some myths and misunderstandings
-have been perpetuated in prior literature. This introduces a risk of
-misapplications of the technique, and complicates manuscript review and
-assessment of the quality of published studies. The present paper seeks
-to address these challenges by introducing updated guidelines for
-estimating and reporting LCA, based on current best practice.
-Furthermore, meeting the increasing demand for open source research
-software, this paper introduces new functionality in the R-package
-\texttt{tidySEM} (\textbf{vanlissaTidySEMTidyStructural2022?}) to perform LCA using free
-open source software (FOSS).
+analysis. Vermunt et al. (2004) defined LCA more generally as
+virtually any statistical model where ``some of the parameters {[}\ldots{]}
+differ across unobserved subgroups''.
+
+Despite the popularity of LCA,
+two challenges impede broader adoption and correct application of the method.
+The first challenge is that most existing software that implements these models is commercial and closed-source.
+The second challenge is that there is a lack of standards for estimating and reporting LCA, and some myths and misunderstandings have been perpetuated in prior literature.
+This introduces a risk of misapplications of the technique,
+and complicates manuscript review and assessment of the quality of published studies.
+The present paper seeks to address these challenges by introducing updated guidelines for estimating and reporting LCA,
+based on current best practices.
+Furthermore, meeting the increasing demand for open source research software,
+this paper introduces new functionality in the R-package \texttt{tidySEM} (Van Lissa, 2022b) to perform LCA using free open source software (FOSS).
 
@@
 
-Despite the popularity of LCA, most existing software is commercial and
-closed-source, for example, Mplus (\textbf{muthenMplusUserGuide1998?}), for which
-the R-package \texttt{MplusAutomation} offers augmented LCA functionality
-(\textbf{hallquistMplusAutomationPackageFacilitating2018?}), and LatentGOLD
-{[}REF{]}.
-A
-downside of commercial software is that its use is restricted to
-individuals and institutions who can afford a license. Another downside
-is that the proprietary source code cannot be audited, debugged, or
-enhanced by third parties. This goes against open science principles
-(\textbf{lamprechtFAIRPrinciplesResearch2019?}), incurs a risk that mistakes in
-the source code go unnoticed, and curbs progress as independent
-developers cannot fix bugs or add functionality.
-
-There are several compelling reasons to prefer free open source software
-(FOSS) instead. FOSS is freely available for anyone to use, modify, and
-distribute, which means that scientists can access and use a wide range
-of powerful tools without financial impediments. This is invaluable for
-reducing gatekeeping of researchers with limited budgets, such as those
-from developing countries. Furthermore, FOSS promotes collaboration
-within the scientific community, as its users can contribute to the
-development, maintenance, and support of the software. Finally, FOSS
-promotes transparency and reproducibility: Because the source code for
-FOSS is accessible, researchers can review and verify the methods and
-algorithms used, which helps to ensure the integrity and reliability of
-research findings. Overall, FOSS has the potential to significantly
-improve the efficiency and effectiveness of scientific research, while
-also promoting transparency, collaboration, and inclusion within the
-scientific community.
-
-Notable FOSS for LCA includes mclust (Scrucca, Fop, Murphy, \& Raftery, 2016), depmixS4
-(Visser \& Speekenbrink, 2010), and OpenMx (Neale et al., 2016). A crucial limitation of
-these existing packages is that some have limited functionality (e.g.,
-Mclust), or implement specific innovations (e.g., depmixS4). OpenMx is
-unique in that, like Mplus, it implement LCA in the context of a fully
-featured structural equation modeling framework. Remaining obstacles to
-the use of \texttt{OpenMx} is that it is not very user-friendly and its
-functionality for LCA is poorly documented.
-
-These limitations are addressed by the \texttt{tidySEM} package, which provide
-a user-friendly interface for LCA in \texttt{OpenMx}. The package also includes
-convenience functions to obtain relevant statistics, tests and plots in
-accordance with best practices in estimation and reporting LCA, as
-described in this paper. The primary purpose of the \texttt{tidySEM} package is
-to provide a tidy workflow for generating, estimating, reporting, and
-plotting structural equation models in a software-agnostic way. The user
-interface is simple, as models can be specified using the widely used
-\texttt{lavaan} syntax (Rosseel, 2012). The package contains
-user-friendly wrapper functions to conduct LCA with sensible defaults,
-but the resulting models can still be fully customized (low barrier,
-high ceiling). Specifically, LCA models constructed in \texttt{tidySEM} inherit
-from \texttt{OpenMx}' format \texttt{MxModel}, and remain compatible with downstream
-functions that process such models. Being FOSS, \texttt{tidySEM} facilitates
-fully-reproducible analyses and contributes to open science. It was
-developed with open science in mind, and adheres to the FAIR software
-principles (Findable, Accessible, Interoperable, and Reusable, \textbf{lamprechtFAIRPrinciplesResearch2019?}), and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{OpenSSF Best
-Practices} for
-FOSS software.
+Despite the popularity of LCA,
+most existing software is commercial and closed-source,
+for example, \emph{Mplus} (Muthén \& Muthén, 1998), for which the R-package \texttt{MplusAutomation} offers augmented LCA functionality (Hallquist, Wiley, \& Van Lissa, 2018), and \emph{Latent GOLD} (Vermunt \& Magidson, 2000).
+A downside of commercial software is that its use is restricted to individuals and institutions that can afford a license.
+Another downside is that proprietary source code cannot be audited, debugged, or enhanced by third parties.
+This goes against open science principles (Lamprecht et al., 2020),
+incurs a risk that mistakes in the source code may go unnoticed,
+and curbs progress as independent developers cannot fix bugs or add functionality.
+
+There are several compelling reasons to prefer free open source software instead.
+FOSS is freely available for anyone to use, modify, and distribute, which means that scientists can access and use a wide range of powerful tools without financial impediments.
+This is invaluable for reducing gatekeeping of researchers with limited budgets, such as those from developing countries.
+Furthermore, FOSS promotes collaboration within the scientific community,
+as its users can contribute to the development, maintenance, and support of the software.
+Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
+researchers can review and verify the methods and algorithms used,
+which helps to ensure the integrity and reliability of research findings.
+Overall, FOSS has the potential to significantly improve the efficiency and effectiveness of scientific research, while also promoting transparency, collaboration, and inclusivity within the scientific community.
+
+Notable FOSS for LCA includes \texttt{mclust} (Scrucca, Fop, Murphy, \& Raftery, 2016), \texttt{depmixS4} (Visser \& Speekenbrink, 2010), and \texttt{OpenMx} (Neale et al., 2016).
+A crucial limitation of these existing packages is that some have limited functionality (e.g., \texttt{mclust}),
+or implement specific innovations (e.g., \texttt{depmixS4}).
+\texttt{OpenMx} is unique in that, like proprietary \emph{Mplus}, it implements LCA in the context of a fully featured structural equation modeling framework.
+A major downside of \texttt{OpenMx} is that it is not very user-friendly and its functionality is poorly documented.
+
+These limitations are addressed by the \texttt{tidySEM} package,
+which provides a user-friendly interface for LCA in \texttt{OpenMx}.
+The package also includes convenience functions to obtain relevant statistics, tests and plots in accordance with best practices in estimation and reporting of LCA, as described in this paper.
+The primary purpose of the \texttt{tidySEM} package is to provide a tidy workflow for generating, estimating, reporting, and plotting structural equation models in a software-agnostic way.
+The user interface is simple, as models can be specified using the widely used \texttt{lavaan} syntax (Rosseel, 2012).
+The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
+but the resulting models can still be fully customized making it a low threshold - high ceiling package.
+For instance, LCA models constructed in \texttt{tidySEM} inherit the \texttt{MxModel} format from \texttt{OpenMx} which renders them compatible with various legacy functions from \texttt{OpenMx} and related LCA ecosystem .
+Since \texttt{tidySEM} was developed with open science in mind, it facilitates fully-reproducible analyses and adheres to the \emph{FAIR (Findable, Accessible, Interoperable, and Reusable) software principles} (Lamprecht et al., 2020),
+and the \href{https://bestpractices.coreinfrastructure.org/en/criteria}{\emph{OpenSSF Best Practices}} for FOSS software.
 
@@
 
-LCA refers to methods for estimating unobserved groups based on a
-parametric model of observed indicators of group membership
-(\textbf{vermuntLatentClassAnalysis2004?}). It has become popular across
-scientific fields, under a number of different names; most notably
-(finite Gaussian) mixture modeling and latent profile analysis. The
-concept of LCA can be understood in different ways. Generally speaking,
-LCA assumes that the study population is composed of \(K\) subpopulations.
-It further assumes that the observed data are a mixture of data
-originating from those subpopulations; hence the name mixture model.
-Consider the simplest possible univariate ``model'', which is a normal
-distribution. This model has two parameters: the mean and the variance.
-LCA aims to estimate the values of those parameters across a known
-number of classes \(K\), as well as the probability of every individual
-\(i\) belonging to classes \(1 \ldots k\). Commonly, the same model is
-estimated across all classes, but with different parameters for each
-class (i.e., class-specific means and variances).
+LCA refers to methods for estimating unobserved
+groups based on a parametric model of observed indicators of group
+membership (Vermunt et al., 2004).
+It has become popular across scientific fields,
+under a number of different names;
+most notably (\emph{finite Gaussian}) \emph{mixture modeling} and \emph{latent profile analysis}.
+The concept of LCA can be understood in different ways.
+Generally speaking, LCA assumes that the study population is composed of \(k\)
+subpopulations.
+It further assumes that the observed data are
+a mixture of observations originating from those subpopulations;
+hence the name \emph{mixture model}.
+Consider the simplest possible
+univariate model, which is a normal distribution.
+This model has two parameters: the mean and the variance.
+LCA aims to estimate the values of those parameters across \(k\) classes,
+as well as estimate the probability of every individual observation \(i\) belonging to each of the \(1 \ldots k\) classes.
+Commonly, the same model is fitted to all \(k\) classes,
+but each class is characterized by a unique set of parameters for that model (i.e., class-specific means and variances).
 
@@
 When estimating a two-class mixture model, the detective observes that
-the model estimates the mean shoe size of the two groups are equal to
+the model estimates the mean shoe size of the two groups as equal to
 9.22 and
@@
 
-Another way to conceptualize LCA is by analogy to a measurement model in
-structural equation modeling (\textbf{molenaarArbitraryNatureLatent1994?}). A
-mixture model is like confirmatory factor analysis, except with a
-categorical instead of a continuous latent variable. One difference
-between the two techniques is that factor analysis can be considered as
-a way to group observed \emph{variables} into latent constructs, and LCA
-groups \emph{individuals} into classes. While factor analysis seeks to
-describe the relations between variables, LCA seeks to describe
-unobserved groups and classify individual observations to those groups.
-In
-line with this distinction, LCA (mixture models) is referred to as a
-``person-centered'' technique, and factor analysis as a
-``variable-centered'' technique (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+Another way to conceptualize LCA is through an analogy to a measurement model in structural equation modeling (Molenaar \& Eye, 1994).
+A mixture model is like confirmatory factor analysis,
+except for the fact that the latent variable is categorical instead of continuous.
+Another difference between the two techniques is that factor analysis can group observed \emph{variables} into latent constructs, while LCA can group \emph{observations} into classes.
+While factor analysis seeks to describe the relations between variables,
+LCA seeks to describe unobserved groups and classify individual observations to those groups.
+In line with this distinction, LCA is referred to as a ``person-centered'' technique, and
+factor analysis as a ``variable-centered'' technique
+(Masyn, 2013; Nylund-Gibson \& Choi, 2018).
 
@@
 thought of as similar to a multi-group structural equation model. The
-main distinction is that group membership is not known a-priori, but
-estimated from the data. Whereas in a multi-group model, the data are
-split by group and treated as independent samples, in LCA, all cases
-contribute to the estimation of all parameters in all groups. The
-relative contribution of each case to the parameters of each group is
-determined by that case's posterior probability of belonging to that
-group.
-
-When the focus is on each individual's estimated class membership, LCA
-can be thought of as a type of clustering algorithm that corrects for
-(random) measurement error. Specifically, the posterior class
-probability that an individual belongs to a latent class can be computed
-from the likelihood of that individual's observed data under given the
-class-specific model. These posterior class probabilities can be used to
-weight follow-up analyses, or individuals can be assigned to classes
-based on their highest class probability. If the classes are clearly
-distinct, measurement error is low, and the latter approach may be
-acceptable. This perspective on mixture modeling is sometimes referred
-to as ``model-based clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016). Many clustering algorithms apply some recursive
-splitting algorithm to the data. By contrast ``model-based'' clustering
-refers to the fact that LCA estimates cluster membership based on a
-parametric model. Since a parametric model estimates only a fixed number
-of parameters, it can be a relatively parsimonious solution compared to
-non-parametric techniques.
+main distinction is that group membership is not known a-priori, but estimated from the data.
+Whereas in a multi-group model, the data are split by group and treated
+as independent samples, in LCA, all cases contribute to the estimation
+of all parameters in all groups. The relative contribution of each case
+to the parameters of each group is determined by that case's \emph{posterior
+probability} of belonging to that group.
+
+When the focus is on each individual's estimated class membership,
+LCA can be thought of as a type of clustering algorithm that corrects for random measurement error by default, and is able to correct for systematic measurement error if it is included in the model.
+Specifically, the posterior class probability that an individual belongs to a latent class
+can be computed from the likelihood of that individual's observed data given the class-specific model.
+These posterior class probabilities can be used to weight follow-up analyses,
+or to assign individuals to classes based on their highest class probability.
+If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
+This perspective on mixture modeling is sometimes referred to as ``model-based
+clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016).
+Many clustering algorithms apply some recursive splitting algorithm to the data. By
+contrast ``model-based'' clustering refers to the fact that LCA estimates
+cluster membership based on a parametric model.
+Since a parametric model estimates only a fixed number of parameters,
+it can be a relatively parsimonious solution compared to non-parametric techniques.
 
 Finally, in the context of machine learning, LCA can be considered as an
-\emph{unsupervised} learning method for \emph{classification}
-(Figueiredo \& Jain, 2002). The term \emph{unsupervised} refers to the
-fact that the outcome variable -- true class membership -- is not known,
-and the term \emph{classification} refers to the fact that the algorithm is
-predicting a categorical outcome: class membership.
+\emph{unsupervised} learning method for \emph{classification} (Figueiredo \& Jain, 2002).
+The term \emph{unsupervised} refers to the fact that the outcome variable --
+true class membership -- is not known, and the term \emph{classification}
+refers to the fact that the algorithm is predicting a categorical
+outcome: class membership.
 
@@
 
-The term LCA was initially conceived to refer to analyses with
-categorical, usually binary indicators (Collins \& Lanza, 2009). It was
-later generalized to refer to any technique that estimates a categorical
-latent variable based on a parametric model of observed indicators, as
-it only refers to the purpose of the analysis, and does not imply
-restrictions to the model used, or the level of measurement of the
-indicators (Jeroen K. Vermunt et al., 2004). To prevent ambiguity, we refer to LCA
-with binary or ordinal indicators as \emph{LCA with ordinal indicators}.
-Given the abundance of redundant terms in use for closely related types
-of LCA models, we here provide a rudimentary taxonomy.
+The term LCA was initially conceived to refer to analyses with categorical,
+usually binary indicators (Collins \& Lanza, 2009).
+It was later generalized to refer to any technique that estimates a categorical latent variable based on a
+parametric model of observed indicators, as it only refers
+to the purpose of the analysis, and does not imply restrictions to the
+model used, or the level of measurement of the indicators (Vermunt et al., 2004).
+To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators},
+and the sole \emph{LCA} expression as an umbrella term.
+Given the abundance of redundant terms in use for closely related types of LCA models,
+we here provide a rudimentary taxonomy.
 
 A common type of LCA is the \emph{finite Gaussian mixture model}; a
-univariate analysis where the observed distribution of a single
-continuous variable is assumed to result from a mixture of a known
-number of normal (Gaussian) distributions. The parameters of a finite
-Gaussian mixture model are the means and variances of these underlying
-normal distributions. The analysis of shoe sizes presented earlier is a
+univariate analysis where the observed distribution of a single variable
+is assumed to result from a mixture of a known number of normal
+(Gaussian) distributions. The parameters of a finite Gaussian mixture
+model are the means and variances of these underlying normal
+distributions. The analysis of shoe sizes presented earlier is a
 canonical example of this type of analysis. In the multivariate case,
@@
 are the means, variances, and covariances between the indicators (which
-can be standardized to obtain correlations). These parameters can be
-estimated freely, or set to be constrained across classes
-(Collins \& Lanza, 2009).
+can be standardized to obtain correlations). These parameters can be estimated
+freely, or set to be constrained across classes (Collins \& Lanza, 2009).
 
@@
 indicators. Conditional independence means that, after class membership
-is accounted for, the residual covariances/correlations between
-indicators are assumed to be zero. In some cases, such constraints will
-be inappropriate; for instance when the cohesion between indicators is
-expected to differ between classes (Collins \& Lanza, 2009). Consider, for
-example, a mixture model of ocean plastic particles, which found two
-classes of particles based on length and width: A class of larger
-particles with a low correlation between length and width, and a class
-of smaller particles with a high correlation between length and width.
+is accounted for,
+the residual covariances/correlations between indicators are
+assumed to be zero.
+In some cases, such constraints
+will be inappropriate; for instance when the cohesion between indicators
+is expected to differ between classes (Collins \& Lanza, 2009).
+Consider, for example, a mixture model of ocean plastic particles, which found two classes of particles based on length and width:
+a class of larger particles with a low correlation between length and width,
+and a class of smaller particles with a high correlation between length and width.
 The reason for this difference in correlations makes theoretical sense:
-the large particles were heterogenous in shape, whereas the smaller
-particles had been polished to a more uniform (rounded) shape by the
-elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).
-
-A longitudinal extension of LCA is the Hidden Markov Model (HMM, also
-called Latent Transition Analysis), which estimates the change in class
-membership over time with the markov assumption over time. The markov
-assumption states that the class membership at time \(t\) is only affected
+the large particles were heterogenous in shape,
+whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).
+
+A longitudinal extension of LCA is the \emph{Hidden Markov Model} (\emph{HMM}, also
+called \emph{Latent Transition Analysis}), which estimates the change in class
+membership over time under the Markov assumption. The Markov
+assumption states that class membership at time \(t\) is only affected
 by the class membership at time \(t-1\). The structural part of the model
-is comprise of the initial states (proportion of the sample in each
+is comprised of the initial states (proportion of the sample in each
 class at the first time point), and transition probabilities
-(probabilities to switch class membership across time). The transition
+(probabilities of class membership switching across time). The transition
 probabilities are estimated as multinomial logistic regressions between
-class memberships (Jeroen K. Vermunt, 2010).
-
-Finally, some LCA models use latent variables as indicators. A common
-application of this approach is in longitudinal research with repeated
-measures of a construct. Examples of this approach include \emph{growth
-mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
-(Jung \& Wickrama, 2008). These techniques estimate a latent growth
-model to describe grouped trajectories over time. The indicators of
-class membership in a GMM are the intercepts, and variances, and
-covariances of the latent growth variables that describe individual
-trajectories on the observed variables. For linear trajectories, these
-latent variables are an intercept and slope. The free variances allow
-trajectories to vary within classes. The LCGA differs from the GMM in
-that it assumes homogeneity of trajectories within classes. This is
-achieved by restricting the (co)variances of the growth parameters to
-zero. Any variance in the indicators not explained by the class-specific
+class memberships (Vermunt, 2010).
+
+Finally, some LCA models use latent variables as indicators.
+A common application of this approach is in
+longitudinal research with repeated measures of a construct.
+Examples of this approach include
+\emph{growth mixture models} (GMM) and \emph{latent class growth analyses} (LCGA)
+(Jung \& Wickrama, 2008).
+These techniques estimate a latent growth model
+to describe grouped trajectories over time.
+The indicators of class membership in a GMM are the
+intercepts, variances, and covariances of the latent growth variables that describe individual trajectories on the observed variables.
+For linear trajectories, these latent variables are the intercept and slope.
+The free variances allow trajectories to vary within classes. The LCGA differs from the GMM in that it assumes within-class homogeneity of trajectories.
+This is achieved by restricting the (co)variances of the growth parameters to zero.
+Any variance in the indicators not explained by the class-specific
 latent trajectories is assumed to be error variance.
@@
 
-Although LCA is often discussed as an exploratory analysis technique, it
-can also be used in a confirmatory manner. Given that LCA is similar to
-confirmatory factor analysis, it can be used to similar ends: To assess
-whether a theoretical measurement model holds. This use case is relevant
-when testing a theory that postulates the existence of a categorical
-latent variable. For example, \emph{identity status theory} posits that, at
-any given point in time, adolescents reside in one of four identity
-statuses (Marcia, 1966). LCA can be used to identify these
-four statuses based on observed indicators (e.g., self-reported identity
-exploration and commitment)
-(\textbf{luyckxDevelopmentalTypologiesIdentity2008?}). Similarly, \emph{personality
-type theory} states that dimensional differences in personality can
-largely be explained by an undercontrolled, overcontrolled, and
-resilient type, which could also be restored using LCA
-(\textbf{donnellanResilientOvercontrolledUndercontrolled2010?}).
-
-One challenge is that, unlike confirmatory factor analysis, LCA does not
-provide absolute fit indices (see Model Fit Indices). Thus, to test a
-theory, one can ascertain that the data are better described by the
-number of classes dictated by theory than by a different number of
-classes. Furthermore, it would be possible to test whether the observed
-pattern of class-specific means corresponds to a hypothesized pattern;
-either qualitatively or quantitatively, using significance tests for
-pre-specified values (see Inference).
-
-Appendix A is a tutorial to confirmatory LCA in \texttt{tidySEM}, based on a
-recent study that tested the theory that identities of Belgian
-adolescents with migration backgrounds could be summarized as those
-relating to their heritage, national and regional identity
-(Maene, D'hondt, Van Lissa, Thijs, \& Stevens, 2022).
+Although LCA is often discussed as an exploratory analysis technique,
+it can also be used in a confirmatory manner.
+Given that LCA is similar to confirmatory factor analysis,
+it can be used to similar ends:
+to assess whether a theoretical measurement model holds.
+This use case is relevant when testing a theory that postulates the existence of a categorical latent variable.
+For example, \emph{identity status theory} posits that, at any given point in time, adolescents reside in one of four identity statuses (Marcia, 1966).
+LCA can be used to identify these four statuses based on observed indicators (e.g., self-reported identity exploration and commitment, Luyckx, Schwartz, Goossens, Soenens, \& Beyers, 2008).
+Similarly, \emph{personality type theory} states that dimensional differences in personality can largely be explained by an undercontrolled, overcontrolled, and resilient personality types, which could also be restored using LCA (Donnellan \& Robins, 2010).
+
+One challenge is that, unlike confirmatory factor analysis, LCA does not provide absolute fit indices (see Model Fit Indices).
+Thus, to test a theory, one can ascertain that the data are better described by the number of classes dictated by theory than by a different number of classes.
+Furthermore, it would be possible to test whether the observed pattern of class-specific means corresponds to a hypothesized pattern;
+either qualitatively or quantitatively, using significance tests for pre-specified values (see Inference).
+
+Appendix A showcases a step-by-step tutorial to confirmatory LCA in \texttt{tidySEM},
+based on a recent study that tested a theory that identities of Belgian adolescents with migration backgrounds could be summarized as those relating to their heritage, national and regional identities (Maene, D'hondt, Van Lissa, Thijs, \& Stevens, 2022).
 
@@
 
-Another use case is to explore whether a population is comprised of
-latent subpopulations, and if so, how many. The primary focus in this
-use case is on class enumeration. For example,
-one study explored heterogeneity in the population of
-military service members and veterans with prior suicide attempts, and
-identified two latent profiles characterized by high versus low suicide
-risk based on self-reported suicidal ideation and information about
-prior attempted methods and severity (Gromatsky et al., 2022).
-Knowing that the population of suicidal service members is heterogeneous
-can inform future research, policy, and care provided. Appendix B offers
-a tutorial for exploratory LCA, based on the aforementioned study of the
-distribution of ocean microplastics (Alkema et al., 2022).
+Another use case of LCA is to explore whether a population is comprised of latent subpopulations, and if so, how many.
+The primary focus in this use case is on class enumeration.
+For example,
+one study explored heterogeneity in the population of military service members and veterans with prior suicide attempts,
+and identified two latent profiles characterized by high versus low suicide risk based on self-reported suicidal ideation and prior attempt methods and severity
+(Gromatsky et al., 2022).
+Knowing that the population of suicidal service members is heterogeneous can inform future research, policy, and care.
+Appendix B offers a tutorial for exploratory LCA,
+based on the aforementioned study of the distribution of ocean microplastics (Alkema et al., 2022).
 
@@
 
-LCA can also be used as a measurement model, similar to confirmatory
-factor analysis. This is useful when class membership is measured
-imperfectly by several indicators: LCA can partial out measurement error
-and restore most likely class membership. For example, one study found
-discrepancies between self-reported employment status and official
-register data (Pavlopoulos \& Vermunt, 2015). While it might be tempting
-to assume that register data is free of error, this assumption was shown
-to be false. A mixture model was used to estimate employment status,
-correcting for measurement error, from both self-report and registry
-data. The study investigated random and systematic measurement error in
-both indicators, which reveal strengths and weaknesses of both types of
-assessment.
-
-This example also illustrates that LCA allows us to evaluate the
-reliability of different indicators (Geiser \& Wurpts, 2014).
-A
-continuous indicator that discriminates well between classes will show
-large differences in class-specific means; an ordinal indicator will
-show very high or very low conditional response probabilities for some
-classes but not for others. This information can be used to select the
-indicators that are most diagnostic of class membership. For an example
-of LCA with a focus on differential item functioning, consider a study
-on university admission tests of Saudi Arabian students
-(Tsaousis, Sideridis, \& AlGhamdi, 2020).
+LCA can also be used as a measurement model, similar to confirmatory factor analysis.
+This is useful when class membership is measured imperfectly by several indicators:
+LCA can partial out measurement error and restore most likely class membership.
+For example, one study found discrepancies between self-reported employment status and official register data (Pavlopoulos \& Vermunt, 2015).
+While it might be tempting to assume that register data is free of error, this assumption was shown to be false.
+A mixture model was used to estimate employment status, correcting for measurement error, from both self-report and registry data.
+The study investigated random and systematic measurement error in the indicatorss stemming from the two registers,
+which revealed strengths and weaknesses of both types of assessment.
+
+This example also illustrates that LCA allows us to evaluate the reliability of different types of indicators (Geiser \& Wurpts, 2014).
+As a rule, a continuous indicator that discriminates well between classes will show large differences in class-specific means.
+An ordinal indicator with the same property will show very high or very low conditional response probabilities for some classes but not for others.
+This information can be used to select the indicators that are most diagnostic of class membership.
+For an example of LCA with a focus on differential item functioning,
+see a study on university admission tests of Saudi Arabian students (Tsaousis, Sideridis, \& AlGhamdi, 2020).
 
@@
 
-LCA can also be used to reduce high-dimensional data to just a few
-prototypes. Factor analysis is also sometimes used for dimension
-reduction, but only accounts for linear covariance between indicators,
-whereas LCA can accommodate complex - but discrete - patterns. For
-example, imagine that there exist two indicators - \(x\) and \(y\), and some
-individuals score high on both, some score low on both, and some score
-high on \(x\) but low on \(y\). Factor analysis would only be able to
-capture the differences between the first two groups well; LCA captures
-differences between all three. Note that this is a pragmatic use of LCA;
-there is no theory about the existence of a categorical latent variable
-(cf.~Testing theory). Consequently, the class solution should be treated
-as descriptive of the sample, and not be ``reified'' - treated as evidence
-of the existence of a categorical latent variable.
-
-The use of LCA for dimension reduction is common in longitudinal
-research, where one developmental process is summarized using latent
-class growth analysis (LCGA), and the resulting categories are then used
-as a moderator of another developmental process in a multi-group model.
-For example, one study conducted a LCGA of adolescent empathy
-development on two dimensions of empathy and found three groups: high,
-average, and low empathy (Van Lissa et al., 2015). These groups were
-then used as moderator of a latent growth analysis of adolescent- and
-parent-reported conflict. Results showed that the high-empathy group
-evidenced greater adolescent-parent agreement about the incidence of
-conflict than the other groups, and that the low empathy group had more
-conflict with parents than the other groups. Note that both these key
-findings are non-linear; modeling them in a single-group model would be
-difficult. LCGA greatly simplifies model specification and
-interpretation, albeit at a cost of some loss of information and,
-potentially, generalizability. Appendix C is a tutorial on LCGA, based
-on ongoing research on trajectories of depressive symptoms in military
-service members after deployment.
-
-The need for dimension reduction can be especially pertinent when using
-ordinal indicators. For example, one study measured fifteen dichotomous
-health indicators among injured military personnel
-(MacGregor et al., 2021). Combinatorics informs us that fifteen
+LCA can also be used to reduce high-dimensional data
+to just a few prototypes.
+Factor analysis is also sometimes used for dimension reduction,
+but only accounts for linear covariance between indicators,
+whereas LCA can accommodate complex - but discrete - patterns.
+For example, imagine that there exist two indicators - \(x\) and \(y\). Some individuals score high on both indicators, some score low on both, and other score high on \(x\) but low on \(y\).
+Factor analysis would only be able to capture the differences between the first two groups well, but LCA will capture differences between all three classes.
+Note, however, that this is a pragmatic use of LCA.
+There is no theory about the existence of a categorical latent variable (cf.~Testing theory).
+Consequently, the final class solution should be treated as descriptive of the sample.
+It should not be reified or treated as evidence of the existence of a categorical latent variable without accumulation of additional empirical evidence and deliberation on the findings.
+The discovered classes could indeed prove to be theoretically justified and thus inform the theory building process, but a single exploratory fining is never sufficient for reaching such conclusions.
+
+The use of LCA for dimension reduction is common in longitudinal research,
+where one developmental process is summarized using latent class growth analysis (LCGA),
+and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
+For example, one study conducted a LCGA of adolescent empathy development on two dimensions of empathy identifying: high, average, and low empathy classes (Van Lissa et al., 2015).
+The membership to these three groups was then used as a moderator in a latent growth analysis of adolescent- and parent-reported conflict.
+The results showed that the high-empathy group evidenced greater adolescent-parent agreement about the incidence of conflict than other groups,
+and the low empathy group had more conflict with parents than other groups.
+Note that as both of these effects were non-linear,
+modeling them in a single-group framework would introduce unnecessary bias.
+LCGA greatly simplifies model specification and interpretation,
+albeit at a cost of some loss of information, and potentially generalizability.
+Appendix C is a tutorial on LCGA,
+based on ongoing research on trajectories of depressive symptoms in military service members after deployment.
+
+The need for dimension reduction can be especially pertinent when using ordinal indicators.
+For example, one study measured
+fifteen dichotomous health indicators among injured military personnel (MacGregor et al., 2021).
+Combinatorics informs us that fifteen
 dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
-However, LCA was able to reduce response patterns to five clinically
-meaningful latent classes.
+However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data.
 
@@
 
-Another use case is to estimate individuals' unobserved class membership
-based on observed indicators. The shoe size example in Figure
-\ref{fig:shoedens} is a rudimentary illustration of this application.
-In applied research, LCA is often used for classification in clinical
-contexts. For example, LCA was used to determine clinical cut-off scores
-for diagnosing whooping cough (Baughman, Bisgard, Lynn, \& Meade, 2006). Another study
-used LCA to not only classify the sample used to estimate the model, but
-also future cases {[}REF Zegwaard et al, in prep{]}. The study identified
-four types of care providers among those who supported close kin with
-mental health problems. Importantly, the LCA model was embedded in an
-interactive app that could be used by healthcare professionals to
-determine the most likely class membership for new care providers, to
-provide more targeted support. To obtain predicted class membership
-based on an existing LCA model, the authors computed the likelihood of
-data from a new participant under the multivariate normal distributions
-estimated within all latent classes.
-These applications of LCA aid the clinical decision making process, and
-help decide whether additional support is warranted, or what type of
-intervention is appropriate.
+Another use case of LCA is to estimate individuals' unobserved class membership based on observed indicators.
+The shoe size example in Figure \ref{fig:shoedens}
+is a rudimentary illustration of such application.
+In applied research, LCA is often used for classification in clinical contexts.
+For example, LCA was used to determine clinical cut-off scores for diagnosing whooping cough (Baughman, Bisgard, Lynn, \& Meade, 2006).
+Another study used LCA not only to classify the sample used to estimate the model, but also predict future cases {[}REF Zegwaard et al, in prep{]}.
+The study identified four types of care providers among those who supported close kin with mental health problems.
+Importantly, the LCA model was embedded in an interactive app that could be used by healthcare professionals to determine the most likely class membership for new care providers and pave the way to providing them more targeted support.
+To obtain predicted class membership based on an existing LCA model,
+the authors computed the likelihood of data from a new participant under the multivariate normal distributions estimated within all latent classes.
+These applications of LCA aid the clinical decision making process,
+and help decide whether additional support is warranted,
+or what type of intervention is appropriate.
 
@@
 
-Finally, LCA can be used to deal with data which violate certain
-assumptions. For example, zero-inflated and hurdle models are special
-cases of LCA for dealing data with a very high proportion of zeros
-(extreme right skew) (\textbf{baughmanMixtureModelFramework2007?}). These models
-estimate one class to explain the ``true zeros'', and another class with a
-different distribution to explain all other values, which can include
-zero. Note that this LCA has different models for both classes, and thus
-falls beyond the scope of this paper.
-For instance, the aforementioned study on marine plastic noted that many
-prior studies had treated the distribution of particles as
-(multivariate) normal, despite clear deviations from normality
-(Alkema et al., 2022). A mixture of two normal distributions with
-distinct variances described the observed distribution better.
+Finally, LCA can be used to deal with data which violate certain assumptions.
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) {[}Baughman (2007); magnus\_multidimensional\_2021{]}.
+These models estimate one class to explain the ``true zeros'',
+and another class with a different distribution to explain all other values, which can include zero.
+Note that this LCA has different models for both classes,
+and thus falls beyond the scope of this paper.
+For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
+A mixture of two normal distributions with distinct variances described the observed distribution better.
 
@@
 \ref{tab:checkest}, are rooted in existing recommendations for best
-practices for estimating specific subtypes of LCA, including latent
+practices for estimating specific subtypes of LCA,
+including latent
 class growth analysis (Van De Schoot, Sijbrandij, Winter, Depaoli, \& Vermunt, 2017) and latent
 class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
-These were generalized to be relevant to all types of LCA, and updated
-to current best practices, as explained below.
+These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.
 
@@
 patterns and violations of assumptions that had not been considered
-prior to data collection. The function \texttt{descriptives(data)} provides
-extensive descriptive statistics, including the number of unique values,
-variance of continuous and categorical variables, missingness, skew and
-kurtosis (peakedness of the distribution). However, sample-level
-descriptive statistics are of limited value when the goal of a study is
-to identify subsamples using LCA. Plots convey additional information;
-specifically, density plots for continuous variables, and bar charts for
-categorical ones. Density plots can also aid in the choice of the number
-of classes, as explained in the section on visualization.
-
-Particular points of attention include measurement level of the
-indicators. Recall that \texttt{tidySEM} can estimate mixture models with
-either numeric (type \texttt{numeric} and \texttt{integer}) or \texttt{ordered} indicators.
-The column \texttt{type} should thus only contain these types. Also check that
-the number of \texttt{unique} values match the intended measurement level.
-Numeric variables with few unique values are typically better modeled as
-ordinal. Our experience consulting on LCA methods and moderating the
-\texttt{tidyLPA} Google group suggest that the misapplication of mixture models
-to ordinal (e.g., Likert-type) indicators is the most common source of
-user error. Whereas it has been argued that some parametric methods are
-robust when scales with 7+ response categories are treated as continuous
-(e.g., Norman, 2010), it is very unlikely that an ordinal
-variable can be modeled as a \emph{mixture} of multiple normal distributions.
-To illustrate the problem, consider what happens when estimating a
-5-class solution on a single 5-point Likert scale indicator. Each
-class-specific mean would describe a single response category, and a
-class-specific variance component would be empirically underidentified.
-In sum, variables with few unique values might not be suitable for
-Gaussian mixture modeling; LCA with ordinal indicators is more
-appropriate.
+prior to data collection.
+The function \texttt{descriptives(data)} provides extensive descriptive statistics,
+including the number of unique values,
+variance of continuous and categorical variables,
+missingness, skew and kurtosis (peakedness of the distribution).
+However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
+Plots convey additional information; specifically, density plots for continuous variables, and bar charts
+for categorical ones.
+Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.
+
+Particular points of attention include measurement level of the indicators.
+Recall that \texttt{tidySEM} can estimate mixture models with either numeric (type \texttt{numeric} and \texttt{integer}) or \texttt{ordered} indicators.
+The column \texttt{type} should thus only contain these types.
+Also check that the number of \texttt{unique} values match the intended measurement level.
+Numeric variables with few unique values are typically better modeled as ordinal.
+Our experience consulting on LCA methods
+and moderating the \texttt{tidyLPA} Google group suggest that the
+misapplication of mixture models to ordinal (e.g., Likert-type)
+indicators is the most common source of user error. Whereas it has been
+argued that some parametric methods are robust when scales with 7+
+response categories are treated as continuous {[}e.g., Norman (2010); rhemtulla\_when\_2012{]},
+it is very unlikely
+that an ordinal variable can be modeled as a \emph{mixture} of multiple
+normal distributions.
+To illustrate the problem, consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
+Each class-specific mean would describe a single response
+category,
+and a class-specific variance component would be empirically underidentified.
+In sum, variables with few unique values might not be suitable for Gaussian mixture modeling;
+LCA with ordinal indicators is more appropriate.
 
@@
 
-LCA can accommodate variables of different measurement levels:
-continuous, binary, nominal, and ordinal. Numeric variables should have
-an interval measurement level, such that it is sensible to estimate
-means and (co)variances. For the remaining three measurement levels,
-some preprocessing is required. The exact data types accepted by
-\texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}. To
-correctly specify ordered variables such as Likert scales, use the
-function \texttt{OpenMx::mxFactor()}. Binary variables, such as presence or
-absence of some symptom, are also treated as ordinal, with one category
-coded as zero and the other as one. Nominal variables are converted to
-binary indicators of group membership via dummy coding, for example,
-using \texttt{mx\_dummies()}.
-
-A second consideration in data preprocessing is that models with
-indicators on different scales may present with convergence issues. The
-reason for this is that the parameter space is high-dimensional with
-many potential local optima. If one indicator is on a much larger scale
-than another, the parameter space may be more extensively explored for
-that parameter. The problem can be resolved by transforming the
-indicators to place them on (roughly) similar scales. The most common
-transformation is standardization, which involves subtracting the mean
-of each variable from each value and dividing by the standard deviation.
+LCA can accommodate variables of different measurement levels: continuous, binary, nominal, and ordinal.
+Numeric variables should have an interval measurement level,
+such that it is sensible to estimate means and (co)variances.
+For the remaining three measurement levels,
+some preprocessing is required.
+The exact data types accepted by \texttt{tidySEM} at the time of writing are \texttt{numeric} and \texttt{ordered}.
+To correctly specify ordered variables
+such as Likert scales,
+use the function \texttt{OpenMx::mxFactor()}.
+Binary variables,
+such as presence or absence of some symptom, are also treated as ordinal, with one category coded as zero and the other as one.
+Nominal variables are converted to binary indicators of group membership via dummy coding, for example, using \texttt{mx\_dummies()}.
+
+A second consideration in data preprocessing is that models with indicators on different scales may present with convergence issues.
+The reason for this is that the parameter space is high-dimensional with many potential local optima.
+If one indicator is on a much larger scale than another, the parameter space may be more extensively explored for that parameter.
+The problem can be resolved by transforming the indicators to place them on (roughly) similar scales.
+The most common transformation is standardization, which involves subtracting the mean of each variable from each value and dividing by the standard deviation.
 In R, this transformation is applied using the \texttt{scale()} function.
-Standardization retains all available information, and model parameters
-can be transformed back to the original scale of the indicators by
-reversing the transformation. The transformed variable has a mean of
-zero and a standard deviation of one. Note that standardization is
-typically not necessary if indicators are already on the same, or
-similar, scales.
-
-A third consideration in preprocessing relates to the univariate
-distributions of indicators. For example, the \texttt{plas\_depression} data
-show extreme right skew in depressive symptoms: most respondents report
-few symptoms, and few respondents report many symptoms. Similarly, the
-\texttt{alkema\_microplastics} data are extremely right-skewed: Very small
-plastic fragments abound in the oceans, and large pieces are rare. These
-distributional characteristics must be taken into account either in
-preprocessing, or model specification. In preprocessing, the data can be
-transformed to reduce skew. After reducing skew, a simpler LCA model may
-fit the data well. Alternatively, one can specify a model that
-accommodates the distributional idiosyncrasies. For example, allowing
-free variances across classes might result in one class with a near-zero
-mean and small variance to account for the fact that most respondents
-have few depressive symptoms, and one class with an elevated mean and
-large variance to account for all respondents with non-zero symptoms.
+Standardization retains all available information,
+and model parameters can be transformed back to the original scale of the indicators by reversing the transformation.
+The transformed variable has a mean of zero and a standard deviation of one.
+Note that standardization is typically not necessary if indicators are already on the same, or similar, scales.
+
+A third consideration in preprocessing relates to the univariate distributions of indicators.
+For example, the \texttt{plas\_depression} data show extreme right skew in depressive symptoms:
+most respondents report few symptoms,
+and few respondents report many symptoms.
+Similarly, the \texttt{alkema\_microplastics} data are extremely right-skewed: Very small plastic fragments abound in the oceans, and large pieces are rare.
+These distributional characteristics must be taken into account either in preprocessing, or model specification.
+In preprocessing, the data can be transformed to reduce skew.
+After reducing skew, a simpler LCA model may fit the data well.
+Alternatively,
+one can specify a model that accommodates the distributional idiosyncrasies.
+For example, allowing free variances across classes might result in one class with a near-zero mean and small variance to account for the fact that most respondents have few depressive symptoms,
+and one class with an elevated mean and large variance to account for all respondents with non-zero symptoms.
 
@@
 missing data and reporting how missingness was handled
-(Van De Schoot et al., 2017). While we concur that
-investigating missingness is due diligence, it is important to emphasize
-that missingness is adequately handled by default in \texttt{OpenMx}, which is
-the computational backend of \texttt{tidySEM}. Its Full Information Maximum
-Likelihood (FIML) estimator makes use of all available information
-without imputing missing values, by estimating the model based on all
-available information for each subject
-(Enders, 2022; \textbf{endersRelativePerformanceFull2001?}). FIML is a
-best-practice solution for handling missing data; at least on par with
-multiple imputation (Lee \& Shi, 2021). Multiple imputation is less
-suitable to LCA for two reasons. First, because LCA methods are often
-computationally intensive, and conducting them on several imputed
-datasets multiplies the load. Second, because many multiple imputation
-methods assume normality; an assumption likely to be violated when data
-come from a heterogeneous population.
-
-Three types of missingness have been distinguished in the literature
-(Enders, 2022; Rubin, 1976): random missingness
-(MCAR); missingness contingent on \emph{observed} variables that can be known
-and controlled for (MAR); and missingness related to unobserved
-variables, which can be neither known nor controlled for (MNAR). A
-so-called ``MCAR'' test is often reported (Jamshidian \& Jalal, 2010), but
-note that this name is somewhat misleading, as a significant test
-indicates that missingness is MAR, but a non-significant test statistic
-does not distinguish between MCAR or MNAR. As the classic MCAR test
-relies on the comparison of variances across groups with different
-patterns of missing data, it assumes normality (Little, 1988). This
-assumption is tenuous in the context of LCA. A non-parametric MCAR test
-may be more suitable (Jamshidian \& Jalal, 2010). For this tutorial, the
-lead author has implemented this test in the \texttt{mice} package as
-\texttt{mice::mcar()}. Our recommendation is to inspect and report the
-proportion of missingness per variable (e.g., using
-\texttt{tidySEM::descriptives()}), and report the \texttt{mice::mcar()} test. As FIML
-estimation assumes that missingness is either MCAR or MAR, one would
-proceed with FIML regardless of the outcome.
-
-One minor concern is that \texttt{tidySEM} uses a different algorithm to
-determine starting values for complete data (K-means) and for missing
-data (hierarchical clustering). This should not be a problem, however,
-as the global optimizer used to estimate the model should be independent
-of starting values (\textbf{coranaMinimizingMultimodalFunctions1987?}). A second
-concern is that FIML is not suitable for cases with complete missing
-data. If such cases must be included, their values must be imputed
-first.
+(Van De Schoot et al., 2017).
+While we concur that investigating missingness is due diligence,
+it is important to emphasize that missingness is adequately handled by default in \texttt{OpenMx}, which is the computational backend of \texttt{tidySEM}.
+Its Full Information Maximum Likelihood (FIML) estimator makes use of all available
+information without imputing missing values, by estimating the model based on all
+available information for each subject {[}C. Enders and Bandalos (2001); enders\_applied\_2022{]}.
+FIML is a best-practice solution for handling missing data;
+at least on par with multiple imputation (Lee \& Shi, 2021).
+Multiple imputation is less suitable to LCA for two reasons. First,
+because LCA methods are often computationally intensive,
+and conducting them on several imputed datasets multiplies the load.
+Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population, so the imputation model could bias the results as it does not account for the expected heterogeneity.
+
+Three types of missingness have
+been distinguished in the literature (C. K. Enders, 2022; Rubin, 1976): random missingness (MCAR);
+missingness contingent on \emph{observed} variables that can be known and controlled for (MAR);
+and missingness related to unobserved variables,
+which can be neither known nor controlled for (MNAR).
+A so-called ``MCAR'' test is often reported (Jamshidian \& Jalal, 2010),
+but note that this name is somewhat misleading,
+as a significant test indicates that missingness is MAR,
+but a non-significant test statistic does not distinguish between MCAR or MNAR.
+As the classic MCAR test relies on the comparison of variances across groups with different patterns of missing data, it assumes normality (Little, 1988).
+This assumption is tenuous in the context of LCA.
+A non-parametric MCAR test may be more suitable (Jamshidian \& Jalal, 2010).
+For this tutorial,
+the lead author has implemented this test in the \texttt{mice} package as \texttt{mice::mcar()}.
+Our recommendation is to inspect and report the proportion of missingness per variable (e.g., using \texttt{tidySEM::descriptives()}),
+and report the \texttt{mice::mcar()} test, as way to identify variables that can help the parameter recovery instead of evidence of MCAR or MNAR.
+As FIML estimation assumes that missingness is either MCAR or MAR,
+one would proceed with FIML regardless of the outcome.
+
+One minor concern is that \texttt{tidySEM} uses a different algorithm to determine starting values for complete data (K-means) and for missing data (hierarchical clustering).
+This should not be a problem, however, as the global optimizer used to estimate the model should be independent of starting values (Corana, Marchesi, Martini, \& Ridella, 1987).
+A second concern is that FIML is not suitable for cases with complete missing data.
+If such cases must be included, their values must be imputed first.
 
@@
 
-In principle, LCA can be based on any parametric model. At present,
-\texttt{tidySEM} only facilitates estimation of models that can be specified in
-\texttt{lavaan} syntax. The available model parameters depend on the
-measurement level of the indicators. Continuous indicators can be
-parametrized in terms of means, variances, and covariances. LCGA and GMM
-have these same parameters, but with respect to the latent growth
-variables. Ordinal indicators can be parametrized differently; the
-default approach in \texttt{tidySEM} assumes that each ordinal indicator
-reflects an underlying standard normal distribution,
-\(N(\mu = 0, \sigma = 1)\). The parameters are ``thresholds'' that
-correspond to quantiles of this distribution, estimated based on the
-proportion of individuals in each of the response categories of the
-indicator variable. For example, a binary indicator has a single
-threshold that distinguishes the two response categories. If responses
-are distributed 50/50, then the corresponding threshold would be
-\(t_1 = 0.00\). If the responses are distributed 60/40, then the resulting
-threshold would be \(t_1 = 0.25\). All of the aforementioned parameters
-can be freely estimated, or constrained across classes, or fixed (e.g.,
-to zero).
-
-Prior literature emphasized the importance of considering alternative
-model specifications
-(\textbf{vandeschootGRoLTSChecklistGuidelinesReporting2017?}). What has remained
-underemphasized is that alternative model specifications should be
-approached differently for exploratory versus confirmatory analyses (see
-Use Cases). Most prior literature has emphasized exploratory LCA (e.g., Nylund, Asparouhov, \& Muthén, 2007). In this context, model specification typically
-consists of an exhaustive grid search along several model specifications
-and varying numbers of classes. A ``final'' model is then selected based
-on a combination of fit indices, significance tests, and
-interpretability (see Class Enumeration).
-
-Confirmatory LCA does not require such an extensive search over the
-model space. If hypothesized models have been specified a priori, and
-possibly even preregistered, it makes sense to focus on these models,
-rather than proceed in a purely data-driven manner. Nonetheless, the
-theoretical model specification should still be compared to a few
-alternatives because LCA lacks objective fit indices. The fit of a
-theoretical model can thus only be assessed by comparison to other
-candidates. A one-class model is a sensible benchmark to test whether
-the theoretical model fits better than a model that assumes a
-homogeneous population. Other candidates for comparison include
-competing theoretical models and those with a few more or less classes
-than the hypothesized number. If the theoretical model has much better
-fit than other models, for example, based on IC weights, this provides
-evidence for the theory. If another model fits substantially better,
-this provides evidence against the theory. If fit differences are small,
-the theoretical model may still be preferred, as theory is a relevant
-consideration in model selection (\textbf{jungIntroductionLatentClass2008?}). The
-fact that the theoretical model received little differential support
-should of course be discussed.
-
-A second consideration in model specification is the risk of overfitting
-(\textbf{hastieElementsStatisticalLearning2009?}). Overfitting occurs when a
-model has many parameters relative to the number of observations,
-causing it to capture idiosyncratic noise in the sample and limiting its
-generalizability to new samples. As explained before, the number of
-parameters in LCA scales with the number of estimated classes.
-Consequently, LCA with many classes have a potentially very high number
-of parameters. In Bayesian mixture modeling, it was observed that if the
-estimated number of classes exceeded the true number, the number of
-observations assigned to these classes tended towards zero as sample
-size increased (\textbf{rousseauAsymptoticBehaviourPosterior2011?}). Although it
-is not known whether frequentist LCA exhibits the same behavior, low
-class counts are still an indicator of potential overfitting. The
-section on Class Enumeration addresses ways to guard against
-overfitting.
-
-Models can be specified in \texttt{tidySEM} in different ways. The wrapper
-functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for
-continuous and ordinal indicators respectively, without needing to
-specify model syntax. The function \texttt{mx\_profiles()} has two arguments to
-free elements of the covariance matrix: \texttt{variances} and \texttt{covariances}.
-By default, \texttt{mx\_profiles()} fixes covariances to zero, and thus performs
-latent profile analysis. Finally, function \texttt{mx\_mixture()} accepts a more
-flexible \texttt{model} argument, which can be a character string specifying a
-model in \texttt{lavaan} syntax, which is translated to \texttt{OpenMx} models via the
-function \texttt{as\_ram()}, or a list of \texttt{OpenMx} models.
+In principle, LCA can be based on any parametric model.
+At present, \texttt{tidySEM} only facilitates estimation of models that can be specified in \texttt{lavaan} syntax.
+The available model parameters depend on the measurement level of the indicators.
+Continuous indicators can be parametrized in terms of means, variances, and covariances.
+LCGA and GMM have these same parameters,
+but with respect to the latent growth variables.
+Ordinal indicators are parametrized differently;
+the default approach in \texttt{tidySEM} assumes that each ordinal indicator reflects an underlying standard normal distribution, \(N(\mu = 0, \sigma = 1)\).
+The parameters are ``thresholds'' that correspond to quantiles of this distribution, estimated based on the proportion of individuals in each of the
+response categories of the indicator variable.
+For example, a binary
+indicator has a single threshold that distinguishes the two response
+categories. If responses are distributed 50/50, then the corresponding
+threshold would be \(t_1 = 0.00\). If the responses are distributed 60/40,
+then the resulting threshold would be \(t_1 = 0.25\).
+All of the aforementioned parameters can be freely estimated,
+or constrained across classes, or fixed (e.g., to zero).
+
+Prior literature emphasized the importance of considering alternative model specifications (Van De Schoot et al., 2017).
+What has remained underemphasized is that alternative model specifications should be approached differently for exploratory versus confirmatory analyses (see Use Cases).
+Most prior literature has emphasized exploratory LCA (e.g., Nylund, Asparouhov, \& Muthén, 2007).
+In this context, model specification typically consists of an exhaustive grid search along several model specifications and varying numbers of classes.
+A ``final'' model is then selected based on a combination of fit indices,
+significance tests,
+and interpretability (see Class Enumeration).
+
+Confirmatory LCA does not require such an extensive search over the model space.
+If hypothesized models have been specified a priori,
+and possibly even preregistered,
+it makes sense to focus on these models, rather than proceed in a purely data-driven manner.
+Nonetheless, the theoretical model specification should still be compared to a few alternatives because LCA lacks objective fit indices.
+The fit of a theoretical model can thus only be assessed by comparison to other candidates.
+A one-class model is a sensible benchmark
+to test whether the theoretical model fits better than a model that assumes a homogeneous population.
+Other candidates for comparison include competing theoretical models and those with a few more or less classes than the hypothesized number.
+If the theoretical model has much better fit than other models,
+for example, based on IC weights,
+this provides evidence for the theory.
+If another model fits substantially better,
+this provides evidence against the theory.
+If fit differences are small, the theoretical model may still be preferred, as theory is a relevant consideration in model selection (Jung \& Wickrama, 2008).
+The fact that the theoretical model received little differential support should of course be discussed.
+
+A second consideration in model specification is the risk of overfitting (Hastie, Tibshirani, \& Friedman, 2009).
+Overfitting occurs when a model has many parameters relative to the number of observations,
+causing it to capture idiosyncratic noise in the sample and limiting its generalizability to new samples.
+As explained before, the number of parameters in LCA scales with the number of estimated classes.
+Consequently, LCA with many classes have a potentially very high number of parameters.
+In Bayesian mixture modeling,
+it was observed that if the estimated number of classes exceeded the true number, the number of observations assigned to these classes tended towards zero as sample size increased (Rousseau \& Mengersen, 2011).
+Although it is not known whether frequentist LCA exhibits the same behavior,
+low class counts are still an indicator of potential overfitting.
+The section on Class Enumeration addresses ways to guard against overfitting.
+
+Models can be specified in \texttt{tidySEM} in different ways.
+The wrapper functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
+The function \texttt{mx\_profiles()} has two arguments to free elements of the covariance matrix: \texttt{variances} and \texttt{covariances}.
+By default, \texttt{mx\_profiles()} fixes covariances to zero, and thus performs latent profile analysis.
+Finally, function \texttt{mx\_mixture()} accepts a more flexible \texttt{model} argument, which can be a character string specifying a model in \texttt{lavaan} syntax,
+which is translated to \texttt{OpenMx} models via the function \texttt{as\_ram()},
+or a list of \texttt{OpenMx} models.
 
@@
 LCA parameters and model fit statistics can be estimated in different
-ways. The choice of estimator depends on the presence of missing values,
-sample size, number of indicators, and available computational resources
-(Weller, Bowen, \& Faubert, 2020). LCA in \texttt{tidySEM} uses maximum likelihood
-estimation (ML), which can be conceptually understood as follows:
-Imagine we are estimating two parameters, e.g.~the class-specific means
-\(\mu_c\) on a continuous indicator for two classes (ignoring the variance
-for now). ML attempts to find a combination of values for these two
-parameters that minimizes minus two times the log likelihood (\(-2LL\)) of
-the data.
-To understand the ML optimization procedure, imagine a three-dimensional
-landscape: The X and Y dimensions represent potential values of the
-class-specific means, \(X = \mu_1\) and \(Y = \mu_2\), and the Z-dimension
-is determined by \(Z = -2LL\). The optimizer must find the deepest
-``valley'' in this landscape, which reflects the combination of \(\mu_1\)
-and \(\mu_2\) that minimizes the \(-2LL\). The ML optimizer behaves somewhat
-like a marble dropped in this landscape. It is dropped at some random
-point in space, and rolls into the nearest valley (a combination of
-values for \(\mu_1\) and \(\mu_2\) that has a low \(-2LL\)). The challenge is
-that LCA models are often highly parametrized. This means that the
-landscape is not three-dimensional, but (very) high-dimensional, with
-many valleys and sparse data to provide information about which valley
-is deepest. If the marble rolls into one of these valleys, it will
-settle there and not climb out again. The estimator has ``converged''.
-
-A unique challenge in LCA is that not only model parameters, but also
-class membership must be estimated. Parameters in all classes are
-estimated from the same dataset, but for each class, observations are
-weighted by their probability of being in that class - which is in turn
-computed based on the class-specific parameters. The resulting
-chicken-and-egg problem is solved by the expectation-maximization (EM)
-algorithm: It alternates between estimating model parameters,
+ways.
+The choice of estimator depends on the presence of missing
+values, sample size, number of indicators, and available computational
+resources (Weller, Bowen, \& Faubert, 2020).
+LCA in \texttt{tidySEM} uses maximum likelihood estimation (ML), which can be conceptually understood as follows:
+Imagine we are estimating two
+parameters, e.g.~the class-specific means \(\mu_c\) on a continuous indicator for two classes
+(ignoring the variance for now).
+ML attempts to find a combination of values for these two parameters that minimizes minus two times the log likelihood (\(-2LL\)) of the data.
+To understand the ML optimization procedure,
+imagine a three-dimensional landscape:
+The X and Y dimensions represent potential values of the class-specific means, \(X = \mu_1\) and \(Y = \mu_2\), and the Z-dimension is determined by \(Z = -2LL\).
+The optimizer must find the
+deepest ``valley'' in this landscape,
+which reflects the combination of \(\mu_1\) and \(\mu_2\) that minimizes the \(-2LL\).
+The ML
+optimizer behaves somewhat like a marble dropped in this landscape. It
+is dropped at some random point in space, and rolls into the nearest valley (a combination of values for \(\mu_1\) and \(\mu_2\) that has a low \(-2LL\)).
+The challenge is that LCA models are often highly parametrized.
+This means that the landscape is not three-dimensional, but (very) high-dimensional,
+with many valleys and sparse data to provide information about which valley is deepest.
+If the marble rolls into one of these valleys,
+it will settle there and not climb out again.
+The estimator has ``converged''.
+
+A unique challenge in LCA is that not only model parameters,
+but also class membership must be estimated.
+Parameters in all classes are estimated from the same dataset,
+but for each class,
+observations are weighted by their probability of being in that class - which is in turn computed based on the class-specific parameters.
+The resulting chicken-and-egg problem is solved by the expectation-maximization (EM) algorithm:
+It alternates between estimating model parameters,
 calculating posterior class probabilities from those parameters,
-re-estimating model parameters using those new class probabilities as
-weights, and so on, until both have converged.
-
-A risk with this approach is that the optimizer might get stuck in a
-shallower valley (a ``local optimum''), and never discovers the deepest
-valley (the ``global optimum'', or best solution). One solution to this
-problem is to drop many marbles at random places, compare their final
-\(-2LL\) values, choose the solution with the lowest \(-2LL\), and make sure
-that several marbles replicated this solution. This is the ``random
-starts'' approach implemented in, e.g., Mplus.
+re-estimating model parameters using those new class probabilities as weights, and so on, until both have converged.
+
+A risk with this approach is that the optimizer might get stuck in a shallower valley (a ``local optimum''),
+and never discovers the deepest valley (the ``global optimum'', or best solution).
+One solution to this problem is to drop many marbles at random places,
+compare their final \(-2LL\) values,
+choose the solution
+with the lowest \(-2LL\), and make sure that several marbles replicated
+this solution.
+This is the ``random starts'' approach implemented in, e.g., Mplus.
 
@@
 computationally expensive to run this many replications. Moreover,
-because the algorithm begins with random starting values, it is
-inefficient because many marbles are likely to start far away from any
-valley. Additionally, if all sets of starting values are close to a
-specific local optimum, there is a risk that the global optimum is never
-found. An alternative solution that overcomes these challenges is to use
-a global optimizer, like simulated annealing (SA, \textbf{coranaMinimizingMultimodalFunctions1987?}). SA does not behave quite like
-the marbles of the random starts approach; instead, it iteratively
-considers some ``destination'' in the landscape, and compares its \(-2LL\)
-to the current one. If the destination \(-2LL\) is lower, the estimator
-moves there. However, if the destination \(-2LL\) is \emph{higher}, it still
-moves there occasionally, with a certain probability. Think of this as
-occasionally flicking the marble, to see if it rolls back into the
-current valley, or finds another, even deeper one. This property allows
-SA it to escape local optima and find the global optimum.
-
-Since SA is a global optimizer, it should be independent of starting
-values. With this in mind, it is efficient to start estimation from a
-``reasonable solution'', instead of a random starting point. For example,
-if we assume that the different classes are likely to have different
-mean values on the indicators, then a nonparametric clustering
-algorithm - like K-means or hierarchical clustering - can be used to
-determine these cluster centroids (see also \textbf{biernackiChoosingStartingValues2003?}). By default, \texttt{tidySEM} derives
-starting values using K-means clustering {[}REF{]}, computing the mixture
-model starting values by treating the K-means solution as multi-group
-model. Next, SA is used to find the global optimum solution. Finally, SA
-is followed up with a short run of the ML algorithm, as ML inherently
-produces an asymptotic covariance matrix for the parameters that can be
-used to compute standard errors. Note that these defaults can be
-manually overridden.
-
-In practice, models do not always converge. Reasons may include local
-optima, bad starting values, incorrect model specification (the model
-may be too complex for the data, or conversely, too simple to capture
-important patterns), or poor data quality (small data sets, or
-distributions that are difficult to model). Convergence problems may be
-indicated by errors and warnings from the estimator, or they may be
-suspected, for example, when results change if the analyses are
-reproduced.
-
-Nonconvergence may be due to bad model specification. Consider, for
-example, an LCA with multiple ordinal indicators. If certain
-combinations of responses never occur together, in general or within one
-of the classes, this can result in extreme conditional item
-probabilities (exactly 0 or 1). Such extreme values impede model
-convergence. This problem could be addressed by trying alternative model
-specifications (e.g., fewer classes), or by merging categories of the
-ordinal indicators with very few responses. Similarly problems occur
-with continuous indicators. For instance, consider a zero-inflated
-indicator and a two class LPA. To account for the excess zeroes, one
-class will have a mean of zero. It is difficult to estimate the standard
-deviation for this class, however, because all of its members have the
-same value. If standard deviations are constrained to be equal across
-classes, this model will likely not converge. This problem can be
-addressed by considering more appropriate model specifications (e.g.,
-free class variances).
-
-Nonconvergence due to bad starting values can be overcome by manually
-specifying starting values: Specify the argument \texttt{run\ =\ FALSE} in the
-call to \texttt{mx\_mixture()} and related functions. This returns the model
-with default starting values, which can be updated. Nonconvergence and
-local optima may also be overcome by re-estimating the model while
-permuting the starting values. The \texttt{OpenMx} package contains a family of
-functions for this purpose: \texttt{mxTryHard()} and \texttt{mxTryHardWideSearch()}
-for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal
-indicators. When poor data quality is to blame, the only solution is to
-collect more, or higher quality, data.
+because the algorithm begins with random starting values,
+it is inefficient because many marbles are likely to start far away from any valley.
+Additionally, if all sets of starting values are close to a specific local optimum,
+there is a risk that the global optimum is never found.
+An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing (SA, Corana et al., 1987).
+SA does not behave quite like the marbles of the random starts approach;
+instead, it iteratively considers some ``destination'' in the landscape,
+and compares its \(-2LL\) to the current one.
+If the destination \(-2LL\) is lower,
+the estimator moves there.
+However, if the destination \(-2LL\) is \emph{higher},
+it still moves there occasionally, with a certain probability.
+Think of this as occasionally flicking the marble,
+to see if it rolls back into the current valley, or finds another,
+even deeper one.
+This property allows SA it to escape local optima and find the global optimum.
+
+Since SA is a global optimizer, it should be independent of starting values.
+With this in mind, it is efficient to start estimation from a ``reasonable solution'', instead of a random starting point.
+For example, if we assume that the different classes are likely to have different mean values on the indicators,
+then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids (see also Biernacki, Celeux, \& Govaert, 2003).
+By default, \texttt{tidySEM} derives starting values using K-means clustering (Likas, Vlassis, \& J. Verbeek, 2003),
+computing the mixture model starting values by treating the K-means solution as multi-group model.
+Next, SA is used to find the global optimum solution.
+Finally, SA is followed up with a short run of the ML algorithm,
+as ML inherently produces an asymptotic
+covariance matrix for the parameters that can be used to compute
+standard errors.
+Note that these defaults can be manually overridden.
+
+In practice, models do not always converge.
+Reasons may include local optima,
+bad starting values,
+incorrect model specification (the model may be too complex for the data, or conversely, too simple to capture important patterns),
+or poor data quality (small data sets, or distributions that are difficult to model).
+Convergence problems may be indicated by errors and warnings from the estimator,
+or they may be suspected, for example,
+when results change if the analyses are reproduced.
+
+Nonconvergence may be due to bad model specification.
+Consider, for example, an LCA with multiple ordinal indicators.
+If certain combinations of responses never occur together,
+in general or within one of the classes,
+this can result in extreme conditional item probabilities (exactly 0 or 1).
+Such extreme values impede model convergence.
+This problem could be addressed by trying alternative model specifications (e.g., fewer classes),
+or by merging categories of the ordinal indicators with very few responses.
+Similarly problems occur with continuous indicators.
+For instance, consider a zero-inflated indicator and a two class LPA.
+To account for the excess zeroes, one class will have a mean of zero.
+It is difficult to estimate the standard deviation for this class, however, because all of its members have the same value.
+If standard deviations are constrained to be equal across classes,
+this model will likely not converge.
+This problem can be addressed by considering more appropriate model specifications (e.g., free class variances).
+
+Nonconvergence due to bad starting values
+can be overcome by manually specifying starting values:
+Specify the argument \texttt{run\ =\ FALSE} in the call to \texttt{mx\_mixture()} and related functions.
+This returns the model with default starting values, which can be updated.
+Nonconvergence and local optima may also be overcome by re-estimating the model while permuting the starting values.
+The \texttt{OpenMx} package contains a family of functions for this purpose: \texttt{mxTryHard()} and \texttt{mxTryHardWideSearch()} for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal indicators.
+When poor data quality is to blame, the only solution is to collect more, or higher quality, data.
 
@@
 
-Class enumeration refers to the process of determining the appropriate
-number of classes. It is approached differently for exploratory versus
-confirmatory LCA. In exploratory LCA, the set of models typically
-consists of a sequence from 1 to the maximum number of classes \(K\). The
-number \(K\) may be chosen a-priori or dictated by the data. It is
-sensible to limit \(K\) to the maximum number of classes that result in
-valid and interpretable solutions (also see Classification Diagnostics).
-Above a certain number of classes, convergence problems typically arise,
-or the proportion of cases assigned to the smallest classes may be low
-relative to the number of class-specific parameters. In Bayesian LCA, it
-has been observed that, when the estimated number of classes exceeds the
-true number, the proportion of cases assigned to the excessive classes
-converges to zero as the sample size increases
-(\textbf{rousseauAsymptoticBehaviourPosterior2011?}). It is not known whether
-this also happens in frequentist LCA; nonetheless, near-zero class
-memberships should be taken into account when determining the maximum
-\(K\).
-
-In confirmatory LCA, the set of models with different numbers of classes
-to be compared is typically smaller. Its purpose is to provide context
-for the relative fit of the theoretical model, or to compare multiple
-theoretical models.
-
-In class enumeration, the set of models is compared on several
-statistical and substantive criteria. In exploratory LCA, these criteria
-may be decisive in selecting a final model; in the confirmatory case,
-the burden of proof is reversed: there must be convincing evidence
-against the theoretical model in order to reject it. Statistical
-criteria to consider in class enumeration include model fit indices and
+Class enumeration refers to the process of determining the appropriate number of classes.
+It is approached differently for exploratory versus confirmatory LCA.
+In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes \(k\).
+The number \(k\) may be chosen a-priori or dictated by the data.
+It is sensible to limit \(k\) to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
+Above a certain number of classes,
+convergence problems typically arise,
+or the proportion of cases assigned to the smallest classes may be low relative to the number of class-specific parameters.
+In Bayesian LCA, it has been observed that, when the estimated number of classes exceeds the true number,
+the proportion of cases assigned to the excessive classes converges to zero as the sample size increases (Rousseau \& Mengersen, 2011).
+It is not known whether this also happens in frequentist LCA;
+nonetheless, near-zero class memberships should be taken into account when determining the maximum \(k\).
+
+In confirmatory LCA,
+the set of models with different numbers of classes to be compared is typically smaller.
+Its purpose is to provide context for the relative fit of the theoretical model, or to compare multiple theoretical models.
+
+In class enumeration, the set of models is compared on several statistical and substantive criteria.
+In exploratory LCA, these criteria may be decisive in selecting a final model;
+in the confirmatory case, the burden of proof is reversed: there must be convincing evidence against the theoretical model in order to reject it.
+Statistical criteria to consider in class enumeration include model fit indices and
 likelihood ratio tests (Masyn, 2013; Weller et al., 2020).
-Substantive criteria include convergence and model identification, class
-separability, and interpretability. We address each of these criteria
-below.
+Substantive criteria include convergence and model identification,
+class separability,
+and interpretability.
+We address each of these criteria below.
 
@@
 
-As there are no absolute fit indices for LCA, information criteria are
-typically used to assess relative fit. Information criteria are computed
-from the \(-2LL\), which is the most basic fit measure, testing the
-overall model fit if the model properly represents the data. Information
-criteria add a penalty for model complexity to the \(-2LL\), and thus
-balance fit and complexity. The lower the value of an information
-criterion, the better the overall fit of the model. The original
-information criterion is the Akaike Information Criterion (AIC),
-computed as \(-2LL+2p\), where \(p\) is the number of parameters. The
-Bayesian Information Criterion (BIC) multiplies the penalty for
-complexity with a function of the sample size, \(p \ln(n)\), such that
-complex models are penalized more in smaller samples. Finally, the
-so-called \emph{sample size adjusted} BIC (saBIC) implements a different
-ad-hoc penalty for sample size, \(\ln(\frac{n + 2}{24})\). Although a
-simulation study showed that the saBIC performed well for class
-enumeration, its penalty is not based on theory {[}REF{]}. BIC also performs
-well and is theoretically substantiated; as such, it may be the most
-appropriate information criterion to use for model comparison
-(Masyn, 2013; Nylund-Gibson \& Choi, 2018). All three ICs are
-available in the output of \texttt{table\_fit()}.
+As there are no absolute fit indices for LCA,
+information criteria are typically used to assess relative fit.
+Information criteria are computed from the \(-2LL\),
+which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
+Information criteria add a penalty for model complexity to the \(-2LL\), and thus balance fit and complexity.
+The lower
+the value of an information criterion, the better the overall fit of the
+model.
+The original information criterion is the Akaike Information Criterion (AIC),
+computed as \(-2LL+2p\), where \(p\) is the number of parameters.
+The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
+such that complex models are penalized more in smaller samples.
+Finally,
+the so-called \emph{sample size adjusted} BIC (saBIC) implements a different ad-hoc penalty for sample size,
+\(\ln(\frac{n + 2}{24})\).
+Although simulation studies have showed that the saBIC performed well for class enumeration (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013),
+its penalty is not based on theory.
+BIC also performs well and is theoretically substantiated;
+as such, it may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+All three ICs are available in the output of \texttt{table\_fit()}.
 
 Information criteria may occasionally contradict each other, so it is
-important to identify a suitable strategy to reconcile them. One option
-is to select a specific fit index, such as the BIC, before seeing the
-results. This choice can be preregistered to avoid concerns about cherry
-picking. Another option is to always prefer the most parsimonious model
-that has best fit according to any of the available fit indices. Other
-approach is to select several competing model based on the ICs, and
-evaluate the theoretical intepretation. Yet another option is to
-incorporate information from multiple fit indices using the analytic
-hierarchy process (Akogul \& Erisoglu, 2016), which is implemented in
-\texttt{tidyLPA} (\textbf{rosenbergTidyLPAPackageEasily2018?}). Finally, one can use a
-scree plot to determine the inflection point at which additional classes
-contribute little to further decreases of the ICs
-(Nylund-Gibson \& Choi, 2018). The preferred number of classes is at the
-inflection point where the slope of the curve is clearly leveling off.
-Adding this many classes results in substantial decreases of the IC, and
-adding further classes results in relatively small decreases.
-
-Given a specific IC, it is also possible to calculate so-called
-``IC-weights'', which indicate the relative support the data provide for
-each model in the set (\textbf{wagenmakersAICModelSelection2004?}). This is
-particularly useful for confirmatory LCA, as it allows one to quantify
-the support for the theoretical model. IC-weights are calculated by
-calling \texttt{ic\_weights()} on the model fit table. Note, however, that
-IC-weights will (strongly) favor the model with the lowest IC value; the
-scree plot approach above may be more suitable to select the most
-parsimonious model with adequate fit.
+important to identify a suitable strategy to reconcile them.
+One option is to select a specific fit index, such as the BIC,
+before seeing the results.
+This choice can be preregistered to avoid concerns about cherry picking.
+Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
+Other approach is to select several competing model based on the ICs, and evaluate the theoretical interpretation.
+Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016),
+which is implemented in \texttt{tidyLPA} (Rosenberg, Beymer, Anderson, Van Lissa, \& Schmidt, 2018).
+Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs (Nylund-Gibson \& Choi, 2018).
+The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
+Adding this many classes results in substantial decreases of the IC,
+and adding further classes results in relatively small decreases.
+
+Given a specific IC,
+it is also possible to calculate so-called ``IC-weights'',
+which indicate the relative support the data provide for each model in the set (Wagenmakers \& Farrell, 2004).
+This is particularly useful for confirmatory LCA,
+as it allows one to quantify the support for the theoretical model.
+IC-weights are calculated by calling \texttt{ic\_weights()} on the model fit table.
+Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
+the scree plot approach above may be more suitable to select the most parsimonious model with adequate fit.
 
@@
 
-When two models are nested, the significance of the difference in fit
-between these models can be tested using a Likelihood Ratio test (LR).
-The test statistic is computed as the -2 times the log of the ratio of
-likelihoods, and its degrees of freedom are equal to the difference in
-the degrees of freedom of the two compared models. Since a \(k-1\)-class
-model is nested in a \(k\)-class model, it is possible to test the
-difference in model fit using a LR test. One challenge is that the LR
-test statistic is not asymptotically \(\chi^2\) distributed for LCA. The
-so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem
-by applying an ad-hoc correction to the LR {[}REF lo 2001{]}. This test is
-implemented in \texttt{tidySEM}, and is part of the default output of
-\texttt{table\_fit()}, comparing subsequent rows of the table.
-
-Another solution is to derive an empirical sampling distribution for the
-LR difference. The so-called bootstrapped Likelihood Ratio Test (BLRT)
-does so by simulating data from the \(k-1\)-class model \(B\) times (e.g.,
-1000), and fitting the \(k\)- and \(k-1\)-class models to each simulated
-dataset {[}REF Nylund et al., 2007{]}. The p-value of this test is the
-proportion of \(B\) in which the LR of the models fit to simulated data
-exceeds the LR observed in the real data. Simulation studies suggest
-that the BLRT has greater statistical power than the LRT, although both
-performed comparably well at detecting the true number of classes {[}REF
-Nylund 2007, REF Tein 2013{]}. A crucial limitation of the BLRT is that it
-is very computationally expensive. With this in mind, \texttt{table\_fit()} does
-not provide the BLRT by default. Specific models can be compared using
-the function \texttt{BLRT()}.
-
-Anecdotal experience as a statistical co-author on LCA studies suggests
-that both these tests tend to be liberal, returning significant p-values
-for models with more classes even when other indicators suggest that
-those models are overfitted (e.g., near-zero class counts). Other
-authors have reported similar experiences
-(\textbf{sinhaPractitionerGuideLatent2021?}), which matches the differences
-between LRT and ICs from information theory as LRT tests and compare the
-models based on in sample predictive accuracy, and ICs evaluates and
-compare the models out-of-sample predictive accuracy which helps to
-control for overfitting (McElreath, 2020). This suggests
-that, like existing ICs, LR tests might not adequately balance fit and
-complexity in LCA. Better solutions for class enumeration are thus an
-important target for future research.
-
-A promising novel solution is the so-called \emph{lazy bootstrap}, which is
-much more efficient than the BLRT because it is based on sample
-statistics of model-implied data generated from the \(k-1\) and \(k\)-class
-models, rather than on fit statistics of LCA models fit to those data
-(Kollenburg, Mulder, \& Vermunt, 2018). This simulation-based approach to inference
-is similar to the posterior predictive p-value in Bayesian statistics,
-which describes how well each model is able to reproduce the observed
-correlation matrix (see McElreath, 2020). One version of the
-lazy bootstrap is already implemented in the commercial program
-latentGOLD {[}REF{]}. At the time of writing, we are developing a more
-general simulation-based likelihood ratio test (SBLRT) for class
-enumeration. Future versions of \texttt{tidySEM} will likely incorporate this
-test.
+When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
+The test statistic is computed as the -2 times the log of the ratio of likelihoods,
+and its degrees of freedom are equal to the difference in the degrees of freedom of the two compared models.
+Since a \(k-1\)-class model is nested in a \(k\)-class model,
+it is possible to test the difference in model fit using a LR test.
+One challenge is that the LR test statistic is not asymptotically \(\chi^2\) distributed for LCA.
+The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR (Lo, Mendell, \& Rubin, 2001).
+This test is implemented in \texttt{tidySEM}, and is part of the default output of \texttt{table\_fit()}, comparing subsequent rows of the table.
+
+Another solution is to derive an empirical sampling distribution for the LR difference.
+The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the \(k-1\)-class model \(B\) times (e.g., 1000),
+and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset (Nylund et al., 2007).
+The p-value of this test is the proportion of \(B\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
+Simulation studies suggest that the BLRT has greater statistical power than the LRT,
+although both performed comparably well at detecting the true number of classes (Nylund et al., 2007; Tein et al., 2013).
+A crucial limitation of the BLRT is that it is very computationally expensive.
+With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
+Specific models can be compared using the function \texttt{BLRT()}.
+
+Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
+returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
+Other authors have reported similar experiences (Sinha, Calfee, \& Delucchi, 2021), which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting (McElreath, 2020).
+This suggests that, like existing ICs,
+LR tests might not adequately balance fit and complexity in LCA.
+Better solutions for class enumeration are thus an important target for future research.
+
+A promising novel solution is the so-called \emph{lazy bootstrap},
+which is much more efficient than the BLRT because it is based on sample statistics of model-implied data generated from the \(k-1\) and \(k\)-class models,
+rather than on fit statistics of LCA models fit to those data (Kollenburg, Mulder, \& Vermunt, 2018).
+This simulation-based approach to inference is similar to the posterior predictive p-value in Bayesian statistics,
+which describes how well each model is able to reproduce the observed correlation matrix (see McElreath, 2020).
+One version of the lazy bootstrap is already implemented in the commercial program \emph{Latent GOLD} (Vermunt \& Magidson, 2021).
+At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
+Future versions of \texttt{tidySEM} will likely incorporate this test.
 
@@
 
-Classification diagnostics assess whether the classes identified by LCA
-are distinct and clearly separable. These diagnostics are not model fit
-indices and should not be used for class enumeration, because a model
-can fit the data well and still show poor latent class separation
-(Masyn, 2013). However, classification diagnostics can be used to
-eliminate models: If one analysis goal is to interpret properties of
-(members of) classes, it makes sense to only consider models that
-classify well (Nylund-Gibson \& Choi, 2018). Classification diagnostics
-should thus not be used for model selection, but they can help restrict
-the search space of acceptable models.
-
-All classification diagnostics are derived from the posterior
-classification probabilities: an \(n \times k\) matrix called \(P\), with
-the probability that every individual \(n\) belongs to latent class \(k\).
-When classification accuracy is high, and classes are distinct, each
-individual's posterior class probabilities are high for one class, and
-low for the remaining classes. The individual classification
-probabilities \(P\) are obtained by running
-\texttt{class\_prob(res,\ type\ =\ "individual")}. Note that one column is added to
-the matrix, with each individual's most likely class membership \(M\). \(M\)
-is obtained by assigning each individual to the class corresponding to
-their maximum posterior class probability.
-Predicted class
-membership \(M\) is often of interest to researchers who wish to carry out
-follow up analyses. Note, however, that follow-up analyses should take
-classification error into account. The function \texttt{BCH()} does this
-automatically.
-
-The matrix \(P\) can be summarized in different way. For example, the
-estimated frequency table of the latent categorical variable can be
-obtained by summing the posterior probabilities by class (i.e., the
-column sums of \(P\)). This frequency table \(F_P\) is obtained by calling
-\texttt{class\_prob(res,\ type\ =\ "sum.posterior")}. These numbers are fractional
-because this table takes classification error into account. Individuals
-can contribute partially to multiple classes.
-
-Second, most likely class membership based on the highest posterior
-class probability, \(F_M\), is obtained by running
-\texttt{class\_prob(res,\ "sum.mostlikely")}. This is the frequency table of the
-variable \(M\). Unlike \(F_P\), \(F_M\) ignores classification error. If every
-observation is classified with perfect accuracy, \(F_M\) and \(F_P\) are
-identical. Part of the output of \texttt{table\_fit()} for LCA analyses are the
-minimum and maximum percentage of the sample assigned to a particular
-class, \texttt{n\_min} and \texttt{n\_max}, derived from \(F_M\). This is reported by
-default because the smallest class proportion, \texttt{n\_min}, is of particular
-interest. When the size of the smallest class is not much larger than
-the number of parameters per class, proceed with caution: The model
-might not be locally identified in that class
-(\textbf{depaoliMixtureClassRecovery2013?}). Existing rules of thumb have focused
-on smallest class size in terms of percentage of the total sample size.
-This is not particularly meaningful, however, as the primary concern is
-that small classes may comprise too few observations to accurately
-estimate class parameters. Thus, a better rule of thumb might be to
-ascertain that the smallest class has many more observations than the
-average number of parameters per class. Another good practice is to
-ensure that the smallest class has a unique and meaningful substantive
-interpretation. If the smallest class does not comprise at least a few
-observations per parameter, or is redundant with another larger class, a
-simpler solution is preferable.
-
-To assess classification accuracy, we turn to the third diagnostic
-table: The average posterior probabilities by most likely class
-membership, \(P_{p_{(M = m)}}\),
-\texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}: for participants assigned to
-one particular class (rows) according to \(M\), this table gives the mean
-probability of being assigned to any class (columns). This table is
-obtained by calculating the column means of \(P\) for the \(k\) subsets of
-observations with most likely class of \(M = 1 \ldots k\). The diagonal
-represents the average predictive probability of a the subjects related
-to their most likely class, indicating the certainty of class assignment
-for the subjects. Some guidelines have being suggested for this
-diagonal, like for them to be higher than 0.7 (\textbf{masyn2013?}), but we
-suggest researchers to be cautions about these guidelines as they cannot
-be generalized and should not be use as strict cutoffs to defined ``good''
-or ``bad'' classification.
-
-The table of classification probabilities for the most likely latent
-class \(M\) by latent class, \(P_{p_{(M = m | C = c)}}\) is similar to the
-aforementioned table \(P_{p_{(M = m)}}\), except that it accounts for
-classification uncertainty. If \(C\) is the true class of an observation,
-which is imperfectly measured, and \(M\) is the most likely class, then
-this table shows the probability of an observation being assigned to
-class \(m\) in \(M\), given that its true class is \(c\) in \(C\), or
-\(p(M=m|C=c)\). Values on the diagonal can thus be conceptualized as the
-reliability with which each level of the categorical latent variable is
-measured by \(M\), and off-diagonal elements can be conceptualized as
-measurement error. The minimum and maximum of the diagonal of this table
-are given by default as part of the output of \texttt{table\_fit()} for LCA
-analyses (\texttt{prob\_min} and \texttt{prob\_max}). These probabilities are of
-particular interest because they conveniently summarize classification
-accuracy per class. Both probabilities should be high, as this means
-that all classes are reliably classified. If the minimum probability is
-low, there is at least one class with poor classification accuracy.
-
-Finally, class separability can be summarized in a single statistic: the
-entropy criterion {[}REF{]}.
-In
-physics, high entropy refers to maximum randomness or uncertainty, and
-low entropy corresponds to a strict arrangement of the units of study.
-Applied to LCA, high entropy means that classes are completely
-indistinguishable; every individual has equal probability of being in
-any class. Low entropy, by contrast, means that classes are fully
-separated: every individual has a near-one probability of being in one
-class only, and near-zero probability of being in any other class.
-Crucially, in \texttt{tidySEM}, the entropy criterion is defined as 1-entropy
-(a convention originating in Mplus). Its interpretation is reversed; 0
-means the model classification is no better than chance, and 1 means
-perfect classification. There are no rules of thumb for entropy, because
-entropy is also sensitive to the sample size (entropy tends to be lower
-in larger samples), number of latent classes (entropy tends to be lower
-for higher number of classes), and number of indicators. Like other
-classification diagnostics, entropy is not a model fit measure and
-should not be used for model selection (Masyn, 2013). However, it
-can be used to disqualify certain solutions if the goal is to identify
-clearly distinct classes.
+Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
+These diagnostics are not model fit indices and should not be used for class enumeration,
+because a model can fit the data well and still show poor latent class separation (Masyn, 2013).
+However, classification diagnostics can be used to eliminate models:
+If one analysis goal is to interpret properties of (members of) classes,
+it makes sense to only consider models that classify well (Nylund-Gibson \& Choi, 2018).
+Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.
+
+All classification diagnostics are derived from the posterior classification probabilities: an \(n \times k\) matrix called \(P\), with the probability that every individual \(n\) belongs to latent class \(k\).
+When classification accuracy is high, and classes are distinct,
+each individual's posterior class probabilities are high for one class, and low for the remaining classes.
+The individual classification probabilities \(P\) are obtained by running \texttt{class\_prob(res,\ type\ =\ "individual")}.
+Note that one column is added to the matrix, with each individual's most likely class membership \(M\).
+\(M\) is obtained by assigning each individual to the class corresponding to their maximum posterior class probability.
+Predicted class membership \(M\) is often of interest to researchers who wish to carry out follow up analyses.
+Note, however, that follow-up analyses should take classification error into account.
+The function \texttt{BCH()} does this automatically.
+
+The matrix \(P\) can be summarized in different way.
+For example, the estimated frequency table of the latent categorical variable can be obtained by summing the posterior probabilities by class (i.e., the column sums of \(P\)).
+This frequency table \(F_P\) is obtained by calling \texttt{class\_prob(res,\ type\ =\ "sum.posterior"}.
+These numbers are fractional because this table takes classification error into account.
+Individuals can contribute partially to multiple classes.
+
+Second, most likely class membership based on the highest posterior class probability, \(F_M\), is obtained by running \texttt{class\_prob(res,\ "sum.mostlikely")}.
+This is the frequency table of the variable \(M\).
+Unlike \(F_P\), \(F_M\) ignores classification error.
+If every observation is classified with perfect accuracy,
+\(F_M\) and \(F_P\) are identical.
+Part of the output of \texttt{table\_fit()} for LCA analyses
+are the minimum and maximum percentage of the sample assigned to a particular class, \texttt{n\_min} and \texttt{n\_max}, derived from \(F_M\).
+This is reported by default because the smallest class proportion, \texttt{n\_min}, is of particular interest.
+When the size of the smallest class is not much larger than the number of parameters per class,
+proceed with caution:
+The model might not be locally identified in that class (Depaoli, 2013).
+Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
+This is not particularly meaningful,
+however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
+Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
+Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
+If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.
+
+To assess classification accuracy, we turn to the third diagnostic table:
+The average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\), \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}:
+for participants assigned to one particular class (rows) according to \(M\),
+this table gives the mean probability of being assigned to any class (columns).
+This table is obtained by calculating the column means of \(P\) for the
+\(k\) subsets of observations with most likely class of \(M = 1 \ldots k\).
+The diagonal represents the average predictive probability of a the subjects related to their most likely class, indicating the certainty of class assignment for the subjects.
+Some guidelines have being suggested for this diagonal, like for them to be higher than 0.7 (Masyn, 2013), but we suggest researchers to be cautions about these guidelines as they cannot be generalized and should not be use as strict cutoffs as evidence ``good'' or ``bad'' classification.
+
+The table of classification probabilities for the most likely latent class \(M\) by latent class, \(P_{p_{(M = m | C = c)}}\) is similar to the aforementioned table \(P_{p_{(M = m)}}\),
+except that it accounts for classification uncertainty.
+If \(C\) is the true class of an observation, which is imperfectly measured,
+and \(M\) is the most likely class,
+then this table shows the probability of an observation being assigned to class \(m\) in \(M\), given that its true class is \(c\) in \(C\), or \(p(M=m|C=c)\).
+Values on the diagonal can thus be conceptualized as the reliability with which each level of the categorical latent variable is measured by \(M\),
+and off-diagonal elements can be conceptualized as measurement error.
+The minimum and maximum of the diagonal of this table are given by default as part of the output of \texttt{table\_fit()} for LCA analyses (\texttt{prob\_min} and \texttt{prob\_max}).
+These probabilities are of particular interest because they conveniently summarize classification accuracy per class.
+Both probabilities should be high, as this means that all classes are reliably classified.
+If the minimum probability is low,
+there is at least one class with poor classification accuracy.
+
+Finally, class separability can be summarized in a single statistic:
+the entropy criterion (Celeux \& Soromenho, 1996).
+Celeux, G., \& Soromenho, G. (1996).
+In physics, high entropy refers to maximum randomness or uncertainty,
+and low entropy corresponds to a strict arrangement of the units of study.
+Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
+Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
+Crucially, in \texttt{tidySEM}, the entropy criterion is defined as 1-entropy (a convention originating in \emph{Mplus}).
+Its interpretation is reversed;
+0 means the model classification is no better than chance,
+and 1 means perfect classification.
+There are no rules of thumb for entropy,
+because entropy is also sensitive to the sample size (entropy tends to be lower in larger samples),
+number of latent classes (entropy tends to be lower for higher number of classes),
+and number of indicators.
+Like other classification diagnostics,
+entropy is not a model fit measure and should not be used for model selection (Masyn, 2013).
+However, it can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
 
@@
 
-Class names should be chosen to accurately reflect theoretically
-relevant characteristics of group membership. These class names should
-aid the interpretation of the class solutions and serve as a shorthand
-throughout the manuscript. For example, one study used a dual-trajectory
-LCGA of cognitive and affective empathy development with gender as a
-covariate. Although this model has dozens of parameters, the resulting
-classes were labeled low, average, and high empathy because level
-differences were the most distinguishing feature. However, overly
-simplistic and general class names can be misleading, a phenomenon known
-as the \emph{naming fallacy} (Weller et al., 2020). For example, it would be
-misleading to call one group ``low empathy'' and another group ``medium
-empathy'' if the empathy differences were non-significant, or if other
-differences, like the effect of the covariate gender, were more
-substantial than level differences. One way to avoid the naming fallacy
-is comprehensively reporting model parameters, and explicating that
-class names are just a shorthand.
-
-When labeling classes, it is important to note that LCA models are
-identified up to the class order.
-This
-means that the order of classes is arbitrary and may change if the
-analysis is replicated. A simple solution to ensure replicable results
-is setting a random seed before fitting the model, using \texttt{set.seed()}.
-This will ensure that the classes stay in the same order upon
-replication, but that may not be the preferred order for, e.g.,
-interpreting the results. To address this problem, the function
-\texttt{mx\_switch\_labels()} can be used. By default, the function orders
-classes from largest to smallest class count - but classes can also be
-ordered by values of a specific parameter (e.g., the mean intercept in a
-latent class growth model). Note that label switching is a much bigger
-problem when using Bayesian estimation, where it can occur during model
-estimation. This is beyond the scope of the present paper.
+Class names should be chosen to accurately reflect theoretically relevant characteristics of group membership.
+These class names should aid the interpretation of the class solutions and serve as a shorthand throughout the manuscript.
+For example, one study used a dual-trajectory LCGA of cognitive and affective empathy development with gender as a covariate.
+Although this model has dozens of parameters,
+the resulting classes were labeled low, average, and high empathy because level differences were the most distinguishing feature.
+However, overly simplistic and general class names can be misleading, a phenomenon known as the \emph{naming fallacy} (Weller et al., 2020).
+For example, it would be misleading to call one group ``low empathy'' and another group ``medium empathy'' if the empathy differences were non-significant,
+or if other differences, like the effect of the covariate gender, were more substantial than level differences.
+One way to avoid the naming fallacy is comprehensively reporting model parameters,
+and explicating that class names are just a shorthand.
+
+When labeling classes, it is important to note that LCA models are identified up to the class order.
+This means that the order of classes is arbitrary and may change if the analysis is replicated.
+A simple solution to ensure replicable results is setting a random seed before fitting the model, using \texttt{set.seed()}.
+This will ensure that the classes stay in the same order upon replication,
+but that may not be the preferred order for, e.g., interpreting the results.
+To address this problem, the function \texttt{mx\_switch\_labels()} can be used.
+By default, the function orders classes from largest to smallest class count - but classes can also be ordered by values of a specific parameter (e.g., the mean intercept in a latent class growth model).
+Note that label switching is a much bigger problem when using Bayesian estimation,
+where it can occur during model estimation.
+This is beyond the scope of the present paper.
 
@@
 
-LCA is rarely performed in isolation; often, researchers wish to relate
-class membership to auxiliary variables, or even auxiliary models. It is
-important to keep in mind that model-implied class membership is subject
-to classification error. Treating class membership as an observed
-variable in follow-up analyses disregards this error. The results will
-be biased, except when class separation is very high - as indicated by
-high (near 1) entropy and minimum posterior classification probability.
+LCA is rarely performed in isolation;
+often, researchers wish to relate class membership to auxiliary variables, or even auxiliary models.
+It is important to keep in mind that model-implied class membership is subject to classification error.
+Treating class membership as an observed variable in follow-up analyses disregards this error.
+The results will be biased,
+except when class separation is very high - as indicated by high (near 1) entropy and minimum posterior classification probability.
 It is therefore important to account for classification error.
 
-There are several different ways to do follow-up analyses in latent
-class analysis (LCA) while accounting for classification error. At
-present, the so-called BCH method is considered one of the best
-approaches {[}after its creators, REF{]}. It has been implemented in
-\texttt{tidySEM}. The advantage of the BCH method is that it not only allows
-for the comparison of individual variables, but also entire auxiliary
-models across latent classes.
+There are several different ways to do follow-up analyses in latent class analysis (LCA) while accounting for classification error.
+At present, the so-called BCH method is considered one of the best approaches (Bolck, Croon, \& Hagenaars, 2004).
+It has been implemented in \texttt{tidySEM}.
+The advantage of the BCH method is that it not only allows for the comparison of individual variables, but also entire auxiliary models across latent classes.
 
@@
 Among studies using LCA, reporting practices vary significantly
-(Weller et al., 2020). Various authors have tried to improve and
+(Weller et al., 2020).
+Various authors have tried to improve and
 standardize ways of reporting LCA (e.g. Masyn, 2013; Weller et al., 2020), but more work is needed.
@@
 
-Journals and funding bodies increasingly require research to be open and
-reproducible. For instance, many journals (and the APA) have adopted the
-TOP guidelines, which recommend comprehensive citation of literature,
-software, and data sources; sharing data and analysis code required to
-exactly reproduce analyses, and pre-registering study plans prior to
-data collection or analysis (\textbf{nosekTransparencyOpennessPromotion2016a?}).
-A key advantage of using open source software for LCA is that, compared
-to commercial software, it enables much greater adherence to open
-science principles. Sharing code has limited utility if it can only be
-evaluated in commercial software, as this restricts the ability to
-reproduce analyses to license holders, and prevents auditing research
-because the underlying source code is proprietary. The \texttt{tidySEM} package
-enables performing LCA in free open source software, thus increasing the
-utility of shared code.
-
-Adopting FOSS makes it possible to conduct a fully reproducible LCA
-analysis compliant with best practices in open science. The Workflow for
-Open Reproducible Code in Science describes how to create a fully
-reproducible research archive, and the \texttt{worcs} R-package automates many
-of the steps involved (\textbf{vanlissaWORCSWorkflowOpen2021a?}).
-This workflow is based on three principles: First, the paper (or
-analysis section) is written as a \emph{dynamic document} that contains both
-prose and analysis code. Results are automatically rendered as text,
-tables, and plots. Second, the research archive is \emph{version controlled},
-which means that the entire historical record of the project is
-retained. This version controlled archive can be made public. Third,
-\emph{dependency management} ensures that all software used is documented,
-including its version number and provenance. Without dependency
-management, code might not work (for example, because it is not clear
-where certain functions come from), or its results may change when some
-dependencies are updated. By combining \texttt{tidySEM} and \texttt{worcs}, it is
-possible to conduct a fully reproducible LCA analysis compliant with
-best practices in open science.
-
-An additional argument for sharing data in the context of LCA is that
-reproducing LCA analyses requires individual participant data, not just
-sufficient statistics like many other methods (i.e., a correlation or
-covariance matrix). If the original data cannot be shared, sharing
-synthetic data at least enables reproducing and reusing analysis code.
-The \texttt{worcs} package provides a generic, non-parametric method to create
-a synthetic copy of data when data cannot be shared. It is also possible
-to synthesize data directly from the LCA model using \texttt{mxGenerateData()},
+Journals and funding bodies increasingly require research to be open and reproducible.
+For instance, many journals (and the APA) have adopted the TOP guidelines, which recommend comprehensive citation of literature, software, and data sources;
+sharing data and analysis code required to exactly reproduce analyses,
+and pre-registering study plans prior to data collection or analysis (Nosek et al., 2016).
+A key advantage of using open source software for LCA is that, compared to commercial software, it enables much greater adherence to open science principles.
+Sharing code has limited utility if it can only be evaluated in commercial software,
+as this restricts the ability to reproduce analyses to license holders,
+and prevents auditing research because the underlying source code is proprietary.
+The \texttt{tidySEM} package enables performing LCA in free open source software,
+thus increasing the utility of shared code.
+
+Adopting FOSS makes it possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
+The Workflow for Open Reproducible Code in Science describes how to create a fully reproducible research archive,
+and the \texttt{worcs} R-package automates many of the steps involved (Van Lissa et al., 2021).
+This workflow is based on three principles: First,
+the paper (or analysis section) is written as a \emph{dynamic document} that contains both prose and analysis code.
+Results are automatically rendered as text, tables, and plots.
+Second,
+the research archive is \emph{version controlled}, which means that the entire historical record of the project is retained.
+This version controlled archive can be made public.
+Third, \emph{dependency management} ensures that all software used is documented, including its version number and provenance.
+Without dependency management, code might not work (for example, because it is not clear where certain functions come from),
+or its results may change when some dependencies are updated.
+By combining \texttt{tidySEM} and \texttt{worcs},
+it is possible to conduct a fully reproducible LCA analysis compliant with best practices in open science.
+
+An additional argument for sharing data in the context of LCA is that reproducing LCA analyses requires individual participant data,
+not just sufficient statistics like many other methods (i.e., a correlation or covariance matrix).
+If the original data cannot be shared,
+sharing synthetic data at least enables reproducing and reusing analysis code.
+The \texttt{worcs} package provides a generic, non-parametric method to create a synthetic copy of data when data cannot be shared.
+It is also possible to synthesize data directly from the LCA model using \texttt{mxGenerateData()},
 which often provides a better basis for reproducing the LCA.
@@
 
-LCA analyses inherently involve some subjectivity, because of the
-potential for disagreement between criteria for class enumeration and
-the lack of objective fit indices. In our experience, reviewers (and
-authors) often disagree about which criteria should be used.
-Preregistering the class enumeration criteria as part of the analysis
-plan reduces disagreement and eliminates concerns about cherry picking
-fit measures after the results are known. It is particularly important
-to pre-specify the criteria that will be used to determine the maximum
-number of classes, because there is a risk that information criteria and
-likelihood ratio tests will suggest a larger number of classes, even if
-there are concerns about overfitting, classification accuracy, or model
-interpretability.
-
-The aforementioned Workflow for Open Reproducible Code in Science
-facilitates creating a preregistration. More interestingly, it also
-enables creating a \emph{Preregistration As Code} (PAC): A draft paper with
-reproducible analysis section based on fake (synthetic or simulated)
-data (\textbf{peikertReproducibleResearchTutorial2021?}). When the real data are
-accessed or collected, the analysis is simply reproduced with this new
-data file, and the results in the paper are automatically updated. PACs
-are quite unambiguous; the selection of the final model can even be
-automated by hard-coding the criteria into the analysis section. For an
-example of PAC, see (\textbf{vanlissaComplementingPreregisteredConfirmatory2022?}).
+LCA analyses inherently involve some subjectivity,
+because of the potential for disagreement between criteria for class enumeration and the lack of objective fit indices.
+In our experience, reviewers (and authors) often disagree about which criteria should be used.
+Preregistering the class enumeration criteria as part of the analysis plan reduces disagreement and eliminates concerns about cherry picking fit measures after the results are known.
+It is particularly important to pre-specify the criteria that will be used to determine the maximum number of classes,
+because there is a risk that information criteria and likelihood ratio tests will suggest a larger number of classes, even if there are concerns about overfitting, classification accuracy, or model interpretability.
+
+The aforementioned Workflow for Open Reproducible Code in Science facilitates creating a preregistration.
+More interestingly, it also enables creating a \emph{Preregistration As Code} (PAC):
+A draft paper with reproducible analysis section based on fake (synthetic or simulated) data (Peikert, Van Lissa, \& Brandmaier, 2021).
+When the real data are accessed or collected,
+the analysis is simply reproduced with this new data file,
+and the results in the paper are automatically updated.
+PACs are quite unambiguous; the selection of the final model can even be automated by hard-coding the criteria into the analysis section.
+For an example of PAC, see Van Lissa (2022a).
 
@@
 
-The function \texttt{table\_fit()} reports an overview of fit indices,
-customized for LCA. In reporting model fit, it is important, firstly, to
-report the effective sample size, which should be the same for all
-models in order to compare relative fit measures. Secondly, it should be
-clearly reported, for example, in the strategy of analyses, which
-parameters are estimated across and within classes. In case of doubt, a
-full overview of the parameters is given by the \texttt{coef()} and
-\texttt{table\_results()} functions. The number of parameters for each model
-should be included in the fit table. Thirdly, it is important that
-readers can assess the relative fit of different solutions. The most
-fundamental fit measure is the \(-2LL\), as this is the quantity optimized
-during estimation. From these statistics, information criteria can be
-computed. By default, \texttt{table\_fit()} includes the AIC, BIC, and saBIC. We
-suggest to report at least the BIC, as it is ubiquitous, grounded in
-theory, and performs well in class enumeration.
-
-Classification diagnostics should be reported, as they may determine
-whether models are excluded from consideration. Of particular interest
-is the diagonal of either the average posterior probabilities by most
-likely class membership, or of the most likely class membership by true
-class membership. The former indicates the average probability that
-individuals belong to the class they were assigned to, and the latter
-indicates the probability that individuals are assigned to a class if
-they actually belong to that class. By default, \texttt{table\_fit()} reports
-the latter as \texttt{prob\_min} and \texttt{prob\_max}. The entropy is also of interest
-as an overall measure of class separation. Finally, the range of
-proportions of cases assigned to each class is important; the minimum
-proportion of cases per class can be used to diagnose classes that are
-too small to be locally identified or practically relevant (see \textbf{depaoliMixtureClassRecovery2013?}). By default, \texttt{table\_fit()} reports the
-minimum and maximum proportion of cases, based on most likely class
-membership. Fifth, the LMR or BLRT can be reported if it is part of the
-class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}. By default,
-\texttt{table\_fit()} reports the VLM for subsequent models in the table.
+The function \texttt{table\_fit()} reports an overview of fit indices, customized for LCA.
+In reporting model fit, it is important, firstly, to report the effective sample size, which should be the same for all models in order to compare relative fit measures.
+Secondly, it should be clearly reported,
+for example, in the strategy of analyses,
+which parameters are estimated across and within classes.
+In case of doubt, a full overview of the parameters is given by the \texttt{coef()} and \texttt{table\_results()} functions.
+The number of parameters for each model should be included in the fit table.
+Thirdly, it is important that readers can assess the relative fit of different solutions.
+The most fundamental fit measure is the \(-2LL\),
+as this is the quantity optimized during estimation.
+From these statistics, information criteria can be computed.
+By default, \texttt{table\_fit()} includes the AIC, BIC, and saBIC.
+We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.
+
+Classification diagnostics should be reported,
+as they may determine whether models are excluded from consideration.
+Of particular interest is the diagonal of either the average posterior probabilities by most likely class membership,
+or of the most likely class membership by true class membership.
+The former indicates the average probability that individuals belong to the class they were assigned to,
+and the latter indicates the probability that individuals are assigned to a class if they actually belong to that class.
+By default, \texttt{table\_fit()} reports the latter as \texttt{prob\_min} and \texttt{prob\_max}.
+The entropy is also of interest as an overall measure of class separation.
+Finally, the range of proportions of cases assigned to each class is important;
+the minimum proportion of cases per class can be used to diagnose classes that are too small to be locally identified or practically relevant (see Depaoli, 2013).
+By default, \texttt{table\_fit()} reports the minimum and maximum proportion of cases,
+based on most likely class membership.
+Fifth, the LMR or BLRT can be reported if it is part of the class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}.
+By default, \texttt{table\_fit()} reports the VLM for subsequent models in the table.
 
@@
 
-Reporting LCA results involves reporting the class proportions, obtained
-via \texttt{class\_prob(res,\ "sum.posterior")}, and model parameters, obtained
-via \texttt{table\_results(res)}. The class proportions are obtained by
-standardizing the class weights: A \(k\)-class model has \(k-1\) free class
-weights, and one weight constrained to 1 for model identification. The
-free weights are included in the model results, but are redundant with
-the aforementioned class proportions. Other model parameters may include
-means, (co)variances, regression coefficients (rarely), and thresholds
-for ordinal indicators. It is good practice to report the point estimate
-and confidence interval of these parameters. Although it is conventional
-to report p-values as well, note that the default p-values tests the
-hypothesis that the parameter is equal to zero. This hypothesis may be
-irrelevant in LCA when the focus is on parameter differences between
-classes. Reporting standard errors allows readers to compute p-values
-for other more informative hypotheses. Thresholds are more readily
-interpretable when converted to conditional item probabilities, using
-the function \texttt{table\_prob()}. These indicate the probability of an item
-being endorsed given that the observation belongs to a particular latent
-(Geiser, 2012).
+Reporting LCA results involves reporting the class proportions,
+obtained via \texttt{class\_prob(res,\ "sum.posterior")},
+and model parameters, obtained via \texttt{table\_results(res)}.
+The class proportions are obtained by standardizing the class weights:
+A \(k\)-class model has \(k-1\) free class weights,
+and one weight constrained to 1 for model identification.
+The free weights are included in the model results,
+but are redundant with the aforementioned class proportions.
+Other model parameters may include means, (co)variances, regression coefficients (rarely), and thresholds for ordinal indicators.
+It is good practice to report the point estimate and confidence interval of these parameters.
+Although it is conventional to report p-values as well,
+note that the default p-values tests the hypothesis that the parameter is equal to zero.
+This hypothesis may be irrelevant in LCA when the focus is on parameter differences between classes.
+Reporting standard errors allows readers to compute p-values for other more informative hypotheses.
+Thresholds are more readily interpretable when converted to conditional item probabilities, using the function \texttt{table\_prob()}.
+These indicate the probability of an item being
+endorsed given that the observation belongs to a particular latent (Geiser, 2012).
 
@@
 
-As mentioned before, the results table included null-hypothesis tests
-for all parameters by default. Whether these tests are informative
-depends on the study goals and scale of the data. For example, if LCA
-indicators are mean-centered, then the p-values of class-specific means
-indicate whether the mean in that class differs significantly from the
-sample average. It is also possible to test more informative hypotheses,
-using the function \texttt{wald\_test()}. Equality-constrained informative
-hypotheses are specified using the \texttt{bain} syntax, and evaluated on the
-model object. This enables testing theoretical hypotheses - for example,
-if specific values for class-specific means had been hypothesized. The
-same function allows testing parameters for equality, either within or
-across classes.
+As mentioned before, the results table included null-hypothesis tests for all parameters by default.
+Whether these tests are informative depends on the study goals and scale of the data.
+For example, if LCA indicators are mean-centered,
+then the p-values of class-specific means indicate whether the mean in that class differs significantly from the sample average.
+It is also possible to test more informative hypotheses,
+using the function \texttt{wald\_test()}.
+Equality-constrained informative hypotheses are specified using the \texttt{bain} syntax,
+and evaluated on the model object.
+This enables testing theoretical hypotheses - for example, if specific values for class-specific means had been hypothesized.
+The same function allows testing parameters for equality, either within or across classes.
 
@@
 
-In an informal review of published graphics for LCA, we found that most
-visualizations focused on the estimated model parameters, displaying
-class means as points connected by lines, with point shape and linetype
-indicating class membership. Note that the lines introduce spurious
-information. Furthermore, they omit uncertainty about class means and do
-not illustrate how well the model describes the data.
-
-The Grammar of Graphics addresses best practices in visualization, which
-we summarize here with specific emphasis on the application to LCA. Most
-importantly, good graphics start with the \emph{data} in mind. Visualizing
-raw data allows one to assess class separability, and which aspects of
-the data are captured well or poorly by the model. A second
-consideration are which \emph{aesthetics} to use. In a color plot, class
-membership can be indicated by color. In a black and white plot, it can
-be indicated by linetype and point shape. A third consideration is which
-geometric objects (geoms) to use. As mentioned before, common practice
-is to plot parameter estimates as points, and connect these points with
-lines. Points are suitable geoms to visualize point estimates and raw
-data. For continuous indicators, additional relevant geoms include error
-bars for standard errors and standard deviations, and density plots to
-visualize the distribution of raw data (smoothed with a kernel density
-function). In two dimensions, (co)variances can be visualized as
-ellipses. Finally, when plotting multiple class solutions, it makes
-sense to \emph{facet} a plot into multiple panels by class solution. The
-plots in \texttt{tidySEM} are based on best practices as outlined in the
-grammar of graphics, and are implemented in \texttt{ggplot2}. As such, these
-plots can be further modified just like any other \texttt{ggplot2} object. One
-remaining limitation is that these best practices have only been
-implemented for LCA with continuous indicators, in \texttt{plot\_profiles()},
-\texttt{plot\_density()} and \texttt{plot\_bivariate()}, and for LCGA in
-\texttt{plot\_growth()}. Plots for LCA with ordinal indicators are still
-rudimentary, more work remains to be done in this area.
+In an informal review of published graphics for LCA,
+we found that most visualizations focused on the estimated model parameters,
+displaying class means as points connected by lines, with point shape and linetype indicating class membership.
+Note that the lines introduce spurious information.
+Furthermore, they omit uncertainty about class means and do not illustrate how well the model describes the data.
+
+The Grammar of Graphics addresses best practices in visualization, which we summarize here with specific emphasis on the application to LCA.
+Most importantly, good graphics start with the \emph{data} in mind.
+Visualizing raw data allows one to assess class separability,
+and which aspects of the data are captured well or poorly by the model.
+A second consideration are which \emph{aesthetics} to use.
+In a color plot, class membership can be indicated by color.
+In a black and white plot, it can be indicated by linetype and point shape.
+A third consideration is which geometric objects (geoms) to use.
+As mentioned before, common practice is to plot parameter estimates as points, and connect these points with lines.
+Points are suitable geoms to visualize point estimates and raw data.
+For continuous indicators, additional relevant geoms include error bars for standard errors and standard deviations,
+and density plots to visualize the distribution of raw data (smoothed with a kernel density function).
+In two dimensions, (co)variances can be visualized as ellipses.
+Finally, when plotting multiple class solutions,
+it makes sense to \emph{facet} a plot into multiple panels by class solution.
+The plots in \texttt{tidySEM} are based on best practices as outlined in the grammar of graphics,
+and are implemented in \texttt{ggplot2}.
+As such, these plots can be further modified just like any other \texttt{ggplot2} object.
+One remaining limitation is that these best practices have only been implemented for LCA with continuous indicators, in \texttt{plot\_profiles()}, \texttt{plot\_density()} and \texttt{plot\_bivariate()}, and for LCGA in \texttt{plot\_growth()}.
+Plots for LCA with ordinal indicators are still rudimentary, more work remains to be done in this area.
 
@@
 
-Based on our experience in statistical consultations, there are a few
-common pitfalls and misunderstandings about LCA. First, perhaps the most
-common problem we see in consultations is treating ordinal data as
-continuous. As explained previously, LCA for ordinal indicators should
-be used for Likert-type scales and other data with few unique values. A
-second common pitfall is omitting the 1-class solution from class
-enumeration. If a one-class solution has better or similar fit to models
-with a higher number of classes, LCA may not be appropriate. A one-class
-solution should thus always be included as a benchmark. Third, although
-it is widely understood that classification diagnostics (e.g., entropy,
-posterior probability, class counts) are not class enumeration criteria,
-they still have a role in class enumeration. Specifically, they can be
-used to exclude solutions that do not meet predefined criteria from
-consideration. Fourth, previous reporting standards called for reporting
-of sample descriptive statistics (see \textbf{schreiberLatentClassAnalysis2017?}). While we agree that this is due
-dilligence, sample descriptives have two limitations in the context of
-LCA: they are not sufficient to check or reproduce the analysis, and
-many use cases of LCA assume that the population is heterogeneous, which
-limits the utility of whole-sample descriptives. These two limitations
-can be addressed by data sharing and reporting descriptive statistics
-for each latent class separately. Fifth, a recent publication mistakenly
-stated that an assumption of mixture models is that observed indicators
-are normally distributed (Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020). This is incorrect, as
-evident from the shoe size example: When the number of classes is
-greater than one, Gaussian mixture models assume that the distribution
-of observed indicators is a \emph{mixture of multiple normal distributions}.
-If the indicators were normally
-distributed, there would be no classes to extract as the whole
-population would belong to a single (homogeneous) class. Finally, the
-same paper suggested that robust standard errors should be used when
-indicators are not normally distributed (Spurk et al., 2020). This is
-also incorrect; to our knowledge, there is no literature on the merits
-of robust standard errors for LCA. Finally, we caution against step-wise
-class enumeration, whereby additional models are estimated until the
-selected class enumeration criteria no longer show improvement of model
-fit (e.g., see Spurk et al., 2020). This practice is problematic,
-because it is possible that an even higher number of classes would lead
-to a (significant) improvement in fit. Consider this thought experiment:
-there exist two groups of three relatively similar classes each (six in
-total). Step-wise class enumeration shows improvement in model fit going
-from a 1-class to 2-class model: the two classes capture the difference
-between the two groups of three classes. Then, no noteworthy improvement
-in model fit is found for adding more classes, until the 6-class
-solution is reached. In the step-wise approach, the analysis would be
-terminated after the 2-class model - even though the correct number of
-classes is 6.
+Based on our experience in statistical consultations,
+there are a few common pitfalls and misunderstandings about LCA.
+First, perhaps the most common problem we see in consultations is treating ordinal data as continuous.
+As explained previously, LCA for ordinal indicators should be used for Likert-type scales and other data with few unique values.
+A second common pitfall is omitting the 1-class solution from class enumeration.
+If a one-class solution has better or similar fit to models with a higher number of classes, LCA may not be appropriate.
+A one-class solution should thus always be included as a benchmark.
+Third, although it is widely understood that classification diagnostics (e.g., entropy, posterior probability, class counts) are not class enumeration criteria,
+they still have a role in class enumeration.
+Specifically, they can be used to exclude solutions that do not meet predefined criteria from consideration.
+Fourth, previous reporting standards called for reporting of sample descriptive statistics (see Schreiber, 2017).
+While we agree that this is due dilligence,
+sample descriptives have two limitations in the context of LCA:
+they are not sufficient to check or reproduce the analysis,
+and many use cases of LCA assume that the population is heterogeneous, which limits the utility of whole-sample descriptives.
+These two limitations can be addressed by data sharing and reporting descriptive statistics for each latent class separately.
+Fifth, a recent publication mistakenly stated that an assumption of mixture
+models is that observed indicators are normally distributed
+(Spurk, Hirschi, Wang, Valero, \& Kauffeld, 2020).
+This is incorrect, as evident from the shoe size example:
+When the number of classes is greater than one,
+Gaussian mixture models assume that the distribution of observed indicators is a \emph{mixture of multiple normal distributions}.
+If the indicators were normally distributed,
+there would be no classes to extract as the whole population
+would belong to a single (homogeneous) class.
+Finally, the same paper suggested that robust standard errors
+should be used when indicators are not normally distributed
+(Spurk et al., 2020).
+This is also incorrect;
+to our knowledge, there is no literature on the merits of robust standard errors for LCA.
+Finally, we caution against step-wise class enumeration,
+whereby additional models are estimated until the selected class enumeration criteria no longer show improvement of model fit (e.g., see Spurk et al., 2020).
+This practice is problematic,
+because it is possible that an even higher number of classes would lead to a (significant) improvement in fit.
+Consider this thought experiment: there exist two groups of three relatively similar classes each (six in total).
+Step-wise class enumeration shows improvement in model fit going from a 1-class to 2-class model: the two classes capture the difference between the two groups of three classes.
+Then, no noteworthy improvement in model fit is found for adding more classes, until the 6-class solution is reached.
+In the step-wise approach, the analysis would be terminated after the 2-class model - even though the correct number of classes is 6.
 
@@
 
-The best practices discussed in this paper are summarized in the
-SMART-LCA Checklist: Standards for More Accuracy in Reporting of
-different Types of Latent Class Analysis.
+The best practices discussed in this paper are summarized in the SMART-LCA Checklist: Standards for More Accuracy in Reporting of different Types of Latent Class Analysis.
 
@@
   \item
-    Any other reasons why the observed (multivariate) distribution
-    could not be described by the LCA (e.g., extreme skew)?
+    Any other reasons why the observed (multivariate) distribution could not be described by the LCA (e.g., extreme skew)?
   \end{itemize}
@@
   \item
-    In exploratory LCA, compare all sensible alternative
-    specifications
-  \item
-    In confirmatory LCA, justify alternative models on theoretical
-    grounds
-  \item
-    Check that the observed (multivariate) distribution could
-    theoretically be captured by the specified model (see \#1).
+    In exploratory LCA, compare all sensible alternative specifications
+  \item
+    In confirmatory LCA, justify alternative models on theoretical grounds
+  \item
+    Check that the observed (multivariate) distribution could theoretically be captured by the specified model (see \#1).
   \end{itemize}
@@
   \item
-    Justify the maximum number of classes \(K\)
-  \item
-    In exploratory LCA, estimate 1:\(K\) classes
-  \item
-    In confirmatory LCA, estimate the theoretical number of classes,
-    a 1-class model, and optionally other numbers of classes for
-    comparison
+    Justify the maximum number of classes \(k\)
+  \item
+    In exploratory LCA, estimate 1:\(k\) classes
+  \item
+    In confirmatory LCA, estimate the theoretical number of classes, a 1-class model, and optionally other numbers of classes for comparison
   \end{itemize}
 \item
-  Justify the (preregistered) criteria used for class enumeration,
-  which can include:
+  Justify the (preregistered) criteria used for class enumeration, which can include:
 
@@
 \item
-  Justify the (preregistered) criteria used to eliminate models from
-  consideration, including:
+  Justify the (preregistered) criteria used to eliminate models from consideration, including:
 
@@
   \item
-    Local identifiability (acceptable number of observations per
-    parameter in each class)
+    Local identifiability (acceptable number of observations per parameter in each class)
   \item
@@
   \item
-    All parameters of the chosen model (means, (co)variances,
-    thresholds)
+    All parameters of the chosen model (means, (co)variances, thresholds)
   \item
@@
   \item
-    Show multivariate distributions for multivariate models, if
-    possible
+    Show multivariate distributions for multivariate models, if possible
   \end{itemize}
@@
   \item
-    If entropy \textgreater{} .95 and posterior classification probability for
-    each class \textgreater{} .95, this is less crucial
+    If entropy \textgreater{} .95 and posterior classification probability for each class \textgreater{} .95, this is less crucial
   \end{itemize}
@@
 (2016).
-A {Comparison} of {Information} {Criteria} in {Clustering} {Based} on {Mixture} of {Multivariate} {Normal} {Distributions}.
+A comparison of information criteria in clustering based on mixture of multivariate normal distributions.
 \emph{Mathematical and Computational Applications}, \emph{21}(3), 34.
@@
 (2022).
-Maximizing {Realism}: {Mapping} {Plastic} {Particles} at the {Ocean} {Surface} {Using} {Mixtures} of {Normal} {Distributions}.
+Maximizing realism: Mapping plastic particles at the ocean surface using mixtures of normal distributions.
 \emph{Environmental Science \& Technology}, \emph{56}(22), 15552--15562.
 \url{https://doi.org/10.1021/acs.est.2c03559}
+
+\leavevmode\vadjust pre{\hypertarget{ref-baughman_mixture_2007}{}}%
+Baughman, A.
+L.
+(2007).
+Mixture model framework facilitates understanding of zero-inflated and hurdle models for count data.
+\emph{Journal of Biopharmaceutical Statistics}, \emph{17}(5), 943--946.
+\url{https://doi.org/10.1080/10543400701514098}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-biernacki_choosing_2003}{}}%
+Biernacki, C., Celeux, G., \& Govaert, G.
+(2003).
+Choosing starting values for the {EM} algorithm for getting the highest likelihood in multivariate gaussian mixture models.
+\emph{Computational Statistics \& Data Analysis}, \emph{41}(3), 561--575.
+\url{https://doi.org/10.1016/S0167-9473(02)00163-9}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bolck_estimating_2004}{}}%
+Bolck, A., Croon, M., \& Hagenaars, J.
+(2004).
+Estimating latent structure models with categorical variables: One-step versus three-step estimators.
+\emph{Political Analysis}, \emph{12}(1), 3--27.
+\url{https://doi.org/10.1093/pan/mph001}
+
+\leavevmode\vadjust pre{\hypertarget{ref-celeux_entropy_1996}{}}%
+Celeux, G., \& Soromenho, G.
+(1996).
+An entropy criterion for assessing the number of clusters in a mixture model.
+\emph{Journal of Classification}, \emph{13}(2), 195--212.
+\url{https://doi.org/10.1007/BF01246098}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chen_efficacy_2017}{}}%
+Chen, Q., Luo, W., Palardy, G.
+J., Glaman, R., \& McEnturff, A.
+(2017).
+The efficacy of common fit indices for enumerating classes in growth mixture models when nested data structure is ignored: A monte carlo study.
+\emph{{SAGE} Open}, \emph{7}(1), 215824401770045.
+\url{https://doi.org/10.1177/2158244017700459}
+
 \leavevmode\vadjust pre{\hypertarget{ref-collins_latent_2009}{}}%
-Collins, L. M., \& Lanza, S. T. (2009). \emph{Latent {Class} and {Latent} {Transition} {Analysis}: {With} {Applications} in the {Social}, {Behavioral}, and {Health} {Sciences}}. John Wiley \& Sons.
+Collins, L. M., \& Lanza, S. T. (2009). \emph{Latent class and latent transition analysis: With applications in the social, behavioral, and health sciences}. John Wiley \& Sons.
+
+\leavevmode\vadjust pre{\hypertarget{ref-corana_minimizing_1987}{}}%
+Corana, A., Marchesi, M., Martini, C., \& Ridella, S.
+(1987).
+Minimizing multimodal functions of continuous variables with the {``simulated annealing''} algorithm---corrigenda for this article is available here.
+\emph{{ACM} Transactions on Mathematical Software}, \emph{13}(3), 262--280.
+\url{https://doi.org/10.1145/29380.29864}
+
+\leavevmode\vadjust pre{\hypertarget{ref-depaoli_mixture_2013}{}}%
+Depaoli, S.
+(2013).
+Mixture class recovery in {GMM} under varying degrees of class separation: Frequentist versus bayesian estimation.
+\emph{Psychological Methods}, \emph{18}(2), 186--219.
+\url{https://doi.org/10.1037/a0031609}
+
+\leavevmode\vadjust pre{\hypertarget{ref-donnellan_resilient_2010}{}}%
+Donnellan, M.
+B., \& Robins, R.
+W.
+(2010).
+Resilient, overcontrolled, and undercontrolled personality types: Issues and controversies: Personality types.
+\emph{Social and Personality Psychology Compass}, \emph{4}(11), 1070--1083.
+\url{https://doi.org/10.1111/j.1751-9004.2010.00313.x}
 
 \leavevmode\vadjust pre{\hypertarget{ref-enders_applied_2022}{}}%
-Enders, C. K. (2022). \emph{Applied {Missing} {Data} {Analysis}}. Guilford Publications.
+Enders, C. K. (2022). \emph{Applied missing data analysis}. Guilford Publications.
+
+\leavevmode\vadjust pre{\hypertarget{ref-enders_relative_2001}{}}%
+Enders, C., \& Bandalos, D.
+(2001).
+The relative performance of full information maximum likelihood estimation for missing data in structural equation models.
+\emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{8}(3), 430--457.
+\url{https://doi.org/10.1207/S15328007SEM0803_5}
 
@@
 Unsupervised learning of finite mixture models.
-\emph{IEEE Transactions on Pattern Analysis and Machine Intelligence}, \emph{24}(3), 381--396.
+\emph{{IEEE} Transactions on Pattern Analysis and Machine Intelligence}, \emph{24}(3), 381--396.
 \url{https://doi.org/10.1109/34.990138}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-geiser_data_2012}{}}%
-Geiser, C. (2012). \emph{Data {Analysis} with {Mplus}}. Guilford Press.
+Geiser, C. (2012). \emph{Data analysis with mplus}. Guilford Press.
 
@@
 Is adding more indicators to a latent class analysis beneficial or detrimental?
-{Results} of a {Monte} {Carlo} study.
+Results of a monte carlo study.
 \emph{Frontiers in Psychology: Quantitative Psychology and Measurement}, \emph{5}.
@@
 (2022).
-Characteristics of suicide attempts associated with lethality and method: {A} latent class analysis of the {Military} {Suicide} {Research} {Consortium}.
+Characteristics of suicide attempts associated with lethality and method: A latent class analysis of the military suicide research consortium.
 \emph{Journal of Psychiatric Research}, \emph{149}, 54--61.
+\url{https://doi.org/10.1016/j.jpsychires.2022.02.016}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hallquist_mplusautomation_2018}{}}%
+Hallquist, M., Wiley, J., \& Van Lissa, C.
+J.
+(2018).
+{MplusAutomation}: An r package for facilitating large-scale latent variable analyses in mplus (Version 0.7-3).
+Retrieved from \url{https://CRAN.R-project.org/package=MplusAutomation}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hastie_elements_2009}{}}%
+Hastie, T., Tibshirani, R., \& Friedman, J. (2009). \emph{The elements of statistical learning: Data mining, inference, and prediction} (Second). New York: Springer.
 
 \leavevmode\vadjust pre{\hypertarget{ref-hennig_handbook_2015}{}}%
-Hennig, C., Meila, M., Murtagh, F., \& Rocci, R. (2015). \emph{Handbook of {Cluster} {Analysis}}. 28.
+Hennig, C., Meila, M., Murtagh, F., \& Rocci, R. (2015). \emph{Handbook of cluster analysis}. 28.
 
@@
 (2010).
-Tests of {Homoscedasticity}, {Normality}, and {Missing} {Completely} at {Random} for {Incomplete} {Multivariate} {Data}.
+Tests of homoscedasticity, normality, and missing completely at random for incomplete multivariate data.
 \emph{Psychometrika}, \emph{75}(4), 649--674.
@@
 Jung, T., \& Wickrama, K.
-a.
+A.
 S.
 (2008).
-An {Introduction} to {Latent} {Class} {Growth} {Analysis} and {Growth} {Mixture} {Modeling}.
+An introduction to latent class growth analysis and growth mixture modeling.
 \emph{Social and Personality Psychology Compass}, \emph{2}(1), 302--317.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-van_kollenburg_lazy_2018}{}}%
-Kollenburg, G. H. van, Mulder, J., \& Vermunt, J. K. (2018). \emph{The {Lazy} {Bootstrap}. {A} {Fast} {Resampling} {Method} for {Evaluating} {Latent} {Class} {Model} {Fit}}. 23.
+Kollenburg, G. H. van, Mulder, J., \& Vermunt, J. K. (2018). \emph{The lazy bootstrap. A fast resampling method for evaluating latent class model fit}. 23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lamprecht_towards_2020}{}}%
+Lamprecht, AL., Garcia, L., Kuzak, M., Martinez, C., Arcila, R., Martin Del Pico, E., \ldots{} Capella-Gutierrez, S.
+(2020).
+Towards {FAIR} principles for research software.
+\emph{Data Science}, \emph{3}(1), 37--59.
+\url{https://doi.org/10.3233/DS-190026}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-likas_global_2003}{}}%
+Likas, A., Vlassis, N., \& J. Verbeek, J. (2003). The global k-means clustering algorithm. \emph{Pattern Recognition}, \emph{36}(2), 451--461. \url{https://doi.org/10.1016/S0031-3203(02)00060-2}
+
 \leavevmode\vadjust pre{\hypertarget{ref-little_test_1988}{}}%
@@
 (1988).
-A {Test} of {Missing} {Completely} at {Random} for {Multivariate} {Data} with {Missing} {Values}.
+A test of missing completely at random for multivariate data with missing values.
 \emph{Journal of the American Statistical Association}, \emph{83}(404), pp.
@@
 \url{https://doi.org/10.2307/2290157}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lo_testing_2001}{}}%
+Lo, Y., Mendell, N. R., \& Rubin, D. B. (2001). Testing the number of components in a normal mixture. \emph{Biometrika}, \emph{88}(3), 767--778. \url{https://doi.org/10.1093/biomet/88.3.767}
+
+\leavevmode\vadjust pre{\hypertarget{ref-luyckx_developmental_2008}{}}%
+Luyckx, K., Schwartz, S.
+J., Goossens, L., Soenens, B., \& Beyers, W.
+(2008).
+Developmental typologies of identity formation and adjustment in female emerging adults: A latent class growth analysis approach.
+\emph{Journal of Research on Adolescence}, \emph{18}(4), 595--619.
+\url{https://doi.org/10.1111/j.1532-7795.2008.00573.x}
 
@@
 (2022).
-Perceived {Teacher} {Discrimination} and {Depressive} {Feelings} in {Adolescents}: {The} {Role} of {National}, {Regional}, and {Heritage} {Identities} in {Flemish} {Schools}.
+Perceived teacher discrimination and depressive feelings in adolescents: The role of national, regional, and heritage identities in flemish schools.
 \emph{Journal of Youth and Adolescence}, \emph{51}(12), 2281--2293.
@@
 (2013).
-Latent {Class} {Analysis} and {Finite} {Mixture} {Modeling}.
-In \emph{The {Oxford} {Handbook} of {Quantitative} {Methods}}: \emph{Vol.} \emph{2: Statistical Analysis} (p.
+Latent class analysis and finite mixture modeling.
+In \emph{The oxford handbook of quantitative methods}: \emph{Vol.} \emph{2: Statistical Analysis} (p.
 551).
@@
 (2020).
-\emph{Statistical {Rethinking} {\textbar} {A} {Bayesian} {Course} with {Examples} in {R} and {STAN}} (2nd ed.).
-Chapman; Hall/CRC.
+\emph{Statistical rethinking {\textbar} a bayesian course with examples in r and {STAN}} (2nd ed.).
+Chapman; Hall/{CRC}.
 Retrieved from \url{https://www.taylorfrancis.com/books/mono/10.1201/9780429029608/statistical-rethinking-richard-mcelreath}
+
+\leavevmode\vadjust pre{\hypertarget{ref-molenaar_arbitrary_1994}{}}%
+Molenaar, P.
+C.
+M., \& Eye, A.
+von.
+(1994).
+On the arbitrary nature of latent variables.
+In \emph{Latent variables analysis: Applications for developmental research} (pp.
+226--242).
+Thousand Oaks, {CA}, {US}: Sage Publications, Inc.
+
+\leavevmode\vadjust pre{\hypertarget{ref-muthen_mplus_1998}{}}%
+Muthén, L. K., \& Muthén, B. O. (1998). \emph{Mplus user's guide}. Los Angeles, {CA}: Muthén \& Muthén.
 
@@
 (2016).
-{OpenMx} 2.0: {Extended} {Structural} {Equation} and {Statistical} {Modeling}.
+{OpenMx} 2.0: Extended structural equation and statistical modeling.
 \emph{Psychometrika}, \emph{81}(2), 535--549.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-nosek_transparency_2016}{}}%
+Nosek, B.
+A., Alter, G., Banks, G.
+C., Borsboom, D., Bowman, S., Breckler, S., \ldots{} DeHaven, A.
+C.
+(2016).
+\emph{Transparency and openness promotion ({TOP}) guidelines}.
+{OSF} Preprints.
+\url{https://doi.org/10.31219/osf.io/vj54c}
+
 \leavevmode\vadjust pre{\hypertarget{ref-nylund_deciding_2007}{}}%
@@
 (2007).
-Deciding on the {Number} of {Classes} in {Latent} {Class} {Analysis} and {Growth} {Mixture} {Modeling}: {A} {Monte} {Carlo} {Simulation} {Study}.
+Deciding on the number of classes in latent class analysis and growth mixture modeling: A monte carlo simulation study.
 \emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{14}(4), 535--569.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pavlopoulos_measuring_2015}{}}%
-Pavlopoulos, D., \& Vermunt, J. K. (2015). \emph{Measuring temporary employment. {Do} survey or register data tell the truth?} 37.
+Pavlopoulos, D., \& Vermunt, J. K. (2015). \emph{Measuring temporary employment. Do survey or register data tell the truth?} 37.
+
+\leavevmode\vadjust pre{\hypertarget{ref-peikert_reproducible_2021}{}}%
+Peikert, A., Van Lissa, C.
+J., \& Brandmaier, A.
+M.
+(2021).
+Reproducible research in r: A tutorial on how to do the same thing more than once.
+\emph{Psych}, \emph{3}(4), 836--867.
+\url{https://doi.org/10.3390/psych3040053}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rosenberg_tidylpa_2018}{}}%
+Rosenberg, J., Beymer, P., Anderson, D., Van Lissa, C.
+J., \& Schmidt, J.
+(2018).
+{tidyLPA}: An r package to easily carry out latent profile analysis ({LPA}) using open-source or commercial software.
+\emph{Journal of Open Source Software}, \emph{3}(30), 978.
+\url{https://doi.org/10.21105/joss.00978}
 
 \leavevmode\vadjust pre{\hypertarget{ref-rosseel_lavaan_2012}{}}%
-Rosseel, Y. (2012). Lavaan: {An} {R} {Package} for {Structural} {Equation} {Modeling}. \emph{Journal of Statistical Software}, \emph{48}, 1--36. \url{https://doi.org/10.18637/jss.v048.i02}
+Rosseel, Y. (2012). Lavaan: An r package for structural equation modeling. \emph{Journal of Statistical Software}, \emph{48}, 1--36. \url{https://doi.org/10.18637/jss.v048.i02}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rousseau_asymptotic_2011}{}}%
+Rousseau, J., \& Mengersen, K.
+(2011).
+Asymptotic behaviour of the posterior distribution in overfitted mixture models.
+\emph{Journal of the Royal Statistical Society: Series B (Statistical Methodology)}, \emph{73}(5), 689--710.
+\url{https://doi.org/10.1111/j.1467-9868.2011.00781.x}
 
 \leavevmode\vadjust pre{\hypertarget{ref-rubin_inference_1976}{}}%
-Rubin, D. B. (1976). Inference and {Missing} {Data}. \emph{Biometrika}, \emph{63}(3), 581--592. \url{https://doi.org/10.2307/2335739}
+Rubin, D. B. (1976). Inference and missing data. \emph{Biometrika}, \emph{63}(3), 581--592. \url{https://doi.org/10.2307/2335739}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schreiber_latent_2017}{}}%
+Schreiber, J.
+B.
+(2017).
+Latent class analysis: An example for reporting results.
+\emph{Research in Social and Administrative Pharmacy}, \emph{13}(6), 1196--1201.
+\url{https://doi.org/10.1016/j.sapharm.2016.11.011}
 
@@
 (2016).
-\href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5096736}{Mclust 5: {Clustering}, {Classification} and {Density} {Estimation} {Using} {Gaussian} {Finite} {Mixture} {Models}}.
+\href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5096736}{Mclust 5: Clustering, classification and density estimation using gaussian finite mixture models}.
 \emph{The R Journal}, \emph{8}(1), 289--317.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sinha_practitioners_2021}{}}%
+Sinha, P., Calfee, C.
+S., \& Delucchi, K.
+L.
+(2021).
+Practitioner's guide to latent class analysis: Methodological considerations and common pitfalls.
+\emph{Critical Care Medicine}, \emph{49}(1), e63--e79.
+\url{https://doi.org/10.1097/CCM.0000000000004710}
 
@@
 (2020).
-Latent profile analysis: {A} review and {``how to''} guide of its application within vocational behavior research.
+Latent profile analysis: A review and {``how to''} guide of its application within vocational behavior research.
 \emph{Journal of Vocational Behavior}, \emph{120}, 103445.
 \url{https://doi.org/10.1016/j.jvb.2020.103445}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tein_statistical_2013}{}}%
+Tein, JY., Coxe, S., \& Cham, H.
+(2013).
+Statistical power to detect the correct number of classes in latent profile analysis.
+\emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{20}(4), 640--657.
+\url{https://doi.org/10.1080/10705511.2013.824781}
 
@@
 (2020).
-Measurement {Invariance} and {Differential} {Item} {Functioning} {Across} {Gender} {Within} a {Latent} {Class} {Analysis} {Framework}: {Evidence} {From} a {High}-{Stakes} {Test} for {University} {Admission} in {Saudi} {Arabia}.
+Measurement invariance and differential item functioning across gender within a latent class analysis framework: Evidence from a high-stakes test for university admission in saudi arabia.
 \emph{Frontiers in Psychology}, \emph{11}.
@@
 (2017).
-The {GRoLTS}-{Checklist}: {Guidelines} for {Reporting} on {Latent} {Trajectory} {Studies}.
+The {GRoLTS}-checklist: Guidelines for reporting on latent trajectory studies.
 \emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{24}(3), 451--467.
 \url{https://doi.org/10.1080/10705511.2016.1247646}
+
+\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_complementing_2022}{}}%
+Van Lissa, C.
+J.
+(2022a).
+Complementing preregistered confirmatory analyses with rigorous, reproducible exploration using machine learning.
+\emph{Religion, Brain \& Behavior}, \emph{0}(0), 1--5.
+\url{https://doi.org/10.1080/2153599X.2022.2070254}
+
+\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_tidysem_2022}{}}%
+Van Lissa, C. J. (2022b). {tidySEM}: Tidy structural equation modeling (Version 0.2.4). Retrieved from \href{https://www.github.com/cjvanlissa/tidySEM}{www.github.com/cjvanlissa/tidySEM}
+
+\leavevmode\vadjust pre{\hypertarget{ref-van_lissa_worcs_2021}{}}%
+Van Lissa, C.
+J., Brandmaier, A.
+M., Brinkman, L., Lamprecht, AL., Peikert, A., Struiksma, M.
+E., \& Vreede, B.
+M.
+I.
+(2021).
+{WORCS}: A workflow for open reproducible code in science.
+\emph{Data Science}, \emph{4}(1), 29--49.
+\url{https://doi.org/10.3233/DS-210031}
 
@@
 (2015).
-Divergence {Between} {Adolescent} and {Parental} {Perceptions} of {Conflict} in {Relationship} to {Adolescent} {Empathy} {Development}.
+Divergence between adolescent and parental perceptions of conflict in relationship to adolescent empathy development.
 \emph{Journal of Youth and Adolescence}, \emph{44}(1), 48--61.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vermunt_longitudinal_2010}{}}%
-Vermunt, Jeroen K.
+Vermunt, J.
+K.
 (2010).
-Longitudinal {Research} {Using} {Mixture} {Models}.
+Longitudinal research using mixture models.
 In K.
-Montfort, J.
+van Montfort, J.
 H.
+L.
 Oud, \& A.
-Satorra (Eds.), \emph{Longitudinal {Research} with {Latent} {Variables}} (pp.
+Satorra (Eds.), \emph{Longitudinal research with latent variables} (pp.
 119--152).
-Berlin/Heidelberg: Springer.
+Berlin, Heidelberg: Springer.
 \url{https://doi.org/10.1007/978-3-642-11760-2_4}
 
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_latent_2000}{}}%
+Vermunt, J. K., \& Magidson, J. (2000). Latent {GOLD} 2.0 user's guide. \emph{Statistical Innovations Inc.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vermunt_upgrade_2021}{}}%
+Vermunt, J. K., \& Magidson, J. (2021). Upgrade manual for latent {GOLD} basic, advanced, syntax, and choice version 6.0. \emph{Statistical Innovations Inc.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-vermunt_latent_2004}{}}%
-Vermunt, Jeroen K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.
+Vermunt, J.
+K., Magidson, J., Lewis-Beck, M., Bryman, A., Liao, T.
 F., \& Department of Methodology and Statistics.
@@
 Latent class analysis.
-In \emph{The {Sage} encyclopedia of social sciences research methods} (pp.
+In \emph{The sage encyclopedia of social sciences research methods} (pp.
 549--553).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-depmixS4}{}}%
-Visser, I., \& Speekenbrink, M.
-(2010).
-{depmixS4}: An {R} package for hidden markov models.
-\emph{Journal of Statistical Software}, \emph{36}(7), 1--21.
-Retrieved from \url{https://www.jstatsoft.org/v36/i07/}
+\leavevmode\vadjust pre{\hypertarget{ref-visser_depmixs4_2010}{}}%
+Visser, I., \& Speekenbrink, M. (2010). {depmixS}4: An r package for hidden markov models. \emph{Journal of Statistical Software}, \emph{36}, 1--21. \url{https://doi.org/10.18637/jss.v036.i07}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wagenmakers_aic_2004}{}}%
+Wagenmakers, EJ., \& Farrell, S. (2004). {AIC} model selection using akaike weights. \emph{Psychonomic Bulletin \& Review}, \emph{11}(1), 192--196. \url{https://doi.org/10.3758/BF03206482}
 
@@
 (2020).
-Latent {Class} {Analysis}: {A} {Guide} to {Best} {Practice}.
+Latent class analysis: A guide to best practice.
 \emph{Journal of Black Psychology}, \emph{46}(4), 287--311.

### 2023-02-16 manuscript.Rmd
Update pt.2
Covered the second part of the tutorial. Added missing citations, rephrased many things, wrote conclusion.

@@
 keywords          : "latent class analysis, mixture models, best practices, free open source software, tidySEM"
-wordcount         : "9001"
+wordcount         : "12158"
 bibliography      : "mixture.bib"
@@
 A mixture model was used to estimate employment status, correcting for measurement error, from both self-report and registry data.
-The study investigated random and systematic measurement error in the indicatorss stemming from the two registers,
+The study investigated random and systematic measurement error in the indicators stemming from the two registers,
 which revealed strengths and weaknesses of both types of assessment.
@@
 The use of LCA for dimension reduction is common in longitudinal research,
-where one developmental process is summarized using latent class growth analysis (LCGA),
+where one developmental process is summarized using *latent class growth analysis* (*LCGA*),
 and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
@@
 dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
-However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data.
+However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data,
+which makes intervention-planning significantly more manageable.
 
@@
 
-### Violations of normality
+### Violations of Normality
 
 Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007; magnus_multidimensional_2021].
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007; @magnus_multidimensional_2022].
 These models estimate one class to explain the "true zeros",
@@
 The best practices in estimation, as outlined in Table
-\@ref(tab:checkest), are rooted in existing recommendations for best
-practices for estimating specific subtypes of LCA,
+\@ref(tab:checkest-tab), are rooted in existing recommendations 
+for estimating specific subtypes of LCA,
 including latent
@@
 class analysis with ordinal indicators [e.g., @nylund-gibson_ten_2018].
-These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.
+These guidelines were generalized to be relevant to all types of LCA, 
+and updated to current best practices, as explained below.
 <!-- In order to aid researchers working with latent trajectory models, -->
@@
 
-```{r checkest, include=TRUE}
-tab_check <- data.frame(Item = c(
+```{r checkest-tab, include=TRUE}
+item <- data.frame(Item = c(
   "Examining Observed Data",
@@
 ))
-tab_check$`#` = paste0(1:length(tab_check$Item), ".")
-kbl(tab_check[, c("#", "Item")])
+
+item$`#` = paste0(1:length(item$Item), ".")
+knitr::kable(item[, c("#", "Item")], caption = "")
+
 ```
@@
 variance of continuous and categorical variables,
-missingness, skew and kurtosis (peakedness of the distribution). 
+their missingness, skew and kurtosis. 
 However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
@@
 for categorical ones.
-Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.
+Density plots can also aid in the choice of the number of classes, as explained in the Visualization section.
 
 Particular points of attention include measurement level of the indicators.
-Recall that `tidySEM` can estimate mixture models with either numeric (type `numeric` and `integer`) or `ordered` indicators.
+Recall that `tidySEM` can estimate mixture models with either numeric (`numeric` or `integer`) or `ordered` type indicators.
 The column `type` should thus only contain these types.
@@
 normal distributions.
-To illustrate the problem, consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
+Consider what happens when estimating a 5-class solution on a single 5-point Likert scale indicator.
 Each class-specific mean would describe a single response
@@
 and a class-specific variance component would be empirically underidentified.
-In sum, variables with few unique values might not be suitable for Gaussian mixture modeling;
+In sum, variables with few unique values might not be suitable for Gaussian mixture modeling.
 LCA with ordinal indicators is more appropriate.
@@
 
-### Data preprocessing
+### Data Preprocessing
 
@@
 Numeric variables should have an interval measurement level,
-such that it is sensible to estimate means and (co)variances.
+such that it is sensible to estimate theirmeans and (co)variances.
 For the remaining three measurement levels,
@@
 A third consideration in preprocessing relates to the univariate distributions of indicators.
-For example, the `plas_depression` data show extreme right skew in depressive symptoms:
+For example, the `plas_depression` data shows extreme right skew in depressive symptoms:
 most respondents report few symptoms,
@@
 information without imputing missing values, by estimating the model based on all
-available information for each subject [@enders_relative_2001; enders_applied_2022].
+available information for each subject [@enders_relative_2001; @enders_applied_2022].
 FIML is a best-practice solution for handling missing data;
@@
 and conducting them on several imputed datasets multiplies the load.
-Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population, so the imputation model could bias the results as it does not account for the expected heterogeneity.
+Second, because many multiple imputation methods assume normality; an assumption likely to be violated when data come from a heterogeneous population, the imputation model could bias the results as it may account for the expected heterogeneity.
 
@@
 
-A second consideration in model specification is the risk of overfitting [@hastie_elements_2009].
+The second consideration in model specification is the risk of overfitting [@hastie_elements_2009].
 Overfitting occurs when a model has many parameters relative to the number of observations,
@@
 
-As there are no absolute fit indices for LCA,
-information criteria are typically used to assess relative fit.
-Information criteria are computed from the $-2LL$,
+Due to the lack of absolute fit indices for LCA, the assessment of relative fit
+relies on information criteria.
+Information criteria (ICs) are computed from the $-2LL$,
 which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
-Information criteria add a penalty for model complexity to the $-2LL$, and thus balance fit and complexity.
-The lower
-the value of an information criterion, the better the overall fit of the
+In addition to capturing model fit, ICs add a penalty for model complexity thus
+preferring more model fit and parsimony.
+The lower the value of an information criterion, the better the overall fit of the
 model.
-The original information criterion is the Akaike Information Criterion (AIC),
-computed as $-2LL+2p$, where $p$ is the number of parameters.
-The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
+The *Akaike Information Criterion* (*AIC*) is the original information criterion.
+It is computed as $-2LL+2p$, where $p$ is the number of parameters.
+The *Bayesian Information Criterion* (*BIC*) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
 such that complex models are penalized more in smaller samples.
 Finally,
-the so-called *sample size adjusted* BIC (saBIC) implements a different ad-hoc penalty for sample size,
-$\ln(\frac{n + 2}{24})$.
-Although simulation studies have showed that the saBIC performed well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
-its penalty is not based on theory.
-BIC also performs well and is theoretically substantiated;
-as such, it may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
-All three ICs are available in the output of `table_fit()`.
-
-Information criteria may occasionally contradict each other, so it is
+the so-called *sample size adjusted BIC* (*saBIC*) implements a different ad-hoc penalty for sample size,
+$\ln(\frac{n + 2}{24})$. Although simulation studies have showen that the saBIC performs well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
+a limitation is that its penalty is not based on theory. BIC also performs well and is theoretically substantiated.
+As such, BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
+In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.
+
+ICs may occasionally contradict each other, so it is
 important to identify a suitable strategy to reconcile them.
@@
 Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices. 
-Other approach is to select several competing model based on the ICs, and evaluate the theoretical interpretation.
+Alternatively, one may select several competing models based on information criteria, 
+and evaluate them based on the theoretical interpretation of the identified classes.
 Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process [@akogul_comparison_2016],
@@
 Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs [@nylund-gibson_ten_2018].
-The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
-Adding this many classes results in substantial decreases of the IC,
-and adding further classes results in relatively small decreases.
-
-Given a specific IC,
-it is also possible to calculate so-called "IC-weights",
-which indicate the relative support the data provide for each model in the set [@wagenmakers_aic_2004].
+The preferred number of classes can be found at the inflection point where the slope of the curve is clearly leveling off.
+Adding this many classes results in substantial decreases of the ICs,
+and adding additional classes may add too much complexity to the model while not lowering the IC value.
+
+Given a specific information criterion, it is also possible to calculate so-called "IC-weights",
+which indicate the relative support the data provides for each model in the set [@wagenmakers_aic_2004].
 This is particularly useful for confirmatory LCA,
 as it allows one to quantify the support for the theoretical model.
-IC-weights are calculated by calling `ic_weights()` on the model fit table.
+In `tidySEM`, IC-weights are calculated by calling `ic_weights()` on the model fit table.
 Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
@@
 
-When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
+When two models are nested, the significance of the difference in fit between these models can be tested using a *Likelihood Ratio* (*LR*) *test*.
 The test statistic is computed as the -2 times the log of the ratio of likelihoods,
@@
 One challenge is that the LR test statistic is not asymptotically $\chi^2$ distributed for LCA.
-The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR [@lo_testing_2001].
-This test is implemented in `tidySEM`, and is part of the default output of `table_fit()`, comparing subsequent rows of the table.
+The so-called *Lo-Mendell-Rubin (LMR) adjusted LR test* overcomes this problem by applying an ad-hoc correction to the LR [@lo_testing_2001].
+This test is implemented in `tidySEM`, and is a part of the default output of `table_fit()`.
 
 Another solution is to derive an empirical sampling distribution for the LR difference.
-The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the $k-1$-class model $B$ times (e.g., 1000),
+The so-called *Bootstrapped Likelihood Ratio Test (BLRT)* does so by simulating data from the $k-1$-class model $b$ times (e.g., 1000),
 and fitting the $k$- and $k-1$-class models to each simulated dataset [@nylund_deciding_2007]. 
-The p-value of this test is the proportion of $B$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
+The p-value of this test is the proportion of $b$ in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
 Simulation studies suggest that the BLRT has greater statistical power than the LRT,
-although both performed comparably well at detecting the true number of classes [@nylund_deciding_2007; @tein_statistical_2013].
+although both perform comparably well at detecting the true number of classes [@nylund_deciding_2007; @tein_statistical_2013].
 A crucial limitation of the BLRT is that it is very computationally expensive.
@@
 
-Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
+The first author's anecdotal experience as a statistical co-author on LCA studies suggests that both aforementioned tests tend to be liberal,
 returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
 Other authors have reported similar experiences [@sinha_practitioners_2021], which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting [@mcelreath_statistical_2020].
-This suggests that, like existing ICs,
-LR tests might not adequately balance fit and complexity in LCA.
+This suggests that, like existing ICs, LR tests might not adequately balance fit and complexity in LCA.
 Better solutions for class enumeration are thus an important target for future research.
@@
 which describes how well each model is able to reproduce the observed correlation matrix [see @mcelreath_statistical_2020].
-One version of the lazy bootstrap is already implemented in the commercial program *Latent GOLD* [@vermunt_upgrade_2021].
-At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
+A version of the lazy bootstrap is already implemented in the commercial program *Latent GOLD* [@vermunt_upgrade_2021].
+At the time of writing, we are developing a more general *simulation-based likelihood ratio test* (*SBLRT*) for class enumeration.
 Future versions of `tidySEM` will likely incorporate this test.
@@
 Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
-These diagnostics are not model fit indices and should not be used for class enumeration,
-because a model can fit the data well and still show poor latent class separation [@masyn_latent_2013].
-However, classification diagnostics can be used to eliminate models:
-If one analysis goal is to interpret properties of (members of) classes,
-it makes sense to only consider models that classify well [@nylund-gibson_ten_2018].
+These diagnostics are not model fit indices and should never be used for class enumeration.
+This is because a model can fit the data well and still exhibit poor separation between the latent classes [@masyn_latent_2013].
+Instead, classification diagnostics can be used to eliminate models:
+when the goal is to interpret properties of classes or their members,
+it makes sense to only consider models which are successful at separating latent classes [@nylund-gibson_ten_2018].
 Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.
 
-All classification diagnostics are derived from the posterior classification probabilities: an $n \times k$ matrix called $P$, with the probability that every individual $n$ belongs to latent class $k$.
+All classification diagnostics are derived from the posterior classification probabilities: an $n \times k$ matrix called $P$, with the probability that every individual $i$ belongs to latent class $k$
 When classification accuracy is high, and classes are distinct,
 each individual's posterior class probabilities are high for one class, and low for the remaining classes.
-The individual classification probabilities $P$ are obtained by running `class_prob(res, type = "individual")`.
+In `tidySEM`, individual classification probabilities $P$ are obtained by running `class_prob(res, type = "individual")`.
 Note that one column is added to the matrix, with each individual's most likely class membership $M$.
@@
 Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
+<!-- DA: what is many more? When making a rule a thumb, you have to eyeball a quantity and recommend it for users without statistical intuition-->
 Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
@@
 To assess classification accuracy, we turn to the third diagnostic table:
-The average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$, `class_prob(res, type = "avg.mostlikely")`:
-for participants assigned to one particular class (rows) according to $M$,
+the average posterior probabilities by most likely class membership, $P_{p_{(M = m)}}$
+which can be obtained using `class_prob(res, type = "avg.mostlikely")`.
+For participants assigned to one particular class (rows) according to $M$,
 this table gives the mean probability of being assigned to any class (columns).
@@
 $k$ subsets of observations with most likely class of $M = 1 \ldots k$.
-The diagonal represents the average predictive probability of a the subjects related to their most likely class, indicating the certainty of class assignment for the subjects.
-Some guidelines have being suggested for this diagonal, like for them to be higher than 0.7 [@masyn_latent_2013], but we suggest researchers to be cautions about these guidelines as they cannot be generalized and should not be use as strict cutoffs as evidence "good" or "bad" classification.
-
-The table of classification probabilities for the most likely latent class $M$ by latent class, $P_{p_{(M = m | C = c)}}$ is similar to the aforementioned table $P_{p_{(M = m)}}$,
+The diagonal represents the average predictive probability of the subjects related 
+to their most likely class, indicating the certainty of class assignment.
+Some guidelines have being suggested for this diagonal, 
+such as its values having to be higher than 0.7 [@masyn_latent_2013]. 
+We, however, advise caution as these guidelines cannot be generalized to all situations,
+and should not be used as strict cutoffs communication the classification quality.
+
+The table of classification probabilities for the most likely latent class $M$ by latent class, 
+$P_{p_{(M = m | C = c)}}$ is similar to the aforementioned $P_{p_{(M = m)}}$ table
 except that it accounts for classification uncertainty.
@@
 and $M$ is the most likely class,
-then this table shows the probability of an observation being assigned to class $m$ in $M$, given that its true class is $c$ in $C$, or $p(M=m|C=c)$.
+then this table shows the probability of an observation being assigned to class $m$ in $M$, given that its true class is $c$ in $C$, or $P(M=m|C=c)$.
 Values on the diagonal can thus be conceptualized as the reliability with which each level of the categorical latent variable is measured by $M$,
-and off-diagonal elements can be conceptualized as measurement error.
+and off-diagonal elements can be conceptualized as the measurement error.
 <!-- Mauricio: Can you fact-check this?!? I really struglle here. MGV: this looks good -->
-The minimum and maximum of the diagonal of this table are given by default as part of the output of `table_fit()` for LCA analyses (`prob_min` and `prob_max`).
+The minimum and maximum of the diagonal of this table are given by default as a part of the output of `table_fit()` for LCA analyses (`prob_min` and `prob_max`).
 These probabilities are of particular interest because they conveniently summarize classification accuracy per class.
@@
 the entropy criterion [@celeux_entropy_1996].
-Celeux, G., & Soromenho, G. (1996). 
 In physics, high entropy refers to maximum randomness or uncertainty,
 and low entropy corresponds to a strict arrangement of the units of study.
-Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
-Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
+Applied to LCA, high entropy means that classes are completely indistinguishable; 
+every individual has an equal probability of being in any class.
+Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in a single class, and a near-zero probability of being in any other class.
 Crucially, in `tidySEM`, the entropy criterion is defined as 1-entropy (a convention originating in *Mplus*).
-Its interpretation is reversed;
+Its interpretation is reversed:
 0 means the model classification is no better than chance,
 and 1 means perfect classification. 
-There are no rules of thumb for entropy,
-because entropy is also sensitive to the sample size (entropy tends to be lower in larger samples),
-number of latent classes (entropy tends to be lower for higher number of classes), 
-and number of indicators.
+There are no rules of thumb for entropy. As a measure, it will vary based on data characteristics:
+entropy tends to be lower in larger samples, datasets with a large number of indicators and latent classes.
 Like other classification diagnostics,
 entropy is not a model fit measure and should not be used for model selection [@masyn_latent_2013].
-However, it can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
+It can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
 
@@
 These class names should aid the interpretation of the class solutions and serve as a shorthand throughout the manuscript.
-For example, one study used a dual-trajectory LCGA of cognitive and affective empathy development with gender as a covariate.
+For example, a study used a dual-trajectory LCGA of cognitive and affective empathy development with gender as a covariate.
 Although this model has dozens of parameters,
@@
 However, overly simplistic and general class names can be misleading, a phenomenon known as the *naming fallacy* [@weller_latent_2020].
-For example, it would be misleading to call one group "low empathy" and another group "medium empathy" if the empathy differences were non-significant,
-or if other differences, like the effect of the covariate gender, were more substantial than level differences.
-One way to avoid the naming fallacy is comprehensively reporting model parameters,
-and explicating that class names are just a shorthand.
+For example, it would be misleading to call one group "low empathy" and another group "medium empathy" if the empathy differences are non-significant,
+or if other differences, like the effect of the covariate gender, were more substantial than empathy level differences.
+A way to avoid the naming fallacy is to comprehensively report model parameters,
+and explicate that class names are just a shorthand which may not capture the full ground-truth complexity.
 
@@
 This means that the order of classes is arbitrary and may change if the analysis is replicated.
-A simple solution to ensure replicable results is setting a random seed before fitting the model, using `set.seed()`.
+A simple solution to ensure replicable results is setting a random seed before fitting the model by using `set.seed()`.
 This will ensure that the classes stay in the same order upon replication, 
@@
 To address this problem, the function `mx_switch_labels()` can be used.
-By default, the function orders classes from largest to smallest class count - but classes can also be ordered by values of a specific parameter (e.g., the mean intercept in a latent class growth model).
+By default, the function orders classes from largest to smallest class count - but classes can also be ordered by their values on a specific parameter (e.g., the mean intercept in a latent class growth model).
 Note that label switching is a much bigger problem when using Bayesian estimation,
@@
 
-LCA is rarely performed in isolation;
-often, researchers wish to relate class membership to auxiliary variables, or even auxiliary models.
+LCA is rarely performed in isolation.
+Often, researchers wish to relate class membership to auxiliary variables, or even auxiliary models.
 It is important to keep in mind that model-implied class membership is subject to classification error.
 Treating class membership as an observed variable in follow-up analyses disregards this error.
-The results will be biased,
-except when class separation is very high - as indicated by high (near 1) entropy and minimum posterior classification probability.
+The resulting models will be biased, except when class separation is very high 
+- as indicated by entropy values and minimum posterior classification probability approaching 1.
 It is therefore important to account for classification error.
 
-There are several different ways to do follow-up analyses in latent class analysis (LCA) while accounting for classification error.
-At present, the so-called BCH method is considered one of the best approaches [@bolck_estimating_2004].
-It has been implemented in `tidySEM`.
-The advantage of the BCH method is that it not only allows for the comparison of individual variables, but also entire auxiliary models across latent classes.
+There are several different ways to do follow-up analyses in LCA while accounting for classification error.
+At present, the so-called *BCH method* is considered one of the best approaches [@bolck_estimating_2004].
+It has been implemented in `tidySEM`. The advantage of the BCH method is that it not only allows for the comparison of individual variables, 
+but also entire auxiliary models across latent classes.
 
@@
 @weller_latent_2020], but more work is needed. 
+In this section, we provide a comprehensive set of guidelines which are evidence-based and 
+adhere to the state-of-the-art practices and trends in science.
 
@@
 In our experience, reviewers (and authors) often disagree about which criteria should be used.
-Preregistering the class enumeration criteria as part of the analysis plan reduces disagreement and eliminates concerns about cherry picking fit measures after the results are known.
+Preregistering the class enumeration criteria as a part of the analysis plan reduces disagreement and eliminates concerns about cherry picking fit measures after the results are known.
 It is particularly important to pre-specify the criteria that will be used to determine the maximum number of classes,
@@
 based on most likely class membership.
-Fifth, the LMR or BLRT can be reported if it is part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
+Fifth, the LMR or BLRT can be reported if it is a part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
 By default, `table_fit()` reports the VLM for subsequent models in the table.
 
-### Reporting results
+<!-- DA: We do not introduce VLM, what does it stand for? -->
+
+### Reporting Results
 
@@
 The class proportions are obtained by standardizing the class weights:
-A $k$-class model has $k-1$ free class weights,
+a $k$-class model has $k-1$ free class weights,
 and one weight constrained to 1 for model identification.
@@
 but are redundant with the aforementioned class proportions.
-Other model parameters may include means, (co)variances, regression coefficients (rarely), and thresholds for ordinal indicators.
+Other model parameters may include means, (co)variances, thresholds for ordinal indicators, and (rarely) regression coefficients.
 It is good practice to report the point estimate and confidence interval of these parameters.
@@
 we found that most visualizations focused on the estimated model parameters,
-displaying class means as points connected by lines, with point shape and linetype indicating class membership.
+displaying class means as points connected by lines, with point, shape and line type indicating class membership.
 Note that the lines introduce spurious information.
@@
 
-The Grammar of Graphics addresses best practices in visualization, which we summarize here with specific emphasis on the application to LCA.
+The Grammar of Graphics [@wilkinson_grammar_2005; @wickham_layered_2010] 
+addresses best practices in visualization, 
+which we summarize here with specific emphasis on the application to LCA.
 Most importantly, good graphics start with the *data* in mind.
@@
 In a black and white plot, it can be indicated by linetype and point shape.
-A third consideration is which geometric objects (geoms) to use.
-As mentioned before, common practice is to plot parameter estimates as points, and connect these points with lines.
+A third consideration is which *geometric objects* (*geoms*) to use.
+As mentioned before, a common practice is to plot parameter estimates as points,
+and connect these points with lines.
 Points are suitable geoms to visualize point estimates and raw data.
@@
 it makes sense to *facet* a plot into multiple panels by class solution.
-The plots in `tidySEM` are based on best practices as outlined in the grammar of graphics,
-and are implemented in `ggplot2`.
+The plots in `tidySEM` are based on best grammar of graphics practices,
+and are implemented in `ggplot2` (@wickham_ggplot2_2016).
 As such, these plots can be further modified just like any other `ggplot2` object.
@@
 Fourth, previous reporting standards called for reporting of sample descriptive statistics [see @schreiber_latent_2017].
-While we agree that this is due dilligence,
+While we agree that this is due diligence,
 sample descriptives have two limitations in the context of LCA:
@@
 
-The best practices discussed in this paper are summarized in the SMART-LCA Checklist: Standards for More Accuracy in Reporting of different Types of Latent Class Analysis.
+The best practices discussed in this paper are summarized in the *SMART-LCA Checklist*: *Standards for More Accuracy in Reporting of different Types of Latent Class Analysis*.
 
@@
 
-Write me
+We set ourselves several goals with this tutorial:
+
+(1) Providing a comprehensive primer of LCA models
+
+We defined the most prominent models in the LCA family and showcased various use cases where researchers can benefit from using
+mixture models. The Appendices of the present work provide step-by-step instructions to multiple types of LCA models.
+
+(2) Providing guidance with respect to best practices in estimation and reporting of mixture models.
+
+We expanded on several existing guidelines so that they encompass both the full variety of LCA models, 
+as well as embody the highest scientific standards of the day. These guidelines are intended to be used 
+by researchers aiming to perform robust and insightful analyses, as well as evaluators wishing to assess the quality and validity of preexisting work.
+
+(3) Creating a simultaneously most powerful and simple R package for fitting LCA models
+
+We discussed the full breadth of LCA tools and showcased how these can be implemented in a low threshold - high ceiling environment.
+Beginners will find the `tidySEM` syntax highly intuitive and easy to master, while more advanced users will enjoy its compatibility and integration with various preexisting packages.
+Both classes of users will benefit from the fact that `tidySEM` provides a free open source access to state-of-the-art LCA models and tools.
+
+With all of this said, we hope that our work will have lowered the barrier of entry to the exciting world of latent class analysis and incentivize more applied and theoretical research in the field.
 

### 2023-02-16 manuscript.tex

@@
 {}{\PatchFailed}
-\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 9001}
+\keywords{latent class analysis, mixture models, best practices, free open source software, tidySEM\newline\indent Word count: 12158}
 \usepackage{csquotes}
@@
 The use of LCA for dimension reduction is common in longitudinal research,
-where one developmental process is summarized using latent class growth analysis (LCGA),
+where one developmental process is summarized using \emph{latent class growth analysis} (\emph{LCGA}),
 and the resulting categories are then used as a moderator of another developmental process in a multi-group model.
@@
 dichotomous items have \(2^{15}\) or \(32,768\) unique symptom combinations.
-However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data.
+However, LCA was able to reduce response patterns to five clinically meaningful latent classes supported by the data,
+which makes intervention-planning significantly more manageable.
 
@@
 \hypertarget{violations-of-normality}{%
-\subsubsection{Violations of normality}\label{violations-of-normality}}
+\subsubsection{Violations of Normality}\label{violations-of-normality}}
 
 Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) {[}Baughman (2007); magnus\_multidimensional\_2021{]}.
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (Baughman, 2007; Magnus \& Garnier-Villarreal, 2022).
 These models estimate one class to explain the ``true zeros'',
@@
 The best practices in estimation, as outlined in Table
-\ref{tab:checkest}, are rooted in existing recommendations for best
-practices for estimating specific subtypes of LCA,
+\ref{tab:checkest-tab}, are rooted in existing recommendations
+for estimating specific subtypes of LCA,
 including latent
@@
 class analysis with ordinal indicators (e.g., Nylund-Gibson \& Choi, 2018).
-These were generalized to be relevant to all types of LCA, and updated to current best practices, as explained below.
+These guidelines were generalized to be relevant to all types of LCA,
+and updated to current best practices, as explained below.
 
@@
 
+\begin{table}
+
+\caption{\label{tab:checkest-tab}}
+\centering
 \begin{tabular}[t]{l|l}
@@
 \end{tabular}
+\end{table}
 
@@
 variance of continuous and categorical variables,
-missingness, skew and kurtosis (peakedness of the distribution).
+their missingness, skew and kurtosis.
 However, sample-level descriptive statistics are of limited value when the goal of a study is to identify subsamples using LCA.
@@
 for categorical ones.
-Density plots can also aid in the choice of the number of classes, as explained in the section on visualization.
+Density plots can also aid in the choice of the number of classes, as explained in the Visualization section.
 
 Particular points of attention include measurement level of the indicators.
-Recall that \texttt{tidySEM} can estimate mixture models with either numeric (type \texttt{numeric} and \texttt{integer}) or \texttt{ordered} indicators.
+Recall that \texttt{tidySEM} can estimate mixture models with either numeric (\texttt{numeric} or \texttt{integer}) or \texttt{ordered} type indicators.
 The column \texttt{type} should thus only contain these types.
@@
 \hypertarget{data-preprocessing}{%
-\subsubsection{Data preprocessing}\label{data-preprocessing}}
+\subsubsection{Data Preprocessing}\label{data-preprocessing}}
 
@@
 A third consideration in preprocessing relates to the univariate distributions of indicators.
-For example, the \texttt{plas\_depression} data show extreme right skew in depressive symptoms:
+For example, the \texttt{plas\_depression} data shows extreme right skew in depressive symptoms:
 most respondents report few symptoms,
@@
 information without imputing missing values, by estimating the model based on all
-available information for each subject {[}C. Enders and Bandalos (2001); enders\_applied\_2022{]}.
+available information for each subject (C. K. Enders, 2022; C. Enders \& Bandalos, 2001).
 FIML is a best-practice solution for handling missing data;
@@
 
-A second consideration in model specification is the risk of overfitting (Hastie, Tibshirani, \& Friedman, 2009).
+The second consideration in model specification is the risk of overfitting (Hastie, Tibshirani, \& Friedman, 2009).
 Overfitting occurs when a model has many parameters relative to the number of observations,
@@
 
-As there are no absolute fit indices for LCA,
-information criteria are typically used to assess relative fit.
-Information criteria are computed from the \(-2LL\),
+Due to the lack of absolute fit indices for LCA, the assessment of relative fit
+relies on information criteria.
+Information criteria (ICs) are computed from the \(-2LL\),
 which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
-Information criteria add a penalty for model complexity to the \(-2LL\), and thus balance fit and complexity.
-The lower
-the value of an information criterion, the better the overall fit of the
+In addition to capturing model fit, ICs add a penalty for model complexity thus
+preferring more model fit and parsimony.
+The lower the value of an information criterion, the better the overall fit of the
 model.
-The original information criterion is the Akaike Information Criterion (AIC),
-computed as \(-2LL+2p\), where \(p\) is the number of parameters.
-The Bayesian Information Criterion (BIC) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
+The \emph{Akaike Information Criterion} (\emph{AIC}) is the original information criterion.
+It is computed as \(-2LL+2p\), where \(p\) is the number of parameters.
+The \emph{Bayesian Information Criterion} (\emph{BIC}) multiplies the penalty for complexity with a function of the sample size, \(p \ln(n)\),
 such that complex models are penalized more in smaller samples.
 Finally,
-the so-called \emph{sample size adjusted} BIC (saBIC) implements a different ad-hoc penalty for sample size,
-\(\ln(\frac{n + 2}{24})\).
-Although simulation studies have showed that the saBIC performed well for class enumeration (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013),
-its penalty is not based on theory.
-BIC also performs well and is theoretically substantiated;
-as such, it may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
-All three ICs are available in the output of \texttt{table\_fit()}.
-
-Information criteria may occasionally contradict each other, so it is
+the so-called \emph{sample size adjusted BIC} (\emph{saBIC}) implements a different ad-hoc penalty for sample size,
+\(\ln(\frac{n + 2}{24})\). Although simulation studies have showen that the saBIC performs well for class enumeration (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013),
+a limitation is that its penalty is not based on theory. BIC also performs well and is theoretically substantiated.
+As such, BIC may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+In \texttt{tidySEM}, all three ICs are available as a part of the output of \texttt{table\_fit()}.
+
+ICs may occasionally contradict each other, so it is
 important to identify a suitable strategy to reconcile them.
@@
 Another option is to always prefer the most parsimonious model that has best fit according to any of the available fit indices.
-Other approach is to select several competing model based on the ICs, and evaluate the theoretical interpretation.
+Alternatively, one may select several competing models based on information criteria,
+and evaluate them based on the theoretical interpretation of the identified classes.
 Yet another option is to incorporate information from multiple fit indices using the analytic hierarchy process (Akogul \& Erisoglu, 2016),
@@
 Finally, one can use a scree plot to determine the inflection point at which additional classes contribute little to further decreases of the ICs (Nylund-Gibson \& Choi, 2018).
-The preferred number of classes is at the inflection point where the slope of the curve is clearly leveling off.
-Adding this many classes results in substantial decreases of the IC,
-and adding further classes results in relatively small decreases.
-
-Given a specific IC,
-it is also possible to calculate so-called ``IC-weights'',
-which indicate the relative support the data provide for each model in the set (Wagenmakers \& Farrell, 2004).
+The preferred number of classes can be found at the inflection point where the slope of the curve is clearly leveling off.
+Adding this many classes results in substantial decreases of the ICs,
+and adding additional classes may add too much complexity to the model while not lowering the IC value.
+
+Given a specific information criterion, it is also possible to calculate so-called ``IC-weights'',
+which indicate the relative support the data provides for each model in the set (Wagenmakers \& Farrell, 2004).
 This is particularly useful for confirmatory LCA,
 as it allows one to quantify the support for the theoretical model.
-IC-weights are calculated by calling \texttt{ic\_weights()} on the model fit table.
+In \texttt{tidySEM}, IC-weights are calculated by calling \texttt{ic\_weights()} on the model fit table.
 Note, however, that IC-weights will (strongly) favor the model with the lowest IC value;
@@
 
-When two models are nested, the significance of the difference in fit between these models can be tested using a Likelihood Ratio test (LR).
+When two models are nested, the significance of the difference in fit between these models can be tested using a \emph{Likelihood Ratio} (\emph{LR}) \emph{test}.
 The test statistic is computed as the -2 times the log of the ratio of likelihoods,
@@
 One challenge is that the LR test statistic is not asymptotically \(\chi^2\) distributed for LCA.
-The so-called Lo-Mendell-Rubin adjusted LR test (LMR) overcomes this problem by applying an ad-hoc correction to the LR (Lo, Mendell, \& Rubin, 2001).
-This test is implemented in \texttt{tidySEM}, and is part of the default output of \texttt{table\_fit()}, comparing subsequent rows of the table.
+The so-called \emph{Lo-Mendell-Rubin (LMR) adjusted LR test} overcomes this problem by applying an ad-hoc correction to the LR (Lo, Mendell, \& Rubin, 2001).
+This test is implemented in \texttt{tidySEM}, and is a part of the default output of \texttt{table\_fit()}.
 
 Another solution is to derive an empirical sampling distribution for the LR difference.
-The so-called bootstrapped Likelihood Ratio Test (BLRT) does so by simulating data from the \(k-1\)-class model \(B\) times (e.g., 1000),
+The so-called \emph{Bootstrapped Likelihood Ratio Test (BLRT)} does so by simulating data from the \(k-1\)-class model \(b\) times (e.g., 1000),
 and fitting the \(k\)- and \(k-1\)-class models to each simulated dataset (Nylund et al., 2007).
-The p-value of this test is the proportion of \(B\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
+The p-value of this test is the proportion of \(b\) in which the LR of the models fit to simulated data exceeds the LR observed in the real data.
 Simulation studies suggest that the BLRT has greater statistical power than the LRT,
-although both performed comparably well at detecting the true number of classes (Nylund et al., 2007; Tein et al., 2013).
+although both perform comparably well at detecting the true number of classes (Nylund et al., 2007; Tein et al., 2013).
 A crucial limitation of the BLRT is that it is very computationally expensive.
@@
 
-Anecdotal experience as a statistical co-author on LCA studies suggests that both these tests tend to be liberal,
+The first author's anecdotal experience as a statistical co-author on LCA studies suggests that both aforementioned tests tend to be liberal,
 returning significant p-values for models with more classes even when other indicators suggest that those models are overfitted (e.g., near-zero class counts).
 Other authors have reported similar experiences (Sinha, Calfee, \& Delucchi, 2021), which matches the differences between LRT and ICs from information theory as LRT tests and compare the models based on in sample predictive accuracy, and ICs evaluates and compare the models out-of-sample predictive accuracy which helps to control for overfitting (McElreath, 2020).
-This suggests that, like existing ICs,
-LR tests might not adequately balance fit and complexity in LCA.
+This suggests that, like existing ICs, LR tests might not adequately balance fit and complexity in LCA.
 Better solutions for class enumeration are thus an important target for future research.
@@
 which describes how well each model is able to reproduce the observed correlation matrix (see McElreath, 2020).
-One version of the lazy bootstrap is already implemented in the commercial program \emph{Latent GOLD} (Vermunt \& Magidson, 2021).
-At the time of writing, we are developing a more general simulation-based likelihood ratio test (SBLRT) for class enumeration.
+A version of the lazy bootstrap is already implemented in the commercial program \emph{Latent GOLD} (Vermunt \& Magidson, 2021).
+At the time of writing, we are developing a more general \emph{simulation-based likelihood ratio test} (\emph{SBLRT}) for class enumeration.
 Future versions of \texttt{tidySEM} will likely incorporate this test.
@@
 Classification diagnostics assess whether the classes identified by LCA are distinct and clearly separable.
-These diagnostics are not model fit indices and should not be used for class enumeration,
-because a model can fit the data well and still show poor latent class separation (Masyn, 2013).
-However, classification diagnostics can be used to eliminate models:
-If one analysis goal is to interpret properties of (members of) classes,
-it makes sense to only consider models that classify well (Nylund-Gibson \& Choi, 2018).
+These diagnostics are not model fit indices and should never be used for class enumeration.
+This is because a model can fit the data well and still exhibit poor separation between the latent classes (Masyn, 2013).
+Instead, classification diagnostics can be used to eliminate models:
+when the goal is to interpret properties of classes or their members,
+it makes sense to only consider models which are successful at separating latent classes (Nylund-Gibson \& Choi, 2018).
 Classification diagnostics should thus not be used for model selection, but they can help restrict the search space of acceptable models.
 
-All classification diagnostics are derived from the posterior classification probabilities: an \(n \times k\) matrix called \(P\), with the probability that every individual \(n\) belongs to latent class \(k\).
+All classification diagnostics are derived from the posterior classification probabilities: an \(n \times k\) matrix called \(P\), with the probability that every individual \(i\) belongs to latent class \(k\)
 When classification accuracy is high, and classes are distinct,
 each individual's posterior class probabilities are high for one class, and low for the remaining classes.
-The individual classification probabilities \(P\) are obtained by running \texttt{class\_prob(res,\ type\ =\ "individual")}.
+In \texttt{tidySEM}, individual classification probabilities \(P\) are obtained by running \texttt{class\_prob(res,\ type\ =\ "individual")}.
 Note that one column is added to the matrix, with each individual's most likely class membership \(M\).
@@
 To assess classification accuracy, we turn to the third diagnostic table:
-The average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\), \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}:
-for participants assigned to one particular class (rows) according to \(M\),
+the average posterior probabilities by most likely class membership, \(P_{p_{(M = m)}}\)
+which can be obtained using \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}.
+For participants assigned to one particular class (rows) according to \(M\),
 this table gives the mean probability of being assigned to any class (columns).
@@
 \(k\) subsets of observations with most likely class of \(M = 1 \ldots k\).
-The diagonal represents the average predictive probability of a the subjects related to their most likely class, indicating the certainty of class assignment for the subjects.
-Some guidelines have being suggested for this diagonal, like for them to be higher than 0.7 (Masyn, 2013), but we suggest researchers to be cautions about these guidelines as they cannot be generalized and should not be use as strict cutoffs as evidence ``good'' or ``bad'' classification.
-
-The table of classification probabilities for the most likely latent class \(M\) by latent class, \(P_{p_{(M = m | C = c)}}\) is similar to the aforementioned table \(P_{p_{(M = m)}}\),
+The diagonal represents the average predictive probability of the subjects related
+to their most likely class, indicating the certainty of class assignment.
+Some guidelines have being suggested for this diagonal,
+such as its values having to be higher than 0.7 (Masyn, 2013).
+We, however, advise caution as these guidelines cannot be generalized to all situations,
+and should not be used as strict cutoffs communication the classification quality.
+
+The table of classification probabilities for the most likely latent class \(M\) by latent class,
+\(P_{p_{(M = m | C = c)}}\) is similar to the aforementioned \(P_{p_{(M = m)}}\) table
 except that it accounts for classification uncertainty.
@@
 and \(M\) is the most likely class,
-then this table shows the probability of an observation being assigned to class \(m\) in \(M\), given that its true class is \(c\) in \(C\), or \(p(M=m|C=c)\).
+then this table shows the probability of an observation being assigned to class \(m\) in \(M\), given that its true class is \(c\) in \(C\), or \(P(M=m|C=c)\).
 Values on the diagonal can thus be conceptualized as the reliability with which each level of the categorical latent variable is measured by \(M\),
-and off-diagonal elements can be conceptualized as measurement error.
-The minimum and maximum of the diagonal of this table are given by default as part of the output of \texttt{table\_fit()} for LCA analyses (\texttt{prob\_min} and \texttt{prob\_max}).
+and off-diagonal elements can be conceptualized as the measurement error.
+The minimum and maximum of the diagonal of this table are given by default as a part of the output of \texttt{table\_fit()} for LCA analyses (\texttt{prob\_min} and \texttt{prob\_max}).
 These probabilities are of particular interest because they conveniently summarize classification accuracy per class.
@@
 the entropy criterion (Celeux \& Soromenho, 1996).
-Celeux, G., \& Soromenho, G. (1996).
 In physics, high entropy refers to maximum randomness or uncertainty,
 and low entropy corresponds to a strict arrangement of the units of study.
-Applied to LCA, high entropy means that classes are completely indistinguishable; every individual has equal probability of being in any class.
-Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in one class only, and near-zero probability of being in any other class.
+Applied to LCA, high entropy means that classes are completely indistinguishable;
+every individual has an equal probability of being in any class.
+Low entropy, by contrast, means that classes are fully separated: every individual has a near-one probability of being in a single class, and a near-zero probability of being in any other class.
 Crucially, in \texttt{tidySEM}, the entropy criterion is defined as 1-entropy (a convention originating in \emph{Mplus}).
-Its interpretation is reversed;
+Its interpretation is reversed:
 0 means the model classification is no better than chance,
 and 1 means perfect classification.
-There are no rules of thumb for entropy,
-because entropy is also sensitive to the sample size (entropy tends to be lower in larger samples),
-number of latent classes (entropy tends to be lower for higher number of classes),
-and number of indicators.
+There are no rules of thumb for entropy. As a measure, it will vary based on data characteristics:
+entropy tends to be lower in larger samples, datasets with a large number of indicators and latent classes.
 Like other classification diagnostics,
 entropy is not a model fit measure and should not be used for model selection (Masyn, 2013).
-However, it can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
+It can be used to disqualify certain solutions if the goal is to identify clearly distinct classes.
 
@@
 However, overly simplistic and general class names can be misleading, a phenomenon known as the \emph{naming fallacy} (Weller et al., 2020).
-For example, it would be misleading to call one group ``low empathy'' and another group ``medium empathy'' if the empathy differences were non-significant,
-or if other differences, like the effect of the covariate gender, were more substantial than level differences.
-One way to avoid the naming fallacy is comprehensively reporting model parameters,
-and explicating that class names are just a shorthand.
+For example, it would be misleading to call one group ``low empathy'' and another group ``medium empathy'' if the empathy differences are non-significant,
+or if other differences, like the effect of the covariate gender, were more substantial than empathy level differences.
+A way to avoid the naming fallacy is to comprehensively report model parameters,
+and explicate that class names are just a shorthand which may not capture the full ground-truth complexity.
 
@@
 This means that the order of classes is arbitrary and may change if the analysis is replicated.
-A simple solution to ensure replicable results is setting a random seed before fitting the model, using \texttt{set.seed()}.
+A simple solution to ensure replicable results is setting a random seed before fitting the model by using \texttt{set.seed()}.
 This will ensure that the classes stay in the same order upon replication,
@@
 To address this problem, the function \texttt{mx\_switch\_labels()} can be used.
-By default, the function orders classes from largest to smallest class count - but classes can also be ordered by values of a specific parameter (e.g., the mean intercept in a latent class growth model).
+By default, the function orders classes from largest to smallest class count - but classes can also be ordered by their values on a specific parameter (e.g., the mean intercept in a latent class growth model).
 Note that label switching is a much bigger problem when using Bayesian estimation,
@@
 
-LCA is rarely performed in isolation;
-often, researchers wish to relate class membership to auxiliary variables, or even auxiliary models.
+LCA is rarely performed in isolation.
+Often, researchers wish to relate class membership to auxiliary variables, or even auxiliary models.
 It is important to keep in mind that model-implied class membership is subject to classification error.
 Treating class membership as an observed variable in follow-up analyses disregards this error.
-The results will be biased,
-except when class separation is very high - as indicated by high (near 1) entropy and minimum posterior classification probability.
+The resulting models will be biased, except when class separation is very high
+- as indicated by entropy values and minimum posterior classification probability approaching 1.
 It is therefore important to account for classification error.
 
-There are several different ways to do follow-up analyses in latent class analysis (LCA) while accounting for classification error.
-At present, the so-called BCH method is considered one of the best approaches (Bolck, Croon, \& Hagenaars, 2004).
-It has been implemented in \texttt{tidySEM}.
-The advantage of the BCH method is that it not only allows for the comparison of individual variables, but also entire auxiliary models across latent classes.
+There are several different ways to do follow-up analyses in LCA while accounting for classification error.
+At present, the so-called \emph{BCH method} is considered one of the best approaches (Bolck, Croon, \& Hagenaars, 2004).
+It has been implemented in \texttt{tidySEM}. The advantage of the BCH method is that it not only allows for the comparison of individual variables,
+but also entire auxiliary models across latent classes.
 
@@
 standardize ways of reporting LCA (e.g. Masyn, 2013; Weller et al., 2020), but more work is needed.
+In this section, we provide a comprehensive set of guidelines which are evidence-based and
+adhere to the state-of-the-art practices and trends in science.
 
@@
 based on most likely class membership.
-Fifth, the LMR or BLRT can be reported if it is part of the class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}.
+Fifth, the LMR or BLRT can be reported if it is a part of the class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}.
 By default, \texttt{table\_fit()} reports the VLM for subsequent models in the table.
@@
 \hypertarget{reporting-results}{%
-\subsubsection{Reporting results}\label{reporting-results}}
+\subsubsection{Reporting Results}\label{reporting-results}}
 
@@
 The class proportions are obtained by standardizing the class weights:
-A \(k\)-class model has \(k-1\) free class weights,
+a \(k\)-class model has \(k-1\) free class weights,
 and one weight constrained to 1 for model identification.
@@
 
-The Grammar of Graphics addresses best practices in visualization, which we summarize here with specific emphasis on the application to LCA.
+The Grammar of Graphics (Wickham, 2010; Wilkinson, 2005)
+addresses best practices in visualization,
+which we summarize here with specific emphasis on the application to LCA.
 Most importantly, good graphics start with the \emph{data} in mind.
@@
 In a black and white plot, it can be indicated by linetype and point shape.
-A third consideration is which geometric objects (geoms) to use.
-As mentioned before, common practice is to plot parameter estimates as points, and connect these points with lines.
+A third consideration is which \emph{geometric objects} (\emph{geoms}) to use.
+As mentioned before, a common practice is to plot parameter estimates as points,
+and connect these points with lines.
 Points are suitable geoms to visualize point estimates and raw data.
@@
 it makes sense to \emph{facet} a plot into multiple panels by class solution.
-The plots in \texttt{tidySEM} are based on best practices as outlined in the grammar of graphics,
-and are implemented in \texttt{ggplot2}.
+The plots in \texttt{tidySEM} are based on best grammar of graphics practices,
+and are implemented in \texttt{ggplot2} (Wickham (2016)).
 As such, these plots can be further modified just like any other \texttt{ggplot2} object.
@@
 Fourth, previous reporting standards called for reporting of sample descriptive statistics (see Schreiber, 2017).
-While we agree that this is due dilligence,
+While we agree that this is due diligence,
 sample descriptives have two limitations in the context of LCA:
@@
 
-The best practices discussed in this paper are summarized in the SMART-LCA Checklist: Standards for More Accuracy in Reporting of different Types of Latent Class Analysis.
+The best practices discussed in this paper are summarized in the \emph{SMART-LCA Checklist}: \emph{Standards for More Accuracy in Reporting of different Types of Latent Class Analysis}.
 
@@
 
-Write me
+We set ourselves several goals with this tutorial:
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\tightlist
+\item
+  Providing a comprehensive primer of LCA models
+\end{enumerate}
+
+We defined the most prominent models in the LCA family and showcased various use cases where researchers can benefit from using
+mixture models. The Appendices of the present work provide step-by-step instructions to multiple types of LCA models.
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\setcounter{enumi}{1}
+\tightlist
+\item
+  Providing guidance with respect to best practices in estimation and reporting of mixture models.
+\end{enumerate}
+
+We expanded on several existing guidelines so that they encompass both the full variety of LCA models,
+as well as embody the highest scientific standards of the day. These guidelines are intended to be used
+by researchers aiming to perform robust and insightful analyses, as well as evaluators wishing to assess the quality and validity of preexisting work.
+
+\begin{enumerate}
+\def\labelenumi{(\arabic{enumi})}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  Creating a simultaneously most powerful and simple R package for fitting LCA models
+\end{enumerate}
+
+We discussed the full breadth of LCA tools and showcased how these can be implemented in a low threshold - high ceiling environment.
+Beginners will find the \texttt{tidySEM} syntax highly intuitive and easy to master, while more advanced users will enjoy its compatibility and integration with various preexisting packages.
+Both classes of users will benefit from the fact that \texttt{tidySEM} provides a free open source access to state-of-the-art LCA models and tools.
+
+With all of this said, we hope that our work will have lowered the barrier of entry to the exciting world of latent class analysis and incentivize more applied and theoretical research in the field.
 
@@
 \url{https://doi.org/10.1007/s10964-022-01665-7}
+
+\leavevmode\vadjust pre{\hypertarget{ref-magnus_multidimensional_2022}{}}%
+Magnus, B.
+E., \& Garnier-Villarreal, M.
+(2022).
+A multidimensional zero-inflated graded response model for ordinal symptom data.
+\emph{Psychological Methods}, \emph{27}, 261--279.
+\url{https://doi.org/10.1037/met0000395}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_layered_2010}{}}%
+Wickham, H. (2010). A layered grammar of graphics. \emph{Journal of Computational and Graphical Statistics}, \emph{19}(1), 3--28. \url{https://doi.org/10.1198/jcgs.2009.07098}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_ggplot2_2016}{}}%
+Wickham, H. (2016). \emph{ggplot2: Elegant graphics for data analysis}. Springer-Vertlag New York. Retrieved from \url{https://ggplot2.tidyverse.org}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wilkinson_grammar_2005}{}}%
+Wilkinson, L. (2005). \emph{The grammar of graphics}. New York: Springer-Verlag. \url{https://doi.org/10.1007/0-387-28695-0}
+
 \end{CSLReferences}

### 2023-02-23 manuscript.Rmd
edits from to do list, MGV

@@
 <!-- CJ: This example is not perfect, we need another example still.. -->
-For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
-A mixture of two normal distributions with distinct variances described the observed distribution better.
+For instance, a depression questionnaire PH-9 have found 2 distinct groups, ones with no symptoms, and one with symptoms.
+A mixture model can model the "presence" and "frequency" of symptoms as separate classes, instead of assuming that the sample follows a single continuous underlying distribution [@magnus_multidimensional_2022].
+<!-- the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
+A mixture of two normal distributions with distinct variances described the observed distribution better. -->
 <!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical.
@@
 then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
-By default, `tidySEM` derives starting values using K-means clustering [@likas_global_2003],
+By default, `tidySEM` derives starting values using K-means clustering [@likas_global_2003; @Forgy1965ClusterAO; @Hartigan_kmeans],
 computing the mixture model starting values by treating the K-means solution as multi-group model.
@@
 the so-called *sample size adjusted BIC* (*saBIC*) implements a different ad-hoc penalty for sample size,
-$\ln(\frac{n + 2}{24})$. Although simulation studies have showen that the saBIC performs well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
-a limitation is that its penalty is not based on theory. BIC also performs well and is theoretically substantiated.
+$\ln(\frac{n + 2}{24})$. Although simulation studies have shown that the saBIC performs well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
+a limitation is that its penalty is not based on theory, but on the Minimum Description Length (MDL) as the shortes set of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987].
+BIC also performs well and is theoretically substantiated.
 As such, BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
@@
 however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
-Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
-<!-- DA: what is many more? When making a rule a thumb, you have to eyeball a quantity and recommend it for users without statistical intuition-->
+Thus, a better rule of thumb might be to ascertain that the smallest class has at least as many observations as the average number of parameters per class.
+<!-- DA: what is many more? When making a rule a thumb, you have to eyeball a quantity and recommend it for users without statistical intuition
+MGV: here I changed it "at least as many", as a low bound recommendation-->
 Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
@@
 which can be obtained using `class_prob(res, type = "avg.mostlikely")`.
-For participants assigned to one particular class (rows) according to $M$,
-this table gives the mean probability of being assigned to any class (columns).
+For participants assigned to one particular class (rows) according to $M$ (most likely class based on predicted probabilities),
+this table gives the mean probability of being assigned to each class (columns).
 This table is obtained by calculating the column means of $P$ for the 
@@
 Fifth, the LMR or BLRT can be reported if it is a part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
-By default, `table_fit()` reports the VLM for subsequent models in the table.
+By default, `table_fit()` reports the LMR (also called Vuong’s robust likelihood-ratio test, VLMR) for subsequent models in the table.
 

### 2023-03-24 manuscript.Rmd
update title page, add authorship CREDIT

@@
 ---
-title             : "Best Practices in Latent Class Analysis using Free Open Source Software"
-shorttitle        : "BEST PRACTICES IN LATENT CLASS ANALYSIS"
+title             : "Best Practices in Latent Class Analysis using the Free Open Source Software"
+shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
 author: 
   - name          : "Caspar J. van Lissa"
-    affiliation   : "1,2"
+    affiliation   : "1"
     corresponding : yes    # Define only one corresponding author
-    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
-    email         : "c.j.vanlissa@gmail.com"
+    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
+    email: "c.j.vanlissa@tilburguniversity.edu"
+    role:         # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
+      - Conceptualization
+      - Software
+      - Supervision
+      - Writing – original draft
+      - Writing – review & editing
   - name          : "Mauricio Garnier-Villarreal"
+    affiliation   : "2"
+    corresponding : no   
+    email         : "m.garniervillarreal@vu.nl"
+    role:
+      - Writing – review & editing
+  - name          : "Daniel Anadria"
     affiliation   : "3"
     corresponding : no
-    email         : "m.garniervillarreal@vu.nl"
-  - name          : "Daniel Anadria"
-    affiliation   : "1"
-    corresponding : no
-    address       : "Padualaan 14, 3584CH Utrecht, The Netherlands"
     email         : "d.anadria@uu.nl"
+    role:
+      - Writing – original draft
+      - Writing – review & editing
 affiliation:
   - id            : "1"
+    institution   : "Tilburg University, Methodology & Statistics"
+  - id            : "2"
+    institution   : "Vrije Universiteit Amsterdam, Sociology"
+  - id            : "3"
     institution   : "Utrecht University, Methodology & Statistics"
-  - id            : "2"
-    institution   : "Open Science Community Utrecht"
-  - id            : "3"
-    institution   : "Vrije Universiteit Amsterdam, Sociology"
 author_note: |

### 2023-03-24 manuscript.Rmd
change argument for BIC

@@
   describes best practices for the estimation and reporting of LCA,
-  using free and open source software in R. To this end, 
+  using the free and open source software in R. To this end, 
   we introduce new functionality for estimating and reporting mixture models
@@
 Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007; @magnus_multidimensional_2022].
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007].
 These models estimate one class to explain the "true zeros",
@@
 <!-- CJ: This example is not perfect, we need another example still.. -->
-For instance, a depression questionnaire PH-9 have found 2 distinct groups, ones with no symptoms, and one with symptoms.
-A mixture model can model the "presence" and "frequency" of symptoms as separate classes, instead of assuming that the sample follows a single continuous underlying distribution [@magnus_multidimensional_2022].
+One example of mixture models for violations of normality 
+is in modeling responses to the PH-9, a questionnaire of depressive symptoms.
+Researchers observed two distinct groups: One with no symptoms on most indicators (zero inflation),
+and one with non-zero symptoms [@magnus_multidimensional_2022].
+As this response pattern violates the assumption of multivariate normality,
+the authors used a mixture model to capture the "presence" and "frequency" of symptoms as separate classes.
 <!-- the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
@@
 then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
-By default, `tidySEM` derives starting values using K-means clustering [@likas_global_2003; @Forgy1965ClusterAO; @Hartigan_kmeans],
+By default, `tidySEM` derives starting values using K-means clustering [@Hartigan_kmeans],
 computing the mixture model starting values by treating the K-means solution as multi-group model.
@@
 Finally,
-the so-called *sample size adjusted BIC* (*saBIC*) implements a different ad-hoc penalty for sample size,
-$\ln(\frac{n + 2}{24})$. Although simulation studies have shown that the saBIC performs well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
-a limitation is that its penalty is not based on theory, but on the Minimum Description Length (MDL) as the shortes set of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987].
-BIC also performs well and is theoretically substantiated.
-As such, BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
+the so-called *sample size adjusted BIC* (*saBIC*), $\ln(\frac{n + 2}{24})$,
+implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987].
+Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
+but tends to select too many classes when the number of the number of classes and the sample size are large [@tein_statistical_2013; @chen_efficacy_2017].
+The BIC has been found to perform consistently well across simulation studies,
+and is among the most widely known information criteria.
+As such, BIC may be a sensible default criterion for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
 In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.
@@
 By default, `table_fit()` includes the AIC, BIC, and saBIC.
-We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.
+We suggest to report at least the BIC, as it is ubiquitous, well-understood,
+and performs well in class enumeration.
 

### 2023-04-10 manuscript.Rmd

@@
 This test is implemented in `tidySEM`, and is a part of the default output of `table_fit()`.
+<!-- A note on ‘Testing the number of components in a normal mixture’ Get access Arrow -->
+<!-- Neal O. Jeffries -->
+<!-- Biometrika, Volume 90, Issue 4, December 2003, Pages 991–994, https://doi.org/10.1093/biomet/90.4.991 -->
+<!-- [LMR]requires conditions that are generally not met when the null hypothesis holds. Consequently, the result is not proven and simulations suggest that it may not be correct. -->
 
@@
 
+There are known limitations to both aforementioned tests;
+on the one hand, the LRT is known to contain mathematical inaccuracies [@jeffriesNoteTestingNumber2003a], yet still shows reasonable performance in simulation studies.
+On the other hand, the BLRT is particularly sensitive to model misspecification and is thus not appropriate for exploratory applications [@nylund_deciding_2007].
 The first author's anecdotal experience as a statistical co-author on LCA studies suggests that both aforementioned tests tend to be liberal,
@@
 Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
-This is not particularly meaningful,
-however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
-Thus, a better rule of thumb might be to ascertain that the smallest class has at least as many observations as the average number of parameters per class.
-<!-- DA: what is many more? When making a rule a thumb, you have to eyeball a quantity and recommend it for users without statistical intuition
-MGV: here I changed it "at least as many", as a low bound recommendation-->
+A major concern with these rules of thumb is that it ignores the absolute sample size within each class.
+Small classes may comprise too few observations to accurately estimate class parameters.
+Thus, a better rule of thumb might be to ascertain that the smallest class is sufficiently large;
+both in terms of absolute sample size and in terms of the number of observations per estimated parameter.
+Well-established rules of thumb suggest ratios of 10 or 20 observations per parameter [@klinePrinciplesPracticeStructural2016; @jacksonRevisitingSampleSize2003].
 Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
-If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.
+If the smallest class has too few observations per parameter, or its substantive interpretation is redundant with another class,
+a simpler solution with fewer classes or a simpler class-specific model is preferable.
 
@@
 which can be obtained using `class_prob(res, type = "avg.mostlikely")`.
-For participants assigned to one particular class (rows) according to $M$ (most likely class based on predicted probabilities),
+For participants assigned to one particular class (rows) according to $M$ (most likely class according to posterior class probabilities),
 this table gives the mean probability of being assigned to each class (columns).
@@
 Fifth, the LMR or BLRT can be reported if it is a part of the class enumeration strategy, using `lr_lmr()` and `BLRT()`.
-By default, `table_fit()` reports the LMR (also called Vuong’s robust likelihood-ratio test, VLMR) for subsequent models in the table.
-
-<!-- DA: We do not introduce VLM, what does it stand for? -->
+By default, `table_fit()` reports the LMR for subsequent models in the table.
 

### 2023-04-10 manuscript.Rmd
Final Proofread pt. 1
Minor spelling corrections.
Added one comment.

Part 2 coming by early afternoon today

@@
 as its users can contribute to the development, maintenance, and support of the software.
-Finally, FOSS promotes tranparency and reproducibility: Because the source code for FOSS is accessible,
+Finally, FOSS promotes transparency and reproducibility: Because the source code for FOSS is accessible,
 researchers can review and verify the methods and algorithms used,
@@
 or implement specific innovations (e.g., `depmixS4`).
+<!-- DA: I don't understand the message for `depmixS4`.
+Why is 'implementing specific innovations' a limitation?
+If this is a euphemism, we can express the limitation more directly, or qualify these innovations.
+-->
 `OpenMx` is unique in that, like proprietary *Mplus*, it implements LCA in the context of a fully featured structural equation modeling framework.
@@
 The distribution is evidently bimodal, which bodes well for the intended
-mixture model. In this case, the number of classes is known a-priori.
+mixture model. In this case, the number of classes is known a priori.
 When estimating a two-class mixture model, the detective observes that
@@
 thought of as similar to a multi-group structural equation model. The
-main distinction is that group membership is not known a-priori, but estimated from the data.
+main distinction is that group membership is not known a priori, but estimated from the data.
 Whereas in a multi-group model, the data are split by group and treated
@@
 In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes $k$. 
-The number $k$ may be chosen a-priori or dictated by the data.
+The number $k$ may be chosen a priori or dictated by the data.
 It is sensible to limit $k$ to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
@@
 <!-- 1. We do not consider solutions with minimum posterior classification probability < .90 because poor class separability compromises interpretability of the results -->
-<!-- Nevertheless, researchers can set an a-priori threshold for the minimum proportion of cases per class, -->
+<!-- Nevertheless, researchers can set an a priori threshold for the minimum proportion of cases per class, -->
 <!-- for example by considering only solutions with a number of participants at least five times as large as the number of parameters per class. -->

### 2023-04-11 manuscript.Rmd
Final Proofread pt.2
Consistency checks, proofreading, updated table with sections

Third and final proofread to be pushed in by the end of workday today

@@
 Why is 'implementing specific innovations' a limitation?
-If this is a euphemism, we can express the limitation more directly, or qualify these innovations.
+If this is a euphemism, we can express the limitation more directly, i.e.
+qualify these innovations, e.g.
+[...] which are not software-agnostic (if that is the limitation).
 -->
@@
 The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
-but the resulting models can still be fully customized making it a low threshold - high ceiling package.
+but the resulting models can still be fully customized making it a low threshold high ceiling package.
 For instance, LCA models constructed in `tidySEM` inherit the `MxModel` format from `OpenMx` which renders them compatible with various legacy functions from `OpenMx` and related LCA ecosystem .
@@
 or to assign individuals to classes based on their highest class probability.
-If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
-This perspective on mixture modeling is sometimes referred to as "model-based
-clustering" [@hennig_handbook_2015; @scrucca_mclust_2016]. 
-Many clustering algorithms apply some recursive splitting algorithm to the data. By
-contrast "model-based" clustering refers to the fact that LCA estimates
+The latter approach may be acceptable when the classes are clearly distinct and the measurement error is low. 
+This perspective on mixture modeling is sometimes referred to as *model-based
+clustering* [@hennig_handbook_2015; @scrucca_mclust_2016]. 
+Many clustering algorithms apply some recursive splitting algorithm to the data. 
+By contrast, model-based clustering refers to the fact that LCA estimates
 cluster membership based on a parametric model.
-Since a parametric model estimates only a fixed number of parameters,
-it can be a relatively parsimonious solution compared to non-parametric techniques.
+Since parametric models estimate only a fixed number of parameters,
+they can be a relatively parsimonious solution compared to non-parametric techniques.
 
@@
 
-The term LCA was initially conceived to refer to analyses with categorical,
+Initially, the term LCA was conceived to refer to analyses with categorical,
 usually binary indicators [@collins_latent_2009].
-It was later generalized to refer to any technique that estimates a categorical latent variable based on a
-parametric model of observed indicators, as it only refers
+Later, it became generalized to refer to any technique that estimates a categorical latent variable based on a
+parametric model of observed indicators, as the term only refers
 to the purpose of the analysis, and does not imply restrictions to the
@@
 To prevent ambiguity, we refer to LCA with binary or ordinal indicators as *LCA with ordinal indicators*,
-and the sole *LCA* expression as an umbrella term.
+and use *LCA* as an umbrella term for the family techniques.
 Given the abundance of redundant terms in use for closely related types of LCA models,
@@
 
-A common type of LCA is the *finite Gaussian mixture model*; a
+A common type of LCA is the *finite Gaussian mixture model*, a
 univariate analysis where the observed distribution of a single variable
@@
 assumed to be zero.
-In some cases, such constraints
+In some cases such constraints
 will be inappropriate; for instance when the cohesion between indicators
@@
 the large particles were heterogenous in shape,
-whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements [@alkema_maximizing_2022].
+whereas the smaller particles had been polished to a more uniform, rounded shape by the elements [@alkema_maximizing_2022].
 
@@
 called *Latent Transition Analysis*), which estimates the change in class
-membership over time under the Markov assumption. The Markov
+membership over time under the *Markov assumption*. The Markov
 assumption states that class membership at time $t$ is only affected
@@
 
-### Testing theory
+### Testing Theory
 
@@
 
-### Measurement model
+### Measurement Model
 
@@
 
-### Dimension reduction
+### Dimension Reduction
 
@@
 These guidelines were generalized to be relevant to all types of LCA, 
-and updated to current best practices, as explained below.
+and updated to current best practices showcasing `tidySEM` syntax.
 <!-- In order to aid researchers working with latent trajectory models, -->
@@
   "Examining Observed Data",
+  "Data Preprocessing",
   "Handling Missing Data",
-  "Alternative Model Specifications",
-  "Software",
-  "Algorithm",
+  "Model Specification",
+  "Estimation and Convergence",
   "Class Enumeration",
   "Model Fit Indices",
+  "Nested Model Tests",
   "Classification Diagnostics",
-  "Interpreting Class Solutions",
-  "Label switching"
-
+  "Labelling Classes",
+  "Follow-up Analyses"
 ))
@@
 Numeric variables should have an interval measurement level,
-such that it is sensible to estimate theirmeans and (co)variances.
+such that it is sensible to estimate their means and (co)variances.
 For the remaining three measurement levels,

### 2023-04-11 manuscript.tex
Final Proofread pt.2
Consistency checks, proofreading, updated table with sections

Third and final proofread to be pushed in by the end of workday today

@@
 \hypersetup{
-  pdftitle={Best Practices in Latent Class Analysis using Free Open Source Software},
-  pdfauthor={Caspar J. van Lissa1,2, Mauricio Garnier-Villarreal3, \& Daniel Anadria1},
+  pdftitle={Best Practices in Latent Class Analysis using the Free Open Source Software},
+  pdfauthor={Caspar J. van Lissa1, Mauricio Garnier-Villarreal2, \& Daniel Anadria3},
   pdflang={en-EN},
@@
 
-\title{Best Practices in Latent Class Analysis using Free Open Source Software}
-\author{Caspar J. van Lissa\textsuperscript{1,2}, Mauricio Garnier-Villarreal\textsuperscript{3}, \& Daniel Anadria\textsuperscript{1}}
+\title{Best Practices in Latent Class Analysis using the Free Open Source Software}
+\author{Caspar J. van Lissa\textsuperscript{1}, Mauricio Garnier-Villarreal\textsuperscript{2}, \& Daniel Anadria\textsuperscript{3}}
 \date{}
 
-\shorttitle{BEST PRACTICES IN LATENT CLASS ANALYSIS}
+\shorttitle{BEST PRACTICES LATENT CLASS ANALYSIS}
 
@@
 
+The authors made the following contributions.
+Caspar J.
+van Lissa: Conceptualization, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Mauricio Garnier-Villarreal: Writing -- review \& editing; Daniel Anadria: Writing -- original draft, Writing -- review \& editing.
+
 Correspondence concerning this article should be addressed to Caspar J.
-van Lissa, Padualaan 14, 3584CH Utrecht, The Netherlands.
-E-mail: \href{mailto:c.j.vanlissa@gmail.com}{\nolinkurl{c.j.vanlissa@gmail.com}}
+van Lissa, Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands.
+E-mail: \href{mailto:c.j.vanlissa@tilburguniversity.edu}{\nolinkurl{c.j.vanlissa@tilburguniversity.edu}}
 
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Utrecht University, Methodology \& Statistics\\\textsuperscript{2} Open Science Community Utrecht\\\textsuperscript{3} Vrije Universiteit Amsterdam, Sociology}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Tilburg University, Methodology \& Statistics\\\textsuperscript{2} Vrije Universiteit Amsterdam, Sociology\\\textsuperscript{3} Utrecht University, Methodology \& Statistics}
 
@@
 describes best practices for the estimation and reporting of LCA,
-using free and open source software in R. To this end,
+using the free and open source software in R. To this end,
 we introduce new functionality for estimating and reporting mixture models
@@
 The package contains user-friendly wrapper functions to conduct LCA with sensible defaults,
-but the resulting models can still be fully customized making it a low threshold - high ceiling package.
+but the resulting models can still be fully customized making it a low threshold high ceiling package.
 For instance, LCA models constructed in \texttt{tidySEM} inherit the \texttt{MxModel} format from \texttt{OpenMx} which renders them compatible with various legacy functions from \texttt{OpenMx} and related LCA ecosystem .
@@
 or to assign individuals to classes based on their highest class probability.
-If the classes are clearly distinct, measurement error is low, and the latter approach may be acceptable.
-This perspective on mixture modeling is sometimes referred to as ``model-based
-clustering'' (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016).
-Many clustering algorithms apply some recursive splitting algorithm to the data. By
-contrast ``model-based'' clustering refers to the fact that LCA estimates
+The latter approach may be acceptable when the classes are clearly distinct and the measurement error is low.
+This perspective on mixture modeling is sometimes referred to as \emph{model-based
+clustering} (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca et al., 2016).
+Many clustering algorithms apply some recursive splitting algorithm to the data.
+By contrast, model-based clustering refers to the fact that LCA estimates
 cluster membership based on a parametric model.
-Since a parametric model estimates only a fixed number of parameters,
-it can be a relatively parsimonious solution compared to non-parametric techniques.
+Since parametric models estimate only a fixed number of parameters,
+they can be a relatively parsimonious solution compared to non-parametric techniques.
 
@@
 
-The term LCA was initially conceived to refer to analyses with categorical,
+Initially, the term LCA was conceived to refer to analyses with categorical,
 usually binary indicators (Collins \& Lanza, 2009).
-It was later generalized to refer to any technique that estimates a categorical latent variable based on a
-parametric model of observed indicators, as it only refers
+Later, it became generalized to refer to any technique that estimates a categorical latent variable based on a
+parametric model of observed indicators, as the term only refers
 to the purpose of the analysis, and does not imply restrictions to the
@@
 To prevent ambiguity, we refer to LCA with binary or ordinal indicators as \emph{LCA with ordinal indicators},
-and the sole \emph{LCA} expression as an umbrella term.
+and use \emph{LCA} as an umbrella term for the family techniques.
 Given the abundance of redundant terms in use for closely related types of LCA models,
@@
 
-A common type of LCA is the \emph{finite Gaussian mixture model}; a
+A common type of LCA is the \emph{finite Gaussian mixture model}, a
 univariate analysis where the observed distribution of a single variable
@@
 the large particles were heterogenous in shape,
-whereas the smaller particles had been polished to a more uniform (rounded) shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).
+whereas the smaller particles had been polished to a more uniform, rounded shape by the elements (Alkema, Van Lissa, Kooi, \& Koelmans, 2022).
 
@@
 called \emph{Latent Transition Analysis}), which estimates the change in class
-membership over time under the Markov assumption. The Markov
+membership over time under the \emph{Markov assumption}. The Markov
 assumption states that class membership at time \(t\) is only affected
@@
 \hypertarget{testing-theory}{%
-\subsubsection{Testing theory}\label{testing-theory}}
+\subsubsection{Testing Theory}\label{testing-theory}}
 
@@
 \hypertarget{measurement-model}{%
-\subsubsection{Measurement model}\label{measurement-model}}
+\subsubsection{Measurement Model}\label{measurement-model}}
 
@@
 \hypertarget{dimension-reduction}{%
-\subsubsection{Dimension reduction}\label{dimension-reduction}}
+\subsubsection{Dimension Reduction}\label{dimension-reduction}}
 
@@
 Finally, LCA can be used to deal with data which violate certain assumptions.
-For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (Baughman, 2007; Magnus \& Garnier-Villarreal, 2022).
+For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) (Baughman, 2007).
 These models estimate one class to explain the ``true zeros'',
@@
 and thus falls beyond the scope of this paper.
-For instance, the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality (Alkema et al., 2022).
-A mixture of two normal distributions with distinct variances described the observed distribution better.
+One example of mixture models for violations of normality
+is in modeling responses to the PH-9, a questionnaire of depressive symptoms.
+Researchers observed two distinct groups: One with no symptoms on most indicators (zero inflation),
+and one with non-zero symptoms (Magnus \& Garnier-Villarreal, 2022).
+As this response pattern violates the assumption of multivariate normality,
+the authors used a mixture model to capture the ``presence'' and ``frequency'' of symptoms as separate classes.
 
@@
 These guidelines were generalized to be relevant to all types of LCA,
-and updated to current best practices, as explained below.
+and updated to current best practices showcasing \texttt{tidySEM} syntax.
 
@@
 \hline
-2. & Handling Missing Data\\
+2. & Data Preprocessing\\
 \hline
-3. & Alternative Model Specifications\\
+3. & Handling Missing Data\\
 \hline
-4. & Software\\
+4. & Model Specification\\
 \hline
-5. & Algorithm\\
+5. & Estimation and Convergence\\
 \hline
@@
 \hline
-8. & Classification Diagnostics\\
+8. & Nested Model Tests\\
 \hline
-9. & Interpreting Class Solutions\\
+9. & Classification Diagnostics\\
 \hline
-10. & Label switching\\
+10. & Labelling Classes\\
+\hline
+11. & Follow-up Analyses\\
 \hline
@@
 then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids (see also Biernacki, Celeux, \& Govaert, 2003).
-By default, \texttt{tidySEM} derives starting values using K-means clustering (Likas, Vlassis, \& J. Verbeek, 2003),
+By default, \texttt{tidySEM} derives starting values using K-means clustering (Hartigan \& Wong, 1979),
 computing the mixture model starting values by treating the K-means solution as multi-group model.
@@
 In exploratory LCA, the set of models typically consists of a sequence from 1 to the maximum number of classes \(k\).
-The number \(k\) may be chosen a-priori or dictated by the data.
+The number \(k\) may be chosen a priori or dictated by the data.
 It is sensible to limit \(k\) to the maximum number of classes that result in valid and interpretable solutions (also see Classification Diagnostics).
@@
 Finally,
-the so-called \emph{sample size adjusted BIC} (\emph{saBIC}) implements a different ad-hoc penalty for sample size,
-\(\ln(\frac{n + 2}{24})\). Although simulation studies have showen that the saBIC performs well for class enumeration (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013),
-a limitation is that its penalty is not based on theory. BIC also performs well and is theoretically substantiated.
-As such, BIC may be the most appropriate information criterion to use for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
+the so-called \emph{sample size adjusted BIC} (\emph{saBIC}), \(\ln(\frac{n + 2}{24})\),
+implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data (Rissanen, 1983; Sclove, 1987).
+Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
+but tends to select too many classes when the number of the number of classes and the sample size are large (Chen, Luo, Palardy, Glaman, \& McEnturff, 2017; Tein, Coxe, \& Cham, 2013).
+The BIC has been found to perform consistently well across simulation studies,
+and is among the most widely known information criteria.
+As such, BIC may be a sensible default criterion for model comparison (Masyn, 2013; Nylund-Gibson \& Choi, 2018).
 In \texttt{tidySEM}, all three ICs are available as a part of the output of \texttt{table\_fit()}.
@@
 
+There are known limitations to both aforementioned tests;
+on the one hand, the LRT is known to contain mathematical inaccuracies (Jeffries, 2003), yet still shows reasonable performance in simulation studies.
+On the other hand, the BLRT is particularly sensitive to model misspecification and is thus not appropriate for exploratory applications (Nylund et al., 2007).
 The first author's anecdotal experience as a statistical co-author on LCA studies suggests that both aforementioned tests tend to be liberal,
@@
 Existing rules of thumb have focused on smallest class size in terms of percentage of the total sample size.
-This is not particularly meaningful,
-however, as the primary concern is that small classes may comprise too few observations to accurately estimate class parameters.
-Thus, a better rule of thumb might be to ascertain that the smallest class has many more observations than the average number of parameters per class.
+A major concern with these rules of thumb is that it ignores the absolute sample size within each class.
+Small classes may comprise too few observations to accurately estimate class parameters.
+Thus, a better rule of thumb might be to ascertain that the smallest class is sufficiently large;
+both in terms of absolute sample size and in terms of the number of observations per estimated parameter.
+Well-established rules of thumb suggest ratios of 10 or 20 observations per parameter (Jackson, 2003; Kline, 2016).
 Another good practice is to ensure that the smallest class has a unique and meaningful substantive interpretation.
-If the smallest class does not comprise at least a few observations per parameter, or is redundant with another larger class, a simpler solution is preferable.
+If the smallest class has too few observations per parameter, or its substantive interpretation is redundant with another class,
+a simpler solution with fewer classes or a simpler class-specific model is preferable.
 
@@
 which can be obtained using \texttt{class\_prob(res,\ type\ =\ "avg.mostlikely")}.
-For participants assigned to one particular class (rows) according to \(M\),
-this table gives the mean probability of being assigned to any class (columns).
+For participants assigned to one particular class (rows) according to \(M\) (most likely class according to posterior class probabilities),
+this table gives the mean probability of being assigned to each class (columns).
 This table is obtained by calculating the column means of \(P\) for the
@@
 By default, \texttt{table\_fit()} includes the AIC, BIC, and saBIC.
-We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.
+We suggest to report at least the BIC, as it is ubiquitous, well-understood,
+and performs well in class enumeration.
 
@@
 Fifth, the LMR or BLRT can be reported if it is a part of the class enumeration strategy, using \texttt{lr\_lmr()} and \texttt{BLRT()}.
-By default, \texttt{table\_fit()} reports the VLM for subsequent models in the table.
+By default, \texttt{table\_fit()} reports the LMR for subsequent models in the table.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Hartigan_kmeans}{}}%
+Hartigan, J.
+A., \& Wong, M.
+A.
+(1979).
+Algorithm AS 136: A k-means clustering algorithm.
+\emph{Journal of the Royal Statistical Society.
+Series C (Applied Statistics)}, \emph{28}(1), 100--108.
+Retrieved from \url{http://www.jstor.org/stable/2346830}
+
 \leavevmode\vadjust pre{\hypertarget{ref-hastie_elements_2009}{}}%
@@
 Hennig, C., Meila, M., Murtagh, F., \& Rocci, R. (2015). \emph{Handbook of cluster analysis}. 28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jacksonRevisitingSampleSize2003}{}}%
+Jackson, D.
+L.
+(2003).
+Revisiting {Sample Size} and {Number} of {Parameter Estimates}: {Some Support} for the {N}:q {Hypothesis}.
+\emph{Structural Equation Modeling: A Multidisciplinary Journal}, \emph{10}(1), 128--141.
+\url{https://doi.org/10.1207/S15328007SEM1001_6}
 
@@
 \url{https://doi.org/10.1007/s11336-010-9175-3}
+
+\leavevmode\vadjust pre{\hypertarget{ref-jeffriesNoteTestingNumber2003a}{}}%
+Jeffries, N.
+O.
+(2003).
+A {Note} on '{Testing} the {Number} of {Components} in a {Normal Mixture}'.
+\emph{Biometrika}, \emph{90}(4), 991--994.
+Retrieved from \url{https://www.jstor.org/stable/30042105}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-klinePrinciplesPracticeStructural2016}{}}%
+Kline, R. B. (2016). \emph{Principles and {Practice} of {Structural Equation Modeling}} (Fourth). {New York}: {The Guilford Press}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-van_kollenburg_lazy_2018}{}}%
@@
 \url{https://doi.org/10.1037/met0000381}
-
-\leavevmode\vadjust pre{\hypertarget{ref-likas_global_2003}{}}%
-Likas, A., Vlassis, N., \& J. Verbeek, J. (2003). The global k-means clustering algorithm. \emph{Pattern Recognition}, \emph{36}(2), 451--461. \url{https://doi.org/10.1016/S0031-3203(02)00060-2}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Rissanen_1985}{}}%
+Rissanen, J.
+(1983).
+{A Universal Prior for Integers and Estimation by Minimum Description Length}.
+\emph{The Annals of Statistics}, \emph{11}(2), 416--431.
+\url{https://doi.org/10.1214/aos/1176346150}
+
 \leavevmode\vadjust pre{\hypertarget{ref-rosenberg_tidylpa_2018}{}}%
@@
 \url{https://doi.org/10.1016/j.sapharm.2016.11.011}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sclove_application_1987}{}}%
+Sclove, S. L. (1987). Application of model-selection criteria to some problems in multivariate analysis. \emph{Psychometrika}, \emph{52}(3), 333--343. \url{https://doi.org/10.1007/BF02294360}
 

### 2023-04-11 manuscript.Rmd
Final Proofread pt.3
Minor adjustments to phrasing etc.

@@
 argued that some parametric methods are robust when scales with 7+
-response categories are treated as continuous [e.g., @norman_likert_2010; rhemtulla_when_2012],
+response categories are treated as continuous [e.g., @norman_likert_2010; @rhemtulla_when_2012],
 it is very unlikely
@@
 As explained before, the number of parameters in LCA scales with the number of estimated classes.
-Consequently, LCA with many classes have a potentially very high number of parameters.
+Consequently, LCA with many classes have a large number of parameters.
 In Bayesian mixture modeling,
@@
 
-Models can be specified in `tidySEM` in different ways.
+In `tidySEM` models can be specified in different ways.
 The wrapper functions `mx_profiles()` and `mx_lca()` construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
@@
 there is a risk that the global optimum is never found.
-An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing [SA, @corana_minimizing_1987].
+An alternative solution that overcomes these challenges is to use a global optimizer, like *simulated annealing* [*SA*, @corana_minimizing_1987].
 SA does not behave quite like the marbles of the random starts approach;
@@
 The `OpenMx` package contains a family of functions for this purpose: `mxTryHard()` and `mxTryHardWideSearch()` for continuous indicators, and `mxTryHardOrdinal()` for ordinal indicators.
-When poor data quality is to blame, the only solution is to collect more, or higher quality, data.
+When poor data quality is to blame, the only solution is to collect more, or higher quality data.
 
@@
 likelihood ratio tests [@weller_latent_2020; @masyn_latent_2013].
-Substantive criteria include convergence and model identification,
+Substantive criteria include model convergence and identification, and
 class separability,
@@
 relies on information criteria.
-Information criteria (ICs) are computed from the $-2LL$,
+*Information criteria* (*ICs*) are computed from the $-2LL$,
 which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
@@
 
-Given a specific information criterion, it is also possible to calculate so-called "IC-weights",
-which indicate the relative support the data provides for each model in the set [@wagenmakers_aic_2004].
+Given a specific information criterion, it is also possible to calculate *IC-weights*,
+which indicate the relative support the data provides for each model under consideration [@wagenmakers_aic_2004].
 This is particularly useful for confirmatory LCA,
@@
 With this in mind, `table_fit()` does not provide the BLRT by default.
-Specific models can be compared using the function `BLRT()`.
+Instead, specific models can be compared using the function `BLRT()`.
 

### 2023-04-11 manuscript.tex
Final Proofread pt.3
Minor adjustments to phrasing etc.

@@
 argued that some parametric methods are robust when scales with 7+
-response categories are treated as continuous {[}e.g., Norman (2010); rhemtulla\_when\_2012{]},
+response categories are treated as continuous (e.g., Norman, 2010; Rhemtulla, Brosseau-Liard, \& Savalei, 2012),
 it is very unlikely
@@
 
-Models can be specified in \texttt{tidySEM} in different ways.
+In \texttt{tidySEM} models can be specified in different ways.
 The wrapper functions \texttt{mx\_profiles()} and \texttt{mx\_lca()} construct LCA models for continuous and ordinal indicators respectively, without needing to specify model syntax.
@@
 there is a risk that the global optimum is never found.
-An alternative solution that overcomes these challenges is to use a global optimizer, like simulated annealing (SA, Corana et al., 1987).
+An alternative solution that overcomes these challenges is to use a global optimizer, like \emph{simulated annealing} (\emph{SA}, Corana et al., 1987).
 SA does not behave quite like the marbles of the random starts approach;
@@
 The \texttt{OpenMx} package contains a family of functions for this purpose: \texttt{mxTryHard()} and \texttt{mxTryHardWideSearch()} for continuous indicators, and \texttt{mxTryHardOrdinal()} for ordinal indicators.
-When poor data quality is to blame, the only solution is to collect more, or higher quality, data.
+When poor data quality is to blame, the only solution is to collect more, or higher quality data.
 
@@
 likelihood ratio tests (Masyn, 2013; Weller et al., 2020).
-Substantive criteria include convergence and model identification,
+Substantive criteria include model convergence and identification, and
 class separability,
@@
 relies on information criteria.
-Information criteria (ICs) are computed from the \(-2LL\),
+\emph{Information criteria} (\emph{ICs}) are computed from the \(-2LL\),
 which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
@@
 
-Given a specific information criterion, it is also possible to calculate so-called ``IC-weights'',
-which indicate the relative support the data provides for each model in the set (Wagenmakers \& Farrell, 2004).
+Given a specific information criterion, it is also possible to calculate \emph{IC-weights},
+which indicate the relative support the data provides for each model under consideration (Wagenmakers \& Farrell, 2004).
 This is particularly useful for confirmatory LCA,
@@
 With this in mind, \texttt{table\_fit()} does not provide the BLRT by default.
-Specific models can be compared using the function \texttt{BLRT()}.
+Instead, specific models can be compared using the function \texttt{BLRT()}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-rhemtulla_when_2012}{}}%
+Rhemtulla, M., Brosseau-Liard, P.
+É., \& Savalei, V.
+(2012).
+When can categorical variables be treated as continuous?
+A comparison of robust continuous and categorical {SEM} estimation methods under suboptimal conditions.
+\emph{Psychological Methods}, \emph{17}, 354--373.
+\url{https://doi.org/10.1037/a0029315}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Rissanen_1985}{}}%
